### 2015-09-28 doc/dispRity.demo.Rnw
vignette INIT

\documentclass{article}
\usepackage[sc]{mathpazo}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}
\usepackage{url}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=true,bookmarksnumbered=true,bookmarksopen=true,bookmarksopenlevel=2,
 breaklinks=false,pdfborder={0 0 1},backref=false,colorlinks=false]
 {hyperref}
\hypersetup{
 pdfstartview={XYZ null null 1}}

\newcommand{\dispRity}{\texttt{dispRity} }
\newcommand{\R}{\texttt{R} }

\begin{document}
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.path='figure/minimal-', fig.align='center', fig.show='hold')
options(formatR.arrow=TRUE,width=90)
@

\title{dispRity demo}

\author{Thomas Guillerme}

\maketitle

This a quick demo to go through the beta version of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}).
Many parts are still missing (see section \ref{whatsleft}) but this version should be running all right and give an idea of what's implemented in the package.

\section{Before starting}

\dispRity is a package for calculating disparity in \R.
What is disparity? Well that's a more complex question...
To keep it short, this package allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.
% More explanations to put here?

\subsection{Installation}
You can install this package pretty easily if you use the latest version of \R and \texttt{devtools}.
Just copy past the following
<<installation, eval=FALSE>>=
install.packages("devtools")
library(devtools)
install_github("TGuillerme/dispRity", ref="release")
library(dispRity)
@

Note that we use here the \texttt{release} branch which is version \Sexpr{packageVersion("dispRity")}.
If you want the piping-hot (and full of bugs) version, change the option \texttt{ref="release"} to \texttt{ref="master"}.

This package depends heavily on the \texttt{ape} package as well as the nice function \texttt{timeSliceTree::paleotree}.
However, I recommend to also install the packages \texttt{geomorph} or \texttt{Claddis} if you want to try this example on your own morphometric or cladistic data.

\subsection{Getting the data}
You'll need to do this and that

\section{A super quick go through}
This package allows to:
\begin{enumerate}
    \item Splitting the data
    \item Bootstrapping the data
    \item Calculating disparity
    \item Summarising the results
\end{enumerate}

<<Quick go, fig.width=4, fig.height=4, out.width='.4\\linewidth'>>=
## Separate your ordinated data into different series for example
1+1
## Bootstrap this data
1+1
## Calculate a disparity metric from it
1+1
# Summarise and plot the disparity
summary(1)
plot(1)
@

\section{In more details}

\subsection{Splitting the data}

\subsection{Bootstrapping the data}

\subsection{Calculating disparity}

\subsection{Summarising the results}

\section{What's left?}
\label{whatsleft}

\end{document}

### 2015-09-28 doc/dispRity.demo.tex
vignette INIT

\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[sc]{mathpazo}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}
\usepackage{url}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=true,bookmarksnumbered=true,bookmarksopen=true,bookmarksopenlevel=2,
 breaklinks=false,pdfborder={0 0 1},backref=false,colorlinks=false]
 {hyperref}
\hypersetup{
 pdfstartview={XYZ null null 1}}

\newcommand{\dispRity}{\texttt{dispRity} }
\newcommand{\R}{\texttt{R} }
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

\title{dispRity demo}

\author{Thomas Guillerme}

\maketitle

This a quick demo to go through the beta version of the \dispRity package (v.0.1.0).
Many parts are still missing (see section \ref{whatsleft}) but this version should be running all right and give an idea of what's implemented in the package.

\section{Before starting}

\dispRity is a package for calculating disparity in \R.
What is disparity? Well that's a more complex question...
To keep it short, this package allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.
% More explanations to put here?

\subsection{Installation}
You can install this package pretty easily if you use the latest version of \R and \texttt{devtools}.
Just copy past the following
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{install.packages}\hlstd{(}\hlstr{"devtools"}\hlstd{)}
\hlkwd{library}\hlstd{(devtools)}
\hlkwd{install_github}\hlstd{(}\hlstr{"TGuillerme/dispRity"}\hlstd{,} \hlkwc{ref}\hlstd{=}\hlstr{"release"}\hlstd{)}
\hlkwd{library}\hlstd{(dispRity)}
\end{alltt}
\end{kframe}
\end{knitrout}

Note that we use here the \texttt{release} branch which is version 0.1.0.
If you want the piping-hot (and full of bugs) version, change the option \texttt{ref="release"} to \texttt{ref="master"}.

This package depends heavily on the \texttt{ape} package as well as the nice function \texttt{timeSliceTree::paleotree}.
However, I recommend to also install the packages \texttt{geomorph} or \texttt{Claddis} if you want to try this example on your own morphometric or cladistic data.

\subsection{Getting the data}
You'll need to do this and that

\section{A super quick go through}
This package allows to:
\begin{enumerate}
    \item Separate your ordinated data into different series for example
    \item Bootstrap this data
    \item Calculate a disparity metric from it
    \item Summarise and plot the disparity
\end{enumerate}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Separate your ordinated data into different series for example}
\hlnum{1}\hlopt{+}\hlnum{1}
\end{alltt}
\begin{verbatim}
## [1] 2
\end{verbatim}
\begin{alltt}
\hlcom{## Bootstrap this data}
\hlnum{1}\hlopt{+}\hlnum{1}
\end{alltt}
\begin{verbatim}
## [1] 2
\end{verbatim}
\begin{alltt}
\hlcom{## Calculate a disparity metric from it}
\hlnum{1}\hlopt{+}\hlnum{1}
\end{alltt}
\begin{verbatim}
## [1] 2
\end{verbatim}
\begin{alltt}
\hlcom{# Summarise and plot the disparity}
\hlkwd{summary}\hlstd{(}\hlnum{1}\hlstd{)}
\end{alltt}
\begin{verbatim}
##    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
##       1       1       1       1       1       1
\end{verbatim}
\begin{alltt}
\hlkwd{plot}\hlstd{(}\hlnum{1}\hlstd{)}
\end{alltt}
\end{kframe}

{\centering \includegraphics[width=.4\linewidth]{figure/minimal-Quick_go-1}

}

\end{knitrout}

\section{What's left?}
\label{whatsleft}

\end{document}

### 2015-09-29 doc/dispRity.demo.Rnw
Updated vignette

@@
 
-\subsection{Getting the data}
+\subsection{Which data?}
 You'll need to do this and that
+%To develop
+
+<<data, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+## Loading demo and the package data
+library(dispRity)
+
+## An ordinated matrix with the tips only
+data(BeckLee_mat50)
+dim(BeckLee_mat50)
+head(BeckLee_mat50[,1:5])
+
+## An ordinated matrix with the tips and the nodes (named nX)
+data(BeckLee_mat99)
+dim(BeckLee_mat99)
+head(BeckLee_mat99[,1:5], 2)
+tail(BeckLee_mat99[,1:5], 2)
+
+## A list of first and last occurence data for some fossils
+data(BeckLee_ages)
+head(BeckLee_ages)
+
+## And a phylogeny:
+data(BeckLee_tree)
+plot(BeckLee_tree) ; nodelabels(cex=0.8)
+@
 
@@
 
-<<Quick go, fig.width=4, fig.height=4, out.width='.4\\linewidth'>>=
-## Separate your ordinated data into different series for example
-1+1
-## Bootstrap this data
-1+1
-## Calculate a disparity metric from it
-1+1
-# Summarise and plot the disparity
-summary(1)
-plot(1)
+<<Quick go, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+## Splitting the data
+sliced_data <- time.series(BeckLee_mat99, BeckLee_tree, method = "continuous",
+    model = "acctran", time = 5, FADLAD = BeckLee_ages)
+
+## Bootstrapping the data
+bootstrapped_data <- boot.matrix(sliced_data, 100)
+
+## Calculating disparity
+sum_of_ranges <- dispRity(bootstrapped_data, metric = c(sum, range))
+
+## Summarising the results
+summary(sum_of_ranges)
+plot(sum_of_ranges, type = "continuous")
 @
@@
 
+\subsection{The \dispRity objects}
+Disparity analysis involve a lot of shuffling around with many matrices (especially when bootstrapping the data) which can be a bit impractical to visualise and quickly jam your \R console.
+For example, try to use \texttt{str} to show the structure of the object created above:
+
+<<dispRity class1, eval=FALSE>>=
+str(sum_of_ranges)
+@
+
+Therefore this package proposes a new class of object called \dispRity (how unexpected!) that allow to use a \texttt{S3 print.dispRity} method.
+In other words, when you are creating objects using the \dispRity functions, you can quickly and easily have a look at them by just printing it.
+For example:
+
+<<dispRity class2>>=
+## Which class is the sum_of_ranges object?
+class(sum_of_ranges)
+
+## We can summarise it using the S3 method print.dispRity
+sum_of_ranges 
+@
+
+Note that each object output from different functions will show summarise slightly different informations.
+
 \subsection{Splitting the data}
+One of the first functionality of this package is to facilitate splitting the ordinated matrix in subsamples.
+The original ordinated matrix represents the total any-o-space.
+
+\subsubsection{Custom splitting}
+\texttt{cust.series} is a fairly straightforward function and allows to split the data according to a factor determined by the user.
+For example, here we can split the matrix based on phylogeny: let's classify in two different groups the crown and stem mammals present in the tree.
+
+<<cust.series2>>=
+## We want to separate the species around the node 71 (see phylogeny above).
+## All the descendant are crown and all the ancestors are stem
+crown <- extract.clade(BeckLee_tree, node = 71)$tip.label
+stem <- drop.tip(BeckLee_tree, tip = crown)$tip.label
+
+## We then have to feed this information in a data.frame with one column
+factors <- as.data.frame(
+    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
+    ncol = 1, dimnames = list(c(crown, stem))))
+
+## Then we can use the customised series function
+crown_stem <- cust.series(BeckLee_mat50, factors)
+
+## This created a dispRity object containing two series: crown and stem
+class(crown_stem)
+crown_stem
+
+## This object contains three elements
+names(crown_stem)
+
+## With "data" being the list sub-matrices
+str(crown_stem$data)
+
+## "taxa" being the list of taxa in the original ordinated matrix
+str(crown_stem$taxa)
+
+## "series" containing information on the series type (custom) and names (crown and stem)
+crown_stem$series
+@
+
+Of course, using phylogeny as a factor is just an example.
+Any type of factors can be used in this example (e.g. body mass categories, habitat, diet, etc...).
+Which leads us to another type of category used a lot in disparity-through-time analysis: time!
+
+\subsubsection{Time splitting}
+\texttt{time.series} is a function that allows to split the matrix in different ways using time as a category.
+Two types of time series splitting can be performed with the \dispRity package by using the \texttt{method} option in \texttt{time.series}.
+\begin{enumerate}
+\item the classical discrete time splitting (or time-binning) using \texttt{method = "discrete"};
+\item or a new continuous time splitting (or time-slicing) using \texttt{method = "continuous"}.
+\end{enumerate}
 

### 2015-09-29 doc/dispRity.demo.tex
Updated vignette

@@
 
-\subsection{Getting the data}
+\subsection{Which data?}
 You'll need to do this and that
+%To develop
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{## Loading demo and the package data}
+\hlkwd{library}\hlstd{(dispRity)}
+\end{alltt}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: paleotree}}\begin{alltt}
+\hlcom{## An ordinated matrix with the tips only}
+\hlkwd{data}\hlstd{(BeckLee_mat50)}
+\hlkwd{dim}\hlstd{(BeckLee_mat50)}
+\end{alltt}
+\begin{verbatim}
+## [1] 50 48
+\end{verbatim}
+\begin{alltt}
+\hlkwd{head}\hlstd{(BeckLee_mat50[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{])}
+\end{alltt}
+\begin{verbatim}
+##                    [,1]          [,2]        [,3]       [,4]      [,5]
+## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
+## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
+## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
+## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
+## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
+## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273
+\end{verbatim}
+\begin{alltt}
+\hlcom{## An ordinated matrix with the tips and the nodes (named nX)}
+\hlkwd{data}\hlstd{(BeckLee_mat99)}
+\hlkwd{dim}\hlstd{(BeckLee_mat99)}
+\end{alltt}
+\begin{verbatim}
+## [1] 99 97
+\end{verbatim}
+\begin{alltt}
+\hlkwd{head}\hlstd{(BeckLee_mat99[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{],} \hlnum{2}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+##                  [,1]       [,2]       [,3]       [,4]        [,5]
+## Cimolestes -0.6082437 -0.0323683 0.08458885 -0.4338448 -0.30536875
+## Maelestes  -0.5730206 -0.2840361 0.01308847 -0.1258848  0.06123611
+\end{verbatim}
+\begin{alltt}
+\hlkwd{tail}\hlstd{(BeckLee_mat99[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{],} \hlnum{2}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+##            [,1]      [,2]       [,3]       [,4]       [,5]
+## n48 -0.05529018 0.4799330 0.04118477 0.04944912 -0.3558830
+## n49 -0.13067785 0.4478168 0.11956268 0.13800340 -0.3222785
+\end{verbatim}
+\begin{alltt}
+\hlcom{## A list of first and last occurence data for some fossils}
+\hlkwd{data}\hlstd{(BeckLee_ages)}
+\hlkwd{head}\hlstd{(BeckLee_ages)}
+\end{alltt}
+\begin{verbatim}
+##             FAD  LAD
+## Adapis     37.2 36.8
+## Asioryctes 83.6 72.1
+## Leptictis  33.9 33.3
+## Miacis     49.0 46.7
+## Mimotona   61.6 59.2
+## Notharctus 50.2 47.0
+\end{verbatim}
+\begin{alltt}
+\hlcom{## And a phylogeny:}
+\hlkwd{data}\hlstd{(BeckLee_tree)}
+\hlkwd{plot}\hlstd{(BeckLee_tree) ;} \hlkwd{nodelabels}\hlstd{(}\hlkwc{cex}\hlstd{=}\hlnum{0.8}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=.8\linewidth]{figure/minimal-data-1}
+
+}
+
+\end{knitrout}
 
@@
 \begin{enumerate}
-    \item Separate your ordinated data into different series for example
-    \item Bootstrap this data
-    \item Calculate a disparity metric from it
-    \item Summarise and plot the disparity
+    \item Splitting the data
+    \item Bootstrapping the data
+    \item Calculating disparity
+    \item Summarising the results
 \end{enumerate}
@@
 \begin{alltt}
-\hlcom{## Separate your ordinated data into different series for example}
-\hlnum{1}\hlopt{+}\hlnum{1}
+\hlcom{## Splitting the data}
+\hlstd{sliced_data} \hlkwb{<-} \hlkwd{time.series}\hlstd{(BeckLee_mat99, BeckLee_tree,} \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
+    \hlkwc{model} \hlstd{=} \hlstr{"acctran"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{FADLAD} \hlstd{= BeckLee_ages)}
+\end{alltt}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Some tips have FAD/LAD and are assumed to interval single points in time.}}\begin{alltt}
+\hlcom{## Bootstrapping the data}
+\hlstd{bootstrapped_data} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(sliced_data,} \hlnum{100}\hlstd{)}
+
+\hlcom{## Calculating disparity}
+\hlstd{sum_of_ranges} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(bootstrapped_data,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, range))}
+
+\hlcom{## Summarising the results}
+\hlkwd{summary}\hlstd{(sum_of_ranges)}
 \end{alltt}
 \begin{verbatim}
-## [1] 2
+##      series  n   mean   2.5%    25%    75%  97.5%
+## 1 133.51104  3 -0.348 -0.397 -0.350 -0.349 -0.230
+## 2 100.13328 14  0.101 -0.252 -0.046  0.214  0.291
+## 3  66.75552 23 -0.059 -0.163 -0.112  0.031  0.115
+## 4  33.37776 11  0.008 -0.134 -0.086  0.132  0.217
+## 5         0 10  0.009 -0.134 -0.086  0.140  0.187
 \end{verbatim}
 \begin{alltt}
-\hlcom{## Bootstrap this data}
-\hlnum{1}\hlopt{+}\hlnum{1}
-\end{alltt}
-\begin{verbatim}
-## [1] 2
-\end{verbatim}
-\begin{alltt}
-\hlcom{## Calculate a disparity metric from it}
-\hlnum{1}\hlopt{+}\hlnum{1}
-\end{alltt}
-\begin{verbatim}
-## [1] 2
-\end{verbatim}
-\begin{alltt}
-\hlcom{# Summarise and plot the disparity}
-\hlkwd{summary}\hlstd{(}\hlnum{1}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-##    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-##       1       1       1       1       1       1
-\end{verbatim}
-\begin{alltt}
-\hlkwd{plot}\hlstd{(}\hlnum{1}\hlstd{)}
+\hlkwd{plot}\hlstd{(sum_of_ranges,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{)}
 \end{alltt}
@@
 
-{\centering \includegraphics[width=.4\linewidth]{figure/minimal-Quick_go-1}
+{\centering \includegraphics[width=.8\linewidth]{figure/minimal-Quick_go-1}
 
@@
 
+\section{In more details}
+
+\subsection{The \dispRity objects}
+Disparity analysis involve a lot of shuffling around with many matrices (especially when bootstrapping the data) which can be a bit impractical to visualise and quickly jam your \R console.
+For example, try to use \texttt{str} to show the structure of the object created above:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{str}\hlstd{(sum_of_ranges)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+Therefore this package proposes a new class of object called \dispRity (how unexpected!) that allow to use a \texttt{S3 print.dispRity} method.
+In other words, when you are creating objects using the \dispRity functions, you can quickly and easily have a look at them by just printing it.
+For example:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{## Which class is the sum_of_ranges object?}
+\hlkwd{class}\hlstd{(sum_of_ranges)}
+\end{alltt}
+\begin{verbatim}
+## [1] "dispRity"
+\end{verbatim}
+\begin{alltt}
+\hlcom{## We can summarise it using the S3 method print.dispRity}
+\hlstd{sum_of_ranges}
+\end{alltt}
+\begin{verbatim}
+## Disparity measurements across 5 series for 99 taxa. 
+## Series:
+## 133.51104, 100.13328, 66.75552, 33.37776, 0.
+## Disparity calculated as: sum range for 97 dimensions.
+## Data was split using continuous method.
+## Data was bootstrapped 100 times, using the full bootstrap method.
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+Note that each object output from different functions will show summarise slightly different informations.
+
+\subsection{Splitting the data}
+One of the first functionality of this package is to facilitate splitting the ordinated matrix in subsamples.
+The original ordinated matrix represents the total any-o-space.
+
+\subsubsection{Custom splitting}
+\texttt{cust.series} is a fairly straightforward function and allows to split the data according to a factor determined by the user.
+For example, here we can split the matrix based on phylogeny: let's classify in two different groups the crown and stem mammals present in the tree.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{## We want to separate the species around the node 71 (see phylogeny above).}
+\hlcom{## All the descendant are crown and all the ancestors are stem}
+\hlstd{crown} \hlkwb{<-} \hlkwd{extract.clade}\hlstd{(BeckLee_tree,} \hlkwc{node} \hlstd{=} \hlnum{71}\hlstd{)}\hlopt{$}\hlstd{tip.label}
+\hlstd{stem} \hlkwb{<-} \hlkwd{drop.tip}\hlstd{(BeckLee_tree,} \hlkwc{tip} \hlstd{= crown)}\hlopt{$}\hlstd{tip.label}
+
+\hlcom{## We then have to feed this information in a data.frame with one column}
+\hlstd{factors} \hlkwb{<-} \hlkwd{as.data.frame}\hlstd{(}
+    \hlkwd{matrix}\hlstd{(}\hlkwc{data} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{rep}\hlstd{(}\hlstr{"crown"}\hlstd{,} \hlkwd{length}\hlstd{(crown)),} \hlkwd{rep}\hlstd{(}\hlstr{"stem"}\hlstd{,} \hlkwd{length}\hlstd{(stem))),}
+    \hlkwc{ncol} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{dimnames} \hlstd{=} \hlkwd{list}\hlstd{(}\hlkwd{c}\hlstd{(crown, stem))))}
+
+\hlcom{## Then we can use the customised series function}
+\hlstd{crown_stem} \hlkwb{<-} \hlkwd{cust.series}\hlstd{(BeckLee_mat50, factors)}
+
+\hlcom{## This created a dispRity object containing two series: crown and stem}
+\hlkwd{class}\hlstd{(crown_stem)}
+\end{alltt}
+\begin{verbatim}
+## [1] "dispRity"
+\end{verbatim}
+\begin{alltt}
+\hlstd{crown_stem}
+\end{alltt}
+\begin{verbatim}
+## 2 custom series for 50 taxa. 
+## Series:
+## crown, stem.
+\end{verbatim}
+\begin{alltt}
+\hlcom{## This object contains three elements}
+\hlkwd{names}\hlstd{(crown_stem)}
+\end{alltt}
+\begin{verbatim}
+## [1] "data"   "taxa"   "series"
+\end{verbatim}
+\begin{alltt}
+\hlcom{## With "data" being the list sub-matrices}
+\hlkwd{str}\hlstd{(crown_stem}\hlopt{$}\hlstd{data)}
+\end{alltt}
+\begin{verbatim}
+## List of 2
+##  $ crown: num [1:30, 1:48] 0.3079 0.6531 0.5089 -0.1652 -0.0419 ...
+##   ..- attr(*, "dimnames")=List of 2
+##   .. ..$ : chr [1:30] "Dasypodidae" "Bradypus" "Myrmecophagidae" "Todralestes" ...
+##   .. ..$ : NULL
+##  $ stem : num [1:20, 1:48] -0.739 -0.68 -0.511 -0.477 -0.473 ...
+##   ..- attr(*, "dimnames")=List of 2
+##   .. ..$ : chr [1:20] "Daulestes" "Bulaklestes" "Uchkudukodon" "Kennalestes" ...
+##   .. ..$ : NULL
+\end{verbatim}
+\begin{alltt}
+\hlcom{## "taxa" being the list of taxa in the original ordinated matrix}
+\hlkwd{str}\hlstd{(crown_stem}\hlopt{$}\hlstd{taxa)}
+\end{alltt}
+\begin{verbatim}
+##  chr [1:50] "Cimolestes" "Maelestes" "Batodon" "Bulaklestes" "Daulestes" ...
+\end{verbatim}
+\begin{alltt}
+\hlcom{## "series" containing information on the series type (custom) and names (crown and stem)}
+\hlstd{crown_stem}\hlopt{$}\hlstd{series}
+\end{alltt}
+\begin{verbatim}
+## [1] "custom" "crown"  "stem"
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+Of course, using phylogeny as a factor is just an example.
+Any type of factors can be used in this example (e.g. body mass categories, habitat, diet, etc...).
+Which leads us to another type of category used a lot in disparity-through-time analysis: time!
+
+\subsubsection{Time splitting}
+\texttt{time.series} is a function that allows to split the matrix in different ways using time as a category.
+Two types of time series splitting can be performed with the \dispRity package by using the \texttt{method} option in \texttt{time.series}.
+\begin{enumerate}
+\item the classical discrete time splitting (or time-binning) using \texttt{method = "discrete"};
+\item or a new continuous time splitting (or time-slicing) using \texttt{method = "continuous"}.
+\end{enumerate}
+
+\subsection{Bootstrapping the data}
+
+\subsection{Calculating disparity}
+
+\subsection{Summarising the results}
+
 \section{What's left?}

### 2015-09-29 doc/dispRity.demo.Rnw
dispRity demo draft

@@
 \subsection{Which data?}
-You'll need to do this and that
-%To develop
+In this demo we are going to use the data from Beck and Lee 2014 \cite{beckancient2014}:
 
@@
 
+However, I greatly encourage you to use your own data (after all, that's the point of an \R package).
+You will simply need a ordinated distance matrix (make sure that the number of column is at maximum the number of rows minus one) and a phylogenetic tree (with the tips matching the ordinated matrix).
+Just replace all the \texttt{BeckLee} mentions in this demo by your very own data!
+
 \section{A super quick go through}
-This package allows to:
+This package allows the following:
 \begin{enumerate}
@@
 ## Calculating disparity
-sum_of_ranges <- dispRity(bootstrapped_data, metric = c(sum, range))
+sum_of_ranges <- dispRity(bootstrapped_data, metric = c(sum, ranges))
 
@@
 @
+
+Et voil\`{a}...
 
@@
 \texttt{time.series} is a function that allows to split the matrix in different ways using time as a category.
+This functions needs as input an ordinated matrix and a matching phylogenetic tree.
 Two types of time series splitting can be performed with the \dispRity package by using the \texttt{method} option in \texttt{time.series}.
@@
 \end{enumerate}
+For the time-slicing method details, see Guillerme and Cooper \textit{in prep.} \cite{GuillermeSTD}.
+Once the method is chosen, the function asks to specify the \texttt{time} which can be given as a vector of \texttt{numeric} values for:
+\begin{enumerate}
+\item defining the boundaries of the time bins (when \texttt{method = "discrete"});
+\item defining the time-slicing places (when using \texttt{method = "continuous"}).
+\end{enumerate}
+Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a fix number of equidistant time-bins/slices.
+Additionally, we can also input a data frame containing the first/last occurrence data (\texttt{FADLAD}) for taxa that span over a longer time than the tips/nodes age.
+
+Let's see and example for time-binning (\texttt{method = "discrete"}):
+
+<<discrete.series>>=
+## Let's generate three time bins containing the data present every 40 Ma
+time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
+    method = "discrete", time = c(120, 80, 40, 0))
+
+## Note that the function provides a warning saying that tips where single
+## points in time (no FAD/LAD information). We can fix that by adding the
+## age data for the taxa that have some longer occurrence spans.
+time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
+    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
+
+## To entirely avoid the warning we could collect the occurrence span data
+## for all the taxa but that's not necessary. The function automatically
+## assumes no occurrence span time (i.e. single points in time) for all taxa
+## by default.
+
+time_bins
+@
+
+This generated indeed a list of 3 sub matrices.
+Note that we can also generate equivalent results by just telling the function that we want three time-bins (series) as follow:
+
+<<discrete.series2>>=
+time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete", time = 3)
+@
+We now have three time bins of 44.50368 Ma each. But hey, that's a bit silly, let's stick to bins of 40 Ma.
+Note that we can also ask to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between taxa AND nodes.
+
+For the time-slicing method (\texttt{method = "continuous"}), the idea is pretty similar.
+We just need to provide a matrix that contains the ordinated distance between taxa AND nodes and an evolutionary model via the \texttt{model} argument.
+For now they are four evolutionary models implemented \cite{GuillermeSTD}:
+\begin{enumerate}
+\item \texttt{"acctran"} where the data chosen on each time slice is always the one of the offspring
+\item \texttt{"deltran"} where the data chosen on each time slice is always the one of the descendant
+\item \texttt{"punctuated"} where the data chosen on each time slice is randomly chosen between the offspring or the descendant
+\item \texttt{"gradual"} where the data chosen on each time slice is either the offspring or the descendant depending on branch length
+\end{enumerate}
+
+<<continuous.series>>=
+## Let's generate four time slices every 40 Ma
+time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
+    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
+    FADLAD = BeckLee_ages)
+
+time_slices
+
+## Note that in the same way as for the discrete method, we can also
+## automatically generate the slices
+time.series(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
+    model = "gradual", time = 4)
+## But that gives the same silly results (slicing ~ 44.50368 Ma)
+@
 
 \subsection{Bootstrapping the data}
+To measure how robust is the data, we can perform bootstrap pseudo-replications or rarefaction analysis to make sure the results are not dependant on weird outliers.
+The \texttt{boot.matrix} allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
+The default options will bootstrap the matrix 1000 times with no rarefaction.
+
+<<boot.matrix1>>=
+boot.matrix(data = BeckLee_mat50)
+## As we can wee, the output is also a dispRity object that is summarised
+## automatically, giving information on the data as well as the number of
+## bootstraps and the bootstraps methods.
+@
+
+This function proposes the user to specify the bootstrap algorithm with the \texttt{boot.type} method.
+Currently two algorithms are implemented:
+\begin{enumerate}
+\item \texttt{"full"} where the bootstrapping is entirely stochastic (all the data is bootstrapped)
+\item \texttt{"single"} where only one random taxa is replaced by one other random taxa each pseudo-replication
+\end{enumerate}
+This function also has a \texttt{rarefaction} option to rarefy the data (how unexpected!).
+The default status is \texttt{FALSE} but it can be set to \texttt{TRUE} to fully rarefy the data (i.e.
+remove $n$ taxa for the number of pseudo-replicates, where $n$ varies from the maximum number of taxa present in the dataset to a minimum of 3 taxa).
+It can also be set to a fix \texttt{numeric} value (or a set of numeric values).
+Finally, the last option from this function, \texttt{rm.last.axis}, is whether to remove a certain amount of axis of the ordinated data.
+This option can be logical where \texttt{FALSE} (default) will not remove any axis and \texttt{TRUE} will remove the 5\% last axis or it can be set to a user defined proportion.
+This is a common practice in disparity-through-time analysis but it is hard to justify statistically.
+Here is a series of examples:
+
+<<boot.matrix2>>=
+## Bootstrapping with the single bootstrap method
+boot.matrix(BeckLee_mat50, boot.type = "single")
+## Bootstrapping with the full rarefaction
+boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
+## Or with a set number of rarefaction levels
+boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8,3))
+## And removing the last axis (default)
+boot.matrix(BeckLee_mat50, rm.last.axis = TRUE)
+## Or with a fix value (50%)
+boot.matrix(BeckLee_mat50, rm.last.axis = 0.5)
+@
+
+Okay, that's all well and good but this function can also deal with a list of matrices or with a \dispRity object output from the \texttt{cust.series} or \texttt{time.series} functions.
+Like this:
+
+<<boot.matrix3>>=
+#Let's do a full rarefaction on the crown/stem data for later
+crown_stemBS <- boot.matrix(crown_stem, rarefaction = TRUE)
+## And just bootstrapping for the time binning/slicing
+time_binsBS <- boot.matrix(time_bins)
+time_slicesBS <- boot.matrix(time_slices)
+
+## And just to remind us what's in each objects
+crown_stemBS
+time_binsBS
+time_slicesBS
+@
 
 \subsection{Calculating disparity}
+Until now, all the functions where merely splitting and bootstrapping matrices.
+Now we're going to see the functionalities of the core function of this package: the \dispRity function.
+This function is a modulable function that allow to simply (and quickly!) calculate disparity from a matrix.
+Wait, what's modulable? and back to the introduction, what's disparity?
+
+Because disparity can be measured in many ways, this function is just a tool to measure the disparity \textit{as defined by the user} (and here's where the modulable part comes in).
+In fact, the \dispRity function intakes two main arguments: the data and the disparity metric.
+The disparity metric boils down to a single value that describes the matrix.
+Some are really classic like the sum or the products of the ranges or the variances of the matrix columns \cite{Wills1994} (i.e. the any-o-space dimensions).
+However, many more exist like the median distances between the taxa and the any-o-space centroid \cite{GuillermeSTD}.
+
+One can usually decompose the disparity metrics into two elements:
+\begin{enumerate}
+\item the \textbf{class metric} that is typically a descriptor of the matrix.
+For example describing the ranges of each column in the matrix or the euclidean distance between each row and the centroid of the matrix.
+\item the \textbf{summary metric} that is typically a summarise of the class metric values. For example, the sum of the ranges or the median euclidean distance. 
+\end{enumerate}
+Basically the combination can be infinite between the class and summary metrics.
+For example, people might want to measure the median variance of the axis or the product of the distance from the centroid.
+However, it is probable that some metrics are better to reflect some biological aspects of the any-o-space than others...
+
+In practice, the \dispRity function intakes a pair of class and summary metrics as a definition of disparity.
+Several of these metrics are implemented in other packages (like \texttt{stats::median}, \texttt{base::sum}, etc.) and this package proposes several metrics listed in \texttt{dispRity.metric}:
+
+<<dispRity.metric>>=
+?dispRity.metric
+@
+
+It is even possible to use your very own class and summary metrics!
+This will be actually heavily encouraged and facilitate with the \texttt{make.metric} function in a future version (see \ref{whatsleft}).
+
+To use these metrics pairs in the \dispRity function, it's pretty easy:
+
+<<dispRity1>>=
+dispRity(BeckLee_mat50, metric = c(sum, ranges))
+## Yep, that's it. The order of the metrics does not matter (the function detects
+## automatically which one is the class and the summary metric) so the combinations
+## are high!
+dispRity(BeckLee_mat50, metric = c(sum, variances))
+dispRity(BeckLee_mat50, metric = c(prod, centroids))
+dispRity(BeckLee_mat50, metric = c(mode.val, ranges))
+@
+
+Of course, this function can take an ordinated matrix but, more interesting it can also deal with a list of matrices (from the \texttt{cust.series} or \texttt{time.series} functions for example) or a bootstrapped or/and rarefied matrix (from the \texttt{boot.matrix} function for example).
+
+<<dispRity2>>=
+## For example, let's calculate the sum of ranges from our crown/stem mammals
+dispRity(crown_stem, metric = c(sum, ranges))
+
+## Or even better, from the bootstrapped data
+disp_crown_stemBS <- dispRity(crown_stemBS, metric = c(sum, ranges))
+## Note that the bootstrapped + rarefied data takes some time to run (it's
+## calculating disparity from 46 matrices!)
+disp_time_binsBS <- dispRity(time_binsBS, metric = c(sum, ranges))
+disp_time_slicesBS <- dispRity(time_slicesBS, metric = c(sum, ranges))
+@
+
+Again, note that the results are \dispRity objects and can displayed through the S3 \texttt{print.dispRity} method to remind users what are in these objects.
 
 \subsection{Summarising the results}
+Finally, after calculating disparity, probably the main thing we want to do is actually see the values of the disparity metrics, either in a table or in a plot.
+Well, that again is easily doable by using the S3 \texttt{summary.dispRity} and \texttt{plot.dispRity} functions.
+
+\subsubsection{\texttt{summary} function}
+This function intakes a \dispRity object plus various options namely the \texttt{CI} values for the confidence intervals levels; the \texttt{cent.tend} for the central tendency to use for summarising the results and to "graphical" options which are whether to recall the \dispRity options and how much digits are wanted in the results.
+
+<<summary>>=
+## Let's have a look at the disparity-through-time using in the time-binned data
+summary(disp_time_binsBS)
+
+## The CIs are calculated as 50 and 95 and the central tendency is the mean by default.
+## We can change that easily by specifying different values in the options
+summary(disp_time_binsBS, CI = 25, cent.tend = median, rounding = 0)
+
+## Also, if we happen to have forgotten what was in the disp_time_binsBS object (as in which
+## options where used to modify the original matrix) we can use the recall option
+summary(disp_time_binsBS, recall = TRUE)
+
+## Note that the recall just prints text. The output of summary.dispRity is always a data.frame
+summary_table <- summary(disp_time_binsBS, recall = TRUE)
+class(summary_table)
+
+## Finally we can see the results of the rarefaction analysis for the crown/stem data
+head(summary(disp_crown_stemBS))
+## This outputs a longer table with all the variations of the number of crown/stem taxa
+## from the maximum (30 and 20) to the minimum (3)
+@
+
+\subsubsection{\texttt{plot} function}
+Finally, this function allows to visualise the disparity results in a often nicer fashion than just a table (even though it tends to be the exact same data).
+The \texttt{plot.dispRity} option intakes the same options as \texttt{summary.dispRity} along side with various graphical options described in the function manual.
+Here, let's just have a look a few of these options.
+
+Let's start with plotting the difference in disparity between the crown and the stem mammals.
+For this we are going to simply specify the type of plots using the \texttt{type} option that allows to choose between the \texttt{"continuous"} or \texttt{"discrete"} method.
+In our case, we are interested in looking at the results in a discrete way.
+
+<<plot1, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+plot(disp_crown_stemBS, type = "discrete")
+@
+
+The grey squares represent the confidence intervals (50 in dark grey and 90 in light grey) and the dot represents the mean. As we can see, the disparity seems a bit higher in crown mammals.
+Now that can be due just to the difference in the number of species in both crown and stem mammals.
+By the way, how many taxa where in each group? We can look at that by using the \texttt{diversity} option.
+
+<<plot2, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+plot(disp_crown_stemBS, type = "discrete", diversity = TRUE)
+@
+
+This shows that the number of taxa is higher (30) in crown mammals than in stem ones (20), that is why it is maybe interesting to look at the rarefied data using the \texttt{rarefaction} option.
+This option can be either \texttt{FALSE} (default) for always plotting the maximum number of taxa per series or \texttt{TRUE} to always plot the minimum.
+Additionally, it can be a specific number to plot a specific number of taxa.
+Let say we want both groups to have 20 taxa maximum.
+
+<<plot3, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, rarefaction = 20)
+@
+
+We can see now that the confidence intervals overlap more than previously
+To understand how the number of taxa affects each series, we can use the \texttt{rarefaction = "plot"} function to plot the rarefaction curves.
+
+<<plot4, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, rarefaction = "plot")
+@
+
+The different dashed lines represent the different confidence intervals.
+
+Additionally, for the \texttt{type = "discrete"} option, we might also decide to plot many boxplots (which can be a bit messy).
+The \texttt{discrete\_type} option allows to switch between \texttt{"box"} or \texttt{"line"} for showing the results in one or the other format.
+Let's look at that with the time binned data (even though there is only three time bins).
+
+<<plot5, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+plot(disp_time_binsBS, type = "discrete", discrete_type = "line")
+@
+
+Finally, we can use the \texttt{type = "continuous"} option that comes in handy for plotting continuous data like in the time sliced analysis (obviously...).
+
+<<plot6, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+plot(disp_time_slicesBS, type = "continuous")
+@
+
+I encourage you to play with the graphical options to have some prettier results.
+Note that most of the options from \texttt{plot} can be passed to \texttt{plot.dispRity} via \texttt{...}.
+
+<<plot7, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+op<-par(bty="n")
+plot(disp_time_slicesBS, CI = c(50, 75, 95), cent.tend = median, type = "continuous",
+    diversity = TRUE, col = c("black", "red", "orange", "yellow"), ylab = c("Disparity", "Diversity"),
+    xlab = "Time (in in units from past to present)", time.series = FALSE, main = "Many options...")
+par(op)
+@
 
@@
 
+\bibliographystyle{vancouver}
+\bibliography{References}
+
 \end{document}

### 2015-09-29 doc/dispRity.demo.tex
dispRity demo draft

@@
 \hlcom{## Calculating disparity}
-\hlstd{sum_of_ranges} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(bootstrapped_data,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, range))}
+\hlstd{sum_of_ranges} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(bootstrapped_data,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
 
@@
 ##      series  n  mean  2.5%   25%   75% 97.5%
-## 1 133.51104  3 -0.348 -0.397 -0.350 -0.349 -0.230
-## 2 100.13328 14  0.101 -0.252 -0.046  0.214  0.291
-## 3  66.75552 23 -0.059 -0.163 -0.112  0.031  0.115
-## 4  33.37776 11  0.008 -0.134 -0.086  0.132  0.217
-## 5         0 10  0.009 -0.134 -0.086  0.140  0.187
+## 1 133.51104  3 17.57  0.00 15.51 24.35 24.35
+## 2 100.13328 14 51.76 45.88 50.30 53.78 56.73
+## 3  66.75552 23 64.41 59.89 62.89 66.17 68.61
+## 4  33.37776 11 42.57 36.65 40.95 44.39 46.82
+## 5         0 10 40.53 34.39 38.60 42.76 45.36
 \end{verbatim}
@@
 \end{knitrout}
+
+Et voil\`{a}...
 
@@
 ## 133.51104, 100.13328, 66.75552, 33.37776, 0.
-## Disparity calculated as: sum range for 97 dimensions.
+## Disparity calculated as: sum ranges for 97 dimensions.
 ## Data was split using continuous method.
@@
 \end{enumerate}
+For the time-slicing method details, see Guillerme and Cooper \textit{in prep.} \cite{GuillermeSTD}.
+Once the method is chosen, the function asks to specify the \texttt{time} which can be given as a vector of \texttt{numeric} values for:
+\begin{enumerate}
+\item defining the boundaries of the time bins (when \texttt{method = "discrete"});
+\item defining the time-slicing places (when using \texttt{method = "continuous"}).
+\end{enumerate}
+Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a fix number of equidistant time-bins/slices.
+Additionally, we can also input a data frame containing the first/last occurrence data (\texttt{FADLAD}) for taxa that span over a longer time than the tips/nodes age.
+
+Let's see and example for time-binning (\texttt{method = "discrete"}):
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{## Let's generate three time bins containing the data present every 40 Ma}
+\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat50,} \hlkwc{tree} \hlstd{= BeckLee_tree,}
+    \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{))}
+\end{alltt}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# No FADLAD table has been provided so every tip is assumed to interval single points in time.}}\begin{alltt}
+\hlcom{## Note that the function provides a warning saying that tips where single}
+\hlcom{## points in time (no FAD/LAD information). We can fix that by adding the}
+\hlcom{## age data for the taxa that have some longer occurrence spans.}
+\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat50,} \hlkwc{tree} \hlstd{= BeckLee_tree,}
+    \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{),} \hlkwc{FADLAD} \hlstd{= BeckLee_ages)}
+\end{alltt}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Some tips have FAD/LAD and are assumed to interval single points in time.}}\begin{alltt}
+\hlcom{## To entirely avoid the warning we could collect the occurrence span data}
+\hlcom{## for all the taxa but that's not necessary. The function automatically}
+\hlcom{## assumes no occurrence span time (i.e. single points in time) for all taxa}
+\hlcom{## by default.}
+
+\hlstd{time_bins}
+\end{alltt}
+\begin{verbatim}
+## 3 discrete series for 50 taxa. 
+## Series:
+## 120-80, 80-40, 40-0.
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+This generated indeed a list of 3 sub matrices.
+Note that we can also generate equivalent results by just telling the function that we want three time-bins (series) as follow:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat50,} \hlkwc{tree} \hlstd{= BeckLee_tree,} \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{3}\hlstd{)}
+\end{alltt}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# No FADLAD table has been provided so every tip is assumed to interval single points in time.}}\begin{verbatim}
+## 3 discrete series for 50 taxa. 
+## Series:
+## 133.51104-89.00736, 89.00736-44.50368, 44.50368-0.
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+We now have three time bins of 44.50368 Ma each. But hey, that's a bit silly, let's stick to bins of 40 Ma.
+Note that we can also ask to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between taxa AND nodes.
+
+For the time-slicing method (\texttt{method = "continuous"}), the idea is pretty similar.
+We just need to provide a matrix that contains the ordinated distance between taxa AND nodes and an evolutionary model via the \texttt{model} argument.
+For now they are four evolutionary models implemented \cite{GuillermeSTD}:
+\begin{enumerate}
+\item \texttt{"acctran"} where the data chosen on each time slice is always the one of the offspring
+\item \texttt{"deltran"} where the data chosen on each time slice is always the one of the descendant
+\item \texttt{"punctuated"} where the data chosen on each time slice is randomly chosen between the offspring or the descendant
+\item \texttt{"gradual"} where the data chosen on each time slice is either the offspring or the descendant depending on branch length
+\end{enumerate}
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{## Let's generate four time slices every 40 Ma}
+\hlstd{time_slices} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat99,} \hlkwc{tree} \hlstd{= BeckLee_tree,}
+    \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,} \hlkwc{model} \hlstd{=} \hlstr{"gradual"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{),}
+    \hlkwc{FADLAD} \hlstd{= BeckLee_ages)}
+\end{alltt}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Some tips have FAD/LAD and are assumed to interval single points in time.}}\begin{alltt}
+\hlstd{time_slices}
+\end{alltt}
+\begin{verbatim}
+## 4 continuous series for 99 taxa. 
+## Series:
+## 120, 80, 40, 0.
+\end{verbatim}
+\begin{alltt}
+\hlcom{## Note that in the same way as for the discrete method, we can also}
+\hlcom{## automatically generate the slices}
+\hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat99,} \hlkwc{tree} \hlstd{= BeckLee_tree,} \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
+    \hlkwc{model} \hlstd{=} \hlstr{"gradual"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{4}\hlstd{)}
+\end{alltt}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# No FADLAD table has been provided so every tip is assumed to interval single points in time.}}\begin{verbatim}
+## 4 continuous series for 99 taxa. 
+## Series:
+## 133.51104, 89.00736, 44.50368, 0.
+\end{verbatim}
+\begin{alltt}
+\hlcom{## But that gives the same silly results (slicing ~ 44.50368 Ma)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
 \subsection{Bootstrapping the data}
+To measure how robust is the data, we can perform bootstrap pseudo-replications or rarefaction analysis to make sure the results are not dependant on weird outliers.
+The \texttt{boot.matrix} allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
+The default options will bootstrap the matrix 1000 times with no rarefaction.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{boot.matrix}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat50)}
+\end{alltt}
+\begin{verbatim}
+## Bootstrapped ordinated matrix with 50 taxa. 
+## 1 unnamed series.
+## Data was bootstrapped 1000 times, using the full bootstrap method.
+\end{verbatim}
+\begin{alltt}
+\hlcom{## As we can wee, the output is also a dispRity object that is summarised}
+\hlcom{## automatically, giving information on the data as well as the number of}
+\hlcom{## bootstraps and the bootstraps methods.}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+This function proposes the user to specify the bootstrap algorithm with the \texttt{boot.type} method.
+Currently two algorithms are implemented:
+\begin{enumerate}
+\item \texttt{"full"} where the bootstrapping is entirely stochastic (all the data is bootstrapped)
+\item \texttt{"single"} where only one random taxa is replaced by one other random taxa each pseudo-replication
+\end{enumerate}
+This function also has a \texttt{rarefaction} option to rarefy the data (how unexpected!).
+The default status is \texttt{FALSE} but it can be set to \texttt{TRUE} to fully rarefy the data (i.e.
+remove $n$ taxa for the number of pseudo-replicates, where $n$ varies from the maximum number of taxa present in the dataset to a minimum of 3 taxa).
+It can also be set to a fix \texttt{numeric} value (or a set of numeric values).
+Finally, the last option from this function, \texttt{rm.last.axis}, is whether to remove a certain amount of axis of the ordinated data.
+This option can be logical where \texttt{FALSE} (default) will not remove any axis and \texttt{TRUE} will remove the 5\% last axis or it can be set to a user defined proportion.
+This is a common practice in disparity-through-time analysis but it is hard to justify statistically.
+Here is a series of examples:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{## Bootstrapping with the single bootstrap method}
+\hlkwd{boot.matrix}\hlstd{(BeckLee_mat50,} \hlkwc{boot.type} \hlstd{=} \hlstr{"single"}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+## Bootstrapped ordinated matrix with 50 taxa. 
+## 1 unnamed series.
+## Data was bootstrapped 1000 times, using the single bootstrap method.
+\end{verbatim}
+\begin{alltt}
+\hlcom{## Bootstrapping with the full rarefaction}
+\hlkwd{boot.matrix}\hlstd{(BeckLee_mat50,} \hlkwc{bootstraps} \hlstd{=} \hlnum{20}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+## Bootstrapped ordinated matrix with 50 taxa. 
+## 1 unnamed series.
+## Data was bootstrapped 20 times, using the full bootstrap method.
+## Data was fully rarefied (down to 3 taxa).
+\end{verbatim}
+\begin{alltt}
+\hlcom{## Or with a set number of rarefaction levels}
+\hlkwd{boot.matrix}\hlstd{(BeckLee_mat50,} \hlkwc{bootstraps} \hlstd{=} \hlnum{20}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{6}\hlopt{:}\hlnum{8}\hlstd{,}\hlnum{3}\hlstd{))}
+\end{alltt}
+\begin{verbatim}
+## Bootstrapped ordinated matrix with 50 taxa. 
+## 1 unnamed series.
+## Data was bootstrapped 20 times, using the full bootstrap method.
+## Data was rarefied with a maximum of 6, 7, 8 and 3 taxa.
+\end{verbatim}
+\begin{alltt}
+\hlcom{## And removing the last axis (default)}
+\hlkwd{boot.matrix}\hlstd{(BeckLee_mat50,} \hlkwc{rm.last.axis} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+## Bootstrapped ordinated matrix with 50 taxa. 
+## 1 unnamed series.
+## Data was bootstrapped 1000 times, using the full bootstrap method. 
+## The 6 last axis have been removed from the original data.
+\end{verbatim}
+\begin{alltt}
+\hlcom{## Or with a fix value (50%)}
+\hlkwd{boot.matrix}\hlstd{(BeckLee_mat50,} \hlkwc{rm.last.axis} \hlstd{=} \hlnum{0.5}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+## Bootstrapped ordinated matrix with 50 taxa. 
+## 1 unnamed series.
+## Data was bootstrapped 1000 times, using the full bootstrap method. 
+## The 35 last axis have been removed from the original data.
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+Okay, that's all well and good but this function can also deal with a list of matrices or with a \dispRity object output from the \texttt{cust.series} or \texttt{time.series} functions.
+Like this:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{#Let's do a full rarefaction on the crown/stem data for later}
+\hlstd{crown_stemBS} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(crown_stem,} \hlkwc{rarefaction} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\hlcom{## And just bootstrapping for the time binning/slicing}
+\hlstd{time_binsBS} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(time_bins)}
+\hlstd{time_slicesBS} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(time_slices)}
+
+\hlcom{## And just to remind us what's in each objects}
+\hlstd{crown_stemBS}
+\end{alltt}
+\begin{verbatim}
+## Bootstrapped ordinated matrix with 50 taxa. 
+## Series:
+## crown, stem.
+## Data was split using custom method.
+## Data was bootstrapped 1000 times, using the full bootstrap method.
+## Data was fully rarefied (down to 3 taxa).
+\end{verbatim}
+\begin{alltt}
+\hlstd{time_binsBS}
+\end{alltt}
+\begin{verbatim}
+## Bootstrapped ordinated matrix with 50 taxa. 
+## Series:
+## 120-80, 80-40, 40-0.
+## Data was split using discrete method.
+## Data was bootstrapped 1000 times, using the full bootstrap method.
+\end{verbatim}
+\begin{alltt}
+\hlstd{time_slicesBS}
+\end{alltt}
+\begin{verbatim}
+## Bootstrapped ordinated matrix with 99 taxa. 
+## Series:
+## 120, 80, 40, 0.
+## Data was split using continuous method.
+## Data was bootstrapped 1000 times, using the full bootstrap method.
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
 
 \subsection{Calculating disparity}
+Until now, all the functions where merely splitting and bootstrapping matrices.
+Now we're going to see the functionalities of the core function of this package: the \dispRity function.
+This function is a modulable function that allow to simply (and quickly!) calculate disparity from a matrix.
+Wait, what's modulable? and back to the introduction, what's disparity?
+
+Because disparity can be measured in many ways, this function is just a tool to measure the disparity \textit{as defined by the user} (and here's where the modulable part comes in).
+In fact, the \dispRity function intakes two main arguments: the data and the disparity metric.
+The disparity metric boils down to a single value that describes the matrix.
+Some are really classic like the sum or the products of the ranges or the variances of the matrix columns \cite{Wills1994} (i.e. the any-o-space dimensions).
+However, many more exist like the median distances between the taxa and the any-o-space centroid \cite{GuillermeSTD}.
+
+One can usually decompose the disparity metrics into two elements:
+\begin{enumerate}
+\item the \textbf{class metric} that is typically a descriptor of the matrix.
+For example describing the ranges of each column in the matrix or the euclidean distance between each row and the centroid of the matrix.
+\item the \textbf{summary metric} that is typically a summarise of the class metric values. For example, the sum of the ranges or the median euclidean distance. 
+\end{enumerate}
+Basically the combination can be infinite between the class and summary metrics.
+For example, people might want to measure the median variance of the axis or the product of the distance from the centroid.
+However, it is probable that some metrics are better to reflect some biological aspects of the any-o-space than others...
+
+In practice, the \dispRity function intakes a pair of class and summary metrics as a definition of disparity.
+Several of these metrics are implemented in other packages (like \texttt{stats::median}, \texttt{base::sum}, etc.) and this package proposes several metrics listed in \texttt{dispRity.metric}:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlopt{?}\hlstd{dispRity.metric}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+It is even possible to use your very own class and summary metrics!
+This will be actually heavily encouraged and facilitate with the \texttt{make.metric} function in a future version (see \ref{whatsleft}).
+
+To use these metrics pairs in the \dispRity function, it's pretty easy:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{dispRity}\hlstd{(BeckLee_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
+\end{alltt}
+\begin{verbatim}
+## Disparity measurements across 1 series for 50 taxa. 
+## 1 unnamed series.
+## Disparity calculated as: sum ranges for 48 dimensions.
+\end{verbatim}
+\begin{alltt}
+\hlcom{## Yep, that's it. The order of the metrics does not matter (the function detects}
+\hlcom{## automatically which one is the class and the summary metric) so the combinations}
+\hlcom{## are high!}
+\hlkwd{dispRity}\hlstd{(BeckLee_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, variances))}
+\end{alltt}
+\begin{verbatim}
+## Disparity measurements across 1 series for 50 taxa. 
+## 1 unnamed series.
+## Disparity calculated as: sum variances for 48 dimensions.
+\end{verbatim}
+\begin{alltt}
+\hlkwd{dispRity}\hlstd{(BeckLee_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(prod, centroids))}
+\end{alltt}
+\begin{verbatim}
+## Disparity measurements across 1 series for 50 taxa. 
+## 1 unnamed series.
+## Disparity calculated as: prod centroids for 48 dimensions.
+\end{verbatim}
+\begin{alltt}
+\hlkwd{dispRity}\hlstd{(BeckLee_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(mode.val, ranges))}
+\end{alltt}
+\begin{verbatim}
+## Disparity measurements across 1 series for 50 taxa. 
+## 1 unnamed series.
+## Disparity calculated as: mode.val ranges for 48 dimensions.
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+Of course, this function can take an ordinated matrix but, more interesting it can also deal with a list of matrices (from the \texttt{cust.series} or \texttt{time.series} functions for example) or a bootstrapped or/and rarefied matrix (from the \texttt{boot.matrix} function for example).
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{## For example, let's calculate the sum of ranges from our crown/stem mammals}
+\hlkwd{dispRity}\hlstd{(crown_stem,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
+\end{alltt}
+\begin{verbatim}
+## Disparity measurements across 2 series for 50 taxa. 
+## Series:
+## crown, stem.
+## Disparity calculated as: sum ranges for 48 dimensions.
+## Data was split using custom method.
+\end{verbatim}
+\begin{alltt}
+\hlcom{## Or even better, from the bootstrapped data}
+\hlstd{disp_crown_stemBS} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(crown_stemBS,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
+\hlcom{## Note that the bootstrapped + rarefied data takes some time to run (it's}
+\hlcom{## calculating disparity from 46 matrices!)}
+\hlstd{disp_time_binsBS} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(time_binsBS,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
+\hlstd{disp_time_slicesBS} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(time_slicesBS,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+Again, note that the results are \dispRity objects and can displayed through the S3 \texttt{print.dispRity} method to remind users what are in these objects.
 
 \subsection{Summarising the results}
+Finally, after calculating disparity, probably the main thing we want to do is actually see the values of the disparity metrics, either in a table or in a plot.
+Well, that again is easily doable by using the S3 \texttt{summary.dispRity} and \texttt{plot.dispRity} functions.
+
+\subsubsection{\texttt{summary} function}
+This function intakes a \dispRity object plus various options namely the \texttt{CI} values for the confidence intervals levels; the \texttt{cent.tend} for the central tendency to use for summarising the results and to "graphical" options which are whether to recall the \dispRity options and how much digits are wanted in the results.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{## Let's have a look at the disparity-through-time using in the time-binned data}
+\hlkwd{summary}\hlstd{(disp_time_binsBS)}
+\end{alltt}
+\begin{verbatim}
+##   series  n  mean  2.5%   25%   75% 97.5%
+## 1 120-80  8 20.20 16.58 19.28 21.64 23.39
+## 2  80-40 27 33.97 31.65 33.29 34.73 35.78
+## 3   40-0 16 29.00 26.24 28.19 29.92 31.30
+\end{verbatim}
+\begin{alltt}
+\hlcom{## The CIs are calculated as 50 and 95 and the central tendency is the mean by default.}
+\hlcom{## We can change that easily by specifying different values in the options}
+\hlkwd{summary}\hlstd{(disp_time_binsBS,} \hlkwc{CI} \hlstd{=} \hlnum{25}\hlstd{,} \hlkwc{cent.tend} \hlstd{= median,} \hlkwc{rounding} \hlstd{=} \hlnum{0}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+##   series  n median 37.5% 62.5%
+## 1 120-80  8     20    20    21
+## 2  80-40 27     34    34    34
+## 3   40-0 16     29    29    29
+\end{verbatim}
+\begin{alltt}
+\hlcom{## Also, if we happen to have forgotten what was in the disp_time_binsBS object (as in which}
+\hlcom{## options where used to modify the original matrix) we can use the recall option}
+\hlkwd{summary}\hlstd{(disp_time_binsBS,} \hlkwc{recall} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+## Disparity calculated as: sum ranges for 48 dimensions.
+## Data was split using discrete method.
+## Data was bootstrapped 1000 times, using the full bootstrap method.
+##   series  n  mean  2.5%   25%   75% 97.5%
+## 1 120-80  8 20.20 16.58 19.28 21.64 23.39
+## 2  80-40 27 33.97 31.65 33.29 34.73 35.78
+## 3   40-0 16 29.00 26.24 28.19 29.92 31.30
+\end{verbatim}
+\begin{alltt}
+\hlcom{## Note that the recall just prints text. The output of summary.dispRity is always a data.frame}
+\hlstd{summary_table} \hlkwb{<-} \hlkwd{summary}\hlstd{(disp_time_binsBS,} \hlkwc{recall} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+## Disparity calculated as: sum ranges for 48 dimensions.
+## Data was split using discrete method.
+## Data was bootstrapped 1000 times, using the full bootstrap method.
+\end{verbatim}
+\begin{alltt}
+\hlkwd{class}\hlstd{(summary_table)}
+\end{alltt}
+\begin{verbatim}
+## [1] "data.frame"
+\end{verbatim}
+\begin{alltt}
+\hlcom{## Finally we can see the results of the rarefaction analysis for the crown/stem data}
+\hlkwd{head}\hlstd{(}\hlkwd{summary}\hlstd{(disp_crown_stemBS))}
+\end{alltt}
+\begin{verbatim}
+##   series n  mean  2.5%   25%   75% 97.5%
+## 1  crown 3 15.17 10.04 15.16 16.04 16.64
+## 2  crown 4 18.39 14.83 18.34 19.42 19.95
+## 3  crown 5 20.83 18.04 19.68 21.83 22.42
+## 4  crown 6 22.70 19.20 21.75 23.76 24.41
+## 5  crown 7 24.15 21.01 23.47 25.29 26.11
+## 6  crown 8 25.44 22.06 24.77 26.48 27.43
+\end{verbatim}
+\begin{alltt}
+\hlcom{## This outputs a longer table with all the variations of the number of crown/stem taxa}
+\hlcom{## from the maximum (30 and 20) to the minimum (3)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\subsubsection{\texttt{plot} function}
+Finally, this function allows to visualise the disparity results in a often nicer fashion than just a table (even though it tends to be the exact same data).
+The \texttt{plot.dispRity} option intakes the same options as \texttt{summary.dispRity} along side with various graphical options described in the function manual.
+Here, let's just have a look a few of these options.
+
+Let's start with plotting the difference in disparity between the crown and the stem mammals.
+For this we are going to simply specify the type of plots using the \texttt{type} option that allows to choose between the \texttt{"continuous"} or \texttt{"discrete"} method.
+In our case, we are interested in looking at the results in a discrete way.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot1-1}
+
+}
+
+\end{knitrout}
+
+The grey squares represent the confidence intervals (50 in dark grey and 90 in light grey) and the dot represents the mean. As we can see, the disparity seems a bit higher in crown mammals.
+Now that can be due just to the difference in the number of species in both crown and stem mammals.
+By the way, how many taxa where in each group? We can look at that by using the \texttt{diversity} option.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot2-1}
+
+}
+
+\end{knitrout}
+
+This shows that the number of taxa is higher (30) in crown mammals than in stem ones (20), that is why it is maybe interesting to look at the rarefied data using the \texttt{rarefaction} option.
+This option can be either \texttt{FALSE} (default) for always plotting the maximum number of taxa per series or \texttt{TRUE} to always plot the minimum.
+Additionally, it can be a specific number to plot a specific number of taxa.
+Let say we want both groups to have 20 taxa maximum.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlnum{20}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot3-1}
+
+}
+
+\end{knitrout}
+
+We can see now that the confidence intervals overlap more than previously
+To understand how the number of taxa affects each series, we can use the \texttt{rarefaction = "plot"} function to plot the rarefaction curves.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlstr{"plot"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot4-1}
+
+}
+
+\end{knitrout}
+
+The different dashed lines represent the different confidence intervals.
+
+Additionally, for the \texttt{type = "discrete"} option, we might also decide to plot many boxplots (which can be a bit messy).
+The \texttt{discrete\_type} option allows to switch between \texttt{"box"} or \texttt{"line"} for showing the results in one or the other format.
+Let's look at that with the time binned data (even though there is only three time bins).
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{plot}\hlstd{(disp_time_binsBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{discrete_type} \hlstd{=} \hlstr{"line"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot5-1}
+
+}
+
+\end{knitrout}
+
+Finally, we can use the \texttt{type = "continuous"} option that comes in handy for plotting continuous data like in the time sliced analysis (obviously...).
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{plot}\hlstd{(disp_time_slicesBS,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot6-1}
+
+}
+
+\end{knitrout}
+
+I encourage you to play with the graphical options to have some prettier results.
+Note that most of the options from \texttt{plot} can be passed to \texttt{plot.dispRity} via \texttt{...}.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{op}\hlkwb{<-}\hlkwd{par}\hlstd{(}\hlkwc{bty}\hlstd{=}\hlstr{"n"}\hlstd{)}
+\hlkwd{plot}\hlstd{(disp_time_slicesBS,} \hlkwc{CI} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{50}\hlstd{,} \hlnum{75}\hlstd{,} \hlnum{95}\hlstd{),} \hlkwc{cent.tend} \hlstd{= median,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
+    \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"black"}\hlstd{,} \hlstr{"red"}\hlstd{,} \hlstr{"orange"}\hlstd{,} \hlstr{"yellow"}\hlstd{),} \hlkwc{ylab} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"Disparity"}\hlstd{,} \hlstr{"Diversity"}\hlstd{),}
+    \hlkwc{xlab} \hlstd{=} \hlstr{"Time (in in units from past to present)"}\hlstd{,} \hlkwc{time.series} \hlstd{=} \hlnum{FALSE}\hlstd{,} \hlkwc{main} \hlstd{=} \hlstr{"Many options..."}\hlstd{)}
+\hlkwd{par}\hlstd{(op)}
+\end{alltt}
+\end{kframe}
+
+{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot7-1}
+
+}
+
+\end{knitrout}
 

### 2015-09-30 doc/dispRity.demo.Rnw
demo vignette updated

@@
 
-<<Quick go, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+<<Quick go, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
 ## Splitting the data
@@
 <<boot.matrix3>>=
-#Let's do a full rarefaction on the crown/stem data for later
+## Let's do a full rarefaction on the crown/stem data for later
 crown_stemBS <- boot.matrix(crown_stem, rarefaction = TRUE)
@@
 ## calculating disparity from 46 matrices!)
-disp_time_binsBS <- dispRity(time_binsBS, metric = c(sum, ranges))
-disp_time_slicesBS <- dispRity(time_slicesBS, metric = c(sum, ranges))
+
+## Let's calculate a different metric for the disparity-through-time analysis
+disp_time_binsBS <- dispRity(time_binsBS, metric = c(median, centroids))
+disp_time_slicesBS <- dispRity(time_slicesBS, metric = c(median, centroids))
 @
@@
 
-<<plot1, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
-plot(disp_crown_stemBS, type = "discrete")
+<<plot1, fig.width=15, fig.height=5, out.width='1\\linewidth'>>=
+## Graphical options
+quartz(width = 15, height = 5) ; par(mfrow = (c(1,3)), bty = "n")
+
+## Plotting the score for both groups
+plot(disp_crown_stemBS, type = "discrete" , ylim = c(25,40))
+
+## Same one with the number of taxa for each data set
+plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, ylim = c(25,40))
+
+## Same one with rarefied data (note how that affects the diversity data as well!)
+plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, rarefaction = 20, ylim = c(25,40))
 @
@@
 
-<<plot2, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
-plot(disp_crown_stemBS, type = "discrete", diversity = TRUE)
-@
-
 This shows that the number of taxa is higher (30) in crown mammals than in stem ones (20), that is why it is maybe interesting to look at the rarefied data using the \texttt{rarefaction} option.
@@
 
-<<plot3, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
-plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, rarefaction = 20)
-@
-
 We can see now that the confidence intervals overlap more than previously
@@
 
-<<plot4, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+<<plot2, fig.width=5, fig.height=8, out.width='0.5\\linewidth'>>=
+## Graphical options
+par(bty = "n")
+
+## The rarefaction curves
 plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, rarefaction = "plot")
+## Note that the function automatically draws the curves for each series and splits
+## the plot screen accordingly.
 @
@@
 
-<<plot5, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+<<plot3, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
+## Graphical options
+quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
+
+## The disparity-through-time data in a time-binned way
 plot(disp_time_binsBS, type = "discrete", discrete_type = "line")
+
+## The disparity-through-time data in a time-sliced ways
+plot(disp_time_slicesBS, type = "continuous")
+
 @
@@
 
-<<plot6, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
-plot(disp_time_slicesBS, type = "continuous")
+<<plot4, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
+## Graphical options
+par(bty = "n")
+
+## First, plotting the continuous data
+plot(disp_time_slicesBS, type = "continuous", ylim=c(0.5,2))
+
+## Second, adding the discrete data
+par(new = TRUE)
+plot(disp_time_binsBS, type = "discrete", discrete_type = "line", ylim=c(0.5,2), xlab="", ylab="")
 @
@@
 
-<<plot7, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
-op<-par(bty="n")
+<<plot5, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
+## Graphical options
+par(bty = "n")
+
+## A plot with many options!
 plot(disp_time_slicesBS, CI = c(50, 75, 95), cent.tend = median, type = "continuous",
@@
     xlab = "Time (in in units from past to present)", time.series = FALSE, main = "Many options...")
-par(op)
 @
@@
 \label{whatsleft}
+\subsection{Some statistics}
+\subsection{Users defined metrics}
+\subsection{More user defined stuff}
 

### 2015-09-30 doc/dispRity.demo.tex
demo vignette updated

@@
 ##      series  n  mean  2.5%   25%   75% 97.5%
-## 1 133.51104  3 17.57  0.00 15.51 24.35 24.35
-## 2 100.13328 14 51.76 45.88 50.30 53.78 56.73
-## 3  66.75552 23 64.41 59.89 62.89 66.17 68.61
-## 4  33.37776 11 42.57 36.65 40.95 44.39 46.82
-## 5         0 10 40.53 34.39 38.60 42.76 45.36
+## 1 133.51104  3 17.81  0.00 15.51 24.35 24.35
+## 2 100.13328 14 52.05 45.80 49.90 54.50 57.67
+## 3  66.75552 23 64.80 59.23 63.20 66.57 69.25
+## 4  33.37776 11 42.95 37.94 41.40 44.77 47.43
+## 5         0 10 40.94 35.13 39.22 42.69 46.13
 \end{verbatim}
@@
 
-{\centering \includegraphics[width=.8\linewidth]{figure/minimal-Quick_go-1}
+{\centering \includegraphics[width=.5\linewidth]{figure/minimal-Quick_go-1}
 
@@
 \begin{alltt}
-\hlcom{#Let's do a full rarefaction on the crown/stem data for later}
+\hlcom{## Let's do a full rarefaction on the crown/stem data for later}
 \hlstd{crown_stemBS} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(crown_stem,} \hlkwc{rarefaction} \hlstd{=} \hlnum{TRUE}\hlstd{)}
@@
 \hlcom{## calculating disparity from 46 matrices!)}
-\hlstd{disp_time_binsBS} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(time_binsBS,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
-\hlstd{disp_time_slicesBS} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(time_slicesBS,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
+
+\hlcom{## Let's calculate a different metric for the disparity-through-time analysis}
+\hlstd{disp_time_binsBS} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(time_binsBS,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(median, centroids))}
+\hlstd{disp_time_slicesBS} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(time_slicesBS,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(median, centroids))}
 \end{alltt}
@@
 ##   series  n  mean  2.5%   25%   75% 97.5%
-## 1 120-80  8 20.20 16.58 19.28 21.64 23.39
-## 2  80-40 27 33.97 31.65 33.29 34.73 35.78
-## 3   40-0 16 29.00 26.24 28.19 29.92 31.30
+## 1 120-80  8 1.114 0.915 1.080 1.167 1.235
+## 2  80-40 27 1.322 1.274 1.308 1.338 1.358
+## 3   40-0 16 1.324 1.248 1.306 1.345 1.372
 \end{verbatim}
@@
 ##   series  n median 37.5% 62.5%
-## 1 120-80  8     20    20    21
-## 2  80-40 27     34    34    34
-## 3   40-0 16     29    29    29
+## 1 120-80  8      1     1     1
+## 2  80-40 27      1     1     1
+## 3   40-0 16      1     1     1
 \end{verbatim}
@@
 \begin{verbatim}
-## Disparity calculated as: sum ranges for 48 dimensions.
+## Disparity calculated as: median centroids for 48 dimensions.
 ## Data was split using discrete method.
@@
 ##   series n  mean  2.5%   25%   75% 97.5%
-## 1  crown 3 15.17 10.04 15.16 16.04 16.64
-## 2  crown 4 18.39 14.83 18.34 19.42 19.95
-## 3  crown 5 20.83 18.04 19.68 21.83 22.42
-## 4  crown 6 22.70 19.20 21.75 23.76 24.41
-## 5  crown 7 24.15 21.01 23.47 25.29 26.11
-## 6  crown 8 25.44 22.06 24.77 26.48 27.43
+## 1  crown 3 15.19 10.19 15.17 16.08 16.66
+## 2  crown 4 18.47 14.97 18.40 19.41 19.89
+## 3  crown 5 20.83 17.99 19.66 21.81 22.46
+## 4  crown 6 22.60 18.84 21.68 23.77 24.51
+## 5  crown 7 24.23 21.16 23.49 25.32 26.15
+## 6  crown 8 25.46 22.40 24.82 26.51 27.42
 \end{verbatim}
@@
 \begin{alltt}
-\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{)}
+\hlcom{## Graphical options}
+\hlkwd{quartz}\hlstd{(}\hlkwc{width} \hlstd{=} \hlnum{15}\hlstd{,} \hlkwc{height} \hlstd{=} \hlnum{5}\hlstd{) ;} \hlkwd{par}\hlstd{(}\hlkwc{mfrow} \hlstd{= (}\hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,}\hlnum{3}\hlstd{)),} \hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
+
+\hlcom{## Plotting the score for both groups}
+\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"} \hlstd{,} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{25}\hlstd{,}\hlnum{40}\hlstd{))}
+
+\hlcom{## Same one with the number of taxa for each data set}
+\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{25}\hlstd{,}\hlnum{40}\hlstd{))}
+
+\hlcom{## Same one with rarefied data (note how that affects the diversity data as well!)}
+\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlnum{20}\hlstd{,} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{25}\hlstd{,}\hlnum{40}\hlstd{))}
 \end{alltt}
@@
 
-{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot1-1}
+{\centering \includegraphics[width=1\linewidth]{figure/minimal-plot1-1}
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot2-1}
-
-}
-
-\end{knitrout}
-
 This shows that the number of taxa is higher (30) in crown mammals than in stem ones (20), that is why it is maybe interesting to look at the rarefied data using the \texttt{rarefaction} option.
@@
 Let say we want both groups to have 20 taxa maximum.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlnum{20}\hlstd{)}
-\end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot3-1}
-
-}
-
-\end{knitrout}
 
@@
 \begin{alltt}
+\hlcom{## Graphical options}
+\hlkwd{par}\hlstd{(}\hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
+
+\hlcom{## The rarefaction curves}
 \hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlstr{"plot"}\hlstd{)}
+\hlcom{## Note that the function automatically draws the curves for each series and splits}
+\hlcom{## the plot screen accordingly.}
 \end{alltt}
@@
 
-{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot4-1}
+{\centering \includegraphics[width=0.5\linewidth]{figure/minimal-plot2-1}
 
@@
 \begin{alltt}
+\hlcom{## Graphical options}
+\hlkwd{quartz}\hlstd{(}\hlkwc{width} \hlstd{=} \hlnum{10}\hlstd{,} \hlkwc{height} \hlstd{=} \hlnum{5}\hlstd{) ;} \hlkwd{par}\hlstd{(}\hlkwc{mfrow} \hlstd{= (}\hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,}\hlnum{2}\hlstd{)),} \hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
+
+\hlcom{## The disparity-through-time data in a time-binned way}
 \hlkwd{plot}\hlstd{(disp_time_binsBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{discrete_type} \hlstd{=} \hlstr{"line"}\hlstd{)}
+
+\hlcom{## The disparity-through-time data in a time-sliced ways}
+\hlkwd{plot}\hlstd{(disp_time_slicesBS,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{)}
 \end{alltt}
@@
 
-{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot5-1}
+{\centering \includegraphics[width=1\linewidth]{figure/minimal-plot3-1}
 
@@
 \begin{alltt}
-\hlkwd{plot}\hlstd{(disp_time_slicesBS,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{)}
+\hlcom{## Graphical options}
+\hlkwd{par}\hlstd{(}\hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
+
+\hlcom{## First, plotting the continuous data}
+\hlkwd{plot}\hlstd{(disp_time_slicesBS,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{,} \hlkwc{ylim}\hlstd{=}\hlkwd{c}\hlstd{(}\hlnum{0.5}\hlstd{,}\hlnum{2}\hlstd{))}
+
+\hlcom{## Second, adding the discrete data}
+\hlkwd{par}\hlstd{(}\hlkwc{new} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\hlkwd{plot}\hlstd{(disp_time_binsBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{discrete_type} \hlstd{=} \hlstr{"line"}\hlstd{,} \hlkwc{ylim}\hlstd{=}\hlkwd{c}\hlstd{(}\hlnum{0.5}\hlstd{,}\hlnum{2}\hlstd{),} \hlkwc{xlab}\hlstd{=}\hlstr{""}\hlstd{,} \hlkwc{ylab}\hlstd{=}\hlstr{""}\hlstd{)}
 \end{alltt}
@@
 
-{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot6-1}
+{\centering \includegraphics[width=.5\linewidth]{figure/minimal-plot4-1}
 
@@
 \begin{alltt}
-\hlstd{op}\hlkwb{<-}\hlkwd{par}\hlstd{(}\hlkwc{bty}\hlstd{=}\hlstr{"n"}\hlstd{)}
+\hlcom{## Graphical options}
+\hlkwd{par}\hlstd{(}\hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
+
+\hlcom{## A plot with many options!}
 \hlkwd{plot}\hlstd{(disp_time_slicesBS,} \hlkwc{CI} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{50}\hlstd{,} \hlnum{75}\hlstd{,} \hlnum{95}\hlstd{),} \hlkwc{cent.tend} \hlstd{= median,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
@@
     \hlkwc{xlab} \hlstd{=} \hlstr{"Time (in in units from past to present)"}\hlstd{,} \hlkwc{time.series} \hlstd{=} \hlnum{FALSE}\hlstd{,} \hlkwc{main} \hlstd{=} \hlstr{"Many options..."}\hlstd{)}
-\hlkwd{par}\hlstd{(op)}
 \end{alltt}
@@
 
-{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot7-1}
+{\centering \includegraphics[width=.5\linewidth]{figure/minimal-plot5-1}
 

### 2015-10-01 doc/dispRity.demo.tex
Updated demo vignette

@@
 
-\dispRity is a package for calculating disparity in \R.
-What is disparity? Well that's a more complex question...
+\dispRity is a package for calculating disparity in \texttt{R}.
 To keep it short, this package allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.
-% More explanations to put here?
+These ordinated matrices can be called the morphospace (when using morphometric data) or the cladisto-space (for cladistic data) and will represent the total spread of the forms available in the dataset.
+However, keep in mind that if you are using other type of data, say ecological data for example, then the ordinated matrix will represent the eco-space which represents the total spread of ecological traits available in the dataset.
+In the rest of this demo, I will sometimes refer to the ordinated matrix as the any-o-space where you can replace \textit{any} by any prefix you want your space to be based on!
 
 \subsection{Installation}
-You can install this package pretty easily if you use the latest version of \R and \texttt{devtools}.
-Just copy past the following
+You can install this package easily if you use the latest version of \R and \texttt{devtools}.
+
 \begin{knitrout}
@@
 \hlkwd{library}\hlstd{(devtools)}
-\hlkwd{install_github}\hlstd{(}\hlstr{"TGuillerme/dispRity"}\hlstd{,} \hlkwc{ref}\hlstd{=}\hlstr{"release"}\hlstd{)}
+\hlkwd{install_github}\hlstd{(}\hlstr{"TGuillerme/dispRity"}\hlstd{,} \hlkwc{ref} \hlstd{=} \hlstr{"release"}\hlstd{)}
 \hlkwd{library}\hlstd{(dispRity)}
@@
 
-Note that we use here the \texttt{release} branch which is version 0.1.0.
-If you want the piping-hot (and full of bugs) version, change the option \texttt{ref="release"} to \texttt{ref="master"}.
-
-This package depends heavily on the \texttt{ape} package as well as the nice function \texttt{timeSliceTree::paleotree}.
-However, I recommend to also install the packages \texttt{geomorph} or \texttt{Claddis} if you want to try this example on your own morphometric or cladistic data.
+Note that we use the \texttt{release} branch here which is version 0.1.0.
+If you want the piping-hot (and full of bugs) version, change the option \texttt{ref = "release"} to \texttt{ref = "master"}.
+This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.
+Future version will just require the specific function but this version (0.1.0) will install \texttt{paleotree} as a dependency (which is not that bad since it is a super useful package anyway!).
+Additionally, I warmly recommend to install the packages \texttt{geomorph} or/and \texttt{Claddis} if you want to try this demo with your very own ordinated morphometric or cladistic data.
 
 \subsection{Which data?}
-In this demo we are going to use the data from Beck and Lee 2014 \cite{beckancient2014}:
+In this demo we are going to use a subset of the ordinated cladistic data from Beck and Lee 2014 \cite{beckancient2014}:
 
@@
 \begin{alltt}
-\hlcom{## A list of first and last occurence data for some fossils}
+\hlcom{## A list of first and last occurrences data for some fossils}
 \hlkwd{data}\hlstd{(BeckLee_ages)}
@@
 However, I greatly encourage you to use your own data (after all, that's the point of an \R package).
-You will simply need a ordinated distance matrix (make sure that the number of column is at maximum the number of rows minus one) and a phylogenetic tree (with the tips matching the ordinated matrix).
-Just replace all the \texttt{BeckLee} mentions in this demo by your very own data!
+You will simply need (1) an ordinated distance matrix: make sure that the number of column is at maximum the number of rows minus one.
+And secondly (2) a phylogenetic tree with the tips matching the ordinated matrix and a \texttt{\$root.time} element for determining the age of the tree.
+You can use the function \texttt{tree.age} provided in this package but not described in this demo for generating the root.time automatically (see \texttt{?tree.age}).
+In the following examples, just replace all the \texttt{BeckLee} mentions in this demo by your very own data:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{## My ordinated matrix with tips only}
+\hlstd{MyData_mat50} \hlkwb{<-} \hlstd{BeckLee_mat50}
+
+\hlcom{## My ordinated matrix with tips and nodes}
+\hlstd{MyData_mat99} \hlkwb{<-} \hlstd{BeckLee_mat99}
+
+\hlcom{## My first and last occurence data}
+\hlstd{MyData_ages} \hlkwb{<-} \hlstd{BeckLee_ages}
+
+\hlcom{## My phylogenetic tree}
+\hlstd{MyData_tree} \hlkwb{<-} \hlstd{BeckLee_tree}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
 \section{A super quick go through}
-This package allows the following:
+Without much explanation, here's what this package allows to run:
 \begin{enumerate}
@@
 \hlcom{## Splitting the data}
-\hlstd{sliced_data} \hlkwb{<-} \hlkwd{time.series}\hlstd{(BeckLee_mat99, BeckLee_tree,} \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
-    \hlkwc{model} \hlstd{=} \hlstr{"acctran"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{FADLAD} \hlstd{= BeckLee_ages)}
+\hlstd{sliced_data} \hlkwb{<-} \hlkwd{time.series}\hlstd{(MyData_mat99, MyData_tree,} \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
+    \hlkwc{model} \hlstd{=} \hlstr{"acctran"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{FADLAD} \hlstd{= MyData_ages)}
 \end{alltt}
@@
 ##      series  n  mean  2.5%   25%   75% 97.5%
-## 1 133.51104  3 17.81  0.00 15.51 24.35 24.35
-## 2 100.13328 14 52.05 45.80 49.90 54.50 57.67
-## 3  66.75552 23 64.80 59.23 63.20 66.57 69.25
-## 4  33.37776 11 42.95 37.94 41.40 44.77 47.43
-## 5         0 10 40.94 35.13 39.22 42.69 46.13
+## 1 133.51104  3 16.67  0.00 15.51 18.67 24.35
+## 2 100.13328 14 51.48 45.03 49.29 53.62 56.54
+## 3  66.75552 23 64.77 60.09 63.26 66.34 69.58
+## 4  33.37776 11 42.41 35.87 40.94 44.40 46.98
+## 5         0 10 40.20 34.46 38.38 42.46 44.57
 \end{verbatim}
@@
 
-Et voil\`{a}...
+\textit{Et voil\`{a}}, that's pretty much it...
 
 \section{In more details}
+In this section we are going to look at each functionality in more details. Of course, at this stage of the development of the package, any suggestion is most welcome.
+Please create an issue on the \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} or even just drop me an email at \href{mailto:guillert@tcd.ie}{guillert@tcd.ie}.
 
@@
 \hlkwd{str}\hlstd{(sum_of_ranges)}
+\hlcom{## That's a 2512 lines output!}
 \end{alltt}
@@
 
-Therefore this package proposes a new class of object called \dispRity (how unexpected!) that allow to use a \texttt{S3 print.dispRity} method.
-In other words, when you are creating objects using the \dispRity functions, you can quickly and easily have a look at them by just printing it.
+Therefore this package proposes a new class of object called \dispRity (how unexpected!) that allow to use a S3 \texttt{print.dispRity} method.
+In other words, when you are creating objects using the \dispRity functions, you can quickly and easily have a look at them by just calling the object (it will automatically print it via \texttt{print.dispRity}).
 For example:
@@
 \end{verbatim}
+\begin{alltt}
+\hlcom{## Some attentive ape package users will note the similarities with a phylo object...}
+\end{alltt}
 \end{kframe}
@@
 
-Note that each object output from different functions will show summarise slightly different informations.
+Note that each object output from the different functions detailed below will display slightly different summarised informations (depending on the object content).
 
@@
 One of the first functionality of this package is to facilitate splitting the ordinated matrix in subsamples.
-The original ordinated matrix represents the total any-o-space.
+In fact, in many disparity analysis, one will analysis differences in subsamples of the any-o-space.
+For example, the subsamples can be composed of elements at a certain point in time (disparity-through-time analysis), or elements with some traits in common (e.g. habitat, phylogeny, etc.).
 
 \subsubsection{Custom splitting}
-\texttt{cust.series} is a fairly straightforward function and allows to split the data according to a factor determined by the user.
+\texttt{cust.series} is a function that allows to split the ordinated matrix according to a factor determined by the user.
 For example, here we can split the matrix based on phylogeny: let's classify in two different groups the crown and stem mammals present in the tree.
@@
 \hlcom{## We want to separate the species around the node 71 (see phylogeny above).}
-\hlcom{## All the descendant are crown and all the ancestors are stem}
-\hlstd{crown} \hlkwb{<-} \hlkwd{extract.clade}\hlstd{(BeckLee_tree,} \hlkwc{node} \hlstd{=} \hlnum{71}\hlstd{)}\hlopt{$}\hlstd{tip.label}
-\hlstd{stem} \hlkwb{<-} \hlkwd{drop.tip}\hlstd{(BeckLee_tree,} \hlkwc{tip} \hlstd{= crown)}\hlopt{$}\hlstd{tip.label}
+\hlcom{## All the descendant of these node are crown and all the ancestors are stem}
+\hlstd{crown} \hlkwb{<-} \hlkwd{extract.clade}\hlstd{(MyData_tree,} \hlkwc{node} \hlstd{=} \hlnum{71}\hlstd{)}\hlopt{$}\hlstd{tip.label}
+\hlstd{stem} \hlkwb{<-} \hlkwd{drop.tip}\hlstd{(MyData_tree,} \hlkwc{tip} \hlstd{= crown)}\hlopt{$}\hlstd{tip.label}
 
-\hlcom{## We then have to feed this information in a data.frame with one column}
+\hlcom{## We then have to feed this information in a data frame with one column}
 \hlstd{factors} \hlkwb{<-} \hlkwd{as.data.frame}\hlstd{(}
@@
 
-\hlcom{## Then we can use the customised series function}
-\hlstd{crown_stem} \hlkwb{<-} \hlkwd{cust.series}\hlstd{(BeckLee_mat50, factors)}
+\hlcom{## We then can use the customised series function}
+\hlstd{crown_stem} \hlkwb{<-} \hlkwd{cust.series}\hlstd{(MyData_mat50, factors)}
 
@@
 \begin{alltt}
-\hlcom{## "series" containing information on the series type (custom) and names (crown and stem)}
+\hlcom{## and finally, "series" containing information on the series type (custom) and names}
+\hlcom{## (crown and stem)}
 \hlstd{crown_stem}\hlopt{$}\hlstd{series}
@@
 Any type of factors can be used in this example (e.g. body mass categories, habitat, diet, etc...).
-Which leads us to another type of category used a lot in disparity-through-time analysis: time!
 
@@
 This functions needs as input an ordinated matrix and a matching phylogenetic tree.
-Two types of time series splitting can be performed with the \dispRity package by using the \texttt{method} option in \texttt{time.series}.
+Two types of time series splitting can be performed by using the \texttt{method} option:
 \begin{enumerate}
@@
 For the time-slicing method details, see Guillerme and Cooper \textit{in prep.} \cite{GuillermeSTD}.
-Once the method is chosen, the function asks to specify the \texttt{time} which can be given as a vector of \texttt{numeric} values for:
+Once the method is chosen, the function intakes the \texttt{time} argument which can be a vector of \texttt{numeric} values for:
 \begin{enumerate}
@@
 \hlcom{## Let's generate three time bins containing the data present every 40 Ma}
-\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat50,} \hlkwc{tree} \hlstd{= BeckLee_tree,}
+\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50,} \hlkwc{tree} \hlstd{= MyData_tree,}
     \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{))}
@@
 \hlcom{## age data for the taxa that have some longer occurrence spans.}
-\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat50,} \hlkwc{tree} \hlstd{= BeckLee_tree,}
-    \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{),} \hlkwc{FADLAD} \hlstd{= BeckLee_ages)}
+\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50,} \hlkwc{tree} \hlstd{= MyData_tree,}
+    \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{),} \hlkwc{FADLAD} \hlstd{= MyData_ages)}
 \end{alltt}
@@
 \begin{alltt}
-\hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat50,} \hlkwc{tree} \hlstd{= BeckLee_tree,} \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{3}\hlstd{)}
+\hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50,} \hlkwc{tree} \hlstd{= MyData_tree,} \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{3}\hlstd{)}
 \end{alltt}
@@
 \end{knitrout}
+
 We now have three time bins of 44.50368 Ma each. But hey, that's a bit silly, let's stick to bins of 40 Ma.
-Note that we can also ask to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between taxa AND nodes.
+When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1\% of the the tree length as long as at least three elements are present at that point in time (an extra 1\% is added until reaching the three elements).
+We can also ask to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between tips AND nodes.
 
 For the time-slicing method (\texttt{method = "continuous"}), the idea is pretty similar.
-We just need to provide a matrix that contains the ordinated distance between taxa AND nodes and an evolutionary model via the \texttt{model} argument.
-For now they are four evolutionary models implemented \cite{GuillermeSTD}:
+We need to provide a matrix that contains the ordinated distance between taxa AND nodes and an evolutionary model via the \texttt{model} argument.
+For now they are four evolutionary models implemented \cite{GuillermeSTD} (but see section \ref{whatsleft}):
 \begin{enumerate}
@@
 \begin{alltt}
-\hlcom{## Let's generate four time slices every 40 Ma}
-\hlstd{time_slices} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat99,} \hlkwc{tree} \hlstd{= BeckLee_tree,}
+\hlcom{## Let's generate four time slices every 40 Ma assuming a gradual evolution model}
+\hlstd{time_slices} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat99,} \hlkwc{tree} \hlstd{= MyData_tree,}
     \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,} \hlkwc{model} \hlstd{=} \hlstr{"gradual"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{),}
-    \hlkwc{FADLAD} \hlstd{= BeckLee_ages)}
+    \hlkwc{FADLAD} \hlstd{= MyData_ages)}
 \end{alltt}
@@
 \hlcom{## automatically generate the slices}
-\hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat99,} \hlkwc{tree} \hlstd{= BeckLee_tree,} \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
+\hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat99,} \hlkwc{tree} \hlstd{= MyData_tree,} \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
     \hlkwc{model} \hlstd{=} \hlstr{"gradual"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{4}\hlstd{)}
@@
 \begin{alltt}
-\hlcom{## But that gives the same silly results (slicing ~ 44.50368 Ma)}
+\hlcom{## But that gives the same silly results (slicing every 44.50368 Ma)}
 \end{alltt}
@@
 \subsection{Bootstrapping the data}
+After separating the any-o-space into submatrices, one can directly calculate disparity within each subsample or first test the robustness of the data.
 To measure how robust is the data, we can perform bootstrap pseudo-replications or rarefaction analysis to make sure the results are not dependant on weird outliers.
@@
 \begin{alltt}
-\hlkwd{boot.matrix}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat50)}
+\hlkwd{boot.matrix}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50)}
 \end{alltt}
@@
 \hlcom{## As we can wee, the output is also a dispRity object that is summarised}
-\hlcom{## automatically, giving information on the data as well as the number of}
+\hlcom{## automatically, and gives information on the data as well as the number of}
 \hlcom{## bootstraps and the bootstraps methods.}
@@
 This function proposes the user to specify the bootstrap algorithm with the \texttt{boot.type} method.
-Currently two algorithms are implemented:
+Currently two algorithms are implemented (but see section \ref{whatsleft}):
 \begin{enumerate}
@@
 \hlcom{## Bootstrapping with the single bootstrap method}
-\hlkwd{boot.matrix}\hlstd{(BeckLee_mat50,} \hlkwc{boot.type} \hlstd{=} \hlstr{"single"}\hlstd{)}
+\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{boot.type} \hlstd{=} \hlstr{"single"}\hlstd{)}
 \end{alltt}
@@
 \hlcom{## Bootstrapping with the full rarefaction}
-\hlkwd{boot.matrix}\hlstd{(BeckLee_mat50,} \hlkwc{bootstraps} \hlstd{=} \hlnum{20}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{bootstraps} \hlstd{=} \hlnum{20}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlnum{TRUE}\hlstd{)}
 \end{alltt}
@@
 \hlcom{## Or with a set number of rarefaction levels}
-\hlkwd{boot.matrix}\hlstd{(BeckLee_mat50,} \hlkwc{bootstraps} \hlstd{=} \hlnum{20}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{6}\hlopt{:}\hlnum{8}\hlstd{,}\hlnum{3}\hlstd{))}
+\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{bootstraps} \hlstd{=} \hlnum{20}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{6}\hlopt{:}\hlnum{8}\hlstd{,}\hlnum{3}\hlstd{))}
 \end{alltt}
@@
 \hlcom{## And removing the last axis (default)}
-\hlkwd{boot.matrix}\hlstd{(BeckLee_mat50,} \hlkwc{rm.last.axis} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{rm.last.axis} \hlstd{=} \hlnum{TRUE}\hlstd{)}
 \end{alltt}
@@
 \hlcom{## Or with a fix value (50%)}
-\hlkwd{boot.matrix}\hlstd{(BeckLee_mat50,} \hlkwc{rm.last.axis} \hlstd{=} \hlnum{0.5}\hlstd{)}
+\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{rm.last.axis} \hlstd{=} \hlnum{0.5}\hlstd{)}
 \end{alltt}
@@
 
-Okay, that's all well and good but this function can also deal with a list of matrices or with a \dispRity object output from the \texttt{cust.series} or \texttt{time.series} functions.
+Of course, one could be interested in directly supplying the subsampled matrices generated above directly to this function.
+In fact, it can also deal with a list of matrices or with a \dispRity object output from the \texttt{cust.series} or \texttt{time.series} functions.
 Like this:
@@
 \subsection{Calculating disparity}
-Until now, all the functions where merely splitting and bootstrapping matrices.
 Now we're going to see the functionalities of the core function of this package: the \dispRity function.
 This function is a modulable function that allow to simply (and quickly!) calculate disparity from a matrix.
-Wait, what's modulable? and back to the introduction, what's disparity?
 
-Because disparity can be measured in many ways, this function is just a tool to measure the disparity \textit{as defined by the user} (and here's where the modulable part comes in).
+Because disparity can be measured in many ways, this function is a tool to measure disparity \textit{as defined by the user} (and here's where the modulable part comes in).
 In fact, the \dispRity function intakes two main arguments: the data and the disparity metric.
 The disparity metric boils down to a single value that describes the matrix.
-Some are really classic like the sum or the products of the ranges or the variances of the matrix columns \cite{Wills1994} (i.e. the any-o-space dimensions).
+Some are classic like the sum or the product of the ranges or the variances of the matrix columns (i.e. the any-o-space axis/dimensions) \cite{Wills1994}.
 However, many more exist like the median distances between the taxa and the any-o-space centroid \cite{GuillermeSTD}.
@@
 \begin{enumerate}
-\item the \textbf{class metric} that is typically a descriptor of the matrix.
-For example describing the ranges of each column in the matrix or the euclidean distance between each row and the centroid of the matrix.
-\item the \textbf{summary metric} that is typically a summarise of the class metric values. For example, the sum of the ranges or the median euclidean distance. 
+\item the \textbf{class metric} that is a descriptor of the matrix.
+For example describing the ranges of each column in the matrix or the euclidean distances between each row and the centroid of the matrix.
+\item the \textbf{summary metric} that is a summary of the class metric values. For example, the sum of the ranges or the median of the euclidean distances. 
 \end{enumerate}
 Basically the combination can be infinite between the class and summary metrics.
-For example, people might want to measure the median variance of the axis or the product of the distance from the centroid.
-However, it is probable that some metrics are better to reflect some biological aspects of the any-o-space than others...
+For example, people might want to measure the median variances of the axis or the product of the distances from the centroid.
+However, it is probable that some metrics are better to reflect some biological aspects of the any-o-space than others (see section \ref{whatsleft})...
 
 In practice, the \dispRity function intakes a pair of class and summary metrics as a definition of disparity.
-Several of these metrics are implemented in other packages (like \texttt{stats::median}, \texttt{base::sum}, etc.) and this package proposes several metrics listed in \texttt{dispRity.metric}:
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlopt{?}\hlstd{dispRity.metric}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
+Several of these metrics are implemented in other packages (like \texttt{stats::median}, \texttt{base::sum}, etc.) and this package proposes several metrics listed in \texttt{dispRity.metric} (see \texttt{?dispRity.metric}).
 It is even possible to use your very own class and summary metrics!
@@
 \begin{alltt}
-\hlkwd{dispRity}\hlstd{(BeckLee_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
+\hlkwd{dispRity}\hlstd{(MyData_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
 \end{alltt}
@@
 \hlcom{## automatically which one is the class and the summary metric) so the combinations}
-\hlcom{## are high!}
-\hlkwd{dispRity}\hlstd{(BeckLee_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, variances))}
+\hlcom{## are numerous!}
+\hlkwd{dispRity}\hlstd{(MyData_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, variances))}
 \end{alltt}
@@
 \begin{alltt}
-\hlkwd{dispRity}\hlstd{(BeckLee_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(prod, centroids))}
+\hlkwd{dispRity}\hlstd{(MyData_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(prod, centroids))}
 \end{alltt}
@@
 \begin{alltt}
-\hlkwd{dispRity}\hlstd{(BeckLee_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(mode.val, ranges))}
+\hlkwd{dispRity}\hlstd{(MyData_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(mode.val, ranges))}
 \end{alltt}
@@
 \end{verbatim}
+\begin{alltt}
+\hlcom{## Or funny user defined ones!}
+\hlstd{total.range} \hlkwb{<-} \hlkwa{function}\hlstd{(}\hlkwc{X}\hlstd{)} \hlkwd{abs}\hlstd{(}\hlkwd{range}\hlstd{(X)[}\hlnum{1}\hlstd{]}\hlopt{-}\hlkwd{range}\hlstd{(X)[}\hlnum{2}\hlstd{])}
+\hlkwd{dispRity}\hlstd{(MyData_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(total.range, ranges))}
+\end{alltt}
+\begin{verbatim}
+## Disparity measurements across 1 series for 50 taxa. 
+## 1 unnamed series.
+## Disparity calculated as: total.range ranges for 48 dimensions.
+\end{verbatim}
+\begin{alltt}
+\hlcom{## That would be disparity as the range of the axis ranges!}
+\end{alltt}
 \end{kframe}
@@
 
-Of course, this function can take an ordinated matrix but, more interesting it can also deal with a list of matrices (from the \texttt{cust.series} or \texttt{time.series} functions for example) or a bootstrapped or/and rarefied matrix (from the \texttt{boot.matrix} function for example).
+Of course, this function can take a simple ordinated matrix but, more interesting it can also deal with a list of matrices (from the \texttt{cust.series} or \texttt{time.series} functions for example) or a bootstrapped or/and rarefied matrix (from the \texttt{boot.matrix} function for example).
 
@@
 Again, note that the results are \dispRity objects and can displayed through the S3 \texttt{print.dispRity} method to remind users what are in these objects.
+Also note that an even faster parallel version will be developed in the future (see section \ref{whatsleft}).
 
 \subsection{Summarising the results}
-Finally, after calculating disparity, probably the main thing we want to do is actually see the values of the disparity metrics, either in a table or in a plot.
-Well, that again is easily doable by using the S3 \texttt{summary.dispRity} and \texttt{plot.dispRity} functions.
+As shown in these examples above however, even though the disparity has been calculated in many way, the function doesn't directly outputs the results, rather just a summary of the analysis.
+To really display the results we can use the S3 \texttt{summary.dispRity} and \texttt{plot.dispRity} functions.
 
 \subsubsection{\texttt{summary} function}
-This function intakes a \dispRity object plus various options namely the \texttt{CI} values for the confidence intervals levels; the \texttt{cent.tend} for the central tendency to use for summarising the results and to "graphical" options which are whether to recall the \dispRity options and how much digits are wanted in the results.
+This function intakes a \dispRity object plus various options namely the \texttt{CI} values for the confidence intervals levels; the \texttt{cent.tend} for the central tendency to use for summarising the results and two \textit{visual} options which are whether to recall the \dispRity options and how much digits are wanted in the results.
 
@@
 ##   series  n  mean  2.5%   25%   75% 97.5%
-## 1 120-80  8 1.114 0.915 1.080 1.167 1.235
-## 2  80-40 27 1.322 1.274 1.308 1.338 1.358
-## 3   40-0 16 1.324 1.248 1.306 1.345 1.372
+## 1 120-80  8 1.110 0.899 1.075 1.162 1.225
+## 2  80-40 27 1.322 1.275 1.308 1.337 1.361
+## 3   40-0 16 1.324 1.253 1.306 1.347 1.373
 \end{verbatim}
@@
 \begin{alltt}
-\hlcom{## Also, if we happen to have forgotten what was in the disp_time_binsBS object (as in which}
-\hlcom{## options where used to modify the original matrix) we can use the recall option}
+\hlcom{## Also, if we happen to have forgotten what was in the disp_time_binsBS object (as in}
+\hlcom{## which options where used to modify the original matrix) we can use the recall option:}
 \hlkwd{summary}\hlstd{(disp_time_binsBS,} \hlkwc{recall} \hlstd{=} \hlnum{TRUE}\hlstd{)}
@@
 ##   series  n  mean  2.5%   25%   75% 97.5%
-## 1 120-80  8 1.114 0.915 1.080 1.167 1.235
-## 2  80-40 27 1.322 1.274 1.308 1.338 1.358
-## 3   40-0 16 1.324 1.248 1.306 1.345 1.372
+## 1 120-80  8 1.110 0.899 1.075 1.162 1.225
+## 2  80-40 27 1.322 1.275 1.308 1.337 1.361
+## 3   40-0 16 1.324 1.253 1.306 1.347 1.373
 \end{verbatim}
 \begin{alltt}
-\hlcom{## Note that the recall just prints text. The output of summary.dispRity is always a data.frame}
+\hlcom{## Note that the recall just prints text. The output of summary.dispRity is always a}
+\hlcom{## data.frame}
 \hlstd{summary_table} \hlkwb{<-} \hlkwd{summary}\hlstd{(disp_time_binsBS,} \hlkwc{recall} \hlstd{=} \hlnum{TRUE}\hlstd{)}
@@
 ##   series n  mean  2.5%   25%   75% 97.5%
-## 1  crown 3 15.19 10.19 15.17 16.08 16.66
-## 2  crown 4 18.47 14.97 18.40 19.41 19.89
-## 3  crown 5 20.83 17.99 19.66 21.81 22.46
-## 4  crown 6 22.60 18.84 21.68 23.77 24.51
-## 5  crown 7 24.23 21.16 23.49 25.32 26.15
-## 6  crown 8 25.46 22.40 24.82 26.51 27.42
+## 1  crown 3 15.22 10.27 15.20 16.05 16.71
+## 2  crown 4 18.33 14.82 18.19 19.36 19.94
+## 3  crown 5 20.75 17.74 19.48 21.81 22.40
+## 4  crown 6 22.71 19.18 21.75 23.79 24.45
+## 5  crown 7 24.14 20.97 23.47 25.29 26.04
+## 6  crown 8 25.34 21.71 24.65 26.47 27.44
 \end{verbatim}
@@
 \subsubsection{\texttt{plot} function}
-Finally, this function allows to visualise the disparity results in a often nicer fashion than just a table (even though it tends to be the exact same data).
+This final function allows to visualise the disparity results in a often nicer fashion than just a table (even though the exact same data is displayed).
 The \texttt{plot.dispRity} option intakes the same options as \texttt{summary.dispRity} along side with various graphical options described in the function manual.
@@
 
+The grey squares represent the confidence intervals (50 in dark grey and 90 in light grey) and the dot represents the mean.
+However, it might be interesting to see how many taxa are present in each group.
+This can be done via the \texttt{diversity} option.
+A third interesting option is to plot the rarefied data if a rarefaction was applied on the data with the \texttt{boot.matrix} function.
+For displaying the rarefied data we can use the \texttt{rarefaction} option.
+This option can be either \texttt{FALSE} (default) for always plotting the maximum number of taxa per series or \texttt{TRUE} to always plot the minimum.
+Additionally, it can be a specific number to plot a specific number of taxa.
+Let say we want both groups to have 20 taxa maximum.
+We can see now that the confidence intervals overlap more than previously
+
 \begin{knitrout}
@@
 \hlkwd{quartz}\hlstd{(}\hlkwc{width} \hlstd{=} \hlnum{15}\hlstd{,} \hlkwc{height} \hlstd{=} \hlnum{5}\hlstd{) ;} \hlkwd{par}\hlstd{(}\hlkwc{mfrow} \hlstd{= (}\hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,}\hlnum{3}\hlstd{)),} \hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
-
 \hlcom{## Plotting the score for both groups}
 \hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"} \hlstd{,} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{25}\hlstd{,}\hlnum{40}\hlstd{))}
-
 \hlcom{## Same one with the number of taxa for each data set}
 \hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{25}\hlstd{,}\hlnum{40}\hlstd{))}
-
 \hlcom{## Same one with rarefied data (note how that affects the diversity data as well!)}
-\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlnum{20}\hlstd{,} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{25}\hlstd{,}\hlnum{40}\hlstd{))}
+\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlnum{20}\hlstd{,}
+    \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{25}\hlstd{,}\hlnum{40}\hlstd{))}
 \end{alltt}
@@
 
-The grey squares represent the confidence intervals (50 in dark grey and 90 in light grey) and the dot represents the mean. As we can see, the disparity seems a bit higher in crown mammals.
-Now that can be due just to the difference in the number of species in both crown and stem mammals.
-By the way, how many taxa where in each group? We can look at that by using the \texttt{diversity} option.
-
-This shows that the number of taxa is higher (30) in crown mammals than in stem ones (20), that is why it is maybe interesting to look at the rarefied data using the \texttt{rarefaction} option.
-This option can be either \texttt{FALSE} (default) for always plotting the maximum number of taxa per series or \texttt{TRUE} to always plot the minimum.
-Additionally, it can be a specific number to plot a specific number of taxa.
-Let say we want both groups to have 20 taxa maximum.
-
-We can see now that the confidence intervals overlap more than previously
 To understand how the number of taxa affects each series, we can use the \texttt{rarefaction = "plot"} function to plot the rarefaction curves.
@@
 \hlkwd{par}\hlstd{(}\hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
-
 \hlcom{## The rarefaction curves}
@@
 
-The different dashed lines represent the different confidence intervals.
+The different dashed lines represent the different confidence intervals around each rarefaction curve.
 
@@
 Let's look at that with the time binned data (even though there is only three time bins).
+Finally, we can use the \texttt{type = "continuous"} option that comes in handy for plotting continuous data like in the time sliced analysis (obviously...).
 
@@
 \hlkwd{quartz}\hlstd{(}\hlkwc{width} \hlstd{=} \hlnum{10}\hlstd{,} \hlkwc{height} \hlstd{=} \hlnum{5}\hlstd{) ;} \hlkwd{par}\hlstd{(}\hlkwc{mfrow} \hlstd{= (}\hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,}\hlnum{2}\hlstd{)),} \hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
-
-\hlcom{## The disparity-through-time data in a time-binned way}
+\hlcom{## The disparity-through-time data in a time-binned way (with lines rather than}
+\hlcom{## boxes)}
 \hlkwd{plot}\hlstd{(disp_time_binsBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{discrete_type} \hlstd{=} \hlstr{"line"}\hlstd{)}
-
 \hlcom{## The disparity-through-time data in a time-sliced ways}
@@
 
-Finally, we can use the \texttt{type = "continuous"} option that comes in handy for plotting continuous data like in the time sliced analysis (obviously...).
+We can even plot both results (discrete and continuous) on the same graph:
 
@@
 \hlkwd{par}\hlstd{(}\hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
-
 \hlcom{## First, plotting the continuous data}
 \hlkwd{plot}\hlstd{(disp_time_slicesBS,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{,} \hlkwc{ylim}\hlstd{=}\hlkwd{c}\hlstd{(}\hlnum{0.5}\hlstd{,}\hlnum{2}\hlstd{))}
-
 \hlcom{## Second, adding the discrete data}
 \hlkwd{par}\hlstd{(}\hlkwc{new} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\hlkwd{plot}\hlstd{(disp_time_binsBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{discrete_type} \hlstd{=} \hlstr{"line"}\hlstd{,} \hlkwc{ylim}\hlstd{=}\hlkwd{c}\hlstd{(}\hlnum{0.5}\hlstd{,}\hlnum{2}\hlstd{),} \hlkwc{xlab}\hlstd{=}\hlstr{""}\hlstd{,} \hlkwc{ylab}\hlstd{=}\hlstr{""}\hlstd{)}
+\hlkwd{plot}\hlstd{(disp_time_binsBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{discrete_type} \hlstd{=} \hlstr{"line"}\hlstd{,} \hlkwc{ylim}\hlstd{=}\hlkwd{c}\hlstd{(}\hlnum{0.5}\hlstd{,}\hlnum{2}\hlstd{),}
+    \hlkwc{xlab}\hlstd{=}\hlstr{""}\hlstd{,} \hlkwc{ylab}\hlstd{=}\hlstr{""}\hlstd{)}
 \end{alltt}
@@
 
+Note that the results vary in values but not in pattern.
+The change in values might be due to the fact that nodes data are not included in the discrete analysis.
+
 I encourage you to play with the graphical options to have some prettier results.
@@
 \hlkwd{par}\hlstd{(}\hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
-
 \hlcom{## A plot with many options!}
-\hlkwd{plot}\hlstd{(disp_time_slicesBS,} \hlkwc{CI} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{50}\hlstd{,} \hlnum{75}\hlstd{,} \hlnum{95}\hlstd{),} \hlkwc{cent.tend} \hlstd{= median,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
-    \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"black"}\hlstd{,} \hlstr{"red"}\hlstd{,} \hlstr{"orange"}\hlstd{,} \hlstr{"yellow"}\hlstd{),} \hlkwc{ylab} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"Disparity"}\hlstd{,} \hlstr{"Diversity"}\hlstd{),}
-    \hlkwc{xlab} \hlstd{=} \hlstr{"Time (in in units from past to present)"}\hlstd{,} \hlkwc{time.series} \hlstd{=} \hlnum{FALSE}\hlstd{,} \hlkwc{main} \hlstd{=} \hlstr{"Many options..."}\hlstd{)}
+\hlkwd{plot}\hlstd{(disp_time_slicesBS,} \hlkwc{CI} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{seq}\hlstd{(}\hlkwc{from}\hlstd{=}\hlnum{10}\hlstd{,} \hlkwc{to}\hlstd{=}\hlnum{100}\hlstd{,} \hlkwc{by}\hlstd{=}\hlnum{10}\hlstd{)),} \hlkwc{cent.tend} \hlstd{= sd,}
+    \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlstr{"log"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"black"}\hlstd{,} \hlkwd{rainbow}\hlstd{(}\hlnum{10}\hlstd{)),}
+    \hlkwc{ylab} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"Disparity"}\hlstd{,} \hlstr{"log(Diversity)"}\hlstd{),} \hlkwc{xlab} \hlstd{=} \hlstr{"Time (in in units from
+    past to present)"}\hlstd{,} \hlkwc{time.series} \hlstd{=} \hlnum{FALSE}\hlstd{,} \hlkwc{main} \hlstd{=} \hlstr{"Many options..."}\hlstd{)}
 \end{alltt}
@@
 
-{\centering \includegraphics[width=.5\linewidth]{figure/minimal-plot5-1}
+{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot5-1}
 
@@
 \label{whatsleft}
+As stated at the start of the demo, this version 0.1.0 is still in development and many parts are missing.
+Here are the new functionalities that will be implemented before the proper release version (v.1).
+
 \subsection{Some statistics}
+Some people might get interested in p-values and other statsy numbers.
+Because the format of the \dispRity objects is a bit special, I am currently developing a \texttt{dispRity.test} function that will allow to intake your favourite statistical test and a \dispRity object.
+I will also propose several inbuilt test such as the time lag t-test \cite{GuillermeSTD}.
+Any other suggestions are most welcome!
+
 \subsection{Users defined metrics}
+To help people develop their own disparity metrics (both class and summary) and to make sure they will be usable in the \dispRity function, I am developing a \texttt{make.metric} function that will test the metric in two ways:
+(1) is the output sensible (defined by the user)?
+(2) does it compute properly in the \dispRity function?
+
+Along side with developing this package, I also aim to test the efficiency of some metrics in describing some aspects of the any-o-space so that it could be used as guidelines for disparity analysis
+
 \subsection{More user defined stuff}
+I intend also develop functions to help users to develop their own algorithms for the bootstrap method (via \texttt{make.boot}) or the evolutionary models (via \texttt{make.model}).
+Both functions will provide similar testing as the \texttt{make.metric} function.
+
+\subsection{Faster!}
+Finally, for long analysis, I intend to develop a parallel running version of the package.
+In fact, most of the internal functions are base on \texttt{lapply} functions that can be easily passed to \texttt{snow::parLapply} or similar parallel functions.
 

### 2015-10-01 doc/dispRity-tutorial.Rnw
Renamed Rnw file

--- 2015-10-01 doc/dispRity.demo.tex
@@
-\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
+\documentclass{article}
 \usepackage[sc]{mathpazo}
@@
 \newcommand{\R}{\texttt{R} }
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+
 \begin{document}
+<<setup, include=FALSE, cache=FALSE>>=
+library(knitr)
+# set global chunk options
+opts_chunk$set(fig.path='figure/minimal-', fig.align='center', fig.show='hold')
+options(formatR.arrow=TRUE,width=90)
+@
 
@@
 
-This a quick demo to go through the beta version of the \dispRity package (v.0.1.0).
+This a quick demo to go through the beta version of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}).
 Many parts are still missing (see section \ref{whatsleft}) but this version should be running all right and give an idea of what's implemented in the package.
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{install.packages}\hlstd{(}\hlstr{"devtools"}\hlstd{)}
-\hlkwd{library}\hlstd{(devtools)}
-\hlkwd{install_github}\hlstd{(}\hlstr{"TGuillerme/dispRity"}\hlstd{,} \hlkwc{ref} \hlstd{=} \hlstr{"release"}\hlstd{)}
-\hlkwd{library}\hlstd{(dispRity)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
+<<installation, eval=FALSE>>=
+install.packages("devtools")
+library(devtools)
+install_github("TGuillerme/dispRity", ref = "release")
+library(dispRity)
+@
 
-Note that we use the \texttt{release} branch here which is version 0.1.0.
+Note that we use the \texttt{release} branch here which is version \Sexpr{packageVersion("dispRity")}.
 If you want the piping-hot (and full of bugs) version, change the option \texttt{ref = "release"} to \texttt{ref = "master"}.
 This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.
-Future version will just require the specific function but this version (0.1.0) will install \texttt{paleotree} as a dependency (which is not that bad since it is a super useful package anyway!).
+Future version will just require the specific function but this version (\Sexpr{packageVersion("dispRity")}) will install \texttt{paleotree} as a dependency (which is not that bad since it is a super useful package anyway!).
 Additionally, I warmly recommend to install the packages \texttt{geomorph} or/and \texttt{Claddis} if you want to try this demo with your very own ordinated morphometric or cladistic data.
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## Loading demo and the package data}
-\hlkwd{library}\hlstd{(dispRity)}
-\end{alltt}
+<<data, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+## Loading demo and the package data
+library(dispRity)
 
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: paleotree}}\begin{alltt}
-\hlcom{## An ordinated matrix with the tips only}
-\hlkwd{data}\hlstd{(BeckLee_mat50)}
-\hlkwd{dim}\hlstd{(BeckLee_mat50)}
-\end{alltt}
-\begin{verbatim}
-## [1] 50 48
-\end{verbatim}
-\begin{alltt}
-\hlkwd{head}\hlstd{(BeckLee_mat50[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{])}
-\end{alltt}
-\begin{verbatim}
-##                    [,1]          [,2]        [,3]       [,4]      [,5]
-## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
-## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
-## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
-## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
-## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
-## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273
-\end{verbatim}
-\begin{alltt}
-\hlcom{## An ordinated matrix with the tips and the nodes (named nX)}
-\hlkwd{data}\hlstd{(BeckLee_mat99)}
-\hlkwd{dim}\hlstd{(BeckLee_mat99)}
-\end{alltt}
-\begin{verbatim}
-## [1] 99 97
-\end{verbatim}
-\begin{alltt}
-\hlkwd{head}\hlstd{(BeckLee_mat99[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{],} \hlnum{2}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-##                  [,1]       [,2]       [,3]       [,4]        [,5]
-## Cimolestes -0.6082437 -0.0323683 0.08458885 -0.4338448 -0.30536875
-## Maelestes  -0.5730206 -0.2840361 0.01308847 -0.1258848  0.06123611
-\end{verbatim}
-\begin{alltt}
-\hlkwd{tail}\hlstd{(BeckLee_mat99[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{],} \hlnum{2}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-##            [,1]      [,2]       [,3]       [,4]       [,5]
-## n48 -0.05529018 0.4799330 0.04118477 0.04944912 -0.3558830
-## n49 -0.13067785 0.4478168 0.11956268 0.13800340 -0.3222785
-\end{verbatim}
-\begin{alltt}
-\hlcom{## A list of first and last occurrences data for some fossils}
-\hlkwd{data}\hlstd{(BeckLee_ages)}
-\hlkwd{head}\hlstd{(BeckLee_ages)}
-\end{alltt}
-\begin{verbatim}
-##             FAD  LAD
-## Adapis     37.2 36.8
-## Asioryctes 83.6 72.1
-## Leptictis  33.9 33.3
-## Miacis     49.0 46.7
-## Mimotona   61.6 59.2
-## Notharctus 50.2 47.0
-\end{verbatim}
-\begin{alltt}
-\hlcom{## And a phylogeny:}
-\hlkwd{data}\hlstd{(BeckLee_tree)}
-\hlkwd{plot}\hlstd{(BeckLee_tree) ;} \hlkwd{nodelabels}\hlstd{(}\hlkwc{cex}\hlstd{=}\hlnum{0.8}\hlstd{)}
-\end{alltt}
-\end{kframe}
+## An ordinated matrix with the tips only
+data(BeckLee_mat50)
+dim(BeckLee_mat50)
+head(BeckLee_mat50[,1:5])
 
-{\centering \includegraphics[width=.8\linewidth]{figure/minimal-data-1}
+## An ordinated matrix with the tips and the nodes (named nX)
+data(BeckLee_mat99)
+dim(BeckLee_mat99)
+head(BeckLee_mat99[,1:5], 2)
+tail(BeckLee_mat99[,1:5], 2)
 
-}
+## A list of first and last occurrences data for some fossils
+data(BeckLee_ages)
+head(BeckLee_ages)
 
-\end{knitrout}
+## And a phylogeny:
+data(BeckLee_tree)
+plot(BeckLee_tree) ; nodelabels(cex=0.8)
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## My ordinated matrix with tips only}
-\hlstd{MyData_mat50} \hlkwb{<-} \hlstd{BeckLee_mat50}
+<<Setup the data>>=
+## My ordinated matrix with tips only
+MyData_mat50 <- BeckLee_mat50
 
-\hlcom{## My ordinated matrix with tips and nodes}
-\hlstd{MyData_mat99} \hlkwb{<-} \hlstd{BeckLee_mat99}
+## My ordinated matrix with tips and nodes
+MyData_mat99 <- BeckLee_mat99
 
-\hlcom{## My first and last occurence data}
-\hlstd{MyData_ages} \hlkwb{<-} \hlstd{BeckLee_ages}
+## My first and last occurence data
+MyData_ages <- BeckLee_ages
 
-\hlcom{## My phylogenetic tree}
-\hlstd{MyData_tree} \hlkwb{<-} \hlstd{BeckLee_tree}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
+## My phylogenetic tree
+MyData_tree <- BeckLee_tree
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## Splitting the data}
-\hlstd{sliced_data} \hlkwb{<-} \hlkwd{time.series}\hlstd{(MyData_mat99, MyData_tree,} \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
-    \hlkwc{model} \hlstd{=} \hlstr{"acctran"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{FADLAD} \hlstd{= MyData_ages)}
-\end{alltt}
+<<Quick go, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
+## Splitting the data
+sliced_data <- time.series(MyData_mat99, MyData_tree, method = "continuous",
+    model = "acctran", time = 5, FADLAD = MyData_ages)
 
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Some tips have FAD/LAD and are assumed to interval single points in time.}}\begin{alltt}
-\hlcom{## Bootstrapping the data}
-\hlstd{bootstrapped_data} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(sliced_data,} \hlnum{100}\hlstd{)}
+## Bootstrapping the data
+bootstrapped_data <- boot.matrix(sliced_data, 100)
 
-\hlcom{## Calculating disparity}
-\hlstd{sum_of_ranges} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(bootstrapped_data,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
+## Calculating disparity
+sum_of_ranges <- dispRity(bootstrapped_data, metric = c(sum, ranges))
 
-\hlcom{## Summarising the results}
-\hlkwd{summary}\hlstd{(sum_of_ranges)}
-\end{alltt}
-\begin{verbatim}
-##      series  n  mean  2.5%   25%   75% 97.5%
-## 1 133.51104  3 16.67  0.00 15.51 18.67 24.35
-## 2 100.13328 14 51.48 45.03 49.29 53.62 56.54
-## 3  66.75552 23 64.77 60.09 63.26 66.34 69.58
-## 4  33.37776 11 42.41 35.87 40.94 44.40 46.98
-## 5         0 10 40.20 34.46 38.38 42.46 44.57
-\end{verbatim}
-\begin{alltt}
-\hlkwd{plot}\hlstd{(sum_of_ranges,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{)}
-\end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=.5\linewidth]{figure/minimal-Quick_go-1}
-
-}
-
-\end{knitrout}
+## Summarising the results
+summary(sum_of_ranges)
+plot(sum_of_ranges, type = "continuous")
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{str}\hlstd{(sum_of_ranges)}
-\hlcom{## That's a 2512 lines output!}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
+<<dispRity class1, eval=FALSE>>=
+str(sum_of_ranges)
+## That's a 2512 lines output!
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## Which class is the sum_of_ranges object?}
-\hlkwd{class}\hlstd{(sum_of_ranges)}
-\end{alltt}
-\begin{verbatim}
-## [1] "dispRity"
-\end{verbatim}
-\begin{alltt}
-\hlcom{## We can summarise it using the S3 method print.dispRity}
-\hlstd{sum_of_ranges}
-\end{alltt}
-\begin{verbatim}
-## Disparity measurements across 5 series for 99 taxa. 
-## Series:
-## 133.51104, 100.13328, 66.75552, 33.37776, 0.
-## Disparity calculated as: sum ranges for 97 dimensions.
-## Data was split using continuous method.
-## Data was bootstrapped 100 times, using the full bootstrap method.
-\end{verbatim}
-\begin{alltt}
-\hlcom{## Some attentive ape package users will note the similarities with a phylo object...}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
+<<dispRity class2>>=
+## Which class is the sum_of_ranges object?
+class(sum_of_ranges)
+
+## We can summarise it using the S3 method print.dispRity
+sum_of_ranges
+## Some attentive ape package users will note the similarities with a phylo object...
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## We want to separate the species around the node 71 (see phylogeny above).}
-\hlcom{## All the descendant of these node are crown and all the ancestors are stem}
-\hlstd{crown} \hlkwb{<-} \hlkwd{extract.clade}\hlstd{(MyData_tree,} \hlkwc{node} \hlstd{=} \hlnum{71}\hlstd{)}\hlopt{$}\hlstd{tip.label}
-\hlstd{stem} \hlkwb{<-} \hlkwd{drop.tip}\hlstd{(MyData_tree,} \hlkwc{tip} \hlstd{= crown)}\hlopt{$}\hlstd{tip.label}
+<<cust.series2>>=
+## We want to separate the species around the node 71 (see phylogeny above).
+## All the descendant of these node are crown and all the ancestors are stem
+crown <- extract.clade(MyData_tree, node = 71)$tip.label
+stem <- drop.tip(MyData_tree, tip = crown)$tip.label
 
-\hlcom{## We then have to feed this information in a data frame with one column}
-\hlstd{factors} \hlkwb{<-} \hlkwd{as.data.frame}\hlstd{(}
-    \hlkwd{matrix}\hlstd{(}\hlkwc{data} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{rep}\hlstd{(}\hlstr{"crown"}\hlstd{,} \hlkwd{length}\hlstd{(crown)),} \hlkwd{rep}\hlstd{(}\hlstr{"stem"}\hlstd{,} \hlkwd{length}\hlstd{(stem))),}
-    \hlkwc{ncol} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{dimnames} \hlstd{=} \hlkwd{list}\hlstd{(}\hlkwd{c}\hlstd{(crown, stem))))}
+## We then have to feed this information in a data frame with one column
+factors <- as.data.frame(
+    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
+    ncol = 1, dimnames = list(c(crown, stem))))
 
-\hlcom{## We then can use the customised series function}
-\hlstd{crown_stem} \hlkwb{<-} \hlkwd{cust.series}\hlstd{(MyData_mat50, factors)}
+## We then can use the customised series function
+crown_stem <- cust.series(MyData_mat50, factors)
 
-\hlcom{## This created a dispRity object containing two series: crown and stem}
-\hlkwd{class}\hlstd{(crown_stem)}
-\end{alltt}
-\begin{verbatim}
-## [1] "dispRity"
-\end{verbatim}
-\begin{alltt}
-\hlstd{crown_stem}
-\end{alltt}
-\begin{verbatim}
-## 2 custom series for 50 taxa. 
-## Series:
-## crown, stem.
-\end{verbatim}
-\begin{alltt}
-\hlcom{## This object contains three elements}
-\hlkwd{names}\hlstd{(crown_stem)}
-\end{alltt}
-\begin{verbatim}
-## [1] "data"   "taxa"   "series"
-\end{verbatim}
-\begin{alltt}
-\hlcom{## With "data" being the list sub-matrices}
-\hlkwd{str}\hlstd{(crown_stem}\hlopt{$}\hlstd{data)}
-\end{alltt}
-\begin{verbatim}
-## List of 2
-##  $ crown: num [1:30, 1:48] 0.3079 0.6531 0.5089 -0.1652 -0.0419 ...
-##   ..- attr(*, "dimnames")=List of 2
-##   .. ..$ : chr [1:30] "Dasypodidae" "Bradypus" "Myrmecophagidae" "Todralestes" ...
-##   .. ..$ : NULL
-##  $ stem : num [1:20, 1:48] -0.739 -0.68 -0.511 -0.477 -0.473 ...
-##   ..- attr(*, "dimnames")=List of 2
-##   .. ..$ : chr [1:20] "Daulestes" "Bulaklestes" "Uchkudukodon" "Kennalestes" ...
-##   .. ..$ : NULL
-\end{verbatim}
-\begin{alltt}
-\hlcom{## "taxa" being the list of taxa in the original ordinated matrix}
-\hlkwd{str}\hlstd{(crown_stem}\hlopt{$}\hlstd{taxa)}
-\end{alltt}
-\begin{verbatim}
-##  chr [1:50] "Cimolestes" "Maelestes" "Batodon" "Bulaklestes" "Daulestes" ...
-\end{verbatim}
-\begin{alltt}
-\hlcom{## and finally, "series" containing information on the series type (custom) and names}
-\hlcom{## (crown and stem)}
-\hlstd{crown_stem}\hlopt{$}\hlstd{series}
-\end{alltt}
-\begin{verbatim}
-## [1] "custom" "crown"  "stem"
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+## This created a dispRity object containing two series: crown and stem
+class(crown_stem)
+crown_stem
+
+## This object contains three elements
+names(crown_stem)
+
+## With "data" being the list sub-matrices
+str(crown_stem$data)
+
+## "taxa" being the list of taxa in the original ordinated matrix
+str(crown_stem$taxa)
+
+## and finally, "series" containing information on the series type (custom) and names
+## (crown and stem)
+crown_stem$series
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## Let's generate three time bins containing the data present every 40 Ma}
-\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50,} \hlkwc{tree} \hlstd{= MyData_tree,}
-    \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{))}
-\end{alltt}
+<<discrete.series>>=
+## Let's generate three time bins containing the data present every 40 Ma
+time_bins <- time.series(data = MyData_mat50, tree = MyData_tree,
+    method = "discrete", time = c(120, 80, 40, 0))
 
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# No FADLAD table has been provided so every tip is assumed to interval single points in time.}}\begin{alltt}
-\hlcom{## Note that the function provides a warning saying that tips where single}
-\hlcom{## points in time (no FAD/LAD information). We can fix that by adding the}
-\hlcom{## age data for the taxa that have some longer occurrence spans.}
-\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50,} \hlkwc{tree} \hlstd{= MyData_tree,}
-    \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{),} \hlkwc{FADLAD} \hlstd{= MyData_ages)}
-\end{alltt}
+## Note that the function provides a warning saying that tips where single
+## points in time (no FAD/LAD information). We can fix that by adding the
+## age data for the taxa that have some longer occurrence spans.
+time_bins <- time.series(data = MyData_mat50, tree = MyData_tree,
+    method = "discrete", time = c(120, 80, 40, 0), FADLAD = MyData_ages)
 
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Some tips have FAD/LAD and are assumed to interval single points in time.}}\begin{alltt}
-\hlcom{## To entirely avoid the warning we could collect the occurrence span data}
-\hlcom{## for all the taxa but that's not necessary. The function automatically}
-\hlcom{## assumes no occurrence span time (i.e. single points in time) for all taxa}
-\hlcom{## by default.}
+## To entirely avoid the warning we could collect the occurrence span data
+## for all the taxa but that's not necessary. The function automatically
+## assumes no occurrence span time (i.e. single points in time) for all taxa
+## by default.
 
-\hlstd{time_bins}
-\end{alltt}
-\begin{verbatim}
-## 3 discrete series for 50 taxa. 
-## Series:
-## 120-80, 80-40, 40-0.
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+time_bins
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50,} \hlkwc{tree} \hlstd{= MyData_tree,} \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{3}\hlstd{)}
-\end{alltt}
-
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# No FADLAD table has been provided so every tip is assumed to interval single points in time.}}\begin{verbatim}
-## 3 discrete series for 50 taxa. 
-## Series:
-## 133.51104-89.00736, 89.00736-44.50368, 44.50368-0.
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+<<discrete.series2>>=
+time.series(data = MyData_mat50, tree = MyData_tree, method = "discrete", time = 3)
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## Let's generate four time slices every 40 Ma assuming a gradual evolution model}
-\hlstd{time_slices} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat99,} \hlkwc{tree} \hlstd{= MyData_tree,}
-    \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,} \hlkwc{model} \hlstd{=} \hlstr{"gradual"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{),}
-    \hlkwc{FADLAD} \hlstd{= MyData_ages)}
-\end{alltt}
+<<continuous.series>>=
+## Let's generate four time slices every 40 Ma assuming a gradual evolution model
+time_slices <- time.series(data = MyData_mat99, tree = MyData_tree, 
+    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
+    FADLAD = MyData_ages)
 
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Some tips have FAD/LAD and are assumed to interval single points in time.}}\begin{alltt}
-\hlstd{time_slices}
-\end{alltt}
-\begin{verbatim}
-## 4 continuous series for 99 taxa. 
-## Series:
-## 120, 80, 40, 0.
-\end{verbatim}
-\begin{alltt}
-\hlcom{## Note that in the same way as for the discrete method, we can also}
-\hlcom{## automatically generate the slices}
-\hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat99,} \hlkwc{tree} \hlstd{= MyData_tree,} \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
-    \hlkwc{model} \hlstd{=} \hlstr{"gradual"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{4}\hlstd{)}
-\end{alltt}
+time_slices
 
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# No FADLAD table has been provided so every tip is assumed to interval single points in time.}}\begin{verbatim}
-## 4 continuous series for 99 taxa. 
-## Series:
-## 133.51104, 89.00736, 44.50368, 0.
-\end{verbatim}
-\begin{alltt}
-\hlcom{## But that gives the same silly results (slicing every 44.50368 Ma)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
+## Note that in the same way as for the discrete method, we can also
+## automatically generate the slices
+time.series(data = MyData_mat99, tree = MyData_tree, method = "continuous",
+    model = "gradual", time = 4)
+## But that gives the same silly results (slicing every 44.50368 Ma)
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{boot.matrix}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50)}
-\end{alltt}
-\begin{verbatim}
-## Bootstrapped ordinated matrix with 50 taxa. 
-## 1 unnamed series.
-## Data was bootstrapped 1000 times, using the full bootstrap method.
-\end{verbatim}
-\begin{alltt}
-\hlcom{## As we can wee, the output is also a dispRity object that is summarised}
-\hlcom{## automatically, and gives information on the data as well as the number of}
-\hlcom{## bootstraps and the bootstraps methods.}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
+<<boot.matrix1>>=
+boot.matrix(data = MyData_mat50)
+## As we can wee, the output is also a dispRity object that is summarised
+## automatically, and gives information on the data as well as the number of
+## bootstraps and the bootstraps methods.
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## Bootstrapping with the single bootstrap method}
-\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{boot.type} \hlstd{=} \hlstr{"single"}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-## Bootstrapped ordinated matrix with 50 taxa. 
-## 1 unnamed series.
-## Data was bootstrapped 1000 times, using the single bootstrap method.
-\end{verbatim}
-\begin{alltt}
-\hlcom{## Bootstrapping with the full rarefaction}
-\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{bootstraps} \hlstd{=} \hlnum{20}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-## Bootstrapped ordinated matrix with 50 taxa. 
-## 1 unnamed series.
-## Data was bootstrapped 20 times, using the full bootstrap method.
-## Data was fully rarefied (down to 3 taxa).
-\end{verbatim}
-\begin{alltt}
-\hlcom{## Or with a set number of rarefaction levels}
-\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{bootstraps} \hlstd{=} \hlnum{20}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{6}\hlopt{:}\hlnum{8}\hlstd{,}\hlnum{3}\hlstd{))}
-\end{alltt}
-\begin{verbatim}
-## Bootstrapped ordinated matrix with 50 taxa. 
-## 1 unnamed series.
-## Data was bootstrapped 20 times, using the full bootstrap method.
-## Data was rarefied with a maximum of 6, 7, 8 and 3 taxa.
-\end{verbatim}
-\begin{alltt}
-\hlcom{## And removing the last axis (default)}
-\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{rm.last.axis} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-## Bootstrapped ordinated matrix with 50 taxa. 
-## 1 unnamed series.
-## Data was bootstrapped 1000 times, using the full bootstrap method. 
-## The 6 last axis have been removed from the original data.
-\end{verbatim}
-\begin{alltt}
-\hlcom{## Or with a fix value (50%)}
-\hlkwd{boot.matrix}\hlstd{(MyData_mat50,} \hlkwc{rm.last.axis} \hlstd{=} \hlnum{0.5}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-## Bootstrapped ordinated matrix with 50 taxa. 
-## 1 unnamed series.
-## Data was bootstrapped 1000 times, using the full bootstrap method. 
-## The 35 last axis have been removed from the original data.
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+<<boot.matrix2>>=
+## Bootstrapping with the single bootstrap method
+boot.matrix(MyData_mat50, boot.type = "single")
+## Bootstrapping with the full rarefaction
+boot.matrix(MyData_mat50, bootstraps = 20, rarefaction = TRUE)
+## Or with a set number of rarefaction levels
+boot.matrix(MyData_mat50, bootstraps = 20, rarefaction = c(6:8,3))
+## And removing the last axis (default)
+boot.matrix(MyData_mat50, rm.last.axis = TRUE)
+## Or with a fix value (50%)
+boot.matrix(MyData_mat50, rm.last.axis = 0.5)
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## Let's do a full rarefaction on the crown/stem data for later}
-\hlstd{crown_stemBS} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(crown_stem,} \hlkwc{rarefaction} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\hlcom{## And just bootstrapping for the time binning/slicing}
-\hlstd{time_binsBS} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(time_bins)}
-\hlstd{time_slicesBS} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(time_slices)}
+<<boot.matrix3>>=
+## Let's do a full rarefaction on the crown/stem data for later
+crown_stemBS <- boot.matrix(crown_stem, rarefaction = TRUE)
+## And just bootstrapping for the time binning/slicing
+time_binsBS <- boot.matrix(time_bins)
+time_slicesBS <- boot.matrix(time_slices)
 
-\hlcom{## And just to remind us what's in each objects}
-\hlstd{crown_stemBS}
-\end{alltt}
-\begin{verbatim}
-## Bootstrapped ordinated matrix with 50 taxa. 
-## Series:
-## crown, stem.
-## Data was split using custom method.
-## Data was bootstrapped 1000 times, using the full bootstrap method.
-## Data was fully rarefied (down to 3 taxa).
-\end{verbatim}
-\begin{alltt}
-\hlstd{time_binsBS}
-\end{alltt}
-\begin{verbatim}
-## Bootstrapped ordinated matrix with 50 taxa. 
-## Series:
-## 120-80, 80-40, 40-0.
-## Data was split using discrete method.
-## Data was bootstrapped 1000 times, using the full bootstrap method.
-\end{verbatim}
-\begin{alltt}
-\hlstd{time_slicesBS}
-\end{alltt}
-\begin{verbatim}
-## Bootstrapped ordinated matrix with 99 taxa. 
-## Series:
-## 120, 80, 40, 0.
-## Data was split using continuous method.
-## Data was bootstrapped 1000 times, using the full bootstrap method.
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+## And just to remind us what's in each objects
+crown_stemBS
+time_binsBS
+time_slicesBS
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{dispRity}\hlstd{(MyData_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
-\end{alltt}
-\begin{verbatim}
-## Disparity measurements across 1 series for 50 taxa. 
-## 1 unnamed series.
-## Disparity calculated as: sum ranges for 48 dimensions.
-\end{verbatim}
-\begin{alltt}
-\hlcom{## Yep, that's it. The order of the metrics does not matter (the function detects}
-\hlcom{## automatically which one is the class and the summary metric) so the combinations}
-\hlcom{## are numerous!}
-\hlkwd{dispRity}\hlstd{(MyData_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, variances))}
-\end{alltt}
-\begin{verbatim}
-## Disparity measurements across 1 series for 50 taxa. 
-## 1 unnamed series.
-## Disparity calculated as: sum variances for 48 dimensions.
-\end{verbatim}
-\begin{alltt}
-\hlkwd{dispRity}\hlstd{(MyData_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(prod, centroids))}
-\end{alltt}
-\begin{verbatim}
-## Disparity measurements across 1 series for 50 taxa. 
-## 1 unnamed series.
-## Disparity calculated as: prod centroids for 48 dimensions.
-\end{verbatim}
-\begin{alltt}
-\hlkwd{dispRity}\hlstd{(MyData_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(mode.val, ranges))}
-\end{alltt}
-\begin{verbatim}
-## Disparity measurements across 1 series for 50 taxa. 
-## 1 unnamed series.
-## Disparity calculated as: mode.val ranges for 48 dimensions.
-\end{verbatim}
-\begin{alltt}
-\hlcom{## Or funny user defined ones!}
-\hlstd{total.range} \hlkwb{<-} \hlkwa{function}\hlstd{(}\hlkwc{X}\hlstd{)} \hlkwd{abs}\hlstd{(}\hlkwd{range}\hlstd{(X)[}\hlnum{1}\hlstd{]}\hlopt{-}\hlkwd{range}\hlstd{(X)[}\hlnum{2}\hlstd{])}
-\hlkwd{dispRity}\hlstd{(MyData_mat50,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(total.range, ranges))}
-\end{alltt}
-\begin{verbatim}
-## Disparity measurements across 1 series for 50 taxa. 
-## 1 unnamed series.
-## Disparity calculated as: total.range ranges for 48 dimensions.
-\end{verbatim}
-\begin{alltt}
-\hlcom{## That would be disparity as the range of the axis ranges!}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
+<<dispRity1>>=
+dispRity(MyData_mat50, metric = c(sum, ranges))
+## Yep, that's it. The order of the metrics does not matter (the function detects
+## automatically which one is the class and the summary metric) so the combinations
+## are numerous!
+dispRity(MyData_mat50, metric = c(sum, variances))
+dispRity(MyData_mat50, metric = c(prod, centroids))
+dispRity(MyData_mat50, metric = c(mode.val, ranges))
+## Or funny user defined ones!
+total.range <- function(X) abs(range(X)[1]-range(X)[2])
+dispRity(MyData_mat50, metric = c(total.range, ranges))
+## That would be disparity as the range of the axis ranges!
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## For example, let's calculate the sum of ranges from our crown/stem mammals}
-\hlkwd{dispRity}\hlstd{(crown_stem,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
-\end{alltt}
-\begin{verbatim}
-## Disparity measurements across 2 series for 50 taxa. 
-## Series:
-## crown, stem.
-## Disparity calculated as: sum ranges for 48 dimensions.
-## Data was split using custom method.
-\end{verbatim}
-\begin{alltt}
-\hlcom{## Or even better, from the bootstrapped data}
-\hlstd{disp_crown_stemBS} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(crown_stemBS,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, ranges))}
-\hlcom{## Note that the bootstrapped + rarefied data takes some time to run (it's}
-\hlcom{## calculating disparity from 46 matrices!)}
+<<dispRity2>>=
+## For example, let's calculate the sum of ranges from our crown/stem mammals
+dispRity(crown_stem, metric = c(sum, ranges))
+## Or even better, from the bootstrapped data
+disp_crown_stemBS <- dispRity(crown_stemBS, metric = c(sum, ranges))
+## Note that the bootstrapped + rarefied data takes some time to run (it's
+## calculating disparity from 46 matrices!)
 
-\hlcom{## Let's calculate a different metric for the disparity-through-time analysis}
-\hlstd{disp_time_binsBS} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(time_binsBS,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(median, centroids))}
-\hlstd{disp_time_slicesBS} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(time_slicesBS,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(median, centroids))}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
+## Let's calculate a different metric for the disparity-through-time analysis
+disp_time_binsBS <- dispRity(time_binsBS, metric = c(median, centroids))
+disp_time_slicesBS <- dispRity(time_slicesBS, metric = c(median, centroids))
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## Let's have a look at the disparity-through-time using in the time-binned data}
-\hlkwd{summary}\hlstd{(disp_time_binsBS)}
-\end{alltt}
-\begin{verbatim}
-##   series  n  mean  2.5%   25%   75% 97.5%
-## 1 120-80  8 1.110 0.899 1.075 1.162 1.225
-## 2  80-40 27 1.322 1.275 1.308 1.337 1.361
-## 3   40-0 16 1.324 1.253 1.306 1.347 1.373
-\end{verbatim}
-\begin{alltt}
-\hlcom{## The CIs are calculated as 50 and 95 and the central tendency is the mean by default.}
-\hlcom{## We can change that easily by specifying different values in the options}
-\hlkwd{summary}\hlstd{(disp_time_binsBS,} \hlkwc{CI} \hlstd{=} \hlnum{25}\hlstd{,} \hlkwc{cent.tend} \hlstd{= median,} \hlkwc{rounding} \hlstd{=} \hlnum{0}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-##   series  n median 37.5% 62.5%
-## 1 120-80  8      1     1     1
-## 2  80-40 27      1     1     1
-## 3   40-0 16      1     1     1
-\end{verbatim}
-\begin{alltt}
-\hlcom{## Also, if we happen to have forgotten what was in the disp_time_binsBS object (as in}
-\hlcom{## which options where used to modify the original matrix) we can use the recall option:}
-\hlkwd{summary}\hlstd{(disp_time_binsBS,} \hlkwc{recall} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-## Disparity calculated as: median centroids for 48 dimensions.
-## Data was split using discrete method.
-## Data was bootstrapped 1000 times, using the full bootstrap method.
-##   series  n  mean  2.5%   25%   75% 97.5%
-## 1 120-80  8 1.110 0.899 1.075 1.162 1.225
-## 2  80-40 27 1.322 1.275 1.308 1.337 1.361
-## 3   40-0 16 1.324 1.253 1.306 1.347 1.373
-\end{verbatim}
-\begin{alltt}
-\hlcom{## Note that the recall just prints text. The output of summary.dispRity is always a}
-\hlcom{## data.frame}
-\hlstd{summary_table} \hlkwb{<-} \hlkwd{summary}\hlstd{(disp_time_binsBS,} \hlkwc{recall} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-## Disparity calculated as: median centroids for 48 dimensions.
-## Data was split using discrete method.
-## Data was bootstrapped 1000 times, using the full bootstrap method.
-\end{verbatim}
-\begin{alltt}
-\hlkwd{class}\hlstd{(summary_table)}
-\end{alltt}
-\begin{verbatim}
-## [1] "data.frame"
-\end{verbatim}
-\begin{alltt}
-\hlcom{## Finally we can see the results of the rarefaction analysis for the crown/stem data}
-\hlkwd{head}\hlstd{(}\hlkwd{summary}\hlstd{(disp_crown_stemBS))}
-\end{alltt}
-\begin{verbatim}
-##   series n  mean  2.5%   25%   75% 97.5%
-## 1  crown 3 15.22 10.27 15.20 16.05 16.71
-## 2  crown 4 18.33 14.82 18.19 19.36 19.94
-## 3  crown 5 20.75 17.74 19.48 21.81 22.40
-## 4  crown 6 22.71 19.18 21.75 23.79 24.45
-## 5  crown 7 24.14 20.97 23.47 25.29 26.04
-## 6  crown 8 25.34 21.71 24.65 26.47 27.44
-\end{verbatim}
-\begin{alltt}
-\hlcom{## This outputs a longer table with all the variations of the number of crown/stem taxa}
-\hlcom{## from the maximum (30 and 20) to the minimum (3)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
+<<summary>>=
+## Let's have a look at the disparity-through-time using in the time-binned data
+summary(disp_time_binsBS)
+## The CIs are calculated as 50 and 95 and the central tendency is the mean by default.
+## We can change that easily by specifying different values in the options
+summary(disp_time_binsBS, CI = 25, cent.tend = median, rounding = 0)
+## Also, if we happen to have forgotten what was in the disp_time_binsBS object (as in
+## which options where used to modify the original matrix) we can use the recall option:
+summary(disp_time_binsBS, recall = TRUE)
+## Note that the recall just prints text. The output of summary.dispRity is always a
+## data.frame
+summary_table <- summary(disp_time_binsBS, recall = TRUE)
+class(summary_table)
+
+## Finally we can see the results of the rarefaction analysis for the crown/stem data
+head(summary(disp_crown_stemBS))
+## This outputs a longer table with all the variations of the number of crown/stem taxa
+## from the maximum (30 and 20) to the minimum (3)
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## Graphical options}
-\hlkwd{quartz}\hlstd{(}\hlkwc{width} \hlstd{=} \hlnum{15}\hlstd{,} \hlkwc{height} \hlstd{=} \hlnum{5}\hlstd{) ;} \hlkwd{par}\hlstd{(}\hlkwc{mfrow} \hlstd{= (}\hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,}\hlnum{3}\hlstd{)),} \hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
-\hlcom{## Plotting the score for both groups}
-\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"} \hlstd{,} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{25}\hlstd{,}\hlnum{40}\hlstd{))}
-\hlcom{## Same one with the number of taxa for each data set}
-\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{25}\hlstd{,}\hlnum{40}\hlstd{))}
-\hlcom{## Same one with rarefied data (note how that affects the diversity data as well!)}
-\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlnum{20}\hlstd{,}
-    \hlkwc{ylim} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{25}\hlstd{,}\hlnum{40}\hlstd{))}
-\end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=1\linewidth]{figure/minimal-plot1-1}
-
-}
-
-\end{knitrout}
+<<plot1, fig.width=15, fig.height=5, out.width='1\\linewidth'>>=
+## Graphical options
+quartz(width = 15, height = 5) ; par(mfrow = (c(1,3)), bty = "n")
+## Plotting the score for both groups
+plot(disp_crown_stemBS, type = "discrete" , ylim = c(25,40))
+## Same one with the number of taxa for each data set
+plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, ylim = c(25,40))
+## Same one with rarefied data (note how that affects the diversity data as well!)
+plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, rarefaction = 20,
+    ylim = c(25,40))
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## Graphical options}
-\hlkwd{par}\hlstd{(}\hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
-\hlcom{## The rarefaction curves}
-\hlkwd{plot}\hlstd{(disp_crown_stemBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{rarefaction} \hlstd{=} \hlstr{"plot"}\hlstd{)}
-\hlcom{## Note that the function automatically draws the curves for each series and splits}
-\hlcom{## the plot screen accordingly.}
-\end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=0.5\linewidth]{figure/minimal-plot2-1}
-
-}
-
-\end{knitrout}
+<<plot2, fig.width=5, fig.height=8, out.width='0.5\\linewidth'>>=
+## Graphical options
+par(bty = "n")
+## The rarefaction curves
+plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, rarefaction = "plot")
+## Note that the function automatically draws the curves for each series and splits
+## the plot screen accordingly.
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## Graphical options}
-\hlkwd{quartz}\hlstd{(}\hlkwc{width} \hlstd{=} \hlnum{10}\hlstd{,} \hlkwc{height} \hlstd{=} \hlnum{5}\hlstd{) ;} \hlkwd{par}\hlstd{(}\hlkwc{mfrow} \hlstd{= (}\hlkwd{c}\hlstd{(}\hlnum{1}\hlstd{,}\hlnum{2}\hlstd{)),} \hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
-\hlcom{## The disparity-through-time data in a time-binned way (with lines rather than}
-\hlcom{## boxes)}
-\hlkwd{plot}\hlstd{(disp_time_binsBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{discrete_type} \hlstd{=} \hlstr{"line"}\hlstd{)}
-\hlcom{## The disparity-through-time data in a time-sliced ways}
-\hlkwd{plot}\hlstd{(disp_time_slicesBS,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{)}
-\end{alltt}
-\end{kframe}
+<<plot3, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
+## Graphical options
+quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
+## The disparity-through-time data in a time-binned way (with lines rather than
+## boxes)
+plot(disp_time_binsBS, type = "discrete", discrete_type = "line")
+## The disparity-through-time data in a time-sliced ways
+plot(disp_time_slicesBS, type = "continuous")
 
-{\centering \includegraphics[width=1\linewidth]{figure/minimal-plot3-1}
-
-}
-
-\end{knitrout}
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## Graphical options}
-\hlkwd{par}\hlstd{(}\hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
-\hlcom{## First, plotting the continuous data}
-\hlkwd{plot}\hlstd{(disp_time_slicesBS,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{,} \hlkwc{ylim}\hlstd{=}\hlkwd{c}\hlstd{(}\hlnum{0.5}\hlstd{,}\hlnum{2}\hlstd{))}
-\hlcom{## Second, adding the discrete data}
-\hlkwd{par}\hlstd{(}\hlkwc{new} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\hlkwd{plot}\hlstd{(disp_time_binsBS,} \hlkwc{type} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{discrete_type} \hlstd{=} \hlstr{"line"}\hlstd{,} \hlkwc{ylim}\hlstd{=}\hlkwd{c}\hlstd{(}\hlnum{0.5}\hlstd{,}\hlnum{2}\hlstd{),}
-    \hlkwc{xlab}\hlstd{=}\hlstr{""}\hlstd{,} \hlkwc{ylab}\hlstd{=}\hlstr{""}\hlstd{)}
-\end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=.5\linewidth]{figure/minimal-plot4-1}
-
-}
-
-\end{knitrout}
+<<plot4, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
+## Graphical options
+par(bty = "n")
+## First, plotting the continuous data
+plot(disp_time_slicesBS, type = "continuous", ylim=c(0.5,2))
+## Second, adding the discrete data
+par(new = TRUE)
+plot(disp_time_binsBS, type = "discrete", discrete_type = "line", ylim=c(0.5,2),
+    xlab="", ylab="")
+@
 
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{## Graphical options}
-\hlkwd{par}\hlstd{(}\hlkwc{bty} \hlstd{=} \hlstr{"n"}\hlstd{)}
-\hlcom{## A plot with many options!}
-\hlkwd{plot}\hlstd{(disp_time_slicesBS,} \hlkwc{CI} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{seq}\hlstd{(}\hlkwc{from}\hlstd{=}\hlnum{10}\hlstd{,} \hlkwc{to}\hlstd{=}\hlnum{100}\hlstd{,} \hlkwc{by}\hlstd{=}\hlnum{10}\hlstd{)),} \hlkwc{cent.tend} \hlstd{= sd,}
-    \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{,} \hlkwc{diversity} \hlstd{=} \hlstr{"log"}\hlstd{,} \hlkwc{col} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"black"}\hlstd{,} \hlkwd{rainbow}\hlstd{(}\hlnum{10}\hlstd{)),}
-    \hlkwc{ylab} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"Disparity"}\hlstd{,} \hlstr{"log(Diversity)"}\hlstd{),} \hlkwc{xlab} \hlstd{=} \hlstr{"Time (in in units from
-    past to present)"}\hlstd{,} \hlkwc{time.series} \hlstd{=} \hlnum{FALSE}\hlstd{,} \hlkwc{main} \hlstd{=} \hlstr{"Many options..."}\hlstd{)}
-\end{alltt}
-\end{kframe}
-
-{\centering \includegraphics[width=.8\linewidth]{figure/minimal-plot5-1}
-
-}
-
-\end{knitrout}
+<<plot5, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+## Graphical options
+par(bty = "n")
+## A plot with many options!
+plot(disp_time_slicesBS, CI = c(seq(from=10, to=100, by=10)), cent.tend = sd,
+    type = "continuous", diversity = "log", col = c("black", rainbow(10)),
+    ylab = c("Disparity", "log(Diversity)"), xlab = "Time (in in units from
+    past to present)", time.series = FALSE, main = "Many options...")
+@
 
@@
 \label{whatsleft}
-As stated at the start of the demo, this version 0.1.0 is still in development and many parts are missing.
+As stated at the start of the demo, this version \Sexpr{packageVersion("dispRity")} is still in development and many parts are missing.
 Here are the new functionalities that will be implemented before the proper release version (v.1).

### 2015-10-06 doc/dispRity-tutorial.Rnw
Minor updates on tutorial

@@
 
-This a quick demo to go through the beta version of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}).
+This is a quick demo to go through the beta version of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}).
 Many parts are still missing (see section \ref{whatsleft}) but this version should be running all right and give an idea of what's implemented in the package.
@@
 library(dispRity)
-
 ## An ordinated matrix with the tips only
@@
 head(BeckLee_mat50[,1:5])
-
 ## An ordinated matrix with the tips and the nodes (named nX)
@@
 tail(BeckLee_mat99[,1:5], 2)
-
 ## A list of first and last occurrences data for some fossils
@@
 head(BeckLee_ages)
-
 ## And a phylogeny:
@@
 MyData_mat50 <- BeckLee_mat50
-
 ## My ordinated matrix with tips and nodes
 MyData_mat99 <- BeckLee_mat99
-
 ## My first and last occurence data
 MyData_ages <- BeckLee_ages
-
 ## My phylogenetic tree
@@
 sliced_data <- time.series(MyData_mat99, MyData_tree, method = "continuous",
-    model = "acctran", time = 5, FADLAD = MyData_ages)
-
+    model = "acctran", time = 9, FADLAD = MyData_ages)
 ## Bootstrapping the data
 bootstrapped_data <- boot.matrix(sliced_data, 100)
-
 ## Calculating disparity
 sum_of_ranges <- dispRity(bootstrapped_data, metric = c(sum, ranges))
-
 ## Summarising the results
 summary(sum_of_ranges)
-plot(sum_of_ranges, type = "continuous")
+plot(sum_of_ranges)
 @
@@
 class(sum_of_ranges)
-
 ## We can summarise it using the S3 method print.dispRity
@@
 stem <- drop.tip(MyData_tree, tip = crown)$tip.label
-
 ## We then have to feed this information in a data frame with one column
@@
     ncol = 1, dimnames = list(c(crown, stem))))
-
 ## We then can use the customised series function
 crown_stem <- cust.series(MyData_mat50, factors)
-
 ## This created a dispRity object containing two series: crown and stem
@@
 crown_stem
-
 ## This object contains three elements
 names(crown_stem)
-
 ## With "data" being the list sub-matrices
 str(crown_stem$data)
-
 ## "taxa" being the list of taxa in the original ordinated matrix
 str(crown_stem$taxa)
-
 ## and finally, "series" containing information on the series type (custom) and names
@@
     method = "discrete", time = c(120, 80, 40, 0))
-
 ## Note that the function provides a warning saying that tips where single
@@
     method = "discrete", time = c(120, 80, 40, 0), FADLAD = MyData_ages)
-
 ## To entirely avoid the warning we could collect the occurrence span data
@@
 ## by default.
-
 time_bins
@@
 time_slices
-
 ## Note that in the same way as for the discrete method, we can also
@@
 time_slicesBS <- boot.matrix(time_slices)
-
 ## And just to remind us what's in each objects
@@
 In our case, we are interested in looking at the results in a discrete way.
+Note that the plot \texttt{type} can be left missing.
+The function will then use \texttt{"discrete"} as default or \texttt{"continuous"} if the input data came from the \texttt{time.series} function with \texttt{method = "continuous"}.
 
@@
 ## Plotting the score for both groups
-plot(disp_crown_stemBS, type = "discrete" , ylim = c(25,40))
+plot(disp_crown_stemBS, ylim = c(25,40))
 ## Same one with the number of taxa for each data set
-plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, ylim = c(25,40))
+plot(disp_crown_stemBS, diversity = TRUE, ylim = c(25,40))
 ## Same one with rarefied data (note how that affects the diversity data as well!)
-plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, rarefaction = 20,
-    ylim = c(25,40))
+plot(disp_crown_stemBS, diversity = TRUE, rarefaction = 20, ylim = c(25,40))
 @
@@
 ## The rarefaction curves
-plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, rarefaction = "plot")
+plot(disp_crown_stemBS, diversity = TRUE, rarefaction = "plot")
 ## Note that the function automatically draws the curves for each series and splits
@@
 ## boxes)
-plot(disp_time_binsBS, type = "discrete", discrete_type = "line")
+plot(disp_time_binsBS, discrete_type = "line")
 ## The disparity-through-time data in a time-sliced ways
-plot(disp_time_slicesBS, type = "continuous")
+plot(disp_time_slicesBS)
 
@@
 ## First, plotting the continuous data
-plot(disp_time_slicesBS, type = "continuous", ylim=c(0.5,2))
+plot(disp_time_slicesBS, ylim=c(0.5,2))
 ## Second, adding the discrete data
 par(new = TRUE)
-plot(disp_time_binsBS, type = "discrete", discrete_type = "line", ylim=c(0.5,2),
-    xlab="", ylab="")
+plot(disp_time_binsBS, discrete_type = "line", ylim=c(0.5,2), xlab="", ylab="")
 @

### 2015-10-06 doc/dispRity-ecology-demo.Rnw
Added ecological data example

\documentclass{article}
\usepackage[sc]{mathpazo}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}
\usepackage{url}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=true,bookmarksnumbered=true,bookmarksopen=true,bookmarksopenlevel=2,
 breaklinks=false,pdfborder={0 0 1},backref=false,colorlinks=false]
 {hyperref}
\hypersetup{
 pdfstartview={XYZ null null 1}}

\newcommand{\dispRity}{\texttt{dispRity} }
\newcommand{\R}{\texttt{R} }

\begin{document}
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.path='figure/minimal-', fig.align='center', fig.show='hold')
options(formatR.arrow=TRUE,width=90)
@

\title{dispRity demo for ecological data}

\author{Thomas Guillerme}

\maketitle

This is a quick demo for using the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) in ecological analysis.
See the \href{https://github.com/TGuillerme/dispRity/blob/master/dispRity-tutorial.0.1.0.pdf}{dispRity-tutorial} vignette for a general demo of the \dispRity package.

\section{Before starting}

\dispRity is a package for calculating disparity in \texttt{R}.
To keep it short, this package allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.
These ordinated matrices can be called the morphospace (when using morphometric data) or the cladisto-space (for cladistic data) and will represent the total spread of the forms available in the dataset.
However, keep in mind that if you are using other type of data, say ecological data for example, then the ordinated matrix will represent the eco-space which represents the total spread of ecological traits available in the dataset.
In the rest of this demo, I will sometimes refer to the ordinated matrix as the any-o-space where you can replace \textit{any} by any prefix you want your space to be based on!

\subsection{Installation}
You can install this package easily if you use the latest version of \R and \texttt{devtools}.

<<installation, eval=FALSE>>=
install.packages("devtools")
library(devtools)
install_github("TGuillerme/dispRity", ref = "master")
library(dispRity)
@

WARNING: this is going to load the latest version, it might contain bugs.
A further version of this vignette will be based on the release version.

Note that we use the \texttt{release} branch here which is version \Sexpr{packageVersion("dispRity")}.
If you want the piping-hot (and full of bugs) version, change the option \texttt{ref = "release"} to \texttt{ref = "master"}.
This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.

\subsection{Data}

<<data, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## Loading demo and the package data
library(dispRity)
## let's use the matrix from McClean (unpubl.)
data(McClean_data)
## This dataset contains an ordinated matrix
ord_matrix <- McClean_data$ordination
## As well as two list of factors for separating the rows of the matrix
treatments <- McClean_data$treatment
depth <- McClean_data$depth
@

\section{A classic PCA}
A classical way to represent this ordinated data will be to use PCA plots

<<PCAplot, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## The x and y axis represent the two first dimensions of the PCA
x <- ord_matrix[, 1]
y <- ord_matrix[, 2]
## The colors will represent the treatments
cols <- sub("treat1", "red", treatments)
cols <- sub("treat2", "blue", cols)
## The symbols will represent the depth
pchs <- sub(1, 16, depth)
pchs <- as.numeric(sub(2, 17, pchs))
## Graphical option
par(bty = "n")
## A classic PCA plot
plot(x, y, col = cols, pch = pchs, xlab = "PCA1", ylab = "PCA2",
    xlim = range(x) + c(0, 100))
@

However, the problem, is it ignores the 18 other axis of the ordination and the PCA axis 1 and 2 do not represent a biological reality \textit{per se} but more some ordinations of correlations between the data and some factors.
Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset.

\section{Splitting the data}

First let's make a factor table.

<<factors>>=
## Making the factor table
factors <- as.data.frame(matrix(data = c(treatments, depth), nrow = nrow(ord_matrix),
    ncol = 2, byrow = FALSE, dimnames = list(rownames(ord_matrix))))
names(factors)<-c("Treat", "Depth")
## And here is what it looks like
head(factors)
@

Second, let's split the data according to these factors to create the subsamples of the ordinated space by using the \texttt{cust.series} function:

<<cust.series>>=
## Splitting the ordinated space into four subsamples
customised_series <- cust.series(ord_matrix, factors)
## Note that the output of dispRity functions are dispRity objects
class(customised_series)
## These objects are automatically printed in a summary method (calling S3 print.dispRity)
## giving information about the object
customised_series
@

For more details on the \dispRity objects, see the \href{https://github.com/TGuillerme/dispRity/blob/master/dispRity-tutorial.0.1.0.pdf}{dispRity-tutorial} vignette.
Basically the idea is to avoid jamming the \R console such as when using:

<<dispRity class, eval=FALSE>>=
## Summarise the object
str(customised_series)
@

\subsection{Calculating disparity}

Now we're going to see the functionalities of the core function of this package: the \dispRity function.
This function is a modulable function that allow to simply (and quickly!) calculate disparity from a matrix.
Disparity can be calculated in many ways, this function is a tool to measure disparity \textit{as defined by the user} (and here's where the modulable part comes in).
For more details on disparity, see the \href{https://github.com/TGuillerme/dispRity/blob/master/dispRity-tutorial.0.1.0.pdf}{dispRity-tutorial} vignette.

One can usually decompose the disparity metrics into two elements:
\begin{enumerate}
\item the \textbf{class metric} that is a descriptor of the matrix. For example describing the ranges of each column in the matrix or the euclidean distances between each row and the centroid of the matrix.
\item the \textbf{summary metric} that is a summary of the class metric values. For example, the sum of the ranges or the median of the euclidean distances. 
\end{enumerate}
Basically the combination can be infinite between the class and summary metrics.
For example, people might want to measure the median variances of the axis or the product of the distances from the centroid.
However, it is probable that some metrics are better to reflect some biological aspects of the any-o-space than others (see section \ref{whatsleft})...

In practice, the \dispRity function intakes a pair of class and summary metrics as a definition of disparity.
Several of these metrics are implemented in other packages (like \texttt{stats::median}, \texttt{base::sum}, etc.) and this package proposes several metrics listed in \texttt{dispRity.metric} (see \texttt{?dispRity.metric}).
It is even possible to use your very own class and summary metrics!
This will be actually heavily encouraged and facilitate with the \texttt{make.metric} function in a future version (see \ref{whatsleft}).

To use these metrics pairs in the \dispRity function, it's pretty easy:

<<dispRity1>>=
## For example, let's calculate the median distance between each plot and the
## centroid of the ordinated space
disparity <- dispRity(customised_series, metric = c(median, centroids))
## Note that disparity is a dispRity object and printing it just gives details
## on the object, not the results. We need to use summary.dispRity (S3) to get
## the results.
summary(disparity)
@

Note that we calculated the median distance between plots and the centroid but the output displays mean values.
This is because summary will, by default, summarise the data using the mean value.
Here the mean represents the mean of the median distance for each series which is a bit useless (i.e. the mean of one value is that same value).
We can display the median as well using:

<<dispRity2>>=
summary(disparity, cent.tend=median)
## Or even the product! It won't affect the results
summary(disparity, cent.tend=prod)
@

This is because we did calculate the noise within our data.
We can classically do so by bootstrapping the data!

\subsection{Bootstrapping the data}

The \dispRity pacakge also provides easy way to bootstrap the data via the \texttt{boot.matrix} function.
We can even rarefy the data to see the effect of the number of plots per series:
For more details on the bootstrapping options, see the \href{https://github.com/TGuillerme/dispRity/blob/master/dispRity-tutorial.0.1.0.pdf}{dispRity-tutorial} vignette.

<<boot.matrix>>=
bootstrapped_data <- boot.matrix(customised_series, bootstraps=100)
rarefied_data <- boot.matrix(customised_series, bootstraps=100, rarefaction=TRUE)
## Note that the output is a dispRity object giving some details on the series and the bootstraps
bootstrapped_data
@

We can now rerun a more robust disparity analysis:

<<dispRity3>>=
disparity_BS <- dispRity(bootstrapped_data, metric = c(median, centroids))
disparity_rare <- dispRity(rarefied_data, metric = c(median, centroids))
## Note that calculation time is increased!
@

We can now summarise the data using various options such as the confidence intervals levels and the central tendency.

<<summary>>=
## The default:
summary(disparity_BS)
## The CIs are calculated as 50 and 95 and the central tendency is the mean by default.
## But we can specify different options
summary(disparity_BS, CI=90, cent.tend=median)

## Finally we can see the results of the rarefaction analysis:
head(summary(disparity_rare))
## This outputs a longer table with all the variations of plots down to 3 plots per series.
@

<<disparity plots, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## Graphical option
par(bty = "n")
## Plotting the score for each groups
plot(disparity_BS)
@

<<rare plots, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
## Graphical options
quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
## The same but looking at the number of plots
plot(disparity_rare, diversity = TRUE)
## With the same number of plots per group
plot(disparity_rare, diversity = TRUE, rarefaction = 17)

@

<<rare plots2, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## Graphical option
par(bty = "n")
## Plotting the rarefaction curves
plot(disparity_rare, rarefaction = "plot")
@

\end{document}

### 2015-10-06 doc/dispRity-ecology-demo.Rnw
ecology demo

@@
 This is a quick demo for using the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) in ecological analysis.
-See the \href{https://github.com/TGuillerme/dispRity/blob/master/dispRity-tutorial.0.1.0.pdf}{dispRity-tutorial} vignette for a general demo of the \dispRity package.
+See the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos} for a general demo of the \dispRity package.
 
@@
 To keep it short, this package allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.
-These ordinated matrices can be called the morphospace (when using morphometric data) or the cladisto-space (for cladistic data) and will represent the total spread of the forms available in the dataset.
-However, keep in mind that if you are using other type of data, say ecological data for example, then the ordinated matrix will represent the eco-space which represents the total spread of ecological traits available in the dataset.
-In the rest of this demo, I will sometimes refer to the ordinated matrix as the any-o-space where you can replace \textit{any} by any prefix you want your space to be based on!
+These ordinated matrices can represent the any-o-space of possibilities of your analyses (e.g. for morphological data, the morphospace represents all the possible morphologies).
 
@@
 
-WARNING: this is going to load the latest version, it might contain bugs.
-A further version of this vignette will be based on the release version.
-
 Note that we use the \texttt{release} branch here which is version \Sexpr{packageVersion("dispRity")}.
@@
 
+For this example with ecological data we are going to use data from McClean (unpubl.) that contains an ordination of a distance matrix between different experimental plots.
+
 <<data, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
@@
 data(McClean_data)
-## This dataset contains an ordinated matrix
+## This dataset contains an ordinated matrix (in 20 dimensions) of the distance
+## between experimental plots.
 ord_matrix <- McClean_data$ordination
-## As well as two list of factors for separating the rows of the matrix
+## As well as two list of different factors affecting each experimental plot:
+## the treatment and the depth.
 treatments <- McClean_data$treatment
@@
 \section{A classic PCA}
-A classical way to represent this ordinated data will be to use PCA plots
+A classical way to represent this ordinated data will be to use PCA plots.
 
@@
 
+This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix.
 However, the problem, is it ignores the 18 other axis of the ordination and the PCA axis 1 and 2 do not represent a biological reality \textit{per se} but more some ordinations of correlations between the data and some factors.
-Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset.
+Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a \textit{n}-dimensional object.
+In practice, we might be interested in looking how some experimental treatment for example, will affect the position of our experimental plots in this \textit{n}-dimensional object.
+For example: do the experimental plots shift in some specific space of the \textit{n}-dimensional object when depth increases?
 
@@
 
-First let's make a factor table.
+As a first split, one might want to split the data (i.e. the \textit{n}-dimensional object) into subsamples that we want to compare.
+First let's make a factor table:
 
@@
 
-For more details on the \dispRity objects, see the \href{https://github.com/TGuillerme/dispRity/blob/master/dispRity-tutorial.0.1.0.pdf}{dispRity-tutorial} vignette.
+For more details on the \dispRity objects, see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}.
 Basically the idea is to avoid jamming the \R console such as when using:
@@
 Disparity can be calculated in many ways, this function is a tool to measure disparity \textit{as defined by the user} (and here's where the modulable part comes in).
-For more details on disparity, see the \href{https://github.com/TGuillerme/dispRity/blob/master/dispRity-tutorial.0.1.0.pdf}{dispRity-tutorial} vignette.
+For more details on disparity, see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}.
 
@@
 For example, people might want to measure the median variances of the axis or the product of the distances from the centroid.
-However, it is probable that some metrics are better to reflect some biological aspects of the any-o-space than others (see section \ref{whatsleft})...
+However, it is probable that some metrics are better to reflect some biological aspects of the any-o-space than others...
 
@@
 It is even possible to use your very own class and summary metrics!
-This will be actually heavily encouraged and facilitate with the \texttt{make.metric} function in a future version (see \ref{whatsleft}).
+This will be actually heavily encouraged and facilitate with the \texttt{make.metric} function in a future version.
 
@@
 We can even rarefy the data to see the effect of the number of plots per series:
-For more details on the bootstrapping options, see the \href{https://github.com/TGuillerme/dispRity/blob/master/dispRity-tutorial.0.1.0.pdf}{dispRity-tutorial} vignette.
+For more details on the bootstrapping options, see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}.
 
@@
 
+\subsection{Summarising the data}
+
 We can now summarise the data using various options such as the confidence intervals levels and the central tendency.
@@
 
+Finally we can also plot the results using the simple \texttt{plot.dispRity S3} method:
+
 <<disparity plots, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
@@
 @
+
+Or have a look at the effect of the number of experimental plots:
 
@@
 plot(disparity_rare, diversity = TRUE, rarefaction = 17)
+@
 
-@
+Or event have a look at the rarefaction curves:
 

### 2015-10-07 doc/dispRity-ecology-demo.Rnw
Updated ecological example

@@
 
-Note that we use the \texttt{release} branch here which is version \Sexpr{packageVersion("dispRity")}.
-If you want the piping-hot (and full of bugs) version, change the option \texttt{ref = "release"} to \texttt{ref = "master"}.
+Note that we use the \texttt{master} branch here which is version \Sexpr{packageVersion("dispRity")}.
+This is the piping-hot latest version.
+If you want a more stable (and tested version), change \texttt{ref = "master"} to \texttt{ref = "release"}.
 This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.
@@
 ## The colors will represent the treatments
-cols <- sub("treat1", "red", treatments)
-cols <- sub("treat2", "blue", cols)
+cols <- sub("a", "red", treatments)
+cols <- sub("b", "blue", cols)
 ## The symbols will represent the depth
@@
 Several of these metrics are implemented in other packages (like \texttt{stats::median}, \texttt{base::sum}, etc.) and this package proposes several metrics listed in \texttt{dispRity.metric} (see \texttt{?dispRity.metric}).
-It is even possible to use your very own class and summary metrics!
+But it is even possible to use your very own class and summary metrics! Or even any metric you decide makes sense (note that this is not yet implemented in v.\Sexpr{packageVersion("dispRity")}).
 This will be actually heavily encouraged and facilitate with the \texttt{make.metric} function in a future version.

### 2015-10-07 tests/testthat/make.data/make.data.R

@@ -94,7 +94,7 @@ McClean_distance<-as.matrix(dist(McClean_matrix, method="euclidean"))
 McClean_ordination<-cmdscale(McClean_distance, eig = TRUE, k = nrow(McClean_distance)-1)$points
 
 #Factors
-treatment<-c("treat1","treat1","treat2","treat2","treat1","treat1","treat1","treat2","treat2","treat1","treat1","treat2","treat2","treat2","treat2","treat2","treat2","treat2","treat1","treat2","treat2","treat1","treat2","treat2","treat2","treat2","treat1","treat1","treat1","treat2","treat2","treat1","treat1","treat1","treat1","treat1","treat1","treat1","treat1","treat1")
+treatment<-c("a","a","b","b","a","a","a","b","b","a","a","b","b","b","b","b","b","b","a","b","b","a","b","b","b","b","a","a","a","b","b","a","a","a","a","a","a","a","a","a")
 depth<-c(1,2,1,2,1,1,2,1,2,1,2,1,2,1,2,1,2,2,2,1,2,1,1,2,1,2,1,1,2,1,2,1,2,1,2,1,1,1,1,1)
 
 #Output list

### 2015-10-08 doc/dispRity-tutorial.Rnw
Updated demos

@@
 And secondly (2) a phylogenetic tree with the tips matching the ordinated matrix and a \texttt{\$root.time} element for determining the age of the tree.
-You can use the function \texttt{tree.age} provided in this package but not described in this demo for generating the root.time automatically (see \texttt{?tree.age}).
+You can use the function \texttt{tree.age} provided in this package but not described in this demo for generating the \texttt{\$root.time} automatically (see \texttt{?tree.age}).
 In the following examples, just replace all the \texttt{BeckLee} mentions in this demo by your very own data:
@@
 str(sum_of_ranges)
-## That's a 2512 lines output!
+## That's a more than 4000 lines of output
 @
@@
     matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
-    ncol = 1, dimnames = list(c(crown, stem))))
+    ncol = 1, dimnames = list(c(crown, stem), "Group")))
 ## We then can use the customised series function
@@
 Any type of factors can be used in this example (e.g. body mass categories, habitat, diet, etc...).
+See the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos} for other specific examples.
 
@@
 Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a fix number of equidistant time-bins/slices.
-Additionally, we can also input a data frame containing the first/last occurrence data (\texttt{FADLAD}) for taxa that span over a longer time than the tips/nodes age.
+Additionally, we can also input a data frame containing the First/Last Occurrence Data (\texttt{FADLAD}) for taxa that span over a longer time than the tips/nodes age.
 
@@
 We now have three time bins of 44.50368 Ma each. But hey, that's a bit silly, let's stick to bins of 40 Ma.
-When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1\% of the the tree length as long as at least three elements are present at that point in time (an extra 1\% is added until reaching the three elements).
-We can also ask to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between tips AND nodes.
+When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1\% of the the tree length as long as at least three elements are present at that point in time (an extra 1\% is added until reaching the required minimum of three elements).
+We can also ask to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between tips \textit{AND} nodes.
 
@@
     FADLAD = MyData_ages)
-
 time_slices
@@
 The default status is \texttt{FALSE} but it can be set to \texttt{TRUE} to fully rarefy the data (i.e.
-remove $n$ taxa for the number of pseudo-replicates, where $n$ varies from the maximum number of taxa present in the dataset to a minimum of 3 taxa).
+remove $n$ taxa for the number of pseudo-replicates, where $n$ varies from the maximum number of taxa present in the dataset to a minimum of 3 elements).
 It can also be set to a fix \texttt{numeric} value (or a set of numeric values).
@@
 In fact, the \dispRity function intakes two main arguments: the data and the disparity metric.
+%To change in future version!
 The disparity metric boils down to a single value that describes the matrix.
@@
 \subsubsection{\texttt{summary} function}
-This function intakes a \dispRity object plus various options namely the \texttt{CI} values for the confidence intervals levels; the \texttt{cent.tend} for the central tendency to use for summarising the results and two \textit{visual} options which are whether to recall the \dispRity options and how much digits are wanted in the results.
+This function intakes a \dispRity object plus various options namely the \texttt{quantile} values for the confidence intervals levels; the \texttt{cent.tend} for the central tendency to use for summarising the results and two \textit{visual} options which are whether to recall the \dispRity options and how much digits are wanted in the results.
 
@@
 summary(disp_time_binsBS)
-## The CIs are calculated as 50 and 95 and the central tendency is the mean by default.
+## The 50 and 95 quantiles are calculated and the central tendency is the mean by default.
+## Note that the function also displays the observed disparity (non-bootstrapped).
 ## We can change that easily by specifying different values in the options
-summary(disp_time_binsBS, CI = 25, cent.tend = median, rounding = 0)
+summary(disp_time_binsBS, quantile = 88, cent.tend = sd, rounding = 1)
 ## Also, if we happen to have forgotten what was in the disp_time_binsBS object (as in
@@
 ## This outputs a longer table with all the variations of the number of crown/stem taxa
-## from the maximum (30 and 20) to the minimum (3)
+## from the maximum (30 and 20) to the minimum (3). Note that the observed disparity
+## is NA for rarefied data.
 @
@@
 However, it might be interesting to see how many taxa are present in each group.
-This can be done via the \texttt{diversity} option.
+This can be done via the \texttt{elements} option.
 A third interesting option is to plot the rarefied data if a rarefaction was applied on the data with the \texttt{boot.matrix} function.
@@
 ## Same one with the number of taxa for each data set
-plot(disp_crown_stemBS, diversity = TRUE, ylim = c(25,40))
-## Same one with rarefied data (note how that affects the diversity data as well!)
-plot(disp_crown_stemBS, diversity = TRUE, rarefaction = 20, ylim = c(25,40))
+plot(disp_crown_stemBS, elements = TRUE, ylim = c(25,40))
+## Same one with rarefied data (note how that affects the elements data as well!)
+plot(disp_crown_stemBS, elements = TRUE, rarefaction = 20, ylim = c(25,40))
 @
@@
 ## The rarefaction curves
-plot(disp_crown_stemBS, diversity = TRUE, rarefaction = "plot")
+plot(disp_crown_stemBS, elements = TRUE, rarefaction = "plot")
 ## Note that the function automatically draws the curves for each series and splits
@@
 The \texttt{discrete\_type} option allows to switch between \texttt{"box"} or \texttt{"line"} for showing the results in one or the other format.
+We can also plot the observed data when existing (i.e. not with some rarefaction levels) by using \texttt{observed = TRUE}.
 Let's look at that with the time binned data (even though there is only three time bins).
@@
 ## The disparity-through-time data in a time-binned way (with lines rather than
-## boxes)
-plot(disp_time_binsBS, discrete_type = "line")
+## boxes) with the observed data as crosses
+plot(disp_time_binsBS, discrete_type = "line", observed = TRUE)
 ## The disparity-through-time data in a time-sliced ways
@@
 ## A plot with many options!
-plot(disp_time_slicesBS, CI = c(seq(from=10, to=100, by=10)), cent.tend = sd,
-    type = "continuous", diversity = "log", col = c("black", rainbow(10)),
+plot(disp_time_slicesBS, quantile = c(seq(from=10, to=100, by=10)), cent.tend = sd,
+    type = "continuous", elements = "log", col = c("black", rainbow(10)),
     ylab = c("Disparity", "log(Diversity)"), xlab = "Time (in in units from
-    past to present)", time.series = FALSE, main = "Many options...")
+    past to present)", time.series = FALSE, observed = TRUE, main = "Many options...")
 @

### 2015-10-08 doc/dispRity-ecology-demo.Rnw

@@
 
-<<data, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+<<data>>=
 ## Loading demo and the package data
@@
 summary(disparity_BS)
-## The CIs are calculated as 50 and 95 and the central tendency is the mean by default.
+## The quantiles are calculated as 50 and 95 and the central tendency is the mean by default.
 ## But we can specify different options
-summary(disparity_BS, CI=90, cent.tend=median)
+summary(disparity_BS, quantile=90, cent.tend=median)
 
@@
 ## The same but looking at the number of plots
-plot(disparity_rare, diversity = TRUE)
+plot(disparity_rare, elements = TRUE)
 ## With the same number of plots per group
-plot(disparity_rare, diversity = TRUE, rarefaction = 17)
+plot(disparity_rare, elements = TRUE, rarefaction = 17)
 @

### 2015-10-09 doc/dispRity-ecology-demo.Rnw

@@
 library(devtools)
-install_github("TGuillerme/dispRity", ref = "master")
+install_github("TGuillerme/dispRity", ref = "release")
 library(dispRity)
@@
 
-Note that we use the \texttt{master} branch here which is version \Sexpr{packageVersion("dispRity")}.
-This is the piping-hot latest version.
-If you want a more stable (and tested version), change \texttt{ref = "master"} to \texttt{ref = "release"}.
+Note that we use the \texttt{release} branch here which is version 0.1.1.
+If you want the piping-hot (and full of bugs) version, change the option \texttt{ref = "release"} to \texttt{ref = "master"}.
 This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.

### 2015-10-14 doc/dispRity-metrics.Rnw
metric vignette INIT

\documentclass{article}
\usepackage[sc]{mathpazo}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}
\usepackage{url}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=true,bookmarksnumbered=true,bookmarksopen=true,bookmarksopenlevel=2,
 breaklinks=false,pdfborder={0 0 1},backref=false,colorlinks=false]
 {hyperref}
\hypersetup{
 pdfstartview={XYZ null null 1}}

\newcommand{\dispRity}{\texttt{dispRity} }
\newcommand{\R}{\texttt{R} }

\begin{document}
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.path='figure/minimal-', fig.align='center', fig.show='hold')
options(formatR.arrow=TRUE,width=90)
@

\title{dispRity metrics}

\author{Thomas Guillerme}

\maketitle

This vignette details how the disparity metrics are implemented in the \dispRity package (v.\Sexpr{packageVersion("dispRity")}).

\section{Disparity metrics}
There are many ways for measuring disparity.
In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. MDS, PCA, PCO, PCoA).
For example, one can look at ellipsoid hyper-volume of the ordinated space \cite{DonohueDim}, the sum and the product of the ranges and variances of the ordinated space dimensions \cite{Wills1994} or the median position of the elements in the ordinated space relative to its centroid \cite{GuillermeSTD}.
Of course, there is many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks or some being more generalist.

Because of this great diversity of metric, the package \dispRity has not a unique way to measure disparity but rather proposes to facilitate users to define their own disparity metric that will best suit to their particular analysis.
In fact, the core function of the package, \dispRity, allows the user to define any metric in the \texttt{metric} argument.
However, because of the infrastructure of the package (especially the on of the \dispRity objects -- see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}), the \texttt{metric} argument has to follow certain slightly restrictive rules described in detail below.
\begin{enumerate}
    \item \texttt{metric} must be between one or three \texttt{function} objects
    \item \texttt{metric} must intake a \texttt{matrix} or \texttt{vector} object
    \item \texttt{metric} individual functions must be classified in one of the three levels
    \item \texttt{metric} one of the functions must be a level 1 function
\end{enumerate}

\section{The function levels}
The metric functions levels determine the dimensional level of decomposition of the input \texttt{matrix}.
In other words, each level designate the dimensions of the output: either three (a \texttt{matrix}); two (a \texttt{vector}); or one (a single \texttt{numeric} value) dimensions (see \ref{fun_type}).

\begin{figure}[!htbp]
\centering
    \includegraphics[keepaspectratio=true]{dispRity_fun.pdf}
    \caption{Illustration of the different levels of output of a function with an input \texttt{matrix}}
    \label{fun_type}
\end{figure}

\newpage

\subsection{A level 1 function}
A level 1 function for example, will decompose the \texttt{matrix} (or a \texttt{vector} output from a level 2 function, see below) into a single value:

<<level 1 example>>=
## First, let's create a dummy matrix
dummy_matrix <- matrix(rnorm(12), 4, 3)
## Any summary metric such as mean or median are good examples of level 1
## functions as they reduce the matrix to a single dimensions (value):
mean(dummy_matrix)
median(dummy_matrix)
@

\subsection{A level 2 function}
A level 2 function for example, will decompose the \texttt{matrix} into a \texttt{vector}.
Several level 2 functions are implemented in \dispRity (see \texttt{?dispRity.metric}) such as the \texttt{variances} function that calculates the variance of each column of the ordinated matrix (or axis/dimensions):

<<level 2 example>>=
## We can define a different function that outputs the products of the rows:
rows <- function(matrix) apply(matrix, 1, prod)
## Rows is a level 2 function as well (outputs a vector of 4 values):
rows(dummy_matrix)
@

\subsection{A level 3 function}
Finally a level 3 function will transform the \texttt{matrix} into another \texttt{matrix}.
Note that the dimension of the output \texttt{matrix} don't need to match the the input \texttt{matrix}:

<<level 3 example>>=
## The var functions computes the variance/covariance matrix which will be a
## three by three matrix
var(dummy_matrix)
@

\section{\texttt{make.metric}}
Of course, functions can be more complex and involve multiple operations such as the \texttt{centroids} function (see \texttt{?dispRity.metric}) that calculates the euclidean distance between each elements and the ordinated space centroid.
The \texttt{make.metric} function implemented in \dispRity is designed to help testing and finding the level of the functions.
Concretely, This function tests:
\begin{enumerate}
  \item if your function can deal with a matrix as an \texttt{input}.
  \item which level is your function (level 1, 2 or 3, see \texttt{?dispRity.metric}).
  \item whether the function can properly be implemented in the \texttt{dispRity} function.
\end{enumerate}
For example, let's see if the functions described above are the right levels:

<<make.metric example>>=
## First we need to load the package
library(dispRity)
## Which level is the mean function? And is able to be used in the dispRity function?
make.metric(mean)
## Which level is the rows function? And is able to be used in the dispRity function?
make.metric(rows)
## Note that the function also tells us that we will need a level 1 function as well.
## We'll cover this below.
## Which level is the var function? And is able to be used in the dispRity function?
make.metric(var)
@

\section{metric argument in \dispRity}
Of course, functions can be more complex and involve multiple operations such as the \texttt{centroids} function (see \texttt{?dispRity.metric}) that calculates the euclidean distance between each elements and the ordinated space centroid.
The \texttt{make.metric} function implemented in \dispRity is designed to help testing and finding the level of the functions.
Concretely, This function tests:
\begin{enumerate}
  \item if your function can deal with a matrix as an \texttt{input}.
  \item which level is your function (level 1, 2 or 3, see \texttt{?dispRity.metric}).
  \item whether the function can properly be implemented in the \texttt{dispRity} function.
\end{enumerate}
For example, let's see if the functions described above are the right levels:

\bibliographystyle{vancouver}
\bibliography{References}

\end{document}

### 2015-10-15 doc/dispRity-metrics.Rnw
added metric implementation details

@@
 \section{metric argument in \dispRity}
-Of course, functions can be more complex and involve multiple operations such as the \texttt{centroids} function (see \texttt{?dispRity.metric}) that calculates the euclidean distance between each elements and the ordinated space centroid.
-The \texttt{make.metric} function implemented in \dispRity is designed to help testing and finding the level of the functions.
-Concretely, This function tests:
-\begin{enumerate}
-  \item if your function can deal with a matrix as an \texttt{input}.
-  \item which level is your function (level 1, 2 or 3, see \texttt{?dispRity.metric}).
-  \item whether the function can properly be implemented in the \texttt{dispRity} function.
-\end{enumerate}
-For example, let's see if the functions described above are the right levels:
+Using this metric structure, on can easily use any disparity metric in the \dispRity function as follows:
+
+<<dispRity example>>=
+## First we need to load the data
+data(BeckLee_mat50)
+## Measuring disparity as the standard deviation of all the value of the ordinated
+## matrix (level 1 function).
+summary(dispRity(BeckLee_mat50, metric = sd))
+## Measuring disparity as the standard deviation of the variance of each axis of
+## the ordinated matrix (level 1 and level 2 functions).
+summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))
+## Measuring disparity as the standard deviation of the variance of each axis of
+## the variance covariance matrix (level 1, level 2 and level 3 functions).
+summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round=10)
+## Note that the order of each function in the metric argument does not matter,
+## the algorithm always classifies them by levels and computes first level 3,
+## then level 2 and finally level 1.
+summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)), round=10)
+## Both ways output the same disparity values.
+@
 

### 2015-11-14 doc/dispRity-test.Rnw
Test updates
vignette INIT

--- 2015-10-15 doc/dispRity-metrics.Rnw
@@
 
-\title{dispRity metrics}
+\title{Testing differences in \texttt{dispRity}}
 
@@
 
-This vignette details how the disparity metrics are implemented in the \dispRity package (v.\Sexpr{packageVersion("dispRity")}).
+This vignette details how to apply statistical test to \dispRity objects in the \dispRity package (v.\Sexpr{packageVersion("dispRity")}).
 
-\section{Disparity metrics}
-There are many ways for measuring disparity.
-In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. MDS, PCA, PCO, PCoA).
-For example, one can look at ellipsoid hyper-volume of the ordinated space \cite{DonohueDim}, the sum and the product of the ranges and variances of the ordinated space dimensions \cite{Wills1994} or the median position of the elements in the ordinated space relative to its centroid \cite{GuillermeSTD}.
-Of course, there is many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks or some being more generalist.
+The \dispRity package allows to easily apply some statistical (or non statistical!) tests to the bootstrapped disparity calculations.
+This is done in a modulable way, quite similar to the way disparity metrics are implemented in this package (see \href{https://github.com/TGuillerme/dispRity}{the vignette about dispRity metrics}).
 
-Because of this great diversity of metric, the package \dispRity has not a unique way to measure disparity but rather proposes to facilitate users to define their own disparity metric that will best suit to their particular analysis.
-In fact, the core function of the package, \dispRity, allows the user to define any metric in the \texttt{metric} argument.
-However, because of the infrastructure of the package (especially the on of the \dispRity objects -- see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}), the \texttt{metric} argument has to follow certain slightly restrictive rules described in detail below.
-\begin{enumerate}
-    \item \texttt{metric} must be between one or three \texttt{function} objects
-    \item \texttt{metric} must intake a \texttt{matrix} or \texttt{vector} object
-    \item \texttt{metric} individual functions must be classified in one of the three levels
-    \item \texttt{metric} one of the functions must be a level 1 function
-\end{enumerate}
+\section{\texttt{test.dispRity}}
 
-\section{The function levels}
-The metric functions levels determine the dimensional level of decomposition of the input \texttt{matrix}.
-In other words, each level designate the dimensions of the output: either three (a \texttt{matrix}); two (a \texttt{vector}); or one (a single \texttt{numeric} value) dimensions (see \ref{fun_type}).
-
-\begin{figure}[!htbp]
-\centering
-    \includegraphics[keepaspectratio=true]{dispRity_fun.pdf}
-    \caption{Illustration of the different levels of output of a function with an input \texttt{matrix}}
-    \label{fun_type}
-\end{figure}
-
-\newpage
+The \texttt{test.dispRity} function allows to apply a user defined test to the disparity scores of the \dispRity object.
 

### 2015-11-18 doc/dispRity-quick-demo.Rnw
Rewriting vignettes INIT

--- 2015-10-08 doc/dispRity-tutorial.Rnw
@@
 
-\title{dispRity demo}
+\title{\dispRity quick demo}
 
-\author{Thomas Guillerme}
+\author{Thomas Guillerme\\guillert@tcd.ie}
 
@@
 
-This is a quick demo to go through the beta version of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}).
-Many parts are still missing (see section \ref{whatsleft}) but this version should be running all right and give an idea of what's implemented in the package.
+This is quick demo aiming to give a quick overview of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}).
+Please refer to \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes, namely the \dispRity tutorial that explains the functions in more details.
 
-\section{Before starting}
+To keep it short, this package allows to summarise ordinated matrices (e.g.
+MDS, PCA, PCO, PCoA) into values called \textbf{disparity}.
+It then allows to visualise and analyse the disparity of the ordinated matrices.
+
+This demo, which will use dummy data will quickly cover the following points:
+\begin{enumerate}
+\item 
+\end{enumerate}
+
+\section{Installation}
+You can install this package easily if you are using the latest version of \R and \texttt{devtools}.
+
+<<installation, eval=FALSE>>=
+install.packages("devtools")
+library(devtools)
+install_github("TGuillerme/dispRity", ref = "release")
+library(dispRity)
+@
+
+\section{}
 

### 2015-11-18 doc/dispRity-quick-demo.Rnw
demo update

@@
 \begin{enumerate}
-\item 
+\item Installing \dispRity
+\item Ordinated matrices
+\item Creating submatrices
+\item Bootstrapping the data
+\item Calculating disparity
+\item Plotting the results
+\item Testing differences
 \end{enumerate}
 
-\section{Installation}
+\section{Installing \dispRity}
 You can install this package easily if you are using the latest version of \R and \texttt{devtools}.
@@
 
-\section{}
+\section{Ordinated matrices}
+Ordinated matrices are transformed distance matrices based on various methods (e.g. MDS, PCA, PCO, PCoA). For this example, we will just generate \textit{dummy} data with 26 elements and 15 variables.
+
+<<data>>=
+## Loading the package
+library(dispRity)
+## random seed for reproducibility
+set.seed(1)
+
+## This is a multivariate matrix:
+dummy_data <- replicate(15, rnorm(26, mean=sample(1:10)))
+rownames(dummy_data) <- LETTERS
+head(dummy_data[,1:5])
+## We can the calculate the distance matrix:
+dummy_dist <- dist(dummy_data)
+## And ordinate it into 10 dimensions:
+dummy_ord <- cmdscale(dummy_dist, k=10)
+head(dummy_ord)
+@ 
+This ordinated data will constitute our multidimensional space (e.g. the ecospace, morphospace, cladistospace, etc...).
+
+\section{Creating submatrices}
+We can now create two different groups if we are interested in looking at the difference in occupancy of the multidimensional space between the vowels and the consonants.
+
+<<cust.data>>=
+## Creating a data.frame that contains information on the letter type
+factors <- data.frame("type"=rep("consonant", 26), row.names=LETTERS,
+    stringsAsFactors=FALSE)
+factors[c(1,5,9,15,21,25),1] <- rep("vowels", 6)
+
+## Splitting the data into sub-matrices according the the factors data.frame
+data_series <- cust.series(dummy_ord, factors)
+@
+
+Note that from now on, objects are attributed the class \dispRity that summarises the object content.
+
+<<class>>=
+class(data_series)
+str(data_series)
+data_series
+@
+
+\section{Bootstrapping the data}
+We can now bootstrap to get rid of problems from outliers and also rarefy the dataset to have the same number of vowels or consonants.
+
+<<boot.data>>=
+## Simple bootstrapping
+boot_data <- boot.matrix(data_series)
+
+## Bootstrapping and rarefaction to get the same number of elements
+## in both categories (vowels and consonants)
+rare_data <- boot.matrix(data_series, rarefaction=6)
+
+## Again both objects can be quickly summarised by just calling them
+boot_data
+rare_data
+@
+
+\section{Calculating disparity}
+We can calculate disparity as the sum of the variance of each dimension for each sub-matrix to measure the spread of the data for vowels of consonants.
+
+<<disparity>>=
+## Calculating disparity for both objects
+disparity_boot <- dispRity(boot_data, metric=c(sum, variances))
+disparity_rare <- dispRity(rare_data, metric=c(sum, variances))
+
+## Again, printing the objects will just give basic information.
+disparity_boot
+
+## To actually get the results, we can use the summary function
+summary(disparity_boot)
+summary(disparity_rare)
+@
+
+\section{Plotting the results}
+As we can guess from the summary, our results differ between consonants and vowels but consonants seems not affected by sample size.
+But let's visualise it:
+
+<<plot1, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
+## Graphical options
+quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
+## The disparity-through-time data in a time-binned way (with lines rather than
+## boxes) with the observed data as crosses
+plot(disparity_boot, main="raw data", elements=TRUE, ylab=c("disparity", "letters"))
+## The disparity-through-time data in a time-sliced ways
+plot(disparity_rare, main="rarefied data", elements=TRUE, ylab=c("disparity", "letters"))
+@
+
+\section{Testing differences}
+Finally, to be sure if the disparity is different between vowels and consonants or not, we can apply a statistical test to both categories:
+
+<<test>>=
+## Graphical options
+quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
+## The disparity-through-time data in a time-binned way (with lines rather than
+## boxes) with the observed data as crosses
+plot(disparity_boot, main="raw data", elements=TRUE, ylab=c("disparity", "letters"))
+## The disparity-through-time data in a time-sliced ways
+plot(disparity_rare, main="rarefied data", elements=TRUE, ylab=c("disparity", "letters"))
+@
 

### 2015-11-19 doc/dispRity-palaeo-demo.Rnw
demo update
Updated palaeo demo

--- 2015-10-09 doc/dispRity-ecology-demo.Rnw
@@
 
-\title{dispRity demo for ecological data}
+\title{\texttt{dispRity} demo for palaeobiologists}
 
-\author{Thomas Guillerme}
+\author{Thomas Guillerme \\ guillert@tcd.ie}
 
@@
 
-This is a quick demo for using the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) in ecological analysis.
-See the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos} for a general demo of the \dispRity package.
+This demo aims to give quick overview of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) for palaeobiology analysis.
+Please refer to \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes, namely the \dispRity tutorial that explains the functions in more details.
 
-\section{Before starting}
+To keep it short, this package allows to easily perform \textbf{disparity-through-time} analysis.
+This type of analysis often starts by ordinating \textbf{morphometric} or \textbf{cladistic} data into a multidimensional object hereafter called the \textbf{morphospace}.
+One might then be interested how the occupancy of the morphospace changes through time by measuring a summary metric of this morphospace through time, called \textbf{disparity}.
+In the morphospace case, disparity might be seen as a value summarising the diversity of morphologies.
 
-\dispRity is a package for calculating disparity in \texttt{R}.
-To keep it short, this package allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.
-These ordinated matrices can represent the any-o-space of possibilities of your analyses (e.g. for morphological data, the morphospace represents all the possible morphologies).
+This demo showcases a typical disparity-through-time analyse with the following steps:
 
-\subsection{Installation}
-You can install this package easily if you use the latest version of \R and \texttt{devtools}.
+\begin{enumerate}
+\item Installing \dispRity
+\item The morphospace
+\item Splitting the morphospace through time
+\item Bootstrapping the data
+\item Calculating disparity
+\item Plotting the results
+\item Testing differences
+\end{enumerate}
+
+\section{Installing \dispRity}
+You can install this package easily if you are using the latest version of \R and \texttt{devtools}.
 
@@
 
-Note that we use the \texttt{release} branch here which is version 0.1.1.
-If you want the piping-hot (and full of bugs) version, change the option \texttt{ref = "release"} to \texttt{ref = "master"}.
-This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.
+\section{The morphospace}
+In this example, we are going to use a subset of the data from \cite{beckancient2014} that contains an ordinated cladistic matrix and a phylogenetic tree containing both the same 50 taxa along side a table that contains some first and last occurence data for several taxa.
+The ordinated matrix will represent our full morphospace, i.e. all the morphologies that ever existed through time (for this dataset).
 
-\subsection{Data}
-
-For this example with ecological data we are going to use data from McClean (unpubl.) that contains an ordination of a distance matrix between different experimental plots.
-
-<<data>>=
+<<data, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
 ## Loading demo and the package data
 library(dispRity)
-## let's use the matrix from McClean (unpubl.)
-data(McClean_data)
-## This dataset contains an ordinated matrix (in 20 dimensions) of the distance
-## between experimental plots.
-ord_matrix <- McClean_data$ordination
-## As well as two list of different factors affecting each experimental plot:
-## the treatment and the depth.
-treatments <- McClean_data$treatment
-depth <- McClean_data$depth
+
+## The morphospace:
+data(BeckLee_mat50)
+head(BeckLee_mat50[,1:5])
+
+dim(BeckLee_mat50)
+## The morphospace contains 50 taxa and has 48 dimensions (or axis)
+
+## A list of first and last occurrences data for some fossils
+data(BeckLee_ages)
+head(BeckLee_ages)
+
+## And a phylogeny:
+data(BeckLee_tree)
+plot(BeckLee_tree, cex=0.8); axisPhylo(root=140)
 @
 
-\section{A classic PCA}
-A classical way to represent this ordinated data will be to use PCA plots.
+Note that you can have an even nicer tree if you use the \texttt{strap} package:
 
-<<PCAplot, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
-## The x and y axis represent the two first dimensions of the PCA
-x <- ord_matrix[, 1]
-y <- ord_matrix[, 2]
-## The colors will represent the treatments
-cols <- sub("a", "red", treatments)
-cols <- sub("b", "blue", cols)
-## The symbols will represent the depth
-pchs <- sub(1, 16, depth)
-pchs <- as.numeric(sub(2, 17, pchs))
-## Graphical option
-par(bty = "n")
-## A classic PCA plot
-plot(x, y, col = cols, pch = pchs, xlab = "PCA1", ylab = "PCA2",
-    xlim = range(x) + c(0, 100))
+<<installation, eval=FALSE>>=
+install.packages(strap)
+library(strap)
+geoscalePhylo(BeckLee_tree, cex.tip=0.8, cex.ts=0.6)
 @
 
-This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix.
-However, the problem, is it ignores the 18 other axis of the ordination and the PCA axis 1 and 2 do not represent a biological reality \textit{per se} but more some ordinations of correlations between the data and some factors.
-Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a \textit{n}-dimensional object.
-In practice, we might be interested in looking how some experimental treatment for example, will affect the position of our experimental plots in this \textit{n}-dimensional object.
-For example: do the experimental plots shift in some specific space of the \textit{n}-dimensional object when depth increases?
+\section{Splitting the morphospace through time}
 
-\section{Splitting the data}
+\section{Bootstrapping the data}
+\section{Calculating disparity}
+\section{Plotting the results}
+\section{Testing differences}
 
-As a first split, one might want to split the data (i.e. the \textit{n}-dimensional object) into subsamples that we want to compare.
-First let's make a factor table:
-
-<<factors>>=
-## Making the factor table
-factors <- as.data.frame(matrix(data = c(treatments, depth), nrow = nrow(ord_matrix),
-    ncol = 2, byrow = FALSE, dimnames = list(rownames(ord_matrix))))
-names(factors)<-c("Treat", "Depth")
-## And here is what it looks like
-head(factors)
-@
-
-Second, let's split the data according to these factors to create the subsamples of the ordinated space by using the \texttt{cust.series} function:
-
-<<cust.series>>=
-## Splitting the ordinated space into four subsamples
-customised_series <- cust.series(ord_matrix, factors)
-## Note that the output of dispRity functions are dispRity objects
-class(customised_series)
-## These objects are automatically printed in a summary method (calling S3 print.dispRity)
-## giving information about the object
-customised_series
-@
-
-For more details on the \dispRity objects, see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}.
-Basically the idea is to avoid jamming the \R console such as when using:
-
-<<dispRity class, eval=FALSE>>=
-## Summarise the object
-str(customised_series)
-@
-
-\subsection{Calculating disparity}
-
-Now we're going to see the functionalities of the core function of this package: the \dispRity function.
-This function is a modulable function that allow to simply (and quickly!) calculate disparity from a matrix.
-Disparity can be calculated in many ways, this function is a tool to measure disparity \textit{as defined by the user} (and here's where the modulable part comes in).
-For more details on disparity, see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}.
-
-One can usually decompose the disparity metrics into two elements:
-\begin{enumerate}
-\item the \textbf{class metric} that is a descriptor of the matrix.
-For example describing the ranges of each column in the matrix or the euclidean distances between each row and the centroid of the matrix.
-\item the \textbf{summary metric} that is a summary of the class metric values. For example, the sum of the ranges or the median of the euclidean distances. 
-\end{enumerate}
-Basically the combination can be infinite between the class and summary metrics.
-For example, people might want to measure the median variances of the axis or the product of the distances from the centroid.
-However, it is probable that some metrics are better to reflect some biological aspects of the any-o-space than others...
-
-In practice, the \dispRity function intakes a pair of class and summary metrics as a definition of disparity.
-Several of these metrics are implemented in other packages (like \texttt{stats::median}, \texttt{base::sum}, etc.) and this package proposes several metrics listed in \texttt{dispRity.metric} (see \texttt{?dispRity.metric}).
-But it is even possible to use your very own class and summary metrics! Or even any metric you decide makes sense (note that this is not yet implemented in v.\Sexpr{packageVersion("dispRity")}).
-This will be actually heavily encouraged and facilitate with the \texttt{make.metric} function in a future version.
-
-To use these metrics pairs in the \dispRity function, it's pretty easy:
-
-<<dispRity1>>=
-## For example, let's calculate the median distance between each plot and the
-## centroid of the ordinated space
-disparity <- dispRity(customised_series, metric = c(median, centroids))
-## Note that disparity is a dispRity object and printing it just gives details
-## on the object, not the results. We need to use summary.dispRity (S3) to get
-## the results.
-summary(disparity)
-@
-
-Note that we calculated the median distance between plots and the centroid but the output displays mean values.
-This is because summary will, by default, summarise the data using the mean value.
-Here the mean represents the mean of the median distance for each series which is a bit useless (i.e. the mean of one value is that same value).
-We can display the median as well using:
-
-<<dispRity2>>=
-summary(disparity, cent.tend=median)
-## Or even the product! It won't affect the results
-summary(disparity, cent.tend=prod)
-@
-
-This is because we did calculate the noise within our data.
-We can classically do so by bootstrapping the data!
-
-\subsection{Bootstrapping the data}
-
-The \dispRity pacakge also provides easy way to bootstrap the data via the \texttt{boot.matrix} function.
-We can even rarefy the data to see the effect of the number of plots per series:
-For more details on the bootstrapping options, see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}.
-
-<<boot.matrix>>=
-bootstrapped_data <- boot.matrix(customised_series, bootstraps=100)
-rarefied_data <- boot.matrix(customised_series, bootstraps=100, rarefaction=TRUE)
-## Note that the output is a dispRity object giving some details on the series and the bootstraps
-bootstrapped_data
-@
-
-We can now rerun a more robust disparity analysis:
-
-<<dispRity3>>=
-disparity_BS <- dispRity(bootstrapped_data, metric = c(median, centroids))
-disparity_rare <- dispRity(rarefied_data, metric = c(median, centroids))
-## Note that calculation time is increased!
-@
-
-\subsection{Summarising the data}
-
-We can now summarise the data using various options such as the confidence intervals levels and the central tendency.
-
-<<summary>>=
-## The default:
-summary(disparity_BS)
-## The quantiles are calculated as 50 and 95 and the central tendency is the mean by default.
-## But we can specify different options
-summary(disparity_BS, quantile=90, cent.tend=median)
-
-## Finally we can see the results of the rarefaction analysis:
-head(summary(disparity_rare))
-## This outputs a longer table with all the variations of plots down to 3 plots per series.
-@
-
-Finally we can also plot the results using the simple \texttt{plot.dispRity S3} method:
-
-<<disparity plots, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
-## Graphical option
-par(bty = "n")
-## Plotting the score for each groups
-plot(disparity_BS)
-@
-
-Or have a look at the effect of the number of experimental plots:
-
-<<rare plots, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
-## Graphical options
-quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
-## The same but looking at the number of plots
-plot(disparity_rare, elements = TRUE)
-## With the same number of plots per group
-plot(disparity_rare, elements = TRUE, rarefaction = 17)
-@
-
-Or event have a look at the rarefaction curves:
-
-<<rare plots2, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
-## Graphical option
-par(bty = "n")
-## Plotting the rarefaction curves
-plot(disparity_rare, rarefaction = "plot")
-@
+\bibliographystyle{sysbio}
+\bibliography{References}
 

### 2015-11-24 doc/dispRity-palaeo-demo.Rnw
Paleo-demo updated

@@
 \usepackage{geometry}
+\usepackage{natbib}
 \geometry{verbose,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
@@
 
+In this demo we are going to test whether the disparity changed through time in this dataset since the last 100 million years.
+
 \section{Installing \dispRity}
@@
 library(dispRity)
+
+## Setting the random seed for repeatability
+set.seed(123)
 
@@
 
-<<installation, eval=FALSE>>=
+<<strap, eval=FALSE>>=
 install.packages(strap)
@@
 \section{Splitting the morphospace through time} 
+One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller sub-samples that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
+Basically, the full morphospace in the previous step represents the total amount of morphologies in all time and is likely to be greater than any sub-sample of the morphospace.
+
+The \dispRity package provides a \texttt{time.series} function that allows to split the morphospace time slices (using \texttt{method="continuous"}) or into time bins (using \texttt{method="discrete"}).
+In this example, we are going to split the morphospace into 5 equal time bins of 20 million years long from 100 million years ago (Mya) to the present.
+Note that we can also provide here the table containing the first and last occurrences data for some fossils to take into account some fossils that might span between our different bins.
+
+<<time-binning>>=
+## Creating the vector of time bins ages
+time_bins <- rev(seq(from=0, to=100, by=20))
+
+## Splitting the morphospace using the time.series function
+binned_morphospace <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
+    method = "discrete", time = time_bins, inc.nodes = FALSE,
+    FADLAD = BeckLee_ages)
+@
+
+The warning message tells us that some taxa were considered as single points in time because we didn't provide any temporal range information for some taxa.
+The output object is a \dispRity object.
+For details about this object class, please refer to the \href{https://github.com/TGuillerme/dispRity}{other vignettes}.
+In brief however, \dispRity objects are lists of different elements (i.e.
+disparity results, morphospace sub-samples, etc...) that display only a summary of the object when calling the object for visualisation reasons.
+
+<<time-binning2>>=
+## The object class
+class(binned_morphospace)
+
+## What's in the object (messy version for the console!)
+str(binned_morphospace)
+names(binned_morphospace)
+
+## How it's printed by default
+binned_morphospace
+@
+
+Note that these objects will gradually contain more information when reaching higher steps in the disparity-through-time analysis.
 
 \section{Bootstrapping the data}
+Once we obtain our different subsamples, we might want to bootstrap or/and rarefy it (i.e. pseudo-replicating the data).
+The bootstrap will allow us to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of taxa to get rid of eventual sampling problems.
+For that, we can use the \texttt{boot.matrix} function that allows to bootstrap the \dispRity object along with the \texttt{rarefaction} option that will rarefy it.
+
+<<boot.rare>>=
+## Bootstrapping each sub-sample 100 times
+boot_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100)
+
+## Getting the minimum number of rows (=taxa) in the sub-samples
+min(unlist(lapply(binned_morphospace$data, nrow)))
+
+## Bootstrapping each sub-sample 100 times and rarefying them 
+rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
+    rarefaction = 6)
+
+## Note that the output objects are both from the "dispRity" class ...
+class(boot_bin_morphospace); class(rare_bin_morphospace)
+
+## ... and display some more information on the previous step when called
+boot_bin_morphospace
+rare_bin_morphospace
+@
+
 \section{Calculating disparity}
+
+We can now calculate the disparity within each sub-sample along side with some confidence intervals generated by the pseudo-replication step above (bootstrap/rarefaction).
+Disparity can be calculated in many ways and this package allows user to come up with their own disparity metrics.
+For more info, please refer to the \href{https://github.com/TGuillerme/dispRity}{other vignettes}.
+
+In this example, we are going to calculate the spread of the data in each sub-sample by calculating the disparity as the sum of the variance of each dimension of the morphospace in each sub-sample using the \dispRity function.
+
+<<disparity>>=
+## Disparity for the bootstrapped data
+boot_disparity <- dispRity(boot_bin_morphospace, metric = c(sum, variances))
+
+## Disparity for the rarefied data
+rare_disparity <- dispRity(rare_bin_morphospace, metric = c(sum, variances))
+
+## Again, these objects are both from the "dispRity" class and just display
+## some information when called, note the results!
+boot_disparity
+rare_disparity
+
+## To display the actual results we can use the summary function
+summary(boot_disparity)
+summary(rare_disparity)
+
+## See more options on the summary function:
+?summary.dispRity
+@
+
 \section{Plotting the results}
+It is sometimes easier to visualise the results in a plot than in a table.
+For that we can use the \texttt{plot} function to plot the \dispRity objects.
+
+<<plot, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
+## Graphical options
+quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
+
+## Plotting the bootstrapped and rarefied results
+plot(boot_disparity, main = "bootstrapped results")
+plot(rare_disparity, main = "rarefied results")
+
+## See more options on the plot function:
+?plot.dispRity
+@
+
+As we can see in this case, the rarefaction tends to increase the confidence intervals around the mean.
+
 \section{Testing differences}
+Finally, to draw some conclusions from these results, we need to apply some statistical testing.
+As stated before, we wanted to test whether the disparity changed through time in this dataset since the last 100 million years.
+To do so, we can simply apply some comparisons of the means between each time-bins in a sequential manner to see whether the disparity in bin X is equal to the disparity in bin Y which is in turn equal to the disparity in bin Z, etc.
+Because our data is temporally auto-correlated (i.e.
+what happens in bin Y is dependent on what happened in bin X) and was pseudo-replicated (i.e.
+each bootstrap draw creates non-independent sub-samples because they are all based on the same sub-sample), we are going to apply a non-parametric mean comparison: the \texttt{wilcox.test}.
+Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis.
+
+<<testing>>=
+## testing the differences between bins in the bootstrapped dataset.
+test.dispRity(boot_disparity, test = wilcox.test, comparison = "sequential",
+    correction = "bonferroni")
+
+## testing the differences between bins in the rarefied dataset.
+test.dispRity(rare_disparity, test = wilcox.test, comparison = "sequential",
+    correction = "bonferroni")
+@
+
+Here our results show significant changes in disparity through time between all time bins during the last 100 million years.
+However, when looking at the rarefied results, there is no significant difference between the time bins in the Cenozoic (60-40, 40-20 and 20-0 Mya), suggesting that the differences detected in the first test might just be due to the differences in number of taxa sample (respectively \Sexpr{unlist(lapply(binned_morphospace[[1]], nrow))[[3]]}, \Sexpr{unlist(lapply(binned_morphospace[[1]], nrow))[[4]]} and \Sexpr{unlist(lapply(binned_morphospace[[1]], nrow))[[5]]} in each time bin).
 

### 2015-11-24 doc/dispRity-metrics.Rnw
dispRity metric vignette

@@
 
-\title{dispRity metrics}
+\title{\texttt{dispRity} metrics}
 
@@
 In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. MDS, PCA, PCO, PCoA).
-For example, one can look at ellipsoid hyper-volume of the ordinated space \cite{DonohueDim}, the sum and the product of the ranges and variances of the ordinated space dimensions \cite{Wills1994} or the median position of the elements in the ordinated space relative to its centroid \cite{GuillermeSTD}.
-Of course, there is many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks or some being more generalist.
+For example, one can look at ellipsoid hyper-volume of the ordinated space \citep{DonohueDim}, the sum and the product of the ranges and variances of the ordinated space dimensions \citep{Wills1994} or the median position of the elements in the ordinated space relative to its centroid \citep{GuillermeSTD}.
+Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks or some being more generalist.
 
@@
 In fact, the core function of the package, \dispRity, allows the user to define any metric in the \texttt{metric} argument.
-However, because of the infrastructure of the package (especially the on of the \dispRity objects -- see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}), the \texttt{metric} argument has to follow certain slightly restrictive rules described in detail below.
+However, because of the infrastructure of the package (especially the on of the \dispRity objects -- see the \href{https://github.com/TGuillerme/dispRity}{other vignettes}), the \texttt{metric} argument has to follow certain slightly restrictive rules described in detail below.
 \begin{enumerate}
     \item \texttt{metric} must be between one or three \texttt{function} objects
-    \item \texttt{metric} must intake a \texttt{matrix} or \texttt{vector} object
-    \item \texttt{metric} individual functions must be classified in one of the three levels
-    \item \texttt{metric} one of the functions must be a level 1 function
+    \item \texttt{metric} must can have only one argument as input: a \texttt{matrix} or a \texttt{vector}
+    \item \texttt{metric} individual functions must be classified in one of the three levels (see below)
+    \item \texttt{metric} one of the functions must be a level 1 function (see below)
 \end{enumerate}
@@
 ## Any summary metric such as mean or median are good examples of level 1
-## functions as they reduce the matrix to a single dimensions (value):
+## functions as they reduce the matrix to a single dimension (value):
 mean(dummy_matrix)
@@
 \subsection{A level 2 function}
-A level 2 function for example, will decompose the \texttt{matrix} into a \texttt{vector}.
-Several level 2 functions are implemented in \dispRity (see \texttt{?dispRity.metric}) such as the \texttt{variances} function that calculates the variance of each column of the ordinated matrix (or axis/dimensions):
+A level 2 function will decompose the \texttt{matrix} into a \texttt{vector}.
+Several level 2 functions are implemented in \dispRity (see \texttt{?dispRity.metric}) such as the \texttt{variances} function that calculates the variance of each column of the ordinated matrix (i.e.
+it's axis or dimensions):
 
@@
 rows <- function(matrix) apply(matrix, 1, prod)
-## Rows is a level 2 function as well (outputs a vector of 4 values):
+## rows is a level 2 function as well (outputs a vector of 4 values):
 rows(dummy_matrix)
@@
 \begin{enumerate}
-  \item if your function can deal with a matrix as an \texttt{input}.
-  \item which level is your function (level 1, 2 or 3, see \texttt{?dispRity.metric}).
+  \item if your function can deal with a \texttt{matrix} or a \texttt{vector} as an \texttt{input}.
+  \item which level is your function (level 1, 2 or 3, see above).
   \item whether the function can properly be implemented in the \texttt{dispRity} function.
@@
 
-\bibliographystyle{vancouver}
+\bibliographystyle{sysbio}
 \bibliography{References}

### 2015-11-25 doc/dispRity-manual.Rnw
renamed tutorial to manual

\documentclass{article}
\usepackage[sc]{mathpazo}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{natbib}
\usepackage[utf8]{inputenc}
\geometry{verbose,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}
\usepackage{url}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=true,bookmarksnumbered=true,bookmarksopen=true,bookmarksopenlevel=2,
 breaklinks=false,pdfborder={0 0 1},backref=false,colorlinks=false]
 {hyperref}
\hypersetup{
 pdfstartview={XYZ null null 1}}

\newcommand{\dispRity}{\texttt{dispRity} }
\newcommand{\R}{\texttt{R} }

\begin{document}
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.path='figure/minimal-', fig.align='center', fig.show='hold')
options(formatR.arrow=TRUE,width=90)
@

\title{\texttt{dispRity} manual}

\author{Thomas Guillerme}

\maketitle

\dispRity is a package for calculating disparity in \texttt{R}.
It allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.

\tableofcontents

\section{Before starting}
\subsection{Glossary}
Because this package is aimed to be multidisciplinary, many names or terms used in this tutorial might be non-familiar to certain fields.
Here is a list of what are the exact meaning of these term:
\begin{itemize}
\item \textbf{Ordinated space}: it designates the mathematical multidimensional object studied here. In morphometrics, this one is often referred as being the morphospace, or the cladisto-space for cladisticts or the eco-space for ecology, etc. In practice though, this term designates an ordinated matrix where the columns represent the dimensions of the ordinated space (often $>$ 3!) and the rows represent the elements within this space.
\item \textbf{Elements}: it designates the rows of the ordinated space, elements can be either taxa, field sites, countries, etc...
\item \textbf{Dimensions}: it designates the columns of the ordinated space. The dimensions can also be referred to as axis.
\end{itemize}

\subsection{Installation}
You can install this package easily if you use the latest version of \R and \texttt{devtools}.

<<installation, eval=FALSE>>=
install.packages("devtools")
library(devtools)
install_github("TGuillerme/dispRity", ref = "release")
library(dispRity)
@

\noindent Note that we use the \texttt{release} branch here which is version \Sexpr{packageVersion("dispRity")}.
For the piping-hot (but potentially full of bugs) version, one can change the argument \texttt{ref = "release"} to \texttt{ref = "master"}.
This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.

\subsection{Data}
In this tutorial we are going to use a subset of the ordinated cladistic data from \cite{beckancient2014} that contains 50 taxa ordinated using their cladistic distance (i.e. the distance between their discrete morphological characters).
Note that this data is more oriented towards palaebiology analysis but that it can apply to other disciplines.
Please refer to the \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes covering some specific example of the package with ecological data.

<<data, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## Loading the package
library(dispRity)

## Loading the ordinated matrix containing 50 taxa
data(BeckLee_mat50)
dim(BeckLee_mat50)
head(BeckLee_mat50[,1:5])

## Loading another ordinated matrix containing 50 tips + 49 nodes
data(BeckLee_mat99)
dim(BeckLee_mat99)
head(BeckLee_mat99[,1:5], 2)
tail(BeckLee_mat99[,1:5], 2)

## Loading a list of first and last occurrence data
data(BeckLee_ages)
head(BeckLee_ages)

## Loading the phylogeny
data(BeckLee_tree)
plot(BeckLee_tree) ; nodelabels(cex=0.8) ; axisPhylo(root=140)
@

\subsection{A quick go through}
Here is a really crude and quick go through the package to show some of the features of this package.
Note that all these features will be discussed in more details below.

<<Quick go, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
## Splitting the data
sliced_data <- time.series(BeckLee_mat99, BeckLee_tree, method = "continuous",
    model = "acctran", time = rev(seq(from=0, to=130, by=15)), FADLAD = BeckLee_ages)
## Bootstrapping the data
bootstrapped_data <- boot.matrix(sliced_data, 100)
## Calculating disparity
sum_of_variances <- dispRity(bootstrapped_data, metric = c(sum, variances))
## Summarising the results
summary(sum_of_variances)
plot(sum_of_variances)
## Testing some the effect of time on disparity
summary(test.dispRity(sum_of_variances, test = aov, comparisons = "all"))
@

\section{Package specificity}

\subsection{The \dispRity objects}
Disparity analysis can involve a lot of shuffling around with many matrices (especially when bootstrapping the data) which can be a bit impractical to visualise and quickly jam your \R console.
For example, we can have a look at the structure of the object created in the quick example:

<<dispRity class1, eval=FALSE>>=
str(sum_of_variances)
## That's a more than 4500 lines of output!
@

\noindent Therefore this package proposes a new class of object called \dispRity objects.
These objects allow to easily use a S3 method functions such as \texttt{summary.dispRity} (just called as \texttt{summary}; see section \ref{summary}) or \texttt{plot.dispRity} (just called as \texttt{plot}; see section \ref{plot}).
But also, this allows to use the S3 method for printing \dispRity objects via \texttt{print.dispRity} that allows to summarise the content of the objects similar to the \texttt{phylo} class objects (see \texttt{print.phylo::ape}).

<<dispRity class2>>=
## Which class is the sum_of_ranges object?
class(sum_of_ranges)
## We can summarise it using the S3 method print.dispRity
sum_of_ranges
## This displays some information about what's in the object
@

\noindent Note however, that it is always possible to recall the full object using the argument \texttt{all=TRUE}:

<<dispRity class3>>=
## This displays some information about what's in the object
sum_of_ranges
## This displays some information about what's in the object
@

\subsection{Modular functions}

\section{Functions}

\subsection{\texttt{tree.age}}
\subsection{\texttt{cust.series}}
\subsection{\texttt{time.series}}
\subsection{\texttt{boot.matrix}}
\subsection{\texttt{dispRity}}
\subsection{\texttt{summary}}
\label{summary}
\subsection{\texttt{plot}}
\label{plot}
\subsection{\texttt{test.dispRity}}

\bibliographystyle{sysbio}
\bibliography{References}

\end{document}

### 2015-11-25 doc/dispRity-manual.Rnw
manual specificities part

@@
 \item \textbf{Dimensions}: it designates the columns of the ordinated space. The dimensions can also be referred to as axis.
+\item \textbf{Series}: it designates sub-samples of the ordinated space.
+Basically a series contain the same number of dimensions as the morphospace but might contain a smaller number of elements.
+For example, if our ordinated space is composed of birds and mammals (i.e.
+the elements) and 50 dimensions, we can create two series of just mammals or birds as elements (but the same 50 dimensions) to look at the difference in disparity between both groups.
 \end{itemize}
@@
 
-\section{Package specificity}
+\section{Package specificities}
 
@@
 <<dispRity class2>>=
-## Which class is the sum_of_ranges object?
-class(sum_of_ranges)
+## Which class is the sum_of_variances object?
+class(sum_of_variances)
+## What's in the object
+names(sum_of_variances)
 ## We can summarise it using the S3 method print.dispRity
-sum_of_ranges
+sum_of_variances
 ## This displays some information about what's in the object
@@
 
-<<dispRity class3>>=
-## This displays some information about what's in the object
-sum_of_ranges
-## This displays some information about what's in the object
+<<dispRity class3, eval=FALSE>>=
+## Displaying the full object
+print(sum_of_variances, all=TRUE)
+@
+
+\noindent Finally, some utility functions such as \texttt{get.dispRity} or \texttt{extract.dispRity} allows to access to some specific content of the object:
+
+<<dispRity class4>>=
+## Extracting some specific series from the disparity object
+series_1_and_4 <- get.dispRity(sum_of_variances, what=c(1,4))
+series_1_and_4
+## The observed disparity
+extract.dispRity(sum_of_variances)
+## The list of bootstrapped scores of disparity
+str(extract.dispRity(sum_of_variances, observed = FALSE))
 @
@@
 \subsection{Modular functions}
+This package aims to be a modular package where users can personalise some aspects of the package fairly easily.
+In this version \Sexpr{packageVersion("dispRity")} only one function is fully modular (\dispRity; see section \ref{disparity}) but more will be hopefully available in the near future (see section \ref{developments}).
+
+The modular idea is that users can use implemented tools to help facilitating their own personalised function.
+For example, the \dispRity function intake a \texttt{metric} argument designating how disparity should be calculated.
+This argument can take some already implemented functions such as \texttt{mean} but also some completely new ones such as the sum divided by length:
+
+<<dispRity modular>>=
+## Disparity measured as the mean
+summary(dispRity(sliced_data, metric = mean))
+## A function for measuring the sum divided by the length (the mean!)
+sum.by.length <- function(X)  sum(X)/length(X)
+## Disparity measured as the mean divided by the length
+summary(dispRity(sliced_data, metric = sum.by.length))
+## Giving the exact same results!
+@
+
+\noindent For more information concerning this modularity, please refer to the \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for the vignette about the metric implementation in \dispRity.
 
@@
 \subsection{\texttt{dispRity}}
+\label{disparity}
 \subsection{\texttt{summary}}
@@
 
+\section{Developments}
+\label{developments}
+As stated at the start of the demo, this version \Sexpr{packageVersion("dispRity")} is still in development and many parts are missing.
+Here are the new functionalities that will be implemented before the proper release version (v.1).
+
+\subsection{More user defined stuff}
+I intend also develop functions to help users to develop their own algorithms for the bootstrap method (via \texttt{make.boot}) or the evolutionary models (via \texttt{make.model}).
+Both functions will provide similar testing as the \texttt{make.metric} function.
+
+\subsection{Faster!}
+Finally, for long analysis, I intend to develop a parallel running version of the package.
+In fact, most of the internal functions are base on \texttt{lapply} functions that can be easily passed to \texttt{snow::parLapply} or similar parallel functions.
+
 \bibliographystyle{sysbio}

### 2015-11-25 doc/dispRity-manual.Rnw
manual (rough) outline

@@
 
-\subsection{\texttt{tree.age}}
 \subsection{\texttt{cust.series}}
+
+\texttt{cust.series} is a function that allows to split the ordinated matrix according to a factor determined by the user.
+For example, here we can split the matrix based on phylogeny: let's classify in two different groups the crown and stem mammals present in the tree.
+
+<<cust.series2>>=
+## We want to separate the species around the node 71 (see phylogeny above).
+## All the descendant of these node are crown and all the ancestors are stem
+crown <- extract.clade(BeckLee_tree, node = 71)$tip.label
+stem <- drop.tip(BeckLee_tree, tip = crown)$tip.label
+## We then have to feed this information in a data frame with one column
+factors <- as.data.frame(
+    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
+    ncol = 1, dimnames = list(c(crown, stem), "Group")))
+## We then can use the customised series function
+crown_stem <- cust.series(BeckLee_mat50, factors)
+## This created a dispRity object containing two series: crown and stem
+class(crown_stem)
+crown_stem
+## This object contains three elements
+names(crown_stem)
+## With "data" being the list sub-matrices
+str(crown_stem$data)
+## "taxa" being the list of taxa in the original ordinated matrix
+str(crown_stem$taxa)
+## and finally, "series" containing information on the series type (custom) and names
+## (crown and stem)
+crown_stem$series
+@
+
+Of course, using phylogeny as a factor is just an example.
+Any type of factors can be used in this example (e.g. body mass categories, habitat, diet, etc...).
+See the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos} for other specific examples.
+
 \subsection{\texttt{time.series}}
+
+\texttt{time.series} is a function that allows to split the matrix in different ways using time as a category.
+This functions needs as input an ordinated matrix and a matching phylogenetic tree.
+Two types of time series splitting can be performed by using the \texttt{method} option:
+\begin{enumerate}
+\item the classical discrete time splitting (or time-binning) using \texttt{method = "discrete"};
+\item or a new continuous time splitting (or time-slicing) using \texttt{method = "continuous"}.
+\end{enumerate}
+For the time-slicing method details, see Guillerme and Cooper \textit{in prep.} \cite{GuillermeSTD}.
+Once the method is chosen, the function intakes the \texttt{time} argument which can be a vector of \texttt{numeric} values for:
+\begin{enumerate}
+\item defining the boundaries of the time bins (when \texttt{method = "discrete"});
+\item defining the time-slicing places (when using \texttt{method = "continuous"}).
+\end{enumerate}
+Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a fix number of equidistant time-bins/slices.
+Additionally, we can also input a data frame containing the First/Last Occurrence Data (\texttt{FADLAD}) for taxa that span over a longer time than the tips/nodes age.
+
+Let's see and example for time-binning (\texttt{method = "discrete"}):
+
+<<discrete.series>>=
+## Let's generate three time bins containing the data present every 40 Ma
+time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
+    method = "discrete", time = c(120, 80, 40, 0))
+## Note that the function provides a warning saying that tips where single
+## points in time (no FAD/LAD information). We can fix that by adding the
+## age data for the taxa that have some longer occurrence spans.
+time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
+    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
+## To entirely avoid the warning we could collect the occurrence span data
+## for all the taxa but that's not necessary. The function automatically
+## assumes no occurrence span time (i.e. single points in time) for all taxa
+## by default.
+time_bins
+@
+
+This generated indeed a list of 3 sub matrices.
+Note that we can also generate equivalent results by just telling the function that we want three time-bins (series) as follow:
+
+<<discrete.series2>>=
+time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete", time = 3)
+@
+
+We now have three time bins of 44.50368 Ma each. But hey, that's a bit silly, let's stick to bins of 40 Ma.
+When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1\% of the the tree length as long as at least three elements are present at that point in time (an extra 1\% is added until reaching the required minimum of three elements).
+We can also ask to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between tips \textit{AND} nodes.
+
+For the time-slicing method (\texttt{method = "continuous"}), the idea is pretty similar.
+We need to provide a matrix that contains the ordinated distance between taxa AND nodes and an evolutionary model via the \texttt{model} argument.
+For now they are four evolutionary models implemented \cite{GuillermeSTD} (but see section \ref{whatsleft}):
+\begin{enumerate}
+\item \texttt{"acctran"} where the data chosen on each time slice is always the one of the offspring
+\item \texttt{"deltran"} where the data chosen on each time slice is always the one of the descendant
+\item \texttt{"punctuated"} where the data chosen on each time slice is randomly chosen between the offspring or the descendant
+\item \texttt{"gradual"} where the data chosen on each time slice is either the offspring or the descendant depending on branch length
+\end{enumerate}
+
+<<continuous.series>>=
+## Let's generate four time slices every 40 Ma assuming a gradual evolution model
+time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
+    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
+    FADLAD = BeckLee_ages)
+time_slices
+## Note that in the same way as for the discrete method, we can also
+## automatically generate the slices
+time.series(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
+    model = "gradual", time = 4)
+## But that gives the same silly results (slicing every 44.50368 Ma)
+@
+
 \subsection{\texttt{boot.matrix}}
+
+After separating the any-o-space into submatrices, one can directly calculate disparity within each subsample or first test the robustness of the data.
+To measure how robust is the data, we can perform bootstrap pseudo-replications or rarefaction analysis to make sure the results are not dependant on weird outliers.
+The \texttt{boot.matrix} allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
+The default options will bootstrap the matrix 1000 times with no rarefaction.
+
+<<boot.matrix1>>=
+boot.matrix(data = BeckLee_mat50)
+## As we can wee, the output is also a dispRity object that is summarised
+## automatically, and gives information on the data as well as the number of
+## bootstraps and the bootstraps methods.
+@
+
+This function proposes the user to specify the bootstrap algorithm with the \texttt{boot.type} method.
+Currently two algorithms are implemented (but see section \ref{whatsleft}):
+\begin{enumerate}
+\item \texttt{"full"} where the bootstrapping is entirely stochastic (all the data is bootstrapped)
+\item \texttt{"single"} where only one random taxa is replaced by one other random taxa each pseudo-replication
+\end{enumerate}
+This function also has a \texttt{rarefaction} option to rarefy the data (how unexpected!).
+The default status is \texttt{FALSE} but it can be set to \texttt{TRUE} to fully rarefy the data (i.e.
+remove $n$ taxa for the number of pseudo-replicates, where $n$ varies from the maximum number of taxa present in the dataset to a minimum of 3 elements).
+It can also be set to a fix \texttt{numeric} value (or a set of numeric values).
+Finally, the last option from this function, \texttt{rm.last.axis}, is whether to remove a certain amount of axis of the ordinated data.
+This option can be logical where \texttt{FALSE} (default) will not remove any axis and \texttt{TRUE} will remove the 5\% last axis or it can be set to a user defined proportion.
+This is a common practice in disparity-through-time analysis but it is hard to justify statistically.
+Here is a series of examples:
+
+<<boot.matrix2>>=
+## Bootstrapping with the single bootstrap method
+boot.matrix(BeckLee_mat50, boot.type = "single")
+## Bootstrapping with the full rarefaction
+boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
+## Or with a set number of rarefaction levels
+boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8,3))
+## And removing the last axis (default)
+boot.matrix(BeckLee_mat50, rm.last.axis = TRUE)
+## Or with a fix value (50%)
+boot.matrix(BeckLee_mat50, rm.last.axis = 0.5)
+@
+
+Of course, one could be interested in directly supplying the subsampled matrices generated above directly to this function.
+In fact, it can also deal with a list of matrices or with a \dispRity object output from the \texttt{cust.series} or \texttt{time.series} functions.
+Like this:
+
+<<boot.matrix3>>=
+## Let's do a full rarefaction on the crown/stem data for later
+crown_stemBS <- boot.matrix(crown_stem, rarefaction = TRUE)
+## And just bootstrapping for the time binning/slicing
+time_binsBS <- boot.matrix(time_bins)
+time_slicesBS <- boot.matrix(time_slices)
+## And just to remind us what's in each objects
+crown_stemBS
+time_binsBS
+time_slicesBS
+@
+
 \subsection{\texttt{dispRity}}
 \label{disparity}
+
+Now we're going to see the functionalities of the core function of this package: the \dispRity function.
+This function is a modulable function that allow to simply (and quickly!) calculate disparity from a matrix.
+
+Because disparity can be measured in many ways, this function is a tool to measure disparity \textit{as defined by the user} (and here's where the modulable part comes in).
+In fact, the \dispRity function intakes two main arguments: the data and the disparity metric.
+%To change in future version!
+The disparity metric boils down to a single value that describes the matrix.
+Some are classic like the sum or the product of the ranges or the variances of the matrix columns (i.e. the any-o-space axis/dimensions) \cite{Wills1994}.
+However, many more exist like the median distances between the taxa and the any-o-space centroid \cite{GuillermeSTD}.
+
+One can usually decompose the disparity metrics into two elements:
+\begin{enumerate}
+\item the \textbf{class metric} that is a descriptor of the matrix.
+For example describing the ranges of each column in the matrix or the euclidean distances between each row and the centroid of the matrix.
+\item the \textbf{summary metric} that is a summary of the class metric values. For example, the sum of the ranges or the median of the euclidean distances. 
+\end{enumerate}
+Basically the combination can be infinite between the class and summary metrics.
+For example, people might want to measure the median variances of the axis or the product of the distances from the centroid.
+However, it is probable that some metrics are better to reflect some biological aspects of the any-o-space than others (see section \ref{whatsleft})...
+
+In practice, the \dispRity function intakes a pair of class and summary metrics as a definition of disparity.
+Several of these metrics are implemented in other packages (like \texttt{stats::median}, \texttt{base::sum}, etc.) and this package proposes several metrics listed in \texttt{dispRity.metric} (see \texttt{?dispRity.metric}).
+It is even possible to use your very own class and summary metrics!
+This will be actually heavily encouraged and facilitate with the \texttt{make.metric} function in a future version (see \ref{whatsleft}).
+
+To use these metrics pairs in the \dispRity function, it's pretty easy:
+
+<<dispRity1>>=
+dispRity(BeckLee_mat50, metric = c(sum, ranges))
+## Yep, that's it. The order of the metrics does not matter (the function detects
+## automatically which one is the class and the summary metric) so the combinations
+## are numerous!
+dispRity(BeckLee_mat50, metric = c(sum, variances))
+dispRity(BeckLee_mat50, metric = c(prod, centroids))
+dispRity(BeckLee_mat50, metric = c(mode.val, ranges))
+## Or funny user defined ones!
+total.range <- function(X) abs(range(X)[1]-range(X)[2])
+dispRity(BeckLee_mat50, metric = c(total.range, ranges))
+## That would be disparity as the range of the axis ranges!
+@
+
+Of course, this function can take a simple ordinated matrix but, more interesting it can also deal with a list of matrices (from the \texttt{cust.series} or \texttt{time.series} functions for example) or a bootstrapped or/and rarefied matrix (from the \texttt{boot.matrix} function for example).
+
+<<dispRity2>>=
+## For example, let's calculate the sum of ranges from our crown/stem mammals
+dispRity(crown_stem, metric = c(sum, ranges))
+## Or even better, from the bootstrapped data
+disp_crown_stemBS <- dispRity(crown_stemBS, metric = c(sum, ranges))
+## Note that the bootstrapped + rarefied data takes some time to run (it's
+## calculating disparity from 46 matrices!)
+
+## Let's calculate a different metric for the disparity-through-time analysis
+disp_time_binsBS <- dispRity(time_binsBS, metric = c(median, centroids))
+disp_time_slicesBS <- dispRity(time_slicesBS, metric = c(median, centroids))
+@
+
+Again, note that the results are \dispRity objects and can displayed through the S3 \texttt{print.dispRity} method to remind users what are in these objects.
+Also note that an even faster parallel version will be developed in the future (see section \ref{whatsleft}).
+
 \subsection{\texttt{summary}}
 \label{summary}
+
+As shown in these examples above however, even though the disparity has been calculated in many way, the function doesn't directly outputs the results, rather just a summary of the analysis.
+To really display the results we can use the S3 \texttt{summary.dispRity} and \texttt{plot.dispRity} functions.
+
+\subsubsection{\texttt{summary} function}
+This function intakes a \dispRity object plus various options namely the \texttt{quantile} values for the confidence intervals levels; the \texttt{cent.tend} for the central tendency to use for summarising the results and two \textit{visual} options which are whether to recall the \dispRity options and how much digits are wanted in the results.
+
+<<summary>>=
+## Let's have a look at the disparity-through-time using in the time-binned data
+summary(disp_time_binsBS)
+## The 50 and 95 quantiles are calculated and the central tendency is the mean by default.
+## Note that the function also displays the observed disparity (non-bootstrapped).
+## We can change that easily by specifying different values in the options
+summary(disp_time_binsBS, quantile = 88, cent.tend = sd, rounding = 1)
+## Also, if we happen to have forgotten what was in the disp_time_binsBS object (as in
+## which options where used to modify the original matrix) we can use the recall option:
+summary(disp_time_binsBS, recall = TRUE)
+## Note that the recall just prints text. The output of summary.dispRity is always a
+## data.frame
+summary_table <- summary(disp_time_binsBS, recall = TRUE)
+class(summary_table)
+
+## Finally we can see the results of the rarefaction analysis for the crown/stem data
+head(summary(disp_crown_stemBS))
+## This outputs a longer table with all the variations of the number of crown/stem taxa
+## from the maximum (30 and 20) to the minimum (3). Note that the observed disparity
+## is NA for rarefied data.
+@
+
 \subsection{\texttt{plot}}
 \label{plot}
+
+This final function allows to visualise the disparity results in a often nicer fashion than just a table (even though the exact same data is displayed).
+The \texttt{plot.dispRity} option intakes the same options as \texttt{summary.dispRity} along side with various graphical options described in the function manual.
+Here, let's just have a look a few of these options.
+
+Let's start with plotting the difference in disparity between the crown and the stem mammals.
+For this we are going to simply specify the type of plots using the \texttt{type} option that allows to choose between the \texttt{"continuous"} or \texttt{"discrete"} method.
+In our case, we are interested in looking at the results in a discrete way.
+Note that the plot \texttt{type} can be left missing.
+The function will then use \texttt{"discrete"} as default or \texttt{"continuous"} if the input data came from the \texttt{time.series} function with \texttt{method = "continuous"}.
+
+The grey squares represent the confidence intervals (50 in dark grey and 90 in light grey) and the dot represents the mean.
+However, it might be interesting to see how many taxa are present in each group.
+This can be done via the \texttt{elements} option.
+A third interesting option is to plot the rarefied data if a rarefaction was applied on the data with the \texttt{boot.matrix} function.
+For displaying the rarefied data we can use the \texttt{rarefaction} option.
+This option can be either \texttt{FALSE} (default) for always plotting the maximum number of taxa per series or \texttt{TRUE} to always plot the minimum.
+Additionally, it can be a specific number to plot a specific number of taxa.
+Let say we want both groups to have 20 taxa maximum.
+We can see now that the confidence intervals overlap more than previously
+
+<<plot1, fig.width=15, fig.height=5, out.width='1\\linewidth'>>=
+## Graphical options
+quartz(width = 15, height = 5) ; par(mfrow = (c(1,3)), bty = "n")
+## Plotting the score for both groups
+plot(disp_crown_stemBS, ylim = c(25,40))
+## Same one with the number of taxa for each data set
+plot(disp_crown_stemBS, elements = TRUE, ylim = c(25,40))
+## Same one with rarefied data (note how that affects the elements data as well!)
+plot(disp_crown_stemBS, elements = TRUE, rarefaction = 20, ylim = c(25,40))
+@
+
+To understand how the number of taxa affects each series, we can use the \texttt{rarefaction = "plot"} function to plot the rarefaction curves.
+
+<<plot2, fig.width=5, fig.height=8, out.width='0.5\\linewidth'>>=
+## Graphical options
+par(bty = "n")
+## The rarefaction curves
+plot(disp_crown_stemBS, elements = TRUE, rarefaction = "plot")
+## Note that the function automatically draws the curves for each series and splits
+## the plot screen accordingly.
+@
+
+The different dashed lines represent the different confidence intervals around each rarefaction curve.
+
+Additionally, for the \texttt{type = "discrete"} option, we might also decide to plot many boxplots (which can be a bit messy).
+The \texttt{discrete\_type} option allows to switch between \texttt{"box"} or \texttt{"line"} for showing the results in one or the other format.
+We can also plot the observed data when existing (i.e. not with some rarefaction levels) by using \texttt{observed = TRUE}.
+Let's look at that with the time binned data (even though there is only three time bins).
+Finally, we can use the \texttt{type = "continuous"} option that comes in handy for plotting continuous data like in the time sliced analysis (obviously...).
+
+<<plot3, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
+## Graphical options
+quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
+## The disparity-through-time data in a time-binned way (with lines rather than
+## boxes) with the observed data as crosses
+plot(disp_time_binsBS, discrete_type = "line", observed = TRUE)
+## The disparity-through-time data in a time-sliced ways
+plot(disp_time_slicesBS)
+
+@
+
+We can even plot both results (discrete and continuous) on the same graph:
+
+<<plot4, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
+## Graphical options
+par(bty = "n")
+## First, plotting the continuous data
+plot(disp_time_slicesBS, ylim=c(0.5,2))
+## Second, adding the discrete data
+par(new = TRUE)
+plot(disp_time_binsBS, discrete_type = "line", ylim=c(0.5,2), xlab="", ylab="")
+@
+
+Note that the results vary in values but not in pattern.
+The change in values might be due to the fact that nodes data are not included in the discrete analysis.
+
+I encourage you to play with the graphical options to have some prettier results.
+Note that most of the options from \texttt{plot} can be passed to \texttt{plot.dispRity} via \texttt{...}.
+
+<<plot5, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
+## Graphical options
+par(bty = "n")
+## A plot with many options!
+plot(disp_time_slicesBS, quantile = c(seq(from=10, to=100, by=10)), cent.tend = sd,
+    type = "continuous", elements = "log", col = c("black", rainbow(10)),
+    ylab = c("Disparity", "log(Diversity)"), xlab = "Time (in in units from
+    past to present)", time.series = FALSE, observed = TRUE, main = "Many options...")
+@
+
 \subsection{\texttt{test.dispRity}}
+
+\subsection{utilities}
+\subsubsection{\texttt{tree.age}}
+This function allows to calculate the age of each individual nodes and tips in a tree.
+It can either use the root age of the tree (if present as \texttt{\$root.time}) or else calculate the age using a user defined root age via the \texttt{age} argument.
+Also, it is possible to decide whether the time is calculated towards the past (e.g. million years ago) or towards the present (e.g. in time since the origin).
+
+<<tree.age>>=
+## This tree has a root age
+BeckLee_tree$root.time
+## So we can get the age of each tips and nodes directly
+head(tree.age(BeckLee_tree), 5)
+## But we can also decide to make the age relative (between 1 and 0)
+head(tree.age(BeckLee_tree, age = 1), 5)
+## Or even relative, but from the root (i.e. how far are the nodes/tips
+## from the root)
+head(tree.age(BeckLee_tree, age = 1, order = "present"), 5)
+@
 

### 2015-11-25 doc/dispRity-manual.Rnw
time series + bootstrap

@@
 \section{Functions}
+One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter \textit{series}).
+These series correspond at regions of the ordinated space that we want to compare to each other.
 
 \subsection{\texttt{cust.series}}
-
-\texttt{cust.series} is a function that allows to split the ordinated matrix according to a factor determined by the user.
-For example, here we can split the matrix based on phylogeny: let's classify in two different groups the crown and stem mammals present in the tree.
+The \texttt{cust.series} function allows to create these series according to factors determined by the user.
+In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (all the living mammals and their direct ancestor) and the stem mammals (all the mammals that have no direct offspring).
 
@@
     ncol = 1, dimnames = list(c(crown, stem), "Group")))
-## We then can use the customised series function
+
+## We then can use the customised series function to create the two series
+## i.e. the two regions of the ordinated space
 crown_stem <- cust.series(BeckLee_mat50, factors)
+
 ## This created a dispRity object containing two series: crown and stem
@@
 crown_stem
+
 ## This object contains three elements
@@
 
-Of course, using phylogeny as a factor is just an example.
-Any type of factors can be used in this example (e.g. body mass categories, habitat, diet, etc...).
-See the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos} for other specific examples.
-
 \subsection{\texttt{time.series}}
 
-\texttt{time.series} is a function that allows to split the matrix in different ways using time as a category.
-This functions needs as input an ordinated matrix and a matching phylogenetic tree.
-Two types of time series splitting can be performed by using the \texttt{method} option:
+Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
+The \texttt{time.series} function allows to create series that contain all the elements present at specific points in time.
+This functions needs as input an ordinated space and a matching phylogenetic tree.
+Two types of time series can be performed by using the \texttt{method} option:
 \begin{enumerate}
-\item the classical discrete time splitting (or time-binning) using \texttt{method = "discrete"};
-\item or a new continuous time splitting (or time-slicing) using \texttt{method = "continuous"}.
+\item discrete time series (or time-binning) using \texttt{method = "discrete"};
+\item continuous time series (or time-slicing) using \texttt{method = "continuous"}.
 \end{enumerate}
-For the time-slicing method details, see Guillerme and Cooper \textit{in prep.} \cite{GuillermeSTD}.
-Once the method is chosen, the function intakes the \texttt{time} argument which can be a vector of \texttt{numeric} values for:
+For the time-slicing method details see \cite{GuillermeSTD}.
+For both methods, the function intakes the \texttt{time} argument which can be a vector of \texttt{numeric} values for:
 \begin{enumerate}
@@
 \end{enumerate}
-Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a fix number of equidistant time-bins/slices.
-Additionally, we can also input a data frame containing the First/Last Occurrence Data (\texttt{FADLAD}) for taxa that span over a longer time than the tips/nodes age.
+Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a given number of equidistant time-bins/slices.
+Additionally, it is also possible to input a data frame containing the First/Last Occurrence Data (\texttt{FADLAD}) for taxa that span over a longer time than the tips/nodes age.
 
-Let's see and example for time-binning (\texttt{method = "discrete"}):
+Here is an example for \texttt{method = "discrete"}:
 
 <<discrete.series>>=
-## Let's generate three time bins containing the data present every 40 Ma
+## Generating three time bins containing the data present every 40 Ma
 time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
@@
 
-This generated indeed a list of 3 sub matrices.
+\noindent This generated indeed a list of 3 sub matrices.
 Note that we can also generate equivalent results by just telling the function that we want three time-bins (series) as follow:
@@
 <<discrete.series2>>=
+## Automatically generate three equal length bins:
 time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete", time = 3)
 @
+\noindent We now have three time bins of 44.50368 Ma each.
 
-We now have three time bins of 44.50368 Ma each. But hey, that's a bit silly, let's stick to bins of 40 Ma.
-When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1\% of the the tree length as long as at least three elements are present at that point in time (an extra 1\% is added until reaching the required minimum of three elements).
+When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1\% of the tree length as long as at least three elements are present at that point in time (an extra 1\% is added until reaching the required minimum of three elements).
 We can also ask to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between tips \textit{AND} nodes.
 
-For the time-slicing method (\texttt{method = "continuous"}), the idea is pretty similar.
-We need to provide a matrix that contains the ordinated distance between taxa AND nodes and an evolutionary model via the \texttt{model} argument.
-For now they are four evolutionary models implemented \cite{GuillermeSTD} (but see section \ref{whatsleft}):
+For the time-slicing method (\texttt{method = "continuous"}), the idea is really similar.
+This option intakes a matrix that contains the ordinated distance between taxa \textit{AND} nodes and an assumed evolutionary model via the \texttt{model} argument:
 \begin{enumerate}
@@
 <<continuous.series>>=
-## Let's generate four time slices every 40 Ma assuming a gradual evolution model
+## Generating four time slices every 40 Ma assuming a gradual evolution model
 time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
@@
     model = "gradual", time = 4)
-## But that gives the same silly results (slicing every 44.50368 Ma)
 @
@@
 
-After separating the any-o-space into submatrices, one can directly calculate disparity within each subsample or first test the robustness of the data.
-To measure how robust is the data, we can perform bootstrap pseudo-replications or rarefaction analysis to make sure the results are not dependant on weird outliers.
+Once we obtain our different series, we might want to bootstrap or/and rarefy it (i.e. pseudo-replicating the data).
+The bootstrap will allow us to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of elements to get rid of eventual sampling problems.
 The \texttt{boot.matrix} allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
-The default options will bootstrap the matrix 1000 times with no rarefaction.
-
+The default options will bootstrap the matrix 1000 times without rarefaction.
+The number of bootstrap pseudo-replicates can be defined using the \texttt{bootstraps} option (see below)
+.
 <<boot.matrix1>>=
 boot.matrix(data = BeckLee_mat50)
-## As we can wee, the output is also a dispRity object that is summarised
+## As we can see, the output is also a dispRity object that is summarised
 ## automatically, and gives information on the data as well as the number of
@@
 
-This function proposes the user to specify the bootstrap algorithm with the \texttt{boot.type} method.
-Currently two algorithms are implemented (but see section \ref{whatsleft}):
+Additionally, this function allows to control the bootstrap algorithm through the \texttt{boot.type} argument.
+Currently two algorithms are implemented:
 \begin{enumerate}
 \item \texttt{"full"} where the bootstrapping is entirely stochastic (all the data is bootstrapped)
-\item \texttt{"single"} where only one random taxa is replaced by one other random taxa each pseudo-replication
+\item \texttt{"single"} where only one random elements is replaced by one other random elements each pseudo-replication
 \end{enumerate}
-This function also has a \texttt{rarefaction} option to rarefy the data (how unexpected!).
-The default status is \texttt{FALSE} but it can be set to \texttt{TRUE} to fully rarefy the data (i.e.
-remove $n$ taxa for the number of pseudo-replicates, where $n$ varies from the maximum number of taxa present in the dataset to a minimum of 3 elements).
+This function also also to rarefy the data using the \texttt{rarefaction} argument.
+The default argument is \texttt{FALSE} but it can be set to \texttt{TRUE} to fully rarefy the data (i.e.
+remove $n$ elements for the number of pseudo-replicates, where $n$ varies from the maximum number of elements present in the dataset to a minimum of 3 elements).
 It can also be set to a fix \texttt{numeric} value (or a set of numeric values).
-Finally, the last option from this function, \texttt{rm.last.axis}, is whether to remove a certain amount of axis of the ordinated data.
-This option can be logical where \texttt{FALSE} (default) will not remove any axis and \texttt{TRUE} will remove the 5\% last axis or it can be set to a user defined proportion.
-This is a common practice in disparity-through-time analysis but it is hard to justify statistically.
-Here is a series of examples:
+Finally, one last argument, \texttt{rm.last.axis} allows to remove a certain amount of dimensions (or axis) for the ordinated space.
+This can be logical argument where \texttt{FALSE} (default) will not remove any dimension and \texttt{TRUE} will remove the last dimensions that bear up to 5\% of the total ordinated space's variance.
 
@@
 In fact, it can also deal with a list of matrices or with a \dispRity object output from the \texttt{cust.series} or \texttt{time.series} functions.
-Like this:
+let's bootstrap our previously generate series:
 
 <<boot.matrix3>>=
-## Let's do a full rarefaction on the crown/stem data for later
+## Bootstrap and full rarefaction on the crown/stem series
 crown_stemBS <- boot.matrix(crown_stem, rarefaction = TRUE)
-## And just bootstrapping for the time binning/slicing
+## Bootstrap on the time binning/slicing series
 time_binsBS <- boot.matrix(time_bins)
 time_slicesBS <- boot.matrix(time_slices)
-## And just to remind us what's in each objects
+## Note that all these objects are of class dispRity
 crown_stemBS
@@
 <<dispRity2>>=
-## For example, let's calculate the sum of ranges from our crown/stem mammals
-dispRity(crown_stem, metric = c(sum, ranges))
+## For example, let's calculate the sum of variances from our crown/stem mammals
+dispRity(crown_stem, metric = c(sum, variances))
 ## Or even better, from the bootstrapped data
-disp_crown_stemBS <- dispRity(crown_stemBS, metric = c(sum, ranges))
+disp_crown_stemBS <- dispRity(crown_stemBS, metric = c(sum, variances))
 ## Note that the bootstrapped + rarefied data takes some time to run (it's

### 2015-11-26 doc/dispRity-palaeo-demo.Rnw
Updated toc

@@
 
-This demo showcases a typical disparity-through-time analyse with the following steps:
+This demo showcases a typical disparity-through-time analyse: we are going to test whether the disparity changed through time in this dataset since the last 100 million years.
 
-\begin{enumerate}
-\item Installing \dispRity
-\item The morphospace
-\item Splitting the morphospace through time
-\item Bootstrapping the data
-\item Calculating disparity
-\item Plotting the results
-\item Testing differences
-\end{enumerate}
+\tableofcontents
 
-In this demo we are going to test whether the disparity changed through time in this dataset since the last 100 million years.
-
-\section{Installing \dispRity}
+\section{Before starting}
+\subsection{Installing \dispRity}
 You can install this package easily if you are using the latest version of \R and \texttt{devtools}.
@@
 
-\section{The morphospace}
+\subsection{The morphospace}
 In this example, we are going to use a subset of the data from \cite{beckancient2014} that contains an ordinated cladistic matrix and a phylogenetic tree containing both the same 50 taxa along side a table that contains some first and last occurence data for several taxa.
@@
 
-\section{Splitting the morphospace through time} 
+\section{A disparity-through-time analysis}
+\subsection{Splitting the morphospace through time} 
 One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller sub-samples that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
@@
 
-\section{Bootstrapping the data}
+\subsection{Bootstrapping the data}
 Once we obtain our different subsamples, we might want to bootstrap or/and rarefy it (i.e. pseudo-replicating the data).
@@
 
-\section{Calculating disparity}
+\subsection{Calculating disparity}
 
@@
 
-\section{Plotting the results}
+\subsection{Plotting the results}
 It is sometimes easier to visualise the results in a plot than in a table.
@@
 
-\section{Testing differences}
+\subsection{Testing differences}
 Finally, to draw some conclusions from these results, we need to apply some statistical testing.

### 2015-11-26 doc/dispRity-ecology-demo.Rnw
Ecology demo v2 INIT

@@
 
-\title{dispRity demo for ecological data}
+\title{\texttt{dispRity} demo for ecologists}
 
-\author{Thomas Guillerme}
+%1-Introduce the data
+%"the data represents experimental plots in different treatments at two different water depths"
+
+%2-Explain rarefaction better
+%"the data represents experimental plots in different treatments at two different water depths"
+
+%3-Hypothesis
+%We want to know whether the treatments had an effect on the density of different species in the plots (Does treatment alter invertebrate community composition?).
+And does depth have an effect on community composition in the experimental plots?
+Annnnd: Do the treatments affect the communities differently at different depths?
+
+%(Personally I hate them as you know!)  Just use them as a way of visually showing that the 'multivariate' communities in the different treatments/ depths are different or similar and I guess which components explain most of the variance.
+They really aren't that powerful which is why I love your approach.
+You sold it so well to me, I think your best selling points (and indeed perhaps actually including a sentence at the start) are that you don't lose all the dimensions of the data like you do in PCA.
+And that you can keep the important ones in (again like you sold to me, that, some axes may be far more influential than a pca with throw up).
+
+\author{Thomas Guillerme \\ guillert@tcd.ie}
 
@@
 
-This is a quick demo for using the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) in ecological analysis.
-See the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos} for a general demo of the \dispRity package.
+This demo aims to give quick overview of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) for ecological analysis.
+Please refer to \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes, namely the \dispRity tutorial that explains the functions in more details.
+
+To keep it short, this package allows to use all the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
+For example, one might want to know whether the some sort of water treatment alters invertebrate communities and composition in natural habitats.
+
+\tableofcontents
 
 \section{Before starting}
-
-\dispRity is a package for calculating disparity in \texttt{R}.
-To keep it short, this package allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.
-These ordinated matrices can represent the any-o-space of possibilities of your analyses (e.g. for morphological data, the morphospace represents all the possible morphologies).
-
-\subsection{Installation}
-You can install this package easily if you use the latest version of \R and \texttt{devtools}.
+\subsection{Installing \dispRity}
+You can install this package easily if you are using the latest version of \R and \texttt{devtools}.
 
@@
 
-Note that we use the \texttt{release} branch here which is version 0.1.1.
-If you want the piping-hot (and full of bugs) version, change the option \texttt{ref = "release"} to \texttt{ref = "master"}.
-This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.
+This is a quick demo for using the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) in ecological analysis.
+See the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos} for a general demo of the \dispRity package.
 
 \subsection{Data}
-
 For this example with ecological data we are going to use data from McClean (unpubl.) that contains an ordination of a distance matrix between different experimental plots.
@@
 library(dispRity)
+
+## Setting the random seed for repeatability
+set.seed(123)
+
 ## let's use the matrix from McClean (unpubl.)
@@
 
-\section{A classic PCA}
+\section{A classical two dimensional approach}
 A classical way to represent this ordinated data will be to use PCA plots.
@@
 
+However, a common problem in this type of multivariate analysis is that often only a handful of dimensions are studied (usually the ones that bears the most variance of the ordinated matrix).
+
 This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix.
@@
 
-\section{Splitting the data}
+\section{A multidimensional approach with \dispRity}
+\subsection{Splitting the data}
 
@@
 
+\subsection{Testing the hypothesis}
+
+\bibliographystyle{sysbio}
+\bibliography{References}
+
 \end{document}

### 2015-11-26 doc/dispRity-ecology-demo.Rnw
Ecology demo update

@@
 To keep it short, this package allows to use all the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
-For example, one might want to know whether the some sort of water treatment alters invertebrate communities and composition in natural habitats.
+For example, one might want to know whether the sort of water treatment at certain depth alters invertebrate communities and composition in natural habitats.
 
@@
 \subsection{Data}
-For this example with ecological data we are going to use data from McClean (unpubl.) that contains an ordination of a distance matrix between different experimental plots.
+For this example with ecological data we are going to use data from McClean (unpubl.). %INC CITE!
+This data is the ordination of a distance matrix based on nutrient enrichment and depth and freshwater benthic invertebrates.
 
@@
 
-## let's use the matrix from McClean (unpubl.)
+## Loading the data
 data(McClean_data)
-## This dataset contains an ordinated matrix (in 20 dimensions) of the distance
-## between experimental plots.
+
+## This dataset contains an ordinated matrix (20 dimensions) of the distance
+## between 40 experimental plots.
 ord_matrix <- McClean_data$ordination
+
 ## As well as two list of different factors affecting each experimental plot:
@@
 \section{A classical two dimensional approach}
-A classical way to represent this ordinated data will be to use PCA plots.
+A classical way to represent this ordinated data will be to use two dimensional plots.
 
 <<PCAplot, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
-## The x and y axis represent the two first dimensions of the PCA
+## The x and y axis represent the two first dimensions of the ordination
 x <- ord_matrix[, 1]
@@
 pchs <- as.numeric(sub(2, 17, pchs))
+
 ## Graphical option
 par(bty = "n")
-## A classic PCA plot
-plot(x, y, col = cols, pch = pchs, xlab = "PCA1", ylab = "PCA2",
+## A classic 2D ordination plot
+plot(x, y, col = cols, pch = pchs, xlab = "Dimension 1", ylab = "Dimension 2",
     xlim = range(x) + c(0, 100))
@@
 
-However, a common problem in this type of multivariate analysis is that often only a handful of dimensions are studied (usually the ones that bears the most variance of the ordinated matrix).
+This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix (i.e. the first two dimensions).
+However, the problem, is that this plot ignores the 18 other dimensions of the ordination!
+Additionally, these two represented dimensions do not represent a biological reality \textit{per se}; i.e.
+the values on the first dimension do not represent some continuous traits (e.g.
+depth) but just the ordinations of correlations between the data and some factors.
 
-This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix.
-However, the problem, is it ignores the 18 other axis of the ordination and the PCA axis 1 and 2 do not represent a biological reality \textit{per se} but more some ordinations of correlations between the data and some factors.
 Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a \textit{n}-dimensional object.
-In practice, we might be interested in looking how some experimental treatment for example, will affect the position of our experimental plots in this \textit{n}-dimensional object.
-For example: do the experimental plots shift in some specific space of the \textit{n}-dimensional object when depth increases?
+To explore our hypothesis, we could look if there are differences in the hyper-volume that different treatments occupy in the ordinated space to see the effect of water treatments on the communities.
 
@@
 
-As a first split, one might want to split the data (i.e. the \textit{n}-dimensional object) into subsamples that we want to compare.
-First let's make a factor table:
+The first step in such analysis will be to create different series that are two sub-samples of the ordinated space (i.e. the \textit{n}-dimensional object).
+Each of these series contain a certain number of elements (i.e. the 40 experimental field plots) that have some attributes in common.
+In our example, we are going to group the elements according to their depth and treatment.
 
 <<factors>>=
-## Making the factor table
+## Creating the table that contain the elements and their attributes
 factors <- as.data.frame(matrix(data = c(treatments, depth), nrow = nrow(ord_matrix),
@@
 names(factors)<-c("Treat", "Depth")
-## And here is what it looks like
 head(factors)
@@
 
-Second, let's split the data according to these factors to create the subsamples of the ordinated space by using the \texttt{cust.series} function:
+Second, let's split the data according to these factors to create the series of the ordinated space by using the \texttt{cust.series} function:
 
@@
 customised_series <- cust.series(ord_matrix, factors)
+
 ## Note that the output of dispRity functions are dispRity objects
@@
 
-For more details on the \dispRity objects, see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}.
+For more details on the \dispRity objects, see the \href{https://github.com/TGuillerme/dispRity}{\dispRity manual}.
 Basically the idea is to avoid jamming the \R console such as when using:
@@
 <<dispRity class, eval=FALSE>>=
-## Summarise the object
+## Summarising the object
 str(customised_series)
+
+## Displaying the full object
+print.dispRity(customised_series, all=TRUE)
 @
@@
 
-Now we're going to see the functionalities of the core function of this package: the \dispRity function.
-This function is a modulable function that allow to simply (and quickly!) calculate disparity from a matrix.
-Disparity can be calculated in many ways, this function is a tool to measure disparity \textit{as defined by the user} (and here's where the modulable part comes in).
-For more details on disparity, see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}.
+Disparity can be calculated in many ways, and therefore, the \dispRity function allows to measure disparity \textit{as defined by the user}.
+For more details on disparity, see the \href{https://github.com/TGuillerme/dispRity}{other vignettes}.
 
-One can usually decompose the disparity metrics into two elements:
-\begin{enumerate}
-\item the \textbf{class metric} that is a descriptor of the matrix.
-For example describing the ranges of each column in the matrix or the euclidean distances between each row and the centroid of the matrix.
-\item the \textbf{summary metric} that is a summary of the class metric values. For example, the sum of the ranges or the median of the euclidean distances. 
-\end{enumerate}
-Basically the combination can be infinite between the class and summary metrics.
-For example, people might want to measure the median variances of the axis or the product of the distances from the centroid.
-However, it is probable that some metrics are better to reflect some biological aspects of the any-o-space than others...
-
-In practice, the \dispRity function intakes a pair of class and summary metrics as a definition of disparity.
-Several of these metrics are implemented in other packages (like \texttt{stats::median}, \texttt{base::sum}, etc.) and this package proposes several metrics listed in \texttt{dispRity.metric} (see \texttt{?dispRity.metric}).
-But it is even possible to use your very own class and summary metrics! Or even any metric you decide makes sense (note that this is not yet implemented in v.\Sexpr{packageVersion("dispRity")}).
-This will be actually heavily encouraged and facilitate with the \texttt{make.metric} function in a future version.
-
-To use these metrics pairs in the \dispRity function, it's pretty easy:
+In this example, we are going to define disparity as the median distance between the different experimental plots and the centroid of the ordinated space.
+High values of disparity will indicate a general high dispersal from this centroid (i.e. on average, the experimental plots are far apart in the ordinated space).
+We can define the metrics easily in the \dispRity function by feeding them to the \texttt{metric} argument.
+Here we are going to feed the functions \texttt{median::stats} and \texttt{centroids::dispRity} which calculates distances between elements and centroid.
 
 <<dispRity1>>=
-## For example, let's calculate the median distance between each plot and the
-## centroid of the ordinated space
+## Calculating disparity
 disparity <- dispRity(customised_series, metric = c(median, centroids))
+
 ## Note that disparity is a dispRity object and printing it just gives details
@@
 ## the results.
+disparity
 summary(disparity)
@@
 
-Note that we calculated the median distance between plots and the centroid but the output displays mean values.
-This is because summary will, by default, summarise the data using the mean value.
-Here the mean represents the mean of the median distance for each series which is a bit useless (i.e. the mean of one value is that same value).
-We can display the median as well using:
+\subsection{Bootstrapping the data}
+One might also want to bootstrap the data to test the robustness of the measured disparity to outliers.
+Also, as we can see, each series has different numbers of elements.
+It might be interesting to rarefy the data as well to have only series with the same number of elements.
+Both steps are easily doable through the \texttt{boot.matrix} function.
 
-<<dispRity2>>=
-summary(disparity, cent.tend=median)
-## Or even the product! It won't affect the results
-summary(disparity, cent.tend=prod)
+<<boot.matrix>>=
+## Simple bootstrapping (1000 times)
+bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100)
+## Bootstrapping with rarefaction (i.e. only re-sampling 17 elements each time)
+rarefied_data <- boot.matrix(customised_series, bootstraps = 100, rarefaction = 17)
+## Note that the output is a dispRity object giving some details on the series and the bootstraps
+bootstrapped_data
+rarefied_data
 @
 
-This is because we did calculate the noise within our data.
-We can classically do so by bootstrapping the data!
-
-\subsection{Bootstrapping the data}
-
-The \dispRity pacakge also provides easy way to bootstrap the data via the \texttt{boot.matrix} function.
-We can even rarefy the data to see the effect of the number of plots per series:
-For more details on the bootstrapping options, see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}.
-
-<<boot.matrix>>=
-bootstrapped_data <- boot.matrix(customised_series, bootstraps=100)
-rarefied_data <- boot.matrix(customised_series, bootstraps=100, rarefaction=TRUE)
-## Note that the output is a dispRity object giving some details on the series and the bootstraps
-bootstrapped_data
-@
-
-We can now rerun a more robust disparity analysis:
+\noindent We can now rerun a more robust disparity analysis using the bootstrapped data:
 
 <<dispRity3>>=
+## Bootstrapped disparity
 disparity_BS <- dispRity(bootstrapped_data, metric = c(median, centroids))
+## Rarefied disparity
 disparity_rare <- dispRity(rarefied_data, metric = c(median, centroids))
-## Note that calculation time is increased!
 @
@@
 <<summary>>=
-## The default:
+## Summarising disparity (default)
 summary(disparity_BS)
@@
 ## Finally we can see the results of the rarefaction analysis:
-head(summary(disparity_rare))
-## This outputs a longer table with all the variations of plots down to 3 plots per series.
+summary(disparity_rare)
+## Note that for the first three series, there is no observed disparity since the data was
+## rarefied for these series.
 @
 
-Finally we can also plot the results using the simple \texttt{plot.dispRity S3} method:
-
-<<disparity plots, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
-## Graphical option
-par(bty = "n")
-## Plotting the score for each groups
-plot(disparity_BS)
-@
-
-Or have a look at the effect of the number of experimental plots:
+\noindent We can also plot the results and have a look at the effect of the number of experimental plots:
 
@@
 quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
-## The same but looking at the number of plots
-plot(disparity_rare, elements = TRUE)
-## With the same number of plots per group
-plot(disparity_rare, elements = TRUE, rarefaction = 17)
+## Plotting the bootstrapped disparity
+plot(disparity_BS, main="Bootstrapped data")
+## Plotting the rarefied disparity
+plot(disparity_rare, main="Rarefied data")
 @
 
-Or event have a look at the rarefaction curves:
+\noindent As we can see, there seems to be no strong effect of the number of experimental plots in each series (i.e.
+the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
 
-<<rare plots2, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
-## Graphical option
-par(bty = "n")
-## Plotting the rarefaction curves
-plot(disparity_rare, rarefaction = "plot")
+\subsection{Testing the hypothesis}
+Finally, we can test our hypothesis (whether the sort of water treatment at certain depth alters invertebrate communities and composition in natural habitats) by using the \texttt{test.dispRity} function.
+
+<<test>>=
+## Testing the effect of our factors on the bootstrapped data
+summary(test.dispRity(disparity_BS, test = aov, comparisons = "all"))
+## Testing the effect of our factors on the rarefied data
+summary(test.dispRity(disparity_rare, test = aov, comparisons = "all"))
+
+## Post-hoc testing of the effect of the two different factors
+## Note that the comparison contains the list of the pairs of series to compare
+test.dispRity(disparity_BS, test = wilcox.test, comparison = list(c(1,2), c(3,4)))
+## Testing the effect of the two different factors for the rarefied data
+test.dispRity(disparity_rare, test = wilcox.test, comparison = list(c(1,2), c(3,4)))
 @
 
-\subsection{Testing the hypothesis}
+\noindent As we can see, there is strong support for an effect of the treatment and the depth on the median distance between each experimental plots and the centroid of the ordinated space.
+In other words, it seems that with the second treatment and the second level of depth, the invertebrate communities and composition were further apart (i.e.
+more dispersed in the ordinated space) than with the first treatment and the first level of depth.
 
-\bibliographystyle{sysbio}
-\bibliography{References}
+% \bibliographystyle{sysbio}
+% \bibliography{References}
 

### 2015-11-28 doc/dispRity-ecology-demo.Rnw
manuals updates

@@
 This demo aims to give quick overview of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) for ecological analysis.
-Please refer to \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes, namely the \dispRity tutorial that explains the functions in more details.
+Please refer to \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes, namely the \dispRity tutorial that explains the functions in more detail.
 
-To keep it short, this package allows to use all the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
+To keep it short, this package allows the use of all of the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
 For example, one might want to know whether the sort of water treatment at certain depth alters invertebrate communities and composition in natural habitats.
@@
 This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix (i.e. the first two dimensions).
+At a first glance, it seems difficult to see a clear effect from the treatments or the depth since the different experimental plots with the same parameters don't seem to cluster.
 However, the problem, is that this plot ignores the 18 other dimensions of the ordination!
@@
 Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a \textit{n}-dimensional object.
-To explore our hypothesis, we could look if there are differences in the hyper-volume that different treatments occupy in the ordinated space to see the effect of water treatments on the communities.
+To explore our hypothesis, we could look to see if there are differences in the hyper-volume that different treatments occupy in the ordinated space to see the effect of water treatments on the communities.
 

### 2015-11-28 doc/dispRity-palaeo-demo.Rnw
manuals updates

@@
 This demo aims to give quick overview of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) for palaeobiology analysis.
-Please refer to \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes, namely the \dispRity tutorial that explains the functions in more details.
+Please refer to \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes, namely the \dispRity tutorial that explains the functions in more detail.
 

### 2015-11-29 doc/dispRity-manual.Rnw
added random seed to manual
Updated manual

@@
 \item \textbf{Ordinated space}: it designates the mathematical multidimensional object studied here.
-In morphometrics, this one is often referred as being the morphospace, or the cladisto-space for cladisticts or the eco-space for ecology, etc.
-In practice though, this term designates an ordinated matrix where the columns represent the dimensions of the ordinated space (often $>$ 3!) and the rows represent the elements within this space.
+In morphometrics, this one is often referred as being the morphospace.
+However it can also be the the cladisto-space for cladistic data or the eco-space in ecology, etc.
+In practice, this term designates an ordinated matrix where the columns represent the dimensions of the ordinated space (often -- but not necessarily -- $>$ 3!) and the rows represent the elements within this space.
 \item \textbf{Elements}: it designates the rows of the ordinated space, elements can be either taxa, field sites, countries, etc...
@@
 <<installation, eval=FALSE>>=
-install.packages("devtools")
-library(devtools)
+if(!require(devtools)) install.packages("devtools")
 install_github("TGuillerme/dispRity", ref = "release")
@@
 \noindent Note that we use the \texttt{release} branch here which is version \Sexpr{packageVersion("dispRity")}.
-For the piping-hot (but potentially full of bugs) version, one can change the argument \texttt{ref = "release"} to \texttt{ref = "master"}.
+For the piping-hot (but potentially full of bugs) version, you can change the argument \texttt{ref = "release"} to \texttt{ref = "master"}.
 This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.
@@
 \subsection{Data}
-In this tutorial we are going to use a subset of the ordinated cladistic data from \cite{beckancient2014} that contains 50 taxa ordinated using their cladistic distance (i.e.
+In this tutorial we are going to use a subset of the ordinated cladistic data from \cite{beckancient2014} that contains 50 taxa (elements) ordinated using their cladistic distance (i.e.
 the distance between their discrete morphological characters).
 Note that this data is more oriented towards palaebiology analysis but that it can apply to other disciplines.
-Please refer to the \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes covering some specific example of the package with ecological data.
+Please refer to the \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes covering some specific example.
 
@@
 \subsection{A quick go through}
-Here is a really crude and quick go through the package to show some of the features of this package.
+Here is a really crude and quick analysis to go through the package, showing some of its features.
 Note that all these features will be discussed in more details below.
@@
 sliced_data <- time.series(BeckLee_mat99, BeckLee_tree, method = "continuous",
-    model = "acctran", time = rev(seq(from=0, to=130, by=15)), FADLAD = BeckLee_ages)
+    model = "acctran", time = rev(seq(from=0, to=130, by=15)),
+    FADLAD = BeckLee_ages)
 ## Bootstrapping the data
@@
 plot(sum_of_variances)
-## Testing some the effect of time on disparity
+## Testing the effect of time on disparity
 summary(test.dispRity(sum_of_variances, test = aov, comparisons = "all"))
@@
 str(sum_of_variances)
-## That's a more than 4500 lines of output!
+## That's a more than 4000 lines of output!
 @
@@
 <<dispRity class2>>=
-## Which class is the sum_of_variances object?
+## What is the class of the sum_of_variances object?
 class(sum_of_variances)
-## What's in the object
+## What is in the object?
 names(sum_of_variances)
-## We can summarise it using the S3 method print.dispRity
+## Summarising it using the S3 method print.dispRity:
 sum_of_variances
-## This displays some information about what's in the object
+## This displays some basic information about the object content
 @
@@
 ## Displaying the full object
-print(sum_of_variances, all=TRUE)
+print(sum_of_variances, all = TRUE)
 @
@@
 ## Extracting some specific series from the disparity object
-series_1_and_4 <- get.dispRity(sum_of_variances, what=c(1,4))
+series_1_and_4 <- get.dispRity(sum_of_variances, what = c(1,4))
 series_1_and_4
@@
 <<dispRity modular>>=
-## Disparity measured as the mean
+## Disparity measured as the mean value of the ordinated matrix
 summary(dispRity(sliced_data, metric = mean))
 ## A function for measuring the sum divided by the length (the mean!)
-sum.by.length <- function(X)  sum(X)/length(X)
+sum.divide.by.length <- function(X) sum(X)/length(X)
 ## Disparity measured as the mean divided by the length
-summary(dispRity(sliced_data, metric = sum.by.length))
+summary(dispRity(sliced_data, metric = sum.divide.by.length))
 ## Giving the exact same results!
@@
 One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter \textit{series}).
-These series correspond at regions of the ordinated space that we want to compare to each other.
+These series correspond to regions of the ordinated space that we want to compare to each other.
 
@@
 str(crown_stem$data)
-## "taxa" being the list of taxa in the original ordinated matrix
-str(crown_stem$taxa)
-## and finally, "series" containing information on the series type (custom) and names
-## (crown and stem)
+## "elements" being the list of taxa in the original ordinated matrix
+str(crown_stem$elements)
+## and finally, "series" containing information on the series type (custom)
+## and names (crown and stem)
 crown_stem$series
@@
 Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
-The \texttt{time.series} function allows to create series that contain all the elements present at specific points in time.
+The \texttt{time.series} function allows to create series that contain all the elements present at specific points or during specific periods in time.
 This functions needs as input an ordinated space and a matching phylogenetic tree.
@@
 \item defining the boundaries of the time bins (when \texttt{method = "discrete"});
-\item defining the time-slicing places (when using \texttt{method = "continuous"}).
+\item defining the time slices (when using \texttt{method = "continuous"}).
 \end{enumerate}
 Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a given number of equidistant time-bins/slices.
-Additionally, it is also possible to input a data frame containing the First/Last Occurrence Data (\texttt{FADLAD}) for taxa that span over a longer time than the tips/nodes age.
+Additionally, it is also possible to input a data frame containing the First/Last Occurrence Data (\texttt{FAD/LAD}) for taxa that span over a longer time than the tips/nodes age.
 
@@
 ## Automatically generate three equal length bins:
-time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete", time = 3)
+time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+    time = 3)
 @
-\noindent We now have three time bins of 44.50368 Ma each.
+\noindent We now have three time bins of 44.50368 million years each.
 
-When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1\% of the tree length as long as at least three elements are present at that point in time (an extra 1\% is added until reaching the required minimum of three elements).
-We can also ask to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between tips \textit{AND} nodes.
+When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1\% of the tree length as long as at least three elements are present at that point in time.
+The algorithm adds an extra 1\% tree length until reaching the required minimum of three elements.
+It is also possible to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between tips \textit{AND} nodes.
 
@@
 <<continuous.series>>=
-## Generating four time slices every 40 Ma assuming a gradual evolution model
+## Generating four time slices every 40 million years assuming a gradual
+## evolution model
 time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
@@
 
-Once we obtain our different series, we might want to bootstrap or/and rarefy it (i.e. pseudo-replicating the data).
+Once we obtain our different series, we might want to bootstrap and rarefy it (i.e. pseudo-replicating the data).
 The bootstrap will allow us to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of elements to get rid of eventual sampling problems.
@@
 The default options will bootstrap the matrix 1000 times without rarefaction.
-The number of bootstrap pseudo-replicates can be defined using the \texttt{bootstraps} option (see below)
-.
+The number of bootstrap pseudo-replicates can be defined using the \texttt{bootstraps} option (see below).
+
 <<boot.matrix1>>=
@@
 \end{enumerate}
-This function also also to rarefy the data using the \texttt{rarefaction} argument.
+
+This function also allows to rarefy the data using the \texttt{rarefaction} argument.
 The default argument is \texttt{FALSE} but it can be set to \texttt{TRUE} to fully rarefy the data (i.e.
@@
 
-Of course, one could be interested in directly supplying the subsampled matrices generated above directly to this function.
+Of course, one could be interested in directly supplying the sub-sampled matrices generated above directly to this function.
 In fact, it can also deal with a list of matrices or with a \dispRity object output from the \texttt{cust.series} or \texttt{time.series} functions.
-let's bootstrap our previously generate series:
 
@@
 ## Bootstrap and full rarefaction on the crown/stem series
-crown_stemBS <- boot.matrix(crown_stem, rarefaction = TRUE)
+crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
 ## Bootstrap on the time binning/slicing series
-time_binsBS <- boot.matrix(time_bins)
-time_slicesBS <- boot.matrix(time_slices)
+time_binsBS <- boot.matrix(time_bins, bootstraps = 100)
+time_slicesBS <- boot.matrix(time_slices, bootstraps = 100)
 ## Note that all these objects are of class dispRity
@@
 
-Now we're going to see the functionalities of the core function of this package: the \dispRity function.
-This function is a modulable function that allow to simply (and quickly!) calculate disparity from a matrix.
+This function is a modular function that allow to simply (and quickly!) calculate disparity from a matrix.
 
-Because disparity can be measured in many ways, this function is a tool to measure disparity \textit{as defined by the user} (and here's where the modulable part comes in).
+Because disparity can be measured in many ways, this function is a tool to measure disparity \textit{as defined by the user}.
 In fact, the \dispRity function intakes two main arguments: the data and the disparity metric.
-%To change in future version!
-The disparity metric boils down to a single value that describes the matrix.
-Some are classic like the sum or the product of the ranges or the variances of the matrix columns (i.e. the any-o-space axis/dimensions) \cite{Wills1994}.
-However, many more exist like the median distances between the taxa and the any-o-space centroid \cite{GuillermeSTD}.
+The disparity metric is a function or a set of functions that summarises the ordinated matrix to a single value that represents, in our example, the diversity of morphologies.
 
-One can usually decompose the disparity metrics into two elements:
-\begin{enumerate}
-\item the \textbf{class metric} that is a descriptor of the matrix.
-For example describing the ranges of each column in the matrix or the euclidean distances between each row and the centroid of the matrix.
-\item the \textbf{summary metric} that is a summary of the class metric values. For example, the sum of the ranges or the median of the euclidean distances. 
-\end{enumerate}
-Basically the combination can be infinite between the class and summary metrics.
-For example, people might want to measure the median variances of the axis or the product of the distances from the centroid.
-However, it is probable that some metrics are better to reflect some biological aspects of the any-o-space than others (see section \ref{whatsleft})...
+The \dispRity algorithm decomposes the metrics functions into three levels that correspond to the dimensions of the output of each metric function.
+For more details on this algorithm please refer to the \href{https://github.com/TGuillerme/dispRity}{metric vignette}.
 
-In practice, the \dispRity function intakes a pair of class and summary metrics as a definition of disparity.
-Several of these metrics are implemented in other packages (like \texttt{stats::median}, \texttt{base::sum}, etc.) and this package proposes several metrics listed in \texttt{dispRity.metric} (see \texttt{?dispRity.metric}).
-It is even possible to use your very own class and summary metrics!
-This will be actually heavily encouraged and facilitate with the \texttt{make.metric} function in a future version (see \ref{whatsleft}).
-
-To use these metrics pairs in the \dispRity function, it's pretty easy:
+In practice, the \dispRity function intakes one or more functions as a definition of disparity.
+Several of these functions will be already implemented in other packages (such as \texttt{stats::median}, \texttt{base::sum}, etc.); some others are implemented in this package (listed in \texttt{?dispRity.metric}) and finally some others will be defined by the users.
+The \texttt{make.metric} function is designed to help users create and test their own disparity metric functions.
+In practice, the use of these metrics in the \dispRity function, is pretty easy:
 
 <<dispRity1>>=
-dispRity(BeckLee_mat50, metric = c(sum, ranges))
-## Yep, that's it. The order of the metrics does not matter (the function detects
-## automatically which one is the class and the summary metric) so the combinations
-## are numerous!
+dispRity(BeckLee_mat50, metric = mean)
+## This defines disparity as the mean value of the morphospace (using
+## mean::base).
+
+## It is also possible to combine multiple functions:
 dispRity(BeckLee_mat50, metric = c(sum, variances))
+## Defining disparity as the sum (sum::base) of the variances
+## (variances::dispRity) of each dimension of the morphospace.
+
 dispRity(BeckLee_mat50, metric = c(prod, centroids))
-dispRity(BeckLee_mat50, metric = c(mode.val, ranges))
-## Or funny user defined ones!
+## For the product (prod::base) of the distances between each elements and the
+## centroid of the morphospace (centroids::dispRity).
+
+## Or user defined ones:
 total.range <- function(X) abs(range(X)[1]-range(X)[2])
-dispRity(BeckLee_mat50, metric = c(total.range, ranges))
-## That would be disparity as the range of the axis ranges!
+dispRity(BeckLee_mat50, metric = c(total.range, centroids))
+## For the range (total.range, user defined) of the distances between each
+## elements and the centroid of the morphospace (centroids::dispRity).
+
+## Or more complex ones:
+vars1 <- dispRity(BeckLee_mat50, metric = c(sd, variances, var))
+## For the standard deviation (sd::stats) of the variance (variances::dispRity)
+## of each column of the variance/covariance matrix (var::stats).
+
+## Note that the order of the function is not important since the levels of
+## each function are automatically detected by the dispRity algorithm.
+vars2 <- dispRity(BeckLee_mat50, metric = c(variances, var, sd))
+all(summary(vars) == summary(vars2))
 @
 
-Of course, this function can take a simple ordinated matrix but, more interesting it can also deal with a list of matrices (from the \texttt{cust.series} or \texttt{time.series} functions for example) or a bootstrapped or/and rarefied matrix (from the \texttt{boot.matrix} function for example).
+\noindent Note that these functions do not directly output the disparity values but only the summary of the \dispRity objects.
+To display the results, see the section \ref{summary}.
+
+In these examples we used only simple ordinated matrix but of course, it might be more interesting to directly use the \dispRity objects we generated in the steps above.
+Thus we can calculate the bootstrapped and rarefied disparity in each sub-sample of the morphospace.
+In this example we are going to define disparity as being the sum of the variances in each dimension)=.
 
 <<dispRity2>>=
-## For example, let's calculate the sum of variances from our crown/stem mammals
-dispRity(crown_stem, metric = c(sum, variances))
-## Or even better, from the bootstrapped data
+## Disparity in crown and stem mammals:
 disp_crown_stemBS <- dispRity(crown_stemBS, metric = c(sum, variances))
-## Note that the bootstrapped + rarefied data takes some time to run (it's
-## calculating disparity from 46 matrices!)
 
-## Let's calculate a different metric for the disparity-through-time analysis
+## Disparity through time:
 disp_time_binsBS <- dispRity(time_binsBS, metric = c(median, centroids))
@@
 
-Again, note that the results are \dispRity objects and can displayed through the S3 \texttt{print.dispRity} method to remind users what are in these objects.
-Also note that an even faster parallel version will be developed in the future (see section \ref{whatsleft}).
+\noindent Note that the computational time was longer for the rarefied data (\texttt{crown\_stemBS}) since it had to calculate the disparity on all the rarefied matrix, each bootstrapped 100 times.
+A faster parallel version will be implemented in future releases (see section \ref{whatsleft}).
+
+% STOPPED HERE
 
@@
 As shown in these examples above however, even though the disparity has been calculated in many way, the function doesn't directly outputs the results, rather just a summary of the analysis.
-To really display the results we can use the S3 \texttt{summary.dispRity} and \texttt{plot.dispRity} functions.
 
-\subsubsection{\texttt{summary} function}
 This function intakes a \dispRity object plus various options namely the \texttt{quantile} values for the confidence intervals levels; the \texttt{cent.tend} for the central tendency to use for summarising the results and two \textit{visual} options which are whether to recall the \dispRity options and how much digits are wanted in the results.

### 2015-11-30 doc/dispRity-manual.Rnw
added plot and summary to the manual

@@
 ## We want to separate the species around the node 71 (see phylogeny above).
-## All the descendant of these node are crown and all the ancestors are stem
+## All the descendant of these node are crown and all the ancestors are stem.
 crown <- extract.clade(BeckLee_tree, node = 71)$tip.label
@@
 \subsection{\texttt{boot.matrix}}
+\label{rare}
 
@@
 vars2 <- dispRity(BeckLee_mat50, metric = c(variances, var, sd))
-all(summary(vars) == summary(vars2))
+all(summary(vars1) == summary(vars2))
 @
@@
 
-% STOPPED HERE
-
 \subsection{\texttt{summary}}
@@
 
-As shown in these examples above however, even though the disparity has been calculated in many way, the function doesn't directly outputs the results, rather just a summary of the analysis.
-
+This function is a S3 function (\texttt{summary.dispRity}) allowing to summarize the content of \dispRity objects that contain disparity calculations.
 This function intakes a \dispRity object plus various options namely the \texttt{quantile} values for the confidence intervals levels; the \texttt{cent.tend} for the central tendency to use for summarising the results and two \textit{visual} options which are whether to recall the \dispRity options and how much digits are wanted in the results.
@@
 <<summary>>=
-## Let's have a look at the disparity-through-time using in the time-binned data
 summary(disp_time_binsBS)
-## The 50 and 95 quantiles are calculated and the central tendency is the mean by default.
-## Note that the function also displays the observed disparity (non-bootstrapped).
-## We can change that easily by specifying different values in the options
+## By default, the 50 and 95 quantiles and the mean are calculated.
+## Note that the function also displays the observed disparity
+## (non-bootstrapped).
+
+## These arguments can be changed easily as follow
 summary(disp_time_binsBS, quantile = 88, cent.tend = sd, rounding = 1)
-## Also, if we happen to have forgotten what was in the disp_time_binsBS object (as in
-## which options where used to modify the original matrix) we can use the recall option:
+
+## If on happens to forgot what was calculated in the summarised object,
+## it is possible to recall the different steps details using recall:
 summary(disp_time_binsBS, recall = TRUE)
-## Note that the recall just prints text. The output of summary.dispRity is always a
-## data.frame
-summary_table <- summary(disp_time_binsBS, recall = TRUE)
-class(summary_table)
 
-## Finally we can see the results of the rarefaction analysis for the crown/stem data
-head(summary(disp_crown_stemBS))
-## This outputs a longer table with all the variations of the number of crown/stem taxa
-## from the maximum (30 and 20) to the minimum (3). Note that the observed disparity
-## is NA for rarefied data.
+## Note that the information with each different number of elements is
+## displayed for rarefied data:
+summary(disp_crown_stemBS)
+## The observed disparity is NA for rarefied data.
 @
@@
 
-This final function allows to visualise the disparity results in a often nicer fashion than just a table (even though the exact same data is displayed).
-The \texttt{plot.dispRity} option intakes the same options as \texttt{summary.dispRity} along side with various graphical options described in the function manual.
-Here, let's just have a look a few of these options.
+A nicer way to summarise the data is to plot the results using the S3 method \texttt{plot.dispRity}.
+This function intakes the same options as \texttt{summary.dispRity} along side with various graphical options described in the function manual (see \texttt{?plot.dispRity}).
 
-Let's start with plotting the difference in disparity between the crown and the stem mammals.
-For this we are going to simply specify the type of plots using the \texttt{type} option that allows to choose between the \texttt{"continuous"} or \texttt{"discrete"} method.
-In our case, we are interested in looking at the results in a discrete way.
-Note that the plot \texttt{type} can be left missing.
-The function will then use \texttt{"discrete"} as default or \texttt{"continuous"} if the input data came from the \texttt{time.series} function with \texttt{method = "continuous"}.
+Two main categories of plots can be displayed: discrete plots with the argument \texttt{type = "discrete"} and continuous plots with the argument \texttt{type = "continuous"} (see below).
+Note that this argument can be left missing.
+In this case, the algorithm will automatically detect the type of series from the \dispRity object.
 
-The grey squares represent the confidence intervals (50 in dark grey and 90 in light grey) and the dot represents the mean.
-However, it might be interesting to see how many taxa are present in each group.
-This can be done via the \texttt{elements} option.
-A third interesting option is to plot the rarefied data if a rarefaction was applied on the data with the \texttt{boot.matrix} function.
-For displaying the rarefied data we can use the \texttt{rarefaction} option.
-This option can be either \texttt{FALSE} (default) for always plotting the maximum number of taxa per series or \texttt{TRUE} to always plot the minimum.
-Additionally, it can be a specific number to plot a specific number of taxa.
-Let say we want both groups to have 20 taxa maximum.
-We can see now that the confidence intervals overlap more than previously
+The \texttt{type = "discrete"} argument plots superimposed boxes that represent the different levels of confidence intervals and a dot representing the central tendency.
+It is also possible to display the number of elements per series (has a horizontal doted line) using the option \texttt{elements = TRUE}.
+Additionally, when the data is rarefied (see section \ref{rare}), one can also indicate which level of rarefaction to display (i.e.
+display the results for a certain number of elements) by using the \texttt{rarefaction} argument.
 
@@
 quartz(width = 15, height = 5) ; par(mfrow = (c(1,3)), bty = "n")
-## Plotting the score for both groups
-plot(disp_crown_stemBS, ylim = c(25,40))
-## Same one with the number of taxa for each data set
-plot(disp_crown_stemBS, elements = TRUE, ylim = c(25,40))
-## Same one with rarefied data (note how that affects the elements data as well!)
-plot(disp_crown_stemBS, elements = TRUE, rarefaction = 20, ylim = c(25,40))
+
+## Plotting the disparity for both groups
+plot(disp_crown_stemBS, type = "discrete")
+## Same plot but with the number of elements on the right. Note that the type 
+## argument is not necessary since the "discrete" type is automatically detected
+## from the input dispRity object.
+plot(disp_crown_stemBS, elements = TRUE)
+## Rarefied version of the same plot (20 elements per series)
+plot(disp_crown_stemBS, elements = TRUE, rarefaction = 20)
 @
 
-To understand how the number of taxa affects each series, we can use the \texttt{rarefaction = "plot"} function to plot the rarefaction curves.
+\noindent It is also possible to easily look at the behaviour of the disparity metric with the rarefied data (i.e.
+how does the disparity score change when the number of elements goes down) by using the \texttt{rarefaction = "plot"} argument:
 
@@
 plot(disp_crown_stemBS, elements = TRUE, rarefaction = "plot")
-## Note that the function automatically draws the curves for each series and splits
-## the plot screen accordingly.
 @
 
-The different dashed lines represent the different confidence intervals around each rarefaction curve.
+\noindent The different dashed lines represent the different confidence intervals around each rarefaction curve.
+We can see that the disparity metric becomes stable after more than 10 elements.
 
-Additionally, for the \texttt{type = "discrete"} option, we might also decide to plot many boxplots (which can be a bit messy).
-The \texttt{discrete\_type} option allows to switch between \texttt{"box"} or \texttt{"line"} for showing the results in one or the other format.
-We can also plot the observed data when existing (i.e. not with some rarefaction levels) by using \texttt{observed = TRUE}.
-Let's look at that with the time binned data (even though there is only three time bins).
-Finally, we can use the \texttt{type = "continuous"} option that comes in handy for plotting continuous data like in the time sliced analysis (obviously...).
+It is also possible, when using the \texttt{type = "discrete"} argument, to plot lines rather than box by switching the \texttt{discrete\_type} from \texttt{"box"} to \texttt{"line"} to display one dimensional boxplots.
+This option can be useful when displaying many series.
+Additionally, it is possible to plot the observed data when existing (i.e. not when the data is rarefied) by using the option \texttt{observed = TRUE}.
 
-<<plot3, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
+<<plot3, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
 ## Graphical options
-quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
+par(bty = "n")
 ## The disparity-through-time data in a time-binned way (with lines rather than
@@
 plot(disp_time_binsBS, discrete_type = "line", observed = TRUE)
-## The disparity-through-time data in a time-sliced ways
-plot(disp_time_slicesBS)
 
@@
 
-We can even plot both results (discrete and continuous) on the same graph:
+The \texttt{type = "continuous"} argument plots superimposed continuous (along the x axis) polygons that represent the confidence intervals and a continuous line that represents the central tendency of the data.
+Note that, as a S3 method function, \texttt{plot.dispRity} can take the classic graphical options from the \texttt{plot} function or even to add other plots to the previous ones.
 
@@
 par(bty = "n")
-## First, plotting the continuous data
-plot(disp_time_slicesBS, ylim=c(0.5,2))
+
+## First, plotting the continuous data with a y axis limit
+plot(disp_time_slicesBS, type = "continuous", ylim = c(0.5,2))
+## Note that the type = "continuous" argument is not mandatory here.
+
 ## Second, adding the discrete data
@@
 
-Note that the results vary in values but not in pattern.
-The change in values might be due to the fact that nodes data are not included in the discrete analysis.
-
-I encourage you to play with the graphical options to have some prettier results.
+\noindent I encourage you to play with the graphical options to have some prettier results.
 Note that most of the options from \texttt{plot} can be passed to \texttt{plot.dispRity} via \texttt{...}.
@@
 ## A plot with many options!
-plot(disp_time_slicesBS, quantile = c(seq(from=10, to=100, by=10)), cent.tend = sd,
-    type = "continuous", elements = "log", col = c("black", rainbow(10)),
-    ylab = c("Disparity", "log(Diversity)"), xlab = "Time (in in units from
-    past to present)", time.series = FALSE, observed = TRUE, main = "Many options...")
+plot(disp_time_slicesBS, quantile = c(seq(from=10, to=100, by=10)),
+    cent.tend = sd, type = "continuous", elements = "log",
+    col = c("black", rainbow(10)), ylab = c("Disparity", "log(Diversity)"),
+    xlab = "Time (in in units from past to present)", time.series = FALSE,
+    observed = TRUE, main = "Many options...")
 @

### 2015-12-01 doc/dispRity-manual.Rnw

@@
 
-<<plot4, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
+<<plot4, fig.width=8, fig.height=8, out.width='.5\\linewidth'>>=
 ## Graphical options
@@
 
+Finally, the \dispRity package allows to apply some tests to the data in order to properly test hypothesis.
+The function \texttt{test.dispRity} works in a similar way as the \dispRity function: it intakes a \dispRity object (containing disparity measurements), a \texttt{test} and a \texttt{comparison} argument.
+The \texttt{test} argument can be any statistical or non-statistical test to apply to the disparity object.
+It can be common test function (e.g. \texttt{t.test::stats}) or some functions implemented in \dispRity (see \texttt{dispRity.test}) or even user defined functions.
+The \texttt{comparison} argument must indicate the way the test should be applied to the data:
+\begin{itemize}
+\item \texttt{"pairwise"} (default): to compare each series pairwise
+\item \texttt{"referential"}: to compare each series to the first one
+\item \texttt{"sequential"}: to compare each series to the following one
+\item \texttt{"all"}: to compare all the series together (like in analysis of variance)
+\end{itemize}
+It is also possible to input a list of pairs of \texttt{numeric} values or \texttt{characters} matching the series names to create personalised test.
+This function also allows to correct for type I error inflation when using multiple comparisons via the \texttt{correction} argument.
+This argument can be empty (no correction applied) or can contain one of the corrections from the \texttt{p.adjust::stats} function (see \texttt{?p.adjust}).
+
+Note that the \texttt{test.dispRity} algorithm deals with some classic tests outputs (\texttt{h.test}, \texttt{lm} and \texttt{numeric} vector) and summarises the test output.
+It is however possible to get the full detailed output by using the options \texttt{details = TRUE}.
+
+<<test.dispRity>>=
+## Performing a pairwise t-test to the test a difference in disparity between
+## crown and stem mammals (note that "pairwise" is default)
+test.dispRity(disp_crown_stemBS, test = t.test)
+
+## The same test but with the detailed output
+test.dispRity(disp_crown_stemBS, test = t.test, details = TRUE)
+
+## A wilcoxon rank test applied to sliced disparity the data sequentially with
+## the Bonferroni correction.
+test.dispRity(disp_time_slicesBS, test = wilcox.test,
+    comparisons = "sequential", correction = "bonferroni")
+
+## An analysis of variance (aov) applied to the same data
+test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all")
+
+## Note that in this case, the output is the regular aov output:
+slice_aov <- test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all")
+class(slice_aov) ; summary(slice_aov)
+
+## Measuring the overlap between distributions in the time bins (using the
+## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff) with a
+## personalised series comparisons and some additional arguments to pass to the
+## test function (bw = bw.nrd)
+test.dispRity(disp_time_binsBS, test = bhatt.coeff, bw = bw.nrd,
+    comparisons = list(c(1,2), c(1,3), c(2,1), c(3,1)))
+
+@
+
 \subsection{utilities}
@@
 As stated at the start of the demo, this version \Sexpr{packageVersion("dispRity")} is still in development and many parts are missing.
-Here are the new functionalities that will be implemented before the proper release version (v.1).
+Here are the new functionalities that will be implemented in further versions.
 
-\subsection{More user defined stuff}
+\subsection{More user defined functions}
 I intend also develop functions to help users to develop their own algorithms for the bootstrap method (via \texttt{make.boot}) or the evolutionary models (via \texttt{make.model}).
@@
 
+\subsection{Suggestions?}
+If you have any extra suggestions or comments on what has already been developed or will be developed, please send me an email (\href{mailto:guillert@tcd.ie}{guillert@tcd.ie}) or if you are a GitHub user, directly create an issue on the \href{https://github.com/TGuillerme/dispRity}{GitHub page (github.com/TGuillerme/dispRity)}.
+
 \bibliographystyle{sysbio}

### 2015-12-01 doc/dispRity-palaeo-demo.Rnw

@@
 
-This demo showcases a typical disparity-through-time analyse: we are going to test whether the disparity changed through time in this dataset since the last 100 million years.
+This demo showcases a typical disparity-through-time analyse: we are going to test whether the disparity changed through time in a subset of Eutheria since the last 100 million years.
 
@@
 <<strap, eval=FALSE>>=
-install.packages(strap)
-library(strap)
+if(!require(strap)) install.packages("strap")
 geoscalePhylo(BeckLee_tree, cex.tip=0.8, cex.ts=0.6)
@@
 
-The \dispRity package provides a \texttt{time.series} function that allows to split the morphospace time slices (using \texttt{method="continuous"}) or into time bins (using \texttt{method="discrete"}).
+The \dispRity package provides a \texttt{time.series} function that allows to split the morphospace time slices (using \texttt{method = "continuous"}) or into time bins (using \texttt{method = "discrete"}).
 In this example, we are going to split the morphospace into 5 equal time bins of 20 million years long from 100 million years ago (Mya) to the present.
-Note that we can also provide here the table containing the first and last occurrences data for some fossils to take into account some fossils that might span between our different bins.
+Note that we will also provide to the function the table containing the first and last occurrences data for some fossils to take into account the fact that some fossils might span between our different time bins.
 
@@
 
-The warning message tells us that some taxa were considered as single points in time because we didn't provide any temporal range information for some taxa.
+The warning message tells us that some taxa were considered as single points in time because we didn't provide temporal range information for all the taxa: they are therefore considered as unique time points without any occurrence range.
+
 The output object is a \dispRity object.
-For details about this object class, please refer to the \href{https://github.com/TGuillerme/dispRity}{other vignettes}.
-In brief however, \dispRity objects are lists of different elements (i.e.
-disparity results, morphospace sub-samples, etc...) that display only a summary of the object when calling the object for visualisation reasons.
+For details about this object class, please refer to the \href{https://github.com/TGuillerme/dispRity}{the \dispRity manual}.
+In brief, \dispRity objects are lists of different elements (i.e.
+disparity results, morphospace sub-samples, etc...) that display only a summary of the object when calling the object for avoiding jamming the \R console.
 
@@
 
-Note that these objects will gradually contain more information when reaching higher steps in the disparity-through-time analysis.
+Note that these objects will gradually contain more information when reaching the following steps in the disparity-through-time analysis.
 
 \subsection{Bootstrapping the data}
-Once we obtain our different subsamples, we might want to bootstrap or/and rarefy it (i.e. pseudo-replicating the data).
-The bootstrap will allow us to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of taxa to get rid of eventual sampling problems.
-For that, we can use the \texttt{boot.matrix} function that allows to bootstrap the \dispRity object along with the \texttt{rarefaction} option that will rarefy it.
+Once we obtain our different subsamples, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
+The bootstrapping will allow to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of taxa to get rid of eventual sampling problems (i.e.
+more taxa in one bin than the other).
+The \texttt{boot.matrix} function that allows to bootstrap the \dispRity object and the \texttt{rarefaction} option within allows to rarefy it.
 
@@
 
-## ... and display some more information on the previous step when called
+## ... and display some more information on the previous step when printed:
 boot_bin_morphospace
@@
 
-We can now calculate the disparity within each sub-sample along side with some confidence intervals generated by the pseudo-replication step above (bootstrap/rarefaction).
+We can now calculate the disparity within each sub-sample along side with some confidence intervals generated by the pseudo-replication step above (bootstraps/rarefaction).
 Disparity can be calculated in many ways and this package allows user to come up with their own disparity metrics.
-For more info, please refer to the \href{https://github.com/TGuillerme/dispRity}{other vignettes}.
+For more info, please refer to the \href{https://github.com/TGuillerme/dispRity}{\dispRity manual and the metric vignette}.
 
 In this example, we are going to calculate the spread of the data in each sub-sample by calculating the disparity as the sum of the variance of each dimension of the morphospace in each sub-sample using the \dispRity function.
+Thus, in this example, the disparity is defined by the multi-dimensional variance of each sub-sample (i.e. the spread of the taxa within the morphospace).
+Note that this metric comes with a caveat not solved here since it ignores the covariance among the dimensions of the morphospace.
 
@@
 ## Plotting the bootstrapped and rarefied results
-plot(boot_disparity, main = "bootstrapped results")
-plot(rare_disparity, main = "rarefied results")
+plot(boot_disparity, type = "continuous", main = "bootstrapped results")
+plot(rare_disparity, type = "continuous", main = "rarefied results")
 
@@
 As stated before, we wanted to test whether the disparity changed through time in this dataset since the last 100 million years.
-To do so, we can simply apply some comparisons of the means between each time-bins in a sequential manner to see whether the disparity in bin X is equal to the disparity in bin Y which is in turn equal to the disparity in bin Z, etc.
+To do so, we can apply some comparisons of the means between each time-bins in a sequential manner to see whether the disparity in bin X is equal to the disparity in bin Y which is in turn equal to the disparity in bin Z, etc.
 Because our data is temporally auto-correlated (i.e.
-what happens in bin Y is dependent on what happened in bin X) and was pseudo-replicated (i.e.
+what happens in bin Y depends on what happened in bin X) and was pseudo-replicated (i.e.
 each bootstrap draw creates non-independent sub-samples because they are all based on the same sub-sample), we are going to apply a non-parametric mean comparison: the \texttt{wilcox.test}.
-Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis.
+Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis (see \texttt{?p.adjust} for more details).
 

### 2016-03-15 doc/dispRity-ecology-demo.Rnw
Ecological demo vignette update

@@
 
-%1-Introduce the data
-%"the data represents experimental plots in different treatments at two different water depths"
-
-%2-Explain rarefaction better
-%"the data represents experimental plots in different treatments at two different water depths"
-
-%3-Hypothesis
-%We want to know whether the treatments had an effect on the density of different species in the plots (Does treatment alter invertebrate community composition?).
-And does depth have an effect on community composition in the experimental plots?
-Annnnd: Do the treatments affect the communities differently at different depths?
-
-%(Personally I hate them as you know!)  Just use them as a way of visually showing that the 'multivariate' communities in the different treatments/ depths are different or similar and I guess which components explain most of the variance.
-They really aren't that powerful which is why I love your approach.
-You sold it so well to me, I think your best selling points (and indeed perhaps actually including a sentence at the start) are that you don't lose all the dimensions of the data like you do in PCA.
-And that you can keep the important ones in (again like you sold to me, that, some axes may be far more influential than a pca with throw up).
-
 \author{Thomas Guillerme \\ guillert@tcd.ie}
@@
 To keep it short, this package allows the use of all of the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
-For example, one might want to know whether the sort of water treatment at certain depth alters invertebrate communities and composition in natural habitats.
+For example, one might want to know whether a specific kind of water treatment at certain water depth alters invertebrate communities and composition in natural habitats.
 
@@
 @
-
-This is a quick demo for using the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) in ecological analysis.
-See the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos} for a general demo of the \dispRity package.
 
@@
 depth <- McClean_data$depth
+
 @
@@
 <<PCAplot, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
-## The x and y axis represent the two first dimensions of the ordination
+## The x and y axis represent the two first dimensions (i.e. principal
+## components) of the ordination
 x <- ord_matrix[, 1]
@@
 ## A classic 2D ordination plot
-plot(x, y, col = cols, pch = pchs, xlab = "Dimension 1", ylab = "Dimension 2",
+plot(x, y, col = cols, pch = pchs, xlab = "PC 1", ylab = "PC 2",
     xlim = range(x) + c(0, 100))
@@
 This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix (i.e. the first two dimensions).
-At a first glance, it seems difficult to see a clear effect from the treatments or the depth since the different experimental plots with the same parameters don't seem to cluster.
+At a first glance, it seems difficult to see a clear effect from the treatments (blue or red) or the depth (rounds or triangles) since the different experimental plots with the same parameters don't seem to cluster.
 However, the problem, is that this plot ignores the 18 other dimensions of the ordination!
@@
 Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a \textit{n}-dimensional object.
-To explore our hypothesis, we could look to see if there are differences in the hyper-volume that different treatments occupy in the ordinated space to see the effect of water treatments on the communities.
 
@@
     ncol = 2, byrow = FALSE, dimnames = list(rownames(ord_matrix))))
-names(factors)<-c("Treat", "Depth")
+names(factors) <- c("Treat", "Depth")
 head(factors)
@@
 ## Displaying the full object
-print.dispRity(customised_series, all=TRUE)
+print(customised_series, all = TRUE)
 @
@@
 We can define the metrics easily in the \dispRity function by feeding them to the \texttt{metric} argument.
-Here we are going to feed the functions \texttt{median::stats} and \texttt{centroids::dispRity} which calculates distances between elements and centroid.
+Here we are going to feed the functions \texttt{stats::median} and \texttt{dispRity::centroids} which calculates distances between elements and centroid.
 
@@
 <<boot.matrix>>=
-## Simple bootstrapping (1000 times)
+## Simple bootstrapping (100 times)
 bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100)
+
 ## Bootstrapping with rarefaction (i.e. only re-sampling 17 elements each time)
-rarefied_data <- boot.matrix(customised_series, bootstraps = 100, rarefaction = 17)
-## Note that the output is a dispRity object giving some details on the series and the bootstraps
+rarefied_data <- boot.matrix(customised_series, bootstraps = 100,
+    rarefaction = 17)
+
+## Note that the output is a dispRity object giving some details on the series
+## and the bootstraps
 bootstrapped_data
@@
 disparity_BS <- dispRity(bootstrapped_data, metric = c(median, centroids))
+
 ## Rarefied disparity
@@
 summary(disparity_BS)
-## The quantiles are calculated as 50 and 95 and the central tendency is the mean by default.
-## But we can specify different options
-summary(disparity_BS, quantile=90, cent.tend=median)
+
+## The quantiles are calculated as 50 and 95 and the central tendency is the
+## mean by default but we can specify different options
+summary(disparity_BS, quantile = 90, cent.tend = median)
 
@@
 summary(disparity_rare)
-## Note that for the first three series, there is no observed disparity since the data was
-## rarefied for these series.
+
+## Note that for the first three series, there is no observed disparity since
+## the data was rarefied for these series.
 @
@@
 quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
+
 ## Plotting the bootstrapped disparity
-plot(disparity_BS, main="Bootstrapped data")
+plot(disparity_BS, main = "Bootstrapped data")
+
 ## Plotting the rarefied disparity
-plot(disparity_rare, main="Rarefied data")
+plot(disparity_rare, main = "Rarefied data")
 @
@@
 summary(test.dispRity(disparity_BS, test = aov, comparisons = "all"))
+
 ## Testing the effect of our factors on the rarefied data
@@
 ## Note that the comparison contains the list of the pairs of series to compare
-test.dispRity(disparity_BS, test = wilcox.test, comparison = list(c(1,2), c(3,4)))
+test.dispRity(disparity_BS, test = wilcox.test,
+    comparisons = list(c(1,2), c(3,4)))
+
 ## Testing the effect of the two different factors for the rarefied data
-test.dispRity(disparity_rare, test = wilcox.test, comparison = list(c(1,2), c(3,4)))
+test.dispRity(disparity_rare, test = wilcox.test,
+    comparisons = list(c(1,2), c(3,4)))
 @

### 2016-03-15 doc/dispRity-manual.Rnw
Updates on the manual

@@
 \item \textbf{Elements}: it designates the rows of the ordinated space, elements can be either taxa, field sites, countries, etc...
-\item \textbf{Dimensions}: it designates the columns of the ordinated space. The dimensions can also be referred to as axis.
+\item \textbf{Dimensions}: it designates the columns of the ordinated space. The dimensions can also be referred to as axis or principal components.
 \item \textbf{Series}: it designates sub-samples of the ordinated space.
-Basically a series contain the same number of dimensions as the morphospace but might contain a smaller number of elements.
+Basically a series contain the same number of dimensions as the morphospace but may contain a smaller number of elements.
 For example, if our ordinated space is composed of birds and mammals (i.e.
-the elements) and 50 dimensions, we can create two series of just mammals or birds as elements (but the same 50 dimensions) to look at the difference in disparity between both groups.
+the elements) and 50 dimensions, we can create two series of just mammals or birds as elements (but with the same 50 dimensions) to look at the difference in disparity between the two clades.
 \end{itemize}
@@
 For the piping-hot (but potentially full of bugs) version, you can change the argument \texttt{ref = "release"} to \texttt{ref = "master"}.
-This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.
+This package depends mainly on the \texttt{ape} package and uses functions of several other (\texttt{ade4}, \texttt{geometry}, \texttt{grDevices}, \texttt{hypervolume}, \texttt{paleotree} and \texttt{snow}).
 
@@
 data(BeckLee_tree)
-plot(BeckLee_tree) ; nodelabels(cex=0.8) ; axisPhylo(root=140)
+plot(BeckLee_tree, cex = 0.8) ; nodelabels(cex = 0.8) ; axisPhylo(root = 140)
+
 @
@@
     FADLAD = BeckLee_ages)
+
 ## Bootstrapping the data
 bootstrapped_data <- boot.matrix(sliced_data, 100)
+
 ## Calculating disparity
-sum_of_variances <- dispRity(bootstrapped_data, metric = c(sum, variances))
+median_centroids <- dispRity(bootstrapped_data, metric = c(median, centroids))
+
 ## Summarising the results
-summary(sum_of_variances)
-plot(sum_of_variances)
+summary(median_centroids)
+plot(median_centroids)
+
 ## Testing the effect of time on disparity
-summary(test.dispRity(sum_of_variances, test = aov, comparisons = "all"))
+summary(test.dispRity(median_centroids, test = aov, comparisons = "all"))
 @
@@
 \subsection{The \dispRity objects}
-Disparity analysis can involve a lot of shuffling around with many matrices (especially when bootstrapping the data) which can be a bit impractical to visualise and quickly jam your \R console.
+Disparity analysis can involve a lot of shuffling around with many matrices (especially when bootstrapping the data) which can be impractical to visualise and will quickly jam your \R console.
 For example, we can have a look at the structure of the object created in the quick example:
@@
 <<dispRity class1, eval=FALSE>>=
-str(sum_of_variances)
+str(median_centroids)
 ## That's a more than 4000 lines of output!
@@
 
-\noindent Therefore this package proposes a new class of object called \dispRity objects.
-These objects allow to easily use a S3 method functions such as \texttt{summary.dispRity} (just called as \texttt{summary}; see section \ref{summary}) or \texttt{plot.dispRity} (just called as \texttt{plot}; see section \ref{plot}).
-But also, this allows to use the S3 method for printing \dispRity objects via \texttt{print.dispRity} that allows to summarise the content of the objects similar to the \texttt{phylo} class objects (see \texttt{print.phylo::ape}).
+\noindent Therefore this package proposes a specific class of object called \dispRity objects.
+These objects allow to easily use S3 method functions such as \texttt{summary.dispRity} (just called as \texttt{summary}; see section \ref{summary}) or \texttt{plot.dispRity} (just called as \texttt{plot}; see section \ref{plot}).
+But also, this allows to use the S3 method for printing \dispRity objects via \texttt{print.dispRity} that allows to summarise the content of the objects similar to the \texttt{phylo} class objects (see \texttt{ape::print.phylo}).
 
 <<dispRity class2>>=
-## What is the class of the sum_of_variances object?
-class(sum_of_variances)
+## What is the class of the median_centroids object?
+class(median_centroids)
+
 ## What is in the object?
-names(sum_of_variances)
+names(median_centroids)
+
 ## Summarising it using the S3 method print.dispRity:
-sum_of_variances
+median_centroids
 ## This displays some basic information about the object content
@@
 ## Displaying the full object
-print(sum_of_variances, all = TRUE)
+print(median_centroids, all = TRUE)
+## But that's 155091 lines!
 @
@@
 ## Extracting some specific series from the disparity object
-series_1_and_4 <- get.dispRity(sum_of_variances, what = c(1,4))
+series_1_and_4 <- get.dispRity(median_centroids, what = c(1,4))
 series_1_and_4
+
 ## The observed disparity
-extract.dispRity(sum_of_variances)
-## The list of bootstrapped scores of disparity
-str(extract.dispRity(sum_of_variances, observed = FALSE))
+extract.dispRity(median_centroids)
+
+## The distribution of the bootstrapped disparity scores for the first series
+str(extract.dispRity(get.dispRity(median_centroids, what = 1),
+    observed = FALSE))
 @
@@
 This package aims to be a modular package where users can personalise some aspects of the package fairly easily.
-In this version \Sexpr{packageVersion("dispRity")} only one function is fully modular (\dispRity; see section \ref{disparity}) but more will be hopefully available in the near future (see section \ref{developments}).
-
-The modular idea is that users can use implemented tools to help facilitating their own personalised function.
+The modular aspect allows the package users to use several implemented tools to help them creating personalised functions.
 For example, the \dispRity function intake a \texttt{metric} argument designating how disparity should be calculated.
@@
 summary(dispRity(sliced_data, metric = mean))
+
 ## A function for measuring the sum divided by the length (the mean!)
 sum.divide.by.length <- function(X) sum(X)/length(X)
+
 ## Disparity measured as the mean divided by the length
@@
 
-\noindent For more information concerning this modularity, please refer to the \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for the vignette about the metric implementation in \dispRity.
+\noindent For more information, please refer to the \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for the vignette about the metric implementation in \dispRity.
 
 \section{Functions}
+In the following section, different functionalities are detailed.
+The functions are presented in a ``classic'' pipelined study fashion for convenience (i.e. splitting the data, bootstrapping it, calculating disparity, testing differences).
+Note however that most of these functions can be used in any order and that some steps can of course be skipped.
+
 One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter \textit{series}).
@@
 
+\noindent Note that here we separating the taxa into two exclusive groups (i.e. crown mammals can be crown only, same for stem mammals).
+It is however possible to add more factors with overlapping groups:
+
+<<cust.series3>>=
+## Adding all mammals to the a new series (along with crown and stem as before)
+factors2 <- as.data.frame(
+    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem)),
+    rep("Mammalia", Ntip(BeckLee_tree))), ncol = 2,
+    dimnames = list(c(crown, stem), c("Group", "Class")), byrow = FALSE))
+
+## Creating the three series: Group.stem/crown and Class.Mammalia
+cust.series(BeckLee_mat50, factors2)
+@
+
 \subsection{\texttt{time.series}}
@@
 \item defining the boundaries of the time bins (when \texttt{method = "discrete"});
-\item defining the time slices (when using \texttt{method = "continuous"}).
+\item defining the time slices (when \texttt{method = "continuous"}).
 \end{enumerate}
 Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a given number of equidistant time-bins/slices.
-Additionally, it is also possible to input a data frame containing the First/Last Occurrence Data (\texttt{FAD/LAD}) for taxa that span over a longer time than the tips/nodes age.
+Additionally, it is also possible to input a data frame containing the first and last occurrence data (\texttt{FAD/LAD}) for taxa that span over a longer time than the tips/nodes age.
 
@@
 
-<<discrete.series>>=
+<<discrete.series1>>=
 ## Generating three time bins containing the data present every 40 Ma
-time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
-    method = "discrete", time = c(120, 80, 40, 0))
-## Note that the function provides a warning saying that tips where single
-## points in time (no FAD/LAD information). We can fix that by adding the
-## age data for the taxa that have some longer occurrence spans.
+time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+    time = c(120, 80, 40, 0))
+@
+
+\noindent In this example, the taxa where split inside each timebin according to their age.
+Note however that some taxa might span between some time bins and should be included in more than one.
+This is possible by providing a table containing the first and last occurrence data:
+
+<<discrete.series1.1>>=
+## Generating three time bins containing the data present every 40 Ma, including
+## taxa that might span between time bins.
 time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
     method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
-## To entirely avoid the warning we could collect the occurrence span data
-## for all the taxa but that's not necessary. The function automatically
-## assumes no occurrence span time (i.e. single points in time) for all taxa
-## by default.
 time_bins
@@
 @
+
 \noindent We now have three time bins of 44.50368 million years each.
@@
 The algorithm adds an extra 1\% tree length until reaching the required minimum of three elements.
-It is also possible to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between tips \textit{AND} nodes.
+It is also possible to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between tips \textit{and} nodes.
 
 For the time-slicing method (\texttt{method = "continuous"}), the idea is really similar.
-This option intakes a matrix that contains the ordinated distance between taxa \textit{AND} nodes and an assumed evolutionary model via the \texttt{model} argument:
+This option intakes a matrix that contains the ordinated distance between taxa \textit{and} nodes and an assumed evolutionary model via the \texttt{model} argument:
 \begin{enumerate}
@@
 Once we obtain our different series, we might want to bootstrap and rarefy it (i.e. pseudo-replicating the data).
-The bootstrap will allow us to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of elements to get rid of eventual sampling problems.
+The bootstrap will allow us to make each series more robust to outliers and the rarefaction will allow us to compare slices with the same number of elements to get rid of eventual sampling problems.
 The \texttt{boot.matrix} allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
@@
 
-Additionally, this function allows to control the bootstrap algorithm through the \texttt{boot.type} argument.
+This function allows to control the bootstrap algorithm through the \texttt{boot.type} argument.
 Currently two algorithms are implemented:
@@
 remove $n$ elements for the number of pseudo-replicates, where $n$ varies from the maximum number of elements present in the dataset to a minimum of 3 elements).
-It can also be set to a fix \texttt{numeric} value (or a set of numeric values).
+It can also be set to a fix \texttt{numeric} value (or a set of \texttt{numeric} values).
 Finally, one last argument, \texttt{rm.last.axis} allows to remove a certain amount of dimensions (or axis) for the ordinated space.
@@
 boot.matrix(BeckLee_mat50, boot.type = "single")
+
 ## Bootstrapping with the full rarefaction
 boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
+
 ## Or with a set number of rarefaction levels
 boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8,3))
+
 ## And removing the last axis (default)
 boot.matrix(BeckLee_mat50, rm.last.axis = TRUE)
+
 ## Or with a fix value (50%)
@@
 crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
+
 ## Bootstrap on the time binning/slicing series
@@
 time_slicesBS <- boot.matrix(time_slices, bootstraps = 100)
+
 ## Note that all these objects are of class dispRity
@@
 
-This function is a modular function that allow to simply (and quickly!) calculate disparity from a matrix.
+This function is a modular function that allows to simply (and quickly!) calculate disparity from a matrix.
 
@@
 dispRity(BeckLee_mat50, metric = mean)
-## This defines disparity as the mean value of the morphospace (using
-## mean::base).
+## This defines disparity as the mean value of the morphospace
+## (using base::mean).
 
@@
 dispRity(BeckLee_mat50, metric = c(sum, variances))
-## Defining disparity as the sum (sum::base) of the variances
-## (variances::dispRity) of each dimension of the morphospace.
+## Defining disparity as the sum (base::sum) of the variances
+## (dispRity::variances) of each dimension of the morphospace.
 
 dispRity(BeckLee_mat50, metric = c(prod, centroids))
-## For the product (prod::base) of the distances between each elements and the
-## centroid of the morphospace (centroids::dispRity).
+## For the product (base::prod) of the distances between each elements and the
+## centroid of the morphospace (dispRIty::centroids).
 
@@
 ## For the range (total.range, user defined) of the distances between each
-## elements and the centroid of the morphospace (centroids::dispRity).
+## elements and the centroid of the morphospace (dispRity::centroids).
 
@@
 vars1 <- dispRity(BeckLee_mat50, metric = c(sd, variances, var))
-## For the standard deviation (sd::stats) of the variance (variances::dispRity)
-## of each column of the variance/covariance matrix (var::stats).
+## For the standard deviation (stats::sd) of the variance (dispRity::variances)
+## of each column of the variance/covariance matrix (stats::var).
 
@@
 @
+
+%STOPPED HERE
 

### 2016-03-15 doc/dispRity-manual.Rnw
Manual update

@@
 <<boot.matrix1>>=
+## Bootstrapping a matrix
 boot.matrix(data = BeckLee_mat50)
@@
 
-%STOPPED HERE
-
 \noindent Note that these functions do not directly output the disparity values but only the summary of the \dispRity objects.
@@
 Thus we can calculate the bootstrapped and rarefied disparity in each sub-sample of the morphospace.
-In this example we are going to define disparity as being the sum of the variances in each dimension)=.
+In this example we are going to define disparity as being the median distance between each element and each dimensions centroids.
 
@@
 ## Disparity in crown and stem mammals:
-disp_crown_stemBS <- dispRity(crown_stemBS, metric = c(sum, variances))
+disp_crown_stemBS <- dispRity(crown_stemBS, metric = c(median, centroids))
 
@@
 \noindent Note that the computational time was longer for the rarefied data (\texttt{crown\_stemBS}) since it had to calculate the disparity on all the rarefied matrix, each bootstrapped 100 times.
-A faster parallel version will be implemented in future releases (see section \ref{whatsleft}).
+For both the \texttt{boot.matrix} and the \dispRity functions, it is possible to increase computational speed in some cases using the \texttt{parallel} option (see each function's manuals).
 
@@
 ## displayed for rarefied data:
-summary(disp_crown_stemBS)
+tail(summary(disp_crown_stemBS))
 ## The observed disparity is NA for rarefied data.
@@
 
-A nicer way to summarise the data is to plot the results using the S3 method \texttt{plot.dispRity}.
+An alternative way to summarise the data is to plot the results using the S3 method \texttt{plot.dispRity}.
 This function intakes the same options as \texttt{summary.dispRity} along side with various graphical options described in the function manual (see \texttt{?plot.dispRity}).
 
-Two main categories of plots can be displayed: discrete plots with the argument \texttt{type = "discrete"} and continuous plots with the argument \texttt{type = "continuous"} (see below).
+Two main categories of plots can be displayed: continuous plots with the argument \texttt{type = "continuous"} (see below) and discrete plots (default is \texttt{type = "box"}.
 Note that this argument can be left missing.
@@
 
-The \texttt{type = "discrete"} argument plots superimposed boxes that represent the different levels of confidence intervals and a dot representing the central tendency.
+The discrete plot arguments can be either:
+\begin{itemize}
+\item \texttt{type = "box"} for plotting boxplots.
+\item \texttt{type = "polygon"} for plotting boxes that represent the different levels of confidence intervals (with a dot for the central tendency).
+\item \texttt{type = "lines"} which is similar to \texttt{type = "polygon"} but plots lines rather than boxes for the confidence intervals levels.
+\end{itemize}
 It is also possible to display the number of elements per series (has a horizontal doted line) using the option \texttt{elements = TRUE}.
@@
 ## Plotting the disparity for both groups
-plot(disp_crown_stemBS, type = "discrete")
-## Same plot but with the number of elements on the right. Note that the type 
-## argument is not necessary since the "discrete" type is automatically detected
-## from the input dispRity object.
-plot(disp_crown_stemBS, elements = TRUE)
+plot(disp_crown_stemBS, type = "box")
+
+## Same plot but with the number of elements on the right and using the
+## "polygon" plot type.
+plot(disp_crown_stemBS, type = "polygon", elements = TRUE)
 ## Rarefied version of the same plot (20 elements per series)
-plot(disp_crown_stemBS, elements = TRUE, rarefaction = 20)
+plot(disp_crown_stemBS, type = "polygon", elements = TRUE, rarefaction = 20)
 @
@@
 
-It is also possible, when using the \texttt{type = "discrete"} argument, to plot lines rather than box by switching the \texttt{discrete\_type} from \texttt{"box"} to \texttt{"line"} to display one dimensional boxplots.
-This option can be useful when displaying many series.
+It is also possible, when using the \texttt{type = "lines"} argument, to plot lines rather than boxes.
+This option can be useful when displaying many series in a same window.
 Additionally, it is possible to plot the observed data when existing (i.e. not when the data is rarefied) by using the option \texttt{observed = TRUE}.
@@
 ## boxes) with the observed data as crosses
-plot(disp_time_binsBS, discrete_type = "line", observed = TRUE)
+plot(disp_time_binsBS, type = "lines", observed = TRUE)
 
@@
 ## First, plotting the continuous data with a y axis limit
-plot(disp_time_slicesBS, type = "continuous", ylim = c(0.5,2))
+plot(disp_time_slicesBS, type = "continuous")
 ## Note that the type = "continuous" argument is not mandatory here.
-
-## Second, adding the discrete data
-par(new = TRUE)
-plot(disp_time_binsBS, discrete_type = "line", ylim=c(0.5,2), xlab="", ylab="")
 @
@@
 
+\noindent Finally, it is possible to add multiple plots on top of each other:
+
+<<plot5.1, fig.width=8, fig.height=8, out.width='.5\\linewidth'>>=
+## Graphical options
+par(bty = "n")
+
+## First, plotting the continuous data with a y axis limit
+plot(disp_time_slicesBS, ylim = c(0.5,2))
+
+## Second, adding the discrete data
+par(new = TRUE)
+plot(disp_time_binsBS, discrete_type = "line", ylim=c(0.5,2), xlab="", ylab="")
+@
+
 \subsection{\texttt{test.dispRity}}
@@
 Finally, the \dispRity package allows to apply some tests to the data in order to properly test hypothesis.
-The function \texttt{test.dispRity} works in a similar way as the \dispRity function: it intakes a \dispRity object (containing disparity measurements), a \texttt{test} and a \texttt{comparison} argument.
+The function \texttt{test.dispRity} works in a similar way as the \dispRity function: it intakes a \dispRity object (containing disparity measurements), a \texttt{test} and a \texttt{comparisons} argument.
 The \texttt{test} argument can be any statistical or non-statistical test to apply to the disparity object.
-It can be common test function (e.g. \texttt{t.test::stats}) or some functions implemented in \dispRity (see \texttt{dispRity.test}) or even user defined functions.
-The \texttt{comparison} argument must indicate the way the test should be applied to the data:
+It can be common test function (e.g. \texttt{stats::t.test}) or some functions implemented in \dispRity (see \texttt{dispRity.test}) or even user defined functions.
+The \texttt{comparisons} argument must indicate the way the test should be applied to the data:
 \begin{itemize}
@@
 It is also possible to input a list of pairs of \texttt{numeric} values or \texttt{characters} matching the series names to create personalised test.
+Some other specific tests implemented in \dispRity such as the \texttt{dispRity::null.test} and the \texttt{dispRity::sequential.test} have a specific way to be applied to the data and therefore ignore the \texttt{comparisons} argument.
+
 This function also allows to correct for type I error inflation when using multiple comparisons via the \texttt{correction} argument.
-This argument can be empty (no correction applied) or can contain one of the corrections from the \texttt{p.adjust::stats} function (see \texttt{?p.adjust}).
+This argument can be empty (no correction applied) or can contain one of the corrections from the \texttt{stats::p.adjust} function (see \texttt{?p.adjust}).
 
@@
 
-\subsection{Faster!}
-Finally, for long analysis, I intend to develop a parallel running version of the package.
-In fact, most of the internal functions are base on \texttt{lapply} functions that can be easily passed to \texttt{snow::parLapply} or similar parallel functions.
-
 \subsection{Suggestions?}

### 2016-03-17 doc/dispRity-metrics.Rnw
metric update

@@
 In fact, the core function of the package, \dispRity, allows the user to define any metric in the \texttt{metric} argument.
-However, because of the infrastructure of the package (especially the on of the \dispRity objects -- see the \href{https://github.com/TGuillerme/dispRity}{other vignettes}), the \texttt{metric} argument has to follow certain slightly restrictive rules described in detail below.
+However, because of the infrastructure of the package (especially the one of the \dispRity objects -- see the \href{https://github.com/TGuillerme/dispRity}{the \dispRity manual}), the \texttt{metric} argument has to follow certain rules:
 \begin{enumerate}
-    \item \texttt{metric} must be between one or three \texttt{function} objects
-    \item \texttt{metric} must can have only one argument as input: a \texttt{matrix} or a \texttt{vector}
-    \item \texttt{metric} individual functions must be classified in one of the three levels (see below)
-    \item \texttt{metric} one of the functions must be a level 1 function (see below)
+    \item it must be composed of one to three \texttt{function} objects;
+    \item these functions can only have as input a \texttt{matrix} or a \texttt{vector};
+    \item each of these functions must be of one of the three levels described below;
+    \item at least one of the functions must be a level 1 function (see below).
 \end{enumerate}
@@
 \subsection{A level 1 function}
-A level 1 function for example, will decompose the \texttt{matrix} (or a \texttt{vector} output from a level 2 function, see below) into a single value:
+A level 1 function for example, will decompose a \texttt{matrix} or a \texttt{vector} into a single value:
 
@@
 dummy_matrix <- matrix(rnorm(12), 4, 3)
+
 ## Any summary metric such as mean or median are good examples of level 1
-## functions as they reduce the matrix to a single dimension (value):
+## functions as they reduce the matrix to a single dimension (i.e. one value):
 mean(dummy_matrix)
@@
 \subsection{A level 2 function}
-A level 2 function will decompose the \texttt{matrix} into a \texttt{vector}.
-Several level 2 functions are implemented in \dispRity (see \texttt{?dispRity.metric}) such as the \texttt{variances} function that calculates the variance of each column of the ordinated matrix (i.e.
-it's axis or dimensions):
+A level 2 function will decompose a \texttt{matrix} into a \texttt{vector}.
+Several level 2 functions are implemented in \dispRity (see \texttt{?dispRity.metric}) such as the \texttt{variances} function that calculates the variance of each dimension of the ordinated matrix.
 
 <<level 2 example>>=
-## We can define a different function that outputs the products of the rows:
-rows <- function(matrix) apply(matrix, 1, prod)
-## rows is a level 2 function as well (outputs a vector of 4 values):
-rows(dummy_matrix)
+## We can define a function that outputs the products of each row of the matrix:
+prod.rows <- function(matrix) apply(matrix, 1, prod)
+
+## This function is a level 2 function as it reduces the matrix into a two
+## dimensional element (a vector of values):
+prod.rows(dummy_matrix)
 @
@@
 The \texttt{make.metric} function implemented in \dispRity is designed to help testing and finding the level of the functions.
-Concretely, This function tests:
+This function tests:
 \begin{enumerate}
-  \item if your function can deal with a \texttt{matrix} or a \texttt{vector} as an \texttt{input}.
-  \item which level is your function (level 1, 2 or 3, see above).
-  \item whether the function can properly be implemented in the \texttt{dispRity} function.
+  \item if your function can deal with a \texttt{matrix} or a \texttt{vector} as an input;
+  \item which is your function's level according to it's output (level 1, 2 or 3, see above);
+  \item whether the function can properly be implemented in the \texttt{dispRity} function (the function is fed into a \texttt{lapply} loop.
 \end{enumerate}
@@
 library(dispRity)
-## Which level is the mean function? And is able to be used in the dispRity function?
+
+## Which level is the mean function? And is able to be used in the dispRity?
 make.metric(mean)
-## Which level is the rows function? And is able to be used in the dispRity function?
-make.metric(rows)
+
+## Same questions from the prod.rows function:
+make.metric(prod.rows)
 ## Note that the function also tells us that we will need a level 1 function as well.
 ## We'll cover this below.
-## Which level is the var function? And is able to be used in the dispRity function?
+
+## Same questions for the var function:
 make.metric(var)
@@
 data(BeckLee_mat50)
-## Measuring disparity as the standard deviation of all the value of the ordinated
-## matrix (level 1 function).
+
+## Measuring disparity as the standard deviation of all the value of the
+## ordinated matrix (level 1 function).
 summary(dispRity(BeckLee_mat50, metric = sd))
+
 ## Measuring disparity as the standard deviation of the variance of each axis of
@@
 summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))
+
 ## Measuring disparity as the standard deviation of the variance of each axis of
@@
 summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round=10)
-## Note that the order of each function in the metric argument does not matter,
-## the algorithm always classifies them by levels and computes first level 3,
-## then level 2 and finally level 1.
-summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)), round=10)
-## Both ways output the same disparity values.
+
+@
+
+\noindent Note that the order of each function in the metric argument does not matter, the \dispRity function will automatically detect the function levels (using \texttt{make.metric}) and apply them to the data in decreasing order (level 3 $>$ level 2 $>$ level 1).
+
+<<dispRity example2>>=
+## Disparity as the standard deviation of the variance of each axis of the
+## variance covariance matrix:
+disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)),
+  round=10)
+
+## Same but using a different functions order for the metric argument
+disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
+  round=10)
+
+## Both ways output the same disparity values:
+disparity1$observed == disparity2$observed # is TRUE
 @

### 2016-03-17 doc/dispRity-metrics.Rnw
Added proper links to the manual

@@
 In fact, the core function of the package, \dispRity, allows the user to define any metric in the \texttt{metric} argument.
-However, because of the infrastructure of the package (especially the one of the \dispRity objects -- see the \href{https://github.com/TGuillerme/dispRity}{the \dispRity manual}), the \texttt{metric} argument has to follow certain rules:
+However, because of the infrastructure of the package (especially the one of the \dispRity objects -- see the \href{https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf}{\dispRity manual}), the \texttt{metric} argument has to follow certain rules:
 \begin{enumerate}

### 2016-03-17 doc/dispRity-palaeo-demo.Rnw
Minor typo in loading strap (library(strap))
All manual updated

@@
 This demo aims to give quick overview of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) for palaeobiology analysis.
-Please refer to \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes, namely the \dispRity tutorial that explains the functions in more detail.
+Please refer to \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes, namely the \href{https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf}{\dispRity manual} that explains the functions in more detail.
 
@@
 This type of analysis often starts by ordinating \textbf{morphometric} or \textbf{cladistic} data into a multidimensional object hereafter called the \textbf{morphospace}.
-One might then be interested how the occupancy of the morphospace changes through time by measuring a summary metric of this morphospace through time, called \textbf{disparity}.
-In the morphospace case, disparity might be seen as a value summarising the diversity of morphologies.
+One might be interested in studying how the occupancy of the morphospace changed through time by measuring a summary metric of this morphospace through time, called \textbf{disparity}.
+Disparity can then be seen as a value summarising the diversity of morphologies.
 
-This demo showcases a typical disparity-through-time analyse: we are going to test whether the disparity changed through time in a subset of Eutheria since the last 100 million years.
+This demo showcases a typical disparity-through-time analyse: we are going to test whether the disparity changed through time in a subset of Eutheria (Mammalia) since the last 100 million years.
 
@@
 \subsection{The morphospace}
-In this example, we are going to use a subset of the data from \cite{beckancient2014} that contains an ordinated cladistic matrix and a phylogenetic tree containing both the same 50 taxa along side a table that contains some first and last occurence data for several taxa.
+In this example, we are going to use a subset of the data from \cite{beckancient2014} that contains an ordinated cladistic matrix and a phylogenetic tree containing 50 taxa along side with a table containing some first and last occurence data for several taxa.
 The ordinated matrix will represent our full morphospace, i.e. all the morphologies that ever existed through time (for this dataset).
@@
 data(BeckLee_tree)
-plot(BeckLee_tree, cex=0.8); axisPhylo(root=140)
+plot(BeckLee_tree, cex=0.7); axisPhylo(root=140)
 @
@@
 if(!require(strap)) install.packages("strap")
-geoscalePhylo(BeckLee_tree, cex.tip=0.8, cex.ts=0.6)
+library(strap)
+geoscalePhylo(BeckLee_tree, cex.tip=0.7, cex.ts=0.6)
 @
@@
 \subsection{Splitting the morphospace through time} 
-One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller sub-samples that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
-Basically, the full morphospace in the previous step represents the total amount of morphologies in all time and is likely to be greater than any sub-sample of the morphospace.
+One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time seriess that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
+Basically, the full morphospace in the previous step represents the total amount of morphologies in all time and is likely to be greater than any time series of the morphospace.
 
@@
 ## Creating the vector of time bins ages
-time_bins <- rev(seq(from=0, to=100, by=20))
+time_bins <- rev(seq(from = 0, to = 100, by = 20))
 
@@
 
-The warning message tells us that some taxa were considered as single points in time because we didn't provide temporal range information for all the taxa: they are therefore considered as unique time points without any occurrence range.
-
 The output object is a \dispRity object.
-For details about this object class, please refer to the \href{https://github.com/TGuillerme/dispRity}{the \dispRity manual}.
+For details about this object class, please refer to the \href{https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf}{\dispRity manual}.
 In brief, \dispRity objects are lists of different elements (i.e.
-disparity results, morphospace sub-samples, etc...) that display only a summary of the object when calling the object for avoiding jamming the \R console.
+disparity results, morphospace time seriess, etc...) that display only a summary of the object when calling the object for avoiding jamming the \R console.
 
@@
 \subsection{Bootstrapping the data}
-Once we obtain our different subsamples, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
-The bootstrapping will allow to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of taxa to get rid of eventual sampling problems (i.e.
+Once we obtain our different time series, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
+The bootstrapping will allow to make each series more robust to outliers and the rarefaction will allow us to compare slices with the same number of taxa to get rid of eventual sampling problems (i.e.
 more taxa in one bin than the other).
@@
 <<boot.rare>>=
-## Bootstrapping each sub-sample 100 times
+## Bootstrapping each time series 100 times
 boot_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100)
 
-## Getting the minimum number of rows (=taxa) in the sub-samples
+## Getting the minimum number of rows (=taxa) in the time seriess
 min(unlist(lapply(binned_morphospace$data, nrow)))
 
-## Bootstrapping each sub-sample 100 times and rarefying them 
+## Bootstrapping each time series 100 times and rarefying them 
 rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
@@
 
-We can now calculate the disparity within each sub-sample along side with some confidence intervals generated by the pseudo-replication step above (bootstraps/rarefaction).
+We can now calculate the disparity within each time series along side with some confidence intervals generated by the pseudo-replication step above (bootstraps/rarefaction).
 Disparity can be calculated in many ways and this package allows user to come up with their own disparity metrics.
@@
 
-In this example, we are going to calculate the spread of the data in each sub-sample by calculating the disparity as the sum of the variance of each dimension of the morphospace in each sub-sample using the \dispRity function.
-Thus, in this example, the disparity is defined by the multi-dimensional variance of each sub-sample (i.e. the spread of the taxa within the morphospace).
+In this example, we are going to calculate the spread of the data in each time series by calculating the disparity as the sum of the variance of each dimension of the morphospace in each time series using the \dispRity function.
+Thus, in this example, the disparity is defined by the multi-dimensional variance of each time series (i.e. the spread of the taxa within the morphospace).
 Note that this metric comes with a caveat not solved here since it ignores the covariance among the dimensions of the morphospace.
@@
 summary(rare_disparity)
+@
 
+\noindent As for any \R package, you can refer yourself to each individual function for more details.
+
+<<disparity2, eval = FALSE>>=
 ## See more options on the summary function:
@@
 plot(rare_disparity, type = "continuous", main = "rarefied results")
-
-## See more options on the plot function:
-?plot.dispRity
 @
@@
 what happens in bin Y depends on what happened in bin X) and was pseudo-replicated (i.e.
-each bootstrap draw creates non-independent sub-samples because they are all based on the same sub-sample), we are going to apply a non-parametric mean comparison: the \texttt{wilcox.test}.
+each bootstrap draw creates non-independent time seriess because they are all based on the same time series), we are going to apply a non-parametric mean comparison: the \texttt{wilcox.test}.
 Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis (see \texttt{?p.adjust} for more details).

### 2016-04-21 R/dispRity-package.R
Updated the Travis error (PASS) messages

@@ -1,4 +1,4 @@
-#' Measuring disparity in R
+#' Measuring Disparity in R
 #' 
 #' A modular package for measuring disparity from ordinated matrices. Disparity can be calculated from any ordinated matrix. The package provides a set of implemented metrics to measure the ordinated space and allows users to provide and test their own metrics. The package also provides functions for looking at disparity in a serial way (e.g. time series) as well as visualising the results. Finally, this package provides several basic statistical test for disparity analysis.
 #' 

### 2016-04-21 R/dispRity.R

@@ -70,8 +70,10 @@
 #' ## Calculating disparity using one thread
 #' system.time(dispRity(bootstrapped_data, metric = c(sum, variances)))
 #' ## Bootstrapping a series of matrices using 4 threads
-#' system.time(dispRity(bootstrapped_data, metric = c(sum, variances), parallel = c(4, "SOCK")))
-#' # System time is significantly longer! Using parallel is only an improvement for big datasets.
+#' system.time(dispRity(bootstrapped_data, metric = c(sum, variances),
+#'      parallel = c(4, "SOCK")))
+#' ## System time is significantly longer! Using parallel is only an improvement
+#' ## for big datasets.
 #' }
 #' 
 #' @seealso \code{\link{boot.matrix}} for bootstrapping the data; \code{\link{dispRity.metric}} for details on the implemented metrics and \code{\link{summary.dispRity}} for summarising \code{dispRity} objects.

### 2016-04-21 multiple-t.test.R

@@ -1,0 +1,88 @@
+######
+# Applying this to disparity
+######
+# Now let say we want to calculate the disparity
+
+library(dispRity)
+
+# First let's generate two dummy matrices:
+## Generating a dummy ordinated matrix
+ordinated_matrix <- matrix(data = rnorm(180), nrow = 20, ncol = 9,
+     dimnames = list(letters[1:20]))
+## Creating a list of dummy factors (1 or 2)
+factors <- as.data.frame(matrix(data = c(rep(1,10), rep(2,10)), nrow = 20,
+     ncol = 1, dimnames = list(letters[1:20])))
+## Splitting the dummy ordinated matrix
+custom_series <- cust.series(ordinated_matrix, factors)
+
+matrix1 <- custom_series$data$V1.1
+matrix2 <- custom_series$data$V1.2
+
+# now we can calculate some distribution of disparity like the distances from centroids
+disparity1 <- centroids(matrix1)
+disparity2 <- centroids(matrix2)
+
+# One way to look at disparity can be to look a the median distances from centroids and compare them
+median(disparity1) - median(disparity2)
+# And that gives us our difference.
+# The problem is that it's not really statsy. The otherway to look at it could be to look at the difference of the distributions:
+t.test(disparity1, disparity2)
+
+#~~~~~~~
+# That gives us no difference between our two random disparities (as expected from a random normal distribution!)
+# Now the problem here is that we know we have the TRUE answer (simulations!). In reality, and especially with palaeontological data, we might have issues with the data set such as bad sampling, outliers, etc...
+# One way to solve these issues is to pseudo-replicate (bootstrap) the data so that our results are less reliable on errors.
+#~~~~~~~
+
+# Now looking at the same logic as before, we can bootstrap the matrices
+bootstrapped_matrices <- boot.matrix(custom_series)
+
+matrices1 <- bootstrapped_matrices$data$bootstraps$V1.1[[1]]
+matrices2 <- bootstrapped_matrices$data$bootstraps$V1.2[[1]]
+
+# now we can calculate disparity the same way
+disparity1 <- lapply(matrices1, centroids)
+disparity2 <- lapply(matrices2, centroids)
+
+# One way to look at it would be to calculate the distribution of the median distances (i.e. 1000 medians of disparity1 vs 1000 medians of disparity2)
+t.test(unlist(lapply(disparity1, median)), unlist(lapply(disparity2, median)))
+
[…]

### 2016-05-25 doc/dispRity-metrics.Rmd
metrics vignette now available in Rmd

---
title: "dispRity metrics"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: References.bib
bst: sysbio.bst
vignette: >
  %\VignetteIndexEntry{dispRity metric}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette details how the disparity metrics are implemented in the `dispRity` package (v.0.2).

Disparity metrics
=================

There are many ways for measuring disparity. In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. MDS, PCA, PCO, PCoA).
For example, one can look at ellipsoid hyper-volume of the ordinated space [@DonohueDim], the sum and the product of the ranges and variances of the ordinated space dimensions [@Wills1994] or the median position of the elements in the ordinated space relative to its centroid [@GuillermeSTD].
Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks or some being more generalist.

Because of this great diversity of metric, the package `dispRity` has not a unique way to measure disparity but rather proposes to facilitate users to define their own disparity metric that will best suit to their particular analysis.
In fact, the core function of the package, `dispRity`, allows the user to define any metric in the `metric` argument.
However, because of the infrastructure of the package (especially the one of the `dispRity` objects – see the [`dispRity` manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf)), the `metric` argument has to follow certain rules:

1.  it must be composed of one to three `function` objects;

2.  these functions can only have as input a `matrix` or a `vector`;

3.  each of these functions must be of one of the three levels described below;

4.  at least one of the functions must be a level 1 function (see below).

The function levels
===================

The metric functions levels determine the dimensional level of decomposition of the input `matrix`.
In other words, each level designate the dimensions of the output: either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimensions.

![Illustration of the different levels of output of a function with an input `matrix`](dispRity_fun.pdf)

A level 1 function
------------------

A level 1 function for example, will decompose a `matrix` or a `vector` into a single value:

```{r}
# First, let's create a dummy matrix
dummy_matrix <- matrix(rnorm(12), 4, 3)

# Any summary metric such as mean or median are good examples of level 1
# functions as they reduce the matrix to a single dimension (i.e. one value):
mean(dummy_matrix)
median(dummy_matrix)
```

A level 2 function
------------------

A level 2 function will decompose a `matrix` into a `vector`.
Several level 2 functions are implemented in `dispRity`(see `?dispRity.metric`) such as the `variances` function that calculates the variance of each dimension of the ordinated matrix.

```{r}
# We can define a function that outputs the products of each row of the matrix:
prod.rows <- function(matrix) apply(matrix, 1, prod)

# This function is a level 2 function as it reduces the matrix into a two
# dimensional element (a vector of values):
prod.rows(dummy_matrix)
```

A level 3 function
------------------

Finally a level 3 function will transform the `matrix` into another `matrix`.
Note that the dimension of the output `matrix` don't need to match the the input `matrix`:

```{r}
# The var functions computes the variance/covariance matrix which will be a
# three by three matrix
var(dummy_matrix)
```

`make.metric`
=============

Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the euclidean distance between each elements and the ordinated space centroid.
The `make.metric` function implemented in `dispRity`is designed to help testing and finding the level of the functions.
This function tests:

1.  if your function can deal with a `matrix` or a `vector` as an input;

2.  which is your function's level according to it's output (level 1, 2 or 3, see above);

3.  whether the function can properly be implemented in the `dispRity` function (the function is fed into a `lapply` loop).

For example, let's see if the functions described above are the right levels:

```{r}
# First we need to load the package
library(dispRity)

# Which level is the mean function? And is able to be used in the dispRity?
make.metric(mean)

# Same questions from the prod.rows function:
make.metric(prod.rows)

# Same questions for the var function:
make.metric(var)
```

metric argument in `dispRity`
===========================================

Using this metric structure, on can easily use any disparity metric in the `dispRity`function as follows:

```{r}
# First we need to load the data
data(BeckLee_mat50)

# Measuring disparity as the standard deviation of all the value of the
# ordinated matrix (level 1 function).
summary(dispRity(BeckLee_mat50, metric = sd))

# Measuring disparity as the standard deviation of the variance of each axis of
# the ordinated matrix (level 1 and level 2 functions).
summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))

# Measuring disparity as the standard deviation of the variance of each axis of
# the variance covariance matrix (level 1, level 2 and level 3 functions).
summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
```

Note that the order of each function in the metric argument does not matter, the `dispRity`function will automatically detect the function levels (using `make.metric`) and apply them to the data in decreasing order (level 3 > level 2 > level 1).

```{r}
# Disparity as the standard deviation of the variance of each axis of the
# variance covariance matrix:
disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)),
                      round = 10)

# Same but using a different functions order for the metric argument
disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
                      round = 10)

# Both ways output the same disparity values:
disparity1$observed == disparity2$observed # is TRUE
```

What level is the disparity?
------------------

In these examples, we considered disparity to be a single value.
For example, in the previous one, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
This results in single value to be calculated.

In some cases however, it might also be possible to calculate disparity as a distribution.
For example, one of the metrics commonly used in palaeobiology is the sum of variances of each dimensions (`metric = c(sum, variances)`) [@Wills1994].
However, it might be interesting to look directly at the distribution of the variances itself.

```{r}
# A one dimensional disparity
sum_variances1 <- summary(dispRity(BeckLee_mat50, metric = c(sum, variances)))

# A two dimensional disparity
all_variances <- summary(dispRity(BeckLee_mat50, metric = variances))

# Transforming the two dimensional one in a one dimensional one
sum_variances2 <- summary(dispRity(BeckLee_mat50, metric = sum))

# Both ways output the same disparity values:
sum_variances1$observed == sum_variances2$observed
```

# References

### 2016-05-25 doc/dispRity-ecology-demo.Rmd
Red manuals INIT

---
title: "dispRity ecology demo"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: References.bib
bst: sysbio.bst
vignette: >
  %\VignetteIndexEntry{dispRity ecology demo}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This demo aims to give quick overview of the <span>`dispRity`
</span>package (v.0.2) for ecological analysis. Please refer to [GitHub
page:
github.com/TGuillerme/dispRity](https://github.com/TGuillerme/dispRity)
for other vignettes, namely the `dispRity` tutorial that
explains the functions in more detail.

To keep it short, this package allows the use of all of the dimensions
of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis
rather than just a sub-set of dimensions. For example, one might want to
know whether a specific kind of water treatment at certain water depth
alters invertebrate communities and composition in natural habitats.

Before starting
===============

Installing `dispRity` 
-----------------------------------

You can install this package easily if you are using the latest version
of <span>`R` </span>and `devtools`.

` `\
`   `\

Data
----

For this example with ecological data we are going to use data from
McClean (unpubl.). This data is the ordination of a distance matrix
based on nutrient enrichment and depth and freshwater benthic
invertebrates.

\

\
\
\
\
\
\
\
\
`  `\
\
\
\
`  `\
`  `

A classical two dimensional approach
====================================

A classical way to represent this ordinated data will be to use two
dimensional plots.

\
\
`   `\
`   `\
\
`   `\
`   `\
\
`   `\
`   `\
\
\
`  `\
\
`          `\
`         `

![image](figure/minimal-PCAplot-1){width=".8\linewidth"}

This shows the distribution of the experimental plots along the two
first axis of variance of the ordinated distance matrix (i.e. the first
two dimensions). At a first glance, it seems difficult to see a clear
effect from the treatments (blue or red) or the depth (rounds or
triangles) since the different experimental plots with the same
parameters don’t seem to cluster. However, the problem, is that this
plot ignores the 18 other dimensions of the ordination! Additionally,
these two represented dimensions do not represent a biological reality
*per se*; i.e. the values on the first dimension do not represent some
continuous traits (e.g. depth) but just the ordinations of correlations
between the data and some factors.

Therefore, one might want to approach this problem without getting stuck
in only two dimensions and consider the whole dataset as a
*n*-dimensional object.

A multidimensional approach with `dispRity` 
=========================================================

Splitting the data
------------------

The first step in such analysis will be to create different series that
are two sub-samples of the ordinated space (i.e. the *n*-dimensional
object). Each of these series contain a certain number of elements (i.e.
the 40 experimental field plots) that have some attributes in common. In
our example, we are going to group the elements according to their depth
and treatment.

\
`       `\
`            `\
`   `\

    ##     Treat Depth
    ## 1       a     1
    ## 1.1     a     2
    ## 2       b     1
    ## 2.1     b     2
    ## 3       a     1
    ## 3.1     a     1

Second, let’s split the data according to these factors to create the
series of the ordinated space by using the `cust.series` function:

\
`  `\
\
\

    ## [1] "dispRity"

\
\

    ## 4 custom series for 40 elements 
    ## Series:
    ## Treat.a, Treat.b, Depth.1, Depth.2.

For more details on the `dispRity` objects, see the
[<span>`dispRity`
</span>manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf).
Basically the idea is to avoid jamming the <span>`R` </span>console such
as when using:

\
\
\
\
`   `

Calculating disparity
---------------------

Disparity can be calculated in many ways, and therefore, the
`dispRity` function allows to measure disparity *as defined
by the user*. For more details on disparity, see the [other
vignettes](https://github.com/TGuillerme/dispRity).

In this example, we are going to define disparity as the median distance
between the different experimental plots and the centroid of the
ordinated space. High values of disparity will indicate a general high
dispersal from this centroid (i.e. on average, the experimental plots
are far apart in the ordinated space). We can define the metrics easily
in the `dispRity` function by feeding them to the `metric`
argument. Here we are going to feed the functions `stats::median` and
`dispRity::centroids` which calculates distances between elements and
centroid.

\
`     `\
\
\
\
\

    ## Disparity measurements across 4 series for 40 elements 
    ## Series:
    ## Treat.a, Treat.b, Depth.1, Depth.2.
    ## Disparity calculated as: c(median, centroids) for 20 dimensions.
    ## Data was split using custom method.

    ##    series  n observed
    ## 1 Treat.a 21    82.38
    ## 2 Treat.b 19    94.68
    ## 3 Depth.1 23    83.98
    ## 4 Depth.2 17    89.72

Bootstrapping the data
----------------------

One might also want to bootstrap the data to test the robustness of the
measured disparity to outliers. Also, as we can see, each series has
different numbers of elements. It might be interesting to rarefy the
data as well to have only series with the same number of elements. Both
steps are easily doable through the `boot.matrix` function.

\
`     `\
\
\
`     `\
`      `\
\
\
\

    ## Bootstrapped ordinated matrix with 40 elements 
    ## Series:
    ## Treat.a, Treat.b, Depth.1, Depth.2.
    ## Data was split using custom method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.

    ## Bootstrapped ordinated matrix with 40 elements 
    ## Series:
    ## Treat.a, Treat.b, Depth.1, Depth.2.
    ## Data was split using custom method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.
    ## Data was rarefied with a maximum of 17 elements

We can now rerun a more robust disparity analysis using the bootstrapped
data:

\
`     `\
\
\
`     `

Summarising the data
--------------------

We can now summarise the data using various options such as the
confidence intervals levels and the central tendency.

\

    ##    series  n observed  mean  2.5%   25%   75% 97.5%
    ## 1 Treat.a 21    82.38  79.3 64.83 74.45  84.1  95.4
    ## 2 Treat.b 19    94.68 100.2 79.63 90.99 106.8 133.6
    ## 3 Depth.1 23    83.98  82.4 69.04 78.32  85.9  92.7
    ## 4 Depth.2 17    89.72 100.9 69.58 85.03 110.6 163.4

\
\
`     `

    ##    series  n observed median    5%   95%
    ## 1 Treat.a 21    82.38  78.50 68.34  91.2
    ## 2 Treat.b 19    94.68  97.02 83.17 128.3
    ## 3 Depth.1 23    83.98  83.21 72.79  91.3
    ## 4 Depth.2 17    89.72  93.04 75.77 149.4

\

    ##    series  n observed  mean  2.5%   25%   75% 97.5%
    ## 1 Treat.a 17       NA  79.4 67.47 74.52  83.1  96.7
    ## 2 Treat.b 17       NA  99.8 79.24 88.04 106.9 147.4
    ## 3 Depth.1 17       NA  82.4 69.41 77.08  86.3 106.1
    ## 4 Depth.2 17    89.72 102.7 77.36 86.44 113.7 159.0

\

We can also plot the results and have a look at the effect of the number
of experimental plots:

\
`          `\
\
\
`   `\
\
\
`   `

![image](figure/minimal-rare_plots-1){width="1\linewidth"}

As we can see, there seems to be no strong effect of the number of
experimental plots in each series (i.e. the rarefied plot is really
similar to the bootstrapped plot) which is a good thing!

Testing the hypothesis
----------------------

Finally, we can test our hypothesis (whether the sort of water treatment
at certain depth alters invertebrate communities and composition in
natural habitats) by using the `test.dispRity` function.

\
`     `

    ##              Df Sum Sq Mean Sq F value Pr(>F)    
    ## series        3  39184   13061   61.17 <2e-16 ***
    ## Residuals   396  84553     214                   
    ## ---
    ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

\
`     `

    ##              Df Sum Sq Mean Sq F value Pr(>F)    
    ## series        3  42229   14076   63.85 <2e-16 ***
    ## Residuals   396  87301     220                   
    ## ---
    ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

\
\
`  `\
`       `

    ##                      W      p.value
    ## Treat.a - Treat.b  695 7.170082e-26
    ## Depth.1 - Depth.2 2064 7.360379e-13

\
`  `\
`       `

    ##                      W      p.value
    ## Treat.a - Treat.b  880 7.845518e-24
    ## Depth.1 - Depth.2 1599 9.672040e-17

As we can see, there is strong support for an effect of the treatment
and the depth on the median distance between each experimental plots and
the centroid of the ordinated space. In other words, it seems that with
the second treatment and the second level of depth, the invertebrate
communities and composition were further apart (i.e. more dispersed in
the ordinated space) than with the first treatment and the first level
of depth.

### 2016-05-25 doc/dispRity-manual.Rmd
Red manuals INIT

---
title: "dispRity manual"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: References.bib
bst: sysbio.bst
vignette: >
  %\VignetteIndexEntry{dispRity manual}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

`dispRity` is a package for calculating disparity in `R`.
It allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA)
into single values.

Before starting
===============

Glossary
--------

Because this package is aimed to be multidisciplinary, many names or
terms used in this tutorial might be non-familiar to certain fields.
Here is a list of what are the exact meaning of these term:

-   **Ordinated space**: it designates the mathematical multidimensional
    object studied here. In morphometrics, this one is often referred as
    being the morphospace. However it can also be the the cladisto-space
    for cladistic data or the eco-space in ecology, etc. In practice,
    this term designates an ordinated matrix where the columns represent
    the dimensions of the ordinated space (often – but not necessarily –
    $>$ 3!) and the rows represent the elements within this space.

-   **Elements**: it designates the rows of the ordinated space,
    elements can be either taxa, field sites, countries, etc...

-   **Dimensions**: it designates the columns of the ordinated space.
    The dimensions can also be referred to as axis or
    principal components.

-   **Series**: it designates sub-samples of the ordinated space.
    Basically a series contain the same number of dimensions as the
    morphospace but may contain a smaller number of elements. For
    example, if our ordinated space is composed of birds and
    mammals (i.e. the elements) and 50 dimensions, we can create two
    series of just mammals or birds as elements (but with the same
    50 dimensions) to look at the difference in disparity between the
    two clades.

Installation
------------

You can install this package easily if you use the latest version of
<span>`R` </span>and `devtools`.

` `\
`   `\

Note that we use the `release` branch here which is version 0.2. For the
piping-hot (but potentially full of bugs) version, you can change the
argument `ref = release` to `ref = master`. This package depends mainly
on the `ape` package and uses functions of several other (`ade4`,
`geometry`, `grDevices`, `hypervolume`, `paleotree` and `snow`).

Data
----

In this tutorial we are going to use a subset of the ordinated cladistic
data from @beckancient2014 that contains 50 taxa (elements) ordinated
using their cladistic distance (i.e. the distance between their discrete
morphological characters). Note that this data is more oriented towards
palaebiology analysis but that it can apply to other disciplines. Please
refer to the [GitHub page:
github.com/TGuillerme/dispRity](https://github.com/TGuillerme/dispRity)
for other vignettes covering some specific example.

\

\
\
\
\
\

    ## [1] 50 48

    ##                    [,1]          [,2]        [,3]       [,4]      [,5]
    ## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
    ## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
    ## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
    ## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
    ## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
    ## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273

\
\

    ## [1] 99 97

` `

    ##                  [,1]       [,2]       [,3]       [,4]        [,5]
    ## Cimolestes -0.6082437 -0.0323683 0.08458885 -0.4338448 -0.30536875
    ## Maelestes  -0.5730206 -0.2840361 0.01308847 -0.1258848  0.06123611

` `

    ##            [,1]      [,2]       [,3]       [,4]       [,5]
    ## n48 -0.05529018 0.4799330 0.04118477 0.04944912 -0.3558830
    ## n49 -0.13067785 0.4478168 0.11956268 0.13800340 -0.3222785

\
\

    ##             FAD  LAD
    ## Adapis     37.2 36.8
    ## Asioryctes 83.6 72.1
    ## Leptictis  33.9 33.3
    ## Miacis     49.0 46.7
    ## Mimotona   61.6 59.2
    ## Notharctus 50.2 47.0

\
\
`         `

![image](figure/minimal-data-1){width=".8\linewidth"}

A quick go through
------------------

Here is a really crude and quick analysis to go through the package,
showing some of its features. Note that all these features will be
discussed in more details below.

\
`     `\
`           `\
`     `\
\
\
`   `\
\
\
`     `\
\
\

    ##   series  n observed  mean  2.5%   25%   75% 97.5%
    ## 1    120  5    1.494 1.200 0.756 0.854 1.484 1.550
    ## 2    105 11    1.725 1.646 1.481 1.614 1.686 1.770
    ## 3     90 18    1.820 1.775 1.681 1.747 1.806 1.849
    ## 4     75 19    1.923 1.814 1.681 1.778 1.865 1.901
    ## 5     60 20    1.946 1.859 1.751 1.833 1.894 1.925
    ## 6     45 14    1.930 1.849 1.742 1.827 1.883 1.951
    ## 7     30 10    1.910 1.796 1.570 1.774 1.866 1.935
    ## 8     15 10    1.910 1.801 1.594 1.782 1.853 1.914
    ## 9      0 10    1.910 1.815 1.588 1.788 1.865 1.929

\
\
\
`     `

    ##              Df Sum Sq Mean Sq F value Pr(>F)    
    ## series        8  34.47   4.309   279.5 <2e-16 ***
    ## Residuals   891  13.74   0.015                   
    ## ---
    ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

![image](figure/minimal-Quick_go-1){width=".5\linewidth"}

Package specificities
=====================

The `dispRity` objects
-----------------------------------

Disparity analysis can involve a lot of shuffling around with many
matrices (especially when bootstrapping the data) which can be
impractical to visualise and will quickly jam your <span>`R`
</span>console. For example, we can have a look at the structure of the
object created in the quick example:

\

Therefore this package proposes a specific class of object called
`dispRity` objects. These objects allow to easily use S3
method functions such as `summary.dispRity` (just called as `summary`;
see section \[summary\]) or `plot.dispRity` (just called as `plot`; see
section \[plot\]). But also, this allows to use the S3 method for
printing `dispRity` objects via `print.dispRity` that
allows to summarise the content of the objects similar to the `phylo`
class objects (see `ape::print.phylo`).

\

    ## [1] "dispRity"

\

    ## [1] "data"      "disparity" "elements"  "series"    "call"

\

    ## Disparity measurements across 9 series for 99 elements 
    ## Series:
    ## 120, 105, 90, 75, 60, 45 ...
    ## Disparity calculated as: c(median, centroids) for 97 dimensions.
    ## Data was split using continuous method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.

Note however, that it is always possible to recall the full object using
the argument `all=TRUE`:

\
`   `\

Finally, some utility functions such as `get.dispRity` or
`extract.dispRity` allows to access to some specific content of the
object:

\
`     `\

    ## Disparity measurements across 2 series for 24 elements 
    ## Series:
    ## 120, 75.
    ## Disparity calculated as: c(median, centroids) for 97 dimensions.
    ## Data was split using continuous method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.

\

    ##      120      105       90       75       60       45       30       15        0 
    ## 1.493570 1.725280 1.820292 1.922531 1.945952 1.930306 1.910203 1.910203 1.910203

\
`   `\
`      `

    ## List of 1
    ##  $ 120: num [1:100] 1.52 0.786 1.555 1.288 0.891 ...

Modular functions
-----------------

This package aims to be a modular package where users can personalise
some aspects of the package fairly easily. The modular aspect allows the
package users to use several implemented tools to help them creating
personalised functions. For example, the <span>`dispRity`
</span>function intake a `metric` argument designating how disparity
should be calculated. This argument can take some already implemented
functions such as `mean` but also some completely new ones such as the
sum divided by length:

\
`  `

    ##   series  n observed
    ## 1    120  5   -0.005
    ## 2    105 11   -0.010
    ## 3     90 18   -0.010
    ## 4     75 19   -0.001
    ## 5     60 20    0.005
    ## 6     45 14    0.013
    ## 7     30 10    0.017
    ## 8     15 10    0.017
    ## 9      0 10    0.017

\
`   `\
\
\
`  `

    ##   series  n observed
    ## 1    120  5   -0.005
    ## 2    105 11   -0.010
    ## 3     90 18   -0.010
    ## 4     75 19   -0.001
    ## 5     60 20    0.005
    ## 6     45 14    0.013
    ## 7     30 10    0.017
    ## 8     15 10    0.017
    ## 9      0 10    0.017

For more information, please refer to the [GitHub page:
github.com/TGuillerme/dispRity](https://github.com/TGuillerme/dispRity)
for the vignette about the metric implementation in <span>`dispRity`
</span>.

Functions
=========

In the following section, different functionalities are detailed. The
functions are presented in a “classic” pipelined study fashion for
convenience (i.e. splitting the data, bootstrapping it, calculating
disparity, testing differences). Note however that most of these
functions can be used in any order and that some steps can of course be
skipped.

One of the first steps in disparity analysis is to split the ordinated
space into sub-samples of this space (hereafter *series*). These series
correspond to regions of the ordinated space that we want to compare to
each other.

`cust.series`
-------------

The `cust.series` function allows to create these series according to
factors determined by the user. In this example, we can split the matrix
based on the phylogeny and separate to different groups of taxa: the
crown mammals (all the living mammals and their direct ancestor) and the
stem mammals (all the mammals that have no direct offspring).

\
\
`     `\
`    `\
\
`  `\
`         `\
`          `\
\
\
\
`  `\
\
\

    ## [1] "dispRity"

    ## 2 custom series for 50 elements 
    ## Series:
    ## Group.crown, Group.stem.

\

    ## [1] "data"     "elements" "series"

\

    ## List of 2
    ##  $ Group.crown: num [1:30, 1:48] 0.3079 0.6531 0.5089 -0.1652 -0.0419 ...
    ##   ..- attr(*, "dimnames")=List of 2
    ##   .. ..$ : chr [1:30] "Dasypodidae" "Bradypus" "Myrmecophagidae" "Todralestes" ...
    ##   .. ..$ : NULL
    ##  $ Group.stem : num [1:20, 1:48] -0.739 -0.68 -0.511 -0.477 -0.473 ...
    ##   ..- attr(*, "dimnames")=List of 2
    ##   .. ..$ : chr [1:20] "Daulestes" "Bulaklestes" "Uchkudukodon" "Kennalestes" ...
    ##   .. ..$ : NULL

\

    ##  chr [1:50] "Cimolestes" "Maelestes" "Batodon" "Bulaklestes" "Daulestes" ...

\
\

    ## [1] "custom"      "Group.crown" "Group.stem"

Note that here we separating the taxa into two exclusive groups (i.e.
crown mammals can be crown only, same for stem mammals). It is however
possible to add more factors with overlapping groups:

\
`  `\
`         `\
`        `\
`           `\
\
\

    ## 3 custom series for 50 elements 
    ## Series:
    ## Group.crown, Group.stem, Class.Mammalia.

`time.series`
-------------

Another way to split the ordinated space (maybe more relevant to
palaeobiologists) is to do it according to time. The `time.series`
function allows to create series that contain all the elements present
at specific points or during specific periods in time. This functions
needs as input an ordinated space and a matching phylogenetic tree. Two
types of time series can be performed by using the `method` option:

1.  discrete time series (or time-binning) using `method = discrete`;

2.  continuous time series (or time-slicing) using
    `method = continuous`.

For the time-slicing method details see @GuillermeSTD. For both methods,
the function intakes the `time` argument which can be a vector of
`numeric` values for:

1.  defining the boundaries of the time bins (when `method = discrete`);

2.  defining the time slices (when `method = continuous`).

Otherwise, the `time` argument can be set as a single `numeric` value
for automatically generating a given number of equidistant
time-bins/slices. Additionally, it is also possible to input a data
frame containing the first and last occurrence data (`FAD/LAD`) for taxa
that span over a longer time than the tips/nodes age.

Here is an example for `method = discrete`:

\
`      `\
`         `

    ## 3 discrete series for 50 elements 
    ## Series:
    ## 120-80, 80-40, 40-0.

In this example, the taxa where split inside each timebin according to
their age. Note however that some taxa might span between some time bins
and should be included in more than one. This is possible by providing a
table containing the first and last occurrence data:

\
\
`     `\
`              `\

    ## 3 discrete series for 50 elements 
    ## Series:
    ## 120-80, 80-40, 40-0.

This generated indeed a list of 3 sub matrices. Note that we can also
generate equivalent results by just telling the function that we want
three time-bins (series) as follow:

\
`      `\
`      `

    ## 3 discrete series for 50 elements 
    ## Series:
    ## 133.51104-89.00736, 89.00736-44.50368, 44.50368-0.

We now have three time bins of 44.50368 million years each.

When using this method, the oldest boundary of the first bin (or the
first slice, see below) is automatically generated as being the root age
+ 1% of the tree length as long as at least three elements are present
at that point in time. The algorithm adds an extra 1% tree length until
reaching the required minimum of three elements. It is also possible to
include nodes in each bin by using `inc.nodes = TRUE` and providing a
matrix that contains the ordinated distance between tips *and* nodes.

For the time-slicing method (`method = continuous`), the idea is really
similar. This option intakes a matrix that contains the ordinated
distance between taxa *and* nodes and an assumed evolutionary model via
the `model` argument:

1.  `acctran` where the data chosen on each time slice is always the one
    of the offspring

2.  `deltran` where the data chosen on each time slice is always the one
    of the descendant

3.  `punctuated` where the data chosen on each time slice is randomly
    chosen between the offspring or the descendant

4.  `gradual` where the data chosen on each time slice is either the
    offspring or the descendant depending on branch length

\
\
`     `\
`               `\
`     `\

    ## 4 continuous series for 99 elements 
    ## Series:
    ## 120, 80, 40, 0.

\
\
`      `\
`         `

    ## 4 continuous series for 99 elements 
    ## Series:
    ## 133.51104, 89.00736, 44.50368, 0.

`boot.matrix` {#rare}
-------------

Once we obtain our different series, we might want to bootstrap and
rarefy it (i.e. pseudo-replicating the data). The bootstrap will allow
us to make each series more robust to outliers and the rarefaction will
allow us to compare slices with the same number of elements to get rid
of eventual sampling problems. The `boot.matrix` allows to bootstraps
and rarefy ordinated matrices in a fast and easy way. The default
options will bootstrap the matrix 1000 times without rarefaction. The
number of bootstrap pseudo-replicates can be defined using the
`bootstraps` option (see below).

\
` `

    ## Bootstrapped ordinated matrix with 50 elements 
    ## 1
    ## Data was bootstrapped 1000 times, using the full bootstrap method.

\
\

This function allows to control the bootstrap algorithm through the
`boot.type` argument. Currently two algorithms are implemented:

1.  `full` where the bootstrapping is entirely stochastic (all the data
    is bootstrapped)

2.  `single` where only one random elements is replaced by one other
    random elements each pseudo-replication

This function also allows to rarefy the data using the `rarefaction`
argument. The default argument is `FALSE` but it can be set to `TRUE` to
fully rarefy the data (i.e. remove $n$ elements for the number of
pseudo-replicates, where $n$ varies from the maximum number of elements
present in the dataset to a minimum of 3 elements). It can also be set
to a fix `numeric` value (or a set of `numeric` values). Finally, one
last argument, `rm.last.axis` allows to remove a certain amount of
dimensions (or axis) for the ordinated space. This can be logical
argument where `FALSE` (default) will not remove any dimension and
`TRUE` will remove the last dimensions that bear up to 5% of the total
ordinated space’s variance.

\
`   `

    ## Bootstrapped ordinated matrix with 50 elements 
    ## 1
    ## Data was bootstrapped 1000 times, using the single bootstrap method.

\
`      `

    ## Bootstrapped ordinated matrix with 50 elements 
    ## 1
    ## Data was bootstrapped 20 times, using the full bootstrap method.
    ## Data was fully rarefied (down to 3 elements).

\
`      `

    ## Bootstrapped ordinated matrix with 50 elements 
    ## 1
    ## Data was bootstrapped 20 times, using the full bootstrap method.
    ## Data was rarefied with a maximum of 6, 7, 8 and 3 elements

\
`   `

    ## Bootstrapped ordinated matrix with 50 elements 
    ## 1
    ## Data was bootstrapped 1000 times, using the full bootstrap method. 
    ## The 6 last axis were removed from the original ordinated data.

<span>`## Or with a fix value (50%)`</span>\
`   `

    ## Bootstrapped ordinated matrix with 50 elements 
    ## 1
    ## Data was bootstrapped 1000 times, using the full bootstrap method. 
    ## The 35 last axis were removed from the original ordinated data.

Of course, one could be interested in directly supplying the sub-sampled
matrices generated above directly to this function. In fact, it can also
deal with a list of matrices or with a `dispRity` object
output from the `cust.series` or `time.series` functions.

\
`        `\
\
\
`     `\
`     `\
\
\

    ## Bootstrapped ordinated matrix with 50 elements 
    ## Series:
    ## Group.crown, Group.stem.
    ## Data was split using custom method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.
    ## Data was fully rarefied (down to 3 elements).

    ## Bootstrapped ordinated matrix with 50 elements 
    ## Series:
    ## 120-80, 80-40, 40-0.
    ## Data was split using discrete method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.

    ## Bootstrapped ordinated matrix with 99 elements 
    ## Series:
    ## 120, 80, 40, 0.
    ## Data was split using continuous method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.

`dispRity` {#disparity}
----------

This function is a modular function that allows to simply (and quickly!)
calculate disparity from a matrix.

Because disparity can be measured in many ways, this function is a tool
to measure disparity *as defined by the user*. In fact, the
`dispRity` function intakes two main arguments: the data
and the disparity metric. The disparity metric is a function or a set of
functions that summarises the ordinated matrix to a single value that
represents, in our example, the diversity of morphologies.

The `dispRity` algorithm decomposes the metrics functions
into three levels that correspond to the dimensions of the output of
each metric function. For more details on this algorithm please refer to
the [metric vignette](https://github.com/TGuillerme/dispRity).

In practice, the `dispRity` function intakes one or more
functions as a definition of disparity. Several of these functions will
be already implemented in other packages (such as `stats::median`,
`base::sum`, etc.); some others are implemented in this package (listed
in `?dispRity.metric`) and finally some others will be defined by the
users. The `make.metric` function is designed to help users create and
test their own disparity metric functions. In practice, the use of these
metrics in the `dispRity` function, is pretty easy:

`  `

    ## Disparity measurements across 1 series for 50 elements 
    ## 1
    ## Disparity calculated as: mean for 48 dimensions.

\
\
\
\
`   `

    ## Disparity measurements across 1 series for 50 elements 
    ## 1
    ## Disparity calculated as: c(sum, variances) for 48 dimensions.

\
\
\
`   `

    ## Disparity measurements across 1 series for 50 elements 
    ## 1
    ## Disparity calculated as: c(prod, centroids) for 48 dimensions.

\
\
\
\
`   `\
`   `

    ## Disparity measurements across 1 series for 50 elements 
    ## 1
    ## Disparity calculated as: c(total.range, centroids) for 48 dimensions.

\
\
\
\
`     `\
\
\
\
\
\
`     `\
`  `

    ## [1] TRUE

Note that these functions do not directly output the disparity values
but only the summary of the `dispRity` objects. To display
the results, see the section \[summary\].

In these examples we used only simple ordinated matrix but of course, it
might be more interesting to directly use the <span>`dispRity`
</span>objects we generated in the steps above. Thus we can calculate
the bootstrapped and rarefied disparity in each sub-sample of the
morphospace. In this example we are going to define disparity as being
the median distance between each element and each dimensions centroids.

\
`     `\
\
\
`     `\
`     `

Note that the computational time was longer for the rarefied data
(`crown_stemBS`) since it had to calculate the disparity on all the
rarefied matrix, each bootstrapped 100 times. For both the `boot.matrix`
and the `dispRity` functions, it is possible to increase
computational speed in some cases using the `parallel` option (see each
function’s manuals).

`summary`
---------

This function is a S3 function (`summary.dispRity`) allowing to
summarize the content of `dispRity` objects that contain
disparity calculations. This function intakes a <span>`dispRity`
</span>object plus various options namely the `quantile` values for the
confidence intervals levels; the `cent.tend` for the central tendency to
use for summarising the results and two *visual* options which are
whether to recall the `dispRity` options and how much
digits are wanted in the results.

    ##   series  n observed  mean  2.5%   25%   75% 97.5%
    ## 1 120-80  8    1.203 1.103 0.895 1.057 1.178 1.236
    ## 2  80-40 27    1.344 1.322 1.275 1.306 1.339 1.366
    ## 3   40-0 16    1.353 1.323 1.265 1.304 1.346 1.368

\
\
\
\
\
`        `

    ##   series  n observed  sd  6% 94%
    ## 1 120-80  8      1.2 0.1 1.0 1.2
    ## 2  80-40 27      1.3 0.0 1.3 1.4
    ## 3   40-0 16      1.4 0.0 1.3 1.4

\
\
`   `

    ## Disparity calculated as: c(median, centroids) for 48 dimensions.
    ## Data was split using discrete method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.
    ##   series  n observed  mean  2.5%   25%   75% 97.5%
    ## 1 120-80  8    1.203 1.103 0.895 1.057 1.178 1.236
    ## 2  80-40 27    1.344 1.322 1.275 1.306 1.339 1.366
    ## 3   40-0 16    1.353 1.323 1.265 1.304 1.346 1.368

\
\

    ##        series  n observed  mean  2.5%   25%   75% 97.5%
    ## 41 Group.stem 15       NA 1.229 1.146 1.210 1.253 1.282
    ## 42 Group.stem 16       NA 1.232 1.154 1.217 1.250 1.279
    ## 43 Group.stem 17       NA 1.233 1.174 1.218 1.251 1.281
    ## 44 Group.stem 18       NA 1.236 1.175 1.219 1.255 1.282
    ## 45 Group.stem 19       NA 1.239 1.173 1.226 1.256 1.286
    ## 46 Group.stem 20     1.27 1.240 1.185 1.225 1.258 1.280

`plot`
------

An alternative way to summarise the data is to plot the results using
the S3 method `plot.dispRity`. This function intakes the same options as
`summary.dispRity` along side with various graphical options described
in the function manual (see `?plot.dispRity`).

Two main categories of plots can be displayed: continuous plots with the
argument `type = continuous` (see below) and discrete plots (default is
`type = box`. Note that this argument can be left missing. In this case,
the algorithm will automatically detect the type of series from the
`dispRity` object.

The discrete plot arguments can be either:

-   `type = box` for plotting boxplots.

-   `type = polygon` for plotting boxes that represent the different
    levels of confidence intervals (with a dot for the
    central tendency).

-   `type = lines` which is similar to `type = polygon` but plots lines
    rather than boxes for the confidence intervals levels.

It is also possible to display the number of elements per series (has a
horizontal doted line) using the option `elements = TRUE`. Additionally,
when the data is rarefied (see section \[rare\]), one can also indicate
which level of rarefaction to display (i.e. display the results for a
certain number of elements) by using the `rarefaction` argument.

\
`          `\
\
\
`   `\
\
\
\
`      `\
\
`         `

![image](figure/minimal-plot1-1){width="1\linewidth"}

It is also possible to easily look at the behaviour of the disparity
metric with the rarefied data (i.e. how does the disparity score change
when the number of elements goes down) by using the `rarefaction = plot`
argument:

\
`  `\
\
`      `

![image](figure/minimal-plot2-1){width="0.5\linewidth"}

The different dashed lines represent the different confidence intervals
around each rarefaction curve. We can see that the disparity metric
becomes stable after more than 10 elements.

It is also possible, when using the `type = lines` argument, to plot
lines rather than boxes. This option can be useful when displaying many
series in a same window. Additionally, it is possible to plot the
observed data when existing (i.e. not when the data is rarefied) by
using the option `observed = TRUE`.

\
`  `\
\
\
`      `

![image](figure/minimal-plot3-1){width=".8\linewidth"}

The `type = continuous` argument plots superimposed continuous (along
the x axis) polygons that represent the confidence intervals and a
continuous line that represents the central tendency of the data. Note
that, as a S3 method function, `plot.dispRity` can take the classic
graphical options from the `plot` function or even to add other plots to
the previous ones.

\
`  `\
\
\
`   `\

![image](figure/minimal-plot4-1){width=".5\linewidth"}

I encourage you to play with the graphical options to have some prettier
results. Note that most of the options from `plot` can be passed to
`plot.dispRity` via `...`.

\
`  `\
\
`     `\
`           `\
`           `\
`         `\
`         `

![image](figure/minimal-plot5-1){width=".8\linewidth"}

Finally, it is possible to add multiple plots on top of each other:

\
`  `\
\
\
`   `\
\
\
`  `\
`      `

![image](figure/minimal-plot5_1-1){width=".5\linewidth"}

`test.dispRity`
---------------

Finally, the `dispRity` package allows to apply some tests
to the data in order to properly test hypothesis. The function
`test.dispRity` works in a similar way as the <span>`dispRity`
</span>function: it intakes a `dispRity` object (containing
disparity measurements), a `test` and a `comparisons` argument. The
`test` argument can be any statistical or non-statistical test to apply
to the disparity object. It can be common test function (e.g.
`stats::t.test`) or some functions implemented in <span>`dispRity`
</span>(see `dispRity.test`) or even user defined functions. The
`comparisons` argument must indicate the way the test should be applied
to the data:

-   `pairwise` (default): to compare each series pairwise

-   `referential`: to compare each series to the first one

-   `sequential`: to compare each series to the following one

-   `all`: to compare all the series together (like in analysis
    of variance)

It is also possible to input a list of pairs of `numeric` values or
`characters` matching the series names to create personalised test. Some
other specific tests implemented in `dispRity` such as the
`dispRity::null.test` and the `dispRity::sequential.test` have a
specific way to be applied to the data and therefore ignore the
`comparisons` argument.

This function also allows to correct for type I error inflation when
using multiple comparisons via the `correction` argument. This argument
can be empty (no correction applied) or can contain one of the
corrections from the `stats::p.adjust` function (see `?p.adjust`).

Note that the `test.dispRity` algorithm deals with some classic tests
outputs (`h.test`, `lm` and `numeric` vector) and summarises the test
output. It is however possible to get the full detailed output by using
the options `details = TRUE`.

\
\
`  `

    ##                                 t       df      p.value mean of x mean of y
    ## Group.crown - Group.stem 42.25372 164.8723 2.303646e-90   1.36798  1.239652

\
`     `

    ## $`Group.crown - Group.stem`
    ## 
    ##  Welch Two Sample t-test
    ## 
    ## data:  data[[list_of_comp[[1]]]] and data[[list_of_comp[[2]]]]
    ## t = 42.254, df = 164.87, p-value < 2.2e-16
    ## alternative hypothesis: true difference in means is not equal to 0
    ## 95 percent confidence interval:
    ##  0.1223319 0.1343251
    ## sample estimates:
    ## mean of x mean of y 
    ##  1.367980  1.239652

\
\
`  `\
`         `

    ##             W      p.value
    ## 120 - 80    0 7.457311e-34
    ## 80 - 40  4448 5.334298e-01
    ## 40 - 0   1600 2.961960e-16

\
`     `

    ## Call:
    ##    test(formula = data ~ series, data = data)
    ## 
    ## Terms:
    ##                   series Residuals
    ## Sum of Squares  33.50243  10.79609
    ## Deg. of Freedom        3       396
    ## 
    ## Residual standard error: 0.1651147
    ## Estimated effects may be unbalanced

\
`       `\
` `

    ## [1] "aov" "lm"
    ##              Df Sum Sq Mean Sq F value Pr(>F)    
    ## series        3   33.5  11.167   409.6 <2e-16 ***
    ## Residuals   396   10.8   0.027                   
    ## ---
    ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

\
\
\
\
`    `\
`         `

    ##                bhatt.coeff
    ## 120-80 - 80-40  0.04472136
    ## 120-80 - 40-0   0.03741657
    ## 80-40 - 120-80  0.04472136
    ## 40-0 - 120-80   0.03741657

utilities
---------

### `tree.age`

This function allows to calculate the age of each individual nodes and
tips in a tree. It can either use the root age of the tree (if present
as `$root.time`) or else calculate the age using a user defined root age
via the `age` argument. Also, it is possible to decide whether the time
is calculated towards the past (e.g. million years ago) or towards the
present (e.g. in time since the origin).

\

    ## [1] 139.074

\
` `

    ##    ages     elements
    ## 1 90.00    Daulestes
    ## 2 90.00  Bulaklestes
    ## 3 90.00 Uchkudukodon
    ## 4 77.85  Kennalestes
    ## 5 77.85   Asioryctes

\
`    `

    ##    ages     elements
    ## 1 0.647    Daulestes
    ## 2 0.647  Bulaklestes
    ## 3 0.647 Uchkudukodon
    ## 4 0.560  Kennalestes
    ## 5 0.560   Asioryctes

\
\
`       `

    ##        ages     elements
    ## 1 0.3528637    Daulestes
    ## 2 0.3528637  Bulaklestes
    ## 3 0.3528637 Uchkudukodon
    ## 4 0.4402271  Kennalestes
    ## 5 0.4402271   Asioryctes

Developments
============

As stated at the start of the demo, this version 0.2 is still in
development and many parts are missing. Here are the new functionalities
that will be implemented in further versions.

More user defined functions
---------------------------

I intend also develop functions to help users to develop their own
algorithms for the bootstrap method (via `make.boot`) or the
evolutionary models (via `make.model`). Both functions will provide
similar testing as the `make.metric` function.

Suggestions?
------------

If you have any extra suggestions or comments on what has already been
developed or will be developed, please send me an email
(<guillert@tcd.ie>) or if you are a GitHub user, directly create an
issue on the [GitHub page
(github.com/TGuillerme/dispRity)](https://github.com/TGuillerme/dispRity).

### 2016-05-25 doc/dispRity-palaeo-demo.Rmd
Red manuals INIT

---
title: "dispRity palaeo demo"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: References.bib
bst: sysbio.bst
vignette: >
  %\VignetteIndexEntry{dispRity palaeo demo}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This demo aims to give quick overview of the <span>`dispRity`
</span>package (v.0.2) for palaeobiology analysis. Please refer to
[GitHub page:
github.com/TGuillerme/dispRity](https://github.com/TGuillerme/dispRity)
for other vignettes, namely the [<span>`dispRity`
</span>manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf)
that explains the functions in more detail.

To keep it short, this package allows to easily perform
**disparity-through-time** analysis. This type of analysis often starts
by ordinating **morphometric** or **cladistic** data into a
multidimensional object hereafter called the **morphospace**. One might
be interested in studying how the occupancy of the morphospace changed
through time by measuring a summary metric of this morphospace through
time, called **disparity**. Disparity can then be seen as a value
summarising the diversity of morphologies.

This demo showcases a typical disparity-through-time analyse: we are
going to test whether the disparity changed through time in a subset of
Eutheria (Mammalia) since the last 100 million years.

Before starting
===============

Installing `dispRity` 
-----------------------------------

You can install this package easily if you are using the latest version
of <span>`R` </span>and `devtools`.

` `\
`   `\

The morphospace
---------------

In this example, we are going to use a subset of the data from
@beckancient2014 that contains an ordinated cladistic matrix and a
phylogenetic tree containing 50 taxa along side with a table containing
some first and last occurence data for several taxa. The ordinated
matrix will represent our full morphospace, i.e. all the morphologies
that ever existed through time (for this dataset).

\

\
\
\
\
\

    ##                    [,1]          [,2]        [,3]       [,4]      [,5]
    ## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
    ## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
    ## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
    ## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
    ## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
    ## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273

    ## [1] 50 48

\
\
\
\

    ##             FAD  LAD
    ## Adapis     37.2 36.8
    ## Asioryctes 83.6 72.1
    ## Leptictis  33.9 33.3
    ## Miacis     49.0 46.7
    ## Mimotona   61.6 59.2
    ## Notharctus 50.2 47.0

\
\
`  `

![image](figure/minimal-data-1){width=".8\linewidth"}

Note that you can have an even nicer tree if you use the `strap`
package:

` `\
\
`  `

A disparity-through-time analysis
=================================

Splitting the morphospace through time
--------------------------------------

One of the crucial steps in disparity-through-time analysis is to split
the full morphospace into smaller time seriess that contain the total
amount of morphologies at certain points in time (time-slicing) or
during certain periods in time (time-binning). Basically, the full
morphospace in the previous step represents the total amount of
morphologies in all time and is likely to be greater than any time
series of the morphospace.

The `dispRity` package provides a `time.series` function
that allows to split the morphospace time slices (using
`method = continuous`) or into time bins (using `method = discrete`). In
this example, we are going to split the morphospace into 5 equal time
bins of 20 million years long from 100 million years ago (Mya) to the
present. Note that we will also provide to the function the table
containing the first and last occurrences data for some fossils to take
into account the fact that some fossils might span between our different
time bins.

\
`          `\
\
\
`     `\
`           `\
`     `

The output object is a `dispRity` object. For details about
this object class, please refer to the [<span>`dispRity`
</span>manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf).
In brief, `dispRity` objects are lists of different
elements (i.e. disparity results, morphospace time seriess, etc...) that
display only a summary of the object when calling the object for
avoiding jamming the <span>`R` </span>console.

\

    ## [1] "dispRity"

\

    ## List of 3
    ##  $ data    :List of 5
    ##   ..$ 100-80: num [1:8, 1:48] -0.739 -0.68 -0.511 -0.473 -0.459 ...
    ##   .. ..- attr(*, "dimnames")=List of 2
    ##   .. .. ..$ : chr [1:8] "Daulestes" "Bulaklestes" "Uchkudukodon" "Asioryctes" ...
    ##   .. .. ..$ : NULL
    ##   ..$ 80-60 : num [1:15, 1:48] -0.477 -0.473 -0.474 -0.532 -0.409 ...
    ##   .. ..- attr(*, "dimnames")=List of 2
    ##   .. .. ..$ : chr [1:15] "Kennalestes" "Asioryctes" "Ukhaatherium" "Cimolestes" ...
    ##   .. .. ..$ : NULL
    ##   ..$ 60-40 : num [1:13, 1:48] -0.165 0.254 0.622 0.753 0.671 ...
    ##   .. ..- attr(*, "dimnames")=List of 2
    ##   .. .. ..$ : chr [1:13] "Todralestes" "Pezosiren" "Tribosphenomys" "Paramys" ...
    ##   .. .. ..$ : NULL
    ##   ..$ 40-20 : num [1:6, 1:48] -0.121 -0.203 -0.158 0.696 0.355 ...
    ##   .. ..- attr(*, "dimnames")=List of 2
    ##   .. .. ..$ : chr [1:6] "Leptictis" "Dilambdogale" "Widanelfarasia" "Moeritherium" ...
    ##   .. .. ..$ : NULL
    ##   ..$ 20-0  : num [1:10, 1:48] 0.3079 0.6531 0.5089 -0.0419 0.235 ...
    ##   .. ..- attr(*, "dimnames")=List of 2
    ##   .. .. ..$ : chr [1:10] "Dasypodidae" "Bradypus" "Myrmecophagidae" "Potamogalinae" ...
    ##   .. .. ..$ : NULL
    ##  $ elements: chr [1:50] "Cimolestes" "Maelestes" "Batodon" "Bulaklestes" ...
    ##  $ series  : chr [1:6] "discrete" "100-80" "80-60" "60-40" ...
    ##  - attr(*, "class")= chr "dispRity"

    ## [1] "data"     "elements" "series"

\

    ## 5 discrete series for 50 elements 
    ## Series:
    ## 100-80, 80-60, 60-40, 40-20, 20-0.

Note that these objects will gradually contain more information when
reaching the following steps in the disparity-through-time analysis.

Bootstrapping the data
----------------------

Once we obtain our different time series, we can bootstrap and rarefy
them (i.e. pseudo-replicating the data). The bootstrapping will allow to
make each series more robust to outliers and the rarefaction will allow
us to compare slices with the same number of taxa to get rid of eventual
sampling problems (i.e. more taxa in one bin than the other). The
`boot.matrix` function that allows to bootstrap the <span>`dispRity`
</span>object and the `rarefaction` option within allows to rarefy it.

\
`     `\
\
\

    ## [1] 6

\
`     `\
`      `\
\
\
` `

    ## [1] "dispRity"
    ## [1] "dispRity"

\

    ## Bootstrapped ordinated matrix with 50 elements 
    ## Series:
    ## 100-80, 80-60, 60-40, 40-20, 20-0.
    ## Data was split using discrete method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.

    ## Bootstrapped ordinated matrix with 50 elements 
    ## Series:
    ## 100-80, 80-60, 60-40, 40-20, 20-0.
    ## Data was split using discrete method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.
    ## Data was rarefied with a maximum of 6 elements

Calculating disparity
---------------------

We can now calculate the disparity within each time series along side
with some confidence intervals generated by the pseudo-replication step
above (bootstraps/rarefaction). Disparity can be calculated in many ways
and this package allows user to come up with their own disparity
metrics. For more info, please refer to the [<span>`dispRity`
</span>manual and the metric
vignette](https://github.com/TGuillerme/dispRity).

In this example, we are going to calculate the spread of the data in
each time series by calculating the disparity as the sum of the variance
of each dimension of the morphospace in each time series using the
`dispRity` function. Thus, in this example, the disparity
is defined by the multi-dimensional variance of each time series (i.e.
the spread of the taxa within the morphospace). Note that this metric
comes with a caveat not solved here since it ignores the covariance
among the dimensions of the morphospace.

\
`     `\
\
\
`     `\
\
\
\

    ## Disparity measurements across 5 series for 50 elements 
    ## Series:
    ## 100-80, 80-60, 60-40, 40-20, 20-0.
    ## Disparity calculated as: c(sum, variances) for 48 dimensions.
    ## Data was split using discrete method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.

    ## Disparity measurements across 5 series for 50 elements 
    ## Series:
    ## 100-80, 80-60, 60-40, 40-20, 20-0.
    ## Disparity calculated as: c(sum, variances) for 48 dimensions.
    ## Data was split using discrete method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.
    ## Data was rarefied with a maximum of 6 elements

\

    ##   series  n observed  mean  2.5%   25%   75% 97.5%
    ## 1 100-80  8    1.675 1.460 1.087 1.389 1.568 1.648
    ## 2  80-60 15    1.782 1.678 1.538 1.631 1.728 1.792
    ## 3  60-40 13    1.913 1.773 1.607 1.734 1.826 1.886
    ## 4  40-20  6    2.022 1.673 1.212 1.537 1.822 1.942
    ## 5   20-0 10    1.971 1.773 1.598 1.716 1.842 1.890

    ##   series n observed  mean  2.5%   25%   75% 97.5%
    ## 1 100-80 6       NA 1.440 1.067 1.339 1.575 1.726
    ## 2  80-60 6       NA 1.670 1.276 1.614 1.798 1.922
    ## 3  60-40 6       NA 1.763 1.458 1.704 1.846 1.986
    ## 4  40-20 6     2.02 1.721 1.392 1.633 1.828 1.992
    ## 5   20-0 6       NA 1.762 1.459 1.642 1.880 2.009

As for any <span>`R` </span>package, you can refer yourself to each
individual function for more details.

\

Plotting the results
--------------------

It is sometimes easier to visualise the results in a plot than in a
table. For that we can use the `plot` function to plot the
`dispRity` objects.

\
`          `\
\
\
`      `\
`      `

![image](figure/minimal-plot-1){width="1\linewidth"}

As we can see in this case, the rarefaction tends to increase the
confidence intervals around the mean.

Testing differences
-------------------

Finally, to draw some conclusions from these results, we need to apply
some statistical testing. As stated before, we wanted to test whether
the disparity changed through time in this dataset since the last 100
million years. To do so, we can apply some comparisons of the means
between each time-bins in a sequential manner to see whether the
disparity in bin X is equal to the disparity in bin Y which is in turn
equal to the disparity in bin Z, etc. Because our data is temporally
auto-correlated (i.e. what happens in bin Y depends on what happened in
bin X) and was pseudo-replicated (i.e. each bootstrap draw creates
non-independent time seriess because they are all based on the same time
series), we are going to apply a non-parametric mean comparison: the
`wilcox.test`. Also, we are going to apply a p-value correction to avoid
type I error inflation due to performing multiple test for the same
hypothesis (see `?p.adjust` for more details).

\
`     `\
`      `

    ##                   W      p.value
    ## 100-80 - 80-60  471 7.427563e-28
    ## 80-60 - 60-40  1562 1.798899e-16
    ## 60-40 - 40-20  6250 9.061511e-03
    ## 40-20 - 20-0   3725 7.379715e-03

\
`     `\
`      `

    ##                   W      p.value
    ## 100-80 - 80-60 1695 2.717026e-15
    ## 80-60 - 60-40  3424 4.733196e-04
    ## 60-40 - 40-20  5669 4.095267e-01
    ## 40-20 - 20-0   4254 2.740887e-01

Here our results show significant changes in disparity through time
between all time bins during the last 100 million years. However, when
looking at the rarefied results, there is no significant difference
between the time bins in the Cenozoic (60-40, 40-20 and 20-0 Mya),
suggesting that the differences detected in the first test might just be
due to the differences in number of taxa sample (respectively 13, 6 and
10 in each time bin).

### 2016-05-25 doc/dispRity-metrics.Rmd
Metrics manual update

@@
 
+Metrics implemented in `dispRity`
+===========================================
+
+Several disparity metrics are directly implemented in the `dispRity` package.
+The detailed list can be found in `?dispRity.metric` along with some explanations on each metric.
+
+Level | Name | Description | Source | 
+------|------|---------------------------------------------------|--------|
+3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
+3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
+3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
+3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
+3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
+2 | `ranges` | The range of each dimension | `dispRity` |
+2 | `variances` | The variance of each dimension | `dispRity` |
+2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
+1 | `mode.val` | The modal value | `dispRity` |
+
+1: The functions estimation of the eigen value only works for MDS or PCoA ordinations (not PCA).
+
+2: Note that by default, the centroid is the centroid of the elements.
+It can however be fixed to an arbitrary value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
+
+## Equations and implementations:
+Some of the functions described below are directly implemented in the `dispRity` package and do not require any dependencies.
+
+\begin{equation}
+    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
+\end{equation}
+With *k* being the number of dimensions, $\lambda_i$ is the eigen value of each dimension [R implementation line 140](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R).
+
+\begin{equation}
+    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
+\end{equation}
+With *k* being the dimensions [R implementation line 220](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R).
+
+\begin{equation}
+    ranges = |max(k_i) - min(k_i)|
+\end{equation}
+With *k* being the dimensions [R implementation line 100](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R).
+
+\begin{equation}
+    variances = \sigma^{2}{k_i}
+\end{equation}
+With *k* being the dimensions and $\sigma^{2}$ their variance [R implementation line 95](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R).
+
+\begin{equation}
+    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
+\end{equation}
+With *n* being each element in the ordinated space, *k* being its dimensions and $Centroid_{k}$ their mean (or can be set to an arbitrary value) and  [R implementation line 111](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R).
+
+### Volumes and surface metrics
+```{r}
+# First lets create a small dummy ordinate matrix
+set.seed(1)
+dummy_space <- space.maker(10, 5, rnorm)
+
+# Let's then calculate disparity as the ellipsoid volume:
+summary(dispRity(dummy_space, metric = ellipse.volume))
+
+# Which is the equivalent of doing (with some rounding):
+ellipse.volume(dummy_space)
+
+# Now for the convex hull surface:
+summary(dispRity(dummy_space, metric = convhull.surface))
+
+# And for it's volume:
+summary(dispRity(dummy_space, metric = convhull.volume))
+
+# Note that the more the dimensions, the more the volume tends towards 0!
+summary(dispRity(space.maker(10,9, rnorm), metric = convhull.volume))
+
+# The estimation of the hypervolume is a bit slower:
+summary(dispRity(dummy_space, metric = hyper.volume))
+
+# Note that the hypervolume is different of the convex hull volume!
+# This function also intakes more arguments that can be fed to dispRity
+summary(dispRity(dummy_space, metric = hyper.volume,
+                 repsperpoint = 100, quantile = 0.5))
+# Which gives quiet different results!
+```
+
+### Diagonal, ranges and variances
+```{r}
+# Or we can use other more straightforward metrics such as the diagonal of our
+# dummy space (i.e. the longest euclidean distance in our 5 dimensional space)
+summary(dispRity(dummy_space, metric = diagonal))
+
+# Another type of metrics, based on the range are used a lot in palaeobiology.
+summary(dispRity(dummy_space, metric = ranges))
+# Note that this gives a distribution of metrics now (i.e. the range for our 5
+# dimensions).
+# This distribution can be summarised using metrics such as the sum or product
+# of ranges:
+summary(dispRity(dummy_space, metric = c(sum, ranges)))
+summary(dispRity(dummy_space, metric = c(prod, ranges)))
+
+# The same can be done with the variances
+summary(dispRity(dummy_space, metric = variances))
+summary(dispRity(dummy_space, metric = c(sum, variances)))
+summary(dispRity(dummy_space, metric = c(prod, variances)))
+# Note that neither the sum nor the product of the variances are really
+# representative of the variances distribution!
+```
+
+### Centroids metrics
+```{r}
+# The centroid metric allows to calculate the distances between each elements
+# and their centroids:
+summary(dispRity(dummy_space, metric = centroids))
+# Note that this also gives a distribution
+
+# Another interesting use of this function is for looking at both the average
+# spread of the elements (or the entire distribution!) and their position within
+# the dummy space by looking at the distances between the elements and the
+# center of the dummy space (or an arbitrary point)
+
+centroids_to_group <- dispRity(dummy_space, metric = centroids)
+centroids_to_space <- dispRity(dummy_space, metric = centroids,
+                               centroid = c(rep(0, 5)))
+
+# Average spread
+summary(dispRity(centroids_to_group, metric = median))
+# Average position within de dummy space
+summary(dispRity(centroids_to_space, metric = median))
+```
+
 # References

### 2016-05-29 doc/dispRity-ecology-demo.Rmd

@@
 This demo aims to give quick overview of the <span>`dispRity`
-</span>package (v.0.2) for ecological analysis. Please refer to [GitHub
+</span> package (v.0.2) for ecological analysis. Please refer to [GitHub
 page:
@@
 
-` `\
-`   `\
+```{r, eval = FALSE}
+if(!require(devtools)) install.packages("devtools")
+install_github("TGuillerme/dispRity", ref = "release")
+library(dispRity)
+```
 
@@
 
-\
+```{r}
+## Loading demo and the package data
+library(dispRity)
 
-\
-\
-\
-\
-\
-\
-\
-\
-`  `\
-\
-\
-\
-`  `\
-`  `
+## Setting the random seed for repeatability
+set.seed(123)
+
+## Loading the data
+data(McClean_data)
+
+## This dataset contains an ordinated matrix (20 dimensions) of the distance
+## between 40 experimental plots.
+ord_matrix <- McClean_data$ordination
+
+## As well as two list of different factors affecting each experimental plot:
+## the treatment and the depth.
+treatments <- McClean_data$treatment
+depth <- McClean_data$depth
+```
 
@@
 
-\
-\
-`   `\
-`   `\
-\
-`   `\
-`   `\
-\
-`   `\
-`   `\
-\
-\
-`  `\
-\
-`          `\
-`         `
+```{r}
+##<!--, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>= -->
+## The x and y axis represent the two first dimensions (i.e. principal
+## components) of the ordination
+x <- ord_matrix[, 1]
+y <- ord_matrix[, 2]
+## The colors will represent the treatments
+cols <- sub("a", "red", treatments)
+cols <- sub("b", "blue", cols)
+## The symbols will represent the depth
+pchs <- sub(1, 16, depth)
+pchs <- as.numeric(sub(2, 17, pchs))
+
+## Graphical option
+par(bty = "n")
+## A classic 2D ordination plot
+plot(x, y, col = cols, pch = pchs, xlab = "PC 1", ylab = "PC 2",
+    xlim = range(x) + c(0, 100))
+```
 
@@
 
-\
-`       `\
-`            `\
-`   `\
-
-    ##     Treat Depth
-    ## 1       a     1
-    ## 1.1     a     2
-    ## 2       b     1
-    ## 2.1     b     2
-    ## 3       a     1
-    ## 3.1     a     1
+```{r}
+## Creating the table that contain the elements and their attributes
+factors <- as.data.frame(matrix(data = c(treatments, depth), nrow = nrow(ord_matrix),
+    ncol = 2, byrow = FALSE, dimnames = list(rownames(ord_matrix))))
+names(factors) <- c("Treat", "Depth")
+head(factors)
+```
 
@@
 
-\
-`  `\
-\
-\
+```{r}
+## Splitting the ordinated space into four subsamples
+customised_series <- cust.series(ord_matrix, factors)
 
-    ## [1] "dispRity"
-
-\
-\
-
-    ## 4 custom series for 40 elements 
-    ## Series:
-    ## Treat.a, Treat.b, Depth.1, Depth.2.
+## Note that the output of dispRity functions are dispRity objects
+class(customised_series)
+## These objects are automatically printed in a summary method (calling S3 print.dispRity)
+## giving information about the object
+customised_series
+```
 
@@
 
-\
-\
-\
-\
-`   `
+```{r}
+## Summarising the object
+str(customised_series)
+
+## Displaying the full object
+print(customised_series, all = TRUE)
+```
 
@@
 
-\
-`     `\
-\
-\
-\
-\
+```{r}
+## Calculating disparity
+disparity <- dispRity(customised_series, metric = c(median, centroids))
 
-    ## Disparity measurements across 4 series for 40 elements 
-    ## Series:
-    ## Treat.a, Treat.b, Depth.1, Depth.2.
-    ## Disparity calculated as: c(median, centroids) for 20 dimensions.
-    ## Data was split using custom method.
-
-    ##    series  n observed
-    ## 1 Treat.a 21    82.38
-    ## 2 Treat.b 19    94.68
-    ## 3 Depth.1 23    83.98
-    ## 4 Depth.2 17    89.72
+## Note that disparity is a dispRity object and printing it just gives details
+## on the object, not the results. We need to use summary.dispRity (S3) to get
+## the results.
+disparity
+summary(disparity)
+```
 
@@
 
-\
-`     `\
-\
-\
-`     `\
-`      `\
-\
-\
-\
+```{r}
+## Simple bootstrapping (100 times)
+bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100)
 
-    ## Bootstrapped ordinated matrix with 40 elements 
-    ## Series:
-    ## Treat.a, Treat.b, Depth.1, Depth.2.
-    ## Data was split using custom method.
-    ## Data was bootstrapped 100 times, using the full bootstrap method.
+## Bootstrapping with rarefaction (i.e. only re-sampling 17 elements each time)
+rarefied_data <- boot.matrix(customised_series, bootstraps = 100,
+    rarefaction = 17)
 
-    ## Bootstrapped ordinated matrix with 40 elements 
-    ## Series:
-    ## Treat.a, Treat.b, Depth.1, Depth.2.
-    ## Data was split using custom method.
-    ## Data was bootstrapped 100 times, using the full bootstrap method.
-    ## Data was rarefied with a maximum of 17 elements
+## Note that the output is a dispRity object giving some details on the series
+## and the bootstraps
+bootstrapped_data
+rarefied_data
+```
 
@@
 
-\
-`     `\
-\
-\
-`     `
+```{r}
+## Bootstrapped disparity
+disparity_BS <- dispRity(bootstrapped_data, metric = c(median, centroids))
+
+## Rarefied disparity
+disparity_rare <- dispRity(rarefied_data, metric = c(median, centroids))
+```
 
@@
 
-\
+```{r}
+## Summarising disparity (default)
+summary(disparity_BS)
 
-    ##    series  n observed  mean  2.5%   25%   75% 97.5%
-    ## 1 Treat.a 21    82.38  79.3 64.83 74.45  84.1  95.4
-    ## 2 Treat.b 19    94.68 100.2 79.63 90.99 106.8 133.6
-    ## 3 Depth.1 23    83.98  82.4 69.04 78.32  85.9  92.7
-    ## 4 Depth.2 17    89.72 100.9 69.58 85.03 110.6 163.4
+## The quantiles are calculated as 50 and 95 and the central tendency is the
+## mean by default but we can specify different options
+summary(disparity_BS, quantile = 90, cent.tend = median)
 
-\
-\
-`     `
+## Finally we can see the results of the rarefaction analysis:
+summary(disparity_rare)
 
-    ##    series  n observed median    5%   95%
-    ## 1 Treat.a 21    82.38  78.50 68.34  91.2
-    ## 2 Treat.b 19    94.68  97.02 83.17 128.3
-    ## 3 Depth.1 23    83.98  83.21 72.79  91.3
-    ## 4 Depth.2 17    89.72  93.04 75.77 149.4
-
-\
-
-    ##    series  n observed  mean  2.5%   25%   75% 97.5%
-    ## 1 Treat.a 17       NA  79.4 67.47 74.52  83.1  96.7
-    ## 2 Treat.b 17       NA  99.8 79.24 88.04 106.9 147.4
-    ## 3 Depth.1 17       NA  82.4 69.41 77.08  86.3 106.1
-    ## 4 Depth.2 17    89.72 102.7 77.36 86.44 113.7 159.0
-
-\
+## Note that for the first three series, there is no observed disparity since
+## the data was rarefied for these series.
+```
 
@@
 
-\
-`          `\
-\
-\
-`   `\
-\
-\
-`   `
+```{r}
+## Graphical options
+quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
+
+## Plotting the bootstrapped disparity
+plot(disparity_BS, main = "Bootstrapped data")
+
+## Plotting the rarefied disparity
+plot(disparity_rare, main = "Rarefied data")
+```
 
@@
 
-\
-`     `
+```{r}
+## Testing the effect of our factors on the bootstrapped data
+summary(test.dispRity(disparity_BS, test = aov, comparisons = "all"))
 
-    ##              Df Sum Sq Mean Sq F value Pr(>F)    
-    ## series        3  39184   13061   61.17 <2e-16 ***
-    ## Residuals   396  84553     214                   
-    ## ---
-    ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## Testing the effect of our factors on the rarefied data
+summary(test.dispRity(disparity_rare, test = aov, comparisons = "all"))
 
-\
-`     `
+## Post-hoc testing of the effect of the two different factors
+## Note that the comparison contains the list of the pairs of series to compare
+test.dispRity(disparity_BS, test = wilcox.test,
+    comparisons = list(c(1,2), c(3,4)))
 
-    ##              Df Sum Sq Mean Sq F value Pr(>F)    
-    ## series        3  42229   14076   63.85 <2e-16 ***
-    ## Residuals   396  87301     220                   
-    ## ---
-    ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-
-\
-\
-`  `\
-`       `
-
-    ##                      W      p.value
-    ## Treat.a - Treat.b  695 7.170082e-26
-    ## Depth.1 - Depth.2 2064 7.360379e-13
-
-\
-`  `\
-`       `
-
-    ##                      W      p.value
-    ## Treat.a - Treat.b  880 7.845518e-24
-    ## Depth.1 - Depth.2 1599 9.672040e-17
+## Testing the effect of the two different factors for the rarefied data
+test.dispRity(disparity_rare, test = wilcox.test,
+    comparisons = list(c(1,2), c(3,4)))
+```
 

### 2016-05-29 doc/dispRity-manual.Rmd
Added code to the Rmd docs (removed redundant Rnw)

@@
 
-\
+```{r}
+## Loading the package
+library(dispRity)
 
-\
-\
-\
-\
-\
+## Setting the random seed for repeatability
+set.seed(123)
 
-    ## [1] 50 48
+## Loading the ordinated matrix containing 50 taxa
+data(BeckLee_mat50)
+dim(BeckLee_mat50)
+head(BeckLee_mat50[,1:5])
 
-    ##                    [,1]          [,2]        [,3]       [,4]      [,5]
-    ## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
-    ## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
-    ## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
-    ## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
-    ## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
-    ## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273
+## Loading another ordinated matrix containing 50 tips + 49 nodes
+data(BeckLee_mat99)
+dim(BeckLee_mat99)
+head(BeckLee_mat99[,1:5], 2)
+tail(BeckLee_mat99[,1:5], 2)
 
-\
-\
+## Loading a list of first and last occurrence data
+data(BeckLee_ages)
+head(BeckLee_ages)
 
-    ## [1] 99 97
-
-` `
-
-    ##                  [,1]       [,2]       [,3]       [,4]        [,5]
-    ## Cimolestes -0.6082437 -0.0323683 0.08458885 -0.4338448 -0.30536875
-    ## Maelestes  -0.5730206 -0.2840361 0.01308847 -0.1258848  0.06123611
-
-` `
-
-    ##            [,1]      [,2]       [,3]       [,4]       [,5]
-    ## n48 -0.05529018 0.4799330 0.04118477 0.04944912 -0.3558830
-    ## n49 -0.13067785 0.4478168 0.11956268 0.13800340 -0.3222785
-
-\
-\
-
-    ##             FAD  LAD
-    ## Adapis     37.2 36.8
-    ## Asioryctes 83.6 72.1
-    ## Leptictis  33.9 33.3
-    ## Miacis     49.0 46.7
-    ## Mimotona   61.6 59.2
-    ## Notharctus 50.2 47.0
-
-\
-\
-`         `
+## Loading the phylogeny
+data(BeckLee_tree)
+plot(BeckLee_tree, cex = 0.8) ; nodelabels(cex = 0.8) ; axisPhylo(root = 140)
+```
 
@@
 
-\
-`     `\
-`           `\
-`     `\
-\
-\
-`   `\
-\
-\
-`     `\
-\
-\
+```{r}
+## Splitting the data
+sliced_data <- time.series(BeckLee_mat99, BeckLee_tree, method = "continuous",
+    model = "acctran", time = rev(seq(from=0, to=130, by=15)),
+    FADLAD = BeckLee_ages)
 
-    ##   series  n observed  mean  2.5%   25%   75% 97.5%
-    ## 1    120  5    1.494 1.200 0.756 0.854 1.484 1.550
-    ## 2    105 11    1.725 1.646 1.481 1.614 1.686 1.770
-    ## 3     90 18    1.820 1.775 1.681 1.747 1.806 1.849
-    ## 4     75 19    1.923 1.814 1.681 1.778 1.865 1.901
-    ## 5     60 20    1.946 1.859 1.751 1.833 1.894 1.925
-    ## 6     45 14    1.930 1.849 1.742 1.827 1.883 1.951
-    ## 7     30 10    1.910 1.796 1.570 1.774 1.866 1.935
-    ## 8     15 10    1.910 1.801 1.594 1.782 1.853 1.914
-    ## 9      0 10    1.910 1.815 1.588 1.788 1.865 1.929
+## Bootstrapping the data
+bootstrapped_data <- boot.matrix(sliced_data, 100)
 
-\
-\
-\
-`     `
+## Calculating disparity
+median_centroids <- dispRity(bootstrapped_data, metric = c(median, centroids))
 
-    ##              Df Sum Sq Mean Sq F value Pr(>F)    
-    ## series        8  34.47   4.309   279.5 <2e-16 ***
-    ## Residuals   891  13.74   0.015                   
-    ## ---
-    ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## Summarising the results
+summary(median_centroids)
+plot(median_centroids)
+
+## Testing the effect of time on disparity
+summary(test.dispRity(median_centroids, test = aov, comparisons = "all"))
+```
 
@@
 
-\
+```{r}
+str(median_centroids)
+## That's a more than 4000 lines of output!
+```
 
@@
 
-\
+```{r}
+## What is the class of the median_centroids object?
+class(median_centroids)
 
-    ## [1] "dispRity"
+## What is in the object?
+names(median_centroids)
 
-\
+## Summarising it using the S3 method print.dispRity:
+median_centroids
+## This displays some basic information about the object content
+@
 
-    ## [1] "data"      "disparity" "elements"  "series"    "call"
+\noindent Note however, that it is always possible to recall the full object using the argument \texttt{all=TRUE}:
 
-\
-
-    ## Disparity measurements across 9 series for 99 elements 
-    ## Series:
-    ## 120, 105, 90, 75, 60, 45 ...
-    ## Disparity calculated as: c(median, centroids) for 97 dimensions.
-    ## Data was split using continuous method.
-    ## Data was bootstrapped 100 times, using the full bootstrap method.
-
-Note however, that it is always possible to recall the full object using
-the argument `all=TRUE`:
-
-\
-`   `\
+<<dispRity class3, eval=FALSE>>=
+## Displaying the full object
+print(median_centroids, all = TRUE)
+## But that's 155091 lines!
+```
 
@@
 
-\
-`     `\
+```{r}
+## Extracting some specific series from the disparity object
+series_1_and_4 <- get.dispRity(median_centroids, what = c(1,4))
+series_1_and_4
 
-    ## Disparity measurements across 2 series for 24 elements 
-    ## Series:
-    ## 120, 75.
-    ## Disparity calculated as: c(median, centroids) for 97 dimensions.
-    ## Data was split using continuous method.
-    ## Data was bootstrapped 100 times, using the full bootstrap method.
+## The observed disparity
+extract.dispRity(median_centroids)
 
-\
-
-    ##      120      105       90       75       60       45       30       15        0 
-    ## 1.493570 1.725280 1.820292 1.922531 1.945952 1.930306 1.910203 1.910203 1.910203
-
-\
-`   `\
-`      `
-
-    ## List of 1
-    ##  $ 120: num [1:100] 1.52 0.786 1.555 1.288 0.891 ...
+## The distribution of the bootstrapped disparity scores for the first series
+str(extract.dispRity(get.dispRity(median_centroids, what = 1),
+    observed = FALSE))
+```
 
@@
 
-\
-`  `
+```{r}
+## Disparity measured as the mean value of the ordinated matrix
+summary(dispRity(sliced_data, metric = mean))
 
-    ##   series  n observed
-    ## 1    120  5   -0.005
-    ## 2    105 11   -0.010
-    ## 3     90 18   -0.010
-    ## 4     75 19   -0.001
-    ## 5     60 20    0.005
-    ## 6     45 14    0.013
-    ## 7     30 10    0.017
-    ## 8     15 10    0.017
-    ## 9      0 10    0.017
+## A function for measuring the sum divided by the length (the mean!)
+sum.divide.by.length <- function(X) sum(X)/length(X)
 
-\
-`   `\
-\
-\
-`  `
-
-    ##   series  n observed
-    ## 1    120  5   -0.005
-    ## 2    105 11   -0.010
-    ## 3     90 18   -0.010
-    ## 4     75 19   -0.001
-    ## 5     60 20    0.005
-    ## 6     45 14    0.013
-    ## 7     30 10    0.017
-    ## 8     15 10    0.017
-    ## 9      0 10    0.017
+## Disparity measured as the mean divided by the length
+summary(dispRity(sliced_data, metric = sum.divide.by.length))
+## Giving the exact same results!
+```
 
@@
 
-\
-\
-`     `\
-`    `\
-\
-`  `\
-`         `\
-`          `\
-\
-\
-\
-`  `\
-\
-\
+```{r}
+## We want to separate the species around the node 71 (see phylogeny above).
+## All the descendant of these node are crown and all the ancestors are stem.
+crown <- extract.clade(BeckLee_tree, node = 71)$tip.label
+stem <- drop.tip(BeckLee_tree, tip = crown)$tip.label
+## We then have to feed this information in a data frame with one column
+factors <- as.data.frame(
+    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
+    ncol = 1, dimnames = list(c(crown, stem), "Group")))
 
-    ## [1] "dispRity"
+## We then can use the customised series function to create the two series
+## i.e. the two regions of the ordinated space
+crown_stem <- cust.series(BeckLee_mat50, factors)
 
-    ## 2 custom series for 50 elements 
-    ## Series:
-    ## Group.crown, Group.stem.
+## This created a dispRity object containing two series: crown and stem
+class(crown_stem)
+crown_stem
 
-\
-
-    ## [1] "data"     "elements" "series"
-
-\
-
-    ## List of 2
-    ##  $ Group.crown: num [1:30, 1:48] 0.3079 0.6531 0.5089 -0.1652 -0.0419 ...
-    ##   ..- attr(*, "dimnames")=List of 2
-    ##   .. ..$ : chr [1:30] "Dasypodidae" "Bradypus" "Myrmecophagidae" "Todralestes" ...
-    ##   .. ..$ : NULL
-    ##  $ Group.stem : num [1:20, 1:48] -0.739 -0.68 -0.511 -0.477 -0.473 ...
-    ##   ..- attr(*, "dimnames")=List of 2
-    ##   .. ..$ : chr [1:20] "Daulestes" "Bulaklestes" "Uchkudukodon" "Kennalestes" ...
-    ##   .. ..$ : NULL
-
-\
-
-    ##  chr [1:50] "Cimolestes" "Maelestes" "Batodon" "Bulaklestes" "Daulestes" ...
-
-\
-\
-
-    ## [1] "custom"      "Group.crown" "Group.stem"
+## This object contains three elements
+names(crown_stem)
+## With "data" being the list sub-matrices
+str(crown_stem$data)
+## "elements" being the list of taxa in the original ordinated matrix
+str(crown_stem$elements)
+## and finally, "series" containing information on the series type (custom)
+## and names (crown and stem)
+crown_stem$series
+```
 
@@
 
-\
-`  `\
-`         `\
-`        `\
-`           `\
-\
-\
+```{r}
+## Adding all mammals to the a new series (along with crown and stem as before)
+factors2 <- as.data.frame(
+    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem)),
+    rep("Mammalia", Ntip(BeckLee_tree))), ncol = 2,
+    dimnames = list(c(crown, stem), c("Group", "Class")), byrow = FALSE))
 
-    ## 3 custom series for 50 elements 
-    ## Series:
-    ## Group.crown, Group.stem, Class.Mammalia.
+## Creating the three series: Group.stem/crown and Class.Mammalia
+cust.series(BeckLee_mat50, factors2)
+```
 
@@
 
-\
-`      `\
-`         `
+```{r}
+## Generating three time bins containing the data present every 40 Ma
+time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+    time = c(120, 80, 40, 0))
 
-    ## 3 discrete series for 50 elements 
-    ## Series:
-    ## 120-80, 80-40, 40-0.
-
+```
 In this example, the taxa where split inside each timebin according to
@@
 
-\
-\
-`     `\
-`              `\
+```{r}
+## Generating three time bins containing the data present every 40 Ma, including
+## taxa that might span between time bins.
+time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
+    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
+time_bins
 
-    ## 3 discrete series for 50 elements 
-    ## Series:
-    ## 120-80, 80-40, 40-0.
+```
 
@@
 
-\
-`      `\
-`      `
-
-    ## 3 discrete series for 50 elements 
-    ## Series:
-    ## 133.51104-89.00736, 89.00736-44.50368, 44.50368-0.
+```{r}
+## Automatically generate three equal length bins:
+time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+    time = 3)
+```
 
@@
 
-\
-\
-`     `\
-`               `\
-`     `\
-
-    ## 4 continuous series for 99 elements 
-    ## Series:
-    ## 120, 80, 40, 0.
-
-\
-\
-`      `\
-`         `
-
-    ## 4 continuous series for 99 elements 
-    ## Series:
-    ## 133.51104, 89.00736, 44.50368, 0.
+```{r}
+## Generating four time slices every 40 million years assuming a gradual
+## evolution model
+time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
+    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
+    FADLAD = BeckLee_ages)
+time_slices
+## Note that in the same way as for the discrete method, we can also
+## automatically generate the slices
+time.series(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
+    model = "gradual", time = 4)
+```
 
@@
 
-\
-` `
-
-    ## Bootstrapped ordinated matrix with 50 elements 
-    ## 1
-    ## Data was bootstrapped 1000 times, using the full bootstrap method.
-
-\
-\
+```{r}
+## Bootstrapping a matrix
+boot.matrix(data = BeckLee_mat50)
+## As we can see, the output is also a dispRity object that is summarised
+## automatically, and gives information on the data as well as the number of
+## bootstraps and the bootstraps methods.
+```
 
@@
 
-\
-`   `
+```{r}
+## Bootstrapping with the single bootstrap method
+boot.matrix(BeckLee_mat50, boot.type = "single")
 
-    ## Bootstrapped ordinated matrix with 50 elements 
-    ## 1
-    ## Data was bootstrapped 1000 times, using the single bootstrap method.
+## Bootstrapping with the full rarefaction
+boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
 
-\
-`      `
+## Or with a set number of rarefaction levels
+boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8,3))
 
-    ## Bootstrapped ordinated matrix with 50 elements 
-    ## 1
-    ## Data was bootstrapped 20 times, using the full bootstrap method.
-    ## Data was fully rarefied (down to 3 elements).
+## And removing the last axis (default)
+boot.matrix(BeckLee_mat50, rm.last.axis = TRUE)
 
-\
-`      `
-
-    ## Bootstrapped ordinated matrix with 50 elements 
-    ## 1
-    ## Data was bootstrapped 20 times, using the full bootstrap method.
-    ## Data was rarefied with a maximum of 6, 7, 8 and 3 elements
-
-\
-`   `
-
-    ## Bootstrapped ordinated matrix with 50 elements 
-    ## 1
-    ## Data was bootstrapped 1000 times, using the full bootstrap method. 
-    ## The 6 last axis were removed from the original ordinated data.
-
-<span>`## Or with a fix value (50%)`</span>\
-`   `
-
-    ## Bootstrapped ordinated matrix with 50 elements 
-    ## 1
-    ## Data was bootstrapped 1000 times, using the full bootstrap method. 
-    ## The 35 last axis were removed from the original ordinated data.
+## Or with a fix value (50%)
+boot.matrix(BeckLee_mat50, rm.last.axis = 0.5)
+```
 
@@
 
-\
-`        `\
-\
-\
-`     `\
-`     `\
-\
-\
+```{r}
+## Bootstrap and full rarefaction on the crown/stem series
+crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
 
-    ## Bootstrapped ordinated matrix with 50 elements 
-    ## Series:
-    ## Group.crown, Group.stem.
-    ## Data was split using custom method.
-    ## Data was bootstrapped 100 times, using the full bootstrap method.
-    ## Data was fully rarefied (down to 3 elements).
+## Bootstrap on the time binning/slicing series
+time_binsBS <- boot.matrix(time_bins, bootstraps = 100)
+time_slicesBS <- boot.matrix(time_slices, bootstraps = 100)
 
-    ## Bootstrapped ordinated matrix with 50 elements 
-    ## Series:
-    ## 120-80, 80-40, 40-0.
-    ## Data was split using discrete method.
-    ## Data was bootstrapped 100 times, using the full bootstrap method.
-
-    ## Bootstrapped ordinated matrix with 99 elements 
-    ## Series:
-    ## 120, 80, 40, 0.
-    ## Data was split using continuous method.
-    ## Data was bootstrapped 100 times, using the full bootstrap method.
+## Note that all these objects are of class dispRity
+crown_stemBS
+time_binsBS
+time_slicesBS
+```
 
@@
 
-`  `
+```{r}
+dispRity(BeckLee_mat50, metric = mean)
+## This defines disparity as the mean value of the morphospace
+## (using base::mean).
 
-    ## Disparity measurements across 1 series for 50 elements 
-    ## 1
-    ## Disparity calculated as: mean for 48 dimensions.
+## It is also possible to combine multiple functions:
+dispRity(BeckLee_mat50, metric = c(sum, variances))
+## Defining disparity as the sum (base::sum) of the variances
+## (dispRity::variances) of each dimension of the morphospace.
 
-\
-\
-\
-\
-`   `
+dispRity(BeckLee_mat50, metric = c(prod, centroids))
+## For the product (base::prod) of the distances between each elements and the
+## centroid of the morphospace (dispRIty::centroids).
 
-    ## Disparity measurements across 1 series for 50 elements 
-    ## 1
-    ## Disparity calculated as: c(sum, variances) for 48 dimensions.
+## Or user defined ones:
+total.range <- function(X) abs(range(X)[1]-range(X)[2])
+dispRity(BeckLee_mat50, metric = c(total.range, centroids))
+## For the range (total.range, user defined) of the distances between each
+## elements and the centroid of the morphospace (dispRity::centroids).
 
-\
-\
-\
-`   `
+## Or more complex ones:
+vars1 <- dispRity(BeckLee_mat50, metric = c(sd, variances, var))
+## For the standard deviation (stats::sd) of the variance (dispRity::variances)
+## of each column of the variance/covariance matrix (stats::var).
 
-    ## Disparity measurements across 1 series for 50 elements 
-    ## 1
-    ## Disparity calculated as: c(prod, centroids) for 48 dimensions.
-
-\
-\
-\
-\
-`   `\
-`   `
-
-    ## Disparity measurements across 1 series for 50 elements 
-    ## 1
-    ## Disparity calculated as: c(total.range, centroids) for 48 dimensions.
-
-\
-\
-\
-\
-`     `\
-\
-\
-\
-\
-\
-`     `\
-`  `
-
-    ## [1] TRUE
+## Note that the order of the function is not important since the levels of
+## each function are automatically detected by the dispRity algorithm.
+vars2 <- dispRity(BeckLee_mat50, metric = c(variances, var, sd))
+all(summary(vars1) == summary(vars2))
+```
 
@@
 
-\
-`     `\
-\
-\
-`     `\
-`     `
+```{r}
+## Disparity in crown and stem mammals:
+disp_crown_stemBS <- dispRity(crown_stemBS, metric = c(median, centroids))
+
+## Disparity through time:
+disp_time_binsBS <- dispRity(time_binsBS, metric = c(median, centroids))
+disp_time_slicesBS <- dispRity(time_slicesBS, metric = c(median, centroids))
+
+```
 
@@
 
-    ##   series  n observed  mean  2.5%   25%   75% 97.5%
-    ## 1 120-80  8    1.203 1.103 0.895 1.057 1.178 1.236
-    ## 2  80-40 27    1.344 1.322 1.275 1.306 1.339 1.366
-    ## 3   40-0 16    1.353 1.323 1.265 1.304 1.346 1.368
+```{r}
+summary(disp_time_binsBS)
+## By default, the 50 and 95 quantiles and the mean are calculated.
+## Note that the function also displays the observed disparity
+## (non-bootstrapped).
 
-\
-\
-\
-\
-\
-`        `
+## These arguments can be changed easily as follow
+summary(disp_time_binsBS, quantile = 88, cent.tend = sd, rounding = 1)
 
-    ##   series  n observed  sd  6% 94%
-    ## 1 120-80  8      1.2 0.1 1.0 1.2
-    ## 2  80-40 27      1.3 0.0 1.3 1.4
-    ## 3   40-0 16      1.4 0.0 1.3 1.4
+## If on happens to forgot what was calculated in the summarised object,
+## it is possible to recall the different steps details using recall:
+summary(disp_time_binsBS, recall = TRUE)
 
-\
-\
-`   `
-
-    ## Disparity calculated as: c(median, centroids) for 48 dimensions.
-    ## Data was split using discrete method.
-    ## Data was bootstrapped 100 times, using the full bootstrap method.
-    ##   series  n observed  mean  2.5%   25%   75% 97.5%
-    ## 1 120-80  8    1.203 1.103 0.895 1.057 1.178 1.236
-    ## 2  80-40 27    1.344 1.322 1.275 1.306 1.339 1.366
-    ## 3   40-0 16    1.353 1.323 1.265 1.304 1.346 1.368
-
-\
-\
-
-    ##        series  n observed  mean  2.5%   25%   75% 97.5%
-    ## 41 Group.stem 15       NA 1.229 1.146 1.210 1.253 1.282
-    ## 42 Group.stem 16       NA 1.232 1.154 1.217 1.250 1.279
-    ## 43 Group.stem 17       NA 1.233 1.174 1.218 1.251 1.281
-    ## 44 Group.stem 18       NA 1.236 1.175 1.219 1.255 1.282
-    ## 45 Group.stem 19       NA 1.239 1.173 1.226 1.256 1.286
-    ## 46 Group.stem 20     1.27 1.240 1.185 1.225 1.258 1.280
+## Note that the information with each different number of elements is
+## displayed for rarefied data:
+tail(summary(disp_crown_stemBS))
+## The observed disparity is NA for rarefied data.
+```
 
@@
 
-\
-`          `\
-\
-\
-`   `\
-\
-\
-\
-`      `\
-\
-`         `
+```{r}
+## Graphical options
+quartz(width = 15, height = 5) ; par(mfrow = (c(1,3)), bty = "n")
+
+## Plotting the disparity for both groups
+plot(disp_crown_stemBS, type = "box")
+
+## Same plot but with the number of elements on the right and using the
+## "polygon" plot type.
+plot(disp_crown_stemBS, type = "polygon", elements = TRUE)
+## Rarefied version of the same plot (20 elements per series)
+plot(disp_crown_stemBS, type = "polygon", elements = TRUE, rarefaction = 20)
+```
 
@@
 
-\
-`  `\
-\
-`      `
+```{r}
+## Graphical options
+par(bty = "n")
+## The rarefaction curves
+plot(disp_crown_stemBS, elements = TRUE, rarefaction = "plot")
+
+```
 
@@
 
-\
-`  `\
-\
-\
-`      `
+```{r}
+## Graphical options
+par(bty = "n")
+## The disparity-through-time data in a time-binned way (with lines rather than
+## boxes) with the observed data as crosses
+plot(disp_time_binsBS, type = "lines", observed = TRUE)
+```
 
@@
 
-\
-`  `\
-\
-\
-`   `\
+```{r}
+## Graphical options
+par(bty = "n")
+
+## First, plotting the continuous data with a y axis limit
+plot(disp_time_slicesBS, type = "continuous")
+## Note that the type = "continuous" argument is not mandatory here.
+```
 
@@
 
-\
-`  `\
-\
-`     `\
-`           `\
-`           `\
-`         `\
-`         `
+```{r}
+## Graphical options
+par(bty = "n")
+## A plot with many options!
+plot(disp_time_slicesBS, quantile = c(seq(from=10, to=100, by=10)),
+    cent.tend = sd, type = "continuous", elements = "log",
+    col = c("black", rainbow(10)), ylab = c("Disparity", "log(Diversity)"),
+    xlab = "Time (in in units from past to present)", time.series = FALSE,
+    observed = TRUE, main = "Many options...")
+```
 
@@
 
-\
-`  `\
-\
-\
-`   `\
-\
-\
-`  `\
-`      `
+```{r}
+## Graphical options
+par(bty = "n")
+
+## First, plotting the continuous data with a y axis limit
+plot(disp_time_slicesBS, ylim = c(0.5,2))
+
+## Second, adding the discrete data
+par(new = TRUE)
+plot(disp_time_binsBS, discrete_type = "line", ylim=c(0.5,2), xlab="", ylab="")
+```
 
@@
 
-\
-\
-`  `
+```{r}
+## Performing a pairwise t-test to the test a difference in disparity between
+## crown and stem mammals (note that "pairwise" is default)
+test.dispRity(disp_crown_stemBS, test = t.test)
 
-    ##                                 t       df      p.value mean of x mean of y
-    ## Group.crown - Group.stem 42.25372 164.8723 2.303646e-90   1.36798  1.239652
+## The same test but with the detailed output
+test.dispRity(disp_crown_stemBS, test = t.test, details = TRUE)
 
-\
-`     `
+## A wilcoxon rank test applied to sliced disparity the data sequentially with
+## the Bonferroni correction.
+test.dispRity(disp_time_slicesBS, test = wilcox.test,
+    comparisons = "sequential", correction = "bonferroni")
 
-    ## $`Group.crown - Group.stem`
-    ## 
-    ##  Welch Two Sample t-test
-    ## 
-    ## data:  data[[list_of_comp[[1]]]] and data[[list_of_comp[[2]]]]
-    ## t = 42.254, df = 164.87, p-value < 2.2e-16
-    ## alternative hypothesis: true difference in means is not equal to 0
-    ## 95 percent confidence interval:
-    ##  0.1223319 0.1343251
-    ## sample estimates:
-    ## mean of x mean of y 
-    ##  1.367980  1.239652
+## An analysis of variance (aov) applied to the same data
+test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all")
 
-\
-\
-`  `\
-`         `
+## Note that in this case, the output is the regular aov output:
+slice_aov <- test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all")
+class(slice_aov) ; summary(slice_aov)
 
-    ##             W      p.value
-    ## 120 - 80    0 7.457311e-34
-    ## 80 - 40  4448 5.334298e-01
-    ## 40 - 0   1600 2.961960e-16
+## Measuring the overlap between distributions in the time bins (using the
+## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff) with a
+## personalised series comparisons and some additional arguments to pass to the
+## test function (bw = bw.nrd)
+test.dispRity(disp_time_binsBS, test = bhatt.coeff, bw = bw.nrd,
+    comparisons = list(c(1,2), c(1,3), c(2,1), c(3,1)))
 
-\
-`     `
-
-    ## Call:
-    ##    test(formula = data ~ series, data = data)
-    ## 
-    ## Terms:
-    ##                   series Residuals
-    ## Sum of Squares  33.50243  10.79609
-    ## Deg. of Freedom        3       396
-    ## 
-    ## Residual standard error: 0.1651147
-    ## Estimated effects may be unbalanced
-
-\
-`       `\
-` `
-
-    ## [1] "aov" "lm"
-    ##              Df Sum Sq Mean Sq F value Pr(>F)    
-    ## series        3   33.5  11.167   409.6 <2e-16 ***
-    ## Residuals   396   10.8   0.027                   
-    ## ---
-    ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-
-\
-\
-\
-\
-`    `\
-`         `
-
-    ##                bhatt.coeff
-    ## 120-80 - 80-40  0.04472136
-    ## 120-80 - 40-0   0.03741657
-    ## 80-40 - 120-80  0.04472136
-    ## 40-0 - 120-80   0.03741657
+```
 
@@
 
-\
-
-    ## [1] 139.074
-
-\
-` `
-
-    ##    ages     elements
-    ## 1 90.00    Daulestes
-    ## 2 90.00  Bulaklestes
-    ## 3 90.00 Uchkudukodon
-    ## 4 77.85  Kennalestes
-    ## 5 77.85   Asioryctes
-
-\
-`    `
-
-    ##    ages     elements
-    ## 1 0.647    Daulestes
-    ## 2 0.647  Bulaklestes
-    ## 3 0.647 Uchkudukodon
-    ## 4 0.560  Kennalestes
-    ## 5 0.560   Asioryctes
-
-\
-\
-`       `
-
-    ##        ages     elements
-    ## 1 0.3528637    Daulestes
-    ## 2 0.3528637  Bulaklestes
-    ## 3 0.3528637 Uchkudukodon
-    ## 4 0.4402271  Kennalestes
-    ## 5 0.4402271   Asioryctes
+```{r}
+## This tree has a root age
+BeckLee_tree$root.time
+## So we can get the age of each tips and nodes directly
+head(tree.age(BeckLee_tree), 5)
+## But we can also decide to make the age relative (between 1 and 0)
+head(tree.age(BeckLee_tree, age = 1), 5)
+## Or even relative, but from the root (i.e. how far are the nodes/tips
+## from the root)
+head(tree.age(BeckLee_tree, age = 1, order = "present"), 5)
+```
 

### 2016-05-29 doc/dispRity-palaeo-demo.Rmd
Added code to the Rmd docs (removed redundant Rnw)

@@
 
-\
+```{r}
+## Loading demo and the package data
+library(dispRity)
 
-\
-\
-\
-\
-\
+## Setting the random seed for repeatability
+set.seed(123)
 
-    ##                    [,1]          [,2]        [,3]       [,4]      [,5]
-    ## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
-    ## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
-    ## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
-    ## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
-    ## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
-    ## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273
+## The morphospace:
+data(BeckLee_mat50)
+head(BeckLee_mat50[,1:5])
 
-    ## [1] 50 48
+dim(BeckLee_mat50)
+## The morphospace contains 50 taxa and has 48 dimensions (or axis)
 
-\
-\
-\
-\
+## A list of first and last occurrences data for some fossils
+data(BeckLee_ages)
+head(BeckLee_ages)
 
-    ##             FAD  LAD
-    ## Adapis     37.2 36.8
-    ## Asioryctes 83.6 72.1
-    ## Leptictis  33.9 33.3
-    ## Miacis     49.0 46.7
-    ## Mimotona   61.6 59.2
-    ## Notharctus 50.2 47.0
-
-\
-\
-`  `
+## And a phylogeny:
+data(BeckLee_tree)
+plot(BeckLee_tree, cex=0.7); axisPhylo(root=140)
+```
 
@@
 
-` `\
-\
-`  `
+```{r, eval = FALSE}
+if(!require(strap)) install.packages("strap")
+library(strap)
+geoscalePhylo(BeckLee_tree, cex.tip=0.7, cex.ts=0.6)
+```
 
@@
 
-\
-`          `\
-\
-\
-`     `\
-`           `\
-`     `
+```{r}
+## Creating the vector of time bins ages
+time_bins <- rev(seq(from = 0, to = 100, by = 20))
+
+## Splitting the morphospace using the time.series function
+binned_morphospace <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
+    method = "discrete", time = time_bins, inc.nodes = FALSE,
+    FADLAD = BeckLee_ages)
+```
 
@@
 
-\
+```{r}
+## The object class
+class(binned_morphospace)
 
-    ## [1] "dispRity"
+## What's in the object (messy version for the console!)
+str(binned_morphospace)
+names(binned_morphospace)
 
-\
-
-    ## List of 3
-    ##  $ data    :List of 5
-    ##   ..$ 100-80: num [1:8, 1:48] -0.739 -0.68 -0.511 -0.473 -0.459 ...
-    ##   .. ..- attr(*, "dimnames")=List of 2
-    ##   .. .. ..$ : chr [1:8] "Daulestes" "Bulaklestes" "Uchkudukodon" "Asioryctes" ...
-    ##   .. .. ..$ : NULL
-    ##   ..$ 80-60 : num [1:15, 1:48] -0.477 -0.473 -0.474 -0.532 -0.409 ...
-    ##   .. ..- attr(*, "dimnames")=List of 2
-    ##   .. .. ..$ : chr [1:15] "Kennalestes" "Asioryctes" "Ukhaatherium" "Cimolestes" ...
-    ##   .. .. ..$ : NULL
-    ##   ..$ 60-40 : num [1:13, 1:48] -0.165 0.254 0.622 0.753 0.671 ...
-    ##   .. ..- attr(*, "dimnames")=List of 2
-    ##   .. .. ..$ : chr [1:13] "Todralestes" "Pezosiren" "Tribosphenomys" "Paramys" ...
-    ##   .. .. ..$ : NULL
-    ##   ..$ 40-20 : num [1:6, 1:48] -0.121 -0.203 -0.158 0.696 0.355 ...
-    ##   .. ..- attr(*, "dimnames")=List of 2
-    ##   .. .. ..$ : chr [1:6] "Leptictis" "Dilambdogale" "Widanelfarasia" "Moeritherium" ...
-    ##   .. .. ..$ : NULL
-    ##   ..$ 20-0  : num [1:10, 1:48] 0.3079 0.6531 0.5089 -0.0419 0.235 ...
-    ##   .. ..- attr(*, "dimnames")=List of 2
-    ##   .. .. ..$ : chr [1:10] "Dasypodidae" "Bradypus" "Myrmecophagidae" "Potamogalinae" ...
-    ##   .. .. ..$ : NULL
-    ##  $ elements: chr [1:50] "Cimolestes" "Maelestes" "Batodon" "Bulaklestes" ...
-    ##  $ series  : chr [1:6] "discrete" "100-80" "80-60" "60-40" ...
-    ##  - attr(*, "class")= chr "dispRity"
-
-    ## [1] "data"     "elements" "series"
-
-\
-
-    ## 5 discrete series for 50 elements 
-    ## Series:
-    ## 100-80, 80-60, 60-40, 40-20, 20-0.
+## How it's printed by default
+binned_morphospace
+```
 
@@
 
-\
-`     `\
-\
-\
+```{r}
+## Bootstrapping each time series 100 times
+boot_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100)
 
-    ## [1] 6
+## Getting the minimum number of rows (=taxa) in the time seriess
+min(unlist(lapply(binned_morphospace$data, nrow)))
 
-\
-`     `\
-`      `\
-\
-\
-` `
+## Bootstrapping each time series 100 times and rarefying them 
+rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
+    rarefaction = 6)
 
-    ## [1] "dispRity"
-    ## [1] "dispRity"
+## Note that the output objects are both from the "dispRity" class ...
+class(boot_bin_morphospace); class(rare_bin_morphospace)
 
-\
-
-    ## Bootstrapped ordinated matrix with 50 elements 
-    ## Series:
-    ## 100-80, 80-60, 60-40, 40-20, 20-0.
-    ## Data was split using discrete method.
-    ## Data was bootstrapped 100 times, using the full bootstrap method.
-
-    ## Bootstrapped ordinated matrix with 50 elements 
-    ## Series:
-    ## 100-80, 80-60, 60-40, 40-20, 20-0.
-    ## Data was split using discrete method.
-    ## Data was bootstrapped 100 times, using the full bootstrap method.
-    ## Data was rarefied with a maximum of 6 elements
-
+## ... and display some more information on the previous step when printed:
+boot_bin_morphospace
+rare_bin_morphospace
+```
 Calculating disparity
@@
 
-\
-`     `\
-\
-\
-`     `\
-\
-\
-\
+```{r}
+## Disparity for the bootstrapped data
+boot_disparity <- dispRity(boot_bin_morphospace, metric = c(sum, variances))
 
-    ## Disparity measurements across 5 series for 50 elements 
-    ## Series:
-    ## 100-80, 80-60, 60-40, 40-20, 20-0.
-    ## Disparity calculated as: c(sum, variances) for 48 dimensions.
-    ## Data was split using discrete method.
-    ## Data was bootstrapped 100 times, using the full bootstrap method.
+## Disparity for the rarefied data
+rare_disparity <- dispRity(rare_bin_morphospace, metric = c(sum, variances))
 
-    ## Disparity measurements across 5 series for 50 elements 
-    ## Series:
-    ## 100-80, 80-60, 60-40, 40-20, 20-0.
-    ## Disparity calculated as: c(sum, variances) for 48 dimensions.
-    ## Data was split using discrete method.
-    ## Data was bootstrapped 100 times, using the full bootstrap method.
-    ## Data was rarefied with a maximum of 6 elements
+## Again, these objects are both from the "dispRity" class and just display
+## some information when called, note the results!
+boot_disparity
+rare_disparity
 
-\
-
-    ##   series  n observed  mean  2.5%   25%   75% 97.5%
-    ## 1 100-80  8    1.675 1.460 1.087 1.389 1.568 1.648
-    ## 2  80-60 15    1.782 1.678 1.538 1.631 1.728 1.792
-    ## 3  60-40 13    1.913 1.773 1.607 1.734 1.826 1.886
-    ## 4  40-20  6    2.022 1.673 1.212 1.537 1.822 1.942
-    ## 5   20-0 10    1.971 1.773 1.598 1.716 1.842 1.890
-
-    ##   series n observed  mean  2.5%   25%   75% 97.5%
-    ## 1 100-80 6       NA 1.440 1.067 1.339 1.575 1.726
-    ## 2  80-60 6       NA 1.670 1.276 1.614 1.798 1.922
-    ## 3  60-40 6       NA 1.763 1.458 1.704 1.846 1.986
-    ## 4  40-20 6     2.02 1.721 1.392 1.633 1.828 1.992
-    ## 5   20-0 6       NA 1.762 1.459 1.642 1.880 2.009
+## To display the actual results we can use the summary function
+summary(boot_disparity)
+summary(rare_disparity)
+```
 
@@
 
-\
+```{r}
+## See more options on the summary function:
+?summary.dispRity
+```
 
@@
 
-\
-`          `\
-\
-\
-`      `\
-`      `
+```{r}
+## Graphical options
+quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
+
+## Plotting the bootstrapped and rarefied results
+plot(boot_disparity, type = "continuous", main = "bootstrapped results")
+plot(rare_disparity, type = "continuous", main = "rarefied results")
+```
 
@@
 
-\
-`     `\
-`      `
+```{r}
+## testing the differences between bins in the bootstrapped dataset.
+test.dispRity(boot_disparity, test = wilcox.test, comparison = "sequential",
+    correction = "bonferroni")
 
-    ##                   W      p.value
-    ## 100-80 - 80-60  471 7.427563e-28
-    ## 80-60 - 60-40  1562 1.798899e-16
-    ## 60-40 - 40-20  6250 9.061511e-03
-    ## 40-20 - 20-0   3725 7.379715e-03
-
-\
-`     `\
-`      `
-
-    ##                   W      p.value
-    ## 100-80 - 80-60 1695 2.717026e-15
-    ## 80-60 - 60-40  3424 4.733196e-04
-    ## 60-40 - 40-20  5669 4.095267e-01
-    ## 40-20 - 20-0   4254 2.740887e-01
+## testing the differences between bins in the rarefied dataset.
+test.dispRity(rare_disparity, test = wilcox.test, comparison = "sequential",
+    correction = "bonferroni")
+```
 

### 2016-05-29 doc/dispRity-ecology-demo.Rmd
Manuals formating

@@
 
-This demo aims to give quick overview of the <span>`dispRity`
-</span> package (v.0.2) for ecological analysis. Please refer to [GitHub
-page:
-github.com/TGuillerme/dispRity](https://github.com/TGuillerme/dispRity)
-for other vignettes, namely the `dispRity` tutorial that
-explains the functions in more detail.
+This demo aims to give quick overview of the `dispRity` package (v.0.2) for ecological analysis.
+Please refer to [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the `dispRity` manual that explains the functions in more details.
 
-To keep it short, this package allows the use of all of the dimensions
-of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis
-rather than just a sub-set of dimensions. For example, one might want to
-know whether a specific kind of water treatment at certain water depth
-alters invertebrate communities and composition in natural habitats.
+To keep it short, this package allows the use of all of the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
+For example, one might want to know whether a specific kind of water treatment at certain water depth alters invertebrate communities and composition in natural habitats.
 
@@
 
-You can install this package easily if you are using the latest version
-of <span>`R` </span>and `devtools`.
+You can install this package easily if you are using the latest version of `R` (> 3.0.0) and `devtools`.
 
@@
 
-For this example with ecological data we are going to use data from
-McClean (unpubl.). This data is the ordination of a distance matrix
-based on nutrient enrichment and depth and freshwater benthic
-invertebrates.
+For this example with ecological data we are going to use data from McClean (unpubl.).
+This data is the ordination of a distance matrix based on nutrient enrichment and depth and freshwater benthic invertebrates.
 
@@
 
-A classical way to represent this ordinated data will be to use two
-dimensional plots.
+A classical way to represent this ordinated data will be to use two dimensional plots.
 
@@
 
-This shows the distribution of the experimental plots along the two
-first axis of variance of the ordinated distance matrix (i.e. the first
-two dimensions). At a first glance, it seems difficult to see a clear
-effect from the treatments (blue or red) or the depth (rounds or
-triangles) since the different experimental plots with the same
-parameters don’t seem to cluster. However, the problem, is that this
-plot ignores the 18 other dimensions of the ordination! Additionally,
-these two represented dimensions do not represent a biological reality
-*per se*; i.e. the values on the first dimension do not represent some
-continuous traits (e.g. depth) but just the ordinations of correlations
-between the data and some factors.
+This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix (i.e. the first two dimensions).
+At a first glance, it seems difficult to see a clear effect from the treatments (blue or red) or the depth (rounds or triangles) since the different experimental plots with the same parameters don’t seem to cluster.
+However, the problem, is that this plot ignores the 18 other dimensions of the ordination!
+Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e.
+the values on the first dimension do not represent some continuous traits (e.g.
+depth) but just the ordinations of correlations between the data and some factors.
 
-Therefore, one might want to approach this problem without getting stuck
-in only two dimensions and consider the whole dataset as a
-*n*-dimensional object.
+Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.
 
@@
 
-The first step in such analysis will be to create different series that
-are two sub-samples of the ordinated space (i.e. the *n*-dimensional
-object). Each of these series contain a certain number of elements (i.e.
-the 40 experimental field plots) that have some attributes in common. In
-our example, we are going to group the elements according to their depth
-and treatment.
+The first step in such analysis will be to create different series that are two sub-samples of the ordinated space (i.e. the *n*-dimensional object).
+Each of these series contain a certain number of elements (i.e. the 40 experimental field plots) that have some attributes in common.
+In our example, we are going to group the elements according to their depth and treatment.
 
@@
 
-Second, let’s split the data according to these factors to create the
-series of the ordinated space by using the `cust.series` function:
+Second, let’s split the data according to these factors to create the series of the ordinated space by using the `cust.series` function:
 
@@
 
-For more details on the `dispRity` objects, see the
-[<span>`dispRity`
-</span>manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf).
-Basically the idea is to avoid jamming the <span>`R` </span>console such
-as when using:
+For more details on the `dispRity` objects, see the [`dispRity` manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf).
+Basically the idea is to avoid jamming the `R` console such as when using:
 
@@
 
-Disparity can be calculated in many ways, and therefore, the
-`dispRity` function allows to measure disparity *as defined
-by the user*. For more details on disparity, see the [other
-vignettes](https://github.com/TGuillerme/dispRity).
+Disparity can be calculated in many ways, and therefore, the `dispRity` function allows to measure disparity *as defined by the user*.
+For more details on measuring disparity, see the [`dispRity` metric vignette](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.Rmd).
 
-In this example, we are going to define disparity as the median distance
-between the different experimental plots and the centroid of the
-ordinated space. High values of disparity will indicate a general high
-dispersal from this centroid (i.e. on average, the experimental plots
-are far apart in the ordinated space). We can define the metrics easily
-in the `dispRity` function by feeding them to the `metric`
-argument. Here we are going to feed the functions `stats::median` and
-`dispRity::centroids` which calculates distances between elements and
+In this example, we are going to define disparity as the median distance between the different experimental plots and the centroid of the ordinated space.
+High values of disparity will indicate a general high dispersal from this centroid (i.e. on average, the experimental plots are far apart in the ordinated space).
+We can define the metrics easily in the `dispRity` function by feeding them to the `metric` argument.
+Here we are going to feed the functions `stats::median` and `dispRity::centroids` which calculates distances between elements and
 centroid.
@@
 
-One might also want to bootstrap the data to test the robustness of the
-measured disparity to outliers. Also, as we can see, each series has
-different numbers of elements. It might be interesting to rarefy the
-data as well to have only series with the same number of elements. Both
-steps are easily doable through the `boot.matrix` function.
+One might also want to bootstrap the data to test the robustness of the measured disparity to outliers.
+Also, as we can see, each series has different numbers of elements.
+It might be interesting to rarefy the data as well to have only series with the same number of elements.
+Both steps are easily doable through the `boot.matrix` function.
 
@@
 
-We can now rerun a more robust disparity analysis using the bootstrapped
-data:
+We can now rerun a more robust disparity analysis using the bootstrapped data:
 
@@
 
-We can now summarise the data using various options such as the
-confidence intervals levels and the central tendency.
+We can now summarise the data using various options such as the confidence intervals levels and the central tendency.
 
@@
 
-We can also plot the results and have a look at the effect of the number
-of experimental plots:
+We can also plot the results and have a look at the effect of the number of experimental plots:
 
@@
 
-As we can see, there seems to be no strong effect of the number of
-experimental plots in each series (i.e. the rarefied plot is really
-similar to the bootstrapped plot) which is a good thing!
+As we can see, there seems to be no strong effect of the number of experimental plots in each series (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
 
@@
 
-Finally, we can test our hypothesis (whether the sort of water treatment
-at certain depth alters invertebrate communities and composition in
-natural habitats) by using the `test.dispRity` function.
+Finally, we can test our hypothesis (whether the sort of water treatment at certain depth alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.
 
@@
 
-As we can see, there is strong support for an effect of the treatment
-and the depth on the median distance between each experimental plots and
-the centroid of the ordinated space. In other words, it seems that with
-the second treatment and the second level of depth, the invertebrate
-communities and composition were further apart (i.e. more dispersed in
-the ordinated space) than with the first treatment and the first level
-of depth.
+As we can see, there is strong support for an effect of the treatment and the depth on the median distance between each experimental plots and the centroid of the ordinated space.
+In other words, it seems that with the second treatment and the second level of depth, the invertebrate communities and composition were further apart (i.e.
+more dispersed in the ordinated space) than with the first treatment and the first level of depth.

### 2016-05-29 doc/dispRity-manual.Rmd
Manuals formating

@@
 
-`dispRity` is a package for calculating disparity in `R`.
-It allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA)
-into single values.
+`dispRity` is a package for calculating disparity in `R`. It allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.
 
@@
 
-Because this package is aimed to be multidisciplinary, many names or
-terms used in this tutorial might be non-familiar to certain fields.
+Because this package is aimed to be multidisciplinary, many names or terms used in this tutorial might be non-familiar to certain fields.
 Here is a list of what are the exact meaning of these term:
 
--   **Ordinated space**: it designates the mathematical multidimensional
-    object studied here. In morphometrics, this one is often referred as
-    being the morphospace. However it can also be the the cladisto-space
-    for cladistic data or the eco-space in ecology, etc. In practice,
-    this term designates an ordinated matrix where the columns represent
-    the dimensions of the ordinated space (often – but not necessarily –
-    $>$ 3!) and the rows represent the elements within this space.
+-   **Ordinated space**: it designates the mathematical multidimensional object studied here.
+    In morphometrics, this one is often referred as being the morphospace.
+    However it can also be the the cladisto-space for cladistic data or the eco-space in ecology, etc.
+    In practice, this term designates an ordinated matrix where the columns represent the dimensions of the ordinated space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
 
--   **Elements**: it designates the rows of the ordinated space,
-    elements can be either taxa, field sites, countries, etc...
+-   **Elements**: it designates the rows of the ordinated space, elements can be either taxa, field sites, countries, etc...
 
--   **Dimensions**: it designates the columns of the ordinated space.
-    The dimensions can also be referred to as axis or
-    principal components.
+-   **Dimensions**: it designates the columns of the ordinated space. The dimensions can also be referred to as axis or principal components.
 
 -   **Series**: it designates sub-samples of the ordinated space.
-    Basically a series contain the same number of dimensions as the
-    morphospace but may contain a smaller number of elements. For
-    example, if our ordinated space is composed of birds and
-    mammals (i.e. the elements) and 50 dimensions, we can create two
-    series of just mammals or birds as elements (but with the same
-    50 dimensions) to look at the difference in disparity between the
-    two clades.
+    Basically a series contain the same number of dimensions as the morphospace but may contain a smaller number of elements.
+    For example, if our ordinated space is composed of birds and mammals (i.e.
+the elements) and 50 dimensions, we can create two series of just mammals or birds as elements (but with the same 50 dimensions) to look at the difference in disparity between the two clades.
 
@@
 
-You can install this package easily if you use the latest version of
-<span>`R` </span>and `devtools`.
+You can install this package easily if you use the latest version of `R` (> 3.0.0) and `devtools`.
 
@@
 
-Note that we use the `release` branch here which is version 0.2. For the
-piping-hot (but potentially full of bugs) version, you can change the
-argument `ref = release` to `ref = master`. This package depends mainly
-on the `ape` package and uses functions of several other (`ade4`,
-`geometry`, `grDevices`, `hypervolume`, `paleotree` and `snow`).
+Note that we use the `release` branch here which is version 0.2.
+For the piping-hot (but potentially full of bugs) version, you can change the argument `ref = release` to `ref = master`.
+This package depends mainly on the `ape` package and uses functions of several other (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree` and `snow`).
 
@@
 
-In this tutorial we are going to use a subset of the ordinated cladistic
-data from @beckancient2014 that contains 50 taxa (elements) ordinated
-using their cladistic distance (i.e. the distance between their discrete
-morphological characters). Note that this data is more oriented towards
-palaebiology analysis but that it can apply to other disciplines. Please
-refer to the [GitHub page:
-github.com/TGuillerme/dispRity](https://github.com/TGuillerme/dispRity)
-for other vignettes covering some specific example.
+In this tutorial we are going to use a subset of the ordinated cladistic data from @beckancient2014 that contains 50 taxa (elements) ordinated using their cladistic distance (i.e.
+the distance between their discrete morphological characters).
+Note that this data is more oriented towards palaebiology analysis but that it can apply to other disciplines.
+Please refer to the [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes covering some specific example.
 
@@
 
-Here is a really crude and quick analysis to go through the package,
-showing some of its features. Note that all these features will be
-discussed in more details below.
+Here is a really crude and quick analysis to go through the package, showing some of its features. Note that all these features will be discussed in more details below.
 
@@
 
-Disparity analysis can involve a lot of shuffling around with many
-matrices (especially when bootstrapping the data) which can be
-impractical to visualise and will quickly jam your <span>`R`
-</span>console. For example, we can have a look at the structure of the
-object created in the quick example:
+Disparity analysis can involve a lot of shuffling around with many matrices (especially when bootstrapping the data) which can be impractical to visualise and will quickly jam your `R` console.
+For example, we can have a look at the structure of the object created in the quick example:
 
@@
 
-Therefore this package proposes a specific class of object called
-`dispRity` objects. These objects allow to easily use S3
-method functions such as `summary.dispRity` (just called as `summary`;
-see section \[summary\]) or `plot.dispRity` (just called as `plot`; see
-section \[plot\]). But also, this allows to use the S3 method for
-printing `dispRity` objects via `print.dispRity` that
-allows to summarise the content of the objects similar to the `phylo`
-class objects (see `ape::print.phylo`).
+Therefore this package proposes a specific class of object called `dispRity` objects.
+These objects allow to easily use S3 method functions such as `summary.dispRity` (just called as `summary`; see section \[summary\]) or `plot.dispRity` (just called as `plot`; see section \[plot\]).
+But also, this allows to use the S3 method for printing `dispRity` objects via `print.dispRity` that allows to summarise the content of the objects similar to the `phylo` class objects (see `ape::print.phylo`).
 
@@
 ## This displays some basic information about the object content
-@
+```
 
-\noindent Note however, that it is always possible to recall the full object using the argument \texttt{all=TRUE}:
+Note however, that it is always possible to recall the full object using the argument \texttt{all=TRUE}:
 
-<<dispRity class3, eval=FALSE>>=
+```{r, eval=FALSE}
 ## Displaying the full object
@@
 
-Finally, some utility functions such as `get.dispRity` or
-`extract.dispRity` allows to access to some specific content of the
-object:
+Finally, some utility functions such as `get.dispRity` or `extract.dispRity` allows to access to some specific content of the object:
 
@@
 
-This package aims to be a modular package where users can personalise
-some aspects of the package fairly easily. The modular aspect allows the
-package users to use several implemented tools to help them creating
-personalised functions. For example, the <span>`dispRity`
-</span>function intake a `metric` argument designating how disparity
-should be calculated. This argument can take some already implemented
-functions such as `mean` but also some completely new ones such as the
-sum divided by length:
+This package aims to be a modular package where users can personalise some aspects of the package fairly easily.
+The modular aspect allows the package users to use several implemented tools to help them creating personalised functions.
+For example, the `dispRity` function intake a `metric` argument designating how disparity should be calculated.
+This argument can take some already implemented functions such as `mean` but also some completely new ones such as the sum divided by length:
 
@@
 
-For more information, please refer to the [GitHub page:
-github.com/TGuillerme/dispRity](https://github.com/TGuillerme/dispRity)
-for the vignette about the metric implementation in <span>`dispRity`
-</span>.
+For more information on the metrics, please refer to the [`dispRity` metric manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metric.Rmd) for the vignette about the metric implementation in `dispRity`.
 
@@
 
-In the following section, different functionalities are detailed. The
-functions are presented in a “classic” pipelined study fashion for
-convenience (i.e. splitting the data, bootstrapping it, calculating
-disparity, testing differences). Note however that most of these
-functions can be used in any order and that some steps can of course be
-skipped.
+In the following section, different functionalities are detailed.
+The functions are presented in a "classic" pipelined study fashion for convenience (i.e.
+splitting the data, bootstrapping it, calculating disparity, testing differences).
+Note however that most of these functions can be used in any order and that some steps can of course be skipped.
 
-One of the first steps in disparity analysis is to split the ordinated
-space into sub-samples of this space (hereafter *series*). These series
-correspond to regions of the ordinated space that we want to compare to
-each other.
+One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter *series*).
+These series correspond to regions of the ordinated space that we want to compare to each other.
 
@@
 
-The `cust.series` function allows to create these series according to
-factors determined by the user. In this example, we can split the matrix
-based on the phylogeny and separate to different groups of taxa: the
-crown mammals (all the living mammals and their direct ancestor) and the
-stem mammals (all the mammals that have no direct offspring).
+The `cust.series` function allows to create these series according to factors determined by the user.
+In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (all the living mammals and their direct ancestor) and the stem mammals (all the mammals that have no direct offspring).
 
@@
 
-Note that here we separating the taxa into two exclusive groups (i.e.
-crown mammals can be crown only, same for stem mammals). It is however
-possible to add more factors with overlapping groups:
+Note that here we separating the taxa into two exclusive groups (i.e. crown mammals can be crown only, same for stem mammals).
+It is however possible to add more factors with overlapping groups:
 
@@
 
-Another way to split the ordinated space (maybe more relevant to
-palaeobiologists) is to do it according to time. The `time.series`
-function allows to create series that contain all the elements present
-at specific points or during specific periods in time. This functions
-needs as input an ordinated space and a matching phylogenetic tree. Two
-types of time series can be performed by using the `method` option:
+Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
+The `time.series` function allows to create series that contain all the elements present at specific points or during specific periods in time.
+This functions needs as input an ordinated space and a matching phylogenetic tree.
+Two types of time series can be performed by using the `method` option:
 
@@
 
-2.  continuous time series (or time-slicing) using
-    `method = continuous`.
+2.  continuous time series (or time-slicing) using `method = continuous`.
 
-For the time-slicing method details see @GuillermeSTD. For both methods,
-the function intakes the `time` argument which can be a vector of
-`numeric` values for:
+For the time-slicing method details see @GuillermeSTD.
+For both methods, the function intakes the `time` argument which can be a vector of `numeric` values for:
 
@@
 
-Otherwise, the `time` argument can be set as a single `numeric` value
-for automatically generating a given number of equidistant
-time-bins/slices. Additionally, it is also possible to input a data
-frame containing the first and last occurrence data (`FAD/LAD`) for taxa
-that span over a longer time than the tips/nodes age.
+Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
+Additionally, it is also possible to input a data frame containing the first and last occurrence data (`FAD/LAD`) for taxa that span over a longer time than the tips/nodes age.
 
@@
 ```
-In this example, the taxa where split inside each timebin according to
-their age. Note however that some taxa might span between some time bins
-and should be included in more than one. This is possible by providing a
-table containing the first and last occurrence data:
+In this example, the taxa where split inside each timebin according to their age.
+Note however that some taxa might span between some time bins and should be included in more than one.
+This is possible by providing a table containing the first and last occurrence data:
 
@@
 
-This generated indeed a list of 3 sub matrices. Note that we can also
-generate equivalent results by just telling the function that we want
-three time-bins (series) as follow:
+This generated indeed a list of 3 sub matrices.
+Note that we can also generate equivalent results by just telling the function that we want three time-bins (series) as follow:
 
@@
 
-When using this method, the oldest boundary of the first bin (or the
-first slice, see below) is automatically generated as being the root age
-+ 1% of the tree length as long as at least three elements are present
-at that point in time. The algorithm adds an extra 1% tree length until
-reaching the required minimum of three elements. It is also possible to
-include nodes in each bin by using `inc.nodes = TRUE` and providing a
-matrix that contains the ordinated distance between tips *and* nodes.
+When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1% of the tree length as long as at least three elements are present at that point in time.
+The algorithm adds an extra 1% tree length until reaching the required minimum of three elements.
+It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance between tips *and* nodes.
 
-For the time-slicing method (`method = continuous`), the idea is really
-similar. This option intakes a matrix that contains the ordinated
-distance between taxa *and* nodes and an assumed evolutionary model via
-the `model` argument:
+For the time-slicing method (`method = continuous`), the idea is really similar.
+This option intakes a matrix that contains the ordinated distance between taxa *and* nodes and an assumed evolutionary model via the `model` argument:
 
-1.  `acctran` where the data chosen on each time slice is always the one
-    of the offspring
+1.  `acctran` where the data chosen on each time slice is always the one of the offspring
 
-2.  `deltran` where the data chosen on each time slice is always the one
-    of the descendant
+2.  `deltran` where the data chosen on each time slice is always the one of the descendant
 
-3.  `punctuated` where the data chosen on each time slice is randomly
-    chosen between the offspring or the descendant
+3.  `punctuated` where the data chosen on each time slice is randomly chosen between the offspring or the descendant
 
-4.  `gradual` where the data chosen on each time slice is either the
-    offspring or the descendant depending on branch length
+4.  `gradual` where the data chosen on each time slice is either the offspring or the descendant depending on branch length
 
@@
 
-Once we obtain our different series, we might want to bootstrap and
-rarefy it (i.e. pseudo-replicating the data). The bootstrap will allow
-us to make each series more robust to outliers and the rarefaction will
-allow us to compare slices with the same number of elements to get rid
-of eventual sampling problems. The `boot.matrix` allows to bootstraps
-and rarefy ordinated matrices in a fast and easy way. The default
-options will bootstrap the matrix 1000 times without rarefaction. The
-number of bootstrap pseudo-replicates can be defined using the
-`bootstraps` option (see below).
+Once we obtain our different series, we might want to bootstrap and rarefy it (i.e. pseudo-replicating the data).
+The bootstrap will allow us to make each series more robust to outliers and the rarefaction will allow us to compare slices with the same number of elements to get rid of eventual sampling problems.
+The `boot.matrix` allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
+The default options will bootstrap the matrix 1000 times without rarefaction.
+The number of bootstrap pseudo-replicates can be defined using the `bootstraps` option (see below).
 
@@
 
-This function allows to control the bootstrap algorithm through the
-`boot.type` argument. Currently two algorithms are implemented:
+This function allows to control the bootstrap algorithm through the `boot.type` argument.
+Currently two algorithms are implemented:
 
-1.  `full` where the bootstrapping is entirely stochastic (all the data
-    is bootstrapped)
+1.  `full` where the bootstrapping is entirely stochastic (all the data is bootstrapped)
 
-2.  `single` where only one random elements is replaced by one other
-    random elements each pseudo-replication
+2.  `single` where only one random elements is replaced by one other random elements each pseudo-replication
 
-This function also allows to rarefy the data using the `rarefaction`
-argument. The default argument is `FALSE` but it can be set to `TRUE` to
-fully rarefy the data (i.e. remove $n$ elements for the number of
-pseudo-replicates, where $n$ varies from the maximum number of elements
-present in the dataset to a minimum of 3 elements). It can also be set
-to a fix `numeric` value (or a set of `numeric` values). Finally, one
-last argument, `rm.last.axis` allows to remove a certain amount of
-dimensions (or axis) for the ordinated space. This can be logical
-argument where `FALSE` (default) will not remove any dimension and
-`TRUE` will remove the last dimensions that bear up to 5% of the total
-ordinated space’s variance.
+This function also allows to rarefy the data using the `rarefaction` argument.
+The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
+remove $n$ elements for the number of pseudo-replicates, where $n$ varies from the maximum number of elements present in the dataset to a minimum of 3 elements).
+It can also be set to a fix `numeric` value (or a set of `numeric` values).
+Finally, one last argument, `rm.last.axis` allows to remove a certain amount of dimensions (or axis) for the ordinated space.
+This can be logical argument where `FALSE` (default) will not remove any dimension and `TRUE` will remove the last dimensions that bear up to 5% of the total ordinated space’s variance.
 
@@
 
-Of course, one could be interested in directly supplying the sub-sampled
-matrices generated above directly to this function. In fact, it can also
-deal with a list of matrices or with a `dispRity` object
-output from the `cust.series` or `time.series` functions.
+Of course, one could be interested in directly supplying the sub-sampled matrices generated above directly to this function.
+In fact, it can also deal with a list of matrices or with a `dispRity` object output from the `cust.series` or `time.series` functions.
 
@@
 
-This function is a modular function that allows to simply (and quickly!)
-calculate disparity from a matrix.
+This function is a modular function that allows to simply (and quickly!) calculate disparity from a matrix.
 
-Because disparity can be measured in many ways, this function is a tool
-to measure disparity *as defined by the user*. In fact, the
-`dispRity` function intakes two main arguments: the data
-and the disparity metric. The disparity metric is a function or a set of
-functions that summarises the ordinated matrix to a single value that
-represents, in our example, the diversity of morphologies.
+Because disparity can be measured in many ways, this function is a tool to measure disparity *as defined by the user*.
+In fact, the `dispRity` function intakes two main arguments: the data and the disparity metric.
+The disparity metric is a function or a set of functions that summarises the ordinated matrix to a single value that represents, in our example, the diversity of morphologies.
 
-The `dispRity` algorithm decomposes the metrics functions
-into three levels that correspond to the dimensions of the output of
-each metric function. For more details on this algorithm please refer to
-the [metric vignette](https://github.com/TGuillerme/dispRity).
+The `dispRity` algorithm decomposes the metrics functions into three levels that correspond to the dimensions of the output of each metric function.
+For more details on this algorithm please refer to the [metric vignette](https://github.com/TGuillerme/dispRity).
 
-In practice, the `dispRity` function intakes one or more
-functions as a definition of disparity. Several of these functions will
-be already implemented in other packages (such as `stats::median`,
-`base::sum`, etc.); some others are implemented in this package (listed
-in `?dispRity.metric`) and finally some others will be defined by the
-users. The `make.metric` function is designed to help users create and
-test their own disparity metric functions. In practice, the use of these
-metrics in the `dispRity` function, is pretty easy:
+In practice, the `dispRity` function intakes one or more functions as a definition of disparity.
+Several of these functions will be already implemented in other packages (such as `stats::median`, `base::sum`, etc.); some others are implemented in this package (listed in `?dispRity.metric`) and finally some others will be defined by the users.
+The `make.metric` function is designed to help users create and test their own disparity metric functions. In practice, the use of these metrics in the `dispRity` function, is pretty easy:
 
@@
 
-Note that these functions do not directly output the disparity values
-but only the summary of the `dispRity` objects. To display
-the results, see the section \[summary\].
+Note that these functions do not directly output the disparity values but only the summary of the `dispRity` objects.
+To display the results, see the section \[summary\].
 
-In these examples we used only simple ordinated matrix but of course, it
-might be more interesting to directly use the <span>`dispRity`
-</span>objects we generated in the steps above. Thus we can calculate
-the bootstrapped and rarefied disparity in each sub-sample of the
-morphospace. In this example we are going to define disparity as being
-the median distance between each element and each dimensions centroids.
+In these examples we used only simple ordinated matrix but of course, it might be more interesting to directly use the `dispRity` objects we generated in the steps above.
+Thus we can calculate the bootstrapped and rarefied disparity in each sub-sample of the morphospace.
+In this example we are going to define disparity as being the median distance between each element and each dimensions centroids.
 
@@
 
-Note that the computational time was longer for the rarefied data
-(`crown_stemBS`) since it had to calculate the disparity on all the
-rarefied matrix, each bootstrapped 100 times. For both the `boot.matrix`
-and the `dispRity` functions, it is possible to increase
-computational speed in some cases using the `parallel` option (see each
-function’s manuals).
+Note that the computational time was longer for the rarefied data (`crown_stemBS`) since it had to calculate the disparity on all the rarefied matrix, each bootstrapped 100 times.
+For both the `boot.matrix` and the `dispRity` functions, it is possible to increase computational speed in some cases using the `parallel` option (see each function’s manuals).
 
@@
 
-This function is a S3 function (`summary.dispRity`) allowing to
-summarize the content of `dispRity` objects that contain
-disparity calculations. This function intakes a <span>`dispRity`
-</span>object plus various options namely the `quantile` values for the
-confidence intervals levels; the `cent.tend` for the central tendency to
-use for summarising the results and two *visual* options which are
-whether to recall the `dispRity` options and how much
-digits are wanted in the results.
+This function is a S3 function (`summary.dispRity`) allowing to summarize the content of `dispRity` objects that contain disparity calculations.
+This function intakes a `dispRity` object plus various options namely the `quantile` values for the confidence intervals levels; the `cent.tend` for the central tendency to use for summarising the results and two *visual* options which are whether to recall the `dispRity` options and how much digits are wanted in the results.
 
@@
 
-An alternative way to summarise the data is to plot the results using
-the S3 method `plot.dispRity`. This function intakes the same options as
-`summary.dispRity` along side with various graphical options described
-in the function manual (see `?plot.dispRity`).
+An alternative way to summarise the data is to plot the results using the S3 method `plot.dispRity`.
+This function intakes the same options as `summary.dispRity` along side with various graphical options described in the function manual (see `?plot.dispRity`).
 
-Two main categories of plots can be displayed: continuous plots with the
-argument `type = continuous` (see below) and discrete plots (default is
-`type = box`. Note that this argument can be left missing. In this case,
-the algorithm will automatically detect the type of series from the
-`dispRity` object.
+Two main categories of plots can be displayed: continuous plots with the argument `type = continuous` (see below) and discrete plots (default is `type = box`.
+Note that this argument can be left missing. In this case, the algorithm will automatically detect the type of series from the `dispRity` object.
 
@@
 
--   `type = polygon` for plotting boxes that represent the different
-    levels of confidence intervals (with a dot for the
-    central tendency).
+-   `type = polygon` for plotting boxes that represent the different levels of confidence intervals (with a dot for the central tendency).
 
--   `type = lines` which is similar to `type = polygon` but plots lines
-    rather than boxes for the confidence intervals levels.
+-   `type = lines` which is similar to `type = polygon` but plots lines rather than boxes for the confidence intervals levels.
 
-It is also possible to display the number of elements per series (has a
-horizontal doted line) using the option `elements = TRUE`. Additionally,
-when the data is rarefied (see section \[rare\]), one can also indicate
-which level of rarefaction to display (i.e. display the results for a
-certain number of elements) by using the `rarefaction` argument.
+It is also possible to display the number of elements per series (has a horizontal doted line) using the option `elements = TRUE`.
+Additionally, when the data is rarefied (see section \[rare\]), one can also indicate which level of rarefaction to display (i.e.
+display the results for a certain number of elements) by using the `rarefaction` argument.
 
@@
 
-It is also possible to easily look at the behaviour of the disparity
-metric with the rarefied data (i.e. how does the disparity score change
-when the number of elements goes down) by using the `rarefaction = plot`
-argument:
+It is also possible to easily look at the behaviour of the disparity metric with the rarefied data (i.e.
+how does the disparity score change when the number of elements goes down) by using the `rarefaction = plot` argument:
 
@@
 
-The different dashed lines represent the different confidence intervals
-around each rarefaction curve. We can see that the disparity metric
-becomes stable after more than 10 elements.
+The different dashed lines represent the different confidence intervals around each rarefaction curve. We can see that the disparity metric becomes stable after more than 10 elements.
 
-It is also possible, when using the `type = lines` argument, to plot
-lines rather than boxes. This option can be useful when displaying many
-series in a same window. Additionally, it is possible to plot the
-observed data when existing (i.e. not when the data is rarefied) by
-using the option `observed = TRUE`.
+It is also possible, when using the `type = lines` argument, to plot lines rather than boxes.
+This option can be useful when displaying many series in a same window.
+Additionally, it is possible to plot the observed data when existing (i.e.
+not when the data is rarefied) by using the option `observed = TRUE`.
 
@@
 
-The `type = continuous` argument plots superimposed continuous (along
-the x axis) polygons that represent the confidence intervals and a
-continuous line that represents the central tendency of the data. Note
-that, as a S3 method function, `plot.dispRity` can take the classic
-graphical options from the `plot` function or even to add other plots to
-the previous ones.
+The `type = continuous` argument plots superimposed continuous (along the x axis) polygons that represent the confidence intervals and a continuous line that represents the central tendency of the data.
+Note that, as a S3 method function, `plot.dispRity` can take the classic graphical options from the `plot` function or even to add other plots to the previous ones.
 
@@
 
-I encourage you to play with the graphical options to have some prettier
-results. Note that most of the options from `plot` can be passed to
-`plot.dispRity` via `...`.
+I encourage you to play with the graphical options to have some prettier results.
+Note that most of the options from `plot` can be passed to `plot.dispRity` via `...`.
 
@@
 
-Finally, the `dispRity` package allows to apply some tests
-to the data in order to properly test hypothesis. The function
-`test.dispRity` works in a similar way as the <span>`dispRity`
-</span>function: it intakes a `dispRity` object (containing
-disparity measurements), a `test` and a `comparisons` argument. The
-`test` argument can be any statistical or non-statistical test to apply
-to the disparity object. It can be common test function (e.g.
-`stats::t.test`) or some functions implemented in <span>`dispRity`
-</span>(see `dispRity.test`) or even user defined functions. The
-`comparisons` argument must indicate the way the test should be applied
-to the data:
+Finally, the `dispRity` package allows to apply some tests to the data in order to properly test hypothesis.
+The function `test.dispRity` works in a similar way as the `dispRity` function: it intakes a `dispRity` object (containing disparity measurements), a `test` and a `comparisons` argument.
+The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
+It can be common test function (e.g.
+`stats::t.test`) or some functions implemented in `dispRity` (see `dispRity.test`) or even user defined functions.
+The `comparisons` argument must indicate the way the test should be applied to the data:
 
@@
 
--   `all`: to compare all the series together (like in analysis
-    of variance)
+-   `all`: to compare all the series together (like in analysis of variance)
 
-It is also possible to input a list of pairs of `numeric` values or
-`characters` matching the series names to create personalised test. Some
-other specific tests implemented in `dispRity` such as the
-`dispRity::null.test` and the `dispRity::sequential.test` have a
-specific way to be applied to the data and therefore ignore the
-`comparisons` argument.
+It is also possible to input a list of pairs of `numeric` values or `characters` matching the series names to create personalised test.
+Some other specific tests implemented in `dispRity` such as the `dispRity::null.test` and the `dispRity::sequential.test` have a specific way to be applied to the data and therefore ignore the `comparisons` argument.
 
-This function also allows to correct for type I error inflation when
-using multiple comparisons via the `correction` argument. This argument
-can be empty (no correction applied) or can contain one of the
-corrections from the `stats::p.adjust` function (see `?p.adjust`).
+This function also allows to correct for type I error inflation when using multiple comparisons via the `correction` argument.
+This argument can be empty (no correction applied) or can contain one of the corrections from the `stats::p.adjust` function (see `?p.adjust`).
 
-Note that the `test.dispRity` algorithm deals with some classic tests
-outputs (`h.test`, `lm` and `numeric` vector) and summarises the test
-output. It is however possible to get the full detailed output by using
-the options `details = TRUE`.
+Note that the `test.dispRity` algorithm deals with some classic tests outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
+It is however possible to get the full detailed output by using the options `details = TRUE`.
 
@@
 
-This function allows to calculate the age of each individual nodes and
-tips in a tree. It can either use the root age of the tree (if present
-as `$root.time`) or else calculate the age using a user defined root age
-via the `age` argument. Also, it is possible to decide whether the time
-is calculated towards the past (e.g. million years ago) or towards the
-present (e.g. in time since the origin).
+This function allows to calculate the age of each individual nodes and tips in a tree.
+It can either use the root age of the tree (if present as `$root.time`) or else calculate the age using a user defined root age via the `age` argument.
+Also, it is possible to decide whether the time is calculated towards the past (e.g. million years ago) or towards the present (e.g. in time since the origin).
 
@@
 
-As stated at the start of the demo, this version 0.2 is still in
-development and many parts are missing. Here are the new functionalities
-that will be implemented in further versions.
+As stated at the start of the demo, this version 0.2 is still in development and many parts are missing.
+Here are the new functionalities that will be implemented in further versions.
 
@@
 
-I intend also develop functions to help users to develop their own
-algorithms for the bootstrap method (via `make.boot`) or the
-evolutionary models (via `make.model`). Both functions will provide
-similar testing as the `make.metric` function.
+I intend also develop functions to help users to develop their own algorithms for the bootstrap method (via `make.boot`) or the evolutionary models (via `make.model`).
+Both functions will provide similar testing as the `make.metric` function.
 
@@
 
-If you have any extra suggestions or comments on what has already been
-developed or will be developed, please send me an email
-(<guillert@tcd.ie>) or if you are a GitHub user, directly create an
-issue on the [GitHub page
-(github.com/TGuillerme/dispRity)](https://github.com/TGuillerme/dispRity).
+If you have any extra suggestions or comments on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).

### 2016-05-29 doc/dispRity-palaeo-demo.Rmd
Manuals formating

@@
 
-This demo aims to give quick overview of the <span>`dispRity`
-</span>package (v.0.2) for palaeobiology analysis. Please refer to
-[GitHub page:
-github.com/TGuillerme/dispRity](https://github.com/TGuillerme/dispRity)
-for other vignettes, namely the [<span>`dispRity`
-</span>manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf)
-that explains the functions in more detail.
+This demo aims to give quick overview of the `dispRity` package (v.0.2) for palaeobiology analysis.
+Please refer to [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the [`dispRity` manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.Rmd) that explains the functions in more detail.
 
-To keep it short, this package allows to easily perform
-**disparity-through-time** analysis. This type of analysis often starts
-by ordinating **morphometric** or **cladistic** data into a
-multidimensional object hereafter called the **morphospace**. One might
-be interested in studying how the occupancy of the morphospace changed
-through time by measuring a summary metric of this morphospace through
-time, called **disparity**. Disparity can then be seen as a value
-summarising the diversity of morphologies.
+To keep it short, this package allows to easily perform **disparity-through-time** analysis.
+This type of analysis often starts by ordinating **morphometric** or **cladistic** data into a multidimensional object hereafter called the **morphospace**.
+One might be interested in studying how the occupancy of the morphospace changed through time by measuring a summary metric of this morphospace through time, called **disparity**.
+Disparity can then be seen as a value summarising the diversity of morphologies.
 
-This demo showcases a typical disparity-through-time analyse: we are
-going to test whether the disparity changed through time in a subset of
-Eutheria (Mammalia) since the last 100 million years.
+This demo showcases a typical disparity-through-time analyse: we are going to test whether the disparity changed through time in a subset of Eutheria (Mammalia) since the last 100 million years.
 
@@
 
-You can install this package easily if you are using the latest version
-of <span>`R` </span>and `devtools`.
+You can install this package easily if you are using the latest version of `R` and `devtools`.
 
@@
 
-In this example, we are going to use a subset of the data from
-@beckancient2014 that contains an ordinated cladistic matrix and a
-phylogenetic tree containing 50 taxa along side with a table containing
-some first and last occurence data for several taxa. The ordinated
-matrix will represent our full morphospace, i.e. all the morphologies
-that ever existed through time (for this dataset).
+In this example, we are going to use a subset of the data from @beckancient2014 that contains an ordinated cladistic matrix and a
+phylogenetic tree containing 50 taxa along side with a table containing some first and last occurence data for several taxa.
+The ordinated matrix will represent our full morphospace, i.e. all the morphologies that ever existed through time (for this dataset).
 
@@
 
-Note that you can have an even nicer tree if you use the `strap`
-package:
+Note that you can have an even nicer tree if you use the `strap` package:
 
@@
 
-One of the crucial steps in disparity-through-time analysis is to split
-the full morphospace into smaller time seriess that contain the total
-amount of morphologies at certain points in time (time-slicing) or
-during certain periods in time (time-binning). Basically, the full
-morphospace in the previous step represents the total amount of
-morphologies in all time and is likely to be greater than any time
-series of the morphospace.
+One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time seriess that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
+Basically, the full morphospace in the previous step represents the total amount of morphologies in all time and is likely to be greater than any time series of the morphospace.
 
-The `dispRity` package provides a `time.series` function
-that allows to split the morphospace time slices (using
-`method = continuous`) or into time bins (using `method = discrete`). In
-this example, we are going to split the morphospace into 5 equal time
-bins of 20 million years long from 100 million years ago (Mya) to the
-present. Note that we will also provide to the function the table
-containing the first and last occurrences data for some fossils to take
-into account the fact that some fossils might span between our different
-time bins.
+The `dispRity` package provides a `time.series` function that allows to split the morphospace time slices (using `method = continuous`) or into time bins (using `method = discrete`).
+In this example, we are going to split the morphospace into 5 equal time bins of 20 million years long from 100 million years ago (Mya) to the present.
+Note that we will also provide to the function the table containing the first and last occurrences data for some fossils to take into account the fact that some fossils might span between our different time bins.
 
@@
 
-The output object is a `dispRity` object. For details about
-this object class, please refer to the [<span>`dispRity`
-</span>manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf).
-In brief, `dispRity` objects are lists of different
-elements (i.e. disparity results, morphospace time seriess, etc...) that
-display only a summary of the object when calling the object for
-avoiding jamming the <span>`R` </span>console.
+The output object is a `dispRity` object.
+For details about this object class, please refer to the [`dispRity` manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.Rmd).
+In brief, `dispRity` objects are lists of different elements (i.e.
+disparity results, morphospace time seriess, etc...) that display only a summary of the object when calling the object for avoiding jamming the `R` console.
 
@@
 
-Note that these objects will gradually contain more information when
-reaching the following steps in the disparity-through-time analysis.
+Note that these objects will gradually contain more information when reaching the following steps in the disparity-through-time analysis.
 
@@
 
-Once we obtain our different time series, we can bootstrap and rarefy
-them (i.e. pseudo-replicating the data). The bootstrapping will allow to
-make each series more robust to outliers and the rarefaction will allow
-us to compare slices with the same number of taxa to get rid of eventual
-sampling problems (i.e. more taxa in one bin than the other). The
-`boot.matrix` function that allows to bootstrap the <span>`dispRity`
-</span>object and the `rarefaction` option within allows to rarefy it.
+Once we obtain our different time series, we can bootstrap and rarefy them (i.e.
+pseudo-replicating the data).
+The bootstrapping will allow to make each series more robust to outliers and the rarefaction will allow us to compare slices with the same number of taxa to get rid of eventual sampling problems (i.e.
+more taxa in one bin than the other).
+The `boot.matrix` function that allows to bootstrap the `dispRity` object and the `rarefaction` option within allows to rarefy it.
 
@@
 
-We can now calculate the disparity within each time series along side
-with some confidence intervals generated by the pseudo-replication step
-above (bootstraps/rarefaction). Disparity can be calculated in many ways
-and this package allows user to come up with their own disparity
-metrics. For more info, please refer to the [<span>`dispRity`
-</span>manual and the metric
-vignette](https://github.com/TGuillerme/dispRity).
+We can now calculate the disparity within each time series along side with some confidence intervals generated by the pseudo-replication step above (bootstraps/rarefaction).
+Disparity can be calculated in many ways and this package allows user to come up with their own disparity metrics.
+For more info, please refer to the [`dispRity` metric vignette](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.Rmd).
 
-In this example, we are going to calculate the spread of the data in
-each time series by calculating the disparity as the sum of the variance
-of each dimension of the morphospace in each time series using the
-`dispRity` function. Thus, in this example, the disparity
-is defined by the multi-dimensional variance of each time series (i.e.
-the spread of the taxa within the morphospace). Note that this metric
-comes with a caveat not solved here since it ignores the covariance
-among the dimensions of the morphospace.
+In this example, we are going to calculate the spread of the data in each time series by calculating the disparity as the sum of the variance of each dimension of the morphospace in each time series using the `dispRity` function.
+Thus, in this example, the disparity is defined by the multi-dimensional variance of each time series (i.e. the spread of the taxa within the morphospace).
+Note that this metric comes with a caveat not solved here since it ignores the covariance among the dimensions of the morphospace.
 
@@
 
-As for any <span>`R` </span>package, you can refer yourself to each
-individual function for more details.
+As for any `R` package, you can refer yourself to each individual function for more details.
 
@@
 
-It is sometimes easier to visualise the results in a plot than in a
-table. For that we can use the `plot` function to plot the
-`dispRity` objects.
+It is sometimes easier to visualise the results in a plot than in a table.
+For that we can use the `plot` function to plot the `dispRity` objects.
 
@@
 
-As we can see in this case, the rarefaction tends to increase the
-confidence intervals around the mean.
+As we can see in this case, the rarefaction tends to increase the confidence intervals around the mean.
 
@@
 
-Finally, to draw some conclusions from these results, we need to apply
-some statistical testing. As stated before, we wanted to test whether
-the disparity changed through time in this dataset since the last 100
-million years. To do so, we can apply some comparisons of the means
-between each time-bins in a sequential manner to see whether the
-disparity in bin X is equal to the disparity in bin Y which is in turn
-equal to the disparity in bin Z, etc. Because our data is temporally
-auto-correlated (i.e. what happens in bin Y depends on what happened in
-bin X) and was pseudo-replicated (i.e. each bootstrap draw creates
-non-independent time seriess because they are all based on the same time
-series), we are going to apply a non-parametric mean comparison: the
-`wilcox.test`. Also, we are going to apply a p-value correction to avoid
-type I error inflation due to performing multiple test for the same
-hypothesis (see `?p.adjust` for more details).
+Finally, to draw some conclusions from these results, we need to apply some statistical testing.
+As stated before, we wanted to test whether the disparity changed through time in this dataset since the last 100 million years.
+To do so, we can apply some comparisons of the means between each time-bins in a sequential manner to see whether the disparity in bin X is equal to the disparity in bin Y which is in turn equal to the disparity in bin Z, etc.
+Because our data is temporally auto-correlated (i.e.
+what happens in bin Y depends on what happened in bin X) and was pseudo-replicated (i.e.
+each bootstrap draw creates non-independent time seriess because they are all based on the same time series), we are going to apply a non-parametric mean comparison: the `wilcox.test`.
+Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis (see `?p.adjust` for more details).
 
@@
 
-Here our results show significant changes in disparity through time
-between all time bins during the last 100 million years. However, when
-looking at the rarefied results, there is no significant difference
-between the time bins in the Cenozoic (60-40, 40-20 and 20-0 Mya),
-suggesting that the differences detected in the first test might just be
-due to the differences in number of taxa sample (respectively 13, 6 and
-10 in each time bin).
+Here our results show significant changes in disparity through time between all time bins during the last 100 million years.
+However, when looking at the rarefied results, there is no significant difference between the time bins in the Cenozoic (60-40, 40-20 and 20-0 Mya), suggesting that the differences detected in the first test might just be due to the differences in number of taxa sample (respectively 13, 6 and 10 in each time bin).

### 2016-06-10 doc/dispRity-ecology-demo.Rmd
test-disparity problem when distributions are equal
Updated manuals

@@
 
-```{r}
-##<!--, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>= -->
+```{r, fig.width=6, fig.height=6}
 ## The x and y axis represent the two first dimensions (i.e. principal
@@
 ```
-
-![image](figure/minimal-PCAplot-1){width=".8\linewidth"}
 
@@
 
-```{r}
+```{r, fig.width=6, fig.height=6}
 ## Graphical options
@@
 ```
-
-![image](figure/minimal-rare_plots-1){width="1\linewidth"}
 

### 2016-06-10 doc/dispRity-manual.Rmd
Updated manuals

@@
 
-```{r}
+```{r, fig.width=6, fig.height=6}
 ## Loading the package
@@
 
-![image](figure/minimal-data-1){width=".8\linewidth"}
-
 A quick go through
@@
 
-```{r}
+```{r, fig.width=6, fig.height=6}
 ## Splitting the data
@@
 ```
-
-![image](figure/minimal-Quick_go-1){width=".5\linewidth"}
 
@@
 
-![image](figure/minimal-plot1-1){width="1\linewidth"}
-
 It is also possible to easily look at the behaviour of the disparity metric with the rarefied data (i.e.
@@
 ```
-
-![image](figure/minimal-plot2-1){width="0.5\linewidth"}
 
@@
 
-![image](figure/minimal-plot3-1){width=".8\linewidth"}
-
 The `type = continuous` argument plots superimposed continuous (along the x axis) polygons that represent the confidence intervals and a continuous line that represents the central tendency of the data.
@@
 
-![image](figure/minimal-plot4-1){width=".5\linewidth"}
-
 I encourage you to play with the graphical options to have some prettier results.
@@
 
-![image](figure/minimal-plot5-1){width=".8\linewidth"}
-
 Finally, it is possible to add multiple plots on top of each other:
 
-```{r}
+```{r, fig.width=6, fig.height=6}
 ## Graphical options
@@
 ```
-
-![image](figure/minimal-plot5_1-1){width=".5\linewidth"}
 
@@
 ## Measuring the overlap between distributions in the time bins (using the
-## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff) with a
-## personalised series comparisons and some additional arguments to pass to the
-## test function (bw = bw.nrd)
-test.dispRity(disp_time_binsBS, test = bhatt.coeff, bw = bw.nrd,
-    comparisons = list(c(1,2), c(1,3), c(2,1), c(3,1)))
+## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)
+test.dispRity(disp_time_binsBS, test = bhatt.coeff)
 

### 2016-06-15 doc/dispRity-palaeo-demo.Rmd

@@
 
-```{r}
+```{r, fig.width=6, fig.height=6}
 ## Loading demo and the package data
@@
 ```
-
-![image](figure/minimal-data-1){width=".8\linewidth"}
 
@@
 ```
-
-![image](figure/minimal-plot-1){width="1\linewidth"}
 

### 2016-09-06 doc/dispRity-simulate_data.Rmd
simulate data vignette

---
title: "dispRity simulate data"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: References.bib
bst: sysbio.bst
vignette: >
  %\VignetteIndexEntry{dispRity metric}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette details how to simulate morphological data in `dispRity` (v.0.2.1).

Simulating morphological data in `dispRity`
=================

Before starting
===============

Installing `dispRity` 
-----------------------------------

You can install this package easily if you are using the latest version of `R` and `devtools`.

```{r, eval = FALSE}
if(!require(devtools)) install.packages("devtools")
install_github("TGuillerme/dispRity", ref = "release")
library(dispRity)
```

Quick go through
=================================
This is a quick example on how it works

```{r}
library(dispRity)
set.seed(123)
## Setting the starting tree with 15 taxa as a random coalescent tree
my_tree <- rcoal(15)

## Generating a matrix with 100 characters (85% binary and 15% three states) and
## an equal rates model with a gamma rates distribution (0.5, 1) with no 
## invariant characters.
my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)

## The head of the matrix
my_matrix[1:5, 1:10]

## Checking the matrix properties with a quick Maximum Parsimony tree search
check.morpho(my_matrix, my_tree)
```

In details
=================================
Basically, `sim.morpho` is really flexible and intakes a lot of different arguments to allow to simulate realistic matrix.
It has two implemented models: `"ER"` for Equal Rates (M*k* model) or the `"HKY"` one which is the molecular HKY model but transforms pyrines in 0's and purimidines in 1's.
It can also intake some specific distribution for the rates or the substitutions, allowing to modify these basic models.

The `check.morpho` runs a quick MP tree using the `phangorn` parsimony algorithm.
It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g. the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the MP tree and the provided tree.

Good parameters for a good matrix
-----------------------------------

I was testing the effect of different parameters on the ability to generate a ``realistic'' matrix (meaning not to different from the input tree with a consistency and retention index close to what's in the literature).
The matrix seems the most ``realistic'' with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three states characters, a gamma distribution with a shape parameter ($\alpha$) of 0.5 and no scaling ($\beta$ = 1).

Something like that should give a decent matrix:

```{r}
set.seed(1)
## tree
my_tree <- rcoal(15)
## matrix
morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
    rates = c(rgamma, 0.5, 1), invariant = TRUE)
check.morpho(morpho_mat, my_tree)
```

### 2016-11-28 doc/dispRity-manual.Rmd

@@
 
-Glossary
+Glossary <a name="glossary"></a>
 --------
@@
 
-Note that we use the `release` branch here which is version 0.2.
+Note that we use the `release` branch here which is version `0.2`.
 For the piping-hot (but potentially full of bugs) version, you can change the argument `ref = release` to `ref = master`.
@@
 data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.8) ; nodelabels(cex = 0.8) ; axisPhylo(root = 140)
+plot(BeckLee_tree, cex = 0.8) ; nodelabels(cex = 0.7) ; axisPhylo(root = 140)
 ```
@@
 
-```{r}
+```{r, eval = FALSE}
 str(median_centroids)
-## That's a more than 4000 lines of output!
+## That's a more than 5000 lines of output!
 ```
@@
 Therefore this package proposes a specific class of object called `dispRity` objects.
-These objects allow to easily use S3 method functions such as `summary.dispRity` (just called as `summary`; see section \[summary\]) or `plot.dispRity` (just called as `plot`; see section \[plot\]).
+These objects allow to easily use S3 method functions such as `summary.dispRity` (just called as `summary`; see [summary](#summary) section) or `plot.dispRity` (just called as `plot`; see [plot](#plot) section).
 But also, this allows to use the S3 method for printing `dispRity` objects via `print.dispRity` that allows to summarise the content of the objects similar to the `phylo` class objects (see `ape::print.phylo`).
+The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).
 
@@
 
-Note however, that it is always possible to recall the full object using the argument \texttt{all=TRUE}:
+Note however, that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:
 
@@
 print(median_centroids, all = TRUE)
-## But that's 155091 lines!
+## But that's more than 155000 lines!
 ```
 
-Finally, some utility functions such as `get.dispRity` or `extract.dispRity` allows to access to some specific content of the object:
+Some utility functions such as `sort.dispRity`, `get.dispRity` or `extract.dispRity` allows to access or reorder to some specific content of the object:
 
 ```{r}
+## Reordering the series (from the present to the past)
+sort.dispRity(median_centroids, decreasing = TRUE)
+
 ## Extracting some specific series from the disparity object
-series_1_and_4 <- get.dispRity(median_centroids, what = c(1,4))
-series_1_and_4
+get.dispRity(median_centroids, what = c(1,4))
 
-## The observed disparity
-extract.dispRity(median_centroids)
+## Extracting the observed disparity
+extract.dispRity(median_centroids, observed = TRUE)
 
@@
 
-This package aims to be a modular package where users can personalise some aspects of the package fairly easily.
+This package aims to be a *modular package*, that is, it is designed to be fairly easily modified by users.
 The modular aspect allows the package users to use several implemented tools to help them creating personalised functions.
 For example, the `dispRity` function intake a `metric` argument designating how disparity should be calculated.
-This argument can take some already implemented functions such as `mean` but also some completely new ones such as the sum divided by length:
+This argument can take some already implemented functions such as `mean` but it can also intake some completely new and revolutionary ones such as the sum divided by length:
 
@@
 summary(dispRity(sliced_data, metric = sum.divide.by.length))
-## Giving the exact same results!
+## That is a bit like reinventing the wheel...
 ```
@@
 In the following section, different functionalities are detailed.
-The functions are presented in a "classic" pipelined study fashion for convenience (i.e.
-splitting the data, bootstrapping it, calculating disparity, testing differences).
-Note however that most of these functions can be used in any order and that some steps can of course be skipped.
+The functions are presented in a "classic" pipelined study fashion for convenience (i.e. splitting the data, bootstrapping it, calculating disparity, testing differences).
+Note however that most of these functions can be used in any order and that some steps can of course be skipped and that the package contains many more additional functions!
 
-One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter *series*).
+One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter called *series*, see [glossary](#glossary)).
 These series correspond to regions of the ordinated space that we want to compare to each other.
@@
 The `cust.series` function allows to create these series according to factors determined by the user.
-In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (all the living mammals and their direct ancestor) and the stem mammals (all the mammals that have no direct offspring).
+In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (i.e.
+all the living mammals and their direct ancestor) and the stem mammals (i.e.
+all the mammals that have no related living descendants).
 
@@
 crown_stem <- cust.series(BeckLee_mat50, factors)
+```
 
-## This created a dispRity object containing two series: crown and stem
-class(crown_stem)
+The results of this function (like many others presented below) output a `dispRity` object:
+
+```{r}
 crown_stem
@@
 
-Note that here we separating the taxa into two exclusive groups (i.e. crown mammals can be crown only, same for stem mammals).
+Note that here we are separating the taxa into two exclusive groups (i.e. crown mammals can be crown only, same for stem mammals).
 It is however possible to add more factors with overlapping groups:
@@
 
-1.  discrete time series (or time-binning) using `method = discrete`;
+ *  discrete time series (or time-binning) using `method = discrete`;
 
-2.  continuous time series (or time-slicing) using `method = continuous`.
+ *  continuous time series (or time-slicing) using `method = continuous`.
 
@@
 
-1.  defining the boundaries of the time bins (when `method = discrete`);
+ *  defining the boundaries of the time bins (when `method = discrete`);
 
-2.  defining the time slices (when `method = continuous`).
+ *  defining the time slices (when `method = continuous`).
 
 Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
-Additionally, it is also possible to input a data frame containing the first and last occurrence data (`FAD/LAD`) for taxa that span over a longer time than the tips/nodes age.
+Additionally, it is also possible to input a data frame containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age.
 
@@
 ```
-In this example, the taxa where split inside each timebin according to their age.
+
+In this example, the taxa where split inside each time-bin according to their age.
 Note however that some taxa might span between some time bins and should be included in more than one.
@@
 ## taxa that might span between time bins.
-time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
-    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
-time_bins
+(time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
+    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages))
 
@@
 
-1.  `acctran` where the data chosen on each time slice is always the one of the offspring
+ *  `acctran` where the data chosen on each time slice is always the one of the offspring
 
-2.  `deltran` where the data chosen on each time slice is always the one of the descendant
+ *  `deltran` where the data chosen on each time slice is always the one of the descendant
 
-3.  `punctuated` where the data chosen on each time slice is randomly chosen between the offspring or the descendant
+ *  `punctuated` where the data chosen on each time slice is randomly chosen between the offspring or the descendant
 
-4.  `gradual` where the data chosen on each time slice is either the offspring or the descendant depending on branch length
+ *  `gradual` where the data chosen on each time slice is either the offspring or the descendant depending on branch length
 
@@
 ## Generating four time slices every 40 million years assuming a gradual
-## evolution model
-time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
+## evolutionary model
+(time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
     method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-time_slices
+    FADLAD = BeckLee_ages))
 ## Note that in the same way as for the discrete method, we can also
@@
 
-`boot.matrix` {#rare}
+`boot.matrix`
 -------------
 
-Once we obtain our different series, we might want to bootstrap and rarefy it (i.e. pseudo-replicating the data).
+Once we obtain our different series, it is possible to bootstrap and rarefy it (i.e. pseudo-replicating the data).
 The bootstrap will allow us to make each series more robust to outliers and the rarefaction will allow us to compare slices with the same number of elements to get rid of eventual sampling problems.
@@
 boot.matrix(data = BeckLee_mat50)
-## As we can see, the output is also a dispRity object that is summarised
-## automatically, and gives information on the data as well as the number of
-## bootstraps and the bootstraps methods.
 ```
@@
 
-1.  `full` where the bootstrapping is entirely stochastic (all the data is bootstrapped)
+ *  `full` where the bootstrapping is entirely stochastic (all the data is bootstrapped)
 
-2.  `single` where only one random elements is replaced by one other random elements each pseudo-replication
+ *  `single` where only one random elements is replaced by one other random elements each pseudo-replication
 
@@
 The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove $n$ elements for the number of pseudo-replicates, where $n$ varies from the maximum number of elements present in the dataset to a minimum of 3 elements).
+remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in the dataset to a minimum of 3 elements).
 It can also be set to a fix `numeric` value (or a set of `numeric` values).
@@
 This can be logical argument where `FALSE` (default) will not remove any dimension and `TRUE` will remove the last dimensions that bear up to 5% of the total ordinated space’s variance.
+Note that it can also intake a `numerical` value designating the proportion of axis to keep.
 
@@
 ## Bootstrap and full rarefaction on the crown/stem series
-crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
+(crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE))
 
 ## Bootstrap on the time binning/slicing series
-time_binsBS <- boot.matrix(time_bins, bootstraps = 100)
-time_slicesBS <- boot.matrix(time_slices, bootstraps = 100)
-
-## Note that all these objects are of class dispRity
-crown_stemBS
-time_binsBS
-time_slicesBS
+(time_binsBS <- boot.matrix(time_bins, bootstraps = 100))
+(time_slicesBS <- boot.matrix(time_slices, bootstraps = 100))
 ```
 
-`dispRity` {#disparity}
+`dispRity`
 ----------
@@
 In practice, the `dispRity` function intakes one or more functions as a definition of disparity.
-Several of these functions will be already implemented in other packages (such as `stats::median`, `base::sum`, etc.); some others are implemented in this package (listed in `?dispRity.metric`) and finally some others will be defined by the users.
-The `make.metric` function is designed to help users create and test their own disparity metric functions. In practice, the use of these metrics in the `dispRity` function, is pretty easy:
+Several of these functions will be already implemented in other packages (such as `stats::median`, `base::sum`, etc.); some others are implemented in this package (listed in `?dispRity.metric`) and finally some others can be defined by the users.
+The `make.metric` function is designed to help users create and test their own disparity metric functions.
+In practice, the use of these metrics in the `dispRity` function, is pretty easy:
 
@@
 
-Note that these functions do not directly output the disparity values but only the summary of the `dispRity` objects.
-To display the results, see the section \[summary\].
+These functions do not directly output the disparity values but only the summary of the `dispRity` objects.
+To display the results, see the section [summary](#summary).
 
@@
 Thus we can calculate the bootstrapped and rarefied disparity in each sub-sample of the morphospace.
-In this example we are going to define disparity as being the median distance between each element and each dimensions centroids.
+Here we are going to define disparity as being the median distance between each element and each dimensions centroids.
 
@@
 
-`summary`
+`summary` <a name="summary"></a>
 ---------
@@
 
-## If on happens to forgot what was calculated in the summarised object,
+## If one happens to forgot what was calculated in the summarised object,
 ## it is possible to recall the different steps details using recall:
@@
 
-`plot`
+`plot` <a name="plot"></a>
 ------
@@
 ```{r, fig.width=6, fig.height=6}
-## Graphical options
-quartz(width = 15, height = 5) ; par(mfrow = (c(1,3)), bty = "n")
+## Graphical parameters
+op <- par(mfrow = c(2,2), bty = "n")
 
@@
 ## Rarefied version of the same plot (20 elements per series)
-plot(disp_crown_stemBS, type = "polygon", elements = TRUE, rarefaction = 20)
+plot(disp_crown_stemBS, type = "polygon", rarefaction = 20)
 ```
@@
 
--   `pairwise` (default): to compare each series pairwise
+ *   `pairwise` (default): to compare each series pairwise
 
--   `referential`: to compare each series to the first one
+ *  `referential`: to compare each series to the first one
 
--   `sequential`: to compare each series to the following one
+ *  `sequential`: to compare each series to the following one
 
--   `all`: to compare all the series together (like in analysis of variance)
+ *  `all`: to compare all the series together (like in analysis of variance)
 

### 2016-11-28 doc/dispRity-metrics.Rmd

@@
 # A two dimensional disparity
-all_variances <- summary(dispRity(BeckLee_mat50, metric = variances))
+all_variances <- dispRity(BeckLee_mat50, metric = variances)
 
 # Transforming the two dimensional one in a one dimensional one
-sum_variances2 <- summary(dispRity(BeckLee_mat50, metric = sum))
+sum_variances2 <- summary(dispRity(all_variances, metric = sum))
 

### 2016-12-29 doc/dispRity-manual.Rmd
Partially updated manual

@@
 `dispRity` is a package for calculating disparity in `R`. It allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.
+<!-- Add version and table of content here! -->
 
@@
 
-Because this package is aimed to be multidisciplinary, many names or terms used in this tutorial might be non-familiar to certain fields.
+Because this package is aimed for multidisciplinary purpose, some names or terms used in this tutorial might be non-familiar to certain fields.
 Here is a list of what are the exact meaning of these term:
 
--   **Ordinated space**: it designates the mathematical multidimensional object studied here.
+-   **Ordinated space**: designates the mathematical multidimensional object that will be analysed with this package.
     In morphometrics, this one is often referred as being the morphospace.
     However it can also be the the cladisto-space for cladistic data or the eco-space in ecology, etc.
-    In practice, this term designates an ordinated matrix where the columns represent the dimensions of the ordinated space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
+    In practice, this term designates an ordinated matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
 
--   **Elements**: it designates the rows of the ordinated space, elements can be either taxa, field sites, countries, etc...
+-   **Elements**: designates the rows of the ordinated space, elements can be either taxa, field sites, countries, etc...
 
--   **Dimensions**: it designates the columns of the ordinated space. The dimensions can also be referred to as axis or principal components.
+-   **Dimensions**: designates the columns of the ordinated space. The dimensions can also be referred to as axis or principal components in the case of the ordinated space obtained from a PCA.
 
 -   **Series**: it designates sub-samples of the ordinated space.
-    Basically a series contain the same number of dimensions as the morphospace but may contain a smaller number of elements.
+    Basically a series contain the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
     For example, if our ordinated space is composed of birds and mammals (i.e.
@@
 
-Note that we use the `release` branch here which is version `0.2`.
+Note that we use the `release` branch here which is version `0.2`. <!-- add version !-->
 For the piping-hot (but potentially full of bugs) version, you can change the argument `ref = release` to `ref = master`.
-This package depends mainly on the `ape` package and uses functions of several other (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree` and `snow`).
+This package depends mainly on the `ape` package and uses functions of several other (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, and `RCurl`).
 
@@
 
-In this tutorial we are going to use a subset of the ordinated cladistic data from @beckancient2014 that contains 50 taxa (elements) ordinated using their cladistic distance (i.e.
-the distance between their discrete morphological characters).
+In this tutorial we are going to use a subset of the ordinated cladistic data from @beckancient2014 <!-- fix citation -->
+that contains 50 taxa (elements) ordinated using their cladistic distance (i.e. the distance between their discrete morphological characters).
 Note that this data is more oriented towards palaebiology analysis but that it can apply to other disciplines.
@@
 data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.8) ; nodelabels(cex = 0.7) ; axisPhylo(root = 140)
+plot(BeckLee_tree, cex = 0.8) ; axisPhylo(root = 140) ; nodelabels(cex = 0.5)
 ```
+<a name="mammal_tree"></a>
 
@@
 
-Here is a really crude and quick analysis to go through the package, showing some of its features. Note that all these features will be discussed in more details below.
+Here is a really crude and quick analysis to go through the package, showing some of its features.
+Note that all these features will be discussed in more details below.
 
 ```{r, fig.width=6, fig.height=6}
-## Splitting the data
+## Splitting the data into time series
 sliced_data <- time.series(BeckLee_mat99, BeckLee_tree, method = "continuous",
@@
 
-## Bootstrapping the data
+## Bootstrapping the data to measure its variance
 bootstrapped_data <- boot.matrix(sliced_data, 100)
 
-## Calculating disparity
+## Calculating disparity as the median distance from each element to the groups centroids
 median_centroids <- dispRity(bootstrapped_data, metric = c(median, centroids))
 
-## Summarising the results
+## Summarising the disparity results
 summary(median_centroids)
@@
 
-## Testing the effect of time on disparity
+## Testing the effect of time on disparity (note that using an ANOVA violates some assumptions!)
 summary(test.dispRity(median_centroids, test = aov, comparisons = "all"))
@@
 str(median_centroids)
-## That's a more than 5000 lines of output!
+## That's a more than 72 lines of lists of lists of matrices!
 ```
@@
 The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).
+
+<!-- TODO: change to html link -->
 
@@
 
-## Summarising it using the S3 method print.dispRity:
+## Summarising it using the S3 method print.dispRity
 median_centroids
-## This displays some basic information about the object content
 ```
@@
 print(median_centroids, all = TRUE)
-## But that's more than 155000 lines!
+## But that's more than ~ 2500 lines on a 13 inch laptop screen!
 ```
 
-Some utility functions such as `sort.dispRity`, `get.dispRity` or `extract.dispRity` allows to access or reorder to some specific content of the object:
-
-```{r}
-## Reordering the series (from the present to the past)
-sort.dispRity(median_centroids, decreasing = TRUE)
-
-## Extracting some specific series from the disparity object
-get.dispRity(median_centroids, what = c(1,4))
-
-## Extracting the observed disparity
-extract.dispRity(median_centroids, observed = TRUE)
-
-## The distribution of the bootstrapped disparity scores for the first series
-str(extract.dispRity(get.dispRity(median_centroids, what = 1),
-    observed = FALSE))
-```
+The package provides some utilities functions directly associated to the `dispRity` object manipulation such as `sort.dispRity`, `extract.dispRity` or `sort.dispRity` (see the full list in the [`dispRity` utilities](#dispRity.utilities) section).
+These functions come in handy for modifying the `dispRity` object without having to go into it's structure!
 
@@
 
-This package aims to be a *modular package*, that is, it is designed to be fairly easily modified by users.
+This package aims to be a *modular package*.
+Its architecture is designed to be fairly easily modified by users.
 The modular aspect allows the package users to use several implemented tools to help them creating personalised functions.
-For example, the `dispRity` function intake a `metric` argument designating how disparity should be calculated.
-This argument can take some already implemented functions such as `mean` but it can also intake some completely new and revolutionary ones such as the sum divided by length:
+For example, the `dispRity` function intake a `metric` argument that is directly the function for calculating disparity should.
+This argument can take some already implemented functions such as `mean` but it can also intake some completely new ones such as the sum divided by length (even though this example sounds like reinventing the wheel):
 
@@
 
-## A function for measuring the sum divided by the length (the mean!)
-sum.divide.by.length <- function(X) sum(X)/length(X)
+## A function for measuring the sum divided by the length
+sum.divided.by.length <- function(X) sum(X)/length(X)
 
 ## Disparity measured as the mean divided by the length
-summary(dispRity(sliced_data, metric = sum.divide.by.length))
-## That is a bit like reinventing the wheel...
+summary(dispRity(sliced_data, metric = sum.divided.by.length))
 ```
@@
 
+<!-- TODO: change to html link -->
+
 Functions
@@
 
-In the following section, different functionalities are detailed.
+In the following section, different functionalities of the package are detailed.
 The functions are presented in a "classic" pipelined study fashion for convenience (i.e. splitting the data, bootstrapping it, calculating disparity, testing differences).
@@
 
-The `cust.series` function allows to create these series according to factors determined by the user.
-In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (i.e.
-all the living mammals and their direct ancestor) and the stem mammals (i.e.
-all the mammals that have no related living descendants).
+The `cust.series` function allows to create these series according to one or more factors determined by the user.
+In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (all the living mammals and their direct ancestor) and the stem mammals (all the mammals that have no related living descendants).
+In order to do so, we need to separate the tree into two subtrees at the node 71 (see [phylogeny above](#mammal_tree)).
 
 ```{r}
-## We want to separate the species around the node 71 (see phylogeny above).
-## All the descendant of these node are crown and all the ancestors are stem.
+## Selecting the crown and stem mammals
 crown <- extract.clade(BeckLee_tree, node = 71)$tip.label
 stem <- drop.tip(BeckLee_tree, tip = crown)$tip.label
-## We then have to feed this information in a data frame with one column
-factors <- as.data.frame(
+```
+
+We can then feed these lists of species to the `cust.series` function after transforming it into a factorial table:
+
+```{r}
+## Creating the factor table
+series_factors <- as.data.frame(
     matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
     ncol = 1, dimnames = list(c(crown, stem), "Group")))
+head(series_factors)
 
-## We then can use the customised series function to create the two series
-## i.e. the two regions of the ordinated space
-crown_stem <- cust.series(BeckLee_mat50, factors)
+## Creating the two series
+crown_stem <- cust.series(BeckLee_mat50, series_factors)
 ```
 
-The results of this function (like many others presented below) output a `dispRity` object:
+The results of this function (like many others presented below) outputs a `dispRity` object:
 
@@
 crown_stem
+```
 
+The two series represent two groups occupying specific regions of the mammalian morphospace.
+The `dispRity` object itself always contains the original morphospace (`$matrix`) as well as information on which taxa compose the series (`$series`):
+
+```{r}
 ## This object contains three elements
 names(crown_stem)
-## With "data" being the list sub-matrices
-str(crown_stem$data)
-## "elements" being the list of taxa in the original ordinated matrix
-str(crown_stem$elements)
-## and finally, "series" containing information on the series type (custom)
-## and names (crown and stem)
-crown_stem$series
+## With "matrix" being the original matrix (the ordinated space)
+str(crown_stem$matrix)
+## And "series" the elements from this ordinated space into the two sub-spaces
+str(crown_stem$series)
+## Note that the elements here are indicated as numeric values referring to the row numbers in the
+## original matrix
 ```
@@
 ## Adding all mammals to the a new series (along with crown and stem as before)
-factors2 <- as.data.frame(
+series_factors2 <- as.data.frame(
     matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem)),
@@
 ## Creating the three series: Group.stem/crown and Class.Mammalia
-cust.series(BeckLee_mat50, factors2)
+cust.series(BeckLee_mat50, series_factors2)
 ```
@@
 Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
-The `time.series` function allows to create series that contain all the elements present at specific points or during specific periods in time.
+The `time.series` function allows to create series containing all the elements present at specific points in time or during specific periods.
 This functions needs as input an ordinated space and a matching phylogenetic tree.
@@
 For the time-slicing method details see @GuillermeSTD.
+<!-- TODO: link to the time-slicing algorithm explained! -->
 For both methods, the function intakes the `time` argument which can be a vector of `numeric` values for:
@@
     time = c(120, 80, 40, 0))
-
 ```
 
-In this example, the taxa where split inside each time-bin according to their age.
+In this example, the taxa were split inside each time-bin according to their age.
 Note however that some taxa might span between some time bins and should be included in more than one.
@@
 ```{r}
-## Generating three time bins containing the data present every 40 Ma, including
-## taxa that might span between time bins.
+## Generating three time bins containing the data present every 40 Ma, including taxa that might
+## span between time bins.
 (time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
     method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages))
-
 ```
 
-This generated indeed a list of 3 sub matrices.
+This generated indeed a `dispRity` object of 3 series.
 Note that we can also generate equivalent results by just telling the function that we want three time-bins (series) as follow:
@@
 
-For the time-slicing method (`method = continuous`), the idea is really similar.
-This option intakes a matrix that contains the ordinated distance between taxa *and* nodes and an assumed evolutionary model via the `model` argument:
+For the time-slicing method (`method = continuous`), the idea is fairly similar.
+This option, however, requires a matrix that contains the ordinated distance between taxa *and* nodes as well as an extra argument being the assumed evolutionary model (via the `model` argument):
 
@@
 ```{r}
-## Generating four time slices every 40 million years assuming a gradual
-## evolutionary model
+## Generating four time slices every 40 million years assuming a gradual evolutionary model
 (time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
@@
     FADLAD = BeckLee_ages))
-## Note that in the same way as for the discrete method, we can also
-## automatically generate the slices
+## Note that in the same way as before, we can also automatically generate the slices
 time.series(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
@@
 
-Once we obtain our different series, it is possible to bootstrap and rarefy it (i.e. pseudo-replicating the data).
-The bootstrap will allow us to make each series more robust to outliers and the rarefaction will allow us to compare slices with the same number of elements to get rid of eventual sampling problems.
+Once we obtain our different series, it is possible to bootstrap and rarefy them (i.e. pseudo-replicating the data).
+The bootstrap will allow us to make each series more robust to outliers and the rarefaction will allow us to compare series with the same number of elements to get rid of eventual sampling problems.
 The `boot.matrix` allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
-The default options will bootstrap the matrix 1000 times without rarefaction.
-The number of bootstrap pseudo-replicates can be defined using the `bootstraps` option (see below).
+The default options will bootstrap the matrix 1000 times without rarefaction using the "full" bootstrap method (see below).
 
 ```{r}
-## Bootstrapping a matrix
 boot.matrix(data = BeckLee_mat50)
@@
 
-This function allows to control the bootstrap algorithm through the `boot.type` argument.
+The number of bootstrap pseudo-replicates can be defined using the `bootstraps` option.
+The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
 Currently two algorithms are implemented:
@@
 The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in the dataset to a minimum of 3 elements).
-It can also be set to a fix `numeric` value (or a set of `numeric` values).
-Finally, one last argument, `rm.last.axis` allows to remove a certain amount of dimensions (or axis) for the ordinated space.
-This can be logical argument where `FALSE` (default) will not remove any dimension and `TRUE` will remove the last dimensions that bear up to 5% of the total ordinated space’s variance.
-Note that it can also intake a `numerical` value designating the proportion of axis to keep.
+remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in each series to a minimum of 3 elements).
+It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
+One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
+When left missing, all dimensions from the ordinated matrix are used.
 
@@
 
-## And removing the last axis (default)
-boot.matrix(BeckLee_mat50, rm.last.axis = TRUE)
+## Removing the last axis 50% last dimensions
+boot.matrix(BeckLee_mat50, dimensions = 0.5)
 
-## Or with a fix value (50%)
-boot.matrix(BeckLee_mat50, rm.last.axis = 0.5)
+## Keeping the 10 first dimensions
+boot.matrix(BeckLee_mat50, dimensions = 10)
 ```
 
-Of course, one could be interested in directly supplying the sub-sampled matrices generated above directly to this function.
+Of course, one could be interested in directly supplying the series generated above directly to this function.
 In fact, it can also deal with a list of matrices or with a `dispRity` object output from the `cust.series` or `time.series` functions.
@@
 ```{r}
-## Bootstrap and full rarefaction on the crown/stem series
+## Bootstrapping and fully rarefying the crown/stem series
 (crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE))
 
-## Bootstrap on the time binning/slicing series
+## Bootstrapping on the time binning/slicing series
 (time_binsBS <- boot.matrix(time_bins, bootstraps = 100))
@@
 This function is a modular function that allows to simply (and quickly!) calculate disparity from a matrix.
-
 Because disparity can be measured in many ways, this function is a tool to measure disparity *as defined by the user*.
@@
 
+<!-- TODO: change to html link -->
+
 In practice, the `dispRity` function intakes one or more functions as a definition of disparity.
@@
 ```{r}
+## Calculating disparity as the mean value of the ordinated space (using base::mean)
 dispRity(BeckLee_mat50, metric = mean)
-## This defines disparity as the mean value of the morphospace
-## (using base::mean).
 
-## It is also possible to combine multiple functions:
+## Calculating disparity as the sum of variances of the ordinated space using base::sum on the
+## variances and dispRity::variances on each dimension of the ordinated spaces
 dispRity(BeckLee_mat50, metric = c(sum, variances))
-## Defining disparity as the sum (base::sum) of the variances
-## (dispRity::variances) of each dimension of the morphospace.
 
+## Calculating disparity as the product of the centroids (using base::prod and dispRity::centroids)
 dispRity(BeckLee_mat50, metric = c(prod, centroids))
-## For the product (base::prod) of the distances between each elements and the
-## centroid of the morphospace (dispRIty::centroids).
 
-## Or user defined ones:
+## Defining the total range function:
 total.range <- function(X) abs(range(X)[1]-range(X)[2])
+
+## Calculating disparity as the total range of the centroids:
 dispRity(BeckLee_mat50, metric = c(total.range, centroids))
-## For the range (total.range, user defined) of the distances between each
-## elements and the centroid of the morphospace (dispRity::centroids).
+```
 
-## Or more complex ones:
+Note that in practice, the order of the functions in the `metric` argument does not matter:
+
+```{r}
+## Disparity as the standard deviation of the variances of the variance/covariance matrix;
 vars1 <- dispRity(BeckLee_mat50, metric = c(sd, variances, var))
-## For the standard deviation (stats::sd) of the variance (dispRity::variances)
-## of each column of the variance/covariance matrix (stats::var).
 
-## Note that the order of the function is not important since the levels of
-## each function are automatically detected by the dispRity algorithm.
+## Same disparity metric but with the functions ordered differently
 vars2 <- dispRity(BeckLee_mat50, metric = c(variances, var, sd))
+
+## Both versions gives the same results
 all(summary(vars1) == summary(vars2))
@@
 
-In these examples we used only simple ordinated matrix but of course, it might be more interesting to directly use the `dispRity` objects we generated in the steps above.
-Thus we can calculate the bootstrapped and rarefied disparity in each sub-sample of the morphospace.
-Here we are going to define disparity as being the median distance between each element and each dimensions centroids.
+In these examples we used only simple ordinated matrix but of course, it possible to directly use the `dispRity` objects we generated in the steps above.
+Thus we can calculate the bootstrapped and rarefied disparity in each series of the ordinated space.
+Here we are going to define disparity as being the median distance between each element and each dimensions' centre.
 
@@
 disp_time_slicesBS <- dispRity(time_slicesBS, metric = c(median, centroids))
-
 ```
@@
 Note that the computational time was longer for the rarefied data (`crown_stemBS`) since it had to calculate the disparity on all the rarefied matrix, each bootstrapped 100 times.
-For both the `boot.matrix` and the `dispRity` functions, it is possible to increase computational speed in some cases using the `parallel` option (see each function’s manuals).
+For both the `boot.matrix` and the `dispRity` functions, it is possible to increase computational speed in some cases using the `parallel` option (see each [parallel section](#parallel) below).
 
@@
 This function is a S3 function (`summary.dispRity`) allowing to summarize the content of `dispRity` objects that contain disparity calculations.
-This function intakes a `dispRity` object plus various options namely the `quantile` values for the confidence intervals levels; the `cent.tend` for the central tendency to use for summarising the results and two *visual* options which are whether to recall the `dispRity` options and how much digits are wanted in the results.
 
@@
 summary(disp_time_binsBS)
-## By default, the 50 and 95 quantiles and the mean are calculated.
-## Note that the function also displays the observed disparity
-## (non-bootstrapped).
+```
 
-## These arguments can be changed easily as follow
-summary(disp_time_binsBS, quantile = 88, cent.tend = sd, rounding = 1)
+This function intakes a `dispRity` object plus various options namely the `quantile` values for the confidence intervals levels; the `cent.tend` for the 
+central tendency to use for summarising the results and two *visual* options which are whether to recall the `dispRity` options and how much digits are wanted in the results.
+By default the 50 and 95 quantiles and the median are calculated but these options can be modified as follows:
+
+<!-- STOPPED HERE -->
+
+```{r}
+## Summarising the disparity using the 88 quantile and the standard deviation as the central
+## tendency. 
+summary(disp_time_binsBS, quantile = 88, cent.tend = sd)
 
@@
 
-### `tree.age`
+### `dispRity` object utilities  <a name="dispRity.utilities"></a>
+
+export(make.dispRity)
+export(fill.dispRity)
+export(matrix.dispRity)
+export(get.series.dispRity)
+export(extract.dispRity)
+export(scale.dispRity)
+export(sort.dispRity)
+
+### Package general utilities
+
+## `tree.age`
 
@@
 
-Developments
+parallel  <a name="parallel"></a>
+---------
+
+Some of the core functions from the package have a `parallel` option.
+
+Suggestions
 ============
-
-As stated at the start of the demo, this version 0.2 is still in development and many parts are missing.
-Here are the new functionalities that will be implemented in further versions.
-
-More user defined functions
----------------------------
-
-I intend also develop functions to help users to develop their own algorithms for the bootstrap method (via `make.boot`) or the evolutionary models (via `make.model`).
-Both functions will provide similar testing as the `make.metric` function.
-
-Suggestions?
-------------
-
 If you have any extra suggestions or comments on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).

### 2017-01-03 doc/dispRity-manual.Rmd
further man updates

@@
 
-## What is in the object?
+## What does the object contain?
 names(median_centroids)
@@
 str(crown_stem$series)
-## Note that the elements here are indicated as numeric values referring to the row numbers in the
-## original matrix
 ```
@@
 
-This function intakes a `dispRity` object plus various options namely the `quantile` values for the confidence intervals levels; the `cent.tend` for the 
-central tendency to use for summarising the results and two *visual* options which are whether to recall the `dispRity` options and how much digits are wanted in the results.
-By default the 50 and 95 quantiles and the median are calculated but these options can be modified as follows:
-
-<!-- STOPPED HERE -->
+This function intakes a `dispRity` object plus various options namely the `quantile` values for the confidence intervals levels (by default, the 50 and 95 quantiles are calculated); the `cent.tend` for the  central tendency to use for summarising the results (default is `median`); the `rounding` option corresponding to the number of decimals to print (default is `2`) and the `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
+These options can easily be changed from the defaults as follows:
 
 ```{r}
-## Summarising the disparity using the 88 quantile and the standard deviation as the central
-## tendency. 
+## Summarising the results using the 88 quantile and the standard deviation as the central tendency. 
 summary(disp_time_binsBS, quantile = 88, cent.tend = sd)
 
-## If one happens to forgot what was calculated in the summarised object,
-## it is possible to recall the different steps details using recall:
-summary(disp_time_binsBS, recall = TRUE)
+## Printing the details of the dispRity object and rounding the values to the 5th decimal
+summary(disp_time_binsBS, recall = TRUE, rounding = 5)
+```
 
-## Note that the information with each different number of elements is
-## displayed for rarefied data:
-tail(summary(disp_crown_stemBS))
-## The observed disparity is NA for rarefied data.
+> Information about the number of elements per series as well as the observed (i.e. non-bootstrapped) disparity is also calculated.
+This is specifically handy when rarefying the data for example
+
+```{r}
+head(summary(disp_crown_stemBS))
 ```
@@
 
-An alternative way to summarise the data is to plot the results using the S3 method `plot.dispRity`.
+An alternative way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
 This function intakes the same options as `summary.dispRity` along side with various graphical options described in the function manual (see `?plot.dispRity`).
 
-Two main categories of plots can be displayed: continuous plots with the argument `type = continuous` (see below) and discrete plots (default is `type = box`.
-Note that this argument can be left missing. In this case, the algorithm will automatically detect the type of series from the `dispRity` object.
+The plots can be of four different types: `continuous` for displaying continuous disparity curves and `box`, `lines` and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines and confidence interval polygons.
 
-The discrete plot arguments can be either:
-
--   `type = box` for plotting boxplots.
-
--   `type = polygon` for plotting boxes that represent the different levels of confidence intervals (with a dot for the central tendency).
-
--   `type = lines` which is similar to `type = polygon` but plots lines rather than boxes for the confidence intervals levels.
+> This argument can be left missing. In this case, the algorithm will automatically detect the type of series from the `dispRity` object.
 
 It is also possible to display the number of elements per series (has a horizontal doted line) using the option `elements = TRUE`.
-Additionally, when the data is rarefied (see section \[rare\]), one can also indicate which level of rarefaction to display (i.e.
-display the results for a certain number of elements) by using the `rarefaction` argument.
+Additionally, when the data is rarefied, one can also indicate which level of rarefaction to display (i.e. display the results for a certain number of elements) by using the `rarefaction` argument.
 
@@
 op <- par(mfrow = c(2,2), bty = "n")
-
-## Plotting the disparity for both groups
+## Plotting continuous disparity results
+plot(disp_time_binsBS, type = "continuous")
+## Plotting discrete disparity results
 plot(disp_crown_stemBS, type = "box")
-
-## Same plot but with the number of elements on the right and using the
-## "polygon" plot type.
+## Plotting the disparity using lines for the rarefaction level of 20 elements only
+plot(disp_crown_stemBS, type = "lines", rarefaction = 20)
+## Plotting the disparity using polygons while also displaying the number of elements
 plot(disp_crown_stemBS, type = "polygon", elements = TRUE)
-## Rarefied version of the same plot (20 elements per series)
-plot(disp_crown_stemBS, type = "polygon", rarefaction = 20)
 ```
 
+> The `type` argument is not mandatory for the two first plots since both `continuous` and `box` are the default options for respectively continuous and discrete data.
+
 It is also possible to easily look at the behaviour of the disparity metric with the rarefied data (i.e.
-how does the disparity score change when the number of elements goes down) by using the `rarefaction = plot` argument:
+how does the disparity score change when the number of elements goes down) by using the `rarefaction = TRUE` argument:
 
+<!-- 
 ```{r, fig.width=6, fig.height=6}
@@
 ## The rarefaction curves
-plot(disp_crown_stemBS, elements = TRUE, rarefaction = "plot")
+plot(disp_crown_stemBS, elements = TRUE, rarefaction = TRUE)
 
@@
 Some of the core functions from the package have a `parallel` option.
+ -->
 

### 2017-01-06 doc/dispRity-manual.Rmd
manual updates

@@
 
-It is also possible to easily look at the behaviour of the disparity metric with the rarefied data (i.e.
-how does the disparity score change when the number of elements goes down) by using the `rarefaction = TRUE` argument:
-
-<!-- 
-```{r, fig.width=6, fig.height=6}
-## Graphical options
-par(bty = "n")
-## The rarefaction curves
-plot(disp_crown_stemBS, elements = TRUE, rarefaction = TRUE)
-
-```
-
-The different dashed lines represent the different confidence intervals around each rarefaction curve. We can see that the disparity metric becomes stable after more than 10 elements.
-
-It is also possible, when using the `type = lines` argument, to plot lines rather than boxes.
-This option can be useful when displaying many series in a same window.
-Additionally, it is possible to plot the observed data when existing (i.e.
-not when the data is rarefied) by using the option `observed = TRUE`.
-
-```{r}
-## Graphical options
-par(bty = "n")
-## The disparity-through-time data in a time-binned way (with lines rather than
-## boxes) with the observed data as crosses
-plot(disp_time_binsBS, type = "lines", observed = TRUE)
-```
-
-The `type = continuous` argument plots superimposed continuous (along the x axis) polygons that represent the confidence intervals and a continuous line that represents the central tendency of the data.
-Note that, as a S3 method function, `plot.dispRity` can take the classic graphical options from the `plot` function or even to add other plots to the previous ones.
+Since `plot.dispRity` uses the arguments from the generic `plot` method, it is of course possible to change pretty much everything using the regular plot arguments:
 
@@
 par(bty = "n")
-
-## First, plotting the continuous data with a y axis limit
-plot(disp_time_slicesBS, type = "continuous")
-## Note that the type = "continuous" argument is not mandatory here.
+## Plotting the results with some classic options from plot
+plot(disp_time_slicesBS, col = c("blue", "orange", "green"), ylab = c("Some measurement"),
+    xlab = "Some other measurement", main = "Many options...", ylim = c(0,2.1))
+## Adding a legend
+legend("topleft", legend = c("Central tendency", "Confidence interval 1", "Confidence interval 2"),
+    col = c("blue", "orange", "green"), pch = 19)
 ```
 
-I encourage you to play with the graphical options to have some prettier results.
-Note that most of the options from `plot` can be passed to `plot.dispRity` via `...`.
+Additionally to the classic `plot` arguments, the function can also take arguments that are specific to `plot.dispRity` like adding the number of elements or rarefaction level (as described above) but also changing the values of the quantiles to plot as well as the central tendency.
 
@@
 par(bty = "n")
-## A plot with many options!
-plot(disp_time_slicesBS, quantile = c(seq(from=10, to=100, by=10)),
-    cent.tend = sd, type = "continuous", elements = "log",
-    col = c("black", rainbow(10)), ylab = c("Disparity", "log(Diversity)"),
+## Plotting the results with some plot.dispRity arguments
+plot(disp_time_slicesBS, quantile = c(seq(from=10, to=100, by=10)), cent.tend = sd, type = "c",
+    elements = TRUE, col = c("black", rainbow(10)), ylab = c("Disparity", "Diversity"),
     xlab = "Time (in in units from past to present)", time.series = FALSE,
@@
 
-Finally, it is possible to add multiple plots on top of each other:
+> Note that the argument `observed = TRUE` allows to plot the disparity values calculated from the non-bootstrapped data as crosses on the plot.
+
+For comparing results, it is also possible to add a plot to the existent plot by using `add = TRUE`:
 
@@
 par(bty = "n")
+## Plotting the continuous disparity with a fixed y axis
+plot(disp_time_slicesBS, ylim = c(0.5,2))
+## Adding the discrete data using the same fixed y axis to have the results at the same scale
+plot(disp_time_binsBS, discrete_type = "line", ylim=c(0.5,2), xlab="", ylab="", add = TRUE)
+```
 
-## First, plotting the continuous data with a y axis limit
-plot(disp_time_slicesBS, ylim = c(0.5,2))
+<!-- TODO: fix the placement bug with add = TRUE! -->
 
-## Second, adding the discrete data
-par(new = TRUE)
-plot(disp_time_binsBS, discrete_type = "line", ylim=c(0.5,2), xlab="", ylab="")
+Finally, if your data has been fully rarefied, it is also possible to easily look at rarefaction curves by using the `rarefaction = TRUE` argument:
+
+```{r, fig.width=6, fig.height=6}
+## Graphical options
+par(bty = "n")
+## Plotting the rarefaction curves
+plot(disp_crown_stemBS, rarefaction = TRUE)
 ```
+<!--
 

### 2017-01-10 doc/dispRity-manual.Rmd
Manual update for 0.3.0

@@
 
-<!-- TODO: change to html link -->
-
 ```{r}
@@
 For more information on the metrics, please refer to the [`dispRity` metric manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metric.Rmd) for the vignette about the metric implementation in `dispRity`.
-
-<!-- TODO: change to html link -->
 
@@
 
-`time.series`
+`time.series`  <a name="time.series"></a>
 -------------
@@
 The `dispRity` algorithm decomposes the metrics functions into three levels that correspond to the dimensions of the output of each metric function.
-For more details on this algorithm please refer to the [metric vignette](https://github.com/TGuillerme/dispRity).
-
-<!-- TODO: change to html link -->
+For more details on this algorithm please refer to the [metric vignette](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metric.Rmd).
 
@@
 Here we are going to define disparity as being the median distance between each element and each dimensions' centre.
+
+<a name="disparity.calculation"></a>
 
@@
 ## Plotting the disparity using lines for the rarefaction level of 20 elements only
-plot(disp_crown_stemBS, type = "lines", rarefaction = 20)
+plot(disp_crown_stemBS, type = "line", rarefaction = 20)
 ## Plotting the disparity using polygons while also displaying the number of elements
@@
 ## Adding the discrete data using the same fixed y axis to have the results at the same scale
-plot(disp_time_binsBS, discrete_type = "line", ylim=c(0.5,2), xlab="", ylab="", add = TRUE)
+plot(disp_time_binsBS, type = "line", ylim=c(0.5,2), xlab="", ylab="", add = TRUE)
 ```
-
-<!-- TODO: fix the placement bug with add = TRUE! -->
 
@@
 ```
-<!--
 
@@
 
-Finally, the `dispRity` package allows to apply some tests to the data in order to properly test hypothesis.
-The function `test.dispRity` works in a similar way as the `dispRity` function: it intakes a `dispRity` object (containing disparity measurements), a `test` and a `comparisons` argument.
-The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
-It can be common test function (e.g.
-`stats::t.test`) or some functions implemented in `dispRity` (see `dispRity.test`) or even user defined functions.
+Finally, the `dispRity` package allows to apply some test functions to the calculated disparity in order to test hypothesis.
+The function `test.dispRity` works in a similar way as the `dispRity` function: it intakes a `dispRity` object, a `test` and a `comparisons` argument.
+
 The `comparisons` argument must indicate the way the test should be applied to the data:
@@
 
+The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
+It can be common statistical test functions (e.g. `stats::t.test`), some functions implemented in `dispRity` (e.g. see `?null.test`) or any function defined by the user.
+
 This function also allows to correct for type I error inflation when using multiple comparisons via the `correction` argument.
@@
 ```{r}
-## Performing a pairwise t-test to the test a difference in disparity between
-## crown and stem mammals (note that "pairwise" is default)
+## Performing a t-test to the test a difference in disparity between crown and stem mammals
 test.dispRity(disp_crown_stemBS, test = t.test)
 
-## The same test but with the detailed output
+## Performing the same test but with the detailed t.test output
 test.dispRity(disp_crown_stemBS, test = t.test, details = TRUE)
 
-## A wilcoxon rank test applied to sliced disparity the data sequentially with
-## the Bonferroni correction.
-test.dispRity(disp_time_slicesBS, test = wilcox.test,
-    comparisons = "sequential", correction = "bonferroni")
+## Performing a Wilcoxon test applied to sliced disparity the data sequentially with correction
+test.dispRity(disp_time_slicesBS, test = wilcox.test, comparisons = "sequential",
+    correction = "bonferroni")
 
-## An analysis of variance (aov) applied to the same data
-test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all")
+## Measuring the overlap between distributions in the time bins (using the implemented
+## Bhattacharyya Coefficient function - see ?bhatt.coeff)
+test.dispRity(disp_time_binsBS, test = bhatt.coeff)
+```
 
-## Note that in this case, the output is the regular aov output:
-slice_aov <- test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all")
-class(slice_aov) ; summary(slice_aov)
+It is also possible to apply some more *complex* tests that have their own output class (like `stats::aov`).
+The results can then be analysed as usual using the associated `summary` S3 method:
 
-## Measuring the overlap between distributions in the time bins (using the
-## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)
-test.dispRity(disp_time_binsBS, test = bhatt.coeff)
+```{r}
+## Performing and ANOVA applied to the same data
+slice_aov <- (test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all"))
 
+## The output is a regular ANOVA output...
+class(slice_aov)
+
+## That can be summarised using summary
+summary(slice_aov)
 ```
@@
 ---------
+The package also provides some utilities functions to facilitate multidimensional analysis.
 
 ### `dispRity` object utilities  <a name="dispRity.utilities"></a>
+The first series of utilities are functions for manipulating `dispRity` objects:
 
-export(make.dispRity)
-export(fill.dispRity)
-export(matrix.dispRity)
-export(get.series.dispRity)
-export(extract.dispRity)
-export(scale.dispRity)
-export(sort.dispRity)
+#### `make.dispRity`
+This function creates empty `dispRity` objects.
+
+```{r}
+## Creating an empty dispRity object
+make.dispRity()
+## Creating an "empty" dispRity object with a matrix
+(disparity_obj <- make.dispRity(matrix(rnorm(20), 5, 4)))
+```
+
+####  `fill.dispRity`
+This function initialises a `dispRity` object and generating it's call properties.
+
+```{r}
+## The dispRity object's call is indeed empty
+disparity_obj$call
+## Creating an empty dispRity object
+(disparity_obj <- fill.dispRity(disparity_obj))
+## The dipRity object has now the proper minimal attributes
+disparity_obj$call
+```
+
+####  `matrix.dispRity`
+This function extracts a specific matrix from a disparity object. The matrix can be one of the bootstrapped or/and rarefied one.
+
+```{r}
+## Extracting the matrix containing the coordinates for the crown mammals
+str(matrix.dispRity(disp_crown_stemBS, "Group.crown"))
+## Extracting the 3rd boostrapped matrix with the 10th rarefaction level from the same group
+str(matrix.dispRity(disp_crown_stemBS, series = 1, bootstrap = 3, rarefaction = 10))
+```
+
+#### `get.series.dispRity`
+This function creates a dispRity object that contains only elements from one specific series.
+
+```{r}
+## Extracting all the data for the crown mammals
+(crown_mammals <- get.series.dispRity(disp_crown_stemBS, "Group.crown"))
+## The object keeps the properties of the parent object but is composed of only one series
+length(crown_mammals$series)
+```
+
+#### `extract.dispRity`
+This function extracts the calculated disparity values of a specific matrix.
+
+```{r}
+## Extracting the observed disparity (default)
+extract.dispRity(disp_crown_stemBS)
+
+## Extracting the disparity from the bootstrapped values from the 10th rarefaction level from the
+## second series (stem mammals)
+extract.dispRity(disp_crown_stemBS, observed = FALSE, series = 2, rarefaction = 10)
+```
+
+#### `scale.dispRity`
+This is the S3 method of `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object.
+
+```{r}
+## Getting the disparity values of the time series
+summary(disp_time_slicesBS)
+
+## Scaling the same disparity values
+summary(scale(disp_time_slicesBS, scale = TRUE))
+```
+
+#### `sort.dispRity`
+This is the S3 method of `sort` for sorting the series alphabetically (default) or following a specific pattern.
+
+```{r}
+## Sorting the disparity series in inverse alphabetic order
+summary(sort(disp_time_slicesBS, decreasing = TRUE))
+```
 
@@
 
-## `tree.age`
+#### `tree.age`
 
@@
 
+#### `slice.tree`
+
+This function is at the core of `time.series` when using the time slicing method.
+It allows to slice a phylogenetic tree at a certain time point and selecting the nodes or tips following the models detailed in the `time.series` [section above](#time.series).
+
+```{r, fig.width=8, fig.height=6}
+set.seed(1)
+## Generating a random ultrametric tree
+tree <- rcoal(20)
+## Adding some node labels
+tree$node.label <- letters[1:19]
+## Adding a root age (as the maximum tree age)
+tree$root.time <- max(tree.age(tree)$ages)
+
+## Creating a deltran tree at the slice 0.25
+tree_25 <- slice.tree(tree, age = 0.25, "deltran")
+## Creating a deltran tree at the slice 0.50
+tree_50 <- slice.tree(tree, age = 0.50, "acctran")
+
+## Illustrating the results
+par(mfrow = c(1,3))
+plot(tree, main = "Full tree", cex = 1.5)
+nodelabels(tree$node.label, cex = 1.1)
+axisPhylo()
+plot(tree_25, main = "Slice at time 0.25", cex = 1.5)
+axisPhylo()
+plot(tree_50, main = "Slice at time 0.50", cex = 1.5)
+axisPhylo()
+```
+
+#### `sim.morpho`
+This function (and it's associated utilities) allows to generate realistic discrete morphological matrices.
+See the [associated vignette](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-simulate_data.Rmd) for more details.
+
+Other utilities functions are also available within the package and more will be added in the future. For the full list, refer to the package's functions list within `R`.
+
 parallel  <a name="parallel"></a>
@@
 
-Some of the core functions from the package have a `parallel` option.
- -->
+`dispRity` and `boot.matrix` have also `parallel` option.
+This option uses the package `snow` to run the `lapply` loops within the package in parallel.
+In both functions, this option is ignored when missing, otherwise, it takes as option, the number of threads (cores) and the virtual connection process (VCP) name following the function `makeCluster`.
+
+For example, we can speed up the [disparity calculations of the rarefied data](#disparity.calculation) by using 4 different threads and a `"SOCK"` connection (although `"MPI"` is much more recommended but needs more specific installation).
+
+```{r, eval=FALSE}
+dispRity(crown_stemBS, metric = c(median, centroids), parallel = c(4, "SOCK"))
+```
 

### 2017-01-10 doc/dispRity-simulate_data.Rmd

@@
 morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
-    rates = c(rgamma, 0.5, 1), invariant = TRUE)
+    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
 check.morpho(morpho_mat, my_tree)

### 2017-01-13 doc/dispRity-simulate_data.Rmd
Updated data simulation manual

@@
 ---
-
-This vignette details how to simulate morphological data in `dispRity` (v.0.2.1).
 
@@
 =================================
-This is a quick example on how it works
+The function basically intakes a phylogenetic tree, the number of required characters and the evolutionary model and a function from which to draw the rates.
+The package then proposes a function for quickly checking the matrix' phylogenetic signal using parsimony.
 
@@
 Basically, `sim.morpho` is really flexible and intakes a lot of different arguments to allow to simulate realistic matrix.
-It has two implemented models: `"ER"` for Equal Rates (M*k* model) or the `"HKY"` one which is the molecular HKY model but transforms pyrines in 0's and purimidines in 1's.
-It can also intake some specific distribution for the rates or the substitutions, allowing to modify these basic models.
+It has three implemented models: `"ER"` for Equal Rates (the M*k* model); the `"HKY"` one which is the molecular HKY model but transforms pyrines in 0's and purimidines in 1's; or the `"mixed"` model that randomly uses an `"ER"` or and `"HKY"` for the binary characters and `"ER"` for the multistates (>2) characters.
+Both models intakes specific distributions for their rates or substitution models.
+These distributions should be passed to these arguments in the format of `c(sampler_function, distribution_parameters)` where the the sampler function is a the random generation function of that distribution (e.g.
+`rnorm`, `runif`, etc...) and the parameters are any parameters to be passed to this function.
 
-The `check.morpho` runs a quick MP tree using the `phangorn` parsimony algorithm.
+The `check.morpho` runs a quick maximum parsimony tree using the `phangorn` parsimony algorithm.
 It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g.
-the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the MP tree and the provided tree.
+the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree.
 
-Good parameters for a good matrix
+Parameters for a good(ish) matrix
 -----------------------------------
-
-I was testing the effect of different parameters on the ability to generate a ``realistic'' matrix (meaning not to different from the input tree with a consistency and retention index close to what's in the literature).
-The matrix seems the most ``realistic'' with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three states characters, a gamma distribution with a shape parameter ($\alpha$) of 0.5 and no scaling ($\beta$ = 1).
-
-Something like that should give a decent matrix:
+There are many parameters that can realise a ``realistic'' matrix (meaning not to different from the input tree with a consistency and retention index close to what's in the literature) but because of the randomness of the generation (the code generates random matrices after all!) not all end up in creating good matrices.
+The following parameters however, seems to generate a fairly ``realistic'' with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three states characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
 

### 2017-01-17 doc/dispRity-metrics.Rmd
Updated metrics vignette

--- 2016-05-25 doc/dispRity-metrics.Rmd
@@
 
-This vignette details how the disparity metrics are implemented in the `dispRity` package (v.0.2).
-
 Disparity metrics
@@
 
-There are many ways for measuring disparity. In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. MDS, PCA, PCO, PCoA).
-For example, one can look at ellipsoid hyper-volume of the ordinated space [@DonohueDim], the sum and the product of the ranges and variances of the ordinated space dimensions [@Wills1994] or the median position of the elements in the ordinated space relative to its centroid [@GuillermeSTD].
+There are many ways for measuring disparity!
+In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
+For example, one can look at ellipsoid hyper-volume of the ordinated space [@DonohueDim], the sum and the product of the ranges and variances [@Wills1994] or the median position of the elements relative to their centroid [@GuillermeSTD].
 Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks or some being more generalist.
@@
 Because of this great diversity of metric, the package `dispRity` has not a unique way to measure disparity but rather proposes to facilitate users to define their own disparity metric that will best suit to their particular analysis.
-In fact, the core function of the package, `dispRity`, allows the user to define any metric in the `metric` argument.
-However, because of the infrastructure of the package (especially the one of the `dispRity` objects – see the [`dispRity` manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf)), the `metric` argument has to follow certain rules:
+In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
+However, because of the infrastructure of the package (especially the one of the `dispRity` objects – see the [`dispRity` manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.Rmd)), the `metric` argument has to follow certain rules:
 
-1.  it must be composed of one to three `function` objects;
+1.  it must be composed from one to three `function` objects;
 
-2.  these functions can only have as input a `matrix` or a `vector`;
+2.  these functions can only intake a `matrix` or a `vector`;
 
@@
 
-4.  at least one of the functions must be a level 1 function (see below).
+4.  at least one of the functions must be of level 1 or 2 (see below).
 
@@
 
-![Illustration of the different levels of output of a function with an input `matrix`](dispRity_fun.pdf)
+![Illustration of the different levels of output of a function with an input `matrix`](dispRity_fun.png)
 
@@
 
-A level 1 function for example, will decompose a `matrix` or a `vector` into a single value:
+A level 1 function will decompose a `matrix` or a `vector` into a single value:
 
 ```{r}
-# First, let's create a dummy matrix
+## First, let's create a dummy matrix
 dummy_matrix <- matrix(rnorm(12), 4, 3)
 
-# Any summary metric such as mean or median are good examples of level 1
-# functions as they reduce the matrix to a single dimension (i.e. one value):
+## Any summary metric such as mean or median are good examples of level 1
+## functions as they reduce the matrix to a single dimension (i.e. one value):
 mean(dummy_matrix)
@@
 A level 2 function will decompose a `matrix` into a `vector`.
-Several level 2 functions are implemented in `dispRity`(see `?dispRity.metric`) such as the `variances` function that calculates the variance of each dimension of the ordinated matrix.
 
 ```{r}
-# We can define a function that outputs the products of each row of the matrix:
+## We can define a function that outputs the products of each row of the matrix:
 prod.rows <- function(matrix) apply(matrix, 1, prod)
 
-# This function is a level 2 function as it reduces the matrix into a two
-# dimensional element (a vector of values):
+## This function is a level 2 function as it reduces the matrix into a two
+## dimensional element (a vector of values):
 prod.rows(dummy_matrix)
 ```
+
+Several level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` function that calculates the variance of each dimension of the ordinated matrix or the `ranges` one that calculates the range of each dimension.
 
@@
 Finally a level 3 function will transform the `matrix` into another `matrix`.
-Note that the dimension of the output `matrix` don't need to match the the input `matrix`:
+Note that the dimension of the output matrix don't need to match the the input one:
 
 ```{r}
-# The var functions computes the variance/covariance matrix which will be a
-# three by three matrix
+## The var functions computes the variance/covariance matrix which will be a 3x3 matrix
 var(dummy_matrix)
@@
 Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the euclidean distance between each elements and the ordinated space centroid.
-The `make.metric` function implemented in `dispRity`is designed to help testing and finding the level of the functions.
+The `make.metric` function implemented in `dispRity` is designed to help testing and finding the level of the functions.
 This function tests:
@@
 ```{r}
-# First we need to load the package
+## First we need to load the package
 library(dispRity)
 
-# Which level is the mean function? And is able to be used in the dispRity?
+## Which level is the mean function? And is able to be used in the dispRity?
 make.metric(mean)
 
-# Same questions from the prod.rows function:
+## Same questions from the prod.rows function:
 make.metric(prod.rows)
 
-# Same questions for the var function:
+## Same questions for the var function:
 make.metric(var)
@@
 
-Using this metric structure, on can easily use any disparity metric in the `dispRity`function as follows:
+Using this metric structure, we can easily use any disparity metric in the `dispRity`function as follows:
 
 ```{r}
-# First we need to load the data
+## First we need to load the data
 data(BeckLee_mat50)
 
-# Measuring disparity as the standard deviation of all the value of the
-# ordinated matrix (level 1 function).
+## Measuring disparity as the standard deviation of all the value of the
+## ordinated matrix (level 1 function).
 summary(dispRity(BeckLee_mat50, metric = sd))
 
-# Measuring disparity as the standard deviation of the variance of each axis of
-# the ordinated matrix (level 1 and level 2 functions).
+## Measuring disparity as the standard deviation of the variance of each axis of
+## the ordinated matrix (level 1 and level 2 functions).
 summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))
 
-# Measuring disparity as the standard deviation of the variance of each axis of
-# the variance covariance matrix (level 1, level 2 and level 3 functions).
+## Measuring disparity as the standard deviation of the variance of each axis of
+## the variance covariance matrix (level 1, level 2 and level 3 functions).
 summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
@@
 
-Note that the order of each function in the metric argument does not matter, the `dispRity`function will automatically detect the function levels (using `make.metric`) and apply them to the data in decreasing order (level 3 > level 2 > level 1).
+Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function levels (using `make.metric`) and apply them to the data in decreasing order (level 3 > level 2 > level 1).
 
 ```{r}
-# Disparity as the standard deviation of the variance of each axis of the
-# variance covariance matrix:
-disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)),
-                      round = 10)
+## Disparity as the standard deviation of the variance of each axis of the
+## variance covariance matrix:
+disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
 
-# Same but using a different functions order for the metric argument
-disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
-                      round = 10)
+## Same but using a different functions order for the metric argument
+disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)), round = 10)
 
-# Both ways output the same disparity values:
-disparity1$observed == disparity2$observed # is TRUE
+## Both ways output the same disparity values:
+disparity1 == disparity2
 ```
@@
 
-In some cases however, it might also be possible to calculate disparity as a distribution.
+In some cases however, it might also be interesting to calculate disparity as a distribution.
 For example, one of the metrics commonly used in palaeobiology is the sum of variances of each dimensions (`metric = c(sum, variances)`) [@Wills1994].
@@
 ```{r}
-# A one dimensional disparity
-sum_variances1 <- summary(dispRity(BeckLee_mat50, metric = c(sum, variances)))
+## A one dimensional disparity value
+(sum_variances1 <- summary(dispRity(BeckLee_mat50, metric = c(median, variances))))
 
-# A two dimensional disparity
-all_variances <- summary(dispRity(BeckLee_mat50, metric = variances))
+## A two dimensional disparity
+(summary(all_variances <- (dispRity(BeckLee_mat50, metric = variances))))
+```
 
-# Transforming the two dimensional one in a one dimensional one
-sum_variances2 <- summary(dispRity(BeckLee_mat50, metric = sum))
+Looking at all the variances rather than simply their median can tell us that (by the look of it) the variances of each dimensions seems to be normally distributed.
 
-# Both ways output the same disparity values:
-sum_variances1$observed == sum_variances2$observed
+> It is also possible to reuse an already calculated disparity distribution and summarise it using a level 1 metric:
+
+```{r}
+## Transforming the two dimensional one in a one dimensional one
+sum_variances2 <- summary(dispRity(all_variances, metric = median))
+
+## Both ways output the same disparity values:
+sum_variances1 == sum_variances2
+```
+
+Metrics implemented in `dispRity`
+===========================================
+
+Several disparity metrics are directly implemented in the `dispRity` package.
+The detailed list can be found in `?dispRity.metric` along with some explanations on each metric.
+
+Level | Name | Description | Source | 
+------|------|---------------------------------------------------|--------|
+3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
+3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
+3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
+3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
+3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
+2 | `ranges` | The range of each dimension | `dispRity` |
+2 | `variances` | The variance of each dimension | `dispRity` |
+2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
+1 | `mode.val` | The modal value | `dispRity` |
+
+1: The functions estimation of the eigen value only works for MDS or PCoA ordinations (not PCA).
+
+2: Note that by default, the centroid is the centroid of the elements.
+It can however be fixed to an arbitrary value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
+
+## Equations and implementations:
+Some of the functions described below are directly implemented in the `dispRity` package and do not require any dependencies  ([see implementation here](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R)).
+
+\begin{equation}
+    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
+\end{equation}
+With *k* being the number of dimensions, $\lambda_i$ is the eigen value of each dimension.
+
+\begin{equation}
+    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
+\end{equation}
+With *k* being the dimensions.
+
+\begin{equation}
+    ranges = |max(k_i) - min(k_i)|
+\end{equation}
+With *k* being the dimensions.
+
+\begin{equation}
+    variances = \sigma^{2}{k_i}
+\end{equation}
+With *k* being the dimensions and $\sigma^{2}$ their variance.
+
+\begin{equation}
+    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
+\end{equation}
+With *n* being each element in the ordinated space, *k* being its dimensions and $Centroid_{k}$ their mean (or can be set to an arbitrary value) and.
+
+## Metrics run through
+Here is a run through of the main metrics implemented in `dispRity`.
+First, we will create a dummy ordinated space using the `space.maker` utility function:
+
+```{r}
+## Creating a 10*5 normal space
+set.seed(1)
+dummy_space <- space.maker(10, 5, rnorm)
+```
+
+### Volumes and surface metrics
+The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `hyper.volume` all measure the surface or the volume of the ordinated space occupancy:
+
+```{r}
+## Calculating the ellipsoid volume
+summary(dispRity(dummy_space, metric = ellipse.volume))
+```
+> Because there is only one series (i.e. one matrix) in the dispRity object, this operation is the equivalent of doing simply `ellipse.volume(dummy_space)` (with rounding).
+
+```{r}
+## Calculating the convex hull surface
+summary(dispRity(dummy_space, metric = convhull.surface))
+
+## Calculating the convex hull volume
+summary(dispRity(dummy_space, metric = convhull.volume))
+```
+
+The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
+For the full and correct calculation of the volume of the ordinated space, it is possible to us the `hyper.volume` function that intakes more options but takes longer to calculate.
+
+```{r}
+## Calculating the true multidimensional volume
+summary(dispRity(dummy_space, metric = hyper.volume))
+
+## Calculating the volume with different options to be passed to hypervolume::hypervolume
+summary(dispRity(dummy_space, metric = hyper.volume, repsperpoint = 100, quantile = 0.5))
+```
+> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [Curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) which often results in near 0 disparity values.
+
+### Ranges, variances and diagonal
+
+The functions `ranges`, `variances` and `diagonal` all measure properties of the ordinated space base on it's dimensions properties (they are also less affected by the Curse of dimensionality):
+
+`ranges` and `variances` both work on the same principle and measure the range/variance of each dimensions:
+
+```{r}
+## Calculating the ranges of each dimensions in the ordinated space
+ranges(dummy_space)
+
+## Calculating disparity as the distribution of these ranges
+summary(dispRity(dummy_space, metric = ranges))
+
+## Calculating disparity as the sum and the product of these ranges
+summary(dispRity(dummy_space, metric = c(sum, ranges)))
+summary(dispRity(dummy_space, metric = c(prod, ranges)))
+
+## Calculating the variances of each dimensions in the ordinated space
+variances(dummy_space)
+
+## Calculating disparity as the distribution of these variances
+summary(dispRity(dummy_space, metric = variances))
+
+## Calculating disparity as the sum and the product of these variances
+summary(dispRity(dummy_space, metric = c(sum, variances)))
+summary(dispRity(dummy_space, metric = c(prod, variances)))
+```
+
+The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest euclidean distance in our 5 dimensional space):
+
+```{r}
+## Calculating the ordinated space's diagonal
+summary(dispRity(dummy_space, metric = diagonal))
+```
+
+> This metric is only mathematically valid if the dimensions within the space are all orthogonal!
+
+### Centroids metrics
+
+The `centroids` metric allows to measure the position of the different elements compared to a fixed point in the ordinated space.
+By default, this function measures the distance between each elements and the central point between them:
+
+```{r}
+## Calculating the distribution of the distances between each elements and the centroid
+summary(dispRity(dummy_space, metric = centroids))
+
+## Calculating disparity as the median value of these distances
+summary(dispRity(dummy_space, metric = c(median, centroids)))
+```
+
+It is however possible to fix the coordinates of a specific point in the ordinated space as long as it has the correct number of dimensions:
+
+```{r}
+## Calculating the distance between the elements and the origin of the ordinated space
+summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))
+
+## Calculating disparity as the distance between the elements and a specific point
+summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
 ```

### 2017-01-17 doc/dispRity-manual.Rmd
Removed old manual and added table of content

@@
 date: "`r Sys.Date()`"
-output: rmarkdown::html_vignette
+output:
+    rmarkdown::html_vignette:
+        toc: true
+        toc_depth: 2
 bibliography: References.bib

### 2017-01-18 doc/dispRity-manual.Rmd
Updated links in manuals

@@
 
-For more information on the metrics, please refer to the [`dispRity` metric manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metric.Rmd) for the vignette about the metric implementation in `dispRity`.
+For more information on the metrics, please refer to the [`dispRity` metric manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html) for the vignette about the metric implementation in `dispRity`.
 
@@
 The `dispRity` algorithm decomposes the metrics functions into three levels that correspond to the dimensions of the output of each metric function.
-For more details on this algorithm please refer to the [metric vignette](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metric.Rmd).
+For more details on this algorithm please refer to the [metric vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html).
 
@@
 This function (and it's associated utilities) allows to generate realistic discrete morphological matrices.
-See the [associated vignette](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-simulate_data.Rmd) for more details.
+See the [associated vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-simulate_data.html) for more details.
 

### 2017-01-25 doc/dispRity-ecology-demo.Rmd

@@
 
-This demo aims to give quick overview of the `dispRity` package (v.0.2) for ecological analysis.
+This demo aims to give quick overview of the `dispRity` package (v.0.3) for ecological analysis.
 Please refer to [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the `dispRity` manual that explains the functions in more details.
@@
 To keep it short, this package allows the use of all of the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
-For example, one might want to know whether a specific kind of water treatment at certain water depth alters invertebrate communities and composition in natural habitats.
+For example, we might want to know whether a specific kind of water treatment at certain water depth alters invertebrate communities and composition in natural habitats.
 
@@
 
-For this example with ecological data we are going to use data from McClean (unpubl.).
+For this example with ecological data we are going to use data from D. McClean (unpubl.).
 This data is the ordination of a distance matrix based on nutrient enrichment and depth and freshwater benthic invertebrates.
@@
 
-## As well as two list of different factors affecting each experimental plot:
-## the treatment and the depth.
+## As well as two list of different factors affecting each experimental plot: treatment and depth.
 treatments <- McClean_data$treatment
@@
 ```{r, fig.width=6, fig.height=6}
-## The x and y axis represent the two first dimensions (i.e. principal
-## components) of the ordination
+## The x and y axis represent the two first dimensions (principal components) of the ordination
 x <- ord_matrix[, 1]
 y <- ord_matrix[, 2]
-## The colors will represent the treatments
+## Setting the colors that will represent the treatments
 cols <- sub("a", "red", treatments)
 cols <- sub("b", "blue", cols)
-## The symbols will represent the depth
+## Setting the symbols that will represent the depth
 pchs <- sub(1, 16, depth)
@@
 ## A classic 2D ordination plot
-plot(x, y, col = cols, pch = pchs, xlab = "PC 1", ylab = "PC 2",
-    xlim = range(x) + c(0, 100))
+plot(x, y, col = cols, pch = pchs, xlab = "PC 1", ylab = "PC 2", xlim = range(x) + c(0, 100))
 ```
@@
 
-Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.
+Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.
 
@@
 
-The first step in such analysis will be to create different series that are two sub-samples of the ordinated space (i.e. the *n*-dimensional object).
-Each of these series contain a certain number of elements (i.e. the 40 experimental field plots) that have some attributes in common.
+The first step in such analysis will be to create different series that will represent sub-samples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
+Each of these series contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
 In our example, we are going to group the elements according to their depth and treatment.
@@
 ## Splitting the ordinated space into four subsamples
-customised_series <- cust.series(ord_matrix, factors)
-
-## Note that the output of dispRity functions are dispRity objects
-class(customised_series)
-## These objects are automatically printed in a summary method (calling S3 print.dispRity)
-## giving information about the object
-customised_series
+(customised_series <- cust.series(ord_matrix, factors))
 ```
 
-For more details on the `dispRity` objects, see the [`dispRity` manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf).
-Basically the idea is to avoid jamming the `R` console such as when using:
-
-```{r}
-## Summarising the object
-str(customised_series)
-
-## Displaying the full object
-print(customised_series, all = TRUE)
-```
+Note that the output of `dispRity` functions are `dispRity` objects.
+They allow specific attributes to be displayed using the basic S3 functions (`print`, `summary`, `plot`, etc...).
+For more details on the `dispRity` objects, see the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html).
 
@@
 
-Disparity can be calculated in many ways, and therefore, the `dispRity` function allows to measure disparity *as defined by the user*.
-For more details on measuring disparity, see the [`dispRity` metric vignette](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.Rmd).
+Disparity can be calculated in many ways, and therefore, the `dispRity` function allows to measure disparity as defined by the user.
+For more details on measuring disparity, see the [`dispRity` metric vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html).
 
@@
 We can define the metrics easily in the `dispRity` function by feeding them to the `metric` argument.
-Here we are going to feed the functions `stats::median` and `dispRity::centroids` which calculates distances between elements and
-centroid.
+Here we are going to feed the functions `stats::median` and `dispRity::centroids` which calculates distances between elements and their centroid.
 
 ```{r}
-## Calculating disparity
-disparity <- dispRity(customised_series, metric = c(median, centroids))
+## Calculating disparity as the median distance between each elements and their group centroid
+(disparity <- dispRity(customised_series, metric = c(median, centroids)))
+```
 
-## Note that disparity is a dispRity object and printing it just gives details
-## on the object, not the results. We need to use summary.dispRity (S3) to get
-## the results.
-disparity
+Similarly than with `cust.series`, `dispRity` displays a `dispRity` object.
+To actually look at the calculated values, we can summarise the data by simply using `summary`:
+
+```{r}
+## Displaying the summary of the calculated disparity
 summary(disparity)
@@
 
-One might also want to bootstrap the data to test the robustness of the measured disparity to outliers.
+We can also want to bootstrap the data to test the robustness of the measured disparity to outliers.
 Also, as we can see, each series has different numbers of elements.
@@
 ```{r}
-## Simple bootstrapping (100 times)
-bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100)
+## Bootstrapping the customised series 100 times
+(bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100))
 
-## Bootstrapping with rarefaction (i.e. only re-sampling 17 elements each time)
-rarefied_data <- boot.matrix(customised_series, bootstraps = 100,
-    rarefaction = 17)
-
-## Note that the output is a dispRity object giving some details on the series
-## and the bootstraps
-bootstrapped_data
-rarefied_data
+## Bootstrapping the series and using rarefaction (i.e. only re-sampling 17 elements each time)
+(rarefied_data <- boot.matrix(customised_series, bootstraps = 100, rarefaction = 17))
 ```
@@
 ```{r}
-## Bootstrapped disparity
+## Calculating the bootstrapped disparity
 disparity_BS <- dispRity(bootstrapped_data, metric = c(median, centroids))
 
-## Rarefied disparity
+## Calculating the rarefied disparity
 disparity_rare <- dispRity(rarefied_data, metric = c(median, centroids))
@@
 We can now summarise the data using various options such as the confidence intervals levels and the central tendency.
+For that we will use the `summary` function mentioned above using several of the options specific to `dispRity` objects (see more details with `?summary.dispRity`).
 
 ```{r}
-## Summarising disparity (default)
+## Summarising bootstrapped disparity
 summary(disparity_BS)
 
-## The quantiles are calculated as 50 and 95 and the central tendency is the
-## mean by default but we can specify different options
-summary(disparity_BS, quantile = 90, cent.tend = median)
-
-## Finally we can see the results of the rarefaction analysis:
+## Summarising the rarefied disparity
 summary(disparity_rare)
 
-## Note that for the first three series, there is no observed disparity since
-## the data was rarefied for these series.
+## Summarising the same results but using the mean as the central tendency and displaying 
+## the 90th quantile only.
+summary(disparity_rare, quantile = 90, cent.tend = mean)
 ```
 
-We can also plot the results and have a look at the effect of the number of experimental plots:
+We can also plot the results and have a look at the effect of the number of experimental plots using the generic `plot` function and it's associated arguments (see more details with `?plot.dispRity` for the arguments specific to `dispRity` objects).
 
-```{r, fig.width=6, fig.height=6}
+```{r, fig.width=12, fig.height=6}
 ## Graphical options
@@
 ## Note that the comparison contains the list of the pairs of series to compare
-test.dispRity(disparity_BS, test = wilcox.test,
-    comparisons = list(c(1,2), c(3,4)))
+test.dispRity(disparity_BS, test = wilcox.test, comparisons = list(c(1,2), c(3,4)))
 
 ## Testing the effect of the two different factors for the rarefied data
-test.dispRity(disparity_rare, test = wilcox.test,
-    comparisons = list(c(1,2), c(3,4)))
+test.dispRity(disparity_rare, test = wilcox.test, comparisons = list(c(1,2), c(3,4)))
 ```

### 2017-01-25 doc/dispRity-metrics.Rmd

@@
 In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
-However, because of the infrastructure of the package (especially the one of the `dispRity` objects – see the [`dispRity` manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.Rmd)), the `metric` argument has to follow certain rules:
+However, because of the infrastructure of the package (especially the one of the `dispRity` objects – see the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html), the `metric` argument has to follow certain rules:
 

### 2017-03-14 doc/dispRity-manual.Rmd
Minor edits to R functions and documentation

@@
 
-`dispRity` is a package for calculating disparity in `R`. It allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.
+`dispRity` is a package for calculating disparity in `R`. It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values so they can easily be compared. 
 <!-- Add version and table of content here! -->
 
-Before starting
+Before starting # NC: Not sure this qualifier is needed.
 ===============
@@
 
-Because this package is aimed for multidisciplinary purpose, some names or terms used in this tutorial might be non-familiar to certain fields.
-Here is a list of what are the exact meaning of these term:
+This package is designed to be used in multiple disciplines, including palaeonotology, ecology and evolution. Therefore some names or terms used in this tutorial might be unfamiliar.
+Below is a list of the meanings of these terms, in the context of `dispRity`.
 
--   **Ordinated space**: designates the mathematical multidimensional object that will be analysed with this package.
-    In morphometrics, this one is often referred as being the morphospace.
-    However it can also be the the cladisto-space for cladistic data or the eco-space in ecology, etc.
+-   **Ordinated space**. The mathematical multidimensional object that will be analysed with this package.
+    In morphometrics, this is often referred to as the morphospace.
+    However it may also be referred to as the cladisto-space for cladistic data or the eco-space for ecological data etc.
     In practice, this term designates an ordinated matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
 
--   **Elements**: designates the rows of the ordinated space, elements can be either taxa, field sites, countries, etc...
+-   **Elements**. The rows of the ordinated space. Elements can be taxa, field sites, countries etc.
 
--   **Dimensions**: designates the columns of the ordinated space. The dimensions can also be referred to as axis or principal components in the case of the ordinated space obtained from a PCA.
+-   **Dimensions**. The columns of the ordinated space. The dimensions are referred to as axes of variation, or principal components, for ordinated spaces obtained from a PCA.
 
--   **Series**: it designates sub-samples of the ordinated space.
-    Basically a series contain the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
-    For example, if our ordinated space is composed of birds and mammals (i.e.
-the elements) and 50 dimensions, we can create two series of just mammals or birds as elements (but with the same 50 dimensions) to look at the difference in disparity between the two clades.
+-   **Series**. Sub-samples of the ordinated space. #Change to subsample?
+    A series contains the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
+    For example, if our ordinated space is composed of birds and mammals (the elements) and 50 principal components (the dimensions), we can create two series containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
 
-Installation
-------------
+Installation and loading # NC: Maybe this should be first?
+-------------------------
 
@@
 install_github("TGuillerme/dispRity", ref = "release")
+# Load the package
 library(dispRity)
@@
 
-Note that we use the `release` branch here which is version `0.2`. <!-- add version !-->
+Note this uses the `release` branch (version `0.2`). <!-- add version !-->
 For the piping-hot (but potentially full of bugs) version, you can change the argument `ref = release` to `ref = master`.
-This package depends mainly on the `ape` package and uses functions of several other (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, and `RCurl`).
+`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, and `RCurl`).
 
-Data
-----
+Datasets
+--------
+
+`dispRity` contains several datasets to cover the different use cases for the package. 
+These are:
+
+# NC: Describe all datasets briefly?
+- @beckancient2014
+-
+-
+-
 
 In this tutorial we are going to use a subset of the ordinated cladistic data from @beckancient2014 <!-- fix citation -->
-that contains 50 taxa (elements) ordinated using their cladistic distance (i.e. the distance between their discrete morphological characters).
-Note that this data is more oriented towards palaebiology analysis but that it can apply to other disciplines.
-Please refer to the [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes covering some specific example.
+that contains 50 mammalian taxa (elements) ordinated using their cladistic distance (i.e. the distances among their discrete morphological characters).
+For examples using other kinds of datasets please refer to the [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes.
+
+Loading and exploring the Beck and Lee data
+-------------------------------------------
+#NC: Not sure all this is needed here - maybe better to discuss what data is needed for specific analyses?
+Note that for a vignette it doesn't have to be the same example, or an analysis that runs all the way through - just examples of things you can do.
+I feel like if you want to make this multi disciplinary you could describe all four(?) datasets here then use them interchangeably below in the different function examples.
+
+#NC: Also perhaps better to think of a vignette as - here is something you might want to do, here is how to do it in R.
+Rather than here is a function in my package, here is what it does.
+That's what the help files are for...
 
 ```{r, fig.width=6, fig.height=6}
-## Loading the package
-library(dispRity)
-
 ## Setting the random seed for repeatability
@@
 
-A quick go through
-------------------
+Basic use of `dispRity`
+-----------------------
 
-Here is a really crude and quick analysis to go through the package, showing some of its features.
-Note that all these features will be discussed in more details below.
+# NC: I think here what is needed are quick examples for more than one kind of dataset. With explanation! I've made a start below to help.
 
+`dispRity` is very easy to use for a variety of analyses. 
+Below I will demonstrate its basic usage with a variety of datasets.
+Note that the power of `dispRity` lies in the ability to create custom disparity metrics etc.
+More details on these features, and the complexities of some of the functions, are provided later in this vignette.
+
+Example 1: Calculating and visualising disparity-through-time using Beck and Lee 201X data for extinct and extant eutherian mammal genera.
+
+Often in palaeontology and evolutionary biology we are interested in seeing how disparity (morphological diversity) has changed through time along a phylogeny. 
+To do this we need the following steps.
+1. Split the data so that 
+2. Estimate ancestral values
+3. Calculate disparity in each time slice
+4. Bootstrap
+5. Plot the results
+
+We can easily do this using `dispRity`.
+Note that we require the following data: a cladistic matrix, a phylogeny, .... and so on
+
+# Add the data loading bit here I think rather than above. Add setseed for each analysis example? Most people will not do all the examples at once.
+
+# Wonder if there could be a wrapper function for "disparity-through-time"? With options for the metric of course, and bootstrap or not etc.
 ```{r, fig.width=6, fig.height=6}
@@
 
-Package specificities
-=====================
+# NC: Move other examples here
 
-The `dispRity` objects
+Features of the `dispRity` package
+==================================
+
+`dispRity` objects
 -----------------------------------
 
-Disparity analysis can involve a lot of shuffling around with many matrices (especially when bootstrapping the data) which can be impractical to visualise and will quickly jam your `R` console.
-For example, we can have a look at the structure of the object created in the quick example:
+Disparity analysis involves a lot of manipulation of many matrices (especially when bootstrapping) which can be impractical to visualise and will quickly overwhelm your `R` console.
+Even the simple Beck and Lee 201X example above produces an object with > 72 lines of lists of lists of matrices!
 
-```{r, eval = FALSE}
-str(median_centroids)
-## That's a more than 72 lines of lists of lists of matrices!
-```
-
-Therefore this package proposes a specific class of object called `dispRity` objects.
-These objects allow to easily use S3 method functions such as `summary.dispRity` (just called as `summary`; see [summary](#summary) section) or `plot.dispRity` (just called as `plot`; see [plot](#plot) section).
-But also, this allows to use the S3 method for printing `dispRity` objects via `print.dispRity` that allows to summarise the content of the objects similar to the `phylo` class objects (see `ape::print.phylo`).
+Therefore `dispRity` uses a specific class of object called a `dispRity` object.
+These objects allow users to use S3 method functions such as `summary.dispRity` (see [summary](#summary) section), `plot.dispRity` (`plot`; see [plot](#plot) section) and `print.dispRity`. 
+# NC: I've simpiflied this as people who understand will know this already, people that don't won't care and don't need to know.
+`dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g.
+`sort.dispRity`, `extract.dispRity` see the full list in the [`dispRity` utilities](#dispRity.utilities) section).
+These functions modify the `dispRity` object without having to delve into its complex structure!
 The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).
 
+# NC: Explain the point of these examples here... Are they needed?
 ```{r}
@@
 
-Note however, that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:
+Note that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:
 
 ```{r, eval=FALSE}
-## Displaying the full object
+## Display the full object
 print(median_centroids, all = TRUE)
-## But that's more than ~ 2500 lines on a 13 inch laptop screen!
+## This is more than ~ 2500 lines on my 13 inch laptop screen!
 ```
 
-The package provides some utilities functions directly associated to the `dispRity` object manipulation such as `sort.dispRity`, `extract.dispRity` or `sort.dispRity` (see the full list in the [`dispRity` utilities](#dispRity.utilities) section).
-These functions come in handy for modifying the `dispRity` object without having to go into it's structure!
+# Modular functions - 
+# NC: not sure any more of these modular functions are relevant in the vignette? Maybe just mention the metrics? Remember a vignette is for users, not for you to show how the programming works.
 
-Modular functions
------------------
+# NC: I'd include a brief explanation of the include disparity metrics here.
+But yes keep the crazy "level" details for the other vignette on metrics.
+I think you want to avoid having too many separate vignettes really.
 
-This package aims to be a *modular package*.
-Its architecture is designed to be fairly easily modified by users.
-The modular aspect allows the package users to use several implemented tools to help them creating personalised functions.
-For example, the `dispRity` function intake a `metric` argument that is directly the function for calculating disparity should.
-This argument can take some already implemented functions such as `mean` but it can also intake some completely new ones such as the sum divided by length (even though this example sounds like reinventing the wheel):
+Disparity metrics
+---------------------------
+
+#
+
+Custom disparity metrics
+---------------------------
+
+`dispRity` is a *modular package*, i.e. its architecture is designed to be fairly easily modified by users so they can create personalised functions.
+The most relevant example of this is that `dispRity` allows users to implement custom disparity metrics when using the `dispRity` function via the `metric` argument.
+
+As a simple example, if one wanted to use the sum divided by length (i.e. the mean!) as a disparity metric, this could be done as follows.
 
 ```{r}
-## Disparity measured as the mean value of the ordinated matrix
-summary(dispRity(sliced_data, metric = mean))
-
-## A function for measuring the sum divided by the length
+## Create a function for measuring the sum divided by the length
 sum.divided.by.length <- function(X) sum(X)/length(X)
 
-## Disparity measured as the mean divided by the length
+## Calculate disparity using sum.divided.by.length
 summary(dispRity(sliced_data, metric = sum.divided.by.length))
+
+## This should match disparity obtained when we use the function mean...
+summary(dispRity(sliced_data, metric = mean))
 ```
@@
 For more information on the metrics, please refer to the [`dispRity` metric manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html) for the vignette about the metric implementation in `dispRity`.
+# NC: See comment above about moving a lot of this here.
 
@@
 
-`cust.series`
+# NC: Again I think this might be better written as "here are some things you might want to modify, here is how" rather than a function by function approach.
+
+`cust.series` # NC: I'd have called this custom.series. It's only 2 more letters and makes it much more readable, and less about custard.
 -------------
@@
 In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (all the living mammals and their direct ancestor) and the stem mammals (all the mammals that have no related living descendants).
-In order to do so, we need to separate the tree into two subtrees at the node 71 (see [phylogeny above](#mammal_tree)).
+To do so, we need to separate the tree into two subtrees at the node 71 (see [phylogeny above](#mammal_tree)).
 
@@
 We can then feed these lists of species to the `cust.series` function after transforming it into a factorial table:
-
+# NC: This is a fairly horrible set up to go through! Can't you make a function?
 ```{r}

### 2017-03-14 R/bhatt.coeff.R

@@ -2,10 +2,10 @@
 #'
 #' @title Bhattacharrya Coefficient
 #'
-#' @description Calculates the probability of overlap between to distributions.
+#' @description Calculates the probability of overlap between two distributions.
 #'
 #' @param x,y two distributions.
-#' @param bw the bandwith size, either a \code{numeric} or a \code{function} (see \code{\link[stats]{bw.nrd0}}).
+#' @param bw the bandwidth size, either a \code{numeric} or a \code{function} (see \code{\link[stats]{bw.nrd0}}).
 #' @param ... optional arguments to be passed to the \code{bw} argument.
 #'
 #' @examples
@@ -23,13 +23,13 @@
 #' @author Thomas Guillerme
 #' @export
 
-bhatt.coeff<-function(x, y, bw=bw.nrd0, ...) {
-    #SANITIZING
-    #x and y
+bhatt.coeff<-function(x, y, bw = bw.nrd0, ...) {
+    ## Check data
+    ## x and y
     check.class(x, "numeric")
     check.class(y, "numeric")
     
-    #bw
+    ## bw (bandwidth)
     if(class(bw) == "numeric") {
         check.length(bw, 1, " must be either a single numeric value or a function.")
          bw<-round(bw)
@@ -37,35 +37,38 @@ bhatt.coeff<-function(x, y, bw=bw.nrd0, ...) {
         check.class(bw, "function", " must be either a single numeric value or a function.")
     }
 
-    #BHATTACHARYYA COEFFICIENT
-    #sum(sqrt(x relative counts in bin_i * y relative counts in bin_i))
+    ## BHATTACHARYYA COEFFICIENT
+    ## sum(sqrt(x relative counts in bin_i * y relative counts in bin_i))
 
-    #Setting the right number of bins (i)
+    ## Setting the right number of bins (i)
     if(class(bw) == 'function') {
-        #Bin width
-        band.width<-bw(c(x,y), ...)
-        #Bin breaks
-        bin.breaks<-seq(from=min(c(x,y)), to=max(c(x,y)+band.width), by=band.width) #adding an extra bandwith to the max to be sure to include all the data
-        #Number of bins
-        bin.n<-length(bin.breaks)-1
+        ## Bin width
+        band.width <- bw(c(x, y), ...)
+        ## Bin breaks
+        ## adding an extra bandwith to the max to be sure to include all the data
[…]

### 2017-03-16 doc/dispRity-manual.Rmd
Added list of major comments to the desiderata list (from Natalie's comments on vignette)

@@
 
+# Important note:
+> This vignette is being step by step fully rewritten.
+Feel free to comment on the current stages of rewriting (what information would you like to see in here?) by raising and issue in GitHub or sending me an email (<guillert@tcd.ie>).
+Else, you can check the former version of this vignette [here](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html).
+
 `dispRity` is a package for calculating disparity in `R`. It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values so they can easily be compared. 
@@
 
-Before starting # NC: Not sure this qualifier is needed.
-===============
+Installation and loading
+-------------------------
+
+You can install this package easily if you use the latest version of `R` (> 3.0.0) and `devtools`.
+
+```{r, eval = FALSE}
+if(!require(devtools)) install.packages("devtools")
+install_github("TGuillerme/dispRity", ref = "release")
+# Load the package
+library(dispRity)
+```
+
+Note this uses the `release` branch (version `0.3`). <!-- add version !-->
+For the piping-hot (but potentially full of bugs) version, you can change the argument `ref = release` to `ref = master`.
+`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, and `RCurl`).
 
@@
 
-This package is designed to be used in multiple disciplines, including palaeonotology, ecology and evolution. Therefore some names or terms used in this tutorial might be unfamiliar.
+This package is designed to be used in multiple disciplines, including palaeonotology, ecology and evolution.
+Therefore some names or terms used in this tutorial might be unfamiliar.
 Below is a list of the meanings of these terms, in the context of `dispRity`.
@@
 
--   **Series**. Sub-samples of the ordinated space. #Change to subsample?
-    A series contains the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
+-   **Sub-samples**. Sub-samples of the ordinated space.
+    A sub-sample (or series) contains the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
     For example, if our ordinated space is composed of birds and mammals (the elements) and 50 principal components (the dimensions), we can create two series containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
 
-Installation and loading # NC: Maybe this should be first?
--------------------------
-
-You can install this package easily if you use the latest version of `R` (> 3.0.0) and `devtools`.
-
-```{r, eval = FALSE}
-if(!require(devtools)) install.packages("devtools")
-install_github("TGuillerme/dispRity", ref = "release")
-# Load the package
-library(dispRity)
-```
-
-Note this uses the `release` branch (version `0.2`). <!-- add version !-->
-For the piping-hot (but potentially full of bugs) version, you can change the argument `ref = release` to `ref = master`.
-`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, and `RCurl`).
+<!-- TG: good call for changing series to subsamples. NEED TO BE APPLIED TO THE WHOLE PACKAGE! -->
 
@@
 
-# NC: Describe all datasets briefly?
+<!-- # NC: Describe all datasets briefly?
+TG: Fair enough. Goes on my TODO list (is commented out for now).
+We might change it to our own dataset though.
 - @beckancient2014
@@
 -
+ -->
 
@@
 -------------------------------------------
-#NC: Not sure all this is needed here - maybe better to discuss what data is needed for specific analyses?
+<!-- #NC: Not sure all this is needed here - maybe better to discuss what data is needed for specific analyses?
 Note that for a vignette it doesn't have to be the same example, or an analysis that runs all the way through - just examples of things you can do.
@@
 That's what the help files are for...
+-->
 
@@
 
-# NC: I think here what is needed are quick examples for more than one kind of dataset. With explanation! I've made a start below to help.
+<!-- # NC: I think here what is needed are quick examples for more than one kind of dataset. With explanation! I've made a start below to help. -->
 
@@
 
-Example 1: Calculating and visualising disparity-through-time using Beck and Lee 201X data for extinct and extant eutherian mammal genera.
+Example 1: Calculating and visualising disparity-through-time using Beck and Lee 2014 data for extinct and extant eutherian mammal genera.
 
 Often in palaeontology and evolutionary biology we are interested in seeing how disparity (morphological diversity) has changed through time along a phylogeny. 
-To do this we need the following steps.
+To do this we need the following steps:
+
 1. Split the data so that 
@@
 
-We can easily do this using `dispRity`.
+<!-- We can easily do this using `dispRity`.
 Note that we require the following data: a cladistic matrix, a phylogeny, .... and so on
@@
 
-# Wonder if there could be a wrapper function for "disparity-through-time"? With options for the metric of course, and bootstrap or not etc.
+# Wonder if there could be a wrapper function for "disparity-through-time"? With options for the metric of course, and bootstrap or not etc. -->
+
 ```{r, fig.width=6, fig.height=6}
@@
 
-# NC: Move other examples here
+<!-- # NC: Move other examples here -->
 
@@
 Disparity analysis involves a lot of manipulation of many matrices (especially when bootstrapping) which can be impractical to visualise and will quickly overwhelm your `R` console.
-Even the simple Beck and Lee 201X example above produces an object with > 72 lines of lists of lists of matrices!
+Even the simple Beck and Lee 2014 example above produces an object with > 72 lines of lists of lists of matrices!
 
@@
 These objects allow users to use S3 method functions such as `summary.dispRity` (see [summary](#summary) section), `plot.dispRity` (`plot`; see [plot](#plot) section) and `print.dispRity`. 
-# NC: I've simpiflied this as people who understand will know this already, people that don't won't care and don't need to know.
+<!-- # NC: I've simpiflied this as people who understand will know this already, people that don't won't care and don't need to know.
+ -->
  `dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g.
@@
 
-# NC: Explain the point of these examples here... Are they needed?
+<!-- # NC: Explain the point of these examples here... Are they needed? -->
+
 ```{r}
@@
 
+<!-- 
 # Modular functions - 
@@
 I think you want to avoid having too many separate vignettes really.
+-->
 
@@
 For more information on the metrics, please refer to the [`dispRity` metric manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html) for the vignette about the metric implementation in `dispRity`.
-# NC: See comment above about moving a lot of this here.
+<!-- # NC: See comment above about moving a lot of this here. -->
 
@@
 
-# NC: Again I think this might be better written as "here are some things you might want to modify, here is how" rather than a function by function approach.
+<!-- # NC: Again I think this might be better written as "here are some things you might want to modify, here is how" rather than a function by function approach. -->
 
@@
 We can then feed these lists of species to the `cust.series` function after transforming it into a factorial table:
-# NC: This is a fairly horrible set up to go through! Can't you make a function?
+<!-- # NC: This is a fairly horrible set up to go through! Can't you make a function? -->
+
 ```{r}

### 2017-03-23 doc/dispRity-manual.Rmd
Changed "series" to "subsamples" across the whole package!

@@
 
--   **Sub-samples**. Sub-samples of the ordinated space.
-    A sub-sample (or series) contains the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
-    For example, if our ordinated space is composed of birds and mammals (the elements) and 50 principal components (the dimensions), we can create two series containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
-
-<!-- TG: good call for changing series to subsamples. NEED TO BE APPLIED TO THE WHOLE PACKAGE! -->
+-   **Subsamples**. Sub-samples of the ordinated space.
+    A sub-sample (or subsamples) contains the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
+    For example, if our ordinated space is composed of birds and mammals (the elements) and 50 principal components (the dimensions), we can create two subsamples containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
 
@@
 ```{r, fig.width=6, fig.height=6}
-## Splitting the data into time series
-sliced_data <- time.series(BeckLee_mat99, BeckLee_tree, method = "continuous",
+## Splitting the data into time subsamples
+sliced_data <- time.subsamples(BeckLee_mat99, BeckLee_tree, method = "continuous",
     model = "acctran", time = rev(seq(from=0, to=130, by=15)),
@@
 
-One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter called *series*, see [glossary](#glossary)).
-These series correspond to regions of the ordinated space that we want to compare to each other.
+One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter called *subsamples*, see [glossary](#glossary)).
+These subsamples correspond to regions of the ordinated space that we want to compare to each other.
 
@@
 
-`cust.series` # NC: I'd have called this custom.series. It's only 2 more letters and makes it much more readable, and less about custard.
+`custom.subsamples` # NC: I'd have called this custom.subsamples. It's only 2 more letters and makes it much more readable, and less about custard.
 -------------
 
-The `cust.series` function allows to create these series according to one or more factors determined by the user.
+The `custom.subsamples` function allows to create these subsamples according to one or more factors determined by the user.
 In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (all the living mammals and their direct ancestor) and the stem mammals (all the mammals that have no related living descendants).
@@
 
-We can then feed these lists of species to the `cust.series` function after transforming it into a factorial table:
+We can then feed these lists of species to the `custom.subsamples` function after transforming it into a factorial table:
 <!-- # NC: This is a fairly horrible set up to go through! Can't you make a function? -->
@@
 ## Creating the factor table
-series_factors <- as.data.frame(
+subsamples_factors <- as.data.frame(
     matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
     ncol = 1, dimnames = list(c(crown, stem), "Group")))
-head(series_factors)
+head(subsamples_factors)
 
-## Creating the two series
-crown_stem <- cust.series(BeckLee_mat50, series_factors)
+## Creating the two subsamples
+crown_stem <- custom.subsamples(BeckLee_mat50, subsamples_factors)
 ```
@@
 
-The two series represent two groups occupying specific regions of the mammalian morphospace.
-The `dispRity` object itself always contains the original morphospace (`$matrix`) as well as information on which taxa compose the series (`$series`):
+The two subsamples represent two groups occupying specific regions of the mammalian morphospace.
+The `dispRity` object itself always contains the original morphospace (`$matrix`) as well as information on which taxa compose the subsamples (`$subsamples`):
 
@@
 str(crown_stem$matrix)
-## And "series" the elements from this ordinated space into the two sub-spaces
-str(crown_stem$series)
+## And "subsamples" the elements from this ordinated space into the two sub-spaces
+str(crown_stem$subsamples)
 ```
@@
 ```{r}
-## Adding all mammals to the a new series (along with crown and stem as before)
-series_factors2 <- as.data.frame(
+## Adding all mammals to the a new subsamples (along with crown and stem as before)
+subsamples_factors2 <- as.data.frame(
     matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem)),
@@
 
-## Creating the three series: Group.stem/crown and Class.Mammalia
-cust.series(BeckLee_mat50, series_factors2)
+## Creating the three subsamples: Group.stem/crown and Class.Mammalia
+custom.subsamples(BeckLee_mat50, subsamples_factors2)
 ```
 
-`time.series`  <a name="time.series"></a>
+`time.subsamples`  <a name="time.subsamples"></a>
 -------------
@@
 Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
-The `time.series` function allows to create series containing all the elements present at specific points in time or during specific periods.
+The `time.subsamples` function allows to create subsamples containing all the elements present at specific points in time or during specific periods.
 This functions needs as input an ordinated space and a matching phylogenetic tree.
-Two types of time series can be performed by using the `method` option:
+Two types of time subsamples can be performed by using the `method` option:
 
- *  discrete time series (or time-binning) using `method = discrete`;
+ *  discrete time subsamples (or time-binning) using `method = discrete`;
 
- *  continuous time series (or time-slicing) using `method = continuous`.
+ *  continuous time subsamples (or time-slicing) using `method = continuous`.
 
@@
 ## Generating three time bins containing the data present every 40 Ma
-time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
     time = c(120, 80, 40, 0))
@@
 ## span between time bins.
-(time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
+(time_bins <- time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree,
     method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages))
@@
 
-This generated indeed a `dispRity` object of 3 series.
-Note that we can also generate equivalent results by just telling the function that we want three time-bins (series) as follow:
+This generated indeed a `dispRity` object of 3 subsamples.
+Note that we can also generate equivalent results by just telling the function that we want three time-bins (subsamples) as follow:
 
@@
 ## Automatically generate three equal length bins:
-time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
     time = 3)
@@
 ## Generating four time slices every 40 million years assuming a gradual evolutionary model
-(time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
+(time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
     method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
@@
 ## Note that in the same way as before, we can also automatically generate the slices
-time.series(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
+time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
     model = "gradual", time = 4)
@@
 
-Once we obtain our different series, it is possible to bootstrap and rarefy them (i.e. pseudo-replicating the data).
-The bootstrap will allow us to make each series more robust to outliers and the rarefaction will allow us to compare series with the same number of elements to get rid of eventual sampling problems.
+Once we obtain our different subsamples, it is possible to bootstrap and rarefy them (i.e. pseudo-replicating the data).
+The bootstrap will allow us to make each subsamples more robust to outliers and the rarefaction will allow us to compare subsamples with the same number of elements to get rid of eventual sampling problems.
 The `boot.matrix` allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
@@
 The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in each series to a minimum of 3 elements).
+remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in each subsamples to a minimum of 3 elements).
 It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
@@
 
-Of course, one could be interested in directly supplying the series generated above directly to this function.
-In fact, it can also deal with a list of matrices or with a `dispRity` object output from the `cust.series` or `time.series` functions.
+Of course, one could be interested in directly supplying the subsamples generated above directly to this function.
+In fact, it can also deal with a list of matrices or with a `dispRity` object output from the `custom.subsamples` or `time.subsamples` functions.
 
 ```{r}
-## Bootstrapping and fully rarefying the crown/stem series
+## Bootstrapping and fully rarefying the crown/stem subsamples
 (crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE))
 
-## Bootstrapping on the time binning/slicing series
+## Bootstrapping on the time binning/slicing subsamples
 (time_binsBS <- boot.matrix(time_bins, bootstraps = 100))
@@
 In these examples we used only simple ordinated matrix but of course, it possible to directly use the `dispRity` objects we generated in the steps above.
-Thus we can calculate the bootstrapped and rarefied disparity in each series of the ordinated space.
+Thus we can calculate the bootstrapped and rarefied disparity in each subsamples of the ordinated space.
 Here we are going to define disparity as being the median distance between each element and each dimensions' centre.
@@
 
-> Information about the number of elements per series as well as the observed (i.e. non-bootstrapped) disparity is also calculated.
+> Information about the number of elements per subsamples as well as the observed (i.e. non-bootstrapped) disparity is also calculated.
 This is specifically handy when rarefying the data for example
@@
 
-> This argument can be left missing. In this case, the algorithm will automatically detect the type of series from the `dispRity` object.
+> This argument can be left missing. In this case, the algorithm will automatically detect the type of subsamples from the `dispRity` object.
 
-It is also possible to display the number of elements per series (has a horizontal doted line) using the option `elements = TRUE`.
+It is also possible to display the number of elements per subsamples (has a horizontal doted line) using the option `elements = TRUE`.
 Additionally, when the data is rarefied, one can also indicate which level of rarefaction to display (i.e. display the results for a certain number of elements) by using the `rarefaction` argument.
@@
     elements = TRUE, col = c("black", rainbow(10)), ylab = c("Disparity", "Diversity"),
-    xlab = "Time (in in units from past to present)", time.series = FALSE,
+    xlab = "Time (in in units from past to present)", time.subsamples = FALSE,
     observed = TRUE, main = "Many options...")
@@
 
- *   `pairwise` (default): to compare each series pairwise
+ *   `pairwise` (default): to compare each subsamples pairwise
 
- *  `referential`: to compare each series to the first one
+ *  `referential`: to compare each subsamples to the first one
 
- *  `sequential`: to compare each series to the following one
+ *  `sequential`: to compare each subsamples to the following one
 
- *  `all`: to compare all the series together (like in analysis of variance)
+ *  `all`: to compare all the subsamples together (like in analysis of variance)
 
-It is also possible to input a list of pairs of `numeric` values or `characters` matching the series names to create personalised test.
+It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsamples names to create personalised test.
 Some other specific tests implemented in `dispRity` such as the `dispRity::null.test` and the `dispRity::sequential.test` have a specific way to be applied to the data and therefore ignore the `comparisons` argument.
@@
 ### `dispRity` object utilities  <a name="dispRity.utilities"></a>
-The first series of utilities are functions for manipulating `dispRity` objects:
+The first subsamples of utilities are functions for manipulating `dispRity` objects:
 
@@
 ## Extracting the 3rd boostrapped matrix with the 10th rarefaction level from the same group
-str(matrix.dispRity(disp_crown_stemBS, series = 1, bootstrap = 3, rarefaction = 10))
+str(matrix.dispRity(disp_crown_stemBS, subsamples = 1, bootstrap = 3, rarefaction = 10))
 ```
 
-#### `get.series.dispRity`
-This function creates a dispRity object that contains only elements from one specific series.
+#### `get.subsamples.dispRity`
+This function creates a dispRity object that contains only elements from one specific subsamples.
 
@@
 ## Extracting all the data for the crown mammals
-(crown_mammals <- get.series.dispRity(disp_crown_stemBS, "Group.crown"))
-## The object keeps the properties of the parent object but is composed of only one series
-length(crown_mammals$series)
+(crown_mammals <- get.subsamples.dispRity(disp_crown_stemBS, "Group.crown"))
+## The object keeps the properties of the parent object but is composed of only one subsamples
+length(crown_mammals$subsamples)
 ```
@@
 ## Extracting the disparity from the bootstrapped values from the 10th rarefaction level from the
-## second series (stem mammals)
-extract.dispRity(disp_crown_stemBS, observed = FALSE, series = 2, rarefaction = 10)
+## second subsamples (stem mammals)
+extract.dispRity(disp_crown_stemBS, observed = FALSE, subsamples = 2, rarefaction = 10)
 ```
@@
 ```{r}
-## Getting the disparity values of the time series
+## Getting the disparity values of the time subsamples
 summary(disp_time_slicesBS)
@@
 #### `sort.dispRity`
-This is the S3 method of `sort` for sorting the series alphabetically (default) or following a specific pattern.
+This is the S3 method of `sort` for sorting the subsamples alphabetically (default) or following a specific pattern.
 
 ```{r}
-## Sorting the disparity series in inverse alphabetic order
+## Sorting the disparity subsamples in inverse alphabetic order
 summary(sort(disp_time_slicesBS, decreasing = TRUE))
@@
 
-This function is at the core of `time.series` when using the time slicing method.
-It allows to slice a phylogenetic tree at a certain time point and selecting the nodes or tips following the models detailed in the `time.series` [section above](#time.series).
+This function is at the core of `time.subsamples` when using the time slicing method.
+It allows to slice a phylogenetic tree at a certain time point and selecting the nodes or tips following the models detailed in the `time.subsamples` [section above](#time.subsamples).
 

### 2017-03-23 R/boot.matrix.R

@@ -4,7 +4,7 @@
 #'
 #' @usage boot.matrix(data, bootstraps = 1000, rarefaction = FALSE, dimensions, verbose = FALSE, boot.type = "full", parallel)
 #' 
-#' @param data An ordinated matrix of maximal dimensions \eqn{k*(k-1)} or a list of matrices (typically output from \link{time.series} or \link{cust.series}).
+#' @param data An ordinated matrix of maximal dimensions \eqn{k*(k-1)} or a list of matrices (typically output from \link{time.subsamples} or \link{cust.subsamples}).
 #' @param bootstraps The number of bootstrap pseudoreplicates (\code{default = 1000}).
 #' @param rarefaction Either a \code{logical} value whether to fully rarefy the data or a set of \code{numeric} values used to rarefy the data (see details).
 #' @param dimensions Optional, a \code{numeric} value or proportion of the dimensions to keep.
@@ -16,19 +16,19 @@
 #' This function outputs a \code{dispRity} object containing:
 #' \item{data}{A \code{list} of the observed and bootstrapped matrices.}
 #' \item{elements}{A \code{vector} containing all the names of the elements from the original matrix.}
-#' \item{series}{A \code{vector} containing the name of the series (is \code{"1"} if the input was a single \code{matrix}).}
+#' \item{subsamples}{A \code{vector} containing the name of the subsamples (is \code{"1"} if the input was a single \code{matrix}).}
 #' \item{call}{A \code{vector} containing the arguments used for the bootstrapping.}
 #' \code{dispRity} objects can be summarised using \code{print} (S3).
 #'
 #' @details  
-#' \code{rarefaction}: when the input is \code{numeric}, the number of elements is set to the value(s) for each bootstrap. If some series have fewer elements than the rarefaction value, the series is not rarefied.
+#' \code{rarefaction}: when the input is \code{numeric}, the number of elements is set to the value(s) for each bootstrap. If some subsamples have fewer elements than the rarefaction value, the subsamples is not rarefied.
 #' \code{boot.type}: the different bootstrap algorithms are:
 #' \itemize{
 #'   \item \code{"full"}: resamples all the rows of the matrix and replaces them with a new random sample of rows (with \code{replace = TRUE}, meaning all the elements can be duplicated in each bootstrap).
[…]

### 2017-03-23 R/custom.subsamples_fun.R

@@ -2,14 +2,14 @@
 split.elements <- function(one_factor, data) {
 
     ## lapply fun for selecting the elements
-    select.elements <- function(series, one_factor) {
-        return(list("elements" = as.matrix(which(as.character(one_factor) == as.character(levels(as.factor(one_factor))[[series]])))))
+    select.elements <- function(subsamples, one_factor) {
+        return(list("elements" = as.matrix(which(as.character(one_factor) == as.character(levels(as.factor(one_factor))[[subsamples]])))))
     }
 
-    ## Select the elements per series
+    ## Select the elements per subsamples
     selected_elements <- lapply(as.list(1:length(levels(as.factor(one_factor)))), select.elements, one_factor)
 
-    ## Adding the names to the series
+    ## Adding the names to the subsamples
     names(selected_elements) <- levels(as.factor(one_factor))
 
     ## Output

### 2017-03-23 R/dispRity.R

@@ -1,6 +1,6 @@
 #' @title Calculates disparity from an ordinated matrix.
 #'
-#' @description Calculates disparity on an ordinated matrix or series of matrices, where the disparity metric can be user specified.
+#' @description Calculates disparity on an ordinated matrix or subsamples of matrices, where the disparity metric can be user specified.
 #'
 #' @param data An ordinated matrix of maximal dimensions \eqn{k*(k-1)}, or a \code{dispRity} object (see details).
 #' @param metric A vector containing one to three functions. At least of must be a "level 1" or a "level 2" function (see details).
@@ -13,13 +13,13 @@
 #' \item{data}{A \code{list} of the observed and bootstrapped matrices.}
 #' \item{disparity}{A \code{list} of disparity values (containing the observed disparity and the eventual bootstrapped one).}
 #' \item{elements}{A \code{vector} containing all the names of the elements from the original matrix.}
-#' \item{series}{A \code{vector} containing the name of the series (is \code{"1"} if the input was a single \code{matrix}).}
+#' \item{subsamples}{A \code{vector} containing the name of the subsamples (is \code{"1"} if the input was a single \code{matrix}).}
 #' \item{call}{A \code{vector} containing the arguments used for the bootstrapping.}
 #' \code{dispRity} objects can be summarised using \code{print} (S3).
 #' Use \link{summary.dispRity} to summarise the \code{dispRity} object.
 #' 
 #' @details  
-#' The \code{dispRity} object given to the \code{data} argument can be: a list of matrices (typically output from the functions \code{\link{time.series}} or \code{\link{cust.series}}), a bootstrapped matrix output from \code{\link{boot.matrix}} or a list of disparity measurements calculated from this \code{dispRity} function.
+#' The \code{dispRity} object given to the \code{data} argument can be: a list of matrices (typically output from the functions \code{\link{time.subsamples}} or \code{\link{cust.subsamples}}), a bootstrapped matrix output from \code{\link{boot.matrix}} or a list of disparity measurements calculated from this \code{dispRity} function.
 #' 
 #' \code{metric} should be input as a vector of functions.
[…]

### 2017-03-23 R/null.test.R

@@ -25,13 +25,13 @@
 #' results ; plot(results)
 #' 
 #' \dontrun{
-#' ## Running the test on multiple series
-#' ## Generating the series
+#' ## Running the test on multiple subsamples
+#' ## Generating the subsamples
 #' factors <- as.data.frame(matrix(data = c(rep(1, 12), rep(2, 13), rep(3, 12),
 #'      rep(4, 13)), dimnames = list(rownames(BeckLee_mat50))), ncol = 1)
-#' customised_series <- cust.series(BeckLee_mat50, factors)
+#' customised_subsamples <- cust.subsamples(BeckLee_mat50, factors)
 #' ## Bootstrapping the data
-#' bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100)
+#' bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 100)
 #' ## Calculating variances of each dimension
 #' sum_variances <- dispRity(bootstrapped_data, metric = c(sum, variances))
 #' ## Testing against normal distribution
@@ -52,8 +52,8 @@
 # source("null.test_fun.R")
 # data(BeckLee_mat50)
 # factors <- as.data.frame(matrix(data = c(rep(1, 12), rep(2, 13), rep(3, 25)), dimnames =list(rownames(BeckLee_mat50))), ncol = 1)
-# customised_series <- cust.series(BeckLee_mat50, factors)
-# bootstrapped_data <- boot.matrix(customised_series, bootstraps = 10)
+# customised_subsamples <- cust.subsamples(BeckLee_mat50, factors)
+# bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 10)
 # data_single <- dispRity(bootstrapped_data, metric = c(sum, variances))
 # data_multi <- dispRity(bootstrapped_data, metric = variances)
 # single <- dispRity(boot.matrix(BeckLee_mat50, bootstraps = 10), metric = c(sum, variances))
@@ -100,30 +100,30 @@ null.test <- function(data, replicates = 100, null.distrib, null.args = NULL, nu
     #NULL TESTING
 
     #Generating the null models
-    if(length(data$series) != 1) {
-        #Subdivide the data per series
-        sub_data <- lapply(as.list(seq(1:length(data$series))), function(X) get.series.dispRity(data, X))
-        #Apply the data to all series
+    if(length(data$subsamples) != 1) {
+        #Subdivide the data per subsamples
[…]

### 2017-03-23 R/plot.dispRity.R

@@ -3,16 +3,16 @@
 #' @description Plots a \code{dispRity} object.
 #'
 #' @param data A \code{dispRity} object.
-#' @param type Either \code{"continuous"} (\code{"c"}), \code{"box"} (\code{"b"}), \code{"line"} (\code{"l"}) or \code{"polygon"} (\code{"p"}). When unspecified, is set to \code{"continuous"} if \code{\link{time.series}} is used with \code{method = "continuous"}, else is set to \code{"box"}. See details.
+#' @param type Either \code{"continuous"} (\code{"c"}), \code{"box"} (\code{"b"}), \code{"line"} (\code{"l"}) or \code{"polygon"} (\code{"p"}). When unspecified, is set to \code{"continuous"} if \code{\link{time.subsamples}} is used with \code{method = "continuous"}, else is set to \code{"box"}. See details.
 #' @param quantiles The quantiles to display (default is \code{quantiles = c(50, 95)}; is ignored if the \code{dispRity} object is not bootstrapped).
 #' @param cent.tend A function for summarising the bootstrapped disparity values (default is \code{\link[stats]{median}}).
 #' @param rarefaction Either \code{NULL} (default) or \code{FALSE} for not using the rarefaction scores; an \code{numeric} value of the level of rarefaction to plot; or \code{TRUE} for plotting the rarefaction curves.
-#' @param elements \code{logical} whether to plot the number of elements per series.
+#' @param elements \code{logical} whether to plot the number of elements per subsamples.
 #' @param ylim Optional, two \code{numeric} values for the range of the y axis.
 #' @param xlab Optional, a \code{character} string for the caption of the x axis.
 #' @param ylab Optional, one or two (if \code{elements = TRUE}) \code{character} string(s) for the caption of the y axis.
 #' @param col Optional, some \code{character} string(s) for the colour of the graph.
-#' @param time.series \code{logical} whether to handle continuous data from the \code{time.series} function as time (in Ma). When this option is set to TRUE for other \code{type} options, the names of the series are used for the x axis labels.
+#' @param time.subsamples \code{logical} whether to handle continuous data from the \code{time.subsamples} function as time (in Ma). When this option is set to TRUE for other \code{type} options, the names of the subsamples are used for the x axis labels.
[…]

### 2017-03-23 R/sequential.test.R

@@ -3,9 +3,9 @@
 #' @title Sequential linear regressions
 #'
 #' @description Function still in development.
-# ' @description Performs a sequential \code{\link[stats]{glm}} on the series by correcting for time autocorrelation. 
+# ' @description Performs a sequential \code{\link[stats]{glm}} on the subsamples by correcting for time autocorrelation. 
 #'
-# ' @param series time series of which to estimate the slopes sequentially.
+# ' @param subsamples time subsamples of which to estimate the slopes sequentially.
 # ' @param family the family of the \code{\link[stats]{glm}}.
 # ' @param correction optional, which p-value correction to apply (see \code{\link[stats]{p.adjust}}). If missing, no correction is applied.
 # ' @param call optional, a call from a \code{dispRity} object.
@@ -17,31 +17,31 @@
 # ' @examples
 # ' ## Load the Beck & Lee 2014 data
 # ' data(BeckLee_mat50)
-# ' ## Calculating the disparity from a customised series
-# ' ## Generating the series
+# ' ## Calculating the disparity from a customised subsamples
+# ' ## Generating the subsamples
 # ' factors <- as.data.frame(matrix(data = c(rep(1, 12), rep(2, 13), rep(3, 12),
 # '      rep(4, 13)), dimnames = list(rownames(BeckLee_mat50))), ncol = 1)
-# ' customised_series <- cust.series(BeckLee_mat50, factors)
+# ' customised_subsamples <- cust.subsamples(BeckLee_mat50, factors)
 # ' ## Bootstrapping the data
-# ' bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100)
+# ' bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 100)
 # ' ## Calculating variances of each dimension
 # ' dim_variances <- dispRity(bootstrapped_data, metric = variances)
-# ' ## Extracting the disparity values of each series
-# ' series <- extract.dispRity(dim_variances, observed = FALSE,
+# ' ## Extracting the disparity values of each subsamples
+# ' subsamples <- extract.dispRity(dim_variances, observed = FALSE,
 # '      keep.structure = TRUE, concatenate = TRUE)
 # '
-# ' ## Running a gaussian sequential test on the series
-# ' results <- sequential.test(series, family = gaussian)
+# ' ## Running a gaussian sequential test on the subsamples
+# ' results <- sequential.test(subsamples, family = gaussian)
 # ' ## Summarising the results
[…]

### 2017-03-23 R/test.dispRity.R

@@ -4,7 +4,7 @@
 #'
 #' @param data A \code{dispRity} object.
 #' @param test A test \code{function} to apply to the data.
-#' @param comparisons If data contains more than two series, the type of comparisons to apply: either \code{"pairwise"} (default), \code{"referential"}, \code{"sequential"}, \code{"all"} or a list of pairs of series names/number to compare (see details).
+#' @param comparisons If data contains more than two subsamples, the type of comparisons to apply: either \code{"pairwise"} (default), \code{"referential"}, \code{"sequential"}, \code{"all"} or a list of pairs of subsamples names/number to compare (see details).
 #' @param rarefaction A \code{numeric} value indicating whether to use a specific rarefaction level (default = \code{NULL}).
 #' @param correction Which p-value correction to apply to \code{htest} category test (see \code{\link[stats]{p.adjust}}). If missing, no correction is applied.
 #' @param concatenate Logical, whether to concatenate eventual bootstrapped disparity values (\code{TRUE}; default) or to apply the test to each bootstrapped values individually (\code{FALSE}).
@@ -15,11 +15,11 @@
 #' @details  
 #' The \code{comparison} argument can be:
 #' \itemize{
-#'   \item \code{"pairwise"}: pairwise comparisons of all the series (default).
-#'   \item \code{"referential"}: compares the first series to all the others.
-#'   \item \code{"sequential"}: compares each series sequentially (e.g. first against second, second against third, etc.).
-#'   \item \code{"all"}: compares all the series simultaneously to the data (i.e. \code{bootstrapped disparity ~ series names}). This argument is used for \code{\link[stats]{aov}} or \code{\link[stats]{glm}} type tests.
-#'   \item A list of pairs of number of series to compare. Each element of the the list must contain two elements
+#'   \item \code{"pairwise"}: pairwise comparisons of all the subsamples (default).
+#'   \item \code{"referential"}: compares the first subsamples to all the others.
+#'   \item \code{"sequential"}: compares each subsamples sequentially (e.g. first against second, second against third, etc.).
[…]

### 2017-03-23 R/test.dispRity_fun.R

@@ -1,70 +1,70 @@
-#Mapply function for applying the tests to distribution data
+## Mapply function for applying the tests to distribution data
 test.mapply <- function(pair_comparison, data, test, ...) {
-    #Check if any is constant
-    # uniques1 <- which(unlist(lapply(data[[pair_comparison[[1]]]], function(X) return(length(unique(X))))) == 1)
-    # uniques2 <- which(unlist(lapply(data[[pair_comparison[[2]]]], function(X) return(length(unique(X))))) == 1)
+    ## Check if any is constant
+    ##  uniques1 <- which(unlist(lapply(data[[pair_comparison[[1]]]], function(X) return(length(unique(X))))) == 1)
+    ##  uniques2 <- which(unlist(lapply(data[[pair_comparison[[2]]]], function(X) return(length(unique(X))))) == 1)
 
-    # if(is.na(any(match(uniques1, uniques2)))) {
+    ##  if(is.na(any(match(uniques1, uniques2)))) {
         tests <- mapply(test, data[[pair_comparison[[1]]]], data[[pair_comparison[[2]]]], MoreArgs = ..., SIMPLIFY = FALSE)
         return(tests)
-    # } else {
-    #     warning("Pair of series ", paste(pair_comparison, collapse = " and "), " contains not enough unique values.")
-    # }
+    ##  } else {
+    ##      warning("Pair of subsamples ", paste(pair_comparison, collapse = " and "), " contains not enough unique values.")
+    ##  }
 }
 
-#transforming test into a lapply from a given list of comp for multiple distributions
+## transforming test into a lapply from a given list of comp for multiple distributions
 test.list.lapply.distributions <- function(list_of_comp, data, test, ...) {
     output <- lapply(list_of_comp, test.mapply, data, test, ...)
     return(output)
 }
 
-#Creating a list of sequences
+## Creating a list of sequences
 set.sequence <- function(length) {
-    #Sequence of only 2
+    ## Sequence of only 2
     if(length==2) {
         output <- matrix(data=c(1,2), nrow=2, byrow=TRUE)
     } else {
-    #sequence of more
+    ## sequence of more
         output <- matrix(data=c(1:(length-1), 2:length), nrow=2, byrow=TRUE)
     }
     return(output)
 }
 
-#convert a list from character to numeric
+## convert a list from character to numeric
 convert.to.numeric <- function(list, object) {
     return(lapply(list, match, names(object)))
 }
 
-#convert a list from numeric to character
+## convert a list from numeric to character
[…]

### 2017-03-23 R/time.subsamples.R

@@ -1,15 +1,16 @@
-#' @title Separating ordinated data in time series.
+#' @title Separating ordinated data in time subsamples.
+#' @aliases time.series
 #'
-#' @description Splits the ordinated data into a time series list.
+#' @description Splits the ordinated data into a time subsamples list.
 #' 
-#' @usage time.series(data, tree, method, time, model, inc.nodes = FALSE, FADLAD, verbose = FALSE)
+#' @usage time.subsamples(data, tree, method, time, model, inc.nodes = FALSE, FADLAD, verbose = FALSE)
 #'
 #' @param data An ordinated matrix of maximal dimensions \eqn{k*(k-1)}.
 #' @param tree A \code{phylo} object matching the data and with a \code{root.time} element.
-#' @param method The time series method: either \code{"discrete"} (or \code{"d"}) or \code{"continuous"} (or \code{"c"}).
+#' @param method The time subsamples method: either \code{"discrete"} (or \code{"d"}) or \code{"continuous"} (or \code{"c"}).
 #' @param time Either a single \code{integer} for them number of discrete or continuous samples or a \code{vector} containing the age of each sample.
 #' @param model One of the following models: \code{"acctran"}, \code{"deltran"}, \code{"punctuated"} or \code{"gradual"}. Is ignored if \code{method = "discrete"}.
-#' @param inc.nodes A \code{logical} value indicating whether nodes should be included in the time series. Is ignored if \code{method = "continuous"}.
+#' @param inc.nodes A \code{logical} value indicating whether nodes should be included in the time subsamples. Is ignored if \code{method = "continuous"}.
 #' @param FADLAD An optional \code{data.frame} containing the first and last occurrence data.
 #' @param verbose A \code{logical} value indicating whether to be verbose or not. Is ignored if \code{method = "discrete"}.
 #'
@@ -17,11 +18,11 @@
 #' This function outputs a \code{dispRity} object containing:
 #' \item{data}{A \code{list} of the split ordinated data (each element is a \code{matrix}).}
 #' \item{elements}{A \code{vector} containing all the rownames from the input matrix.}
-#' \item{series}{A \code{vector} containing the name of the series.}
+#' \item{subsamples}{A \code{vector} containing the name of the subsamples.}
[…]

### 2017-03-23 tests/testthat/test-morpho.utilities.R

@@ -1,4 +1,4 @@
-#TESTING cust.series
+#TESTING morpho.utilities
 
 context("morpho.utilities")
 

### 2017-03-23 tests/testthat/test-utilities.dispRity.R

@@ -14,7 +14,7 @@ test_that("make.matrix", {
         test1$call
         ,"list")
     expect_is(
-        test1$series
+        test1$subsamples
         ,"list")
 
     test2 <- make.dispRity(data = matrix(rnorm(12), ncol = 3))
@@ -29,7 +29,7 @@ test_that("make.matrix", {
         test2$call
         ,"list")
     expect_is(
-        test2$series
+        test2$subsamples
         ,"list")
     expect_equal(
         length(unlist(test2))
@@ -54,7 +54,7 @@ test_that("fill.dispRity", {
         test$call
         ,"list")
     expect_is(
-        test$series
+        test$subsamples
         ,"list")
 
 
@@ -65,7 +65,7 @@ test_that("fill.dispRity", {
         test$call$dimensions
         , ncol(test$matrix))
     expect_equal(
-        as.vector(test$series[[1]]$elements)
+        as.vector(test$subsamples[[1]]$elements)
         , 1:nrow(test$matrix))
 })
 
@@ -75,13 +75,13 @@ test_that("matrix.dispRity", {
     ## Load the Beck & Lee 2014 data
     data(BeckLee_mat50)
 
-    ## Calculating the disparity from a customised series
-    ## Generating the series
+    ## Calculating the disparity from a customised subsamples
+    ## Generating the subsamples
     factors <- as.data.frame(matrix(data = c(rep(1, nrow(BeckLee_mat50)/2), rep(2, nrow(BeckLee_mat50)/2)), nrow = nrow(BeckLee_mat50), ncol = 1, dimnames = list(rownames(BeckLee_mat50))))
-    customised_series <- cust.series(BeckLee_mat50, factors)
+    customised_subsamples <- custom.subsamples(BeckLee_mat50, factors)
     ## Bootstrapping and rarefying the data
     set.seed(1)
-    dispRity_data <- boot.matrix(customised_series, bootstraps = 100,rarefaction = c(15, 10))
+    dispRity_data <- boot.matrix(customised_subsamples, bootstraps = 100,rarefaction = c(15, 10))
 
 
     expect_error(
@@ -91,40 +91,40 @@ test_that("matrix.dispRity", {
         all(matrix.dispRity(dispRity_data) == BeckLee_mat50)
         )
     expect_equal(
-        dim(matrix.dispRity(dispRity_data, series = 2))
+        dim(matrix.dispRity(dispRity_data, subsamples = 2))
         , c(25, 48))
     expect_equal(
[…]

### 2017-05-28 doc/dispRity-manual.Rmd
Updates on the manual for version 0.4

@@
 
-# Important note:
-> This vignette is being step by step fully rewritten.
-Feel free to comment on the current stages of rewriting (what information would you like to see in here?) by raising and issue in GitHub or sending me an email (<guillert@tcd.ie>).
-Else, you can check the former version of this vignette [here](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html).
+# `dispRity`
 
-`dispRity` is a package for calculating disparity in `R`. It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values so they can easily be compared. 
+This is a package for measuring disparity in `R`. It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values so they can easily be compared.
 <!-- Add version and table of content here! -->
 
-Installation and loading
--------------------------
+# Introduction
+
+## What's `dispRity`
+This is a package for measuring disparity in `R`.
+It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
+Typically, these analysis are used in palaeobiology to study the changes in morphology through time.
+
+## Installing and running the package
 
@@
 ```{r, eval = FALSE}
+## Checking if devtools is already installed
 if(!require(devtools)) install.packages("devtools")
+
+## Installing the latest released version directly from GitHub
 install_github("TGuillerme/dispRity", ref = "release")
-# Load the package
+
+## Loading the package
 library(dispRity)
@@
 
-Note this uses the `release` branch (version `0.3`). <!-- add version !-->
+```{r, eval = TRUE, echo = FALSE, message = FALSE}
+## Checking if Claddis is installed BUILD ONLY
+library(dispRity)
+```
+
+Note this uses the `release` branch (version`0.4`). <!-- add version !-->
 For the piping-hot (but potentially full of bugs) version, you can change the argument `ref = release` to `ref = master`.
@@
 
-Glossary <a name="glossary"></a>
---------
+## Why not CRAN
 
-This package is designed to be used in multiple disciplines, including palaeonotology, ecology and evolution.
-Therefore some names or terms used in this tutorial might be unfamiliar.
-Below is a list of the meanings of these terms, in the context of `dispRity`.
+This package is not readily available on CRAN.
+This is mainly because some parts are still in development and that I prefer the reactivity of GitHub for implementing new suggestions from users.
+However, the package follows the strict (and useful!) CRAN standards via Travis. <!-- add link !-->
+
+## Help
+
+If you need help about the package, I hope that the following manual will be useful.
+If you still need more help, do not hesitate to drop me an email!
+
+Parts of this package are still in development and some other parts are probably not covered.
+Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
+
+# Getting started
+
+## Which data?
+
+Disparity must be measured from ordinated matrices.
+These matrices can be of any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your elements names as rows (taxa, experiments, countries, etc.) and your ordinations axis as columns (the dimensions of your dataset).
+
+### Ordination from `Claddis`
+
+```{r, eval = TRUE, echo = FALSE, message = FALSE}
+## Checking if Claddis is installed BUILD ONLY
+if(!require(Claddis)) install.packages("Claddis")
+```
+`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
+For this, simply input a matrix in the `Claddis` format to the function an it will automatically calculate and ordinate the distances among taxa:
+
+```{r}
+## Ordinating the example data from Claddis
+Claddis.ordination(Michaux1989) 
+```
+
+Note that some several options are available, namely which type of distance should be computed.
+See more info in the function manual (`?Claddis.ordination`).
+Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.
+
+### Ordination from `geomorph`
+
+> TBC!
+
+### General ordinations
+
+If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordinations by using the following functions from the `stats` package.
+Here is how to do it for the following types of matrices:
+
+ * Multivariate matrices
+
+```{r}
+## A multivariate matrix
+head(USArrests)
+
+## Ordinating the matrix using prcomp()
+ordination <- prcomp(USArrests)
+
+## Selecting the ordinated matrix
+ordinated_matrix <- ordination$x
+head(ordinated_matrix)
+```
+This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, have a look at the `?princomp` function.
+
+ * Distance matrices
+
+```{r}
+## A matrix of distances between cities
+str(eurodist)
+
+## Ordinating the matrix using cmdscale() with k = 5 dimensions
+ordinated_matrix <- cmdscale(eurodist, k = 5)
+head(ordinated_matrix)
+```
+This results in a ordinated matrix with European cities as elements and 5 dimensions.
+Note that the number of dimensions should not exceed the number of elements - 1.
+
+Of course any other methods for creating the ordination matrix is totally valid!
+The only requirements for `dispRity` is to use matrices with elements as rows and dimensions as columns...
+
+## Simple analysis
+
+Two functions allow to run templated analysis simply using an ordination matrix: either calculating the disparity through time or the disparity per groups.
+
+### Example data
+
+For illustrating the following functions, we will use data from Beck and Lee 2015. <!-- ADD PAPER CITE HERE -->
+This dataset contains an ordinated matrix of 50 mammals' discrete characters (called `BeckLee_mat50`), a matrix of the same 50 mammals and the estimate discrete data from their descendants (thus 50 + 49 rows, called `BeckLee_mat99`), a data frame containing the ages of each taxa in the dataset (called `BeckLee_ages`) and finally a phylogenetic tree with the relations of the 50 mammals (called `BeckLee_tree`).
+
+```{r, fig.width=6, fig.height=6}
+## Setting the random seed for repeatability
+set.seed(123)
+
+## Loading the ordinated matrices
+data(BeckLee_mat50)
+data(BeckLee_mat99)
+
+## The first 5 taxa and dimensions of the 50 taxa matrix
+head(BeckLee_mat50[,1:5])
+
+## The first 5 taxa and dimensions of the 99 taxa + ancestors matrix
+head(BeckLee_mat99[,1:5])
+tail(BeckLee_mat99[,1:5])
+
+## Loading a list of first and last occurrence data
+data(BeckLee_ages)
+head(BeckLee_ages)
+
+## Loading the phylogeny
+data(BeckLee_tree)
+plot(BeckLee_tree, cex = 0.8) ; axisPhylo(root = 140) ; nodelabels(cex = 0.5)
+```
+
+Note of course that you can use your data obtain from the previous chapter!
+
+### Disparity through time
+
+This
+
+```{r}
+# dispRity.through.time()
+```
+
+### Disparity between groups
+
+```{r}
+# dispRity.per.group()
+```
+
+# Specific tutorials
+
+# Glossary <a name="glossary"></a>
 
@@
 
-Datasets
---------
+# The package's guts
 
-`dispRity` contains several datasets to cover the different use cases for the package. 
-These are:
+## Running `dispRity` in parallel
 
-<!-- # NC: Describe all datasets briefly?
-TG: Fair enough. Goes on my TODO list (is commented out for now).
-We might change it to our own dataset though.
-- @beckancient2014
--
--
--
- -->
+## `dispRity` object structure
 
-In this tutorial we are going to use a subset of the ordinated cladistic data from @beckancient2014 <!-- fix citation -->
-that contains 50 mammalian taxa (elements) ordinated using their cladistic distance (i.e. the distances among their discrete morphological characters).
-For examples using other kinds of datasets please refer to the [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes.
+## Modularity
 
-Loading and exploring the Beck and Lee data
--------------------------------------------
-<!-- #NC: Not sure all this is needed here - maybe better to discuss what data is needed for specific analyses?
-Note that for a vignette it doesn't have to be the same example, or an analysis that runs all the way through - just examples of things you can do.
-I feel like if you want to make this multi disciplinary you could describe all four(?) datasets here then use them interchangeably below in the different function examples.
+### Making your own metrics
 
-#NC: Also perhaps better to think of a vignette as - here is something you might want to do, here is how to do it in R.
-Rather than here is a function in my package, here is what it does.
-That's what the help files are for...
--->
+### Making your own tests
 
-```{r, fig.width=6, fig.height=6}
-## Setting the random seed for repeatability
-set.seed(123)
-
-## Loading the ordinated matrix containing 50 taxa
-data(BeckLee_mat50)
-dim(BeckLee_mat50)
-head(BeckLee_mat50[,1:5])
-
-## Loading another ordinated matrix containing 50 tips + 49 nodes
-data(BeckLee_mat99)
-dim(BeckLee_mat99)
-head(BeckLee_mat99[,1:5], 2)
-tail(BeckLee_mat99[,1:5], 2)
-
-## Loading a list of first and last occurrence data
-data(BeckLee_ages)
-head(BeckLee_ages)
-
-## Loading the phylogeny
-data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.8) ; axisPhylo(root = 140) ; nodelabels(cex = 0.5)
-```
-<a name="mammal_tree"></a>
-
-Basic use of `dispRity`
------------------------
-
-<!-- # NC: I think here what is needed are quick examples for more than one kind of dataset. With explanation! I've made a start below to help. -->
-
-`dispRity` is very easy to use for a variety of analyses. 
-Below I will demonstrate its basic usage with a variety of datasets.
-Note that the power of `dispRity` lies in the ability to create custom disparity metrics etc.
-More details on these features, and the complexities of some of the functions, are provided later in this vignette.
-
-Example 1: Calculating and visualising disparity-through-time using Beck and Lee 2014 data for extinct and extant eutherian mammal genera.
-
-Often in palaeontology and evolutionary biology we are interested in seeing how disparity (morphological diversity) has changed through time along a phylogeny. 
-To do this we need the following steps:
-
-1. Split the data so that 
-2. Estimate ancestral values
-3. Calculate disparity in each time slice
-4. Bootstrap
-5. Plot the results
-
-<!-- We can easily do this using `dispRity`.
-Note that we require the following data: a cladistic matrix, a phylogeny, .... and so on
-
-# Add the data loading bit here I think rather than above. Add setseed for each analysis example? Most people will not do all the examples at once.
-
-# Wonder if there could be a wrapper function for "disparity-through-time"? With options for the metric of course, and bootstrap or not etc. -->
-
-```{r, fig.width=6, fig.height=6}
-## Splitting the data into time subsamples
-sliced_data <- time.subsamples(BeckLee_mat99, BeckLee_tree, method = "continuous",
-    model = "acctran", time = rev(seq(from=0, to=130, by=15)),
-    FADLAD = BeckLee_ages)
-
-## Bootstrapping the data to measure its variance
-bootstrapped_data <- boot.matrix(sliced_data, 100)
-
-## Calculating disparity as the median distance from each element to the groups centroids
-median_centroids <- dispRity(bootstrapped_data, metric = c(median, centroids))
-
-## Summarising the disparity results
-summary(median_centroids)
-plot(median_centroids)
-
-## Testing the effect of time on disparity (note that using an ANOVA violates some assumptions!)
-summary(test.dispRity(median_centroids, test = aov, comparisons = "all"))
-```
-
-<!-- # NC: Move other examples here -->
-
-Features of the `dispRity` package
-==================================
-
-`dispRity` objects
------------------------------------
-
-Disparity analysis involves a lot of manipulation of many matrices (especially when bootstrapping) which can be impractical to visualise and will quickly overwhelm your `R` console.
-Even the simple Beck and Lee 2014 example above produces an object with > 72 lines of lists of lists of matrices!
-
-Therefore `dispRity` uses a specific class of object called a `dispRity` object.
-These objects allow users to use S3 method functions such as `summary.dispRity` (see [summary](#summary) section), `plot.dispRity` (`plot`; see [plot](#plot) section) and `print.dispRity`. 
-<!-- # NC: I've simpiflied this as people who understand will know this already, people that don't won't care and don't need to know.
- -->
- `dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g.
-`sort.dispRity`, `extract.dispRity` see the full list in the [`dispRity` utilities](#dispRity.utilities) section).
-These functions modify the `dispRity` object without having to delve into its complex structure!
-The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).
-
-<!-- # NC: Explain the point of these examples here... Are they needed? -->
-
-```{r}
-## What is the class of the median_centroids object?
-class(median_centroids)
-
-## What does the object contain?
-names(median_centroids)
-
-## Summarising it using the S3 method print.dispRity
-median_centroids
-```
-
-Note that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:
-
-```{r, eval=FALSE}
-## Display the full object
-print(median_centroids, all = TRUE)
-## This is more than ~ 2500 lines on my 13 inch laptop screen!
-```
-
-<!-- 
-# Modular functions - 
-# NC: not sure any more of these modular functions are relevant in the vignette? Maybe just mention the metrics? Remember a vignette is for users, not for you to show how the programming works.
-
-# NC: I'd include a brief explanation of the include disparity metrics here.
-But yes keep the crazy "level" details for the other vignette on metrics.
-I think you want to avoid having too many separate vignettes really.
--->
-
-Disparity metrics
----------------------------
-
-#
-
-Custom disparity metrics
----------------------------
-
-`dispRity` is a *modular package*, i.e. its architecture is designed to be fairly easily modified by users so they can create personalised functions.
-The most relevant example of this is that `dispRity` allows users to implement custom disparity metrics when using the `dispRity` function via the `metric` argument.
-
-As a simple example, if one wanted to use the sum divided by length (i.e. the mean!) as a disparity metric, this could be done as follows.
-
-```{r}
-## Create a function for measuring the sum divided by the length
-sum.divided.by.length <- function(X) sum(X)/length(X)
-
-## Calculate disparity using sum.divided.by.length
-summary(dispRity(sliced_data, metric = sum.divided.by.length))
-
-## This should match disparity obtained when we use the function mean...
-summary(dispRity(sliced_data, metric = mean))
-```
-
-For more information on the metrics, please refer to the [`dispRity` metric manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html) for the vignette about the metric implementation in `dispRity`.
-<!-- # NC: See comment above about moving a lot of this here. -->
-
-Functions
-=========
-
-In the following section, different functionalities of the package are detailed.
-The functions are presented in a "classic" pipelined study fashion for convenience (i.e. splitting the data, bootstrapping it, calculating disparity, testing differences).
-Note however that most of these functions can be used in any order and that some steps can of course be skipped and that the package contains many more additional functions!
-
-One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter called *subsamples*, see [glossary](#glossary)).
-These subsamples correspond to regions of the ordinated space that we want to compare to each other.
-
-<!-- # NC: Again I think this might be better written as "here are some things you might want to modify, here is how" rather than a function by function approach. -->
-
-`custom.subsamples` # NC: I'd have called this custom.subsamples. It's only 2 more letters and makes it much more readable, and less about custard.
--------------
-
-The `custom.subsamples` function allows to create these subsamples according to one or more factors determined by the user.
-In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (all the living mammals and their direct ancestor) and the stem mammals (all the mammals that have no related living descendants).
-To do so, we need to separate the tree into two subtrees at the node 71 (see [phylogeny above](#mammal_tree)).
-
-```{r}
-## Selecting the crown and stem mammals
-crown <- extract.clade(BeckLee_tree, node = 71)$tip.label
-stem <- drop.tip(BeckLee_tree, tip = crown)$tip.label
-```
-
-We can then feed these lists of species to the `custom.subsamples` function after transforming it into a factorial table:
-<!-- # NC: This is a fairly horrible set up to go through! Can't you make a function? -->
-
-```{r}
-## Creating the factor table
-subsamples_factors <- as.data.frame(
-    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
-    ncol = 1, dimnames = list(c(crown, stem), "Group")))
-head(subsamples_factors)
-
-## Creating the two subsamples
-crown_stem <- custom.subsamples(BeckLee_mat50, subsamples_factors)
-```
-
-The results of this function (like many others presented below) outputs a `dispRity` object:
-
-```{r}
-crown_stem
-```
-
-The two subsamples represent two groups occupying specific regions of the mammalian morphospace.
-The `dispRity` object itself always contains the original morphospace (`$matrix`) as well as information on which taxa compose the subsamples (`$subsamples`):
-
-```{r}
-## This object contains three elements
-names(crown_stem)
-## With "matrix" being the original matrix (the ordinated space)
-str(crown_stem$matrix)
-## And "subsamples" the elements from this ordinated space into the two sub-spaces
-str(crown_stem$subsamples)
-```
-
-Note that here we are separating the taxa into two exclusive groups (i.e. crown mammals can be crown only, same for stem mammals).
-It is however possible to add more factors with overlapping groups:
-
-```{r}
-## Adding all mammals to the a new subsamples (along with crown and stem as before)
-subsamples_factors2 <- as.data.frame(
-    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem)),
-    rep("Mammalia", Ntip(BeckLee_tree))), ncol = 2,
-    dimnames = list(c(crown, stem), c("Group", "Class")), byrow = FALSE))
-
-## Creating the three subsamples: Group.stem/crown and Class.Mammalia
-custom.subsamples(BeckLee_mat50, subsamples_factors2)
-```
-
-`time.subsamples`  <a name="time.subsamples"></a>
--------------
-
-Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
-The `time.subsamples` function allows to create subsamples containing all the elements present at specific points in time or during specific periods.
-This functions needs as input an ordinated space and a matching phylogenetic tree.
-Two types of time subsamples can be performed by using the `method` option:
-
- *  discrete time subsamples (or time-binning) using `method = discrete`;
-
- *  continuous time subsamples (or time-slicing) using `method = continuous`.
-
-For the time-slicing method details see @GuillermeSTD.
-<!-- TODO: link to the time-slicing algorithm explained! -->
-For both methods, the function intakes the `time` argument which can be a vector of `numeric` values for:
-
- *  defining the boundaries of the time bins (when `method = discrete`);
-
- *  defining the time slices (when `method = continuous`).
-
-Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
-Additionally, it is also possible to input a data frame containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age.
-
-Here is an example for `method = discrete`:
-
-```{r}
-## Generating three time bins containing the data present every 40 Ma
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-    time = c(120, 80, 40, 0))
-```
-
-In this example, the taxa were split inside each time-bin according to their age.
-Note however that some taxa might span between some time bins and should be included in more than one.
-This is possible by providing a table containing the first and last occurrence data:
-
-```{r}
-## Generating three time bins containing the data present every 40 Ma, including taxa that might
-## span between time bins.
-(time_bins <- time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree,
-    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages))
-```
-
-This generated indeed a `dispRity` object of 3 subsamples.
-Note that we can also generate equivalent results by just telling the function that we want three time-bins (subsamples) as follow:
-
-```{r}
-## Automatically generate three equal length bins:
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-    time = 3)
-```
-
-We now have three time bins of 44.50368 million years each.
-
-When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1% of the tree length as long as at least three elements are present at that point in time.
-The algorithm adds an extra 1% tree length until reaching the required minimum of three elements.
-It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance between tips *and* nodes.
-
-For the time-slicing method (`method = continuous`), the idea is fairly similar.
-This option, however, requires a matrix that contains the ordinated distance between taxa *and* nodes as well as an extra argument being the assumed evolutionary model (via the `model` argument):
-
- *  `acctran` where the data chosen on each time slice is always the one of the offspring
-
- *  `deltran` where the data chosen on each time slice is always the one of the descendant
-
- *  `punctuated` where the data chosen on each time slice is randomly chosen between the offspring or the descendant
-
- *  `gradual` where the data chosen on each time slice is either the offspring or the descendant depending on branch length
-
-```{r}
-## Generating four time slices every 40 million years assuming a gradual evolutionary model
-(time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages))
-## Note that in the same way as before, we can also automatically generate the slices
-time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
-    model = "gradual", time = 4)
-```
-
-`boot.matrix`
--------------
-
-Once we obtain our different subsamples, it is possible to bootstrap and rarefy them (i.e. pseudo-replicating the data).
-The bootstrap will allow us to make each subsamples more robust to outliers and the rarefaction will allow us to compare subsamples with the same number of elements to get rid of eventual sampling problems.
-The `boot.matrix` allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
-The default options will bootstrap the matrix 1000 times without rarefaction using the "full" bootstrap method (see below).
-
-```{r}
-boot.matrix(data = BeckLee_mat50)
-```
-
-The number of bootstrap pseudo-replicates can be defined using the `bootstraps` option.
-The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
-Currently two algorithms are implemented:
-
- *  `full` where the bootstrapping is entirely stochastic (all the data is bootstrapped)
-
- *  `single` where only one random elements is replaced by one other random elements each pseudo-replication
-
-This function also allows to rarefy the data using the `rarefaction` argument.
-The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in each subsamples to a minimum of 3 elements).
-It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
-One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
-When left missing, all dimensions from the ordinated matrix are used.
-
-```{r}
-## Bootstrapping with the single bootstrap method
-boot.matrix(BeckLee_mat50, boot.type = "single")
-
-## Bootstrapping with the full rarefaction
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
-
-## Or with a set number of rarefaction levels
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8,3))
-
-## Removing the last axis 50% last dimensions
-boot.matrix(BeckLee_mat50, dimensions = 0.5)
-
-## Keeping the 10 first dimensions
-boot.matrix(BeckLee_mat50, dimensions = 10)
-```
-
-Of course, one could be interested in directly supplying the subsamples generated above directly to this function.
-In fact, it can also deal with a list of matrices or with a `dispRity` object output from the `custom.subsamples` or `time.subsamples` functions.
-
-```{r}
-## Bootstrapping and fully rarefying the crown/stem subsamples
-(crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE))
-
-## Bootstrapping on the time binning/slicing subsamples
-(time_binsBS <- boot.matrix(time_bins, bootstraps = 100))
-(time_slicesBS <- boot.matrix(time_slices, bootstraps = 100))
-```
-
-`dispRity`
-----------
-
-This function is a modular function that allows to simply (and quickly!) calculate disparity from a matrix.
-Because disparity can be measured in many ways, this function is a tool to measure disparity *as defined by the user*.
-In fact, the `dispRity` function intakes two main arguments: the data and the disparity metric.
-The disparity metric is a function or a set of functions that summarises the ordinated matrix to a single value that represents, in our example, the diversity of morphologies.
-
-The `dispRity` algorithm decomposes the metrics functions into three levels that correspond to the dimensions of the output of each metric function.
-For more details on this algorithm please refer to the [metric vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html).
-
-In practice, the `dispRity` function intakes one or more functions as a definition of disparity.
-Several of these functions will be already implemented in other packages (such as `stats::median`, `base::sum`, etc.); some others are implemented in this package (listed in `?dispRity.metric`) and finally some others can be defined by the users.
-The `make.metric` function is designed to help users create and test their own disparity metric functions.
-In practice, the use of these metrics in the `dispRity` function, is pretty easy:
-
-```{r}
-## Calculating disparity as the mean value of the ordinated space (using base::mean)
-dispRity(BeckLee_mat50, metric = mean)
-
-## Calculating disparity as the sum of variances of the ordinated space using base::sum on the
-## variances and dispRity::variances on each dimension of the ordinated spaces
-dispRity(BeckLee_mat50, metric = c(sum, variances))
-
-## Calculating disparity as the product of the centroids (using base::prod and dispRity::centroids)
-dispRity(BeckLee_mat50, metric = c(prod, centroids))
-
-## Defining the total range function:
-total.range <- function(X) abs(range(X)[1]-range(X)[2])
-
-## Calculating disparity as the total range of the centroids:
-dispRity(BeckLee_mat50, metric = c(total.range, centroids))
-```
-
-Note that in practice, the order of the functions in the `metric` argument does not matter:
-
-```{r}
-## Disparity as the standard deviation of the variances of the variance/covariance matrix;
-vars1 <- dispRity(BeckLee_mat50, metric = c(sd, variances, var))
-
-## Same disparity metric but with the functions ordered differently
-vars2 <- dispRity(BeckLee_mat50, metric = c(variances, var, sd))
-
-## Both versions gives the same results
-all(summary(vars1) == summary(vars2))
-```
-
-These functions do not directly output the disparity values but only the summary of the `dispRity` objects.
-To display the results, see the section [summary](#summary).
-
-In these examples we used only simple ordinated matrix but of course, it possible to directly use the `dispRity` objects we generated in the steps above.
-Thus we can calculate the bootstrapped and rarefied disparity in each subsamples of the ordinated space.
-Here we are going to define disparity as being the median distance between each element and each dimensions' centre.
-
-<a name="disparity.calculation"></a>
-
-```{r}
-## Disparity in crown and stem mammals:
-disp_crown_stemBS <- dispRity(crown_stemBS, metric = c(median, centroids))
-
-## Disparity through time:
-disp_time_binsBS <- dispRity(time_binsBS, metric = c(median, centroids))
-disp_time_slicesBS <- dispRity(time_slicesBS, metric = c(median, centroids))
-```
-
-Note that the computational time was longer for the rarefied data (`crown_stemBS`) since it had to calculate the disparity on all the rarefied matrix, each bootstrapped 100 times.
-For both the `boot.matrix` and the `dispRity` functions, it is possible to increase computational speed in some cases using the `parallel` option (see each [parallel section](#parallel) below).
-
-`summary` <a name="summary"></a>
----------
-
-This function is a S3 function (`summary.dispRity`) allowing to summarize the content of `dispRity` objects that contain disparity calculations.
-
-```{r}
-summary(disp_time_binsBS)
-```
-
-This function intakes a `dispRity` object plus various options namely the `quantile` values for the confidence intervals levels (by default, the 50 and 95 quantiles are calculated); the `cent.tend` for the  central tendency to use for summarising the results (default is `median`); the `rounding` option corresponding to the number of decimals to print (default is `2`) and the `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
-These options can easily be changed from the defaults as follows:
-
-```{r}
-## Summarising the results using the 88 quantile and the standard deviation as the central tendency. 
-summary(disp_time_binsBS, quantile = 88, cent.tend = sd)
-
-## Printing the details of the dispRity object and rounding the values to the 5th decimal
-summary(disp_time_binsBS, recall = TRUE, rounding = 5)
-```
-
-> Information about the number of elements per subsamples as well as the observed (i.e. non-bootstrapped) disparity is also calculated.
-This is specifically handy when rarefying the data for example
-
-```{r}
-head(summary(disp_crown_stemBS))
-```
-
-`plot` <a name="plot"></a>
-------
-
-An alternative way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
-This function intakes the same options as `summary.dispRity` along side with various graphical options described in the function manual (see `?plot.dispRity`).
-
-The plots can be of four different types: `continuous` for displaying continuous disparity curves and `box`, `lines` and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines and confidence interval polygons.
-
-> This argument can be left missing. In this case, the algorithm will automatically detect the type of subsamples from the `dispRity` object.
-
-It is also possible to display the number of elements per subsamples (has a horizontal doted line) using the option `elements = TRUE`.
-Additionally, when the data is rarefied, one can also indicate which level of rarefaction to display (i.e. display the results for a certain number of elements) by using the `rarefaction` argument.
-
-```{r, fig.width=6, fig.height=6}
-## Graphical parameters
-op <- par(mfrow = c(2,2), bty = "n")
-## Plotting continuous disparity results
-plot(disp_time_binsBS, type = "continuous")
-## Plotting discrete disparity results
-plot(disp_crown_stemBS, type = "box")
-## Plotting the disparity using lines for the rarefaction level of 20 elements only
-plot(disp_crown_stemBS, type = "line", rarefaction = 20)
-## Plotting the disparity using polygons while also displaying the number of elements
-plot(disp_crown_stemBS, type = "polygon", elements = TRUE)
-```
-
-> The `type` argument is not mandatory for the two first plots since both `continuous` and `box` are the default options for respectively continuous and discrete data.
-
-Since `plot.dispRity` uses the arguments from the generic `plot` method, it is of course possible to change pretty much everything using the regular plot arguments:
-
-```{r, fig.width=6, fig.height=6}
-## Graphical options
-par(bty = "n")
-## Plotting the results with some classic options from plot
-plot(disp_time_slicesBS, col = c("blue", "orange", "green"), ylab = c("Some measurement"),
-    xlab = "Some other measurement", main = "Many options...", ylim = c(0,2.1))
-## Adding a legend
-legend("topleft", legend = c("Central tendency", "Confidence interval 1", "Confidence interval 2"),
-    col = c("blue", "orange", "green"), pch = 19)
-```
-
-Additionally to the classic `plot` arguments, the function can also take arguments that are specific to `plot.dispRity` like adding the number of elements or rarefaction level (as described above) but also changing the values of the quantiles to plot as well as the central tendency.
-
-```{r, fig.width=6, fig.height=6}
-## Graphical options
-par(bty = "n")
-## Plotting the results with some plot.dispRity arguments
-plot(disp_time_slicesBS, quantile = c(seq(from=10, to=100, by=10)), cent.tend = sd, type = "c",
-    elements = TRUE, col = c("black", rainbow(10)), ylab = c("Disparity", "Diversity"),
-    xlab = "Time (in in units from past to present)", time.subsamples = FALSE,
-    observed = TRUE, main = "Many options...")
-```
-
-> Note that the argument `observed = TRUE` allows to plot the disparity values calculated from the non-bootstrapped data as crosses on the plot.
-
-For comparing results, it is also possible to add a plot to the existent plot by using `add = TRUE`:
-
-```{r, fig.width=6, fig.height=6}
-## Graphical options
-par(bty = "n")
-## Plotting the continuous disparity with a fixed y axis
-plot(disp_time_slicesBS, ylim = c(0.5,2))
-## Adding the discrete data using the same fixed y axis to have the results at the same scale
-plot(disp_time_binsBS, type = "line", ylim=c(0.5,2), xlab="", ylab="", add = TRUE)
-```
-
-Finally, if your data has been fully rarefied, it is also possible to easily look at rarefaction curves by using the `rarefaction = TRUE` argument:
-
-```{r, fig.width=6, fig.height=6}
-## Graphical options
-par(bty = "n")
-## Plotting the rarefaction curves
-plot(disp_crown_stemBS, rarefaction = TRUE)
-```
-
-`test.dispRity`
----------------
-
-Finally, the `dispRity` package allows to apply some test functions to the calculated disparity in order to test hypothesis.
-The function `test.dispRity` works in a similar way as the `dispRity` function: it intakes a `dispRity` object, a `test` and a `comparisons` argument.
-
-The `comparisons` argument must indicate the way the test should be applied to the data:
-
- *   `pairwise` (default): to compare each subsamples pairwise
-
- *  `referential`: to compare each subsamples to the first one
-
- *  `sequential`: to compare each subsamples to the following one
-
- *  `all`: to compare all the subsamples together (like in analysis of variance)
-
-It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsamples names to create personalised test.
-Some other specific tests implemented in `dispRity` such as the `dispRity::null.test` and the `dispRity::sequential.test` have a specific way to be applied to the data and therefore ignore the `comparisons` argument.
-
-The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
-It can be common statistical test functions (e.g. `stats::t.test`), some functions implemented in `dispRity` (e.g. see `?null.test`) or any function defined by the user.
-
-This function also allows to correct for type I error inflation when using multiple comparisons via the `correction` argument.
-This argument can be empty (no correction applied) or can contain one of the corrections from the `stats::p.adjust` function (see `?p.adjust`).
-
-Note that the `test.dispRity` algorithm deals with some classic tests outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
-It is however possible to get the full detailed output by using the options `details = TRUE`.
-
-```{r}
-## Performing a t-test to the test a difference in disparity between crown and stem mammals
-test.dispRity(disp_crown_stemBS, test = t.test)
-
-## Performing the same test but with the detailed t.test output
-test.dispRity(disp_crown_stemBS, test = t.test, details = TRUE)
-
-## Performing a Wilcoxon test applied to sliced disparity the data sequentially with correction
-test.dispRity(disp_time_slicesBS, test = wilcox.test, comparisons = "sequential",
-    correction = "bonferroni")
-
-## Measuring the overlap between distributions in the time bins (using the implemented
-## Bhattacharyya Coefficient function - see ?bhatt.coeff)
-test.dispRity(disp_time_binsBS, test = bhatt.coeff)
-```
-
-It is also possible to apply some more *complex* tests that have their own output class (like `stats::aov`).
-The results can then be analysed as usual using the associated `summary` S3 method:
-
-```{r}
-## Performing and ANOVA applied to the same data
-slice_aov <- (test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all"))
-
-## The output is a regular ANOVA output...
-class(slice_aov)
-
-## That can be summarised using summary
-summary(slice_aov)
-```
-
-utilities
----------
-The package also provides some utilities functions to facilitate multidimensional analysis.
-
-### `dispRity` object utilities  <a name="dispRity.utilities"></a>
-The first subsamples of utilities are functions for manipulating `dispRity` objects:
-
-#### `make.dispRity`
-This function creates empty `dispRity` objects.
-
-```{r}
-## Creating an empty dispRity object
-make.dispRity()
-## Creating an "empty" dispRity object with a matrix
-(disparity_obj <- make.dispRity(matrix(rnorm(20), 5, 4)))
-```
-
-####  `fill.dispRity`
-This function initialises a `dispRity` object and generating it's call properties.
-
-```{r}
-## The dispRity object's call is indeed empty
-disparity_obj$call
-## Creating an empty dispRity object
-(disparity_obj <- fill.dispRity(disparity_obj))
-## The dipRity object has now the proper minimal attributes
-disparity_obj$call
-```
-
-####  `matrix.dispRity`
-This function extracts a specific matrix from a disparity object. The matrix can be one of the bootstrapped or/and rarefied one.
-
-```{r}
-## Extracting the matrix containing the coordinates for the crown mammals
-str(matrix.dispRity(disp_crown_stemBS, "Group.crown"))
-## Extracting the 3rd boostrapped matrix with the 10th rarefaction level from the same group
-str(matrix.dispRity(disp_crown_stemBS, subsamples = 1, bootstrap = 3, rarefaction = 10))
-```
-
-#### `get.subsamples.dispRity`
-This function creates a dispRity object that contains only elements from one specific subsamples.
-
-```{r}
-## Extracting all the data for the crown mammals
-(crown_mammals <- get.subsamples.dispRity(disp_crown_stemBS, "Group.crown"))
-## The object keeps the properties of the parent object but is composed of only one subsamples
-length(crown_mammals$subsamples)
-```
-
-#### `extract.dispRity`
-This function extracts the calculated disparity values of a specific matrix.
-
-```{r}
-## Extracting the observed disparity (default)
-extract.dispRity(disp_crown_stemBS)
-
-## Extracting the disparity from the bootstrapped values from the 10th rarefaction level from the
-## second subsamples (stem mammals)
-extract.dispRity(disp_crown_stemBS, observed = FALSE, subsamples = 2, rarefaction = 10)
-```
-
-#### `scale.dispRity`
-This is the S3 method of `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object.
-
-```{r}
-## Getting the disparity values of the time subsamples
-summary(disp_time_slicesBS)
-
-## Scaling the same disparity values
-summary(scale(disp_time_slicesBS, scale = TRUE))
-```
-
-#### `sort.dispRity`
-This is the S3 method of `sort` for sorting the subsamples alphabetically (default) or following a specific pattern.
-
-```{r}
-## Sorting the disparity subsamples in inverse alphabetic order
-summary(sort(disp_time_slicesBS, decreasing = TRUE))
-```
-
-### Package general utilities
-
-#### `tree.age`
-
-This function allows to calculate the age of each individual nodes and tips in a tree.
-It can either use the root age of the tree (if present as `$root.time`) or else calculate the age using a user defined root age via the `age` argument.
-Also, it is possible to decide whether the time is calculated towards the past (e.g. million years ago) or towards the present (e.g. in time since the origin).
-
-```{r}
-## This tree has a root age
-BeckLee_tree$root.time
-## So we can get the age of each tips and nodes directly
-head(tree.age(BeckLee_tree), 5)
-## But we can also decide to make the age relative (between 1 and 0)
-head(tree.age(BeckLee_tree, age = 1), 5)
-## Or even relative, but from the root (i.e. how far are the nodes/tips
-## from the root)
-head(tree.age(BeckLee_tree, age = 1, order = "present"), 5)
-```
-
-#### `slice.tree`
-
-This function is at the core of `time.subsamples` when using the time slicing method.
-It allows to slice a phylogenetic tree at a certain time point and selecting the nodes or tips following the models detailed in the `time.subsamples` [section above](#time.subsamples).
-
-```{r, fig.width=8, fig.height=6}
-set.seed(1)
-## Generating a random ultrametric tree
-tree <- rcoal(20)
-## Adding some node labels
-tree$node.label <- letters[1:19]
-## Adding a root age (as the maximum tree age)
-tree$root.time <- max(tree.age(tree)$ages)
-
-## Creating a deltran tree at the slice 0.25
-tree_25 <- slice.tree(tree, age = 0.25, "deltran")
-## Creating a deltran tree at the slice 0.50
-tree_50 <- slice.tree(tree, age = 0.50, "acctran")
-
-## Illustrating the results
-par(mfrow = c(1,3))
-plot(tree, main = "Full tree", cex = 1.5)
-nodelabels(tree$node.label, cex = 1.1)
-axisPhylo()
-plot(tree_25, main = "Slice at time 0.25", cex = 1.5)
-axisPhylo()
-plot(tree_50, main = "Slice at time 0.50", cex = 1.5)
-axisPhylo()
-```
-
-#### `sim.morpho`
-This function (and it's associated utilities) allows to generate realistic discrete morphological matrices.
-See the [associated vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-simulate_data.html) for more details.
-
-Other utilities functions are also available within the package and more will be added in the future. For the full list, refer to the package's functions list within `R`.
-
-parallel  <a name="parallel"></a>
----------
-
-`dispRity` and `boot.matrix` have also `parallel` option.
-This option uses the package `snow` to run the `lapply` loops within the package in parallel.
-In both functions, this option is ignored when missing, otherwise, it takes as option, the number of threads (cores) and the virtual connection process (VCP) name following the function `makeCluster`.
-
-For example, we can speed up the [disparity calculations of the rarefied data](#disparity.calculation) by using 4 different threads and a `"SOCK"` connection (although `"MPI"` is much more recommended but needs more specific installation).
-
-```{r, eval=FALSE}
-dispRity(crown_stemBS, metric = c(median, centroids), parallel = c(4, "SOCK"))
-```
-
-Suggestions
-============
-If you have any extra suggestions or comments on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
+## Where is this going?

### 2017-05-29 doc/dispRity-manual.Rmd
Updated manual

@@
 
-```{r, fig.width=6, fig.height=6}
+```{r, fig.width=7, fig.height=7}
 ## Setting the random seed for repeatability
@@
 
-This
+One of the functions of the package is to look at how disparity changes through time.
+This function (and the following one) are the two most straight forward, they analysis the ordinated data in a pipelined way using a lot of default parameters.
+
+For disparity through time analysis, you will need:
+
+  * An ordinated matrix (we covered that above)
+  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element. To give your tree a root time, you can simply do `my_tree$root.time <- my_age`.
+  * A required number of time subsamples (here 3)
+  * Your definition of disparity (here the sum of variances)
+
+Note that in the following example, the matrix and the tree comes from the previous sections
 
 ```{r}
-# dispRity.through.time()
+## Measuring disparity through time
+disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
+                                        time = 3, metric = c(sum, variances))
 ```
+
+When display our disparity data, we can now have information on the operations done to the matrix:
+
+```{r}
+## Printing what's in the disparity data
+disparity_data
+```
+Indeed, we asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times and the metric used was the sum of variances.
+
+We can now analyse this data by summarising and plotting it or performing some statistical tests on it (e.g. a simple `aov`):
+
+```{r, fig.width=7, fig.height=7}
+## Summarising the measurements
+summary(disparity_data)
+
+## Plotting the results
+plot(disparity_data)
+
+## Testing for an eventual difference between the time bins
+disp_aov <- test.dispRity(disparity_data, test = aov, comparisons = "all")
+summary(disp_aov)
+```
+
+Please refer to the specific tutorials below for (much!) more information on the nuts and bolts of the package.
+You can also directly explore the specific functions manuals within R and navigate to they're related functions.
 
@@
 
+One other typical analysis done with disparity is to look at the difference in disparity between several groups.
+
 ```{r}
-# dispRity.per.group()
+## Creating the two groups as a list
+mammal_groups <- list("crown" = c(16, 19:41, 45:50),
+                      "stem" = c(1:15, 17:18, 42:44))
+
+## Measuring disparity for each group
+disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
+                                     metric = c(sum, variances))
+```
+
+Similarly as above, we can display information about both groups by simply calling it and then summarising the data (and plotting it!) and performing a statistical test to compare both groups (here a Wilcoxon test).
+
+```{r, fig.width=7, fig.height=7}
+## Printing what's in the disparity data
+disparity_data
+
+## Summarising the measurements
+summary(disparity_data)
+
+## Plotting the results
+plot(disparity_data)
+
+## Testing for an eventual difference between the groups
+test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
+
 ```
@@
 # Specific tutorials
+
+## Bootstraps and rarefactions
+
+## Time slicing
+
+## Disparity metrics
+
+## Plotting `dispRity` data
+
+## Testing disparity hypothesis
+
+## Disparity as a distribution
+
+## Ecological study
 

### 2017-05-30 doc/dispRity-manual.Rmd
Added sections from previous manuals (not stream-lined)

@@
 Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
+
+## Citations
+
+You can cite both the package or this manual with the following citation:
+> Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646
+
+Note that this citation is only temporary (but can still be used!).
+A future proper version of the latest package release, this manual and an associated methods paper should be submitted soon(ish!).
 
@@
 
+## Time slicing
+
+Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
+The `time.subsamples` function allows to create subsamples containing all the elements present at specific points in time or during specific periods.
+This functions needs as input an ordinated space and a matching phylogenetic tree.
+Two types of time subsamples can be performed by using the `method` option:
+
+ *  discrete time subsamples (or time-binning) using `method = discrete`;
+
+ *  continuous time subsamples (or time-slicing) using `method = continuous`.
+
+For the time-slicing method details see @GuillermeSTD.
+<!-- TODO: link to the time-slicing algorithm explained! -->
+For both methods, the function intakes the `time` argument which can be a vector of `numeric` values for:
+
+ *  defining the boundaries of the time bins (when `method = discrete`);
+
+ *  defining the time slices (when `method = continuous`).
+
+Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
+Additionally, it is also possible to input a data frame containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age.
+
+Here is an example for `method = discrete`:
+
+```{r, eval=FALSE}
+## Generating three time bins containing the data present every 40 Ma
+time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+    time = c(120, 80, 40, 0))
+```
+
+In this example, the taxa were split inside each time-bin according to their age.
+Note however that some taxa might span between some time bins and should be included in more than one.
+This is possible by providing a table containing the first and last occurrence data:
+
+```{r, eval=FALSE}
+## Generating three time bins containing the data present every 40 Ma, including taxa that might
+## span between time bins.
+(time_bins <- time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree,
+    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages))
+```
+
+This generated indeed a `dispRity` object of 3 subsamples.
+Note that we can also generate equivalent results by just telling the function that we want three time-bins (subsamples) as follow:
+
+```{r, eval=FALSE}
+## Automatically generate three equal length bins:
+time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+    time = 3)
+```
+
+We now have three time bins of 44.50368 million years each.
+
+When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1% of the tree length as long as at least three elements are present at that point in time.
+The algorithm adds an extra 1% tree length until reaching the required minimum of three elements.
+It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance between tips *and* nodes.
+
+For the time-slicing method (`method = continuous`), the idea is fairly similar.
+This option, however, requires a matrix that contains the ordinated distance between taxa *and* nodes as well as an extra argument being the assumed evolutionary model (via the `model` argument):
+
+ *  `acctran` where the data chosen on each time slice is always the one of the offspring
+
+ *  `deltran` where the data chosen on each time slice is always the one of the descendant
+
+ *  `punctuated` where the data chosen on each time slice is randomly chosen between the offspring or the descendant
+
+ *  `gradual` where the data chosen on each time slice is either the offspring or the descendant depending on branch length
+
+```{r, eval=FALSE}
+## Generating four time slices every 40 million years assuming a gradual evolutionary model
+(time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
+    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
+    FADLAD = BeckLee_ages))
+## Note that in the same way as before, we can also automatically generate the slices
+time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
+    model = "gradual", time = 4)
+```
+
 ## Bootstraps and rarefactions
 
-## Time slicing
+Once we obtain our different subsamples, it is possible to bootstrap and rarefy them (i.e. pseudo-replicating the data).
+The bootstrap will allow us to make each subsamples more robust to outliers and the rarefaction will allow us to compare subsamples with the same number of elements to get rid of eventual sampling problems.
+The `boot.matrix` allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
+The default options will bootstrap the matrix 1000 times without rarefaction using the "full" bootstrap method (see below).
+
+```{r, eval=FALSE}
+boot.matrix(data = BeckLee_mat50)
+```
+
+The number of bootstrap pseudo-replicates can be defined using the `bootstraps` option.
+The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
+Currently two algorithms are implemented:
+
+ *  `full` where the bootstrapping is entirely stochastic (all the data is bootstrapped)
+
+ *  `single` where only one random elements is replaced by one other random elements each pseudo-replication
+
+This function also allows to rarefy the data using the `rarefaction` argument.
+The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
+remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in each subsamples to a minimum of 3 elements).
+It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
+One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
+When left missing, all dimensions from the ordinated matrix are used.
+
+```{r, eval=FALSE}
+## Bootstrapping with the single bootstrap method
+boot.matrix(BeckLee_mat50, boot.type = "single")
+
+## Bootstrapping with the full rarefaction
+boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
+
+## Or with a set number of rarefaction levels
+boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8,3))
+
+## Removing the last axis 50% last dimensions
+boot.matrix(BeckLee_mat50, dimensions = 0.5)
+
+## Keeping the 10 first dimensions
+boot.matrix(BeckLee_mat50, dimensions = 10)
+```
+
+Of course, one could be interested in directly supplying the subsamples generated above directly to this function.
+In fact, it can also deal with a list of matrices or with a `dispRity` object output from the `custom.subsamples` or `time.subsamples` functions.
+
+```{r, eval=FALSE}
+## Bootstrapping and fully rarefying the crown/stem subsamples
+(crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE))
+
+## Bootstrapping on the time binning/slicing subsamples
+(time_binsBS <- boot.matrix(time_bins, bootstraps = 100))
+(time_slicesBS <- boot.matrix(time_slices, bootstraps = 100))
+```
 
@@
 
+An alternative way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
+This function intakes the same options as `summary.dispRity` along side with various graphical options described in the function manual (see `?plot.dispRity`).
+
+The plots can be of four different types: `continuous` for displaying continuous disparity curves and `box`, `lines` and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines and confidence interval polygons.
+
+> This argument can be left missing. In this case, the algorithm will automatically detect the type of subsamples from the `dispRity` object.
+
+It is also possible to display the number of elements per subsamples (has a horizontal doted line) using the option `elements = TRUE`.
+Additionally, when the data is rarefied, one can also indicate which level of rarefaction to display (i.e. display the results for a certain number of elements) by using the `rarefaction` argument.
+
+```{r, fig.width=6, fig.height=6, eval=FALSE}
+## Graphical parameters
+op <- par(mfrow = c(2,2), bty = "n")
+## Plotting continuous disparity results
+plot(disp_time_binsBS, type = "continuous")
+## Plotting discrete disparity results
+plot(disp_crown_stemBS, type = "box")
+## Plotting the disparity using lines for the rarefaction level of 20 elements only
+plot(disp_crown_stemBS, type = "line", rarefaction = 20)
+## Plotting the disparity using polygons while also displaying the number of elements
+plot(disp_crown_stemBS, type = "polygon", elements = TRUE)
+```
+
+> The `type` argument is not mandatory for the two first plots since both `continuous` and `box` are the default options for respectively continuous and discrete data.
+
+Since `plot.dispRity` uses the arguments from the generic `plot` method, it is of course possible to change pretty much everything using the regular plot arguments:
+
+```{r, fig.width=6, fig.height=6, eval=FALSE}
+## Graphical options
+par(bty = "n")
+## Plotting the results with some classic options from plot
+plot(disp_time_slicesBS, col = c("blue", "orange", "green"), ylab = c("Some measurement"),
+    xlab = "Some other measurement", main = "Many options...", ylim = c(0,2.1))
+## Adding a legend
+legend("topleft", legend = c("Central tendency", "Confidence interval 1", "Confidence interval 2"),
+    col = c("blue", "orange", "green"), pch = 19)
+```
+
+Additionally to the classic `plot` arguments, the function can also take arguments that are specific to `plot.dispRity` like adding the number of elements or rarefaction level (as described above) but also changing the values of the quantiles to plot as well as the central tendency.
+
+```{r, fig.width=6, fig.height=6, eval=FALSE}
+## Graphical options
+par(bty = "n")
+## Plotting the results with some plot.dispRity arguments
+plot(disp_time_slicesBS, quantile = c(seq(from=10, to=100, by=10)), cent.tend = sd, type = "c",
+    elements = TRUE, col = c("black", rainbow(10)), ylab = c("Disparity", "Diversity"),
+    xlab = "Time (in in units from past to present)", time.subsamples = FALSE,
+    observed = TRUE, main = "Many options...")
+```
+
+> Note that the argument `observed = TRUE` allows to plot the disparity values calculated from the non-bootstrapped data as crosses on the plot.
+
+For comparing results, it is also possible to add a plot to the existent plot by using `add = TRUE`:
+
+```{r, fig.width=6, fig.height=6, eval=FALSE}
+## Graphical options
+par(bty = "n")
+## Plotting the continuous disparity with a fixed y axis
+plot(disp_time_slicesBS, ylim = c(0.5,2))
+## Adding the discrete data using the same fixed y axis to have the results at the same scale
+plot(disp_time_binsBS, type = "line", ylim=c(0.5,2), xlab="", ylab="", add = TRUE)
+```
+
+Finally, if your data has been fully rarefied, it is also possible to easily look at rarefaction curves by using the `rarefaction = TRUE` argument:
+
+```{r, fig.width=6, fig.height=6, eval=FALSE}
+## Graphical options
+par(bty = "n")
+## Plotting the rarefaction curves
+plot(disp_crown_stemBS, rarefaction = TRUE)
+```
+
 ## Testing disparity hypothesis
+
+Finally, the `dispRity` package allows to apply some test functions to the calculated disparity in order to test hypothesis.
+The function `test.dispRity` works in a similar way as the `dispRity` function: it intakes a `dispRity` object, a `test` and a `comparisons` argument.
+
+The `comparisons` argument must indicate the way the test should be applied to the data:
+
+ *   `pairwise` (default): to compare each subsamples pairwise
+
+ *  `referential`: to compare each subsamples to the first one
+
+ *  `sequential`: to compare each subsamples to the following one
+
+ *  `all`: to compare all the subsamples together (like in analysis of variance)
+
+It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsamples names to create personalised test.
+Some other specific tests implemented in `dispRity` such as the `dispRity::null.test` and the `dispRity::sequential.test` have a specific way to be applied to the data and therefore ignore the `comparisons` argument.
+
+The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
+It can be common statistical test functions (e.g. `stats::t.test`), some functions implemented in `dispRity` (e.g. see `?null.test`) or any function defined by the user.
+
+This function also allows to correct for type I error inflation when using multiple comparisons via the `correction` argument.
+This argument can be empty (no correction applied) or can contain one of the corrections from the `stats::p.adjust` function (see `?p.adjust`).
+
+Note that the `test.dispRity` algorithm deals with some classic tests outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
+It is however possible to get the full detailed output by using the options `details = TRUE`.
+
+```{r, eval=FALSE}
+## Performing a t-test to the test a difference in disparity between crown and stem mammals
+test.dispRity(disp_crown_stemBS, test = t.test)
+
+## Performing the same test but with the detailed t.test output
+test.dispRity(disp_crown_stemBS, test = t.test, details = TRUE)
+
+## Performing a Wilcoxon test applied to sliced disparity the data sequentially with correction
+test.dispRity(disp_time_slicesBS, test = wilcox.test, comparisons = "sequential",
+    correction = "bonferroni")
+
+## Measuring the overlap between distributions in the time bins (using the implemented
+## Bhattacharyya Coefficient function - see ?bhatt.coeff)
+test.dispRity(disp_time_binsBS, test = bhatt.coeff)
+```
+
+It is also possible to apply some more *complex* tests that have their own output class (like `stats::aov`).
+The results can then be analysed as usual using the associated `summary` S3 method:
+
+```{r, eval=FALSE}
+## Performing and ANOVA applied to the same data
+slice_aov <- (test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all"))
+
+## The output is a regular ANOVA output...
+class(slice_aov)
+
+## That can be summarised using summary
+summary(slice_aov)
+```
 
@@
 ## Ecological study
+
+## Simulating discrete morphological data
+
+### Quick go through
+The function basically intakes a phylogenetic tree, the number of required characters and the evolutionary model and a function from which to draw the rates.
+The package then proposes a function for quickly checking the matrix' phylogenetic signal using parsimony.
+
+```{r, eval=FALSE}
+library(dispRity)
+set.seed(123)
+## Setting the starting tree with 15 taxa as a random coalescent tree
+my_tree <- rcoal(15)
+
+## Generating a matrix with 100 characters (85% binary and 15% three states) and
+## an equal rates model with a gamma rates distribution (0.5, 1) with no 
+## invariant characters.
+my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
+    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)
+
+## The head of the matrix
+my_matrix[1:5, 1:10]
+
+## Checking the matrix properties with a quick Maximum Parsimony tree search
+check.morpho(my_matrix, my_tree)
+```
+
+### In details
+
+Basically, `sim.morpho` is really flexible and intakes a lot of different arguments to allow to simulate realistic matrix.
+It has three implemented models: `"ER"` for Equal Rates (the M*k* model); the `"HKY"` one which is the molecular HKY model but transforms pyrines in 0's and purimidines in 1's; or the `"mixed"` model that randomly uses an `"ER"` or and `"HKY"` for the binary characters and `"ER"` for the multistates (>2) characters.
+Both models intakes specific distributions for their rates or substitution models.
+These distributions should be passed to these arguments in the format of `c(sampler_function, distribution_parameters)` where the the sampler function is a the random generation function of that distribution (e.g.
+`rnorm`, `runif`, etc...) and the parameters are any parameters to be passed to this function.
+
+The `check.morpho` runs a quick maximum parsimony tree using the `phangorn` parsimony algorithm.
+It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g.
+the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree.
+
+### Parameters for a good(ish) matrix
+There are many parameters that can realise a ``realistic'' matrix (meaning not to different from the input tree with a consistency and retention index close to what's in the literature) but because of the randomness of the generation (the code generates random matrices after all!) not all end up in creating good matrices.
+The following parameters however, seems to generate a fairly ``realistic'' with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three states characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
+
+```{r, eval=FALSE}
+set.seed(1)
+## tree
+my_tree <- rcoal(15)
+## matrix
+morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
+    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
+check.morpho(morpho_mat, my_tree)
+```
 
@@
 
-## `dispRity` object structure
+## Manipulating `dispRity` objects
+
+Disparity analysis involves a lot of manipulation of many matrices (especially when bootstrapping) which can be impractical to visualise and will quickly overwhelm your `R` console.
+Even the simple Beck and Lee 2014 example above produces an object with > 72 lines of lists of lists of matrices!
+
+Therefore `dispRity` uses a specific class of object called a `dispRity` object.
+These objects allow users to use S3 method functions such as `summary.dispRity` (see [summary](#summary) section), `plot.dispRity` (`plot`; see [plot](#plot) section) and `print.dispRity`. 
+<!-- # NC: I've simpiflied this as people who understand will know this already, people that don't won't care and don't need to know.
+ -->
+ `dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g.
+`sort.dispRity`, `extract.dispRity` see the full list in the [`dispRity` utilities](#dispRity.utilities) section).
+These functions modify the `dispRity` object without having to delve into its complex structure!
+The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).
+
+<!-- # NC: Explain the point of these examples here... Are they needed? -->
+
+```{r, eval=FALSE}
+## What is the class of the median_centroids object?
+class(median_centroids)
+
+## What does the object contain?
+names(median_centroids)
+
+## Summarising it using the S3 method print.dispRity
+median_centroids
+```
+
+Note that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:
+
+```{r, eval=FALSE}
+## Display the full object
+print(median_centroids, all = TRUE)
+## This is more than ~ 2500 lines on my 13 inch laptop screen!
+```
+
+<!-- 
+# Modular functions - 
+# NC: not sure any more of these modular functions are relevant in the vignette? Maybe just mention the metrics? Remember a vignette is for users, not for you to show how the programming works.
+
+# NC: I'd include a brief explanation of the include disparity metrics here.
+But yes keep the crazy "level" details for the other vignette on metrics.
+I think you want to avoid having too many separate vignettes really.
+-->
+
+## `dispRity` utilities
+
+The package also provides some utilities functions to facilitate multidimensional analysis.
+
+### `dispRity` object utilities  <a name="dispRity.utilities"></a>
+The first subsamples of utilities are functions for manipulating `dispRity` objects:
+
+#### `make.dispRity`
+This function creates empty `dispRity` objects.
+
+```{r, eval=FALSE}
+## Creating an empty dispRity object
+make.dispRity()
+## Creating an "empty" dispRity object with a matrix
+(disparity_obj <- make.dispRity(matrix(rnorm(20), 5, 4)))
+```
+
+####  `fill.dispRity`
+This function initialises a `dispRity` object and generating it's call properties.
+
+```{r, eval=FALSE}
+## The dispRity object's call is indeed empty
+disparity_obj$call
+## Creating an empty dispRity object
+(disparity_obj <- fill.dispRity(disparity_obj))
+## The dipRity object has now the proper minimal attributes
+disparity_obj$call
+```
+
+####  `matrix.dispRity`
+This function extracts a specific matrix from a disparity object. The matrix can be one of the bootstrapped or/and rarefied one.
+
+```{r, eval=FALSE}
+## Extracting the matrix containing the coordinates for the crown mammals
+str(matrix.dispRity(disp_crown_stemBS, "Group.crown"))
+## Extracting the 3rd boostrapped matrix with the 10th rarefaction level from the same group
+str(matrix.dispRity(disp_crown_stemBS, subsamples = 1, bootstrap = 3, rarefaction = 10))
+```
+
+#### `get.subsamples.dispRity`
+This function creates a dispRity object that contains only elements from one specific subsamples.
+
+```{r, eval=FALSE}
+## Extracting all the data for the crown mammals
+(crown_mammals <- get.subsamples.dispRity(disp_crown_stemBS, "Group.crown"))
+## The object keeps the properties of the parent object but is composed of only one subsamples
+length(crown_mammals$subsamples)
+```
+
+#### `extract.dispRity`
+This function extracts the calculated disparity values of a specific matrix.
+
+```{r, eval=FALSE}
+## Extracting the observed disparity (default)
+extract.dispRity(disp_crown_stemBS)
+
+## Extracting the disparity from the bootstrapped values from the 10th rarefaction level from the
+## second subsamples (stem mammals)
+extract.dispRity(disp_crown_stemBS, observed = FALSE, subsamples = 2, rarefaction = 10)
+```
+
+#### `scale.dispRity`
+This is the S3 method of `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object.
+
+```{r, eval=FALSE}
+## Getting the disparity values of the time subsamples
+summary(disp_time_slicesBS)
+
+## Scaling the same disparity values
+summary(scale(disp_time_slicesBS, scale = TRUE))
+```
+
+#### `sort.dispRity`
+This is the S3 method of `sort` for sorting the subsamples alphabetically (default) or following a specific pattern.
+
+```{r, eval=FALSE}
+## Sorting the disparity subsamples in inverse alphabetic order
+summary(sort(disp_time_slicesBS, decreasing = TRUE))
 

### 2017-05-30 doc/dispRity-manual.Rmd
disparity.per.group manual

@@
 
-This is a package for measuring disparity in `R`. It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values so they can easily be compared.
-<!-- Add version and table of content here! -->
+This is a package for measuring disparity in `R`.
+It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values so they can easily be compared.
+This manual is based on the version `1.0`.
 
@@
 You can cite both the package or this manual with the following citation:
+
 > Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646
@@
 
-One other typical analysis done with disparity is to look at the difference in disparity between several groups.
+Alternatively, you can be interested in looking at disparity between groups rather than through time.
+For example, on question could be is there a difference in disparity between two groups?
+
+To perform such analysis, you will need:
+
+ * An ordinated matrix (always!)
+ * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix.
+For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4.
+Note that elements can be present in multiple groups at once.
+ * Your definition of disparity (here the sum of variances)
 

### 2017-06-10 doc/dispRity-manual.Rmd
Manual update

@@
 > Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646
+
+<!-- TG: I'm gonna change that to a newest version with the submission -->
 
@@
 
+This following section contains specific information relative to some functions.
+If one of your questions are not covered in these sections, please refer to the function manuals in R or send me an email. <!-- Add anchor to contacts -->
+
 ## Time slicing
 
-Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
-The `time.subsamples` function allows to create subsamples containing all the elements present at specific points in time or during specific periods.
-This functions needs as input an ordinated space and a matching phylogenetic tree.
+The function `time.subsamples` allows to divide the ordinated matrix into different time subsamples given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
+Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.
+
 Two types of time subsamples can be performed by using the `method` option:
@@
  *  discrete time subsamples (or time-binning) using `method = discrete`;
-
  *  continuous time subsamples (or time-slicing) using `method = continuous`.
@@
  *  defining the boundaries of the time bins (when `method = discrete`);
-
  *  defining the time slices (when `method = continuous`).
@@
 In this example, the taxa were split inside each time-bin according to their age.
-Note however that some taxa might span between some time bins and should be included in more than one.
-This is possible by providing a table containing the first and last occurrence data:
+However, the taxa here are considered like single points in time.
+It is totally possible that the taxa in question could have a longer longevity (e.g. if the taxonomic focus is at the genera level or above) and that it spans across multiple time bins.
+In this case, it is possible to include them in more than one bin by providing a table of First and last occurrence dates (FAD/LAD).
+This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:
 
 ```{r, eval=FALSE}
-## Generating three time bins containing the data present every 40 Ma, including taxa that might
-## span between time bins.
-(time_bins <- time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree,
-    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages))
+## Displaying the table of first and last occurrence dates for each taxa
+head(BeckLee_ages)
+
+## Generating time bins including taxa that might span between them
+time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+    time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
 ```
 
-This generated indeed a `dispRity` object of 3 subsamples.
 Note that we can also generate equivalent results by just telling the function that we want three time-bins (subsamples) as follow:
@@
 
-We now have three time bins of 44.50368 million years each.
-
 When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1% of the tree length as long as at least three elements are present at that point in time.
@@
 For the time-slicing method (`method = continuous`), the idea is fairly similar.
-This option, however, requires a matrix that contains the ordinated distance between taxa *and* nodes as well as an extra argument being the assumed evolutionary model (via the `model` argument):
+This option, however, requires a matrix that contains the ordinated distance between taxa *and* nodes as well as an extra argument being the assumed evolutionary model (via the `model` argument).
+This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node.
+The model can be one of the following:
 
- *  `acctran` where the data chosen on each time slice is always the one of the offspring
-
- *  `deltran` where the data chosen on each time slice is always the one of the descendant
-
- *  `punctuated` where the data chosen on each time slice is randomly chosen between the offspring or the descendant
-
- *  `gradual` where the data chosen on each time slice is either the offspring or the descendant depending on branch length
+ *  `acctran` where the data chosen along the branch is always the one of the offspring
+ *  `deltran` where the data chosen along the branch is always the one of the descendant
+ *  `punctuated` where the data chosen along the branch is randomly chosen between the offspring or the descendant
+ *  `gradual` where the data chosen along the branch is either the offspring or the descendant depending on branch length
 
 ```{r, eval=FALSE}
-## Generating four time slices every 40 million years assuming a gradual evolutionary model
-(time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
+## Generating four time slices every 40 million years under gradual evolution
+time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
     method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages))
-## Note that in the same way as before, we can also automatically generate the slices
-time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
-    model = "gradual", time = 4)
+    FADLAD = BeckLee_ages)
+## Generating four time slices automatically
+time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree,
+    method = "continuous", model = "gradual", time = 4, FADLAD = BeckLee_ages)
 ```
@@
 
-Once we obtain our different subsamples, it is possible to bootstrap and rarefy them (i.e. pseudo-replicating the data).
-The bootstrap will allow us to make each subsamples more robust to outliers and the rarefaction will allow us to compare subsamples with the same number of elements to get rid of eventual sampling problems.
-The `boot.matrix` allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
-The default options will bootstrap the matrix 1000 times without rarefaction using the "full" bootstrap method (see below).
+One important step in analysing ordinated matrices is to pseudo-replicate the data to seem how robust the results of future analysis are sensitive to outliers in the dataset.
+This can be achieved using the function `boot.matrix` to bootstrap or rarefy the ordinated matrix.
+The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):
 
 ```{r, eval=FALSE}
+## Default bootstrapping
 boot.matrix(data = BeckLee_mat50)
@@
 
-The number of bootstrap pseudo-replicates can be defined using the `bootstraps` option.
+The number of bootstrap replicates can be defined using the `bootstraps` option.
 The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
@@
 
- *  `full` where the bootstrapping is entirely stochastic (all the data is bootstrapped)
-
- *  `single` where only one random elements is replaced by one other random elements each pseudo-replication
-
-This function also allows to rarefy the data using the `rarefaction` argument.
-The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in each subsamples to a minimum of 3 elements).
-It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
-One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
-When left missing, all dimensions from the ordinated matrix are used.
+ *  `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any elements drawn from the dat);
+ *  `single` where only one random element is replaced by one other random element for each pseudo-replicate.
 
@@
 boot.matrix(BeckLee_mat50, boot.type = "single")
+```
 
+This function also allows to rarefy the data using the `rarefaction` argument.
+Rarefaction allows to limit the number of elements to be drawn at each bootstrap replication.
+For example, one might be interested in looking at the effect of reducing the number of elements to the results of the analysis.
+
+This can be achieved by using the `rarefaction` option that draws only *n-x* at each bootstrap replicate (where *x* is the number of elements not sampled).
+The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
+remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsamples to a minimum of 3 elements).
+It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
+
+```{r, eval=FALSE}
 ## Bootstrapping with the full rarefaction
@@
 boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8,3))
+```
 
+One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
+When left missing, all dimensions from the ordinated matrix are used.
+
+```{r, eval = FALSE}
 ## Removing the last axis 50% last dimensions
@@
 
-```{r, eval=FALSE}
-## Bootstrapping and fully rarefying the crown/stem subsamples
-(crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE))
+```{r, eval=TRUE}
+## Creating subsamples of crown and stem mammals
+crown_stem <- custom.subsamples(BeckLee_mat50,
+    group = list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44)))
+## Bootstrapping and rarefying these groups
+boot.matrix(crown_stem, bootstraps = 200, rarefaction = TRUE)
 
+## Creating time slices subsamples
+time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
+    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
+    FADLAD = BeckLee_ages)
 ## Bootstrapping on the time binning/slicing subsamples
-(time_binsBS <- boot.matrix(time_bins, bootstraps = 100))
-(time_slicesBS <- boot.matrix(time_slices, bootstraps = 100))
+boot.matrix(time_slices, bootstraps = 1000)
 ```
@@
 
-## Plotting `dispRity` data
+There are many ways for measuring disparity!
+In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
+For example, one can look at ellipsoid hyper-volume of the ordinated space
+<!-- Cite Donohue -->
+, the sum and the product of the ranges and variances
+<!-- Cite Wills -->
+ or the median position of the elements relative to their centroid.
+ <!-- Cite Sive -->
+Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks or some being more generalist.
+
+Because of this great diversity of metric, the package `dispRity` has not a unique way to measure disparity but rather proposes to facilitate users to define their own disparity metric that will best suit to their particular analysis.
+In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
+However the `metric` argument has to follow certain rules:
+
+1.  it must be composed from one to three `function` objects;
+2.  the(se) function(s) must intake as a first argument a `matrix` or a `vector`;
+3.  the(se) function(s) must be of one of the three levels described below;
+4.  at least one of the functions must be of level 1 or 2 (see below).
+
+### The function levels
+
+The metric functions levels determine the "dimensionality" level of decomposition of the input ordinated matrix.
+In other words, each level designate the dimensions of the output: either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimensions.
+
+![Illustration of the different levels of output of a function with an input `matrix`](dispRity_fun.png)
+
+#### A level 1 function
+
+A level 1 function will decompose a `matrix` or a `vector` into a single value:
+
+```{r}
+## Creating a dummy matrix
+dummy_matrix <- matrix(rnorm(12), 4, 3)
+
+## Example of level 1 functions
+mean(dummy_matrix)
+median(dummy_matrix)
+```
+
+Any summary metric such as mean or median are good examples of level 1 functions as they reduce the matrix to a single dimension (i.e. one value).
+
+#### A level 2 function
+
+A level 2 function will decompose a `matrix` into a `vector`.
+
+```{r}
+## Defining the function as the product of rows
+prod.rows <- function(matrix) apply(matrix, 1, prod)
+
+## A level 2 metric
+prod.rows(dummy_matrix)
+```
+
+Several level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` function that calculates the variance of each dimension of the ordinated matrix or the `ranges` one that calculates the range of each dimension.
+
+#### A level 3 function
+
+Finally a level 3 function will transform the matrix into another matrix.
+Note that the dimension of the output matrix don't need to match the the input one:
+
+```{r}
+## A level 3 metric
+var(dummy_matrix)
+
+## A level 3 metric with a forced matrix output
+as.matrix(dist(dummy_matrix))
+```
+
+### `make.metric`
+
+Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the euclidean distance between each elements and the a position in the ordinated space.
+The `make.metric` function implemented in `dispRity` is designed to help testing and finding the level of the functions.
+This function tests:
+
+1.  if your function can deal with a `matrix` or a `vector` as an input;
+2.  which is your function's level according to it's output (level 1, 2 or 3, see above);
+3.  whether the function can properly be implemented in the `dispRity` function (the function is fed into a `lapply` loop).
+
+For example, let's see if the functions described above are the right levels:
+
+```{r}
+## Which level is the mean function? And is able to be used in the dispRity?
+make.metric(mean)
+
+## Same questions from the prod.rows function:
+make.metric(prod.rows)
+
+## Same questions for the var function:
+make.metric(var)
+```
+
+A silent version of the function is also available to be included in your code.
+This can be done using the option `silent = TRUE` and will simply output the level of the metric.
+
+```{r}
+## Testing whether mean is level 1
+if(make.metric(mean, silent = TRUE) != "level1") {
+    message("The metric is not level 1.")
+}
+## Testing whether var is level 1
+if(make.metric(var, silent = TRUE) != "level1") {
+    message("The metric is not level 1.")
+}
+```
+
+### metric argument in `dispRity`
+
+Using this metric structure, we can easily use any disparity metric in the `dispRity` function as follows:
+
+```{r}
+## Measuring disparity as the standard deviation of all the value of the
+## ordinated matrix (level 1 function).
+summary(dispRity(BeckLee_mat50, metric = sd))
+
+## Measuring disparity as the standard deviation of the variance of each axis of
+## the ordinated matrix (level 1 and level 2 functions).
+summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))
+
+## Measuring disparity as the standard deviation of the variance of each axis of
+## the variance covariance matrix (level 1, level 2 and level 3 functions).
+
+summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
+```
+
+Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function levels (using `make.metric`) and apply them to the data in decreasing order (level 3 > level 2 > level 1).
+
+```{r}
+## Disparity as the standard deviation of the variance of each axis of the
+## variance covariance matrix:
+disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
+
+## Same but using a different functions order for the metric argument
+disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)), round = 10)
+
+## Both ways output the same disparity values:
+disparity1 == disparity2
+```
+
+#### What level is the disparity?
+
+In these examples, we considered disparity to be a single value.
+For example, in the previous one, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
+This results in single value to be calculated.
+
+In some cases however, it might also be interesting to calculate disparity as a distribution.
+For example, one of the metrics commonly used in palaeobiology is the sum of variances of each dimensions (`metric = c(sum, variances)`) [@Wills1994].
+However, it might be interesting to look directly at the distribution of the variances itself.
+
+```{r}
+## A one dimensional disparity value
+(sum_variances1 <- summary(dispRity(BeckLee_mat50, metric = c(median, variances))))
+
+## A two dimensional disparity
+(summary(all_variances <- (dispRity(BeckLee_mat50, metric = variances))))
+```
+
+Looking at all the variances rather than simply their median can tell us that (by the look of it) the variances of each dimensions seems to be normally distributed.
+
+> It is also possible to reuse an already calculated disparity distribution and summarise it using a level 1 metric:
+
+```{r}
+## Transforming the two dimensional one in a one dimensional one
+sum_variances2 <- summary(dispRity(all_variances, metric = median))
+
+## Both ways output the same disparity values:
+sum_variances1 == sum_variances2
+```
+
+### Metrics implemented in `dispRity`
+
+Several disparity metrics are directly implemented in the `dispRity` package.
+The detailed list can be found in `?dispRity.metric` along with some explanations on each metric.
+
+Level | Name | Description | Source | 
+------|------|---------------------------------------------------|--------|
+3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
+3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
+3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
+3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
+3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
+2 | `ranges` | The range of each dimension | `dispRity` |
+2 | `variances` | The variance of each dimension | `dispRity` |
+2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
+1 | `mode.val` | The modal value | `dispRity` |
+
+1: The functions estimation of the eigen value only works for MDS or PCoA ordinations (not PCA).
+
+2: Note that by default, the centroid is the centroid of the elements.
+It can however be fixed to an arbitrary value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
+
+### Equations and implementations:
+Some of the functions described below are directly implemented in the `dispRity` package and do not require any dependencies  ([see implementation here](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R)).
+
+\begin{equation}
+    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
+\end{equation}
+With *k* being the number of dimensions, $\lambda_i$ is the eigen value of each dimension.
+
+\begin{equation}
+    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
+\end{equation}
+With *k* being the dimensions.
+
+\begin{equation}
+    ranges = |max(k_i) - min(k_i)|
+\end{equation}
+With *k* being the dimensions.
+
+\begin{equation}
+    variances = \sigma^{2}{k_i}
+\end{equation}
+With *k* being the dimensions and $\sigma^{2}$ their variance.
+
+\begin{equation}
+    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
+\end{equation}
+With *n* being each element in the ordinated space, *k* being its dimensions and $Centroid_{k}$ their mean (or can be set to an arbitrary value) and.
+
+### Metrics run through
+Here is a run through of the main metrics implemented in `dispRity`.
+First, we will create a dummy ordinated space using the `space.maker` utility function:
+
+```{r}
+## Creating a 10*5 normal space
+set.seed(1)
+dummy_space <- space.maker(10, 5, rnorm)
+```
+
+#### Volumes and surface metrics
+The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `hyper.volume` all measure the surface or the volume of the ordinated space occupancy:
+
+```{r}
+## Calculating the ellipsoid volume
+summary(dispRity(dummy_space, metric = ellipse.volume))
+```
+> Because there is only one subsamples (i.e. one matrix) in the dispRity object, this operation is the equivalent of doing simply `ellipse.volume(dummy_space)` (with rounding).
+
+```{r}
+## Calculating the convex hull surface
+summary(dispRity(dummy_space, metric = convhull.surface))
+
+## Calculating the convex hull volume
+summary(dispRity(dummy_space, metric = convhull.volume))
+```
+
+The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
+For the full and correct calculation of the volume of the ordinated space, it is possible to us the `hyper.volume` function that intakes more options but takes longer to calculate.
+
+```{r}
+## Calculating the true multidimensional volume
+summary(dispRity(dummy_space, metric = hyper.volume))
+
+## Calculating the volume with different options to be passed to hypervolume::hypervolume
+summary(dispRity(dummy_space, metric = hyper.volume, repsperpoint = 100, quantile = 0.5))
+```
+> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [Curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) which often results in near 0 disparity values.
+
+#### Ranges, variances and diagonal
+
+The functions `ranges`, `variances` and `diagonal` all measure properties of the ordinated space base on it's dimensions properties (they are also less affected by the Curse of dimensionality):
+
+`ranges` and `variances` both work on the same principle and measure the range/variance of each dimensions:
+
+```{r}
+## Calculating the ranges of each dimensions in the ordinated space
+ranges(dummy_space)
+
+## Calculating disparity as the distribution of these ranges
+summary(dispRity(dummy_space, metric = ranges))
+
+## Calculating disparity as the sum and the product of these ranges
+summary(dispRity(dummy_space, metric = c(sum, ranges)))
+summary(dispRity(dummy_space, metric = c(prod, ranges)))
+
+## Calculating the variances of each dimensions in the ordinated space
+variances(dummy_space)
+
+## Calculating disparity as the distribution of these variances
+summary(dispRity(dummy_space, metric = variances))
+
+## Calculating disparity as the sum and the product of these variances
+summary(dispRity(dummy_space, metric = c(sum, variances)))
+summary(dispRity(dummy_space, metric = c(prod, variances)))
+```
+
+The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest euclidean distance in our 5 dimensional space):
+
+```{r}
+## Calculating the ordinated space's diagonal
+summary(dispRity(dummy_space, metric = diagonal))
+```
+
+> This metric is only mathematically valid if the dimensions within the space are all orthogonal!
+
+#### Centroids metrics
+
+The `centroids` metric allows to measure the position of the different elements compared to a fixed point in the ordinated space.
+By default, this function measures the distance between each elements and the central point between them:
+
+```{r}
+## Calculating the distribution of the distances between each elements and the centroid
+summary(dispRity(dummy_space, metric = centroids))
+
+## Calculating disparity as the median value of these distances
+summary(dispRity(dummy_space, metric = c(median, centroids)))
+```
+
+It is however possible to fix the coordinates of a specific point in the ordinated space as long as it has the correct number of dimensions:
+
+```{r}
+## Calculating the distance between the elements and the origin of the ordinated space
+summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))
+
+## Calculating disparity as the distance between the elements and a specific point
+summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
+```
 

### 2017-06-16 doc/dispRity-manual.Rmd
Updated manual

@@
 ## Plotting the results
-plot(disparity_data)
+plot(disparity_data, type = "continuous")
 
@@
 
-An alternative way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
+## Summarising dispRity data (plots)
+
+Because of its architecture, printing `dispRity` objects only summarise their content but does not print the disparity value measured or associated analysis.
+<!-- link to dispRity objects -->
+To actually summary what's in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphic representation of the results.
+
+### Summarising `dispRity` data
+This function is a S3 function (`summary.dispRity`) allowing to summarise the content of `dispRity` objects that contain disparity calculations.
+
+```{r}
+## Example data from previous sections
+crown_stem <- custom.subsamples(BeckLee_mat50,
+    group = list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44)))
+## Bootstrapping and rarefying these groups
+boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 200, rarefaction = TRUE)
+## Calculate disparity
+disparity_crown_stem <- dispRity(boot_crown_stem, metric = c(sum, variances))
+
+## Creating time slices subsamples
+time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
+    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
+    FADLAD = BeckLee_ages)
+## Bootstrapping on the time binning/slicing subsamples
+boot_time_slices <- boot.matrix(time_slices, bootstraps = 1000)
+## Calculate disparity
+disparity_time_slices <- dispRity(boot_time_slices, metric = c(sum, variances))
+```
+
+This objects are easy to summarise as follow:
+
+```{r}
+## Default summary
+summary(disparity_time_slices)
+```
+
+Information about the number of elements per subsamples as well as the observed (i.e. non-bootstrapped) disparity is also calculated.
+This is specifically handy when rarefying the data for example
+
+```{r}
+head(summary(disparity_crown_stem))
+```
+
+The summary functions can also take various options such as the `quantile` values for the confidence intervals levels (by default, the 50 and 95 quantiles are calculated); the `cent.tend` for the  central tendency to use for summarising the results (default is `median`); the `rounding` option corresponding to the number of decimals to print (default is `2`) and the `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
+These options can easily be changed from the defaults as follows:
+
+```{r}
+## Same but using the 88 quantile and the standard deviation as the summary. 
+summary(disparity_time_slices, quantile = 88, cent.tend = sd)
+
+## Printing the details of the object and rounding the values to the 5th decimal
+summary(disparity_time_slices, recall = TRUE, rounding = 5)
+```
+
+Information about the number of elements per subsamples as well as the observed (i.e. non-bootstrapped) disparity is also calculated.
+This is specifically handy when rarefying the data for example
+
+Note that the summary table is a `data.frame`, hence it is easy to modify as any table using `dplyr`.
+You can also export it in `csv` format using `write.csv` or even directly export into `LaTeX` format using the following;
+
+```{r, eval = FALSE}
+## Loading the xtable package
+require(xtable)
+## Converting the table in LaTeX
+xtable(summary(disparity_time_slices))
+```
+
+### Plotting `dispRity` data
+An alternative (and more fun!) way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
 This function intakes the same options as `summary.dispRity` along side with various graphical options described in the function manual (see `?plot.dispRity`).
@@
 
-```{r, fig.width=6, fig.height=6, eval=FALSE}
+```{r, fig.width=8, fig.height=8}
 ## Graphical parameters
 op <- par(mfrow = c(2,2), bty = "n")
+
 ## Plotting continuous disparity results
-plot(disp_time_binsBS, type = "continuous")
+plot(disparity_time_slices, type = "continuous")
+
 ## Plotting discrete disparity results
-plot(disp_crown_stemBS, type = "box")
-## Plotting the disparity using lines for the rarefaction level of 20 elements only
-plot(disp_crown_stemBS, type = "line", rarefaction = 20)
-## Plotting the disparity using polygons while also displaying the number of elements
-plot(disp_crown_stemBS, type = "polygon", elements = TRUE)
-```
-
-> The `type` argument is not mandatory for the two first plots since both `continuous` and `box` are the default options for respectively continuous and discrete data.
+plot(disparity_crown_stem, type = "box")
+
+## Same but using lines for the rarefaction level of 20 elements only
+plot(disparity_crown_stem, type = "line", rarefaction = 20)
+
+## Same but using polygons while also displaying the number of elements
+plot(disparity_crown_stem, type = "polygon", elements = TRUE)
+
+## Resetting graphical parameters
+par(op)
+```
 
@@
 
-```{r, fig.width=6, fig.height=6, eval=FALSE}
+```{r, fig.width=8, fig.height=8}
 ## Graphical options
 par(bty = "n")
+
 ## Plotting the results with some classic options from plot
-plot(disp_time_slicesBS, col = c("blue", "orange", "green"), ylab = c("Some measurement"),
-    xlab = "Some other measurement", main = "Many options...", ylim = c(0,2.1))
+plot(disparity_time_slices, col = c("blue", "orange", "green"),
+    ylab = c("Some measurement"), xlab = "Some other measurement",
+    main = "Many options...", ylim = c(5, 0), xlim = c(4, 0))
+
 ## Adding a legend
-legend("topleft", legend = c("Central tendency", "Confidence interval 1", "Confidence interval 2"),
+legend("topleft", legend = c("Central tendency",
+                             "Confidence interval 1",
+                            "Confidence interval 2"),
      col = c("blue", "orange", "green"), pch = 19)
+
+## Resetting graphical parameters
+par(op)
 ```
@@
 
-```{r, fig.width=6, fig.height=6, eval=FALSE}
+```{r, fig.width=8, fig.height=8}
 ## Graphical options
 par(bty = "n")
+
 ## Plotting the results with some plot.dispRity arguments
-plot(disp_time_slicesBS, quantile = c(seq(from=10, to=100, by=10)), cent.tend = sd, type = "c",
-    elements = TRUE, col = c("black", rainbow(10)), ylab = c("Disparity", "Diversity"),
-    xlab = "Time (in in units from past to present)", time.subsamples = FALSE,
-    observed = TRUE, main = "Many options...")
+plot(disparity_time_slices, quantile = c(seq(from=10, to=100, by=10)),
+    cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
+    ylab = c("Disparity", "Diversity"), time.subsamples = FALSE,
+    xlab = "Time (in in units from past to present)", observed = TRUE,
+    main = "Many more options...")
+
+## Resetting graphical parameters
+par(op)
 ```
@@
 
-```{r, fig.width=6, fig.height=6, eval=FALSE}
+```{r, fig.width=8, fig.height=8}
 ## Graphical options
 par(bty = "n")
+
 ## Plotting the continuous disparity with a fixed y axis
-plot(disp_time_slicesBS, ylim = c(0.5,2))
-## Adding the discrete data using the same fixed y axis to have the results at the same scale
-plot(disp_time_binsBS, type = "line", ylim=c(0.5,2), xlab="", ylab="", add = TRUE)
-```
+plot(disparity_time_slices, ylim = c(1, 4))
+## Adding the discrete data
+plot(disparity_crown_stem, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
+    add = TRUE)
+
+## Resetting graphical parameters
+par(op)
+```
+
+<!-- Change this example to add time bins! -->
 
@@
 
-```{r, fig.width=6, fig.height=6, eval=FALSE}
+```{r, fig.width=8, fig.height=8}
 ## Graphical options
 par(bty = "n")
+
 ## Plotting the rarefaction curves
-plot(disp_crown_stemBS, rarefaction = TRUE)
+plot(disparity_crown_stem, rarefaction = TRUE)
+
+## Resetting graphical parameters
+par(op)
 ```

### 2017-06-19 doc/dispRity-palaeo-demo.Rmd

@@
 
-This demo aims to give quick overview of the `dispRity` package (v.0.2) for palaeobiology analysis.
-Please refer to [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the [`dispRity` manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.Rmd) that explains the functions in more detail.
+This demo aims to give quick overview of the `dispRity` package (v.0.3) for palaeobiology analysis.
+Please refer to [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html) that explains the functions in more detail.
 
@@
 One might be interested in studying how the occupancy of the morphospace changed through time by measuring a summary metric of this morphospace through time, called **disparity**.
-Disparity can then be seen as a value summarising the diversity of morphologies.
+Disparity can then be seen as a value or a distribution of valuess summarising the diversity of morphologies.
 
@@
 
-In this example, we are going to use a subset of the data from @beckancient2014 that contains an ordinated cladistic matrix and a
-phylogenetic tree containing 50 taxa along side with a table containing some first and last occurence data for several taxa.
-The ordinated matrix will represent our full morphospace, i.e. all the morphologies that ever existed through time (for this dataset).
+In this example, we are going to use a subset of the data from [@beckancient2014] that contains an ordinated cladistic matrix and a
+phylogenetic tree containing 50 taxa along side with a table containing some first and last occurences data for several taxa.
+The ordinated matrix will represent our full morphospace, i.e. all the mammalian morphologies that ever existed through time (for this dataset).
 
@@
 
-## The morphospace:
+## Loading the morphospace:
 data(BeckLee_mat50)
@@
 
-## A list of first and last occurrences data for some fossils
+## Showing a list of first and last occurrences data for some fossils
 data(BeckLee_ages)
@@
 
-## And a phylogeny:
+## Plotting a phylogeny
 data(BeckLee_tree)
@@
 
-Note that you can have an even nicer tree if you use the `strap` package:
+> You can have an even nicer looking tree if you use the `strap` package!
 
@@
 
-One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time seriess that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
-Basically, the full morphospace in the previous step represents the total amount of morphologies in all time and is likely to be greater than any time series of the morphospace.
+One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time subsamples that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
+Basically, the full morphospace in the previous step represents the total amount of morphologies in all time and is likely to be greater than any time subsamples of the morphospace.
 
-The `dispRity` package provides a `time.series` function that allows to split the morphospace time slices (using `method = continuous`) or into time bins (using `method = discrete`).
+The `dispRity` package provides a `time.subsamples` function that allows to split the morphospace time slices (using `method = continuous`) or into time bins (using `method = discrete`).
 In this example, we are going to split the morphospace into 5 equal time bins of 20 million years long from 100 million years ago (Mya) to the present.
-Note that we will also provide to the function the table containing the first and last occurrences data for some fossils to take into account the fact that some fossils might span between our different time bins.
+We will also provide to the function the table containing the first and last occurrences data for some fossils to take into account the fact that some fossils might span between our different time bins.
 
@@
 ## Creating the vector of time bins ages
-time_bins <- rev(seq(from = 0, to = 100, by = 20))
+(time_bins <- rev(seq(from = 0, to = 100, by = 20)))
 
-## Splitting the morphospace using the time.series function
-binned_morphospace <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
+## Splitting the morphospace using the time.subsamples function
+(binned_morphospace <- time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree,
     method = "discrete", time = time_bins, inc.nodes = FALSE,
-    FADLAD = BeckLee_ages)
+    FADLAD = BeckLee_ages))
 ```
@@
 The output object is a `dispRity` object.
-For details about this object class, please refer to the [`dispRity` manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.Rmd).
-In brief, `dispRity` objects are lists of different elements (i.e.
-disparity results, morphospace time seriess, etc...) that display only a summary of the object when calling the object for avoiding jamming the `R` console.
+For details about this object class, please refer to the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html).
+In brief, however, `dispRity` objects are lists of different elements (i.e.
+disparity results, morphospace time subsamples, morphospace attributes, etc.) that display only a summary of the object when calling the object for avoiding jamming the `R` console.
 
 ```{r}
-## The object class
+## Printing the class of the object
 class(binned_morphospace)
 
-## What's in the object (messy version for the console!)
+## Printing the content of the object
 str(binned_morphospace)
@@
 
-## How it's printed by default
+## Printing the object as a dispRity class
 binned_morphospace
@@
 
-Note that these objects will gradually contain more information when reaching the following steps in the disparity-through-time analysis.
+> These objects will gradually contain more information when reaching the following steps in the disparity-through-time analysis.
 
@@
 
-Once we obtain our different time series, we can bootstrap and rarefy them (i.e.
-pseudo-replicating the data).
-The bootstrapping will allow to make each series more robust to outliers and the rarefaction will allow us to compare slices with the same number of taxa to get rid of eventual sampling problems (i.e.
-more taxa in one bin than the other).
+Once we obtain our different time subsamples, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
+The bootstrapping will allow to make each subsamples more robust to outliers and the rarefaction will allow us to compare subsamples with the same number of taxa to get rid of eventual sampling problems (i.e.
+more taxa in one subsamples than the other).
 The `boot.matrix` function that allows to bootstrap the `dispRity` object and the `rarefaction` option within allows to rarefy it.
@@
 ```{r}
-## Bootstrapping each time series 100 times
-boot_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100)
+## Bootstrapping each time subsamples 100 times
+(boot_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100))
 
-## Getting the minimum number of rows (=taxa) in the time seriess
-min(unlist(lapply(binned_morphospace$data, nrow)))
+## Getting the minimum number of rows (=taxa) in the time subsamples
+min(unlist(lapply(boot_bin_morphospace$subsamples, lapply, nrow)))
 
-## Bootstrapping each time series 100 times and rarefying them 
-rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
-    rarefaction = 6)
-
-## Note that the output objects are both from the "dispRity" class ...
-class(boot_bin_morphospace); class(rare_bin_morphospace)
-
-## ... and display some more information on the previous step when printed:
-boot_bin_morphospace
-rare_bin_morphospace
+## Bootstrapping each time subsamples 100 times and rarefying them 
+(rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
+    rarefaction = 6))
 ```
@@
 
-We can now calculate the disparity within each time series along side with some confidence intervals generated by the pseudo-replication step above (bootstraps/rarefaction).
+We can now calculate the disparity within each time subsamples along side with some confidence intervals generated by the pseudo-replication step above (bootstraps/rarefaction).
 Disparity can be calculated in many ways and this package allows user to come up with their own disparity metrics.
-For more info, please refer to the [`dispRity` metric vignette](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.Rmd).
+For more info, please refer to the [`dispRity` metric vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html).
 
-In this example, we are going to calculate the spread of the data in each time series by calculating the disparity as the sum of the variance of each dimension of the morphospace in each time series using the `dispRity` function.
-Thus, in this example, the disparity is defined by the multi-dimensional variance of each time series (i.e. the spread of the taxa within the morphospace).
+In this example, we are going to calculate the spread of the data in each time subsamples by calculating the disparity as the sum of the variance of each dimension of the morphospace in each time subsamples using the `dispRity` function.
+Thus, in this example, the disparity is defined by the multi-dimensional variance of each time subsamples (i.e. the spread of the taxa within the morphospace).
 Note that this metric comes with a caveat not solved here since it ignores the covariance among the dimensions of the morphospace.
@@
 ```{r}
-## Disparity for the bootstrapped data
-boot_disparity <- dispRity(boot_bin_morphospace, metric = c(sum, variances))
+## Calculating disparity for the bootstrapped data
+(boot_disparity <- dispRity(boot_bin_morphospace, metric = c(sum, variances)))
 
-## Disparity for the rarefied data
-rare_disparity <- dispRity(rare_bin_morphospace, metric = c(sum, variances))
+## Calculating disparity for the rarefied data
+(rare_disparity <- dispRity(rare_bin_morphospace, metric = c(sum, variances)))
+```
 
-## Again, these objects are both from the "dispRity" class and just display
-## some information when called, note the results!
-boot_disparity
-rare_disparity
-
-## To display the actual results we can use the summary function
-summary(boot_disparity)
-summary(rare_disparity)
-```
+The `dispRity` function does not actually display the calculated disparity values but rather only the properties of the disparity object (size, subsamples, metric, etc.).
+To display the actual calculated scores, we need to summarise the disparity object using the S3 method `summary` that is applied on a `dispRity` object (see `?summary.dispRity` for more details).
 
@@
 ```{r}
-## See more options on the summary function:
-?summary.dispRity
+## Summarising the disparity results
+summary(boot_disparity)
+summary(rare_disparity)
 ```
+> The summary.dispRity function comes with many options on which values to calculate (central tendency and quantiles) and on how many digits to display.
+Refer to the function's manual for more details.
 
@@
 It is sometimes easier to visualise the results in a plot than in a table.
-For that we can use the `plot` function to plot the `dispRity` objects.
+For that we can use the `plot` S3 function to plot the `dispRity` objects (see `?plot.dispRity` for more details).
 
-```{r, fig.width=6, fig.height=6}
+```{r, fig.width=12, fig.height=6}
 ## Graphical options
@@
 
-As we can see in this case, the rarefaction tends to increase the confidence intervals around the mean.
-
 Testing differences
@@
 
-Finally, to draw some conclusions from these results, we need to apply some statistical testing.
-As stated before, we wanted to test whether the disparity changed through time in this dataset since the last 100 million years.
-To do so, we can apply some comparisons of the means between each time-bins in a sequential manner to see whether the disparity in bin X is equal to the disparity in bin Y which is in turn equal to the disparity in bin Z, etc.
+Finally, to draw some valid conclusions from these results, we can apply some statistical testing to the data.
+We can test, for example, if mammalian disparity changed through time since the last 100 million years (Mya).
+To do so, we can apply some comparisons of the means between each time-bins in a sequential manner to see whether the disparity in bin *n* is equal to the disparity in bin *n+1* which is in turn equal to the disparity in bin *n+2*, etc.
 Because our data is temporally auto-correlated (i.e.
-what happens in bin Y depends on what happened in bin X) and was pseudo-replicated (i.e.
-each bootstrap draw creates non-independent time seriess because they are all based on the same time series), we are going to apply a non-parametric mean comparison: the `wilcox.test`.
+what happens in bin *n+1* depends on what happened in bin *n*) and was pseudo-replicated (i.e.
+each bootstrap draw creates non-independent time subsamples because they are all based on the same time subsamples), we are going to apply a non-parametric mean comparison: the `wilcox.test`.
 Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis (see `?p.adjust` for more details).
@@
 
-Here our results show significant changes in disparity through time between all time bins during the last 100 million years.
-However, when looking at the rarefied results, there is no significant difference between the time bins in the Cenozoic (60-40, 40-20 and 20-0 Mya), suggesting that the differences detected in the first test might just be due to the differences in number of taxa sample (respectively 13, 6 and 10 in each time bin).
+Here our results show significant changes in disparity through time between all time bins (all p-values > 0.05).
+However, when looking at the rarefied results, there is no significant difference between the time bins in the Palaeogene (60-40 to 40-20 Mya), suggesting that the differences detected in the first test might just be due to the differences in number of taxa sample (13 or 6 taxa in each time bin).
+
+# References

### 2017-06-19 doc/dispRity-palaeo-demo.Rmd
Revert "Merge pull request #33 from TGuillerme/master"
This reverts commit 53d9a8f98c5ad8d65cc1f4466045e3da67ef2d8d, reversing
changes made to dac6e1109659bb50d7f261c03a9e0969fe791807.

@@
 
-One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time subsamples that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
-Basically, the full morphospace in the previous step represents the total amount of morphologies in all time and is likely to be greater than any time subsamples of the morphospace.
+One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time series that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
+Basically, the full morphospace in the previous step represents the total amount of morphologies in all time and is likely to be greater than any time series of the morphospace.
 
-The `dispRity` package provides a `time.subsamples` function that allows to split the morphospace time slices (using `method = continuous`) or into time bins (using `method = discrete`).
+The `dispRity` package provides a `time.series` function that allows to split the morphospace time slices (using `method = continuous`) or into time bins (using `method = discrete`).
 In this example, we are going to split the morphospace into 5 equal time bins of 20 million years long from 100 million years ago (Mya) to the present.
@@
 
-## Splitting the morphospace using the time.subsamples function
-(binned_morphospace <- time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree,
+## Splitting the morphospace using the time.series function
+(binned_morphospace <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
     method = "discrete", time = time_bins, inc.nodes = FALSE,
@@
 In brief, however, `dispRity` objects are lists of different elements (i.e.
-disparity results, morphospace time subsamples, morphospace attributes, etc.) that display only a summary of the object when calling the object for avoiding jamming the `R` console.
+disparity results, morphospace time series, morphospace attributes, etc.) that display only a summary of the object when calling the object for avoiding jamming the `R` console.
 
@@
 
-Once we obtain our different time subsamples, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
-The bootstrapping will allow to make each subsamples more robust to outliers and the rarefaction will allow us to compare subsamples with the same number of taxa to get rid of eventual sampling problems (i.e.
-more taxa in one subsamples than the other).
+Once we obtain our different time series, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
+The bootstrapping will allow to make each series more robust to outliers and the rarefaction will allow us to compare series with the same number of taxa to get rid of eventual sampling problems (i.e.
+more taxa in one series than the other).
 The `boot.matrix` function that allows to bootstrap the `dispRity` object and the `rarefaction` option within allows to rarefy it.
@@
 ```{r}
-## Bootstrapping each time subsamples 100 times
+## Bootstrapping each time series 100 times
 (boot_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100))
 
-## Getting the minimum number of rows (=taxa) in the time subsamples
-min(unlist(lapply(boot_bin_morphospace$subsamples, lapply, nrow)))
+## Getting the minimum number of rows (=taxa) in the time series
+min(unlist(lapply(boot_bin_morphospace$series, lapply, nrow)))
 
-## Bootstrapping each time subsamples 100 times and rarefying them 
+## Bootstrapping each time series 100 times and rarefying them 
 (rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
@@
 
-We can now calculate the disparity within each time subsamples along side with some confidence intervals generated by the pseudo-replication step above (bootstraps/rarefaction).
+We can now calculate the disparity within each time series along side with some confidence intervals generated by the pseudo-replication step above (bootstraps/rarefaction).
 Disparity can be calculated in many ways and this package allows user to come up with their own disparity metrics.
@@
 
-In this example, we are going to calculate the spread of the data in each time subsamples by calculating the disparity as the sum of the variance of each dimension of the morphospace in each time subsamples using the `dispRity` function.
-Thus, in this example, the disparity is defined by the multi-dimensional variance of each time subsamples (i.e. the spread of the taxa within the morphospace).
+In this example, we are going to calculate the spread of the data in each time series by calculating the disparity as the sum of the variance of each dimension of the morphospace in each time series using the `dispRity` function.
+Thus, in this example, the disparity is defined by the multi-dimensional variance of each time series (i.e. the spread of the taxa within the morphospace).
 Note that this metric comes with a caveat not solved here since it ignores the covariance among the dimensions of the morphospace.
@@
 
-The `dispRity` function does not actually display the calculated disparity values but rather only the properties of the disparity object (size, subsamples, metric, etc.).
+The `dispRity` function does not actually display the calculated disparity values but rather only the properties of the disparity object (size, series, metric, etc.).
 To display the actual calculated scores, we need to summarise the disparity object using the S3 method `summary` that is applied on a `dispRity` object (see `?summary.dispRity` for more details).
@@
 what happens in bin *n+1* depends on what happened in bin *n*) and was pseudo-replicated (i.e.
-each bootstrap draw creates non-independent time subsamples because they are all based on the same time subsamples), we are going to apply a non-parametric mean comparison: the `wilcox.test`.
+each bootstrap draw creates non-independent time series because they are all based on the same time series), we are going to apply a non-parametric mean comparison: the `wilcox.test`.
 Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis (see `?p.adjust` for more details).

### 2017-06-19 R/dispRity-package.R

@@ -67,7 +67,7 @@ NULL
 #'
 #' And example of \code{dispRity} object.
 #'
-#' This matrix is based on \code{\link{BeckLee}} dataset and split into 7 continuous subsamples (\code{\link{time.subsamples}}).
+#' This matrix is based on \code{\link{BeckLee}} dataset and split into 7 continuous series (\code{\link{time.series}}).
 #' It was bootstrapped 100 times (\code{\link{boot.matrix}}) with four rarefaction levels.
 #' Disparity was calculated as the \code{\link[stats]{median}} of the \code{\link{centroids}} (\code{\link{dispRity}}).
 #'

### 2017-06-19 R/geomorph.ordination.R

@@ -1,45 +1,0 @@
-#' @title Imports data from geomorph
-#'
-#' @description Takes geomorph landmark data and computes both a procrustes analysis and an ordination 
-#'
-#' @param data An array (p x k x n) containing landmark coordinates for a set of specimens.
-#' @param k The number of dimensions in the ordination. If left empty, the number of dimensions is set to number of rows - 1.
-#' @param ... Any optional arguments to be passed to \code{\link[geomorph]{gpagen}}.
-#' 
-# @examples
-# require(geomorph)
-# ## Loading the plethodon dataset
-# data(plethodon)
-# 
-# ## Obtaining the ordination matrix
-# ordination <- geomorph.ordination(plethodon) 
-#
-#' @seealso \code{\link[geomorph]{gpagen}}, \code{\link[stats]{cmdscale}}, \code{\link{custom.subsamples}}, \code{\link{time.subsamples}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}.
-#' 
-
-#source("sanitizing.R")
-geomorph.ordination <- function(data, k, ...) {
-
-    stop("IN DEVELOPMENT")
-
-    match_call <- match.call()
-
-    ## Sanitising
-    ## data
-    check.class(data, "list")
-    if(is.null(data$land)) {
-        stop(paste(match_call$data, "must be a list with a 'land' element."))
-    } else {
-        if(class(data$land) != "array") {
-            stop(paste(match_call$data, "$land must be a (p x k x n) array.", sep = ""))
-        }
-    }
-
-    
-    ## Procrustes
-    Y.gpa <- geomorph::gpagen(data$land)
-
-    ## Ordination?
-
-    return(Y.gpa)
-}

### 2017-06-19 R/make.metric.R

@@ -44,36 +44,33 @@
 #' @author Thomas Guillerme
 
 
-make.metric <- function(fun, ..., silent = FALSE) {
-    ## Sanitizing
-    ## fun
+make.metric<-function(fun, ..., silent = FALSE) {
+    #Sanitizing
+    #fun
     check.class(fun, "function")
 
-    ## Getting the function name
+    #Getting the function name
     match_call<-match.call()
 
-    ## Testing the metric
-    ## making the testing matrix
+    #Testing the metric
+    #making the testing matrix
     matrix <- matrix(rnorm(20), 5,4)
 
-    ## Testing the metric
+    #Testing the metric
     test <- NULL
-    op <- options(warn = 2)
-    test <- try(fun(matrix, ...), silent = TRUE)
+    try(test <- fun(matrix, ...), silent = TRUE)
     #try(test <- fun(matrix), silent = TRUE) ; warning("DEBUG")
-    options(op)
 
-    if(any(test == "try-error")) {
-        if(silent != TRUE) {
-            stop(paste("The provided metric function generated an error or a warning!\nDoes the following works?\n    ", as.expression(match_call$fun),"(matrix(rnorm(9),3,3))\n(the problem can also come from the optional arguments...)", sep = ""))
-        }
+    #Did the test failed?
+    if(is.null(test)) {
+        if(silent != TRUE) stop(paste("The provided function did not output anything!\nDoes the following works?\n", as.expression(match_call$fun),"(matrix(rnorm(9),3,3))", sep = ""))
     } else {
 
         ##########
-        ## What is the output class of the function?
+        #What is the output class of the function?
         ##########
 
-        ## If class is matrix -> level3.fun
+        #If class is matrix -> level3.fun
         if(class(test) == "matrix") {
             fun_type <- "level3"
             if(silent != TRUE) {
@@ -81,15 +78,15 @@ make.metric <- function(fun, ..., silent = FALSE) {
                 cat(paste("\nAdditional level 2 and/or level 1 function(s) will be needed.", sep = ""))
             }
         } else {
-            ## If class is numeric
+            #If class is numeric
             if(class(test) == "numeric") {
-                ## If only one value -> level1.fun
+                #If only one value -> level1.fun
                 if(length(test) == 1) {
                     fun_type <- "level1"
                     if(silent != TRUE) {
                         cat(paste(as.expression(match_call$fun)," outputs a single value.\n", as.expression(match_call$fun), " is detected as being a level 1 function.", sep = ""))
                     }
[…]

### 2017-06-19 R/morpho.utilities.R

@@ -32,18 +32,20 @@
 #' @param matrix a discrete morphological character matrix.
 #' 
 #' @examples
-#' ## A random multi state matrix
-#' matrix <- matrix(sample(c(0,1,2), 100, TRUE), 10, 10)
+# \dontrun{
+#' ## Create a random matrix
+#' tree <- rcoal(10)
+#' ## setting up the parameters
+#' ## A gamma rate distribution with of shape alpha = 0.5
+#' my_rates = c(rgamma, 2, 1)
+#' ## A fixed substitution rate of 2 (T/T ratio in HKY)
+#' my_substitutions = c(runif, 2, 2)
+#' ## Mk matrix (10*50) (for Mkv models)
+#' matrixMk <- sim.morpho(tree, characters = 50, model = "ER", rates = my_rates) 
 #' 
-#' ## Get the contrast matrix
-#' get.contrast.matrix(matrix)
-#' 
-#' ## Adding inapplicable and missing data to the matrix
-#' matrix[sample(1:100, 10)] <- "?"
-#' matrix[sample(1:100, 10)] <- "-"
-#' 
-#' ## Get the contrast matrix
-#' get.contrast.matrix(matrix)
+#' ## Get the matrix's contrast matrix
+#' get.contrast.matrix(matrixMk)
+# }
 #' 
 #' @seealso \code{\link{check.morpho}}
 #' 
@@ -82,7 +84,7 @@ get.contrast.matrix <- function(matrix) {
 #' @description Apply inapplicable characters to discrete morphological matrix.
 #'
 #' @param matrix A discrete morphological matrix.
-#' @param NAs Either a numeric value of how many characters to make inapplicable or vector of characters inapplicability source (either \code{"character"} or \code{"clade"}; see details). The length of this vector must be at maximum half the total number of characters.
+#' @param inapplicables Optional, a vector of characters inapplicability source (either \code{"character"} or \code{"clade"}; see details). The length of this vector must be at maximum half the total number of characters.
 #' @param tree If any inapplicable source is \code{"clade"}, a tree from where to select the clades.
 #' @param invariant Whether to allow invariant sites among the characters with inapplicable data. If \code{invariant = FALSE} the algorithm will try to remove such characters (if possible).
 #' @param verbose Whether to be verbose or not.
@@ -90,67 +92,62 @@ get.contrast.matrix <- function(matrix) {
 #' 
 #' @details
 #' \itemize{
[…]

### 2017-06-19 R/null.test.R

@@ -21,17 +21,17 @@
 #' ## Calculating the disparity as the ellipsoid volume
 #' obs_disparity <- dispRity(BeckLee_mat50, metric = ellipse.volume)
 #' ## Testing against normal distribution
-#' null.test(obs_disparity, replicates = 100, null.distrib = rnorm)
-# results ; plot(results)
+#' results <- null.test(obs_disparity, replicates = 100, null.distrib = rnorm)
+#' results ; plot(results)
 #' 
 #' \dontrun{
-#' ## Running the test on multiple subsamples
-#' ## Generating the subsamples
-#' groups <- as.data.frame(matrix(data = c(rep(1, 12), rep(2, 13), rep(3, 12),
+#' ## Running the test on multiple series
+#' ## Generating the series
+#' factors <- as.data.frame(matrix(data = c(rep(1, 12), rep(2, 13), rep(3, 12),
 #'      rep(4, 13)), dimnames = list(rownames(BeckLee_mat50))), ncol = 1)
-#' customised_subsamples <- cust.subsamples(BeckLee_mat50, groups)
+#' customised_series <- cust.series(BeckLee_mat50, factors)
 #' ## Bootstrapping the data
-#' bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 100)
+#' bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100)
 #' ## Calculating variances of each dimension
 #' sum_variances <- dispRity(bootstrapped_data, metric = c(sum, variances))
 #' ## Testing against normal distribution
@@ -51,9 +51,9 @@
 # source("sanitizing.R")
 # source("null.test_fun.R")
 # data(BeckLee_mat50)
-# groups <- as.data.frame(matrix(data = c(rep(1, 12), rep(2, 13), rep(3, 25)), dimnames =list(rownames(BeckLee_mat50))), ncol = 1)
-# customised_subsamples <- cust.subsamples(BeckLee_mat50, groups)
-# bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 10)
+# factors <- as.data.frame(matrix(data = c(rep(1, 12), rep(2, 13), rep(3, 25)), dimnames =list(rownames(BeckLee_mat50))), ncol = 1)
+# customised_series <- cust.series(BeckLee_mat50, factors)
+# bootstrapped_data <- boot.matrix(customised_series, bootstraps = 10)
[…]

### 2017-06-19 R/plot.dispRity.R

@@ -3,16 +3,16 @@
 #' @description Plots a \code{dispRity} object.
 #'
 #' @param data A \code{dispRity} object.
-#' @param type Either \code{"continuous"} (\code{"c"}), \code{"box"} (\code{"b"}), \code{"line"} (\code{"l"}) or \code{"polygon"} (\code{"p"}). When unspecified, is set to \code{"continuous"} if \code{\link{time.subsamples}} is used with \code{method = "continuous"}, else is set to \code{"box"}. See details.
+#' @param type Either \code{"continuous"} (\code{"c"}), \code{"box"} (\code{"b"}), \code{"line"} (\code{"l"}) or \code{"polygon"} (\code{"p"}). When unspecified, is set to \code{"continuous"} if \code{\link{time.series}} is used with \code{method = "continuous"}, else is set to \code{"box"}. See details.
 #' @param quantiles The quantiles to display (default is \code{quantiles = c(50, 95)}; is ignored if the \code{dispRity} object is not bootstrapped).
 #' @param cent.tend A function for summarising the bootstrapped disparity values (default is \code{\link[stats]{median}}).
 #' @param rarefaction Either \code{NULL} (default) or \code{FALSE} for not using the rarefaction scores; an \code{numeric} value of the level of rarefaction to plot; or \code{TRUE} for plotting the rarefaction curves.
-#' @param elements \code{logical} whether to plot the number of elements per subsamples.
+#' @param elements \code{logical} whether to plot the number of elements per series.
 #' @param ylim Optional, two \code{numeric} values for the range of the y axis.
 #' @param xlab Optional, a \code{character} string for the caption of the x axis.
 #' @param ylab Optional, one or two (if \code{elements = TRUE}) \code{character} string(s) for the caption of the y axis.
 #' @param col Optional, some \code{character} string(s) for the colour of the graph.
-#' @param time.subsamples \code{logical} whether to handle continuous data from the \code{time.subsamples} function as time (in Ma). When this option is set to TRUE for other \code{type} options, the names of the subsamples are used for the x axis labels.
+#' @param time.series \code{logical} whether to handle continuous data from the \code{time.series} function as time (in Ma). When this option is set to TRUE for other \code{type} options, the names of the series are used for the x axis labels.
[…]

### 2017-06-19 R/sanitizing.R

@@ -83,6 +83,36 @@ check.method <- function(argument, all_arguments, msg) {
     }
 }
 
+
+## Cleaning a tree so that the species match with the ones in a table
+clean.tree<-function(tree, table, verbose=FALSE) {
+    missing.species <- comparative.data(tree, data.frame("species"=row.names(table), "dummy"=rnorm(nrow(table)), "dumb"=rnorm(nrow(table))), "species")$dropped
+    if(length(missing.species$tips) != 0) {
+        tree.tmp <- drop.tip(tree, missing.species$tips)
+        if (verbose==TRUE) {
+            cat("Dropped tips:\n")
+            cat(missing.species$tips, sep=", ")
+        }
+        tree <- tree.tmp
+    }
+
+    return(tree)
+}
+
+## Cleaning a table so that the species match with the ones in the tree
+clean.table <- function(table, tree, verbose = FALSE) {
+    missing.species <- comparative.data(tree, data.frame("species" = row.names(table), "dummy" = rnorm(nrow(table)), "dumb" = rnorm(nrow(table))), "species")$dropped
+    if(length(missing.species$unmatched.rows) != 0) {
+        table.tmp <- table[-c(match(missing.species$unmatched.rows, rownames(table))),]
+        if (verbose == TRUE) {
+            cat("Dropped rows:\n")
+            cat(missing.species$unmatched.rows, sep = ", ")
+        }
+        table <- table.tmp
+    }
+    return(table)
+}
+
 ## Transforming a tree to binary with no 0 branch length.
 bin.tree <- function(tree){
     if(!is.binary.tree(tree)) {

### 2017-06-19 R/sequential.test.R

@@ -3,9 +3,9 @@
 #' @title Sequential linear regressions
 #'
 #' @description Function still in development.
-# ' @description Performs a sequential \code{\link[stats]{glm}} on the subsamples by correcting for time autocorrelation. 
+# ' @description Performs a sequential \code{\link[stats]{glm}} on the series by correcting for time autocorrelation. 
 #'
-# ' @param subsamples time subsamples of which to estimate the slopes sequentially.
+# ' @param series time series of which to estimate the slopes sequentially.
 # ' @param family the family of the \code{\link[stats]{glm}}.
 # ' @param correction optional, which p-value correction to apply (see \code{\link[stats]{p.adjust}}). If missing, no correction is applied.
 # ' @param call optional, a call from a \code{dispRity} object.
@@ -17,31 +17,31 @@
 # ' @examples
 # ' ## Load the Beck & Lee 2014 data
 # ' data(BeckLee_mat50)
-# ' ## Calculating the disparity from a customised subsamples
-# ' ## Generating the subsamples
-# ' groups <- as.data.frame(matrix(data = c(rep(1, 12), rep(2, 13), rep(3, 12),
+# ' ## Calculating the disparity from a customised series
+# ' ## Generating the series
+# ' factors <- as.data.frame(matrix(data = c(rep(1, 12), rep(2, 13), rep(3, 12),
 # '      rep(4, 13)), dimnames = list(rownames(BeckLee_mat50))), ncol = 1)
-# ' customised_subsamples <- cust.subsamples(BeckLee_mat50, groups)
+# ' customised_series <- cust.series(BeckLee_mat50, factors)
 # ' ## Bootstrapping the data
-# ' bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 100)
+# ' bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100)
 # ' ## Calculating variances of each dimension
 # ' dim_variances <- dispRity(bootstrapped_data, metric = variances)
-# ' ## Extracting the disparity values of each subsamples
-# ' subsamples <- extract.dispRity(dim_variances, observed = FALSE,
+# ' ## Extracting the disparity values of each series
+# ' series <- extract.dispRity(dim_variances, observed = FALSE,
 # '      keep.structure = TRUE, concatenate = TRUE)
 # '
-# ' ## Running a gaussian sequential test on the subsamples
-# ' results <- sequential.test(subsamples, family = gaussian)
[…]

### 2017-06-19 R/sim.morpho.R

@@ -30,29 +30,28 @@
 #' }
 #' 
 #' @examples
-#' 
-#' set.seed(4)
-#' ## A random tree with 15 tips
-#' tree <- rcoal(15)
-#' ## Setting up the parameters
+#' ## A random tree with 10 tips
+#' tree <- rcoal(10)
+#' ## setting up the parameters
+#' ## A gamma rate distribution with of alpha = 0.5
 #' my_rates = c(rgamma, rate = 10, shape = 5)
+#' ## A fixed substitution rate of 2 (T/T ratio in HKY)
 #' my_substitutions = c(runif, 2, 2)
-#' 
-#' ## HKY binary (15*50)
+#'
+# \dontrun{
+#' ## HKY binary (10*50)
 #' matrixHKY <- sim.morpho(tree, characters = 50, model = "HKY",
 #'      rates = my_rates, substitution = my_substitutions)
-#' 
-#' ## Mk matrix (15*50) (for Mkv models)
+#' ## Mk matrix (10*50) (for Mkv models)
 #' matrixMk <- sim.morpho(tree, characters = 50, model = "ER", rates = my_rates) 
-#' 
-#' ## Mk invariant matrix (15*50) (for Mk models)
+#' ## Mk invariant matrix (10*50) (for Mk models)
 #' matrixMk <- sim.morpho(tree, characters = 50, model = "ER", rates = my_rates,
 #'      invariant = FALSE)
-#'
-#' ## MIXED model invariant matrix (15*50)
+#' ## MIXED model invariant matrix (10*50)
 #' matrixMk <- sim.morpho(tree, characters = 50, model = "MIXED",
 #'      rates = my_rates, substitution = my_substitutions,  invariant = FALSE,
 #'      verbose = TRUE)
+# }
 #' 
 #' @seealso \code{\link{check.morpho}}, \code{\link[ape]{rTraitDisc}}, \code{\link[phyclust]{gen.seq.HKY}}
 #' 

### 2017-06-19 R/summary.dispRity_fun.R

@@ -4,30 +4,30 @@ CI.converter <- function(CI) {
 }
 
 # Wrapping function for summarising a single rarefaction
-get.summary <- function(disparity_subsamples_rare, cent.tend, quantiles) {
+get.summary <- function(disparity_series_rare, cent.tend, quantiles) {
     output <- list()
-    # output$cent_tend <- apply(disparity_subsamples_rare, 1, cent.tend)
+    # output$cent_tend <- apply(disparity_series_rare, 1, cent.tend)
     if(!missing(cent.tend)) {
-        output$cent_tend <- cent.tend(as.vector(disparity_subsamples_rare))
+        output$cent_tend <- cent.tend(as.vector(disparity_series_rare))
     }
-    # output$quantiles <- apply(disparity_subsamples_rare, 1, quantile, probs = CI.converter(quantiles))
+    # output$quantiles <- apply(disparity_series_rare, 1, quantile, probs = CI.converter(quantiles))
     if(!missing(quantiles)) {
-        output$quantiles <- quantile(as.vector(disparity_subsamples_rare), probs = CI.converter(quantiles))
+        output$quantiles <- quantile(as.vector(disparity_series_rare), probs = CI.converter(quantiles))
     }
     return(output)
 }
 
-## Lapply wrapping function for summarising a single subsamples
-lapply.summary <- function(disparity_subsamples, cent.tend, quantiles) {
-    return(lapply(disparity_subsamples[-1], get.summary, cent.tend, quantiles))
+## Lapply wrapping function for summarising a single series
+lapply.summary <- function(disparity_series, cent.tend, quantiles) {
+    return(lapply(disparity_series[-1], get.summary, cent.tend, quantiles))
 }
 
 ## Lapply wrapper for getting elements
-lapply.get.elements <- function(subsamples, bootstrapped = TRUE) {
+lapply.get.elements <- function(series, bootstrapped = TRUE) {
     if(bootstrapped){
-        return(unlist(lapply(subsamples[-1], nrow)))
+        return(unlist(lapply(series[-1], nrow)))
     } else {
-        return(unlist(lapply(subsamples, nrow)))
+        return(unlist(lapply(series, nrow)))
     }
 }
 
@@ -62,7 +62,7 @@ rounding.fun <- function(results_table, rounding, seq.test = FALSE) {
 
     if(rounding != "default") {
         for(column in start_column:ncol(results_table)) {
-            if(class(results_table[,column]) != "group") {
+            if(class(results_table[,column]) != "factor") {
[…]

### 2017-06-19 R/time.series_fun.R

@@ -29,8 +29,8 @@ adjust.FADLAD <- function(FADLAD, tree, data) {
 }
 
 
-## Discrete time subsamples
-time.subsamples.discrete <- function(data, tree, time, FADLAD, inc.nodes) {
+## Discrete time series
+time.series.discrete <- function(data, tree, time, FADLAD, inc.nodes) {
     
     ## lapply fun for getting the interval
     get.interval <- function(interval, time, ages_tree, inc.nodes) {
@@ -63,8 +63,8 @@ time.subsamples.discrete <- function(data, tree, time, FADLAD, inc.nodes) {
     return(interval_elements)
 }
 
-## Continuous time subsamples
-time.subsamples.continuous <- function(data, tree, time, model, FADLAD, verbose) {
+## Continuous time series
+time.series.continuous <- function(data, tree, time, model, FADLAD, verbose) {
     ## lapply fun gor getting the slices
     get.slice <- function(slice, time, model, ages_tree, data, verbose) {
         ## Get the subtree
@@ -115,9 +115,9 @@ time.subsamples.continuous <- function(data, tree, time, model, FADLAD, verbose)
     return(slices_elements)
 }
 
-## Making the origin subsamples for a disparity_object
-make.origin.subsamples <- function(data) {
+## Making the origin series for a disparity_object
+make.origin.series <- function(data) {
     origin <- list("elements" = as.matrix(seq(1:nrow(data))))
-    origin_subsamples <- list("origin" = origin)
-    return(origin_subsamples)
+    origin_series <- list("origin" = origin)
+    return(origin_series)
 }

### 2017-06-19 tests/testthat/make.data/make.data.R

@@ -4,98 +4,98 @@ source("MorphDistMatrix.verbose.R")
 source("anc.state.R")
 source("anc.state_fun.R")
 source("anc.unc.R")
-## matrix
-Nexus_data <- ReadMorphNexus("2014-Beck-ProcB-matrix-morpho.nex")
-Nexus_matrix <- Nexus_data$matrix
-## tree
-tree <- read.nexus("2014-Beck-ProcB-TEM.tre")
+#matrix
+Nexus_data<-ReadMorphNexus("2014-Beck-ProcB-matrix-morpho.nex")
+Nexus_matrix<-Nexus_data$matrix
+#tree
+tree<-read.nexus("2014-Beck-ProcB-TEM.tre")
 
-#######################################
-## Cleaning the matrices and the trees
-#######################################
+######################################
+#Cleaning the matrices and the trees
+######################################
 
-## Removing some species
-tree_tmp <- extract.clade(tree, 133)
-tree_tmp <- drop.tip(tree_tmp, extract.clade(tree_tmp, 127)$tip.label)
-tree_data <- drop.tip(tree_tmp, c("Erinaceus", "Ptilocercus", "Orycteropus", "Microgale"))
+#Removing some species
+tree_tmp<-extract.clade(tree, 133)
+tree_tmp<-drop.tip(tree_tmp, extract.clade(tree_tmp, 127)$tip.label)
+tree_data<-drop.tip(tree_tmp, c("Erinaceus", "Ptilocercus", "Orycteropus", "Microgale"))
 
-## Cleaning the tree and the table
-tree_data <- clean.tree(tree_data, Nexus_matrix)
-table <- clean.table(Nexus_matrix, tree_data)
-Nexus_data$matrix <- table
-tree_data <- lapply.root(tree_data)
+#Cleaning the tree and the table
+tree_data<-clean.tree(tree_data, Nexus_matrix)
+table<-clean.table(Nexus_matrix, tree_data)
+Nexus_data$matrix<-table
+tree_data<-lapply.root(tree_data)
 
-## Removing some characters
-Nexus_data$matrix <- table
-to_remove <- which(apply(Nexus_data$matrix, 2, function(X) length(levels(as.factor(X)))) < 2)
-to_remove <- c(to_remove, which(apply(Nexus_data$matrix, 2, function(X) length(which(is.na(X)))) > 25))
-matrix_tmp <- Nexus_data$matrix[,-c(to_remove)]
[…]

### 2017-06-19 tests/testthat/test-dispRity.R

@@ -5,11 +5,11 @@ context("dispRity")
 data(BeckLee_mat50)
 data(BeckLee_tree)
 data_boot <- boot.matrix(BeckLee_mat50, bootstraps = 11, rarefaction = c(5,6))
-data_subsamples_simple <- time.subsamples(BeckLee_mat50, tree = BeckLee_tree,  method = "discrete", time = c(120,80,40,20))
-data_subsamples_boot <- boot.matrix(data_subsamples_simple, bootstraps = 11, rarefaction = c(5,6))
+data_series_simple <- time.series(BeckLee_mat50, tree = BeckLee_tree,  method = "discrete", time = c(120,80,40,20))
+data_series_boot <- boot.matrix(data_series_simple, bootstraps = 11, rarefaction = c(5,6))
 metric = c(sum, variances)
 verbose = TRUE
-data <- data_subsamples_boot
+data <- data_series_boot
 
 test_that("disparity.bootstraps internal works", {
     data(BeckLee_mat50)
@@ -17,7 +17,7 @@ test_that("disparity.bootstraps internal works", {
     metrics_list <- list("level3.fun" = var, "level2.fun" = NULL, "level1.fun" = NULL)
     matrix_decomposition = TRUE
     ## With matrix decomposition
-    test0 <- disparity.bootstraps.silent(data$subsamples[[1]][[2]], metrics_list, data, matrix_decomposition)
+    test0 <- disparity.bootstraps(data$series[[1]][[2]], metrics_list, data, matrix_decomposition)
 
     ## Should be a array
     expect_is(
@@ -36,7 +36,7 @@ test_that("disparity.bootstraps internal works", {
     metrics_list <- list("level3.fun" = NULL, "level2.fun" = variances, "level1.fun" = NULL)
     matrix_decomposition = TRUE
     ## With matrix decomposition
-    test1 <- disparity.bootstraps.silent(data$subsamples[[1]][[2]], metrics_list, data, matrix_decomposition)
+    test1 <- disparity.bootstraps(data$series[[1]][[2]], metrics_list, data, matrix_decomposition)
 
     ## Should be a matrix
     expect_is(
@@ -55,7 +55,7 @@ test_that("disparity.bootstraps internal works", {
     ## Without matrix decomposition
     matrix_decomposition = FALSE
[…]

### 2017-06-19 doc/dispRity-ecology-demo.Rmd
v0.3.9 pre-release for 0.4

@@
 
-The first step in such analysis will be to create different series that will represent sub-samples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
-Each of these series contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
+The first step in such analysis will be to create different subsamples that will represent sub-samples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
+Each of these subsamples contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
 In our example, we are going to group the elements according to their depth and treatment.
@@
 
-Second, let’s split the data according to these factors to create the series of the ordinated space by using the `cust.series` function:
+Second, let’s split the data according to these factors to create the subsamples of the ordinated space by using the `custom.subsamples` function:
 
@@
 ## Splitting the ordinated space into four subsamples
-(customised_series <- cust.series(ord_matrix, factors))
+(customised_subsamples <- custom.subsamples(ord_matrix, factors))
 ```
@@
 ## Calculating disparity as the median distance between each elements and their group centroid
-(disparity <- dispRity(customised_series, metric = c(median, centroids)))
+(disparity <- dispRity(customised_subsamples, metric = c(median, centroids)))
 ```
 
-Similarly than with `cust.series`, `dispRity` displays a `dispRity` object.
+Similarly than with `custom.subsamples`, `dispRity` displays a `dispRity` object.
 To actually look at the calculated values, we can summarise the data by simply using `summary`:
@@
 We can also want to bootstrap the data to test the robustness of the measured disparity to outliers.
-Also, as we can see, each series has different numbers of elements.
-It might be interesting to rarefy the data as well to have only series with the same number of elements.
+Also, as we can see, each subsamples has different numbers of elements.
+It might be interesting to rarefy the data as well to have only subsamples with the same number of elements.
 Both steps are easily doable through the `boot.matrix` function.
@@
 ```{r}
-## Bootstrapping the customised series 100 times
-(bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100))
+## Bootstrapping the customised subsamples 100 times
+(bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 100))
 
-## Bootstrapping the series and using rarefaction (i.e. only re-sampling 17 elements each time)
-(rarefied_data <- boot.matrix(customised_series, bootstraps = 100, rarefaction = 17))
+## Bootstrapping the subsamples and using rarefaction (i.e. only re-sampling 17 elements each time)
+(rarefied_data <- boot.matrix(customised_subsamples, bootstraps = 100, rarefaction = 17))
 ```
@@
 
-As we can see, there seems to be no strong effect of the number of experimental plots in each series (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
+As we can see, there seems to be no strong effect of the number of experimental plots in each subsamples (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
 
@@
 ## Post-hoc testing of the effect of the two different factors
-## Note that the comparison contains the list of the pairs of series to compare
+## Note that the comparison contains the list of the pairs of subsamples to compare
 test.dispRity(disparity_BS, test = wilcox.test, comparisons = list(c(1,2), c(3,4)))

### 2017-06-19 R/boot.matrix_fun.R

@@ -1,9 +1,10 @@
 ## Full bootstrap replacement 
 boot.full <- function(elements, rarefaction) {
-    return( sample(elements, rarefaction, replace = TRUE) )
+    return(sample(elements, rarefaction, replace = TRUE))
 }
 
-## Single bootstrap: for each bootstrap, select one row and replace it by a randomly chosen left one (for n rows, only one row can be present two times).
+## Single bootstrap: for each bootstrap, select one row and replace it by a 
+## randomly chosen remaining row (for n rows, only one row can be present twice).
 boot.single <- function(elements, rarefaction) {
     ## Rarefy the data
     rarefied_sample <- sample(elements, rarefaction, replace = FALSE)
@@ -14,23 +15,34 @@ boot.single <- function(elements, rarefaction) {
     return(rarefied_sample)
 }
 
-## Performs bootstrap on one series and all rarefaction levels
-replicate.bootstraps <- function(rarefaction, bootstraps, series, boot.type.fun, verbose) {
-    if(verbose) message(".", appendLF = FALSE)
-    return(replicate(bootstraps, boot.type.fun(series$elements, rarefaction)))
+## Performs bootstrap on one subsamples and all rarefaction levels
+replicate.bootstraps.verbose <- function(rarefaction, bootstraps, subsamples, boot.type.fun, verbose) {
+    message(".", appendLF = FALSE)
+    return(replicate(bootstraps, boot.type.fun(subsamples$elements, rarefaction)))
+}
+replicate.bootstraps.silent <- function(rarefaction, bootstraps, subsamples, boot.type.fun) {
+    return(replicate(bootstraps, boot.type.fun(subsamples$elements, rarefaction)))
 }
 
-## Performs bootstrap on multiple series and all rarefaction levels
-bootstrap.wrapper <- function(series, bootstraps, rarefaction, boot.type.fun, verbose) {
-    return(lapply(select.rarefaction(series, rarefaction), replicate.bootstraps, bootstraps, series, boot.type.fun, verbose))
+## Performs bootstrap on multiple subsamples and all rarefaction levels
+bootstrap.wrapper <- function(subsamples, bootstraps, rarefaction, boot.type.fun, verbose) {
+
+    ## Verbose?
+    if(verbose == TRUE){
+        replicate.bootstraps <- replicate.bootstraps.verbose
+    } else {
+        replicate.bootstraps <- replicate.bootstraps.silent
+    }
+
[…]

### 2017-06-19 R/clean.data_fun.R

@@ -1,39 +1,31 @@
-#Cleaning a tree so that the species match with the ones in a table
+## Cleaning a tree so that the species match with the ones in a table
 clean.tree.table <- function(tree, data) {
 
-    #create a dummy data
-    dummy_data <- as.data.frame(data)
+    ## Intersecting names between both data sets
+    matching_names <- intersect(tree$tip.label, rownames(data))
 
-    #Adding a row called taxa_names_column containing the taxa names
-    dummy_data[,ncol(data)+1] <- rownames(data)
-    names(dummy_data)[ncol(data)+1] <- "taxa_names_column"
+    ## Which data is present
+    data_match <- rownames(data) %in% matching_names
 
-    #run caper::comparative.data to check the non matching columns/rows
-    missing <- caper::comparative.data(tree, dummy_data, "taxa_names_column", vcv = FALSE, vcv.dim = 2, na.omit = TRUE, force.root = FALSE, warn.dropped = FALSE, scope = NULL)$dropped
+    ## Which tips are present
+    tips_match <- tree$tip.label %in% matching_names
 
-    #Dropping tips (if necessary)
-    if(length(missing$tips) != 0) {
-        #drop the missing tips
-        tree_tmp <- drop.tip(tree, missing$tips)
-        #save the missing tips names
-        dropped_tips <- missing$tips
-    } else {
-        #No drop needed!
-        tree_tmp <- tree
-        dropped_tips <- NA
-    }
-
-    #Dropping rows (if necessary)
-    if(length(missing$unmatched.rows) != 0) {
-        #Drop the unmatched rows
-        data_tmp <- data[-match(missing$unmatched.rows, dummy_data[,ncol(data)+1]),]
-        #save the dropped rows names
-        dropped_rows <- missing$unmatched.rows
-    } else {
-        #No drop needed!
-        data_tmp <- data
+    ## Matching the data    
+    if(all(data_match)) {
         dropped_rows <- NA
+    } else {
+        rows_numbers <- which(!data_match)
+        dropped_rows <- rownames(data)[rows_numbers]
+        data <- data[-c(rows_numbers),]
     }
 
-    return(list("tree" = tree_tmp, "data" = data_tmp, "dropped_tips" = dropped_tips, "dropped_rows" = dropped_rows))
+    ## Matching the tree
+    if(all(tips_match)) {
+        dropped_tips <- NA
+    } else {
[…]

### 2017-06-19 R/cust.series.R

@@ -1,78 +1,0 @@
-#' @title Separating ordinated data in custom series.
-#'
-#' @description Splits the ordinated data into a customized series list.
-#'
-#' @param data An ordinated matrix of maximal dimensions \eqn{k*(k-1)}.
-#' @param factor A \code{data.frame} with the same \eqn{k} elements as in \code{data} as rownames.
-#'
-#' @return
-#' This function outputs a \code{dispRity} object containing:
-#' \item{data}{A \code{list} of the split ordinated data (each element is a \code{matrix}).}
-#' \item{elements}{A \code{vector} containing all the rownames from the input matrix.}
-#' \item{series}{A \code{vector} containing the name of the series.}
-#' \code{dispRity} objects can be summarised using \code{print} (S3).
-#' 
-#' @details
-#' The customized series can typically be a factor. For a finite number of taxonomic groups, traits, etc.
-#'
-#' @examples
-#' ## Generating a dummy ordinated matrix
-#' ordinated_matrix <- matrix(data = rnorm(90), nrow = 10, ncol = 9,
-#'      dimnames = list(letters[1:10]))
-#' ## Creating a list of dummy factors (1 or 2)
-#' factors <- as.data.frame(matrix(data = c(rep(1,5), rep(2,5)), nrow = 10,
-#'      ncol = 1, dimnames = list(letters[1:10])))
-#' ## Splitting the dummy ordinated matrix
-#' cust.series(ordinated_matrix, factors)
-#'
-#' @seealso \code{\link{time.series}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}.
-#'
-#' @author Thomas Guillerme
-
-## DEBUG
-# warning("DEBUG cust.series")
-# source("sanitizing.R")
-# source("cust.series_fun.R")
-# source("time.series_fun.R")
-# data <- matrix(data = rnorm(90), nrow = 10, ncol = 9, dimnames = list(letters[1:10]))
-# factor <- as.data.frame(matrix(data = c(rep(1,5), rep(2,5)), nrow = 10, ncol = 1, dimnames = list(letters[1:10])))
-
-cust.series <- function(data, factor) {
-    ## ----------------------
-    ##  SANITIZING
-    ## ----------------------
-    ## DATA
-    ## data must be a matrix
-    check.class(data, 'matrix')
-    ## data must be of size k*<=k-1
[…]

### 2017-06-19 R/custom.subsamples.R

@@ -1,0 +1,128 @@
+#' @title Separating ordinated data in custom subsamples.
+#' @aliases cust.series custom.series cust.subsamples
+#'
+#' @description Splits the ordinated data into a customized subsamples list.
+#'
+#' @param data An ordinated matrix of maximal dimensions \eqn{k*(k-1)}.
+#' @param group Either a \code{list} of row numbers or names to be used as different groups or a \code{data.frame} with the same \eqn{k} elements as in \code{data} as rownames.
+#'
+#' @return
+#' This function outputs a \code{dispRity} object containing:
+#' \item{data}{A \code{list} of the split ordinated data (each element is a \code{matrix}).}
+#' \item{elements}{A \code{vector} containing all the rownames from the input matrix.}
+#' \item{subsamples}{A \code{vector} containing the name of the subsamples.}
+#' \code{dispRity} objects can be summarised using \code{print} (S3).
+#' 
+#' @details
+#' Note that every elements from the input data can be assigned to multiple groups!
+#'
+#' @examples
+#' ## Generating a dummy ordinated matrix
+#' ordinated_matrix <- matrix(data = rnorm(90), nrow = 10)
+#' 
+#' ## Splitting the ordinated matrix into two groups using row numbers
+#' custom.subsamples(ordinated_matrix, list(c(1:4), c(5:10)))
+#' 
+#' ## Splitting the ordinated matrix into three groups using row names
+#' ordinated_matrix <- matrix(data = rnorm(90), nrow = 10,
+#'      dimnames = list(letters[1:10]))
+#' custom.subsamples(ordinated_matrix,
+#'      list("A" = c("a", "b", "c", "d"), "B" = c("e", "f", "g", "h", "i", "j"),
+#'           "C" = c("a", "c", "d", "f", "h")))
+#' 
+#' ## Splitting the ordinated matrix into four groups using a data frame
+#' groups <- as.data.frame(matrix(data = c(rep(1,5), rep(2,5), rep(c(1,2), 5)),
+#'      nrow = 10, ncol = 2, dimnames = list(letters[1:10], c("g1", "g2"))))
+#' custom.subsamples(ordinated_matrix, groups)
+#'
+#' @seealso \code{\link{time.subsamples}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}.
+#'
+#' @author Thomas Guillerme
+
+## DEBUG
+# warning("DEBUG cust.subsamples")
[…]

### 2017-06-19 R/summary.dispRity.R

@@ -11,8 +11,8 @@
 #'
 #' @return
 #' A \code{data.frame} with:
-#' \item{series}{the series names.}
-#' \item{n}{the number of elements per series.}
+#' \item{subsamples}{the subsamples names.}
+#' \item{n}{the number of elements per subsamples.}
 #' \item{observed}{the observed disparity or the the observed central tendency (<cent_tend>) of disparity (\code{obs.<cent_tend>}).}
 #' \item{bootstraps...}{if \code{data} is bootstrapped, the bootstrapped disparity's central tendency (\code{bs.<cent_tend>}) and the quantiles of the bootstrapped disparity's (or, if \code{data} is not bootstrapped but disparity is calculated as a distribution - see \code{\link[dispRity]{dispRity}}) - the quantiles of the observed disparity is displayed).}
 #' 
@@ -35,11 +35,11 @@
 # source("make.metric.R")
 # source("summary.dispRity_fun.R")
 # data(BeckLee_mat50)
-# factors <- as.data.frame(matrix(data = c(rep(1, 12), rep(2, 13), rep(3, 12), rep(4, 13)), dimnames = list(rownames(BeckLee_mat50))), ncol = 1)
-# customised_series <- cust.series(BeckLee_mat50, factors)
-# bootstrapped_data <- boot.matrix(customised_series, bootstraps = 3, rarefaction = TRUE)
-# series <- extract.dispRity(sum_of_variances, observed = FALSE, keep.structure = TRUE, concatenate = TRUE)
-# data <- sequential.test(series, family = gaussian, correction = "hommel")
+# groups <- as.data.frame(matrix(data = c(rep(1, 12), rep(2, 13), rep(3, 12), rep(4, 13)), dimnames = list(rownames(BeckLee_mat50))), ncol = 1)
+# customised_subsamples <- cust.subsamples(BeckLee_mat50, groups)
+# bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 3, rarefaction = TRUE)
+# subsamples <- extract.dispRity(sum_of_variances, observed = FALSE, keep.structure = TRUE, concatenate = TRUE)
+# data <- sequential.test(subsamples, family = gaussian, correction = "hommel")
 
 # data <- dispRity(bootstrapped_data, metric = variances)
 
@@ -103,16 +103,16 @@ summary.dispRity <- function(data, quantiles = c(50,95), cent.tend = median, rec
[…]

### 2017-06-19 R/utilities_old.R

@@ -318,94 +318,6 @@
 # }
 
 # ##########################
-# #extract.disp
-# ##########################
-# #extract a series of disparity measurement using a number of taxa (can be max or min) 
-# #----
-# #SYNTAX :
-# #<disp.data> a disparity data.frame with a "time" and a "rarefaction" column name
-# #<rarefaction> which rarefaction value to extract
-# #<plot.format> removes the time column to be in a proper plotting format
-# #----
-# extract.disp<-function(disp.data, rarefaction, plot.format=TRUE) {
-#     #SANITIZING
-#     #disparity
-#     check.class(disp.data, "data.frame")
-#     if(any(is.na(match(c("time", "rarefaction"), colnames(disp.data))))) {
-#         stop("disp.data must have at least one column called 'time' and one called 'rarefaction'.")
-#     }
-
-#     #rarefaction
-#     if(class(rarefaction) != 'numeric') {
-#         #check.class(rarefaction, 'character')
-#         if(rarefaction == "min") {
-#             rar.val<-min(table(disp.data$time))+1
-#             is.fun<-FALSE
-#         } else {
-#             if(rarefaction == "max") {
-#                 is.fun<-TRUE
-#             }
-#         }
-#     } else {
-#         is.fun<-FALSE
-#         rar.val<-rarefaction
-#     }
-
-#     #plot.format
-#     #check.class(plot.format, 'logical')
-
-#     #EXTRACTING THE RIGHT RAREFACTION VALUE
-
-#     #Set the first row
-#     sub_samp<-disp.data[which(disp.data$time == levels(disp.data$time)[1]),]
-#     #Extract the rarefaction level
-#     if(is.fun == TRUE) {
-#         disp.data.sort<-sub_samp[which(sub_samp$rarefaction == max(sub_samp$rarefaction)),]
-#     } else {
-#         #Check if rarefaction level exists
-#         if(length(which(sub_samp$rarefaction == rar.val)) == 1) {
-#             #Extract the value
-#             disp.data.sort<-sub_samp[which(sub_samp$rarefaction == rar.val),]
-#         } else {
-#             #If the rarefaction level doesn't exists, extract the max or min
-#             if(all(rar.val > sub_samp$rarefaction)) {
-#                 disp.data.sort<-sub_samp[which(sub_samp$rarefaction == max(sub_samp$rarefaction)),]
-#             } else {
[…]

### 2017-06-19 tests/testthat/test-char.diff.R

@@ -1,0 +1,75 @@
+context("char.diff")
+
+test_that("char.diff pair", {
+    A <- c(1,0,0,0,0)
+    B <- c(0,1,1,1,1)
+
+    #Difference is 0
+    expect_equal(char.diff(list(A,B)), 0)
+    #Difference is triangular
+    expect_equal(char.diff(list(A,B)), char.diff(list(A,B)))
+
+    C <- c(1,1,0,0,0)
+
+    #Difference is 0.4
+    expect_equal(char.diff(list(A,C)), 0.4)
+    #Difference is triangular
+    expect_equal(char.diff(list(A,C)), char.diff(list(C,A)))
+
+    D <- c(0,1,1,0,0)
+
+    #Difference is 0.4
+    expect_equal(char.diff(list(A,D)), 0.8)
+    #Difference is triangular
+    expect_equal(char.diff(list(A,D)), char.diff(list(D,A)))
+
+    E <- c(1,0,0,1,1)
+
+    #Difference is equal to D
+    expect_equal(char.diff(list(D,E)), 0)
+    #Difference is triangular (with D)
+    expect_equal(char.diff(list(A,E)), char.diff(list(A,D)))
+})
+
+## Matrices                     #A,B,C,D,E
+matrix_binary <- matrix(data = c(1,0,1,0,1,
+                                 0,1,1,1,0,
+                                 0,1,0,1,0,
+                                 0,1,0,0,1,
+                                 0,1,0,0,1), ncol = 5, byrow = TRUE)
+colnames(matrix_binary) <- LETTERS[1:5]
+
+                               #A,B,C,D,E,F,G
+matrix_multi <- matrix(data = c(1,2,0,0,1,2,1,
+                                2,3,1,2,2,0,2,
+                                0,4,2,1,1,2,2,
+                                0,4,0,0,0,1,0,
+                                0,4,0,0,0,1,0), ncol = 7, byrow = TRUE)
+colnames(matrix_multi) <- LETTERS[1:7]
+
+                                #A,B,C,D
+matrix_simple <- matrix(data = c(1,NA,3,NA,
+                                 7,7,2,2,
[…]

### 2017-06-19 tests/testthat/test-summary.dispRity.R

@@ -69,16 +69,16 @@ test_that("Works with bootstraps and rarefaction", {
         , c(45.36, NA))
 })
 
-#Case 4, time series
+#Case 4, time subsamples
 data <- test_data$ord_data_tips
-factor <- as.data.frame(matrix(data = c(rep(1, nrow(data)/2),rep(2, nrow(data)/2)), nrow = nrow(data), ncol = 1))
-rownames(factor) <- rownames(data)
-data <- cust.series(data, factor)
+group <- as.data.frame(matrix(data = c(rep(1, nrow(data)/2),rep(2, nrow(data)/2)), nrow = nrow(data), ncol = 1))
+rownames(group) <- rownames(data)
+data <- custom.subsamples(data, group)
 data <- dispRity(data, metric = c(sum, ranges))
 test <- summary(data)
 
 #Test
-test_that("Works with series", {
+test_that("Works with subsamples", {
     expect_is(
         test, "data.frame"
         )
@@ -93,18 +93,18 @@ test_that("Works with series", {
         ,c(37.00, 37.97))
 })
 
-#Case 5, time series + bootstraps
+#Case 5, time subsamples + bootstraps
 set.seed(1)
 data <- test_data$ord_data_tips
-factor <- as.data.frame(matrix(data = c(rep(1, nrow(data)/2),rep(2, nrow(data)/2)), nrow = nrow(data), ncol = 1))
-rownames(factor) <- rownames(data)
-data <- cust.series(data, factor)
+group <- as.data.frame(matrix(data = c(rep(1, nrow(data)/2),rep(2, nrow(data)/2)), nrow = nrow(data), ncol = 1))
+rownames(group) <- rownames(data)
+data <- custom.subsamples(data, group)
 data <- boot.matrix(data, bootstrap = 5)
 data <- dispRity(data, metric = c(sum, ranges))
 test <- summary(data)
 
 #Test
-test_that("Works with series and bootstraps", {
+test_that("Works with subsamples and bootstraps", {
     expect_is(
         test, "data.frame"
         )
@@ -119,18 +119,18 @@ test_that("Works with series and bootstraps", {
         ,c(32.65, 34.09))
 })
 
-#Case 5, time series + bootstraps + rarefaction
+#Case 5, time subsamples + bootstraps + rarefaction
 set.seed(1)
 data <- test_data$ord_data_tips
[…]

### 2017-06-24 doc/dispRity-manual.Rmd
updates (tests) to manual

@@
 
-You can install this package easily if you use the latest version of `R` (> 3.0.0) and `devtools`.
+You can install this package easily if you use the latest version of `R` (> 3.3.0) and `devtools`.
 
@@
 ## Bootstrapping on the time binning/slicing subsamples
-boot.matrix(time_slices, bootstraps = 1000)
+boot.matrix(time_slices, bootstraps = 100)
 ```
@@
 ## Bootstrapping and rarefying these groups
-boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 200, rarefaction = TRUE)
+boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
 ## Calculate disparity
@@
 ## Bootstrapping on the time binning/slicing subsamples
-boot_time_slices <- boot.matrix(time_slices, bootstraps = 1000)
+boot_time_slices <- boot.matrix(time_slices, bootstraps = 100)
 ## Calculate disparity
 disparity_time_slices <- dispRity(boot_time_slices, metric = c(sum, variances))
+
+## Creating time bins subsamples
+time_bins <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
+    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
+## Bootstrapping on the time binning/slicing subsamples
+boot_time_bins <- boot.matrix(time_bins, bootstraps = 100)
+## Calculate disparity
+disparity_time_bins <- dispRity(boot_time_bins, metric = c(sum, variances))
 ```
@@
 ## Graphical options
-par(bty = "n")
+op <- par(bty = "n")
 
@@
 ## Adding the discrete data
-plot(disparity_crown_stem, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
+plot(disparity_time_bins, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
     add = TRUE)
@@
 
-<!-- Change this example to add time bins! -->
-
 Finally, if your data has been fully rarefied, it is also possible to easily look at rarefaction curves by using the `rarefaction = TRUE` argument:
@@
 
-Finally, the `dispRity` package allows to apply some test functions to the calculated disparity in order to test hypothesis.
+The `dispRity` package allows to apply some test functions to the calculated disparity in order to test hypothesis.
 The function `test.dispRity` works in a similar way as the `dispRity` function: it intakes a `dispRity` object, a `test` and a `comparisons` argument.
@@
 It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsamples names to create personalised test.
-Some other specific tests implemented in `dispRity` such as the `dispRity::null.test` and the `dispRity::sequential.test` have a specific way to be applied to the data and therefore ignore the `comparisons` argument.
+Some other specific tests implemented in `dispRity` such as the `dispRity::null.test` that has a specific way to be applied to the data and therefore ignore the `comparisons` argument. 
+<!-- Add sequential test one day! -->
 
@@
 
-```{r, eval=FALSE}
+```{r}
 ## Performing a t-test to the test a difference in disparity between crown and stem mammals
-test.dispRity(disp_crown_stemBS, test = t.test)
+test.dispRity(disparity_crown_stem, test = t.test)
 
 ## Performing the same test but with the detailed t.test output
-test.dispRity(disp_crown_stemBS, test = t.test, details = TRUE)
+test.dispRity(disparity_crown_stem, test = t.test, details = TRUE)
 
 ## Performing a Wilcoxon test applied to sliced disparity the data sequentially with correction
-test.dispRity(disp_time_slicesBS, test = wilcox.test, comparisons = "sequential",
+test.dispRity(disparity_time_slices, test = wilcox.test, comparisons = "sequential",
     correction = "bonferroni")
@@
 ## Bhattacharyya Coefficient function - see ?bhatt.coeff)
-test.dispRity(disp_time_binsBS, test = bhatt.coeff)
+test.dispRity(disparity_time_bins, test = bhatt.coeff)
 ```
@@
 
-```{r, eval=FALSE}
+```{r}
 ## Performing and ANOVA applied to the same data
-slice_aov <- (test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all"))
-
-## The output is a regular ANOVA output...
+slice_aov <- test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all")
+slice_aov
+
+## The output is a regular ANOVA output
 class(slice_aov)
@@
 ```
+Of course, following the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
+The only condition is that the test can be applied to at least two distributions.
+In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `aov` for example).
 
 ## Disparity as a distribution
+
+Disparity is often regarded as a summary value of the position of the all elements in the ordinated space.
+For example, the sum of variances, the product of ranges or the median distance between the elements and their centroid will summarise disparity as a single value.
+This value can be of course pseudo-replicated (boostraped) in order to obtain a more confident distribution of this summary metric.
+However, another way to perform disparity analysis is to use the whole distribution rather than their summary (e.g. the variances or the ranges).
+
+This is possible in the `dispRity` package by calculating disparity as a level 2 metric only!
+Let's have a look using our previous example of bootstrapped time slices but by measuring the distances between each taxa and their centroid as disparity
+
+```{r}
+## Measuring disparity as a whole distribution
+disparity_centroids <- dispRity(boot_time_slices, metric = centroids)
+```
+
+The resulting disparity object being a level 2, it can easily also be transformed into a level 1 one by for example measuring the median distance of all these distributions:
+
+```{r}
+## Measuring median disparity in each time slice
+disparity_centroids_median <- dispRity(disparity_centroids, metric = median)
+```
+
+And we can now compare the differences between both methods:
+
+```{r}
+## Summarising both disparity measurements:
+## The distributions:
+summary(disparity_centroids)
+## The summary of the distributions (as median)
+summary(disparity_centroids_median)
+```
+
+We can see that the summary message for the distribution is slightly different than before.
+Here `summary` also displays the observed central tendency (i.e. the central tendency of the measured distributions).
+Note that, as expected, it is the same in both metrics!
+
+<!-- STOPPED HERE -->
+
+```{r, fig.width=8, fig.height=8}
+## Graphical parameters
+op <- par(bty = "n", mfrow = c(1,2))
+
+## Plotting both disparity measurements
+plot(disparity_centroids, ylab = "Distribution of all the distances")
+plot(disparity_centroids_median, ylab = "Distribution of the medians of all the distances")
+
+par(op)
+```
+
+The resulted distributions can then be studied the same way as described above.
 
@@
 summary(sort(disp_time_slicesBS, decreasing = TRUE))
+```
 

### 2017-07-02 doc/dispRity-manual.Rmd
Updates to the manual (finished content)

@@
 ```
+
 `dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
@@
 ```{r}
+require(Claddis)
+
 ## Ordinating the example data from Claddis
@@
 
-> TBC!
+```{r, eval = TRUE, echo = FALSE, message = FALSE}
+## Checking if Claddis is installed BUILD ONLY
+if(!require(geomorph)) install.packages("geomorph")
+```
+
+The package can also easily take data from `geomorph` using the `geomorph.ordination` function.
+This function simply take procrustes data an performs an ordination:
+
+```{r}
+require(geomorph)
+
+## Loading the plethodon dataset
+data(plethodon)
+
+## Performing a procrustes transform on the landmarks
+procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)
+
+## Ordinating this data
+geomorph.ordination(procrustes)[1:5,1:5]
+```
+
+Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
+Additionally you can provide to the function a `geomorph.data.frame` object.
+If the latter contains sorting information (i.e. factors), they can be directly used to make a `dispRity` object!
+
+```{r}
+## Using a geomorph.data.frame
+geomorph_df <- geomorph.data.frame(procrustes,
+     species = plethodon$species, site = plethodon$site)
+
+## Ordinating this data and making a disparity object
+geomorph.ordination(geomorph_df)
+```
+
+We will see what are these `dispRity` objects in the sections below...
 
@@
 
+<!-- DATA LOADER -->
+
 ```{r}
@@
 time_bins <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
+    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages,
+    inc.nodes = TRUE)
 ## Bootstrapping on the time binning/slicing subsamples
@@
 ## Performing and ANOVA applied to the same data
-slice_aov <- test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all")
+slice_aov <- test.dispRity(disparity_time_slices, test = aov, comparisons = "all")
 slice_aov
@@
 Here `summary` also displays the observed central tendency (i.e. the central tendency of the measured distributions).
-Note that, as expected, it is the same in both metrics!
-
-<!-- STOPPED HERE -->
+Note that, as expected, this central tendency is the same in both metrics!
+
+Another, maybe more intuitive way, to compare both approaches for measuring disparity is to plot the distributions:
 
@@
 
+```{r}
+## Probability of overlap in the distribution of medians
+test.dispRity(disparity_centroids_median, test = bhatt.coeff)
+```
+
+In this case, we are looking at the probability of overlap of the distribution of median distances from centroids between each pair of time slices.
+In other words, we are measuring whether the medians from each bootstrap pseudo-replicate for each time slice overlap.
+But of course, we might be interested in the actual distribution of the distances from the centroid rather than simply their central tendencies.
+This can be problematic depending on the research question asked since we are effectively comparing non-independent medians distributions (because of the pseudo-replication).
+
+One solution is thus to look at the full distribution:
+
+```{r}
+## Probability of overlap for the full distributions
+test.dispRity(disparity_centroids, test = bhatt.coeff)
+```
+
+These results show the actual overlap between all the measured distances from centroids concatenated across all the bootstraps.
+For example, when comparing the slices 120 and 80, we are effectively comparing the 5 \times 100 distances (the distances of the 5 elements in slice 120 bootstrapped 100 times) to the 19 \times 100 distances from slice 80.
+However, this can also be problematic for some specific tests since the _n_ \times 100 distances are also pseudo-replicates and thus are still not independent.
+
+One second solution is to compare the distributions to each other _per replicates_:
+
+```{r}
+## Boostrapped probability of overlap for the full distributions
+test.dispRity(disparity_centroids, test = bhatt.coeff, concatenate = FALSE)
+```
+
+These results show the median overlap between pairs of distributions in the first column (`bhatt.coeff`) and then the distribution of these overlaps between each par of bootstraps.
+In other words, when two distributions are compared, they are now compared for each bootstrap pseudo-replicate, thus effectively creating a distribution of probabilities of overlap.
+For example, when comparing the slices 120 and 80, we have a mean probability of overlap of 0.28 and a probability between 0.18 and 0.43 in 50% of the pseudo-replicates.
+Note that the quantiles and central tendencies can be modified via the `conc.quantiles` option:
+
 ## Ecological study
 
+Disparity analysis are not restricted to palaeobiology analysis only.
+This package is designed for dealing with ordinated matrix, no matter which data was used in the ordination.
+Here is an short example of disparity analysis using ecological data from D. McClean (unpubl.).
+
+The dataset contains an ordinated matrix of 20 dimensions (columns) for 40 elements (rows).
+The elements are different experimental plots with discrete variations of nutrient enrichment and depth and freshwater benthic invertebrates.
+
+```{r}
+## Loading the data
+data(McClean_data)
+
+## Ordinated matrix
+ordinated_matrix <- McClean_data$ordination
+
+## Treatments and depths
+treatments <- McClean_data$treatment
+depth <- McClean_data$depth
+```
+
+### Classic analysis
+
+A classical way to represent this ordinated data will be to use two dimensional plots.
+
+```{r, fig.width=6, fig.height=6}
+## Setting the colors that will represent the treatments
+cols <- sub("a", "red", treatments)
+cols <- sub("b", "blue", cols)
+## Setting the symbols that will represent the depth
+pchs <- sub(1, 16, depth)
+pchs <- as.numeric(sub(2, 17, pchs))
+
+## Graphical option
+op <- par(bty = "n")
+## A classic 2D ordination plot
+plot(ordinated_matrix[, 1:2], col = cols, pch = pchs,
+     xlab = "PC 1", ylab = "PC 2",
+     xlim = range(ordinated_matrix[, 1:2]) + c(0, 100))
+par(op)
+```
+
+This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix (i.e. the first two dimensions).
+At a first glance, it seems difficult to see a clear effect from the treatments (blue or red) or the depth (rounds or triangles) since the different experimental plots with the same parameters don’t seem to cluster.
+However, the problem, is that this plot ignores the 18 other dimensions of the ordination!
+Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e.
+the values on the first dimension do not represent some continuous traits (e.g.
+depth) but just the ordinations of correlations between the data and some factors.
+
+Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.
+
+### Multidimensional analysis with `dispRity`
+
+#### Splitting the data following the different factors
+
+The first step in such analysis will be to create different subsamples that will represent sub-samples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
+Each of these subsamples contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
+In our example, we are going to group the elements according to their depth and treatment.
+
+```{r}
+## Creating the table that contain the elements and their attributes
+factors <- as.data.frame(matrix(data = c(treatments, depth),
+          nrow = nrow(ordinated_matrix), ncol = 2, byrow = FALSE,
+          dimnames = list(rownames(ordinated_matrix))))
+names(factors) <- c("Treat", "Depth")
+head(factors)
+```
+
+Second, let’s split the data according to these factors to create the subsamples of the ordinated space by using the `custom.subsamples` function:
+
+```{r}
+## Splitting the ordinated space into four subsamples
+(customised_subsamples <- custom.subsamples(ordinated_matrix, factors))
+```
+
+#### Calculating disparity
+
+In this example, we are going to define disparity as the n-dimensional ellipsoid volume (see `?hyper.volume`).
+
+```{r}
+## Calculating disparity as the ellipsoid volume of each group
+disparity <- dispRity(customised_subsamples, metric = diagonal)
+summary(disparity)
+```
+
+We can also want to bootstrap the data to test the robustness of the measured disparity to outliers.
+Also, as we can see, each subsamples has different numbers of elements.
+It might also be interesting to rarefy the data as well to have only subsamples with the same number of elements.
+
+```{r}
+## Bootstrapping the subsamples and using rarefaction
+## (i.e. only re-sampling 17 elements each time)
+bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 100,
+                                 rarefaction = 17)
+```
+
+We can now rerun a more robust disparity analysis using the bootstrapped data:
+
+```{r}
+## Calculating the bootstrapped disparity
+disparity <- dispRity(bootstrapped_data, metric = diagonal)
+```
+
+#### Summarising and displaying the data
+
+Now let's summarise this data:
+
+```{r, fig.width=12, fig.height=6, eval = FALSE}
+## Summary table of the data
+summary(disparity)
+
+## Graphical options
+op <- par(mfrow = (c(1,2)), bty = "n")
+
+## Plotting the bootstrapped disparity
+plot(disparity, main = "Bootstrapped", las = 2, xlab = "")
+
+## Plotting the rarefied disparity
+plot(disparity, rarefaction = 17, main = "Rarefied", las = 2, xlab = "")
+
+par(op)
+```
+
+As we can see, there seems to be no strong effect of the number of experimental plots in each subsamples (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
+
+#### Summarising hypothesis
+
+Finally, we can test our hypothesis (whether the sort of water treatment at certain depth alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.
+
+```{r}
+## Testing the effect of our factors on the bootstrapped data
+summary(test.dispRity(disparity, test = aov, comparisons = "all"))
+```
+
 ## Simulating discrete morphological data
+
+The `dispRity` package also allows to simulate discrete morphological data matrices.
 
@@
 
-```{r, eval=FALSE}
-library(dispRity)
-set.seed(123)
+```{r}
+set.seed(3)
 ## Setting the starting tree with 15 taxa as a random coalescent tree
@@
 
+Note that this example produces a tree with a great consistency index and an identical topology as the random coalescent tree!
+Nearly too good to be true...
+
 ### In details
@@
 
-```{r, eval=FALSE}
-set.seed(1)
+```{r}
+set.seed(0)
 ## tree
@@
 
-## Running `dispRity` in parallel
-
 ## Manipulating `dispRity` objects
@@
 Therefore `dispRity` uses a specific class of object called a `dispRity` object.
-These objects allow users to use S3 method functions such as `summary.dispRity` (see [summary](#summary) section), `plot.dispRity` (`plot`; see [plot](#plot) section) and `print.dispRity`. 
-<!-- # NC: I've simpiflied this as people who understand will know this already, people that don't won't care and don't need to know.
- -->
- `dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g.
-`sort.dispRity`, `extract.dispRity` see the full list in the [`dispRity` utilities](#dispRity.utilities) section).
+These objects allow users to use S3 method functions such as `summary.dispRity`, `plot.dispRity` and `print.dispRity`. 
+`dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g. `sort.dispRity`, `extract.dispRity` see the full list in the next section).
 These functions modify the `dispRity` object without having to delve into its complex structure!
@@
 
-<!-- # NC: Explain the point of these examples here... Are they needed? -->
-
-```{r, eval=FALSE}
+```{r}
+## Loading the example data
+data(disparity)
+
 ## What is the class of the median_centroids object?
-class(median_centroids)
+class(disparity)
 
 ## What does the object contain?
-names(median_centroids)
+names(disparity)
 
 ## Summarising it using the S3 method print.dispRity
-median_centroids
+disparity
 ```
@@
 ## Display the full object
-print(median_centroids, all = TRUE)
-## This is more than ~ 2500 lines on my 13 inch laptop screen!
-```
-
-<!-- 
-# Modular functions - 
-# NC: not sure any more of these modular functions are relevant in the vignette? Maybe just mention the metrics? Remember a vignette is for users, not for you to show how the programming works.
-
-# NC: I'd include a brief explanation of the include disparity metrics here.
-But yes keep the crazy "level" details for the other vignette on metrics.
-I think you want to avoid having too many separate vignettes really.
--->
+print(disparity, all = TRUE)
+## This is more nearly ~ 5000 lines on my 13 inch laptop screen!
+```
 
@@
 
-```{r, eval=FALSE}
+```{r}
 ## Creating an empty dispRity object
 make.dispRity()
+
 ## Creating an "empty" dispRity object with a matrix
@@
 
-```{r, eval=FALSE}
+```{r}
 ## The dispRity object's call is indeed empty
 disparity_obj$call
-## Creating an empty dispRity object
+
+## Filling an empty disparity object (that needs to contain at least a matrix)
 (disparity_obj <- fill.dispRity(disparity_obj))
+
 ## The dipRity object has now the proper minimal attributes
@@
 
-```{r, eval=FALSE}
-## Extracting the matrix containing the coordinates for the crown mammals
-str(matrix.dispRity(disp_crown_stemBS, "Group.crown"))
-## Extracting the 3rd boostrapped matrix with the 10th rarefaction level from the same group
-str(matrix.dispRity(disp_crown_stemBS, subsamples = 1, bootstrap = 3, rarefaction = 10))
+```{r}
+## Extracting the matrix containing the coordinates of the elements at time 50
+str(matrix.dispRity(disparity, "50"))
+
+## Extracting the 3rd boostrapped matrix with the 2nd rarefaction level
+## (15 elements) from the second group (80 Mya)
+str(matrix.dispRity(disparity, subsamples = 1, bootstrap = 3, rarefaction = 2))
 ```
@@
 (crown_mammals <- get.subsamples.dispRity(disp_crown_stemBS, "Group.crown"))
+
 ## The object keeps the properties of the parent object but is composed of only one subsamples
@@
 ## Extracting the observed disparity (default)
-extract.dispRity(disp_crown_stemBS)
-
-## Extracting the disparity from the bootstrapped values from the 10th rarefaction level from the
-## second subsamples (stem mammals)
-extract.dispRity(disp_crown_stemBS, observed = FALSE, subsamples = 2, rarefaction = 10)
+extract.dispRity(disparity)
+
+## Extracting the disparity from the bootstrapped values from the
+## 10th rarefaction level from the second subsamples (80 Mya)
+extract.dispRity(disparity, observed = FALSE, subsamples = 2, rarefaction = 10)
 ```
@@
 ## Getting the disparity values of the time subsamples
-summary(disp_time_slicesBS)
+head(summary(disparity))
 
 ## Scaling the same disparity values
-summary(scale(disp_time_slicesBS, scale = TRUE))
+head(summary(scale(disparity, scale = TRUE)))
+
+## Scaling and centering:
+head(summary(scale(disparity, scale = TRUE, center = TRUE)))
 ```
@@
 ## Sorting the disparity subsamples in inverse alphabetic order
-summary(sort(disp_time_slicesBS, decreasing = TRUE))
-```
-
-## Modularity
+head(summary(sort(disparity, decreasing = TRUE)))
+
+## Customised sorting
+head(summary(sort(disparity, sort = c(7,1,3,4,5,2,6))))
+```
+
+<!-- ## Modularity
 
@@
 ## Where is this going?
+ -->
+<!-- ## Running `dispRity` in parallel
+
+The computationally intensive function `dispRity` have a `parallel` option to speed up their calculations.
+
+This option requires the package `snow` and takes arguments that are to be passed to  -->

### 2017-07-12 paper/dispRity_paper.tex
Paper draft INIT

\documentclass[12pt,letterpaper]{article}
\usepackage{natbib}

%Packages
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{dcolumn}
\newcolumntype{d}[1]{D{.}{.}{#1}}

\pagenumbering{arabic}

%Pagination style and stuff
\linespread{2}
\raggedright
\setlength{\parindent}{0.5in}
\setcounter{secnumdepth}{0} 
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}
%\bibpunct{(}{)}{;}{a}{}{,}

%---------------------------------------------
%
%       START
%
%---------------------------------------------

\begin{document}

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: dispRity package.

\bigskip
\medskip
\begin{center}

\noindent{\Large \bf \texttt{dispRity}: a modular \texttt{R} package for measuring disparity.} 
\bigskip

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$}\\
\noindent {\small \it 
$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\\}
\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
\vspace{1in}

%Line numbering
\modulolinenumbers[1]
\linenumbers

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------

\newpage
\begin{abstract}

\begin{enumerate}
    \item I present \texttt{dispRity}, an \texttt{R} package form measuring disparity from ordinated matrices. 
    \item Disparity designates a suit of metrics to describe an ordinated matrix (also referred as morpho/eco/niche/etc.-space).
    \item The package is based on a highly modular architecture were metrics for measuring disparity and tests for testing evolutionary hypothesis are provided as standalone functions.
    \item This modular architecture allows great plasticity in using this package, namely by letting users define their own metric of disparity.
    \item The package also provides numerous tools for modifying, summarising and plotting \texttt{dispRity} objects.
\end{enumerate}

\end{abstract}

\noindent (Keywords: disparity, ordination, multidimensionality, disparity through time, palaeobiology, ecology)\\

\vspace{1.5in}

\newpage

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------

\section{Introduction}
Disparity = multidimensionality

Disparity can be measured in many ways in palaeobiology (cite many examples) or in ecology (donohue and stuff).

Although some packages have disparity components (Claddis, geomorph, adonis), all have their own definition of disparity and makes the whole work idiosuncratic

Therefore I made this package to help measuring disparity

\section{Description}

\subsubsection{The package's data}

Beck and Lee data
Beck and Lee tree
Beck and Lee ages

disparity  disparity data

\subsubsection{Running example}

\subsection{The \texttt{dispRity} package}

\subsubsection{The package's main functions}

\begin{table}
    \begin{tabular}{ll}
        \hline
        Function & Description \\ 
        \hline
        \texttt{geomorph.ordination} & Imports data from the \texttt{geomorph} package. \\
        \texttt{Claddis.ordination} & Imports data from the \texttt{Claddis} package. \\
        \texttt{custom.subsamples} & Separates ordinated data in custom subsamples. \\
        \texttt{time.subsamples} & Separates ordinated data in time subsamples. \\
        \texttt{boot.matrix} & Bootstraps and rarefies ordinated data or a \texttt{dispRity} object. \\
        \texttt{dispRity} & Calculates disparity from an ordinated matrix or a \texttt{dispRity} object. \\
        \texttt{summary.dispRity} & Summarises a \texttt{dispRity} object. \\
        \texttt{plot.dispRity} & Plots a \texttt{dispRity} object. \\
        \texttt{test.dispRity} & Applies statistical tests to a \texttt{dispRity} object.\\
        \texttt{dispRity.per.group} & Pipelined disparity per groups analysis. \\
        \texttt{dispRity.through.time} & Pipelined disparity through time analysis. \\
        \texttt{dispRity.metrics} & A set of implemented disparity metrics functions.\\
        \texttt{dispRity.utilities} & A set of functions for modifying a \texttt{dispRity} object.\\
        \hline
    \end{tabular}
    \caption{Main functions in the \texttt{dispRity} package.}
\end{table}

\subsection{Fork flow}

Workflow figure

\subsubsection{Input data}

\subsubsection{Manipulating the data}

\subsubsection{Measuring disparity}

\subsubsection{Summarising the data}

Link to graph one two three

\subsubsection{Testing hypothesis}

\subsection{The \texttt{dispRity} object}

Disparity object structure figure

\subsection{Additional functions and data}

The package also contains additional functions not directly linked to the disparity workflow nor to the \texttt{dispRity} object but that can be used as utilities for the studying discrete morphological data.

\begin{table}
    \begin{tabular}{ll}
        \hline
        Function & Description \\ 
        \hline
        \texttt{char.diff} & Calculates character differences. \\
        \texttt{clean.data} & Cleans phylogenetic data (tree and table). \\
        \texttt{sim.morpho} & Simulates morphological discrete data. \\
        \texttt{slice.tree} & Slice through a phylogenetic tree. \\
        \texttt{space.maker} & Creates multidimensional spaces. \\
        \texttt{tree.age} & Calculates the age of nodes and tips in a tree. \\
        \hline
    \end{tabular}
    \caption{Additional functions in the \texttt{dispRity} package.}
\end{table}

Additionally, this package also contains an ecology type data set from McClean

McClean data

\section{Conclusion}

This package is good.

\section{Data availability and reproducibility}
Data are available on Dryad or Figshare.
Code for reproducing the analyses is available on GitHub (\url{github.com/TGuillerme/SpatioTemporal_Disparity}).

\section{Acknowledgments}
Thanks to Dave Bapst, Martin Brazeau, Natalie Cooper, Andrew Jackson, Graeme Lloyd and Emma Sherratt.
I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.

\section{Funding}
This work was funded by the European Research Council under the European Union's Seventh Framework Programme (FP/2007–2013)/ERC Grant Agreement number 311092.

\bibliographystyle{sysbio}
\bibliography{References}

\end{document}

### 2017-07-31 doc/dispRity-manual.Rmd
Disparity manual updated

@@
 
-```{r, eval = FALSE}
+```{r, eval=FALSE}
 ## Checking if devtools is already installed
@@
 library(dispRity)
-```
-
-Note this uses the `release` branch (version`0.4`). <!-- add version !-->
-For the piping-hot (but potentially full of bugs) version, you can change the argument `ref = release` to `ref = master`.
-`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, and `RCurl`).
+## Setting the random seed for repeatability
+set.seed(123)
+```
+
+Note this uses the `release` branch (version`0.4`). <!-- @@@ -->
+For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
+`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, `Claddis`, `geomorph` and `RCurl`).
 
@@
 This package is not readily available on CRAN.
-This is mainly because some parts are still in development and that I prefer the reactivity of GitHub for implementing new suggestions from users.
-However, the package follows the strict (and useful!) CRAN standards via Travis. <!-- add link !-->
+This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
+However, the package follows the strict (and useful!) CRAN standards via Travis.
+
+[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
+[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
+[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
 
@@
 
-If you need help about the package, I hope that the following manual will be useful.
-If you still need more help, do not hesitate to drop me an email!
-
-Parts of this package are still in development and some other parts are probably not covered.
+If you need help about the package, hopefully the following manual will be useful.
+However, parts of this package are still in development and some other parts are probably not covered.
 Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
@@
 
-Note that some several options are available, namely which type of distance should be computed.
+Note that several options are available, namely which type of distance should be computed.
 See more info in the function manual (`?Claddis.ordination`).
@@
 
-The package can also easily take data from `geomorph` using the `geomorph.ordination` function.
-This function simply take procrustes data an performs an ordination:
+You can also easily use data from `geomorph` using the `geomorph.ordination` function.
+This function simply take procrustes data and performs an ordination:
 
@@
 Additionally you can provide to the function a `geomorph.data.frame` object.
-If the latter contains sorting information (i.e. factors), they can be directly used to make a `dispRity` object!
+If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsamples)!
 
@@
 
-We will see what are these `dispRity` objects in the sections below...
+More about these `dispRity` objects below!
 
@@
 ```
+
   This results in a ordinated matrix with European cities as elements and 5 dimensions.
-Note that the number of dimensions should not exceed the number of elements - 1.
-
-Of course any other methods for creating the ordination matrix is totally valid!
-The only requirements for `dispRity` is to use matrices with elements as rows and dimensions as columns...
+
+Of course any other methods for creating the ordination matrix is totally valid - or not use any ordination at all!
+The only requirements for the `dispRity` functions is to use matrices with elements as rows and dimensions as columns...
 
@@
 
-Two functions allow to run templated analysis simply using an ordination matrix: either calculating the disparity through time or the disparity per groups.
+Two functions allow to run pipelined analysis simply using an ordination matrix: either calculating the disparity through time or the disparity per groups.
 
@@
 
-For illustrating the following functions, we will use data from Beck and Lee 2015. <!-- ADD PAPER CITE HERE -->
-This dataset contains an ordinated matrix of 50 mammals' discrete characters (called `BeckLee_mat50`), a matrix of the same 50 mammals and the estimate discrete data from their descendants (thus 50 + 49 rows, called `BeckLee_mat99`), a data frame containing the ages of each taxa in the dataset (called `BeckLee_ages`) and finally a phylogenetic tree with the relations of the 50 mammals (called `BeckLee_tree`).
+For illustrating the following functions, we will use data from @beckancient2014.
+This dataset contains an ordinated matrix from 50 mammals' discrete characters (called `BeckLee_mat50`), another matrix of the same 50 mammals and the estimate discrete data from their descendants (thus 50 + 49 rows, called `BeckLee_mat99`), a data frame containing the ages of each taxa in the dataset (called `BeckLee_ages`) and finally a phylogenetic tree with the relations between the 50 mammals (called `BeckLee_tree`).
 
 ```{r, fig.width=7, fig.height=7}
-## Setting the random seed for repeatability
-set.seed(123)
-
 ## Loading the ordinated matrices
@@
 ## The first 5 taxa and dimensions of the 99 taxa + ancestors matrix
-head(BeckLee_mat99[,1:5])
-tail(BeckLee_mat99[,1:5])
+BeckLee_mat99[c(1,2,98,99),1:5]
 
@@
 
-Note of course that you can use your data obtain from the previous chapter!
+Of course you can use your own data as detailed in the [previous chapter](#which-data).
 
@@
   * A required number of time subsamples (here 3)
-  * Your definition of disparity (here the sum of variances)
-
-Note that in the following example, the matrix and the tree comes from the previous sections
+  * Your very own definition of disparity (here the sum of variances)
+
+In the following example, the matrix and the tree comes from the [previous sections](#example-data).
 
@@
 
-When display our disparity data, we can now have information on the operations done to the matrix:
+This generates a `dispRity` object (see [here](#manipulating-dispRity-objects) for technical details).
+When displayed these `dispRity` objects provides us we information on the operations done to the matrix:
 
@@
 ```
-Indeed, we asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times and the metric used was the sum of variances.
+
+Indeed, we asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
 
@@
 summary(disp_aov)
-```
-
-Please refer to the specific tutorials below for (much!) more information on the nuts and bolts of the package.
+
+## Note that anova assumptions where not checked here!
+```
+
+Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
 You can also directly explore the specific functions manuals within R and navigate to they're related functions.
@@
 
-Similarly as above, we can display information about both groups by simply calling it and then summarising the data (and plotting it!) and performing a statistical test to compare both groups (here a Wilcoxon test).
+We can display information about both groups by simply calling the output variable and then summarising the data (and plotting it!) or/and performing a statistical test to compare both groups (here a Wilcoxon test).
 
@@
 test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
-
 ```
@@
 This following section contains specific information relative to some functions.
-If one of your questions are not covered in these sections, please refer to the function manuals in R or send me an email. <!-- Add anchor to contacts -->
+If one of your questions are not covered in these sections, please refer to the function manuals in R or send me an email (<guillert@tcd.ie>).
+The several tutorials below describe specific functionalities of some specific functions, please always refer to the function manuals for the full function documentation!
 
@@
 
- *  discrete time subsamples (or time-binning) using `method = discrete`;
- *  continuous time subsamples (or time-slicing) using `method = continuous`.
+ *  discrete time subsamples (or time-binning) using `method = discrete`
+ *  continuous time subsamples (or time-slicing) using `method = continuous`
 
 For the time-slicing method details see @GuillermeSTD.
-<!-- TODO: link to the time-slicing algorithm explained! -->
 For both methods, the function intakes the `time` argument which can be a vector of `numeric` values for:
 
- *  defining the boundaries of the time bins (when `method = discrete`);
- *  defining the time slices (when `method = continuous`).
+ *  defining the boundaries of the time bins (when `method = discrete`)
+ *  defining the time slices (when `method = continuous`)
 
@@
 
-```{r, eval=FALSE}
+```{r, eval=TRUE}
 ## Generating three time bins containing the data present every 40 Ma
@@
 
+Note that we can also generate equivalent results by just telling the function that we want three time-bins (subsamples) as follow:
+
+```{r, eval=TRUE}
+## Automatically generate three equal length bins:
+time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+                time = 3)
+```
+
 In this example, the taxa were split inside each time-bin according to their age.
-However, the taxa here are considered like single points in time.
-It is totally possible that the taxa in question could have a longer longevity (e.g. if the taxonomic focus is at the genera level or above) and that it spans across multiple time bins.
-In this case, it is possible to include them in more than one bin by providing a table of First and last occurrence dates (FAD/LAD).
+However, the taxa here are considered as single points in time.
+It is totally possible that some taxa could have had longer longevity and that it spans across multiple time bins.
+In this case, it is possible to include them in more than one bin by providing a table of first and last occurrence dates (FAD/LAD).
 This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:
 
-```{r, eval=FALSE}
+```{r, eval=TRUE}
 ## Displaying the table of first and last occurrence dates for each taxa
@@
 
-Note that we can also generate equivalent results by just telling the function that we want three time-bins (subsamples) as follow:
-
-```{r, eval=FALSE}
-## Automatically generate three equal length bins:
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-    time = 3)
-```
-
-When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1% of the tree length as long as at least three elements are present at that point in time.
+When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age plus 1% of the tree length as long as at least three elements are present at that point in time.
 The algorithm adds an extra 1% tree length until reaching the required minimum of three elements.
@@
 
-```{r, eval=FALSE}
+```{r, eval=TRUE}
 ## Generating four time slices every 40 million years under gradual evolution
@@
     FADLAD = BeckLee_ages)
+
 ## Generating four time slices automatically
@@
 
+## Customised subsamples
+
+Another way of separating elements into different categories is to use customised subsamples as briefly explained [above](#disparity-between-groups).
+This function simply takes the list of elements to put in each group (whether they are the actual element names or their idea).
+
+```{r, eval=TRUE}
+## Creating the two groups as a list
+mammal_groups <- list("crown" = c(16, 19:41, 45:50),
+                      "stem" = c(1:15, 17:18, 42:44))
+
+## Separating the dataset into two different groups
+custom.subsamples(BeckLee_mat50, group = mammal_groups)
+```
+
+Elements can easily be assigned to different groups if necessary!
+
+```{r, eval=FALSE}
+## Creating the three groups as a list
+mammal_groups <- list("crown" = c(16, 19:41, 45:50),
+                      "stem" = c(1:15, 17:18, 42:44).
+                      "all" = c(1:50))
+```
+
 ## Bootstraps and rarefactions
 
-One important step in analysing ordinated matrices is to pseudo-replicate the data to seem how robust the results of future analysis are sensitive to outliers in the dataset.
-This can be achieved using the function `boot.matrix` to bootstrap or rarefy the ordinated matrix.
-The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):
-
-```{r, eval=FALSE}
+One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results of future analysis are sensitive to outliers in the dataset.
+This can be achieved using the function `boot.matrix` to bootstrap or/and rarefy the data.
+The default options will bootstrapping the matrix 100 times without rarefaction using the "full" bootstrap method (see below):
+
+```{r, eval=TRUE}
 ## Default bootstrapping
@@
 
- *  `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any elements drawn from the dat);
- *  `single` where only one random element is replaced by one other random element for each pseudo-replicate.
-
-```{r, eval=FALSE}
+ * `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any *m* elements drawn from the data)
+ * `single` where only one random element is replaced by one other random element for each pseudo-replicate
+
+```{r, eval=TRUE}
 ## Bootstrapping with the single bootstrap method
@@
 Rarefaction allows to limit the number of elements to be drawn at each bootstrap replication.
-For example, one might be interested in looking at the effect of reducing the number of elements to the results of the analysis.
+For example, one might be interested in looking at the effect of reducing the number of elements on the results of an analysis.
 
@@
 
-```{r, eval=FALSE}
+```{r, eval=TRUE}
 ## Bootstrapping with the full rarefaction
@@
 
-```{r, eval = FALSE}
+```{r, eval=TRUE}
 ## Removing the last axis 50% last dimensions
@@
 In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
-For example, one can look at ellipsoid hyper-volume of the ordinated space
-<!-- Cite Donohue -->
-, the sum and the product of the ranges and variances
-<!-- Cite Wills -->
- or the median position of the elements relative to their centroid.
- <!-- Cite Sive -->
+For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @finlay2015morphological.
 Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks or some being more generalist.
@@
 
-Several level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` function that calculates the variance of each dimension of the ordinated matrix or the `ranges` one that calculates the range of each dimension.
+Several level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculates the variance or the range of each dimension of the ordinated matrix respectively.
 
@@
 ## the variance covariance matrix (level 1, level 2 and level 3 functions).
-
 summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
@@
 ## variance covariance matrix:
-disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
+disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)),
+                      round = 10)
 
 ## Same but using a different functions order for the metric argument
-disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)), round = 10)
+disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
+                      round = 10)
 
@@
 
-#### What level is the disparity?
-
 In these examples, we considered disparity to be a single value.
 For example, in the previous one, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
-This results in single value to be calculated.
-
-In some cases however, it might also be interesting to calculate disparity as a distribution.
-For example, one of the metrics commonly used in palaeobiology is the sum of variances of each dimensions (`metric = c(sum, variances)`) [@Wills1994].
-However, it might be interesting to look directly at the distribution of the variances itself.
-
-```{r}
-## A one dimensional disparity value
-(sum_variances1 <- summary(dispRity(BeckLee_mat50, metric = c(median, variances))))
-
-## A two dimensional disparity
-(summary(all_variances <- (dispRity(BeckLee_mat50, metric = variances))))
-```
-
-Looking at all the variances rather than simply their median can tell us that (by the look of it) the variances of each dimensions seems to be normally distributed.
-
-> It is also possible to reuse an already calculated disparity distribution and summarise it using a level 1 metric:
-
-```{r}
-## Transforming the two dimensional one in a one dimensional one
-sum_variances2 <- summary(dispRity(all_variances, metric = median))
-
-## Both ways output the same disparity values:
-sum_variances1 == sum_variances2
-```
+It is however possible to calculate [disparity as a distribution](#disparity-as-a-distribution).
 
@@
 
-1: The functions estimation of the eigen value only works for MDS or PCoA ordinations (not PCA).
+1: This function uses an estimation of the eigen value that only works for MDS or PCoA ordinations (not PCA).
 
@@
 Here is a run through of the main metrics implemented in `dispRity`.
-First, we will create a dummy ordinated space using the `space.maker` utility function:
+First, we will create a dummy ordinated space using the `space.maker` utility function (more about that [here](#space.maker):
 
@@
 
-## Calculating the volume with different options to be passed to hypervolume::hypervolume
-summary(dispRity(dummy_space, metric = hyper.volume, repsperpoint = 100, quantile = 0.5))
+## The volume with different options to be passed to hypervolume::hypervolume
+summary(dispRity(dummy_space, metric = hyper.volume, repsperpoint = 100,
+                 quantile = 0.5))
 ```
@@
 
-> This metric is only mathematically valid if the dimensions within the space are all orthogonal!
+> This metric is only an euclidean diagonal (mathematically valid) if the dimensions within the space are all orthogonal!
 
@@
 ```{r}
-## Calculating the distribution of the distances between each elements and the centroid
+## The distribution of the distances between each elements and the centroid
 summary(dispRity(dummy_space, metric = centroids))
 
-## Calculating disparity as the median value of these distances
+## Disparity as the median value of these distances
 summary(dispRity(dummy_space, metric = c(median, centroids)))
@@
 ```{r}
-## Calculating the distance between the elements and the origin of the ordinated space
+## The distance between the elements and the origin of the ordinated space
 summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))
 
-## Calculating disparity as the distance between the elements and a specific point
+## Disparity as the distance between the elements and a specific point in space
 summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
@@
 
-Because of its architecture, printing `dispRity` objects only summarise their content but does not print the disparity value measured or associated analysis.
-<!-- link to dispRity objects -->
+Because of its architecture, printing `dispRity` objects only summarise their content but does not print the disparity value measured or associated analysis (more about this [here](#manipulating-dispRity-objects)).
 To actually summary what's in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphic representation of the results.
@@
 
-The summary functions can also take various options such as the `quantile` values for the confidence intervals levels (by default, the 50 and 95 quantiles are calculated); the `cent.tend` for the  central tendency to use for summarising the results (default is `median`); the `rounding` option corresponding to the number of decimals to print (default is `2`) and the `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
+The summary functions can also take various options such as:
+
+ * the `quantile` values for the confidence intervals levels (by default, the 50 and 95 quantiles are calculated)
+ * the `cent.tend` for the  central tendency to use for summarising the results (default is `median`)
+ * the `rounding` option corresponding to the number of decimals to print (default is `2`)
+ *  and the `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
+
 These options can easily be changed from the defaults as follows:
@@
 
-```{r, eval = FALSE}
+```{r, eval=FALSE}
 ## Loading the xtable package
@@
 
-The plots can be of four different types: `continuous` for displaying continuous disparity curves and `box`, `lines` and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines and confidence interval polygons.
+The plots can be of four different types:
+
+ * `continuous` for displaying continuous disparity curves
+ * and `box`, `lines` and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines and confidence interval polygons.
 
@@
  * `pairwise` (default): to compare each subsamples pairwise
-
  * `referential`: to compare each subsamples to the first one
-
  * `sequential`: to compare each subsamples to the following one
-
  * `all`: to compare all the subsamples together (like in analysis of variance)
@@
 ```{r}
-## Performing a t-test to the test a difference in disparity between crown and stem mammals
+## T-test to the test a difference in disparity between crown and stem mammals
 test.dispRity(disparity_crown_stem, test = t.test)
@@
 
-## Performing a Wilcoxon test applied to sliced disparity the data sequentially with correction
-test.dispRity(disparity_time_slices, test = wilcox.test, comparisons = "sequential",
-    correction = "bonferroni")
-
-## Measuring the overlap between distributions in the time bins (using the implemented
-## Bhattacharyya Coefficient function - see ?bhatt.coeff)
+## Wilcoxon test applied to sliced disparity the data sequentially with correction
+test.dispRity(disparity_time_slices, test = wilcox.test,
+              comparisons = "sequential", correction = "bonferroni")
+
+## Measuring the overlap between distributions in the time bins (using the
+## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)
 test.dispRity(disparity_time_bins, test = bhatt.coeff)
@@
 ## Performing and ANOVA applied to the same data
-slice_aov <- test.dispRity(disparity_time_slices, test = aov, comparisons = "all")
-slice_aov
+(slice_aov <- test.dispRity(disparity_time_slices, test = aov,
+                            comparisons = "all"))
 
@@
 ```
+
 Of course, following the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
@@
 For example, the sum of variances, the product of ranges or the median distance between the elements and their centroid will summarise disparity as a single value.
-This value can be of course pseudo-replicated (boostraped) in order to obtain a more confident distribution of this summary metric.
+This value can be of course pseudo-replicated (bootstrapped) in order to obtain a more confident distribution of this summary metric.
 However, another way to perform disparity analysis is to use the whole distribution rather than their summary (e.g. the variances or the ranges).
@@
 plot(disparity_centroids, ylab = "Distribution of all the distances")
-plot(disparity_centroids_median, ylab = "Distribution of the medians of all the distances")
+plot(disparity_centroids_median,
+     ylab = "Distribution of the medians of all the distances")
 
@@
 
-```{r, fig.width=12, fig.height=6, eval = FALSE}
+```{r, fig.width=12, fig.height=6, eval=TRUE}
 ## Summary table of the data
@@
 
+### `space.maker`
+Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
+This function allows to simulated multidimensional spaces given a certain number of properties.
+It takes as arguments the number of elements (data points) and dimensions to create the space but also allows more fine tuning data simulation: it is possible to give a specific distribution to each dimensions, provide a correlation matrix to link the dimensions or even providing an *a priori* distribution of the variance per distributions!
+
+```{r, fig.width=8, fig.height=8, eval=TRUE}
+## Graphical options
+op <- par(mfrow = (c(2,2)), bty = "n")
+## Visualising 3D spaces
+require(scatterplot3d)
+
+## A cube space
+scatterplot3d(space.maker(2500, 3, runif), pch = 20,
+              xlab = "", ylab = "", zlab = "",
+              main = "Uniform cube space")
+
+## A plane space
+scatterplot3d(space.maker(2500, 3, c(runif, runif, runif),
+              arguments = list(list(min = 0, max = 0), NULL, NULL)), pch = 20,
+              xlab = "", ylab = "", zlab = "",
+              main = "Uniform plane space")
+
+## An ellipsoid space (=a spheric space with correlation)
+cor_matrix <- matrix(cbind(1,0.8,0.2, 0.8,1,0.7, 0.2,0.7,1), nrow = 3)
+scatterplot3d(space.maker(2500, 3, rnorm, cor.matrix = cor_matrix), pch = 20,
+              xlab = "", ylab = "", zlab = "",
+              main = "Normal ellipsoid space")
+
+## A cylindrical space with a decreasing variance per axis
+scatterplot3d(space.maker(2500, 3, c(rnorm, rnorm, runif),
+              scree = c(0.7, 0.2, 0.1)), pch = 20,
+              xlab = "", ylab = "", zlab = "",
+              main = "Normal cylindrical space")
+
+## Resetting the graphic parameters
+par(op)
+```
+
 # Glossary <a name="glossary"></a>

### 2017-08-04 doc/dispRity-manual.Rmd
Fixed typos etc to 'specific tutorials

@@
 
-## What's `dispRity`
+## What is `dispRity`?
 This is a package for measuring disparity in `R`.
 It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
-Typically, these analysis are used in palaeobiology to study the changes in morphology through time.
+Typically, these analysis are used in palaeobiology and evolutionary biology to study the changes in morphology through time.
+However, there are many more applications in ecology, evolution and beyond.
 
@@
 ## Checking if Claddis is installed BUILD ONLY
+# NC: this isn't checking Claddis...
 library(dispRity)
@@
 
-## Why not CRAN
-
-This package is not readily available on CRAN.
+## Why not CRAN?
+
+This package is not available on CRAN.
 This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
@@
 
-If you need help about the package, hopefully the following manual will be useful.
+If you need help with the package, hopefully the following manual will be useful.
 However, parts of this package are still in development and some other parts are probably not covered.
@@
 
-## Which data?
+## What sort of data does dispRity work with?
 
 Disparity must be measured from ordinated matrices.
-These matrices can be of any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your elements names as rows (taxa, experiments, countries, etc.) and your ordinations axis as columns (the dimensions of your dataset).
-
-### Ordination from `Claddis`
+These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).
+
+### Ordination matrices from `Claddis`
 
@@
 `dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
-For this, simply input a matrix in the `Claddis` format to the function an it will automatically calculate and ordinate the distances among taxa:
+For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:
 
@@
 
-### Ordination from `geomorph`
+### Ordination matrices from `geomorph`
 
 ```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
+## Checking if geomorph is installed BUILD ONLY
 if(!require(geomorph)) install.packages("geomorph")
@@
 You can also easily use data from `geomorph` using the `geomorph.ordination` function.
-This function simply take procrustes data and performs an ordination:
+This function simply takes Procrustes aligned data and performs an ordination:
 
@@
 
-## Performing a procrustes transform on the landmarks
+## Performing a Procrustes transform on the landmarks
 procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)
@@
 Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
-Additionally you can provide to the function a `geomorph.data.frame` object.
+Additionally you can give the function a `geomorph.data.frame` object.
 If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsamples)!
@@
 
-## Ordinating this data and making a disparity object
+## Ordinating this data and making a dispRity object
 geomorph.ordination(geomorph_df)
@@
 
-### General ordinations
-
-If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordinations by using the following functions from the `stats` package.
+### Other kinds of ordination matrices
+
+If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordination matrices by using the following functions from the `stats` package.
 Here is how to do it for the following types of matrices:
 
- * Multivariate matrices
+ * Multivariate matrices (principal components analysis; PCA)
 
@@
 
-## Ordinating the matrix using prcomp()
+## Ordinating the matrix using `prcomp` 
 ordination <- prcomp(USArrests)
@@
 ```
-  This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, have a look at the `?princomp` function.
-
- * Distance matrices
+  This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, see the `?princomp` function.
+
+ * Distance matrices (classical multidimensional scaling; MDS)
 
@@
 
-  This results in a ordinated matrix with European cities as elements and 5 dimensions.
-
-Of course any other methods for creating the ordination matrix is totally valid - or not use any ordination at all!
-The only requirements for the `dispRity` functions is to use matrices with elements as rows and dimensions as columns...
-
-## Simple analysis
-
-Two functions allow to run pipelined analysis simply using an ordination matrix: either calculating the disparity through time or the disparity per groups.
+  This results in a ordinated matrix with European cities as elements and five dimensions.
+
+Of course any other method for creating the ordination matrix is totally valid, you can also not use any ordination at all!
+The only requirements for the `dispRity` functions is that the input is a matrix with elements as rows and dimensions as columns.
+
+## Performing a simple dispRity analysis
+
+Two `dispRity` functions allow users to run an analysis pipeline simply by inputting an ordination matrix.
+These functions allow users to either calculate the disparity through time (`dispRity.through.time`) or the disparity of user-defined groups (`dispRity.per.group`).
 
@@
 
-For illustrating the following functions, we will use data from @beckancient2014.
-This dataset contains an ordinated matrix from 50 mammals' discrete characters (called `BeckLee_mat50`), another matrix of the same 50 mammals and the estimate discrete data from their descendants (thus 50 + 49 rows, called `BeckLee_mat99`), a data frame containing the ages of each taxa in the dataset (called `BeckLee_ages`) and finally a phylogenetic tree with the relations between the 50 mammals (called `BeckLee_tree`).
+To illustrate these functions, we will use data from @beckancient2014.
+This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
 
@@
 
-## The first 5 taxa and dimensions of the 50 taxa matrix
-head(BeckLee_mat50[,1:5])
-
-## The first 5 taxa and dimensions of the 99 taxa + ancestors matrix
-BeckLee_mat99[c(1,2,98,99),1:5]
-
-## Loading a list of first and last occurrence data
+## The first five taxa and dimensions of the 50 taxa matrix
+head(BeckLee_mat50[, 1:5])
+
+## The first five taxa and dimensions of the 99 taxa + ancestors matrix
+BeckLee_mat99[c(1, 2, 98, 99), 1:5]
+
+## Loading a list of first and last occurrence dates for the fossils
 data(BeckLee_ages)
@@
 
-## Loading the phylogeny
+## Loading and plotting the phylogeny
 data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.8) ; axisPhylo(root = 140) ; nodelabels(cex = 0.5)
-```
+plot(BeckLee_tree, cex = 0.8) 
+axisPhylo(root = 140)
+nodelabels(cex = 0.5)
+```
+
+<!-- I removed the semicolons above, it is confusing to novice R users -->
 
 Of course you can use your own data as detailed in the [previous chapter](#which-data).
+<!-- I changed the header so may have broken this link -->
 
@@
 
-One of the functions of the package is to look at how disparity changes through time.
-This function (and the following one) are the two most straight forward, they analysis the ordinated data in a pipelined way using a lot of default parameters.
-
-For disparity through time analysis, you will need:
+The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
+This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
+Of course all the defaults can be changed if required, more on this later.
+
+For a disparity through time analysis, you will need:
 
   * An ordinated matrix (we covered that above)
-  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element. To give your tree a root time, you can simply do `my_tree$root.time <- my_age`.
-  * A required number of time subsamples (here 3)
-  * Your very own definition of disparity (here the sum of variances)
-
-In the following example, the matrix and the tree comes from the [previous sections](#example-data).
+  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element.
+To give your tree a root time (i.e.
+an age for the root), you can simply do `my_tree$root.time <- my_age`.
+  * The required number of time subsamples (here `time = 3`)
+  * Your favourite disparity metric (here the sum of variances)
+
+Using the @beckancient2014 data described [above](#example-data):
 
@@
 This generates a `dispRity` object (see [here](#manipulating-dispRity-objects) for technical details).
-When displayed these `dispRity` objects provides us we information on the operations done to the matrix:
-
-```{r}
-## Printing what's in the disparity data
+When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:
+
+```{r}
+## Print the disparity_data object
 disparity_data
@@
 
-Indeed, we asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
-
-We can now analyse this data by summarising and plotting it or performing some statistical tests on it (e.g. a simple `aov`):
+We asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
+
+We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `aov`):
 
 ```{r, fig.width=7, fig.height=7}
-## Summarising the measurements
+## Summarising disparity through time
 summary(disparity_data)
@@
 
-## Testing for an eventual difference between the time bins
+## Testing for an difference among the time bins
 disp_aov <- test.dispRity(disparity_data, test = aov, comparisons = "all")
@@
 
-## Note that anova assumptions where not checked here!
-```
+## Note that ANOVA assumptions were not checked here!
+```
+
+<!-- NC: Is this using aov or lm?
+Note that aov can only be used with balanced designs, so where the three time bins have equal numbers of species.
+I expect you need to use lm (which also fits an ANOVA as an ANOVA is also a linear model).
+THIS IS IMPORTANT -->
 
 Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
-You can also directly explore the specific functions manuals within R and navigate to they're related functions.
-
-### Disparity between groups
-
-Alternatively, you can be interested in looking at disparity between groups rather than through time.
-For example, on question could be is there a difference in disparity between two groups?
-
-To perform such analysis, you will need:
+You can also directly explore the specific function help files within R and navigate to related functions.
+
+### Disparity among groups
+
+The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
+For example, you could ask if there is a difference in disparity between two groups?
+
+To perform such an analysis, you will need:
 
@@
 Note that elements can be present in multiple groups at once.
- * Your definition of disparity (here the sum of variances)
-
-```{r}
-## Creating the two groups as a list
+ * Your favourite disparity metric (here the sum of variances)
+
+Using the @beckancient2014 data described [above](#example-data):
+
+```{r}
+## Creating the two groups (crown versus stem) as a list
 mammal_groups <- list("crown" = c(16, 19:41, 45:50),
@@
 
-We can display information about both groups by simply calling the output variable and then summarising the data (and plotting it!) or/and performing a statistical test to compare both groups (here a Wilcoxon test).
+We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).
 
 ```{r, fig.width=7, fig.height=7}
-## Printing what's in the disparity data
+## Print the disparity_data object
 disparity_data
 
-## Summarising the measurements
+## Summarising disparity in the different groups
 summary(disparity_data)
@@
 
-## Testing for an eventual difference between the groups
+## Testing for a difference between the groups
 test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
@@
 
-This following section contains specific information relative to some functions.
-If one of your questions are not covered in these sections, please refer to the function manuals in R or send me an email (<guillert@tcd.ie>).
-The several tutorials below describe specific functionalities of some specific functions, please always refer to the function manuals for the full function documentation!
+The following section contains information specific to some functions.
+If any of your questions are not covered in these sections, please refer to the function help files in R, send me an email (<guillert@tcd.ie>), or raise an issue on GitHub.
+The several tutorials below describe specific functionalities of certain functions; please always refer to the function help files for the full function documentation!
 

### 2017-08-04 doc/dispRity-manual.Rmd
Fixed typos etc to 'make.metric'

@@
 
-The function `time.subsamples` allows to divide the ordinated matrix into different time subsamples given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
+The function `time.subsamples` allows users to divide the ordinated matrix into different time subsamples or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
 Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.
@@
 For the time-slicing method details see @GuillermeSTD. 
-For both methods, the function intakes the `time` argument which can be a vector of `numeric` values for:
+<!-- NC: Or potentially this paper we are writing for the PalAss? -->
+For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:
 
@@
 Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
-Additionally, it is also possible to input a data frame containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age.
+Additionally, it is also possible to input a dataframe containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age, so taxa can appear in more than one time bin/slice.
 
@@
 ```{r, eval=TRUE}
-## Generating three time bins containing the data present every 40 Ma
+## Generating three time bins containing the taxa present every 40 Ma
 time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
@@
 
-Note that we can also generate equivalent results by just telling the function that we want three time-bins (subsamples) as follow:
+Note that we can also generate equivalent results by just telling the function that we want three time-bins as follow:
 
@@
 However, the taxa here are considered as single points in time.
-It is totally possible that some taxa could have had longer longevity and that it spans across multiple time bins.
+It is totally possible that some taxa could have had longer longevity and that they exist in multiple time bins.
 In this case, it is possible to include them in more than one bin by providing a table of first and last occurrence dates (FAD/LAD).
@@
 
-When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age plus 1% of the tree length as long as at least three elements are present at that point in time.
-The algorithm adds an extra 1% tree length until reaching the required minimum of three elements.
-It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance between tips *and* nodes.
+When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as the root age plus 1\% of the tree length, as long as at least three elements/taxa are present at that point in time.
+The algorithm adds an extra 1\% tree length until reaching the required minimum of three elements.
+It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance among tips *and* nodes.
 
 For the time-slicing method (`method = continuous`), the idea is fairly similar.
-This option, however, requires a matrix that contains the ordinated distance between taxa *and* nodes as well as an extra argument being the assumed evolutionary model (via the `model` argument).
-This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node.
+This option, however, requires a matrix that contains the ordinated distance among taxa *and* nodes and an extra argument describing the assumed evolutionary model (via the `model` argument).
+This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node, meaning that a decision must be made about what the value for the branch should be.
 The model can be one of the following:
@@
 
+<!--NC: Bit puzzled by the offspring/descendent terminology - I'd assume these were the same thing. By descendant do you mean ancestor? I'm confused  -->
+
 ```{r, eval=TRUE}
-## Generating four time slices every 40 million years under gradual evolution
+## Generating four time slices every 40 million years under a model of gradual evolution
 time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
@@
 This function simply takes the list of elements to put in each group (whether they are the actual element names or their idea).
+
+<!-- NC: Their idea? Huh? Whats that mean? -->
 
@@
 
-One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results of future analysis are sensitive to outliers in the dataset.
-This can be achieved using the function `boot.matrix` to bootstrap or/and rarefy the data.
-The default options will bootstrapping the matrix 100 times without rarefaction using the "full" bootstrap method (see below):
+One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results are, and how sensitive they are to outliers in the dataset.
+This can be achieved using the function `boot.matrix` to bootstrap and/or rarefy the data.
+The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):
 
@@
 
-This function also allows to rarefy the data using the `rarefaction` argument.
-Rarefaction allows to limit the number of elements to be drawn at each bootstrap replication.
-For example, one might be interested in looking at the effect of reducing the number of elements on the results of an analysis.
+This function also allows users to rarefy the data using the `rarefaction` argument.
+Rarefaction allows users to limit the number of elements to be drawn at each bootstrap replication.
+This is useful if, for example, one is interested in looking at the effect of reducing the number of elements on the results of an analysis.
 
@@
 The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsamples to a minimum of 3 elements).
+remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsample to a minimum of three elements).
 It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
@@
 ## Or with a set number of rarefaction levels
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8,3))
+boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8, 3))
 ```
@@
 One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
-When left missing, all dimensions from the ordinated matrix are used.
+When missing, all dimensions from the ordinated matrix are used.
+
+<!-- NC: This is confusing to me. If it's a number > 0, it keeps that number of dimensions, and if < 0 it removes the last dimensions by that %? This is very odd. Please explain -->
 
 ```{r, eval=TRUE}
-## Removing the last axis 50% last dimensions
+## Removing the last 50% of the dimensions
 boot.matrix(BeckLee_mat50, dimensions = 0.5)
 
-## Keeping the 10 first dimensions
+## Keeping the first 10 dimensions
 boot.matrix(BeckLee_mat50, dimensions = 10)
@@
 
-Of course, one could be interested in directly supplying the subsamples generated above directly to this function.
-In fact, it can also deal with a list of matrices or with a `dispRity` object output from the `custom.subsamples` or `time.subsamples` functions.
+Of course, one could directly supply the subsamples generated above (using <!--  NC: which function ? How is this line different from the statement in the line below? -->) to this function.
+It can also deal with a list of matrices or with a `dispRity` object output from the `custom.subsamples` or `time.subsamples` functions.
 
@@
 crown_stem <- custom.subsamples(BeckLee_mat50,
-    group = list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44)))
+                                group = list("crown" = c(16, 19:41, 45:50), 
+                                             "stem" = c(1:15, 17:18, 42:44)))
 ## Bootstrapping and rarefying these groups
@@
 
-## Creating time slices subsamples
+## Creating time slice subsamples
 time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
+                               method = "continuous", model = "gradual", 
+                               time = c(120, 80, 40, 0),
                                FADLAD = BeckLee_ages)
-## Bootstrapping on the time binning/slicing subsamples
+## Bootstrapping the time slice subsamples
 boot.matrix(time_slices, bootstraps = 100)
@@
 
-There are many ways for measuring disparity!
+There are many ways of measuring disparity!
 In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
 For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @finlay2015morphological.
-Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks or some being more generalist.
-
-Because of this great diversity of metric, the package `dispRity` has not a unique way to measure disparity but rather proposes to facilitate users to define their own disparity metric that will best suit to their particular analysis.
+<!-- NC: probably should cite Matt Wills' paper for the centroid thing too as he invented it apparently -->
+Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.
+
+Because of this great diversity of metrics, the package `dispRity` does not have one way to measure disparity but rather proposes to facilitate users in defining their own disparity metric that will best suit their particular analysis.
 In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
@@
 
-1.  it must be composed from one to three `function` objects;
-2.  the(se) function(s) must intake as a first argument a `matrix` or a `vector`;
-3.  the(se) function(s) must be of one of the three levels described below;
-4.  at least one of the functions must be of level 1 or 2 (see below).
-
-### The function levels
-
-The metric functions levels determine the "dimensionality" level of decomposition of the input ordinated matrix.
-In other words, each level designate the dimensions of the output: either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimensions.
-
-![Illustration of the different levels of output of a function with an input `matrix`](dispRity_fun.png)
-
-#### A level 1 function
-
-A level 1 function will decompose a `matrix` or a `vector` into a single value:
+1.  It must be composed from one to three `function` objects;
+2.  The function(s) must take as a first argument a `matrix` or a `vector`;
+3.  The function(s) must be of one of the three dimension-levels described below;
+4.  At least one of the functions must be of dimension-level 1 or 2 (see below).
+
+<!-- NC: We discussed this previously, but I don't like the term levels here.
+It's confusing because it means something specific wrt factors.
+Can we change it to dimension-level?
+That's more descriptive and clearer in my opinion.
+I have changed it below, see what you think -->
+
+### The function dimension-levels
+
+The metric function dimension-levels determine the "dimensionality of decomposition" of the input matrix.
+In other words, each dimension-level designates the dimensions of the output, i.e. either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimension.
+
+![Illustration of the different dimension-levels of functions with an input `matrix`](dispRity_fun.png)
+
+#### Dimension-level 1 functions
+
+A dimension-level 1 function will decompose a `matrix` or a `vector` into a single value:
 
@@
 
-## Example of level 1 functions
+## Example of dimension-level 1 functions
 mean(dummy_matrix)
@@
 
-Any summary metric such as mean or median are good examples of level 1 functions as they reduce the matrix to a single dimension (i.e. one value).
-
-#### A level 2 function
-
-A level 2 function will decompose a `matrix` into a `vector`.
+Any summary metric such as mean or median are good examples of dimension-level 1 functions as they reduce the matrix to a single dimension (i.e. one value).
+
+#### Dimension-level 2 functions
+
+A dimension-level 2 function will decompose a `matrix` into a `vector`.
 
@@
 
-## A level 2 metric
+## A dimension-level 2 metric
 prod.rows(dummy_matrix)
@@
 
-Several level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculates the variance or the range of each dimension of the ordinated matrix respectively.
-
-#### A level 3 function
-
-Finally a level 3 function will transform the matrix into another matrix.
-Note that the dimension of the output matrix don't need to match the the input one:
-
-```{r}
-## A level 3 metric
+Several dimension-level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculate the variance or the range of each dimension of the ordinated matrix respectively.
+
+#### Dimension-level 3 functions
+
+Finally a dimension-level 3 function will transform the matrix into another matrix.
+Note that the dimension of the output matrix doesn't need to match the the input matrix:
+
+```{r}
+## A dimension-level 3 metric
 var(dummy_matrix)
 
-## A level 3 metric with a forced matrix output
+## A dimension-level 3 metric with a forced matrix output
 as.matrix(dist(dummy_matrix))
@@
 
-Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the euclidean distance between each elements and the a position in the ordinated space.
-The `make.metric` function implemented in `dispRity` is designed to help testing and finding the level of the functions.
+Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the Euclidean distance between each elements and the a position in the ordinated space.
+The `make.metric` function implemented in `dispRity` is designed to help testing and finding the dimension-level of the functions.
 This function tests:
@@
 1.  if your function can deal with a `matrix` or a `vector` as an input;
-2.  which is your function's level according to it's output (level 1, 2 or 3, see above);
+2.  which is your function's dimension-level according to it's output (dimension-level 1, 2 or 3, see above);
 3.  whether the function can properly be implemented in the `dispRity` function (the function is fed into a `lapply` loop).
 
-For example, let's see if the functions described above are the right levels:
-
-```{r}
-## Which level is the mean function? And is able to be used in the dispRity?
+For example, let's see if the functions described above are the right dimension-levels:
+
+```{r}
+## Which dimension-level is the mean function? And is able to be used in the dispRity?
 make.metric(mean)
@@
 A silent version of the function is also available to be included in your code.
-This can be done using the option `silent = TRUE` and will simply output the level of the metric.
-
-```{r}
-## Testing whether mean is level 1
-if(make.metric(mean, silent = TRUE) != "level1") {
-    message("The metric is not level 1.")
+This can be done using the option `silent = TRUE` and will simply output the dimension-level of the metric.
+
+```{r}
+## Testing whether mean is dimension-level 1
+if(make.metric(mean, silent = TRUE) != "dimension-level1") {
+    message("The metric is not dimension-level 1.")
 }
-## Testing whether var is level 1
-if(make.metric(var, silent = TRUE) != "level1") {
-    message("The metric is not level 1.")
+## Testing whether var is dimension-level 1
+if(make.metric(var, silent = TRUE) != "dimension-level1") {
+    message("The metric is not dimension-level 1.")
 }
@@
 ## Measuring disparity as the standard deviation of all the value of the
-## ordinated matrix (level 1 function).
+## ordinated matrix (dimension-level 1 function).
 summary(dispRity(BeckLee_mat50, metric = sd))
@@
 ## Measuring disparity as the standard deviation of the variance of each axis of
-## the ordinated matrix (level 1 and level 2 functions).
+## the ordinated matrix (dimension-level 1 and 2 functions).
 summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))
@@
 ## Measuring disparity as the standard deviation of the variance of each axis of
-## the variance covariance matrix (level 1, level 2 and level 3 functions).
+## the variance covariance matrix (dimension-level 1, 2 and 3 functions).
 summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
@@
 
-Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function levels (using `make.metric`) and apply them to the data in decreasing order (level 3 > level 2 > level 1).
+Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > dimension-level 2 > dimension-level 1).
 
@@
 
-This is possible in the `dispRity` package by calculating disparity as a level 2 metric only!
+This is possible in the `dispRity` package by calculating disparity as a dimension-level 2 metric only!
 Let's have a look using our previous example of bootstrapped time slices but by measuring the distances between each taxa and their centroid as disparity
@@
 
-The resulting disparity object being a level 2, it can easily also be transformed into a level 1 one by for example measuring the median distance of all these distributions:
+The resulting disparity object being a dimension-level 2, it can easily also be transformed into a dimension-level 1 one by for example measuring the median distance of all these distributions:
 

### 2017-08-04 doc/dispRity-manual.Rmd
Fixed typos etc to 'Metric run through'

@@
 
-Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the Euclidean distance between each elements and the a position in the ordinated space.
-The `make.metric` function implemented in `dispRity` is designed to help testing and finding the dimension-level of the functions.
+Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the Euclidean distance between each element and the centroid of the ordinated space.
+The `make.metric` function implemented in `dispRity` is designed to help test and find the dimension-level of the functions.
 This function tests:
 
-1.  if your function can deal with a `matrix` or a `vector` as an input;
-2.  which is your function's dimension-level according to it's output (dimension-level 1, 2 or 3, see above);
-3.  whether the function can properly be implemented in the `dispRity` function (the function is fed into a `lapply` loop).
+1.  If your function can deal with a `matrix` or a `vector` as an input;
+2.  Your function's dimension-level according to its output (dimension-level 1, 2 or 3, see above);
+3.  Whether the function can be implemented in the `dispRity` function (the function is fed into a `lapply` loop).
 
@@
 ```{r}
-## Which dimension-level is the mean function? And is able to be used in the dispRity?
+## Which dimension-level is the mean function? And can it be used in dispRity?
 make.metric(mean)
 
-## Same questions from the prod.rows function:
+## Which dimension-level is the prod.rows function? And can it be used in dispRity?
 make.metric(prod.rows)
 
-## Same questions for the var function:
+## Which dimension-level is the var function? And can it be used in dispRity?
 make.metric(var)
@@
 
-A silent version of the function is also available to be included in your code.
+A non verbose version of the function is also available.
 This can be done using the option `silent = TRUE` and will simply output the dimension-level of the metric.
+
+<!-- NC: My terminology changes may have broken this bit -->
 
@@
 
-### metric argument in `dispRity`
+### Metrics in the `dispRity` function
 
@@
 ```{r}
-## Measuring disparity as the standard deviation of all the value of the
+## Measuring disparity as the standard deviation of all the values of the
 ## ordinated matrix (dimension-level 1 function).
@@
 
-## Same but using a different functions order for the metric argument
+## Same as above but using a different function order for the metric argument
 disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
@@
 In these examples, we considered disparity to be a single value.
-For example, in the previous one, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
-It is however possible to calculate [disparity as a distribution](#disparity-as-a-distribution).
+For example, in the previous example, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
+It is, however, possible to calculate [disparity as a distribution](#disparity-as-a-distribution).
 
@@
 
-Several disparity metrics are directly implemented in the `dispRity` package.
-The detailed list can be found in `?dispRity.metric` along with some explanations on each metric.
+Several disparity metrics are implemented in the `dispRity` package.
+The detailed list can be found in `?dispRity.metric` along with some description of each metric.
 
@@
 
-1: This function uses an estimation of the eigen value that only works for MDS or PCoA ordinations (not PCA).
+1: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (not PCA).
 
 2: Note that by default, the centroid is the centroid of the elements.
-It can however be fixed to an arbitrary value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
-
-### Equations and implementations:
-Some of the functions described below are directly implemented in the `dispRity` package and do not require any dependencies  ([see implementation here](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R)).
+It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
+
+### Equations and implementations
+Some of the functions described below are implemented in the `dispRity` package and do not require any other packages to calculate ([see implementation here](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R)).
 
@@
 \end{equation}
-With *k* being the number of dimensions, $\lambda_i$ is the eigen value of each dimension.
+Where *k* is the number of dimensions, and $\lambda_i$ is the eigenvalue of each dimension.
 
@@
 \end{equation}
-With *k* being the dimensions.
+Where *k* is the number of dimensions.
 
@@
 \end{equation}
-With *k* being the dimensions and $\sigma^{2}$ their variance.
+Where *k* is the number of dimensions, and $\sigma^{2}$ is their variance.
 
@@
 \end{equation}
-With *n* being each element in the ordinated space, *k* being its dimensions and $Centroid_{k}$ their mean (or can be set to an arbitrary value) and.
+Where *n* is each element in the ordinated space, *k* is the number of dimensions, and $Centroid_{k}$ is their mean (or can be set to another value).
 

### 2017-08-04 doc/dispRity-manual.Rmd
Fixed typos etc to 'testing hypotheses'

@@
 
- *  discrete time subsamples (or time-binning) using `method = discrete`
- *  continuous time subsamples (or time-slicing) using `method = continuous`
+ *  Discrete time subsamples (or time-binning) using `method = discrete`
+ *  Continuous time subsamples (or time-slicing) using `method = continuous`
 
@@
 
- *  defining the boundaries of the time bins (when `method = discrete`)
- *  defining the time slices (when `method = continuous`)
+ *  Defining the boundaries of the time bins (when `method = discrete`)
+ *  Defining the time slices (when `method = continuous`)
 
@@
 
-### Metrics run through
-Here is a run through of the main metrics implemented in `dispRity`.
-First, we will create a dummy ordinated space using the `space.maker` utility function (more about that [here](#space.maker):
+### Using the different disparity metrics
+Here is a brief demonstration of the main metrics implemented in `dispRity`.
+First, we will create a dummy/simulated ordinated space using the `space.maker` utility function (more about that [here](#space.maker):
 
@@
 
+We will use this simulated space to demonstrate the different metrics.
+
 #### Volumes and surface metrics
-The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `hyper.volume` all measure the surface or the volume of the ordinated space occupancy:
+The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `hyper.volume` all measure the surface or the volume of the ordinated space occupied:
 
@@
 ```
-> Because there is only one subsamples (i.e. one matrix) in the dispRity object, this operation is the equivalent of doing simply `ellipse.volume(dummy_space)` (with rounding).
+> Because there is only one subsample (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with rounding).
 
@@
 The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
-For the full and correct calculation of the volume of the ordinated space, it is possible to us the `hyper.volume` function that intakes more options but takes longer to calculate.
+For the full and correct calculation of the volume of the ordinated space, it is possible to use the `hyper.volume` function that has more options but takes longer to calculate.
 
@@
 ```
-> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [Curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) which often results in near 0 disparity values.
+> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) that often results in near 0 disparity values.
 
@@
 
-The functions `ranges`, `variances` and `diagonal` all measure properties of the ordinated space base on it's dimensions properties (they are also less affected by the Curse of dimensionality):
-
-`ranges` and `variances` both work on the same principle and measure the range/variance of each dimensions:
-
-```{r}
-## Calculating the ranges of each dimensions in the ordinated space
+The functions `ranges`, `variances` and `diagonal` all measure properties of the ordinated space based on its dimensional properties (they are also less affected by the "curse of dimensionality"):
+
+`ranges` and `variances` both work on the same principle and measure the range/variance of each dimension:
+
+```{r}
+## Calculating the ranges of each dimension in the ordinated space
 ranges(dummy_space)
@@
 
-## Calculating the variances of each dimensions in the ordinated space
+## Calculating the variances of each dimension in the ordinated space
 variances(dummy_space)
@@
 
-The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest euclidean distance in our 5 dimensional space):
+The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest Euclidean distance in our five dimensional space):
 
@@
 
-> This metric is only an euclidean diagonal (mathematically valid) if the dimensions within the space are all orthogonal!
-
-#### Centroids metrics
-
-The `centroids` metric allows to measure the position of the different elements compared to a fixed point in the ordinated space.
-By default, this function measures the distance between each elements and the central point between them:
-
-```{r}
-## The distribution of the distances between each elements and the centroid
+> This metric is only a Euclidean diagonal (mathematically valid) if the dimensions within the space are all orthogonal!
+
+#### Centroids metric
+
+The `centroids` metric allows users to measure the position of the different elements compared to a fixed point in the ordinated space.
+By default, this function measures the distance between each element and their centroid (centre point):
+
+```{r}
+## The distribution of the distances between each element and their centroid
 summary(dispRity(dummy_space, metric = centroids))
@@
 
-It is however possible to fix the coordinates of a specific point in the ordinated space as long as it has the correct number of dimensions:
-
-```{r}
-## The distance between the elements and the origin of the ordinated space
+It is however possible to fix the coordinates of the centroid to a specific point in the ordinated space, as long as it has the correct number of dimensions:
+
+```{r}
+## The distance between each element and the origin of the ordinated space
 summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))
 
-## Disparity as the distance between the elements and a specific point in space
+## Disparity as the distance between each element and a specific point in space
 summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
@@
 
-Because of its architecture, printing `dispRity` objects only summarise their content but does not print the disparity value measured or associated analysis (more about this [here](#manipulating-dispRity-objects)).
-To actually summary what's in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphic representation of the results.
+Because of its architecture, printing `dispRity` objects only summarises their content but does not print the disparity value measured or associated analysis (more about this [here](#manipulating-dispRity-objects)).
+To actually see what is in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphical representation of the results.
 
 ### Summarising `dispRity` data
-This function is a S3 function (`summary.dispRity`) allowing to summarise the content of `dispRity` objects that contain disparity calculations.
+This function is an S3 function (`summary.dispRity`) allowing users to summarise the content of `dispRity` objects that contain disparity calculations.
 
@@
 
-## Creating time slices subsamples
+## Creating time slice subsamples
 time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
@@
     FADLAD = BeckLee_ages)
-## Bootstrapping on the time binning/slicing subsamples
+## Bootstrapping the time slice subsamples
 boot_time_slices <- boot.matrix(time_slices, bootstraps = 100)
@@
 
-## Creating time bins subsamples
+## Creating time bin subsamples
 time_bins <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
@@
     inc.nodes = TRUE)
-## Bootstrapping on the time binning/slicing subsamples
+## Bootstrapping the time bin subsamples
 boot_time_bins <- boot.matrix(time_bins, bootstraps = 100)
@@
 
-This objects are easy to summarise as follow:
+These objects are easy to summarise as follows:
 
@@
 
-Information about the number of elements per subsamples as well as the observed (i.e. non-bootstrapped) disparity is also calculated.
-This is specifically handy when rarefying the data for example
+Information about the number of elements in each subsample and the observed (i.e. non-bootstrapped) disparity are also calculated.
+This is specifically handy when rarefying the data for example:
 
@@
 
- * the `quantile` values for the confidence intervals levels (by default, the 50 and 95 quantiles are calculated)
- * the `cent.tend` for the  central tendency to use for summarising the results (default is `median`)
- * the `rounding` option corresponding to the number of decimals to print (default is `2`)
- *  and the `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
+ * `quantile` values for the confidence interval levels (by default, the 50 and 95 quantiles are calculated)
+ * `cent.tend` for the central tendency to use for summarising the results (default is `median`)
+ * rounding` option corresponding to the number of decimal places to print (default is `2`)
+ * `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
 
@@
 ```{r}
-## Same but using the 88 quantile and the standard deviation as the summary. 
+## Same as above but using the 88th quantile and the standard deviation as the summary 
 summary(disparity_time_slices, quantile = 88, cent.tend = sd)
 
-## Printing the details of the object and rounding the values to the 5th decimal
+## Printing the details of the object and rounding the values to the 5th decimal place
 summary(disparity_time_slices, recall = TRUE, rounding = 5)
@@
 
-Information about the number of elements per subsamples as well as the observed (i.e. non-bootstrapped) disparity is also calculated.
-This is specifically handy when rarefying the data for example
-
-Note that the summary table is a `data.frame`, hence it is easy to modify as any table using `dplyr`.
-You can also export it in `csv` format using `write.csv` or even directly export into `LaTeX` format using the following;
+Note that the summary table is a `data.frame`, hence it is as easy to modify as any dataframe using `dplyr`.
+You can also export it in `csv` format using `write.csv` or `write_csv` or even directly export into `LaTeX` format using the following;
 
@@
 An alternative (and more fun!) way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
-This function intakes the same options as `summary.dispRity` along side with various graphical options described in the function manual (see `?plot.dispRity`).
+This function takes the same options as `summary.dispRity` along with various graphical options described in the function help files (see `?plot.dispRity`).
 
@@
  * `continuous` for displaying continuous disparity curves
- * and `box`, `lines` and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines and confidence interval polygons.
-
-> This argument can be left missing. In this case, the algorithm will automatically detect the type of subsamples from the `dispRity` object.
-
-It is also possible to display the number of elements per subsamples (has a horizontal doted line) using the option `elements = TRUE`.
-Additionally, when the data is rarefied, one can also indicate which level of rarefaction to display (i.e. display the results for a certain number of elements) by using the `rarefaction` argument.
+ * `box`, `lines`, and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.
+
+> This argument can be left empty. In this case, the algorithm will automatically detect the type of subsamples from the `dispRity` object and plot accordingly.
+
+It is also possible to display the number of elements in each subsample (as a horizontal dotted line) using the option `elements = TRUE`.
+Additionally, when the data is rarefied, one can indicate which level of rarefaction to display (i.e. only display the results for a certain number of elements) by using the `rarefaction` argument.
 
@@
 ## Graphical parameters
-op <- par(mfrow = c(2,2), bty = "n")
+op <- par(mfrow = c(2, 2), bty = "n")
 
@@
 
-## Same but using lines for the rarefaction level of 20 elements only
+## As above but using lines for the rarefaction level of 20 elements only
 plot(disparity_crown_stem, type = "line", rarefaction = 20)
 
-## Same but using polygons while also displaying the number of elements
+## As above but using polygons while also displaying the number of elements
 plot(disparity_crown_stem, type = "polygon", elements = TRUE)
@@
 
-Additionally to the classic `plot` arguments, the function can also take arguments that are specific to `plot.dispRity` like adding the number of elements or rarefaction level (as described above) but also changing the values of the quantiles to plot as well as the central tendency.
+In addition to the classic `plot` arguments, the function can also take arguments that are specific to `plot.dispRity` like adding the number of elements or rarefaction level (as described above), and also changing the values of the quantiles to plot as well as the central tendency.
 
@@
 ## Plotting the results with some plot.dispRity arguments
-plot(disparity_time_slices, quantile = c(seq(from=10, to=100, by=10)),
+plot(disparity_time_slices, quantile = c(seq(from = 10, to = 100, by = 10)),
     cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
@@
 
-## Testing disparity hypothesis
+## Testing disparity hypotheses
 
@@
 ## Customised sorting
-head(summary(sort(disparity, sort = c(7,1,3,4,5,2,6))))
+head(summary(sort(disparity, sort = c(7, 1, 3, 4, 5, 2, 6))))
 ```

### 2017-08-08 paper/dispRity_paper.tex
paper intro updated

@@
 
-\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$}\\
+\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$, Mark Puttick$^2$, Natalie Cooper$^3$}\\
 \noindent {\small \it 
@@
 \begin{enumerate}
-    \item I present \texttt{dispRity}, an \texttt{R} package form measuring disparity from ordinated matrices. 
+    \item We present \texttt{dispRity}, an \texttt{R} package in multidimensional spaces. 
     \item Disparity designates a suit of metrics to describe an ordinated matrix (also referred as morpho/eco/niche/etc.-space).
@@
 \section{Introduction}
-Disparity = multidimensionality
 
-Disparity can be measured in many ways in palaeobiology (cite many examples) or in ecology (donohue and stuff).
+Multidimensionality is a major aspect of natural sciences were variables co-vary through time and space from the organism to the environmental level.
 
-Although some packages have disparity components (Claddis, geomorph, adonis), all have their own definition of disparity and makes the whole work idiosuncratic
+In palaeobiology multidimensionality is a popular way to study morphological diversity and comes as a interesting alternative to unidimensional taxonomic diversity analysis \citep{Hopkins2017}.
 
-Therefore I made this package to help measuring disparity
+Although the concept of morphological disparity is well defined since more than two decades \citep[e.g.][]{gould1991disparity,Foote01071994,Foote29111996}, its specific definition varies widely among authors.
+In fact disparity Disparity is by nature multidimensional \citep{lloyd2016estimating} and thus there are many ways to measure it.
+
+Hereafter we will make a clear distinction between the multidimensional space which is a specific mathematical object and the disparity, which is a metric describing or summarising one or more aspects of this space.
+Both can be defined in various ways depending on the authors, for example, disparity is defined as the weighted mean pairwise dissimilarity in \cite{Close2015} or the ellipsoid volume in \cite{DonohueDim} and the multidimensional space is defined as the morphospace in \cite{raup1966geometric} and the morpho-functional space in \cite{diaz2016global}.
+
+The multidimensional space can be defined in many ways and arise from many mathematical transformations of the data such as the pairwise distance matrix \citep{Close2015}, a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, a multidimensional scaling \citep[MDS;][]{DonohueDim}, etc.
+Similarly, disparity metrics \citep[or indices;][]{Hopkins2017} can defined in many ways \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015,diaz2016global}.
+Finally, difference between disparity metrics can also be measured in many ways: using NPMANOVA \citep[e.g.][]{Brusatte12092008}, multidimensional permutation test \citep[e.g.][]{diaz2016global} or even simple confidence interval overlap \citep[e.g.][]{halliday2016eutherian}.
+
+This variety of definitions and analysis have been developed in an equal variety of softwares such as \texttt{GINGKO} in javascript \citep{bouxin2005ginkgo,de2007ginkgo} or \texttt{geomorph} \citep{adams2013geomorph,adams2017geometric}, \texttt{Claddis} \citep{Claddis}, or \texttt{vegan} \citep{oksanen2007vegan} in R \citep{R}.
+This results in the need to learn different languages (or at least - when restricted to R - different packages with different standards) as well as making analysis sometimes idiosyncratic and often complex to repeat since they are based on a particular feature from a particular software.
+For example, in the excellent and widely used \texttt{geomorph} package morphological disparity analysis can be ran using the \texttt{morphol.disparity} function.
+Unfortunately, however, the multidimensional space can only be defined as the ordination of the procrustes transform of geometric morphometric landmarks, the disparity can only be defined as the Procrustes variance and the difference between groups can only be measured through permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
+
+This package, is the first, to our knowledge, to be entirely dedicate to multidimensional analysis.
+It has a highly modular architecture and allow users to define their own multidimensional space, use their own definition of the disparity metric and measure the differences between metrics in their own terms.
+Furthermore, both a specific design and set of functions (through the \texttt{dispRity} object) allows to design analysis into easy and repeatable pipelines.
 

### 2017-08-09 doc/dispRity-manual.Rmd

@@
 
-The `dispRity` package allows to apply some test functions to the calculated disparity in order to test hypothesis.
-The function `test.dispRity` works in a similar way as the `dispRity` function: it intakes a `dispRity` object, a `test` and a `comparisons` argument.
-
-The `comparisons` argument must indicate the way the test should be applied to the data:
-
- * `pairwise` (default): to compare each subsamples pairwise
- * `referential`: to compare each subsamples to the first one
- * `sequential`: to compare each subsamples to the following one
+The `dispRity` package allows users to apply statistical tests to the calculated disparity to test various hypotheses.
+The function `test.dispRity` works in a similar way to the `dispRity` function: it takes a `dispRity` object, a `test` and a `comparisons` argument.
+
+The `comparisons` argument indicates the way the test should be applied to the data:
+
+ * `pairwise` (default): to compare each subsample in a pairwise manner
+ * `referential`: to compare each subsample to the first subsample
+ * `sequential`: to compare each subsample to the following subsample
  * `all`: to compare all the subsamples together (like in analysis of variance)
 
-It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsamples names to create personalised test.
-Some other specific tests implemented in `dispRity` such as the `dispRity::null.test` that has a specific way to be applied to the data and therefore ignore the `comparisons` argument. 
+It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsample names to create personalised tests.
+Some other tests implemented in `dispRity` such as the `dispRity::null.test` have a specific way they are applied to the data and therefore ignore the `comparisons` argument. 
 <!-- Add sequential test one day! -->
@@
 The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
-It can be common statistical test functions (e.g. `stats::t.test`), some functions implemented in `dispRity` (e.g. see `?null.test`) or any function defined by the user.
-
-This function also allows to correct for type I error inflation when using multiple comparisons via the `correction` argument.
+It can be a common statistical test function (e.g. `stats::t.test`), a function implemented in `dispRity` (e.g. see `?null.test`) or any function defined by the user.
+
+This function also allows users to correct for Type I error inflation (false positives) when using multiple comparisons via the `correction` argument. 
 This argument can be empty (no correction applied) or can contain one of the corrections from the `stats::p.adjust` function (see `?p.adjust`).
 
-Note that the `test.dispRity` algorithm deals with some classic tests outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
-It is however possible to get the full detailed output by using the options `details = TRUE`.
-
-```{r}
-## T-test to the test a difference in disparity between crown and stem mammals
+Note that the `test.dispRity` algorithm deals with some classical test outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
+It is, however, possible to get the full detailed output by using the options `details = TRUE`.
+
+```{r}
+## T-test to test for a difference in disparity between crown and stem mammals
 test.dispRity(disparity_crown_stem, test = t.test)
@@
 
-## Wilcoxon test applied to sliced disparity the data sequentially with correction
+## Wilcoxon test applied to time sliced disparity with sequential comparisons, with Bonferroni correction
 test.dispRity(disparity_time_slices, test = wilcox.test,
@@
 
-It is also possible to apply some more *complex* tests that have their own output class (like `stats::aov`).
+It is also possible to apply some more *complex* tests that have their own output classes (like `stats::aov`).
+
+<!-- NC: again this  needs to be lm, aov is not appropriate. Will let you fix this -->
 The results can then be analysed as usual using the associated `summary` S3 method:
@@
 
-## That can be summarised using summary
+## This output can be summarised using summary
 summary(slice_aov)
@@
 
-Of course, following the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
+Of course, due to the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
 The only condition is that the test can be applied to at least two distributions.
-In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `aov` for example).
+In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm` for example).
 
@@
 For example, the sum of variances, the product of ranges or the median distance between the elements and their centroid will summarise disparity as a single value.
-This value can be of course pseudo-replicated (bootstrapped) in order to obtain a more confident distribution of this summary metric.
-However, another way to perform disparity analysis is to use the whole distribution rather than their summary (e.g. the variances or the ranges).
+This value can be pseudo-replicated (bootstrapped) to obtain a distribution of the summary metric with estimated error.
+However, another way to perform disparity analysis is to use the *whole distribution* rather than just a summary metric (e.g. the variances or the ranges).
 
 This is possible in the `dispRity` package by calculating disparity as a dimension-level 2 metric only!
-Let's have a look using our previous example of bootstrapped time slices but by measuring the distances between each taxa and their centroid as disparity
+Let's have a look using our previous example of bootstrapped time slices but by measuring the distances between each taxon and their centroid as disparity.
+
+<!-- NC: I think when these prelim steps have been done a long time earlier, and/or are not built into the package, you probably want to include the code to get boot_time_slices here too.
+Most people won't read the manual cover to cover in one go -->
 
@@
 
-The resulting disparity object being a dimension-level 2, it can easily also be transformed into a dimension-level 1 one by for example measuring the median distance of all these distributions:
+The resulting disparity object being of dimension-level 2, it can easily be transformed into a dimension-level 1 one by, for example, measuring the median distance of all these distributions:
 
@@
 
-And we can now compare the differences between both methods:
+And we can now compare the differences between these methods:
 
@@
 ## Graphical parameters
-op <- par(bty = "n", mfrow = c(1,2))
+op <- par(bty = "n", mfrow = c(1, 2))
 
@@
 
-The resulted distributions can then be studied the same way as described above.
+We can then test for differences in the resulting distributions using `test.dispRity` and the `bhatt.coeff` test as described above.
 
@@
 
-In this case, we are looking at the probability of overlap of the distribution of median distances from centroids between each pair of time slices.
+In this case, we are looking at the probability of overlap of the distribution of median distances from centroids among each pair of time slices.
 In other words, we are measuring whether the medians from each bootstrap pseudo-replicate for each time slice overlap.
@@
 
-One solution is thus to look at the full distribution:
+One solution, therefore, is to look at the full distribution:
 
@@
 
-These results show the actual overlap between all the measured distances from centroids concatenated across all the bootstraps.
-For example, when comparing the slices 120 and 80, we are effectively comparing the 5 \times 100 distances (the distances of the 5 elements in slice 120 bootstrapped 100 times) to the 19 \times 100 distances from slice 80.
+These results show the actual overlap among all the measured distances from centroids concatenated across all the bootstraps.
+For example, when comparing the slices 120 and 80, we are effectively comparing the 5 \times 100 distances (the distances of the five elements in slice 120 bootstrapped 100 times) to the 19 \times 100 distances from slice 80.
 However, this can also be problematic for some specific tests since the _n_ \times 100 distances are also pseudo-replicates and thus are still not independent.
 
-One second solution is to compare the distributions to each other _per replicates_:
+A second solution is to compare the distributions to each other _for each replicate_:
 
@@
 
-These results show the median overlap between pairs of distributions in the first column (`bhatt.coeff`) and then the distribution of these overlaps between each par of bootstraps.
+These results show the median overlap among pairs of distributions in the first column (`bhatt.coeff`) and then the distribution of these overlaps among each pair of bootstraps.
 In other words, when two distributions are compared, they are now compared for each bootstrap pseudo-replicate, thus effectively creating a distribution of probabilities of overlap.
-For example, when comparing the slices 120 and 80, we have a mean probability of overlap of 0.28 and a probability between 0.18 and 0.43 in 50% of the pseudo-replicates.
-Note that the quantiles and central tendencies can be modified via the `conc.quantiles` option:
-
-## Ecological study
-
-Disparity analysis are not restricted to palaeobiology analysis only.
+For example, when comparing the slices 120 and 80, we have a mean probability of overlap of 0.28 and a probability between 0.18 and 0.43 in 50\% of the pseudo-replicates.
+Note that the quantiles and central tendencies can be modified via the `conc.quantiles` option.
+
+## Ecological studies of disparity
+
+Disparity analyses are not restricted to palaeobiology analysis only.
 This package is designed for dealing with ordinated matrix, no matter which data was used in the ordination.

### 2017-08-09 doc/dispRity-manual.Rmd
Updates following Natalie's issues

@@
 
-Disparity must be measured from ordinated matrices.
+Disparity can be estimated from pretty much any matrix.
+Classically, however, it is measured from ordinated matrices.
 These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).
@@
 
-<!-- I removed the semicolons above, it is confusing to novice R users -->
-
-Of course you can use your own data as detailed in the [previous chapter](#which-data).
-<!-- I changed the header so may have broken this link -->
+Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).
 
@@
 
-We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `aov`):
+We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):
 
@@
 ## Testing for an difference among the time bins
-disp_aov <- test.dispRity(disparity_data, test = aov, comparisons = "all")
-summary(disp_aov)
+disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
+summary(disp_lm)
 
@@
 
- * An ordinated matrix (always!)
+ * An matrix with rows as elements and columns as dimensions (always!)
  * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix.
@@
 
-The function `time.subsamples` allows users to divide the ordinated matrix into different time subsamples or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
+The function `time.subsamples` allows users to divide the matrix into different time subsamples or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
 Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.
@@
 
-For the time-slicing method details see @GuillermeSTD. 
-<!-- NC: Or potentially this paper we are writing for the PalAss? -->
+For the time-slicing method details see @GuillermeSTD. <!-- @@@ Change cite appropriately! -->
+<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
 For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:
@@
 
- *  `acctran` where the data chosen along the branch is always the one of the offspring
- *  `deltran` where the data chosen along the branch is always the one of the descendant
- *  `punctuated` where the data chosen along the branch is randomly chosen between the offspring or the descendant
- *  `gradual` where the data chosen along the branch is either the offspring or the descendant depending on branch length
-
-<!--NC: Bit puzzled by the offspring/descendent terminology - I'd assume these were the same thing. By descendant do you mean ancestor? I'm confused  -->
+ *  `acctran` where the data chosen along the branch is always the one of the descendant
+ *  `deltran` where the data chosen along the branch is always the one of the ancestor
+ *  `punctuated` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
+ *  `gradual` where the data chosen along the branch is either the descendant or the ancestor depending on branch length
 
@@
 
-Another way of separating elements into different categories is to use customised subsamples as briefly explained [above](#disparity-between-groups).
-This function simply takes the list of elements to put in each group (whether they are the actual element names or their idea).
-
-<!-- NC: Their idea? Huh? Whats that mean? -->
+Another way of separating elements into different categories is to use customised subsamples as briefly explained [above](#disparity-among-groups).
+This function simply takes the list of elements to put in each group (whether they are the actual element names or their position in the matrix).
 
@@
 
-<!-- NC: This is confusing to me. If it's a number > 0, it keeps that number of dimensions, and if < 0 it removes the last dimensions by that %? This is very odd. Please explain -->
-
 ```{r, eval=TRUE}
-## Removing the last 50% of the dimensions
+## Using the first 50% of the dimensions
 boot.matrix(BeckLee_mat50, dimensions = 0.5)
 
-## Keeping the first 10 dimensions
+## Using the first 10 dimensions
 boot.matrix(BeckLee_mat50, dimensions = 10)
@@
 
-Of course, one could directly supply the subsamples generated above (using <!--  NC: which function ? How is this line different from the statement in the line below? -->) to this function.
-It can also deal with a list of matrices or with a `dispRity` object output from the `custom.subsamples` or `time.subsamples` functions.
+Of course, one could directly supply the subsamples generated above (using `time.subsamples` or `custom.subsamples`) to this function.
 
@@
                                FADLAD = BeckLee_ages)
+
 ## Bootstrapping the time slice subsamples
@@
 In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
-For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @finlay2015morphological.
-<!-- NC: probably should cite Matt Wills' paper for the centroid thing too as he invented it apparently -->
+For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @Wills1994.
+<!-- NC: probably should cite Matt Wills' paper for the centroid thing too as he invented it apparently; TG: ok fixed.
+Though one can argue Euclides invented it before him (and probably Gnut-the-enlighted-cavemen before-->
 Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.
@@
 
-<!-- NC: We discussed this previously, but I don't like the term levels here.
-It's confusing because it means something specific wrt factors.
-Can we change it to dimension-level?
-That's more descriptive and clearer in my opinion.
-I have changed it below, see what you think -->
-
 ### The function dimension-levels
@@
 
-<!-- NC: My terminology changes may have broken this bit -->
-
 ```{r}
 ## Testing whether mean is dimension-level 1
-if(make.metric(mean, silent = TRUE) != "dimension-level1") {
+if(make.metric(mean, silent = TRUE) != "level1") {
     message("The metric is not dimension-level 1.")
@@
 ## Testing whether var is dimension-level 1
-if(make.metric(var, silent = TRUE) != "dimension-level1") {
+if(make.metric(var, silent = TRUE) != "level1") {
     message("The metric is not dimension-level 1.")
@@
 
-Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > dimension-level 2 > dimension-level 1).
+Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > 2 > 1).
 
@@
 
-It is also possible to apply some more *complex* tests that have their own output classes (like `stats::aov`).
-
-<!-- NC: again this  needs to be lm, aov is not appropriate. Will let you fix this -->
+It is also possible to apply some more *complex* tests that have their own output classes (like `stats::lm`).
+
 The results can then be analysed as usual using the associated `summary` S3 method:
@@
 ```{r}
-## Performing and ANOVA applied to the same data
-(slice_aov <- test.dispRity(disparity_time_slices, test = aov,
+## Performing and linear model applied to the same data
+(slice_lm <- test.dispRity(disparity_time_slices, test = lm,
                             comparisons = "all"))
 
-## The output is a regular ANOVA output
-class(slice_aov)
+## The output is a regular `lm` output
+class(slice_lm)
 
 ## This output can be summarised using summary
-summary(slice_aov)
+summary(slice_lm)
 ```

### 2017-08-09 R/make.metric.R

@@ -1,41 +1,41 @@
 #' @title Creating disparity metrics
 #'
-#' @description Testing the level of disparity metrics
+#' @description Testing the dimension-level of disparity metrics
 #'
 #' @param fun Your very own \code{function}!
 #' @param ... Some arguments to be passed to \code{fun}.
-#' @param silent \code{logical}; if \code{FALSE} (default), the function will be verbose and give no output; if \code{TRUE}, the function will only output the function level.
+#' @param silent \code{logical}; if \code{FALSE} (default), the function will be verbose and give no output; if \code{TRUE}, the function will only output the function's dimension-level.
 #'
 #' @details
 #' This function tests:
 #' \itemize{
 #'   \item 1: if your function can deal with a matrix as an \code{input}.
-#'   \item 2: which level is your function (level 1, 2 or 3, see \code{\link{dispRity.metric}}).
+#'   \item 2: which dimension-level is your function (1, 2 or 3, see \code{\link{dispRity.metric}}).
 #'   \item 3: whether the function can properly be implemented in the \code{dispRity} function.
 #' }
 #' The 3 different metric levels correspond to the dimensions of the output and are:
 #' \itemize{
-#'   \item "level 1": for functions that decompose a \code{matrix} into a single value.
-#'   \item "level 2": for functions that decompose a \code{matrix} into a \code{vector}.
-#'   \item "level 3": for functions that transform the \code{matrix} into another \code{matrix}.
+#'   \item "dimension-level 1": for functions that decompose a \code{matrix} into a single value.
+#'   \item "dimension-level 2": for functions that decompose a \code{matrix} into a \code{vector}.
+#'   \item "dimension-level 3": for functions that transform the \code{matrix} into another \code{matrix}.
 #' }
-#' For example, the disparity metric \code{\link[base]{sum}} of \code{\link[dispRity]{variances}} is composed of two metric levels:
+#' For example, the disparity metric \code{\link[base]{sum}} of \code{\link[dispRity]{variances}} is composed of two metric dimension-levels:
 #' \itemize{
-#'   \item The \code{\link[dispRity]{variances}} (level 2) that calculates the variances per column in a matrix (aggregates a \code{matrix} into a \code{vector}).
-#'   \item The \code{\link[base]{sum}} (level 1) that transform the \code{vector} of variances into a single value.
[…]

### 2017-08-09 doc/dispRity-manual.Rmd
updated vignette

@@
 
-## The volume with different options to be passed to hypervolume::hypervolume
-summary(dispRity(dummy_space, metric = hyper.volume, repsperpoint = 100,
-                 quantile = 0.5))
 ```

### 2017-08-10 doc/dispRity-manual.Rmd
Removed comments that have been dealt with including amusingly snarky replies

@@
 
-<!-- NC: Is this using aov or lm?
-Note that aov can only be used with balanced designs, so where the three time bins have equal numbers of species.
-I expect you need to use lm (which also fits an ANOVA as an ANOVA is also a linear model).
-THIS IS IMPORTANT -->
-
 Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
@@
 For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @Wills1994.
-<!-- NC: probably should cite Matt Wills' paper for the centroid thing too as he invented it apparently; TG: ok fixed.
-Though one can argue Euclides invented it before him (and probably Gnut-the-enlighted-cavemen before-->
 Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.
@@
 
-<!-- NC: I think when these prelim steps have been done a long time earlier, and/or are not built into the package, you probably want to include the code to get boot_time_slices here too.
-Most people won't read the manual cover to cover in one go -->
-
 ```{r}
@@
 
-The resulting disparity object being of dimension-level 2, it can easily be transformed into a dimension-level 1 one by, for example, measuring the median distance of all these distributions:
+The resulting disparity object is of dimension-level 2, so it can easily be transformed into a dimension-level 1 object by, for example, measuring the median distance of all these distributions:
 
@@
 
-Disparity analyses are not restricted to palaeobiology analysis only.
-This package is designed for dealing with ordinated matrix, no matter which data was used in the ordination.
-Here is an short example of disparity analysis using ecological data from D. McClean (unpubl.).
+Disparity analyses are not restricted to palaeobiology analysis.
+This package is designed for dealing with ordinated matrices, no matter which data was used in the ordination.
+Here is an short example of disparity analysis using ecological data from Deirdre McClean (unpubl.).
 
 The dataset contains an ordinated matrix of 20 dimensions (columns) for 40 elements (rows).
-The elements are different experimental plots with discrete variations of nutrient enrichment and depth and freshwater benthic invertebrates.
-
+The elements are different experimental plots with discrete variations of nutrient enrichment, depth, and freshwater benthic invertebrates.
+<!-- NC: Explain a little more about what the data are here as I don't entirely get it. EG what's a treatement etc? Just a really simple explanataion is perfect -->
 ```{r}

### 2017-08-10 doc/dispRity-manual.Rmd
Edits down to simulating discrete data

@@
 ## Ecological studies of disparity
+<!-- NC: I wonder if this might be better just as a "hey we can do ecology stuff too and here is a link to a demo" rather than a new section?
+The manual is loooooooooong so splitting some stuff out might be good?
+-->
 
@@
 <!-- NC: Explain a little more about what the data are here as I don't entirely get it. EG what's a treatement etc? Just a really simple explanataion is perfect -->
+
 ```{r}
@@
 
-### Classic analysis
-
-A classical way to represent this ordinated data will be to use two dimensional plots.
+### Classic ecological "disparity" analysis
+
+A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different experimental plots differ in their values for nutrient enrichment, depth, and freshwater benthic invertebrates.
+<!-- NC: or similar just to make it clearer -->
 
@@
 
-This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix (i.e. the first two dimensions).
-At a first glance, it seems difficult to see a clear effect from the treatments (blue or red) or the depth (rounds or triangles) since the different experimental plots with the same parameters don’t seem to cluster.
-However, the problem, is that this plot ignores the 18 other dimensions of the ordination!
+This shows the distribution of the experimental plots along the two first axis of variation of the ordinated distance matrix (i.e. the first two dimensions).
+At a first glance, it seems difficult to see a clear effect of the treatments (blue or red) or the depth (rounds or triangles) as the different experimental plots with the same parameters don't seem to cluster together.
+However, this may be because this plot ignores the 18 other dimensions of the ordination!
 Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e.
-the values on the first dimension do not represent some continuous traits (e.g.
-depth) but just the ordinations of correlations between the data and some factors.
+the values on the first dimension do not represent a continuous trait (e.g.
+depth), instead they just represent the ordinations of correlations between the data and some factors.
 
@@
 
-#### Splitting the data following the different factors
-
-The first step in such analysis will be to create different subsamples that will represent sub-samples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
-Each of these subsamples contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
+#### Splitting the data with respect to the different factors
+
+The first step is to create different subsamples that represent subsamples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
+Each of these subsamples will contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
 In our example, we are going to group the elements according to their depth and treatment.
@@
 ```{r}
-## Creating the table that contain the elements and their attributes
+## Creating a table that contain the elements and their attributes
 factors <- as.data.frame(matrix(data = c(treatments, depth),
@@
 
-Second, let’s split the data according to these factors to create the subsamples of the ordinated space by using the `custom.subsamples` function:
+Second, let’s split the data according by depth and treatment to create the subsamples of the ordinated space by using the `custom.subsamples` function:
 
@@
 
-We can also want to bootstrap the data to test the robustness of the measured disparity to outliers.
-Also, as we can see, each subsamples has different numbers of elements.
-It might also be interesting to rarefy the data as well to have only subsamples with the same number of elements.
+We can also bootstrap the data to test the robustness of the measured disparity to outliers.
+Also, as we can see, each subsample has different numbers of elements.
+It might also be interesting to rarefy the data to have only subsamples with the same number of elements.
 
@@
 
-We can now rerun a more robust disparity analysis using the bootstrapped data:
+We can now run a more robust disparity analysis using the bootstrapped data:
 
@@
 ## Graphical options
-op <- par(mfrow = (c(1,2)), bty = "n")
+op <- par(mfrow = (c(1, 2)), bty = "n")
 
@@
 
-As we can see, there seems to be no strong effect of the number of experimental plots in each subsamples (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
-
-#### Summarising hypothesis
-
-Finally, we can test our hypothesis (whether the sort of water treatment at certain depth alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.
-
-```{r}
-## Testing the effect of our factors on the bootstrapped data
+As we can see, there seems to be no strong effect of the number of experimental plots in each subsample (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
+
+#### Testing our hypotheses
+
+Finally, we can test our hypothesis (whether the water treatment at certain depths alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.
+<!-- Ah so this is what you are testing! Make this clearer in the intro bit to this section -->
+
+```{r}
+## Testing the effect of treatment and depth on the bootstrapped data
 summary(test.dispRity(disparity, test = aov, comparisons = "all"))
@@
 
-The `dispRity` package also allows to simulate discrete morphological data matrices.
+The `dispRity` package also allows users to simulate discrete morphological data matrices.
 

### 2017-08-10 doc/dispRity-manual.Rmd
Edits to package guts

@@
 The `dispRity` package also allows users to simulate discrete morphological data matrices.
-
-### Quick go through
-The function basically intakes a phylogenetic tree, the number of required characters and the evolutionary model and a function from which to draw the rates.
-The package then proposes a function for quickly checking the matrix' phylogenetic signal using parsimony.
+In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
+The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
+To understand these methods please refer to the phylogenetics literature.
+<!-- NC: I wonder if you need to cite some papers here to explain how people can find out what a suitable model is etc? Or do we just assume everyone knows about making phylogenies? -->
 
@@
 set.seed(3)
-## Setting the starting tree with 15 taxa as a random coalescent tree
+## Simulating a starting tree with 15 taxa as a random coalescent tree
 my_tree <- rcoal(15)
 
-## Generating a matrix with 100 characters (85% binary and 15% three states) and
-## an equal rates model with a gamma rates distribution (0.5, 1) with no 
+## Generating a matrix with 100 characters (85% binary and 15% three state) and
+## an equal rates model with a gamma rate distribution (0.5, 1) with no 
 ## invariant characters.
@@
 
-## The head of the matrix
+## The first few lines of the matrix
 my_matrix[1:5, 1:10]
@@
 
-Note that this example produces a tree with a great consistency index and an identical topology as the random coalescent tree!
+Note that this example produces a tree with a great consistency index and an identical topology to the random coalescent tree!
 Nearly too good to be true...
 
-### In details
-
-Basically, `sim.morpho` is really flexible and intakes a lot of different arguments to allow to simulate realistic matrix.
-It has three implemented models: `"ER"` for Equal Rates (the M*k* model); the `"HKY"` one which is the molecular HKY model but transforms pyrines in 0's and purimidines in 1's; or the `"mixed"` model that randomly uses an `"ER"` or and `"HKY"` for the binary characters and `"ER"` for the multistates (>2) characters.
-Both models intakes specific distributions for their rates or substitution models.
+### A more detailed descriptions
+
+`sim.morpho` is really flexible and takes many different arguments to allow users to simulate realistic matrices.
+It has three implemented models: `"ER"` for Equal Rates (the M*k* model); `"HKY"`, the molecular HKY model but transforms pyrines in 0's and pyrimidines in 1's; or the `"mixed"` model that randomly uses an `"ER"` and/or `"HKY"` for the binary characters and `"ER"` for the multistate (>2) characters.
+Both models take specific distributions for their rate or substitution models.
 These distributions should be passed to these arguments in the format of `c(sampler_function, distribution_parameters)` where the the sampler function is a the random generation function of that distribution (e.g.
@@
 
-The `check.morpho` runs a quick maximum parsimony tree using the `phangorn` parsimony algorithm.
+`check.morpho` runs a quick Maximum Parsimony tree search using the `phangorn` parsimony algorithm.
 It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g.
-the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree.
-
-### Parameters for a good(ish) matrix
-There are many parameters that can realise a ``realistic'' matrix (meaning not to different from the input tree with a consistency and retention index close to what's in the literature) but because of the randomness of the generation (the code generates random matrices after all!) not all end up in creating good matrices.
-The following parameters however, seems to generate a fairly ``realistic'' with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three states characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
+the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree to determine how different they are.
+
+### Parameters for a realistic(ish) matrix
+There are many parameters that can create a ``realistic'' matrix (i.e.
+not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all end up creating "good" matrices.
+The following parameters however, seem to generate fairly ``realistic'' matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
+<!-- NC: Check that I haven't changed what you mean here where i swapped good for realistic. Realistic is a better term. Or be clearer what you mena by "good" -->
 
@@
 Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
-This function allows to simulated multidimensional spaces given a certain number of properties.
-It takes as arguments the number of elements (data points) and dimensions to create the space but also allows more fine tuning data simulation: it is possible to give a specific distribution to each dimensions, provide a correlation matrix to link the dimensions or even providing an *a priori* distribution of the variance per distributions!
+This function allows users to simulate multidimensional spaces with a certain number of properties.
+It takes as arguments the number of elements (data points) and dimensions to create the space but also permits more fine tuning in the data simulation: it is possible to give a specific distribution to each dimension, provide a correlation matrix to link the dimensions or even provide an *a priori* distribution of the variance for each distributions!
+<!-- ooh fancy! -->
 
@@
 ## Graphical options
-op <- par(mfrow = (c(2,2)), bty = "n")
+op <- par(mfrow = (c(2, 2)), bty = "n")
 ## Visualising 3D spaces
@@
 ## An ellipsoid space (=a spheric space with correlation)
-cor_matrix <- matrix(cbind(1,0.8,0.2, 0.8,1,0.7, 0.2,0.7,1), nrow = 3)
+cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3)
 scatterplot3d(space.maker(2500, 3, rnorm, cor.matrix = cor_matrix), pch = 20,

### 2017-08-10 doc/dispRity-manual.Rmd
Finish edits to manual

@@
 
-# The package's guts
+# The dispRity package's guts
 
@@
 
-The package also provides some utilities functions to facilitate multidimensional analysis.
+The package also provides some utility functions to facilitate multidimensional analysis.
 
 ### `dispRity` object utilities  <a name="dispRity.utilities"></a>
-The first subsamples of utilities are functions for manipulating `dispRity` objects:
+The first set of utilities are functions for manipulating `dispRity` objects:
 
@@
 ####  `fill.dispRity`
-This function initialises a `dispRity` object and generating it's call properties.
+This function initialises a `dispRity` object and generates its call properties.
 
@@
 
-## The dipRity object has now the proper minimal attributes
+## The dipRity object has now the correct minimal attributes
 disparity_obj$call
@@
 ####  `matrix.dispRity`
-This function extracts a specific matrix from a disparity object. The matrix can be one of the bootstrapped or/and rarefied one.
+This function extracts a specific matrix from a disparity object.
+The matrix can be one of the bootstrapped matrices or/and a rarefied matrix.
 
@@
 
-## Extracting the 3rd boostrapped matrix with the 2nd rarefaction level
+## Extracting the 3rd bootstrapped matrix with the 2nd rarefaction level
 ## (15 elements) from the second group (80 Mya)

### 2017-08-10 doc/dispRity-ecology-demo.Rmd
Minor edits to reflect changes in main manual

@@
 
-This demo aims to give quick overview of the `dispRity` package (v.0.3) for ecological analysis.
-Please refer to [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the `dispRity` manual that explains the functions in more details.
+This demo aims to give a quick overview of the `dispRity` package (v.0.3) for multidimensional ecological analyses.
+Please refer to [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the `dispRity` manual that explains the functions in more detail.
 
-To keep it short, this package allows the use of all of the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
-For example, we might want to know whether a specific kind of water treatment at certain water depth alters invertebrate communities and composition in natural habitats.
+To keep it short, this package allows user to include **all** the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a subset of dimensions.
+<!-- For example, we might want to know whether a specific kind of water treatment at certain water depth alters invertebrate communities and composition in natural habitats. -->
 
@@
 For this example with ecological data we are going to use data from D. McClean (unpubl.).
-This data is the ordination of a distance matrix based on nutrient enrichment and depth and freshwater benthic invertebrates.
+The dataset contains an ordinated matrix of 20 dimensions (columns) for 40 elements (rows).
+The elements are different experimental plots with discrete variations of nutrient enrichment, depth, and freshwater benthic invertebrates.
+<!-- NC: Explain a little more about what the data are here as I don't entirely get it. EG what's a treatement etc? Just a really simple explanataion is perfect -->
 
@@
 
-## As well as two list of different factors affecting each experimental plot: treatment and depth.
+## Create lists of the different factors affecting each experimental plot: treatment and depth.
 treatments <- McClean_data$treatment
@@
 
-A classical way to represent this ordinated data will be to use two dimensional plots.
+A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different experimental plots differ in their values for nutrient enrichment, depth, and freshwater benthic invertebrates.
+<!-- NC: or similar just to make it clearer -->
 
@@
 ## A classic 2D ordination plot
-plot(x, y, col = cols, pch = pchs, xlab = "PC 1", ylab = "PC 2", xlim = range(x) + c(0, 100))
+plot(x, y, col = cols, pch = pchs, xlab = "PC 1", ylab = "PC 2", 
+     xlim = range(x) + c(0, 100))
 ```
 
-This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix (i.e. the first two dimensions).
-At a first glance, it seems difficult to see a clear effect from the treatments (blue or red) or the depth (rounds or triangles) since the different experimental plots with the same parameters don’t seem to cluster.
-However, the problem, is that this plot ignores the 18 other dimensions of the ordination!
+This shows the distribution of the experimental plots along the two first axis of variation of the ordinated distance matrix (i.e. the first two dimensions).
+At a first glance, it seems difficult to see a clear effect of the treatments (blue or red) or the depth (rounds or triangles) as the different experimental plots with the same parameters don't seem to cluster together.
+However, this may be because this plot ignores the 18 other dimensions of the ordination!
 Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e.
-the values on the first dimension do not represent some continuous traits (e.g.
-depth) but just the ordinations of correlations between the data and some factors.
+the values on the first dimension do not represent a continuous trait (e.g.
+depth), instead they just represent the ordinations of correlations between the data and some factors.
 
@@
 
-The first step in such analysis will be to create different subsamples that will represent sub-samples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
-Each of these subsamples contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
+The first step is to create different subsamples that represent subsamples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
+Each of these subsamples will contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
 In our example, we are going to group the elements according to their depth and treatment.
@@
 They allow specific attributes to be displayed using the basic S3 functions (`print`, `summary`, `plot`, etc...).
-For more details on the `dispRity` objects, see the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html).
+For more details on `dispRity` objects, see the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html).
 
@@
 
-Disparity can be calculated in many ways, and therefore, the `dispRity` function allows to measure disparity as defined by the user.
+Disparity can be calculated in many ways, therefore the `dispRity` function allows users to define their own measure of disparity.
 For more details on measuring disparity, see the [`dispRity` metric vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html).
 
+<!-- NC: In the manual you use hypervolume - not sure if it matters? -->
 In this example, we are going to define disparity as the median distance between the different experimental plots and the centroid of the ordinated space.
-High values of disparity will indicate a general high dispersal from this centroid (i.e. on average, the experimental plots are far apart in the ordinated space).
+High values of disparity will indicate a generally high spread of points from this centroid (i.e. on average, the experimental plots are far apart in the ordinated space).
 We can define the metrics easily in the `dispRity` function by feeding them to the `metric` argument.
@@
 
-Similarly than with `custom.subsamples`, `dispRity` displays a `dispRity` object.
+Similarly to the `custom.subsamples` function, `dispRity` displays a `dispRity` object.
 To actually look at the calculated values, we can summarise the data by simply using `summary`:
@@
 
-We can also want to bootstrap the data to test the robustness of the measured disparity to outliers.
-Also, as we can see, each subsamples has different numbers of elements.
-It might be interesting to rarefy the data as well to have only subsamples with the same number of elements.
+We can also bootstrap the data to test the robustness of the measured disparity to outliers.
+Also, as we can see, each subsample has different numbers of elements.
+It might also be interesting to rarefy the data to have only subsamples with the same number of elements.
 Both steps are easily doable through the `boot.matrix` function.
@@
 
-## Bootstrapping the subsamples and using rarefaction (i.e. only re-sampling 17 elements each time)
+## Bootstrapping the subsamples and using rarefaction 
+## (i.e. only re-sampling 17 elements each time)
 (rarefied_data <- boot.matrix(customised_subsamples, bootstraps = 100, rarefaction = 17))
@@
 
-Summarising the data
---------------------
+Summarising the results
+-----------------------
 
-We can now summarise the data using various options such as the confidence intervals levels and the central tendency.
-For that we will use the `summary` function mentioned above using several of the options specific to `dispRity` objects (see more details with `?summary.dispRity`).
+We can now summarise the results using various options such as the confidence intervals and the central tendency.
+For that we use the `summary` function mentioned above, using several of the options specific to `dispRity` objects (see more details with `?summary.dispRity`).
 
@@
 
-We can also plot the results and have a look at the effect of the number of experimental plots using the generic `plot` function and it's associated arguments (see more details with `?plot.dispRity` for the arguments specific to `dispRity` objects).
+We can also plot the results and look at the effect of the number of experimental plots using the generic `plot` function and its associated arguments (see more details with `?plot.dispRity` for the arguments specific to `dispRity` objects).
 
@@
 
-As we can see, there seems to be no strong effect of the number of experimental plots in each subsamples (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
+As we can see, there seems to be no strong effect of the number of experimental plots in each subsample (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
 
@@
 
-Finally, we can test our hypothesis (whether the sort of water treatment at certain depth alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.
+Finally, we can test our hypothesis (whether the water treatment at certain depths alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.
 
@@
 
-As we can see, there is strong support for an effect of the treatment and the depth on the median distance between each experimental plots and the centroid of the ordinated space.
+As we can see, there is strong support for an effect of the treatment and the depth on the median distance between each experimental plot and the centroid of the ordinated space.
 In other words, it seems that with the second treatment and the second level of depth, the invertebrate communities and composition were further apart (i.e.
 more dispersed in the ordinated space) than with the first treatment and the first level of depth.
+
+<!-- NC: Some comment here about how this is different to the result you get with 2D methods alone? -->

### 2017-08-10 doc/dispRity-palaeo-demo.Rmd
Minor edits to conform to manual

--- 2017-06-19 doc/dispRity-palaeo-demo.Rmd
@@
 
-This demo aims to give quick overview of the `dispRity` package (v.0.3) for palaeobiology analysis.
-Please refer to [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html) that explains the functions in more detail.
+This demo aims to give quick overview of the `dispRity` package (v.0.3) for palaeobiology analyses of disparity, including disparity through time analyses.
+Please refer to my [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html) that explains the functions in more detail.
 
-To keep it short, this package allows to easily perform **disparity-through-time** analysis.
+In brief, this package allows users to easily perform **disparity-through-time** analysis.
 This type of analysis often starts by ordinating **morphometric** or **cladistic** data into a multidimensional object hereafter called the **morphospace**.
-One might be interested in studying how the occupancy of the morphospace changed through time by measuring a summary metric of this morphospace through time, called **disparity**.
-Disparity can then be seen as a value or a distribution of valuess summarising the diversity of morphologies.
+One might be interested in studying how the occupancy of the morphospace has changed through time by measuring a summary metric of this morphospace through time, called **disparity**.
+Disparity can then be seen as a value or a distribution of values summarising the diversity of morphologies.
 
-This demo showcases a typical disparity-through-time analyse: we are going to test whether the disparity changed through time in a subset of Eutheria (Mammalia) since the last 100 million years.
+This demo showcases a typical disparity-through-time analysis: we are going to test whether the disparity changed through time in a subset of eutherian mammals from the last 100 million years.
 
@@
 
-In this example, we are going to use a subset of the data from [@beckancient2014] that contains an ordinated cladistic matrix and a
-phylogenetic tree containing 50 taxa along side with a table containing some first and last occurences data for several taxa.
+In this example, we are going to use a subset of the data from [@beckancient2014]. 
+This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
 The ordinated matrix will represent our full morphospace, i.e. all the mammalian morphologies that ever existed through time (for this dataset).
@@
 
-## Loading the morphospace:
+## Loading the ordinated matrix/morphospace:
 data(BeckLee_mat50)
@@
 dim(BeckLee_mat50)
-## The morphospace contains 50 taxa and has 48 dimensions (or axis)
+## The morphospace contains 50 taxa and has 48 dimensions (or axes)
 
@@
 data(BeckLee_tree)
-plot(BeckLee_tree, cex=0.7); axisPhylo(root=140)
+plot(BeckLee_tree, cex = 0.7)
+axisPhylo(root = 140)
 ```
@@
 library(strap)
-geoscalePhylo(BeckLee_tree, cex.tip=0.7, cex.ts=0.6)
+geoscalePhylo(BeckLee_tree, cex.tip = 0.7, cex.ts = 0.6)
 ```
@@
 
-One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time subsamples that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
-Basically, the full morphospace in the previous step represents the total amount of morphologies in all time and is likely to be greater than any time subsamples of the morphospace.
+One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time subsamples that contain the total number of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
+Basically, the full morphospace represents the total number of morphologies across all time and will be greater than any of the time subsamples of the morphospace.
 
-The `dispRity` package provides a `time.subsamples` function that allows to split the morphospace time slices (using `method = continuous`) or into time bins (using `method = discrete`).
-In this example, we are going to split the morphospace into 5 equal time bins of 20 million years long from 100 million years ago (Mya) to the present.
-We will also provide to the function the table containing the first and last occurrences data for some fossils to take into account the fact that some fossils might span between our different time bins.
+The `dispRity` package provides a `time.subsamples` function that allows users to split the morphospace into time slices (using `method = continuous`) or into time bins (using `method = discrete`).
+In this example, we are going to split the morphospace into five equal time bins of 20 million years long from 100 million years ago to the present.
+We will also provide to the function a table containing the first and last occurrences dates for some fossils to take into account that some fossils might occur in several of our different time bins.
 
@@
 In brief, however, `dispRity` objects are lists of different elements (i.e.
-disparity results, morphospace time subsamples, morphospace attributes, etc.) that display only a summary of the object when calling the object for avoiding jamming the `R` console.
+disparity results, morphospace time subsamples, morphospace attributes, etc.) that display only a summary of the object when calling the object to avoiding filling the `R` console with superfluous output.
 
@@
 
-> These objects will gradually contain more information when reaching the following steps in the disparity-through-time analysis.
+> These objects will gradually contain more information when completing the following steps in the disparity-through-time analysis.
 
@@
 Once we obtain our different time subsamples, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
-The bootstrapping will allow to make each subsamples more robust to outliers and the rarefaction will allow us to compare subsamples with the same number of taxa to get rid of eventual sampling problems (i.e.
-more taxa in one subsamples than the other).
-The `boot.matrix` function that allows to bootstrap the `dispRity` object and the `rarefaction` option within allows to rarefy it.
+The bootstrapping allows us to make each subsample more robust to outliers and the rarefaction allows us to compare subsamples with the same number of taxa to remove sampling biases (i.e.
+more taxa in one subsample than the others).
+The `boot.matrix` function bootstraps the `dispRity` object and the `rarefaction` option within performs rarefaction.
 
 ```{r}
-## Bootstrapping each time subsamples 100 times
+## Bootstrapping each time subsample 100 times
 (boot_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100))
 
-## Getting the minimum number of rows (=taxa) in the time subsamples
+## Getting the minimum number of rows (i.e. taxa) in the time subsamples
 min(unlist(lapply(boot_bin_morphospace$subsamples, lapply, nrow)))
 
-## Bootstrapping each time subsamples 100 times and rarefying them 
+## Bootstrapping each time subsample 100 times and rarefying them 
 (rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
@@
 ```
+
 Calculating disparity
@@
 
-We can now calculate the disparity within each time subsamples along side with some confidence intervals generated by the pseudo-replication step above (bootstraps/rarefaction).
-Disparity can be calculated in many ways and this package allows user to come up with their own disparity metrics.
-For more info, please refer to the [`dispRity` metric vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html).
+We can now calculate the disparity within each time subsamples along with some confidence intervals generated by the pseudoreplication step above (bootstraps/rarefaction).
+Disparity can be calculated in many ways and this package allows users to come up with their own disparity metrics.
+For more details, please refer to the [`dispRity` metric vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html).
 
-In this example, we are going to calculate the spread of the data in each time subsamples by calculating the disparity as the sum of the variance of each dimension of the morphospace in each time subsamples using the `dispRity` function.
-Thus, in this example, the disparity is defined by the multi-dimensional variance of each time subsamples (i.e. the spread of the taxa within the morphospace).
-Note that this metric comes with a caveat not solved here since it ignores the covariance among the dimensions of the morphospace.
+In this example, we are going to calculate the spread of the data in each time subsample by calculating disparity as the sum of the variance of each dimension of the morphospace in each time subsample using the `dispRity` function.
+Thus, in this example, disparity is defined by the multi-dimensional variance of each time subsample (i.e. the spread of the taxa within the morphospace).
+Note that this metric comes with a caveat (not solved here) since it ignores covariances among the dimensions of the morphospace.
+We use this here because it is a standard metric used in disparity-through-time analysis.
+<!-- NC: reference? -->
 
@@
 The `dispRity` function does not actually display the calculated disparity values but rather only the properties of the disparity object (size, subsamples, metric, etc.).
-To display the actual calculated scores, we need to summarise the disparity object using the S3 method `summary` that is applied on a `dispRity` object (see `?summary.dispRity` for more details).
+To display the actual calculated scores, we need to summarise the disparity object using the S3 method `summary` that is applied to a `dispRity` object (see `?summary.dispRity` for more details).
 
-As for any `R` package, you can refer yourself to each individual function for more details.
+As for any `R` package, you can refer to the help files for each individual function for more details.
 
@@
 
-Finally, to draw some valid conclusions from these results, we can apply some statistical testing to the data.
-We can test, for example, if mammalian disparity changed through time since the last 100 million years (Mya).
-To do so, we can apply some comparisons of the means between each time-bins in a sequential manner to see whether the disparity in bin *n* is equal to the disparity in bin *n+1* which is in turn equal to the disparity in bin *n+2*, etc.
-Because our data is temporally auto-correlated (i.e.
-what happens in bin *n+1* depends on what happened in bin *n*) and was pseudo-replicated (i.e.
-each bootstrap draw creates non-independent time subsamples because they are all based on the same time subsamples), we are going to apply a non-parametric mean comparison: the `wilcox.test`.
-Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis (see `?p.adjust` for more details).
+Finally, to draw some valid conclusions from these results, we can apply some statistical tests.
+We can test, for example, if mammalian disparity changed significantly through time over the last 100 million years.
+To do so, we can compare the means of each time-bin in a sequential manner to see whether the disparity in bin *n* is equal to the disparity in bin *n+1*, and whether this is in turn equal to the disparity in bin *n+2*, etc.
+Because our data is temporally autocorrelated (i.e.
+what happens in bin *n+1* depends on what happened in bin *n*) and pseudoreplicated (i.e.
+each bootstrap draw creates non-independent time subsamples because they are all based on the same time subsamples), we apply a non-parametric mean comparison: the `wilcox.test`.
+Also, we need to apply a p-value correction (e.g. Bonferroni correction) to correct for multiple testing (see `?p.adjust` for more details).
 
 ```{r}
-## testing the differences between bins in the bootstrapped dataset.
+## Testing the differences between bins in the bootstrapped dataset.
 test.dispRity(boot_disparity, test = wilcox.test, comparison = "sequential",
@@
 
-## testing the differences between bins in the rarefied dataset.
+## Testing the differences between bins in the rarefied dataset.
 test.dispRity(rare_disparity, test = wilcox.test, comparison = "sequential",
@@
 
-Here our results show significant changes in disparity through time between all time bins (all p-values > 0.05).
-However, when looking at the rarefied results, there is no significant difference between the time bins in the Palaeogene (60-40 to 40-20 Mya), suggesting that the differences detected in the first test might just be due to the differences in number of taxa sample (13 or 6 taxa in each time bin).
+Here our results show significant changes in disparity through time between all time bins (all p-values < 0.05).
+However, when looking at the rarefied results, there is no significant difference between the time bins in the Palaeogene (60-40 to 40-20 Mya), suggesting that the differences detected in the first test might just be due to the differences in number of taxa sampled (13 or 6 taxa) in each time bin.
 
 # References 
+<!-- Add these -->

### 2017-08-10 doc/dispRity-simulate_data.Rmd
Updated with details from manual

--- 2017-01-10 doc/dispRity-simulate_data.Rmd
@@
 
-This vignette details how to simulate morphological data in `dispRity` (v.0.2.1).
-
 Simulating morphological data in `dispRity`
 =================
+
+The `dispRity` package allows users to simulate discrete morphological data matrices.
+In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
+The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
 
@@
 
-Quick go through
+A simple example
 =================================
-This is a quick example on how it works
 
@@
 
-## Generating a matrix with 100 characters (85% binary and 15% three states) and
-## an equal rates model with a gamma rates distribution (0.5, 1) with no 
+## Generating a matrix with 100 characters (85% binary and 15% three state) and
+## an equal rates model with a gamma rate distribution (0.5, 1) with no 
 ## invariant characters.
@@
 
-In details
+A more detailed description
 =================================
-Basically, `sim.morpho` is really flexible and intakes a lot of different arguments to allow to simulate realistic matrix.
-It has two implemented models: `"ER"` for Equal Rates (M*k* model) or the `"HKY"` one which is the molecular HKY model but transforms pyrines in 0's and purimidines in 1's.
-It can also intake some specific distribution for the rates or the substitutions, allowing to modify these basic models.
+`sim.morpho` is really flexible and takes many different arguments to allow users to simulate realistic matrices.
+It has three implemented models: `"ER"` for Equal Rates (the M*k* model); `"HKY"`, the molecular HKY model but transforms pyrines in 0's and pyrimidines in 1's; or the `"mixed"` model that randomly uses an `"ER"` and/or `"HKY"` for the binary characters and `"ER"` for the multistate (>2) characters.
+Both models take specific distributions for their rate or substitution models.
+These distributions should be passed to these arguments in the format of `c(sampler_function, distribution_parameters)` where the the sampler function is a the random generation function of that distribution (e.g.
+`rnorm`, `runif`, etc...) and the parameters are any parameters to be passed to this function.
 
-The `check.morpho` runs a quick MP tree using the `phangorn` parsimony algorithm.
+`check.morpho` runs a quick Maximum Parsimony tree search using the `phangorn` parsimony algorithm.
 It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g.
-the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the MP tree and the provided tree.
+the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree to determine how different they are.
 
-Good parameters for a good matrix
------------------------------------
-
-I was testing the effect of different parameters on the ability to generate a ``realistic'' matrix (meaning not to different from the input tree with a consistency and retention index close to what's in the literature).
-The matrix seems the most ``realistic'' with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three states characters, a gamma distribution with a shape parameter ($\alpha$) of 0.5 and no scaling ($\beta$ = 1).
-
-Something like that should give a decent matrix:
+Parameters for a realistic(ish) matrix
+===========================================
+There are many parameters that can create a ``realistic'' matrix (i.e.
+not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all end up creating "good" matrices.
+The following parameters however, seem to generate fairly ``realistic'' matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
+<!-- NC: Check that I haven't changed what you mean here where i swapped good for realistic. Realistic is a better term. Or be clearer what you mena by "good" -->
 
@@
 ```
+
+`space.maker`
+=================================
+Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
+This function allows users to simulate multidimensional spaces with a certain number of properties.
+It takes as arguments the number of elements (data points) and dimensions to create the space but also permits more fine tuning in the data simulation: it is possible to give a specific distribution to each dimension, provide a correlation matrix to link the dimensions or even provide an *a priori* distribution of the variance for each distributions!
+<!-- added this in from the manual -->
+
+```{r, fig.width=8, fig.height=8, eval=TRUE}
+## Graphical options
+op <- par(mfrow = (c(2, 2)), bty = "n")
+## Visualising 3D spaces
+require(scatterplot3d)
+
+## A cube space
+scatterplot3d(space.maker(2500, 3, runif), pch = 20,
+              xlab = "", ylab = "", zlab = "",
+              main = "Uniform cube space")
+
+## A plane space
+scatterplot3d(space.maker(2500, 3, c(runif, runif, runif),
+              arguments = list(list(min = 0, max = 0), NULL, NULL)), pch = 20,
+              xlab = "", ylab = "", zlab = "",
+              main = "Uniform plane space")
+
+## An ellipsoid space (=a spheric space with correlation)
+cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3)
+scatterplot3d(space.maker(2500, 3, rnorm, cor.matrix = cor_matrix), pch = 20,
+              xlab = "", ylab = "", zlab = "",
+              main = "Normal ellipsoid space")
+
+## A cylindrical space with a decreasing variance per axis
+scatterplot3d(space.maker(2500, 3, c(rnorm, rnorm, runif),
+              scree = c(0.7, 0.2, 0.1)), pch = 20,
+              xlab = "", ylab = "", zlab = "",
+              main = "Normal cylindrical space")
+
+## Resetting the graphic parameters
+par(op)
+```

### 2017-08-12 gitbook/03_specific-tutorials.Rmd
Git book HTML version of manual

--- 2017-08-10 doc/dispRity-manual.Rmd
@@
 ---
-title: "dispRity manual"
-author: "Thomas Guillerme"
+title: "Details of specific functions"
+author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
 date: "`r Sys.Date()`"
 output:
-    rmarkdown::html_vignette:
-        toc: true
-        toc_depth: 2
-bibliography: References.bib
-bst: sysbio.bst
-vignette: >
-  %\VignetteIndexEntry{dispRity manual}
-  %\VignetteEngine{knitr::rmarkdown}
-  \usepackage[utf8]{inputenc}
+  html_document: default
+  pdf_document: default
 ---
 
-# `dispRity`
-
-This is a package for measuring disparity in `R`.
-It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values so they can easily be compared.
-This manual is based on the version `1.0`.
-
-# Introduction
-
-## What is `dispRity`?
-This is a package for measuring disparity in `R`.
-It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
-Typically, these analysis are used in palaeobiology and evolutionary biology to study the changes in morphology through time.
-However, there are many more applications in ecology, evolution and beyond.
-
-## Installing and running the package
-
-You can install this package easily if you use the latest version of `R` (> 3.3.0) and `devtools`.
-
-```{r, eval=FALSE}
-## Checking if devtools is already installed
-if(!require(devtools)) install.packages("devtools")
-
-## Installing the latest released version directly from GitHub
-install_github("TGuillerme/dispRity", ref = "release")
-
-## Loading the package
-library(dispRity)
-```
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
-# NC: this isn't checking Claddis...
-library(dispRity)
-## Setting the random seed for repeatability
-set.seed(123)
-```
-
-Note this uses the `release` branch (version`0.4`). <!-- @@@ -->
-For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
-`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, `Claddis`, `geomorph` and `RCurl`).
-
-## Why not CRAN?
-
-This package is not available on CRAN.
-This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
-However, the package follows the strict (and useful!) CRAN standards via Travis.
-
-[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
-[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
-[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
-
-## Help
-
-If you need help with the package, hopefully the following manual will be useful.
-However, parts of this package are still in development and some other parts are probably not covered.
-Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
-
-## Citations
-
-You can cite both the package or this manual with the following citation:
-
-> Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646
-
-<!-- TG: I'm gonna change that to a newest version with the submission -->
-
-Note that this citation is only temporary (but can still be used!).
-A future proper version of the latest package release, this manual and an associated methods paper should be submitted soon(ish!).
-
-# Getting started
-
-## What sort of data does dispRity work with?
-
-Disparity can be estimated from pretty much any matrix.
-Classically, however, it is measured from ordinated matrices.
-These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).
-
-### Ordination matrices from `Claddis`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
-if(!require(Claddis)) install.packages("Claddis")
-```
-
-`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
-For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:
-
-```{r}
-require(Claddis)
-
-## Ordinating the example data from Claddis
-Claddis.ordination(Michaux1989) 
-```
-
-Note that several options are available, namely which type of distance should be computed.
-See more info in the function manual (`?Claddis.ordination`).
-Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.
-
-### Ordination matrices from `geomorph`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if geomorph is installed BUILD ONLY
-if(!require(geomorph)) install.packages("geomorph")
-```
-
-You can also easily use data from `geomorph` using the `geomorph.ordination` function.
-This function simply takes Procrustes aligned data and performs an ordination:
-
-```{r}
-require(geomorph)
-
-## Loading the plethodon dataset
-data(plethodon)
-
-## Performing a Procrustes transform on the landmarks
-procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)
-
-## Ordinating this data
-geomorph.ordination(procrustes)[1:5,1:5]
-```
-
-Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
-Additionally you can give the function a `geomorph.data.frame` object.
-If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsamples)!
-
-```{r}
-## Using a geomorph.data.frame
-geomorph_df <- geomorph.data.frame(procrustes,
-     species = plethodon$species, site = plethodon$site)
-
-## Ordinating this data and making a dispRity object
-geomorph.ordination(geomorph_df)
-```
-
-More about these `dispRity` objects below!
-
-### Other kinds of ordination matrices
-
-If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordination matrices by using the following functions from the `stats` package.
-Here is how to do it for the following types of matrices:
-
- * Multivariate matrices (principal components analysis; PCA)
-
-```{r}
-## A multivariate matrix
-head(USArrests)
-
-## Ordinating the matrix using `prcomp` 
-ordination <- prcomp(USArrests)
-
-## Selecting the ordinated matrix
-ordinated_matrix <- ordination$x
-head(ordinated_matrix)
-```
-  This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, see the `?princomp` function.
-
- * Distance matrices (classical multidimensional scaling; MDS)
-
-```{r}
-## A matrix of distances between cities
-str(eurodist)
-
-## Ordinating the matrix using cmdscale() with k = 5 dimensions 
-ordinated_matrix <- cmdscale(eurodist, k = 5)
-head(ordinated_matrix)
-```
-
-  This results in a ordinated matrix with European cities as elements and five dimensions.
-
-Of course any other method for creating the ordination matrix is totally valid, you can also not use any ordination at all!
-The only requirements for the `dispRity` functions is that the input is a matrix with elements as rows and dimensions as columns.
-
-## Performing a simple dispRity analysis
-
-Two `dispRity` functions allow users to run an analysis pipeline simply by inputting an ordination matrix.
-These functions allow users to either calculate the disparity through time (`dispRity.through.time`) or the disparity of user-defined groups (`dispRity.per.group`).
-
-### Example data
-
-To illustrate these functions, we will use data from @beckancient2014.
-This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
-
-```{r, fig.width=7, fig.height=7}
-## Loading the ordinated matrices
-data(BeckLee_mat50)
-data(BeckLee_mat99)
-
-## The first five taxa and dimensions of the 50 taxa matrix
-head(BeckLee_mat50[, 1:5])
-
-## The first five taxa and dimensions of the 99 taxa + ancestors matrix
-BeckLee_mat99[c(1, 2, 98, 99), 1:5]
-
-## Loading a list of first and last occurrence dates for the fossils
-data(BeckLee_ages)
-head(BeckLee_ages)
-
-## Loading and plotting the phylogeny
-data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.8) 
-axisPhylo(root = 140)
-nodelabels(cex = 0.5)
-```
-
-Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).
-
-### Disparity through time
-
-The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
-This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
-Of course all the defaults can be changed if required, more on this later.
-
-For a disparity through time analysis, you will need:
-
-  * An ordinated matrix (we covered that above)
-  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element.
-To give your tree a root time (i.e.
-an age for the root), you can simply do `my_tree$root.time <- my_age`.
-  * The required number of time subsamples (here `time = 3`)
-  * Your favourite disparity metric (here the sum of variances)
-
-Using the @beckancient2014 data described [above](#example-data):
-
-```{r}
-## Measuring disparity through time
-disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
-                                        time = 3, metric = c(sum, variances))
-```
-
-This generates a `dispRity` object (see [here](#manipulating-dispRity-objects) for technical details).
-When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:
-
-```{r}
-## Print the disparity_data object
-disparity_data
-```
-
-We asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
-
-We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):
-
-```{r, fig.width=7, fig.height=7}
-## Summarising disparity through time
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data, type = "continuous")
-
-## Testing for an difference among the time bins
-disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
-summary(disp_lm)
-
-## Note that ANOVA assumptions were not checked here!
-```
-
-Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
-You can also directly explore the specific function help files within R and navigate to related functions.
-
-### Disparity among groups
-
-The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
-For example, you could ask if there is a difference in disparity between two groups?
-
-To perform such an analysis, you will need:
-
- * An matrix with rows as elements and columns as dimensions (always!)
- * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix.
-For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4.
-Note that elements can be present in multiple groups at once.
- * Your favourite disparity metric (here the sum of variances)
-
-Using the @beckancient2014 data described [above](#example-data):
-
-```{r}
-## Creating the two groups (crown versus stem) as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44))
-
-## Measuring disparity for each group
-disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
-                                     metric = c(sum, variances))
-```
-
-We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).
-
-```{r, fig.width=7, fig.height=7}
-## Print the disparity_data object
-disparity_data
-
-## Summarising disparity in the different groups
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data)
-
-## Testing for a difference between the groups
-test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
-```
-
-# Specific tutorials
+# Details of specific functions
 
@@
 
-#### Summarising and displaying the data
+#### Summarising and displaying the results
 
@@
 
-### A more detailed descriptions
+### A more detailed description
 
@@
 ```
-
-# Glossary <a name="glossary"></a>
-
--   **Ordinated space**. The mathematical multidimensional object that will be analysed with this package.
-    In morphometrics, this is often referred to as the morphospace.
-    However it may also be referred to as the cladisto-space for cladistic data or the eco-space for ecological data etc.
-    In practice, this term designates an ordinated matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
-
--   **Elements**. The rows of the ordinated space. Elements can be taxa, field sites, countries etc.
-
--   **Dimensions**. The columns of the ordinated space. The dimensions are referred to as axes of variation, or principal components, for ordinated spaces obtained from a PCA.
-
--   **Subsamples**. Sub-samples of the ordinated space.
-    A sub-sample (or subsamples) contains the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
-    For example, if our ordinated space is composed of birds and mammals (the elements) and 50 principal components (the dimensions), we can create two subsamples containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
-
-# The package's guts
-
-## Manipulating `dispRity` objects
-
-Disparity analysis involves a lot of manipulation of many matrices (especially when bootstrapping) which can be impractical to visualise and will quickly overwhelm your `R` console.
-Even the simple Beck and Lee 2014 example above produces an object with > 72 lines of lists of lists of matrices!
-
-Therefore `dispRity` uses a specific class of object called a `dispRity` object.
-These objects allow users to use S3 method functions such as `summary.dispRity`, `plot.dispRity` and `print.dispRity`. 
-`dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g. `sort.dispRity`, `extract.dispRity` see the full list in the next section).
-These functions modify the `dispRity` object without having to delve into its complex structure!
-The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).
-
-```{r}
-## Loading the example data
-data(disparity)
-
-## What is the class of the median_centroids object?
-class(disparity)
-
-## What does the object contain?
-names(disparity)
-
-## Summarising it using the S3 method print.dispRity
-disparity
-```
-
-Note that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:
-
-```{r, eval=FALSE}
-## Display the full object
-print(disparity, all = TRUE)
-## This is more nearly ~ 5000 lines on my 13 inch laptop screen!
-```
-
-## `dispRity` utilities
-
-The package also provides some utilities functions to facilitate multidimensional analysis.
-
-### `dispRity` object utilities  <a name="dispRity.utilities"></a>
-The first subsamples of utilities are functions for manipulating `dispRity` objects:
-
-#### `make.dispRity`
-This function creates empty `dispRity` objects.
-
-```{r}
-## Creating an empty dispRity object
-make.dispRity()
-
-## Creating an "empty" dispRity object with a matrix
-(disparity_obj <- make.dispRity(matrix(rnorm(20), 5, 4)))
-```
-
-####  `fill.dispRity`
-This function initialises a `dispRity` object and generating it's call properties.
-
-```{r}
-## The dispRity object's call is indeed empty
-disparity_obj$call
-
-## Filling an empty disparity object (that needs to contain at least a matrix)
-(disparity_obj <- fill.dispRity(disparity_obj))
-
-## The dipRity object has now the proper minimal attributes
-disparity_obj$call
-```
-
-####  `matrix.dispRity`
-This function extracts a specific matrix from a disparity object. The matrix can be one of the bootstrapped or/and rarefied one.
-
-```{r}
-## Extracting the matrix containing the coordinates of the elements at time 50
-str(matrix.dispRity(disparity, "50"))
-
-## Extracting the 3rd boostrapped matrix with the 2nd rarefaction level
-## (15 elements) from the second group (80 Mya)
-str(matrix.dispRity(disparity, subsamples = 1, bootstrap = 3, rarefaction = 2))
-```
-
-#### `get.subsamples.dispRity`
-This function creates a dispRity object that contains only elements from one specific subsamples.
-
-```{r, eval=FALSE}
-## Extracting all the data for the crown mammals
-(crown_mammals <- get.subsamples.dispRity(disp_crown_stemBS, "Group.crown"))
-
-## The object keeps the properties of the parent object but is composed of only one subsamples
-length(crown_mammals$subsamples)
-```
-
-#### `extract.dispRity`
-This function extracts the calculated disparity values of a specific matrix.
-
-```{r, eval=FALSE}
-## Extracting the observed disparity (default)
-extract.dispRity(disparity)
-
-## Extracting the disparity from the bootstrapped values from the
-## 10th rarefaction level from the second subsamples (80 Mya)
-extract.dispRity(disparity, observed = FALSE, subsamples = 2, rarefaction = 10)
-```
-
-#### `scale.dispRity`
-This is the S3 method of `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object.
-
-```{r, eval=FALSE}
-## Getting the disparity values of the time subsamples
-head(summary(disparity))
-
-## Scaling the same disparity values
-head(summary(scale(disparity, scale = TRUE)))
-
-## Scaling and centering:
-head(summary(scale(disparity, scale = TRUE, center = TRUE)))
-```
-
-#### `sort.dispRity`
-This is the S3 method of `sort` for sorting the subsamples alphabetically (default) or following a specific pattern.
-
-```{r, eval=FALSE}
-## Sorting the disparity subsamples in inverse alphabetic order
-head(summary(sort(disparity, decreasing = TRUE)))
-
-## Customised sorting
-head(summary(sort(disparity, sort = c(7, 1, 3, 4, 5, 2, 6))))
-```
-
-<!-- ## Modularity
-
-### Making your own metrics
-
-### Making your own tests
-
-## Where is this going?
- -->
-<!-- ## Running `dispRity` in parallel
-
-The computationally intensive function `dispRity` have a `parallel` option to speed up their calculations.
-
-This option requires the package `snow` and takes arguments that are to be passed to  -->

### 2017-08-12 gitbook/05_palaeo-demo.Rmd
Git book HTML version of manual

--- 2017-08-10 doc/dispRity-palaeo-demo.Rmd
@@
 ---
-title: "dispRity palaeo demo"
-author: "Thomas Guillerme"
+title: "Palaeobiology demo: disparity-through-time and within groups"
+author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
 date: "`r Sys.Date()`"
-output: rmarkdown::html_vignette
-bibliography: References.bib
-bst: sysbio.bst
-vignette: >
-  %\VignetteIndexEntry{dispRity palaeo demo}
-  %\VignetteEngine{knitr::rmarkdown}
-  \usepackage[utf8]{inputenc}
+output:
+  html_document: default
+  pdf_document: default
 ---
+
+# Palaeobiology demo: disparity-through-time and within groups
 
@@
 
-Before starting
-===============
+## Before starting
 
-Installing `dispRity` 
------------------------------------
+### Installing `dispRity`
 
@@
 
-The morphospace
----------------
+### The morphospace
 
@@
 
-A disparity-through-time analysis
-=================================
+## A disparity-through-time analysis
 
-Splitting the morphospace through time
---------------------------------------
+### Splitting the morphospace through time
 
@@
 
-Bootstrapping the data
-----------------------
+### Bootstrapping the data
 
@@
 
-Calculating disparity
----------------------
+### Calculating disparity
 
@@
 
-Plotting the results
---------------------
+### Plotting the results
 
@@
 
-# References 
+### References 
 <!-- Add these -->

### 2017-08-12 gitbook/02_getting-started.Rmd

--- 2017-08-10 doc/dispRity-manual.Rmd
@@
 ---
-title: "dispRity manual"
-author: "Thomas Guillerme"
+title: "Getting started with dispRity"
+author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
 date: "`r Sys.Date()`"
 output:
-    rmarkdown::html_vignette:
-        toc: true
-        toc_depth: 2
-bibliography: References.bib
-bst: sysbio.bst
-vignette: >
-  %\VignetteIndexEntry{dispRity manual}
-  %\VignetteEngine{knitr::rmarkdown}
-  \usepackage[utf8]{inputenc}
+  html_document: default
+  pdf_document: default
 ---
 
-# `dispRity`
-
-This is a package for measuring disparity in `R`.
-It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values so they can easily be compared.
-This manual is based on the version `1.0`.
-
-# Introduction
-
-## What is `dispRity`?
-This is a package for measuring disparity in `R`.
-It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
-Typically, these analysis are used in palaeobiology and evolutionary biology to study the changes in morphology through time.
-However, there are many more applications in ecology, evolution and beyond.
-
-## Installing and running the package
-
-You can install this package easily if you use the latest version of `R` (> 3.3.0) and `devtools`.
-
-```{r, eval=FALSE}
-## Checking if devtools is already installed
-if(!require(devtools)) install.packages("devtools")
-
-## Installing the latest released version directly from GitHub
-install_github("TGuillerme/dispRity", ref = "release")
-
-## Loading the package
-library(dispRity)
-```
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
-# NC: this isn't checking Claddis...
-library(dispRity)
-## Setting the random seed for repeatability
-set.seed(123)
-```
-
-Note this uses the `release` branch (version`0.4`). <!-- @@@ -->
-For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
-`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, `Claddis`, `geomorph` and `RCurl`).
-
-## Why not CRAN?
-
-This package is not available on CRAN.
-This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
-However, the package follows the strict (and useful!) CRAN standards via Travis.
-
-[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
-[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
-[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
-
-## Help
-
-If you need help with the package, hopefully the following manual will be useful.
-However, parts of this package are still in development and some other parts are probably not covered.
-Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
-
-## Citations
-
-You can cite both the package or this manual with the following citation:
-
-> Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646
-
-<!-- TG: I'm gonna change that to a newest version with the submission -->
-
-Note that this citation is only temporary (but can still be used!).
-A future proper version of the latest package release, this manual and an associated methods paper should be submitted soon(ish!).
-
-# Getting started
+# Getting started with dispRity
 
@@
 These functions allow users to either calculate the disparity through time (`dispRity.through.time`) or the disparity of user-defined groups (`dispRity.per.group`).
+
+**IMPORTANT**
+
+Note that `disparity.through.time` and `disparity.per.group` are wrapper functions (i.e.
+they incorporate lots of other functions) that allow users to run a basic disparity-through-time, or disparity among groups, analysis without too much effort.
+As such they have a lot of defaults. 
+These are described in the help files for the functions that are used to make the wrapper functions, and not described in the help files for `disparity.through.time` and `disparity.per.group`. 
+These defaults are perfect for **data exploration**, but for a proper analysis you should consider the **best parameters for your question and data**. 
+For example, which metric should you use? How many bootstraps do you require? What model of evolution is most appropriate if you are time slicing? Should you rarefy the data?
+See `custom.subsamples`, `boot.matrix`, `dispRity.metric`, \code{\link{summary.dispRity}}, `plot.dispRity` for more details of the defaults used in each of these functions.
+Note that any of these defaults can be changed within the `disparity.through.time` or `disparity.per.group` functions.
+<!-- NC: Add links to these sections of the manual -->
 
@@
 ```
-
-# Specific tutorials
-
-The following section contains information specific to some functions.
-If any of your questions are not covered in these sections, please refer to the function help files in R, send me an email (<guillert@tcd.ie>), or raise an issue on GitHub.
-The several tutorials below describe specific functionalities of certain functions; please always refer to the function help files for the full function documentation!
-
-## Time slicing
-
-The function `time.subsamples` allows users to divide the matrix into different time subsamples or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
-Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.
-
-Two types of time subsamples can be performed by using the `method` option:
-
- *  Discrete time subsamples (or time-binning) using `method = discrete`
- *  Continuous time subsamples (or time-slicing) using `method = continuous`
-
-For the time-slicing method details see @GuillermeSTD. <!-- @@@ Change cite appropriately! -->
-<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
-For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:
-
- *  Defining the boundaries of the time bins (when `method = discrete`)
- *  Defining the time slices (when `method = continuous`)
-
-Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
-Additionally, it is also possible to input a dataframe containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age, so taxa can appear in more than one time bin/slice.
-
-Here is an example for `method = discrete`:
-
-```{r, eval=TRUE}
-## Generating three time bins containing the taxa present every 40 Ma
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0))
-```
-
-Note that we can also generate equivalent results by just telling the function that we want three time-bins as follow:
-
-```{r, eval=TRUE}
-## Automatically generate three equal length bins:
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = 3)
-```
-
-In this example, the taxa were split inside each time-bin according to their age.
-However, the taxa here are considered as single points in time.
-It is totally possible that some taxa could have had longer longevity and that they exist in multiple time bins.
-In this case, it is possible to include them in more than one bin by providing a table of first and last occurrence dates (FAD/LAD).
-This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:
-
-```{r, eval=TRUE}
-## Displaying the table of first and last occurrence dates for each taxa
-head(BeckLee_ages)
-
-## Generating time bins including taxa that might span between them
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
-```
-
-When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as the root age plus 1\% of the tree length, as long as at least three elements/taxa are present at that point in time.
-The algorithm adds an extra 1\% tree length until reaching the required minimum of three elements.
-It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance among tips *and* nodes.
-
-For the time-slicing method (`method = continuous`), the idea is fairly similar.
-This option, however, requires a matrix that contains the ordinated distance among taxa *and* nodes and an extra argument describing the assumed evolutionary model (via the `model` argument).
-This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node, meaning that a decision must be made about what the value for the branch should be.
-The model can be one of the following:
-
- *  `acctran` where the data chosen along the branch is always the one of the descendant
- *  `deltran` where the data chosen along the branch is always the one of the ancestor
- *  `punctuated` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
- *  `gradual` where the data chosen along the branch is either the descendant or the ancestor depending on branch length
-
-```{r, eval=TRUE}
-## Generating four time slices every 40 million years under a model of gradual evolution
-time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-
-## Generating four time slices automatically
-time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree,
-    method = "continuous", model = "gradual", time = 4, FADLAD = BeckLee_ages)
-```
-
-## Customised subsamples
-
-Another way of separating elements into different categories is to use customised subsamples as briefly explained [above](#disparity-among-groups).
-This function simply takes the list of elements to put in each group (whether they are the actual element names or their position in the matrix).
-
-```{r, eval=TRUE}
-## Creating the two groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44))
-
-## Separating the dataset into two different groups
-custom.subsamples(BeckLee_mat50, group = mammal_groups)
-```
-
-Elements can easily be assigned to different groups if necessary!
-
-```{r, eval=FALSE}
-## Creating the three groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44).
-                      "all" = c(1:50))
-```
-
-## Bootstraps and rarefactions
-
-One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results are, and how sensitive they are to outliers in the dataset.
-This can be achieved using the function `boot.matrix` to bootstrap and/or rarefy the data.
-The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):
-
-```{r, eval=TRUE}
-## Default bootstrapping
-boot.matrix(data = BeckLee_mat50)
-```
-
-The number of bootstrap replicates can be defined using the `bootstraps` option.
-The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
-Currently two algorithms are implemented:
-
- * `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any *m* elements drawn from the data)
- * `single` where only one random element is replaced by one other random element for each pseudo-replicate
-
-```{r, eval=TRUE}
-## Bootstrapping with the single bootstrap method
-boot.matrix(BeckLee_mat50, boot.type = "single")
-```
-
-This function also allows users to rarefy the data using the `rarefaction` argument.
-Rarefaction allows users to limit the number of elements to be drawn at each bootstrap replication.
-This is useful if, for example, one is interested in looking at the effect of reducing the number of elements on the results of an analysis.
-
-This can be achieved by using the `rarefaction` option that draws only *n-x* at each bootstrap replicate (where *x* is the number of elements not sampled).
-The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsample to a minimum of three elements).
-It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
-
-```{r, eval=TRUE}
-## Bootstrapping with the full rarefaction
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
-
-## Or with a set number of rarefaction levels
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8, 3))
-```
-
-One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
-When missing, all dimensions from the ordinated matrix are used.
-
-```{r, eval=TRUE}
-## Using the first 50% of the dimensions
-boot.matrix(BeckLee_mat50, dimensions = 0.5)
-
-## Using the first 10 dimensions
-boot.matrix(BeckLee_mat50, dimensions = 10)
-```
-
-Of course, one could directly supply the subsamples generated above (using `time.subsamples` or `custom.subsamples`) to this function.
-
-```{r, eval=TRUE}
-## Creating subsamples of crown and stem mammals
-crown_stem <- custom.subsamples(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot.matrix(crown_stem, bootstraps = 200, rarefaction = TRUE)
-
-## Creating time slice subsamples
-time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-                               method = "continuous", model = "gradual", 
-                               time = c(120, 80, 40, 0),
-                               FADLAD = BeckLee_ages)
-
-## Bootstrapping the time slice subsamples
-boot.matrix(time_slices, bootstraps = 100)
-```
-
-## Disparity metrics
-
-There are many ways of measuring disparity!
-In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
-For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @Wills1994.
-Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.
-
-Because of this great diversity of metrics, the package `dispRity` does not have one way to measure disparity but rather proposes to facilitate users in defining their own disparity metric that will best suit their particular analysis.
-In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
-However the `metric` argument has to follow certain rules:
-
-1.  It must be composed from one to three `function` objects;
-2.  The function(s) must take as a first argument a `matrix` or a `vector`;
-3.  The function(s) must be of one of the three dimension-levels described below;
-4.  At least one of the functions must be of dimension-level 1 or 2 (see below).
-
-### The function dimension-levels
-
-The metric function dimension-levels determine the "dimensionality of decomposition" of the input matrix.
-In other words, each dimension-level designates the dimensions of the output, i.e. either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimension.
-
-![Illustration of the different dimension-levels of functions with an input `matrix`](dispRity_fun.png)
-
-#### Dimension-level 1 functions
-
-A dimension-level 1 function will decompose a `matrix` or a `vector` into a single value:
-
-```{r}
-## Creating a dummy matrix
-dummy_matrix <- matrix(rnorm(12), 4, 3)
-
-## Example of dimension-level 1 functions
-mean(dummy_matrix)
-median(dummy_matrix)
-```
-
-Any summary metric such as mean or median are good examples of dimension-level 1 functions as they reduce the matrix to a single dimension (i.e. one value).
-
-#### Dimension-level 2 functions
-
-A dimension-level 2 function will decompose a `matrix` into a `vector`.
-
-```{r}
-## Defining the function as the product of rows
-prod.rows <- function(matrix) apply(matrix, 1, prod)
-
-## A dimension-level 2 metric
-prod.rows(dummy_matrix)
-```
-
-Several dimension-level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculate the variance or the range of each dimension of the ordinated matrix respectively.
-
-#### Dimension-level 3 functions
-
-Finally a dimension-level 3 function will transform the matrix into another matrix.
-Note that the dimension of the output matrix doesn't need to match the the input matrix:
-
-```{r}
-## A dimension-level 3 metric
-var(dummy_matrix)
-
-## A dimension-level 3 metric with a forced matrix output
-as.matrix(dist(dummy_matrix))
-```
-
-### `make.metric`
-
-Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the Euclidean distance between each element and the centroid of the ordinated space.
-The `make.metric` function implemented in `dispRity` is designed to help test and find the dimension-level of the functions.
-This function tests:
-
-1.  If your function can deal with a `matrix` or a `vector` as an input;
-2.  Your function's dimension-level according to its output (dimension-level 1, 2 or 3, see above);
-3.  Whether the function can be implemented in the `dispRity` function (the function is fed into a `lapply` loop).
-
-For example, let's see if the functions described above are the right dimension-levels:
-
-```{r}
-## Which dimension-level is the mean function? And can it be used in dispRity?
-make.metric(mean)
-
-## Which dimension-level is the prod.rows function? And can it be used in dispRity?
-make.metric(prod.rows)
-
-## Which dimension-level is the var function? And can it be used in dispRity?
-make.metric(var)
-```
-
-A non verbose version of the function is also available.
-This can be done using the option `silent = TRUE` and will simply output the dimension-level of the metric.
-
-```{r}
-## Testing whether mean is dimension-level 1
-if(make.metric(mean, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-## Testing whether var is dimension-level 1
-if(make.metric(var, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-```
-
-### Metrics in the `dispRity` function
-
-Using this metric structure, we can easily use any disparity metric in the `dispRity` function as follows:
-
-```{r}
-## Measuring disparity as the standard deviation of all the values of the
-## ordinated matrix (dimension-level 1 function).
-summary(dispRity(BeckLee_mat50, metric = sd))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the ordinated matrix (dimension-level 1 and 2 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the variance covariance matrix (dimension-level 1, 2 and 3 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
-```
-
-Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > 2 > 1).
-
-```{r}
-## Disparity as the standard deviation of the variance of each axis of the
-## variance covariance matrix:
-disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)),
-                      round = 10)
-
-## Same as above but using a different function order for the metric argument
-disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
-                      round = 10)
-
-## Both ways output the same disparity values:
-disparity1 == disparity2
-```
-
-In these examples, we considered disparity to be a single value.
-For example, in the previous example, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
-It is, however, possible to calculate [disparity as a distribution](#disparity-as-a-distribution).
-
-### Metrics implemented in `dispRity`
-
-Several disparity metrics are implemented in the `dispRity` package.
-The detailed list can be found in `?dispRity.metric` along with some description of each metric.
-
-Level | Name | Description | Source | 
-------|------|---------------------------------------------------|--------|
-3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
-3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
-3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
-2 | `ranges` | The range of each dimension | `dispRity` |
-2 | `variances` | The variance of each dimension | `dispRity` |
-2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
-1 | `mode.val` | The modal value | `dispRity` |
-
-1: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (not PCA).
-
-2: Note that by default, the centroid is the centroid of the elements.
-It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
-
-### Equations and implementations
-Some of the functions described below are implemented in the `dispRity` package and do not require any other packages to calculate ([see implementation here](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R)).
-
-\begin{equation}
-    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
-\end{equation}
-Where *k* is the number of dimensions, and $\lambda_i$ is the eigenvalue of each dimension.
-
-\begin{equation}
-    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
-\end{equation}
-Where *k* is the number of dimensions.
-
-\begin{equation}
-    ranges = |max(k_i) - min(k_i)|
-\end{equation}
-Where *k* is the number of dimensions.
-
-\begin{equation}
-    variances = \sigma^{2}{k_i}
-\end{equation}
-Where *k* is the number of dimensions, and $\sigma^{2}$ is their variance.
-
-\begin{equation}
-    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
-\end{equation}
-Where *n* is each element in the ordinated space, *k* is the number of dimensions, and $Centroid_{k}$ is their mean (or can be set to another value).
-
-### Using the different disparity metrics
-Here is a brief demonstration of the main metrics implemented in `dispRity`.
-First, we will create a dummy/simulated ordinated space using the `space.maker` utility function (more about that [here](#space.maker):
-
-```{r}
-## Creating a 10*5 normal space
-set.seed(1)
-dummy_space <- space.maker(10, 5, rnorm)
-```
-
-We will use this simulated space to demonstrate the different metrics.
-
-#### Volumes and surface metrics
-The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `hyper.volume` all measure the surface or the volume of the ordinated space occupied:
-
-```{r}
-## Calculating the ellipsoid volume
-summary(dispRity(dummy_space, metric = ellipse.volume))
-```
-> Because there is only one subsample (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with rounding).
-
-```{r}
-## Calculating the convex hull surface
-summary(dispRity(dummy_space, metric = convhull.surface))
-
-## Calculating the convex hull volume
-summary(dispRity(dummy_space, metric = convhull.volume))
-```
-
-The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
-For the full and correct calculation of the volume of the ordinated space, it is possible to use the `hyper.volume` function that has more options but takes longer to calculate.
-
-```{r}
-## Calculating the true multidimensional volume
-summary(dispRity(dummy_space, metric = hyper.volume))
-
-```
-> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) that often results in near 0 disparity values.
-
-#### Ranges, variances and diagonal
-
-The functions `ranges`, `variances` and `diagonal` all measure properties of the ordinated space based on its dimensional properties (they are also less affected by the "curse of dimensionality"):
-
-`ranges` and `variances` both work on the same principle and measure the range/variance of each dimension:
-
-```{r}
-## Calculating the ranges of each dimension in the ordinated space
-ranges(dummy_space)
-
-## Calculating disparity as the distribution of these ranges
-summary(dispRity(dummy_space, metric = ranges))
-
-## Calculating disparity as the sum and the product of these ranges
-summary(dispRity(dummy_space, metric = c(sum, ranges)))
-summary(dispRity(dummy_space, metric = c(prod, ranges)))
-
-## Calculating the variances of each dimension in the ordinated space
-variances(dummy_space)
-
-## Calculating disparity as the distribution of these variances
-summary(dispRity(dummy_space, metric = variances))
-
-## Calculating disparity as the sum and the product of these variances
-summary(dispRity(dummy_space, metric = c(sum, variances)))
-summary(dispRity(dummy_space, metric = c(prod, variances)))
-```
-
-The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest Euclidean distance in our five dimensional space):
-
-```{r}
-## Calculating the ordinated space's diagonal
-summary(dispRity(dummy_space, metric = diagonal))
-```
-
-> This metric is only a Euclidean diagonal (mathematically valid) if the dimensions within the space are all orthogonal!
-
-#### Centroids metric
-
-The `centroids` metric allows users to measure the position of the different elements compared to a fixed point in the ordinated space.
-By default, this function measures the distance between each element and their centroid (centre point):
-
-```{r}
-## The distribution of the distances between each element and their centroid
-summary(dispRity(dummy_space, metric = centroids))
-
-## Disparity as the median value of these distances
-summary(dispRity(dummy_space, metric = c(median, centroids)))
-```
-
-It is however possible to fix the coordinates of the centroid to a specific point in the ordinated space, as long as it has the correct number of dimensions:
-
-```{r}
-## The distance between each element and the origin of the ordinated space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))
-
-## Disparity as the distance between each element and a specific point in space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
-```
-
-## Summarising dispRity data (plots)
-
-Because of its architecture, printing `dispRity` objects only summarises their content but does not print the disparity value measured or associated analysis (more about this [here](#manipulating-dispRity-objects)).
-To actually see what is in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphical representation of the results.
-
-### Summarising `dispRity` data
-This function is an S3 function (`summary.dispRity`) allowing users to summarise the content of `dispRity` objects that contain disparity calculations.
-
-<!-- DATA LOADER -->
-
-```{r}
-## Example data from previous sections
-crown_stem <- custom.subsamples(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
-## Calculate disparity
-disparity_crown_stem <- dispRity(boot_crown_stem, metric = c(sum, variances))
-
-## Creating time slice subsamples
-time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-## Bootstrapping the time slice subsamples
-boot_time_slices <- boot.matrix(time_slices, bootstraps = 100)
-## Calculate disparity
-disparity_time_slices <- dispRity(boot_time_slices, metric = c(sum, variances))
-
-## Creating time bin subsamples
-time_bins <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages,
-    inc.nodes = TRUE)
-## Bootstrapping the time bin subsamples
-boot_time_bins <- boot.matrix(time_bins, bootstraps = 100)
-## Calculate disparity
-disparity_time_bins <- dispRity(boot_time_bins, metric = c(sum, variances))
-```
-
-These objects are easy to summarise as follows:
-
-```{r}
-## Default summary
-summary(disparity_time_slices)
-```
-
-Information about the number of elements in each subsample and the observed (i.e. non-bootstrapped) disparity are also calculated.
-This is specifically handy when rarefying the data for example:
-
-```{r}
-head(summary(disparity_crown_stem))
-```
-
-The summary functions can also take various options such as:
-
- * `quantile` values for the confidence interval levels (by default, the 50 and 95 quantiles are calculated)
- * `cent.tend` for the central tendency to use for summarising the results (default is `median`)
- * rounding` option corresponding to the number of decimal places to print (default is `2`)
- * `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
-
-These options can easily be changed from the defaults as follows:
-
-```{r}
-## Same as above but using the 88th quantile and the standard deviation as the summary 
-summary(disparity_time_slices, quantile = 88, cent.tend = sd)
-
-## Printing the details of the object and rounding the values to the 5th decimal place
-summary(disparity_time_slices, recall = TRUE, rounding = 5)
-```
-
-Note that the summary table is a `data.frame`, hence it is as easy to modify as any dataframe using `dplyr`.
-You can also export it in `csv` format using `write.csv` or `write_csv` or even directly export into `LaTeX` format using the following;
-
-```{r, eval=FALSE}
-## Loading the xtable package
-require(xtable)
-## Converting the table in LaTeX
-xtable(summary(disparity_time_slices))
-```
-
-### Plotting `dispRity` data
-An alternative (and more fun!) way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
-This function takes the same options as `summary.dispRity` along with various graphical options described in the function help files (see `?plot.dispRity`).
-
-The plots can be of four different types:
-
- * `continuous` for displaying continuous disparity curves
- * `box`, `lines`, and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.
-
-> This argument can be left empty. In this case, the algorithm will automatically detect the type of subsamples from the `dispRity` object and plot accordingly.
-
-It is also possible to display the number of elements in each subsample (as a horizontal dotted line) using the option `elements = TRUE`.
-Additionally, when the data is rarefied, one can indicate which level of rarefaction to display (i.e. only display the results for a certain number of elements) by using the `rarefaction` argument.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(mfrow = c(2, 2), bty = "n")
-
-## Plotting continuous disparity results
-plot(disparity_time_slices, type = "continuous")
-
-## Plotting discrete disparity results
-plot(disparity_crown_stem, type = "box")
-
-## As above but using lines for the rarefaction level of 20 elements only
-plot(disparity_crown_stem, type = "line", rarefaction = 20)
-
-## As above but using polygons while also displaying the number of elements
-plot(disparity_crown_stem, type = "polygon", elements = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Since `plot.dispRity` uses the arguments from the generic `plot` method, it is of course possible to change pretty much everything using the regular plot arguments:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some classic options from plot
-plot(disparity_time_slices, col = c("blue", "orange", "green"),
-    ylab = c("Some measurement"), xlab = "Some other measurement",
-    main = "Many options...", ylim = c(5, 0), xlim = c(4, 0))
-
-## Adding a legend
-legend("topleft", legend = c("Central tendency",
-                             "Confidence interval 1",
-                            "Confidence interval 2"),
-      col = c("blue", "orange", "green"), pch = 19)
-
-## Resetting graphical parameters
-par(op)
-```
-
-In addition to the classic `plot` arguments, the function can also take arguments that are specific to `plot.dispRity` like adding the number of elements or rarefaction level (as described above), and also changing the values of the quantiles to plot as well as the central tendency.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some plot.dispRity arguments
-plot(disparity_time_slices, quantile = c(seq(from = 10, to = 100, by = 10)),
-    cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
-    ylab = c("Disparity", "Diversity"), time.subsamples = FALSE,
-    xlab = "Time (in in units from past to present)", observed = TRUE,
-    main = "Many more options...")
-
-## Resetting graphical parameters
-par(op)
-```
-
-> Note that the argument `observed = TRUE` allows to plot the disparity values calculated from the non-bootstrapped data as crosses on the plot.
-
-For comparing results, it is also possible to add a plot to the existent plot by using `add = TRUE`:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the continuous disparity with a fixed y axis
-plot(disparity_time_slices, ylim = c(1, 4))
-## Adding the discrete data
-plot(disparity_time_bins, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
-    add = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Finally, if your data has been fully rarefied, it is also possible to easily look at rarefaction curves by using the `rarefaction = TRUE` argument:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the rarefaction curves
-plot(disparity_crown_stem, rarefaction = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-## Testing disparity hypotheses
-
-The `dispRity` package allows users to apply statistical tests to the calculated disparity to test various hypotheses.
-The function `test.dispRity` works in a similar way to the `dispRity` function: it takes a `dispRity` object, a `test` and a `comparisons` argument.
-
-The `comparisons` argument indicates the way the test should be applied to the data:
-
- * `pairwise` (default): to compare each subsample in a pairwise manner
- * `referential`: to compare each subsample to the first subsample
- * `sequential`: to compare each subsample to the following subsample
- * `all`: to compare all the subsamples together (like in analysis of variance)
-
-It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsample names to create personalised tests.
-Some other tests implemented in `dispRity` such as the `dispRity::null.test` have a specific way they are applied to the data and therefore ignore the `comparisons` argument. 
-<!-- Add sequential test one day! -->
-
-The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
-It can be a common statistical test function (e.g. `stats::t.test`), a function implemented in `dispRity` (e.g. see `?null.test`) or any function defined by the user.
-
-This function also allows users to correct for Type I error inflation (false positives) when using multiple comparisons via the `correction` argument. 
-This argument can be empty (no correction applied) or can contain one of the corrections from the `stats::p.adjust` function (see `?p.adjust`).
-
-Note that the `test.dispRity` algorithm deals with some classical test outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
-It is, however, possible to get the full detailed output by using the options `details = TRUE`.
-
-```{r}
-## T-test to test for a difference in disparity between crown and stem mammals
-test.dispRity(disparity_crown_stem, test = t.test)
-
-## Performing the same test but with the detailed t.test output
-test.dispRity(disparity_crown_stem, test = t.test, details = TRUE)
-
-## Wilcoxon test applied to time sliced disparity with sequential comparisons, with Bonferroni correction
-test.dispRity(disparity_time_slices, test = wilcox.test,
-              comparisons = "sequential", correction = "bonferroni")
-
-## Measuring the overlap between distributions in the time bins (using the
-## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)
-test.dispRity(disparity_time_bins, test = bhatt.coeff)
-```
-
-It is also possible to apply some more *complex* tests that have their own output classes (like `stats::lm`).
-
-The results can then be analysed as usual using the associated `summary` S3 method:
-
-```{r}
-## Performing and linear model applied to the same data
-(slice_lm <- test.dispRity(disparity_time_slices, test = lm,
-                            comparisons = "all"))
-
-## The output is a regular `lm` output
-class(slice_lm)
-
-## This output can be summarised using summary
-summary(slice_lm)
-```
-
-Of course, due to the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
-The only condition is that the test can be applied to at least two distributions.
-In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm` for example).
-
-## Disparity as a distribution
-
-Disparity is often regarded as a summary value of the position of the all elements in the ordinated space.
-For example, the sum of variances, the product of ranges or the median distance between the elements and their centroid will summarise disparity as a single value.
-This value can be pseudo-replicated (bootstrapped) to obtain a distribution of the summary metric with estimated error.
-However, another way to perform disparity analysis is to use the *whole distribution* rather than just a summary metric (e.g. the variances or the ranges).
-
-This is possible in the `dispRity` package by calculating disparity as a dimension-level 2 metric only!
-Let's have a look using our previous example of bootstrapped time slices but by measuring the distances between each taxon and their centroid as disparity.
-
-```{r}
-## Measuring disparity as a whole distribution
-disparity_centroids <- dispRity(boot_time_slices, metric = centroids)
-```
-
-The resulting disparity object is of dimension-level 2, so it can easily be transformed into a dimension-level 1 object by, for example, measuring the median distance of all these distributions:
-
-```{r}
-## Measuring median disparity in each time slice
-disparity_centroids_median <- dispRity(disparity_centroids, metric = median)
-```
-
-And we can now compare the differences between these methods:
-
-```{r}
-## Summarising both disparity measurements:
-## The distributions:
-summary(disparity_centroids)
-## The summary of the distributions (as median)
-summary(disparity_centroids_median)
-```
-
-We can see that the summary message for the distribution is slightly different than before.
-Here `summary` also displays the observed central tendency (i.e. the central tendency of the measured distributions).
-Note that, as expected, this central tendency is the same in both metrics!
-
-Another, maybe more intuitive way, to compare both approaches for measuring disparity is to plot the distributions:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(bty = "n", mfrow = c(1, 2))
-
-## Plotting both disparity measurements
-plot(disparity_centroids, ylab = "Distribution of all the distances")
-plot(disparity_centroids_median,
-     ylab = "Distribution of the medians of all the distances")
-
-par(op)
-```
-
-We can then test for differences in the resulting distributions using `test.dispRity` and the `bhatt.coeff` test as described above.
-
-```{r}
-## Probability of overlap in the distribution of medians
-test.dispRity(disparity_centroids_median, test = bhatt.coeff)
-```
-
-In this case, we are looking at the probability of overlap of the distribution of median distances from centroids among each pair of time slices.
-In other words, we are measuring whether the medians from each bootstrap pseudo-replicate for each time slice overlap.
-But of course, we might be interested in the actual distribution of the distances from the centroid rather than simply their central tendencies.
-This can be problematic depending on the research question asked since we are effectively comparing non-independent medians distributions (because of the pseudo-replication).
-
-One solution, therefore, is to look at the full distribution:
-
-```{r}
-## Probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff)
-```
-
-These results show the actual overlap among all the measured distances from centroids concatenated across all the bootstraps.
-For example, when comparing the slices 120 and 80, we are effectively comparing the 5 \times 100 distances (the distances of the five elements in slice 120 bootstrapped 100 times) to the 19 \times 100 distances from slice 80.
-However, this can also be problematic for some specific tests since the _n_ \times 100 distances are also pseudo-replicates and thus are still not independent.
-
-A second solution is to compare the distributions to each other _for each replicate_:
-
-```{r}
-## Boostrapped probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff, concatenate = FALSE)
-```
-
-These results show the median overlap among pairs of distributions in the first column (`bhatt.coeff`) and then the distribution of these overlaps among each pair of bootstraps.
-In other words, when two distributions are compared, they are now compared for each bootstrap pseudo-replicate, thus effectively creating a distribution of probabilities of overlap.
-For example, when comparing the slices 120 and 80, we have a mean probability of overlap of 0.28 and a probability between 0.18 and 0.43 in 50\% of the pseudo-replicates.
-Note that the quantiles and central tendencies can be modified via the `conc.quantiles` option.
-
-## Ecological studies of disparity
-<!-- NC: I wonder if this might be better just as a "hey we can do ecology stuff too and here is a link to a demo" rather than a new section?
-The manual is loooooooooong so splitting some stuff out might be good?
--->
-
-Disparity analyses are not restricted to palaeobiology analysis.
-This package is designed for dealing with ordinated matrices, no matter which data was used in the ordination.
-Here is an short example of disparity analysis using ecological data from Deirdre McClean (unpubl.).
-
-The dataset contains an ordinated matrix of 20 dimensions (columns) for 40 elements (rows).
-The elements are different experimental plots with discrete variations of nutrient enrichment, depth, and freshwater benthic invertebrates.
-<!-- NC: Explain a little more about what the data are here as I don't entirely get it. EG what's a treatement etc? Just a really simple explanataion is perfect -->
-
-```{r}
-## Loading the data
-data(McClean_data)
-
-## Ordinated matrix
-ordinated_matrix <- McClean_data$ordination
-
-## Treatments and depths
-treatments <- McClean_data$treatment
-depth <- McClean_data$depth
-```
-
-### Classic ecological "disparity" analysis
-
-A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different experimental plots differ in their values for nutrient enrichment, depth, and freshwater benthic invertebrates.
-<!-- NC: or similar just to make it clearer -->
-
-```{r, fig.width=6, fig.height=6}
-## Setting the colors that will represent the treatments
-cols <- sub("a", "red", treatments)
-cols <- sub("b", "blue", cols)
-## Setting the symbols that will represent the depth
-pchs <- sub(1, 16, depth)
-pchs <- as.numeric(sub(2, 17, pchs))
-
-## Graphical option
-op <- par(bty = "n")
-## A classic 2D ordination plot
-plot(ordinated_matrix[, 1:2], col = cols, pch = pchs,
-     xlab = "PC 1", ylab = "PC 2",
-     xlim = range(ordinated_matrix[, 1:2]) + c(0, 100))
-par(op)
-```
-
-This shows the distribution of the experimental plots along the two first axis of variation of the ordinated distance matrix (i.e. the first two dimensions).
-At a first glance, it seems difficult to see a clear effect of the treatments (blue or red) or the depth (rounds or triangles) as the different experimental plots with the same parameters don't seem to cluster together.
-However, this may be because this plot ignores the 18 other dimensions of the ordination!
-Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e.
-the values on the first dimension do not represent a continuous trait (e.g.
-depth), instead they just represent the ordinations of correlations between the data and some factors.
-
-Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.
-
-### Multidimensional analysis with `dispRity`
-
-#### Splitting the data with respect to the different factors
-
-The first step is to create different subsamples that represent subsamples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
-Each of these subsamples will contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
-In our example, we are going to group the elements according to their depth and treatment.
-
-```{r}
-## Creating a table that contain the elements and their attributes
-factors <- as.data.frame(matrix(data = c(treatments, depth),
-          nrow = nrow(ordinated_matrix), ncol = 2, byrow = FALSE,
-          dimnames = list(rownames(ordinated_matrix))))
-names(factors) <- c("Treat", "Depth")
-head(factors)
-```
-
-Second, let’s split the data according by depth and treatment to create the subsamples of the ordinated space by using the `custom.subsamples` function:
-
-```{r}
-## Splitting the ordinated space into four subsamples
-(customised_subsamples <- custom.subsamples(ordinated_matrix, factors))
-```
-
-#### Calculating disparity
-
-In this example, we are going to define disparity as the n-dimensional ellipsoid volume (see `?hyper.volume`).
-
-```{r}
-## Calculating disparity as the ellipsoid volume of each group
-disparity <- dispRity(customised_subsamples, metric = diagonal)
-summary(disparity)
-```
-
-We can also bootstrap the data to test the robustness of the measured disparity to outliers.
-Also, as we can see, each subsample has different numbers of elements.
-It might also be interesting to rarefy the data to have only subsamples with the same number of elements.
-
-```{r}
-## Bootstrapping the subsamples and using rarefaction
-## (i.e. only re-sampling 17 elements each time)
-bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 100,
-                                 rarefaction = 17)
-```
-
-We can now run a more robust disparity analysis using the bootstrapped data:
-
-```{r}
-## Calculating the bootstrapped disparity
-disparity <- dispRity(bootstrapped_data, metric = diagonal)
-```
-
-#### Summarising and displaying the data
-
-Now let's summarise this data:
-
-```{r, fig.width=12, fig.height=6, eval=TRUE}
-## Summary table of the data
-summary(disparity)
-
-## Graphical options
-op <- par(mfrow = (c(1, 2)), bty = "n")
-
-## Plotting the bootstrapped disparity
-plot(disparity, main = "Bootstrapped", las = 2, xlab = "")
-
-## Plotting the rarefied disparity
-plot(disparity, rarefaction = 17, main = "Rarefied", las = 2, xlab = "")
-
-par(op)
-```
-
-As we can see, there seems to be no strong effect of the number of experimental plots in each subsample (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
-
-#### Testing our hypotheses
-
-Finally, we can test our hypothesis (whether the water treatment at certain depths alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.
-<!-- Ah so this is what you are testing! Make this clearer in the intro bit to this section -->
-
-```{r}
-## Testing the effect of treatment and depth on the bootstrapped data
-summary(test.dispRity(disparity, test = aov, comparisons = "all"))
-```
-
-## Simulating discrete morphological data
-
-The `dispRity` package also allows users to simulate discrete morphological data matrices.
-
-### Quick go through
-The function basically intakes a phylogenetic tree, the number of required characters and the evolutionary model and a function from which to draw the rates.
-The package then proposes a function for quickly checking the matrix' phylogenetic signal using parsimony.
-
-```{r}
-set.seed(3)
-## Setting the starting tree with 15 taxa as a random coalescent tree
-my_tree <- rcoal(15)
-
-## Generating a matrix with 100 characters (85% binary and 15% three states) and
-## an equal rates model with a gamma rates distribution (0.5, 1) with no 
-## invariant characters.
-my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
-    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)
-
-## The head of the matrix
-my_matrix[1:5, 1:10]
-
-## Checking the matrix properties with a quick Maximum Parsimony tree search
-check.morpho(my_matrix, my_tree)
-```
-
-Note that this example produces a tree with a great consistency index and an identical topology as the random coalescent tree!
-Nearly too good to be true...
-
-### In details
-
-Basically, `sim.morpho` is really flexible and intakes a lot of different arguments to allow to simulate realistic matrix.
-It has three implemented models: `"ER"` for Equal Rates (the M*k* model); the `"HKY"` one which is the molecular HKY model but transforms pyrines in 0's and purimidines in 1's; or the `"mixed"` model that randomly uses an `"ER"` or and `"HKY"` for the binary characters and `"ER"` for the multistates (>2) characters.
-Both models intakes specific distributions for their rates or substitution models.
-These distributions should be passed to these arguments in the format of `c(sampler_function, distribution_parameters)` where the the sampler function is a the random generation function of that distribution (e.g.
-`rnorm`, `runif`, etc...) and the parameters are any parameters to be passed to this function.
-
-The `check.morpho` runs a quick maximum parsimony tree using the `phangorn` parsimony algorithm.
-It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g.
-the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree.
-
-### Parameters for a good(ish) matrix
-There are many parameters that can realise a ``realistic'' matrix (meaning not to different from the input tree with a consistency and retention index close to what's in the literature) but because of the randomness of the generation (the code generates random matrices after all!) not all end up in creating good matrices.
-The following parameters however, seems to generate a fairly ``realistic'' with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three states characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
-
-```{r}
-set.seed(0)
-## tree
-my_tree <- rcoal(15)
-## matrix
-morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
-    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
-check.morpho(morpho_mat, my_tree)
-```
-
-### `space.maker`
-Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
-This function allows to simulated multidimensional spaces given a certain number of properties.
-It takes as arguments the number of elements (data points) and dimensions to create the space but also allows more fine tuning data simulation: it is possible to give a specific distribution to each dimensions, provide a correlation matrix to link the dimensions or even providing an *a priori* distribution of the variance per distributions!
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(mfrow = (c(2,2)), bty = "n")
-## Visualising 3D spaces
-require(scatterplot3d)
-
-## A cube space
-scatterplot3d(space.maker(2500, 3, runif), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Uniform cube space")
-
-## A plane space
-scatterplot3d(space.maker(2500, 3, c(runif, runif, runif),
-              arguments = list(list(min = 0, max = 0), NULL, NULL)), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Uniform plane space")
-
-## An ellipsoid space (=a spheric space with correlation)
-cor_matrix <- matrix(cbind(1,0.8,0.2, 0.8,1,0.7, 0.2,0.7,1), nrow = 3)
-scatterplot3d(space.maker(2500, 3, rnorm, cor.matrix = cor_matrix), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Normal ellipsoid space")
-
-## A cylindrical space with a decreasing variance per axis
-scatterplot3d(space.maker(2500, 3, c(rnorm, rnorm, runif),
-              scree = c(0.7, 0.2, 0.1)), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Normal cylindrical space")
-
-## Resetting the graphic parameters
-par(op)
-```
-
-# Glossary <a name="glossary"></a>
-
--   **Ordinated space**. The mathematical multidimensional object that will be analysed with this package.
-    In morphometrics, this is often referred to as the morphospace.
-    However it may also be referred to as the cladisto-space for cladistic data or the eco-space for ecological data etc.
-    In practice, this term designates an ordinated matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
-
--   **Elements**. The rows of the ordinated space. Elements can be taxa, field sites, countries etc.
-
--   **Dimensions**. The columns of the ordinated space. The dimensions are referred to as axes of variation, or principal components, for ordinated spaces obtained from a PCA.
-
--   **Subsamples**. Sub-samples of the ordinated space.
-    A sub-sample (or subsamples) contains the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
-    For example, if our ordinated space is composed of birds and mammals (the elements) and 50 principal components (the dimensions), we can create two subsamples containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
-
-# The package's guts
-
-## Manipulating `dispRity` objects
-
-Disparity analysis involves a lot of manipulation of many matrices (especially when bootstrapping) which can be impractical to visualise and will quickly overwhelm your `R` console.
-Even the simple Beck and Lee 2014 example above produces an object with > 72 lines of lists of lists of matrices!
-
-Therefore `dispRity` uses a specific class of object called a `dispRity` object.
-These objects allow users to use S3 method functions such as `summary.dispRity`, `plot.dispRity` and `print.dispRity`. 
-`dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g. `sort.dispRity`, `extract.dispRity` see the full list in the next section).
-These functions modify the `dispRity` object without having to delve into its complex structure!
-The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).
-
-```{r}
-## Loading the example data
-data(disparity)
-
-## What is the class of the median_centroids object?
-class(disparity)
-
-## What does the object contain?
-names(disparity)
-
-## Summarising it using the S3 method print.dispRity
-disparity
-```
-
-Note that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:
-
-```{r, eval=FALSE}
-## Display the full object
-print(disparity, all = TRUE)
-## This is more nearly ~ 5000 lines on my 13 inch laptop screen!
-```
-
-## `dispRity` utilities
-
-The package also provides some utilities functions to facilitate multidimensional analysis.
-
-### `dispRity` object utilities  <a name="dispRity.utilities"></a>
-The first subsamples of utilities are functions for manipulating `dispRity` objects:
-
-#### `make.dispRity`
-This function creates empty `dispRity` objects.
-
-```{r}
-## Creating an empty dispRity object
-make.dispRity()
-
-## Creating an "empty" dispRity object with a matrix
-(disparity_obj <- make.dispRity(matrix(rnorm(20), 5, 4)))
-```
-
-####  `fill.dispRity`
-This function initialises a `dispRity` object and generating it's call properties.
-
-```{r}
-## The dispRity object's call is indeed empty
-disparity_obj$call
-
-## Filling an empty disparity object (that needs to contain at least a matrix)
-(disparity_obj <- fill.dispRity(disparity_obj))
-
-## The dipRity object has now the proper minimal attributes
-disparity_obj$call
-```
-
-####  `matrix.dispRity`
-This function extracts a specific matrix from a disparity object. The matrix can be one of the bootstrapped or/and rarefied one.
-
-```{r}
-## Extracting the matrix containing the coordinates of the elements at time 50
-str(matrix.dispRity(disparity, "50"))
-
-## Extracting the 3rd boostrapped matrix with the 2nd rarefaction level
-## (15 elements) from the second group (80 Mya)
-str(matrix.dispRity(disparity, subsamples = 1, bootstrap = 3, rarefaction = 2))
-```
-
-#### `get.subsamples.dispRity`
-This function creates a dispRity object that contains only elements from one specific subsamples.
-
-```{r, eval=FALSE}
-## Extracting all the data for the crown mammals
-(crown_mammals <- get.subsamples.dispRity(disp_crown_stemBS, "Group.crown"))
-
-## The object keeps the properties of the parent object but is composed of only one subsamples
-length(crown_mammals$subsamples)
-```
-
-#### `extract.dispRity`
-This function extracts the calculated disparity values of a specific matrix.
-
-```{r, eval=FALSE}
-## Extracting the observed disparity (default)
-extract.dispRity(disparity)
-
-## Extracting the disparity from the bootstrapped values from the
-## 10th rarefaction level from the second subsamples (80 Mya)
-extract.dispRity(disparity, observed = FALSE, subsamples = 2, rarefaction = 10)
-```
-
-#### `scale.dispRity`
-This is the S3 method of `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object.
-
-```{r, eval=FALSE}
-## Getting the disparity values of the time subsamples
-head(summary(disparity))
-
-## Scaling the same disparity values
-head(summary(scale(disparity, scale = TRUE)))
-
-## Scaling and centering:
-head(summary(scale(disparity, scale = TRUE, center = TRUE)))
-```
-
-#### `sort.dispRity`
-This is the S3 method of `sort` for sorting the subsamples alphabetically (default) or following a specific pattern.
-
-```{r, eval=FALSE}
-## Sorting the disparity subsamples in inverse alphabetic order
-head(summary(sort(disparity, decreasing = TRUE)))
-
-## Customised sorting
-head(summary(sort(disparity, sort = c(7, 1, 3, 4, 5, 2, 6))))
-```
-
-<!-- ## Modularity
-
-### Making your own metrics
-
-### Making your own tests
-
-## Where is this going?
- -->
-<!-- ## Running `dispRity` in parallel
-
-The computationally intensive function `dispRity` have a `parallel` option to speed up their calculations.
-
-This option requires the package `snow` and takes arguments that are to be passed to  -->

### 2017-08-12 gitbook/index.Rmd

--- 2017-08-10 doc/dispRity-manual.Rmd
@@
 title: "dispRity manual"
-author: "Thomas Guillerme"
+author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
 date: "`r Sys.Date()`"
-output:
-    rmarkdown::html_vignette:
-        toc: true
-        toc_depth: 2
-bibliography: References.bib
-bst: sysbio.bst
-vignette: >
-  %\VignetteIndexEntry{dispRity manual}
-  %\VignetteEngine{knitr::rmarkdown}
-  \usepackage[utf8]{inputenc}
+site: bookdown::bookdown_site
+output: bookdown::gitbook
+documentclass: book
+link-citations: yes
+github-repo: TGuillerme/dispRity
+description: "dispRity R package vignette"
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
 
@@
 This manual is based on the version `1.0`.
-
-# Introduction
 
@@
 
-You can install this package easily if you use the latest version of `R` (> 3.3.0) and `devtools`.
+You can install this package easily if you use the latest version of `R` (> 4.0) and `devtools`.
 
@@
 A future proper version of the latest package release, this manual and an associated methods paper should be submitted soon(ish!).
-
-# Getting started
-
-## What sort of data does dispRity work with?
-
-Disparity can be estimated from pretty much any matrix.
-Classically, however, it is measured from ordinated matrices.
-These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).
-
-### Ordination matrices from `Claddis`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
-if(!require(Claddis)) install.packages("Claddis")
-```
-
-`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
-For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:
-
-```{r}
-require(Claddis)
-
-## Ordinating the example data from Claddis
-Claddis.ordination(Michaux1989) 
-```
-
-Note that several options are available, namely which type of distance should be computed.
-See more info in the function manual (`?Claddis.ordination`).
-Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.
-
-### Ordination matrices from `geomorph`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if geomorph is installed BUILD ONLY
-if(!require(geomorph)) install.packages("geomorph")
-```
-
-You can also easily use data from `geomorph` using the `geomorph.ordination` function.
-This function simply takes Procrustes aligned data and performs an ordination:
-
-```{r}
-require(geomorph)
-
-## Loading the plethodon dataset
-data(plethodon)
-
-## Performing a Procrustes transform on the landmarks
-procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)
-
-## Ordinating this data
-geomorph.ordination(procrustes)[1:5,1:5]
-```
-
-Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
-Additionally you can give the function a `geomorph.data.frame` object.
-If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsamples)!
-
-```{r}
-## Using a geomorph.data.frame
-geomorph_df <- geomorph.data.frame(procrustes,
-     species = plethodon$species, site = plethodon$site)
-
-## Ordinating this data and making a dispRity object
-geomorph.ordination(geomorph_df)
-```
-
-More about these `dispRity` objects below!
-
-### Other kinds of ordination matrices
-
-If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordination matrices by using the following functions from the `stats` package.
-Here is how to do it for the following types of matrices:
-
- * Multivariate matrices (principal components analysis; PCA)
-
-```{r}
-## A multivariate matrix
-head(USArrests)
-
-## Ordinating the matrix using `prcomp` 
-ordination <- prcomp(USArrests)
-
-## Selecting the ordinated matrix
-ordinated_matrix <- ordination$x
-head(ordinated_matrix)
-```
-  This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, see the `?princomp` function.
-
- * Distance matrices (classical multidimensional scaling; MDS)
-
-```{r}
-## A matrix of distances between cities
-str(eurodist)
-
-## Ordinating the matrix using cmdscale() with k = 5 dimensions 
-ordinated_matrix <- cmdscale(eurodist, k = 5)
-head(ordinated_matrix)
-```
-
-  This results in a ordinated matrix with European cities as elements and five dimensions.
-
-Of course any other method for creating the ordination matrix is totally valid, you can also not use any ordination at all!
-The only requirements for the `dispRity` functions is that the input is a matrix with elements as rows and dimensions as columns.
-
-## Performing a simple dispRity analysis
-
-Two `dispRity` functions allow users to run an analysis pipeline simply by inputting an ordination matrix.
-These functions allow users to either calculate the disparity through time (`dispRity.through.time`) or the disparity of user-defined groups (`dispRity.per.group`).
-
-### Example data
-
-To illustrate these functions, we will use data from @beckancient2014.
-This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
-
-```{r, fig.width=7, fig.height=7}
-## Loading the ordinated matrices
-data(BeckLee_mat50)
-data(BeckLee_mat99)
-
-## The first five taxa and dimensions of the 50 taxa matrix
-head(BeckLee_mat50[, 1:5])
-
-## The first five taxa and dimensions of the 99 taxa + ancestors matrix
-BeckLee_mat99[c(1, 2, 98, 99), 1:5]
-
-## Loading a list of first and last occurrence dates for the fossils
-data(BeckLee_ages)
-head(BeckLee_ages)
-
-## Loading and plotting the phylogeny
-data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.8) 
-axisPhylo(root = 140)
-nodelabels(cex = 0.5)
-```
-
-Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).
-
-### Disparity through time
-
-The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
-This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
-Of course all the defaults can be changed if required, more on this later.
-
-For a disparity through time analysis, you will need:
-
-  * An ordinated matrix (we covered that above)
-  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element.
-To give your tree a root time (i.e.
-an age for the root), you can simply do `my_tree$root.time <- my_age`.
-  * The required number of time subsamples (here `time = 3`)
-  * Your favourite disparity metric (here the sum of variances)
-
-Using the @beckancient2014 data described [above](#example-data):
-
-```{r}
-## Measuring disparity through time
-disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
-                                        time = 3, metric = c(sum, variances))
-```
-
-This generates a `dispRity` object (see [here](#manipulating-dispRity-objects) for technical details).
-When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:
-
-```{r}
-## Print the disparity_data object
-disparity_data
-```
-
-We asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
-
-We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):
-
-```{r, fig.width=7, fig.height=7}
-## Summarising disparity through time
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data, type = "continuous")
-
-## Testing for an difference among the time bins
-disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
-summary(disp_lm)
-
-## Note that ANOVA assumptions were not checked here!
-```
-
-Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
-You can also directly explore the specific function help files within R and navigate to related functions.
-
-### Disparity among groups
-
-The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
-For example, you could ask if there is a difference in disparity between two groups?
-
-To perform such an analysis, you will need:
-
- * An matrix with rows as elements and columns as dimensions (always!)
- * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix.
-For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4.
-Note that elements can be present in multiple groups at once.
- * Your favourite disparity metric (here the sum of variances)
-
-Using the @beckancient2014 data described [above](#example-data):
-
-```{r}
-## Creating the two groups (crown versus stem) as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44))
-
-## Measuring disparity for each group
-disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
-                                     metric = c(sum, variances))
-```
-
-We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).
-
-```{r, fig.width=7, fig.height=7}
-## Print the disparity_data object
-disparity_data
-
-## Summarising disparity in the different groups
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data)
-
-## Testing for a difference between the groups
-test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
-```
-
-# Specific tutorials
-
-The following section contains information specific to some functions.
-If any of your questions are not covered in these sections, please refer to the function help files in R, send me an email (<guillert@tcd.ie>), or raise an issue on GitHub.
-The several tutorials below describe specific functionalities of certain functions; please always refer to the function help files for the full function documentation!
-
-## Time slicing
-
-The function `time.subsamples` allows users to divide the matrix into different time subsamples or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
-Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.
-
-Two types of time subsamples can be performed by using the `method` option:
-
- *  Discrete time subsamples (or time-binning) using `method = discrete`
- *  Continuous time subsamples (or time-slicing) using `method = continuous`
-
-For the time-slicing method details see @GuillermeSTD. <!-- @@@ Change cite appropriately! -->
-<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
-For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:
-
- *  Defining the boundaries of the time bins (when `method = discrete`)
- *  Defining the time slices (when `method = continuous`)
-
-Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
-Additionally, it is also possible to input a dataframe containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age, so taxa can appear in more than one time bin/slice.
-
-Here is an example for `method = discrete`:
-
-```{r, eval=TRUE}
-## Generating three time bins containing the taxa present every 40 Ma
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0))
-```
-
-Note that we can also generate equivalent results by just telling the function that we want three time-bins as follow:
-
-```{r, eval=TRUE}
-## Automatically generate three equal length bins:
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = 3)
-```
-
-In this example, the taxa were split inside each time-bin according to their age.
-However, the taxa here are considered as single points in time.
-It is totally possible that some taxa could have had longer longevity and that they exist in multiple time bins.
-In this case, it is possible to include them in more than one bin by providing a table of first and last occurrence dates (FAD/LAD).
-This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:
-
-```{r, eval=TRUE}
-## Displaying the table of first and last occurrence dates for each taxa
-head(BeckLee_ages)
-
-## Generating time bins including taxa that might span between them
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
-```
-
-When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as the root age plus 1\% of the tree length, as long as at least three elements/taxa are present at that point in time.
-The algorithm adds an extra 1\% tree length until reaching the required minimum of three elements.
-It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance among tips *and* nodes.
-
-For the time-slicing method (`method = continuous`), the idea is fairly similar.
-This option, however, requires a matrix that contains the ordinated distance among taxa *and* nodes and an extra argument describing the assumed evolutionary model (via the `model` argument).
-This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node, meaning that a decision must be made about what the value for the branch should be.
-The model can be one of the following:
-
- *  `acctran` where the data chosen along the branch is always the one of the descendant
- *  `deltran` where the data chosen along the branch is always the one of the ancestor
- *  `punctuated` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
- *  `gradual` where the data chosen along the branch is either the descendant or the ancestor depending on branch length
-
-```{r, eval=TRUE}
-## Generating four time slices every 40 million years under a model of gradual evolution
-time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-
-## Generating four time slices automatically
-time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree,
-    method = "continuous", model = "gradual", time = 4, FADLAD = BeckLee_ages)
-```
-
-## Customised subsamples
-
-Another way of separating elements into different categories is to use customised subsamples as briefly explained [above](#disparity-among-groups).
-This function simply takes the list of elements to put in each group (whether they are the actual element names or their position in the matrix).
-
-```{r, eval=TRUE}
-## Creating the two groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44))
-
-## Separating the dataset into two different groups
-custom.subsamples(BeckLee_mat50, group = mammal_groups)
-```
-
-Elements can easily be assigned to different groups if necessary!
-
-```{r, eval=FALSE}
-## Creating the three groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44).
-                      "all" = c(1:50))
-```
-
-## Bootstraps and rarefactions
-
-One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results are, and how sensitive they are to outliers in the dataset.
-This can be achieved using the function `boot.matrix` to bootstrap and/or rarefy the data.
-The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):
-
-```{r, eval=TRUE}
-## Default bootstrapping
-boot.matrix(data = BeckLee_mat50)
-```
-
-The number of bootstrap replicates can be defined using the `bootstraps` option.
-The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
-Currently two algorithms are implemented:
-
- * `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any *m* elements drawn from the data)
- * `single` where only one random element is replaced by one other random element for each pseudo-replicate
-
-```{r, eval=TRUE}
-## Bootstrapping with the single bootstrap method
-boot.matrix(BeckLee_mat50, boot.type = "single")
-```
-
-This function also allows users to rarefy the data using the `rarefaction` argument.
-Rarefaction allows users to limit the number of elements to be drawn at each bootstrap replication.
-This is useful if, for example, one is interested in looking at the effect of reducing the number of elements on the results of an analysis.
-
-This can be achieved by using the `rarefaction` option that draws only *n-x* at each bootstrap replicate (where *x* is the number of elements not sampled).
-The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsample to a minimum of three elements).
-It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
-
-```{r, eval=TRUE}
-## Bootstrapping with the full rarefaction
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
-
-## Or with a set number of rarefaction levels
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8, 3))
-```
-
-One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
-When missing, all dimensions from the ordinated matrix are used.
-
-```{r, eval=TRUE}
-## Using the first 50% of the dimensions
-boot.matrix(BeckLee_mat50, dimensions = 0.5)
-
-## Using the first 10 dimensions
-boot.matrix(BeckLee_mat50, dimensions = 10)
-```
-
-Of course, one could directly supply the subsamples generated above (using `time.subsamples` or `custom.subsamples`) to this function.
-
-```{r, eval=TRUE}
-## Creating subsamples of crown and stem mammals
-crown_stem <- custom.subsamples(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot.matrix(crown_stem, bootstraps = 200, rarefaction = TRUE)
-
-## Creating time slice subsamples
-time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-                               method = "continuous", model = "gradual", 
-                               time = c(120, 80, 40, 0),
-                               FADLAD = BeckLee_ages)
-
-## Bootstrapping the time slice subsamples
-boot.matrix(time_slices, bootstraps = 100)
-```
-
-## Disparity metrics
-
-There are many ways of measuring disparity!
-In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
-For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @Wills1994.
-Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.
-
-Because of this great diversity of metrics, the package `dispRity` does not have one way to measure disparity but rather proposes to facilitate users in defining their own disparity metric that will best suit their particular analysis.
-In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
-However the `metric` argument has to follow certain rules:
-
-1.  It must be composed from one to three `function` objects;
-2.  The function(s) must take as a first argument a `matrix` or a `vector`;
-3.  The function(s) must be of one of the three dimension-levels described below;
-4.  At least one of the functions must be of dimension-level 1 or 2 (see below).
-
-### The function dimension-levels
-
-The metric function dimension-levels determine the "dimensionality of decomposition" of the input matrix.
-In other words, each dimension-level designates the dimensions of the output, i.e. either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimension.
-
-![Illustration of the different dimension-levels of functions with an input `matrix`](dispRity_fun.png)
-
-#### Dimension-level 1 functions
-
-A dimension-level 1 function will decompose a `matrix` or a `vector` into a single value:
-
-```{r}
-## Creating a dummy matrix
-dummy_matrix <- matrix(rnorm(12), 4, 3)
-
-## Example of dimension-level 1 functions
-mean(dummy_matrix)
-median(dummy_matrix)
-```
-
-Any summary metric such as mean or median are good examples of dimension-level 1 functions as they reduce the matrix to a single dimension (i.e. one value).
-
-#### Dimension-level 2 functions
-
-A dimension-level 2 function will decompose a `matrix` into a `vector`.
-
-```{r}
-## Defining the function as the product of rows
-prod.rows <- function(matrix) apply(matrix, 1, prod)
-
-## A dimension-level 2 metric
-prod.rows(dummy_matrix)
-```
-
-Several dimension-level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculate the variance or the range of each dimension of the ordinated matrix respectively.
-
-#### Dimension-level 3 functions
-
-Finally a dimension-level 3 function will transform the matrix into another matrix.
-Note that the dimension of the output matrix doesn't need to match the the input matrix:
-
-```{r}
-## A dimension-level 3 metric
-var(dummy_matrix)
-
-## A dimension-level 3 metric with a forced matrix output
-as.matrix(dist(dummy_matrix))
-```
-
-### `make.metric`
-
-Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the Euclidean distance between each element and the centroid of the ordinated space.
-The `make.metric` function implemented in `dispRity` is designed to help test and find the dimension-level of the functions.
-This function tests:
-
-1.  If your function can deal with a `matrix` or a `vector` as an input;
-2.  Your function's dimension-level according to its output (dimension-level 1, 2 or 3, see above);
-3.  Whether the function can be implemented in the `dispRity` function (the function is fed into a `lapply` loop).
-
-For example, let's see if the functions described above are the right dimension-levels:
-
-```{r}
-## Which dimension-level is the mean function? And can it be used in dispRity?
-make.metric(mean)
-
-## Which dimension-level is the prod.rows function? And can it be used in dispRity?
-make.metric(prod.rows)
-
-## Which dimension-level is the var function? And can it be used in dispRity?
-make.metric(var)
-```
-
-A non verbose version of the function is also available.
-This can be done using the option `silent = TRUE` and will simply output the dimension-level of the metric.
-
-```{r}
-## Testing whether mean is dimension-level 1
-if(make.metric(mean, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-## Testing whether var is dimension-level 1
-if(make.metric(var, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-```
-
-### Metrics in the `dispRity` function
-
-Using this metric structure, we can easily use any disparity metric in the `dispRity` function as follows:
-
-```{r}
-## Measuring disparity as the standard deviation of all the values of the
-## ordinated matrix (dimension-level 1 function).
-summary(dispRity(BeckLee_mat50, metric = sd))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the ordinated matrix (dimension-level 1 and 2 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the variance covariance matrix (dimension-level 1, 2 and 3 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
-```
-
-Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > 2 > 1).
-
-```{r}
-## Disparity as the standard deviation of the variance of each axis of the
-## variance covariance matrix:
-disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)),
-                      round = 10)
-
-## Same as above but using a different function order for the metric argument
-disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
-                      round = 10)
-
-## Both ways output the same disparity values:
-disparity1 == disparity2
-```
-
-In these examples, we considered disparity to be a single value.
-For example, in the previous example, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
-It is, however, possible to calculate [disparity as a distribution](#disparity-as-a-distribution).
-
-### Metrics implemented in `dispRity`
-
-Several disparity metrics are implemented in the `dispRity` package.
-The detailed list can be found in `?dispRity.metric` along with some description of each metric.
-
-Level | Name | Description | Source | 
-------|------|---------------------------------------------------|--------|
-3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
-3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
-3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
-2 | `ranges` | The range of each dimension | `dispRity` |
-2 | `variances` | The variance of each dimension | `dispRity` |
-2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
-1 | `mode.val` | The modal value | `dispRity` |
-
-1: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (not PCA).
-
-2: Note that by default, the centroid is the centroid of the elements.
-It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
-
-### Equations and implementations
-Some of the functions described below are implemented in the `dispRity` package and do not require any other packages to calculate ([see implementation here](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R)).
-
-\begin{equation}
-    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
-\end{equation}
-Where *k* is the number of dimensions, and $\lambda_i$ is the eigenvalue of each dimension.
-
-\begin{equation}
-    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
-\end{equation}
-Where *k* is the number of dimensions.
-
-\begin{equation}
-    ranges = |max(k_i) - min(k_i)|
-\end{equation}
-Where *k* is the number of dimensions.
-
-\begin{equation}
-    variances = \sigma^{2}{k_i}
-\end{equation}
-Where *k* is the number of dimensions, and $\sigma^{2}$ is their variance.
-
-\begin{equation}
-    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
-\end{equation}
-Where *n* is each element in the ordinated space, *k* is the number of dimensions, and $Centroid_{k}$ is their mean (or can be set to another value).
-
-### Using the different disparity metrics
-Here is a brief demonstration of the main metrics implemented in `dispRity`.
-First, we will create a dummy/simulated ordinated space using the `space.maker` utility function (more about that [here](#space.maker):
-
-```{r}
-## Creating a 10*5 normal space
-set.seed(1)
-dummy_space <- space.maker(10, 5, rnorm)
-```
-
-We will use this simulated space to demonstrate the different metrics.
-
-#### Volumes and surface metrics
-The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `hyper.volume` all measure the surface or the volume of the ordinated space occupied:
-
-```{r}
-## Calculating the ellipsoid volume
-summary(dispRity(dummy_space, metric = ellipse.volume))
-```
-> Because there is only one subsample (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with rounding).
-
-```{r}
-## Calculating the convex hull surface
-summary(dispRity(dummy_space, metric = convhull.surface))
-
-## Calculating the convex hull volume
-summary(dispRity(dummy_space, metric = convhull.volume))
-```
-
-The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
-For the full and correct calculation of the volume of the ordinated space, it is possible to use the `hyper.volume` function that has more options but takes longer to calculate.
-
-```{r}
-## Calculating the true multidimensional volume
-summary(dispRity(dummy_space, metric = hyper.volume))
-
-```
-> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) that often results in near 0 disparity values.
-
-#### Ranges, variances and diagonal
-
-The functions `ranges`, `variances` and `diagonal` all measure properties of the ordinated space based on its dimensional properties (they are also less affected by the "curse of dimensionality"):
-
-`ranges` and `variances` both work on the same principle and measure the range/variance of each dimension:
-
-```{r}
-## Calculating the ranges of each dimension in the ordinated space
-ranges(dummy_space)
-
-## Calculating disparity as the distribution of these ranges
-summary(dispRity(dummy_space, metric = ranges))
-
-## Calculating disparity as the sum and the product of these ranges
-summary(dispRity(dummy_space, metric = c(sum, ranges)))
-summary(dispRity(dummy_space, metric = c(prod, ranges)))
-
-## Calculating the variances of each dimension in the ordinated space
-variances(dummy_space)
-
-## Calculating disparity as the distribution of these variances
-summary(dispRity(dummy_space, metric = variances))
-
-## Calculating disparity as the sum and the product of these variances
-summary(dispRity(dummy_space, metric = c(sum, variances)))
-summary(dispRity(dummy_space, metric = c(prod, variances)))
-```
-
-The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest Euclidean distance in our five dimensional space):
-
-```{r}
-## Calculating the ordinated space's diagonal
-summary(dispRity(dummy_space, metric = diagonal))
-```
-
-> This metric is only a Euclidean diagonal (mathematically valid) if the dimensions within the space are all orthogonal!
-
-#### Centroids metric
-
-The `centroids` metric allows users to measure the position of the different elements compared to a fixed point in the ordinated space.
-By default, this function measures the distance between each element and their centroid (centre point):
-
-```{r}
-## The distribution of the distances between each element and their centroid
-summary(dispRity(dummy_space, metric = centroids))
-
-## Disparity as the median value of these distances
-summary(dispRity(dummy_space, metric = c(median, centroids)))
-```
-
-It is however possible to fix the coordinates of the centroid to a specific point in the ordinated space, as long as it has the correct number of dimensions:
-
-```{r}
-## The distance between each element and the origin of the ordinated space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))
-
-## Disparity as the distance between each element and a specific point in space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
-```
-
-## Summarising dispRity data (plots)
-
-Because of its architecture, printing `dispRity` objects only summarises their content but does not print the disparity value measured or associated analysis (more about this [here](#manipulating-dispRity-objects)).
-To actually see what is in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphical representation of the results.
-
-### Summarising `dispRity` data
-This function is an S3 function (`summary.dispRity`) allowing users to summarise the content of `dispRity` objects that contain disparity calculations.
-
-<!-- DATA LOADER -->
-
-```{r}
-## Example data from previous sections
-crown_stem <- custom.subsamples(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
-## Calculate disparity
-disparity_crown_stem <- dispRity(boot_crown_stem, metric = c(sum, variances))
-
-## Creating time slice subsamples
-time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-## Bootstrapping the time slice subsamples
-boot_time_slices <- boot.matrix(time_slices, bootstraps = 100)
-## Calculate disparity
-disparity_time_slices <- dispRity(boot_time_slices, metric = c(sum, variances))
-
-## Creating time bin subsamples
-time_bins <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages,
-    inc.nodes = TRUE)
-## Bootstrapping the time bin subsamples
-boot_time_bins <- boot.matrix(time_bins, bootstraps = 100)
-## Calculate disparity
-disparity_time_bins <- dispRity(boot_time_bins, metric = c(sum, variances))
-```
-
-These objects are easy to summarise as follows:
-
-```{r}
-## Default summary
-summary(disparity_time_slices)
-```
-
-Information about the number of elements in each subsample and the observed (i.e. non-bootstrapped) disparity are also calculated.
-This is specifically handy when rarefying the data for example:
-
-```{r}
-head(summary(disparity_crown_stem))
-```
-
-The summary functions can also take various options such as:
-
- * `quantile` values for the confidence interval levels (by default, the 50 and 95 quantiles are calculated)
- * `cent.tend` for the central tendency to use for summarising the results (default is `median`)
- * rounding` option corresponding to the number of decimal places to print (default is `2`)
- * `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
-
-These options can easily be changed from the defaults as follows:
-
-```{r}
-## Same as above but using the 88th quantile and the standard deviation as the summary 
-summary(disparity_time_slices, quantile = 88, cent.tend = sd)
-
-## Printing the details of the object and rounding the values to the 5th decimal place
-summary(disparity_time_slices, recall = TRUE, rounding = 5)
-```
-
-Note that the summary table is a `data.frame`, hence it is as easy to modify as any dataframe using `dplyr`.
-You can also export it in `csv` format using `write.csv` or `write_csv` or even directly export into `LaTeX` format using the following;
-
-```{r, eval=FALSE}
-## Loading the xtable package
-require(xtable)
-## Converting the table in LaTeX
-xtable(summary(disparity_time_slices))
-```
-
-### Plotting `dispRity` data
-An alternative (and more fun!) way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
-This function takes the same options as `summary.dispRity` along with various graphical options described in the function help files (see `?plot.dispRity`).
-
-The plots can be of four different types:
-
- * `continuous` for displaying continuous disparity curves
- * `box`, `lines`, and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.
-
-> This argument can be left empty. In this case, the algorithm will automatically detect the type of subsamples from the `dispRity` object and plot accordingly.
-
-It is also possible to display the number of elements in each subsample (as a horizontal dotted line) using the option `elements = TRUE`.
-Additionally, when the data is rarefied, one can indicate which level of rarefaction to display (i.e. only display the results for a certain number of elements) by using the `rarefaction` argument.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(mfrow = c(2, 2), bty = "n")
-
-## Plotting continuous disparity results
-plot(disparity_time_slices, type = "continuous")
-
-## Plotting discrete disparity results
-plot(disparity_crown_stem, type = "box")
-
-## As above but using lines for the rarefaction level of 20 elements only
-plot(disparity_crown_stem, type = "line", rarefaction = 20)
-
-## As above but using polygons while also displaying the number of elements
-plot(disparity_crown_stem, type = "polygon", elements = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Since `plot.dispRity` uses the arguments from the generic `plot` method, it is of course possible to change pretty much everything using the regular plot arguments:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some classic options from plot
-plot(disparity_time_slices, col = c("blue", "orange", "green"),
-    ylab = c("Some measurement"), xlab = "Some other measurement",
-    main = "Many options...", ylim = c(5, 0), xlim = c(4, 0))
-
-## Adding a legend
-legend("topleft", legend = c("Central tendency",
-                             "Confidence interval 1",
-                            "Confidence interval 2"),
-      col = c("blue", "orange", "green"), pch = 19)
-
-## Resetting graphical parameters
-par(op)
-```
-
-In addition to the classic `plot` arguments, the function can also take arguments that are specific to `plot.dispRity` like adding the number of elements or rarefaction level (as described above), and also changing the values of the quantiles to plot as well as the central tendency.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some plot.dispRity arguments
-plot(disparity_time_slices, quantile = c(seq(from = 10, to = 100, by = 10)),
-    cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
-    ylab = c("Disparity", "Diversity"), time.subsamples = FALSE,
-    xlab = "Time (in in units from past to present)", observed = TRUE,
-    main = "Many more options...")
-
-## Resetting graphical parameters
-par(op)
-```
-
-> Note that the argument `observed = TRUE` allows to plot the disparity values calculated from the non-bootstrapped data as crosses on the plot.
-
-For comparing results, it is also possible to add a plot to the existent plot by using `add = TRUE`:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the continuous disparity with a fixed y axis
-plot(disparity_time_slices, ylim = c(1, 4))
-## Adding the discrete data
-plot(disparity_time_bins, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
-    add = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Finally, if your data has been fully rarefied, it is also possible to easily look at rarefaction curves by using the `rarefaction = TRUE` argument:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the rarefaction curves
-plot(disparity_crown_stem, rarefaction = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-## Testing disparity hypotheses
-
-The `dispRity` package allows users to apply statistical tests to the calculated disparity to test various hypotheses.
-The function `test.dispRity` works in a similar way to the `dispRity` function: it takes a `dispRity` object, a `test` and a `comparisons` argument.
-
-The `comparisons` argument indicates the way the test should be applied to the data:
-
- * `pairwise` (default): to compare each subsample in a pairwise manner
- * `referential`: to compare each subsample to the first subsample
- * `sequential`: to compare each subsample to the following subsample
- * `all`: to compare all the subsamples together (like in analysis of variance)
-
-It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsample names to create personalised tests.
-Some other tests implemented in `dispRity` such as the `dispRity::null.test` have a specific way they are applied to the data and therefore ignore the `comparisons` argument. 
-<!-- Add sequential test one day! -->
-
-The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
-It can be a common statistical test function (e.g. `stats::t.test`), a function implemented in `dispRity` (e.g. see `?null.test`) or any function defined by the user.
-
-This function also allows users to correct for Type I error inflation (false positives) when using multiple comparisons via the `correction` argument. 
-This argument can be empty (no correction applied) or can contain one of the corrections from the `stats::p.adjust` function (see `?p.adjust`).
-
-Note that the `test.dispRity` algorithm deals with some classical test outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
-It is, however, possible to get the full detailed output by using the options `details = TRUE`.
-
-```{r}
-## T-test to test for a difference in disparity between crown and stem mammals
-test.dispRity(disparity_crown_stem, test = t.test)
-
-## Performing the same test but with the detailed t.test output
-test.dispRity(disparity_crown_stem, test = t.test, details = TRUE)
-
-## Wilcoxon test applied to time sliced disparity with sequential comparisons, with Bonferroni correction
-test.dispRity(disparity_time_slices, test = wilcox.test,
-              comparisons = "sequential", correction = "bonferroni")
-
-## Measuring the overlap between distributions in the time bins (using the
-## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)
-test.dispRity(disparity_time_bins, test = bhatt.coeff)
-```
-
-It is also possible to apply some more *complex* tests that have their own output classes (like `stats::lm`).
-
-The results can then be analysed as usual using the associated `summary` S3 method:
-
-```{r}
-## Performing and linear model applied to the same data
-(slice_lm <- test.dispRity(disparity_time_slices, test = lm,
-                            comparisons = "all"))
-
-## The output is a regular `lm` output
-class(slice_lm)
-
-## This output can be summarised using summary
-summary(slice_lm)
-```
-
-Of course, due to the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
-The only condition is that the test can be applied to at least two distributions.
-In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm` for example).
-
-## Disparity as a distribution
-
-Disparity is often regarded as a summary value of the position of the all elements in the ordinated space.
-For example, the sum of variances, the product of ranges or the median distance between the elements and their centroid will summarise disparity as a single value.
-This value can be pseudo-replicated (bootstrapped) to obtain a distribution of the summary metric with estimated error.
-However, another way to perform disparity analysis is to use the *whole distribution* rather than just a summary metric (e.g. the variances or the ranges).
-
-This is possible in the `dispRity` package by calculating disparity as a dimension-level 2 metric only!
-Let's have a look using our previous example of bootstrapped time slices but by measuring the distances between each taxon and their centroid as disparity.
-
-```{r}
-## Measuring disparity as a whole distribution
-disparity_centroids <- dispRity(boot_time_slices, metric = centroids)
-```
-
-The resulting disparity object is of dimension-level 2, so it can easily be transformed into a dimension-level 1 object by, for example, measuring the median distance of all these distributions:
-
-```{r}
-## Measuring median disparity in each time slice
-disparity_centroids_median <- dispRity(disparity_centroids, metric = median)
-```
-
-And we can now compare the differences between these methods:
-
-```{r}
-## Summarising both disparity measurements:
-## The distributions:
-summary(disparity_centroids)
-## The summary of the distributions (as median)
-summary(disparity_centroids_median)
-```
-
-We can see that the summary message for the distribution is slightly different than before.
-Here `summary` also displays the observed central tendency (i.e. the central tendency of the measured distributions).
-Note that, as expected, this central tendency is the same in both metrics!
-
-Another, maybe more intuitive way, to compare both approaches for measuring disparity is to plot the distributions:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(bty = "n", mfrow = c(1, 2))
-
-## Plotting both disparity measurements
-plot(disparity_centroids, ylab = "Distribution of all the distances")
-plot(disparity_centroids_median,
-     ylab = "Distribution of the medians of all the distances")
-
-par(op)
-```
-
-We can then test for differences in the resulting distributions using `test.dispRity` and the `bhatt.coeff` test as described above.
-
-```{r}
-## Probability of overlap in the distribution of medians
-test.dispRity(disparity_centroids_median, test = bhatt.coeff)
-```
-
-In this case, we are looking at the probability of overlap of the distribution of median distances from centroids among each pair of time slices.
-In other words, we are measuring whether the medians from each bootstrap pseudo-replicate for each time slice overlap.
-But of course, we might be interested in the actual distribution of the distances from the centroid rather than simply their central tendencies.
-This can be problematic depending on the research question asked since we are effectively comparing non-independent medians distributions (because of the pseudo-replication).
-
-One solution, therefore, is to look at the full distribution:
-
-```{r}
-## Probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff)
-```
-
-These results show the actual overlap among all the measured distances from centroids concatenated across all the bootstraps.
-For example, when comparing the slices 120 and 80, we are effectively comparing the 5 \times 100 distances (the distances of the five elements in slice 120 bootstrapped 100 times) to the 19 \times 100 distances from slice 80.
-However, this can also be problematic for some specific tests since the _n_ \times 100 distances are also pseudo-replicates and thus are still not independent.
-
-A second solution is to compare the distributions to each other _for each replicate_:
-
-```{r}
-## Boostrapped probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff, concatenate = FALSE)
-```
-
-These results show the median overlap among pairs of distributions in the first column (`bhatt.coeff`) and then the distribution of these overlaps among each pair of bootstraps.
-In other words, when two distributions are compared, they are now compared for each bootstrap pseudo-replicate, thus effectively creating a distribution of probabilities of overlap.
-For example, when comparing the slices 120 and 80, we have a mean probability of overlap of 0.28 and a probability between 0.18 and 0.43 in 50\% of the pseudo-replicates.
-Note that the quantiles and central tendencies can be modified via the `conc.quantiles` option.
-
-## Ecological studies of disparity
-<!-- NC: I wonder if this might be better just as a "hey we can do ecology stuff too and here is a link to a demo" rather than a new section?
-The manual is loooooooooong so splitting some stuff out might be good?
--->
-
-Disparity analyses are not restricted to palaeobiology analysis.
-This package is designed for dealing with ordinated matrices, no matter which data was used in the ordination.
-Here is an short example of disparity analysis using ecological data from Deirdre McClean (unpubl.).
-
-The dataset contains an ordinated matrix of 20 dimensions (columns) for 40 elements (rows).
-The elements are different experimental plots with discrete variations of nutrient enrichment, depth, and freshwater benthic invertebrates.
-<!-- NC: Explain a little more about what the data are here as I don't entirely get it. EG what's a treatement etc? Just a really simple explanataion is perfect -->
-
-```{r}
-## Loading the data
-data(McClean_data)
-
-## Ordinated matrix
-ordinated_matrix <- McClean_data$ordination
-
-## Treatments and depths
-treatments <- McClean_data$treatment
-depth <- McClean_data$depth
-```
-
-### Classic ecological "disparity" analysis
-
-A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different experimental plots differ in their values for nutrient enrichment, depth, and freshwater benthic invertebrates.
-<!-- NC: or similar just to make it clearer -->
-
-```{r, fig.width=6, fig.height=6}
-## Setting the colors that will represent the treatments
-cols <- sub("a", "red", treatments)
-cols <- sub("b", "blue", cols)
-## Setting the symbols that will represent the depth
-pchs <- sub(1, 16, depth)
-pchs <- as.numeric(sub(2, 17, pchs))
-
-## Graphical option
-op <- par(bty = "n")
-## A classic 2D ordination plot
-plot(ordinated_matrix[, 1:2], col = cols, pch = pchs,
-     xlab = "PC 1", ylab = "PC 2",
-     xlim = range(ordinated_matrix[, 1:2]) + c(0, 100))
-par(op)
-```
-
-This shows the distribution of the experimental plots along the two first axis of variation of the ordinated distance matrix (i.e. the first two dimensions).
-At a first glance, it seems difficult to see a clear effect of the treatments (blue or red) or the depth (rounds or triangles) as the different experimental plots with the same parameters don't seem to cluster together.
-However, this may be because this plot ignores the 18 other dimensions of the ordination!
-Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e.
-the values on the first dimension do not represent a continuous trait (e.g.
-depth), instead they just represent the ordinations of correlations between the data and some factors.
-
-Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.
-
-### Multidimensional analysis with `dispRity`
-
-#### Splitting the data with respect to the different factors
-
-The first step is to create different subsamples that represent subsamples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
-Each of these subsamples will contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
-In our example, we are going to group the elements according to their depth and treatment.
-
-```{r}
-## Creating a table that contain the elements and their attributes
-factors <- as.data.frame(matrix(data = c(treatments, depth),
-          nrow = nrow(ordinated_matrix), ncol = 2, byrow = FALSE,
-          dimnames = list(rownames(ordinated_matrix))))
-names(factors) <- c("Treat", "Depth")
-head(factors)
-```
-
-Second, let’s split the data according by depth and treatment to create the subsamples of the ordinated space by using the `custom.subsamples` function:
-
-```{r}
-## Splitting the ordinated space into four subsamples
-(customised_subsamples <- custom.subsamples(ordinated_matrix, factors))
-```
-
-#### Calculating disparity
-
-In this example, we are going to define disparity as the n-dimensional ellipsoid volume (see `?hyper.volume`).
-
-```{r}
-## Calculating disparity as the ellipsoid volume of each group
-disparity <- dispRity(customised_subsamples, metric = diagonal)
-summary(disparity)
-```
-
-We can also bootstrap the data to test the robustness of the measured disparity to outliers.
-Also, as we can see, each subsample has different numbers of elements.
-It might also be interesting to rarefy the data to have only subsamples with the same number of elements.
-
-```{r}
-## Bootstrapping the subsamples and using rarefaction
-## (i.e. only re-sampling 17 elements each time)
-bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 100,
-                                 rarefaction = 17)
-```
-
-We can now run a more robust disparity analysis using the bootstrapped data:
-
-```{r}
-## Calculating the bootstrapped disparity
-disparity <- dispRity(bootstrapped_data, metric = diagonal)
-```
-
-#### Summarising and displaying the data
-
-Now let's summarise this data:
-
-```{r, fig.width=12, fig.height=6, eval=TRUE}
-## Summary table of the data
-summary(disparity)
-
-## Graphical options
-op <- par(mfrow = (c(1, 2)), bty = "n")
-
-## Plotting the bootstrapped disparity
-plot(disparity, main = "Bootstrapped", las = 2, xlab = "")
-
-## Plotting the rarefied disparity
-plot(disparity, rarefaction = 17, main = "Rarefied", las = 2, xlab = "")
-
-par(op)
-```
-
-As we can see, there seems to be no strong effect of the number of experimental plots in each subsample (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
-
-#### Testing our hypotheses
-
-Finally, we can test our hypothesis (whether the water treatment at certain depths alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.
-<!-- Ah so this is what you are testing! Make this clearer in the intro bit to this section -->
-
-```{r}
-## Testing the effect of treatment and depth on the bootstrapped data
-summary(test.dispRity(disparity, test = aov, comparisons = "all"))
-```
-
-## Simulating discrete morphological data
-
-The `dispRity` package also allows users to simulate discrete morphological data matrices.
-
-### Quick go through
-The function basically intakes a phylogenetic tree, the number of required characters and the evolutionary model and a function from which to draw the rates.
-The package then proposes a function for quickly checking the matrix' phylogenetic signal using parsimony.
-
-```{r}
-set.seed(3)
-## Setting the starting tree with 15 taxa as a random coalescent tree
-my_tree <- rcoal(15)
-
-## Generating a matrix with 100 characters (85% binary and 15% three states) and
-## an equal rates model with a gamma rates distribution (0.5, 1) with no 
-## invariant characters.
-my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
-    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)
-
-## The head of the matrix
-my_matrix[1:5, 1:10]
-
-## Checking the matrix properties with a quick Maximum Parsimony tree search
-check.morpho(my_matrix, my_tree)
-```
-
-Note that this example produces a tree with a great consistency index and an identical topology as the random coalescent tree!
-Nearly too good to be true...
-
-### In details
-
-Basically, `sim.morpho` is really flexible and intakes a lot of different arguments to allow to simulate realistic matrix.
-It has three implemented models: `"ER"` for Equal Rates (the M*k* model); the `"HKY"` one which is the molecular HKY model but transforms pyrines in 0's and purimidines in 1's; or the `"mixed"` model that randomly uses an `"ER"` or and `"HKY"` for the binary characters and `"ER"` for the multistates (>2) characters.
-Both models intakes specific distributions for their rates or substitution models.
-These distributions should be passed to these arguments in the format of `c(sampler_function, distribution_parameters)` where the the sampler function is a the random generation function of that distribution (e.g.
-`rnorm`, `runif`, etc...) and the parameters are any parameters to be passed to this function.
-
-The `check.morpho` runs a quick maximum parsimony tree using the `phangorn` parsimony algorithm.
-It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g.
-the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree.
-
-### Parameters for a good(ish) matrix
-There are many parameters that can realise a ``realistic'' matrix (meaning not to different from the input tree with a consistency and retention index close to what's in the literature) but because of the randomness of the generation (the code generates random matrices after all!) not all end up in creating good matrices.
-The following parameters however, seems to generate a fairly ``realistic'' with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three states characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
-
-```{r}
-set.seed(0)
-## tree
-my_tree <- rcoal(15)
-## matrix
-morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
-    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
-check.morpho(morpho_mat, my_tree)
-```
-
-### `space.maker`
-Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
-This function allows to simulated multidimensional spaces given a certain number of properties.
-It takes as arguments the number of elements (data points) and dimensions to create the space but also allows more fine tuning data simulation: it is possible to give a specific distribution to each dimensions, provide a correlation matrix to link the dimensions or even providing an *a priori* distribution of the variance per distributions!
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(mfrow = (c(2,2)), bty = "n")
-## Visualising 3D spaces
-require(scatterplot3d)
-
-## A cube space
-scatterplot3d(space.maker(2500, 3, runif), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Uniform cube space")
-
-## A plane space
-scatterplot3d(space.maker(2500, 3, c(runif, runif, runif),
-              arguments = list(list(min = 0, max = 0), NULL, NULL)), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Uniform plane space")
-
-## An ellipsoid space (=a spheric space with correlation)
-cor_matrix <- matrix(cbind(1,0.8,0.2, 0.8,1,0.7, 0.2,0.7,1), nrow = 3)
-scatterplot3d(space.maker(2500, 3, rnorm, cor.matrix = cor_matrix), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Normal ellipsoid space")
-
-## A cylindrical space with a decreasing variance per axis
-scatterplot3d(space.maker(2500, 3, c(rnorm, rnorm, runif),
-              scree = c(0.7, 0.2, 0.1)), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Normal cylindrical space")
-
-## Resetting the graphic parameters
-par(op)
-```
-
-# Glossary <a name="glossary"></a>
-
--   **Ordinated space**. The mathematical multidimensional object that will be analysed with this package.
-    In morphometrics, this is often referred to as the morphospace.
-    However it may also be referred to as the cladisto-space for cladistic data or the eco-space for ecological data etc.
-    In practice, this term designates an ordinated matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
-
--   **Elements**. The rows of the ordinated space. Elements can be taxa, field sites, countries etc.
-
--   **Dimensions**. The columns of the ordinated space. The dimensions are referred to as axes of variation, or principal components, for ordinated spaces obtained from a PCA.
-
--   **Subsamples**. Sub-samples of the ordinated space.
-    A sub-sample (or subsamples) contains the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
-    For example, if our ordinated space is composed of birds and mammals (the elements) and 50 principal components (the dimensions), we can create two subsamples containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
-
-# The package's guts
-
-## Manipulating `dispRity` objects
-
-Disparity analysis involves a lot of manipulation of many matrices (especially when bootstrapping) which can be impractical to visualise and will quickly overwhelm your `R` console.
-Even the simple Beck and Lee 2014 example above produces an object with > 72 lines of lists of lists of matrices!
-
-Therefore `dispRity` uses a specific class of object called a `dispRity` object.
-These objects allow users to use S3 method functions such as `summary.dispRity`, `plot.dispRity` and `print.dispRity`. 
-`dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g. `sort.dispRity`, `extract.dispRity` see the full list in the next section).
-These functions modify the `dispRity` object without having to delve into its complex structure!
-The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).
-
-```{r}
-## Loading the example data
-data(disparity)
-
-## What is the class of the median_centroids object?
-class(disparity)
-
-## What does the object contain?
-names(disparity)
-
-## Summarising it using the S3 method print.dispRity
-disparity
-```
-
-Note that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:
-
-```{r, eval=FALSE}
-## Display the full object
-print(disparity, all = TRUE)
-## This is more nearly ~ 5000 lines on my 13 inch laptop screen!
-```
-
-## `dispRity` utilities
-
-The package also provides some utilities functions to facilitate multidimensional analysis.
-
-### `dispRity` object utilities  <a name="dispRity.utilities"></a>
-The first subsamples of utilities are functions for manipulating `dispRity` objects:
-
-#### `make.dispRity`
-This function creates empty `dispRity` objects.
-
-```{r}
-## Creating an empty dispRity object
-make.dispRity()
-
-## Creating an "empty" dispRity object with a matrix
-(disparity_obj <- make.dispRity(matrix(rnorm(20), 5, 4)))
-```
-
-####  `fill.dispRity`
-This function initialises a `dispRity` object and generating it's call properties.
-
-```{r}
-## The dispRity object's call is indeed empty
-disparity_obj$call
-
-## Filling an empty disparity object (that needs to contain at least a matrix)
-(disparity_obj <- fill.dispRity(disparity_obj))
-
-## The dipRity object has now the proper minimal attributes
-disparity_obj$call
-```
-
-####  `matrix.dispRity`
-This function extracts a specific matrix from a disparity object. The matrix can be one of the bootstrapped or/and rarefied one.
-
-```{r}
-## Extracting the matrix containing the coordinates of the elements at time 50
-str(matrix.dispRity(disparity, "50"))
-
-## Extracting the 3rd boostrapped matrix with the 2nd rarefaction level
-## (15 elements) from the second group (80 Mya)
-str(matrix.dispRity(disparity, subsamples = 1, bootstrap = 3, rarefaction = 2))
-```
-
-#### `get.subsamples.dispRity`
-This function creates a dispRity object that contains only elements from one specific subsamples.
-
-```{r, eval=FALSE}
-## Extracting all the data for the crown mammals
-(crown_mammals <- get.subsamples.dispRity(disp_crown_stemBS, "Group.crown"))
-
-## The object keeps the properties of the parent object but is composed of only one subsamples
-length(crown_mammals$subsamples)
-```
-
-#### `extract.dispRity`
-This function extracts the calculated disparity values of a specific matrix.
-
-```{r, eval=FALSE}
-## Extracting the observed disparity (default)
-extract.dispRity(disparity)
-
-## Extracting the disparity from the bootstrapped values from the
-## 10th rarefaction level from the second subsamples (80 Mya)
-extract.dispRity(disparity, observed = FALSE, subsamples = 2, rarefaction = 10)
-```
-
-#### `scale.dispRity`
-This is the S3 method of `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object.
-
-```{r, eval=FALSE}
-## Getting the disparity values of the time subsamples
-head(summary(disparity))
-
-## Scaling the same disparity values
-head(summary(scale(disparity, scale = TRUE)))
-
-## Scaling and centering:
-head(summary(scale(disparity, scale = TRUE, center = TRUE)))
-```
-
-#### `sort.dispRity`
-This is the S3 method of `sort` for sorting the subsamples alphabetically (default) or following a specific pattern.
-
-```{r, eval=FALSE}
-## Sorting the disparity subsamples in inverse alphabetic order
-head(summary(sort(disparity, decreasing = TRUE)))
-
-## Customised sorting
-head(summary(sort(disparity, sort = c(7, 1, 3, 4, 5, 2, 6))))
-```
-
-<!-- ## Modularity
-
-### Making your own metrics
-
-### Making your own tests
-
-## Where is this going?
- -->
-<!-- ## Running `dispRity` in parallel
-
-The computationally intensive function `dispRity` have a `parallel` option to speed up their calculations.
-
-This option requires the package `snow` and takes arguments that are to be passed to  -->

### 2017-08-12 paper/dispRity_paper.tex
updated draft

@@
 
+\newcommand{\disp}{\texttt{dispRity} }
+
 \begin{document}
@@
 
-\noindent{\Large \bf \texttt{dispRity}: a modular \texttt{R} package for measuring disparity.} 
+\noindent{\Large \bf \disp: a modular \texttt{R} package for measuring disparity.} 
 \bigskip
 
-\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$, Mark Puttick$^2$, Natalie Cooper$^3$}\\
+\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$}\\
 \noindent {\small \it 
@@
 \begin{enumerate}
-    \item We present \texttt{dispRity}, an \texttt{R} package in multidimensional spaces. 
+    \item We present \disp, an \texttt{R} package in multidimensional spaces. 
     \item Disparity designates a suit of metrics to describe an ordinated matrix (also referred as morpho/eco/niche/etc.-space).
@@
     \item This modular architecture allows great plasticity in using this package, namely by letting users define their own metric of disparity.
-    \item The package also provides numerous tools for modifying, summarising and plotting \texttt{dispRity} objects.
+    \item The package also provides numerous tools for modifying, summarising and plotting \disp objects.
 \end{enumerate}
@@
 
-Multidimensionality is a major aspect of natural sciences were variables co-vary through time and space from the organism to the environmental level.
+% Multidimensionality is a major aspect of natural sciences were variables co-vary through time and space from the organism to the environmental level.
 
-In palaeobiology multidimensionality is a popular way to study morphological diversity and comes as a interesting alternative to unidimensional taxonomic diversity analysis \citep{Hopkins2017}.
+% In palaeobiology multidimensionality is a popular way to study morphological diversity and comes as a interesting alternative to unidimensional taxonomic diversity analysis \citep{Hopkins2017}.
 
-Although the concept of morphological disparity is well defined since more than two decades \citep[e.g.][]{gould1991disparity,Foote01071994,Foote29111996}, its specific definition varies widely among authors.
-In fact disparity Disparity is by nature multidimensional \citep{lloyd2016estimating} and thus there are many ways to measure it.
+% Although the concept of morphological disparity is well defined since more than two decades \citep[e.g.][]{gould1991disparity,Foote01071994,Foote29111996}, its specific definition varies widely among authors.
+% In fact disparity Disparity is by nature multidimensional \citep{lloyd2016estimating} and thus there are many ways to measure it.
+
+%TG: rewrite all that using bits and blobs from previous intro
+% Biological data are complex; understanding the ecology and evolution of species often requires that we analyse multiple variables that covary with each other, and through space and time.
+% Multivariate analyses aim to capture and incorporate this multidimensional complexity, while providing outputs that are interpretable without needing to visualise n-dimensions simultaneously.
+% Key multidimensional features of species that have important roles in ecology and evolution include morphology, functional traits. etc. etc.
+% Many of these can be represented as matrices, these can be used in our package.
+% In the interests of clarity/brievety, we here focus on just one kind of data: morphological diversity.
+
+% In evolutionary biology and palaeontology, morphological diversity is often referred to as disparity.
+% DESCRIBE COOL USES OF DISPARITY ANALYSES. WHY IS DISPARITY AWESOME.
+% Although disparity is commonly studied, its definition varies widely across authors due to the myriad ways it can be measured.
+% LIST WAYS IT IS MEASURED.
+% This is further complicated by confusion over whether diparity refers to the metric describing or summarising one or more aspects of this space, or to the multidimensional space which is a specific mathematical object.
+% This multidimensional space can also be defined in multiple ways.
+% LIST ways
+
+% In theory, this multitude of ways to measure/define disparity is not an issue. [the problem is that people aren't clear about it]
+% The ideal solution is to choose either the most appropriate method for your question or data, or to apply several definitions and compare across these.[check biogeoBEARS]
+% Unfortuantely, in practice this is hampered by existing software implementations.
+% Package maintainers/software developers choose their preferred definition of disparity and method for measuring it, and then enforce this on users by allowing zero flexibility.
+% This can lead to a chain of inappropriate analyses led by everyone just using the package that exists and the method within it, and in the worst case, using the defaults of the package. Oh no!
+% The dispRity package will solve this problem by providing a completely flexible framework for studing disparity.
+% It implements all commonly used metrics and definitions of disparity, as well as providing a simple interface for users to implement their own disparity functions.
+% The package is described here for use with morphological diversity data, but the functions are equally applicable to ecological contexts such as Diaz paper, Deirdre's paper etc.
+
+Disparity is really cool:
 
@@
 It has a highly modular architecture and allow users to define their own multidimensional space, use their own definition of the disparity metric and measure the differences between metrics in their own terms.
-Furthermore, both a specific design and set of functions (through the \texttt{dispRity} object) allows to design analysis into easy and repeatable pipelines.
+Furthermore, both a specific design and set of functions (through the \disp object) allows to design analysis into easy and repeatable pipelines.
 
 \section{Description}
-
-\subsubsection{The package's data}
-
-Beck and Lee data
-Beck and Lee tree
-Beck and Lee ages
-
-disparity  disparity data
-
-\subsubsection{Running example}
-
-\subsection{The \texttt{dispRity} package}
-
-\subsubsection{The package's main functions}
+All the following functionalities of the package described below are based on the \disp class object.
+This object class allows to repeat and streamline disparity analysis by being past from one function to the other (see Example section).
+Specificities of the \disp object can be found in @@@ as well as series of utilities functions allowing to modify such object (see @@@).
 
@@
         \texttt{time.subsamples} & Separates ordinated data in time subsamples. \\
-        \texttt{boot.matrix} & Bootstraps and rarefies ordinated data or a \texttt{dispRity} object. \\
-        \texttt{dispRity} & Calculates disparity from an ordinated matrix or a \texttt{dispRity} object. \\
-        \texttt{summary.dispRity} & Summarises a \texttt{dispRity} object. \\
-        \texttt{plot.dispRity} & Plots a \texttt{dispRity} object. \\
-        \texttt{test.dispRity} & Applies statistical tests to a \texttt{dispRity} object.\\
+        \texttt{boot.matrix} & Bootstraps and rarefies ordinated data or a \disp object. \\
+        \disp & Calculates disparity from an ordinated matrix or a \disp object. \\
+        \texttt{summary.dispRity} & Summarises a \disp object. \\
+        \texttt{plot.dispRity} & Plots a \disp object. \\
+        \texttt{test.dispRity} & Applies statistical tests to a \disp object.\\
         \texttt{dispRity.per.group} & Pipelined disparity per groups analysis. \\
         \texttt{dispRity.through.time} & Pipelined disparity through time analysis. \\
-        \texttt{dispRity.metrics} & A set of implemented disparity metrics functions.\\
-        \texttt{dispRity.utilities} & A set of functions for modifying a \texttt{dispRity} object.\\
         \hline
     \end{tabular}
-    \caption{Main functions in the \texttt{dispRity} package.}
+    \caption{Main functions in the \disp package.}
 \end{table}
 
-\subsection{Fork flow}
+\subsection{Measuring disparity}
+As hinted in the package name, the main functionality of this package is to measure disparity in a multidimensional space.
+Again, within this package, disparity is defined as a description or a summary of one or more aspects of the multidimensional space.
 
-Workflow figure
+The \disp function allows to measure disparity from any matrix (of class \texttt{matrix} or \texttt{data.frame}) where the columns correspond to the different dimensions of the multidimensional space (e.g.
+the ordination axis or any variable that composes the space) and the rows correspond to the elements present in the multidimensional space (e.g.
+species, study sites, etc...).
 
-\subsubsection{Input data}
+Disparity can then be defined by the user as any function or combination of functions that can either transform the matrix into:
+\begin{itemize}
+    \item Another matrix (a Dimensions-level 3 function)
+    \item A vector (a Dimensions-level 2 function)
+    \item A single value (a Dimensions-level 1 function)
+\end{itemize}
 
-\subsubsection{Manipulating the data}
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{../doc/dispRity_fun.pdf} 
+\caption{Illustration of the different metric dimensions levels in the \disp package}
+\label{Fig:levels}
+\end{figure}
 
-\subsubsection{Measuring disparity}
+The function(s) can then be passed to the \texttt{metric} argument to measure a specific aspect of the multidimensional space:
+There is no limitation on the metric dimensions only that at least one dimensions level 2 or 1 are present in the definition (i.e.
+disparity can only be defined as distribution - a vector - or a single value).
 
-\subsubsection{Summarising the data}
-
-Link to graph one two three
-
-\subsubsection{Testing hypothesis}
-
-\subsection{The \texttt{dispRity} object}
-
-Disparity object structure figure
-
-\subsection{Additional functions and data}
-
-The package also contains additional functions not directly linked to the disparity workflow nor to the \texttt{dispRity} object but that can be used as utilities for the studying discrete morphological data.
+Some metrics are implemented in the package
 
 \begin{table}
-    \begin{tabular}{ll}
+\center
+    \begin{tabular}{r|c|c|c|c}
+        name & description & dimension level & definition & source \\
         \hline
-        Function & Description \\ 
-        \hline
-        \texttt{char.diff} & Calculates character differences. \\
-        \texttt{clean.data} & Cleans phylogenetic data (tree and table). \\
-        \texttt{sim.morpho} & Simulates morphological discrete data. \\
-        \texttt{slice.tree} & Slice through a phylogenetic tree. \\
-        \texttt{space.maker} & Creates multidimensional spaces. \\
-        \texttt{tree.age} & Calculates the age of nodes and tips in a tree. \\
-        \hline
+        \texttt{ellipse.volume} & The volume of the ellipsoid of the space & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
+        \texttt{convhull.surface} & 1 & 1 & 1 & 1\\
+        \texttt{convhull.volume} & 1 & 1 & 1 & 1\\
+        \texttt{hypervolume} & 0 & 0 & 0 & 1\\
+        \texttt{diagonal} & 0 & 0 & 0 & 1\\
+        \texttt{ranges} & 0 & 0 & 0 & 1\\
+        \texttt{variances} & 0 & 0 & 0 & 1\\
+        \texttt{centroids} & 0 & 0 & 0 & 1\\
+        \texttt{mode.val} & 0 & 0 & 0 & 1\\
     \end{tabular}
-    \caption{Additional functions in the \texttt{dispRity} package.}
+    \caption{Where $k$ is the number of dimensions, $n$ the number of elements, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean.}
+    \label{Tab:metrics}
 \end{table}
 
-Additionally, this package also contains an ecology type data set from McClean
+% Level | Name | Description | Source | 
+% ------|------|---------------------------------------------------|--------|
+% 3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
+% 3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
+% 3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
+% 3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
+% 3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
+% 2 | `ranges` | The range of each dimension | `dispRity` |
+% 2 | `variances` | The variance of each dimension | `dispRity` |
+% 2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
+% 1 | `mode.val` | The modal value | `dispRity` |
 
-McClean data
+% 1: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (not PCA).
 
-\section{Conclusion}
+% 2: Note that by default, the centroid is the centroid of the elements.
+% It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
+%     ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
+%     diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
+%     ranges = |max(k_i) - min(k_i)|
+%     variances = \sigma^{2}{k_i}
+%     centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
 
-This package is good.
+\subsection{Splitting the multidimensional space into subsets}
+A common analysis using multidimensional space consists into subdividing it into smaller samples, typically to compare to each other.
+For example, one can be interested in looking how a specific sub-samples of the space compares in disparity from another (e.g.
+difference among groups) or, how different sub-samples change sequentially (e.g.
+through time).
+In essence, the original multidimensional space (the input matrix) corresponds to overall space.
+For example, if the multidimensional space is a morphospace, it corresponds to all the possible morphologies observed in the matrix.
+Conversely, sub-samples of this space correspond to parts of the space with pooled characteristics.
+For example, still in a morphospace with time sub-samples, each sub-samples correspond to all the observed morphologies during some time period.
 
-\section{Data availability and reproducibility}
-Data are available on Dryad or Figshare.
-Code for reproducing the analyses is available on GitHub (\url{github.com/TGuillerme/SpatioTemporal_Disparity}).
+These divisions can readily be created in the package using the \texttt{custom.subsamples} or \texttt{time.subsamples} functions.
+The first takes as argument the matrix defining the multidimensional space and then a list of elements to group into different sub-samples.
+The second also takes a multidimensional space and an additional argument on the age of the elements and then sorts them per time categories based on the following methods:
+
+\begin{itemize}
+    \item Discrete time subsamples: where the elements are grouped by age in time bins with corresponding ages.
+    \item Continuous time subsamples: where the elements are defined as being the elements present at specific points in time (see @@@).
+\end{itemize}
+
+When using the \texttt{time.subsamples} function, it is also possible to provide a list of first and last occurence data for taxa spanning across specific time periods.
+This allows species with a longer timespan to be present in multiple time sub-samples.
+
+The continuous time sub-samples method works by using a phylogeny and looking at which taxa are present at any specific point in time.
+This method thus requires a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
+
+The slicing works by selecting any tips (and nodes if \texttt{inc.nodes = TRUE}) at the specific time slice.
+When a slice occurs not on a tip or a node, four methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice.
+The models are:
+\begin{itemize}
+    \item ``acctran'' (accelerated transformation) where the data chosen along the branch is always the one of the descendant
+    \item ``deltran'' (delayed transformation) where the data chosen along the branch is always the one of the ancestor
+    \item ``punctuated'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor
+    \item ``gradual'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length
+\end{itemize}
+
+\subsection{Bootstrapping and rarefying}
+When measuring disparity, it is important to take sampling into account: for example, observed disparity in the whole multidimensional space can simply due to factors like the specific selection of elements (both their number or the presence of outliers).
+To take these sources of error into account, it is common practice to bootstrap the data (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different sub-samples).
+
+\subsubsection{Bootstraps}
+In disparity analysis, if disparity is defined as a dimension-level 1 metric (i.e.
+a metric summarising the multidimensional space into a single value), it can be useful to bootstrap this value in order to have a distribution on which to perform statistical analysis.
+This can be achieved by using the \texttt{boot.matrix} function pseudo-replicates the multidimensional space following these two algorithms:
+
+\begin{itemize}
+    \item ``full'' algorithm: where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data)
+    \item ``single'' algorithm: where only one random element is replaced by one other random element for each pseudo-replicate
+\end{itemize}
+
+\noindent Note that the ``single'' algorithm is somewhat akin to jackknife re-sampling where one element is duplicated rather than removed.
+
+In the case of disparity being a dimension-level 1 metric, it is important to note that the distribution resulting from a bootstrap will be a pseudo-replicated one and thus might not be suitable for some specific parametric statistical tests.
+Conversely, it is possible to use a dimension-level 2 metric on a bootstrapped multidimensional space to avoid this problem: the disparity metric is then measured on a pseudo-replicated multidimensional space rather than the metric itself being pseudo-replicated.
+
+\subsubsection{Rarefactions}
+The same \texttt{boot.matrix} function also allows to rarefy the multidimensional space or its sub-samples.
+Rarefaction can be useful if, for example, one is interested in looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of sub-samples wit the same number of elements in each.
+In practice, rarefaction allows users to set a limit to the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).
+
+Additionally, this argument, is set to \texttt{TRUE} also allows to fully rarefy the sub-samples (i.e.
+removing $x$ elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsample to a minimum of three elements).
+
+\subsection{Interpreting results}
+Once disparity has been calculated (without or without sub-sampling, bootstrapping and rarefaction), the resulted \disp object can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity}.
+Finally, the results can be analysed using the \texttt{test.dispRity} function.
+
+\subsubsection{Summary and plot}
+The summary and plot functions allow to summarise the contents of the \disp objects.
+These functions take as common arguments which central tendency (e.g.
+if central tendency is the mean and disparity was calculated as the sum of the variances for each dimensions, the central disparity tendency would be the mean of the sum of the variances) and the quantiles to represent (in the same example, if the quantile chosen is the 50th, these functions will also display 50\% of the sum of variances around the mean).
+
+Additionally, the \texttt{plot.dispRity} function allows to graphically represent the summarised results using different representations:
+
+\begin{itemize}
+    \item ``continuous'' for displaying continuous disparity curves (in the case of disparity through time analysis for example)
+    \item ``box'', ``lines'', or ``polygons'' to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.
+\end{itemize}
+
+Additional arguments can also be used such as \texttt{observed} to display the observed disparity (i.e.
+non-bootstrapped), \texttt{add} to add the plot in the current quartz window, or \texttt{rarefaction} to only plot the disparity for a specific number of elements and of course, any other normal graphical options can also be passed to this function (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...).
+Note that if the disparity has been calculated on fully rarefied data (using \texttt{boot.matrix(data, rarefaction = TRUE)} prior to calculating disparity), the rarefaction curves for the whole multidimensional space (or its sub-samples) can be plotted using \texttt{rarefaction = ``plot''}.
+
+\subsubsection{test}
+The \texttt{test.dispRity} function allows to test hypothesis on the disparity data (e.g. by comparing the different sub-samples between each other).
+Similarly to the \disp function described above, this one can take any test defined by the user.
+The \texttt{comparison} arguments allows to indicate in which order (if any) should the tests be applied:
+
+\begin{itemize}
+    \item ``pairwise'' for applying pairwise comparisons of all the subsamples (default).
+    \item ``referential'' for comparing the first subsample to all the others.
+    \item ``sequential'' for comparing each subsample sequentially (e.g. first against second, second against third, etc.).
+    \item ``all'' for comparing all the subsamples simultaneously to the data (i.e.
+\texttt{disparity $\mathtt{\sim}$ subsamples}).
+This argument can be used for \texttt{lm} or \texttt{glm} type tests.
+    \item or any list of pairs (or more) of number of sub-samples to compare (for example, for comparing the first to the sixth subsample and the fifth to the second: \texttt{comparison = list(c(1, 6), c(5, 2))})
+\end{itemize}
+
+Some tests are implemented within the package such as the Bhattacharrya Coefficient \citep[\texttt{bhatt.coeff}][]{Bhattacharyya,GuillermeCooper} or a permutation test based on null hypothesised multidimensional space following \cite{diaz2016global} (\texttt{null.test}).
+
+This function also allows additionally argument such as \texttt{rarefaction} (as described above) or \texttt{correction} to adjust eventual p-values when multiple parametric tests are ran.
+
+\section{Examples}
+Here I provide to classic example of disparity analysis using some of the functions described above.
+The data to reproduce these examples is available within the package and thus don't need to be downloaded:
+
+\subsubsection{\disp data}
+The package contains a dataset that is a subset from \cite{beckancient2014} including:
+
+\begin{itemize}
+    \item \texttt{BeckLee\_mat50}: an ordinated matrix for 50 mammals based on the distance between discrete morphological characters.
+    \item \texttt{BeckLee\_mat99}: an ordinated matrix for the same 50 mammals as in \texttt{BeckLee\_mat50} but also containing the reconstruction of their 49 ancestors.
+    \item \texttt{BeckLee\_tree}: a chronogram with the 50 mammals species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
+    \item \texttt{BeckLee\_ages}: the first and last occurrence data for 14 of the mammals species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
+    \item \texttt{disparity}: a pre-made \disp object based on the data above.
+\end{itemize}
+
+\subsubsection{Typical disparity through type analysis}
+
+\texttt{dispRity.through.time(BeckLee\_mat50, )}
+
+This function is a wrapping function that is the equivalent to:
+
+\texttt{dispRity(boot.matrix(time.subsamples(BeckLee\_mat50, )))}
+
+% The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
+% This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
+% Of course all the defaults can be changed if required, more on this later.
+
+% For a disparity through time analysis, you will need:
+
+%   * An ordinated matrix (we covered that above)
+%   * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element.
+To give your tree a root time (i.e.
+an age for the root), you can simply do `my_tree\$root.time <- my\_age`.
+%   * The required number of time subsamples (here `time = 3`)
+%   * Your favourite disparity metric (here the sum of variances)
+
+% Using the @beckancient2014 data described [above](#example-data):
+
+% ## Measuring disparity through time
+% disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
+%                                         time = 3, metric = c(sum, variances))
+% ```
+
+% This generates a `dispRity` object (see [here](#manipulating-dispRity-objects) for technical details).
+% When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:
+
+% ```{r}
+% ## Print the disparity_data object
+% disparity_data
+% ```
+
+% We asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
+
+% We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):
+
+% ```{r, fig.width=7, fig.height=7}
+% ## Summarising disparity through time
+% summary(disparity_data)
+
+% ## Plotting the results
+% plot(disparity_data, type = "continuous")
+
+% ## Testing for an difference among the time bins
+% disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
+% summary(disp_lm)
+
+% ## Note that ANOVA assumptions were not checked here!
+% ```
+
+% Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
+% You can also directly explore the specific function help files within R and navigate to related functions.
+
+\subsection{Typical disparity per groups analysis}
+
+% The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
+% For example, you could ask if there is a difference in disparity between two groups?
+
+% To perform such an analysis, you will need:
+
+%  * An matrix with rows as elements and columns as dimensions (always!)
+%  * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix.
+For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4.
+Note that elements can be present in multiple groups at once.
+%  * Your favourite disparity metric (here the sum of variances)
+
+% Using the @beckancient2014 data described [above](#example-data):
+
+% ```{r}
+% ## Creating the two groups (crown versus stem) as a list
+% mammal_groups <- list("crown" = c(16, 19:41, 45:50),
+%                       "stem" = c(1:15, 17:18, 42:44))
+
+% ## Measuring disparity for each group
+% disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
+%                                      metric = c(sum, variances))
+% ```
+
+% We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).
+
+% ```{r, fig.width=7, fig.height=7}
+% ## Print the disparity_data object
+% disparity_data
+
+% ## Summarising disparity in the different groups
+% summary(disparity_data)
+
+% ## Plotting the results
+% plot(disparity_data)
+
+% ## Testing for a difference between the groups
+% test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
+% ```
+
+\section{Supplementary informations}
+Supplementary information concerning each function's functionalities can be found in R in the individual function manuals as well as in the vignettes available online at @@@.
+
+-Miscellaneous (maybe conclusion)
+Simulations
+Future directions
+
+Conclusion
+
+% \subsubsection{The package's data}
+
+% Beck and Lee data
+% Beck and Lee tree
+% Beck and Lee ages
+
+% disparity  disparity data
+
+% \subsubsection{Running example}
+
+% \subsection{The \disp package}
+
+% \subsubsection{The package's main functions}
+
+% \subsection{Fork flow}
+
+% Workflow figure
+
+% \subsubsection{Input data}
+
+% \subsubsection{Manipulating the data}
+
+% \subsubsection{Measuring disparity}
+
+% \subsubsection{Summarising the data}
+
+% Link to graph one two three
+
+% \subsubsection{Testing hypothesis}
+
+% \subsection{The \disp object}
+
+% Disparity object structure figure
+
+% \subsection{Additional functions and data}
+
+% The package also contains additional functions not directly linked to the disparity workflow nor to the \disp object but that can be used as utilities for the studying discrete morphological data.
+
+% \begin{table}
+%     \begin{tabular}{ll}
+%         \hline
+%         Function & Description \\ 
+%         \hline
+%         \texttt{char.diff} & Calculates character differences. \\
+%         \texttt{clean.data} & Cleans phylogenetic data (tree and table). \\
+%         \texttt{sim.morpho} & Simulates morphological discrete data. \\
+%         \texttt{slice.tree} & Slice through a phylogenetic tree. \\
+%         \texttt{space.maker} & Creates multidimensional spaces. \\
+%         \texttt{tree.age} & Calculates the age of nodes and tips in a tree. \\
+%         \hline
+%     \end{tabular}
+%     \caption{Additional functions in the \disp package.}
+% \end{table}
+
+% Additionally, this package also contains an ecology type data set from McClean
+
+% McClean data
+
+% \section{Conclusion}
+
+% This package is good.
+
+% \section{Data availability and reproducibility}
+% Data are available on Dryad or Figshare.
+% Code for reproducing the analyses is available on GitHub (\url{github.com/TGuillerme/SpatioTemporal_Disparity}).
 
 \section{Acknowledgments}
-Thanks to Dave Bapst, Martin Brazeau, Natalie Cooper, Andrew Jackson, Graeme Lloyd and Emma Sherratt.
+Many thanks to Natalie Cooper for encouraging the development of this package and helping for the redaction of this paper and the package manuals.
+Thanks to Dave Bapst, Martin Brazeau, Andrew Jackson, Graeme Lloyd and Emma Sherratt for specific implementation ideas.
 I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.

### 2017-08-15 inst/gitbook/01_glossary.Rmd
moved manuals to inst/

--- 2017-08-10 doc/dispRity-manual.Rmd
@@
 ---
-title: "dispRity manual"
-author: "Thomas Guillerme"
+title: "Glossary"
+author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
 date: "`r Sys.Date()`"
 output:
-    rmarkdown::html_vignette:
-        toc: true
-        toc_depth: 2
-bibliography: References.bib
-bst: sysbio.bst
-vignette: >
-  %\VignetteIndexEntry{dispRity manual}
-  %\VignetteEngine{knitr::rmarkdown}
-  \usepackage[utf8]{inputenc}
+  html_document: default
+  pdf_document: default
 ---
 
-# `dispRity`
-
-This is a package for measuring disparity in `R`.
-It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values so they can easily be compared.
-This manual is based on the version `1.0`.
-
-# Introduction
-
-## What is `dispRity`?
-This is a package for measuring disparity in `R`.
-It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
-Typically, these analysis are used in palaeobiology and evolutionary biology to study the changes in morphology through time.
-However, there are many more applications in ecology, evolution and beyond.
-
-## Installing and running the package
-
-You can install this package easily if you use the latest version of `R` (> 3.3.0) and `devtools`.
-
-```{r, eval=FALSE}
-## Checking if devtools is already installed
-if(!require(devtools)) install.packages("devtools")
-
-## Installing the latest released version directly from GitHub
-install_github("TGuillerme/dispRity", ref = "release")
-
-## Loading the package
-library(dispRity)
-```
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
-# NC: this isn't checking Claddis...
-library(dispRity)
-## Setting the random seed for repeatability
-set.seed(123)
-```
-
-Note this uses the `release` branch (version`0.4`). <!-- @@@ -->
-For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
-`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, `Claddis`, `geomorph` and `RCurl`).
-
-## Why not CRAN?
-
-This package is not available on CRAN.
-This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
-However, the package follows the strict (and useful!) CRAN standards via Travis.
-
-[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
-[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
-[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
-
-## Help
-
-If you need help with the package, hopefully the following manual will be useful.
-However, parts of this package are still in development and some other parts are probably not covered.
-Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
-
-## Citations
-
-You can cite both the package or this manual with the following citation:
-
-> Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646
-
-<!-- TG: I'm gonna change that to a newest version with the submission -->
-
-Note that this citation is only temporary (but can still be used!).
-A future proper version of the latest package release, this manual and an associated methods paper should be submitted soon(ish!).
-
-# Getting started
-
-## What sort of data does dispRity work with?
-
-Disparity can be estimated from pretty much any matrix.
-Classically, however, it is measured from ordinated matrices.
-These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).
-
-### Ordination matrices from `Claddis`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
-if(!require(Claddis)) install.packages("Claddis")
-```
-
-`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
-For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:
-
-```{r}
-require(Claddis)
-
-## Ordinating the example data from Claddis
-Claddis.ordination(Michaux1989) 
-```
-
-Note that several options are available, namely which type of distance should be computed.
-See more info in the function manual (`?Claddis.ordination`).
-Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.
-
-### Ordination matrices from `geomorph`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if geomorph is installed BUILD ONLY
-if(!require(geomorph)) install.packages("geomorph")
-```
-
-You can also easily use data from `geomorph` using the `geomorph.ordination` function.
-This function simply takes Procrustes aligned data and performs an ordination:
-
-```{r}
-require(geomorph)
-
-## Loading the plethodon dataset
-data(plethodon)
-
-## Performing a Procrustes transform on the landmarks
-procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)
-
-## Ordinating this data
-geomorph.ordination(procrustes)[1:5,1:5]
-```
-
-Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
-Additionally you can give the function a `geomorph.data.frame` object.
-If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsamples)!
-
-```{r}
-## Using a geomorph.data.frame
-geomorph_df <- geomorph.data.frame(procrustes,
-     species = plethodon$species, site = plethodon$site)
-
-## Ordinating this data and making a dispRity object
-geomorph.ordination(geomorph_df)
-```
-
-More about these `dispRity` objects below!
-
-### Other kinds of ordination matrices
-
-If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordination matrices by using the following functions from the `stats` package.
-Here is how to do it for the following types of matrices:
-
- * Multivariate matrices (principal components analysis; PCA)
-
-```{r}
-## A multivariate matrix
-head(USArrests)
-
-## Ordinating the matrix using `prcomp` 
-ordination <- prcomp(USArrests)
-
-## Selecting the ordinated matrix
-ordinated_matrix <- ordination$x
-head(ordinated_matrix)
-```
-  This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, see the `?princomp` function.
-
- * Distance matrices (classical multidimensional scaling; MDS)
-
-```{r}
-## A matrix of distances between cities
-str(eurodist)
-
-## Ordinating the matrix using cmdscale() with k = 5 dimensions 
-ordinated_matrix <- cmdscale(eurodist, k = 5)
-head(ordinated_matrix)
-```
-
-  This results in a ordinated matrix with European cities as elements and five dimensions.
-
-Of course any other method for creating the ordination matrix is totally valid, you can also not use any ordination at all!
-The only requirements for the `dispRity` functions is that the input is a matrix with elements as rows and dimensions as columns.
-
-## Performing a simple dispRity analysis
-
-Two `dispRity` functions allow users to run an analysis pipeline simply by inputting an ordination matrix.
-These functions allow users to either calculate the disparity through time (`dispRity.through.time`) or the disparity of user-defined groups (`dispRity.per.group`).
-
-### Example data
-
-To illustrate these functions, we will use data from @beckancient2014.
-This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
-
-```{r, fig.width=7, fig.height=7}
-## Loading the ordinated matrices
-data(BeckLee_mat50)
-data(BeckLee_mat99)
-
-## The first five taxa and dimensions of the 50 taxa matrix
-head(BeckLee_mat50[, 1:5])
-
-## The first five taxa and dimensions of the 99 taxa + ancestors matrix
-BeckLee_mat99[c(1, 2, 98, 99), 1:5]
-
-## Loading a list of first and last occurrence dates for the fossils
-data(BeckLee_ages)
-head(BeckLee_ages)
-
-## Loading and plotting the phylogeny
-data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.8) 
-axisPhylo(root = 140)
-nodelabels(cex = 0.5)
-```
-
-Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).
-
-### Disparity through time
-
-The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
-This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
-Of course all the defaults can be changed if required, more on this later.
-
-For a disparity through time analysis, you will need:
-
-  * An ordinated matrix (we covered that above)
-  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element.
-To give your tree a root time (i.e.
-an age for the root), you can simply do `my_tree$root.time <- my_age`.
-  * The required number of time subsamples (here `time = 3`)
-  * Your favourite disparity metric (here the sum of variances)
-
-Using the @beckancient2014 data described [above](#example-data):
-
-```{r}
-## Measuring disparity through time
-disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
-                                        time = 3, metric = c(sum, variances))
-```
-
-This generates a `dispRity` object (see [here](#manipulating-dispRity-objects) for technical details).
-When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:
-
-```{r}
-## Print the disparity_data object
-disparity_data
-```
-
-We asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
-
-We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):
-
-```{r, fig.width=7, fig.height=7}
-## Summarising disparity through time
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data, type = "continuous")
-
-## Testing for an difference among the time bins
-disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
-summary(disp_lm)
-
-## Note that ANOVA assumptions were not checked here!
-```
-
-Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
-You can also directly explore the specific function help files within R and navigate to related functions.
-
-### Disparity among groups
-
-The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
-For example, you could ask if there is a difference in disparity between two groups?
-
-To perform such an analysis, you will need:
-
- * An matrix with rows as elements and columns as dimensions (always!)
- * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix.
-For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4.
-Note that elements can be present in multiple groups at once.
- * Your favourite disparity metric (here the sum of variances)
-
-Using the @beckancient2014 data described [above](#example-data):
-
-```{r}
-## Creating the two groups (crown versus stem) as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44))
-
-## Measuring disparity for each group
-disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
-                                     metric = c(sum, variances))
-```
-
-We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).
-
-```{r, fig.width=7, fig.height=7}
-## Print the disparity_data object
-disparity_data
-
-## Summarising disparity in the different groups
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data)
-
-## Testing for a difference between the groups
-test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
-```
-
-# Specific tutorials
-
-The following section contains information specific to some functions.
-If any of your questions are not covered in these sections, please refer to the function help files in R, send me an email (<guillert@tcd.ie>), or raise an issue on GitHub.
-The several tutorials below describe specific functionalities of certain functions; please always refer to the function help files for the full function documentation!
-
-## Time slicing
-
-The function `time.subsamples` allows users to divide the matrix into different time subsamples or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
-Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.
-
-Two types of time subsamples can be performed by using the `method` option:
-
- *  Discrete time subsamples (or time-binning) using `method = discrete`
- *  Continuous time subsamples (or time-slicing) using `method = continuous`
-
-For the time-slicing method details see @GuillermeSTD. <!-- @@@ Change cite appropriately! -->
-<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
-For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:
-
- *  Defining the boundaries of the time bins (when `method = discrete`)
- *  Defining the time slices (when `method = continuous`)
-
-Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
-Additionally, it is also possible to input a dataframe containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age, so taxa can appear in more than one time bin/slice.
-
-Here is an example for `method = discrete`:
-
-```{r, eval=TRUE}
-## Generating three time bins containing the taxa present every 40 Ma
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0))
-```
-
-Note that we can also generate equivalent results by just telling the function that we want three time-bins as follow:
-
-```{r, eval=TRUE}
-## Automatically generate three equal length bins:
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = 3)
-```
-
-In this example, the taxa were split inside each time-bin according to their age.
-However, the taxa here are considered as single points in time.
-It is totally possible that some taxa could have had longer longevity and that they exist in multiple time bins.
-In this case, it is possible to include them in more than one bin by providing a table of first and last occurrence dates (FAD/LAD).
-This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:
-
-```{r, eval=TRUE}
-## Displaying the table of first and last occurrence dates for each taxa
-head(BeckLee_ages)
-
-## Generating time bins including taxa that might span between them
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
-```
-
-When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as the root age plus 1\% of the tree length, as long as at least three elements/taxa are present at that point in time.
-The algorithm adds an extra 1\% tree length until reaching the required minimum of three elements.
-It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance among tips *and* nodes.
-
-For the time-slicing method (`method = continuous`), the idea is fairly similar.
-This option, however, requires a matrix that contains the ordinated distance among taxa *and* nodes and an extra argument describing the assumed evolutionary model (via the `model` argument).
-This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node, meaning that a decision must be made about what the value for the branch should be.
-The model can be one of the following:
-
- *  `acctran` where the data chosen along the branch is always the one of the descendant
- *  `deltran` where the data chosen along the branch is always the one of the ancestor
- *  `punctuated` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
- *  `gradual` where the data chosen along the branch is either the descendant or the ancestor depending on branch length
-
-```{r, eval=TRUE}
-## Generating four time slices every 40 million years under a model of gradual evolution
-time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-
-## Generating four time slices automatically
-time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree,
-    method = "continuous", model = "gradual", time = 4, FADLAD = BeckLee_ages)
-```
-
-## Customised subsamples
-
-Another way of separating elements into different categories is to use customised subsamples as briefly explained [above](#disparity-among-groups).
-This function simply takes the list of elements to put in each group (whether they are the actual element names or their position in the matrix).
-
-```{r, eval=TRUE}
-## Creating the two groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44))
-
-## Separating the dataset into two different groups
-custom.subsamples(BeckLee_mat50, group = mammal_groups)
-```
-
-Elements can easily be assigned to different groups if necessary!
-
-```{r, eval=FALSE}
-## Creating the three groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44).
-                      "all" = c(1:50))
-```
-
-## Bootstraps and rarefactions
-
-One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results are, and how sensitive they are to outliers in the dataset.
-This can be achieved using the function `boot.matrix` to bootstrap and/or rarefy the data.
-The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):
-
-```{r, eval=TRUE}
-## Default bootstrapping
-boot.matrix(data = BeckLee_mat50)
-```
-
-The number of bootstrap replicates can be defined using the `bootstraps` option.
-The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
-Currently two algorithms are implemented:
-
- * `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any *m* elements drawn from the data)
- * `single` where only one random element is replaced by one other random element for each pseudo-replicate
-
-```{r, eval=TRUE}
-## Bootstrapping with the single bootstrap method
-boot.matrix(BeckLee_mat50, boot.type = "single")
-```
-
-This function also allows users to rarefy the data using the `rarefaction` argument.
-Rarefaction allows users to limit the number of elements to be drawn at each bootstrap replication.
-This is useful if, for example, one is interested in looking at the effect of reducing the number of elements on the results of an analysis.
-
-This can be achieved by using the `rarefaction` option that draws only *n-x* at each bootstrap replicate (where *x* is the number of elements not sampled).
-The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsample to a minimum of three elements).
-It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
-
-```{r, eval=TRUE}
-## Bootstrapping with the full rarefaction
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
-
-## Or with a set number of rarefaction levels
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8, 3))
-```
-
-One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
-When missing, all dimensions from the ordinated matrix are used.
-
-```{r, eval=TRUE}
-## Using the first 50% of the dimensions
-boot.matrix(BeckLee_mat50, dimensions = 0.5)
-
-## Using the first 10 dimensions
-boot.matrix(BeckLee_mat50, dimensions = 10)
-```
-
-Of course, one could directly supply the subsamples generated above (using `time.subsamples` or `custom.subsamples`) to this function.
-
-```{r, eval=TRUE}
-## Creating subsamples of crown and stem mammals
-crown_stem <- custom.subsamples(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot.matrix(crown_stem, bootstraps = 200, rarefaction = TRUE)
-
-## Creating time slice subsamples
-time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-                               method = "continuous", model = "gradual", 
-                               time = c(120, 80, 40, 0),
-                               FADLAD = BeckLee_ages)
-
-## Bootstrapping the time slice subsamples
-boot.matrix(time_slices, bootstraps = 100)
-```
-
-## Disparity metrics
-
-There are many ways of measuring disparity!
-In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
-For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @Wills1994.
-Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.
-
-Because of this great diversity of metrics, the package `dispRity` does not have one way to measure disparity but rather proposes to facilitate users in defining their own disparity metric that will best suit their particular analysis.
-In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
-However the `metric` argument has to follow certain rules:
-
-1.  It must be composed from one to three `function` objects;
-2.  The function(s) must take as a first argument a `matrix` or a `vector`;
-3.  The function(s) must be of one of the three dimension-levels described below;
-4.  At least one of the functions must be of dimension-level 1 or 2 (see below).
-
-### The function dimension-levels
-
-The metric function dimension-levels determine the "dimensionality of decomposition" of the input matrix.
-In other words, each dimension-level designates the dimensions of the output, i.e. either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimension.
-
-![Illustration of the different dimension-levels of functions with an input `matrix`](dispRity_fun.png)
-
-#### Dimension-level 1 functions
-
-A dimension-level 1 function will decompose a `matrix` or a `vector` into a single value:
-
-```{r}
-## Creating a dummy matrix
-dummy_matrix <- matrix(rnorm(12), 4, 3)
-
-## Example of dimension-level 1 functions
-mean(dummy_matrix)
-median(dummy_matrix)
-```
-
-Any summary metric such as mean or median are good examples of dimension-level 1 functions as they reduce the matrix to a single dimension (i.e. one value).
-
-#### Dimension-level 2 functions
-
-A dimension-level 2 function will decompose a `matrix` into a `vector`.
-
-```{r}
-## Defining the function as the product of rows
-prod.rows <- function(matrix) apply(matrix, 1, prod)
-
-## A dimension-level 2 metric
-prod.rows(dummy_matrix)
-```
-
-Several dimension-level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculate the variance or the range of each dimension of the ordinated matrix respectively.
-
-#### Dimension-level 3 functions
-
-Finally a dimension-level 3 function will transform the matrix into another matrix.
-Note that the dimension of the output matrix doesn't need to match the the input matrix:
-
-```{r}
-## A dimension-level 3 metric
-var(dummy_matrix)
-
-## A dimension-level 3 metric with a forced matrix output
-as.matrix(dist(dummy_matrix))
-```
-
-### `make.metric`
-
-Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the Euclidean distance between each element and the centroid of the ordinated space.
-The `make.metric` function implemented in `dispRity` is designed to help test and find the dimension-level of the functions.
-This function tests:
-
-1.  If your function can deal with a `matrix` or a `vector` as an input;
-2.  Your function's dimension-level according to its output (dimension-level 1, 2 or 3, see above);
-3.  Whether the function can be implemented in the `dispRity` function (the function is fed into a `lapply` loop).
-
-For example, let's see if the functions described above are the right dimension-levels:
-
-```{r}
-## Which dimension-level is the mean function? And can it be used in dispRity?
-make.metric(mean)
-
-## Which dimension-level is the prod.rows function? And can it be used in dispRity?
-make.metric(prod.rows)
-
-## Which dimension-level is the var function? And can it be used in dispRity?
-make.metric(var)
-```
-
-A non verbose version of the function is also available.
-This can be done using the option `silent = TRUE` and will simply output the dimension-level of the metric.
-
-```{r}
-## Testing whether mean is dimension-level 1
-if(make.metric(mean, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-## Testing whether var is dimension-level 1
-if(make.metric(var, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-```
-
-### Metrics in the `dispRity` function
-
-Using this metric structure, we can easily use any disparity metric in the `dispRity` function as follows:
-
-```{r}
-## Measuring disparity as the standard deviation of all the values of the
-## ordinated matrix (dimension-level 1 function).
-summary(dispRity(BeckLee_mat50, metric = sd))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the ordinated matrix (dimension-level 1 and 2 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the variance covariance matrix (dimension-level 1, 2 and 3 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
-```
-
-Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > 2 > 1).
-
-```{r}
-## Disparity as the standard deviation of the variance of each axis of the
-## variance covariance matrix:
-disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)),
-                      round = 10)
-
-## Same as above but using a different function order for the metric argument
-disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
-                      round = 10)
-
-## Both ways output the same disparity values:
-disparity1 == disparity2
-```
-
-In these examples, we considered disparity to be a single value.
-For example, in the previous example, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
-It is, however, possible to calculate [disparity as a distribution](#disparity-as-a-distribution).
-
-### Metrics implemented in `dispRity`
-
-Several disparity metrics are implemented in the `dispRity` package.
-The detailed list can be found in `?dispRity.metric` along with some description of each metric.
-
-Level | Name | Description | Source | 
-------|------|---------------------------------------------------|--------|
-3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
-3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
-3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
-2 | `ranges` | The range of each dimension | `dispRity` |
-2 | `variances` | The variance of each dimension | `dispRity` |
-2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
-1 | `mode.val` | The modal value | `dispRity` |
-
-1: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (not PCA).
-
-2: Note that by default, the centroid is the centroid of the elements.
-It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
-
-### Equations and implementations
-Some of the functions described below are implemented in the `dispRity` package and do not require any other packages to calculate ([see implementation here](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R)).
-
-\begin{equation}
-    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
-\end{equation}
-Where *k* is the number of dimensions, and $\lambda_i$ is the eigenvalue of each dimension.
-
-\begin{equation}
-    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
-\end{equation}
-Where *k* is the number of dimensions.
-
-\begin{equation}
-    ranges = |max(k_i) - min(k_i)|
-\end{equation}
-Where *k* is the number of dimensions.
-
-\begin{equation}
-    variances = \sigma^{2}{k_i}
-\end{equation}
-Where *k* is the number of dimensions, and $\sigma^{2}$ is their variance.
-
-\begin{equation}
-    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
-\end{equation}
-Where *n* is each element in the ordinated space, *k* is the number of dimensions, and $Centroid_{k}$ is their mean (or can be set to another value).
-
-### Using the different disparity metrics
-Here is a brief demonstration of the main metrics implemented in `dispRity`.
-First, we will create a dummy/simulated ordinated space using the `space.maker` utility function (more about that [here](#space.maker):
-
-```{r}
-## Creating a 10*5 normal space
-set.seed(1)
-dummy_space <- space.maker(10, 5, rnorm)
-```
-
-We will use this simulated space to demonstrate the different metrics.
-
-#### Volumes and surface metrics
-The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `hyper.volume` all measure the surface or the volume of the ordinated space occupied:
-
-```{r}
-## Calculating the ellipsoid volume
-summary(dispRity(dummy_space, metric = ellipse.volume))
-```
-> Because there is only one subsample (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with rounding).
-
-```{r}
-## Calculating the convex hull surface
-summary(dispRity(dummy_space, metric = convhull.surface))
-
-## Calculating the convex hull volume
-summary(dispRity(dummy_space, metric = convhull.volume))
-```
-
-The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
-For the full and correct calculation of the volume of the ordinated space, it is possible to use the `hyper.volume` function that has more options but takes longer to calculate.
-
-```{r}
-## Calculating the true multidimensional volume
-summary(dispRity(dummy_space, metric = hyper.volume))
-
-```
-> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) that often results in near 0 disparity values.
-
-#### Ranges, variances and diagonal
-
-The functions `ranges`, `variances` and `diagonal` all measure properties of the ordinated space based on its dimensional properties (they are also less affected by the "curse of dimensionality"):
-
-`ranges` and `variances` both work on the same principle and measure the range/variance of each dimension:
-
-```{r}
-## Calculating the ranges of each dimension in the ordinated space
-ranges(dummy_space)
-
-## Calculating disparity as the distribution of these ranges
-summary(dispRity(dummy_space, metric = ranges))
-
-## Calculating disparity as the sum and the product of these ranges
-summary(dispRity(dummy_space, metric = c(sum, ranges)))
-summary(dispRity(dummy_space, metric = c(prod, ranges)))
-
-## Calculating the variances of each dimension in the ordinated space
-variances(dummy_space)
-
-## Calculating disparity as the distribution of these variances
-summary(dispRity(dummy_space, metric = variances))
-
-## Calculating disparity as the sum and the product of these variances
-summary(dispRity(dummy_space, metric = c(sum, variances)))
-summary(dispRity(dummy_space, metric = c(prod, variances)))
-```
-
-The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest Euclidean distance in our five dimensional space):
-
-```{r}
-## Calculating the ordinated space's diagonal
-summary(dispRity(dummy_space, metric = diagonal))
-```
-
-> This metric is only a Euclidean diagonal (mathematically valid) if the dimensions within the space are all orthogonal!
-
-#### Centroids metric
-
-The `centroids` metric allows users to measure the position of the different elements compared to a fixed point in the ordinated space.
-By default, this function measures the distance between each element and their centroid (centre point):
-
-```{r}
-## The distribution of the distances between each element and their centroid
-summary(dispRity(dummy_space, metric = centroids))
-
-## Disparity as the median value of these distances
-summary(dispRity(dummy_space, metric = c(median, centroids)))
-```
-
-It is however possible to fix the coordinates of the centroid to a specific point in the ordinated space, as long as it has the correct number of dimensions:
-
-```{r}
-## The distance between each element and the origin of the ordinated space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))
-
-## Disparity as the distance between each element and a specific point in space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
-```
-
-## Summarising dispRity data (plots)
-
-Because of its architecture, printing `dispRity` objects only summarises their content but does not print the disparity value measured or associated analysis (more about this [here](#manipulating-dispRity-objects)).
-To actually see what is in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphical representation of the results.
-
-### Summarising `dispRity` data
-This function is an S3 function (`summary.dispRity`) allowing users to summarise the content of `dispRity` objects that contain disparity calculations.
-
-<!-- DATA LOADER -->
-
-```{r}
-## Example data from previous sections
-crown_stem <- custom.subsamples(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
-## Calculate disparity
-disparity_crown_stem <- dispRity(boot_crown_stem, metric = c(sum, variances))
-
-## Creating time slice subsamples
-time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-## Bootstrapping the time slice subsamples
-boot_time_slices <- boot.matrix(time_slices, bootstraps = 100)
-## Calculate disparity
-disparity_time_slices <- dispRity(boot_time_slices, metric = c(sum, variances))
-
-## Creating time bin subsamples
-time_bins <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages,
-    inc.nodes = TRUE)
-## Bootstrapping the time bin subsamples
-boot_time_bins <- boot.matrix(time_bins, bootstraps = 100)
-## Calculate disparity
-disparity_time_bins <- dispRity(boot_time_bins, metric = c(sum, variances))
-```
-
-These objects are easy to summarise as follows:
-
-```{r}
-## Default summary
-summary(disparity_time_slices)
-```
-
-Information about the number of elements in each subsample and the observed (i.e. non-bootstrapped) disparity are also calculated.
-This is specifically handy when rarefying the data for example:
-
-```{r}
-head(summary(disparity_crown_stem))
-```
-
-The summary functions can also take various options such as:
-
- * `quantile` values for the confidence interval levels (by default, the 50 and 95 quantiles are calculated)
- * `cent.tend` for the central tendency to use for summarising the results (default is `median`)
- * rounding` option corresponding to the number of decimal places to print (default is `2`)
- * `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
-
-These options can easily be changed from the defaults as follows:
-
-```{r}
-## Same as above but using the 88th quantile and the standard deviation as the summary 
-summary(disparity_time_slices, quantile = 88, cent.tend = sd)
-
-## Printing the details of the object and rounding the values to the 5th decimal place
-summary(disparity_time_slices, recall = TRUE, rounding = 5)
-```
-
-Note that the summary table is a `data.frame`, hence it is as easy to modify as any dataframe using `dplyr`.
-You can also export it in `csv` format using `write.csv` or `write_csv` or even directly export into `LaTeX` format using the following;
-
-```{r, eval=FALSE}
-## Loading the xtable package
-require(xtable)
-## Converting the table in LaTeX
-xtable(summary(disparity_time_slices))
-```
-
-### Plotting `dispRity` data
-An alternative (and more fun!) way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
-This function takes the same options as `summary.dispRity` along with various graphical options described in the function help files (see `?plot.dispRity`).
-
-The plots can be of four different types:
-
- * `continuous` for displaying continuous disparity curves
- * `box`, `lines`, and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.
-
-> This argument can be left empty. In this case, the algorithm will automatically detect the type of subsamples from the `dispRity` object and plot accordingly.
-
-It is also possible to display the number of elements in each subsample (as a horizontal dotted line) using the option `elements = TRUE`.
-Additionally, when the data is rarefied, one can indicate which level of rarefaction to display (i.e. only display the results for a certain number of elements) by using the `rarefaction` argument.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(mfrow = c(2, 2), bty = "n")
-
-## Plotting continuous disparity results
-plot(disparity_time_slices, type = "continuous")
-
-## Plotting discrete disparity results
-plot(disparity_crown_stem, type = "box")
-
-## As above but using lines for the rarefaction level of 20 elements only
-plot(disparity_crown_stem, type = "line", rarefaction = 20)
-
-## As above but using polygons while also displaying the number of elements
-plot(disparity_crown_stem, type = "polygon", elements = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Since `plot.dispRity` uses the arguments from the generic `plot` method, it is of course possible to change pretty much everything using the regular plot arguments:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some classic options from plot
-plot(disparity_time_slices, col = c("blue", "orange", "green"),
-    ylab = c("Some measurement"), xlab = "Some other measurement",
-    main = "Many options...", ylim = c(5, 0), xlim = c(4, 0))
-
-## Adding a legend
-legend("topleft", legend = c("Central tendency",
-                             "Confidence interval 1",
-                            "Confidence interval 2"),
-      col = c("blue", "orange", "green"), pch = 19)
-
-## Resetting graphical parameters
-par(op)
-```
-
-In addition to the classic `plot` arguments, the function can also take arguments that are specific to `plot.dispRity` like adding the number of elements or rarefaction level (as described above), and also changing the values of the quantiles to plot as well as the central tendency.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some plot.dispRity arguments
-plot(disparity_time_slices, quantile = c(seq(from = 10, to = 100, by = 10)),
-    cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
-    ylab = c("Disparity", "Diversity"), time.subsamples = FALSE,
-    xlab = "Time (in in units from past to present)", observed = TRUE,
-    main = "Many more options...")
-
-## Resetting graphical parameters
-par(op)
-```
-
-> Note that the argument `observed = TRUE` allows to plot the disparity values calculated from the non-bootstrapped data as crosses on the plot.
-
-For comparing results, it is also possible to add a plot to the existent plot by using `add = TRUE`:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the continuous disparity with a fixed y axis
-plot(disparity_time_slices, ylim = c(1, 4))
-## Adding the discrete data
-plot(disparity_time_bins, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
-    add = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Finally, if your data has been fully rarefied, it is also possible to easily look at rarefaction curves by using the `rarefaction = TRUE` argument:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the rarefaction curves
-plot(disparity_crown_stem, rarefaction = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-## Testing disparity hypotheses
-
-The `dispRity` package allows users to apply statistical tests to the calculated disparity to test various hypotheses.
-The function `test.dispRity` works in a similar way to the `dispRity` function: it takes a `dispRity` object, a `test` and a `comparisons` argument.
-
-The `comparisons` argument indicates the way the test should be applied to the data:
-
- * `pairwise` (default): to compare each subsample in a pairwise manner
- * `referential`: to compare each subsample to the first subsample
- * `sequential`: to compare each subsample to the following subsample
- * `all`: to compare all the subsamples together (like in analysis of variance)
-
-It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsample names to create personalised tests.
-Some other tests implemented in `dispRity` such as the `dispRity::null.test` have a specific way they are applied to the data and therefore ignore the `comparisons` argument. 
-<!-- Add sequential test one day! -->
-
-The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
-It can be a common statistical test function (e.g. `stats::t.test`), a function implemented in `dispRity` (e.g. see `?null.test`) or any function defined by the user.
-
-This function also allows users to correct for Type I error inflation (false positives) when using multiple comparisons via the `correction` argument. 
-This argument can be empty (no correction applied) or can contain one of the corrections from the `stats::p.adjust` function (see `?p.adjust`).
-
-Note that the `test.dispRity` algorithm deals with some classical test outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
-It is, however, possible to get the full detailed output by using the options `details = TRUE`.
-
-```{r}
-## T-test to test for a difference in disparity between crown and stem mammals
-test.dispRity(disparity_crown_stem, test = t.test)
-
-## Performing the same test but with the detailed t.test output
-test.dispRity(disparity_crown_stem, test = t.test, details = TRUE)
-
-## Wilcoxon test applied to time sliced disparity with sequential comparisons, with Bonferroni correction
-test.dispRity(disparity_time_slices, test = wilcox.test,
-              comparisons = "sequential", correction = "bonferroni")
-
-## Measuring the overlap between distributions in the time bins (using the
-## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)
-test.dispRity(disparity_time_bins, test = bhatt.coeff)
-```
-
-It is also possible to apply some more *complex* tests that have their own output classes (like `stats::lm`).
-
-The results can then be analysed as usual using the associated `summary` S3 method:
-
-```{r}
-## Performing and linear model applied to the same data
-(slice_lm <- test.dispRity(disparity_time_slices, test = lm,
-                            comparisons = "all"))
-
-## The output is a regular `lm` output
-class(slice_lm)
-
-## This output can be summarised using summary
-summary(slice_lm)
-```
-
-Of course, due to the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
-The only condition is that the test can be applied to at least two distributions.
-In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm` for example).
-
-## Disparity as a distribution
-
-Disparity is often regarded as a summary value of the position of the all elements in the ordinated space.
-For example, the sum of variances, the product of ranges or the median distance between the elements and their centroid will summarise disparity as a single value.
-This value can be pseudo-replicated (bootstrapped) to obtain a distribution of the summary metric with estimated error.
-However, another way to perform disparity analysis is to use the *whole distribution* rather than just a summary metric (e.g. the variances or the ranges).
-
-This is possible in the `dispRity` package by calculating disparity as a dimension-level 2 metric only!
-Let's have a look using our previous example of bootstrapped time slices but by measuring the distances between each taxon and their centroid as disparity.
-
-```{r}
-## Measuring disparity as a whole distribution
-disparity_centroids <- dispRity(boot_time_slices, metric = centroids)
-```
-
-The resulting disparity object is of dimension-level 2, so it can easily be transformed into a dimension-level 1 object by, for example, measuring the median distance of all these distributions:
-
-```{r}
-## Measuring median disparity in each time slice
-disparity_centroids_median <- dispRity(disparity_centroids, metric = median)
-```
-
-And we can now compare the differences between these methods:
-
-```{r}
-## Summarising both disparity measurements:
-## The distributions:
-summary(disparity_centroids)
-## The summary of the distributions (as median)
-summary(disparity_centroids_median)
-```
-
-We can see that the summary message for the distribution is slightly different than before.
-Here `summary` also displays the observed central tendency (i.e. the central tendency of the measured distributions).
-Note that, as expected, this central tendency is the same in both metrics!
-
-Another, maybe more intuitive way, to compare both approaches for measuring disparity is to plot the distributions:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(bty = "n", mfrow = c(1, 2))
-
-## Plotting both disparity measurements
-plot(disparity_centroids, ylab = "Distribution of all the distances")
-plot(disparity_centroids_median,
-     ylab = "Distribution of the medians of all the distances")
-
-par(op)
-```
-
-We can then test for differences in the resulting distributions using `test.dispRity` and the `bhatt.coeff` test as described above.
-
-```{r}
-## Probability of overlap in the distribution of medians
-test.dispRity(disparity_centroids_median, test = bhatt.coeff)
-```
-
-In this case, we are looking at the probability of overlap of the distribution of median distances from centroids among each pair of time slices.
-In other words, we are measuring whether the medians from each bootstrap pseudo-replicate for each time slice overlap.
-But of course, we might be interested in the actual distribution of the distances from the centroid rather than simply their central tendencies.
-This can be problematic depending on the research question asked since we are effectively comparing non-independent medians distributions (because of the pseudo-replication).
-
-One solution, therefore, is to look at the full distribution:
-
-```{r}
-## Probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff)
-```
-
-These results show the actual overlap among all the measured distances from centroids concatenated across all the bootstraps.
-For example, when comparing the slices 120 and 80, we are effectively comparing the 5 \times 100 distances (the distances of the five elements in slice 120 bootstrapped 100 times) to the 19 \times 100 distances from slice 80.
-However, this can also be problematic for some specific tests since the _n_ \times 100 distances are also pseudo-replicates and thus are still not independent.
-
-A second solution is to compare the distributions to each other _for each replicate_:
-
-```{r}
-## Boostrapped probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff, concatenate = FALSE)
-```
-
-These results show the median overlap among pairs of distributions in the first column (`bhatt.coeff`) and then the distribution of these overlaps among each pair of bootstraps.
-In other words, when two distributions are compared, they are now compared for each bootstrap pseudo-replicate, thus effectively creating a distribution of probabilities of overlap.
-For example, when comparing the slices 120 and 80, we have a mean probability of overlap of 0.28 and a probability between 0.18 and 0.43 in 50\% of the pseudo-replicates.
-Note that the quantiles and central tendencies can be modified via the `conc.quantiles` option.
-
-## Ecological studies of disparity
-<!-- NC: I wonder if this might be better just as a "hey we can do ecology stuff too and here is a link to a demo" rather than a new section?
-The manual is loooooooooong so splitting some stuff out might be good?
--->
-
-Disparity analyses are not restricted to palaeobiology analysis.
-This package is designed for dealing with ordinated matrices, no matter which data was used in the ordination.
-Here is an short example of disparity analysis using ecological data from Deirdre McClean (unpubl.).
-
-The dataset contains an ordinated matrix of 20 dimensions (columns) for 40 elements (rows).
-The elements are different experimental plots with discrete variations of nutrient enrichment, depth, and freshwater benthic invertebrates.
-<!-- NC: Explain a little more about what the data are here as I don't entirely get it. EG what's a treatement etc? Just a really simple explanataion is perfect -->
-
-```{r}
-## Loading the data
-data(McClean_data)
-
-## Ordinated matrix
-ordinated_matrix <- McClean_data$ordination
-
-## Treatments and depths
-treatments <- McClean_data$treatment
-depth <- McClean_data$depth
-```
-
-### Classic ecological "disparity" analysis
-
-A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different experimental plots differ in their values for nutrient enrichment, depth, and freshwater benthic invertebrates.
-<!-- NC: or similar just to make it clearer -->
-
-```{r, fig.width=6, fig.height=6}
-## Setting the colors that will represent the treatments
-cols <- sub("a", "red", treatments)
-cols <- sub("b", "blue", cols)
-## Setting the symbols that will represent the depth
-pchs <- sub(1, 16, depth)
-pchs <- as.numeric(sub(2, 17, pchs))
-
-## Graphical option
-op <- par(bty = "n")
-## A classic 2D ordination plot
-plot(ordinated_matrix[, 1:2], col = cols, pch = pchs,
-     xlab = "PC 1", ylab = "PC 2",
-     xlim = range(ordinated_matrix[, 1:2]) + c(0, 100))
-par(op)
-```
-
-This shows the distribution of the experimental plots along the two first axis of variation of the ordinated distance matrix (i.e. the first two dimensions).
-At a first glance, it seems difficult to see a clear effect of the treatments (blue or red) or the depth (rounds or triangles) as the different experimental plots with the same parameters don't seem to cluster together.
-However, this may be because this plot ignores the 18 other dimensions of the ordination!
-Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e.
-the values on the first dimension do not represent a continuous trait (e.g.
-depth), instead they just represent the ordinations of correlations between the data and some factors.
-
-Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.
-
-### Multidimensional analysis with `dispRity`
-
-#### Splitting the data with respect to the different factors
-
-The first step is to create different subsamples that represent subsamples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
-Each of these subsamples will contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
-In our example, we are going to group the elements according to their depth and treatment.
-
-```{r}
-## Creating a table that contain the elements and their attributes
-factors <- as.data.frame(matrix(data = c(treatments, depth),
-          nrow = nrow(ordinated_matrix), ncol = 2, byrow = FALSE,
-          dimnames = list(rownames(ordinated_matrix))))
-names(factors) <- c("Treat", "Depth")
-head(factors)
-```
-
-Second, let’s split the data according by depth and treatment to create the subsamples of the ordinated space by using the `custom.subsamples` function:
-
-```{r}
-## Splitting the ordinated space into four subsamples
-(customised_subsamples <- custom.subsamples(ordinated_matrix, factors))
-```
-
-#### Calculating disparity
-
-In this example, we are going to define disparity as the n-dimensional ellipsoid volume (see `?hyper.volume`).
-
-```{r}
-## Calculating disparity as the ellipsoid volume of each group
-disparity <- dispRity(customised_subsamples, metric = diagonal)
-summary(disparity)
-```
-
-We can also bootstrap the data to test the robustness of the measured disparity to outliers.
-Also, as we can see, each subsample has different numbers of elements.
-It might also be interesting to rarefy the data to have only subsamples with the same number of elements.
-
-```{r}
-## Bootstrapping the subsamples and using rarefaction
-## (i.e. only re-sampling 17 elements each time)
-bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 100,
-                                 rarefaction = 17)
-```
-
-We can now run a more robust disparity analysis using the bootstrapped data:
-
-```{r}
-## Calculating the bootstrapped disparity
-disparity <- dispRity(bootstrapped_data, metric = diagonal)
-```
-
-#### Summarising and displaying the data
-
-Now let's summarise this data:
-
-```{r, fig.width=12, fig.height=6, eval=TRUE}
-## Summary table of the data
-summary(disparity)
-
-## Graphical options
-op <- par(mfrow = (c(1, 2)), bty = "n")
-
-## Plotting the bootstrapped disparity
-plot(disparity, main = "Bootstrapped", las = 2, xlab = "")
-
-## Plotting the rarefied disparity
-plot(disparity, rarefaction = 17, main = "Rarefied", las = 2, xlab = "")
-
-par(op)
-```
-
-As we can see, there seems to be no strong effect of the number of experimental plots in each subsample (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
-
-#### Testing our hypotheses
-
-Finally, we can test our hypothesis (whether the water treatment at certain depths alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.
-<!-- Ah so this is what you are testing! Make this clearer in the intro bit to this section -->
-
-```{r}
-## Testing the effect of treatment and depth on the bootstrapped data
-summary(test.dispRity(disparity, test = aov, comparisons = "all"))
-```
-
-## Simulating discrete morphological data
-
-The `dispRity` package also allows users to simulate discrete morphological data matrices.
-In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
-The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
-To understand these methods please refer to the phylogenetics literature.
-<!-- NC: I wonder if you need to cite some papers here to explain how people can find out what a suitable model is etc? Or do we just assume everyone knows about making phylogenies? -->
-
-```{r}
-set.seed(3)
-## Simulating a starting tree with 15 taxa as a random coalescent tree
-my_tree <- rcoal(15)
-
-## Generating a matrix with 100 characters (85% binary and 15% three state) and
-## an equal rates model with a gamma rate distribution (0.5, 1) with no 
-## invariant characters.
-my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
-    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)
-
-## The first few lines of the matrix
-my_matrix[1:5, 1:10]
-
-## Checking the matrix properties with a quick Maximum Parsimony tree search
-check.morpho(my_matrix, my_tree)
-```
-
-Note that this example produces a tree with a great consistency index and an identical topology to the random coalescent tree!
-Nearly too good to be true...
-
-### A more detailed descriptions
-
-`sim.morpho` is really flexible and takes many different arguments to allow users to simulate realistic matrices.
-It has three implemented models: `"ER"` for Equal Rates (the M*k* model); `"HKY"`, the molecular HKY model but transforms pyrines in 0's and pyrimidines in 1's; or the `"mixed"` model that randomly uses an `"ER"` and/or `"HKY"` for the binary characters and `"ER"` for the multistate (>2) characters.
-Both models take specific distributions for their rate or substitution models.
-These distributions should be passed to these arguments in the format of `c(sampler_function, distribution_parameters)` where the the sampler function is a the random generation function of that distribution (e.g.
-`rnorm`, `runif`, etc...) and the parameters are any parameters to be passed to this function.
-
-`check.morpho` runs a quick Maximum Parsimony tree search using the `phangorn` parsimony algorithm.
-It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g.
-the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree to determine how different they are.
-
-### Parameters for a realistic(ish) matrix
-There are many parameters that can create a ``realistic'' matrix (i.e.
-not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all end up creating "good" matrices.
-The following parameters however, seem to generate fairly ``realistic'' matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
-<!-- NC: Check that I haven't changed what you mean here where i swapped good for realistic. Realistic is a better term. Or be clearer what you mena by "good" -->
-
-```{r}
-set.seed(0)
-## tree
-my_tree <- rcoal(15)
-## matrix
-morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
-    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
-check.morpho(morpho_mat, my_tree)
-```
-
-### `space.maker`
-Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
-This function allows users to simulate multidimensional spaces with a certain number of properties.
-It takes as arguments the number of elements (data points) and dimensions to create the space but also permits more fine tuning in the data simulation: it is possible to give a specific distribution to each dimension, provide a correlation matrix to link the dimensions or even provide an *a priori* distribution of the variance for each distributions!
-<!-- ooh fancy! -->
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(mfrow = (c(2, 2)), bty = "n")
-## Visualising 3D spaces
-require(scatterplot3d)
-
-## A cube space
-scatterplot3d(space.maker(2500, 3, runif), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Uniform cube space")
-
-## A plane space
-scatterplot3d(space.maker(2500, 3, c(runif, runif, runif),
-              arguments = list(list(min = 0, max = 0), NULL, NULL)), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Uniform plane space")
-
-## An ellipsoid space (=a spheric space with correlation)
-cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3)
-scatterplot3d(space.maker(2500, 3, rnorm, cor.matrix = cor_matrix), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Normal ellipsoid space")
-
-## A cylindrical space with a decreasing variance per axis
-scatterplot3d(space.maker(2500, 3, c(rnorm, rnorm, runif),
-              scree = c(0.7, 0.2, 0.1)), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Normal cylindrical space")
-
-## Resetting the graphic parameters
-par(op)
-```
-
-# Glossary <a name="glossary"></a>
+# Glossary
 
@@
     For example, if our ordinated space is composed of birds and mammals (the elements) and 50 principal components (the dimensions), we can create two subsamples containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
-
-# The package's guts
-
-## Manipulating `dispRity` objects
-
-Disparity analysis involves a lot of manipulation of many matrices (especially when bootstrapping) which can be impractical to visualise and will quickly overwhelm your `R` console.
-Even the simple Beck and Lee 2014 example above produces an object with > 72 lines of lists of lists of matrices!
-
-Therefore `dispRity` uses a specific class of object called a `dispRity` object.
-These objects allow users to use S3 method functions such as `summary.dispRity`, `plot.dispRity` and `print.dispRity`. 
-`dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g. `sort.dispRity`, `extract.dispRity` see the full list in the next section).
-These functions modify the `dispRity` object without having to delve into its complex structure!
-The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).
-
-```{r}
-## Loading the example data
-data(disparity)
-
-## What is the class of the median_centroids object?
-class(disparity)
-
-## What does the object contain?
-names(disparity)
-
-## Summarising it using the S3 method print.dispRity
-disparity
-```
-
-Note that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:
-
-```{r, eval=FALSE}
-## Display the full object
-print(disparity, all = TRUE)
-## This is more nearly ~ 5000 lines on my 13 inch laptop screen!
-```
-
-## `dispRity` utilities
-
-The package also provides some utilities functions to facilitate multidimensional analysis.
-
-### `dispRity` object utilities  <a name="dispRity.utilities"></a>
-The first subsamples of utilities are functions for manipulating `dispRity` objects:
-
-#### `make.dispRity`
-This function creates empty `dispRity` objects.
-
-```{r}
-## Creating an empty dispRity object
-make.dispRity()
-
-## Creating an "empty" dispRity object with a matrix
-(disparity_obj <- make.dispRity(matrix(rnorm(20), 5, 4)))
-```
-
-####  `fill.dispRity`
-This function initialises a `dispRity` object and generating it's call properties.
-
-```{r}
-## The dispRity object's call is indeed empty
-disparity_obj$call
-
-## Filling an empty disparity object (that needs to contain at least a matrix)
-(disparity_obj <- fill.dispRity(disparity_obj))
-
-## The dipRity object has now the proper minimal attributes
-disparity_obj$call
-```
-
-####  `matrix.dispRity`
-This function extracts a specific matrix from a disparity object. The matrix can be one of the bootstrapped or/and rarefied one.
-
-```{r}
-## Extracting the matrix containing the coordinates of the elements at time 50
-str(matrix.dispRity(disparity, "50"))
-
-## Extracting the 3rd boostrapped matrix with the 2nd rarefaction level
-## (15 elements) from the second group (80 Mya)
-str(matrix.dispRity(disparity, subsamples = 1, bootstrap = 3, rarefaction = 2))
-```
-
-#### `get.subsamples.dispRity`
-This function creates a dispRity object that contains only elements from one specific subsamples.
-
-```{r, eval=FALSE}
-## Extracting all the data for the crown mammals
-(crown_mammals <- get.subsamples.dispRity(disp_crown_stemBS, "Group.crown"))
-
-## The object keeps the properties of the parent object but is composed of only one subsamples
-length(crown_mammals$subsamples)
-```
-
-#### `extract.dispRity`
-This function extracts the calculated disparity values of a specific matrix.
-
-```{r, eval=FALSE}
-## Extracting the observed disparity (default)
-extract.dispRity(disparity)
-
-## Extracting the disparity from the bootstrapped values from the
-## 10th rarefaction level from the second subsamples (80 Mya)
-extract.dispRity(disparity, observed = FALSE, subsamples = 2, rarefaction = 10)
-```
-
-#### `scale.dispRity`
-This is the S3 method of `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object.
-
-```{r, eval=FALSE}
-## Getting the disparity values of the time subsamples
-head(summary(disparity))
-
-## Scaling the same disparity values
-head(summary(scale(disparity, scale = TRUE)))
-
-## Scaling and centering:
-head(summary(scale(disparity, scale = TRUE, center = TRUE)))
-```
-
-#### `sort.dispRity`
-This is the S3 method of `sort` for sorting the subsamples alphabetically (default) or following a specific pattern.
-
-```{r, eval=FALSE}
-## Sorting the disparity subsamples in inverse alphabetic order
-head(summary(sort(disparity, decreasing = TRUE)))
-
-## Customised sorting
-head(summary(sort(disparity, sort = c(7, 1, 3, 4, 5, 2, 6))))
-```
-
-<!-- ## Modularity
-
-### Making your own metrics
-
-### Making your own tests
-
-## Where is this going?
- -->
-<!-- ## Running `dispRity` in parallel
-
-The computationally intensive function `dispRity` have a `parallel` option to speed up their calculations.
-
-This option requires the package `snow` and takes arguments that are to be passed to  -->

### 2017-08-15 inst/gitbook/02_getting-started.Rmd
moved manuals to inst/

--- 2017-08-12 gitbook/02_getting-started.Rmd
@@
 they incorporate lots of other functions) that allow users to run a basic disparity-through-time, or disparity among groups, analysis without too much effort.
-As such they have a lot of defaults. 
+As such they use a lot of default options. 
 These are described in the help files for the functions that are used to make the wrapper functions, and not described in the help files for `disparity.through.time` and `disparity.per.group`. 
-These defaults are perfect for **data exploration**, but for a proper analysis you should consider the **best parameters for your question and data**. 
-For example, which metric should you use? How many bootstraps do you require? What model of evolution is most appropriate if you are time slicing? Should you rarefy the data?
-See `custom.subsamples`, `boot.matrix`, `dispRity.metric`, \code{\link{summary.dispRity}}, `plot.dispRity` for more details of the defaults used in each of these functions.
-Note that any of these defaults can be changed within the `disparity.through.time` or `disparity.per.group` functions.
-<!-- NC: Add links to these sections of the manual -->
+These defaults are good enough for **data exploration**, but for a proper analysis you should consider the **best parameters for your question and data**. 
+For example, which metric should you use?
+How many bootstraps do you require?
+What model of evolution is most appropriate if you are time slicing?
+Should you rarefy the data?
+See [`time.subsamples`](#time-slicing), [`custom.subsamples`](#customised-subsamples), [`boot.matrix`](#bootstraps-and-rarefactions) and [`dispRity.metric`](#disparity-metrics) for more details of the defaults used in each of these functions.
+Note that any of these default arguments can be changed within the `disparity.through.time` or `disparity.per.group` functions.
 

### 2017-08-15 inst/gitbook/03_specific-tutorials.Rmd
moved manuals to inst/

--- 2017-08-12 gitbook/03_specific-tutorials.Rmd
@@
 These results show the actual overlap among all the measured distances from centroids concatenated across all the bootstraps.
-For example, when comparing the slices 120 and 80, we are effectively comparing the 5 \times 100 distances (the distances of the five elements in slice 120 bootstrapped 100 times) to the 19 \times 100 distances from slice 80.
-However, this can also be problematic for some specific tests since the _n_ \times 100 distances are also pseudo-replicates and thus are still not independent.
+For example, when comparing the slices 120 and 80, we are effectively comparing the 5 $\times$ 100 distances (the distances of the five elements in slice 120 bootstrapped 100 times) to the 19 $\times$ 100 distances from slice 80.
+However, this can also be problematic for some specific tests since the _n_ $\times$ 100 distances are also pseudo-replicates and thus are still not independent.
 

### 2017-08-15 inst/gitbook/04_guts.Rmd
moved manuals to inst/

--- 2017-08-10 doc/dispRity-manual.Rmd
@@
 ---
-title: "dispRity manual"
-author: "Thomas Guillerme"
+title: "The guts of the dispRity package"
+author: "Thomas Guillerme (guillert@tcd.ie)"
 date: "`r Sys.Date()`"
 output:
-    rmarkdown::html_vignette:
-        toc: true
-        toc_depth: 2
-bibliography: References.bib
-bst: sysbio.bst
-vignette: >
-  %\VignetteIndexEntry{dispRity manual}
-  %\VignetteEngine{knitr::rmarkdown}
-  \usepackage[utf8]{inputenc}
+  html_document: default
+  pdf_document: default
 ---
 
-# `dispRity`
-
-This is a package for measuring disparity in `R`.
-It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values so they can easily be compared.
-This manual is based on the version `1.0`.
-
-# Introduction
-
-## What is `dispRity`?
-This is a package for measuring disparity in `R`.
-It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
-Typically, these analysis are used in palaeobiology and evolutionary biology to study the changes in morphology through time.
-However, there are many more applications in ecology, evolution and beyond.
-
-## Installing and running the package
-
-You can install this package easily if you use the latest version of `R` (> 3.3.0) and `devtools`.
-
-```{r, eval=FALSE}
-## Checking if devtools is already installed
-if(!require(devtools)) install.packages("devtools")
-
-## Installing the latest released version directly from GitHub
-install_github("TGuillerme/dispRity", ref = "release")
-
-## Loading the package
-library(dispRity)
-```
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
-# NC: this isn't checking Claddis...
-library(dispRity)
-## Setting the random seed for repeatability
-set.seed(123)
-```
-
-Note this uses the `release` branch (version`0.4`). <!-- @@@ -->
-For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
-`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, `Claddis`, `geomorph` and `RCurl`).
-
-## Why not CRAN?
-
-This package is not available on CRAN.
-This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
-However, the package follows the strict (and useful!) CRAN standards via Travis.
-
-[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
-[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
-[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
-
-## Help
-
-If you need help with the package, hopefully the following manual will be useful.
-However, parts of this package are still in development and some other parts are probably not covered.
-Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
-
-## Citations
-
-You can cite both the package or this manual with the following citation:
-
-> Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646
-
-<!-- TG: I'm gonna change that to a newest version with the submission -->
-
-Note that this citation is only temporary (but can still be used!).
-A future proper version of the latest package release, this manual and an associated methods paper should be submitted soon(ish!).
-
-# Getting started
-
-## What sort of data does dispRity work with?
-
-Disparity can be estimated from pretty much any matrix.
-Classically, however, it is measured from ordinated matrices.
-These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).
-
-### Ordination matrices from `Claddis`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
-if(!require(Claddis)) install.packages("Claddis")
-```
-
-`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
-For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:
-
-```{r}
-require(Claddis)
-
-## Ordinating the example data from Claddis
-Claddis.ordination(Michaux1989) 
-```
-
-Note that several options are available, namely which type of distance should be computed.
-See more info in the function manual (`?Claddis.ordination`).
-Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.
-
-### Ordination matrices from `geomorph`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if geomorph is installed BUILD ONLY
-if(!require(geomorph)) install.packages("geomorph")
-```
-
-You can also easily use data from `geomorph` using the `geomorph.ordination` function.
-This function simply takes Procrustes aligned data and performs an ordination:
-
-```{r}
-require(geomorph)
-
-## Loading the plethodon dataset
-data(plethodon)
-
-## Performing a Procrustes transform on the landmarks
-procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)
-
-## Ordinating this data
-geomorph.ordination(procrustes)[1:5,1:5]
-```
-
-Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
-Additionally you can give the function a `geomorph.data.frame` object.
-If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsamples)!
-
-```{r}
-## Using a geomorph.data.frame
-geomorph_df <- geomorph.data.frame(procrustes,
-     species = plethodon$species, site = plethodon$site)
-
-## Ordinating this data and making a dispRity object
-geomorph.ordination(geomorph_df)
-```
-
-More about these `dispRity` objects below!
-
-### Other kinds of ordination matrices
-
-If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordination matrices by using the following functions from the `stats` package.
-Here is how to do it for the following types of matrices:
-
- * Multivariate matrices (principal components analysis; PCA)
-
-```{r}
-## A multivariate matrix
-head(USArrests)
-
-## Ordinating the matrix using `prcomp` 
-ordination <- prcomp(USArrests)
-
-## Selecting the ordinated matrix
-ordinated_matrix <- ordination$x
-head(ordinated_matrix)
-```
-  This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, see the `?princomp` function.
-
- * Distance matrices (classical multidimensional scaling; MDS)
-
-```{r}
-## A matrix of distances between cities
-str(eurodist)
-
-## Ordinating the matrix using cmdscale() with k = 5 dimensions 
-ordinated_matrix <- cmdscale(eurodist, k = 5)
-head(ordinated_matrix)
-```
-
-  This results in a ordinated matrix with European cities as elements and five dimensions.
-
-Of course any other method for creating the ordination matrix is totally valid, you can also not use any ordination at all!
-The only requirements for the `dispRity` functions is that the input is a matrix with elements as rows and dimensions as columns.
-
-## Performing a simple dispRity analysis
-
-Two `dispRity` functions allow users to run an analysis pipeline simply by inputting an ordination matrix.
-These functions allow users to either calculate the disparity through time (`dispRity.through.time`) or the disparity of user-defined groups (`dispRity.per.group`).
-
-### Example data
-
-To illustrate these functions, we will use data from @beckancient2014.
-This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
-
-```{r, fig.width=7, fig.height=7}
-## Loading the ordinated matrices
-data(BeckLee_mat50)
-data(BeckLee_mat99)
-
-## The first five taxa and dimensions of the 50 taxa matrix
-head(BeckLee_mat50[, 1:5])
-
-## The first five taxa and dimensions of the 99 taxa + ancestors matrix
-BeckLee_mat99[c(1, 2, 98, 99), 1:5]
-
-## Loading a list of first and last occurrence dates for the fossils
-data(BeckLee_ages)
-head(BeckLee_ages)
-
-## Loading and plotting the phylogeny
-data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.8) 
-axisPhylo(root = 140)
-nodelabels(cex = 0.5)
-```
-
-Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).
-
-### Disparity through time
-
-The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
-This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
-Of course all the defaults can be changed if required, more on this later.
-
-For a disparity through time analysis, you will need:
-
-  * An ordinated matrix (we covered that above)
-  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element.
-To give your tree a root time (i.e.
-an age for the root), you can simply do `my_tree$root.time <- my_age`.
-  * The required number of time subsamples (here `time = 3`)
-  * Your favourite disparity metric (here the sum of variances)
-
-Using the @beckancient2014 data described [above](#example-data):
-
-```{r}
-## Measuring disparity through time
-disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
-                                        time = 3, metric = c(sum, variances))
-```
-
-This generates a `dispRity` object (see [here](#manipulating-dispRity-objects) for technical details).
-When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:
-
-```{r}
-## Print the disparity_data object
-disparity_data
-```
-
-We asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
-
-We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):
-
-```{r, fig.width=7, fig.height=7}
-## Summarising disparity through time
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data, type = "continuous")
-
-## Testing for an difference among the time bins
-disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
-summary(disp_lm)
-
-## Note that ANOVA assumptions were not checked here!
-```
-
-Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
-You can also directly explore the specific function help files within R and navigate to related functions.
-
-### Disparity among groups
-
-The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
-For example, you could ask if there is a difference in disparity between two groups?
-
-To perform such an analysis, you will need:
-
- * An matrix with rows as elements and columns as dimensions (always!)
- * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix.
-For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4.
-Note that elements can be present in multiple groups at once.
- * Your favourite disparity metric (here the sum of variances)
-
-Using the @beckancient2014 data described [above](#example-data):
-
-```{r}
-## Creating the two groups (crown versus stem) as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44))
-
-## Measuring disparity for each group
-disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
-                                     metric = c(sum, variances))
-```
-
-We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).
-
-```{r, fig.width=7, fig.height=7}
-## Print the disparity_data object
-disparity_data
-
-## Summarising disparity in the different groups
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data)
-
-## Testing for a difference between the groups
-test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
-```
-
-# Specific tutorials
-
-The following section contains information specific to some functions.
-If any of your questions are not covered in these sections, please refer to the function help files in R, send me an email (<guillert@tcd.ie>), or raise an issue on GitHub.
-The several tutorials below describe specific functionalities of certain functions; please always refer to the function help files for the full function documentation!
-
-## Time slicing
-
-The function `time.subsamples` allows users to divide the matrix into different time subsamples or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
-Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.
-
-Two types of time subsamples can be performed by using the `method` option:
-
- *  Discrete time subsamples (or time-binning) using `method = discrete`
- *  Continuous time subsamples (or time-slicing) using `method = continuous`
-
-For the time-slicing method details see @GuillermeSTD. <!-- @@@ Change cite appropriately! -->
-<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
-For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:
-
- *  Defining the boundaries of the time bins (when `method = discrete`)
- *  Defining the time slices (when `method = continuous`)
-
-Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
-Additionally, it is also possible to input a dataframe containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age, so taxa can appear in more than one time bin/slice.
-
-Here is an example for `method = discrete`:
-
-```{r, eval=TRUE}
-## Generating three time bins containing the taxa present every 40 Ma
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0))
-```
-
-Note that we can also generate equivalent results by just telling the function that we want three time-bins as follow:
-
-```{r, eval=TRUE}
-## Automatically generate three equal length bins:
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = 3)
-```
-
-In this example, the taxa were split inside each time-bin according to their age.
-However, the taxa here are considered as single points in time.
-It is totally possible that some taxa could have had longer longevity and that they exist in multiple time bins.
-In this case, it is possible to include them in more than one bin by providing a table of first and last occurrence dates (FAD/LAD).
-This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:
-
-```{r, eval=TRUE}
-## Displaying the table of first and last occurrence dates for each taxa
-head(BeckLee_ages)
-
-## Generating time bins including taxa that might span between them
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
-```
-
-When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as the root age plus 1\% of the tree length, as long as at least three elements/taxa are present at that point in time.
-The algorithm adds an extra 1\% tree length until reaching the required minimum of three elements.
-It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance among tips *and* nodes.
-
-For the time-slicing method (`method = continuous`), the idea is fairly similar.
-This option, however, requires a matrix that contains the ordinated distance among taxa *and* nodes and an extra argument describing the assumed evolutionary model (via the `model` argument).
-This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node, meaning that a decision must be made about what the value for the branch should be.
-The model can be one of the following:
-
- *  `acctran` where the data chosen along the branch is always the one of the descendant
- *  `deltran` where the data chosen along the branch is always the one of the ancestor
- *  `punctuated` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
- *  `gradual` where the data chosen along the branch is either the descendant or the ancestor depending on branch length
-
-```{r, eval=TRUE}
-## Generating four time slices every 40 million years under a model of gradual evolution
-time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-
-## Generating four time slices automatically
-time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree,
-    method = "continuous", model = "gradual", time = 4, FADLAD = BeckLee_ages)
-```
-
-## Customised subsamples
-
-Another way of separating elements into different categories is to use customised subsamples as briefly explained [above](#disparity-among-groups).
-This function simply takes the list of elements to put in each group (whether they are the actual element names or their position in the matrix).
-
-```{r, eval=TRUE}
-## Creating the two groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44))
-
-## Separating the dataset into two different groups
-custom.subsamples(BeckLee_mat50, group = mammal_groups)
-```
-
-Elements can easily be assigned to different groups if necessary!
-
-```{r, eval=FALSE}
-## Creating the three groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44).
-                      "all" = c(1:50))
-```
-
-## Bootstraps and rarefactions
-
-One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results are, and how sensitive they are to outliers in the dataset.
-This can be achieved using the function `boot.matrix` to bootstrap and/or rarefy the data.
-The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):
-
-```{r, eval=TRUE}
-## Default bootstrapping
-boot.matrix(data = BeckLee_mat50)
-```
-
-The number of bootstrap replicates can be defined using the `bootstraps` option.
-The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
-Currently two algorithms are implemented:
-
- * `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any *m* elements drawn from the data)
- * `single` where only one random element is replaced by one other random element for each pseudo-replicate
-
-```{r, eval=TRUE}
-## Bootstrapping with the single bootstrap method
-boot.matrix(BeckLee_mat50, boot.type = "single")
-```
-
-This function also allows users to rarefy the data using the `rarefaction` argument.
-Rarefaction allows users to limit the number of elements to be drawn at each bootstrap replication.
-This is useful if, for example, one is interested in looking at the effect of reducing the number of elements on the results of an analysis.
-
-This can be achieved by using the `rarefaction` option that draws only *n-x* at each bootstrap replicate (where *x* is the number of elements not sampled).
-The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsample to a minimum of three elements).
-It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
-
-```{r, eval=TRUE}
-## Bootstrapping with the full rarefaction
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
-
-## Or with a set number of rarefaction levels
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8, 3))
-```
-
-One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
-When missing, all dimensions from the ordinated matrix are used.
-
-```{r, eval=TRUE}
-## Using the first 50% of the dimensions
-boot.matrix(BeckLee_mat50, dimensions = 0.5)
-
-## Using the first 10 dimensions
-boot.matrix(BeckLee_mat50, dimensions = 10)
-```
-
-Of course, one could directly supply the subsamples generated above (using `time.subsamples` or `custom.subsamples`) to this function.
-
-```{r, eval=TRUE}
-## Creating subsamples of crown and stem mammals
-crown_stem <- custom.subsamples(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot.matrix(crown_stem, bootstraps = 200, rarefaction = TRUE)
-
-## Creating time slice subsamples
-time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-                               method = "continuous", model = "gradual", 
-                               time = c(120, 80, 40, 0),
-                               FADLAD = BeckLee_ages)
-
-## Bootstrapping the time slice subsamples
-boot.matrix(time_slices, bootstraps = 100)
-```
-
-## Disparity metrics
-
-There are many ways of measuring disparity!
-In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
-For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @Wills1994.
-Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.
-
-Because of this great diversity of metrics, the package `dispRity` does not have one way to measure disparity but rather proposes to facilitate users in defining their own disparity metric that will best suit their particular analysis.
-In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
-However the `metric` argument has to follow certain rules:
-
-1.  It must be composed from one to three `function` objects;
-2.  The function(s) must take as a first argument a `matrix` or a `vector`;
-3.  The function(s) must be of one of the three dimension-levels described below;
-4.  At least one of the functions must be of dimension-level 1 or 2 (see below).
-
-### The function dimension-levels
-
-The metric function dimension-levels determine the "dimensionality of decomposition" of the input matrix.
-In other words, each dimension-level designates the dimensions of the output, i.e. either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimension.
-
-![Illustration of the different dimension-levels of functions with an input `matrix`](dispRity_fun.png)
-
-#### Dimension-level 1 functions
-
-A dimension-level 1 function will decompose a `matrix` or a `vector` into a single value:
-
-```{r}
-## Creating a dummy matrix
-dummy_matrix <- matrix(rnorm(12), 4, 3)
-
-## Example of dimension-level 1 functions
-mean(dummy_matrix)
-median(dummy_matrix)
-```
-
-Any summary metric such as mean or median are good examples of dimension-level 1 functions as they reduce the matrix to a single dimension (i.e. one value).
-
-#### Dimension-level 2 functions
-
-A dimension-level 2 function will decompose a `matrix` into a `vector`.
-
-```{r}
-## Defining the function as the product of rows
-prod.rows <- function(matrix) apply(matrix, 1, prod)
-
-## A dimension-level 2 metric
-prod.rows(dummy_matrix)
-```
-
-Several dimension-level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculate the variance or the range of each dimension of the ordinated matrix respectively.
-
-#### Dimension-level 3 functions
-
-Finally a dimension-level 3 function will transform the matrix into another matrix.
-Note that the dimension of the output matrix doesn't need to match the the input matrix:
-
-```{r}
-## A dimension-level 3 metric
-var(dummy_matrix)
-
-## A dimension-level 3 metric with a forced matrix output
-as.matrix(dist(dummy_matrix))
-```
-
-### `make.metric`
-
-Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the Euclidean distance between each element and the centroid of the ordinated space.
-The `make.metric` function implemented in `dispRity` is designed to help test and find the dimension-level of the functions.
-This function tests:
-
-1.  If your function can deal with a `matrix` or a `vector` as an input;
-2.  Your function's dimension-level according to its output (dimension-level 1, 2 or 3, see above);
-3.  Whether the function can be implemented in the `dispRity` function (the function is fed into a `lapply` loop).
-
-For example, let's see if the functions described above are the right dimension-levels:
-
-```{r}
-## Which dimension-level is the mean function? And can it be used in dispRity?
-make.metric(mean)
-
-## Which dimension-level is the prod.rows function? And can it be used in dispRity?
-make.metric(prod.rows)
-
-## Which dimension-level is the var function? And can it be used in dispRity?
-make.metric(var)
-```
-
-A non verbose version of the function is also available.
-This can be done using the option `silent = TRUE` and will simply output the dimension-level of the metric.
-
-```{r}
-## Testing whether mean is dimension-level 1
-if(make.metric(mean, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-## Testing whether var is dimension-level 1
-if(make.metric(var, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-```
-
-### Metrics in the `dispRity` function
-
-Using this metric structure, we can easily use any disparity metric in the `dispRity` function as follows:
-
-```{r}
-## Measuring disparity as the standard deviation of all the values of the
-## ordinated matrix (dimension-level 1 function).
-summary(dispRity(BeckLee_mat50, metric = sd))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the ordinated matrix (dimension-level 1 and 2 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the variance covariance matrix (dimension-level 1, 2 and 3 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
-```
-
-Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > 2 > 1).
-
-```{r}
-## Disparity as the standard deviation of the variance of each axis of the
-## variance covariance matrix:
-disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)),
-                      round = 10)
-
-## Same as above but using a different function order for the metric argument
-disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
-                      round = 10)
-
-## Both ways output the same disparity values:
-disparity1 == disparity2
-```
-
-In these examples, we considered disparity to be a single value.
-For example, in the previous example, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
-It is, however, possible to calculate [disparity as a distribution](#disparity-as-a-distribution).
-
-### Metrics implemented in `dispRity`
-
-Several disparity metrics are implemented in the `dispRity` package.
-The detailed list can be found in `?dispRity.metric` along with some description of each metric.
-
-Level | Name | Description | Source | 
-------|------|---------------------------------------------------|--------|
-3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
-3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
-3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
-2 | `ranges` | The range of each dimension | `dispRity` |
-2 | `variances` | The variance of each dimension | `dispRity` |
-2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
-1 | `mode.val` | The modal value | `dispRity` |
-
-1: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (not PCA).
-
-2: Note that by default, the centroid is the centroid of the elements.
-It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
-
-### Equations and implementations
-Some of the functions described below are implemented in the `dispRity` package and do not require any other packages to calculate ([see implementation here](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R)).
-
-\begin{equation}
-    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
-\end{equation}
-Where *k* is the number of dimensions, and $\lambda_i$ is the eigenvalue of each dimension.
-
-\begin{equation}
-    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
-\end{equation}
-Where *k* is the number of dimensions.
-
-\begin{equation}
-    ranges = |max(k_i) - min(k_i)|
-\end{equation}
-Where *k* is the number of dimensions.
-
-\begin{equation}
-    variances = \sigma^{2}{k_i}
-\end{equation}
-Where *k* is the number of dimensions, and $\sigma^{2}$ is their variance.
-
-\begin{equation}
-    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
-\end{equation}
-Where *n* is each element in the ordinated space, *k* is the number of dimensions, and $Centroid_{k}$ is their mean (or can be set to another value).
-
-### Using the different disparity metrics
-Here is a brief demonstration of the main metrics implemented in `dispRity`.
-First, we will create a dummy/simulated ordinated space using the `space.maker` utility function (more about that [here](#space.maker):
-
-```{r}
-## Creating a 10*5 normal space
-set.seed(1)
-dummy_space <- space.maker(10, 5, rnorm)
-```
-
-We will use this simulated space to demonstrate the different metrics.
-
-#### Volumes and surface metrics
-The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `hyper.volume` all measure the surface or the volume of the ordinated space occupied:
-
-```{r}
-## Calculating the ellipsoid volume
-summary(dispRity(dummy_space, metric = ellipse.volume))
-```
-> Because there is only one subsample (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with rounding).
-
-```{r}
-## Calculating the convex hull surface
-summary(dispRity(dummy_space, metric = convhull.surface))
-
-## Calculating the convex hull volume
-summary(dispRity(dummy_space, metric = convhull.volume))
-```
-
-The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
-For the full and correct calculation of the volume of the ordinated space, it is possible to use the `hyper.volume` function that has more options but takes longer to calculate.
-
-```{r}
-## Calculating the true multidimensional volume
-summary(dispRity(dummy_space, metric = hyper.volume))
-
-```
-> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) that often results in near 0 disparity values.
-
-#### Ranges, variances and diagonal
-
-The functions `ranges`, `variances` and `diagonal` all measure properties of the ordinated space based on its dimensional properties (they are also less affected by the "curse of dimensionality"):
-
-`ranges` and `variances` both work on the same principle and measure the range/variance of each dimension:
-
-```{r}
-## Calculating the ranges of each dimension in the ordinated space
-ranges(dummy_space)
-
-## Calculating disparity as the distribution of these ranges
-summary(dispRity(dummy_space, metric = ranges))
-
-## Calculating disparity as the sum and the product of these ranges
-summary(dispRity(dummy_space, metric = c(sum, ranges)))
-summary(dispRity(dummy_space, metric = c(prod, ranges)))
-
-## Calculating the variances of each dimension in the ordinated space
-variances(dummy_space)
-
-## Calculating disparity as the distribution of these variances
-summary(dispRity(dummy_space, metric = variances))
-
-## Calculating disparity as the sum and the product of these variances
-summary(dispRity(dummy_space, metric = c(sum, variances)))
-summary(dispRity(dummy_space, metric = c(prod, variances)))
-```
-
-The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest Euclidean distance in our five dimensional space):
-
-```{r}
-## Calculating the ordinated space's diagonal
-summary(dispRity(dummy_space, metric = diagonal))
-```
-
-> This metric is only a Euclidean diagonal (mathematically valid) if the dimensions within the space are all orthogonal!
-
-#### Centroids metric
-
-The `centroids` metric allows users to measure the position of the different elements compared to a fixed point in the ordinated space.
-By default, this function measures the distance between each element and their centroid (centre point):
-
-```{r}
-## The distribution of the distances between each element and their centroid
-summary(dispRity(dummy_space, metric = centroids))
-
-## Disparity as the median value of these distances
-summary(dispRity(dummy_space, metric = c(median, centroids)))
-```
-
-It is however possible to fix the coordinates of the centroid to a specific point in the ordinated space, as long as it has the correct number of dimensions:
-
-```{r}
-## The distance between each element and the origin of the ordinated space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))
-
-## Disparity as the distance between each element and a specific point in space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
-```
-
-## Summarising dispRity data (plots)
-
-Because of its architecture, printing `dispRity` objects only summarises their content but does not print the disparity value measured or associated analysis (more about this [here](#manipulating-dispRity-objects)).
-To actually see what is in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphical representation of the results.
-
-### Summarising `dispRity` data
-This function is an S3 function (`summary.dispRity`) allowing users to summarise the content of `dispRity` objects that contain disparity calculations.
-
-<!-- DATA LOADER -->
-
-```{r}
-## Example data from previous sections
-crown_stem <- custom.subsamples(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
-## Calculate disparity
-disparity_crown_stem <- dispRity(boot_crown_stem, metric = c(sum, variances))
-
-## Creating time slice subsamples
-time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-## Bootstrapping the time slice subsamples
-boot_time_slices <- boot.matrix(time_slices, bootstraps = 100)
-## Calculate disparity
-disparity_time_slices <- dispRity(boot_time_slices, metric = c(sum, variances))
-
-## Creating time bin subsamples
-time_bins <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages,
-    inc.nodes = TRUE)
-## Bootstrapping the time bin subsamples
-boot_time_bins <- boot.matrix(time_bins, bootstraps = 100)
-## Calculate disparity
-disparity_time_bins <- dispRity(boot_time_bins, metric = c(sum, variances))
-```
-
-These objects are easy to summarise as follows:
-
-```{r}
-## Default summary
-summary(disparity_time_slices)
-```
-
-Information about the number of elements in each subsample and the observed (i.e. non-bootstrapped) disparity are also calculated.
-This is specifically handy when rarefying the data for example:
-
-```{r}
-head(summary(disparity_crown_stem))
-```
-
-The summary functions can also take various options such as:
-
- * `quantile` values for the confidence interval levels (by default, the 50 and 95 quantiles are calculated)
- * `cent.tend` for the central tendency to use for summarising the results (default is `median`)
- * rounding` option corresponding to the number of decimal places to print (default is `2`)
- * `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
-
-These options can easily be changed from the defaults as follows:
-
-```{r}
-## Same as above but using the 88th quantile and the standard deviation as the summary 
-summary(disparity_time_slices, quantile = 88, cent.tend = sd)
-
-## Printing the details of the object and rounding the values to the 5th decimal place
-summary(disparity_time_slices, recall = TRUE, rounding = 5)
-```
-
-Note that the summary table is a `data.frame`, hence it is as easy to modify as any dataframe using `dplyr`.
-You can also export it in `csv` format using `write.csv` or `write_csv` or even directly export into `LaTeX` format using the following;
-
-```{r, eval=FALSE}
-## Loading the xtable package
-require(xtable)
-## Converting the table in LaTeX
-xtable(summary(disparity_time_slices))
-```
-
-### Plotting `dispRity` data
-An alternative (and more fun!) way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
-This function takes the same options as `summary.dispRity` along with various graphical options described in the function help files (see `?plot.dispRity`).
-
-The plots can be of four different types:
-
- * `continuous` for displaying continuous disparity curves
- * `box`, `lines`, and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.
-
-> This argument can be left empty. In this case, the algorithm will automatically detect the type of subsamples from the `dispRity` object and plot accordingly.
-
-It is also possible to display the number of elements in each subsample (as a horizontal dotted line) using the option `elements = TRUE`.
-Additionally, when the data is rarefied, one can indicate which level of rarefaction to display (i.e. only display the results for a certain number of elements) by using the `rarefaction` argument.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(mfrow = c(2, 2), bty = "n")
-
-## Plotting continuous disparity results
-plot(disparity_time_slices, type = "continuous")
-
-## Plotting discrete disparity results
-plot(disparity_crown_stem, type = "box")
-
-## As above but using lines for the rarefaction level of 20 elements only
-plot(disparity_crown_stem, type = "line", rarefaction = 20)
-
-## As above but using polygons while also displaying the number of elements
-plot(disparity_crown_stem, type = "polygon", elements = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Since `plot.dispRity` uses the arguments from the generic `plot` method, it is of course possible to change pretty much everything using the regular plot arguments:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some classic options from plot
-plot(disparity_time_slices, col = c("blue", "orange", "green"),
-    ylab = c("Some measurement"), xlab = "Some other measurement",
-    main = "Many options...", ylim = c(5, 0), xlim = c(4, 0))
-
-## Adding a legend
-legend("topleft", legend = c("Central tendency",
-                             "Confidence interval 1",
-                            "Confidence interval 2"),
-      col = c("blue", "orange", "green"), pch = 19)
-
-## Resetting graphical parameters
-par(op)
-```
-
-In addition to the classic `plot` arguments, the function can also take arguments that are specific to `plot.dispRity` like adding the number of elements or rarefaction level (as described above), and also changing the values of the quantiles to plot as well as the central tendency.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some plot.dispRity arguments
-plot(disparity_time_slices, quantile = c(seq(from = 10, to = 100, by = 10)),
-    cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
-    ylab = c("Disparity", "Diversity"), time.subsamples = FALSE,
-    xlab = "Time (in in units from past to present)", observed = TRUE,
-    main = "Many more options...")
-
-## Resetting graphical parameters
-par(op)
-```
-
-> Note that the argument `observed = TRUE` allows to plot the disparity values calculated from the non-bootstrapped data as crosses on the plot.
-
-For comparing results, it is also possible to add a plot to the existent plot by using `add = TRUE`:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the continuous disparity with a fixed y axis
-plot(disparity_time_slices, ylim = c(1, 4))
-## Adding the discrete data
-plot(disparity_time_bins, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
-    add = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Finally, if your data has been fully rarefied, it is also possible to easily look at rarefaction curves by using the `rarefaction = TRUE` argument:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the rarefaction curves
-plot(disparity_crown_stem, rarefaction = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-## Testing disparity hypotheses
-
-The `dispRity` package allows users to apply statistical tests to the calculated disparity to test various hypotheses.
-The function `test.dispRity` works in a similar way to the `dispRity` function: it takes a `dispRity` object, a `test` and a `comparisons` argument.
-
-The `comparisons` argument indicates the way the test should be applied to the data:
-
- * `pairwise` (default): to compare each subsample in a pairwise manner
- * `referential`: to compare each subsample to the first subsample
- * `sequential`: to compare each subsample to the following subsample
- * `all`: to compare all the subsamples together (like in analysis of variance)
-
-It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsample names to create personalised tests.
-Some other tests implemented in `dispRity` such as the `dispRity::null.test` have a specific way they are applied to the data and therefore ignore the `comparisons` argument. 
-<!-- Add sequential test one day! -->
-
-The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
-It can be a common statistical test function (e.g. `stats::t.test`), a function implemented in `dispRity` (e.g. see `?null.test`) or any function defined by the user.
-
-This function also allows users to correct for Type I error inflation (false positives) when using multiple comparisons via the `correction` argument. 
-This argument can be empty (no correction applied) or can contain one of the corrections from the `stats::p.adjust` function (see `?p.adjust`).
-
-Note that the `test.dispRity` algorithm deals with some classical test outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
-It is, however, possible to get the full detailed output by using the options `details = TRUE`.
-
-```{r}
-## T-test to test for a difference in disparity between crown and stem mammals
-test.dispRity(disparity_crown_stem, test = t.test)
-
-## Performing the same test but with the detailed t.test output
-test.dispRity(disparity_crown_stem, test = t.test, details = TRUE)
-
-## Wilcoxon test applied to time sliced disparity with sequential comparisons, with Bonferroni correction
-test.dispRity(disparity_time_slices, test = wilcox.test,
-              comparisons = "sequential", correction = "bonferroni")
-
-## Measuring the overlap between distributions in the time bins (using the
-## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)
-test.dispRity(disparity_time_bins, test = bhatt.coeff)
-```
-
-It is also possible to apply some more *complex* tests that have their own output classes (like `stats::lm`).
-
-The results can then be analysed as usual using the associated `summary` S3 method:
-
-```{r}
-## Performing and linear model applied to the same data
-(slice_lm <- test.dispRity(disparity_time_slices, test = lm,
-                            comparisons = "all"))
-
-## The output is a regular `lm` output
-class(slice_lm)
-
-## This output can be summarised using summary
-summary(slice_lm)
-```
-
-Of course, due to the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
-The only condition is that the test can be applied to at least two distributions.
-In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm` for example).
-
-## Disparity as a distribution
-
-Disparity is often regarded as a summary value of the position of the all elements in the ordinated space.
-For example, the sum of variances, the product of ranges or the median distance between the elements and their centroid will summarise disparity as a single value.
-This value can be pseudo-replicated (bootstrapped) to obtain a distribution of the summary metric with estimated error.
-However, another way to perform disparity analysis is to use the *whole distribution* rather than just a summary metric (e.g. the variances or the ranges).
-
-This is possible in the `dispRity` package by calculating disparity as a dimension-level 2 metric only!
-Let's have a look using our previous example of bootstrapped time slices but by measuring the distances between each taxon and their centroid as disparity.
-
-```{r}
-## Measuring disparity as a whole distribution
-disparity_centroids <- dispRity(boot_time_slices, metric = centroids)
-```
-
-The resulting disparity object is of dimension-level 2, so it can easily be transformed into a dimension-level 1 object by, for example, measuring the median distance of all these distributions:
-
-```{r}
-## Measuring median disparity in each time slice
-disparity_centroids_median <- dispRity(disparity_centroids, metric = median)
-```
-
-And we can now compare the differences between these methods:
-
-```{r}
-## Summarising both disparity measurements:
-## The distributions:
-summary(disparity_centroids)
-## The summary of the distributions (as median)
-summary(disparity_centroids_median)
-```
-
-We can see that the summary message for the distribution is slightly different than before.
-Here `summary` also displays the observed central tendency (i.e. the central tendency of the measured distributions).
-Note that, as expected, this central tendency is the same in both metrics!
-
-Another, maybe more intuitive way, to compare both approaches for measuring disparity is to plot the distributions:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(bty = "n", mfrow = c(1, 2))
-
-## Plotting both disparity measurements
-plot(disparity_centroids, ylab = "Distribution of all the distances")
-plot(disparity_centroids_median,
-     ylab = "Distribution of the medians of all the distances")
-
-par(op)
-```
-
-We can then test for differences in the resulting distributions using `test.dispRity` and the `bhatt.coeff` test as described above.
-
-```{r}
-## Probability of overlap in the distribution of medians
-test.dispRity(disparity_centroids_median, test = bhatt.coeff)
-```
-
-In this case, we are looking at the probability of overlap of the distribution of median distances from centroids among each pair of time slices.
-In other words, we are measuring whether the medians from each bootstrap pseudo-replicate for each time slice overlap.
-But of course, we might be interested in the actual distribution of the distances from the centroid rather than simply their central tendencies.
-This can be problematic depending on the research question asked since we are effectively comparing non-independent medians distributions (because of the pseudo-replication).
-
-One solution, therefore, is to look at the full distribution:
-
-```{r}
-## Probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff)
-```
-
-These results show the actual overlap among all the measured distances from centroids concatenated across all the bootstraps.
-For example, when comparing the slices 120 and 80, we are effectively comparing the 5 \times 100 distances (the distances of the five elements in slice 120 bootstrapped 100 times) to the 19 \times 100 distances from slice 80.
-However, this can also be problematic for some specific tests since the _n_ \times 100 distances are also pseudo-replicates and thus are still not independent.
-
-A second solution is to compare the distributions to each other _for each replicate_:
-
-```{r}
-## Boostrapped probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff, concatenate = FALSE)
-```
-
-These results show the median overlap among pairs of distributions in the first column (`bhatt.coeff`) and then the distribution of these overlaps among each pair of bootstraps.
-In other words, when two distributions are compared, they are now compared for each bootstrap pseudo-replicate, thus effectively creating a distribution of probabilities of overlap.
-For example, when comparing the slices 120 and 80, we have a mean probability of overlap of 0.28 and a probability between 0.18 and 0.43 in 50\% of the pseudo-replicates.
-Note that the quantiles and central tendencies can be modified via the `conc.quantiles` option.
-
-## Ecological studies of disparity
-<!-- NC: I wonder if this might be better just as a "hey we can do ecology stuff too and here is a link to a demo" rather than a new section?
-The manual is loooooooooong so splitting some stuff out might be good?
--->
-
-Disparity analyses are not restricted to palaeobiology analysis.
-This package is designed for dealing with ordinated matrices, no matter which data was used in the ordination.
-Here is an short example of disparity analysis using ecological data from Deirdre McClean (unpubl.).
-
-The dataset contains an ordinated matrix of 20 dimensions (columns) for 40 elements (rows).
-The elements are different experimental plots with discrete variations of nutrient enrichment, depth, and freshwater benthic invertebrates.
-<!-- NC: Explain a little more about what the data are here as I don't entirely get it. EG what's a treatement etc? Just a really simple explanataion is perfect -->
-
-```{r}
-## Loading the data
-data(McClean_data)
-
-## Ordinated matrix
-ordinated_matrix <- McClean_data$ordination
-
-## Treatments and depths
-treatments <- McClean_data$treatment
-depth <- McClean_data$depth
-```
-
-### Classic ecological "disparity" analysis
-
-A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different experimental plots differ in their values for nutrient enrichment, depth, and freshwater benthic invertebrates.
-<!-- NC: or similar just to make it clearer -->
-
-```{r, fig.width=6, fig.height=6}
-## Setting the colors that will represent the treatments
-cols <- sub("a", "red", treatments)
-cols <- sub("b", "blue", cols)
-## Setting the symbols that will represent the depth
-pchs <- sub(1, 16, depth)
-pchs <- as.numeric(sub(2, 17, pchs))
-
-## Graphical option
-op <- par(bty = "n")
-## A classic 2D ordination plot
-plot(ordinated_matrix[, 1:2], col = cols, pch = pchs,
-     xlab = "PC 1", ylab = "PC 2",
-     xlim = range(ordinated_matrix[, 1:2]) + c(0, 100))
-par(op)
-```
-
-This shows the distribution of the experimental plots along the two first axis of variation of the ordinated distance matrix (i.e. the first two dimensions).
-At a first glance, it seems difficult to see a clear effect of the treatments (blue or red) or the depth (rounds or triangles) as the different experimental plots with the same parameters don't seem to cluster together.
-However, this may be because this plot ignores the 18 other dimensions of the ordination!
-Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e.
-the values on the first dimension do not represent a continuous trait (e.g.
-depth), instead they just represent the ordinations of correlations between the data and some factors.
-
-Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.
-
-### Multidimensional analysis with `dispRity`
-
-#### Splitting the data with respect to the different factors
-
-The first step is to create different subsamples that represent subsamples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
-Each of these subsamples will contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
-In our example, we are going to group the elements according to their depth and treatment.
-
-```{r}
-## Creating a table that contain the elements and their attributes
-factors <- as.data.frame(matrix(data = c(treatments, depth),
-          nrow = nrow(ordinated_matrix), ncol = 2, byrow = FALSE,
-          dimnames = list(rownames(ordinated_matrix))))
-names(factors) <- c("Treat", "Depth")
-head(factors)
-```
-
-Second, let’s split the data according by depth and treatment to create the subsamples of the ordinated space by using the `custom.subsamples` function:
-
-```{r}
-## Splitting the ordinated space into four subsamples
-(customised_subsamples <- custom.subsamples(ordinated_matrix, factors))
-```
-
-#### Calculating disparity
-
-In this example, we are going to define disparity as the n-dimensional ellipsoid volume (see `?hyper.volume`).
-
-```{r}
-## Calculating disparity as the ellipsoid volume of each group
-disparity <- dispRity(customised_subsamples, metric = diagonal)
-summary(disparity)
-```
-
-We can also bootstrap the data to test the robustness of the measured disparity to outliers.
-Also, as we can see, each subsample has different numbers of elements.
-It might also be interesting to rarefy the data to have only subsamples with the same number of elements.
-
-```{r}
-## Bootstrapping the subsamples and using rarefaction
-## (i.e. only re-sampling 17 elements each time)
-bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 100,
-                                 rarefaction = 17)
-```
-
-We can now run a more robust disparity analysis using the bootstrapped data:
-
-```{r}
-## Calculating the bootstrapped disparity
-disparity <- dispRity(bootstrapped_data, metric = diagonal)
-```
-
-#### Summarising and displaying the data
-
-Now let's summarise this data:
-
-```{r, fig.width=12, fig.height=6, eval=TRUE}
-## Summary table of the data
-summary(disparity)
-
-## Graphical options
-op <- par(mfrow = (c(1, 2)), bty = "n")
-
-## Plotting the bootstrapped disparity
-plot(disparity, main = "Bootstrapped", las = 2, xlab = "")
-
-## Plotting the rarefied disparity
-plot(disparity, rarefaction = 17, main = "Rarefied", las = 2, xlab = "")
-
-par(op)
-```
-
-As we can see, there seems to be no strong effect of the number of experimental plots in each subsample (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
-
-#### Testing our hypotheses
-
-Finally, we can test our hypothesis (whether the water treatment at certain depths alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.
-<!-- Ah so this is what you are testing! Make this clearer in the intro bit to this section -->
-
-```{r}
-## Testing the effect of treatment and depth on the bootstrapped data
-summary(test.dispRity(disparity, test = aov, comparisons = "all"))
-```
-
-## Simulating discrete morphological data
-
-The `dispRity` package also allows users to simulate discrete morphological data matrices.
-In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
-The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
-To understand these methods please refer to the phylogenetics literature.
-<!-- NC: I wonder if you need to cite some papers here to explain how people can find out what a suitable model is etc? Or do we just assume everyone knows about making phylogenies? -->
-
-```{r}
-set.seed(3)
-## Simulating a starting tree with 15 taxa as a random coalescent tree
-my_tree <- rcoal(15)
-
-## Generating a matrix with 100 characters (85% binary and 15% three state) and
-## an equal rates model with a gamma rate distribution (0.5, 1) with no 
-## invariant characters.
-my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
-    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)
-
-## The first few lines of the matrix
-my_matrix[1:5, 1:10]
-
-## Checking the matrix properties with a quick Maximum Parsimony tree search
-check.morpho(my_matrix, my_tree)
-```
-
-Note that this example produces a tree with a great consistency index and an identical topology to the random coalescent tree!
-Nearly too good to be true...
-
-### A more detailed descriptions
-
-`sim.morpho` is really flexible and takes many different arguments to allow users to simulate realistic matrices.
-It has three implemented models: `"ER"` for Equal Rates (the M*k* model); `"HKY"`, the molecular HKY model but transforms pyrines in 0's and pyrimidines in 1's; or the `"mixed"` model that randomly uses an `"ER"` and/or `"HKY"` for the binary characters and `"ER"` for the multistate (>2) characters.
-Both models take specific distributions for their rate or substitution models.
-These distributions should be passed to these arguments in the format of `c(sampler_function, distribution_parameters)` where the the sampler function is a the random generation function of that distribution (e.g.
-`rnorm`, `runif`, etc...) and the parameters are any parameters to be passed to this function.
-
-`check.morpho` runs a quick Maximum Parsimony tree search using the `phangorn` parsimony algorithm.
-It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g.
-the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree to determine how different they are.
-
-### Parameters for a realistic(ish) matrix
-There are many parameters that can create a ``realistic'' matrix (i.e.
-not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all end up creating "good" matrices.
-The following parameters however, seem to generate fairly ``realistic'' matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
-<!-- NC: Check that I haven't changed what you mean here where i swapped good for realistic. Realistic is a better term. Or be clearer what you mena by "good" -->
-
-```{r}
-set.seed(0)
-## tree
-my_tree <- rcoal(15)
-## matrix
-morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
-    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
-check.morpho(morpho_mat, my_tree)
-```
-
-### `space.maker`
-Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
-This function allows users to simulate multidimensional spaces with a certain number of properties.
-It takes as arguments the number of elements (data points) and dimensions to create the space but also permits more fine tuning in the data simulation: it is possible to give a specific distribution to each dimension, provide a correlation matrix to link the dimensions or even provide an *a priori* distribution of the variance for each distributions!
-<!-- ooh fancy! -->
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(mfrow = (c(2, 2)), bty = "n")
-## Visualising 3D spaces
-require(scatterplot3d)
-
-## A cube space
-scatterplot3d(space.maker(2500, 3, runif), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Uniform cube space")
-
-## A plane space
-scatterplot3d(space.maker(2500, 3, c(runif, runif, runif),
-              arguments = list(list(min = 0, max = 0), NULL, NULL)), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Uniform plane space")
-
-## An ellipsoid space (=a spheric space with correlation)
-cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3)
-scatterplot3d(space.maker(2500, 3, rnorm, cor.matrix = cor_matrix), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Normal ellipsoid space")
-
-## A cylindrical space with a decreasing variance per axis
-scatterplot3d(space.maker(2500, 3, c(rnorm, rnorm, runif),
-              scree = c(0.7, 0.2, 0.1)), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Normal cylindrical space")
-
-## Resetting the graphic parameters
-par(op)
-```
-
-# Glossary <a name="glossary"></a>
-
--   **Ordinated space**. The mathematical multidimensional object that will be analysed with this package.
-    In morphometrics, this is often referred to as the morphospace.
-    However it may also be referred to as the cladisto-space for cladistic data or the eco-space for ecological data etc.
-    In practice, this term designates an ordinated matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
-
--   **Elements**. The rows of the ordinated space. Elements can be taxa, field sites, countries etc.
-
--   **Dimensions**. The columns of the ordinated space. The dimensions are referred to as axes of variation, or principal components, for ordinated spaces obtained from a PCA.
-
--   **Subsamples**. Sub-samples of the ordinated space.
-    A sub-sample (or subsamples) contains the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
-    For example, if our ordinated space is composed of birds and mammals (the elements) and 50 principal components (the dimensions), we can create two subsamples containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
-
-# The dispRity package's guts
+# The guts of the dispRity package
 

### 2017-08-15 inst/gitbook/_book/manual.tex
moved manuals to inst/

\documentclass[]{book}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={dispRity manual},
            pdfauthor={Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{dispRity manual}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Thomas Guillerme
(\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}) and Natalie
Cooper
(\href{mailto:natalie.cooper@nhm.ac.uk}{\nolinkurl{natalie.cooper@nhm.ac.uk}})}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{2017-08-15}

\usepackage{booktabs}

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}{Lemma}[chapter]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{corollary}{Corollary}[chapter]
\newtheorem{proposition}{Proposition}[chapter]
\theoremstyle{definition}
\newtheorem{example}{Example}[chapter]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\begin{document}
\maketitle

{
\setcounter{tocdepth}{1}
\tableofcontents
}
\chapter{\texorpdfstring{\texttt{dispRity}}{dispRity}}\label{disprity}

This is a package for measuring disparity in \texttt{R}. It allows users
to summarise ordinated matrices (e.g.~MDS, PCA, PCO, PCoA) into single
values so they can easily be compared. This manual is based on the
version \texttt{1.0}.

\section{\texorpdfstring{What is
\texttt{dispRity}?}{What is dispRity?}}\label{what-is-disprity}

This is a package for measuring disparity in \texttt{R}. It allows users
to summarise ordinated matrices (e.g.~MDS, PCA, PCO, PCoA) to perform
some multidimensional analysis. Typically, these analysis are used in
palaeobiology and evolutionary biology to study the changes in
morphology through time. However, there are many more applications in
ecology, evolution and beyond.

\section{Installing and running the
package}\label{installing-and-running-the-package}

You can install this package easily if you use the latest version of
\texttt{R} (\textgreater{} 4.0) and \texttt{devtools}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Checking if devtools is already installed}
\ControlFlowTok{if}\NormalTok{(}\OperatorTok{!}\KeywordTok{require}\NormalTok{(devtools)) }\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}

\NormalTok{## Installing the latest released version directly from GitHub}
\KeywordTok{install_github}\NormalTok{(}\StringTok{"TGuillerme/dispRity"}\NormalTok{, }\DataTypeTok{ref =} \StringTok{"release"}\NormalTok{)}

\NormalTok{## Loading the package}
\KeywordTok{library}\NormalTok{(dispRity)}
\end{Highlighting}
\end{Shaded}

Note this uses the \texttt{release} branch (version\texttt{0.4}). For
the piping-hot (but potentially unstable) version, you can change the
argument \texttt{ref\ =\ release} to \texttt{ref\ =\ master}.
\texttt{dispRity} depends mainly on the \texttt{ape} package and uses
functions from several other packages (\texttt{ade4}, \texttt{geometry},
\texttt{grDevices}, \texttt{hypervolume}, \texttt{paleotree},
\texttt{snow}, \texttt{Claddis}, \texttt{geomorph} and \texttt{RCurl}).

\section{Why not CRAN?}\label{why-not-cran}

This package is not available on CRAN. This is mainly because some parts
are still in development and that the reactivity of GitHub is better for
implementing new suggestions from users. However, the package follows
the strict (and useful!) CRAN standards via Travis. \#\# Help

If you need help with the package, hopefully the following manual will
be useful. However, parts of this package are still in development and
some other parts are probably not covered. Thus if you have suggestions
or comments on on what has already been developed or will be developed,
please send me an email
(\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}) or if you
are a GitHub user, directly create an issue on the
\href{https://github.com/TGuillerme/dispRity}{GitHub page}.

\section{Citations}\label{citations}

You can cite both the package or this manual with the following
citation:

\begin{quote}
Guillerme, T. (2016). dispRity: a package for measuring disparity in R.
Zenodo. 10.5281/zenodo.55646
\end{quote}

Note that this citation is only temporary (but can still be used!). A
future proper version of the latest package release, this manual and an
associated methods paper should be submitted soon(ish!).

\chapter{Glossary}\label{glossary}

\begin{itemize}
\item
  \textbf{Ordinated space}. The mathematical multidimensional object
  that will be analysed with this package. In morphometrics, this is
  often referred to as the morphospace. However it may also be referred
  to as the cladisto-space for cladistic data or the eco-space for
  ecological data etc. In practice, this term designates an ordinated
  matrix where the columns represent the dimensions of the space (often
  -- but not necessarily - \textgreater{} 3!) and the rows represent the
  elements within this space.
\item
  \textbf{Elements}. The rows of the ordinated space. Elements can be
  taxa, field sites, countries etc.
\item
  \textbf{Dimensions}. The columns of the ordinated space. The
  dimensions are referred to as axes of variation, or principal
  components, for ordinated spaces obtained from a PCA.
\item
  \textbf{Subsamples}. Sub-samples of the ordinated space. A sub-sample
  (or subsamples) contains the same number of dimensions as the
  ordinated space but may contain a smaller subset of elements. For
  example, if our ordinated space is composed of birds and mammals (the
  elements) and 50 principal components (the dimensions), we can create
  two subsamples containing just mammals or birds, but with the same 50
  dimensions, to compare disparity in the two clades.
\end{itemize}

\chapter{Getting started with
dispRity}\label{getting-started-with-disprity}

\section{What sort of data does dispRity work
with?}\label{what-sort-of-data-does-disprity-work-with}

Disparity can be estimated from pretty much any matrix. Classically,
however, it is measured from ordinated matrices. These matrices can be
from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they
have your element names as rows (taxa, experiments, countries etc.) and
your ordination axes as columns (the dimensions of your dataset).

\subsection{\texorpdfstring{Ordination matrices from
\texttt{Claddis}}{Ordination matrices from Claddis}}\label{ordination-matrices-from-claddis}

\begin{verbatim}
## Warning: package 'phytools' was built under R version 3.4.1
\end{verbatim}

\texttt{dispRity} package can easily take data from \texttt{Claddis}
using the \texttt{Claddis.ordination} function. For this, simply input a
matrix in the \texttt{Claddis} format to the function and it will
automatically calculate and ordinate the distances among taxa:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{require}\NormalTok{(Claddis)}

\NormalTok{## Ordinating the example data from Claddis}
\KeywordTok{Claddis.ordination}\NormalTok{(Michaux1989) }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                      [,1]          [,2]       [,3]
## Ancilla      7.252259e-17  4.154578e-01  0.2534942
## Turrancilla -5.106645e-01 -4.566150e-16 -0.2534942
## Ancillista   5.106645e-01 -8.153839e-16 -0.2534942
## Amalda      -3.207162e-16 -4.154578e-01  0.2534942
\end{verbatim}

Note that several options are available, namely which type of distance
should be computed. See more info in the function manual
(\texttt{?Claddis.ordination}). Alternatively, it is of course also
possible to manual calculate the ordination matrix using the functions
\texttt{Claddis::MorphDistMatrix} and \texttt{stats::cmdscale}.

\subsection{\texorpdfstring{Ordination matrices from
\texttt{geomorph}}{Ordination matrices from geomorph}}\label{ordination-matrices-from-geomorph}

You can also easily use data from \texttt{geomorph} using the
\texttt{geomorph.ordination} function. This function simply takes
Procrustes aligned data and performs an ordination:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{require}\NormalTok{(geomorph)}

\NormalTok{## Loading the plethodon dataset}
\KeywordTok{data}\NormalTok{(plethodon)}

\NormalTok{## Performing a Procrustes transform on the landmarks}
\NormalTok{procrustes <-}\StringTok{ }\KeywordTok{gpagen}\NormalTok{(plethodon}\OperatorTok{$}\NormalTok{land, }\DataTypeTok{PrinAxes =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{print.progress =} \OtherTok{FALSE}\NormalTok{)}

\NormalTok{## Ordinating this data}
\KeywordTok{geomorph.ordination}\NormalTok{(procrustes)[}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{,}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                PC1        PC2           PC3          PC4          PC5
## [1,] -0.0369931363 0.05118247 -0.0016971082 -0.003128809 -0.010936371
## [2,] -0.0007493738 0.05942082  0.0001371715 -0.002768680 -0.008117383
## [3,]  0.0056004654 0.07419599 -0.0052612103 -0.005034566 -0.002746592
## [4,] -0.0134808572 0.06463959 -0.0458436015 -0.007887369  0.009816827
## [5,] -0.0334696244 0.06863518  0.0136292041  0.007359409  0.022347225
\end{verbatim}

Options for the ordination (from \texttt{?prcomp}) can be directly
passed to this function to perform customised ordinations. Additionally
you can give the function a \texttt{geomorph.data.frame} object. If the
latter contains sorting information (i.e.~factors), they can be directly
used to make a customised \texttt{dispRity} object
\protect\hyperlink{customised-subsamples}{customised \texttt{dispRity}
object}!

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Using a geomorph.data.frame}
\NormalTok{geomorph_df <-}\StringTok{ }\KeywordTok{geomorph.data.frame}\NormalTok{(procrustes,}
     \DataTypeTok{species =}\NormalTok{ plethodon}\OperatorTok{$}\NormalTok{species, }\DataTypeTok{site =}\NormalTok{ plethodon}\OperatorTok{$}\NormalTok{site)}

\NormalTok{## Ordinating this data and making a dispRity object}
\KeywordTok{geomorph.ordination}\NormalTok{(geomorph_df)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 4 customised subsamples for 40 elements:
##     species.Jord, species.Teyah, site.Allo, site.Symp.
\end{verbatim}

More about these \texttt{dispRity} objects below!

\subsection{Other kinds of ordination
matrices}\label{other-kinds-of-ordination-matrices}

If you are not using the packages mentioned above (\texttt{Claddis} and
\texttt{geomorph}) you can easily make your own ordination matrices by
using the following functions from the \texttt{stats} package. Here is
how to do it for the following types of matrices:

\begin{itemize}
\tightlist
\item
  Multivariate matrices (principal components analysis; PCA)
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## A multivariate matrix}
\KeywordTok{head}\NormalTok{(USArrests)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##            Murder Assault UrbanPop Rape
## Alabama      13.2     236       58 21.2
## Alaska       10.0     263       48 44.5
## Arizona       8.1     294       80 31.0
## Arkansas      8.8     190       50 19.5
## California    9.0     276       91 40.6
## Colorado      7.9     204       78 38.7
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Ordinating the matrix using `prcomp` }
\NormalTok{ordination <-}\StringTok{ }\KeywordTok{prcomp}\NormalTok{(USArrests)}

\NormalTok{## Selecting the ordinated matrix}
\NormalTok{ordinated_matrix <-}\StringTok{ }\NormalTok{ordination}\OperatorTok{$}\NormalTok{x}
\KeywordTok{head}\NormalTok{(ordinated_matrix)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                  PC1        PC2        PC3        PC4
## Alabama     64.80216 -11.448007 -2.4949328 -2.4079009
## Alaska      92.82745 -17.982943 20.1265749  4.0940470
## Arizona    124.06822   8.830403 -1.6874484  4.3536852
## Arkansas    18.34004 -16.703911  0.2101894  0.5209936
## California 107.42295  22.520070  6.7458730  2.8118259
## Colorado    34.97599  13.719584 12.2793628  1.7214637
\end{verbatim}

This results in a ordinated matrix with US states as elements and four
dimensions (PC 1 to 4). For an alternative method, see the
\texttt{?princomp} function.

\begin{itemize}
\tightlist
\item
  Distance matrices (classical multidimensional scaling; MDS)
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## A matrix of distances between cities}
\KeywordTok{str}\NormalTok{(eurodist)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Class 'dist'  atomic [1:210] 3313 2963 3175 3339 2762 ...
##   ..- attr(*, "Size")= num 21
##   ..- attr(*, "Labels")= chr [1:21] "Athens" "Barcelona" "Brussels" "Calais" ...
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Ordinating the matrix using cmdscale() with k = 5 dimensions }
\NormalTok{ordinated_matrix <-}\StringTok{ }\KeywordTok{cmdscale}\NormalTok{(eurodist, }\DataTypeTok{k =} \DecValTok{5}\NormalTok{)}
\KeywordTok{head}\NormalTok{(ordinated_matrix)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                 [,1]      [,2]       [,3]       [,4]       [,5]
## Athens    2290.27468 1798.8029   53.79314 -103.82696 -156.95511
## Barcelona -825.38279  546.8115 -113.85842   84.58583  291.44076
## Brussels    59.18334 -367.0814  177.55291   38.79751  -95.62045
## Calais     -82.84597 -429.9147  300.19274  106.35369 -180.44614
## Cherbourg -352.49943 -290.9084  457.35294  111.44915 -417.49668
## Cologne    293.68963 -405.3119  360.09323 -636.20238  159.39266
\end{verbatim}

This results in a ordinated matrix with European cities as elements and
five dimensions.

Of course any other method for creating the ordination matrix is totally
valid, you can also not use any ordination at all! The only requirements
for the \texttt{dispRity} functions is that the input is a matrix with
elements as rows and dimensions as columns.

\section{Performing a simple dispRity
analysis}\label{performing-a-simple-disprity-analysis}

Two \texttt{dispRity} functions allow users to run an analysis pipeline
simply by inputting an ordination matrix. These functions allow users to
either calculate the disparity through time
(\texttt{dispRity.through.time}) or the disparity of user-defined groups
(\texttt{dispRity.per.group}).

\textbf{IMPORTANT}

Note that \texttt{disparity.through.time} and
\texttt{disparity.per.group} are wrapper functions (i.e.~they
incorporate lots of other functions) that allow users to run a basic
disparity-through-time, or disparity among groups, analysis without too
much effort. As such they use a lot of default options. These are
described in the help files for the functions that are used to make the
wrapper functions, and not described in the help files for
\texttt{disparity.through.time} and \texttt{disparity.per.group}. These
defaults are good enough for \textbf{data exploration}, but for a proper
analysis you should consider the \textbf{best parameters for your
question and data}. For example, which metric should you use? How many
bootstraps do you require? What model of evolution is most appropriate
if you are time slicing? Should you rarefy the data? See
\protect\hyperlink{time-slicing}{\texttt{time.subsamples}},
\protect\hyperlink{customised-subsamples}{\texttt{custom.subsamples}},
\protect\hyperlink{bootstraps-and-rarefactions}{\texttt{boot.matrix}}
and \protect\hyperlink{disparity-metrics}{\texttt{dispRity.metric}} for
more details of the defaults used in each of these functions. Note that
any of these default arguments can be changed within the
\texttt{disparity.through.time} or \texttt{disparity.per.group}
functions.

\hypertarget{example-data}{\subsection{Example
data}\label{example-data}}

To illustrate these functions, we will use data from
\citet{beckancient2014}. This dataset contains an ordinated matrix of 50
discrete characters from mammals (\texttt{BeckLee\_mat50}), another
matrix of the same 50 mammals and the estimated discrete data characters
of their descendants (thus 50 + 49 rows, \texttt{BeckLee\_mat99}), a
dataframe containing the ages of each taxon in the dataset
(\texttt{BeckLee\_ages}) and finally a phylogenetic tree with the
relationships among the 50 mammals (\texttt{BeckLee\_tree}).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading the ordinated matrices}
\KeywordTok{data}\NormalTok{(BeckLee_mat50)}
\KeywordTok{data}\NormalTok{(BeckLee_mat99)}

\NormalTok{## The first five taxa and dimensions of the 50 taxa matrix}
\KeywordTok{head}\NormalTok{(BeckLee_mat50[, }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{])}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                    [,1]          [,2]        [,3]       [,4]      [,5]
## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The first five taxa and dimensions of the 99 taxa + ancestors matrix}
\NormalTok{BeckLee_mat99[}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{98}\NormalTok{, }\DecValTok{99}\NormalTok{), }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                   [,1]       [,2]       [,3]        [,4]        [,5]
## Cimolestes -0.60824375 -0.0323683 0.08458885 -0.43384481 -0.30536875
## Maelestes  -0.57302058 -0.2840361 0.01308847 -0.12588477  0.06123611
## n48        -0.05529018  0.4799330 0.04118477  0.04944912 -0.35588301
## n49        -0.13067785  0.4478168 0.11956268  0.13800340 -0.32227852
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading a list of first and last occurrence dates for the fossils}
\KeywordTok{data}\NormalTok{(BeckLee_ages)}
\KeywordTok{head}\NormalTok{(BeckLee_ages)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##             FAD  LAD
## Adapis     37.2 36.8
## Asioryctes 83.6 72.1
## Leptictis  33.9 33.3
## Miacis     49.0 46.7
## Mimotona   61.6 59.2
## Notharctus 50.2 47.0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading and plotting the phylogeny}
\KeywordTok{data}\NormalTok{(BeckLee_tree)}
\KeywordTok{plot}\NormalTok{(BeckLee_tree, }\DataTypeTok{cex =} \FloatTok{0.8}\NormalTok{) }
\KeywordTok{axisPhylo}\NormalTok{(}\DataTypeTok{root =} \DecValTok{140}\NormalTok{)}
\KeywordTok{nodelabels}\NormalTok{(}\DataTypeTok{cex =} \FloatTok{0.5}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-10-1.pdf}

Of course you can use your own data as detailed in the
\protect\hyperlink{What-sort-of-data-does-dispRity-work-with}{previous
chapter}.

\subsection{Disparity through time}\label{disparity-through-time}

The \texttt{dispRity.through.time} function calculates disparity through
time, a common analysis in palaeontology. This function (and the
following one) uses an analysis pipeline with a lot of default
parameters to make the analysis as simple as possible. Of course all the
defaults can be changed if required, more on this later.

For a disparity through time analysis, you will need:

\begin{itemize}
\tightlist
\item
  An ordinated matrix (we covered that above)
\item
  A phylogenetic tree: this must be a \texttt{phylo} object (from the
  \texttt{ape} package) and needs a \texttt{root.time} element. To give
  your tree a root time (i.e.~an age for the root), you can simply do
  \texttt{my\_tree\$root.time\ \textless{}-\ my\_age}.
\item
  The required number of time subsamples (here \texttt{time\ =\ 3})
\item
  Your favourite disparity metric (here the sum of variances)
\end{itemize}

Using the \citet{beckancient2014} data described
\protect\hyperlink{example-data}{above}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring disparity through time}
\NormalTok{disparity_data <-}\StringTok{ }\KeywordTok{dispRity.through.time}\NormalTok{(BeckLee_mat50, BeckLee_tree,}
                                        \DataTypeTok{time =} \DecValTok{3}\NormalTok{, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances))}
\end{Highlighting}
\end{Shaded}

This generates a \texttt{dispRity} object (see
\protect\hyperlink{manipulating-dispRity-objects}{here} for technical
details). When displayed, these \texttt{dispRity} objects provide us
with information on the operations done to the matrix:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Print the disparity_data object}
\NormalTok{disparity_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 3 discrete time subsamples for 50 elements with 48 dimensions:
##     133.51104 - 89.00736, 89.00736 - 44.50368, 44.50368 - 0.
## Data was bootstrapped 100 times (method:"full").
## Disparity was calculated as: metric.
\end{verbatim}

We asked for three subsamples (evenly spread across the age of the
tree), the data was bootstrapped 100 times (default) and the metric used
was the sum of variances.

We can now summarise or plot the \texttt{disparity\_data} object, or
perform statistical tests on it (e.g.~a simple \texttt{lm}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Summarising disparity through time}
\KeywordTok{summary}\NormalTok{(disparity_data)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##             subsamples  n   obs bs.median  2.5%   25%   75% 97.5%
## 1 133.51104 - 89.00736  5 1.575     1.305 0.729 1.118 1.420 1.509
## 2  89.00736 - 44.50368 29 1.922     1.867 1.775 1.830 1.889 1.922
## 3         44.50368 - 0 16 1.990     1.871 1.716 1.831 1.914 1.942
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Plotting the results}
\KeywordTok{plot}\NormalTok{(disparity_data, }\DataTypeTok{type =} \StringTok{"continuous"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-13-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Testing for an difference among the time bins}
\NormalTok{disp_lm <-}\StringTok{ }\KeywordTok{test.dispRity}\NormalTok{(disparity_data, }\DataTypeTok{test =}\NormalTok{ lm, }\DataTypeTok{comparisons =} \StringTok{"all"}\NormalTok{)}
\KeywordTok{summary}\NormalTok{(disp_lm)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Call:
## test(formula = data ~ subsamples, data = data)
## 
## Residuals:
##      Min       1Q   Median       3Q      Max 
## -0.56623 -0.04160  0.01049  0.05507  0.31886 
## 
## Coefficients:
##                               Estimate Std. Error t value Pr(>|t|)    
## (Intercept)                    1.25647    0.01270   98.97   <2e-16 ***
## subsamples44.50368 - 0         0.60863    0.01795   33.90   <2e-16 ***
## subsamples89.00736 - 44.50368  0.60169    0.01795   33.51   <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 0.127 on 297 degrees of freedom
## Multiple R-squared:  0.8361, Adjusted R-squared:  0.835 
## F-statistic: 757.5 on 2 and 297 DF,  p-value: < 2.2e-16
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Note that ANOVA assumptions were not checked here!}
\end{Highlighting}
\end{Shaded}

Please refer to the \protect\hyperlink{specific-tutorial}{specific
tutorials} for (much!) more information on the nuts and bolts of the
package. You can also directly explore the specific function help files
within R and navigate to related functions.

\hypertarget{disparity-among-groups}{\subsection{Disparity among
groups}\label{disparity-among-groups}}

The \texttt{dispRity.per.group} function is used if you are interested
in looking at disparity among groups rather than through time. For
example, you could ask if there is a difference in disparity between two
groups?

To perform such an analysis, you will need:

\begin{itemize}
\tightlist
\item
  An matrix with rows as elements and columns as dimensions (always!)
\item
  A list of group members: this list should be a list of numeric vectors
  or names corresponding to the row names in the matrix. For example
  \texttt{list("a"\ =\ c(1,2),\ "b"\ =\ c(3,4))} will create a group
  \emph{a} containing elements 1 and 2 from the matrix and a group
  \emph{b} containing elements 3 and 4. Note that elements can be
  present in multiple groups at once.
\item
  Your favourite disparity metric (here the sum of variances)
\end{itemize}

Using the \citet{beckancient2014} data described
\protect\hyperlink{example-data}{above}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating the two groups (crown versus stem) as a list}
\NormalTok{mammal_groups <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\StringTok{"crown"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{16}\NormalTok{, }\DecValTok{19}\OperatorTok{:}\DecValTok{41}\NormalTok{, }\DecValTok{45}\OperatorTok{:}\DecValTok{50}\NormalTok{),}
                      \StringTok{"stem"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{15}\NormalTok{, }\DecValTok{17}\OperatorTok{:}\DecValTok{18}\NormalTok{, }\DecValTok{42}\OperatorTok{:}\DecValTok{44}\NormalTok{))}

\NormalTok{## Measuring disparity for each group}
\NormalTok{disparity_data <-}\StringTok{ }\KeywordTok{dispRity.per.group}\NormalTok{(BeckLee_mat50, }\DataTypeTok{group =}\NormalTok{ mammal_groups,}
                                     \DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances))}
\end{Highlighting}
\end{Shaded}

We can display the disparity of both groups by simply looking at the
output variable (\texttt{disparity\_data}) and then summarising the
\texttt{disparity\_data} object and plotting it, and/or by performing a
statistical test to compare disparity across the groups (here a Wilcoxon
test).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Print the disparity_data object}
\NormalTok{disparity_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 2 customised subsamples for 50 elements with 48 dimensions:
##     crown, stem.
## Data was bootstrapped 100 times (method:"full").
## Disparity was calculated as: metric.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Summarising disparity in the different groups}
\KeywordTok{summary}\NormalTok{(disparity_data)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs bs.median  2.5%   25%   75% 97.5%
## 1      crown 30 1.995     1.936 1.866 1.914 1.947 1.971
## 2       stem 20 1.715     1.632 1.541 1.604 1.667 1.695
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Plotting the results}
\KeywordTok{plot}\NormalTok{(disparity_data)}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-15-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Testing for a difference between the groups}
\KeywordTok{test.dispRity}\NormalTok{(disparity_data, }\DataTypeTok{test =}\NormalTok{ wilcox.test, }\DataTypeTok{details =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## $`crown : stem`
## $`crown : stem`[[1]]
## 
##  Wilcoxon rank sum test with continuity correction
## 
## data:  dots[[1L]][[1L]] and dots[[2L]][[1L]]
## W = 10000, p-value < 2.2e-16
## alternative hypothesis: true location shift is not equal to 0
\end{verbatim}

\chapter{Details of specific
functions}\label{details-of-specific-functions}

The following section contains information specific to some functions.
If any of your questions are not covered in these sections, please refer
to the function help files in R, send me an email
(\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}), or raise
an issue on GitHub. The several tutorials below describe specific
functionalities of certain functions; please always refer to the
function help files for the full function documentation!

\hypertarget{time-slicing}{\section{Time slicing}\label{time-slicing}}

The function \texttt{time.subsamples} allows users to divide the matrix
into different time subsamples or slices given a dated phylogeny that
contains all the elements (i.e.~taxa) from the matrix. Each subsample
generated by this function will then contain all the elements present at
a specific point in time or during a specific period in time.

Two types of time subsamples can be performed by using the
\texttt{method} option:

\begin{itemize}
\tightlist
\item
  Discrete time subsamples (or time-binning) using
  \texttt{method\ =\ discrete}
\item
  Continuous time subsamples (or time-slicing) using
  \texttt{method\ =\ continuous}
\end{itemize}

For the time-slicing method details see \citet{GuillermeSTD}. For both
methods, the function takes the \texttt{time} argument which can be a
vector of \texttt{numeric} values for:

\begin{itemize}
\tightlist
\item
  Defining the boundaries of the time bins (when
  \texttt{method\ =\ discrete})
\item
  Defining the time slices (when \texttt{method\ =\ continuous})
\end{itemize}

Otherwise, the \texttt{time} argument can be set as a single
\texttt{numeric} value for automatically generating a given number of
equidistant time-bins/slices. Additionally, it is also possible to input
a dataframe containing the first and last occurrence data (FAD/LAD) for
taxa that span over a longer time than the given tips/nodes age, so taxa
can appear in more than one time bin/slice.

Here is an example for \texttt{method\ =\ discrete}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Generating three time bins containing the taxa present every 40 Ma}
\KeywordTok{time.subsamples}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat50, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }\DataTypeTok{method =} \StringTok{"discrete"}\NormalTok{,}
                \DataTypeTok{time =} \KeywordTok{c}\NormalTok{(}\DecValTok{120}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{0}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 3 discrete time subsamples for 50 elements:
##     120 - 80, 80 - 40, 40 - 0.
\end{verbatim}

Note that we can also generate equivalent results by just telling the
function that we want three time-bins as follow:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Automatically generate three equal length bins:}
\KeywordTok{time.subsamples}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat50, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }\DataTypeTok{method =} \StringTok{"discrete"}\NormalTok{,}
                \DataTypeTok{time =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 3 discrete time subsamples for 50 elements:
##     133.51104 - 89.00736, 89.00736 - 44.50368, 44.50368 - 0.
\end{verbatim}

In this example, the taxa were split inside each time-bin according to
their age. However, the taxa here are considered as single points in
time. It is totally possible that some taxa could have had longer
longevity and that they exist in multiple time bins. In this case, it is
possible to include them in more than one bin by providing a table of
first and last occurrence dates (FAD/LAD). This table should have the
taxa names as row names and two columns for respectively the first and
last occurrence age:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Displaying the table of first and last occurrence dates for each taxa}
\KeywordTok{head}\NormalTok{(BeckLee_ages)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##             FAD  LAD
## Adapis     37.2 36.8
## Asioryctes 83.6 72.1
## Leptictis  33.9 33.3
## Miacis     49.0 46.7
## Mimotona   61.6 59.2
## Notharctus 50.2 47.0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Generating time bins including taxa that might span between them}
\KeywordTok{time.subsamples}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat50, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }\DataTypeTok{method =} \StringTok{"discrete"}\NormalTok{,}
                \DataTypeTok{time =} \KeywordTok{c}\NormalTok{(}\DecValTok{120}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{0}\NormalTok{), }\DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 3 discrete time subsamples for 50 elements:
##     120 - 80, 80 - 40, 40 - 0.
\end{verbatim}

When using this method, the oldest boundary of the first bin (or the
first slice, see below) is automatically generated as the root age plus
1\% of the tree length, as long as at least three elements/taxa are
present at that point in time. The algorithm adds an extra 1\% tree
length until reaching the required minimum of three elements. It is also
possible to include nodes in each bin by using
\texttt{inc.nodes\ =\ TRUE} and providing a matrix that contains the
ordinated distance among tips \emph{and} nodes.

For the time-slicing method (\texttt{method\ =\ continuous}), the idea
is fairly similar. This option, however, requires a matrix that contains
the ordinated distance among taxa \emph{and} nodes and an extra argument
describing the assumed evolutionary model (via the \texttt{model}
argument). This model argument is used when the time slice occurs along
a branch of the tree rather than on a tip or a node, meaning that a
decision must be made about what the value for the branch should be. The
model can be one of the following:

\begin{itemize}
\tightlist
\item
  \texttt{acctran} where the data chosen along the branch is always the
  one of the descendant
\item
  \texttt{deltran} where the data chosen along the branch is always the
  one of the ancestor
\item
  \texttt{punctuated} where the data chosen along the branch is randomly
  chosen between the descendant or the ancestor
\item
  \texttt{gradual} where the data chosen along the branch is either the
  descendant or the ancestor depending on branch length
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Generating four time slices every 40 million years under a model of gradual evolution}
\KeywordTok{time.subsamples}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat99, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }
    \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"gradual"}\NormalTok{, }\DataTypeTok{time =} \KeywordTok{c}\NormalTok{(}\DecValTok{120}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{0}\NormalTok{),}
    \DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 4 continuous (gradual) time subsamples for 99 elements:
##     120, 80, 40, 0.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Generating four time slices automatically}
\KeywordTok{time.subsamples}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat99, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree,}
    \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"gradual"}\NormalTok{, }\DataTypeTok{time =} \DecValTok{4}\NormalTok{, }\DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 4 continuous (gradual) time subsamples for 99 elements:
##     133.51104, 89.00736, 44.50368, 0.
\end{verbatim}

\hypertarget{customised-subsamples}{\section{Customised
subsamples}\label{customised-subsamples}}

Another way of separating elements into different categories is to use
customised subsamples as briefly explained
\protect\hyperlink{disparity-among-groups}{above}. This function simply
takes the list of elements to put in each group (whether they are the
actual element names or their position in the matrix).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating the two groups as a list}
\NormalTok{mammal_groups <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\StringTok{"crown"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{16}\NormalTok{, }\DecValTok{19}\OperatorTok{:}\DecValTok{41}\NormalTok{, }\DecValTok{45}\OperatorTok{:}\DecValTok{50}\NormalTok{),}
                      \StringTok{"stem"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{15}\NormalTok{, }\DecValTok{17}\OperatorTok{:}\DecValTok{18}\NormalTok{, }\DecValTok{42}\OperatorTok{:}\DecValTok{44}\NormalTok{))}

\NormalTok{## Separating the dataset into two different groups}
\KeywordTok{custom.subsamples}\NormalTok{(BeckLee_mat50, }\DataTypeTok{group =}\NormalTok{ mammal_groups)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 2 customised subsamples for 50 elements:
##     crown, stem.
\end{verbatim}

Elements can easily be assigned to different groups if necessary!

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating the three groups as a list}
\NormalTok{mammal_groups <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\StringTok{"crown"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{16}\NormalTok{, }\DecValTok{19}\OperatorTok{:}\DecValTok{41}\NormalTok{, }\DecValTok{45}\OperatorTok{:}\DecValTok{50}\NormalTok{),}
                      \StringTok{"stem"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{15}\NormalTok{, }\DecValTok{17}\OperatorTok{:}\DecValTok{18}\NormalTok{, }\DecValTok{42}\OperatorTok{:}\DecValTok{44}\NormalTok{).}
                      \StringTok{"all"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{50}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\hypertarget{bootstraps-and-rarefactions}{\section{Bootstraps and
rarefactions}\label{bootstraps-and-rarefactions}}

One important step in analysing ordinated matrices is to
pseudo-replicate the data to see how robust the results are, and how
sensitive they are to outliers in the dataset. This can be achieved
using the function \texttt{boot.matrix} to bootstrap and/or rarefy the
data. The default options will bootstrap the matrix 100 times without
rarefaction using the ``full'' bootstrap method (see below):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Default bootstrapping}
\KeywordTok{boot.matrix}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat50)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 50 elements with 48 dimensions.
## Data was bootstrapped 100 times (method:"full").
\end{verbatim}

The number of bootstrap replicates can be defined using the
\texttt{bootstraps} option. The method can be modified by controlling
which bootstrap algorithm to use through the \texttt{boot.type}
argument. Currently two algorithms are implemented:

\begin{itemize}
\tightlist
\item
  \texttt{full} where the bootstrapping is entirely stochastic (\emph{n}
  elements are replaced by any \emph{m} elements drawn from the data)
\item
  \texttt{single} where only one random element is replaced by one other
  random element for each pseudo-replicate
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Bootstrapping with the single bootstrap method}
\KeywordTok{boot.matrix}\NormalTok{(BeckLee_mat50, }\DataTypeTok{boot.type =} \StringTok{"single"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 50 elements with 48 dimensions.
## Data was bootstrapped 100 times (method:"single").
\end{verbatim}

This function also allows users to rarefy the data using the
\texttt{rarefaction} argument. Rarefaction allows users to limit the
number of elements to be drawn at each bootstrap replication. This is
useful if, for example, one is interested in looking at the effect of
reducing the number of elements on the results of an analysis.

This can be achieved by using the \texttt{rarefaction} option that draws
only \emph{n-x} at each bootstrap replicate (where \emph{x} is the
number of elements not sampled). The default argument is \texttt{FALSE}
but it can be set to \texttt{TRUE} to fully rarefy the data (i.e.~remove
\emph{x} elements for the number of pseudo-replicates, where \emph{x}
varies from the maximum number of elements present in each subsample to
a minimum of three elements). It can also be set to one or more
\texttt{numeric} values to only rarefy to the corresponding number of
elements.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Bootstrapping with the full rarefaction}
\KeywordTok{boot.matrix}\NormalTok{(BeckLee_mat50, }\DataTypeTok{bootstraps =} \DecValTok{20}\NormalTok{, }\DataTypeTok{rarefaction =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 50 elements with 48 dimensions.
## Data was bootstrapped 20 times (method:"full") and fully rarefied.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Or with a set number of rarefaction levels}
\KeywordTok{boot.matrix}\NormalTok{(BeckLee_mat50, }\DataTypeTok{bootstraps =} \DecValTok{20}\NormalTok{, }\DataTypeTok{rarefaction =} \KeywordTok{c}\NormalTok{(}\DecValTok{6}\OperatorTok{:}\DecValTok{8}\NormalTok{, }\DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 50 elements with 48 dimensions.
## Data was bootstrapped 20 times (method:"full") and rarefied to 6, 7, 8, 3 elements.
\end{verbatim}

One additional important argument is \texttt{dimensions} that specifies
how many dimensions from the matrix should be used for further analysis.
When missing, all dimensions from the ordinated matrix are used.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Using the first 50% of the dimensions}
\KeywordTok{boot.matrix}\NormalTok{(BeckLee_mat50, }\DataTypeTok{dimensions =} \FloatTok{0.5}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 50 elements with 24 dimensions.
## Data was bootstrapped 100 times (method:"full").
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Using the first 10 dimensions}
\KeywordTok{boot.matrix}\NormalTok{(BeckLee_mat50, }\DataTypeTok{dimensions =} \DecValTok{10}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 50 elements with 10 dimensions.
## Data was bootstrapped 100 times (method:"full").
\end{verbatim}

Of course, one could directly supply the subsamples generated above
(using \texttt{time.subsamples} or \texttt{custom.subsamples}) to this
function.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating subsamples of crown and stem mammals}
\NormalTok{crown_stem <-}\StringTok{ }\KeywordTok{custom.subsamples}\NormalTok{(BeckLee_mat50,}
                                \DataTypeTok{group =} \KeywordTok{list}\NormalTok{(}\StringTok{"crown"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{16}\NormalTok{, }\DecValTok{19}\OperatorTok{:}\DecValTok{41}\NormalTok{, }\DecValTok{45}\OperatorTok{:}\DecValTok{50}\NormalTok{), }
                                             \StringTok{"stem"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{15}\NormalTok{, }\DecValTok{17}\OperatorTok{:}\DecValTok{18}\NormalTok{, }\DecValTok{42}\OperatorTok{:}\DecValTok{44}\NormalTok{)))}
\NormalTok{## Bootstrapping and rarefying these groups}
\KeywordTok{boot.matrix}\NormalTok{(crown_stem, }\DataTypeTok{bootstraps =} \DecValTok{200}\NormalTok{, }\DataTypeTok{rarefaction =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 2 customised subsamples for 50 elements with 48 dimensions:
##     crown, stem.
## Data was bootstrapped 200 times (method:"full") and fully rarefied.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating time slice subsamples}
\NormalTok{time_slices <-}\StringTok{ }\KeywordTok{time.subsamples}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat99, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }
                               \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"gradual"}\NormalTok{, }
                               \DataTypeTok{time =} \KeywordTok{c}\NormalTok{(}\DecValTok{120}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{0}\NormalTok{),}
                               \DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages)}

\NormalTok{## Bootstrapping the time slice subsamples}
\KeywordTok{boot.matrix}\NormalTok{(time_slices, }\DataTypeTok{bootstraps =} \DecValTok{100}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 4 continuous (gradual) time subsamples for 99 elements with 97 dimensions:
##     120, 80, 40, 0.
## Data was bootstrapped 100 times (method:"full").
\end{verbatim}

\hypertarget{disparity-metrics}{\section{Disparity
metrics}\label{disparity-metrics}}

There are many ways of measuring disparity! In brief, disparity is a
summary metric that will represent an aspect of an ordinated space
(e.g.~a MDS, PCA, PCO, PCoA). For example, one can look at ellipsoid
hyper-volume of the ordinated space \citet{DonohueDim}, the sum and the
product of the ranges and variances \citet{Wills1994} or the median
position of the elements relative to their centroid \citet{Wills1994}.
Of course, there are many more examples of metrics one can use for
describing some aspect of the ordinated space, with some performing
better than other ones at particular descriptive tasks, and some being
more generalist.

Because of this great diversity of metrics, the package
\texttt{dispRity} does not have one way to measure disparity but rather
proposes to facilitate users in defining their own disparity metric that
will best suit their particular analysis. In fact, the core function of
the package, \texttt{dispRity}, allows the user to define any metric
with the \texttt{metric} argument. However the \texttt{metric} argument
has to follow certain rules:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  It must be composed from one to three \texttt{function} objects;
\item
  The function(s) must take as a first argument a \texttt{matrix} or a
  \texttt{vector};
\item
  The function(s) must be of one of the three dimension-levels described
  below;
\item
  At least one of the functions must be of dimension-level 1 or 2 (see
  below).
\end{enumerate}

\subsection{The function
dimension-levels}\label{the-function-dimension-levels}

The metric function dimension-levels determine the ``dimensionality of
decomposition'' of the input matrix. In other words, each
dimension-level designates the dimensions of the output, i.e.~either
three (a \texttt{matrix}); two (a \texttt{vector}); or one (a single
\texttt{numeric} value) dimension.

\begin{figure}
\centering
\includegraphics{dispRity_fun.png}
\caption{Illustration of the different dimension-levels of functions
with an input \texttt{matrix}}
\end{figure}

\subsubsection{Dimension-level 1
functions}\label{dimension-level-1-functions}

A dimension-level 1 function will decompose a \texttt{matrix} or a
\texttt{vector} into a single value:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating a dummy matrix}
\NormalTok{dummy_matrix <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\KeywordTok{rnorm}\NormalTok{(}\DecValTok{12}\NormalTok{), }\DecValTok{4}\NormalTok{, }\DecValTok{3}\NormalTok{)}

\NormalTok{## Example of dimension-level 1 functions}
\KeywordTok{mean}\NormalTok{(dummy_matrix)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] -0.3227241
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{median}\NormalTok{(dummy_matrix)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] -0.2690165
\end{verbatim}

Any summary metric such as mean or median are good examples of
dimension-level 1 functions as they reduce the matrix to a single
dimension (i.e.~one value).

\subsubsection{Dimension-level 2
functions}\label{dimension-level-2-functions}

A dimension-level 2 function will decompose a \texttt{matrix} into a
\texttt{vector}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Defining the function as the product of rows}
\NormalTok{prod.rows <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(matrix) }\KeywordTok{apply}\NormalTok{(matrix, }\DecValTok{1}\NormalTok{, prod)}

\NormalTok{## A dimension-level 2 metric}
\KeywordTok{prod.rows}\NormalTok{(dummy_matrix)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] -1.2630799  0.2148864 -0.1797556 -0.2421790
\end{verbatim}

Several dimension-level 2 functions are implemented in \texttt{dispRity}
(see \texttt{?dispRity.metric}) such as the \texttt{variances} or
\texttt{ranges} functions that calculate the variance or the range of
each dimension of the ordinated matrix respectively.

\subsubsection{Dimension-level 3
functions}\label{dimension-level-3-functions}

Finally a dimension-level 3 function will transform the matrix into
another matrix. Note that the dimension of the output matrix doesn't
need to match the the input matrix:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## A dimension-level 3 metric}
\KeywordTok{var}\NormalTok{(dummy_matrix)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##          [,1]      [,2]      [,3]
## [1,] 1.916420 1.9501955 0.2907060
## [2,] 1.950196 3.2948635 0.3958234
## [3,] 0.290706 0.3958234 0.6704976
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## A dimension-level 3 metric with a forced matrix output}
\KeywordTok{as.matrix}\NormalTok{(}\KeywordTok{dist}\NormalTok{(dummy_matrix))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##          1        2        3        4
## 1 0.000000 5.140790 4.005068 3.120827
## 2 5.140790 0.000000 2.179318 2.982267
## 3 4.005068 2.179318 0.000000 2.174896
## 4 3.120827 2.982267 2.174896 0.000000
\end{verbatim}

\subsection{\texorpdfstring{\texttt{make.metric}}{make.metric}}\label{make.metric}

Of course, functions can be more complex and involve multiple operations
such as the \texttt{centroids} function (see \texttt{?dispRity.metric})
that calculates the Euclidean distance between each element and the
centroid of the ordinated space. The \texttt{make.metric} function
implemented in \texttt{dispRity} is designed to help test and find the
dimension-level of the functions. This function tests:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  If your function can deal with a \texttt{matrix} or a \texttt{vector}
  as an input;
\item
  Your function's dimension-level according to its output
  (dimension-level 1, 2 or 3, see above);
\item
  Whether the function can be implemented in the \texttt{dispRity}
  function (the function is fed into a \texttt{lapply} loop).
\end{enumerate}

For example, let's see if the functions described above are the right
dimension-levels:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Which dimension-level is the mean function? And can it be used in dispRity?}
\KeywordTok{make.metric}\NormalTok{(mean)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## mean outputs a single value.
## mean is detected as being a dimension-level 1 function.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Which dimension-level is the prod.rows function? And can it be used in dispRity?}
\KeywordTok{make.metric}\NormalTok{(prod.rows)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## prod.rows outputs a matrix object.
## prod.rows is detected as being a dimension-level 2 function.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Which dimension-level is the var function? And can it be used in dispRity?}
\KeywordTok{make.metric}\NormalTok{(var)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## var outputs a matrix object.
## var is detected as being a dimension-level 3 function.
## Additional dimension-level 2 and/or 1 function(s) will be needed.
\end{verbatim}

A non verbose version of the function is also available. This can be
done using the option \texttt{silent\ =\ TRUE} and will simply output
the dimension-level of the metric.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Testing whether mean is dimension-level 1}
\ControlFlowTok{if}\NormalTok{(}\KeywordTok{make.metric}\NormalTok{(mean, }\DataTypeTok{silent =} \OtherTok{TRUE}\NormalTok{) }\OperatorTok{!=}\StringTok{ "level1"}\NormalTok{) \{}
    \KeywordTok{message}\NormalTok{(}\StringTok{"The metric is not dimension-level 1."}\NormalTok{)}
\NormalTok{\}}
\NormalTok{## Testing whether var is dimension-level 1}
\ControlFlowTok{if}\NormalTok{(}\KeywordTok{make.metric}\NormalTok{(var, }\DataTypeTok{silent =} \OtherTok{TRUE}\NormalTok{) }\OperatorTok{!=}\StringTok{ "level1"}\NormalTok{) \{}
    \KeywordTok{message}\NormalTok{(}\StringTok{"The metric is not dimension-level 1."}\NormalTok{)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## The metric is not dimension-level 1.
\end{verbatim}

\subsection{\texorpdfstring{Metrics in the \texttt{dispRity}
function}{Metrics in the dispRity function}}\label{metrics-in-the-disprity-function}

Using this metric structure, we can easily use any disparity metric in
the \texttt{dispRity} function as follows:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring disparity as the standard deviation of all the values of the}
\NormalTok{## ordinated matrix (dimension-level 1 function).}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(BeckLee_mat50, }\DataTypeTok{metric =}\NormalTok{ sd))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs
## 1          1 50 0.201
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring disparity as the standard deviation of the variance of each axis of}
\NormalTok{## the ordinated matrix (dimension-level 1 and 2 functions).}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(BeckLee_mat50, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sd, variances)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs
## 1          1 50 0.028
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring disparity as the standard deviation of the variance of each axis of}
\NormalTok{## the variance covariance matrix (dimension-level 1, 2 and 3 functions).}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(BeckLee_mat50, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sd, variances, var)), }\DataTypeTok{round =} \DecValTok{10}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n          obs
## 1          1 50 0.0001025857
\end{verbatim}

Note that the order of each function in the metric argument does not
matter, the \texttt{dispRity} function will automatically detect the
function dimension-levels (using \texttt{make.metric}) and apply them to
the data in decreasing order (dimension-level 3 \textgreater{} 2
\textgreater{} 1).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Disparity as the standard deviation of the variance of each axis of the}
\NormalTok{## variance covariance matrix:}
\NormalTok{disparity1 <-}\StringTok{ }\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(BeckLee_mat50, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sd, variances, var)),}
                      \DataTypeTok{round =} \DecValTok{10}\NormalTok{)}

\NormalTok{## Same as above but using a different function order for the metric argument}
\NormalTok{disparity2 <-}\StringTok{ }\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(BeckLee_mat50, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(variances, sd, var)),}
                      \DataTypeTok{round =} \DecValTok{10}\NormalTok{)}

\NormalTok{## Both ways output the same disparity values:}
\NormalTok{disparity1 }\OperatorTok{==}\StringTok{ }\NormalTok{disparity2}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      subsamples    n  obs
## [1,]       TRUE TRUE TRUE
\end{verbatim}

In these examples, we considered disparity to be a single value. For
example, in the previous example, we defined disparity as the standard
deviation of the variances of each column of the variance/covariance
matrix (\texttt{metric\ =\ c(variances,\ sd,\ var)}). It is, however,
possible to calculate
\protect\hyperlink{disparity-as-a-distribution}{disparity as a
distribution}.

\subsection{\texorpdfstring{Metrics implemented in
\texttt{dispRity}}{Metrics implemented in dispRity}}\label{metrics-implemented-in-disprity}

Several disparity metrics are implemented in the \texttt{dispRity}
package. The detailed list can be found in \texttt{?dispRity.metric}
along with some description of each metric.

\begin{longtable}[]{@{}llll@{}}
\toprule
\begin{minipage}[b]{0.08\columnwidth}\raggedright\strut
Level\strut
\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\raggedright\strut
Name\strut
\end{minipage} & \begin{minipage}[b]{0.61\columnwidth}\raggedright\strut
Description\strut
\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\raggedright\strut
Source\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
3\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{ellipse.volume}1\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The volume of the ellipsoid of the space\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\citep{DonohueDim}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
3\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{convhull.surface}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The surface of the convex hull formed by all the elements\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\href{https://cran.r-project.org/web/packages/geometry/index.html}{\texttt{geometry}}\texttt{::convhulln}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
3\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{convhull.volume}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The volume of the convex hull formed by all the elements\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\href{https://cran.r-project.org/web/packages/geometry/index.html}{\texttt{geometry}}\texttt{::convhulln}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
3\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{hypervolume}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The volume of the ordinated space\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\href{https://cran.r-project.org/web/packages/hypervolume/index.html}{\texttt{hypervolume}}\texttt{::hypervolume}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
3\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{diagonal}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The longest distance in the ordinated space (like the diagonal in two
dimensions)\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\texttt{dispRity}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
2\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{ranges}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The range of each dimension\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\texttt{dispRity}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
2\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{variances}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The variance of each dimension\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\texttt{dispRity}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
2\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{centroids}2\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The distance between each element and the centroid of the ordinated
space\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\texttt{dispRity}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
1\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{mode.val}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The modal value\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\texttt{dispRity}\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

1: This function uses an estimation of the eigenvalue that only works
for MDS or PCoA ordinations (not PCA).

2: Note that by default, the centroid is the centroid of the elements.
It can, however, be fixed to a different value by using the
\texttt{centroid} argument
\texttt{centroids(space,\ centroid\ =\ rep(0,\ ncol(space)))}, for
example the origin of the ordinated space.

\subsection{Equations and
implementations}\label{equations-and-implementations}

Some of the functions described below are implemented in the
\texttt{dispRity} package and do not require any other packages to
calculate
(\href{https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R}{see
implementation here}).

\begin{equation}
    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
\end{equation}

Where \emph{k} is the number of dimensions, and \(\lambda_i\) is the
eigenvalue of each dimension.

\begin{equation}
    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
\end{equation}

Where \emph{k} is the number of dimensions.

\begin{equation}
    ranges = |max(k_i) - min(k_i)|
\end{equation}

Where \emph{k} is the number of dimensions.

\begin{equation}
    variances = \sigma^{2}{k_i}
\end{equation}

Where \emph{k} is the number of dimensions, and \(\sigma^{2}\) is their
variance.

\begin{equation}
    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
\end{equation}

Where \emph{n} is each element in the ordinated space, \emph{k} is the
number of dimensions, and \(Centroid_{k}\) is their mean (or can be set
to another value).

\subsection{Using the different disparity
metrics}\label{using-the-different-disparity-metrics}

Here is a brief demonstration of the main metrics implemented in
\texttt{dispRity}. First, we will create a dummy/simulated ordinated
space using the \texttt{space.maker} utility function (more about that
\protect\hyperlink{space.maker}{here}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating a 10*5 normal space}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{dummy_space <-}\StringTok{ }\KeywordTok{space.maker}\NormalTok{(}\DecValTok{10}\NormalTok{, }\DecValTok{5}\NormalTok{, rnorm)}
\end{Highlighting}
\end{Shaded}

We will use this simulated space to demonstrate the different metrics.

\subsubsection{Volumes and surface
metrics}\label{volumes-and-surface-metrics}

The functions \texttt{ellipse.volume}, \texttt{convhull.surface},
\texttt{convhull.volume} and \texttt{hyper.volume} all measure the
surface or the volume of the ordinated space occupied:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the ellipsoid volume}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ ellipse.volume))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs
## 1          1 10 257.8
\end{verbatim}

\begin{quote}
Because there is only one subsample (i.e.~one matrix) in the dispRity
object, this operation is the equivalent of
\texttt{ellipse.volume(dummy\_space)} (with rounding).
\end{quote}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the convex hull surface}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ convhull.surface))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs
## 1          1 10 11.91
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the convex hull volume}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ convhull.volume))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs
## 1          1 10 1.031
\end{verbatim}

The convex hull functions make a (good) estimation of the
multidimensional properties of the ordinated space. For the full and
correct calculation of the volume of the ordinated space, it is possible
to use the \texttt{hyper.volume} function that has more options but
takes longer to calculate.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the true multidimensional volume}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ hyper.volume))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in hypervolume::hypervolume(matrix, method = method, ...): Log number of observations (1.61) is less than or equal to the number of dimensions (4).
## You may not have enough data to accurately estimate a hypervolume with this dimensionality.
## Consider reducing the dimensionality of the analysis.
\end{verbatim}

\begin{verbatim}
## Warning in hypervolume::hypervolume(matrix, method = method, ...): Log number of observations (2.30) is less than or equal to the number of dimensions (5).
## You may not have enough data to accurately estimate a hypervolume with this dimensionality.
## Consider reducing the dimensionality of the analysis.
\end{verbatim}

\begin{verbatim}
##   subsamples  n   obs
## 1          1 10 408.2
\end{verbatim}

\begin{quote}
Cautionary note: measuring volumes in a high number of dimensions can be
strongly affected by the
\href{https://en.wikipedia.org/wiki/Curse_of_dimensionality}{curse of
dimensionality} that often results in near 0 disparity values.
\end{quote}

\subsubsection{Ranges, variances and
diagonal}\label{ranges-variances-and-diagonal}

The functions \texttt{ranges}, \texttt{variances} and \texttt{diagonal}
all measure properties of the ordinated space based on its dimensional
properties (they are also less affected by the ``curse of
dimensionality''):

\texttt{ranges} and \texttt{variances} both work on the same principle
and measure the range/variance of each dimension:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the ranges of each dimension in the ordinated space}
\KeywordTok{ranges}\NormalTok{(dummy_space)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 2.430909 3.726481 2.908329 2.735739 1.588603
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity as the distribution of these ranges}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ ranges))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n obs.median  2.5%   25%   75% 97.5%
## 1          1 10      2.736 1.673 2.431 2.908 3.645
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity as the sum and the product of these ranges}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, ranges)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs
## 1          1 10 13.39
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(prod, ranges)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs
## 1          1 10 114.5
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the variances of each dimension in the ordinated space}
\KeywordTok{variances}\NormalTok{(dummy_space)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.6093144 1.1438620 0.9131859 0.6537768 0.3549372
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity as the distribution of these variances}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ variances))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n obs.median 2.5%   25%   75% 97.5%
## 1          1 10      0.654 0.38 0.609 0.913 1.121
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity as the sum and the product of these variances}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs
## 1          1 10 3.675
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(prod, variances)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs
## 1          1 10 0.148
\end{verbatim}

The \texttt{diagonal} function measures the multidimensional diagonal of
the whole space (i.e.~in our case the longest Euclidean distance in our
five dimensional space):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the ordinated space's diagonal}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ diagonal))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs
## 1          1 10 3.659
\end{verbatim}

\begin{quote}
This metric is only a Euclidean diagonal (mathematically valid) if the
dimensions within the space are all orthogonal!
\end{quote}

\subsubsection{Centroids metric}\label{centroids-metric}

The \texttt{centroids} metric allows users to measure the position of
the different elements compared to a fixed point in the ordinated space.
By default, this function measures the distance between each element and
their centroid (centre point):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The distribution of the distances between each element and their centroid}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ centroids))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n obs.median  2.5%   25%   75% 97.5%
## 1          1 10      1.435 0.788 1.267 1.993 3.167
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Disparity as the median value of these distances}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(median, centroids)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs
## 1          1 10 1.435
\end{verbatim}

It is however possible to fix the coordinates of the centroid to a
specific point in the ordinated space, as long as it has the correct
number of dimensions:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The distance between each element and the origin of the ordinated space}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ centroids, }\DataTypeTok{centroid =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n obs.median  2.5% 25%   75% 97.5%
## 1          1 10      1.487 0.785 1.2 2.044 3.176
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Disparity as the distance between each element and a specific point in space}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ centroids, }\DataTypeTok{centroid =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{,}\DecValTok{4}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n obs.median  2.5%   25%   75% 97.5%
## 1          1 10      5.489 4.293 5.032 6.155 6.957
\end{verbatim}

\section{Summarising dispRity data
(plots)}\label{summarising-disprity-data-plots}

Because of its architecture, printing \texttt{dispRity} objects only
summarises their content but does not print the disparity value measured
or associated analysis (more about this
\protect\hyperlink{manipulating-dispRity-objects}{here}). To actually
see what is in a dispRity object, one can either use the
\texttt{summary} function for visualising the data in a table or
\texttt{plot} to have a graphical representation of the results.

\subsection{\texorpdfstring{Summarising \texttt{dispRity}
data}{Summarising dispRity data}}\label{summarising-disprity-data}

This function is an S3 function (\texttt{summary.dispRity}) allowing
users to summarise the content of \texttt{dispRity} objects that contain
disparity calculations.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Example data from previous sections}
\NormalTok{crown_stem <-}\StringTok{ }\KeywordTok{custom.subsamples}\NormalTok{(BeckLee_mat50,}
                                \DataTypeTok{group =} \KeywordTok{list}\NormalTok{(}\StringTok{"crown"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{16}\NormalTok{, }\DecValTok{19}\OperatorTok{:}\DecValTok{41}\NormalTok{, }\DecValTok{45}\OperatorTok{:}\DecValTok{50}\NormalTok{), }
                                             \StringTok{"stem"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{15}\NormalTok{, }\DecValTok{17}\OperatorTok{:}\DecValTok{18}\NormalTok{, }\DecValTok{42}\OperatorTok{:}\DecValTok{44}\NormalTok{)))}
\NormalTok{## Bootstrapping and rarefying these groups}
\NormalTok{boot_crown_stem <-}\StringTok{ }\KeywordTok{boot.matrix}\NormalTok{(crown_stem, }\DataTypeTok{bootstraps =} \DecValTok{100}\NormalTok{, }\DataTypeTok{rarefaction =} \OtherTok{TRUE}\NormalTok{)}
\NormalTok{## Calculate disparity}
\NormalTok{disparity_crown_stem <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(boot_crown_stem, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances))}

\NormalTok{## Creating time slice subsamples}
\NormalTok{time_slices <-}\StringTok{ }\KeywordTok{time.subsamples}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat99, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }
    \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"gradual"}\NormalTok{, }\DataTypeTok{time =} \KeywordTok{c}\NormalTok{(}\DecValTok{120}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{0}\NormalTok{),}
    \DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages)}
\NormalTok{## Bootstrapping the time slice subsamples}
\NormalTok{boot_time_slices <-}\StringTok{ }\KeywordTok{boot.matrix}\NormalTok{(time_slices, }\DataTypeTok{bootstraps =} \DecValTok{100}\NormalTok{)}
\NormalTok{## Calculate disparity}
\NormalTok{disparity_time_slices <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(boot_time_slices, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances))}

\NormalTok{## Creating time bin subsamples}
\NormalTok{time_bins <-}\StringTok{ }\KeywordTok{time.subsamples}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat99, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }
    \DataTypeTok{method =} \StringTok{"discrete"}\NormalTok{, }\DataTypeTok{time =} \KeywordTok{c}\NormalTok{(}\DecValTok{120}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{0}\NormalTok{), }\DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages,}
    \DataTypeTok{inc.nodes =} \OtherTok{TRUE}\NormalTok{)}
\NormalTok{## Bootstrapping the time bin subsamples}
\NormalTok{boot_time_bins <-}\StringTok{ }\KeywordTok{boot.matrix}\NormalTok{(time_bins, }\DataTypeTok{bootstraps =} \DecValTok{100}\NormalTok{)}
\NormalTok{## Calculate disparity}
\NormalTok{disparity_time_bins <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(boot_time_bins, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances))}
\end{Highlighting}
\end{Shaded}

These objects are easy to summarise as follows:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Default summary}
\KeywordTok{summary}\NormalTok{(disparity_time_slices)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs bs.median  2.5%   25%   75% 97.5%
## 1        120  5 2.823     2.295 1.292 2.039 2.598 2.829
## 2         80 19 3.233     3.079 2.834 2.996 3.138 3.278
## 3         40 15 3.421     3.211 2.909 3.103 3.309 3.518
## 4          0 10 4.055     3.684 3.247 3.557 3.783 3.940
\end{verbatim}

Information about the number of elements in each subsample and the
observed (i.e.~non-bootstrapped) disparity are also calculated. This is
specifically handy when rarefying the data for example:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(disparity_crown_stem))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs bs.median  2.5%   25%   75% 97.5%
## 1      crown 30 1.995     1.926 1.863 1.912 1.948 1.965
## 2      crown 29    NA     1.931 1.878 1.909 1.944 1.967
## 3      crown 28    NA     1.931 1.882 1.908 1.948 1.978
## 4      crown 27    NA     1.930 1.859 1.908 1.946 1.966
## 5      crown 26    NA     1.928 1.867 1.907 1.948 1.978
## 6      crown 25    NA     1.934 1.853 1.904 1.948 1.974
\end{verbatim}

The summary functions can also take various options such as:

\begin{itemize}
\tightlist
\item
  \texttt{quantile} values for the confidence interval levels (by
  default, the 50 and 95 quantiles are calculated)
\item
  \texttt{cent.tend} for the central tendency to use for summarising the
  results (default is \texttt{median})
\item
  rounding\texttt{option\ corresponding\ to\ the\ number\ of\ decimal\ places\ to\ print\ (default\ is}2`)
\item
  \texttt{recall} option for printing the call of the \texttt{dispRity}
  object as well (default is \texttt{FALSE})
\end{itemize}

These options can easily be changed from the defaults as follows:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Same as above but using the 88th quantile and the standard deviation as the summary }
\KeywordTok{summary}\NormalTok{(disparity_time_slices, }\DataTypeTok{quantile =} \DecValTok{88}\NormalTok{, }\DataTypeTok{cent.tend =}\NormalTok{ sd)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs bs.sd    6%   94%
## 1        120  5 2.823 0.398 1.658 2.705
## 2         80 19 3.233 0.116 2.896 3.243
## 3         40 15 3.421 0.161 2.967 3.485
## 4          0 10 4.055 0.184 3.342 3.907
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Printing the details of the object and rounding the values to the 5th decimal place}
\KeywordTok{summary}\NormalTok{(disparity_time_slices, }\DataTypeTok{recall =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{rounding =} \DecValTok{5}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 4 continuous (gradual) time subsamples for 99 elements with 97 dimensions:
##     120, 80, 40, 0.
## Data was bootstrapped 100 times (method:"full").
## Disparity was calculated as: c(sum, variances).
\end{verbatim}

\begin{verbatim}
##   subsamples  n     obs bs.median    2.5%     25%     75%   97.5%
## 1        120  5 2.82292   2.29498 1.29247 2.03886 2.59781 2.82881
## 2         80 19 3.23312   3.07949 2.83426 2.99627 3.13758 3.27800
## 3         40 15 3.42091   3.21071 2.90858 3.10280 3.30911 3.51829
## 4          0 10 4.05457   3.68433 3.24678 3.55676 3.78275 3.93967
\end{verbatim}

Note that the summary table is a \texttt{data.frame}, hence it is as
easy to modify as any dataframe using \texttt{dplyr}. You can also
export it in \texttt{csv} format using \texttt{write.csv} or
\texttt{write\_csv} or even directly export into \texttt{LaTeX} format
using the following;

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading the xtable package}
\KeywordTok{require}\NormalTok{(xtable)}
\NormalTok{## Converting the table in LaTeX}
\KeywordTok{xtable}\NormalTok{(}\KeywordTok{summary}\NormalTok{(disparity_time_slices))}
\end{Highlighting}
\end{Shaded}

\subsection{\texorpdfstring{Plotting \texttt{dispRity}
data}{Plotting dispRity data}}\label{plotting-disprity-data}

An alternative (and more fun!) way to display the calculated disparity
is to plot the results using the S3 method \texttt{plot.dispRity}. This
function takes the same options as \texttt{summary.dispRity} along with
various graphical options described in the function help files (see
\texttt{?plot.dispRity}).

The plots can be of four different types:

\begin{itemize}
\tightlist
\item
  \texttt{continuous} for displaying continuous disparity curves
\item
  \texttt{box}, \texttt{lines}, and \texttt{polygons} to display
  discrete disparity results in respectively a boxplot, confidence
  interval lines, and confidence interval polygons.
\end{itemize}

\begin{quote}
This argument can be left empty. In this case, the algorithm will
automatically detect the type of subsamples from the \texttt{dispRity}
object and plot accordingly.
\end{quote}

It is also possible to display the number of elements in each subsample
(as a horizontal dotted line) using the option
\texttt{elements\ =\ TRUE}. Additionally, when the data is rarefied, one
can indicate which level of rarefaction to display (i.e.~only display
the results for a certain number of elements) by using the
\texttt{rarefaction} argument.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical parameters}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting continuous disparity results}
\KeywordTok{plot}\NormalTok{(disparity_time_slices, }\DataTypeTok{type =} \StringTok{"continuous"}\NormalTok{)}

\NormalTok{## Plotting discrete disparity results}
\KeywordTok{plot}\NormalTok{(disparity_crown_stem, }\DataTypeTok{type =} \StringTok{"box"}\NormalTok{)}

\NormalTok{## As above but using lines for the rarefaction level of 20 elements only}
\KeywordTok{plot}\NormalTok{(disparity_crown_stem, }\DataTypeTok{type =} \StringTok{"line"}\NormalTok{, }\DataTypeTok{rarefaction =} \DecValTok{20}\NormalTok{)}

\NormalTok{## As above but using polygons while also displaying the number of elements}
\KeywordTok{plot}\NormalTok{(disparity_crown_stem, }\DataTypeTok{type =} \StringTok{"polygon"}\NormalTok{, }\DataTypeTok{elements =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-47-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Resetting graphical parameters}
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

Since \texttt{plot.dispRity} uses the arguments from the generic
\texttt{plot} method, it is of course possible to change pretty much
everything using the regular plot arguments:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting the results with some classic options from plot}
\KeywordTok{plot}\NormalTok{(disparity_time_slices, }\DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\StringTok{"blue"}\NormalTok{, }\StringTok{"orange"}\NormalTok{, }\StringTok{"green"}\NormalTok{),}
    \DataTypeTok{ylab =} \KeywordTok{c}\NormalTok{(}\StringTok{"Some measurement"}\NormalTok{), }\DataTypeTok{xlab =} \StringTok{"Some other measurement"}\NormalTok{,}
    \DataTypeTok{main =} \StringTok{"Many options..."}\NormalTok{, }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{5}\NormalTok{, }\DecValTok{0}\NormalTok{), }\DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{, }\DecValTok{0}\NormalTok{))}

\NormalTok{## Adding a legend}
\KeywordTok{legend}\NormalTok{(}\StringTok{"topleft"}\NormalTok{, }\DataTypeTok{legend =} \KeywordTok{c}\NormalTok{(}\StringTok{"Central tendency"}\NormalTok{,}
                             \StringTok{"Confidence interval 1"}\NormalTok{,}
                            \StringTok{"Confidence interval 2"}\NormalTok{),}
      \DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\StringTok{"blue"}\NormalTok{, }\StringTok{"orange"}\NormalTok{, }\StringTok{"green"}\NormalTok{), }\DataTypeTok{pch =} \DecValTok{19}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-48-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Resetting graphical parameters}
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

In addition to the classic \texttt{plot} arguments, the function can
also take arguments that are specific to \texttt{plot.dispRity} like
adding the number of elements or rarefaction level (as described above),
and also changing the values of the quantiles to plot as well as the
central tendency.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting the results with some plot.dispRity arguments}
\KeywordTok{plot}\NormalTok{(disparity_time_slices, }\DataTypeTok{quantile =} \KeywordTok{c}\NormalTok{(}\KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{10}\NormalTok{, }\DataTypeTok{to =} \DecValTok{100}\NormalTok{, }\DataTypeTok{by =} \DecValTok{10}\NormalTok{)),}
    \DataTypeTok{cent.tend =}\NormalTok{ sd, }\DataTypeTok{type =} \StringTok{"c"}\NormalTok{, }\DataTypeTok{elements =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\StringTok{"black"}\NormalTok{, }\KeywordTok{rainbow}\NormalTok{(}\DecValTok{10}\NormalTok{)),}
    \DataTypeTok{ylab =} \KeywordTok{c}\NormalTok{(}\StringTok{"Disparity"}\NormalTok{, }\StringTok{"Diversity"}\NormalTok{), }\DataTypeTok{time.subsamples =} \OtherTok{FALSE}\NormalTok{,}
    \DataTypeTok{xlab =} \StringTok{"Time (in in units from past to present)"}\NormalTok{, }\DataTypeTok{observed =} \OtherTok{TRUE}\NormalTok{,}
    \DataTypeTok{main =} \StringTok{"Many more options..."}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-49-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Resetting graphical parameters}
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

\begin{quote}
Note that the argument \texttt{observed\ =\ TRUE} allows to plot the
disparity values calculated from the non-bootstrapped data as crosses on
the plot.
\end{quote}

For comparing results, it is also possible to add a plot to the existent
plot by using \texttt{add\ =\ TRUE}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting the continuous disparity with a fixed y axis}
\KeywordTok{plot}\NormalTok{(disparity_time_slices, }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{4}\NormalTok{))}
\NormalTok{## Adding the discrete data}
\KeywordTok{plot}\NormalTok{(disparity_time_bins, }\DataTypeTok{type =} \StringTok{"line"}\NormalTok{, }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{4}\NormalTok{), }\DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{,}
    \DataTypeTok{add =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-50-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Resetting graphical parameters}
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

Finally, if your data has been fully rarefied, it is also possible to
easily look at rarefaction curves by using the
\texttt{rarefaction\ =\ TRUE} argument:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting the rarefaction curves}
\KeywordTok{plot}\NormalTok{(disparity_crown_stem, }\DataTypeTok{rarefaction =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-51-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Resetting graphical parameters}
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

\section{Testing disparity
hypotheses}\label{testing-disparity-hypotheses}

The \texttt{dispRity} package allows users to apply statistical tests to
the calculated disparity to test various hypotheses. The function
\texttt{test.dispRity} works in a similar way to the \texttt{dispRity}
function: it takes a \texttt{dispRity} object, a \texttt{test} and a
\texttt{comparisons} argument.

The \texttt{comparisons} argument indicates the way the test should be
applied to the data:

\begin{itemize}
\tightlist
\item
  \texttt{pairwise} (default): to compare each subsample in a pairwise
  manner
\item
  \texttt{referential}: to compare each subsample to the first subsample
\item
  \texttt{sequential}: to compare each subsample to the following
  subsample
\item
  \texttt{all}: to compare all the subsamples together (like in analysis
  of variance)
\end{itemize}

It is also possible to input a list of pairs of \texttt{numeric} values
or \texttt{characters} matching the subsample names to create
personalised tests. Some other tests implemented in \texttt{dispRity}
such as the \texttt{dispRity::null.test} have a specific way they are
applied to the data and therefore ignore the \texttt{comparisons}
argument.

The \texttt{test} argument can be any statistical or non-statistical
test to apply to the disparity object. It can be a common statistical
test function (e.g. \texttt{stats::t.test}), a function implemented in
\texttt{dispRity} (e.g.~see \texttt{?null.test}) or any function defined
by the user.

This function also allows users to correct for Type I error inflation
(false positives) when using multiple comparisons via the
\texttt{correction} argument. This argument can be empty (no correction
applied) or can contain one of the corrections from the
\texttt{stats::p.adjust} function (see \texttt{?p.adjust}).

Note that the \texttt{test.dispRity} algorithm deals with some classical
test outputs (\texttt{h.test}, \texttt{lm} and \texttt{numeric} vector)
and summarises the test output. It is, however, possible to get the full
detailed output by using the options \texttt{details\ =\ TRUE}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## T-test to test for a difference in disparity between crown and stem mammals}
\KeywordTok{test.dispRity}\NormalTok{(disparity_crown_stem, }\DataTypeTok{test =}\NormalTok{ t.test)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [[1]]
##              statistic
## crown : stem  55.60162
## 
## [[2]]
##              parameter
## crown : stem  164.6932
## 
## [[3]]
##                    p.value
## crown : stem 1.201122e-108
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Performing the same test but with the detailed t.test output}
\KeywordTok{test.dispRity}\NormalTok{(disparity_crown_stem, }\DataTypeTok{test =}\NormalTok{ t.test, }\DataTypeTok{details =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## $`crown : stem`
## $`crown : stem`[[1]]
## 
##  Welch Two Sample t-test
## 
## data:  dots[[1L]][[1L]] and dots[[2L]][[1L]]
## t = 55.602, df = 164.69, p-value < 2.2e-16
## alternative hypothesis: true difference in means is not equal to 0
## 95 percent confidence interval:
##  0.2849688 0.3059530
## sample estimates:
## mean of x mean of y 
##  1.926163  1.630702
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Wilcoxon test applied to time sliced disparity with sequential comparisons, with Bonferroni correction}
\KeywordTok{test.dispRity}\NormalTok{(disparity_time_slices, }\DataTypeTok{test =}\NormalTok{ wilcox.test,}
              \DataTypeTok{comparisons =} \StringTok{"sequential"}\NormalTok{, }\DataTypeTok{correction =} \StringTok{"bonferroni"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [[1]]
##          statistic
## 120 : 80        13
## 80 : 40       2497
## 40 : 0         426
## 
## [[2]]
##               p.value
## 120 : 80 1.130227e-33
## 80 : 40  2.904335e-09
## 40 : 0   1.624453e-28
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring the overlap between distributions in the time bins (using the}
\NormalTok{## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)}
\KeywordTok{test.dispRity}\NormalTok{(disparity_time_bins, }\DataTypeTok{test =}\NormalTok{ bhatt.coeff)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                    bhatt.coeff
## 120 - 80 : 80 - 40   0.0000000
## 120 - 80 : 40 - 0    0.0000000
## 80 - 40 : 40 - 0     0.4195376
\end{verbatim}

It is also possible to apply some more \emph{complex} tests that have
their own output classes (like \texttt{stats::lm}).

The results can then be analysed as usual using the associated
\texttt{summary} S3 method:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Performing and linear model applied to the same data}
\NormalTok{(slice_lm <-}\StringTok{ }\KeywordTok{test.dispRity}\NormalTok{(disparity_time_slices, }\DataTypeTok{test =}\NormalTok{ lm,}
                            \DataTypeTok{comparisons =} \StringTok{"all"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Call:
## test(formula = data ~ subsamples, data = data)
## 
## Coefficients:
##   (Intercept)  subsamples120   subsamples40   subsamples80  
##        3.6571        -1.4046        -0.4481        -0.5863
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The output is a regular `lm` output}
\KeywordTok{class}\NormalTok{(slice_lm)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "lm"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## This output can be summarised using summary}
\KeywordTok{summary}\NormalTok{(slice_lm)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Call:
## test(formula = data ~ subsamples, data = data)
## 
## Residuals:
##      Min       1Q   Median       3Q      Max 
## -1.26522 -0.10976  0.01515  0.13109  0.58161 
## 
## Coefficients:
##               Estimate Std. Error t value Pr(>|t|)    
## (Intercept)    3.65712    0.02406  151.99   <2e-16 ***
## subsamples120 -1.40459    0.03403  -41.28   <2e-16 ***
## subsamples40  -0.44812    0.03403  -13.17   <2e-16 ***
## subsamples80  -0.58634    0.03403  -17.23   <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 0.2406 on 396 degrees of freedom
## Multiple R-squared:  0.818,  Adjusted R-squared:  0.8166 
## F-statistic: 593.1 on 3 and 396 DF,  p-value: < 2.2e-16
\end{verbatim}

Of course, due to the modular design of the package, tests can always be
made by the user (the same way disparity metrics can be user made). The
only condition is that the test can be applied to at least two
distributions. In practice, the \texttt{test.dispRity} function will
pass the calculated disparity data (distributions) to the provided
function in either pairs of distributions (if the \texttt{comparisons}
argument is set to \texttt{pairwise}, \texttt{referential} or
\texttt{sequential}) or a table containing all the distributions
(\texttt{comparisons\ =\ all}; this should be in the same format as data
passed to \texttt{lm} for example).

\hypertarget{disparity-as-a-distribution}{\section{Disparity as a
distribution}\label{disparity-as-a-distribution}}

Disparity is often regarded as a summary value of the position of the
all elements in the ordinated space. For example, the sum of variances,
the product of ranges or the median distance between the elements and
their centroid will summarise disparity as a single value. This value
can be pseudo-replicated (bootstrapped) to obtain a distribution of the
summary metric with estimated error. However, another way to perform
disparity analysis is to use the \emph{whole distribution} rather than
just a summary metric (e.g.~the variances or the ranges).

This is possible in the \texttt{dispRity} package by calculating
disparity as a dimension-level 2 metric only! Let's have a look using
our previous example of bootstrapped time slices but by measuring the
distances between each taxon and their centroid as disparity.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring disparity as a whole distribution}
\NormalTok{disparity_centroids <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(boot_time_slices, }\DataTypeTok{metric =}\NormalTok{ centroids)}
\end{Highlighting}
\end{Shaded}

The resulting disparity object is of dimension-level 2, so it can easily
be transformed into a dimension-level 1 object by, for example,
measuring the median distance of all these distributions:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring median disparity in each time slice}
\NormalTok{disparity_centroids_median <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(disparity_centroids, }\DataTypeTok{metric =}\NormalTok{ median)}
\end{Highlighting}
\end{Shaded}

And we can now compare the differences between these methods:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Summarising both disparity measurements:}
\NormalTok{## The distributions:}
\KeywordTok{summary}\NormalTok{(disparity_centroids)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n obs.median bs.median  2.5%   25%   75% 97.5%
## 1        120  5      1.508     1.324 0.536 1.069 1.562 1.868
## 2         80 19      1.790     1.702 1.393 1.597 1.815 1.943
## 3         40 15      1.719     1.703 1.387 1.581 1.857 2.076
## 4          0 10      1.910     1.807 1.366 1.700 1.971 2.085
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The summary of the distributions (as median)}
\KeywordTok{summary}\NormalTok{(disparity_centroids_median)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs bs.median  2.5%   25%   75% 97.5%
## 1        120  5 1.508     1.354 0.508 0.915 1.481 1.615
## 2         80 19 1.790     1.703 1.581 1.664 1.738 1.794
## 3         40 15 1.719     1.706 1.545 1.646 1.761 1.832
## 4          0 10 1.910     1.809 1.594 1.784 1.857 1.940
\end{verbatim}

We can see that the summary message for the distribution is slightly
different than before. Here \texttt{summary} also displays the observed
central tendency (i.e.~the central tendency of the measured
distributions). Note that, as expected, this central tendency is the
same in both metrics!

Another, maybe more intuitive way, to compare both approaches for
measuring disparity is to plot the distributions:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical parameters}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{, }\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}

\NormalTok{## Plotting both disparity measurements}
\KeywordTok{plot}\NormalTok{(disparity_centroids, }\DataTypeTok{ylab =} \StringTok{"Distribution of all the distances"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(disparity_centroids_median,}
     \DataTypeTok{ylab =} \StringTok{"Distribution of the medians of all the distances"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-57-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

We can then test for differences in the resulting distributions using
\texttt{test.dispRity} and the \texttt{bhatt.coeff} test as described
above.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Probability of overlap in the distribution of medians}
\KeywordTok{test.dispRity}\NormalTok{(disparity_centroids_median, }\DataTypeTok{test =}\NormalTok{ bhatt.coeff)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##          bhatt.coeff
## 120 : 80  0.15380832
## 120 : 40  0.23690683
## 120 : 0   0.09196152
## 80 : 40   0.92367172
## 80 : 0    0.46576289
## 40 : 0    0.65992821
\end{verbatim}

In this case, we are looking at the probability of overlap of the
distribution of median distances from centroids among each pair of time
slices. In other words, we are measuring whether the medians from each
bootstrap pseudo-replicate for each time slice overlap. But of course,
we might be interested in the actual distribution of the distances from
the centroid rather than simply their central tendencies. This can be
problematic depending on the research question asked since we are
effectively comparing non-independent medians distributions (because of
the pseudo-replication).

One solution, therefore, is to look at the full distribution:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Probability of overlap for the full distributions}
\KeywordTok{test.dispRity}\NormalTok{(disparity_centroids, }\DataTypeTok{test =}\NormalTok{ bhatt.coeff)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##          bhatt.coeff
## 120 : 80   0.6401042
## 120 : 40   0.6402923
## 120 : 0    0.5787247
## 80 : 40    0.9423481
## 80 : 0     0.8444910
## 40 : 0     0.9392977
\end{verbatim}

These results show the actual overlap among all the measured distances
from centroids concatenated across all the bootstraps. For example, when
comparing the slices 120 and 80, we are effectively comparing the 5
\(\times\) 100 distances (the distances of the five elements in slice
120 bootstrapped 100 times) to the 19 \(\times\) 100 distances from
slice 80. However, this can also be problematic for some specific tests
since the \emph{n} \(\times\) 100 distances are also pseudo-replicates
and thus are still not independent.

A second solution is to compare the distributions to each other
\emph{for each replicate}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Boostrapped probability of overlap for the full distributions}
\KeywordTok{test.dispRity}\NormalTok{(disparity_centroids, }\DataTypeTok{test =}\NormalTok{ bhatt.coeff, }\DataTypeTok{concatenate =} \OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##          bhatt.coeff      2.5%       25%       75%     97.5%
## 120 : 80   0.2534144 0.0000000 0.1450953 0.3554093 0.5615906
## 120 : 40   0.2726082 0.0000000 0.1632993 0.3942394 0.6435771
## 120 : 0    0.2172550 0.0000000 0.0000000 0.3464102 0.5913591
## 80 : 40    0.6016956 0.3097953 0.4779726 0.7208207 0.8306227
## 80 : 0     0.4770117 0.1382736 0.3964076 0.5772107 0.7169318
## 40 : 0     0.5327868 0.1724549 0.3887699 0.6761873 0.8781747
\end{verbatim}

These results show the median overlap among pairs of distributions in
the first column (\texttt{bhatt.coeff}) and then the distribution of
these overlaps among each pair of bootstraps. In other words, when two
distributions are compared, they are now compared for each bootstrap
pseudo-replicate, thus effectively creating a distribution of
probabilities of overlap. For example, when comparing the slices 120 and
80, we have a mean probability of overlap of 0.28 and a probability
between 0.18 and 0.43 in 50\% of the pseudo-replicates. Note that the
quantiles and central tendencies can be modified via the
\texttt{conc.quantiles} option.

\section{Ecological studies of
disparity}\label{ecological-studies-of-disparity}

Disparity analyses are not restricted to palaeobiology analysis. This
package is designed for dealing with ordinated matrices, no matter which
data was used in the ordination. Here is an short example of disparity
analysis using ecological data from Deirdre McClean (unpubl.).

The dataset contains an ordinated matrix of 20 dimensions (columns) for
40 elements (rows). The elements are different experimental plots with
discrete variations of nutrient enrichment, depth, and freshwater
benthic invertebrates.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading the data}
\KeywordTok{data}\NormalTok{(McClean_data)}

\NormalTok{## Ordinated matrix}
\NormalTok{ordinated_matrix <-}\StringTok{ }\NormalTok{McClean_data}\OperatorTok{$}\NormalTok{ordination}

\NormalTok{## Treatments and depths}
\NormalTok{treatments <-}\StringTok{ }\NormalTok{McClean_data}\OperatorTok{$}\NormalTok{treatment}
\NormalTok{depth <-}\StringTok{ }\NormalTok{McClean_data}\OperatorTok{$}\NormalTok{depth}
\end{Highlighting}
\end{Shaded}

\subsection{\texorpdfstring{Classic ecological ``disparity''
analysis}{Classic ecological disparity analysis}}\label{classic-ecological-disparity-analysis}

A classical way to represent this ordinated data would be to use two
dimensional plots to look at how the different experimental plots differ
in their values for nutrient enrichment, depth, and freshwater benthic
invertebrates.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Setting the colors that will represent the treatments}
\NormalTok{cols <-}\StringTok{ }\KeywordTok{sub}\NormalTok{(}\StringTok{"a"}\NormalTok{, }\StringTok{"red"}\NormalTok{, treatments)}
\NormalTok{cols <-}\StringTok{ }\KeywordTok{sub}\NormalTok{(}\StringTok{"b"}\NormalTok{, }\StringTok{"blue"}\NormalTok{, cols)}
\NormalTok{## Setting the symbols that will represent the depth}
\NormalTok{pchs <-}\StringTok{ }\KeywordTok{sub}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{16}\NormalTok{, depth)}
\NormalTok{pchs <-}\StringTok{ }\KeywordTok{as.numeric}\NormalTok{(}\KeywordTok{sub}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{17}\NormalTok{, pchs))}

\NormalTok{## Graphical option}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}
\NormalTok{## A classic 2D ordination plot}
\KeywordTok{plot}\NormalTok{(ordinated_matrix[, }\DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{], }\DataTypeTok{col =}\NormalTok{ cols, }\DataTypeTok{pch =}\NormalTok{ pchs,}
     \DataTypeTok{xlab =} \StringTok{"PC 1"}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{"PC 2"}\NormalTok{,}
     \DataTypeTok{xlim =} \KeywordTok{range}\NormalTok{(ordinated_matrix[, }\DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{]) }\OperatorTok{+}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{100}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-62-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

This shows the distribution of the experimental plots along the two
first axis of variation of the ordinated distance matrix (i.e.~the first
two dimensions). At a first glance, it seems difficult to see a clear
effect of the treatments (blue or red) or the depth (rounds or
triangles) as the different experimental plots with the same parameters
don't seem to cluster together. However, this may be because this plot
ignores the 18 other dimensions of the ordination! Additionally, these
two represented dimensions do not represent a biological reality
\emph{per se}; i.e.~the values on the first dimension do not represent a
continuous trait (e.g.~depth), instead they just represent the
ordinations of correlations between the data and some factors.

Therefore, we might want to approach this problem without getting stuck
in only two dimensions and consider the whole dataset as a
\emph{n}-dimensional object.

\subsection{\texorpdfstring{Multidimensional analysis with
\texttt{dispRity}}{Multidimensional analysis with dispRity}}\label{multidimensional-analysis-with-disprity}

\subsubsection{Splitting the data with respect to the different
factors}\label{splitting-the-data-with-respect-to-the-different-factors}

The first step is to create different subsamples that represent
subsamples of the ordinated space (i.e.~sub-regions within the
\emph{n}-dimensional object). Each of these subsamples will contain a
certain number of elements (i.e.~a subset of the 40 experimental field
plots) that have some attributes in common. In our example, we are going
to group the elements according to their depth and treatment.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating a table that contain the elements and their attributes}
\NormalTok{factors <-}\StringTok{ }\KeywordTok{as.data.frame}\NormalTok{(}\KeywordTok{matrix}\NormalTok{(}\DataTypeTok{data =} \KeywordTok{c}\NormalTok{(treatments, depth),}
          \DataTypeTok{nrow =} \KeywordTok{nrow}\NormalTok{(ordinated_matrix), }\DataTypeTok{ncol =} \DecValTok{2}\NormalTok{, }\DataTypeTok{byrow =} \OtherTok{FALSE}\NormalTok{,}
          \DataTypeTok{dimnames =} \KeywordTok{list}\NormalTok{(}\KeywordTok{rownames}\NormalTok{(ordinated_matrix))))}
\KeywordTok{names}\NormalTok{(factors) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"Treat"}\NormalTok{, }\StringTok{"Depth"}\NormalTok{)}
\KeywordTok{head}\NormalTok{(factors)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##     Treat Depth
## 1       a     1
## 1.1     a     2
## 2       b     1
## 2.1     b     2
## 3       a     1
## 3.1     a     1
\end{verbatim}

Second, let's split the data according by depth and treatment to create
the subsamples of the ordinated space by using the
\texttt{custom.subsamples} function:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Splitting the ordinated space into four subsamples}
\NormalTok{(customised_subsamples <-}\StringTok{ }\KeywordTok{custom.subsamples}\NormalTok{(ordinated_matrix, factors))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 4 customised subsamples for 40 elements:
##     Treat.a, Treat.b, Depth.1, Depth.2.
\end{verbatim}

\subsubsection{Calculating disparity}\label{calculating-disparity}

In this example, we are going to define disparity as the n-dimensional
ellipsoid volume (see \texttt{?hyper.volume}).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity as the ellipsoid volume of each group}
\NormalTok{disparity <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(customised_subsamples, }\DataTypeTok{metric =}\NormalTok{ diagonal)}
\KeywordTok{summary}\NormalTok{(disparity)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs
## 1    Treat.a 21 36.89
## 2    Treat.b 19 40.99
## 3    Depth.1 23 36.38
## 4    Depth.2 17 40.74
\end{verbatim}

We can also bootstrap the data to test the robustness of the measured
disparity to outliers. Also, as we can see, each subsample has different
numbers of elements. It might also be interesting to rarefy the data to
have only subsamples with the same number of elements.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Bootstrapping the subsamples and using rarefaction}
\NormalTok{## (i.e. only re-sampling 17 elements each time)}
\NormalTok{bootstrapped_data <-}\StringTok{ }\KeywordTok{boot.matrix}\NormalTok{(customised_subsamples, }\DataTypeTok{bootstraps =} \DecValTok{100}\NormalTok{,}
                                 \DataTypeTok{rarefaction =} \DecValTok{17}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We can now run a more robust disparity analysis using the bootstrapped
data:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the bootstrapped disparity}
\NormalTok{disparity <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(bootstrapped_data, }\DataTypeTok{metric =}\NormalTok{ diagonal)}
\end{Highlighting}
\end{Shaded}

\subsubsection{Summarising and displaying the
results}\label{summarising-and-displaying-the-results}

Now let's summarise this data:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Summary table of the data}
\KeywordTok{summary}\NormalTok{(disparity)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs bs.median  2.5%   25%   75% 97.5%
## 1    Treat.a 21 36.89     35.01 31.32 32.88 35.44 36.11
## 2    Treat.a 17    NA     33.69 30.38 31.91 34.97 35.98
## 3    Treat.b 19 40.99     38.19 34.27 36.75 39.25 40.12
## 4    Treat.b 17    NA     37.52 33.38 36.08 38.85 39.80
## 5    Depth.1 23 36.38     34.44 31.86 33.63 35.01 35.52
## 6    Depth.1 17    NA     33.38 30.56 32.28 34.09 35.18
## 7    Depth.2 17 40.74     37.30 34.26 36.35 38.62 39.74
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =}\NormalTok{ (}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{)), }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting the bootstrapped disparity}
\KeywordTok{plot}\NormalTok{(disparity, }\DataTypeTok{main =} \StringTok{"Bootstrapped"}\NormalTok{, }\DataTypeTok{las =} \DecValTok{2}\NormalTok{, }\DataTypeTok{xlab =} \StringTok{""}\NormalTok{)}

\NormalTok{## Plotting the rarefied disparity}
\KeywordTok{plot}\NormalTok{(disparity, }\DataTypeTok{rarefaction =} \DecValTok{17}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Rarefied"}\NormalTok{, }\DataTypeTok{las =} \DecValTok{2}\NormalTok{, }\DataTypeTok{xlab =} \StringTok{""}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-68-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

As we can see, there seems to be no strong effect of the number of
experimental plots in each subsample (i.e.~the rarefied plot is really
similar to the bootstrapped plot) which is a good thing!

\subsubsection{Testing our hypotheses}\label{testing-our-hypotheses}

Finally, we can test our hypothesis (whether the water treatment at
certain depths alters invertebrate communities and composition in
natural habitats) by using the \texttt{test.dispRity} function.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Testing the effect of treatment and depth on the bootstrapped data}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{test.dispRity}\NormalTok{(disparity, }\DataTypeTok{test =}\NormalTok{ aov, }\DataTypeTok{comparisons =} \StringTok{"all"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      Length Class Mode
## [1,] 13     aov   list
\end{verbatim}

\section{Simulating discrete morphological
data}\label{simulating-discrete-morphological-data}

The \texttt{dispRity} package also allows users to simulate discrete
morphological data matrices. In brief, the function \texttt{sim.morpho}
takes a phylogenetic tree, the number of required characters, the
evolutionary model, and a function from which to draw the rates. The
package also contains a function for quickly checking the matrix's
phylogenetic signal (as defined in systematics not phylogenetic
comparative methods) using parsimony. To understand these methods please
refer to the phylogenetics literature.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{3}\NormalTok{)}
\NormalTok{## Simulating a starting tree with 15 taxa as a random coalescent tree}
\NormalTok{my_tree <-}\StringTok{ }\KeywordTok{rcoal}\NormalTok{(}\DecValTok{15}\NormalTok{)}

\NormalTok{## Generating a matrix with 100 characters (85% binary and 15% three state) and}
\NormalTok{## an equal rates model with a gamma rate distribution (0.5, 1) with no }
\NormalTok{## invariant characters.}
\NormalTok{my_matrix <-}\StringTok{ }\KeywordTok{sim.morpho}\NormalTok{(}\DataTypeTok{tree =}\NormalTok{ my_tree, }\DataTypeTok{characters =} \DecValTok{100}\NormalTok{, }\DataTypeTok{states =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.85}\NormalTok{,}
    \FloatTok{0.15}\NormalTok{), }\DataTypeTok{rates =} \KeywordTok{c}\NormalTok{(rgamma, }\FloatTok{0.5}\NormalTok{, }\DecValTok{1}\NormalTok{), }\DataTypeTok{invariant =} \OtherTok{FALSE}\NormalTok{)}

\NormalTok{## The first few lines of the matrix}
\NormalTok{my_matrix[}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##     [,1] [,2] [,3] [,4] [,5] [,6] [,7] [,8] [,9] [,10]
## t15 "1"  "1"  "0"  "1"  "1"  "2"  "1"  "1"  "0"  "0"  
## t12 "1"  "1"  "0"  "1"  "1"  "2"  "1"  "1"  "0"  "0"  
## t14 "1"  "1"  "0"  "1"  "1"  "2"  "1"  "1"  "0"  "0"  
## t6  "1"  "1"  "0"  "1"  "1"  "2"  "1"  "1"  "0"  "0"  
## t3  "0"  "0"  "1"  "0"  "1"  "0"  "0"  "2"  "1"  "0"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Checking the matrix properties with a quick Maximum Parsimony tree search}
\KeywordTok{check.morpho}\NormalTok{(my_matrix, my_tree)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                                     
## Maximum parsimony        139.0000000
## Consistency index          0.7625899
## Retention index            0.8881356
## Robinson-Foulds distance   0.0000000
\end{verbatim}

Note that this example produces a tree with a great consistency index
and an identical topology to the random coalescent tree! Nearly too good
to be true\ldots{}

\subsection{A more detailed
description}\label{a-more-detailed-description}

\texttt{sim.morpho} is really flexible and takes many different
arguments to allow users to simulate realistic matrices. It has three
implemented models: \texttt{"ER"} for Equal Rates (the M\emph{k} model);
\texttt{"HKY"}, the molecular HKY model but transforms pyrines in 0's
and pyrimidines in 1's; or the \texttt{"mixed"} model that randomly uses
an \texttt{"ER"} and/or \texttt{"HKY"} for the binary characters and
\texttt{"ER"} for the multistate (\textgreater{}2) characters. Both
models take specific distributions for their rate or substitution
models. These distributions should be passed to these arguments in the
format of \texttt{c(sampler\_function,\ distribution\_parameters)} where
the the sampler function is a the random generation function of that
distribution (e.g. \texttt{rnorm}, \texttt{runif}, etc\ldots{}) and the
parameters are any parameters to be passed to this function.

\texttt{check.morpho} runs a quick Maximum Parsimony tree search using
the \texttt{phangorn} parsimony algorithm. It quickly calculates the
parsimony score, the consistency and retention indices and, if a tree is
provided (e.g.~the tree used to generate the matrix) it calculates the
Robinson-Foulds distance between the most parsimonious tree and the
provided tree to determine how different they are.

\subsection{Parameters for a realistic(ish)
matrix}\label{parameters-for-a-realisticish-matrix}

There are many parameters that can create a
\texttt{realistic\textquotesingle{}\textquotesingle{}\ matrix\ (i.e.\ not\ too\ different\ from\ the\ input\ tree\ with\ a\ consistency\ and\ retention\ index\ close\ to\ what\ is\ seen\ in\ the\ literature)\ but\ because\ of\ the\ randomness\ of\ the\ matrix\ generation\ not\ all\ end\ up\ creating\ "good"\ matrices.\ The\ following\ parameters\ however,\ seem\ to\ generate\ fairly}realistic''
matrices with a starting coalescent tree, equal rates model with 0.85
binary characters and 0.15 three state characters, a gamma distribution
with a shape parameter (\(\alpha\)) of 5 and no scaling (\(\beta\) = 1)
with a rate of 100.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
\NormalTok{## tree}
\NormalTok{my_tree <-}\StringTok{ }\KeywordTok{rcoal}\NormalTok{(}\DecValTok{15}\NormalTok{)}
\NormalTok{## matrix}
\NormalTok{morpho_mat <-}\StringTok{ }\KeywordTok{sim.morpho}\NormalTok{(my_tree, }\DataTypeTok{characters =} \DecValTok{100}\NormalTok{, }\DataTypeTok{model =} \StringTok{"ER"}\NormalTok{,}
    \DataTypeTok{rates =} \KeywordTok{c}\NormalTok{(rgamma, }\DataTypeTok{rate =} \DecValTok{100}\NormalTok{, }\DataTypeTok{shape =} \DecValTok{5}\NormalTok{), }\DataTypeTok{invariant =} \OtherTok{FALSE}\NormalTok{)}
\KeywordTok{check.morpho}\NormalTok{(morpho_mat, my_tree)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                                     
## Maximum parsimony        104.0000000
## Consistency index          0.0000000
## Retention index            0.7886179
## Robinson-Foulds distance   0.0000000
\end{verbatim}

\hypertarget{space.maker}{\subsection{\texorpdfstring{\texttt{space.maker}}{space.maker}}\label{space.maker}}

Another way to simulate data is to directly simulate an ordinated space
with the \texttt{space.maker} function. This function allows users to
simulate multidimensional spaces with a certain number of properties. It
takes as arguments the number of elements (data points)~and dimensions
to create the space but also permits more fine tuning in the data
simulation: it is possible to give a specific distribution to each
dimension, provide a correlation matrix to link the dimensions or even
provide an \emph{a priori} distribution of the variance for each
distributions!

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =}\NormalTok{ (}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{)), }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}
\NormalTok{## Visualising 3D spaces}
\KeywordTok{require}\NormalTok{(scatterplot3d)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Loading required package: scatterplot3d
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## A cube space}
\KeywordTok{scatterplot3d}\NormalTok{(}\KeywordTok{space.maker}\NormalTok{(}\DecValTok{2500}\NormalTok{, }\DecValTok{3}\NormalTok{, runif), }\DataTypeTok{pch =} \DecValTok{20}\NormalTok{,}
              \DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{, }\DataTypeTok{zlab =} \StringTok{""}\NormalTok{,}
              \DataTypeTok{main =} \StringTok{"Uniform cube space"}\NormalTok{)}

\NormalTok{## A plane space}
\KeywordTok{scatterplot3d}\NormalTok{(}\KeywordTok{space.maker}\NormalTok{(}\DecValTok{2500}\NormalTok{, }\DecValTok{3}\NormalTok{, }\KeywordTok{c}\NormalTok{(runif, runif, runif),}
              \DataTypeTok{arguments =} \KeywordTok{list}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{min =} \DecValTok{0}\NormalTok{, }\DataTypeTok{max =} \DecValTok{0}\NormalTok{), }\OtherTok{NULL}\NormalTok{, }\OtherTok{NULL}\NormalTok{)), }\DataTypeTok{pch =} \DecValTok{20}\NormalTok{,}
              \DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{, }\DataTypeTok{zlab =} \StringTok{""}\NormalTok{,}
              \DataTypeTok{main =} \StringTok{"Uniform plane space"}\NormalTok{)}

\NormalTok{## An ellipsoid space (=a spheric space with correlation)}
\NormalTok{cor_matrix <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\KeywordTok{cbind}\NormalTok{(}\DecValTok{1}\NormalTok{, }\FloatTok{0.8}\NormalTok{, }\FloatTok{0.2}\NormalTok{, }\FloatTok{0.8}\NormalTok{, }\DecValTok{1}\NormalTok{, }\FloatTok{0.7}\NormalTok{, }\FloatTok{0.2}\NormalTok{, }\FloatTok{0.7}\NormalTok{, }\DecValTok{1}\NormalTok{), }\DataTypeTok{nrow =} \DecValTok{3}\NormalTok{)}
\KeywordTok{scatterplot3d}\NormalTok{(}\KeywordTok{space.maker}\NormalTok{(}\DecValTok{2500}\NormalTok{, }\DecValTok{3}\NormalTok{, rnorm, }\DataTypeTok{cor.matrix =}\NormalTok{ cor_matrix), }\DataTypeTok{pch =} \DecValTok{20}\NormalTok{,}
              \DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{, }\DataTypeTok{zlab =} \StringTok{""}\NormalTok{,}
              \DataTypeTok{main =} \StringTok{"Normal ellipsoid space"}\NormalTok{)}

\NormalTok{## A cylindrical space with a decreasing variance per axis}
\KeywordTok{scatterplot3d}\NormalTok{(}\KeywordTok{space.maker}\NormalTok{(}\DecValTok{2500}\NormalTok{, }\DecValTok{3}\NormalTok{, }\KeywordTok{c}\NormalTok{(rnorm, rnorm, runif),}
              \DataTypeTok{scree =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.7}\NormalTok{, }\FloatTok{0.2}\NormalTok{, }\FloatTok{0.1}\NormalTok{)), }\DataTypeTok{pch =} \DecValTok{20}\NormalTok{,}
              \DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{, }\DataTypeTok{zlab =} \StringTok{""}\NormalTok{,}
              \DataTypeTok{main =} \StringTok{"Normal cylindrical space"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-72-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Resetting the graphic parameters}
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

\chapter{The guts of the dispRity
package}\label{the-guts-of-the-disprity-package}

\section{\texorpdfstring{Manipulating \texttt{dispRity}
objects}{Manipulating dispRity objects}}\label{manipulating-disprity-objects}

Disparity analysis involves a lot of manipulation of many matrices
(especially when bootstrapping) which can be impractical to visualise
and will quickly overwhelm your \texttt{R} console. Even the simple Beck
and Lee 2014 example above produces an object with \textgreater{} 72
lines of lists of lists of matrices!

Therefore \texttt{dispRity} uses a specific class of object called a
\texttt{dispRity} object. These objects allow users to use S3 method
functions such as \texttt{summary.dispRity}, \texttt{plot.dispRity} and
\texttt{print.dispRity}. \texttt{dispRity} also contains various utility
functions that manipulate the \texttt{dispRity} object (e.g.
\texttt{sort.dispRity}, \texttt{extract.dispRity} see the full list in
the next section). These functions modify the \texttt{dispRity} object
without having to delve into its complex structure! The full structure
of a \texttt{dispRity} object is detailed
\href{https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md}{here}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading the example data}
\KeywordTok{data}\NormalTok{(disparity)}

\NormalTok{## What is the class of the median_centroids object?}
\KeywordTok{class}\NormalTok{(disparity)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "dispRity"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## What does the object contain?}
\KeywordTok{names}\NormalTok{(disparity)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "matrix"     "call"       "subsamples" "disparity"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Summarising it using the S3 method print.dispRity}
\NormalTok{disparity}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 7 continuous (acctran) time subsamples for 99 elements with 97 dimensions:
##      90, 80, 70, 60, 50 ...
## Data was bootstrapped 100 times (method:"full") and rarefied to 20, 15, 10, 5 elements.
## Disparity was calculated as: c(median, centroids).
\end{verbatim}

Note that it is always possible to recall the full object using the
argument \texttt{all\ =\ TRUE} in \texttt{print.dispRity}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Display the full object}
\KeywordTok{print}\NormalTok{(disparity, }\DataTypeTok{all =} \OtherTok{TRUE}\NormalTok{)}
\NormalTok{## This is more nearly ~ 5000 lines on my 13 inch laptop screen!}
\end{Highlighting}
\end{Shaded}

\section{\texorpdfstring{\texttt{dispRity}
utilities}{dispRity utilities}}\label{disprity-utilities}

The package also provides some utility functions to facilitate
multidimensional analysis.

\subsection{\texorpdfstring{\texttt{dispRity} object utilities
}{dispRity object utilities }}\label{disprity-object-utilities}

The first set of utilities are functions for manipulating
\texttt{dispRity} objects:

\subsubsection{\texorpdfstring{\texttt{make.dispRity}}{make.dispRity}}\label{make.disprity}

This function creates empty \texttt{dispRity} objects.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating an empty dispRity object}
\KeywordTok{make.dispRity}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Empty dispRity object.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating an "empty" dispRity object with a matrix}
\NormalTok{(disparity_obj <-}\StringTok{ }\KeywordTok{make.dispRity}\NormalTok{(}\KeywordTok{matrix}\NormalTok{(}\KeywordTok{rnorm}\NormalTok{(}\DecValTok{20}\NormalTok{), }\DecValTok{5}\NormalTok{, }\DecValTok{4}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Empty dispRity object.
\end{verbatim}

\subsubsection{\texorpdfstring{\texttt{fill.dispRity}}{fill.dispRity}}\label{fill.disprity}

This function initialises a \texttt{dispRity} object and generates its
call properties.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The dispRity object's call is indeed empty}
\NormalTok{disparity_obj}\OperatorTok{$}\NormalTok{call}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## list()
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Filling an empty disparity object (that needs to contain at least a matrix)}
\NormalTok{(disparity_obj <-}\StringTok{ }\KeywordTok{fill.dispRity}\NormalTok{(disparity_obj))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 elements with 4 dimensions.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The dipRity object has now the correct minimal attributes}
\NormalTok{disparity_obj}\OperatorTok{$}\NormalTok{call}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## $dimensions
## [1] 4
\end{verbatim}

\subsubsection{\texorpdfstring{\texttt{matrix.dispRity}}{matrix.dispRity}}\label{matrix.disprity}

This function extracts a specific matrix from a disparity object. The
matrix can be one of the bootstrapped matrices or/and a rarefied matrix.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Extracting the matrix containing the coordinates of the elements at time 50}
\KeywordTok{str}\NormalTok{(}\KeywordTok{matrix.dispRity}\NormalTok{(disparity, }\StringTok{"50"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  num [1:18, 1:97] -0.1038 0.2844 0.2848 0.0927 0.1619 ...
##  - attr(*, "dimnames")=List of 2
##   ..$ : chr [1:18] "Leptictis" "Dasypodidae" "n24" "Potamogalinae" ...
##   ..$ : NULL
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Extracting the 3rd bootstrapped matrix with the 2nd rarefaction level}
\NormalTok{## (15 elements) from the second group (80 Mya)}
\KeywordTok{str}\NormalTok{(}\KeywordTok{matrix.dispRity}\NormalTok{(disparity, }\DataTypeTok{subsamples =} \DecValTok{1}\NormalTok{, }\DataTypeTok{bootstrap =} \DecValTok{3}\NormalTok{, }\DataTypeTok{rarefaction =} \DecValTok{2}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  num [1:15, 1:97] -0.7161 0.3496 -0.573 -0.0445 -0.1427 ...
##  - attr(*, "dimnames")=List of 2
##   ..$ : chr [1:15] "n7" "n34" "Maelestes" "n20" ...
##   ..$ : NULL
\end{verbatim}

\subsubsection{\texorpdfstring{\texttt{get.subsamples.dispRity}}{get.subsamples.dispRity}}\label{get.subsamples.disprity}

This function creates a dispRity object that contains only elements from
one specific subsamples.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Extracting all the data for the crown mammals}
\NormalTok{(crown_mammals <-}\StringTok{ }\KeywordTok{get.subsamples.dispRity}\NormalTok{(disp_crown_stemBS, }\StringTok{"Group.crown"}\NormalTok{))}

\NormalTok{## The object keeps the properties of the parent object but is composed of only one subsamples}
\KeywordTok{length}\NormalTok{(crown_mammals}\OperatorTok{$}\NormalTok{subsamples)}
\end{Highlighting}
\end{Shaded}

\subsubsection{\texorpdfstring{\texttt{extract.dispRity}}{extract.dispRity}}\label{extract.disprity}

This function extracts the calculated disparity values of a specific
matrix.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Extracting the observed disparity (default)}
\KeywordTok{extract.dispRity}\NormalTok{(disparity)}

\NormalTok{## Extracting the disparity from the bootstrapped values from the}
\NormalTok{## 10th rarefaction level from the second subsamples (80 Mya)}
\KeywordTok{extract.dispRity}\NormalTok{(disparity, }\DataTypeTok{observed =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{subsamples =} \DecValTok{2}\NormalTok{, }\DataTypeTok{rarefaction =} \DecValTok{10}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\subsubsection{\texorpdfstring{\texttt{scale.dispRity}}{scale.dispRity}}\label{scale.disprity}

This is the S3 method of \texttt{scale} (scaling and/or centring) that
can be applied to the disparity data of a \texttt{dispRity} object.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Getting the disparity values of the time subsamples}
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(disparity))}

\NormalTok{## Scaling the same disparity values}
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(}\KeywordTok{scale}\NormalTok{(disparity, }\DataTypeTok{scale =} \OtherTok{TRUE}\NormalTok{)))}

\NormalTok{## Scaling and centering:}
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(}\KeywordTok{scale}\NormalTok{(disparity, }\DataTypeTok{scale =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{center =} \OtherTok{TRUE}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\subsubsection{\texorpdfstring{\texttt{sort.dispRity}}{sort.dispRity}}\label{sort.disprity}

This is the S3 method of \texttt{sort} for sorting the subsamples
alphabetically (default) or following a specific pattern.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Sorting the disparity subsamples in inverse alphabetic order}
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(}\KeywordTok{sort}\NormalTok{(disparity, }\DataTypeTok{decreasing =} \OtherTok{TRUE}\NormalTok{)))}

\NormalTok{## Customised sorting}
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(}\KeywordTok{sort}\NormalTok{(disparity, }\DataTypeTok{sort =} \KeywordTok{c}\NormalTok{(}\DecValTok{7}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{3}\NormalTok{, }\DecValTok{4}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{))))}
\end{Highlighting}
\end{Shaded}

\chapter{Palaeobiology demo: disparity-through-time and within
groups}\label{palaeobiology-demo-disparity-through-time-and-within-groups}

This demo aims to give quick overview of the \texttt{dispRity} package
(v.0.3) for palaeobiology analyses of disparity, including disparity
through time analyses. Please refer to my
\href{https://github.com/TGuillerme/dispRity}{GitHub page} for other
vignettes, namely the
\href{http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html}{\texttt{dispRity}
manual} that explains the functions in more detail.

In brief, this package allows users to easily perform
\textbf{disparity-through-time} analysis. This type of analysis often
starts by ordinating \textbf{morphometric} or \textbf{cladistic} data
into a multidimensional object hereafter called the
\textbf{morphospace}. One might be interested in studying how the
occupancy of the morphospace has changed through time by measuring a
summary metric of this morphospace through time, called
\textbf{disparity}. Disparity can then be seen as a value or a
distribution of values summarising the diversity of morphologies.

This demo showcases a typical disparity-through-time analysis: we are
going to test whether the disparity changed through time in a subset of
eutherian mammals from the last 100 million years.

\section{Before starting}\label{before-starting}

\subsection{\texorpdfstring{Installing
\texttt{dispRity}}{Installing dispRity}}\label{installing-disprity}

You can install this package easily if you are using the latest version
of \texttt{R} and \texttt{devtools}.

\begin{Shaded}
\begin{Highlighting}[]
\ControlFlowTok{if}\NormalTok{(}\OperatorTok{!}\KeywordTok{require}\NormalTok{(devtools)) }\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}
\KeywordTok{install_github}\NormalTok{(}\StringTok{"TGuillerme/dispRity"}\NormalTok{, }\DataTypeTok{ref =} \StringTok{"release"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(dispRity)}
\end{Highlighting}
\end{Shaded}

\subsection{The morphospace}\label{the-morphospace}

In this example, we are going to use a subset of the data from
\citep{beckancient2014}. This dataset contains an ordinated matrix of 50
discrete characters from mammals (\texttt{BeckLee\_mat50}), another
matrix of the same 50 mammals and the estimated discrete data characters
of their descendants (thus 50 + 49 rows, \texttt{BeckLee\_mat99}), a
dataframe containing the ages of each taxon in the dataset
(\texttt{BeckLee\_ages}) and finally a phylogenetic tree with the
relationships among the 50 mammals (\texttt{BeckLee\_tree}). The
ordinated matrix will represent our full morphospace, i.e.~all the
mammalian morphologies that ever existed through time (for this
dataset).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading demo and the package data}
\KeywordTok{library}\NormalTok{(dispRity)}

\NormalTok{## Setting the random seed for repeatability}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}

\NormalTok{## Loading the ordinated matrix/morphospace:}
\KeywordTok{data}\NormalTok{(BeckLee_mat50)}
\KeywordTok{head}\NormalTok{(BeckLee_mat50[,}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{])}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                    [,1]          [,2]        [,3]       [,4]      [,5]
## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{dim}\NormalTok{(BeckLee_mat50)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 50 48
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The morphospace contains 50 taxa and has 48 dimensions (or axes)}

\NormalTok{## Showing a list of first and last occurrences data for some fossils}
\KeywordTok{data}\NormalTok{(BeckLee_ages)}
\KeywordTok{head}\NormalTok{(BeckLee_ages)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##             FAD  LAD
## Adapis     37.2 36.8
## Asioryctes 83.6 72.1
## Leptictis  33.9 33.3
## Miacis     49.0 46.7
## Mimotona   61.6 59.2
## Notharctus 50.2 47.0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Plotting a phylogeny}
\KeywordTok{data}\NormalTok{(BeckLee_tree)}
\KeywordTok{plot}\NormalTok{(BeckLee_tree, }\DataTypeTok{cex =} \FloatTok{0.7}\NormalTok{)}
\KeywordTok{axisPhylo}\NormalTok{(}\DataTypeTok{root =} \DecValTok{140}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-83-1.pdf}

\begin{quote}
You can have an even nicer looking tree if you use the \texttt{strap}
package!
\end{quote}

\begin{Shaded}
\begin{Highlighting}[]
\ControlFlowTok{if}\NormalTok{(}\OperatorTok{!}\KeywordTok{require}\NormalTok{(strap)) }\KeywordTok{install.packages}\NormalTok{(}\StringTok{"strap"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(strap)}
\KeywordTok{geoscalePhylo}\NormalTok{(BeckLee_tree, }\DataTypeTok{cex.tip =} \FloatTok{0.7}\NormalTok{, }\DataTypeTok{cex.ts =} \FloatTok{0.6}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\section{A disparity-through-time
analysis}\label{a-disparity-through-time-analysis}

\subsection{Splitting the morphospace through
time}\label{splitting-the-morphospace-through-time}

One of the crucial steps in disparity-through-time analysis is to split
the full morphospace into smaller time subsamples that contain the total
number of morphologies at certain points in time (time-slicing) or
during certain periods in time (time-binning). Basically, the full
morphospace represents the total number of morphologies across all time
and will be greater than any of the time subsamples of the morphospace.

The \texttt{dispRity} package provides a \texttt{time.subsamples}
function that allows users to split the morphospace into time slices
(using \texttt{method\ =\ continuous}) or into time bins (using
\texttt{method\ =\ discrete}). In this example, we are going to split
the morphospace into five equal time bins of 20 million years long from
100 million years ago to the present. We will also provide to the
function a table containing the first and last occurrences dates for
some fossils to take into account that some fossils might occur in
several of our different time bins.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating the vector of time bins ages}
\NormalTok{(time_bins <-}\StringTok{ }\KeywordTok{rev}\NormalTok{(}\KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{0}\NormalTok{, }\DataTypeTok{to =} \DecValTok{100}\NormalTok{, }\DataTypeTok{by =} \DecValTok{20}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 100  80  60  40  20   0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Splitting the morphospace using the time.subsamples function}
\NormalTok{(binned_morphospace <-}\StringTok{ }\KeywordTok{time.subsamples}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat50, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree,}
    \DataTypeTok{method =} \StringTok{"discrete"}\NormalTok{, }\DataTypeTok{time =}\NormalTok{ time_bins, }\DataTypeTok{inc.nodes =} \OtherTok{FALSE}\NormalTok{,}
    \DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 discrete time subsamples for 50 elements:
##     100 - 80, 80 - 60, 60 - 40, 40 - 20, 20 - 0.
\end{verbatim}

The output object is a \texttt{dispRity} object. For details about this
object class, please refer to the
\href{http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html}{\texttt{dispRity}
manual}. In brief, however, \texttt{dispRity} objects are lists of
different elements (i.e.~disparity results, morphospace time subsamples,
morphospace attributes, etc.) that display only a summary of the object
when calling the object to avoiding filling the \texttt{R} console with
superfluous output.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Printing the class of the object}
\KeywordTok{class}\NormalTok{(binned_morphospace)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "dispRity"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Printing the content of the object}
\KeywordTok{str}\NormalTok{(binned_morphospace)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## List of 3
##  $ matrix    : num [1:50, 1:48] -0.532 -0.409 -0.692 -0.68 -0.739 ...
##   ..- attr(*, "dimnames")=List of 2
##   .. ..$ : chr [1:50] "Cimolestes" "Maelestes" "Batodon" "Bulaklestes" ...
##   .. ..$ : NULL
##  $ call      :List of 1
##   ..$ subsamples: chr "discrete"
##  $ subsamples:List of 5
##   ..$ 100 - 80:List of 1
##   .. ..$ elements: int [1:8, 1] 5 4 6 8 43 10 11 42
##   ..$ 80 - 60 :List of 1
##   .. ..$ elements: int [1:15, 1] 7 8 9 1 2 3 12 13 14 44 ...
##   ..$ 60 - 40 :List of 1
##   .. ..$ elements: int [1:13, 1] 41 49 24 25 26 27 28 21 22 19 ...
##   ..$ 40 - 20 :List of 1
##   .. ..$ elements: int [1:6, 1] 15 39 40 35 23 47
##   ..$ 20 - 0  :List of 1
##   .. ..$ elements: int [1:10, 1] 36 37 38 32 33 34 50 48 29 30
##  - attr(*, "class")= chr "dispRity"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{names}\NormalTok{(binned_morphospace)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "matrix"     "call"       "subsamples"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Printing the object as a dispRity class}
\NormalTok{binned_morphospace}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 discrete time subsamples for 50 elements:
##     100 - 80, 80 - 60, 60 - 40, 40 - 20, 20 - 0.
\end{verbatim}

\begin{quote}
These objects will gradually contain more information when completing
the following steps in the disparity-through-time analysis.
\end{quote}

\subsection{Bootstrapping the data}\label{bootstrapping-the-data}

Once we obtain our different time subsamples, we can bootstrap and
rarefy them (i.e.~pseudo-replicating the data). The bootstrapping allows
us to make each subsample more robust to outliers and the rarefaction
allows us to compare subsamples with the same number of taxa to remove
sampling biases (i.e.~more taxa in one subsample than the others). The
\texttt{boot.matrix} function bootstraps the \texttt{dispRity} object
and the \texttt{rarefaction} option within performs rarefaction.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Bootstrapping each time subsample 100 times}
\NormalTok{(boot_bin_morphospace <-}\StringTok{ }\KeywordTok{boot.matrix}\NormalTok{(binned_morphospace, }\DataTypeTok{bootstraps =} \DecValTok{100}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 discrete time subsamples for 50 elements with 48 dimensions:
##     100 - 80, 80 - 60, 60 - 40, 40 - 20, 20 - 0.
## Data was bootstrapped 100 times (method:"full").
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Getting the minimum number of rows (i.e. taxa) in the time subsamples}
\KeywordTok{min}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(}\KeywordTok{lapply}\NormalTok{(boot_bin_morphospace}\OperatorTok{$}\NormalTok{subsamples, lapply, nrow)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 6
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Bootstrapping each time subsample 100 times and rarefying them }
\NormalTok{(rare_bin_morphospace <-}\StringTok{ }\KeywordTok{boot.matrix}\NormalTok{(binned_morphospace, }\DataTypeTok{bootstraps =} \DecValTok{100}\NormalTok{,}
    \DataTypeTok{rarefaction =} \DecValTok{6}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 discrete time subsamples for 50 elements with 48 dimensions:
##     100 - 80, 80 - 60, 60 - 40, 40 - 20, 20 - 0.
## Data was bootstrapped 100 times (method:"full") and rarefied to 6 elements.
\end{verbatim}

\subsection{Calculating disparity}\label{calculating-disparity-1}

We can now calculate the disparity within each time subsamples along
with some confidence intervals generated by the pseudoreplication step
above (bootstraps/rarefaction). Disparity can be calculated in many ways
and this package allows users to come up with their own disparity
metrics. For more details, please refer to the
\href{http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html}{\texttt{dispRity}
metric vignette}.

In this example, we are going to calculate the spread of the data in
each time subsample by calculating disparity as the sum of the variance
of each dimension of the morphospace in each time subsample using the
\texttt{dispRity} function. Thus, in this example, disparity is defined
by the multi-dimensional variance of each time subsample (i.e.~the
spread of the taxa within the morphospace). Note that this metric comes
with a caveat (not solved here) since it ignores covariances among the
dimensions of the morphospace. We use this here because it is a standard
metric used in disparity-through-time analysis.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity for the bootstrapped data}
\NormalTok{(boot_disparity <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(boot_bin_morphospace, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 discrete time subsamples for 50 elements with 48 dimensions:
##     100 - 80, 80 - 60, 60 - 40, 40 - 20, 20 - 0.
## Data was bootstrapped 100 times (method:"full").
## Disparity was calculated as: c(sum, variances).
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity for the rarefied data}
\NormalTok{(rare_disparity <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(rare_bin_morphospace, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 discrete time subsamples for 50 elements with 48 dimensions:
##     100 - 80, 80 - 60, 60 - 40, 40 - 20, 20 - 0.
## Data was bootstrapped 100 times (method:"full") and rarefied to 6 elements.
## Disparity was calculated as: c(sum, variances).
\end{verbatim}

The \texttt{dispRity} function does not actually display the calculated
disparity values but rather only the properties of the disparity object
(size, subsamples, metric, etc.). To display the actual calculated
scores, we need to summarise the disparity object using the S3 method
\texttt{summary} that is applied to a \texttt{dispRity} object (see
\texttt{?summary.dispRity} for more details).

As for any \texttt{R} package, you can refer to the help files for each
individual function for more details.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Summarising the disparity results}
\KeywordTok{summary}\NormalTok{(boot_disparity)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs bs.median  2.5%   25%   75% 97.5%
## 1   100 - 80  8 1.675     1.488 1.087 1.389 1.568 1.648
## 2    80 - 60 15 1.782     1.679 1.538 1.631 1.728 1.792
## 3    60 - 40 13 1.913     1.772 1.607 1.734 1.826 1.886
## 4    40 - 20  6 2.022     1.707 1.212 1.537 1.822 1.942
## 5     20 - 0 10 1.971     1.794 1.598 1.716 1.842 1.890
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(rare_disparity)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsamples  n   obs bs.median  2.5%   25%   75% 97.5%
## 1   100 - 80  8 1.675     1.484 1.194 1.400 1.547 1.636
## 2   100 - 80  6    NA     1.477 0.993 1.361 1.569 1.698
## 3    80 - 60 15 1.782     1.674 1.517 1.600 1.725 1.793
## 4    80 - 60  6    NA     1.655 1.299 1.532 1.754 1.882
## 5    60 - 40 13 1.913     1.767 1.601 1.714 1.829 1.861
## 6    60 - 40  6    NA     1.787 1.314 1.672 1.879 1.984
## 7    40 - 20  6 2.022     1.736 1.281 1.603 1.822 1.948
## 8     20 - 0 10 1.971     1.807 1.595 1.729 1.856 1.917
## 9     20 - 0  6    NA     1.790 1.435 1.718 1.873 1.995
\end{verbatim}

\begin{quote}
The summary.dispRity function comes with many options on which values to
calculate (central tendency and quantiles) and on how many digits to
display. Refer to the function's manual for more details.
\end{quote}

\subsection{Plotting the results}\label{plotting-the-results}

It is sometimes easier to visualise the results in a plot than in a
table. For that we can use the \texttt{plot} S3 function to plot the
\texttt{dispRity} objects (see \texttt{?plot.dispRity} for more
details).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\KeywordTok{quartz}\NormalTok{(}\DataTypeTok{width =} \DecValTok{10}\NormalTok{, }\DataTypeTok{height =} \DecValTok{5}\NormalTok{) ; }\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =}\NormalTok{ (}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{)), }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting the bootstrapped and rarefied results}
\KeywordTok{plot}\NormalTok{(boot_disparity, }\DataTypeTok{type =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{main =} \StringTok{"bootstrapped results"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(rare_disparity, }\DataTypeTok{type =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{main =} \StringTok{"rarefied results"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{manual_files/figure-latex/unnamed-chunk-90-1.pdf}

\section{Testing differences}\label{testing-differences}

Finally, to draw some valid conclusions from these results, we can apply
some statistical tests. We can test, for example, if mammalian disparity
changed significantly through time over the last 100 million years. To
do so, we can compare the means of each time-bin in a sequential manner
to see whether the disparity in bin \emph{n} is equal to the disparity
in bin \emph{n+1}, and whether this is in turn equal to the disparity in
bin \emph{n+2}, etc. Because our data is temporally autocorrelated
(i.e.~what happens in bin \emph{n+1} depends on what happened in bin
\emph{n}) and pseudoreplicated (i.e.~each bootstrap draw creates
non-independent time subsamples because they are all based on the same
time subsamples), we apply a non-parametric mean comparison: the
\texttt{wilcox.test}. Also, we need to apply a p-value correction
(e.g.~Bonferroni correction) to correct for multiple testing (see
\texttt{?p.adjust} for more details).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Testing the differences between bins in the bootstrapped dataset.}
\KeywordTok{test.dispRity}\NormalTok{(boot_disparity, }\DataTypeTok{test =}\NormalTok{ wilcox.test, }\DataTypeTok{comparison =} \StringTok{"sequential"}\NormalTok{,}
    \DataTypeTok{correction =} \StringTok{"bonferroni"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [[1]]
##                    statistic
## 100 - 80 : 80 - 60       471
## 80 - 60 : 60 - 40       1562
## 60 - 40 : 40 - 20       6250
## 40 - 20 : 20 - 0        3725
## 
## [[2]]
##                         p.value
## 100 - 80 : 80 - 60 7.427563e-28
## 80 - 60 : 60 - 40  1.798899e-16
## 60 - 40 : 40 - 20  9.061511e-03
## 40 - 20 : 20 - 0   7.379715e-03
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Testing the differences between bins in the rarefied dataset.}
\KeywordTok{test.dispRity}\NormalTok{(rare_disparity, }\DataTypeTok{test =}\NormalTok{ wilcox.test, }\DataTypeTok{comparison =} \StringTok{"sequential"}\NormalTok{,}
    \DataTypeTok{correction =} \StringTok{"bonferroni"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [[1]]
##                    statistic
## 100 - 80 : 80 - 60       662
## 80 - 60 : 60 - 40       1814
## 60 - 40 : 40 - 20       5752
## 40 - 20 : 20 - 0        3621
## 
## [[2]]
##                         p.value
## 100 - 80 : 80 - 60 1.214988e-25
## 80 - 60 : 60 - 40  2.823697e-14
## 60 - 40 : 40 - 20  2.653018e-01
## 40 - 20 : 20 - 0   3.026079e-03
\end{verbatim}

Here our results show significant changes in disparity through time
between all time bins (all p-values \textless{} 0.05). However, when
looking at the rarefied results, there is no significant difference
between the time bins in the Palaeogene (60-40 to 40-20 Mya), suggesting
that the differences detected in the first test might just be due to the
differences in number of taxa sampled (13 or 6 taxa) in each time bin.

\subsection{References}\label{references}

\end{document}

### 2017-08-15 inst/gitbook/index.Rmd
moved manuals to inst/

--- 2017-08-12 gitbook/index.Rmd
@@
 However, the package follows the strict (and useful!) CRAN standards via Travis.
-
+<!-- badges out start
 [![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
@@
 [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
-
+badges out end   -->
 ## Help

### 2017-08-15 inst/gitbook/02_getting-started.Rmd
book update

@@
 
-# Getting started with dispRity
+# Getting started with `dispRity`
 
-## What sort of data does dispRity work with?
+## What sort of data does `dispRity` work with?
 
@@
 
-To illustrate these functions, we will use data from @beckancient2014.
+To illustrate these functions, we will use data from Beck and Lee (2014).
 This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
@@
 To give your tree a root time (i.e.
-an age for the root), you can simply do `my_tree$root.time <- my_age`.
+an age for the root), you can simply do\\ `my_tree$root.time <- my_age`.
   * The required number of time subsamples (here `time = 3`)
@@
 
-Using the @beckancient2014 data described [above](#example-data):
+Using the Beck and Lee (2014) data described [above](#example-data):
 
@@
 
-This generates a `dispRity` object (see [here](#manipulating-dispRity-objects) for technical details).
+This generates a `dispRity` object (see [here](#guts) for technical details).
 When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:
@@
 summary(disp_lm)
-
-## Note that ANOVA assumptions were not checked here!
 ```

### 2017-08-15 inst/gitbook/03_specific-tutorials.Rmd
book update

@@
 
+Before each section, make sure you loaded the Beck and Lee (2014) data (see [example data](#example-data) for more details).
+
+```{r}
+## Loading the data
+data(BeckLee_mat50) ; data(BeckLee_mat99)
+data(BeckLee_tree) ; data(BeckLee_ages)
+```
+
 ## Time slicing
@@
 
-For the time-slicing method details see @GuillermeSTD. <!-- @@@ Change cite appropriately! -->
+For the time-slicing method details see Cooper and Guillerme (in prep.). <!-- @@@ Change cite appropriately! -->
 <!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
@@
 In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
-For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @Wills1994.
+For example, one can look at ellipsoid hyper-volume of the ordinated space (Donohue _et al._ 2013), the sum and the product of the ranges and variances (Wills _et al._ 1994) or the median position of the elements relative to their centroid (Wills _et al._ 1994).
 Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.
@@
 ------|------|---------------------------------------------------|--------|
-3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
-3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
-3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
+1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
+1 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
+1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
+1 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
+1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
 2 | `ranges` | The range of each dimension | `dispRity` |
@@
 This function is an S3 function (`summary.dispRity`) allowing users to summarise the content of `dispRity` objects that contain disparity calculations.
-
-<!-- DATA LOADER -->
 
@@
 
+Here we are using the variables generated in the [section above](#summarising-dispRity-data-plots):
+
 ```{r}
@@
 
-## Wilcoxon test applied to time sliced disparity with sequential comparisons, with Bonferroni correction
+## Wilcoxon test applied to time sliced disparity with sequential comparisons,
+## with Bonferroni correction
 test.dispRity(disparity_time_slices, test = wilcox.test,
@@
 This is possible in the `dispRity` package by calculating disparity as a dimension-level 2 metric only!
-Let's have a look using our previous example of bootstrapped time slices but by measuring the distances between each taxon and their centroid as disparity.
+Let's have a look using our [previous example](#summarising-dispRity-data-plots) of bootstrapped time slices but by measuring the distances between each taxon and their centroid as disparity.
 
@@
 Note that the quantiles and central tendencies can be modified via the `conc.quantiles` option.
-
-## Ecological studies of disparity
-<!-- NC: I wonder if this might be better just as a "hey we can do ecology stuff too and here is a link to a demo" rather than a new section?
-The manual is loooooooooong so splitting some stuff out might be good?
--->
-
-Disparity analyses are not restricted to palaeobiology analysis.
-This package is designed for dealing with ordinated matrices, no matter which data was used in the ordination.
-Here is an short example of disparity analysis using ecological data from Deirdre McClean (unpubl.).
-
-The dataset contains an ordinated matrix of 20 dimensions (columns) for 40 elements (rows).
-The elements are different experimental plots with discrete variations of nutrient enrichment, depth, and freshwater benthic invertebrates.
-<!-- NC: Explain a little more about what the data are here as I don't entirely get it. EG what's a treatement etc? Just a really simple explanataion is perfect -->
-
-```{r}
-## Loading the data
-data(McClean_data)
-
-## Ordinated matrix
-ordinated_matrix <- McClean_data$ordination
-
-## Treatments and depths
-treatments <- McClean_data$treatment
-depth <- McClean_data$depth
-```
-
-### Classic ecological "disparity" analysis
-
-A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different experimental plots differ in their values for nutrient enrichment, depth, and freshwater benthic invertebrates.
-<!-- NC: or similar just to make it clearer -->
-
-```{r, fig.width=6, fig.height=6}
-## Setting the colors that will represent the treatments
-cols <- sub("a", "red", treatments)
-cols <- sub("b", "blue", cols)
-## Setting the symbols that will represent the depth
-pchs <- sub(1, 16, depth)
-pchs <- as.numeric(sub(2, 17, pchs))
-
-## Graphical option
-op <- par(bty = "n")
-## A classic 2D ordination plot
-plot(ordinated_matrix[, 1:2], col = cols, pch = pchs,
-     xlab = "PC 1", ylab = "PC 2",
-     xlim = range(ordinated_matrix[, 1:2]) + c(0, 100))
-par(op)
-```
-
-This shows the distribution of the experimental plots along the two first axis of variation of the ordinated distance matrix (i.e. the first two dimensions).
-At a first glance, it seems difficult to see a clear effect of the treatments (blue or red) or the depth (rounds or triangles) as the different experimental plots with the same parameters don't seem to cluster together.
-However, this may be because this plot ignores the 18 other dimensions of the ordination!
-Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e.
-the values on the first dimension do not represent a continuous trait (e.g.
-depth), instead they just represent the ordinations of correlations between the data and some factors.
-
-Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.
-
-### Multidimensional analysis with `dispRity`
-
-#### Splitting the data with respect to the different factors
-
-The first step is to create different subsamples that represent subsamples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
-Each of these subsamples will contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
-In our example, we are going to group the elements according to their depth and treatment.
-
-```{r}
-## Creating a table that contain the elements and their attributes
-factors <- as.data.frame(matrix(data = c(treatments, depth),
-          nrow = nrow(ordinated_matrix), ncol = 2, byrow = FALSE,
-          dimnames = list(rownames(ordinated_matrix))))
-names(factors) <- c("Treat", "Depth")
-head(factors)
-```
-
-Second, let’s split the data according by depth and treatment to create the subsamples of the ordinated space by using the `custom.subsamples` function:
-
-```{r}
-## Splitting the ordinated space into four subsamples
-(customised_subsamples <- custom.subsamples(ordinated_matrix, factors))
-```
-
-#### Calculating disparity
-
-In this example, we are going to define disparity as the n-dimensional ellipsoid volume (see `?hyper.volume`).
-
-```{r}
-## Calculating disparity as the ellipsoid volume of each group
-disparity <- dispRity(customised_subsamples, metric = diagonal)
-summary(disparity)
-```
-
-We can also bootstrap the data to test the robustness of the measured disparity to outliers.
-Also, as we can see, each subsample has different numbers of elements.
-It might also be interesting to rarefy the data to have only subsamples with the same number of elements.
-
-```{r}
-## Bootstrapping the subsamples and using rarefaction
-## (i.e. only re-sampling 17 elements each time)
-bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 100,
-                                 rarefaction = 17)
-```
-
-We can now run a more robust disparity analysis using the bootstrapped data:
-
-```{r}
-## Calculating the bootstrapped disparity
-disparity <- dispRity(bootstrapped_data, metric = diagonal)
-```
-
-#### Summarising and displaying the results
-
-Now let's summarise this data:
-
-```{r, fig.width=12, fig.height=6, eval=TRUE}
-## Summary table of the data
-summary(disparity)
-
-## Graphical options
-op <- par(mfrow = (c(1, 2)), bty = "n")
-
-## Plotting the bootstrapped disparity
-plot(disparity, main = "Bootstrapped", las = 2, xlab = "")
-
-## Plotting the rarefied disparity
-plot(disparity, rarefaction = 17, main = "Rarefied", las = 2, xlab = "")
-
-par(op)
-```
-
-As we can see, there seems to be no strong effect of the number of experimental plots in each subsample (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!
-
-#### Testing our hypotheses
-
-Finally, we can test our hypothesis (whether the water treatment at certain depths alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.
-<!-- Ah so this is what you are testing! Make this clearer in the intro bit to this section -->
-
-```{r}
-## Testing the effect of treatment and depth on the bootstrapped data
-summary(test.dispRity(disparity, test = aov, comparisons = "all"))
-```
 

### 2017-08-15 inst/gitbook/06_ecology-demo.Rmd
book update

---
title: "dispRity ecology demo"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: References.bib
bst: sysbio.bst
vignette: >
  %\VignetteIndexEntry{dispRity ecology demo}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This is an example of typical disparity analysis that can be performed in ecology.

# Data
For this example, we will use the famous `iris` inbuilt data set

```{r}
data(iris)
```

This data contains petal and sepal length for 150 individual plants sorted into three species.

```{r}
## Separating the species
species <- iris[,5]
## Which species?
unique(species)

## Separating the petal/sepal length
measurements <- iris[,1:4]
head(measurements)
```

We can then ordinate the data using a PCA (`prcomp` function) thus defining our four dimensional space as the poetically named petal-space.

```{r}
## Ordinating the data
ordination <- prcomp(measurements)

## The petal-space
petal_space <- ordination$x

## Adding the elements names to the petal-space (the individuals IDs)
rownames(petal_space) <- 1:nrow(petal_space)
```

# Classic analysis
A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different species are distributed in the petal-space.

```{r, fig.width=6, fig.height=6}
## Measuring the variance on each axis
variances <- apply(petal_space, 2, var)
variances <- variances/sum(variances)

## Graphical option
par(bty = "n")

## A classic 2D ordination plot
plot(petal_space[, 1], petal_space[, 2], col = species,
    xlab = paste0("PC 1 (", round(variances[1], 2), ")"),
    ylab = paste0("PC 2 (", round(variances[2], 2), ")"))
```

This shows the distribution of the different species in the petal-space along the two first axis of variation.
This is a pretty standard way to visualise the multidimensional space and further analysis might be necessary to test wether the groups are different such as a linear discriminant analysis (LDA).
However, in this case we are ignoring the two other dimensions of the ordination!
If we look at the two other axis we see a totally different result:

```{r, fig.width=6, fig.height=6}
## Plotting the two second axis of the petal-space
plot(petal_space[, 3], petal_space[, 4], col = species,
    xlab = paste0("PC 3 (", round(variances[3], 2), ")"),
    ylab = paste0("PC 4 (", round(variances[4], 2), ")"))
```

Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e. the values on the first dimension do not represent a continuous trait (e.g. petal length), instead they just represent the ordinations of correlations between the data and some factors.

Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.

# A multidimensional approach with `dispRity`

The first step is to create different subsamples that represent subsamples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
Each of these subsamples will contain only the individuals of a specific species.

```{r}
## Creating the table that contain the elements and their attributes
petal_subsamples <- custom.subsamples(petal_space, group = list(
                                "setosa" = which(species == "setosa"),
                                "versicolor" = which(species == "versicolor"),
                                "virginica" = which(species == "virginica")))

## Visualising the dispRity object content
petal_subsamples
```

This created a `dispRity` object (more about that [here](#guts)) with three subsamples corresponding to each subspecies.

## Bootstrapping the data
We can the bootstrap the subsamples to be able test the robustness of the measured disparity to outliers.
We can do that using the default options of `boot.matrix` (more about that [here](#bootstraps-and-rarefactions)):

```{r}
## Bootstrapping the data
(petal_bootstrapped <- boot.matrix(petal_subsamples))
```

## Calculating disparity

Disparity can be calculated in many ways, therefore the `dispRity` function allows users to define their own measure of disparity.
For more details on measuring disparity, see the [dispRity metrics section](#disparity-metrics).

In this example, we are going to define disparity as the median distance between the different individuals and the centroid of the ordinated space.
High values of disparity will indicate a generally high spread of points from this centroid (i.e. on average, the individuals are far apart in the ordinated space).
We can define the metrics easily in the `dispRity` function by feeding them to the `metric` argument.
Here we are going to feed the functions `stats::median` and `dispRity::centroids` which calculates distances between elements and their centroid.

```{r}
## Calculating disparity as the median distance between each elements and
## the centroid of the petal-space
(petal_disparity <- dispRity(petal_bootstrapped, metric = c(median, centroids)))
```

## Summarising the results (plot)
Similarly to the `custom.subsamples` and `boot.matrix` function, `dispRity` displays a `dispRity` object.
But we are definitely more interested in actually look at the calculated values.

First we can summarise the data in a table by simply using `summary`:

```{r}
## Displaying the summary of the calculated disparity
summary(petal_disparity)
```

We can also plot the results in a similar way:

```{r, fig.width=6, fig.height=6}
## Graphical options
par(bty = "n")

## Plotting the disparity in the petal_space
plot(petal_disparity)
```

Now contrary to simply plotting the two first axis of the PCA where we saw that the species have a different position in the two first petal-space, we can now also see that they occupy this space clearly differently!

## Testing hypothesis

Finally we can test our hypothesis that we guessed from the disparity plot (that some groups occupy different volume of the petal-space) by using the `test.dispRity` option.

```{r}
## Fitting a linear model to our data
disparity_lm <- test.dispRity(petal_disparity, test = lm, comparisons = "all")

## Testing whether there is a difference in disparity between the different species
summary(aov(disparity_lm))

## Post-hoc testing of the differences between species (corrected for multiple tests)
test.dispRity(petal_disparity, test = t.test, correction = "bonferroni")
```

We can now see that there is a significant difference in petal-space occupancy between all species of iris.

### Setting up a multidimensional null-hypothesis

One other series of test can be done on the shape of the petal-space.
Using a MCMC permutation test we can simulate a petal-space with specific properties and see if our observed petal-space matches these properties (similarly to Diaz _et al._ 2016):

```{r, fig.width=6, fig.height=6}
## Testing against a uniform distribution
disparity_uniform <- null.test(petal_disparity, replicates = 200,
    null.distrib = runif, scale = FALSE)
plot(disparity_uniform)
```

```{r, fig.width=6, fig.height=6}
## Testing against a normal distribution
disparity_normal <- null.test(petal_disparity, replicates = 200,
    null.distrib = rnorm, scale = TRUE)
plot(disparity_normal)
```

In both cases we can see that our petal-space is not entirely normal or uniform.
This is expected because of the simplicity of these parameters.

### 2017-08-15 inst/gitbook/index.Rmd
book update

--- 2017-08-12 gitbook/index.Rmd
@@
 This is a package for measuring disparity in `R`.
-It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values so they can easily be compared.
-This manual is based on the version `1.0`.
+It allows users to summarise matrices as representations as multidimensional spaces (namely from ordinated matrices from MDS, PCA, PCO, PCoA) into a single value or distribution describing a specific aspect of this multidimensional space (the disparity).
+This manual is based on the version `0.4`.
 
@@
 ```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
-# NC: this isn't checking Claddis...
+## loading the package and setting up the start seed.
 library(dispRity)
@@
 
+<!-- badges out start -->
 [![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
@@
 [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
+<!-- badges out end   -->
 
@@
 
-<!-- TG: I'm gonna change that to a newest version with the submission -->
-
 Note that this citation is only temporary (but can still be used!).

### 2017-08-15 paper/dispRity_paper.tex
Minor fixes for 0.4 release

@@
 
-Disparity is really cool:
+Biological data are complex; understanding the ecology and evolution of species often requires that we analyse multiple variables that covary with each other, and through space and time.
+One solution to this problem is to analyse data in a multivariate framework.
+Multivariate analyses aim to capture and incorporate this multidimensional complexity, while providing outputs that are interpretable in a physical world of only three dimensions.
+Classical multidimensional species features that have been analysed through multivariate data ranges from their morphology \citep{raup1966geometric}, to their functional traits \citep{diaz2016global}.
+However, such analysis are not limited at the species level and can also be applied to ecosystems \citep{DonohueDim}.
 
-Hereafter we will make a clear distinction between the multidimensional space which is a specific mathematical object and the disparity, which is a metric describing or summarising one or more aspects of this space.
+Key multidimensional features of species that have important roles in ecology and evolution include morphology, functional traits. etc. etc.
+Many of these can be represented as matrices, these can be used in our package.
+In the interests of clarity/brievety, we here focus on just one kind of data: morphological diversity.
+
+Furthermore, we will also make a clear distinction between the multidimensional space which is a specific mathematical object and the disparity, which is a metric describing or summarising one or more aspects of this space.
+
 Both can be defined in various ways depending on the authors, for example, disparity is defined as the weighted mean pairwise dissimilarity in \cite{Close2015} or the ellipsoid volume in \cite{DonohueDim} and the multidimensional space is defined as the morphospace in \cite{raup1966geometric} and the morpho-functional space in \cite{diaz2016global}.
@@
 
-% \subsubsection{The package's data}
-
-% Beck and Lee data
-% Beck and Lee tree
-% Beck and Lee ages
-
-% disparity  disparity data
-
-% \subsubsection{Running example}
-
-% \subsection{The \disp package}
-
-% \subsubsection{The package's main functions}
-
-% \subsection{Fork flow}
-
-% Workflow figure
-
-% \subsubsection{Input data}
-
-% \subsubsection{Manipulating the data}
-
-% \subsubsection{Measuring disparity}
-
-% \subsubsection{Summarising the data}
-
-% Link to graph one two three
-
-% \subsubsection{Testing hypothesis}
-
-% \subsection{The \disp object}
-
-% Disparity object structure figure
-
-% \subsection{Additional functions and data}
-
-% The package also contains additional functions not directly linked to the disparity workflow nor to the \disp object but that can be used as utilities for the studying discrete morphological data.
-
-% \begin{table}
-%     \begin{tabular}{ll}
-%         \hline
-%         Function & Description \\ 
-%         \hline
-%         \texttt{char.diff} & Calculates character differences. \\
-%         \texttt{clean.data} & Cleans phylogenetic data (tree and table). \\
-%         \texttt{sim.morpho} & Simulates morphological discrete data. \\
-%         \texttt{slice.tree} & Slice through a phylogenetic tree. \\
-%         \texttt{space.maker} & Creates multidimensional spaces. \\
-%         \texttt{tree.age} & Calculates the age of nodes and tips in a tree. \\
-%         \hline
-%     \end{tabular}
-%     \caption{Additional functions in the \disp package.}
-% \end{table}
-
-% Additionally, this package also contains an ecology type data set from McClean
-
-% McClean data
-
-% \section{Conclusion}
-
-% This package is good.
-
 % \section{Data availability and reproducibility}

### 2017-08-15 inst/gitbook/01_glossary.Rmd

@@
 
--   **Ordinated space**. The mathematical multidimensional object that will be analysed with this package.
+-   **Multidimensional space**. The mathematical multidimensional object that will be analysed with this package.
     In morphometrics, this is often referred to as the morphospace.
     However it may also be referred to as the cladisto-space for cladistic data or the eco-space for ecological data etc.
-    In practice, this term designates an ordinated matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
+    In practice, this term designates a matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
 
--   **Elements**. The rows of the ordinated space. Elements can be taxa, field sites, countries etc.
+-   **Elements**. The rows of the multidimensional space matrix. Elements can be taxa, field sites, countries etc.
 
--   **Dimensions**. The columns of the ordinated space. The dimensions are referred to as axes of variation, or principal components, for ordinated spaces obtained from a PCA.
+-   **Dimensions**.
+The columns of the multidimensional space matrix.
+The dimensions can be referred to as axes of variation, or principal components, for ordinated spaces obtained from a PCA for example.
 
--   **Subsamples**. Sub-samples of the ordinated space.
-    A sub-sample (or subsamples) contains the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
-    For example, if our ordinated space is composed of birds and mammals (the elements) and 50 principal components (the dimensions), we can create two subsamples containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
+-   **Subsamples**. Subsamples of the multidimensional space.
+    A subsample (or subsamples) contains the same number of dimensions as the space but may contain a smaller subset of elements.
+    For example, if our space is composed of birds and mammals (the elements) and 50 principal components of variation (the dimensions), we can create two subsamples containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.

### 2017-08-15 inst/gitbook/06_ecology-demo.Rmd

@@
 
+# Ecology demo
+
 This is an example of typical disparity analysis that can be performed in ecology.
 
-# Data
+## Data
 For this example, we will use the famous `iris` inbuilt data set
@@
 
-# Classic analysis
+## Classic analysis
 A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different species are distributed in the petal-space.
@@
 
-# A multidimensional approach with `dispRity`
+## A multidimensional approach with `dispRity`
 
@@
 
-## Bootstrapping the data
+### Bootstrapping the data
 We can the bootstrap the subsamples to be able test the robustness of the measured disparity to outliers.
@@
 
-## Calculating disparity
+### Calculating disparity
 
@@
 
-## Summarising the results (plot)
+### Summarising the results (plot)
 Similarly to the `custom.subsamples` and `boot.matrix` function, `dispRity` displays a `dispRity` object.
@@
 
-## Testing hypothesis
+### Testing hypothesis
 
@@
 
-### Setting up a multidimensional null-hypothesis
+#### Setting up a multidimensional null-hypothesis
 

### 2017-08-15 inst/gitbook/07_future_directions.Rmd

---
title: "Future directions and references"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: References.bib
bst: sysbio.bst
vignette: >
  %\VignetteIndexEntry{future directions}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Future directions

## More tests!

Some more tests are being developed such as a `sequential.test` to run sequential linear models for testing hypothesis in disparity through time or a `model.test` developed in collaboration with [Mark Puttick](https://puttickbiology.wordpress.com/) to fit modes of evolution to disparity curves.
Stay tuned!

## Faster disparity calculations

I am slowly implementing parallel disparity calculation as well as C implementations of some disparity metrics to increase significantly improve the speed of the `dispRity` function.

## More modularity

I am equally slowly developing functions to allow more of the options in the package to be modular (in the same way as the `metric` argument in `dispRity`).
The next arguments to benefit this increased modularity will be `time.subsamples`'s `model` argument and `boot.matrix`'s `type` argument.

# References

* Beck, R. M., & Lee, M. S. (2014). Ancient dates or accelerated rates? Morphological clocks and the antiquity of placental mammals. Proceedings of the Royal Society of London B: Biological Sciences, 281(1793), 20141278.
* Cooper, N., & Guillerme, T. (in prep.). Coming soonish!.
* Diaz, S., Kattge, J., Cornelissen, J.H., Wright, I.J., Lavorel, S., Dray, S., Reu, B., Kleyer, M., Wirth, C., Prentice, I.C. and Garnier, E. (2016). The global spectrum of plant form and function. Nature, 529(7585), 167.
* Donohue, I., Petchey, O.L., Montoya, J.M., Jackson, A.L., McNally, L., Viana, M., Healy, K., Lurgi, M., O'Connor, N.E. and Emmerson, M.C, (2013). On the dimensionality of ecological stability. Ecology letters, 16(4), 421-429.
* Wills, M. A., Briggs, D. E., & Fortey, R. A. (1994). Disparity as an evolutionary index: a comparison of Cambrian and Recent arthropods. Paleobiology, 20(2), 93-130.

### 2017-08-21 inst/gitbook/03_specific-tutorials.Rmd
Gitbook!
removed hyper.volume

@@
 1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-1 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
+<!-- 1 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |-->
 1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
@@
 #### Volumes and surface metrics
-The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `hyper.volume` all measure the surface or the volume of the ordinated space occupied:
+The functions `ellipse.volume`, `convhull.surface` and `convhull.volume`  all measure the surface or the volume of the ordinated space occupied:
 
@@
 The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
-For the full and correct calculation of the volume of the ordinated space, it is possible to use the `hyper.volume` function that has more options but takes longer to calculate.
 
-```{r}
-## Calculating the true multidimensional volume
-summary(dispRity(dummy_space, metric = hyper.volume))
-
-```
 > Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) that often results in near 0 disparity values.
@@
 
-### `space.maker`
+## Simulating multidimensional spaces
 Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.

### 2017-08-21 paper/dispRity_paper.tex

@@
 The \disp function allows to measure disparity from any matrix (of class \texttt{matrix} or \texttt{data.frame}) where the columns correspond to the different dimensions of the multidimensional space (e.g.
-the ordination axis or any variable that composes the space) and the rows correspond to the elements present in the multidimensional space (e.g.
-species, study sites, etc...).
+the ordination axis or any variable composing the space) and the rows correspond to the elements present in the multidimensional space (e.g.
+species, individuals, study sites, etc...).
 
@@
 \centering
-   \includegraphics[width=1\textwidth]{../doc/dispRity_fun.pdf} 
+   \includegraphics[width=1\textwidth]{../inst/gitbook/dispRity_fun.pdf} 
 \caption{Illustration of the different metric dimensions levels in the \disp package}
@@
 \center
-    \begin{tabular}{r|c|c|c|c}
+    \begin{tabular}{r|l|c|c|c}
         name & description & dimension level & definition & source \\
         \hline
-        \texttt{ellipse.volume} & The volume of the ellipsoid of the space & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
-        \texttt{convhull.surface} & 1 & 1 & 1 & 1\\
-        \texttt{convhull.volume} & 1 & 1 & 1 & 1\\
-        \texttt{hypervolume} & 0 & 0 & 0 & 1\\
-        \texttt{diagonal} & 0 & 0 & 0 & 1\\
-        \texttt{ranges} & 0 & 0 & 0 & 1\\
-        \texttt{variances} & 0 & 0 & 0 & 1\\
-        \texttt{centroids} & 0 & 0 & 0 & 1\\
-        \texttt{mode.val} & 0 & 0 & 0 & 1\\
+        \texttt{ellipse.volume}$^1$ & The volume of the ellipsoid around the elements & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
+        \texttt{convhull.surface} & The surface of the convex hull formed by the elements & 1 & NA & \texttt{geometry::convhulln} \\
+        \texttt{convhull.volume} & The volume of the convex hull formed by the elements & 1 & NA & \texttt{geometry::convhulln} \\
+        \texttt{hypervolume} & The volume of the multidimensional space space  & 1 & NA & \texttt{hypervolume::hypervolume} \\
+        \texttt{diagonal} & The longest Eucliden distance between any pair of elements & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & \texttt{dispRity::diagonal} \\
+        \texttt{mode.val} & The modal value of a vector or a matrix & 1 & NA & \texttt{dispRity::mode.val}\\
+        \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & \texttt{dispRity::ranges} \\
+        \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & \texttt{dispRity::variances} \\
+        \texttt{centroids} & The distance between each element and a fixed point$^2$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & \texttt{dispRity::centroids} \\
     \end{tabular}
-    \caption{Where $k$ is the number of dimensions, $n$ the number of elements, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean.}
+    \caption{Where $k$ is the number of dimensions, $n$ the number of elements, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean.
+$^1$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA); $^2$ by default that point is the centroid of the elements.}
     \label{Tab:metrics}
 \end{table}
-
-% Level | Name | Description | Source | 
-% ------|------|---------------------------------------------------|--------|
-% 3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
-% 3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-% 3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-% 3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
-% 3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
-% 2 | `ranges` | The range of each dimension | `dispRity` |
-% 2 | `variances` | The variance of each dimension | `dispRity` |
-% 2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
-% 1 | `mode.val` | The modal value | `dispRity` |
-
-% 1: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (not PCA).
-
-% 2: Note that by default, the centroid is the centroid of the elements.
-% It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
-%     ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
-%     diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
-%     ranges = |max(k_i) - min(k_i)|
-%     variances = \sigma^{2}{k_i}
-%     centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
 
@@
 
-\texttt{dispRity.through.time(BeckLee\_mat50, )}
+As an example, one can define their multidimensional space as being the morphospace of an ordination of the distance between discrete morphological characters for 50 mammals species \citep[from][]{beckancient2014}.
+Then, one can define disparity as the sum of the variances on each dimensions \citep{Wills1994} that will represent an aspect of the the volume of the morphospace (i.e.
+a large sum of variances approximates a large morphospace).
+One can then be interested in how this approximation of the volume of the morphospace changes throughout different time periods (say from the end of the Cretaceous to the end of the Palaeogene).
 
-This function is a wrapping function that is the equivalent to:
+\texttt{disparity <- dispRity.through.time(data = BeckLee\_mat50, time = c(90, 70, 50, 30), metric = c(sum, variances))}
++ method = discrete
++ tree = tree
 
-\texttt{dispRity(boot.matrix(time.subsamples(BeckLee\_mat50, )))}
+\noindent Note that this function is a wrapping function that is the equivalent to:
+
+\texttt{disparity <- dispRity(boot.matrix(time.subsamples(data = BeckLee\_mat50, time = c(90, 70, 50, 30))), metric = c(sum, variances))}
+
+\noindent Which allows a finer tuning of each optional argument in each of the three functions \texttt{time.subsamples}, \texttt{boot.matrix} and \disp.
+The three arguments here are pretty self explanatory: \texttt{data = BeckLee\_mat50} is our definition of the multidimensional space, \texttt{time = c(90, 70, 50, 30)} indicates the boundaries of the different time bins and \texttt{metric = c(sum, variances)} is our definition of disparity.
+
+This function returns a \disp object that does not directly gives us the results (the disparity in each time bin) but rather summarises the content of the \disp object.
+To visualise the disparity values, one can use the \texttt{summary} or/and \texttt{plot} options:
+
+\texttt{summary(disparity)}
+
+\texttt{plot(disparity)}
 

### 2017-08-21 paper/dispRity_paper.tex
Finished example first try

@@
         \texttt{convhull.volume} & The volume of the convex hull formed by the elements & 1 & NA & \texttt{geometry::convhulln} \\
-        \texttt{hypervolume} & The volume of the multidimensional space space  & 1 & NA & \texttt{hypervolume::hypervolume} \\
         \texttt{diagonal} & The longest Eucliden distance between any pair of elements & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & \texttt{dispRity::diagonal} \\
@@
 
-\subsubsection{Typical disparity through type analysis}
+\subsection{Typical disparity per groups analysis}
 
@@
 a large sum of variances approximates a large morphospace).
-One can then be interested in how this approximation of the volume of the morphospace changes throughout different time periods (say from the end of the Cretaceous to the end of the Palaeogene).
+One can then be interested in whether the approximation of this volume is different in the living mammals and their descendants (crown) or the fossil mammals with no living descendants (stem).
+These two groups can be defined as follow (where the integers are the row number of the species in the matrix):
 
-\texttt{disparity <- dispRity.through.time(data = BeckLee\_mat50, time = c(90, 70, 50, 30), metric = c(sum, variances))}
-+ method = discrete
-+ tree = tree
+\texttt{> mammal\_groups <- list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44))}
+
+\noindent It is then possible to measure the disparity between the two groups as follow:
+
+\texttt{> disparity <- dispRity.per.group(data = BeckLee\_mat50, group = mammal\_groups, metric = c(sum, variances))}
 
@@
 
-\texttt{disparity <- dispRity(boot.matrix(time.subsamples(data = BeckLee\_mat50, time = c(90, 70, 50, 30))), metric = c(sum, variances))}
+\texttt{> subsamples <- custom.subsamples(data = BeckLee\_mat50, group = mammal\_groups)}\\
+\texttt{> bootstraps <- boot.matrix(subsamples)}\\
+\texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\
+
+\noindent Which allows a finer tuning of each optional argument in each of the three functions \texttt{custom.subsamples}, \texttt{boot.matrix} and \disp.
+The three arguments here are pretty self explanatory: \texttt{data = BeckLee\_mat50} is our definition of the multidimensional space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity.
+
+This function returns a \disp object that does not directly gives us the results (the disparity in each time bin) but rather summarises the content of the \disp object.
+
+\noindent \texttt{ ---- dispRity object ---- }\\
+\noindent \texttt{2 customised subsamples for 50 elements with 48 dimensions:}\\
+\noindent \texttt{    crown, stem.}\\
+\noindent \texttt{Data was bootstrapped 100 times (method:"full").}\\
+\noindent \texttt{Disparity was calculated as: c(sum, variances).}\\
+
+\bigskip
+As indicated, the \disp object contains 2 customised subsamples (called ``crown'' and ``stem'') from a morphospace made of 48 dimensions and 50 elements.
+The \disp object also displays information on the number and method of bootstrap replicates as well as how disparity was measured.
+To visualise the actual disparity values, one can use the \texttt{summary} or/and \texttt{plot} options (table \ref{Tab:summary_group} and figure \ref{Fig:plot_group}):
+
+\texttt{> summary(disparity)}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{rlrrrrrrr}
+  \hline
+ & subsamples & n & obs & bs.median & 2.5\% & 25\% & 75\% & 97.5\% \\ 
+  \hline
+1 & crown &  30 & 2.00 & 1.93 & 1.87 & 1.92 & 1.95 & 1.98 \\ 
+  2 & stem &  20 & 1.72 & 1.63 & 1.53 & 1.60 & 1.66 & 1.69 \\ 
+   \hline
+\end{tabular}
+\caption{Summarising a \disp object (disparity per groups)}
+\label{Tab:summary_group}
+\end{table}
+
+\texttt{> plot(disparity)}
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{plot_example_group.pdf} 
+\caption{\disp plot of disparity differences between groups}
+\label{Fig:plot_group}
+\end{figure}
+
+As we can see from the summary table and the plot, there seems to be a significant difference in morphospace volume occupation between the two groups.
+It is possible to test this hypothesis by using, for example, a non-parametric Wilcoxon test (\texttt{stats::wilcox.test}):
+
+\texttt{> test.dispRity(disparity, test = wilcox.test)}
+
+\noindent \texttt{[[1]]}\\
+\noindent \texttt{             statistic}\\
+\noindent \texttt{crown : stem     10000}\\
+\noindent \texttt{ }\\
+\noindent \texttt{[[2]]}\\
+\noindent \texttt{                  p.value}\\
+\noindent \texttt{crown : stem 2.562144e-34}\\
+
+\bigskip
+As indicated by the p.value, there is a significant difference in disparity between both groups.
+
+\subsubsection{Typical disparity through type analysis}
+
+Another type of disparity analysis can be disparity through time analysis.
+Using the same definition of the multidimensional space and disparity as in the previous example, one can be interested in how disparity changed through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).
+
+\texttt{> time\_bins <- c(100.5, 66, 56, 33.9)}
+
+It is then possible to measure disparity through time using the following function:
+
+\texttt{> disparity <- dispRity.through.time(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, metric = c(sum, variances))}
+
+\noindent Note that this function is a wrapping function that is the equivalent to:
+
+\texttt{> subsamples <- time.subsamples(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins)}\\
+\texttt{> bootstraps <- boot.matrix(subsamples)}\\
+\texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\
 
 \noindent Which allows a finer tuning of each optional argument in each of the three functions \texttt{time.subsamples}, \texttt{boot.matrix} and \disp.
-The three arguments here are pretty self explanatory: \texttt{data = BeckLee\_mat50} is our definition of the multidimensional space, \texttt{time = c(90, 70, 50, 30)} indicates the boundaries of the different time bins and \texttt{metric = c(sum, variances)} is our definition of disparity.
+The arguments \texttt{data = BeckLee\_mat50} and \texttt{metric = c(sum, variances)} are the same as in the example above.
+However, in this type of analysis, one also needs to have additional arguments: the \texttt{time = time\_bins} indicates the boundaries of the different time bins and the \texttt{tree = BeckLee\_tree} argument provides information on the age of each element.
 
-This function returns a \disp object that does not directly gives us the results (the disparity in each time bin) but rather summarises the content of the \disp object.
-To visualise the disparity values, one can use the \texttt{summary} or/and \texttt{plot} options:
+Again, the resulting \disp object can be summarised and plotted (table \ref{Tab:summary_time} and figure \ref{Fig:plot_time}):
 
-\texttt{summary(disparity)}
+\texttt{> summary(disparity)}
 
-\texttt{plot(disparity)}
+% latex table generated in R 3.4.1 by xtable 1.8-2 package
+% Mon Aug 21 12:21:36 2017
+\begin{table}[ht]
+\centering
+\begin{tabular}{rlrrrrrrr}
+  \hline
+ & subsamples & n & obs & bs.median & 2.5\% & 25\% & 75\% & 97.5\% \\ 
+  \hline
+1 & 100.5 - 66 &  15 & 1.67 & 1.55 & 1.40 & 1.51 & 1.58 & 1.65 \\ 
+  2 & 66 - 56 &   9 & 1.88 & 1.69 & 1.43 & 1.63 & 1.77 & 1.83 \\ 
+  3 & 56 - 33.9 &  13 & 1.96 & 1.83 & 1.62 & 1.77 & 1.86 & 1.90 \\ 
+   \hline
+\end{tabular}
+\caption{Summarising a \disp object (disparity through time)}
+\label{Tab:summary_time}
+\end{table}
 
-% The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
-% This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
-% Of course all the defaults can be changed if required, more on this later.
+\texttt{> plot(disparity, type = 'countinuous')}
 
-% For a disparity through time analysis, you will need:
+\noindent Note that many plot options specific to \disp objects are available such as plotting disparity in a ``continuous'' fashion (inferring disparity between the time bins).
 
-%   * An ordinated matrix (we covered that above)
-%   * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element.
-To give your tree a root time (i.e.
-an age for the root), you can simply do `my_tree\$root.time <- my\_age`.
-%   * The required number of time subsamples (here `time = 3`)
-%   * Your favourite disparity metric (here the sum of variances)
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{plot_example_time.pdf} 
+\caption{\disp plot of disparity through time}
+\label{Fig:plot_time}
+\end{figure}
 
-% Using the @beckancient2014 data described [above](#example-data):
+From the plot and the summary, disparity seems to increase through time from the Cretaceous to the Eocene.
+Similarly to the example above, it is also possible to statistically test this hypothesis by running a simple ANOVA for example (Table \ref{Tab:anova}):
 
-% ## Measuring disparity through time
-% disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
-%                                         time = 3, metric = c(sum, variances))
-% ```
+\texttt{> anova(test.dispRity(disparity, test = lm, comparisons = "all"))}
 
-% This generates a `dispRity` object (see [here](#manipulating-dispRity-objects) for technical details).
-% When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrrrr}
+  \hline
+ & Df & Sum Sq & Mean Sq & F value & Pr($>$F) \\ 
+  \hline
+subsamples & 2 & 3.14 & 1.57 & 177.84 & 0.0000 \\ 
+  Residuals & 297 & 2.62 & 0.01 &  &  \\ 
+   \hline
+\end{tabular}
+\caption{Anova performed on a \disp object}
+\label{Tab:anova}
+\end{table}
 
-% ```{r}
-% ## Print the disparity_data object
-% disparity_data
-% ```
+\section{Additional informations}
+\subsection{Manuals and vignette}
+Supplementary information about the package is readily available on the project page \url{https://TGuillerme/dispRity}.
+Additionally, information concerning each functions can be found in R in the individual function manuals an a vignette with more detailed information is available online at \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}.
+This vignette contains substantial information on the functionalities of the package not described in this succinct description.
+It also contains a more detailed tutorial of a ``classic'' disparity analysis in palaeobiology as well as an introduction to the use of this package in ecology or other disciplines.
 
-% We asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
+\subsection{Data simulations}
+This package also contains a series of simulation functionality for generating random discrete morphological matrices (\texttt{sim.morpho}) or random multidimensional spaces (\texttt{space.maker}).
+This functions are based on a similar modular architecture as the \disp functions allowing users to provide their own distribution descriptions as function as parameters for the simulation.
+For example \texttt{stats::rnorm} can be provided for drawing normal characters rates with \texttt{sim.morpho} or normally distributed multidimensional space with \texttt{space.maker}.
+The discrete morphological data simulations are based on protocols from \cite{GuillermeCooper}, \cite{OReilly20160081} and \cite{puttick2017uncertain}. 
+The multidimensional space simulations are based on the methods from \cite{diaz2016global}.
+Both functionalities are described in more details in the package vignette.
 
-% We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):
+\subsection{Further directions}
+Further functionalities and improvement on current functionalities will be implemented in future versions.
+In fact, optimised architecture and C implementations are being currently developed for faster calculations and lower memory footprints for most computational heavy functions (e.g.
+\disp or \texttt{time.subsamples}).
+New functionalities currently being developed will include time sequential linear models (for accurately measuring directionality in disparity changes between time slices) or mode of disparity evolution testing (allowing to fit different models such as Brownian Motion or Ornstein-Uhlenbeck to disparity curves).
 
-% ```{r, fig.width=7, fig.height=7}
-% ## Summarising disparity through time
-% summary(disparity_data)
+\section{Conclusion}
+The \disp package is based on a highly modular architecture allowing researchers to simply define both their multidimensional space and their disparity metric for efficiently analysing multivariate data.
+The \disp object allows to easily pipeline disparity analysis from the data input (the matrix) to publication standard results (tables, plots, hypothesis testing).
+With the help of this package, researchers will thus be able to run multivariate analysis in an easy way while still allowing a maximum of plasticity necessary to the specificity of such analysis.
 
-% ## Plotting the results
-% plot(disparity_data, type = "continuous")
-
-% ## Testing for an difference among the time bins
-% disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
-% summary(disp_lm)
-
-% ## Note that ANOVA assumptions were not checked here!
-% ```
-
-% Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
-% You can also directly explore the specific function help files within R and navigate to related functions.
-
-\subsection{Typical disparity per groups analysis}
-
-% The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
-% For example, you could ask if there is a difference in disparity between two groups?
-
-% To perform such an analysis, you will need:
-
-%  * An matrix with rows as elements and columns as dimensions (always!)
-%  * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix.
-For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4.
-Note that elements can be present in multiple groups at once.
-%  * Your favourite disparity metric (here the sum of variances)
-
-% Using the @beckancient2014 data described [above](#example-data):
-
-% ```{r}
-% ## Creating the two groups (crown versus stem) as a list
-% mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-%                       "stem" = c(1:15, 17:18, 42:44))
-
-% ## Measuring disparity for each group
-% disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
-%                                      metric = c(sum, variances))
-% ```
-
-% We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).
-
-% ```{r, fig.width=7, fig.height=7}
-% ## Print the disparity_data object
-% disparity_data
-
-% ## Summarising disparity in the different groups
-% summary(disparity_data)
-
-% ## Plotting the results
-% plot(disparity_data)
-
-% ## Testing for a difference between the groups
-% test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
-% ```
-
-\section{Supplementary informations}
-Supplementary information concerning each function's functionalities can be found in R in the individual function manuals as well as in the vignettes available online at @@@.
-
--Miscellaneous (maybe conclusion)
-Simulations
-Future directions
-
-Conclusion
-
-% \section{Data availability and reproducibility}
-% Data are available on Dryad or Figshare.
-% Code for reproducing the analyses is available on GitHub (\url{github.com/TGuillerme/SpatioTemporal_Disparity}).
+\section{Data availability and reproducibility}
+All the code and the data for using the package and it's vignette examples is readily available in the package or on GitHub \url{github.com/TGuillerme/dispRity}.
 

### 2017-08-23 paper/dispRity_paper.tex
editors suggestions
Updates to manuscript

@@
 \begin{enumerate}
-    \item We present \disp, an \texttt{R} package in multidimensional spaces. 
-    \item Disparity designates a suit of metrics to describe an ordinated matrix (also referred as morpho/eco/niche/etc.-space).
-    \item The package is based on a highly modular architecture were metrics for measuring disparity and tests for testing evolutionary hypothesis are provided as standalone functions.
-    \item This modular architecture allows great plasticity in using this package, namely by letting users define their own metric of disparity.
-    \item The package also provides numerous tools for modifying, summarising and plotting \disp objects.
+    \item We present \disp, an \texttt{R} package in for multidimensional spaces analysis.
+    \item Multidimensional spaces (e.g. morpho/eco/etc.-space) can described using a suit of metrics (e.g. disparity metrics) an many statistical tests.
+    \item The package provides a flexible framework were the space, metric and test can be easily defined by the user to analyse their specific ecological or evolutionary hypothesis.
+    \item Its architecture is based on the \disp object with a variety of tools for modifying, summarising or plotting it.
+    \item The package's modularity and simple interface allows to run easily reproducible multidimensional analysis.
 \end{enumerate}
@@
 
-% Multidimensionality is a major aspect of natural sciences were variables co-vary through time and space from the organism to the environmental level.
+% Multidimensionality
+Biological data are complex.
+To understand the ecology and evolution of species we must use multiple variables that inevitably covary with each other through time and space.
+One solution to this problem is to analyse this data in a multivariate framework \citep[e.g.][]{price2015predation,diaz2016global}.
+Such analysis aim to capture complex biological multidimensionality while still providing outputs that are interpretable in our physical world restricted to three dimensions.
+These biological variables are as diverse as morphological traits (discrete, e.g.
+\citealt{brusattedinosaur2012}; or continuous, e.g.
+\citealt{price2015predation}), life history traits \citep[e.g.][]{diaz2016global}, ecosystem properties \citep[e.g.][]{DonohueDim}, etc.
+And can be used to answer specific questions such as changes in morphological diversity through time \citep{Close2015}, testing competitive replacement scenarios \citep{Brusatte12092008}, what is the relation between form and function in organisms \citep{diaz2016global} or even describing the entirety of possible shapes for a group of organisms \citep{raup1966geometric}.
 
-% In palaeobiology multidimensionality is a popular way to study morphological diversity and comes as a interesting alternative to unidimensional taxonomic diversity analysis \citep{Hopkins2017}.
+% Variety of multidimensional spaces, metrics and tests
+In all these analysis, each of these multivariate traits form different multidimensional spaces represented as matrices where the columns are the dimensions (or some transformation thereof) and each rows are the elements available within this space.
+These multidimensional spaces can be defined in many ways such as a pairwise distance matrix \citep{Close2015}, a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, a multidimensional scaling \citep[MDS;][]{DonohueDim}, etc.
+Once this multidimensional space is defined, say morphospace (if its variables are base on morphology), it is then possible to measure a specific aspect of it, say its morphological diversity (more commonly by as disparity; \citealt{Hopkins2017}).
+This specific measurement (hereafter referred to as disparity metrics) can in turn be defined in many different ways depending on the specific question \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015}.
+Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical test such as non-parametric multivariate analyses of variance \citep[e.g.][]{Brusatte12092008}, multidimensional permutation test \citep[e.g.][]{diaz2016global} or even simply by looking at the confidence interval overlap between disparity measurements \citep[e.g.][]{halliday2016eutherian}.
 
-% Although the concept of morphological disparity is well defined since more than two decades \citep[e.g.][]{gould1991disparity,Foote01071994,Foote29111996}, its specific definition varies widely among authors.
-% In fact disparity Disparity is by nature multidimensional \citep{lloyd2016estimating} and thus there are many ways to measure it.
+% Problems with this variety
+In theory, this multitude of ways to generate multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods \citep[such as][for biogreography]{matzke2013biogeobears}.
+In practice, however, this is hampered by existing software implementations.
+Even though many software exist for multidimensional analysis \citep[e.g.][]{bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric}, package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric and then enforce this on users by allowing few to no flexibility.
+For example, in the excellent and widely used \texttt{geomorph} package morphological disparity analysis can be ran using the \texttt{morphol.disparity} function defining the multidimensional space as the ordination of the Procrustes transform of morphometric data, the disparity metric as the Procrustes variance and permutation tests can be used \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
+This can lead to a chain of inappropriate analyses led by users just using the existing software when their question/data is not appropriate. %Oh no!
 
-%TG: rewrite all that using bits and blobs from previous intro
-% Biological data are complex; understanding the ecology and evolution of species often requires that we analyse multiple variables that covary with each other, and through space and time.
-% Multivariate analyses aim to capture and incorporate this multidimensional complexity, while providing outputs that are interpretable without needing to visualise n-dimensions simultaneously.
-% Key multidimensional features of species that have important roles in ecology and evolution include morphology, functional traits. etc. etc.
-% Many of these can be represented as matrices, these can be used in our package.
-% In the interests of clarity/brievety, we here focus on just one kind of data: morphological diversity.
-
-% In evolutionary biology and palaeontology, morphological diversity is often referred to as disparity.
-% DESCRIBE COOL USES OF DISPARITY ANALYSES. WHY IS DISPARITY AWESOME.
-% Although disparity is commonly studied, its definition varies widely across authors due to the myriad ways it can be measured.
-% LIST WAYS IT IS MEASURED.
-% This is further complicated by confusion over whether diparity refers to the metric describing or summarising one or more aspects of this space, or to the multidimensional space which is a specific mathematical object.
-% This multidimensional space can also be defined in multiple ways.
-% LIST ways
-
-% In theory, this multitude of ways to measure/define disparity is not an issue. [the problem is that people aren't clear about it]
-% The ideal solution is to choose either the most appropriate method for your question or data, or to apply several definitions and compare across these.[check biogeoBEARS]
-% Unfortuantely, in practice this is hampered by existing software implementations.
-% Package maintainers/software developers choose their preferred definition of disparity and method for measuring it, and then enforce this on users by allowing zero flexibility.
-% This can lead to a chain of inappropriate analyses led by everyone just using the package that exists and the method within it, and in the worst case, using the defaults of the package. Oh no!
-% The dispRity package will solve this problem by providing a completely flexible framework for studing disparity.
-% It implements all commonly used metrics and definitions of disparity, as well as providing a simple interface for users to implement their own disparity functions.
-% The package is described here for use with morphological diversity data, but the functions are equally applicable to ecological contexts such as Diaz paper, Deirdre's paper etc.
-
-Biological data are complex; understanding the ecology and evolution of species often requires that we analyse multiple variables that covary with each other, and through space and time.
-One solution to this problem is to analyse data in a multivariate framework.
-Multivariate analyses aim to capture and incorporate this multidimensional complexity, while providing outputs that are interpretable in a physical world of only three dimensions.
-Classical multidimensional species features that have been analysed through multivariate data ranges from their morphology \citep{raup1966geometric}, to their functional traits \citep{diaz2016global}.
-However, such analysis are not limited at the species level and can also be applied to ecosystems \citep{DonohueDim}.
-
-Key multidimensional features of species that have important roles in ecology and evolution include morphology, functional traits. etc. etc.
-Many of these can be represented as matrices, these can be used in our package.
-In the interests of clarity/brievety, we here focus on just one kind of data: morphological diversity.
-
-Furthermore, we will also make a clear distinction between the multidimensional space which is a specific mathematical object and the disparity, which is a metric describing or summarising one or more aspects of this space.
-
-Both can be defined in various ways depending on the authors, for example, disparity is defined as the weighted mean pairwise dissimilarity in \cite{Close2015} or the ellipsoid volume in \cite{DonohueDim} and the multidimensional space is defined as the morphospace in \cite{raup1966geometric} and the morpho-functional space in \cite{diaz2016global}.
-
-The multidimensional space can be defined in many ways and arise from many mathematical transformations of the data such as the pairwise distance matrix \citep{Close2015}, a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, a multidimensional scaling \citep[MDS;][]{DonohueDim}, etc.
-Similarly, disparity metrics \citep[or indices;][]{Hopkins2017} can defined in many ways \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015,diaz2016global}.
-Finally, difference between disparity metrics can also be measured in many ways: using NPMANOVA \citep[e.g.][]{Brusatte12092008}, multidimensional permutation test \citep[e.g.][]{diaz2016global} or even simple confidence interval overlap \citep[e.g.][]{halliday2016eutherian}.
-
-This variety of definitions and analysis have been developed in an equal variety of softwares such as \texttt{GINGKO} in javascript \citep{bouxin2005ginkgo,de2007ginkgo} or \texttt{geomorph} \citep{adams2013geomorph,adams2017geometric}, \texttt{Claddis} \citep{Claddis}, or \texttt{vegan} \citep{oksanen2007vegan} in R \citep{R}.
-This results in the need to learn different languages (or at least - when restricted to R - different packages with different standards) as well as making analysis sometimes idiosyncratic and often complex to repeat since they are based on a particular feature from a particular software.
-For example, in the excellent and widely used \texttt{geomorph} package morphological disparity analysis can be ran using the \texttt{morphol.disparity} function.
-Unfortunately, however, the multidimensional space can only be defined as the ordination of the procrustes transform of geometric morphometric landmarks, the disparity can only be defined as the Procrustes variance and the difference between groups can only be measured through permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
-
-This package, is the first, to our knowledge, to be entirely dedicate to multidimensional analysis.
-It has a highly modular architecture and allow users to define their own multidimensional space, use their own definition of the disparity metric and measure the differences between metrics in their own terms.
-Furthermore, both a specific design and set of functions (through the \disp object) allows to design analysis into easy and repeatable pipelines.
+% Here comes the solution!
+The aim of the \disp package is to solve such problems by providing a completely flexible framework for studying multidimensional data.
+In fact, this package is based on a highly modular architecture where each crucial decision in multidimensional analysis (which data, which metric and which test) can be provided by the user.
+It implements many commonly used metrics and definitions of disparity as well as providing a simple interface for users to implement their own disparity functions.
+The package is described here for the use of discrete morphological data diversity analysis but can be generalised to any type of multidimensional data (see the package vignette for more examples).
 
 \section{Description}
-All the following functionalities of the package described below are based on the \disp class object.
-This object class allows to repeat and streamline disparity analysis by being past from one function to the other (see Example section).
-Specificities of the \disp object can be found in @@@ as well as series of utilities functions allowing to modify such object (see @@@).
+In brief, the package intakes a \texttt{matrix} object (the multidimensional space), calculates some aspect of it (the disparity metric) and allows to analyse the resulting \disp object (see Supplementary @@@ for details on the object content).
+Some additional functions allow to modify the multidimensional space such as dividing it per group or through time or/and bootstrapping it (see Fig. \ref{Fig:workflow} and Table \ref{Tab:main_fun}).
+See below for a more detailed description of features.
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{workflowsvg.pdf} 
+\caption{\disp package workflow: rectangles represent matrices; ellipse represent functions; plain black arrows indicate input/output; dashed grey arrows indicate output (though the summary, plot, and test function can not be applied if no disparity has been calculated).}
+\label{Fig:workflow}
+\end{figure}
 
@@
         \texttt{Claddis.ordination} & Imports data from the \texttt{Claddis} package. \\
-        \texttt{custom.subsamples} & Separates ordinated data in custom subsamples. \\
-        \texttt{time.subsamples} & Separates ordinated data in time subsamples. \\
-        \texttt{boot.matrix} & Bootstraps and rarefies ordinated data or a \disp object. \\
-        \disp & Calculates disparity from an ordinated matrix or a \disp object. \\
+        \texttt{custom.subsamples} & Separates data in custom subsamples. \\
+        \texttt{time.subsamples} & Separates data in time subsamples. \\
+        \texttt{boot.matrix} & Bootstraps and rarefies data or a \disp object. \\
+        \disp & Calculates disparity from an matrix or a \disp object. \\
         \texttt{summary.dispRity} & Summarises a \disp object. \\
@@
     \caption{Main functions in the \disp package.}
+    \label{Tab:main_fun}
 \end{table}
@@
 \subsection{Measuring disparity}
-As hinted in the package name, the main functionality of this package is to measure disparity in a multidimensional space.
-Again, within this package, disparity is defined as a description or a summary of one or more aspects of the multidimensional space.
+The \disp function allows to measure disparity from any multidimensional space (of class \texttt{matrix} or \texttt{data.frame}) where the columns correspond dimensions and the rows correspond to the elements present in the space.
+The disparity metric (i.e.
+the aspect of interest of the space) is passed through the \texttt{metric} argument and is defined by the user as any function or combination of functions that can either transform the matrix into:
 
-The \disp function allows to measure disparity from any matrix (of class \texttt{matrix} or \texttt{data.frame}) where the columns correspond to the different dimensions of the multidimensional space (e.g.
-the ordination axis or any variable composing the space) and the rows correspond to the elements present in the multidimensional space (e.g.
-species, individuals, study sites, etc...).
-
-Disparity can then be defined by the user as any function or combination of functions that can either transform the matrix into:
 \begin{itemize}
-    \item Another matrix (a Dimensions-level 3 function)
-    \item A vector (a Dimensions-level 2 function)
-    \item A single value (a Dimensions-level 1 function)
+    \item Another matrix (a dimensions-level 3 function - e.g. a variance-covariance matrix; \texttt{stats::cor})
+    \item A vector (a dimensions-level 2 function - e.g. the variance of each dimension; \texttt{dispRity::variances} - see below)
+    \item A single value (a dimensions-level 1 function - e.g. the overall standard deviation; \texttt{stats::sd})
 \end{itemize}
@@
 \centering
-   \includegraphics[width=1\textwidth]{../inst/gitbook/dispRity_fun.pdf} 
-\caption{Illustration of the different metric dimensions levels in the \disp package}
+   \includegraphics[width=0.8\textwidth]{../inst/gitbook/dispRity_fun.pdf} 
+\caption{Illustration of the different metric dimensions levels in the \disp package. In this example, each cell correspond to a single value (e.g. a 8 $\times$ 7 matrix or a vector of 8 elements)}.
 \label{Fig:levels}
@@
 
-The function(s) can then be passed to the \texttt{metric} argument to measure a specific aspect of the multidimensional space:
-There is no limitation on the metric dimensions only that at least one dimensions level 2 or 1 are present in the definition (i.e.
-disparity can only be defined as distribution - a vector - or a single value).
-
-Some metrics are implemented in the package
+The metrics can be functions from previous packages, defined by the user or from the \disp package (see Table \ref{Tab:metrics} for the currently implemented metrics).
+When multiple functions are passed the \texttt{metric} argument, the functions are sorted per dimension level and applied in decreasing order to the data.
+For example, if the metric is defined as \texttt{metric = c(mean, range)}, the \texttt{range} function (dimension level 2) is first applied to the data and the function \texttt{mean} is then applied to the result (e.g.
+\texttt{mean(range(data))}).
+There is no limitation on the metric dimensions only that at least one dimensions level 2 or 1 are present in the metric definition (i.e.
+disparity can only be defined as distribution or a single value).
 
 \begin{table}
-\center
-    \begin{tabular}{r|l|c|c|c}
-        name & description & dimension level & definition & source \\
+\resizebox{\textwidth}{!}
+{%
+    \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
+        name & description & dimen-sion level & definition & source \\
         \hline
-        \texttt{ellipse.volume}$^1$ & The volume of the ellipsoid around the elements & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
-        \texttt{convhull.surface} & The surface of the convex hull formed by the elements & 1 & NA & \texttt{geometry::convhulln} \\
-        \texttt{convhull.volume} & The volume of the convex hull formed by the elements & 1 & NA & \texttt{geometry::convhulln} \\
-        \texttt{diagonal} & The longest Eucliden distance between any pair of elements & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & \texttt{dispRity::diagonal} \\
-        \texttt{mode.val} & The modal value of a vector or a matrix & 1 & NA & \texttt{dispRity::mode.val}\\
+        \texttt{ellipse. volume}$^1$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
+        \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
+        \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
+        \texttt{diagonal} & The longest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & \texttt{dispRity::diagonal} \\
+        \texttt{mode.val} & The modal value & 1 & NA & \texttt{dispRity::mode.val}\\
         \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & \texttt{dispRity::ranges} \\
@@
     \end{tabular}
-    \caption{Where $k$ is the number of dimensions, $n$ the number of elements, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean.
-$^1$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA); $^2$ by default that point is the centroid of the elements.}
+}%
+    \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean.
+$^1$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA); $^2$ by default that point is the centroid of the elements.}}
     \label{Tab:metrics}
@@
 
+%STOPPED HERE
+
 \subsection{Splitting the multidimensional space into subsets}
-A common analysis using multidimensional space consists into subdividing it into smaller samples, typically to compare to each other.
-For example, one can be interested in looking how a specific sub-samples of the space compares in disparity from another (e.g.
+Prior to calculating disparity, the multidimensional space can be subdivided into subsamples, typically to be compared to each other.
+For example, one can be interested in looking how the disparity of a specific subsample of the space compared to another (e.g.
 difference among groups) or, how different sub-samples change sequentially (e.g.
 through time).
-In essence, the original multidimensional space (the input matrix) corresponds to overall space.
-For example, if the multidimensional space is a morphospace, it corresponds to all the possible morphologies observed in the matrix.
-Conversely, sub-samples of this space correspond to parts of the space with pooled characteristics.
-For example, still in a morphospace with time sub-samples, each sub-samples correspond to all the observed morphologies during some time period.
+In essence, the original space corresponds to overall space (e.g. a morphospace contains all the possible observed morphologies).
+Conversely, subsamples correspond to parts of the space with pooled characteristics (i.e. only a subset of the total morphospace).
 
@@
 
-\subsubsection{test}
+\subsubsection{testing hypothesis}
 The \texttt{test.dispRity} function allows to test hypothesis on the disparity data (e.g. by comparing the different sub-samples between each other).
@@
 The \disp object also displays information on the number and method of bootstrap replicates as well as how disparity was measured.
-To visualise the actual disparity values, one can use the \texttt{summary} or/and \texttt{plot} options (table \ref{Tab:summary_group} and figure \ref{Fig:plot_group}):
+To visualise the actual disparity values, one can use the \texttt{summary} or/and \texttt{plot} options (Table \ref{Tab:summary_group} and Fig. \ref{Fig:plot_group}):
 
@@
 
-Again, the resulting \disp object can be summarised and plotted (table \ref{Tab:summary_time} and figure \ref{Fig:plot_time}):
+Again, the resulting \disp object can be summarised and plotted (Table \ref{Tab:summary_time} and Fig. \ref{Fig:plot_time}):
 
@@
 
+%Editors:
+%Sam Price, Daniele Silvestro
+
+%Reviewers
+%Graeme Lloyd, Dave Bapst, Emma Sherratt
+
 \bibliographystyle{sysbio}

### 2017-08-24 paper/CoverLetter.tex
cover letter

\documentclass[11pt]{letter}
\usepackage[a4paper,left=2.5cm, right=2.5cm, top=1cm, bottom=1cm]{geometry}
\usepackage[osf]{mathpazo}
\signature{Thomas Guillerme}
\address{Imperial College London \\Silwood Park Campus \\Buckhurst Road \\Ascot SL57PY, United Kingdom \\guillert@tcd.ie}
\longindentation=0pt
\begin{document}

\begin{letter}{}
\opening{Dear Editors,}

Studying multivariate data as an alternative to univariate data has become increasingly popular in the last decades both in ecology and evolution.
This includes, for example, studying form and function (e.g. Diaz et al., 2016, \textit{Nature}) or morphological diversity (i.e. disparity - e.g. Brusatte et al. 2012, \textit{Science}).
These methods have in common to use a multidimensional space (through ordination or other techniques) representing the combination of all possible observed traits and then measuring some aspect of the space (e.g. its hyper-volume) for different groups and compare them between each other or to a null hypothesis.

Hence there exist a multitude of ways to define a multidimensional space, a metric measuring a specific aspect thereof and a test for many hypothesis.
In theory this vast tool-kit should allow evolutionary biologists and ecologists to do many things.
There are numerous implementation of such analysis pipeline in numerous R packages (e.g. \texttt{Claddis}, \texttt{geomorph}, \texttt{ade4}, \texttt{vegan}, etc.) or even in other languages (e.g. \texttt{GINGKO} in javascript).
However, each of these implementations are particular to the software/package author's definition of the multidimensional space, metric and test.
This leads researchers to often use these tool-kits by default hampering them to use the best definitions of the space/metric/test fitted to their particular question.
At worst this leads in turn to wrong statistic claims, and, at best to time consuming code retro-engineering.

In this application manuscript entitled ``dispRity: an R package for measuring disparity'', I present a complete and flexible R package allowing to run multidimensional analysis where each parameter (the space, the metric and the test) can be easily defined by the user.
In fact, this package provides a tidy and modular environment where users can use their own multidimensional space defined as a matrix and apply to it any metric and any test (provided as functions).
This allows for highly modular multidimensional analysis and the tidy interface allows easily reproducible analysis.

I look forward to hearing from you soon,

\closing{Yours sincerely,}

\end{letter}
\end{document}

%Editors:
%Sam Price, Daniele Silvestro

%Reviewers
%Graeme Lloyd, Dave Bapst, Emma Sherratt

### 2017-08-24 paper/dispRity_paper.tex
Updated paper

@@
 \section{Description}
-In brief, the package intakes a \texttt{matrix} object (the multidimensional space), calculates some aspect of it (the disparity metric) and allows to analyse the resulting \disp object (see Supplementary @@@ for details on the object content).
+In brief, the package intakes a \texttt{matrix} object (the multidimensional space), calculates some aspect of it (the disparity metric) and allows to analyse the resulting \disp object (see Manuals and Vignettes below).
 Some additional functions allow to modify the multidimensional space such as dividing it per group or through time or/and bootstrapping it (see Fig. \ref{Fig:workflow} and Table \ref{Tab:main_fun}).
@@
 
-%STOPPED HERE
-
 \subsection{Splitting the multidimensional space into subsets}
@@
 
-These divisions can readily be created in the package using the \texttt{custom.subsamples} or \texttt{time.subsamples} functions.
-The first takes as argument the matrix defining the multidimensional space and then a list of elements to group into different sub-samples.
-The second also takes a multidimensional space and an additional argument on the age of the elements and then sorts them per time categories based on the following methods:
+This can be done using the \texttt{custom.subsamples} or \texttt{time.subsamples} functions.
+The first takes as argument the matrix defining the multidimensional space and a list of elements to group into different sub-samples.
+The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time sub-samples for example) and which subsamples to create: (1) discrete time subsamples (or time-binning) or (2) continuous time subsamples (or time-slicing).
 
-\begin{itemize}
-    \item Discrete time subsamples: where the elements are grouped by age in time bins with corresponding ages.
-    \item Continuous time subsamples: where the elements are defined as being the elements present at specific points in time (see @@@).
-\end{itemize}
-
-When using the \texttt{time.subsamples} function, it is also possible to provide a list of first and last occurence data for taxa spanning across specific time periods.
-This allows species with a longer timespan to be present in multiple time sub-samples.
-
-The continuous time sub-samples method works by using a phylogeny and looking at which taxa are present at any specific point in time.
-This method thus requires a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
-
-The slicing works by selecting any tips (and nodes if \texttt{inc.nodes = TRUE}) at the specific time slice.
-When a slice occurs not on a tip or a node, four methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice.
-The models are:
+The time-bining method groups elements by specific age range.
+The time slicing method works by using a phylogeny and looking at which taxa are present at any specific point in time.
+This method thus requires the nodes to be part of the multidimensional space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
+When a slice occurs not on a tip or a node, four methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
 \begin{itemize}
@@
 \subsection{Bootstrapping and rarefying}
-When measuring disparity, it is important to take sampling into account: for example, observed disparity in the whole multidimensional space can simply due to factors like the specific selection of elements (both their number or the presence of outliers).
-To take these sources of error into account, it is common practice to bootstrap the data (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different sub-samples).
+Disparity measurement can be influenced by sampling \citep{Butler2012}.
+To take this source of bias into account, it is common practice to bootstrap the multidimensional space (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different sub-samples).
+Additionally, if disparity is defined as a dimension-level 1 metric, it can be useful to measure it on bootstrapped data in order to have a distribution on which to perform statistical analysis.
 
-\subsubsection{Bootstraps}
-In disparity analysis, if disparity is defined as a dimension-level 1 metric (i.e.
-a metric summarising the multidimensional space into a single value), it can be useful to bootstrap this value in order to have a distribution on which to perform statistical analysis.
-This can be achieved by using the \texttt{boot.matrix} function pseudo-replicates the multidimensional space following these two algorithms:
+Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the multidimensional space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' one where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).
 
-\begin{itemize}
-    \item ``full'' algorithm: where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data)
-    \item ``single'' algorithm: where only one random element is replaced by one other random element for each pseudo-replicate
-\end{itemize}
+% In the case of disparity being a dimension-level 1 metric, it is important to note that the distribution resulting from a bootstrap will be a pseudo-replicated one and thus might not be suitable for some specific parametric statistical tests.
+% Conversely, it is possible to use a dimension-level 2 metric on a bootstrapped multidimensional space to avoid this problem: the disparity metric is then measured on a pseudo-replicated multidimensional space rather than the metric itself being pseudo-replicated.
 
-\noindent Note that the ``single'' algorithm is somewhat akin to jackknife re-sampling where one element is duplicated rather than removed.
-
-In the case of disparity being a dimension-level 1 metric, it is important to note that the distribution resulting from a bootstrap will be a pseudo-replicated one and thus might not be suitable for some specific parametric statistical tests.
-Conversely, it is possible to use a dimension-level 2 metric on a bootstrapped multidimensional space to avoid this problem: the disparity metric is then measured on a pseudo-replicated multidimensional space rather than the metric itself being pseudo-replicated.
-
-\subsubsection{Rarefactions}
-The same \texttt{boot.matrix} function also allows to rarefy the multidimensional space or its sub-samples.
-Rarefaction can be useful if, for example, one is interested in looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of sub-samples wit the same number of elements in each.
-In practice, rarefaction allows users to set a limit to the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).
-
-Additionally, this argument, is set to \texttt{TRUE} also allows to fully rarefy the sub-samples (i.e.
-removing $x$ elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsample to a minimum of three elements).
+Similarly, rarefaction can be achieved through the same \texttt{boot.matrix} function.
+Rarefaction can be useful when looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of sub-samples wit the same number of elements in each.
+In practice, rarefaction limits the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).
 
 \subsection{Interpreting results}
-Once disparity has been calculated (without or without sub-sampling, bootstrapping and rarefaction), the resulted \disp object can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity}.
-Finally, the results can be analysed using the \texttt{test.dispRity} function.
+The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity} (provided disparity was calculated).
+These results can then be analysed using the \texttt{test.dispRity} function for comparing sub-samples or testing hypothesis.
 
 \subsubsection{Summary and plot}
-The summary and plot functions allow to summarise the contents of the \disp objects.
-These functions take as common arguments which central tendency (e.g.
+These \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow to set which central tendency and which quantiles should be represented.
+% as common arguments which central tendency (e.g.
 if central tendency is the mean and disparity was calculated as the sum of the variances for each dimensions, the central disparity tendency would be the mean of the sum of the variances) and the quantiles to represent (in the same example, if the quantile chosen is the 50th, these functions will also display 50\% of the sum of variances around the mean).
-
-Additionally, the \texttt{plot.dispRity} function allows to graphically represent the summarised results using different representations:
-
-\begin{itemize}
-    \item ``continuous'' for displaying continuous disparity curves (in the case of disparity through time analysis for example)
-    \item ``box'', ``lines'', or ``polygons'' to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.
-\end{itemize}
-
-Additional arguments can also be used such as \texttt{observed} to display the observed disparity (i.e.
-non-bootstrapped), \texttt{add} to add the plot in the current quartz window, or \texttt{rarefaction} to only plot the disparity for a specific number of elements and of course, any other normal graphical options can also be passed to this function (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...).
-Note that if the disparity has been calculated on fully rarefied data (using \texttt{boot.matrix(data, rarefaction = TRUE)} prior to calculating disparity), the rarefaction curves for the whole multidimensional space (or its sub-samples) can be plotted using \texttt{rarefaction = ``plot''}.
+Additionally, the \texttt{plot.dispRity} function allows to graphically represent the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves (in the case of disparity through time analysis for example) and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
+Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e.
+non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for
+The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...).
+% Note that if the disparity has been calculated on fully rarefied data (using \texttt{boot.matrix(data, rarefaction = TRUE)} prior to calculating disparity), the rarefaction curves for the whole multidimensional space (or its sub-samples) can be plotted using \texttt{rarefaction = ``plot''}.
 
@@
 Similarly to the \disp function described above, this one can take any test defined by the user.
-The \texttt{comparison} arguments allows to indicate in which order (if any) should the tests be applied:
-
-\begin{itemize}
-    \item ``pairwise'' for applying pairwise comparisons of all the subsamples (default).
-    \item ``referential'' for comparing the first subsample to all the others.
-    \item ``sequential'' for comparing each subsample sequentially (e.g. first against second, second against third, etc.).
-    \item ``all'' for comparing all the subsamples simultaneously to the data (i.e.
-\texttt{disparity $\mathtt{\sim}$ subsamples}).
-This argument can be used for \texttt{lm} or \texttt{glm} type tests.
-    \item or any list of pairs (or more) of number of sub-samples to compare (for example, for comparing the first to the sixth subsample and the fifth to the second: \texttt{comparison = list(c(1, 6), c(5, 2))})
-\end{itemize}
+The \texttt{comparison} arguments allows to indicate in which order (if any) should the tests be applied to the subsamples: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subsample to all the others; (3) ``sequential'' for comparing subsamples sequentially (e.g.
+first against second, second against third, etc.); (4) `all'' for comparing all the subsamples simultaneously (i.e.
+\texttt{disparity $\mathtt{\sim}$ subsamples} - this can be used for \texttt{lm} or \texttt{glm}) or (5) any list of pairs (or more) of sub-samples to compare,
 
 Some tests are implemented within the package such as the Bhattacharrya Coefficient \citep[\texttt{bhatt.coeff}][]{Bhattacharyya,GuillermeCooper} or a permutation test based on null hypothesised multidimensional space following \cite{diaz2016global} (\texttt{null.test}).
-
-This function also allows additionally argument such as \texttt{rarefaction} (as described above) or \texttt{correction} to adjust eventual p-values when multiple parametric tests are ran.
+Additionally, this function also allows additionally argument such as \texttt{rarefaction} (as described above) or \texttt{correction} to adjust eventual p-values when multiple parametric tests are ran.
 
 \section{Examples}
-Here I provide to classic example of disparity analysis using some of the functions described above.
-The data to reproduce these examples is available within the package and thus don't need to be downloaded:
+For brevity reason, the following example is based on a classical palaeobiology morphological disparity analysis.
+Note the more examples are available in the package manual \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}.
 
@@
     \item \texttt{BeckLee\_mat50}: an ordinated matrix for 50 mammals based on the distance between discrete morphological characters.
-    \item \texttt{BeckLee\_mat99}: an ordinated matrix for the same 50 mammals as in \texttt{BeckLee\_mat50} but also containing the reconstruction of their 49 ancestors.
+    \item \texttt{BeckLee\_mat99}: the same matrix \texttt{BeckLee\_mat50} but also containing the reconstruction of their 49 ancestors.
     \item \texttt{BeckLee\_tree}: a chronogram with the 50 mammals species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
@@
 
+In this example, the multidimensional space is thus defined as a morphospace: the ordination of the distance between discrete morphological characters for 50 mammals species \citep[from][]{beckancient2014}.
+Additionally, we can define disparity as the sum of the variances on each dimensions \citep{Wills1994} that will represent an aspect of the the volume of the morphospace (i.e.
+a large sum of variances approximates a large morphospace and vice versa).
+
 \subsection{Typical disparity per groups analysis}
-
-As an example, one can define their multidimensional space as being the morphospace of an ordination of the distance between discrete morphological characters for 50 mammals species \citep[from][]{beckancient2014}.
-Then, one can define disparity as the sum of the variances on each dimensions \citep{Wills1994} that will represent an aspect of the the volume of the morphospace (i.e.
-a large sum of variances approximates a large morphospace).
-One can then be interested in whether the approximation of this volume is different in the living mammals and their descendants (crown) or the fossil mammals with no living descendants (stem).
+Here, we can be interested in whether the approximation of the volume within the morphospace (the disparity as defined above) is different in the living mammals and their descendants (crown) or the fossil mammals with no living descendants (stem).
 These two groups can be defined as follow (where the integers are the row number of the species in the matrix):
 
-\texttt{> mammal\_groups <- list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44))}
+\noindent \texttt{> mammal\_groups <- list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44))}
 
@@
 
-\texttt{> disparity <- dispRity.per.group(data = BeckLee\_mat50, group = mammal\_groups, metric = c(sum, variances))}
+\noindent \texttt{> disparity <- dispRity.per.group(data = BeckLee\_mat50, group = mammal\_groups, metric = c(sum, variances))}
 
@@
 
-\texttt{> subsamples <- custom.subsamples(data = BeckLee\_mat50, group = mammal\_groups)}\\
-\texttt{> bootstraps <- boot.matrix(subsamples)}\\
-\texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\
+\noindent \texttt{> subsamples <- custom.subsamples(data = BeckLee\_mat50, group = mammal\_groups)}\\
+\noindent \texttt{> bootstraps <- boot.matrix(subsamples)}\\
+\noindent \texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\
 
-\noindent Which allows a finer tuning of each optional argument in each of the three functions \texttt{custom.subsamples}, \texttt{boot.matrix} and \disp.
-The three arguments here are pretty self explanatory: \texttt{data = BeckLee\_mat50} is our definition of the multidimensional space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity.
+\noindent Which allows a finer tuning of the optional argument in each functions.
+The three arguments here are defined as follow: \texttt{data = BeckLee\_mat50} is our definition of the multidimensional space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity.
 
-This function returns a \disp object that does not directly gives us the results (the disparity in each time bin) but rather summarises the content of the \disp object.
+This function returns a \disp object that summarises the disparity analysis:
 
+\noindent \texttt{> disparity}
 \noindent \texttt{ ---- dispRity object ---- }\\
@@
 \bigskip
-As indicated, the \disp object contains 2 customised subsamples (called ``crown'' and ``stem'') from a morphospace made of 48 dimensions and 50 elements.
-The \disp object also displays information on the number and method of bootstrap replicates as well as how disparity was measured.
+As indicated, the \disp object contains 2 customised subsamples (called ``crown'' and ``stem'') from a morphospace made of 50 elements for 48 dimensions.
+The \disp object also displays information on the number and method of the bootstrap replicates as well as the definition of disparity.
 To visualise the actual disparity values, one can use the \texttt{summary} or/and \texttt{plot} options (Table \ref{Tab:summary_group} and Fig. \ref{Fig:plot_group}):
 
-\texttt{> summary(disparity)}
+\noindent \texttt{> summary(disparity)}
 
@@
 
-\texttt{> plot(disparity)}
+\noindent \texttt{> plot(disparity)}
 
@@
 
-\texttt{> test.dispRity(disparity, test = wilcox.test)}
+\noindent \texttt{> test.dispRity(disparity, test = wilcox.test)}
 
@@
 
-Another type of disparity analysis can be disparity through time analysis.
-Using the same definition of the multidimensional space and disparity as in the previous example, one can be interested in how disparity changed through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).
+Another type of disparity analysis can be a disparity through time analysis.
+Using the same definition of the multidimensional space and disparity as in the previous example, one can be interested in how disparity changes through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).
 
-\texttt{> time\_bins <- c(100.5, 66, 56, 33.9)}
+\noindent \texttt{> time\_bins <- c(100.5, 66, 56, 33.9)}
 
@@
 
-\texttt{> disparity <- dispRity.through.time(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, metric = c(sum, variances))}
+\noindent \texttt{> disparity <- dispRity.through.time(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, metric = c(sum, variances))}
 
@@
 
-\texttt{> subsamples <- time.subsamples(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins)}\\
-\texttt{> bootstraps <- boot.matrix(subsamples)}\\
-\texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\
+\noindent \texttt{> subsamples <- time.subsamples(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, method = 'discrete')}\\
+\noindent \texttt{> bootstraps <- boot.matrix(subsamples)}\\
+\noindent \texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\
 
-\noindent Which allows a finer tuning of each optional argument in each of the three functions \texttt{time.subsamples}, \texttt{boot.matrix} and \disp.
-The arguments \texttt{data = BeckLee\_mat50} and \texttt{metric = c(sum, variances)} are the same as in the example above.
-However, in this type of analysis, one also needs to have additional arguments: the \texttt{time = time\_bins} indicates the boundaries of the different time bins and the \texttt{tree = BeckLee\_tree} argument provides information on the age of each element.
+\noindent The arguments \texttt{data = BeckLee\_mat50} and \texttt{metric = c(sum, variances)} are the same as in the example above.
+However, in this type of analysis, we also needs to have additional arguments: the \texttt{time = time\_bins} indicates the boundaries of the different time bins, the \texttt{tree = BeckLee\_tree} argument provides information on the age of each element and \texttt{method = 'discrete'} indicates that the data is time-binned.
 
@@
 
-% latex table generated in R 3.4.1 by xtable 1.8-2 package
-% Mon Aug 21 12:21:36 2017
 \begin{table}[ht]
@@
 
-From the plot and the summary, disparity seems to increase through time from the Cretaceous to the Eocene.
-Similarly to the example above, it is also possible to statistically test this hypothesis by running a simple ANOVA for example (Table \ref{Tab:anova}):
+From the plot and the summary, disparity seems to change (increase) through time from the Cretaceous to the Eocene.
+Similarly to the example above, it is also possible to statistically test this hypothesis by running an ANOVA for example (Table \ref{Tab:anova}):
 
-\texttt{> anova(test.dispRity(disparity, test = lm, comparisons = "all"))}
+\texttt{> anova(test.dispRity(disparity, test = lm, comparisons = 'all'))}
 
@@
 \subsection{Manuals and vignette}
-Supplementary information about the package is readily available on the project page \url{https://TGuillerme/dispRity}.
-Additionally, information concerning each functions can be found in R in the individual function manuals an a vignette with more detailed information is available online at \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}.
-This vignette contains substantial information on the functionalities of the package not described in this succinct description.
-It also contains a more detailed tutorial of a ``classic'' disparity analysis in palaeobiology as well as an introduction to the use of this package in ecology or other disciplines.
+Supplementary information about the package is available on the project page \url{https://TGuillerme/dispRity}.
+Information concerning each functions can be found in R in the individual function manuals and a package manual with more detailed information is available online at \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}.
+This manual contains substantially more information and detailed example including a tutorial for a ``classic'' disparity analysis in palaeobiology as well as an introduction to the use of this package in ecology or other disciplines.
 
@@
 This package also contains a series of simulation functionality for generating random discrete morphological matrices (\texttt{sim.morpho}) or random multidimensional spaces (\texttt{space.maker}).
-This functions are based on a similar modular architecture as the \disp functions allowing users to provide their own distribution descriptions as function as parameters for the simulation.
-For example \texttt{stats::rnorm} can be provided for drawing normal characters rates with \texttt{sim.morpho} or normally distributed multidimensional space with \texttt{space.maker}.
+These functions are based on a similar modular architecture as the \disp functions allowing users to provide their own distribution descriptions as parameters for the simulation.
+For example, \texttt{stats::rnorm} can be provided as an argument for drawing normal characters rates with \texttt{sim.morpho} or normally distributed multidimensional space with \texttt{space.maker}.
 The discrete morphological data simulations are based on protocols from \cite{GuillermeCooper}, \cite{OReilly20160081} and \cite{puttick2017uncertain}. 
 The multidimensional space simulations are based on the methods from \cite{diaz2016global}.
-Both functionalities are described in more details in the package vignette.
+Both functionalities are described in more details in the package manual.
 
-\subsection{Further directions}
-Further functionalities and improvement on current functionalities will be implemented in future versions.
-In fact, optimised architecture and C implementations are being currently developed for faster calculations and lower memory footprints for most computational heavy functions (e.g.
+% \subsection{Further directions} TG: maybe this section is useless
+% Further functionalities and improvement on current functionalities will be implemented in future versions.
+% In fact, optimised architecture and a C implementations are being currently developed for faster calculations and lower memory footprints for most computational heavy functions (e.g.
 \disp or \texttt{time.subsamples}).
-New functionalities currently being developed will include time sequential linear models (for accurately measuring directionality in disparity changes between time slices) or mode of disparity evolution testing (allowing to fit different models such as Brownian Motion or Ornstein-Uhlenbeck to disparity curves).
+% New functionalities currently being developed will include time sequential linear models (for accurately measuring directionality in disparity changes between time slices) or mode of disparity evolution testing (allowing to fit different models such as Brownian Motion or Ornstein-Uhlenbeck to disparity curves).
 
@@
 
-%Editors:
-%Sam Price, Daniele Silvestro
-
-%Reviewers
-%Graeme Lloyd, Dave Bapst, Emma Sherratt
-
 \bibliographystyle{sysbio}

### 2017-08-24 paper/CoverLetter.tex
Updated cover letter

@@
 
-Hence there exist a multitude of ways to define a multidimensional space, a metric measuring a specific aspect thereof and a test for many hypothesis.
-In theory this vast tool-kit should allow evolutionary biologists and ecologists to do many things.
-There are numerous implementation of such analysis pipeline in numerous R packages (e.g.
+There exist a multitude of ways to define a multidimensional space, a metric measuring a specific aspect thereof and a test for many hypothesis.
+There are numerous implementation of such analysis pipeline in several R packages (e.g.
 \texttt{Claddis}, \texttt{geomorph}, \texttt{ade4}, \texttt{vegan}, etc.) or even in other languages (e.g.
@@
 However, each of these implementations are particular to the software/package author's definition of the multidimensional space, metric and test.
-This leads researchers to often use these tool-kits by default hampering them to use the best definitions of the space/metric/test fitted to their particular question.
-At worst this leads in turn to wrong statistic claims, and, at best to time consuming code retro-engineering.
+This leads researchers to often use these tool-kits by default, thus hampering them to use the best definitions of the space/metric/test fitted to their particular question.
+At worst this leads to wrong statistic claims, and, at best to time consuming code retro-engineering.
 
-In this application manuscript entitled ``dispRity: an R package for measuring disparity'', I present a complete and flexible R package allowing to run multidimensional analysis where each parameter (the space, the metric and the test) can be easily defined by the user.
+In this application manuscript entitled ``dispRity: an R package for measuring disparity'', I present a complete and flexible R package allowing to run multidimensional analysis where each parameter can be easily defined by the user (the space, the metric and the test).
 In fact, this package provides a tidy and modular environment where users can use their own multidimensional space defined as a matrix and apply to it any metric and any test (provided as functions).
 This allows for highly modular multidimensional analysis and the tidy interface allows easily reproducible analysis.
+Finally, this package comes with an in depth gitbook manual covering most functionalities and being updated regularly following users requests.
 

### 2017-08-25 paper/CoverLetter.tex

--- 2017-08-24 paper/CoverLetter.tex
@@
 
-Studying multivariate data as an alternative to univariate data has become increasingly popular in the last decades both in ecology and evolution.
+Studying multivariate data as an alternative to univariate data has become increasingly popular in the last decades in both ecology and evolution.
 This includes, for example, studying form and function (e.g. Diaz et al., 2016, \textit{Nature}) or morphological diversity (i.e. disparity - e.g. Brusatte et al. 2012, \textit{Science}).
-These methods have in common to use a multidimensional space (through ordination or other techniques) representing the combination of all possible observed traits and then measuring some aspect of the space (e.g.
-its hyper-volume) for different groups and compare them between each other or to a null hypothesis.
+These methods all use a multidimensional space (through ordination or other techniques) representing the combination of all possible observed traits and then measuring some aspect of the space (e.g.
+its hypervolume) for different groups, and comparing these to each other or to a null hypothesis.
 
-Hence there exist a multitude of ways to define a multidimensional space, a metric measuring a specific aspect thereof and a test for many hypothesis.
-In theory this vast tool-kit should allow evolutionary biologists and ecologists to do many things.
-There are numerous implementation of such analysis pipeline in numerous R packages (e.g.
-\texttt{Claddis}, \texttt{geomorph}, \texttt{ade4}, \texttt{vegan}, etc.) or even in other languages (e.g.
-\texttt{GINGKO} in javascript).
-However, each of these implementations are particular to the software/package author's definition of the multidimensional space, metric and test.
-This leads researchers to often use these tool-kits by default hampering them to use the best definitions of the space/metric/test fitted to their particular question.
-At worst this leads in turn to wrong statistic claims, and, at best to time consuming code retro-engineering.
+There are a multitude of ways of carrying out each of these steps, and numerous software implementations in several (e.g.
+in R: \texttt{Claddis}, \texttt{geomorph}, \texttt{ade4}, \texttt{vegan}, in javascript: \texttt{GINGKO}).
+However, each of these implementations is particular to the package author's definition of the multidimensional space, metric and test.
+This leads researchers to use these toolkits by default, thus preventing them from using the best definitions of the space/metric/test for their particular question or dataset.
+At worst this leads to incorrect statistic claims, and, at best to time consuming code retro-engineering.
 
-In this application manuscript entitled ``dispRity: an R package for measuring disparity'', I present a complete and flexible R package allowing to run multidimensional analysis where each parameter (the space, the metric and the test) can be easily defined by the user.
-In fact, this package provides a tidy and modular environment where users can use their own multidimensional space defined as a matrix and apply to it any metric and any test (provided as functions).
+In this application manuscript entitled ``dispRity: an R package for measuring disparity'', I present a complete and flexible R package for running multidimensional analysis where each parameter can be easily defined by the user (the space, the metric and the test).
+This package provides a tidy and modular environment where users can use their own multidimensional space defined as a matrix and apply to it any metric and any test (provided as functions).
 This allows for highly modular multidimensional analysis and the tidy interface allows easily reproducible analysis.
+All functions have documentation and unit tests. 
+Finally, this package comes with an in depth Gitbook manual covering most functionalities that will be updated regularly following user requests.
+
+% NC: Add maybe somewhere how many people are already using the package? The package is likely to be well cited as it is already being used in at least XX projects that I am aware of.
 

### 2017-08-25 paper/dispRity_paper.tex

@@
 \newpage
-\begin{abstract}
+\begin{abstract} % NC: This needs some work, I think it could be a little longer.
+Really need to sell the problem the package solves, what questions are asked with it etc.
+Remember editors ONLY see this initially, so if it is crap it won't go out to review.
 
 \begin{enumerate}
-    \item We present \disp, an \texttt{R} package in for multidimensional spaces analysis.
+    \item I present \disp, an \texttt{R} package for summarising multidimensional spaces.
     \item Multidimensional spaces (e.g. morpho/eco/etc.-space) can described using a suit of metrics (e.g. disparity metrics) an many statistical tests.
-    \item The package provides a flexible framework were the space, metric and test can be easily defined by the user to analyse their specific ecological or evolutionary hypothesis.
+    \item The package provides a flexible framework where the space, metric and test can be easily defined by the user to analyse their specific ecological or evolutionary hypothesis.
     \item Its architecture is based on the \disp object with a variety of tools for modifying, summarising or plotting it.
-    \item The package's modularity and simple interface allows to run easily reproducible multidimensional analysis.
+    \item The package's modularity and simple interface allows users to easily run reproducible multidimensional analyses.
 \end{enumerate}
@@
 
-\noindent (Keywords: disparity, ordination, multidimensionality, disparity through time, palaeobiology, ecology)\\
+\noindent (Keywords: disparity, ordination, multidimensionality, disparity-through-time, palaeobiology, ecology)\\
 
@@
 To understand the ecology and evolution of species we must use multiple variables that inevitably covary with each other through time and space.
-One solution to this problem is to analyse this data in a multivariate framework \citep[e.g.][]{price2015predation,diaz2016global}.
-Such analysis aim to capture complex biological multidimensionality while still providing outputs that are interpretable in our physical world restricted to three dimensions.
-These biological variables are as diverse as morphological traits (discrete, e.g.
-\citealt{brusattedinosaur2012}; or continuous, e.g.
-\citealt{price2015predation}), life history traits \citep[e.g.][]{diaz2016global}, ecosystem properties \citep[e.g.][]{DonohueDim}, etc.
-And can be used to answer specific questions such as changes in morphological diversity through time \citep{Close2015}, testing competitive replacement scenarios \citep{Brusatte12092008}, what is the relation between form and function in organisms \citep{diaz2016global} or even describing the entirety of possible shapes for a group of organisms \citep{raup1966geometric}.
+One solution to this problem is to analyse these data in a multivariate framework \citep[e.g.][]{price2015predation,diaz2016global}.
+Such analyses aim to capture the complex multidimensionality of biological data, while still providing outputs that are interpretable in our physical world restricted to three dimensions.
+These multivariate analyses can be used to investigate changes in morphological diversity through time \citep{Close2015}, competitive replacement scenarios \citep{Brusatte12092008}, relationships among form and function \citep{diaz2016global} and even to describe the entirety of possible shapes for a group of organisms \citep{raup1966geometric}.
+The biological variables in such analyses are equally diverse, including morphological traits (discrete traits like the presence or absence of a character, e.g.
+\citealt{brusattedinosaur2012}; and continuous traits such as lengths, e.g.
+\citealt{price2015predation}), life history traits \citep[e.g.][]{diaz2016global}, and even ecosystem properties \citep[e.g.][]{DonohueDim}.
+% NC: Best to avoid etc.
+where possible
 
 % Variety of multidimensional spaces, metrics and tests
-In all these analysis, each of these multivariate traits form different multidimensional spaces represented as matrices where the columns are the dimensions (or some transformation thereof) and each rows are the elements available within this space.
-These multidimensional spaces can be defined in many ways such as a pairwise distance matrix \citep{Close2015}, a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, a multidimensional scaling \citep[MDS;][]{DonohueDim}, etc.
-Once this multidimensional space is defined, say morphospace (if its variables are base on morphology), it is then possible to measure a specific aspect of it, say its morphological diversity (more commonly by as disparity; \citealt{Hopkins2017}).
-This specific measurement (hereafter referred to as disparity metrics) can in turn be defined in many different ways depending on the specific question \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015}.
-Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical test such as non-parametric multivariate analyses of variance \citep[e.g.][]{Brusatte12092008}, multidimensional permutation test \citep[e.g.][]{diaz2016global} or even simply by looking at the confidence interval overlap between disparity measurements \citep[e.g.][]{halliday2016eutherian}.
+In all these analysis, each set of multivariate traits forms a multidimensional space, usually represented as matrices where the columns are the dimensions (or some transformation thereof) and rows are the elements (often species or individuals) within this space.
+These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix \citep{Close2015}, or as outputs from a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric},or a multidimensional scaling \citep[MDS;][]{DonohueDim}, among others.
+The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
+For example, when using morphological traits the multivariate space will be a morphospace, when using ecological traits it may be referred to as an ecospace or trait space.
+Once the multidimensional space is defined it is then possible to measure a specific aspect of it.
+This specific measurement (hereafter referred to as a 'disparity metric') can in turn be defined in many different ways depending on the specific question being investigated \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015}.
+Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance \citep[e.g.][]{Brusatte12092008}, multidimensional permutation tests \citep[e.g.][]{diaz2016global} or even simply by looking at the confidence interval overlaps between disparity measurements \citep[e.g.][]{halliday2016eutherian}.
+In summary, there are many different ways to perform each step of a multidimensional analysis, making analyses of complexity ever more complex.
 
 % Problems with this variety
-In theory, this multitude of ways to generate multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods \citep[such as][for biogreography]{matzke2013biogeobears}.
+In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods \citep[such as][for biogeography]{matzke2013biogeobears}.
 In practice, however, this is hampered by existing software implementations.
-Even though many software exist for multidimensional analysis \citep[e.g.][]{bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric}, package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric and then enforce this on users by allowing few to no flexibility.
-For example, in the excellent and widely used \texttt{geomorph} package morphological disparity analysis can be ran using the \texttt{morphol.disparity} function defining the multidimensional space as the ordination of the Procrustes transform of morphometric data, the disparity metric as the Procrustes variance and permutation tests can be used \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
-This can lead to a chain of inappropriate analyses led by users just using the existing software when their question/data is not appropriate. %Oh no!
+Although many software packages exist for multidimensional analysis \citep[e.g.][]{bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric}, package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric and then enforce this on users by allowing little to no flexibility.
+For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of morphometric data, the disparity metric as the Procrustes variance, and using permutation tests to test hypotheses \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
+This is ideal in some cases, but in others may not be the most appropriate analysis.
+This can lead to inappropriate analyses by users confined by the existing software implementations. %Oh no!
 
@@
 The aim of the \disp package is to solve such problems by providing a completely flexible framework for studying multidimensional data.
-In fact, this package is based on a highly modular architecture where each crucial decision in multidimensional analysis (which data, which metric and which test) can be provided by the user.
-It implements many commonly used metrics and definitions of disparity as well as providing a simple interface for users to implement their own disparity functions.
+This package is based on a highly modular architecture where each crucial decision in multidimensional analysis (which data, which metric and which test) can be provided by the user.
+It implements many commonly used metrics and definitions of disparity, as well as providing a simple interface for users to implement their own disparity functions.
 The package is described here for the use of discrete morphological data diversity analysis but can be generalised to any type of multidimensional data (see the package vignette for more examples).
@@
 \section{Description}
-In brief, the package intakes a \texttt{matrix} object (the multidimensional space), calculates some aspect of it (the disparity metric) and allows to analyse the resulting \disp object (see Manuals and Vignettes below).
-Some additional functions allow to modify the multidimensional space such as dividing it per group or through time or/and bootstrapping it (see Fig. \ref{Fig:workflow} and Table \ref{Tab:main_fun}).
-See below for a more detailed description of features.
+In brief, the package takes a \texttt{matrix} object (the multidimensional space), calculates some aspect of it (the disparity metric) and allows users to analyse the resulting \disp object (see Manuals and Vignettes below).
+Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow} and Table \ref{Tab:main_fun}).
+See below for a more detailed description of the package features.
 
@@
    \includegraphics[width=1\textwidth]{workflowsvg.pdf} 
-\caption{\disp package workflow: rectangles represent matrices; ellipse represent functions; plain black arrows indicate input/output; dashed grey arrows indicate output (though the summary, plot, and test function can not be applied if no disparity has been calculated).}
+\caption{\disp package workflow: rectangles represent matrices; ellipses represent functions; plain black arrows indicate input/output; dashed grey arrows indicate output (though the summary, plot, and test function cannot be applied if no disparity has been calculated).}
 \label{Fig:workflow}
@@
         \texttt{Claddis.ordination} & Imports data from the \texttt{Claddis} package. \\
-        \texttt{custom.subsamples} & Separates data in custom subsamples. \\
-        \texttt{time.subsamples} & Separates data in time subsamples. \\
+        \texttt{custom.subsamples} & Separates data into custom subsamples. \\
+        \texttt{time.subsamples} & Separates data into time subsamples. \\
         \texttt{boot.matrix} & Bootstraps and rarefies data or a \disp object. \\
@@
         \texttt{test.dispRity} & Applies statistical tests to a \disp object.\\
-        \texttt{dispRity.per.group} & Pipelined disparity per groups analysis. \\
-        \texttt{dispRity.through.time} & Pipelined disparity through time analysis. \\
+        \texttt{dispRity.per.group} & Pipeline for disparity within groups analysis. \\
+        \texttt{dispRity.through.time} & Pipeline for disparity-through-time analysis. \\
         \hline
@@
 \subsection{Measuring disparity}
-The \disp function allows to measure disparity from any multidimensional space (of class \texttt{matrix} or \texttt{data.frame}) where the columns correspond dimensions and the rows correspond to the elements present in the space.
+The \disp function measures disparity of any multidimensional space (of class \texttt{matrix} or \texttt{data.frame}) where the columns correspond to the dimensions and the rows correspond to the elements present in the space.
 The disparity metric (i.e.
@@
 \begin{itemize}
-    \item Another matrix (a dimensions-level 3 function - e.g. a variance-covariance matrix; \texttt{stats::cor})
-    \item A vector (a dimensions-level 2 function - e.g. the variance of each dimension; \texttt{dispRity::variances} - see below)
-    \item A single value (a dimensions-level 1 function - e.g. the overall standard deviation; \texttt{stats::sd})
+    \item Another matrix (a dimension-level 3 function - e.g. a variance-covariance matrix; \texttt{stats::cor})
+    \item A vector (a dimension-level 2 function - e.g. the variance of each dimension; \texttt{dispRity::variances} - see below)
+    \item A single value (a dimension-level 1 function - e.g. the overall standard deviation; \texttt{stats::sd})
 \end{itemize}
@@
    \includegraphics[width=0.8\textwidth]{../inst/gitbook/dispRity_fun.pdf} 
-\caption{Illustration of the different metric dimensions levels in the \disp package. In this example, each cell correspond to a single value (e.g. a 8 $\times$ 7 matrix or a vector of 8 elements)}.
+\caption{Illustration of the different metric dimension-levels in the \disp package.
+In this example, each cell correspond to a single value (e.g.
+a 8 $\times$ 7 matrix or a vector of eight elements)}.
 \label{Fig:levels}
@@
 
-The metrics can be functions from previous packages, defined by the user or from the \disp package (see Table \ref{Tab:metrics} for the currently implemented metrics).
-When multiple functions are passed the \texttt{metric} argument, the functions are sorted per dimension level and applied in decreasing order to the data.
-For example, if the metric is defined as \texttt{metric = c(mean, range)}, the \texttt{range} function (dimension level 2) is first applied to the data and the function \texttt{mean} is then applied to the result (e.g.
+The disparity metrics can be functions from other packages, defined by the user or from the \disp package (see Table \ref{Tab:metrics} for the currently implemented metrics).
+When multiple functions are passed to the \texttt{metric} argument, the functions are sorted by dimension-level and applied in decreasing order to the data.
+For example, if the metric is defined as \texttt{metric = c(mean, range)}, the \texttt{range} function (dimension-level 2) is first applied to the data and the function \texttt{mean} is then applied to the result (e.g.
 \texttt{mean(range(data))}).
-There is no limitation on the metric dimensions only that at least one dimensions level 2 or 1 are present in the metric definition (i.e.
+There is no limitation on the metric dimensions only that at least one dimension-level 2 or 1 function must be present in the metric definition (i.e.
 disparity can only be defined as distribution or a single value).
@@
         \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
-        \texttt{diagonal} & The longest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & \texttt{dispRity::diagonal} \\
+        \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & \texttt{dispRity::diagonal} \\
         \texttt{mode.val} & The modal value & 1 & NA & \texttt{dispRity::mode.val}\\
@@
 Prior to calculating disparity, the multidimensional space can be subdivided into subsamples, typically to be compared to each other.
-For example, one can be interested in looking how the disparity of a specific subsample of the space compared to another (e.g.
-difference among groups) or, how different sub-samples change sequentially (e.g.
+For example, one may be interested in seeing how the disparity of a specific subsample of the space compares to another (e.g.
+difference among groups) or, how different subsamples change sequentially (e.g.
 through time).
@@
 
-This can be done using the \texttt{custom.subsamples} or \texttt{time.subsamples} functions.
-The first takes as argument the matrix defining the multidimensional space and a list of elements to group into different sub-samples.
-The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time sub-samples for example) and which subsamples to create: (1) discrete time subsamples (or time-binning) or (2) continuous time subsamples (or time-slicing).
+This splitting can be done using the \texttt{custom.subsamples} or \texttt{time.subsamples} functions.
+The first takes as argument the matrix defining the multidimensional space and a list of elements to group into different subsamples.
+The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time subsamples for example) and which subsamples to create: (1) discrete time subsamples (or time-binning) or (2) continuous time subsamples (or time-slicing).
 
-The time-bining method groups elements by specific age range.
-The time slicing method works by using a phylogeny and looking at which taxa are present at any specific point in time.
+The time-binning method groups elements by specific age range.
+The time-slicing method works by using a phylogeny and looking at which taxa are present at any specific point in time.
 This method thus requires the nodes to be part of the multidimensional space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
@@
 Disparity measurement can be influenced by sampling \citep{Butler2012}.
-To take this source of bias into account, it is common practice to bootstrap the multidimensional space (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different sub-samples).
-Additionally, if disparity is defined as a dimension-level 1 metric, it can be useful to measure it on bootstrapped data in order to have a distribution on which to perform statistical analysis.
+To take this source of bias into account, it is common practice to bootstrap the multidimensional space (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different subsamples).
+Additionally, if disparity is defined as a dimension-level 1 metric, it can be useful to measure it on bootstrapped data to obtain a distribution on which to perform statistical analyses.
 
-Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the multidimensional space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' one where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).
+Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the multidimensional space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).
 
@@
 Similarly, rarefaction can be achieved through the same \texttt{boot.matrix} function.
-Rarefaction can be useful when looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of sub-samples wit the same number of elements in each.
+Rarefaction can be useful when looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of subsamples with the same number of elements in each.
 In practice, rarefaction limits the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).
@@
 The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity} (provided disparity was calculated).
-These results can then be analysed using the \texttt{test.dispRity} function for comparing sub-samples or testing hypothesis.
+These results can then be analysed using the \texttt{test.dispRity} function for comparing subsamples or testing hypotheses.
 
-\subsubsection{Summary and plot}
-These \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow to set which central tendency and which quantiles should be represented.
+\subsubsection{Summarising and plotting}
+The \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow users to set which central tendency and which quantiles should be represented.
 % as common arguments which central tendency (e.g.
 if central tendency is the mean and disparity was calculated as the sum of the variances for each dimensions, the central disparity tendency would be the mean of the sum of the variances) and the quantiles to represent (in the same example, if the quantile chosen is the 50th, these functions will also display 50\% of the sum of variances around the mean).
-Additionally, the \texttt{plot.dispRity} function allows to graphically represent the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves (in the case of disparity through time analysis for example) and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
+Additionally, the \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves (in the case of disparity-through-time analysis for example) and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
 Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e.
-non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for
-The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...).
+non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for % NC: for...what????
+The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base R.
 % Note that if the disparity has been calculated on fully rarefied data (using \texttt{boot.matrix(data, rarefaction = TRUE)} prior to calculating disparity), the rarefaction curves for the whole multidimensional space (or its sub-samples) can be plotted using \texttt{rarefaction = ``plot''}.
 
-\subsubsection{testing hypothesis}
-The \texttt{test.dispRity} function allows to test hypothesis on the disparity data (e.g. by comparing the different sub-samples between each other).
-Similarly to the \disp function described above, this one can take any test defined by the user.
-The \texttt{comparison} arguments allows to indicate in which order (if any) should the tests be applied to the subsamples: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subsample to all the others; (3) ``sequential'' for comparing subsamples sequentially (e.g.
+\subsubsection{Testing hypotheses}
+The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data (e.g. by comparing the different subsamples to each other).
+Similarly to the \disp function described above, this function can take any test defined by the user or built into base R.
+The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsamples: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subsample to all the others; (3) ``sequential'' for comparing subsamples sequentially (e.g.
 first against second, second against third, etc.); (4) `all'' for comparing all the subsamples simultaneously (i.e.
-\texttt{disparity $\mathtt{\sim}$ subsamples} - this can be used for \texttt{lm} or \texttt{glm}) or (5) any list of pairs (or more) of sub-samples to compare,
+\texttt{disparity $\mathtt{\sim}$ subsamples} - this can be used for \texttt{lm} or \texttt{glm}) or (5) any list of pairs (or more) of subsamples to compare.
 
 Some tests are implemented within the package such as the Bhattacharrya Coefficient \citep[\texttt{bhatt.coeff}][]{Bhattacharyya,GuillermeCooper} or a permutation test based on null hypothesised multidimensional space following \cite{diaz2016global} (\texttt{null.test}).
-Additionally, this function also allows additionally argument such as \texttt{rarefaction} (as described above) or \texttt{correction} to adjust eventual p-values when multiple parametric tests are ran.
+Additionally, this function also allows additional arguments such as \texttt{rarefaction} (as described above) or \texttt{correction} to adjust p-values when using multiple parametric tests.
 
-\section{Examples}
-For brevity reason, the following example is based on a classical palaeobiology morphological disparity analysis.
-Note the more examples are available in the package manual \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}.
+\section{Examples} % NC: I think it would be nice to add a short paragraph describing the cool science question you are asking here, why it is interesting etc.
+Otherwise this is quite a dry software example that doesn't really convince me I need your cool new package.
+The following example is based on a classical palaeobiology morphological disparity analysis.
+Note more examples are available in the package manual \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}.
 
 \subsubsection{\disp data}
-The package contains a dataset that is a subset from \cite{beckancient2014} including:
+The package contains a dataset that is a subset from \cite{beckancient2014} and includes:
 
@@
     \item \texttt{BeckLee\_mat99}: the same matrix \texttt{BeckLee\_mat50} but also containing the reconstruction of their 49 ancestors.
-    \item \texttt{BeckLee\_tree}: a chronogram with the 50 mammals species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
-    \item \texttt{BeckLee\_ages}: the first and last occurrence data for 14 of the mammals species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
+    \item \texttt{BeckLee\_tree}: a chronogram with the 50 mammal species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
+    \item \texttt{BeckLee\_ages}: the first and last occurrence data for 14 of the mammal species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
     \item \texttt{disparity}: a pre-made \disp object based on the data above.
@@
 
-In this example, the multidimensional space is thus defined as a morphospace: the ordination of the distance between discrete morphological characters for 50 mammals species \citep[from][]{beckancient2014}.
-Additionally, we can define disparity as the sum of the variances on each dimensions \citep{Wills1994} that will represent an aspect of the the volume of the morphospace (i.e.
+In this example, the multidimensional space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
+Additionally, we can define disparity as the sum of the variances on each dimension \citep{Wills1994} that will represent an aspect of the the volume of the morphospace (i.e.
 a large sum of variances approximates a large morphospace and vice versa).
 
-\subsection{Typical disparity per groups analysis}
-Here, we can be interested in whether the approximation of the volume within the morphospace (the disparity as defined above) is different in the living mammals and their descendants (crown) or the fossil mammals with no living descendants (stem).
-These two groups can be defined as follow (where the integers are the row number of the species in the matrix):
+\subsection{Typical disparity among groups analysis}
+Here, we may be interested in whether the approximation of the volume within the morphospace (the disparity as defined above) is different in living mammals and their descendants (crown) comparaed to the fossil mammals with no living descendants (stem).
+These two groups can be defined as follows (where the integers are the row number of the species in the matrix):
 
@@
 
-\noindent It is then possible to measure the disparity between the two groups as follow:
+\noindent It is then possible to measure the disparity between the two groups as follows:
 
@@
 
-\noindent Note that this function is a wrapping function that is the equivalent to:
+\noindent Note that this function is a wrapper function that is the equivalent to:
 
@@
 
-\noindent Which allows a finer tuning of the optional argument in each functions.
-The three arguments here are defined as follow: \texttt{data = BeckLee\_mat50} is our definition of the multidimensional space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity.
+\noindent Which allows a finer tuning of the optional arguments in each function.
+The three arguments here are defined as follows: \texttt{data = BeckLee\_mat50} is our definition of the multidimensional space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity.
 
@@
 \bigskip
-As indicated, the \disp object contains 2 customised subsamples (called ``crown'' and ``stem'') from a morphospace made of 50 elements for 48 dimensions.
+As indicated, the \disp object contains two customised subsamples (called ``crown'' and ``stem'') from a morphospace made of 50 elements for 48 dimensions.
 The \disp object also displays information on the number and method of the bootstrap replicates as well as the definition of disparity.
@@
 
-As we can see from the summary table and the plot, there seems to be a significant difference in morphospace volume occupation between the two groups.
+As we can see from the summary table and the plot, there seems to be a significant difference in morphospace volume occupied between the two groups.
 It is possible to test this hypothesis by using, for example, a non-parametric Wilcoxon test (\texttt{stats::wilcox.test}):
@@
 \bigskip
-As indicated by the p.value, there is a significant difference in disparity between both groups.
+As indicated by the p value, there is a significant difference in disparity between the groups.
 
-\subsubsection{Typical disparity through type analysis}
+\subsubsection{Typical disparity-through-time analysis}
 
-Another type of disparity analysis can be a disparity through time analysis.
-Using the same definition of the multidimensional space and disparity as in the previous example, one can be interested in how disparity changes through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).
+% NC: Again a short description of the actual question being asked here and why it is cool.
+Another common type of disparity analysis is a disparity-through-time analysis.
+Using the same definition of the multidimensional space and disparity as in the previous example, we may be interested in how disparity changes through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).
 
@@
 
-It is then possible to measure disparity through time using the following function:
+It is then possible to measure disparity-through-time using the following function:
 
@@
 \noindent The arguments \texttt{data = BeckLee\_mat50} and \texttt{metric = c(sum, variances)} are the same as in the example above.
-However, in this type of analysis, we also needs to have additional arguments: the \texttt{time = time\_bins} indicates the boundaries of the different time bins, the \texttt{tree = BeckLee\_tree} argument provides information on the age of each element and \texttt{method = 'discrete'} indicates that the data is time-binned.
+However, in this type of analysis, we also need to have additional arguments: the \texttt{time = time\_bins} indicates the boundaries of the different time bins, the \texttt{tree = BeckLee\_tree} argument provides information on the age of each element and \texttt{method = 'discrete'} indicates that the data is time-binned.
 
@@
 
-\texttt{> plot(disparity, type = 'countinuous')}
+\texttt{> plot(disparity, type = 'continuous')}
 
@@
    \includegraphics[width=1\textwidth]{plot_example_time.pdf} 
-\caption{\disp plot of disparity through time}
+\caption{\disp plot of disparity-through-time}
 \label{Fig:plot_time}
@@
 From the plot and the summary, disparity seems to change (increase) through time from the Cretaceous to the Eocene.
-Similarly to the example above, it is also possible to statistically test this hypothesis by running an ANOVA for example (Table \ref{Tab:anova}):
+Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, an ANOVA (Table \ref{Tab:anova}):
 
@@
 \end{tabular}
-\caption{Anova performed on a \disp object}
+\caption{ANOVA performed on a \disp object}
 \label{Tab:anova}
@@
 
-\section{Additional informations}
+\section{Additional information}
 \subsection{Manuals and vignette}
 Supplementary information about the package is available on the project page \url{https://TGuillerme/dispRity}.
-Information concerning each functions can be found in R in the individual function manuals and a package manual with more detailed information is available online at \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}.
-This manual contains substantially more information and detailed example including a tutorial for a ``classic'' disparity analysis in palaeobiology as well as an introduction to the use of this package in ecology or other disciplines.
+Information concerning each function can be found in R in the individual function manuals and a package manual with more detailed information is available online at \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}.
+This manual contains substantially more information and detailed examples including a tutorial for a ``classic'' disparity analysis in palaeobiology as well as an introduction to the use of this package in ecology or other disciplines.
 
 \subsection{Data simulations}
-This package also contains a series of simulation functionality for generating random discrete morphological matrices (\texttt{sim.morpho}) or random multidimensional spaces (\texttt{space.maker}).
-These functions are based on a similar modular architecture as the \disp functions allowing users to provide their own distribution descriptions as parameters for the simulation.
+This package also contains functions for simulating random discrete morphological matrices (\texttt{sim.morpho}) or random multidimensional spaces (\texttt{space.maker}).
+These functions are based on a similar modular architecture as that used by the \disp functions, allowing users to provide their own distribution parameters for the simulations.
 For example, \texttt{stats::rnorm} can be provided as an argument for drawing normal characters rates with \texttt{sim.morpho} or normally distributed multidimensional space with \texttt{space.maker}.
@@
 \section{Conclusion}
-The \disp package is based on a highly modular architecture allowing researchers to simply define both their multidimensional space and their disparity metric for efficiently analysing multivariate data.
-The \disp object allows to easily pipeline disparity analysis from the data input (the matrix) to publication standard results (tables, plots, hypothesis testing).
-With the help of this package, researchers will thus be able to run multivariate analysis in an easy way while still allowing a maximum of plasticity necessary to the specificity of such analysis.
+The \disp package is based on a highly modular architecture allowing researchers to simply define both their multidimensional space and their disparity metric to efficiently analyse multivariate data.
+The \disp object allows users to easily pipeline disparity analysis from the data input (the matrix) to publication standard results (tables, plots, hypothesis testing).
+With the help of this package, researchers will be able to run multivariate analysis in an easy way while still allowing the flexibility necessary to such analyses.
 
 \section{Data availability and reproducibility}
-All the code and the data for using the package and it's vignette examples is readily available in the package or on GitHub \url{github.com/TGuillerme/dispRity}.
+All the code and the data for using the package and its vignette with examples are available in the R package help files or on GitHub \url{github.com/TGuillerme/dispRity}.
 
 \section{Acknowledgments}
-Many thanks to Natalie Cooper for encouraging the development of this package and helping for the redaction of this paper and the package manuals.
+Many thanks to Natalie Cooper for encouraging the development of this package and helping with the writing of this paper and the package manuals.
 Thanks to Dave Bapst, Martin Brazeau, Andrew Jackson, Graeme Lloyd and Emma Sherratt for specific implementation ideas.

### 2017-08-25 paper/CoverLetter.tex
Minor changes to cover letter

@@
 
-There are a multitude of ways of carrying out each of these steps, and numerous software implementations in several (e.g.
-in R: \texttt{Claddis}, \texttt{geomorph}, \texttt{ade4}, \texttt{vegan}, in javascript: \texttt{GINGKO}).
+There are a multitude of ways of carrying out each of these steps, and numerous software implementations in several languages (e.g.
+in R: \texttt{Claddis}, \texttt{geomorph}, \texttt{ade4}, \texttt{vegan}, or in javascript: \texttt{GINGKO}).
 However, each of these implementations is particular to the package author's definition of the multidimensional space, metric and test.
@@
 All functions have documentation and unit tests. 
-Finally, this package comes with an in depth Gitbook manual covering most functionalities that will be updated regularly following user requests.
-
-% NC: Add maybe somewhere how many people are already using the package? The package is likely to be well cited as it is already being used in at least XX projects that I am aware of.
+Additionally, this package comes with an in depth Gitbook manual covering most functionalities that will be updated regularly following user requests. 
+Finally, the package is likely to be well cited as it is already being used in at least five projects that I am aware of.
 

### 2017-08-25 paper/CoverLetter.tex
Courried package name

@@
 
-In this application manuscript entitled ``dispRity: an R package for measuring disparity'', I present a complete and flexible R package for running multidimensional analysis where each parameter can be easily defined by the user (the space, the metric and the test).
+In this application manuscript entitled ``\texttt{dispRity}: an R package for measuring disparity'', I present a complete and flexible R package for running multidimensional analysis where each parameter can be easily defined by the user (the space, the metric and the test).
 This package provides a tidy and modular environment where users can use their own multidimensional space defined as a matrix and apply to it any metric and any test (provided as functions).

### 2017-08-25 paper/dispRity_paper.tex
Minor presubmission edits

@@
+% 2961w - 30p
+
 \documentclass[12pt,letterpaper]{article}
@@
 \newpage
-\begin{abstract} % NC: This needs some work, I think it could be a little longer.
-Really need to sell the problem the package solves, what questions are asked with it etc.
-Remember editors ONLY see this initially, so if it is crap it won't go out to review.
+\begin{abstract}
 
     \begin{enumerate}
-    \item I present \disp, an \texttt{R} package for summarising multidimensional spaces.
-    \item Multidimensional spaces (e.g. morpho/eco/etc.-space) can described using a suit of metrics (e.g. disparity metrics) an many statistical tests.
-    \item The package provides a flexible framework where the space, metric and test can be easily defined by the user to analyse their specific ecological or evolutionary hypothesis.
-    \item Its architecture is based on the \disp object with a variety of tools for modifying, summarising or plotting it.
-    \item The package's modularity and simple interface allows users to easily run reproducible multidimensional analyses.
+        \item Biological data is multivariate in essence: many traits in organisms co-vary with each other in space and time.
+        This causes biologists to either reduce these variables in a manageable number of variables or, increasingly, to use multivariate toolkits.
+        One such toolkit is based on creating a multidimensional space where the variables are the axis of this space.
+        It is then possible to use a summary metrics for measuring some properties of this space for some group of species and compare them.
+        For example, if studying morphology, one can create a morphospace for two groups of species, measure some aspect of these groups (such as their spread in the space - often termed disparity) and then test whether they have a different disparity.
+
+        \item There are as many definition of these multidimensional spaces, metrics and tests as there are questions that can be tackled with such methods.
+        Many of these methods are implemented in specific software or R packages.
+        However, the definition of the space, metric and test is often dependent on the software/package's author's point of view or specific question.
+        This can unfortunately hamper researchers to apply different methods that suits best their specific questions.
+
+        \item Here I present the \disp package, a flexible R package for running multidimensional analysis.
+        This package allows users to define each step of the analysis (whether it is the space, the metric or the test) through a highly modular architecture where each definition can be passed as a function.
+        It also provides a tidy interface through the \disp object, allowing to easily run reproducible multivariate analysis.
+
+        \item The \disp package also comes with an extend manual regularly updated following users questions or suggestions.
+        Furthermore, the package contains some simulation tools (e.g to simulate complex multidimensional space or morphological data).
+        Finally it also contains a suit of utilities to work with \disp objects aimed at helping users to develop their own multidimensional metrics and/or tests.
+
     \end{enumerate}
@@
 \citealt{brusattedinosaur2012}; and continuous traits such as lengths, e.g.
-\citealt{price2015predation}), life history traits \citep[e.g.][]{diaz2016global}, and even ecosystem properties \citep[e.g.][]{DonohueDim}.
-% NC: Best to avoid etc.
-where possible
+\citealt{price2015predation}), life history traits \citep[e.g.][]{diaz2016global}, or even ecosystem properties \citep[e.g.][]{DonohueDim}.
 
@@
 In all these analysis, each set of multivariate traits forms a multidimensional space, usually represented as matrices where the columns are the dimensions (or some transformation thereof) and rows are the elements (often species or individuals) within this space.
-These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix \citep{Close2015}, or as outputs from a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric},or a multidimensional scaling \citep[MDS;][]{DonohueDim}, among others.
+These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix \citep{Close2015}, or as outputs from a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, or a multidimensional scaling \citep[MDS;][]{DonohueDim}, among others.
 The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
@@
 Once the multidimensional space is defined it is then possible to measure a specific aspect of it.
-This specific measurement (hereafter referred to as a 'disparity metric') can in turn be defined in many different ways depending on the specific question being investigated \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015}.
+This specific measurement (hereafter referred to as a ``disparity metric'') can in turn be defined in many different ways depending on the specific question being investigated \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015}.
 Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance \citep[e.g.][]{Brusatte12092008}, multidimensional permutation tests \citep[e.g.][]{diaz2016global} or even simply by looking at the confidence interval overlaps between disparity measurements \citep[e.g.][]{halliday2016eutherian}.
@@
 Although many software packages exist for multidimensional analysis \citep[e.g.][]{bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric}, package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric and then enforce this on users by allowing little to no flexibility.
-For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of morphometric data, the disparity metric as the Procrustes variance, and using permutation tests to test hypotheses \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
+For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of morphometric data, the disparity metric as the Procrustes variance, and uses a permutation tests to test hypotheses \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
 This is ideal in some cases, but in others may not be the most appropriate analysis.
@@
 Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow} and Table \ref{Tab:main_fun}).
-See below for a more detailed description of the package features.
 
@@
         \texttt{time.subsamples} & Separates data into time subsamples. \\
-        \texttt{boot.matrix} & Bootstraps and rarefies data or a \disp object. \\
+        \texttt{boot.matrix} & Bootstraps and rarefies a matrix or a \disp object. \\
         \disp & Calculates disparity from an matrix or a \disp object. \\
@@
 In this example, each cell correspond to a single value (e.g.
-a 8 $\times$ 7 matrix or a vector of eight elements)}.
+a 8 $\times$ 7 matrix or a vector of eight elements).}.
 \label{Fig:levels}
@@
 
-\begin{table}
-\resizebox{\textwidth}{!}
-{%
-    \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
-        name & description & dimen-sion level & definition & source \\
-        \hline
-        \texttt{ellipse. volume}$^1$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
-        \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
-        \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
-        \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & \texttt{dispRity::diagonal} \\
-        \texttt{mode.val} & The modal value & 1 & NA & \texttt{dispRity::mode.val}\\
-        \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & \texttt{dispRity::ranges} \\
-        \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & \texttt{dispRity::variances} \\
-        \texttt{centroids} & The distance between each element and a fixed point$^2$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & \texttt{dispRity::centroids} \\
-    \end{tabular}
-}%
-    \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean.
+% \begin{table}
+% \resizebox{\textwidth}{!}
+% {%
+%     \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
+%         name & description & dimen-sion level & definition & source \\
+%         \hline
+%         \texttt{ellipse. volume}$^1$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
+%         \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
+%         \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
+%         \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & \texttt{dispRity::diagonal} \\
+%         \texttt{mode.val} & The modal value & 1 & NA & \texttt{dispRity::mode.val}\\
+%         \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & \texttt{dispRity::ranges} \\
+%         \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & \texttt{dispRity::variances} \\
+%         \texttt{centroids} & The distance between each element and a fixed point$^2$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & \texttt{dispRity::centroids} \\
+%     \end{tabular}
+% }%
+%     \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean.
 $^1$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA); $^2$ by default that point is the centroid of the elements.}}
-    \label{Tab:metrics}
-\end{table}
+%     \label{Tab:metrics}
+% \end{table}
 
@@
 through time).
-In essence, the original space corresponds to overall space (e.g. a morphospace contains all the possible observed morphologies).
+In essence, the original space corresponds to the overall space (e.g. a morphospace contains all the possible observed morphologies).
 Conversely, subsamples correspond to parts of the space with pooled characteristics (i.e. only a subset of the total morphospace).
@@
 Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e.
-non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for % NC: for...what????
+non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for a certain number of elements (i.e.
+the rarefaction level).
 The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base R.
@@
 The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data (e.g. by comparing the different subsamples to each other).
-Similarly to the \disp function described above, this function can take any test defined by the user or built into base R.
+Similarly to the \disp function described above, this function can take any test defined by the user or from other R packages.
 The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsamples: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subsample to all the others; (3) ``sequential'' for comparing subsamples sequentially (e.g.
@@
 
-Some tests are implemented within the package such as the Bhattacharrya Coefficient \citep[\texttt{bhatt.coeff}][]{Bhattacharyya,GuillermeCooper} or a permutation test based on null hypothesised multidimensional space following \cite{diaz2016global} (\texttt{null.test}).
+Some tests are implemented within the package such as the Bhattacharrya Coefficient \citep[\texttt{bhatt.coeff};][]{Bhattacharyya,GuillermeCooper} or a permutation test based on null hypothesised multidimensional space following \cite{diaz2016global} (\texttt{null.test}).
 Additionally, this function also allows additional arguments such as \texttt{rarefaction} (as described above) or \texttt{correction} to adjust p-values when using multiple parametric tests.
 
-\section{Examples} % NC: I think it would be nice to add a short paragraph describing the cool science question you are asking here, why it is interesting etc.
-Otherwise this is quite a dry software example that doesn't really convince me I need your cool new package.
+\section{Examples}
+Multivariate analysis can be really useful for looking multiple aspects of organisms diversity together.
+For example, one classic univariate trait used in ecology and evolution can be the sheer number of species present in a ecosystem or/and through time.
+This trait can be useful to test whether some ecosystems or/and time periods have more species diversity than others.
+In this case one can then test some hypothesis such as: environments with bigger altitudes gradients contain more species.
+
+Conversely, in the case of species diversity, one can also look at another independent multivariate trait: the diversity of morphologies \citep[or disparity;][]{ruta2013}.
+Using disparity, it is then also possible to assess whether one ecosystem or/and time period display more morphological variation.
+Following the previous example, if disparity is also higher with bigger altitudes gradients, one could argue that the species richness is caused by an increased number of morphologies adapted to more ecological niches (or something akin).
+
 The following example is based on a classical palaeobiology morphological disparity analysis.
-Note more examples are available in the package manual \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}.
+Note that more examples are available in the package manual (\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).
 
@@
 \subsection{Typical disparity among groups analysis}
-Here, we may be interested in whether the approximation of the volume within the morphospace (the disparity as defined above) is different in living mammals and their descendants (crown) comparaed to the fossil mammals with no living descendants (stem).
-These two groups can be defined as follows (where the integers are the row number of the species in the matrix):
+One typical question with such analysis would be to test whether two groups of species have a different disparity.
+For example, using the data described above, we can test whether the crown mammals (the living mammals and their descendants) are more diverse in term of morphology than the stem mammals (the fossil mammals with no living descendants).
+In other words, whether the approximation of the volume within the morphospace (the disparity as defined above) is different in crown or stem mammals.
 
-\noindent \texttt{> mammal\_groups <- list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44))}
+\noindent These two groups can be defined as follows (where the integers are the row number of the species in the matrix):
+
+\noindent \texttt{> mammal\_groups <- list('crown' = c(16, 19:41, 45:50), 'stem' = c(1:15, 17:18, 42:44))}
 
@@
 
-\noindent \texttt{> disparity}
+\noindent \texttt{> disparity}\\
 \noindent \texttt{ ---- dispRity object ---- }\\
 \noindent \texttt{2 customised subsamples for 50 elements with 48 dimensions:}\\
-\noindent \texttt{    crown, stem.}\\
+          \texttt{crown, stem.}\\
 \noindent \texttt{Data was bootstrapped 100 times (method:"full").}\\
@@
 To visualise the actual disparity values, one can use the \texttt{summary} or/and \texttt{plot} options (Table \ref{Tab:summary_group} and Fig. \ref{Fig:plot_group}):
+
+% mammal_groups <- list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44))
+% subsamples <- custom.subsamples(data = BeckLee_mat50, group = mammal_groups)
+% bootstraps <- boot.matrix(subsamples)
+% disparity <- dispRity(bootstraps, metric = c(sum, variances))
 
@@
 \end{tabular}
-\caption{Summarising a \disp object (disparity per groups)}
+\caption{Summarising a \disp object (disparity per groups).
+$n$ is the number of elements per subsamples, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 95\% are the confidence intervals.}
 \label{Tab:summary_group}
@@
    \includegraphics[width=1\textwidth]{plot_example_group.pdf} 
-\caption{\disp plot of disparity differences between groups}
+\caption{\disp plot of disparity differences between groups.}
 \label{Fig:plot_group}
@@
 
-As we can see from the summary table and the plot, there seems to be a significant difference in morphospace volume occupied between the two groups.
+As we can see from the summary table (Table \ref{Tab:summary_group}) and the plot (\ref{Fig:plot_group}), there seems to be a significant difference in morphospace volume occupied between the two groups.
 It is possible to test this hypothesis by using, for example, a non-parametric Wilcoxon test (\texttt{stats::wilcox.test}):
 
-\noindent \texttt{> test.dispRity(disparity, test = wilcox.test)}
+\noindent \texttt{> test.dispRity(disparity, test = wilcox.test)}\\
 
@@
 As indicated by the p value, there is a significant difference in disparity between the groups.
+Relating back to our question: yes, crown mammals display a higher diversity in morphologies than their stem counterparts (in this example and dataset!).
 
 \subsubsection{Typical disparity-through-time analysis}
-
-% NC: Again a short description of the actual question being asked here and why it is cool.
-Another common type of disparity analysis is a disparity-through-time analysis.
-Using the same definition of the multidimensional space and disparity as in the previous example, we may be interested in how disparity changes through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).
+A following up question to this observation could be to test whether this difference is due to an overall change in disparity through time or not.
+In other words, did disparity increased through time in mammals globally or is it only between stem and crown mammals.
+Using the same definition of the multidimensional space and disparity as in the previous example, we can measure, for example, changes in disparity through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).
 
@@
 
-\texttt{> summary(disparity)}
+\noindent \texttt{> summary(disparity)}
 
@@
 \end{tabular}
-\caption{Summarising a \disp object (disparity through time)}
+\caption{Summarising a \disp object (disparity through time).
+$n$ is the number of elements per subsamples, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 95\% are the confidence intervals.}
 \label{Tab:summary_time}
@@
 
-\texttt{> plot(disparity, type = 'continuous')}
+\noindent \texttt{> plot(disparity, type = 'continuous')}
 
@@
    \includegraphics[width=1\textwidth]{plot_example_time.pdf} 
-\caption{\disp plot of disparity-through-time}
+\caption{\disp plot of disparity-through-time.
+The black line represents the median disparity (median sum of variances), the dark grey and light surfaces represent respectively the 50\% an 95\% confidence intervals.}
 \label{Fig:plot_time}
@@
 
-\texttt{> anova(test.dispRity(disparity, test = lm, comparisons = 'all'))}
+\noindent \texttt{> anova(test.dispRity(disparity, test = lm, comparisons = 'all'))}
 
@@
 \end{tabular}
-\caption{ANOVA performed on a \disp object}
+\caption{ANOVA performed on a \disp object.}
 \label{Tab:anova}
@@
 
+To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
+
 \section{Additional information}
 \subsection{Manuals and vignette}
-Supplementary information about the package is available on the project page \url{https://TGuillerme/dispRity}.
-Information concerning each function can be found in R in the individual function manuals and a package manual with more detailed information is available online at \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}.
+Supplementary information about the package is available on the project page (\url{https://TGuillerme/dispRity}).
+Information concerning each function can be found in R in the individual function manuals and a package manual with more detailed information is available online (\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).
 This manual contains substantially more information and detailed examples including a tutorial for a ``classic'' disparity analysis in palaeobiology as well as an introduction to the use of this package in ecology or other disciplines.
@@
 
-\section{Data availability and reproducibility}
-All the code and the data for using the package and its vignette with examples are available in the R package help files or on GitHub \url{github.com/TGuillerme/dispRity}.
+% \section{Data availability and reproducibility}
+% All the code and the data for using the package and its vignette with examples are available in the R package help files or on GitHub \url{github.com/TGuillerme/dispRity}.
 
@@
 Many thanks to Natalie Cooper for encouraging the development of this package and helping with the writing of this paper and the package manuals.
-Thanks to Dave Bapst, Martin Brazeau, Andrew Jackson, Graeme Lloyd and Emma Sherratt for specific implementation ideas.
+Thanks to Martin Brazeau and Andrew Jackson for specific implementation ideas. %TG: add Graeme Lloyd, Dave Bapst and Emma Sherratt
 I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.

### 2017-11-06 paper/response_to_reviewers.tex
response to reviewers INIT

\documentclass[12pt,letterpaper]{article}

%Packages
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{color}
\usepackage[usenames,dvipsnames]{xcolor}
\pagenumbering{arabic}

%Pagination style and stuff
%\linespread{2}

\raggedright
\setlength{\parindent}{0.5in}
\setcounter{secnumdepth}{0} 
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}

\setlength\parindent{0pt}

\begin{document}

\textbf{RE: MEE-17-08-729 }\\
\bigskip
Dear Dr Vamosi,\\
\bigskip

I am very grateful to the four referees for their helpful and constructive comments, which we believe have helped us to significantly improve the \texttt{dispRity} package and its associated description paper.
I have taken all of their comments on board, and respond to their points below.
For improving clarity in this document, reviewers comments are displayed in blue, my responses in normal text and the changes to the manuscript in italic.
Also, I've uploaded two versions of the revised manuscript: both have the exact same content but one has all the changes and additions to the main text highlighted in yellow to help with reviewing.

% \item{\textcolor{blue}{\textbf{Graphical abstract:} Current graphical abstract does not capture the content of the article for readers in a single glance.
% The authors should not use red and green together.
% Figure 5 illustrates topology distance with different inference methods and varying proportions of missing data in rows vs columns and would make a better graphical abstract.
% For the graphical abstract, Fig 5 needs to be made clearer in one glance what is going on.
% More details are given in the comments on that figure below.}}

\section{Reviewer 1}

\begin{enumerate}

\item{\textcolor{blue}{While I understand that the author is writing a package that has many diverse capabilities and is presented to be useful to a wide range of disciplines, it is too vague in too many places to be helpful in its current state.
For example, the line in the abstract “It is then possible to use a summary metrics for measuring some properties of this space for some group of species and compare them.”.
“Some properties” is too vague without any qualifiers to be useful. But I believe that with revision, taking into account the issues of clarity, this will be an important addition to the available analytical tools for Biologists.}}

%Increase precision in the abstract

\item{\textcolor{blue}{My main concern is that what disparity actually means is left to the reader’s imagination – it is not defined or discussed anywhere.
Disparity simply means lack of similarity or equality. While the author is probably considering the dispersion of points in a multidimensional space as a measure of dissimilarity, it is important to realise the ambiguity of the word and more explanation is needed early in the introduction (and foremost in the abstract).
Specifically, the lines 58-61 need to be revised, as they simply don’t convey to the reader what is being presented here as disparity.
Say what “specific aspect of it [multidimensional space]” you are referring to.
Also, the different definitions of disparity alluded to on lines 61-63 should be summarised.
Furthermore, the manuscript is lacking a table summarising the metrics of disparity that this package can work with (which I suspect it was supposed to have).}}

%Precision up on what is disparity

\item{\textcolor{blue}{Another concern I have is about how this R package sits within the current analytical toolkit available.
How does this new package relate to existing packages available for R?
Claddis and geomorph are mentioned and discussed because they deal with bespoke data formats that need to be dealt with prior to analysis.
But are there others?
How does the package relate to the ‘vegan’ R package for multivariate analysis of ecological communities?
Within there are metrics for community dissimilarity, metrics from ordinations etc.
Vegan is a popular package for analysis of multivariate datasets and seems to be relevant here with similar functionalities but not discussed.
Similarly, the R package ‘diverse’ has a function called disparity, and the ‘geiger’ function disparity implements the disparity-through-time. }}

%TG: add more functions cross over with vegan and geiger

\end{enumerate}

\subsection{Reviewer 1 - Minor comments}

\begin{enumerate}

\item{\textcolor{blue}{The paper keywords are narrow and suggest a very particular use of disparity analyses (morphological disparity). The paper suggests there are broader applications and so these should be revised to include the diverse applications that the paper suggests. }}
\item{\textcolor{blue}{Line 4 – changes ‘in’ to ‘to}}
\item{\textcolor{blue}{Line 6 – changes ‘axis’ to ‘axes’}}
\item{\textcolor{blue}{Line 84 – Explain what the Procrustes variance is. And when is it appropriate and when is not appropriate? Be more specific.}}
\item{\textcolor{blue}{Lines 98-100 – Suggest changing this to be specific: “calculates a disparity metric from the space”, and “analyse the resulting disparity object through hypothesis testing and visualisation”}}
\item{\textcolor{blue}{Line 116 – “Table ??” = should have been caught during proofreading. Furthermore, there is no table included in the manuscript that details the implemented metrics, and this is an important part of the paper sadly omitted in this submission.}}
\item{\textcolor{blue}{Lines 356-357 – This final sentence of the conclusion section is vague and does not add to the manuscript.}}
\item{\textcolor{blue}{Figure 1 – Procrustes is a proper noun and requires a capital p. Is the Procrustes matrix a “Procrustes distance matrix” or a “Procrustes residuals matrix”? i.e. is it the shape variables after a Procrustes superimposition (latter) or the distances between individuals after a Procrustes superimposition (former)? Both are suitable inputs for ordination.}}
\item{\textcolor{blue}{Figure 2 makes little sense without a summary of the kinds of disparity metrics available to the user, and needs more explanation with examples for each dimension level.}}
\item{\textcolor{blue}{Table 1 – change to disparity-within-groups with hyphens as DTT below. Also add “shape” before data for geomorph.ordination, and perhaps a similar qualifier for Claddis.ordination.}}

\end{enumerate}

\section{Reviewer 2}

\begin{enumerate}
\item{\textcolor{blue}{Firstly, features available in the package but not currently fully explained/highlighted.
Really the only major omission here is a more detailed description of the tools avilable under the "Data Simulations" section of the manuscript.
I suspect many users will find this the *most* useful feature of the package and the author should consider unpacking this section a bit more.
Perhaps adding an additional multipanel figure exhibitng some simulated spaces.}}

% Add simulation description

\item{\textcolor{blue}{Secondly, problematic implementations.
Here there is again only one major issue, and that concerns the time-slicing options.
Initially it should be noted by the author how useful this feature could be in generating time series of disparity that are long enough for time series statistics to be applied.
Currently most palaeobiologists are stuck using binned data (e.g., Figure 4) which tends to lead to time series of insufficient length (here N = 3) to explore, for example, correlations with climate proxies (or some other hypothesised driver of morphological diversity).
Time-slicing could revolutionise the field by opening up data sets to such analyses and an example alongside the binning in Figure 4 should be considered.
However, the current implemented options are, in my view, deeply problematic.
Specifically, dispRity offers four options (L145-156), but in effect these all boil down to treating a point along a branch as either identical to the descendant (end of branch) or the ancestor (start of branch).
However, I think most biologists would treat such a value as more likely to be intermediate between these extremes and possessing some form of uncertianty depending on the generating model's assumptions (e.g., Brownian motion).
I don't mind offering a punctuated model (descendant value propagated back along branch) but some form of intermediate value should also exist.
Without this option data are likely to be biased, e.g., towards generating more "jagged" time series and perhaps more significiant slice-to-slice changes (using the packages "test" functions) than are realistic (ie.e, type I errors).}}

%TG: Add a modeling function for time-slicing

\item{\textcolor{blue}{An additional more minor implementation issue here concerns the "subsamples". Firstly, I would urge the author to reconsider this name as it implies subsampling (e.g., rarefaction) and not what is intended, i.e., subsets of the data (the term used on L124) or groups of taxa. However, a more pressing issue is that I do not see any restrictions on overlap between these "subsamples" and indeed in the case of time-binning this is suggested as a "pro" (L138). Although in cases such as time-binning it is obviously sensible to allow taxa to appear in multiple "subsamples" this changes the assumptions of a statistical test looking for differences between samples, i.e., as overlap increases so would type II errors. I suggest the author consider at least a warning to users in such circumstances.}}

%TG: add a warning to time subsamples? Change subsamples to subsets?

\item{\textcolor{blue}{Finally, additions the author could consider adding to the package. These are mostly minor, but more numerous. Some of these are simple caveats for limitations to the package. For example, disparity metrics can be generated from, for example, a discrete-character taxon matrix directly, or the distance matrix given to the ordination (e.g., the weighted mean pairwise distance metric used in Close et al. 2015, cited). Because the package only deals with matrices generated by the ordination itself these metrics are beyond its scope, although are of potential value to disparity analysis more broadly. Specifically, the Close et al. (2015) metric can be applied to an incomplete matrix, but principal coordinates/classic multidimensional scaling requires a complete distance matrix and hence taxa would have to be pruned to generate an ordination, but can be retained if using a pre-ordination metric. In a similar vein, although the package allows bootstrapping of the space it could also be useful for the raw data itself to be bootstrapped, e.g., the set of measurements used to generate a principal components analysis. It might be helpful for the author just to make it clear (with a sentence or two) that these are things a user might want to attempt, but the package cannot (reasonably) do.}}

%TG: add any matrix usage (done?) + boostrapping the raw data?

\item{\textcolor{blue}{Picking up on the first point from above, there are also additional post-ordination disparity metrics that have been used in the literature but are not available here. For example, hypercube volume, minimum spanning tree length etc. I would also urge the author to cite the work of Mike Foote, who made some early important contributions to disparity in general, and has written papers that discuss some of these potential metrics (e.g., Foote 1997, Annual Review of Ecology and Systematics, 28, 129–). I believe Wills (2001, cited) also discusses a long list of metrics.}}

%TG: implement more metrics

\item{\textcolor{blue}{The package also employs a rarefaction function, but rarefaction can be problematic in the sense that rarefaction curves can sometimes cross such that the sampling level used to make comparisons can effect which of two or more subsets is considered the most disparate. This is why algorithms designed to sample "fairly" rather than equally have been explored, see for example Kotric and Knoll (2015, Paleobiology, 41, 68–). A user may also wish to rarefy by some external data source, such as number of occurrences, or geographic regions. These are obviously major additions, but are something the author might wish to add in future.}}

% %TG: adding rarefaction algorithms.

\item{\textcolor{blue}{Other minor additions the author may also wish to consider are the permutation test from Brusatte et al. (2014, Current Biology, 24, 2386-), a means of adding the geologic timescale to time series plots (e.g., using the geoscale package), and options for the user to only use a limited set of ordination axes (matrix columns), e.g., to avoid using higher axes with negative eigenvalues.}}

\end{enumerate}

\subsection{Reviewer 2 - Minor comments}

\begin{enumerate}

\item{\textcolor{blue}{L4 - "in" should be "to"? "of variables" is then redundant and can be removed.}}
\item{\textcolor{blue}{L6 - "are the axis of this space" could just be "are the axes".}}
\item{\textcolor{blue}{L7 - "a" here should be removed.}}
\item{\textcolor{blue}{L17 - "suits best" should be "best suits".}}
\item{\textcolor{blue}{L23 - "allowing to" should be "allowing users to".}}
\item{\textcolor{blue}{L28 - "suit" should be "suite".}}
\item{\textcolor{blue}{L54 - Technically a principal coordinates analysis *is* classic multidimensional scaling (CMDS). The author may also want to cite use of non-metric multidimensional scaling (e.g., Liow 2004, American Naturalist, 164, 431–).}}
\item{\textcolor{blue}{L80 - I'm not sure I would use the term "enforce" here. I think many of us write packages that serve our own needs, but the great thing about languages like R is that it is flexible and code can be borrowed and appropriated etc. Essentially I think the author may unintentionally disparage other package authors here. Besides which, I do not feel that dispRity is competing with other packages, rather it is supplemental to them. I.e., to repeat my comment from above I see dispRity as sitting at the end of a range of data analysis pipelines with other packages sitting in front of it.}}
\item{\textcolor{blue}{L84 - Redundant "a".}}
\item{\textcolor{blue}{L116 - "??"!}}
\item{\textcolor{blue}{L130 - I don't think the term "possible observed" makes sense. Either the author means all the observed morphologies *or* all the possible morphologies. I suggest the former is safer as I think it likely that possible morphologies introduce sufficient complexity we can never really say this is true. For example, if we alter the number of elements in our sample (removing a species say) the ordination/morphospace we produce will differ, but the range of possible morphologies should not.}}
\item{\textcolor{blue}{L134 - Should probably be "an argument".}}
\item{\textcolor{blue}{L136 - It is not immediately obvious to me why a phylogeny is required here. This only becomes clearer later (time-slicing) so maybe a "heads up" parenthetic (i.e., "see below") could be added here.}}
\item{\textcolor{blue}{L208 - "looking at"?}}
\item{\textcolor{blue}{L213 - Should be "altitudinal"?}}
\item{\textcolor{blue}{L245-247 - I don't think these definitions for crown and stem are quite right. However, for the audience of MEE I suspect there is no need to define them so these parenthetic statements can just be cut.}}
\item{\textcolor{blue}{L252 - It might be more helpful to define these using the phylogeny as this is a more generally applicable approach than manually delimiting row numbers (where human error can easily creep in or the inclusion of exclusion of data can mess things up). I.e., something that identifies descendants of the crown node and then a simple setdiff() between these and all possible names to get the non-crown (stem) taxa).}}
\item{\textcolor{blue}{L296 - Should be "increase".}}
\item{\textcolor{blue}{L324 - Figure 4 caption should be "elements per subsample" and "50\% and 95\%".}}
\item{\textcolor{blue}{L357 - Should be "for such analyses".}}

\end{enumerate}

\section{Reviewer 3 (Michael Collyer)}

\begin{enumerate}

\item{\textcolor{blue}{The author presents a simple example and defers much of the many details users will encounter to his impermanent sources (vignettes, webpages), which I believe to be wise, as the R package will evolve.  I believe the purpose of a manuscript like this is illustrate its purpose and utility rather than serve as an instructional manual.  I believe the author has found the appropriate balance of instructional content between this manuscript and his other sources.}}

\item{\textcolor{blue}{The author acknowledged assistance with writing (suspect English is a second language) and the writing was great in the paper, but less so in the abstract.  Having someone double check the grammar in the abstract would be beneficial.  Also, line 116 refers to table ??.}}

\item{\textcolor{blue}{In the examples, replacing objects is not a good idea.  (This is also a problem in the help file examples in the R package.).  Remember that novice R users will use this software and might develop the habit of overwriting objects in the global environment, as updates.  It is possible to update objects with new names and retain the history during analyses, which is preferred in most cases and essential in others (for example, when a different analysis requires going back to a previously defined object.)  I recommend providing unique names.  For example, disparity is defined twice as an object on lines 255 and 261.  On line 261 it could be defined as DisparityB, and subsequently used without overwriting the object produced from a different function in line 255.}}

% %TG: rename objects in the example

\item{\textcolor{blue}{Relatedly, one should not name objects the same name as functions.  This is simply bad news.  It is not something to encourage users to do.  Using "Disparity" rather than the function name, disparity, accomplishes the same thing without potentially causing problems down the line.  (The author will field requests for help from users, so this can be considered a pitfall avoided beforehand.)  This is discussed some here: https://stackoverflow.com/questions/6135868/in-r-what-exactly-is-the-problem-with-having-variables-with-the-same-name-as-ba}}

\item{\textcolor{blue}{It took me a bit of sleuthing for how the ANOVA on line 326 was performed. 
If I have one major concern, it is here. 
Rather than develop an, e.g., anova.test.dispRity S3 generic function, the author was able to use anova.lm by developing a way to coerce the output into a class "lm" object. 
This is clever but perhaps dangerous. 
For example, if one compares the results in Tables 3 and 4, they seem largely disparate. 
I would expect a statistically significant result, as one 95\% CI from Table 3 only slightly overlaps with the others, but an effect size of F = 177.84?? 
This is highly irregular, and I believe it is because of what the "lm" coercion does and what is actually tested here. 
My discomfort should have been readily apparent. 
The ANOVA df signals that what is performed is a parametric ANOVA on the bootstrapped group disparities across bootstrap iterations, as if these values are independent observations. 
The danger here must be appreciated.
Increasing the number of bootstrap iterations effectively and artificially increases statistical power (which I would argue should not be an objective). 
The ANOVA statistics are not based on the data but on the bootstrap results, which is not at all conventional (I could not find a source to confirm this approach). 
The normal procedure is to calculate a statistic, like an F statistic, in every bootstrap iteration and then draw inference from the distribution of that statistic, much like the "percentile bootstrap" (discussed some here:% \url{https://stats.stackexchange.com/questions/20701/computing-p-value-using-bootstrap-with-r/277391#277391}).
I believe that an ANOVA consistent with the confidence interval approach in Table 3 would bootstrap values in every permutation representing a null model (e.g., the values sampled come from a population with consistent mean and variance), calculate an F statistic based on among- and within-group disparities, then draw inference from the observed F value in the distribution of random outcomes.}}

% %TG: ANOVA problem

% As a minimum, this analysis should be called something different than " ANOVA performed on a dispRity object", which is rather terse, and I would be loathe to suggest that it tests the null hypothesis that disparity does not change through time, as suggested by the author.  In fact, it simply doesn't.  I came up with this simple example to illustrate why, according to the dispRity guidance:

% n <- 40
% p <- 10
% g <- as.data.frame(as.factor(rep(1:2, n)[1:n])) # generate groups
% Y <- matrix(rnorm(n*p), n, p)# generate random data with no pattern
% rownames(Y) <- rownames(g)
% cs <- custom.subsamples(Y, g) # customized subsamples
% bd <- boot.matrix(cs, bootstraps = 100) # bootstrapped data
% sv <- dispRity(bd, metric = c(sum, variances)) # sum_variances, as per directions

% anova(test.dispRity(sv, lm, "all")) # ALWAYS. A. SIGNIFICANT! RESULT!

% To state this plainly, garbage in equals a pattern out.  This happened with every permutation I ran and the result was usually a huge effect size similar to that reported in the manuscript.

\textcolor{blue}{I can see no reason to perform ANOVA like this.  I'm not sure what it helps.  I stopped here in terms of my sleuthing but I wonder what other analytical short-cuts like this could cause problems. This is important!  As the author noted, and I agree, there is a paucity of development in this area.  One of the reasons might be that it takes time to bench test methods.  Surveying our functions in geomorph will reveal many methodological papers cited along with functions.  We release functions once we feel they are theoretically sound.  Introducing methods that have not been vetted will lead users to make dangerously incorrect inferences.  This is a difficult job but it has to be done.  While I'm excited to see this package released, to release it with known flawed analytical functions introduces a much larger mess to clean up later.  My recommendations are these:}

\begin{itemize}
\item \textcolor{blue}{1) Remove these ANOVA results from the manuscript, which do not enhance the confidence intervals much at all.}

\item \textcolor{blue}{2) Consider removing underdeveloped functions or arguments from your package until you have had sufficient time to test them.  Release them along with subsequent manuscripts that show you have tested type 1 error rates and statistical power.}

\item \textcolor{blue}{3) Or (and less recommended), put very stark and clear warnings that statistical results produced by some of these functions have not been tested and the user bares some risk making inferences from them.}

\end{itemize}

\item{\textcolor{blue}{I also have a recommendation for the package, itself, based on my experiences. 
The author should strive to make arguments and function details clearer to users, especially as using dispRity requires a lot of integration between different functions. 
Also, it is a bit of an art, but one to be practiced, to use print.\_ S3 generic functions. 
I tried to compare some of the dispRity functions to morphol.disparity in geomorph for known data sets. 
Below are results (which show that t.tests on disparities seem to produce similar results to morphol.disparity, as expected). 
I use this juxtaposition to demonstrate a subtle difference. 
A novice user will have no idea what [[1]] and [[2]] means as output, and even though I consider myself fairly experienced with R, I had to go run a t.test on some data to remind myself that "statistic" is the t-stat and "parameter" means the degrees of freedom. 
I'm not sure what the confidence intervals on df should inspire but I respect that in this case - unlike my criticism above - it appears that the bootstrapped outcomes are on all values provided by a t.test in each iteration. 
Regardless, this complexity begs for direction. 
Although the geomorph results by comparison were tedious to achieve, I can attest that it saved a lot of time spent working with users to understand results. 
The author should remember that once this manuscript is published, this R package becomes a dedication to the researchers who use it. 
Cleaning up some of these details now will make it easier to remain dedicated later.}}

\end{enumerate}

% I hope these comments are helpful and I look forward to seeing this package fully developed!  My recommendation would be to accept this manuscript with minor revision, if not for the one major issue I had with the ANOVA method.  I hope I have made my concern clear and it is otherwise clear that I believe this manuscript would be a nice contribution to MEE upon revision.

% Sincerely,
% Michael Collyer

% test.dispRity(disparity_var, test = t.test, comparisons = "pairwise",
% +               concatenate = FALSE, correction = "bonferroni",
% +               conc.quantiles = c(mean, c(95)))
% [[1]]
%                                                   statistic       2.5%     97.5%
% plethodon$species.Jord : plethodon$species.Teyah -0.06351374 -0.6744747 0.5683461

% [[2]]
%                                                 parameter     2.5%    97.5%
% plethodon$species.Jord : plethodon$species.Teyah  43.15056 36.62182 45.99247

% [[3]]
%  p.value      2.5%     97.5% 
% 0.7956904 0.4691708 0.9976864

% morphol.disparity(Y~groups , ~groups, data = gdf)
%  |=================================================================================================================| 100%

% Call:
% morphol.disparity(f1 = Y ~ groups, groups = ~groups, data = gdf)

% Randomized Residual Permutation Procedure Used
% 1000 Permutations

% Procrustes variances for defined groups
%       Jord       Teyah 
% 0.004149811 0.004234284

% Pairwise absolute differences between variances
%              Jord        Teyah
% Jord  0.000000e+00 8.447277e-05
% Teyah 8.447277e-05 0.000000e+00

% P-Values
%       Jord Teyah
% Jord  1.000 0.876
% Teyah 0.876 1.000

\section{Reviewer 4 (Gavin Simpson)}

\begin{enumerate}

\item{\textcolor{blue}{I have several concerns regarding the content in the manuscript, however, primarily how it is pitched.
It seems unclear, to me at least, who the intended audience is?
At various points I get the impression that the author intends for the package to be useful, and therefore, presents it, to readers beyond those studying morphometry etc.
There are frequent references to ecology and related disciplines where similar approaches are used.
Yet, those in these broader fields will struggle to follow what is presented as the manuscript exclusively uses "morphospaces" and related terminology that will only be relevant to a specific subset of readers.}

\textcolor{blue}{If you want to aim for a more general audience, consider rewriting the manuscript in more general terminology; samples in rows, variables in columns, for example.
Perhaps include a table to relate terminology used in morphometry with that more commonly encountered in community ecology.
Consider adding an ecological example and removing, if length is an issue, content on the manual and vignette --- we don't really need a paragraph on that, just a sentence to highlight it exists.}}

% %TG: adding a table of terms correspondance
% %TG: adding an ecology example (removing manual and vignette description)

\item{\textcolor{blue}{The referencing is poor; why cite Close et al for the concept of pairwise distances?
I'm pretty sure PCO existed before Brusatte et al, and PCA certainly before 2012, etc.
Throughout I was frustrated with this; the referencing ignores a century of work outside this specific field.
I appreciate there merit in citing relevant literature for your field, but that doesn't absolve one of a duty to accurately attribute ideas to those that came up with them.
If you are unfamiliar with developments in ideas beyond your particular field, I would suggest Legendre \& Legendre's *Numerical Ecology* textbook as a good source for original work relevant to what is presented here.}}

% %TG: source orginal papers

\item{\textcolor{blue}{The comparison or critique of packages that, in the eyes of the author, restrict users grated a little with me. As a developer of one of the packages named, I feel that particular critique is unwarranted for several reasons;
\begin{itemize}
\item i) **vegan** doesn't restrict you at all, as with anything in R you are free to pick and choose ordination methods, dissimilarity measures etc and then do what you want with the scores,
\item ii) **vegan** doesn't provide disparity measures because it was never intended for such usage, but it is general enough that a user could do whatever they want,
\item and iii) like any software, the **dispaRity** author's software included, you have to provide defaults and **vegan**'s defaults reflect the intended audience of community ecology; that's a feature, as it is of **dispaRity**. As such the argument set up on page 6 is some straw man and unnecessary to motivate your package.
\end{itemize}
}}

% %TG: Tone down the critics to other packages

\item{\textcolor{blue}{You don't really define disparity anywhere, except in very loose terms.
This needs making more explicit; I appreciate that disparity is a broad, nebulous concept, but you need to help the reader that may not be familiar with your subject area.
Either that or as mentioned above, refocus the paper on those readers already versed in this particular type of analysis.}}

\item{\textcolor{blue}{Something that I did miss, which is at the forefront of my mind as a lot of work has gone into **vegan** in this area, is how you handle negative eigenvalues and their eigenvectors?
Typically, one needs to account for these issues, which readily arise when embedding data in an Euclidean space (via PCO) using a non-metric dissimilarity.
Does **dispaRity** handle such problems, and if so how?
If not, is this because those steps are left up to the user to deal with?
If so, there are some issues that cannot be corrected by the user without strongly biasing the results of the analysis.
This is one area where there is the potential for the ideal of being accommodating of users can cause problems downstream if you are not in control of issues like non-metric dissimilarities.
An example where it needs specific handling is in multivariate dispersions (distances around centroids), but even simple ordinations based on dissimilarities require special handling.}}

\end{enumerate}

\subsection{Reviewer 4 - Minor comments}

\begin{enumerate}

\item{\textcolor{blue}{L17: insert "ability" between "researchers" and "to", and swap order of "suits" and "best" to read "best suits".}}

\item{\textcolor{blue}{L50: Here is an opportunity to bridge to the community ecology field and make more general statements regarding multivariate data; rows and samples or observations, columns are variables or some function of variables such as an embedding (PCO) or an ordination.}}

\item{\textcolor{blue}{L51--54: here referencing is very poor. Please cite relevant original research. Also, expand on what you mean by MDS; is this different from PCO, which also goes by this name --- do you mean NMDS here?}}

\item{\textcolor{blue}{L65--64 here are additional areas where referencing is poor. Consider Marti Anderson's work on multivariate ANOVA \& dispersions as earlier work, for example:
\begin{itemize}
\item Anderson, M. J. A new method for non-parametric multivariate analysis of variance. Austral Ecol. 26, 32–46 (2001)
\item Anderson, M. J. Distance-based tests for homogeneity of multivariate dispersions. Biometrics 62, 245–253 (2006).
\item Anderson, M. J., Ellingsen, K. E. \& McArdle, B. H. Multivariate dispersion as a measure of beta diversity. Ecol. Lett. 9, 683–693 (2006).
\end{itemize}
}}

\item{\textcolor{blue}{For permutation tests, significant work has been done in this area by Cajo ter Braak, principally in his Canoco software and related publications.}}

\item{\textcolor{blue}{Also comparing overlap in CIs is a terrible way to perform inference; for 95\% confidence intervals, if the intervals for two estimated means are as close as possible but not overlapping, the p value of the difference means test is 0.005, not 0.05. To achieve 0.05 you can allow considerable overlap, but how much is tricky to discern by eye. As such you are at risk of making poor choices on the basis of perceived significance.}}

\item{\textcolor{blue}{L116 There is either a missing reference or a missing table, I presume the latter as no existing table matches the textual description in and around this line.}}

\item{\textcolor{blue}{L134 Needs "an" between "as" and "argument"}}

\item{\textcolor{blue}{L210 delete "sheer"}}

\item{\textcolor{blue}{L213 "longer" not "bigger" and "altitude" not "altitudes"}}

\item{\textcolor{blue}{L218 "longer" not "bigger"}}

\item{\textcolor{blue}{L252 and throughout the code: insert straight quotes. There is a good chance these curly quotes will get replaced with some non-ASCII character and be impossible to paste into R.}}

\item{\textcolor{blue}{In the example around L236--249, can you justify the metric used here? Is the sum of variances along components of an ordination a valid measure of the dispersion in multivariate space of samples? You don't, or I didn't see, reference to original work that justifies this, and it's not something I have come across despite implementing some approaches for testing multivariate dispersion in the **vegan** package. Doesn't this suffer the same problem as confidence bars in an x-y scatterplot where the bars have been defined with reference to only one axis at a time? The solution is a confidence ellipse, which is often quite different to the interval implied by separate confidence interval bars for each axis.}}

\item{\textcolor{blue}{L294 replace "following up" with "subsequent"}}

\item{\textcolor{blue}{L323 This statement seems overly vague. What level of confidence is there in this test? The 95\% intervals overlap throughout, and although you don't show then, one can infer that say the 0.2 probability quantile of subsample 2 overlaps with the 0.975 quantile of sub sample 1.}}

\item{\textcolor{blue}{L326 Where do the 300 samples come from as implied by the residual degrees of freedom. Table 3 suggests 15 + 9 + 13 = 37 samples in this analysis. I'm clearly missing something here. Also, this is an omnibus test, which if we believe the printed values (see previous re DF) only says there is some effect of `subsamples`. Can this analysis be extended to point estimates for mean variance (again assuming you can justify this sum of individual variances method) and post-hoc comparisons of pairwise `subsample` differences, which would be quite useful?}}

\item{\textcolor{blue}{If you need to save space to add an ecological example, cut the *Manuals and vignettes* section and also the *Data simulations* section. The later is somewhat tangential to the main thrust of the paper and you cover this in the vignette/bookdown resources.}}

\end{enumerate}

\end{document}

### 2017-11-06 paper/dispRity_paper_highlighted.tex
response to reviewers INIT

--- 2017-08-25 paper/dispRity_paper.tex
@@
 % 2961w - 30p
+
+%---------------------------------------------
+%
+%       THIS IS THE HIGHLIGHTED VERSION.
+%       For the final version, carefully remove:
+%       '\protect'
+%       '\hl{' (don't forget the '}'
+%
+%---------------------------------------------
 
@@
 %Packages
+
+\usepackage{xcolor}
+\usepackage{color,soul}
 \usepackage{pdflscape}
@@
 
-% \begin{table}
-% \resizebox{\textwidth}{!}
-% {%
-%     \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
-%         name & description & dimen-sion level & definition & source \\
-%         \hline
-%         \texttt{ellipse. volume}$^1$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
-%         \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
-%         \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
-%         \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & \texttt{dispRity::diagonal} \\
-%         \texttt{mode.val} & The modal value & 1 & NA & \texttt{dispRity::mode.val}\\
-%         \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & \texttt{dispRity::ranges} \\
-%         \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & \texttt{dispRity::variances} \\
-%         \texttt{centroids} & The distance between each element and a fixed point$^2$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & \texttt{dispRity::centroids} \\
-%     \end{tabular}
-% }%
-%     \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean.
+\begin{table}
+\resizebox{\textwidth}{!}
+{%
+    \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
+        name & description & dimen-sion level & definition & source \\
+        \hline
+        \texttt{ellipse. volume}$^1$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
+        \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
+        \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
+        \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & \texttt{dispRity::diagonal} \\
+        \texttt{mode.val} & The modal value & 1 & NA & \texttt{dispRity::mode.val}\\
+        \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & \texttt{dispRity::ranges} \\
+        \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & \texttt{dispRity::variances} \\
+        \texttt{centroids} & The distance between each element and a fixed point$^2$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & \texttt{dispRity::centroids} \\
+    \end{tabular}
+}%
+    \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean.
 $^1$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA); $^2$ by default that point is the centroid of the elements.}}
-%     \label{Tab:metrics}
-% \end{table}
+    \label{Tab:metrics}
+\end{table}
 
@@
 Many thanks to Natalie Cooper for encouraging the development of this package and helping with the writing of this paper and the package manuals.
-Thanks to Martin Brazeau and Andrew Jackson for specific implementation ideas. %TG: add Graeme Lloyd, Dave Bapst and Emma Sherratt
+Thanks to Martin Brazeau and Andrew Jackson for specific implementation ideas \hl{and to Michael Collyer, Gavin Simpson and two other anonymous reviewers for constructive comments on the earlier version of this manuscript}.
+%TG: add Graeme Lloyd, Dave Bapst and Emma Sherratt
 I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.

### 2017-11-06 paper/response_to_reviewers.tex
minor responses for rev 1 and 2

@@
 }}
+
+%Increase keywords
+
 \item{\textcolor{blue}{Line 4 – changes ‘in’ to ‘to}}
+
+I have fixed this typo and removed ``variables'' (reviewer 2).
+
 \item{\textcolor{blue}{Line 6 – changes ‘axis’ to ‘axes’}}
+
+I have fixed this typo and remove ``of this space'' (reviewer 2).
+
 \item{\textcolor{blue}{Line 84 – Explain what the Procrustes variance is. And when is it appropriate and when is not appropriate? Be more specific.}}
+
+% Define Procrustes variance
+
 \item{\textcolor{blue}{Lines 98-100 – Suggest changing this to be specific: “calculates a disparity metric from the space”, and “analyse the resulting disparity object through hypothesis testing and visualisation”}}
+
+I have changed the sentence following the reviewer's suggestion.
+
 \item{\textcolor{blue}{Line 116 – “Table ??” = should have been caught during proofreading.
 Furthermore, there is no table included in the manuscript that details the implemented metrics, and this is an important part of the paper sadly omitted in this submission.}}
+
+I sincerely apologise for the omission of the Table! It is now included in the new version of the manuscript with the proper implemented metrics description.
+
 \item{\textcolor{blue}{Lines 356-357 – This final sentence of the conclusion section is vague and does not add to the manuscript.}}
+
+I removed the last sentence.
+
 \item{\textcolor{blue}{Figure 1 – Procrustes is a proper noun and requires a capital p.
@@
 Both are suitable inputs for ordination.}}
+
+I meant both matrices (as this reviewers rightly points our they are suitable for ordination). I have made this more clear in the figure.
+
 \item{\textcolor{blue}{Figure 2 makes little sense without a summary of the kinds of disparity metrics available to the user, and needs more explanation with examples for each dimension level.}}
+
+I've added a longer explanation of the different dimension-levels in the caption of the figure:
+
+\textit{A dimension-level 3 matrix would be a metric resulting in a matrix (e.g.
+the function \texttt{stats::cor} to calculate the correlation between each dimension), a dimension-level 2 metric would result in a vector (i.e.
+a distribution, e.g.
+\texttt{dispRity::variances} which calculates the variance within each dimension) and a dimension-level 1 metric would result in a single value (e.g.
+\texttt{stats::sd} which calculates the standard deviation of the input matrix).}
+
 \item{\textcolor{blue}{Table 1 – change to disparity-within-groups with hyphens as DTT below.
 Also add “shape” before data for geomorph.ordination, and perhaps a similar qualifier for Claddis.ordination.}}
+
+I've added hyphens to disparity-within-group and ``\textit{shape data}'' and ``\textit{discrete morphological data}'' for the two ordinations functions in the table.
 
@@
 
+This option is implemented in the \texttt{boot.matrix} and \texttt{dispRity} functions allowing users to set a \texttt{dimensions} argument to either a proportion or a value of dimensions to use (e.g.
+\texttt{0.95} or \texttt{10} to use only the 95\% of the dimensions or the 10 first ones respectively).
+I've added a line highlighting this feature that is indeed often used:
+
+l. @@@: \textit{Note that this function also allows to work on only a subset of dimensions via the \texttt{dimensions} argument (e.g. if only the $n$ first dimensions must be considered).}
+
 \end{enumerate}
@@
 \item{\textcolor{blue}{L6 - "are the axis of this space" could just be "are the axes".}}
+
+See response to reviewer 1, minor comment 2 and 3.
+
 \item{\textcolor{blue}{L7 - "a" here should be removed.}}
@@
 \item{\textcolor{blue}{L28 - "suit" should be "suite".}}
+
+I've fixed these typos.
+
 \item{\textcolor{blue}{L54 - Technically a principal coordinates analysis *is* classic multidimensional scaling (CMDS).
 The author may also want to cite use of non-metric multidimensional scaling (e.g., Liow 2004, American Naturalist, 164, 431–).}}
+
+%TG: check paper and cite
+
 \item{\textcolor{blue}{L80 - I'm not sure I would use the term "enforce" here.
@@
 I.e., to repeat my comment from above I see dispRity as sitting at the end of a range of data analysis pipelines with other packages sitting in front of it.}}
+
+%TG: done this down
+
 \item{\textcolor{blue}{L84 - Redundant "a".}}
+
+I've fixed this typo.
+
 \item{\textcolor{blue}{L116 - "??"!}}
+
+Sincerest apologies! See reviewer 1 minor comment 6.
+
 \item{\textcolor{blue}{L130 - I don't think the term "possible observed" makes sense.
@@
 For example, if we alter the number of elements in our sample (removing a species say) the ordination/morphospace we produce will differ, but the range of possible morphologies should not.}}
+
+I agree and have changed it to ``\textit{all the observed morphologies}'' line @@@.
+
 \item{\textcolor{blue}{L134 - Should probably be "an argument".}}
+
+I have changed the start of the sentence to ``\textit{The first argument is the matrix}'' line @@@.
+
 \item{\textcolor{blue}{L136 - It is not immediately obvious to me why a phylogeny is required here.
 This only becomes clearer later (time-slicing) so maybe a "heads up" parenthetic (i.e., "see below") could be added here.}}
+
+I've added ``\textit{see below}'' to the sentence (line @@@).
+Following this comment, I've also changed the code since, indeed, a phylogeny should not always be required!
+The function can now works without it as long as First/Last occurrence data are provided for each taxa.
+
 \item{\textcolor{blue}{L208 - "looking at"?}}
 \item{\textcolor{blue}{L213 - Should be "altitudinal"?}}
+
+I've fixed these typos.
+
 \item{\textcolor{blue}{L245-247 - I don't think these definitions for crown and stem are quite right.
 However, for the audience of MEE I suspect there is no need to define them so these parenthetic statements can just be cut.}}
+
+I've removed the inaccurate definitions in parenthesis.
+
 \item{\textcolor{blue}{L252 - It might be more helpful to define these using the phylogeny as this is a more generally applicable approach than manually delimiting row numbers (where human error can easily creep in or the inclusion of exclusion of data can mess things up).
 I.e., something that identifies descendants of the crown node and then a simple setdiff() between these and all possible names to get the non-crown (stem) taxa).}}
+
+%Implement a crown/stem phylo function
+
 \item{\textcolor{blue}{L296 - Should be "increase".}}
+
+I've fixed this typo.
+
 \item{\textcolor{blue}{L324 - Figure 4 caption should be "elements per subsample" and "50\% and 95\%".}}
+
+%Check! fixed and already
+
 \item{\textcolor{blue}{L357 - Should be "for such analyses".}}
+
+I've removed this last sentence (see Reviewer 1 - minor comment 7).
 

### 2017-11-06 paper/dispRity_paper_highlighted.tex
minor responses for rev 1 and 2

@@
 \usepackage{dcolumn}
+\usepackage{hyphenat}
 \newcolumntype{d}[1]{D{.}{.}{#1}}
@@
         \item Biological data is multivariate in essence: many traits in organisms co-vary with each other in space and time.
-        This causes biologists to either reduce these variables in a manageable number of variables or, increasingly, to use multivariate toolkits.
-        One such toolkit is based on creating a multidimensional space where the variables are the axis of this space.
-        It is then possible to use a summary metrics for measuring some properties of this space for some group of species and compare them.
+        This causes biologists to either reduce these \hl{to} a manageable number of variables or, increasingly, to use multivariate toolkits.
+        One such toolkit is based on creating a multidimensional space where the variables are the \hl{axes}.
+        It is then possible to use summary metrics for measuring some properties of this space for some group of species and compare them.
         For example, if studying morphology, one can create a morphospace for two groups of species, measure some aspect of these groups (such as their spread in the space - often termed disparity) and then test whether they have a different disparity.
@@
         However, the definition of the space, metric and test is often dependent on the software/package's author's point of view or specific question.
-        This can unfortunately hamper researchers to apply different methods that suits best their specific questions.
+        This can unfortunately hamper researchers to apply different methods that \hl{best suits} their specific questions.
 
@@
         This package allows users to define each step of the analysis (whether it is the space, the metric or the test) through a highly modular architecture where each definition can be passed as a function.
-        It also provides a tidy interface through the \disp object, allowing to easily run reproducible multivariate analysis.
+        It also provides a tidy interface through the \disp object, allowing \hl{users} to easily run reproducible multivariate analysis.
 
@@
         Furthermore, the package contains some simulation tools (e.g to simulate complex multidimensional space or morphological data).
-        Finally it also contains a suit of utilities to work with \disp objects aimed at helping users to develop their own multidimensional metrics and/or tests.
+        Finally it also contains a suit\hl{e} of utilities to work with \disp objects aimed at helping users to develop their own multidimensional metrics and/or tests.
 
@@
 Although many software packages exist for multidimensional analysis \citep[e.g.][]{bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric}, package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric and then enforce this on users by allowing little to no flexibility.
-For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of morphometric data, the disparity metric as the Procrustes variance, and uses a permutation tests to test hypotheses \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
+For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of morphometric data, the disparity metric as the Procrustes variance, and uses permutation tests to test hypotheses \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
+%Reviewer 1 minor: Explain what the Procrustes variance is. And when is it appropriate and when is not appropriate? Be more specific.
 This is ideal in some cases, but in others may not be the most appropriate analysis.
@@
 \section{Description}
-In brief, the package takes a \texttt{matrix} object (the multidimensional space), calculates some aspect of it (the disparity metric) and allows users to analyse the resulting \disp object (see Manuals and Vignettes below).
+In brief, the package takes a \texttt{matrix} object (the multidimensional space), \hl{calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation} (see Manuals and Vignettes below).
 Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow} and Table \ref{Tab:main_fun}).
@@
         \hline
-        \texttt{geomorph.ordination} & Imports data from the \texttt{geomorph} package. \\
-        \texttt{Claddis.ordination} & Imports data from the \texttt{Claddis} package. \\
+        \texttt{geomorph.ordination} & Imports \hl{shape} data from the \texttt{geomorph} package. \\
+        \texttt{Claddis.ordination} & Imports \hl{discrete morphological} data from the \texttt{Claddis} package. \\
         \texttt{custom.subsamples} & Separates data into custom subsamples. \\
@@
         \texttt{test.dispRity} & Applies statistical tests to a \disp object.\\
-        \texttt{dispRity.per.group} & Pipeline for disparity within groups analysis. \\
+        \texttt{dispRity.per.group} & Pipeline for \hl{disparity-within-groups} analysis. \\
         \texttt{dispRity.through.time} & Pipeline for disparity-through-time analysis. \\
@@
 In this example, each cell correspond to a single value (e.g.
-a 8 $\times$ 7 matrix or a vector of eight elements).}.
+a 8 $\times$ 7 matrix or a vector of eight elements).
+\hl{A dimension-level 3 matrix would be a metric resulting in a matrix (e.g.
+the function \texttt{stats::cor} to calculate the correlation between each dimension), a dimension-level 2 metric would result in a vector (i.e.
+a distribution, e.g.
+\texttt{dispRity::variances} which calculates the variance within each dimension) and a dimension-level 1 metric would result in a single value (e.g.
+\texttt{stats::sd} which calculates the standard deviation of the input matrix)}}.
 \label{Fig:levels}
@@
 disparity can only be defined as distribution or a single value).
+\hl{Note that this function also allows to work on only a subset of dimensions via the \texttt{dimensions} argument (e.g. if only the $n$ first dimensions must be considered).}
 
@@
         \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
-        \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & \texttt{dispRity::diagonal} \\
-        \texttt{mode.val} & The modal value & 1 & NA & \texttt{dispRity::mode.val}\\
-        \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & \texttt{dispRity::ranges} \\
-        \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & \texttt{dispRity::variances} \\
-        \texttt{centroids} & The distance between each element and a fixed point$^2$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & \texttt{dispRity::centroids} \\
+        \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & This package \\
+        \texttt{mode.val} & The modal value & 1 & NA & This package\\
+        \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & This package \\
+        \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & This package \\
+        \texttt{centroids} & The distance between each element and a fixed point$^2$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & This package \\
+        \texttt{ancestral. distance} & The distance between an element and its ancestor & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{k})^2}}$ & This package \\
     \end{tabular}
@@
 
+%TG: Add the proper sources for the convhulln package
+
 \subsection{Splitting the multidimensional space into subsets}
@@
 through time).
-In essence, the original space corresponds to the overall space (e.g. a morphospace contains all the possible observed morphologies).
+In essence, the original space corresponds to the overall space (e.g. a morphospace contains all the observed morphologies).
 Conversely, subsamples correspond to parts of the space with pooled characteristics (i.e. only a subset of the total morphospace).
@@
 This splitting can be done using the \texttt{custom.subsamples} or \texttt{time.subsamples} functions.
-The first takes as argument the matrix defining the multidimensional space and a list of elements to group into different subsamples.
-The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time subsamples for example) and which subsamples to create: (1) discrete time subsamples (or time-binning) or (2) continuous time subsamples (or time-slicing).
+\hl{The first argument is the matrix} defining the multidimensional space and a list of elements to group into different subsamples.
+The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time subsamples for example, \hl{see below}) and which subsamples to create: (1) discrete time subsamples (or time-binning) or (2) continuous time subsamples (or time-slicing).
 
@@
 \section{Examples}
-Multivariate analysis can be really useful for looking multiple aspects of organisms diversity together.
+Multivariate analysis can be really useful for looking \hl{at} multiple aspects of organisms diversity together.
 For example, one classic univariate trait used in ecology and evolution can be the sheer number of species present in a ecosystem or/and through time.
 This trait can be useful to test whether some ecosystems or/and time periods have more species diversity than others.
-In this case one can then test some hypothesis such as: environments with bigger altitudes gradients contain more species.
+In this case one can then test some hypothesis such as: environments with bigger \hl{altitudinal} gradients contain more species.
 
@@
 Using disparity, it is then also possible to assess whether one ecosystem or/and time period display more morphological variation.
-Following the previous example, if disparity is also higher with bigger altitudes gradients, one could argue that the species richness is caused by an increased number of morphologies adapted to more ecological niches (or something akin).
+Following the previous example, if disparity is also higher with bigger \hl{altitudinal} gradients, one could argue that the species richness is caused by an increased number of morphologies adapted to more ecological niches (or something akin).
 
@@
 One typical question with such analysis would be to test whether two groups of species have a different disparity.
-For example, using the data described above, we can test whether the crown mammals (the living mammals and their descendants) are more diverse in term of morphology than the stem mammals (the fossil mammals with no living descendants).
+For example, using the data described above, we can test whether the crown mammals are more diverse in term of morphology than the stem ones.
 In other words, whether the approximation of the volume within the morphospace (the disparity as defined above) is different in crown or stem mammals.
@@
 A following up question to this observation could be to test whether this difference is due to an overall change in disparity through time or not.
-In other words, did disparity increased through time in mammals globally or is it only between stem and crown mammals.
+In other words, did disparity increase through time in mammals globally or is it only between stem and crown mammals.
 Using the same definition of the multidimensional space and disparity as in the previous example, we can measure, for example, changes in disparity through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).
@@
 \caption{\disp plot of disparity-through-time.
-The black line represents the median disparity (median sum of variances), the dark grey and light surfaces represent respectively the 50\% an 95\% confidence intervals.}
+The black line represents the median disparity (median sum of variances), the dark grey and light surfaces represent respectively the 50\% and 95\% confidence intervals.}
 \label{Fig:plot_time}
@@
 The \disp object allows users to easily pipeline disparity analysis from the data input (the matrix) to publication standard results (tables, plots, hypothesis testing).
-With the help of this package, researchers will be able to run multivariate analysis in an easy way while still allowing the flexibility necessary to such analyses.
 

### 2017-11-08 inst/gitbook/index.Rmd
Note for figshare doi for manual
gitbook update

@@
 [![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
-[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
+[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/master/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
 [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
@@
 
+<!-- > Guillerme, T. & Cooper, N. (2016). The dispRity R package manual. Fighsare XXXX -->
+
 Note that this citation is only temporary (but can still be used!).
-A future proper version of the latest package release, this manual and an associated methods paper should be submitted soon(ish!).
+A proper description of the package is currently in review in Methods in Ecology and Evolution and should be re-submitted this winter.
+Upon acceptance (hopefully!), a proper DOI will be released for this manual and the package description.

### 2017-11-08 paper/response_to_reviewers.tex
updated minor responses

@@
 
-%Increase precision in the abstract
+I have changed the terms used in the abstract and proposed more specific examples for clarity:
+
+\tesxtit{It is then possible to measure divers aspects of the distribution of some observation (e.g.
+species) in this space.
+For example, if studying morphology, one can create a morphospace for two groups of species, measure the multidimensional volume occupied by each of these groups and then test whether these two volumes are significantly different or not.}
+
+The changes have also been made throughout the manuscript as generally suggested by all four reviewers.
+%TODO: link to other clarity changes
 
@@
 
-\item{\textcolor{blue}{Line 4 – changes ‘in’ to ‘to}}
+\item{\textcolor{blue}{Line 4 – changes ``in'' to ``to''. Line 6 – changes ``axis'' to ``axes''}}
+\label{abstract_typo}
 
-I have fixed this typo and removed ``variables'' (reviewer 2).
-
-\item{\textcolor{blue}{Line 6 – changes ‘axis’ to ‘axes’}}
-
-I have fixed this typo and remove ``of this space'' (reviewer 2).
+I have fixed these typos.
 
@@
 Furthermore, there is no table included in the manuscript that details the implemented metrics, and this is an important part of the paper sadly omitted in this submission.}}
+\label{table_missing}
 
@@
 \item{\textcolor{blue}{Lines 356-357 – This final sentence of the conclusion section is vague and does not add to the manuscript.}}
-
+\label{remove_last_sentence}
 I removed the last sentence.
@@
 
-\item{\textcolor{blue}{L4 - "in" should be "to"? "of variables" is then redundant and can be removed.}}
-\item{\textcolor{blue}{L6 - "are the axis of this space" could just be "are the axes".}}
+\item{\textcolor{blue}{L4 - "in" should be "to"? "of variables" is then redundant and can be removed. L6 - "are the axis of this space" could just be "are the axes".}}
 
-See response to reviewer 1, minor comment 2 and 3.
+See Reviewer 1 minor comment \ref{abstract_typo}.
 
-\item{\textcolor{blue}{L7 - "a" here should be removed.}}
-\item{\textcolor{blue}{L17 - "suits best" should be "best suits".}}
-\item{\textcolor{blue}{L23 - "allowing to" should be "allowing users to".}}
-\item{\textcolor{blue}{L28 - "suit" should be "suite".}}
+\item{\textcolor{blue}{L7 - "a" here should be removed. L17 - "suits best" should be "best suits". L23 - "allowing to" should be "allowing users to". L28 - "suit" should be "suite".}}
 
@@
 
-Sincerest apologies! See reviewer 1 minor comment 6.
+Sincerest apologies! See Reviewer 1 minor comment \ref{table_missing}.
 
@@
 
-\item{\textcolor{blue}{L208 - "looking at"?}}
-\item{\textcolor{blue}{L213 - Should be "altitudinal"?}}
+\item{\textcolor{blue}{L208 - "looking at"? L213 - Should be "altitudinal"?}}
+\label{altitude}
 
@@
 
-I've removed this last sentence (see Reviewer 1 - minor comment 7).
+I've removed this last sentence (see Reviewer 1 minor comment \ref{remove_last_sentence}).
 
@@
 
+%TG: explain disparity
+
 \item{\textcolor{blue}{Something that I did miss, which is at the forefront of my mind as a lot of work has gone into **vegan** in this area, is how you handle negative eigenvalues and their eigenvectors?
@@
 An example where it needs specific handling is in multivariate dispersions (distances around centroids), but even simple ordinations based on dissimilarities require special handling.}}
+
+%TG: negative egein values explnations
 
@@
 
+I've fixed these typos.
+
 \item{\textcolor{blue}{L50: Here is an opportunity to bridge to the community ecology field and make more general statements regarding multivariate data; rows and samples or observations, columns are variables or some function of variables such as an embedding (PCO) or an ordination.}}
+
+I've changed the sentence to:
+
+\textit{In all these analysis, each set of multivariate traits forms a multidimensional space.
+This space is represented as a matrix where rows are regarded as samples or observations (e.g.
+a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g.
+an embedding, scaling, ordination, etc.
+of the variables).}
 
@@
 Also, expand on what you mean by MDS; is this different from PCO, which also goes by this name --- do you mean NMDS here?}}
+
+%TODO
 
@@
 }}
+\textcolor{blue}{For permutation tests, significant work has been done in this area by Cajo ter Braak, principally in his Canoco software and related publications.
+Also comparing overlap in CIs is a terrible way to perform inference; for 95\% confidence intervals, if the intervals for two estimated means are as close as possible but not overlapping, the p value of the difference means test is 0.005, not 0.05.
+To achieve 0.05 you can allow considerable overlap, but how much is tricky to discern by eye.
+As such you are at risk of making poor choices on the basis of perceived significance.}
 
-\item{\textcolor{blue}{For permutation tests, significant work has been done in this area by Cajo ter Braak, principally in his Canoco software and related publications.}}
-
-\item{\textcolor{blue}{Also comparing overlap in CIs is a terrible way to perform inference; for 95\% confidence intervals, if the intervals for two estimated means are as close as possible but not overlapping, the p value of the difference means test is 0.005, not 0.05.
-To achieve 0.05 you can allow considerable overlap, but how much is tricky to discern by eye.
-As such you are at risk of making poor choices on the basis of perceived significance.}}
+%TODO I agree with the last point of this comment (that comparing CI's overlap is a terrible method!) -> change to "has been done in variable ways, sometimes leading to inaccuracies,"
 
@@
 
-\item{\textcolor{blue}{L134 Needs "an" between "as" and "argument"}}
+Sincerest apologies! See Reviewer 1 minor comment \ref{table_missing}.
 
-\item{\textcolor{blue}{L210 delete "sheer"}}
+\item{\textcolor{blue}{L134 Needs "an" between "as" and "argument". L210 delete "sheer". L213 "longer" not "bigger" and "altitude" not "altitudes". L218 "longer" not "bigger"}}
 
-\item{\textcolor{blue}{L213 "longer" not "bigger" and "altitude" not "altitudes"}}
-
-\item{\textcolor{blue}{L218 "longer" not "bigger"}}
+I've fixed these typos (see also Reviewer 2 minor comment \ref{altitude}).
 
@@
 There is a good chance these curly quotes will get replaced with some non-ASCII character and be impossible to paste into R.}}
+
+I've converted all quotes in quoted code lines as ASCII straight quotes.
 
@@
 
+%TG: justify from palaeo litterature? + add a note "One of the caveats of such method is that it implicitely ignores the covariance between the ordination axis."
+
 \item{\textcolor{blue}{L294 replace "following up" with "subsequent"}}
+
+I've changed the sentence following this suggestion.
 
@@
 The 95\% intervals overlap throughout, and although you don't show then, one can infer that say the 0.2 probability quantile of subsample 2 overlaps with the 0.975 quantile of sub sample 1.}}
+
+%TODO
 
@@
 
+%TODO correct/explain this 300 subsamples + add a posthoc t-test.
+
 \item{\textcolor{blue}{If you need to save space to add an ecological example, cut the *Manuals and vignettes* section and also the *Data simulations* section.
 The later is somewhat tangential to the main thrust of the paper and you cover this in the vignette/bookdown resources.}}
+
+%TODO cut down the manual and vignette part + simulation part: just link to it?
 

### 2017-11-08 paper/dispRity_paper_highlighted.tex
updated minor responses

@@
 \usepackage{hyphenat}
+\usepackage[T1]{fontenc}
+\usepackage{textcomp}
 \newcolumntype{d}[1]{D{.}{.}{#1}}
@@
         One such toolkit is based on creating a multidimensional space where the variables are the \hl{axes}.
-        It is then possible to use summary metrics for measuring some properties of this space for some group of species and compare them.
-        For example, if studying morphology, one can create a morphospace for two groups of species, measure some aspect of these groups (such as their spread in the space - often termed disparity) and then test whether they have a different disparity.
+        It is then possible to \hl{measure divers aspects of the distribution of some observation (e.g. species) in this space}.
+        For example, if studying morphology, one can create a morphospace for two groups of species, measure \hl{the multidimensional volume occupied by each of these groups and then test whether these two volumes are significantly different or not}.
 
@@
         However, the definition of the space, metric and test is often dependent on the software/package's author's point of view or specific question.
-        This can unfortunately hamper researchers to apply different methods that \hl{best suits} their specific questions.
+        This can unfortunately hamper researchers \hl{ability} to apply different methods that \hl{best suits} their specific questions.
 
@@
 % Variety of multidimensional spaces, metrics and tests
-In all these analysis, each set of multivariate traits forms a multidimensional space, usually represented as matrices where the columns are the dimensions (or some transformation thereof) and rows are the elements (often species or individuals) within this space.
+In all these analysis, each set of multivariate traits forms a multidimensional space.
+\hl{This space is represented as a matrix where rows are regarded as samples or observations (e.g.
+a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g.
+an embedding, scaling, ordination, etc.
+of the variables).}
 These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix \citep{Close2015}, or as outputs from a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, or a multidimensional scaling \citep[MDS;][]{DonohueDim}, among others.
@@
 Multivariate analysis can be really useful for looking \hl{at} multiple aspects of organisms diversity together.
-For example, one classic univariate trait used in ecology and evolution can be the sheer number of species present in a ecosystem or/and through time.
+For example, one classic univariate trait used in ecology and evolution can be the number of species present in a ecosystem or/and through time.
 This trait can be useful to test whether some ecosystems or/and time periods have more species diversity than others.
-In this case one can then test some hypothesis such as: environments with bigger \hl{altitudinal} gradients contain more species.
+In this case one can then test some hypothesis such as: environments with \hl{longer altitudinal} gradients contain more species.
 
@@
 Using disparity, it is then also possible to assess whether one ecosystem or/and time period display more morphological variation.
-Following the previous example, if disparity is also higher with bigger \hl{altitudinal} gradients, one could argue that the species richness is caused by an increased number of morphologies adapted to more ecological niches (or something akin).
+Following the previous example, if disparity is also higher with \hl{longer altitudinal} gradients, one could argue that the species richness is caused by an increased number of morphologies adapted to more ecological niches (or something akin).
 
@@
 
-\noindent \texttt{> mammal\_groups <- list('crown' = c(16, 19:41, 45:50), 'stem' = c(1:15, 17:18, 42:44))}
+\noindent \texttt{> mammal\_groups <- list(\textquotedbl crown\textquotedbl = c(16, 19:41, 45:50), \textquotedbl stem\textquotedbl = c(1:15, 17:18, 42:44))}
 
@@
 \subsubsection{Typical disparity-through-time analysis}
-A following up question to this observation could be to test whether this difference is due to an overall change in disparity through time or not.
+A \hl{subsequent} question to this observation could be to test whether this difference is due to an overall change in disparity through time or not.
 In other words, did disparity increase through time in mammals globally or is it only between stem and crown mammals.
@@
 
-\noindent \texttt{> subsamples <- time.subsamples(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, method = 'discrete')}\\
+\noindent \texttt{> subsamples <- time.subsamples(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
 \noindent \texttt{> bootstraps <- boot.matrix(subsamples)}\\
@@
 
-\noindent \texttt{> plot(disparity, type = 'continuous')}
+\noindent \texttt{> plot(disparity, type = \textquotedbl continuous\textquotedbl)}
 
@@
 
-\noindent \texttt{> anova(test.dispRity(disparity, test = lm, comparisons = 'all'))}
+\noindent \texttt{> anova(test.dispRity(disparity, test = lm, comparisons = \textquotedbl all\textquotedbl))}
 
@@
 %TG: add Graeme Lloyd, Dave Bapst and Emma Sherratt
-I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.
+I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D.
+Brazeau \hl{and from the Discovery Project Grant DP170103227 awarded to Vera Weisbecker}.
 
 \section{Funding}
-This work was funded by the European Research Council under the European Union's Seventh Framework Programme (FP/2007–2013)/ERC Grant Agreement number 311092.
+This work was funded by the European Research Council under the European Union's Seventh Framework Programme (FP/2007–2013)/ERC Grant Agreement number 311092 \hl{and the Discovery Project Grant number DP170103227}.
 

### 2017-11-08 paper/response_to_reviewers.tex
Updated responses

@@
 
-\tesxtit{It is then possible to measure divers aspects of the distribution of some observation (e.g.
+\textit{It is then possible to measure divers aspects of the distribution of some observation (e.g.
 species) in this space.
-For example, if studying morphology, one can create a morphospace for two groups of species, measure the multidimensional volume occupied by each of these groups and then test whether these two volumes are significantly different or not.}
+For example, if studying morphology, one can create a morphospace for two groups of species, measure the multidimensional volume occupied by each of these groups and then test whether these two volumes are significantly different or not.} line @@@
 
@@
 While the author is probably considering the dispersion of points in a multidimensional space as a measure of dissimilarity, it is important to realise the ambiguity of the word and more explanation is needed early in the introduction (and foremost in the abstract).
+%TODO improve disparity definition in the abstract
 Specifically, the lines 58-61 need to be revised, as they simply don’t convey to the reader what is being presented here as disparity.
@@
 %Precision up on what is disparity
+
+As Reviewer 4 (major comment \ref{nebulous}) also underlines: ``disparity is a broad, nebulous concept''.
+This observation was the genesis for this package and I am currently working on a collaborative review to define the borders (or the absence thereof) of this concept.
+I feel that explaining this would be out of the scope of this software description paper.
+However, I have tried to give specific examples each time I refer to disparity as a way to measure these ``specific aspect of the space'' and I have also added a clearer definition of what I mean by disparity throughout this manuscript.
+
+I rewrote the part in the introducing disparity as follows:
+
+\textit{It is then possible to measure how the observations are distributed within this space to answer related questions (e.g.
+“does group A occupies more space than group B?”).
+This requires the definition of a proxy for space occupancy: the disparity metric (or index; Hopkins and Gerber, 2017) which can be measured in a multitude of ways.
+For example, one could use a metric based on the variance or the range of each axis of space (Wills, 2001; Ciampaglio et al., 2001) , a distance (e.g.
+Euclidean) measured between observations (Foote, 1993, 1996) , a more direct approximation of the hyper volume (Cornwell et al., 2006; Donohue et al., 2013) , or many more (e.g.
+Navarro, 2003).} lines @@@
 
@@
 
+Procrustes variance = Sum of the diagonal covariance matrix between each elements divided by the number of elements.
+This is appropriate when looking simply at the volume in the multidimensional space.
+Howver, does not work if we want to look at the distance between two groups or if we want to use another proxy than the procrustes variance for the space occupancy (e.g.
+the ellipsoid volume might and direction might be more appropriate in XXX case).
+
+%From the manual: Procrustes variance is the same sum of the diagonal elements of the group covariance matrix divided by the number of observations in the group
+
+%From Michael Collyer: As the author indicates, in geomorph we limit our disparity measure to Procrustes variance (same as the disparity metric used in dispRity) and hypothesis tests to use randomization tests that randomly permute residuals (not data) over many permutations to generate pairwise differences in disparity.
+
 \item{\textcolor{blue}{Lines 98-100 – Suggest changing this to be specific: “calculates a disparity metric from the space”, and “analyse the resulting disparity object through hypothesis testing and visualisation”}}
@@
 
-I sincerely apologise for the omission of the Table! It is now included in the new version of the manuscript with the proper implemented metrics description.
+I sincerely apologise for the omission of the Table! It is now included in the new version of the manuscript with the currently implemented metrics description.
+I would also like to underline that although not all usual disparity metrics are not represented in the table and in the package (by far!) the idea of the \texttt{dispRity} package and specifically the \texttt{metric} argument in the \texttt{dispRity} function is to allow users to come up with their \textit{own} metric.
+
+For example, if one decides to use the relative ellipsoid hyper volume to the number of elements per group as a disparity metric (i.e.
+this metric could be used as a proxy for the average space occupancy per unit element - e.g.
+per taxon).
+This would not be doable directly by using the implemented \texttt{ellipse.volume} function but would require the user to use a function like:\\
+\texttt{dispRity(my\_data, metric = function(X) ellipse.volume(X)/nrows(X))}\\
 
@@
 \label{remove_last_sentence}
+
 I removed the last sentence.
@@
 
-I meant both matrices (as this reviewers rightly points our they are suitable for ordination). I have made this more clear in the figure.
+I meant both matrices (as this reviewers rightly points out they are suitable for ordination).
+I have made this more clear in the figure.
+On a more general note, the package now allows users to input any matrix (although the package does not provide warranty that the provided matrix is meaningful to the user's specific analysis).
 
@@
 Either that or as mentioned above, refocus the paper on those readers already versed in this particular type of analysis.}}
+\label{nebulous}
 

### 2017-11-08 paper/dispRity_paper_highlighted.tex
Updated responses

@@
 of the variables).}
-These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix \citep{Close2015}, or as outputs from a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, or a multidimensional scaling \citep[MDS;][]{DonohueDim}, among others.
+These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix \citep[][and references therein]{lloyd2016estimating}, or as outputs from a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, or a multidimensional scaling \citep[MDS;][]{DonohueDim}, among others.
 The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
 For example, when using morphological traits the multivariate space will be a morphospace, when using ecological traits it may be referred to as an ecospace or trait space.
-Once the multidimensional space is defined it is then possible to measure a specific aspect of it.
-This specific measurement (hereafter referred to as a ``disparity metric'') can in turn be defined in many different ways depending on the specific question being investigated \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015}.
+
+\hl{It is then possible to measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupies more space than group B?'').
+This requires the definition of a proxy for space occupancy: the disparity metric}
+\citep[or index;][]{Hopkins2017}
+\hl{which can be measured in a multitude of ways.}
+\hl{For example, one could use a metric based on the variance or the range of each axis of space}
+\citep{Wills2001, Ciampaglio2001}
+\hl{, a distance (e.g. Euclidean) measured between observations}
+\citep{foote1993contributions,Foote29111996}
+\hl{, a more direct approximation of the hyper volume }
+\citep{cornwell2006trait,DonohueDim}
+\hl{, or many more }
+\citep[e.g.][]{Navarrro2003MDA}.
+
 Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance \citep[e.g.][]{Brusatte12092008}, multidimensional permutation tests \citep[e.g.][]{diaz2016global} or even simply by looking at the confidence interval overlaps between disparity measurements \citep[e.g.][]{halliday2016eutherian}.
@@
 \section{Description}
-In brief, the package takes a \texttt{matrix} object (the multidimensional space), \hl{calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation} (see Manuals and Vignettes below).
+In brief, the package takes a \texttt{matrix} object (the multidimensional space), 
+\hl{calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation}
+(see Manuals and Vignettes below).
 Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow} and Table \ref{Tab:main_fun}).
@@
    \includegraphics[width=0.8\textwidth]{../inst/gitbook/dispRity_fun.pdf} 
-\caption{Illustration of the different metric dimension-levels in the \disp package.
-In this example, each cell correspond to a single value (e.g.
-a 8 $\times$ 7 matrix or a vector of eight elements).
+\caption{Illustration of the different metric dimension-levels in the \disp package. In this example, each cell correspond to a single value (e.g. a 8 $\times$ 7 matrix or a vector of eight elements).
 \hl{A dimension-level 3 matrix would be a metric resulting in a matrix (e.g.
@@
 This splitting can be done using the \texttt{custom.subsamples} or \texttt{time.subsamples} functions.
-\hl{The first argument is the matrix} defining the multidimensional space and a list of elements to group into different subsamples.
-The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time subsamples for example, \hl{see below}) and which subsamples to create: (1) discrete time subsamples (or time-binning) or (2) continuous time subsamples (or time-slicing).
+\hl{The first argument is the matrix}
+defining the multidimensional space and a list of elements to group into different subsamples.
+The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time subsamples for example,
+\hl{see below}
+) and which subsamples to create: (1) discrete time subsamples (or time-binning) or (2) continuous time subsamples (or time-slicing).
 
@@
 \section{Examples}
-Multivariate analysis can be really useful for looking \hl{at} multiple aspects of organisms diversity together.
+Multivariate analysis can be really useful for looking
+ \hl{at} 
+ multiple aspects of organisms diversity together.
 For example, one classic univariate trait used in ecology and evolution can be the number of species present in a ecosystem or/and through time.
 This trait can be useful to test whether some ecosystems or/and time periods have more species diversity than others.
-In this case one can then test some hypothesis such as: environments with \hl{longer altitudinal} gradients contain more species.
+In this case one can then test some hypothesis such as: environments with
+ \hl{longer altitudinal} 
+ gradients contain more species.
 
@@
 Using disparity, it is then also possible to assess whether one ecosystem or/and time period display more morphological variation.
-Following the previous example, if disparity is also higher with \hl{longer altitudinal} gradients, one could argue that the species richness is caused by an increased number of morphologies adapted to more ecological niches (or something akin).
+Following the previous example, if disparity is also higher with
+ \hl{longer altitudinal} 
+ gradients, one could argue that the species richness is caused by an increased number of morphologies adapted to more ecological niches (or something akin).
 
@@
 \subsubsection{Typical disparity-through-time analysis}
-A \hl{subsequent} question to this observation could be to test whether this difference is due to an overall change in disparity through time or not.
+A 
+\hl{subsequent} 
+question to this observation could be to test whether this difference is due to an overall change in disparity through time or not.
 In other words, did disparity increase through time in mammals globally or is it only between stem and crown mammals.
@@
 Many thanks to Natalie Cooper for encouraging the development of this package and helping with the writing of this paper and the package manuals.
-Thanks to Martin Brazeau and Andrew Jackson for specific implementation ideas \hl{and to Michael Collyer, Gavin Simpson and two other anonymous reviewers for constructive comments on the earlier version of this manuscript}.
-%TG: add Graeme Lloyd, Dave Bapst and Emma Sherratt
-I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D.
-Brazeau \hl{and from the Discovery Project Grant DP170103227 awarded to Vera Weisbecker}.
+Thanks to Martin Brazeau and Andrew Jackson for specific implementation ideas 
+\hl{and to Michael Collyer, Gavin Simpson and two other anonymous reviewers for constructive comments on the earlier version of this manuscript}. %TG: add Graeme Lloyd, Dave Bapst and Emma Sherratt
+I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau 
+\hl{and from the Discovery Project Grant DP170103227 awarded to Vera Weisbecker}.
 
 \section{Funding}
-This work was funded by the European Research Council under the European Union's Seventh Framework Programme (FP/2007–2013)/ERC Grant Agreement number 311092 \hl{and the Discovery Project Grant number DP170103227}.
+This work was funded by the European Research Council under the European Union's Seventh Framework Programme (FP/2007–2013)/ERC Grant Agreement number 311092 
+\hl{and the Discovery Project Grant number DP170103227}.
 

### 2017-11-10 paper/response_to_reviewers.tex
reviewer 1 update

@@
 
-% \item{\textcolor{blue}{\textbf{Graphical abstract:} Current graphical abstract does not capture the content of the article for readers in a single glance.
-% The authors should not use red and green together.
-% Figure 5 illustrates topology distance with different inference methods and varying proportions of missing data in rows vs columns and would make a better graphical abstract.
-% For the graphical abstract, Fig 5 needs to be made clearer in one glance what is going on.
-% More details are given in the comments on that figure below.}}
-
 \section{Reviewer 1}
@@
 The changes have also been made throughout the manuscript as generally suggested by all four reviewers.
-%TODO: link to other clarity changes
 
@@
 While the author is probably considering the dispersion of points in a multidimensional space as a measure of dissimilarity, it is important to realise the ambiguity of the word and more explanation is needed early in the introduction (and foremost in the abstract).
-%TODO improve disparity definition in the abstract
 Specifically, the lines 58-61 need to be revised, as they simply don’t convey to the reader what is being presented here as disparity.
@@
 
-%Precision up on what is disparity
+%TODO improve disparity definition in the abstract
 
@@
 
+%TODO
 %TG: add more functions cross over with vegan and geiger
@@
 
+%TODO
 %Increase keywords
@@
 
-% Define Procrustes variance
+In \texttt{geomorph}, the procrustes variance is defined as the sum of the diagonal covariance matrix between each elements divided by the number of elements (or effectively the sum of the squared ordination values divided by the number of elements - see example in \texttt{R} for \texttt{?dispRity::geomorph.ordination}).
+Using this disparity metric is appropriate for testing volume based hypothesis (e.g.
+Does group A occupies more morphospace than group B?) but does not allow to test position based hypothesis (e.g.
+Does group A occupies a different part of the morphospace than group B).
+For example, one might find out that group A and B have the same volume but that doesn't tell us if they occupy the same position in the morphospace.
 
-Procrustes variance = Sum of the diagonal covariance matrix between each elements divided by the number of elements.
-This is appropriate when looking simply at the volume in the multidimensional space.
-Howver, does not work if we want to look at the distance between two groups or if we want to use another proxy than the procrustes variance for the space occupancy (e.g.
-the ellipsoid volume might and direction might be more appropriate in XXX case).
+I've highlighted this in the text:
 
-%From the manual: Procrustes variance is the same sum of the diagonal elements of the group covariance matrix divided by the number of observations in the group
-
-%From Michael Collyer: As the author indicates, in geomorph we limit our disparity measure to Procrustes variance (same as the disparity metric used in dispRity) and hypothesis tests to use randomization tests that randomly permute residuals (not data) over many permutations to generate pairwise differences in disparity.
+\textit{For example, in the excellent and widely used} \texttt{geomorph} \textit{package, morphological disparity analysis uses the} \texttt{morphol.disparity} \textit{function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance), and uses permutation tests(Zelditch et al., 2012; Adams and Otarola-Castillo, 2013; Adams et al., 2017).
+This is ideal for testing volume based hypothesis (e.g.
+does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g.
+does groups A and B occupy the same location?).}
 

### 2017-11-10 inst/gitbook/04_making-up-data.Rmd
Updated manual character simulation
needs to do morphospace simulations

---
title: "Data simulation features"
author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# Making stuff up!

The `dispRity` package also offers some advanced data simulation features to allow to test hypothesis, explore ordinate-spaces or metrics properties or simply playing around with data!
All the following functions are based on the same modular architecture of the package and therefore can be used with most of the functions of the package.

## Simulating discrete morphological data

The function `sim.morpho` allows to simulate discrete morphological data matrices (sometimes referred to as "cladistic" matrices).
It allows to evolve multiple discrete characters on a given phylogenetic trees, given different models, rates, and states.
It even allows to include proper inapplicable data to make datasets as messy as in real life!

In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
The methods are described in details below

```{r}
set.seed(3)
## Simulating a starting tree with 15 taxa as a random coalescent tree
my_tree <- rcoal(15)

## Generating a matrix with 100 characters (85% binary and 15% three state) and
## an equal rates model with a gamma rate distribution (0.5, 1) with no 
## invariant characters.
my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)

## The first few lines of the matrix
my_matrix[1:5, 1:10]

## Checking the matrix properties with a quick Maximum Parsimony tree search
check.morpho(my_matrix, my_tree)
```

Note that this example produces a tree with a great consistency index and an identical topology to the random coalescent tree!
Nearly too good to be true...

### A more detailed description

The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;OReilly2017].

 * The first `tree` argument will be the tree on which to "evolve" the characters and therefore requires branch length.
 You can generate quick and easy random Yule trees using `ape::rtree(number_of_taxa)` but I would advise to use more realistic trees for more realistic simulations based on more realistic models (really realistic then) using the function `tree.bd` from the [`diversitree`](http://www.zoology.ubc.ca/prog/diversitree/) package [@fitzjohndiversitree2012].
 * The second argument, `character` is the number of characters. Pretty straight forward.
 * The third, `states` is the proportion of characters states above two (yes, the minimum number of states is two). This argument intakes the proportion of *n*-states characters, for example `states = c(0.5,0.3,0.2)` will generate 50% of binary-state characters, 30% of three-state characters and 20% of four-state characters. There is no limit in the number of state characters proportion as long as the total makes up 100%. 
 * The forth, `model` is the evolutionary model for generating the character(s). More about this below.
 * The fifth and sixth, `rates` and `substitution` are the model parameters described below as well.
 * Finally, the two logical arguments, are self explanatory: `invariant` whether to allow invariant characters (i.e. characters that don't change) and `verbose` whether to print the simulation progress on your console.

#### Available evolutionary models

There are currently three evolutionary models implemented in `sim.morpho` but more will come in the future.
Note also that they allow fine tuning parameters making them pretty plastic!

 * `"ER"`: this model allows any number of character states and is based on the Mk model [@lewisa2001]. It assumes a unique overall evolutionary rate equal substitution rate between character states. This model is based on the `ape::rTraitDisc` function.
 * `"HKY"`: this is binary state character model based on the molecular HKY model [@HKY85]. It uses the four molecular states (A,C,G,T) with a unique overall evolutionary rate and a biased substitution rate towards transitions (A <-> G or C <-> T) against transvertions (A <-> C and G <-> T). After evolving the nucleotide, this model transforms them into binary states by converting the purines (A and G) into state 0 and the pyrimidines (C and T) into state 1. This method is based on the `phyclust::seq.gen.HKY` function and was first proposed by @OReilly2016.
 * `"MIXED"`: this model uses a random (uniform) mix between both the `"ER"` and the `"HKY"` models.

The models can take the following parameters:
(1) `rates` is the evolutionary rate (i.e. the rate of changes along a branch: the evolutionary speed) and (2) `substitution` is the frequency of changes between one state or another.
For example if a character can have high probability of changing (the *evolutionary* rate) with, each time a change occurs a probability of changing from state *X* to state *Y* (the *substitution* rate).
Note that in the `"ER"` model, the substitution rate is ignore because... by definition this (substitution) rate is equal!

The parameters arguments `rates` and `substitution` takes a distributions from which to draw the parameters values for each character.
For example, if you want an `"HKY"` model with an evolutionary rate (i.e. speed) drawn from a uniform distribution bounded between 0.001 and 0.005, you can define it as `rates = c(runif, min = 0.001, max = 0.005)`, `runif` being the function for random draws from a uniform distribution and `max` and `min` being the distribution parameters.
These distributions should always be passed in the format `c(random_distribution_function, distribution_parameters)` with the names of the distribution parameters arguments.

#### Checking the results

An additional function, `check.morpho` runs a quick Maximum Parsimony tree search using the `phangorn` parsimony algorithm.
It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g. the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree to determine how different they are.

#### Adding inapplicable characters

Once a matrix is generated, it is possible to apply inapplicable characters to it for increasing realism!
Inapplicable characters are commonly designated as `NA` or simply `-`.
They differ from missing characters `?` in their nature by being inapplicable rather than unknown.
For example, considering a binary character defined as "color of the tail" with the following states "blue" and "red"; on a taxon with no tail, the character should be coded as inapplicable ("`-`") since the state of the character "color of tail" is *known*: it's neither "blue" or "red", it's just not there!
It contrasts with coding it as missing ("`?`" - also called as ambiguous) where the state is *unknown*, for example, the taxon of interest is a fossil where the tail has no colour preserved or is not present at all due to bad conservation!

This type of characters can be added to the simulated matrices using the `apply.NA` function/
It takes, as arguments, the `matrix`, the source of inapplicability (`NAs` - more below), the `tree` used to generate the matrix and the two same `invariant` and `verbose` arguments as defined above.
The `NAs` argument allows two types of sources of inapplicability:

 * `"character"` where the inapplicability is due to the character (e.g. coding a character tail for species with no tail).
 In practice, the algorithm chooses a character *X* as the underlying character (e.g. "presence and absence of tail"), arbitrarily chooses one of the states as "absent" (e.g. 0 = absent) and changes in the next character *Y* any state next to character *X* state 0 into an inapplicable token ("`-`").
 This simulates the inapplicability induced by coding the characters (i.e. not always biological).
 * `"clade"` where the inapplicability is due to evolutionary history (e.g. a clade loosing its tail).
 In practice, the algorithm chooses a random clade in the tree and a random character *Z* and replaces the state of the taxa present in the clade by the inapplicable token ("`-`").
 This simulates the inapplicability induced by evolutionary biology (e.g. the lose of a feature in a clade).

To apply these sources of inapplicability, simply repeat the number of inapplicable sources for the desired number of characters with inapplicable data.

```{r}
## Generating 5 "character" NAs and 10 "clade" NAs
my_matrix_NA <- apply.NA(my_matrix, tree = my_tree,
                         NAs = c(rep("character", 5), rep("clade", 10)))

## The first few lines of the resulting matrix
my_matrix_NA[1:10, 1:10]
```
<!--TG: Add inapplicable paper here -->

### Parameters for a realistic(ish) matrix

There are many parameters that can create a "realistic" matrix (i.e. not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all end up creating "good" matrices.
The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.

```{r}
set.seed(0)
## tree
my_tree <- rcoal(15)
## matrix
morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
check.morpho(morpho_mat, my_tree)
```

## Simulating multidimensional spaces
Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
This function allows users to simulate multidimensional spaces with a certain number of properties.
It takes as arguments the number of elements (data points) and dimensions to create the space but also permits more fine tuning in the data simulation: it is possible to give a specific distribution to each dimension, provide a correlation matrix to link the dimensions or even provide an *a priori* distribution of the variance for each distributions!
<!-- ooh fancy! -->

[@diaz2016global]

```{r, fig.width=8, fig.height=8, eval=TRUE}
## Graphical options
op <- par(mfrow = (c(2, 2)), bty = "n")
## Visualising 3D spaces
require(scatterplot3d)

## A cube space
scatterplot3d(space.maker(2500, 3, runif), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Uniform cube space")

## A plane space
scatterplot3d(space.maker(2500, 3, c(runif, runif, runif),
              arguments = list(list(min = 0, max = 0), NULL, NULL)), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Uniform plane space")

## An ellipsoid space (=a spheric space with correlation)
cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3)
scatterplot3d(space.maker(2500, 3, rnorm, cor.matrix = cor_matrix), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Normal ellipsoid space")

## A cylindrical space with a decreasing variance per axis
scatterplot3d(space.maker(2500, 3, c(rnorm, rnorm, runif),
              scree = c(0.7, 0.2, 0.1)), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Normal cylindrical space")

## Resetting the graphic parameters
par(op)
```

### 2017-11-10 inst/gitbook/05_guts.Rmd
Updated manual character simulation
needs to do morphospace simulations

--- 2017-08-15 inst/gitbook/04_guts.Rmd
@@
 
-# The guts of the dispRity package
+# The guts of the `dispRity` package
 
@@
 
+## The `dispRity` object content
+The functions above are utilities to easily and safely access different elements in the `dispRity` object.
+Alternatively, of course, each elements can be accessed manually.
+Here is an explanation on how it works.
+The `dispRity` object is a `list` of two to four elements, each of which are detailed below:
+
+ * `$matrix`: an object of class `matrix`, the full multidimensional space.
+ * `$call`: an object of class `list` containing information on the `dispRity` object content.
+ * `$subsamples`: an object of class `list` containing the subsamples of the multidimensional space.
+ * `$disparity`: an object of class `list` containing the disparity values.
+
+The `dispRity` object is loosely based on `C` structure objects.
+In fact, it is composed of one unique instance of a matrix (the multidimensional space) upon which the metric function is called via "pointers" to only a certain number of elements and/or dimensions of this matrix.
+This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint:  at any time, only one matrix is present in the `R` environment rather than multiple copies of it for each subsample.
+
+### `$matrix`
+This is the multidimensional space, stored in the `R` environment as a `matrix` object.
+It requires row names but not column names.
+By default, if the row names are missing, `dispRity` function will arbitrarily generate them in numeric order (i.e. `rownames(matrix) <- 1:nrow(matrix)`).
+This element of the `dispRity` object is never modified.
+
+### `$call`
+This element contains the information on the `dispRity` object content.
+It is a `list` that can contain the following:
+
+ * `$call$subsamples`: a vector of `character` with information on the subsamples type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"punctuated"`, `"gradual"`).
+This element generated only once via `time.subsamples()` and `custom.subsamples()`.
+ * `$call$dimensions`: either a single `numeric` value indicating how many dimensions to use or a vector of `numeric` values indicating which specific dimensions to use.
+This element is by default the number of columns in `$matrix` but can be modified through `boot.matrix()` or `dispRity()`.
+ * `$call$bootstrap`: this is a `list` containing three elements:
+    * `[[1]]`: the number of bootstrap replicates (`numeric`)
+    * `[[2]]`: the bootstrap method (`character`)
+    * `[[3]]`: the rarefaction levels (`numeric` vector)
+ * `$call$disparity`: this is a `list` containing one element, `$metric`, that is a `list` containing the different functions passed to the `metric` argument in `dispRity`.
+These are `call` elements and get modified each time the `dispRity` function is used (the first element is the first metric(s), the second, the second metric(s), etc.).
+
+### `$subsamples`
+This element contain the eventual subsamples of the multidimensional space.
+It is a `list` of subsample names.
+Each subsample name is in turn a `list` of at least one element called `elements` which is in turn a `matrix`.
+This `elements` matrix is the raw (observed) elements in the subsamples.
+The `elements` matrix is composed of `numeric` values in one column and _n_ rows (the number of elements in the subsample).
+Each of these values are a "pointer" (`C` inspired) to the element of the `$matrix`.
+For example, lets assume a `dispRity` object called `disparity`, composed of at least one subsamples called `sub1`:
+
+```
+ disparity$subsamples$sub1$elements
+      [,1]
+ [1,]    5
+ [2,]    4
+ [3,]    6
+ [4,]    7
+```
+
+The values in the matrix "point" to the elements in `$matrix`: here, the multidimensional space with only the 4th, 5th, 6th and 7th elements.
+The following elements in `diparity$subsamples$sub1` will correspond to the same "pointers" but drawn from the bootstrap replicates.
+The columns will correspond to different bootstrap replicates.
+For example:
+
+```
+ disparity$subsamples$sub1[[2]]
+      [,1] [,2] [,3] [,4]
+ [1,]   57   43   70    4
+ [2,]   43   44    4    4
+ [3,]   42   84   44    1
+ [4,]   84    7    2   10
+```
+This signifies that we have four bootstrap pseudo-replicates pointing each time to four elements in `$matrix`.
+The next element (`[[3]]`) will be the same for the eventual first rarefaction level (i.e.
+the resulting bootstrap matrix will have _m_ rows where _m_ is the number of elements for this rarefaction level).
+The next element after that (`[[4]]`) will be the same for with an other rarefaction level and so forth...
+
+### `$disparity`
+The `$disparity` element is identical to the `$subsamples` element structure (a list of list(s) containing matrices) but the matrices don't contain "pointers" to `$matrix` but the disparity result of the disparity metric applied to the "pointers".
+For example, in our first example (`$elements`) from above, if the disparity metric is of dimensions level 1, we would have:
+
+```
+ disparity$disparity$sub1$elements
+      [,1]
+ [1,]    1.82
+```
+
+This is the observed disparity (1.82) for the subsample called `sub1`.
+If the disparity metric is of dimension level 2 (say the function `range` that outputs two values), we would have:
+
+```
+ disparity$disparity$sub1$elements
+      [,1]
+ [1,]    0.82
+ [2,]    2.82
+```
+
+The following elements in the list follow the same logic as before: rows are disparity values (one row for a dimension level 1 metric, multiple for a dimensions level 2 metric) and columns are the bootstrap replicates (the bootstrap with all elements followed by the eventual rarefaction levels).
+For example for the bootstrap without rarefaction (second element of the list):
+
+```
+ disparity$disparity$sub1[[2]]
+         [,1]     [,2]     [,3]     [,4]
+[1,] 1.744668 1.777418 1.781624 1.739679 
+```
+
 <!-- ## Modularity

### 2017-11-10 inst/gitbook/06_palaeo-demo.Rmd
Updated manual character simulation
needs to do morphospace simulations

--- 2017-08-12 gitbook/05_palaeo-demo.Rmd
@@
 
-This demo aims to give quick overview of the `dispRity` package (v.0.3) for palaeobiology analyses of disparity, including disparity through time analyses.
-Please refer to my [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html) that explains the functions in more detail.
+This demo aims to give quick overview of the `dispRity` package (v.0.4) for palaeobiology analyses of disparity, including disparity through time analyses.
 
-In brief, this package allows users to easily perform **disparity-through-time** analysis.
-This type of analysis often starts by ordinating **morphometric** or **cladistic** data into a multidimensional object hereafter called the **morphospace**.
-One might be interested in studying how the occupancy of the morphospace has changed through time by measuring a summary metric of this morphospace through time, called **disparity**.
-Disparity can then be seen as a value or a distribution of values summarising the diversity of morphologies.
-
-This demo showcases a typical disparity-through-time analysis: we are going to test whether the disparity changed through time in a subset of eutherian mammals from the last 100 million years.
+This demo showcases a typical disparity-through-time analysis: we are going to test whether the disparity changed through time in a subset of eutherian mammals from the last 100 million years using a dataset from Beck and Lee (2014).
 
@@
 
-### Installing `dispRity`
-
-You can install this package easily if you are using the latest version of `R` and `devtools`.
-
-```{r, eval = FALSE}
-if(!require(devtools)) install.packages("devtools")
-install_github("TGuillerme/dispRity", ref = "release")
-library(dispRity)
-```
-
 ### The morphospace
 
-In this example, we are going to use a subset of the data from [@beckancient2014]. 
-This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
+In this example, we are going to use a subset of the data from Beck and Lee (2014).
+See the [example data](#example-data) description for more details.
+Briefly, this dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
 The ordinated matrix will represent our full morphospace, i.e. all the mammalian morphologies that ever existed through time (for this dataset).
@@
 
-```{r, eval = FALSE}
+```{r, eval = TRUE}
 if(!require(strap)) install.packages("strap")
-library(strap)
-geoscalePhylo(BeckLee_tree, cex.tip = 0.7, cex.ts = 0.6)
+strap::geoscalePhylo(BeckLee_tree, cex.tip = 0.7, cex.ts = 0.6)
 ```
@@
 
-The output object is a `dispRity` object.
-For details about this object class, please refer to the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html).
+The output object is a `dispRity` object (see more about that [here](#The-guts-of-the-dispRity-package).
 In brief, however, `dispRity` objects are lists of different elements (i.e.
@@
 ```{r}
-## Bootstrapping each time subsample 100 times
-(boot_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100))
+## Bootstrapping each time subsample 100 times (default)
+(boot_bin_morphospace <- boot.matrix(binned_morphospace))
 
 ## Getting the minimum number of rows (i.e. taxa) in the time subsamples
-min(unlist(lapply(boot_bin_morphospace$subsamples, lapply, nrow)))
+min(size.subsamples(boot_bin_morphospace))
 
@@
 Disparity can be calculated in many ways and this package allows users to come up with their own disparity metrics.
-For more details, please refer to the [`dispRity` metric vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html).
+For more details, please refer to the [`dispRity` metric section](#disparity-metrics).
 
@@
 Note that this metric comes with a caveat (not solved here) since it ignores covariances among the dimensions of the morphospace.
-We use this here because it is a standard metric used in disparity-through-time analysis.
-<!-- NC: reference? -->
+We use this here because it is a standard metric used in disparity-through-time analysis (Wills _et al._ 1994).
 
@@
 However, when looking at the rarefied results, there is no significant difference between the time bins in the Palaeogene (60-40 to 40-20 Mya), suggesting that the differences detected in the first test might just be due to the differences in number of taxa sampled (13 or 6 taxa) in each time bin.
-
-### References 
-<!-- Add these -->

### 2017-11-10 inst/gitbook/index.Rmd
Updated manual character simulation
needs to do morphospace simulations

@@
 ## What is `dispRity`?
-This is a package for measuring disparity in `R`.
+This is a modular package for measuring disparity in `R`.
 It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
@@
 
+### Modular?
+Because their exist a multitude of ways to measure disparity, each adapted to every specific question, this package uses an easy to modify modular architecture.
+In coding, each module is simply a function or a modification of a function that can be passed to the main functions of the package to tweak it to your proper needs!
+In practice, you will notice throughout this manual that some function can take other functions as arguments: the modular architecture of this package allows you to use any function for these arguments (with some restrictions explained for each specific cases).
+This will allow you to finely tune your multidimensional analysis to the needs of your specific question!
+
 ## Installing and running the package
 
-You can install this package easily if you use the latest version of `R` (> 4.0) and `devtools`.
+You can install this package easily if you use the latest version of `R` (> 3.4.0) and `devtools`.
 
@@
 library(dispRity)
-## Setting the random seed for repeatability
+## Setting a random seed for repeatability
 set.seed(123)

### 2017-11-13 inst/gitbook/03_specific-tutorials.Rmd
Updated manuals

@@
 Note that the quantiles and central tendencies can be modified via the `conc.quantiles` option.
-
-## Simulating discrete morphological data
-
-The `dispRity` package also allows users to simulate discrete morphological data matrices.
-In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
-The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
-To understand these methods please refer to the phylogenetics literature.
-<!-- NC: I wonder if you need to cite some papers here to explain how people can find out what a suitable model is etc? Or do we just assume everyone knows about making phylogenies? -->
-
-```{r}
-set.seed(3)
-## Simulating a starting tree with 15 taxa as a random coalescent tree
-my_tree <- rcoal(15)
-
-## Generating a matrix with 100 characters (85% binary and 15% three state) and
-## an equal rates model with a gamma rate distribution (0.5, 1) with no 
-## invariant characters.
-my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
-    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)
-
-## The first few lines of the matrix
-my_matrix[1:5, 1:10]
-
-## Checking the matrix properties with a quick Maximum Parsimony tree search
-check.morpho(my_matrix, my_tree)
-```
-
-Note that this example produces a tree with a great consistency index and an identical topology to the random coalescent tree!
-Nearly too good to be true...
-
-### A more detailed description
-
-`sim.morpho` is really flexible and takes many different arguments to allow users to simulate realistic matrices.
-It has three implemented models: `"ER"` for Equal Rates (the M*k* model); `"HKY"`, the molecular HKY model but transforms pyrines in 0's and pyrimidines in 1's; or the `"mixed"` model that randomly uses an `"ER"` and/or `"HKY"` for the binary characters and `"ER"` for the multistate (>2) characters.
-Both models take specific distributions for their rate or substitution models.
-These distributions should be passed to these arguments in the format of `c(sampler_function, distribution_parameters)` where the the sampler function is a the random generation function of that distribution (e.g.
-`rnorm`, `runif`, etc...) and the parameters are any parameters to be passed to this function.
-
-`check.morpho` runs a quick Maximum Parsimony tree search using the `phangorn` parsimony algorithm.
-It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g.
-the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree to determine how different they are.
-
-### Parameters for a realistic(ish) matrix
-There are many parameters that can create a ``realistic'' matrix (i.e.
-not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all end up creating "good" matrices.
-The following parameters however, seem to generate fairly ``realistic'' matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
-<!-- NC: Check that I haven't changed what you mean here where i swapped good for realistic. Realistic is a better term. Or be clearer what you mena by "good" -->
-
-```{r}
-set.seed(0)
-## tree
-my_tree <- rcoal(15)
-## matrix
-morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
-    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
-check.morpho(morpho_mat, my_tree)
-```
-
-## Simulating multidimensional spaces
-Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
-This function allows users to simulate multidimensional spaces with a certain number of properties.
-It takes as arguments the number of elements (data points) and dimensions to create the space but also permits more fine tuning in the data simulation: it is possible to give a specific distribution to each dimension, provide a correlation matrix to link the dimensions or even provide an *a priori* distribution of the variance for each distributions!
-<!-- ooh fancy! -->
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(mfrow = (c(2, 2)), bty = "n")
-## Visualising 3D spaces
-require(scatterplot3d)
-
-## A cube space
-scatterplot3d(space.maker(2500, 3, runif), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Uniform cube space")
-
-## A plane space
-scatterplot3d(space.maker(2500, 3, c(runif, runif, runif),
-              arguments = list(list(min = 0, max = 0), NULL, NULL)), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Uniform plane space")
-
-## An ellipsoid space (=a spheric space with correlation)
-cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3)
-scatterplot3d(space.maker(2500, 3, rnorm, cor.matrix = cor_matrix), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Normal ellipsoid space")
-
-## A cylindrical space with a decreasing variance per axis
-scatterplot3d(space.maker(2500, 3, c(rnorm, rnorm, runif),
-              scree = c(0.7, 0.2, 0.1)), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Normal cylindrical space")
-
-## Resetting the graphic parameters
-par(op)
-```

### 2017-11-13 inst/gitbook/04_making-up-data.Rmd
Updated manuals

@@
 They differ from missing characters `?` in their nature by being inapplicable rather than unknown.
-For example, considering a binary character defined as "color of the tail" with the following states "blue" and "red"; on a taxon with no tail, the character should be coded as inapplicable ("`-`") since the state of the character "color of tail" is *known*: it's neither "blue" or "red", it's just not there!
+For example, considering a binary character defined as "colour of the tail" with the following states "blue" and "red"; on a taxa with no tail, the character should be coded as inapplicable ("`-`") since the state of the character "colour of tail" is *known*: it's neither "blue" or "red", it's just not there!
 It contrasts with coding it as missing ("`?`" - also called as ambiguous) where the state is *unknown*, for example, the taxon of interest is a fossil where the tail has no colour preserved or is not present at all due to bad conservation!
@@
 ## The first few lines of the resulting matrix
-my_matrix_NA[1:10, 1:10]
+my_matrix_NA[1:10, 90:100]
 ```
@@
 There are many parameters that can create a "realistic" matrix (i.e.
-not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all end up creating "good" matrices.
+not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all parameters combination end up creating "good" matrices.
 The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
@@
 
+<!-- ## Simulating continuous morphological data -->
+<!-- TG: Add tutorial for evolving "morpho" shapes -->
+
 ## Simulating multidimensional spaces
@@
 This function allows users to simulate multidimensional spaces with a certain number of properties.
-It takes as arguments the number of elements (data points) and dimensions to create the space but also permits more fine tuning in the data simulation: it is possible to give a specific distribution to each dimension, provide a correlation matrix to link the dimensions or even provide an *a priori* distribution of the variance for each distributions!
-<!-- ooh fancy! -->
+For example, it is possible to design a multidimensional space with a specific distribution on each axis, a correlation between the axes and a specific cumulative variance per axis.
+This can be useful for creating ordinated spaces for null hypothesis, for example if you're using the function `null.test` [@diaz2016global].
 
-[@diaz2016global]
+This function takes as arguments the number of elements (data points - `elements` argument) and dimensions (`dimensions` argument) to create the space and the distribution functions to be used for each axis.
+The distributions are passed through the `distribution` argument as... modular functions!
+You can either pass a single distribution function for all the axes (for example `distribution = runif` for all the axis being uniform) or a specific distribution function for each specific axis (for example `distribution = c(runif, rnorm, rgamma))` for the first axis being uniform, the second normal and the third gamma).
+You can of course use your very own functions or use the ones implemented in `dispRity` for more complex ones (see below).
+Specific optional arguments for each of these distributions can be passed as a list via the `arguments` argument.
+
+Furthermore, it is possible to add a correlation matrix to add a correlation between the axis via the `cor.matrix` argument or even a vector of proportion of variance to be bear by each axis via the `scree` argument to simulate realistic ordinated spaces.
+
+Here is a simple two dimensional example:
+
+```{r, fig.width=8, fig.height=8, eval=TRUE}
+## Graphical options
+op <- par(bty = "n")
+
+## A square space
+square_space <- space.maker(100, 2, runif)
+
+## The resulting 2D matrix
+head(square_space)
+
+## Visualising the space
+plot(square_space, pch = 20, xlab = "", ylab = "", main = "Uniform 2D space")
+```
+
+Of course, more complex spaces can be created by changing the distributions, their arguments or adding a correlation matrix or a cumulative variance vector:
+
+```{r}
+## A plane space: uniform with one dimensions equal to 0
+plane_space <- space.maker(2500, 3, c(runif, runif, runif),
+                           arguments = list(list(min = 0, max = 0), NULL, NULL))
+
+## Correlation matrix for a 3D space
+(cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3))
+
+## An ellipsoid space (normal space with correlation)
+ellipse_space <- space.maker(2500, 3, rnorm, cor.matrix = cor_matrix)
+
+## A cylindrical space with decreasing axes variance
+cylindrical_space <- space.maker(2500, 3, c(rnorm, rnorm, runif),
+                                 scree = c(0.7, 0.2, 0.1))
+```
+
+### Personalised dimensions distributions
+
+Following the modular architecture of the package, it is of course possible to pass home made distribution functions to the `distribution` argument.
+For example, the `random.circle` function is a personalised one implemented in `dispRity`.
+This function allows to create circles based on basic trigonometry allowing to axis to covary to produce circle coordinates.
+By default, this function generates two sets of coordinates with a `distribution` argument and a minimum and maximum boundary (`inner` and `outer` respectively) to create nice sharp edges to the circle.
+The maximum boundary is equivalent to the radius of the circle (it removes coordinates beyond the circle radius) and the minimum is equivalent to the radius of a smaller circle with no data (it removes coordinates below this inner circle radius).
+
+```{r, fig.width=8, fig.height=8, eval=TRUE}
+## Graphical options
+op <- par(bty = "n")
+
+## Generating coordinates for a normal circle with a upper boundary of 1
+circle <- random.circle(1000, rnorm, inner = 0, outer = 1)
+
+## Plotting the circle
+plot(circle, xlab = "x", ylab = "y", main = "A normal circle")
+
+## Creating doughnut space (a spherical space with a hole)
+doughnut_space <- space.maker(5000, 3, c(rnorm, random.circle),
+     arguments = list(list(mean = 0), list(runif, inner = 0.5, outer = 1)))
+```
+
+### Visualising the space
+
+I suggest using the excellent `scatterplot3d` package to play around and visualise the simulated spaces:
 
@@
 
-## A cube space
-scatterplot3d(space.maker(2500, 3, runif), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Uniform cube space")
+## The plane space
+scatterplot3d(plane_space, pch = 20, xlab = "", ylab = "", zlab = "",
+              xlim = c(-0.5, 0.5), main = "Plane space")
 
-## A plane space
-scatterplot3d(space.maker(2500, 3, c(runif, runif, runif),
-              arguments = list(list(min = 0, max = 0), NULL, NULL)), pch = 20,
-              xlab = "", ylab = "", zlab = "",
-              main = "Uniform plane space")
-
-## An ellipsoid space (=a spheric space with correlation)
-cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3)
-scatterplot3d(space.maker(2500, 3, rnorm, cor.matrix = cor_matrix), pch = 20,
-              xlab = "", ylab = "", zlab = "",
+## The ellipsoid space
+scatterplot3d(ellipse_space, pch = 20, xlab = "", ylab = "", zlab = "",
               main = "Normal ellipsoid space")
@@
 ## A cylindrical space with a decreasing variance per axis
-scatterplot3d(space.maker(2500, 3, c(rnorm, rnorm, runif),
-              scree = c(0.7, 0.2, 0.1)), pch = 20,
-              xlab = "", ylab = "", zlab = "",
+scatterplot3d(cylindrical_space, pch = 20, xlab = "", ylab = "", zlab = "",
               main = "Normal cylindrical space")
+## Axes have different orders of magnitude
 
-## Resetting the graphic parameters
+## Plotting the doughnut space
+scatterplot3d(doughnut_space[,c(2,1,3)], pch = 20, xlab = "", ylab = "",
+              zlab = "", main = "Doughnut space")
 par(op)
 ```
+
+### Generating realistic spaces
+
+It is possible to generate "realistic" spaces by simply extracting the parameters of an existing space and scaling it up to the simulated space.
+For example, we can extract the parameters of the `BeckLee_mat50` ordinated space and simulate a similar space.
+
+```{r, fig.width=8, fig.height=8, eval=TRUE}
+## Loading the data
+data(BeckLee_mat50)
+
+## Number of dimensions
+obs_dim <- ncol(BeckLee_mat50)
+
+## Observed correlation between the dimensions
+obs_correlations <- cor(BeckLee_mat50)
+
+## Observed mean and standard deviation per axis
+obs_mu_sd_axis <- mapply(function(x,y) list("mean" = x, "sd" = y),
+                         as.list(apply(BeckLee_mat50, 2, mean)),
+                         as.list(apply(BeckLee_mat50, 2, sd)), SIMPLIFY = FALSE)
+
+## Observed overall mean and standard deviation
+obs_mu_sd_glob <- list("mean" = mean(BeckLee_mat50), "sd" = sd(BeckLee_mat50))
+
+## Scaled observed variance per axis (scree plot)
+obs_scree <- variances(BeckLee_mat50)/sum(variances(BeckLee_mat50))
+
+## Generating our simulated space
+simulated_space <- space.maker(1000, dimensions = obs_dimensions, 
+                               distribution = obs_distributions,
+                               arguments = obs_arguments,
+                               cor.matrix = obs_correlations)
+
+## Visualising the fit of our data in the space (in the two first dimensions)
+plot(simulated_space[,1:2], xlab = "PC1", ylab = "PC2")
+points(BeckLee_mat50[,1:2], col = "red", pch = 20)
+legend("topleft", legend = c("observed", "simulated"),
+        pch = c(20,21), col = c("red", "black"))
+```
+
+It is now possible to simulate a space using these observed arguments to test several hypothesis:
+
+ * Is the space uniform or normal?
+ * If the space is normal, is the mean and variance global or specific for each axis?
+
+```{r, fig.width=8, fig.height=8, eval=TRUE}
+## Measuring disparity as the sum of variance
+observed_disp <- dispRity(BeckLee_mat50, metric = c(median, centroids))
+
+## Is the space uniform?
+test_unif <- null.test(observed_disp, null.distrib = runif)
+
+## Is the space normal with a mean of 0 and a sd of 1?
+test_norm1 <- null.test(observed_disp, null.distrib = rnorm)
+
+## Is the space normal with the observed mean and sd and cumulative variance
+test_norm2 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
+                        null.args = rep(list(obs_mu_sd_glob), obs_dim),
+                        null.scree = obs_scree)
+
+## Is the space multiple normal with multiple means and sds and a correlation?
+test_norm3 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
+                        null.args = obs_mu_sd_axis, null.cor = obs_correlations)
+
+## Graphical options
+op <- par(mfrow = (c(2, 2)), bty = "n")
+## Plotting the results
+plot(test_unif, main = "Uniform (0,1)")
+plot(test_norm1, main = "Normal (0,1)")
+plot(test_norm2, main = paste0("Normal (", round(obs_mu_sd_glob[[1]], digit = 3),
+                              ",", round(obs_mu_sd_glob[[2]], digit = 3), ")"))
+plot(test_norm4, main = "Normal (variable + correlation)")
+
+```
+
+If we measure disparity as the median distance from the morphospace centroid, we can explain the distribution of the data as normal with the variable observed mean and standard deviation and with a correlation between the dimensions.

### 2017-11-13 paper/response_to_reviewers.tex

@@
 Also, I've uploaded two versions of the revised manuscript: both have the exact same content but one has all the changes and additions to the main text highlighted in yellow to help with reviewing.
+
+% ~~~~~~~~~~~~~~~~~~~~~~~~
+%
+% REVIEWER 1
+%
+% ~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 
+% ~~~~~~~~~~~~~~~~~~~~~~~~
+%
+% REVIEWER 2
+%
+% ~~~~~~~~~~~~~~~~~~~~~~~~
+
 \section{Reviewer 2}
@@
 
-% Add simulation description
+Reviewer 4 (minor comment \ref{remove_simulations}) suggested completely removing this section which conflicts directly with this suggestion.
+The simulation protocols implemented here are already developed (and tested) in details in previous publications with new novel features attached to future publication.
+For example the discrete data simulation protocol is described and detailed in Guillerme \& Cooper 2016 \textit{Mol Phyl Evol}, O’Reilly et al.
+2016 \textit{Biol Let}, O'Reilly et al.
+2017 \textit{Palaeontology} and Guillerme \& Brazeau \textit{in review}.
+The implemented ordinated space simulations are theoretically described in Diaz et al.
+2016 \textit{Nature} and will be used along with additional simulation features not implemented yet (including shape evolution) as part of a project currently in progress.
+I am therefore inclined to follow reviewer's 4 suggestion rather than this one.
+% TODO change this sentence below:
+However, encouraged by the present reviewer's enthusiastic highlighting of the importance of these features, I've developed the description of these simulation in the package manual and should Reviewer 2 and the editor request it, I am willing to add this description directly to some supplementary materials of the current package description.
+
+% TODO add pdf of simulations description
 
@@
 
-l. @@@: \textit{Note that this function also allows to work on only a subset of dimensions via the \texttt{dimensions} argument (e.g. if only the $n$ first dimensions must be considered).}
+l. @@@: \textit{Note that this function also allows to work on only a subset of dimensions via the} \texttt{dimensions} \textit{argument (e.g. if only the $n$ first dimensions must be considered).}
 
@@
 \end{enumerate}
+
+% ~~~~~~~~~~~~~~~~~~~~~~~~
+%
+% REVIEWER 3
+%
+% ~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 
+% ~~~~~~~~~~~~~~~~~~~~~~~~
+%
+% REVIEWER 4
+%
+% ~~~~~~~~~~~~~~~~~~~~~~~~
+
 \section{Reviewer 4 (Gavin Simpson)}
@@
 The later is somewhat tangential to the main thrust of the paper and you cover this in the vignette/bookdown resources.}}
+\label{remove_simulations}
 

### 2017-11-13 inst/gitbook/04_making-up-data.Rmd

@@
 ## Generating our simulated space
-simulated_space <- space.maker(1000, dimensions = obs_dimensions, 
-                               distribution = obs_distributions,
-                               arguments = obs_arguments,
+simulated_space <- space.maker(1000, dimensions = obs_dim, 
+                               distribution = rep(list(rnorm), obs_dim),
+                               arguments = obs_mu_sd_axis,
                                cor.matrix = obs_correlations)
@@
                               ",", round(obs_mu_sd_glob[[2]], digit = 3), ")"))
-plot(test_norm4, main = "Normal (variable + correlation)")
+plot(test_norm3, main = "Normal (variable + correlation)")
 

### 2017-11-13 paper/dispRity_paper_highlighted.tex

@@
 Although many software packages exist for multidimensional analysis \citep[e.g.][]{bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric}, package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric and then enforce this on users by allowing little to no flexibility.
-For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of morphometric data, the disparity metric as the Procrustes variance, and uses permutation tests to test hypotheses \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
-%Reviewer 1 minor: Explain what the Procrustes variance is. And when is it appropriate and when is not appropriate? Be more specific.
-This is ideal in some cases, but in others may not be the most appropriate analysis.
+For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric
+\hl{as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance)}
+, and uses permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
+\hl{This is ideal for testing volume based hypothesis (e.g.
+does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g.
+does groups A and B occupy the same location?).}
 This can lead to inappropriate analyses by users confined by the existing software implementations. %Oh no!

### 2017-11-15 inst/gitbook/03_specific-tutorials.Rmd
Updated manual

@@
  *  `deltran` where the data chosen along the branch is always the one of the ancestor
- *  `punctuated` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
- *  `gradual` where the data chosen along the branch is either the descendant or the ancestor depending on branch length
+ *  `random` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
+ *  `proximity` where the data chosen along the branch is either the descendant or the ancestor depending on branch length
+ *  `punctuated` where the data chosen along the branch is both the descendant and the ancestor with an even probability
+ *  `gradual` where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length
+
+> Note that the four first models are "fixed": the selected data is always either the one of the descendant or the ancestor.
+The two last models are "probabilistic": the data from both the descendant and the ancestor is used with an associate probability.
+These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants.
 
 ```{r, eval=TRUE}
-## Generating four time slices every 40 million years under a model of gradual evolution
+## Generating four time slices every 40 million years under a model of proximity evolution
 time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
+    method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
     FADLAD = BeckLee_ages)
@@
 time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree,
-    method = "continuous", model = "gradual", time = 4, FADLAD = BeckLee_ages)
+    method = "continuous", model = "proximity", time = 4, FADLAD = BeckLee_ages)
 ```
@@
 time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-                               method = "continuous", model = "gradual", 
+                               method = "continuous", model = "proximity", 
                                time = c(120, 80, 40, 0),
@@
 time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
+    method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
     FADLAD = BeckLee_ages)

### 2017-11-15 inst/gitbook/05_guts.Rmd
Updated manual

@@
 
- * `$call$subsamples`: a vector of `character` with information on the subsamples type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"punctuated"`, `"gradual"`).
+ * `$call$subsamples`: a vector of `character` with information on the subsamples type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"punctuated"`, `"gradual"`).
 This element generated only once via `time.subsamples()` and `custom.subsamples()`.

### 2017-11-15 inst/gitbook/01_glossary.Rmd
Updated book (subsamples -> subsets)

@@
 
--   **Subsamples**. Subsamples of the multidimensional space.
-    A subsample (or subsamples) contains the same number of dimensions as the space but may contain a smaller subset of elements.
-    For example, if our space is composed of birds and mammals (the elements) and 50 principal components of variation (the dimensions), we can create two subsamples containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
+-   **Subsets**. Subsets of the multidimensional space.
+    A subset (or subsets) contains the same number of dimensions as the space but may contain a smaller subset of elements.
+    For example, if our space is composed of birds and mammals (the elements) and 50 principal components of variation (the dimensions), we can create two subsets containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.

### 2017-11-15 inst/gitbook/02_getting-started.Rmd
Updated book (subsamples -> subsets)

@@
 Additionally you can give the function a `geomorph.data.frame` object.
-If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsamples)!
+If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsets)!
 
@@
 Should you rarefy the data?
-See [`time.subsamples`](#time-slicing), [`custom.subsamples`](#customised-subsamples), [`boot.matrix`](#bootstraps-and-rarefactions) and [`dispRity.metric`](#disparity-metrics) for more details of the defaults used in each of these functions.
+See [`time.subsets`](#time-slicing), [`custom.subsets`](#customised-subsets), [`boot.matrix`](#bootstraps-and-rarefactions) and [`dispRity.metric`](#disparity-metrics) for more details of the defaults used in each of these functions.
 Note that any of these default arguments can be changed within the `disparity.through.time` or `disparity.per.group` functions.
@@
 an age for the root), you can simply do\\ `my_tree$root.time <- my_age`.
-  * The required number of time subsamples (here `time = 3`)
+  * The required number of time subsets (here `time = 3`)
   * Your favourite disparity metric (here the sum of variances)
@@
 
-We asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
+We asked for three subsets (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
 

### 2017-11-15 inst/gitbook/03_specific-tutorials.Rmd
Updated book (subsamples -> subsets)

@@
 
-The function `time.subsamples` allows users to divide the matrix into different time subsamples or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
-Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.
+The function `time.subsets` allows users to divide the matrix into different time subsets or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
+Each subset generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.
 
-Two types of time subsamples can be performed by using the `method` option:
+Two types of time subsets can be performed by using the `method` option:
 
- *  Discrete time subsamples (or time-binning) using `method = discrete`
- *  Continuous time subsamples (or time-slicing) using `method = continuous`
+ *  Discrete time subsets (or time-binning) using `method = discrete`
+ *  Continuous time subsets (or time-slicing) using `method = continuous`
 
@@
 ## Generating three time bins containing the taxa present every 40 Ma
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
                 time = c(120, 80, 40, 0))
@@
 ## Automatically generate three equal length bins:
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
                 time = 3)
@@
 ## Generating time bins including taxa that might span between them
-time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
                 time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
@@
 ## Generating four time slices every 40 million years under a model of proximity evolution
-time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
+time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
     method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
@@
 ## Generating four time slices automatically
-time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree,
+time.subsets(data = BeckLee_mat99, tree = BeckLee_tree,
     method = "continuous", model = "proximity", time = 4, FADLAD = BeckLee_ages)
@@
 
-## Customised subsamples
+## Customised subsets
 
-Another way of separating elements into different categories is to use customised subsamples as briefly explained [above](#disparity-among-groups).
+Another way of separating elements into different categories is to use customised subsets as briefly explained [above](#disparity-among-groups).
 This function simply takes the list of elements to put in each group (whether they are the actual element names or their position in the matrix).
@@
 ## Separating the dataset into two different groups
-custom.subsamples(BeckLee_mat50, group = mammal_groups)
+custom.subsets(BeckLee_mat50, group = mammal_groups)
 ```
@@
 The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsample to a minimum of three elements).
+remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subset to a minimum of three elements).
 It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
@@
 
-Of course, one could directly supply the subsamples generated above (using `time.subsamples` or `custom.subsamples`) to this function.
+Of course, one could directly supply the subsets generated above (using `time.subsets` or `custom.subsets`) to this function.
 
 ```{r, eval=TRUE}
-## Creating subsamples of crown and stem mammals
-crown_stem <- custom.subsamples(BeckLee_mat50,
+## Creating subsets of crown and stem mammals
+crown_stem <- custom.subsets(BeckLee_mat50,
                                 group = list("crown" = c(16, 19:41, 45:50), 
@@
 
-## Creating time slice subsamples
-time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
+## Creating time slice subsets
+time_slices <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
                                method = "continuous", model = "proximity", 
@@
 
-## Bootstrapping the time slice subsamples
+## Bootstrapping the time slice subsets
 boot.matrix(time_slices, bootstraps = 100)
@@
 ```
-> Because there is only one subsample (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with rounding).
+> Because there is only one subset (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with rounding).
 
@@
 ## Example data from previous sections
-crown_stem <- custom.subsamples(BeckLee_mat50,
+crown_stem <- custom.subsets(BeckLee_mat50,
                                 group = list("crown" = c(16, 19:41, 45:50), 
@@
 
-## Creating time slice subsamples
-time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
+## Creating time slice subsets
+time_slices <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
     method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
     FADLAD = BeckLee_ages)
-## Bootstrapping the time slice subsamples
+## Bootstrapping the time slice subsets
 boot_time_slices <- boot.matrix(time_slices, bootstraps = 100)
@@
 
-## Creating time bin subsamples
-time_bins <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
+## Creating time bin subsets
+time_bins <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
     method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages,
     inc.nodes = TRUE)
-## Bootstrapping the time bin subsamples
+## Bootstrapping the time bin subsets
 boot_time_bins <- boot.matrix(time_bins, bootstraps = 100)
@@
 
-Information about the number of elements in each subsample and the observed (i.e. non-bootstrapped) disparity are also calculated.
+Information about the number of elements in each subset and the observed (i.e. non-bootstrapped) disparity are also calculated.
 This is specifically handy when rarefying the data for example:
@@
 
-> This argument can be left empty. In this case, the algorithm will automatically detect the type of subsamples from the `dispRity` object and plot accordingly.
+> This argument can be left empty. In this case, the algorithm will automatically detect the type of subsets from the `dispRity` object and plot accordingly.
 
-It is also possible to display the number of elements in each subsample (as a horizontal dotted line) using the option `elements = TRUE`.
+It is also possible to display the number of elements in each subset (as a horizontal dotted line) using the option `elements = TRUE`.
 Additionally, when the data is rarefied, one can indicate which level of rarefaction to display (i.e. only display the results for a certain number of elements) by using the `rarefaction` argument.
@@
     cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
-    ylab = c("Disparity", "Diversity"), time.subsamples = FALSE,
+    ylab = c("Disparity", "Diversity"), time.subsets = FALSE,
     xlab = "Time (in in units from past to present)", observed = TRUE,
@@
 
- * `pairwise` (default): to compare each subsample in a pairwise manner
- * `referential`: to compare each subsample to the first subsample
- * `sequential`: to compare each subsample to the following subsample
- * `all`: to compare all the subsamples together (like in analysis of variance)
+ * `pairwise` (default): to compare each subset in a pairwise manner
+ * `referential`: to compare each subset to the first subset
+ * `sequential`: to compare each subset to the following subset
+ * `all`: to compare all the subsets together (like in analysis of variance)
 
-It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsample names to create personalised tests.
+It is also possible to input a list of pairs of `numeric` values or `characters` matching the subset names to create personalised tests.
 Some other tests implemented in `dispRity` such as the `dispRity::null.test` have a specific way they are applied to the data and therefore ignore the `comparisons` argument. 

### 2017-11-15 inst/gitbook/05_guts.Rmd
Updated book (subsamples -> subsets)

--- 2017-11-10 inst/gitbook/05_guts.Rmd
@@
 ## (15 elements) from the second group (80 Mya)
-str(matrix.dispRity(disparity, subsamples = 1, bootstrap = 3, rarefaction = 2))
+str(matrix.dispRity(disparity, subsets = 1, bootstrap = 3, rarefaction = 2))
 ```
 
-#### `get.subsamples.dispRity`
-This function creates a dispRity object that contains only elements from one specific subsamples.
+#### `get.subsets.dispRity`
+This function creates a dispRity object that contains only elements from one specific subsets.
 
@@
 ## Extracting all the data for the crown mammals
-(crown_mammals <- get.subsamples.dispRity(disp_crown_stemBS, "Group.crown"))
+(crown_mammals <- get.subsets.dispRity(disp_crown_stemBS, "Group.crown"))
 
-## The object keeps the properties of the parent object but is composed of only one subsamples
-length(crown_mammals$subsamples)
+## The object keeps the properties of the parent object but is composed of only one subsets
+length(crown_mammals$subsets)
 ```
@@
 ## Extracting the disparity from the bootstrapped values from the
-## 10th rarefaction level from the second subsamples (80 Mya)
-extract.dispRity(disparity, observed = FALSE, subsamples = 2, rarefaction = 10)
+## 10th rarefaction level from the second subsets (80 Mya)
+extract.dispRity(disparity, observed = FALSE, subsets = 2, rarefaction = 10)
 ```
@@
 ```{r, eval=FALSE}
-## Getting the disparity values of the time subsamples
+## Getting the disparity values of the time subsets
 head(summary(disparity))
@@
 #### `sort.dispRity`
-This is the S3 method of `sort` for sorting the subsamples alphabetically (default) or following a specific pattern.
+This is the S3 method of `sort` for sorting the subsets alphabetically (default) or following a specific pattern.
 
 ```{r, eval=FALSE}
-## Sorting the disparity subsamples in inverse alphabetic order
+## Sorting the disparity subsets in inverse alphabetic order
 head(summary(sort(disparity, decreasing = TRUE)))
@@
  * `$call`: an object of class `list` containing information on the `dispRity` object content.
- * `$subsamples`: an object of class `list` containing the subsamples of the multidimensional space.
+ * `$subsets`: an object of class `list` containing the subsets of the multidimensional space.
  * `$disparity`: an object of class `list` containing the disparity values.
@@
 In fact, it is composed of one unique instance of a matrix (the multidimensional space) upon which the metric function is called via "pointers" to only a certain number of elements and/or dimensions of this matrix.
-This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint:  at any time, only one matrix is present in the `R` environment rather than multiple copies of it for each subsample.
+This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint:  at any time, only one matrix is present in the `R` environment rather than multiple copies of it for each subset.
 
@@
 
- * `$call$subsamples`: a vector of `character` with information on the subsamples type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"punctuated"`, `"gradual"`).
-This element generated only once via `time.subsamples()` and `custom.subsamples()`.
+ * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"punctuated"`, `"gradual"`).
+This element generated only once via `time.subsets()` and `custom.subsets()`.
  * `$call$dimensions`: either a single `numeric` value indicating how many dimensions to use or a vector of `numeric` values indicating which specific dimensions to use.
@@
 
-### `$subsamples`
-This element contain the eventual subsamples of the multidimensional space.
-It is a `list` of subsample names.
-Each subsample name is in turn a `list` of at least one element called `elements` which is in turn a `matrix`.
-This `elements` matrix is the raw (observed) elements in the subsamples.
-The `elements` matrix is composed of `numeric` values in one column and _n_ rows (the number of elements in the subsample).
+### `$subsets`
+This element contain the eventual subsets of the multidimensional space.
+It is a `list` of subset names.
+Each subset name is in turn a `list` of at least one element called `elements` which is in turn a `matrix`.
+This `elements` matrix is the raw (observed) elements in the subsets.
+The `elements` matrix is composed of `numeric` values in one column and _n_ rows (the number of elements in the subset).
 Each of these values are a "pointer" (`C` inspired) to the element of the `$matrix`.
-For example, lets assume a `dispRity` object called `disparity`, composed of at least one subsamples called `sub1`:
+For example, lets assume a `dispRity` object called `disparity`, composed of at least one subsets called `sub1`:
 
 ```
- disparity$subsamples$sub1$elements
+ disparity$subsets$sub1$elements
       [,1]
@@
 The values in the matrix "point" to the elements in `$matrix`: here, the multidimensional space with only the 4th, 5th, 6th and 7th elements.
-The following elements in `diparity$subsamples$sub1` will correspond to the same "pointers" but drawn from the bootstrap replicates.
+The following elements in `diparity$subsets$sub1` will correspond to the same "pointers" but drawn from the bootstrap replicates.
 The columns will correspond to different bootstrap replicates.
@@
 ```
- disparity$subsamples$sub1[[2]]
+ disparity$subsets$sub1[[2]]
       [,1] [,2] [,3] [,4]
@@
 ### `$disparity`
-The `$disparity` element is identical to the `$subsamples` element structure (a list of list(s) containing matrices) but the matrices don't contain "pointers" to `$matrix` but the disparity result of the disparity metric applied to the "pointers".
+The `$disparity` element is identical to the `$subsets` element structure (a list of list(s) containing matrices) but the matrices don't contain "pointers" to `$matrix` but the disparity result of the disparity metric applied to the "pointers".
 For example, in our first example (`$elements`) from above, if the disparity metric is of dimensions level 1, we would have:
@@
 
-This is the observed disparity (1.82) for the subsample called `sub1`.
+This is the observed disparity (1.82) for the subset called `sub1`.
 If the disparity metric is of dimension level 2 (say the function `range` that outputs two values), we would have:

### 2017-11-15 inst/gitbook/06_palaeo-demo.Rmd
Updated book (subsamples -> subsets)

@@
 
-One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time subsamples that contain the total number of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
-Basically, the full morphospace represents the total number of morphologies across all time and will be greater than any of the time subsamples of the morphospace.
+One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time subsets that contain the total number of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
+Basically, the full morphospace represents the total number of morphologies across all time and will be greater than any of the time subsets of the morphospace.
 
-The `dispRity` package provides a `time.subsamples` function that allows users to split the morphospace into time slices (using `method = continuous`) or into time bins (using `method = discrete`).
+The `dispRity` package provides a `time.subsets` function that allows users to split the morphospace into time slices (using `method = continuous`) or into time bins (using `method = discrete`).
 In this example, we are going to split the morphospace into five equal time bins of 20 million years long from 100 million years ago to the present.
@@
 
-## Splitting the morphospace using the time.subsamples function
-(binned_morphospace <- time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree,
+## Splitting the morphospace using the time.subsets function
+(binned_morphospace <- time.subsets(data = BeckLee_mat50, tree = BeckLee_tree,
     method = "discrete", time = time_bins, inc.nodes = FALSE,
@@
 In brief, however, `dispRity` objects are lists of different elements (i.e.
-disparity results, morphospace time subsamples, morphospace attributes, etc.) that display only a summary of the object when calling the object to avoiding filling the `R` console with superfluous output.
+disparity results, morphospace time subsets, morphospace attributes, etc.) that display only a summary of the object when calling the object to avoiding filling the `R` console with superfluous output.
 
@@
 
-Once we obtain our different time subsamples, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
-The bootstrapping allows us to make each subsample more robust to outliers and the rarefaction allows us to compare subsamples with the same number of taxa to remove sampling biases (i.e.
-more taxa in one subsample than the others).
+Once we obtain our different time subsets, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
+The bootstrapping allows us to make each subset more robust to outliers and the rarefaction allows us to compare subsets with the same number of taxa to remove sampling biases (i.e.
+more taxa in one subset than the others).
 The `boot.matrix` function bootstraps the `dispRity` object and the `rarefaction` option within performs rarefaction.
@@
 ```{r}
-## Bootstrapping each time subsample 100 times (default)
+## Bootstrapping each time subset 100 times (default)
 (boot_bin_morphospace <- boot.matrix(binned_morphospace))
 
-## Getting the minimum number of rows (i.e. taxa) in the time subsamples
-min(size.subsamples(boot_bin_morphospace))
+## Getting the minimum number of rows (i.e. taxa) in the time subsets
+min(size.subsets(boot_bin_morphospace))
 
-## Bootstrapping each time subsample 100 times and rarefying them 
+## Bootstrapping each time subset 100 times and rarefying them 
 (rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
@@
 
-We can now calculate the disparity within each time subsamples along with some confidence intervals generated by the pseudoreplication step above (bootstraps/rarefaction).
+We can now calculate the disparity within each time subsets along with some confidence intervals generated by the pseudoreplication step above (bootstraps/rarefaction).
 Disparity can be calculated in many ways and this package allows users to come up with their own disparity metrics.
@@
 
-In this example, we are going to calculate the spread of the data in each time subsample by calculating disparity as the sum of the variance of each dimension of the morphospace in each time subsample using the `dispRity` function.
-Thus, in this example, disparity is defined by the multi-dimensional variance of each time subsample (i.e. the spread of the taxa within the morphospace).
+In this example, we are going to calculate the spread of the data in each time subset by calculating disparity as the sum of the variance of each dimension of the morphospace in each time subset using the `dispRity` function.
+Thus, in this example, disparity is defined by the multi-dimensional variance of each time subset (i.e. the spread of the taxa within the morphospace).
 Note that this metric comes with a caveat (not solved here) since it ignores covariances among the dimensions of the morphospace.
@@
 
-The `dispRity` function does not actually display the calculated disparity values but rather only the properties of the disparity object (size, subsamples, metric, etc.).
+The `dispRity` function does not actually display the calculated disparity values but rather only the properties of the disparity object (size, subsets, metric, etc.).
 To display the actual calculated scores, we need to summarise the disparity object using the S3 method `summary` that is applied to a `dispRity` object (see `?summary.dispRity` for more details).
@@
 what happens in bin *n+1* depends on what happened in bin *n*) and pseudoreplicated (i.e.
-each bootstrap draw creates non-independent time subsamples because they are all based on the same time subsamples), we apply a non-parametric mean comparison: the `wilcox.test`.
+each bootstrap draw creates non-independent time subsets because they are all based on the same time subsets), we apply a non-parametric mean comparison: the `wilcox.test`.
 Also, we need to apply a p-value correction (e.g. Bonferroni correction) to correct for multiple testing (see `?p.adjust` for more details).

### 2017-11-15 inst/gitbook/07_ecology-demo.Rmd
Updated book (subsamples -> subsets)

--- 2017-08-15 inst/gitbook/06_ecology-demo.Rmd
@@
 
-The first step is to create different subsamples that represent subsamples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
-Each of these subsamples will contain only the individuals of a specific species.
+The first step is to create different subsets that represent subsets of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
+Each of these subsets will contain only the individuals of a specific species.
 
@@
 ## Creating the table that contain the elements and their attributes
-petal_subsamples <- custom.subsamples(petal_space, group = list(
+petal_subsets <- custom.subsets(petal_space, group = list(
                                 "setosa" = which(species == "setosa"),
@@
 ## Visualising the dispRity object content
-petal_subsamples
+petal_subsets
 ```
 
-This created a `dispRity` object (more about that [here](#guts)) with three subsamples corresponding to each subspecies.
+This created a `dispRity` object (more about that [here](#guts)) with three subsets corresponding to each subspecies.
 
 ### Bootstrapping the data
-We can the bootstrap the subsamples to be able test the robustness of the measured disparity to outliers.
+We can the bootstrap the subsets to be able test the robustness of the measured disparity to outliers.
 We can do that using the default options of `boot.matrix` (more about that [here](#bootstraps-and-rarefactions)):
@@
 ## Bootstrapping the data
-(petal_bootstrapped <- boot.matrix(petal_subsamples))
+(petal_bootstrapped <- boot.matrix(petal_subsets))
 ```
@@
 ### Summarising the results (plot)
-Similarly to the `custom.subsamples` and `boot.matrix` function, `dispRity` displays a `dispRity` object.
+Similarly to the `custom.subsets` and `boot.matrix` function, `dispRity` displays a `dispRity` object.
 But we are definitely more interested in actually look at the calculated values.

### 2017-11-16 inst/gitbook/03_specific-tutorials.Rmd
Updated manual with new metrics

@@
 ------|------|---------------------------------------------------|--------|
-1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
+2 | `ancestral.dist` | The distance between an element and its ancestor | `dispRity` |
+2 | `centroids`<sup>1</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
 1 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
 1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-<!-- 1 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |-->
 1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
+1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
+1 | `mode.val` | The modal value | `dispRity` |
+1 | `n.ball.volume` | The hyper-spherical (*n*-ball) volume | `dispRity` |
+2 | `pairwise.dist` | The pairwise distances between elements | [`stats`](https://cran.r-project.org/web/packages/stats/index.html)`::dist` |
+2 | `radius` | The radius of each dimensions | `dispRity` |
 2 | `ranges` | The range of each dimension | `dispRity` |
+1 | `span.tree.length` | The minimal spanning tree length | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree` |
 2 | `variances` | The variance of each dimension | `dispRity` |
-2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
-1 | `mode.val` | The modal value | `dispRity` |
 
-1: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (not PCA).
+1: Note that by default, the centroid is the centroid of the elements.
+It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
 
-2: Note that by default, the centroid is the centroid of the elements.
-It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
+2: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (*not* PCA).
 
@@
 \begin{equation}
-    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
+    ancestral.dist = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}
 \end{equation}
-Where *k* is the number of dimensions, and $\lambda_i$ is the eigenvalue of each dimension.
+
+\begin{equation}
+    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
+\end{equation}
 
@@
 \end{equation}
-Where *k* is the number of dimensions.
+
+\begin{equation}
+    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
+\end{equation}
+
+\begin{equation}
+    n.ball.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R
+\end{equation}
+
+\begin{equation}
+    radius = |\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|
+\end{equation}
 
@@
 \end{equation}
-Where *k* is the number of dimensions.
 
@@
 \end{equation}
-Where *k* is the number of dimensions, and $\sigma^{2}$ is their variance.
 
 \begin{equation}
-    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
+    span.tree.length = \sum(\mathrm{branch\ length})
 \end{equation}
-Where *n* is each element in the ordinated space, *k* is the number of dimensions, and $Centroid_{k}$ is their mean (or can be set to another value).
+
+Where *k* is the number of dimensions,
+*n* the number of elements,
+$\Gamma$ is the Gamma distribution,
+$\lambda_i$ is the eigenvalue of each dimensions,
+$\sigma^{2}$ is their variance and
+$Centroid_{k}$ is their mean,
+$Ancestor_{n}$ is the coordinates of the ancestor of element $n$,
+$f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.),
+*R* is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid).
 
@@
 #### Volumes and surface metrics
-The functions `ellipse.volume`, `convhull.surface` and `convhull.volume`  all measure the surface or the volume of the ordinated space occupied:
+The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `n.ball.volume`  all measure the surface or the volume of the ordinated space occupied:
 
@@
 summary(dispRity(dummy_space, metric = convhull.volume))
+
+## Calculating the convex hull volume
+summary(dispRity(dummy_space, metric = n.ball.volume))
 ```
@@
 
-#### Ranges, variances and diagonal
+#### Ranges, variances, radius, pairwise distance, modal value and diagonal
 
-The functions `ranges`, `variances` and `diagonal` all measure properties of the ordinated space based on its dimensional properties (they are also less affected by the "curse of dimensionality"):
+The functions `ranges`, `variances` `radius`, `pairwise.dist`, `mode.val` and `diagonal` all measure properties of the ordinated space based on its dimensional properties (they are also less affected by the "curse of dimensionality"):
 
-`ranges` and `variances` both work on the same principle and measure the range/variance of each dimension:
+`ranges`, `variances` and `radius` work on the same principle and measure the range/variance/radius of each dimension:
 
@@
 summary(dispRity(dummy_space, metric = c(prod, variances)))
+
+## Calculating the radius of each dimension in the ordinated space
+radius(dummy_space)
+
+## By default the radius is the maximum distance from the centre of
+## the dimension. It can however be changed to any function:
+radius(dummy_space, type = min)
+radius(dummy_space, type = mean)
+
+## Calculating disparity as the mean average radius
+summary(dispRity(dummy_space, metric = c(mean, radius), type = mean))
 ```
 
-The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest Euclidean distance in our five dimensional space):
+The pairwise distances uses the function `stats::dist` and can take the normal `dist` options:
+
+```{r}
+## The average pairwise euclidean distance
+summary(dispRity(dummy_space, metric = c(mean, pairwise.dist)))
+
+## The distribution of the Manhattan distances
+summary(dispRity(dummy_space, metric = pairwise.dist, method = "manhattan"))
+```
+
+The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest Euclidean distance in our five dimensional space).
+The `mode.val` function measures the modal value of the matrix:
 
@@
 summary(dispRity(dummy_space, metric = diagonal))
+
+## Calculating the modal value of the matrix
+summary(dispRity(dummy_space, metric = mode.val))
 ```
@@
 
-#### Centroids metric
+#### Centroids and ancestral distance metrics
 
@@
 ```
+
+The `ancestral.dist` metric works on a similar principle as the `centroids` function but changes the centroid to be the coordinates of each element's ancestor.
+Therefore this functions needs a tree and node coordinates as additional arguments:
+
+```{r}
+## A generating a random tree with node labels
+tree <- rtree(5) ; tree$node.label <- paste0("n", 1:4)
+## Adding the tip and node names to the matrix
+dummy_space2 <- dummy_space[-1,]
+rownames(dummy_space2) <- c(tree$tip.label, tree$node.label)
+
+## Calculating all the ancestral nodes
+all_anc_centroids <- nodes.coordinates(dummy_space2, tree, full = TRUE)
+
+## Calculating the distances from the ancestral nodes
+ancestral_dist <- dispRity(dummy_space2, metric = ancestral.dist,
+                           nodes.coords = all_anc_centroids)
+
+## Calculating disparity as the sum of the distances from all the ancestral nodes
+summary(dispRity(ancestral_dist, metric = sum))
+```
+
+#### Minimal spanning tree length
+
+The `span.tree.length` uses the `vegan::spantree` function to heuristically calculate the minimum spanning tree (the shortest multidimensional tree connecting each elements) and calculates its length as the sum of every branch lengths.
+
+```{r}
+## The length of the minimal spanning tree
+summary(dispRity(dummy_space, metric = span.tree.length))
+```
+
+Note that because the solution is heuristic, this metric can take a long time to compute for big matrices.
 
@@
 Note that the quantiles and central tendencies can be modified via the `conc.quantiles` option.
+
+## Disparity from other matrices
+
+In the example so far, disparity was measured from an ordinated multidimensional space (i.e. a PCO of the distances between taxa based on discrete morphological characters).
+This is a common approach in palaeobiology, morphometrics or ecology but ordinated matrices are not mandatory for the `dispRity` package!
+It is totally possible to perform the same analysis detailed above using other types of matrices as long as your elements are rows in your matrix.
+
+For example, we can use the data set `eurodist`, an `R` inbuilt dataset that contains the distances (in km) between European cities.
+We can check for example, if Northern European cities are closer to each other than Southern ones:
+
+```{r}
+## Making the eurodist data set into a matrix (rather than "dist" object)
+eurodist <- as.matrix(eurodist)
+eurodist[1:5, 1:5]
+
+## The two groups of cities
+Northern <- c("Brussels", "Calais", "Cherbourg", "Cologne", "Copenhagen",
+              "Hamburg", "Hook of Holland", "Paris", "Stockholm")
+Southern <- c("Athens", "Barcelona", "Geneva", "Gibraltar", "Lisbon", "Lyons",
+              "Madrid", "Marseilles", "Milan", "Munich", "Rome", "Vienna")
+
+## Creating the subset dispRity object
+eurodist_subsets <- custom.subsets(eurodist, group = list("Northern" = Northern,
+                                                        "Southern" = Southern))
+
+## Bootstrapping and rarefying to 9 elements (the number of Northern cities)
+eurodist_bs <- boot.matrix(eurodist_subsets, rarefaction = 9)
+
+## Measuring disparity as the median distance from group's centroid
+euro_disp <- dispRity(eurodist_bs, metric = c(median, centroids))
+
+## Testing the differences using a simple wilcox.test
+euro_diff <- test.dispRity(euro_disp, test = wilcox.test)
+euro_diff_rar <- test.dispRity(euro_disp, test = wilcox.test, rarefaction = 9)
+```
+
+We can compare this approach to an ordination one:
+
+```{r}
+## Ordinating the eurodist matrix
+euro_ord <- cmdscale(eurodist, k = nrow(eurodist) - 2)
+
+## Calculating disparity on the bootstrapped and rarefied subset data
+euro_ord_disp <- dispRity(boot.matrix(custom.subsets(euro_ord, group =
+        list("Northern" = Northern, "Southern" = Southern)), rarefaction = 9),
+        metric = c(median, centroids))
+
+## Testing the differences using a simple wilcox.test
+euro_ord_diff <- test.dispRity(euro_ord_disp, test = wilcox.test)
+euro_ord_diff_rar <- test.dispRity(euro_ord_disp, test = wilcox.test, rarefaction = 9)
+```
+
+And visualise the differences:
+
+```{r, fig.width=12, fig.height=12}
+## Plotting the differences
+par(mfrow = c(2,2), bty = "n")
+## Plotting the normal disparity
+plot(euro_disp, main = "Distance differences")
+## Adding the p-value
+text(1.5, 4000, paste0("p=",round(euro_diff[[2]][[1]], digit = 5)))
+## Plotting the rarefied disparity
+plot(euro_disp, rarefaction = 9, main = "Distance differences (rarefied)")
+## Adding the p-value
+text(1.5, 4000, paste0("p=",round(euro_diff_rar[[2]][[1]], digit = 5)))
+
+## Plotting the ordinated disparity
+plot(euro_ord_disp, main = "Ordinated differences")
+## Adding the p-value
+text(1.5, 1400, paste0("p=",round(euro_ord_diff[[2]][[1]], digit = 5) ))
+## Plotting the rarefied disparity
+plot(euro_ord_disp, rarefaction = 9, main = "Ordinated differences (rarefied)")
+## Adding the p-value
+text(1.5, 1400, paste0("p=",round(euro_ord_diff_rar[[2]][[1]], digit = 5) ))
+
+```
+
+As expected, the results are pretty similar in pattern but different in terms of scale.
+The median centroids distance is expressed in km in the "Distance differences" plots and in Euclidean units of variation in the "Ordinated differences" plots.

### 2017-11-16 inst/gitbook/index.Rmd
Updated manual with new metrics

@@
 This is a package for measuring disparity in `R`.
-It allows users to summarise matrices as representations as multidimensional spaces (namely from ordinated matrices from MDS, PCA, PCO, PCoA) into a single value or distribution describing a specific aspect of this multidimensional space (the disparity).
-This manual is based on the version `0.4`.
+It allows users to summarise matrices as representations as multidimensional spaces into a single value or distribution describing a specific aspect of this multidimensional space (the disparity).
+Multidimensional spaces can be ordinated matrices from MDS, PCA, PCO, PCoA but the package is *not* restricted to any type of matrices!
+This manual is based on the version `0.4.2`.
 

### 2017-11-16 paper/response_to_reviewers.tex
Updated response

@@
 does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g.
-does groups A and B occupy the same location?).}
+does groups A and B occupy the same location?).} lines @@@
 
@@
 I am therefore inclined to follow reviewer's 4 suggestion rather than this one.
-% TODO change this sentence below:
-However, encouraged by the present reviewer's enthusiastic highlighting of the importance of these features, I've developed the description of these simulation in the package manual and should Reviewer 2 and the editor request it, I am willing to add this description directly to some supplementary materials of the current package description.
 
-% TODO add pdf of simulations description
+However, I have heavily edited the online manual to highlight the simulation functionalities implemented in the \texttt{dispRity} package: \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/making-stuff-up.html}.
+The manual now has a chapter entirely dedicated to the data simulation functions with emphasis on both the methods used and the algorithms implemented (and illustrated with advanced examples).
+%TODO: check typos in there.
 
@@
 Specifically, dispRity offers four options (L145-156), but in effect these all boil down to treating a point along a branch as either identical to the descendant (end of branch) or the ancestor (start of branch).
-However, I think most biologists would treat such a value as more likely to be intermediate between these extremes and possessing some form of uncertianty depending on the generating model's assumptions (e.g., Brownian motion).
+However, I think most biologists would treat such a value as more likely to be intermediate between these extremes and possessing some form of uncertainty depending on the generating model's assumptions (e.g., Brownian motion).
 I don't mind offering a punctuated model (descendant value propagated back along branch) but some form of intermediate value should also exist.
-Without this option data are likely to be biased, e.g., towards generating more "jagged" time series and perhaps more significiant slice-to-slice changes (using the packages "test" functions) than are realistic (ie.e, type I errors).}}
+Without this option data are likely to be biased, e.g., towards generating more "jagged" time series and perhaps more significant slice-to-slice changes (using the packages "test" functions) than are realistic (ie.e, type I errors).}}
 
-%TG: Add a modeling function for time-slicing
+The standard approach in palaeobiology relies on the following pipeline: (i) coding characters for tips; (ii) running ancestral states reconstructions for the nodes; (iii) ordinating the resulting matrix (usually via a PCO); (iv) measuring disparity from there (see Lloyd 2016 \textit{Biol J Lin Soc} for an excellent description of this pipeline).
+I feel that the reviewer's suggestion of proposing more accurate models for ancestral states reconstructions and will technically not be doable after step (iii).
+In fact, the ordination, to be meaningful as a ``morphospace'' should always be done on the full dataset.
+I.e. no additional data should be added to the space \textit{a posteriori}.
+In practice, the \textit{time.subsamples} function can only select between two information: the ancestral node (DELTRAN) and the descendant node/tip (ACCTRAN) along the edge since only the two of them are real objects in the morphospace.
+It is therefore not possible to create an ancestral states reconstruction along the edge after running the ordination.
+
+I have however modified the available models to better describe this limitation and I have added two new models that would approximate ancestral reconstruction along edges
+I have modified the name of the two previous ``punctuated'' and ``gradual'' algorithms to ``random'' and ``proximity'' to reflect more accurately how they select either the ACCTRAN or DELTRAN information.
+I've added two more accurate models (now named ``punctuated'' and ``gradual'') to reflect the changes along the branches.
+These two algorithm now allows to use both the ACCTRAN and DELTRAN information with a linked probability to use either of them per bootstrap draw.
+For example, the ``punctuated'' model will have a 0.5 probability of using either node information for each bootstrap draw; the ``gradual'' model transforms the branch length were the slice occurs into a probability of drawing either of the nodes information (e.g.
+if the slices occurs 3/4 along the branch from A to B, the model will draw the information from A and B with a probability of respectively 0.25 and 0.75).
+For example, the median disparity over 1000 bootstrap pseudo-replication along the edge (AB) will integrate 0.75 information from A and 0.25 from B thus approximating the ``intermediate'' between A and B (when averaged ).
+
+I've added these models descriptions in the manual and in the manuscript:
+
+\begin{itemize}
+    \item \textit{``random'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor}
+    \item \textit{``proximity'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length}
+    \item \textit{``punctuated'' where the data chosen along the branch is both the descendant and the ancestor with an even probability}
+    \item \textit{``gradual'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length}
+\end{itemize}
+
+\textit{Note that the four first models are "fixed": the selected data is always either the one of the descendant or the ancestor.
+The two last models are "probabilistic": the data from both the descendant and the ancestor is used with an associate probability.
+These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).} lines @@@
 
@@
 
-%TG: add a warning to time subsamples? Change subsamples to subsets?
+I've changed ``subsample'' to ``subset'' throughout the package, the manual and the paper.
+
+I also agree with this comment: this method creates a trade-off between the number of slices and the type II errors: %TOCHEK@@@
+(1) too few slices decrease resolution and veer towards the time binning (which could increase "significance" by comparing too different time subsamples);
+(2) too many slices will increase the resolution be will create much more pseudo replication and thus increase type II error.
+
+A more detailed description and test of these models might be out of the scope of this package description paper but is currently in development for a special issue paper to be submitted to \textit{Palaeontology} as part as a special symposium for the Palaeontological Association annual meeting 2017 in collaboration with Natalie Cooper (\url{http://www.palass.org/meetings-events/annual-meeting/2017/annual-meeting-2017-london-symposium-abstracts}).
+
+I've added a warning in the manual and the manuscript:
+
+\textit{Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).} line @@@
 
@@
 
-%TG: add any matrix usage (done?) + boostrapping the raw data?
+I have updated the package's structure to be able to deal with any type of matrix as long as rows are the elements (e.g. taxa).
+It is now possible to perform the whole pipelined analysis (including bootstrapping) on non-ordinated matrices.
+
+I've reflected these changes in the manual with a subsection in the chapter ``Specific tutorials'' and in manuscript as follows:
+
+\textit{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements (e.g.
+the multidimensional space can be a distance matrixl; Close et al., 2015).} @@@
 
@@
 
-%TG: implement more metrics
+I agree, there are many more metrics commonly used for disparity analysis not implemented in this package.
+This was a conscious choice: due to the really high number of metrics available (especially when combining different ones), I designed the package to be modular hence allowing users to implement their own metrics (see response to Reviewer 1 minor comment \ref{table_missing}).
+I have precise this using the hypercube volume example in the manuscript:
+
+\textit{When multiple functions are passed to the} \texttt{metric} \textit{argument, the functions are sorted by dimension-level and applied in decreasing order to the data.
+For example, if the metric is defined as} \texttt{metric = c(prod, ranges)} \textit{(the hyperbox volume), the} \texttt{ranges} \textit{function (dimension-level 2) is first applied to data (calculating the range of each dimension) and the function} \texttt{prod} \textit{is then applied to the results (e.g.} \texttt{prod(ranges(data))}\textit{).
+If one needs to define the metric as something more complex, say the hypercube volume, one can directly pass a function description to the} \texttt{metric} \textit{argument (e.g.} \texttt{metric = function(x) min(ranges(x)) * ncol(x)}\textit{).}
+
+I have however added several additional ones to the newest version of the package including the minimal spanning tree length (\texttt{span.tree.length} - as suggested), the ancestral distance which is equivalent to the distance from centroid but were the ``centroid'' are the coordinates of each element's ancestors (\texttt{ancestral.dist}), the pairwise distances (\texttt{pairwise.dist}), the radius which equivalent to the distance from centroids but for the dimensions (i.e.
+the distance between each dimensions coordinates and the mean dimension's coordinates - \texttt{radius}) and the $n$-ball volume which is an extension of the \texttt{ellipse.volume} function for a sphere (the true $n$-ball volume) or an ellipsoid (\texttt{n.ball.volume}).
 

### 2017-11-16 paper/dispRity_paper_highlighted.tex
Updated response

@@
 Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow} and Table \ref{Tab:main_fun}).
+\hl{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements} \citep[e.g.
+the multidimensional space can be a distance matrix][]{Close2015}.
 
@@
         \texttt{Claddis.ordination} & Imports \hl{discrete morphological} data from the \texttt{Claddis} package. \\
-        \texttt{custom.subsamples} & Separates data into custom subsamples. \\
-        \texttt{time.subsamples} & Separates data into time subsamples. \\
+        \texttt{custom.subsets} & Separates data into custom subsets. \\
+        \texttt{time.subsets} & Separates data into time subsets. \\
         \texttt{boot.matrix} & Bootstraps and rarefies a matrix or a \disp object. \\
@@
 When multiple functions are passed to the \texttt{metric} argument, the functions are sorted by dimension-level and applied in decreasing order to the data.
-For example, if the metric is defined as \texttt{metric = c(mean, range)}, the \texttt{range} function (dimension-level 2) is first applied to the data and the function \texttt{mean} is then applied to the result (e.g.
-\texttt{mean(range(data))}).
+For example, if the metric is defined as \texttt{metric = c(\hl{prod}, ranges)} \hl{(the hyperbox volume)}, the \texttt{ranges} function (dimension-level 2) is first applied to data \hl{(calculating the range of each dimension)} and the function \hl{\texttt{prod}} is then applied to the results (e.g.
+\hl{\texttt{prod(ranges(data))}}).
+\hl{If one needs to define the metric as something more complex, say the hypercube volume, one can directly pass a function description to the \texttt{metric} argument (e.g.
+\texttt{metric = function(x) min(ranges(x)) * ncol(x)}).}
 There is no limitation on the metric dimensions only that at least one dimension-level 2 or 1 function must be present in the metric definition (i.e.
@@
     \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
-        name & description & dimen-sion level & definition & source \\
+        name & description & dim level & definition & source \\
         \hline
-        \texttt{ellipse. volume}$^1$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
+        \texttt{ancestral. dist} & The distance between an element and its ancestor & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}$ & This package \\
+        \texttt{centroids} & The distance between each element and a fixed point$^1$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & This package \\
         \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
@@
         \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & This package \\
+        \texttt{ellipse. volume}$^2$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
         \texttt{mode.val} & The modal value & 1 & NA & This package\\
+        \texttt{n.ball. volume} & The hyper-spherical ($n$-ball) volume & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R$ & This package\\
+        \texttt{pairwise. dist} & The pairwise distances between elements & 2 & NA & \texttt{stats::dist}\\        
+        \texttt{radius} & The radius of each dimensions & 2 & $|\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|$ & This package\\
         \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & This package \\
+        \texttt{span.tree. length} & The minimal spanning tree length & 1 & $\sum(\mathrm{branch\ length})$ & \texttt{vegan::spantree}\\        
         \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & This package \\
-        \texttt{centroids} & The distance between each element and a fixed point$^2$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & This package \\
-        \texttt{ancestral. distance} & The distance between an element and its ancestor & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{k})^2}}$ & This package \\
     \end{tabular}
 }%
-    \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean.
-$^1$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA); $^2$ by default that point is the centroid of the elements.}}
+    % \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g.
+it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}$ - for a hyper-ellipsoid).
+$^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA).}}
     \label{Tab:metrics}
@@
 
-%TG: Add the proper sources for the convhulln package
+\textbf{Table 1 (caption):} Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g.
+it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid).
+$^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA).
+
+%TODO: Add the proper sources for the convhulln and vegan packages
 
 \subsection{Splitting the multidimensional space into subsets}
-Prior to calculating disparity, the multidimensional space can be subdivided into subsamples, typically to be compared to each other.
-For example, one may be interested in seeing how the disparity of a specific subsample of the space compares to another (e.g.
-difference among groups) or, how different subsamples change sequentially (e.g.
+Prior to calculating disparity, the multidimensional space can be subdivided into subsets, typically to be compared to each other.
+For example, one may be interested in seeing how the disparity of a specific subset of the space compares to another (e.g.
+difference among groups) or, how different subsets change sequentially (e.g.
 through time).
 In essence, the original space corresponds to the overall space (e.g. a morphospace contains all the observed morphologies).
-Conversely, subsamples correspond to parts of the space with pooled characteristics (i.e. only a subset of the total morphospace).
+Conversely, subsets correspond to parts of the space with pooled characteristics (i.e. only a subset of the total morphospace).
 
-This splitting can be done using the \texttt{custom.subsamples} or \texttt{time.subsamples} functions.
+This splitting can be done using the \texttt{custom.subsets} or \texttt{time.subsets} functions.
 \hl{The first argument is the matrix}
-defining the multidimensional space and a list of elements to group into different subsamples.
-The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time subsamples for example,
+defining the multidimensional space and a list of elements to group into different subsets.
+The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time subsets for example,
 \hl{see below}
-) and which subsamples to create: (1) discrete time subsamples (or time-binning) or (2) continuous time subsamples (or time-slicing).
+) and which subsets to create: (1) discrete time subsets (or time-binning) or (2) continuous time subsets (or time-slicing).
 
@@
 This method thus requires the nodes to be part of the multidimensional space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
-When a slice occurs not on a tip or a node, four methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
+When a slice occurs not on a tip or a node, \hl{six} methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
 \begin{itemize}
@@
     \item ``deltran'' (delayed transformation) where the data chosen along the branch is always the one of the ancestor
-    \item ``punctuated'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor
-    \item ``gradual'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length
+    \item \hl{``random''} where the data chosen along the branch is randomly chosen between the descendant or the ancestor
+    \item \hl{``proximity''} where the data chosen along the branch is either the descendant or the ancestor depending on branch length
+    \item \hl{``punctuated'' where the data chosen along the branch is both the descendant and the ancestor with an even probability}
+    \item \hl{``gradual'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length}
 \end{itemize}
+
+\hl{Note that the four first models are "fixed": the selected data is always either the one of the descendant or the ancestor.
+The two last models are "probabilistic": the data from both the descendant and the ancestor is used with an associate probability.
+These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).
+Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).}
 
@@
 Disparity measurement can be influenced by sampling \citep{Butler2012}.
-To take this source of bias into account, it is common practice to bootstrap the multidimensional space (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different subsamples).
+To take this source of bias into account, it is common practice to bootstrap the multidimensional space (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different subsets).
 Additionally, if disparity is defined as a dimension-level 1 metric, it can be useful to measure it on bootstrapped data to obtain a distribution on which to perform statistical analyses.
@@
 Similarly, rarefaction can be achieved through the same \texttt{boot.matrix} function.
-Rarefaction can be useful when looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of subsamples with the same number of elements in each.
+Rarefaction can be useful when looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of subsets with the same number of elements in each.
 In practice, rarefaction limits the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).
@@
 The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity} (provided disparity was calculated).
-These results can then be analysed using the \texttt{test.dispRity} function for comparing subsamples or testing hypotheses.
+These results can then be analysed using the \texttt{test.dispRity} function for comparing subsets or testing hypotheses.
 
@@
 \subsubsection{Testing hypotheses}
-The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data (e.g. by comparing the different subsamples to each other).
+The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data (e.g. by comparing the different subsets to each other).
 Similarly to the \disp function described above, this function can take any test defined by the user or from other R packages.
-The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsamples: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subsample to all the others; (3) ``sequential'' for comparing subsamples sequentially (e.g.
-first against second, second against third, etc.); (4) `all'' for comparing all the subsamples simultaneously (i.e.
-\texttt{disparity $\mathtt{\sim}$ subsamples} - this can be used for \texttt{lm} or \texttt{glm}) or (5) any list of pairs (or more) of subsamples to compare.
+The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsets: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subset to all the others; (3) ``sequential'' for comparing subsets sequentially (e.g.
+first against second, second against third, etc.); (4) `all'' for comparing all the subsets simultaneously (i.e.
+\texttt{disparity $\mathtt{\sim}$ subsets} - this can be used for \texttt{lm} or \texttt{glm}) or (5) any list of pairs (or more) of subsets to compare.
 
@@
 
-\noindent \texttt{> subsamples <- custom.subsamples(data = BeckLee\_mat50, group = mammal\_groups)}\\
-\noindent \texttt{> bootstraps <- boot.matrix(subsamples)}\\
+\noindent \texttt{> subsets <- custom.subsets(data = BeckLee\_mat50, group = mammal\_groups)}\\
+\noindent \texttt{> bootstraps <- boot.matrix(subsets)}\\
 \noindent \texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\
@@
 \noindent \texttt{ ---- dispRity object ---- }\\
-\noindent \texttt{2 customised subsamples for 50 elements with 48 dimensions:}\\
+\noindent \texttt{2 customised subsets for 50 elements with 48 dimensions:}\\
           \texttt{crown, stem.}\\
@@
 \bigskip
-As indicated, the \disp object contains two customised subsamples (called ``crown'' and ``stem'') from a morphospace made of 50 elements for 48 dimensions.
+As indicated, the \disp object contains two customised subsets (called ``crown'' and ``stem'') from a morphospace made of 50 elements for 48 dimensions.
 The \disp object also displays information on the number and method of the bootstrap replicates as well as the definition of disparity.
@@
 % mammal_groups <- list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44))
-% subsamples <- custom.subsamples(data = BeckLee_mat50, group = mammal_groups)
-% bootstraps <- boot.matrix(subsamples)
+% subsets <- custom.subsets(data = BeckLee_mat50, group = mammal_groups)
+% bootstraps <- boot.matrix(subsets)
 % disparity <- dispRity(bootstraps, metric = c(sum, variances))
@@
   \hline
- & subsamples & n & obs & bs.median & 2.5\% & 25\% & 75\% & 97.5\% \\ 
+ & subsets & n & obs & bs.median & 2.5\% & 25\% & 75\% & 97.5\% \\ 
   \hline
@@
 \caption{Summarising a \disp object (disparity per groups).
-$n$ is the number of elements per subsamples, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 95\% are the confidence intervals.}
+$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 95\% are the confidence intervals.}
 \label{Tab:summary_group}
@@
 
-\noindent \texttt{> subsamples <- time.subsamples(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
-\noindent \texttt{> bootstraps <- boot.matrix(subsamples)}\\
+\noindent \texttt{> subsets <- time.subsets(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
+\noindent \texttt{> bootstraps <- boot.matrix(subsets)}\\
 \noindent \texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\
@@
 \caption{Summarising a \disp object (disparity through time).
-$n$ is the number of elements per subsamples, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 95\% are the confidence intervals.}
+$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 95\% are the confidence intervals.}
 \label{Tab:summary_time}
@@
   \hline
-subsamples & 2 & 3.14 & 1.57 & 177.84 & 0.0000 \\ 
+subsets & 2 & 3.14 & 1.57 & 177.84 & 0.0000 \\ 
   Residuals & 297 & 2.62 & 0.01 &  &  \\ 
@@
 % In fact, optimised architecture and a C implementations are being currently developed for faster calculations and lower memory footprints for most computational heavy functions (e.g.
-\disp or \texttt{time.subsamples}).
+\disp or \texttt{time.subsets}).
 % New functionalities currently being developed will include time sequential linear models (for accurately measuring directionality in disparity changes between time slices) or mode of disparity evolution testing (allowing to fit different models such as Brownian Motion or Ornstein-Uhlenbeck to disparity curves).

### 2017-11-17 paper/response_to_reviewers.tex
Reviewer 2 major done-ish

@@
 
-% %TG: adding rarefaction algorithms.
+This is again an excellent suggestion!
+I will develop more complex rarefaction algorithms based on sampling and occurrence based on methods applied empirically by Kotric and Knoll 2015 \textit{Palaeobiology} and Butler et al 2012 \textit{Evolution}.
+I've added a note
+
+% %TG: TODO maybe!
 
 \item{\textcolor{blue}{Other minor additions the author may also wish to consider are the permutation test from Brusatte et al.
-(2014, Current Biology, 24, 2386-), a means of adding the geologic timescale to time series plots (e.g., using the geoscale package), and options for the user to only use a limited set of ordination axes (matrix columns), e.g., to avoid using higher axes with negative eigenvalues.}}
+(2014, Current Biology, 24, 2386-), a means of adding the geologic timescale to time series plots (e.g., using the geoscale package)}
+
+I have added an illustrative example in the \texttt{plot.dispRity} manual on how to add disparity data (continuous or discrete) to the \texttt{geoscalePlot} and \texttt{geoscaleBox} functions thus allowing users to make ready-to-publish figures.
+
+\item{\textcolor{blue}{and options for the user to only use a limited set of ordination axes (matrix columns), e.g., to avoid using higher axes with negative eigenvalues.}}
 
@@
 
-l. @@@: \textit{Note that this function also allows to work on only a subset of dimensions via the} \texttt{dimensions} \textit{argument (e.g. if only the $n$ first dimensions must be considered).}
+\textit{Note that this function also allows to work on only a subset of dimensions via the} \texttt{dimensions} \textit{argument (e.g. if only the $n$ first dimensions must be considered).} l. @@@:
 

### 2017-11-17 inst/gitbook/02_getting-started.Rmd
updated manual

@@
 
-To illustrate these functions, we will use data from Beck and Lee (2014).
+To illustrate these functions, we will use data from @beckancient2014.
 This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
@@
 
-Using the Beck and Lee (2014) data described [above](#example-data):
+Using the @beckancient2014 data described [above](#example-data):
 
@@
 ## Creating the two groups (crown versus stem) as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44))
+mammal_groups <- crown.stem(BeckLee_tree, inc.nodes = TRUE)
 

### 2017-11-17 inst/gitbook/06_palaeo-demo.Rmd
updated manual

@@
 
-In this example, we are going to use a subset of the data from Beck and Lee (2014).
+In this example, we are going to use a subset of the data from @beckancient2014.
 See the [example data](#example-data) description for more details.
@@
 Note that this metric comes with a caveat (not solved here) since it ignores covariances among the dimensions of the morphospace.
-We use this here because it is a standard metric used in disparity-through-time analysis (Wills _et al._ 1994).
+We use this here because it is a standard metric used in disparity-through-time analysis [@Wills1994].
 

### 2017-11-17 inst/gitbook/07_ecology-demo.Rmd
updated manual

@@
 One other series of test can be done on the shape of the petal-space.
-Using a MCMC permutation test we can simulate a petal-space with specific properties and see if our observed petal-space matches these properties (similarly to Diaz _et al._ 2016):
+Using a MCMC permutation test we can simulate a petal-space with specific properties and see if our observed petal-space matches these properties (similarly to @diaz2016global):
 

### 2017-11-17 inst/gitbook/08_future_directions.Rmd
Updated book (subsamples -> subsets)
updated manual

--- 2017-08-15 inst/gitbook/07_future_directions.Rmd
@@
 
+## More rarefactions!
+
+I am working on including several "clever" bootstrap and rarefaction algorithms for palaeobiology studies based on geological sampling (following:and Butler et al.
+2012 *Evolution* and Kotric and Knoll 2015 *Palaeobiology*).
+
 ## Faster disparity calculations
@@
 I am equally slowly developing functions to allow more of the options in the package to be modular (in the same way as the `metric` argument in `dispRity`).
-The next arguments to benefit this increased modularity will be `time.subsamples`'s `model` argument and `boot.matrix`'s `type` argument.
+The next arguments to benefit this increased modularity will be `time.subsets`'s `model` argument and `boot.matrix`'s `type` argument.
 
-# References
+## More suggestions?
 
+More generally, I welcome any suggestions or collaborations for developing some additional specific ideas.
+Feel free to contact me via email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
+
+<!-- 
 * Beck, R.
@@
 Ecology letters, 16(4), 421-429.
-* Wills, M. A., Briggs, D. E., & Fortey, R. A. (1994). Disparity as an evolutionary index: a comparison of Cambrian and Recent arthropods. Paleobiology, 20(2), 93-130.
+* Wills, M. A., Briggs, D. E., & Fortey, R. A. (1994). Disparity as an evolutionary index: a comparison of Cambrian and Recent arthropods. Paleobiology, 20(2), 93-130. -->

### 2017-11-17 inst/gitbook/dispRity_manual.Rmd
updated manual

--- 2017-11-16 inst/gitbook/03_specific-tutorials.Rmd
@@
+--- 
+title: "dispRity manual"
+author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
+date: "`r Sys.Date()`"
+site: bookdown::bookdown_site
+output: bookdown::gitbook
+documentclass: book
+link-citations: yes
+github-repo: TGuillerme/dispRity
+description: "dispRity R package vignette"
+---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
+
+# `dispRity`
+
+This is a package for measuring disparity in `R`.
+It allows users to summarise matrices as representations as multidimensional spaces into a single value or distribution describing a specific aspect of this multidimensional space (the disparity).
+Multidimensional spaces can be ordinated matrices from MDS, PCA, PCO, PCoA but the package is *not* restricted to any type of matrices!
+This manual is based on the version `0.4.2`.
+
+## What is `dispRity`?
+This is a modular package for measuring disparity in `R`.
+It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
+Typically, these analysis are used in palaeobiology and evolutionary biology to study the changes in morphology through time.
+However, there are many more applications in ecology, evolution and beyond.
+
+### Modular?
+Because their exist a multitude of ways to measure disparity, each adapted to every specific question, this package uses an easy to modify modular architecture.
+In coding, each module is simply a function or a modification of a function that can be passed to the main functions of the package to tweak it to your proper needs!
+In practice, you will notice throughout this manual that some function can take other functions as arguments: the modular architecture of this package allows you to use any function for these arguments (with some restrictions explained for each specific cases).
+This will allow you to finely tune your multidimensional analysis to the needs of your specific question!
+
+## Installing and running the package
+
+You can install this package easily if you use the latest version of `R` (> 3.4.0) and `devtools`.
+
+```{r, eval=FALSE}
+## Checking if devtools is already installed
+if(!require(devtools)) install.packages("devtools")
+
+## Installing the latest released version directly from GitHub
+install_github("TGuillerme/dispRity", ref = "release")
+
+## Loading the package
+library(dispRity)
+```
+
+```{r, eval = TRUE, echo = FALSE, message = FALSE}
+## loading the package and setting up the start seed.
+library(dispRity)
+## Setting a random seed for repeatability
+set.seed(123)
+```
+
+Note this uses the `release` branch (version`0.4`). <!-- @@@ -->
+For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
+`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, `Claddis`, `geomorph` and `RCurl`).
+
+## Why not CRAN?
+
+This package is not available on CRAN.
+This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
+However, the package follows the strict (and useful!) CRAN standards via Travis.
+
+<!-- badges out start -->
+[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
+[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/master/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
+[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
+<!-- badges out end   -->
+
+## Help
+
+If you need help with the package, hopefully the following manual will be useful.
+However, parts of this package are still in development and some other parts are probably not covered.
+Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
+
+## Citations
+
+You can cite both the package or this manual with the following citation:
+
+> Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646
+
+<!-- > Guillerme, T. & Cooper, N. (2016). The dispRity R package manual. Fighsare XXXX -->
+
+Note that this citation is only temporary (but can still be used!).
+A proper description of the package is currently in review in Methods in Ecology and Evolution and should be re-submitted this winter.
+Upon acceptance (hopefully!), a proper DOI will be released for this manual and the package description.
+
+<!--chapter:end:index.Rmd-->
+
+---
+title: "Glossary"
+author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
+date: "`r Sys.Date()`"
+output:
+  html_document: default
+  pdf_document: default
+---
+
+# Glossary
+
+-   **Multidimensional space**. The mathematical multidimensional object that will be analysed with this package.
+    In morphometrics, this is often referred to as the morphospace.
+    However it may also be referred to as the cladisto-space for cladistic data or the eco-space for ecological data etc.
+    In practice, this term designates a matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
+
+-   **Elements**. The rows of the multidimensional space matrix. Elements can be taxa, field sites, countries etc.
+
+-   **Dimensions**.
+The columns of the multidimensional space matrix.
+The dimensions can be referred to as axes of variation, or principal components, for ordinated spaces obtained from a PCA for example.
+
+-   **Subsets**. Subsets of the multidimensional space.
+    A subset (or subsets) contains the same number of dimensions as the space but may contain a smaller subset of elements.
+    For example, if our space is composed of birds and mammals (the elements) and 50 principal components of variation (the dimensions), we can create two subsets containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
+
+<!--chapter:end:01_glossary.Rmd-->
+
+---
+title: "Getting started with dispRity"
+author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
+date: "`r Sys.Date()`"
+output:
+  html_document: default
+  pdf_document: default
+---
+
+# Getting started with `dispRity`
+
+## What sort of data does `dispRity` work with?
+
+Disparity can be estimated from pretty much any matrix.
+Classically, however, it is measured from ordinated matrices.
+These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).
+
+### Ordination matrices from `Claddis`
+
+```{r, eval = TRUE, echo = FALSE, message = FALSE}
+## Checking if Claddis is installed BUILD ONLY
+if(!require(Claddis)) install.packages("Claddis")
+```
+
+`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
+For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:
+
+```{r}
+require(Claddis)
+
+## Ordinating the example data from Claddis
+Claddis.ordination(Michaux1989) 
+```
+
+Note that several options are available, namely which type of distance should be computed.
+See more info in the function manual (`?Claddis.ordination`).
+Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.
+
+### Ordination matrices from `geomorph`
+
+```{r, eval = TRUE, echo = FALSE, message = FALSE}
+## Checking if geomorph is installed BUILD ONLY
+if(!require(geomorph)) install.packages("geomorph")
+```
+
+You can also easily use data from `geomorph` using the `geomorph.ordination` function.
+This function simply takes Procrustes aligned data and performs an ordination:
+
+```{r}
+require(geomorph)
+
+## Loading the plethodon dataset
+data(plethodon)
+
+## Performing a Procrustes transform on the landmarks
+procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)
+
+## Ordinating this data
+geomorph.ordination(procrustes)[1:5,1:5]
+```
+
+Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
+Additionally you can give the function a `geomorph.data.frame` object.
+If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsets)!
+
+```{r}
+## Using a geomorph.data.frame
+geomorph_df <- geomorph.data.frame(procrustes,
+     species = plethodon$species, site = plethodon$site)
+
+## Ordinating this data and making a dispRity object
+geomorph.ordination(geomorph_df)
+```
+
+More about these `dispRity` objects below!
+
+### Other kinds of ordination matrices
+
+If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordination matrices by using the following functions from the `stats` package.
+Here is how to do it for the following types of matrices:
+
+ * Multivariate matrices (principal components analysis; PCA)
+
+```{r}
+## A multivariate matrix
+head(USArrests)
+
+## Ordinating the matrix using `prcomp` 
+ordination <- prcomp(USArrests)
+
+## Selecting the ordinated matrix
+ordinated_matrix <- ordination$x
+head(ordinated_matrix)
+```
+  This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, see the `?princomp` function.
+
+ * Distance matrices (classical multidimensional scaling; MDS)
+
+```{r}
+## A matrix of distances between cities
+str(eurodist)
+
+## Ordinating the matrix using cmdscale() with k = 5 dimensions 
+ordinated_matrix <- cmdscale(eurodist, k = 5)
+head(ordinated_matrix)
+```
+
+  This results in a ordinated matrix with European cities as elements and five dimensions.
+
+Of course any other method for creating the ordination matrix is totally valid, you can also not use any ordination at all!
+The only requirements for the `dispRity` functions is that the input is a matrix with elements as rows and dimensions as columns.
+
+## Performing a simple dispRity analysis
+
+Two `dispRity` functions allow users to run an analysis pipeline simply by inputting an ordination matrix.
+These functions allow users to either calculate the disparity through time (`dispRity.through.time`) or the disparity of user-defined groups (`dispRity.per.group`).
+
+**IMPORTANT**
+
+Note that `disparity.through.time` and `disparity.per.group` are wrapper functions (i.e.
+they incorporate lots of other functions) that allow users to run a basic disparity-through-time, or disparity among groups, analysis without too much effort.
+As such they use a lot of default options. 
+These are described in the help files for the functions that are used to make the wrapper functions, and not described in the help files for `disparity.through.time` and `disparity.per.group`. 
+These defaults are good enough for **data exploration**, but for a proper analysis you should consider the **best parameters for your question and data**. 
+For example, which metric should you use?
+How many bootstraps do you require?
+What model of evolution is most appropriate if you are time slicing?
+Should you rarefy the data?
+See [`time.subsets`](#time-slicing), [`custom.subsets`](#customised-subsets), [`boot.matrix`](#bootstraps-and-rarefactions) and [`dispRity.metric`](#disparity-metrics) for more details of the defaults used in each of these functions.
+Note that any of these default arguments can be changed within the `disparity.through.time` or `disparity.per.group` functions.
+
+### Example data
+
+To illustrate these functions, we will use data from @beckancient2014.
+This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
+
+```{r, fig.width=7, fig.height=7}
+## Loading the ordinated matrices
+data(BeckLee_mat50)
+data(BeckLee_mat99)
+
+## The first five taxa and dimensions of the 50 taxa matrix
+head(BeckLee_mat50[, 1:5])
+
+## The first five taxa and dimensions of the 99 taxa + ancestors matrix
+BeckLee_mat99[c(1, 2, 98, 99), 1:5]
+
+## Loading a list of first and last occurrence dates for the fossils
+data(BeckLee_ages)
+head(BeckLee_ages)
+
+## Loading and plotting the phylogeny
+data(BeckLee_tree)
+plot(BeckLee_tree, cex = 0.8) 
+axisPhylo(root = 140)
+nodelabels(cex = 0.5)
+```
+
+Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).
+
+### Disparity through time
+
+The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
+This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
+Of course all the defaults can be changed if required, more on this later.
+
+For a disparity through time analysis, you will need:
+
+  * An ordinated matrix (we covered that above)
+  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element.
+To give your tree a root time (i.e.
+an age for the root), you can simply do\\ `my_tree$root.time <- my_age`.
+  * The required number of time subsets (here `time = 3`)
+  * Your favourite disparity metric (here the sum of variances)
+
+Using the Beck and Lee (2014) data described [above](#example-data):
+
+```{r}
+## Measuring disparity through time
+disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
+                                        time = 3, metric = c(sum, variances))
+```
+
+This generates a `dispRity` object (see [here](#guts) for technical details).
+When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:
+
+```{r}
+## Print the disparity_data object
+disparity_data
+```
+
+We asked for three subsets (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
+
+We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):
+
+```{r, fig.width=7, fig.height=7}
+## Summarising disparity through time
+summary(disparity_data)
+
+## Plotting the results
+plot(disparity_data, type = "continuous")
+
+## Testing for an difference among the time bins
+disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
+summary(disp_lm)
+```
+
+Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
+You can also directly explore the specific function help files within R and navigate to related functions.
+
+### Disparity among groups
+
+The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
+For example, you could ask if there is a difference in disparity between two groups?
+
+To perform such an analysis, you will need:
+
+ * An matrix with rows as elements and columns as dimensions (always!)
+ * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix.
+For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4.
+Note that elements can be present in multiple groups at once.
+ * Your favourite disparity metric (here the sum of variances)
+
+Using the @beckancient2014 data described [above](#example-data):
+
+```{r}
+## Creating the two groups (crown versus stem) as a list
+mammal_groups <- crown.stem(BeckLee_tree, inc.nodes = TRUE)
+
+## Measuring disparity for each group
+disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
+                                     metric = c(sum, variances))
+```
+
+We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).
+
+```{r, fig.width=7, fig.height=7}
+## Print the disparity_data object
+disparity_data
+
+## Summarising disparity in the different groups
+summary(disparity_data)
+
+## Plotting the results
+plot(disparity_data)
+
+## Testing for a difference between the groups
+test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
+```
+
+<!--chapter:end:02_getting-started.Rmd-->
+
 ---
@@
 
-Before each section, make sure you loaded the Beck and Lee (2014) data (see [example data](#example-data) for more details).
+Before each section, make sure you loaded the @beckancient2014 data (see [example data](#example-data) for more details).
 
@@
 The median centroids distance is expressed in km in the "Distance differences" plots and in Euclidean units of variation in the "Ordinated differences" plots.
+
+<!--chapter:end:03_specific-tutorials.Rmd-->
+
+---
+title: "Data simulation features"
+author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
+date: "`r Sys.Date()`"
+output:
+  html_document: default
+  pdf_document: default
+---
+
+# Making stuff up!
+
+The `dispRity` package also offers some advanced data simulation features to allow to test hypothesis, explore ordinate-spaces or metrics properties or simply playing around with data!
+All the following functions are based on the same modular architecture of the package and therefore can be used with most of the functions of the package.
+
+## Simulating discrete morphological data
+
+The function `sim.morpho` allows to simulate discrete morphological data matrices (sometimes referred to as "cladistic" matrices).
+It allows to evolve multiple discrete characters on a given phylogenetic trees, given different models, rates, and states.
+It even allows to include proper inapplicable data to make datasets as messy as in real life!
+
+In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
+The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
+The methods are described in details below
+
+```{r}
+set.seed(3)
+## Simulating a starting tree with 15 taxa as a random coalescent tree
+my_tree <- rcoal(15)
+
+## Generating a matrix with 100 characters (85% binary and 15% three state) and
+## an equal rates model with a gamma rate distribution (0.5, 1) with no 
+## invariant characters.
+my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
+    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)
+
+## The first few lines of the matrix
+my_matrix[1:5, 1:10]
+
+## Checking the matrix properties with a quick Maximum Parsimony tree search
+check.morpho(my_matrix, my_tree)
+```
+
+Note that this example produces a tree with a great consistency index and an identical topology to the random coalescent tree!
+Nearly too good to be true...
+
+### A more detailed description
+
+The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;OReilly2017].
+
+ * The first `tree` argument will be the tree on which to "evolve" the characters and therefore requires branch length.
+ You can generate quick and easy random Yule trees using `ape::rtree(number_of_taxa)` but I would advise to use more realistic trees for more realistic simulations based on more realistic models (really realistic then) using the function `tree.bd` from the [`diversitree`](http://www.zoology.ubc.ca/prog/diversitree/) package [@fitzjohndiversitree2012].
+ * The second argument, `character` is the number of characters. Pretty straight forward.
+ * The third, `states` is the proportion of characters states above two (yes, the minimum number of states is two).
+This argument intakes the proportion of *n*-states characters, for example `states = c(0.5,0.3,0.2)` will generate 50% of binary-state characters, 30% of three-state characters and 20% of four-state characters.
+There is no limit in the number of state characters proportion as long as the total makes up 100%.
+ * The forth, `model` is the evolutionary model for generating the character(s). More about this below.
+ * The fifth and sixth, `rates` and `substitution` are the model parameters described below as well.
+ * Finally, the two logical arguments, are self explanatory: `invariant` whether to allow invariant characters (i.e.
+characters that don't change) and `verbose` whether to print the simulation progress on your console.
+
+#### Available evolutionary models
+
+There are currently three evolutionary models implemented in `sim.morpho` but more will come in the future.
+Note also that they allow fine tuning parameters making them pretty plastic!
+
+ * `"ER"`: this model allows any number of character states and is based on the Mk model [@lewisa2001].
+It assumes a unique overall evolutionary rate equal substitution rate between character states.
+This model is based on the `ape::rTraitDisc` function.
+ * `"HKY"`: this is binary state character model based on the molecular HKY model [@HKY85].
+It uses the four molecular states (A,C,G,T) with a unique overall evolutionary rate and a biased substitution rate towards transitions (A <-> G or C <-> T) against transvertions (A <-> C and G <-> T).
+After evolving the nucleotide, this model transforms them into binary states by converting the purines (A and G) into state 0 and the pyrimidines (C and T) into state 1.
+This method is based on the `phyclust::seq.gen.HKY` function and was first proposed by @OReilly2016.
+ * `"MIXED"`: this model uses a random (uniform) mix between both the `"ER"` and the `"HKY"` models.
+
+The models can take the following parameters:
+(1) `rates` is the evolutionary rate (i.e. the rate of changes along a branch: the evolutionary speed) and (2) `substitution` is the frequency of changes between one state or another.
+For example if a character can have high probability of changing (the *evolutionary* rate) with, each time a change occurs a probability of changing from state *X* to state *Y* (the *substitution* rate).
+Note that in the `"ER"` model, the substitution rate is ignore because... by definition this (substitution) rate is equal!
+
+The parameters arguments `rates` and `substitution` takes a distributions from which to draw the parameters values for each character.
+For example, if you want an `"HKY"` model with an evolutionary rate (i.e.
+speed) drawn from a uniform distribution bounded between 0.001 and 0.005, you can define it as `rates = c(runif, min = 0.001, max = 0.005)`, `runif` being the function for random draws from a uniform distribution and `max` and `min` being the distribution parameters.
+These distributions should always be passed in the format `c(random_distribution_function, distribution_parameters)` with the names of the distribution parameters arguments.
+
+#### Checking the results
+
+An additional function, `check.morpho` runs a quick Maximum Parsimony tree search using the `phangorn` parsimony algorithm.
+It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g.
+the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree to determine how different they are.
+
+#### Adding inapplicable characters
+
+Once a matrix is generated, it is possible to apply inapplicable characters to it for increasing realism!
+Inapplicable characters are commonly designated as `NA` or simply `-`.
+They differ from missing characters `?` in their nature by being inapplicable rather than unknown.
+For example, considering a binary character defined as "colour of the tail" with the following states "blue" and "red"; on a taxa with no tail, the character should be coded as inapplicable ("`-`") since the state of the character "colour of tail" is *known*: it's neither "blue" or "red", it's just not there!
+It contrasts with coding it as missing ("`?`" - also called as ambiguous) where the state is *unknown*, for example, the taxon of interest is a fossil where the tail has no colour preserved or is not present at all due to bad conservation!
+
+This type of characters can be added to the simulated matrices using the `apply.NA` function/
+It takes, as arguments, the `matrix`, the source of inapplicability (`NAs` - more below), the `tree` used to generate the matrix and the two same `invariant` and `verbose` arguments as defined above.
+The `NAs` argument allows two types of sources of inapplicability:
+
+ * `"character"` where the inapplicability is due to the character (e.g. coding a character tail for species with no tail).
+ In practice, the algorithm chooses a character *X* as the underlying character (e.g.
+"presence and absence of tail"), arbitrarily chooses one of the states as "absent" (e.g.
+0 = absent) and changes in the next character *Y* any state next to character *X* state 0 into an inapplicable token ("`-`").
+ This simulates the inapplicability induced by coding the characters (i.e. not always biological).
+ * `"clade"` where the inapplicability is due to evolutionary history (e.g. a clade loosing its tail).
+ In practice, the algorithm chooses a random clade in the tree and a random character *Z* and replaces the state of the taxa present in the clade by the inapplicable token ("`-`").
+ This simulates the inapplicability induced by evolutionary biology (e.g. the lose of a feature in a clade).
+
+To apply these sources of inapplicability, simply repeat the number of inapplicable sources for the desired number of characters with inapplicable data.
+
+```{r}
+## Generating 5 "character" NAs and 10 "clade" NAs
+my_matrix_NA <- apply.NA(my_matrix, tree = my_tree,
+                         NAs = c(rep("character", 5), rep("clade", 10)))
+
+## The first few lines of the resulting matrix
+my_matrix_NA[1:10, 90:100]
+```
+<!--TG: Add inapplicable paper here -->
+
+### Parameters for a realistic(ish) matrix
+
+There are many parameters that can create a "realistic" matrix (i.e.
+not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all parameters combination end up creating "good" matrices.
+The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
+
+```{r}
+set.seed(0)
+## tree
+my_tree <- rcoal(15)
+## matrix
+morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
+    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
+check.morpho(morpho_mat, my_tree)
+```
+
+<!-- ## Simulating continuous morphological data -->
+<!-- TG: Add tutorial for evolving "morpho" shapes -->
+
+## Simulating multidimensional spaces
+Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
+This function allows users to simulate multidimensional spaces with a certain number of properties.
+For example, it is possible to design a multidimensional space with a specific distribution on each axis, a correlation between the axes and a specific cumulative variance per axis.
+This can be useful for creating ordinated spaces for null hypothesis, for example if you're using the function `null.test` [@diaz2016global].
+
+This function takes as arguments the number of elements (data points - `elements` argument) and dimensions (`dimensions` argument) to create the space and the distribution functions to be used for each axis.
+The distributions are passed through the `distribution` argument as... modular functions!
+You can either pass a single distribution function for all the axes (for example `distribution = runif` for all the axis being uniform) or a specific distribution function for each specific axis (for example `distribution = c(runif, rnorm, rgamma))` for the first axis being uniform, the second normal and the third gamma).
+You can of course use your very own functions or use the ones implemented in `dispRity` for more complex ones (see below).
+Specific optional arguments for each of these distributions can be passed as a list via the `arguments` argument.
+
+Furthermore, it is possible to add a correlation matrix to add a correlation between the axis via the `cor.matrix` argument or even a vector of proportion of variance to be bear by each axis via the `scree` argument to simulate realistic ordinated spaces.
+
+Here is a simple two dimensional example:
+
+```{r, fig.width=8, fig.height=8, eval=TRUE}
+## Graphical options
+op <- par(bty = "n")
+
+## A square space
+square_space <- space.maker(100, 2, runif)
+
+## The resulting 2D matrix
+head(square_space)
+
+## Visualising the space
+plot(square_space, pch = 20, xlab = "", ylab = "", main = "Uniform 2D space")
+```
+
+Of course, more complex spaces can be created by changing the distributions, their arguments or adding a correlation matrix or a cumulative variance vector:
+
+```{r}
+## A plane space: uniform with one dimensions equal to 0
+plane_space <- space.maker(2500, 3, c(runif, runif, runif),
+                           arguments = list(list(min = 0, max = 0), NULL, NULL))
+
+## Correlation matrix for a 3D space
+(cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3))
+
+## An ellipsoid space (normal space with correlation)
+ellipse_space <- space.maker(2500, 3, rnorm, cor.matrix = cor_matrix)
+
+## A cylindrical space with decreasing axes variance
+cylindrical_space <- space.maker(2500, 3, c(rnorm, rnorm, runif),
+                                 scree = c(0.7, 0.2, 0.1))
+```
+
+### Personalised dimensions distributions
+
+Following the modular architecture of the package, it is of course possible to pass home made distribution functions to the `distribution` argument.
+For example, the `random.circle` function is a personalised one implemented in `dispRity`.
+This function allows to create circles based on basic trigonometry allowing to axis to covary to produce circle coordinates.
+By default, this function generates two sets of coordinates with a `distribution` argument and a minimum and maximum boundary (`inner` and `outer` respectively) to create nice sharp edges to the circle.
+The maximum boundary is equivalent to the radius of the circle (it removes coordinates beyond the circle radius) and the minimum is equivalent to the radius of a smaller circle with no data (it removes coordinates below this inner circle radius).
+
+```{r, fig.width=8, fig.height=8, eval=TRUE}
+## Graphical options
+op <- par(bty = "n")
+
+## Generating coordinates for a normal circle with a upper boundary of 1
+circle <- random.circle(1000, rnorm, inner = 0, outer = 1)
+
+## Plotting the circle
+plot(circle, xlab = "x", ylab = "y", main = "A normal circle")
+
+## Creating doughnut space (a spherical space with a hole)
+doughnut_space <- space.maker(5000, 3, c(rnorm, random.circle),
+     arguments = list(list(mean = 0), list(runif, inner = 0.5, outer = 1)))
+```
+
+### Visualising the space
+
+I suggest using the excellent `scatterplot3d` package to play around and visualise the simulated spaces:
+
+```{r, fig.width=8, fig.height=8, eval=TRUE}
+## Graphical options
+op <- par(mfrow = (c(2, 2)), bty = "n")
+## Visualising 3D spaces
+require(scatterplot3d)
+
+## The plane space
+scatterplot3d(plane_space, pch = 20, xlab = "", ylab = "", zlab = "",
+              xlim = c(-0.5, 0.5), main = "Plane space")
+
+## The ellipsoid space
+scatterplot3d(ellipse_space, pch = 20, xlab = "", ylab = "", zlab = "",
+              main = "Normal ellipsoid space")
+
+## A cylindrical space with a decreasing variance per axis
+scatterplot3d(cylindrical_space, pch = 20, xlab = "", ylab = "", zlab = "",
+              main = "Normal cylindrical space")
+## Axes have different orders of magnitude
+
+## Plotting the doughnut space
+scatterplot3d(doughnut_space[,c(2,1,3)], pch = 20, xlab = "", ylab = "",
+              zlab = "", main = "Doughnut space")
+par(op)
+```
+
+### Generating realistic spaces
+
+It is possible to generate "realistic" spaces by simply extracting the parameters of an existing space and scaling it up to the simulated space.
+For example, we can extract the parameters of the `BeckLee_mat50` ordinated space and simulate a similar space.
+
+```{r, fig.width=8, fig.height=8, eval=TRUE}
+## Loading the data
+data(BeckLee_mat50)
+
+## Number of dimensions
+obs_dim <- ncol(BeckLee_mat50)
+
+## Observed correlation between the dimensions
+obs_correlations <- cor(BeckLee_mat50)
+
+## Observed mean and standard deviation per axis
+obs_mu_sd_axis <- mapply(function(x,y) list("mean" = x, "sd" = y),
+                         as.list(apply(BeckLee_mat50, 2, mean)),
+                         as.list(apply(BeckLee_mat50, 2, sd)), SIMPLIFY = FALSE)
+
+## Observed overall mean and standard deviation
+obs_mu_sd_glob <- list("mean" = mean(BeckLee_mat50), "sd" = sd(BeckLee_mat50))
+
+## Scaled observed variance per axis (scree plot)
+obs_scree <- variances(BeckLee_mat50)/sum(variances(BeckLee_mat50))
+
+## Generating our simulated space
+simulated_space <- space.maker(1000, dimensions = obs_dim, 
+                               distribution = rep(list(rnorm), obs_dim),
+                               arguments = obs_mu_sd_axis,
+                               cor.matrix = obs_correlations)
+
+## Visualising the fit of our data in the space (in the two first dimensions)
+plot(simulated_space[,1:2], xlab = "PC1", ylab = "PC2")
+points(BeckLee_mat50[,1:2], col = "red", pch = 20)
+legend("topleft", legend = c("observed", "simulated"),
+        pch = c(20,21), col = c("red", "black"))
+```
+
+It is now possible to simulate a space using these observed arguments to test several hypothesis:
+
+ * Is the space uniform or normal?
+ * If the space is normal, is the mean and variance global or specific for each axis?
+
+```{r, fig.width=8, fig.height=8, eval=TRUE}
+## Measuring disparity as the sum of variance
+observed_disp <- dispRity(BeckLee_mat50, metric = c(median, centroids))
+
+## Is the space uniform?
+test_unif <- null.test(observed_disp, null.distrib = runif)
+
+## Is the space normal with a mean of 0 and a sd of 1?
+test_norm1 <- null.test(observed_disp, null.distrib = rnorm)
+
+## Is the space normal with the observed mean and sd and cumulative variance
+test_norm2 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
+                        null.args = rep(list(obs_mu_sd_glob), obs_dim),
+                        null.scree = obs_scree)
+
+## Is the space multiple normal with multiple means and sds and a correlation?
+test_norm3 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
+                        null.args = obs_mu_sd_axis, null.cor = obs_correlations)
+
+## Graphical options
+op <- par(mfrow = (c(2, 2)), bty = "n")
+## Plotting the results
+plot(test_unif, main = "Uniform (0,1)")
+plot(test_norm1, main = "Normal (0,1)")
+plot(test_norm2, main = paste0("Normal (", round(obs_mu_sd_glob[[1]], digit = 3),
+                              ",", round(obs_mu_sd_glob[[2]], digit = 3), ")"))
+plot(test_norm3, main = "Normal (variable + correlation)")
+
+```
+
+If we measure disparity as the median distance from the morphospace centroid, we can explain the distribution of the data as normal with the variable observed mean and standard deviation and with a correlation between the dimensions.
+
+<!--chapter:end:04_making-up-data.Rmd-->
+
+---
+title: "The guts of the dispRity package"
+author: "Thomas Guillerme (guillert@tcd.ie)"
+date: "`r Sys.Date()`"
+output:
+  html_document: default
+  pdf_document: default
+---
+
+# The guts of the `dispRity` package
+
+## Manipulating `dispRity` objects
+
+Disparity analysis involves a lot of manipulation of many matrices (especially when bootstrapping) which can be impractical to visualise and will quickly overwhelm your `R` console.
+Even the simple Beck and Lee 2014 example above produces an object with > 72 lines of lists of lists of matrices!
+
+Therefore `dispRity` uses a specific class of object called a `dispRity` object.
+These objects allow users to use S3 method functions such as `summary.dispRity`, `plot.dispRity` and `print.dispRity`. 
+`dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g. `sort.dispRity`, `extract.dispRity` see the full list in the next section).
+These functions modify the `dispRity` object without having to delve into its complex structure!
+The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).
+
+```{r}
+## Loading the example data
+data(disparity)
+
+## What is the class of the median_centroids object?
+class(disparity)
+
+## What does the object contain?
+names(disparity)
+
+## Summarising it using the S3 method print.dispRity
+disparity
+```
+
+Note that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:
+
+```{r, eval=FALSE}
+## Display the full object
+print(disparity, all = TRUE)
+## This is more nearly ~ 5000 lines on my 13 inch laptop screen!
+```
+
+## `dispRity` utilities
+
+The package also provides some utility functions to facilitate multidimensional analysis.
+
+### `dispRity` object utilities  <a name="dispRity.utilities"></a>
+The first set of utilities are functions for manipulating `dispRity` objects:
+
+#### `make.dispRity`
+This function creates empty `dispRity` objects.
+
+```{r}
+## Creating an empty dispRity object
+make.dispRity()
+
+## Creating an "empty" dispRity object with a matrix
+(disparity_obj <- make.dispRity(matrix(rnorm(20), 5, 4)))
+```
+
+####  `fill.dispRity`
+This function initialises a `dispRity` object and generates its call properties.
+
+```{r}
+## The dispRity object's call is indeed empty
+disparity_obj$call
+
+## Filling an empty disparity object (that needs to contain at least a matrix)
+(disparity_obj <- fill.dispRity(disparity_obj))
+
+## The dipRity object has now the correct minimal attributes
+disparity_obj$call
+```
+
+####  `matrix.dispRity`
+This function extracts a specific matrix from a disparity object.
+The matrix can be one of the bootstrapped matrices or/and a rarefied matrix.
+
+```{r}
+## Extracting the matrix containing the coordinates of the elements at time 50
+str(matrix.dispRity(disparity, "50"))
+
+## Extracting the 3rd bootstrapped matrix with the 2nd rarefaction level
+## (15 elements) from the second group (80 Mya)
+str(matrix.dispRity(disparity, subsets = 1, bootstrap = 3, rarefaction = 2))
+```
+
+#### `get.subsets.dispRity`
+This function creates a dispRity object that contains only elements from one specific subsets.
+
+```{r, eval=FALSE}
+## Extracting all the data for the crown mammals
+(crown_mammals <- get.subsets.dispRity(disp_crown_stemBS, "Group.crown"))
+
+## The object keeps the properties of the parent object but is composed of only one subsets
+length(crown_mammals$subsets)
+```
+
+#### `extract.dispRity`
+This function extracts the calculated disparity values of a specific matrix.
+
+```{r, eval=FALSE}
+## Extracting the observed disparity (default)
+extract.dispRity(disparity)
+
+## Extracting the disparity from the bootstrapped values from the
+## 10th rarefaction level from the second subsets (80 Mya)
+extract.dispRity(disparity, observed = FALSE, subsets = 2, rarefaction = 10)
+```
+
+#### `scale.dispRity`
+This is the S3 method of `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object.
+
+```{r, eval=FALSE}
+## Getting the disparity values of the time subsets
+head(summary(disparity))
+
+## Scaling the same disparity values
+head(summary(scale(disparity, scale = TRUE)))
+
+## Scaling and centering:
+head(summary(scale(disparity, scale = TRUE, center = TRUE)))
+```
+
+#### `sort.dispRity`
+This is the S3 method of `sort` for sorting the subsets alphabetically (default) or following a specific pattern.
+
+```{r, eval=FALSE}
+## Sorting the disparity subsets in inverse alphabetic order
+head(summary(sort(disparity, decreasing = TRUE)))
+
+## Customised sorting
+head(summary(sort(disparity, sort = c(7, 1, 3, 4, 5, 2, 6))))
+```
+
+## The `dispRity` object content
+The functions above are utilities to easily and safely access different elements in the `dispRity` object.
+Alternatively, of course, each elements can be accessed manually.
+Here is an explanation on how it works.
+The `dispRity` object is a `list` of two to four elements, each of which are detailed below:
+
+ * `$matrix`: an object of class `matrix`, the full multidimensional space.
+ * `$call`: an object of class `list` containing information on the `dispRity` object content.
+ * `$subsets`: an object of class `list` containing the subsets of the multidimensional space.
+ * `$disparity`: an object of class `list` containing the disparity values.
+
+The `dispRity` object is loosely based on `C` structure objects.
+In fact, it is composed of one unique instance of a matrix (the multidimensional space) upon which the metric function is called via "pointers" to only a certain number of elements and/or dimensions of this matrix.
+This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint:  at any time, only one matrix is present in the `R` environment rather than multiple copies of it for each subset.
+
+### `$matrix`
+This is the multidimensional space, stored in the `R` environment as a `matrix` object.
+It requires row names but not column names.
+By default, if the row names are missing, `dispRity` function will arbitrarily generate them in numeric order (i.e. `rownames(matrix) <- 1:nrow(matrix)`).
+This element of the `dispRity` object is never modified.
+
+### `$call`
+This element contains the information on the `dispRity` object content.
+It is a `list` that can contain the following:
+
+ * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"punctuated"`, `"gradual"`).
+This element generated only once via `time.subsets()` and `custom.subsets()`.
+ * `$call$dimensions`: either a single `numeric` value indicating how many dimensions to use or a vector of `numeric` values indicating which specific dimensions to use.
+This element is by default the number of columns in `$matrix` but can be modified through `boot.matrix()` or `dispRity()`.
+ * `$call$bootstrap`: this is a `list` containing three elements:
+    * `[[1]]`: the number of bootstrap replicates (`numeric`)
+    * `[[2]]`: the bootstrap method (`character`)
+    * `[[3]]`: the rarefaction levels (`numeric` vector)
+ * `$call$disparity`: this is a `list` containing one element, `$metric`, that is a `list` containing the different functions passed to the `metric` argument in `dispRity`.
+These are `call` elements and get modified each time the `dispRity` function is used (the first element is the first metric(s), the second, the second metric(s), etc.).
+
+### `$subsets`
+This element contain the eventual subsets of the multidimensional space.
+It is a `list` of subset names.
+Each subset name is in turn a `list` of at least one element called `elements` which is in turn a `matrix`.
+This `elements` matrix is the raw (observed) elements in the subsets.
+The `elements` matrix is composed of `numeric` values in one column and _n_ rows (the number of elements in the subset).
+Each of these values are a "pointer" (`C` inspired) to the element of the `$matrix`.
+For example, lets assume a `dispRity` object called `disparity`, composed of at least one subsets called `sub1`:
+
+```
+ disparity$subsets$sub1$elements
+      [,1]
+ [1,]    5
+ [2,]    4
+ [3,]    6
+ [4,]    7
+```
+
+The values in the matrix "point" to the elements in `$matrix`: here, the multidimensional space with only the 4th, 5th, 6th and 7th elements.
+The following elements in `diparity$subsets$sub1` will correspond to the same "pointers" but drawn from the bootstrap replicates.
+The columns will correspond to different bootstrap replicates.
+For example:
+
+```
+ disparity$subsets$sub1[[2]]
+      [,1] [,2] [,3] [,4]
+ [1,]   57   43   70    4
+ [2,]   43   44    4    4
+ [3,]   42   84   44    1
+ [4,]   84    7    2   10
+```
+This signifies that we have four bootstrap pseudo-replicates pointing each time to four elements in `$matrix`.
+The next element (`[[3]]`) will be the same for the eventual first rarefaction level (i.e.
+the resulting bootstrap matrix will have _m_ rows where _m_ is the number of elements for this rarefaction level).
+The next element after that (`[[4]]`) will be the same for with an other rarefaction level and so forth...
+
+### `$disparity`
+The `$disparity` element is identical to the `$subsets` element structure (a list of list(s) containing matrices) but the matrices don't contain "pointers" to `$matrix` but the disparity result of the disparity metric applied to the "pointers".
+For example, in our first example (`$elements`) from above, if the disparity metric is of dimensions level 1, we would have:
+
+```
+ disparity$disparity$sub1$elements
+      [,1]
+ [1,]    1.82
+```
+
+This is the observed disparity (1.82) for the subset called `sub1`.
+If the disparity metric is of dimension level 2 (say the function `range` that outputs two values), we would have:
+
+```
+ disparity$disparity$sub1$elements
+      [,1]
+ [1,]    0.82
+ [2,]    2.82
+```
+
+The following elements in the list follow the same logic as before: rows are disparity values (one row for a dimension level 1 metric, multiple for a dimensions level 2 metric) and columns are the bootstrap replicates (the bootstrap with all elements followed by the eventual rarefaction levels).
+For example for the bootstrap without rarefaction (second element of the list):
+
+```
+ disparity$disparity$sub1[[2]]
+         [,1]     [,2]     [,3]     [,4]
+[1,] 1.744668 1.777418 1.781624 1.739679 
+```
+
+<!-- ## Modularity
+
+### Making your own metrics
+
+### Making your own tests
+
+## Where is this going?
+ -->
+<!-- ## Running `dispRity` in parallel
+
+The computationally intensive function `dispRity` have a `parallel` option to speed up their calculations.
+
+This option requires the package `snow` and takes arguments that are to be passed to  -->
+
+<!--chapter:end:05_guts.Rmd-->
+
+---
+title: "Palaeobiology demo: disparity-through-time and within groups"
+author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
+date: "`r Sys.Date()`"
+output:
+  html_document: default
+  pdf_document: default
+---
+
+# Palaeobiology demo: disparity-through-time and within groups
+
+This demo aims to give quick overview of the `dispRity` package (v.0.4) for palaeobiology analyses of disparity, including disparity through time analyses.
+
+This demo showcases a typical disparity-through-time analysis: we are going to test whether the disparity changed through time in a subset of eutherian mammals from the last 100 million years using a dataset from Beck and Lee (2014).
+
+## Before starting
+
+### The morphospace
+
+In this example, we are going to use a subset of the data from @beckancient2014.
+See the [example data](#example-data) description for more details.
+Briefly, this dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
+The ordinated matrix will represent our full morphospace, i.e. all the mammalian morphologies that ever existed through time (for this dataset).
+
+```{r, fig.width=6, fig.height=6}
+## Loading demo and the package data
+library(dispRity)
+
+## Setting the random seed for repeatability
+set.seed(123)
+
+## Loading the ordinated matrix/morphospace:
+data(BeckLee_mat50)
+head(BeckLee_mat50[,1:5])
+
+dim(BeckLee_mat50)
+## The morphospace contains 50 taxa and has 48 dimensions (or axes)
+
+## Showing a list of first and last occurrences data for some fossils
+data(BeckLee_ages)
+head(BeckLee_ages)
+
+## Plotting a phylogeny
+data(BeckLee_tree)
+plot(BeckLee_tree, cex = 0.7)
+axisPhylo(root = 140)
+```
+
+> You can have an even nicer looking tree if you use the `strap` package!
+
+```{r, eval = TRUE}
+if(!require(strap)) install.packages("strap")
+strap::geoscalePhylo(BeckLee_tree, cex.tip = 0.7, cex.ts = 0.6)
+```
+
+## A disparity-through-time analysis
+
+### Splitting the morphospace through time
+
+One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time subsets that contain the total number of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
+Basically, the full morphospace represents the total number of morphologies across all time and will be greater than any of the time subsets of the morphospace.
+
+The `dispRity` package provides a `time.subsets` function that allows users to split the morphospace into time slices (using `method = continuous`) or into time bins (using `method = discrete`).
+In this example, we are going to split the morphospace into five equal time bins of 20 million years long from 100 million years ago to the present.
+We will also provide to the function a table containing the first and last occurrences dates for some fossils to take into account that some fossils might occur in several of our different time bins.
+
+```{r}
+## Creating the vector of time bins ages
+(time_bins <- rev(seq(from = 0, to = 100, by = 20)))
+
+## Splitting the morphospace using the time.subsets function
+(binned_morphospace <- time.subsets(data = BeckLee_mat50, tree = BeckLee_tree,
+    method = "discrete", time = time_bins, inc.nodes = FALSE,
+    FADLAD = BeckLee_ages))
+```
+
+The output object is a `dispRity` object (see more about that [here](#The-guts-of-the-dispRity-package).
+In brief, however, `dispRity` objects are lists of different elements (i.e.
+disparity results, morphospace time subsets, morphospace attributes, etc.) that display only a summary of the object when calling the object to avoiding filling the `R` console with superfluous output.
+
+```{r}
+## Printing the class of the object
+class(binned_morphospace)
+
+## Printing the content of the object
+str(binned_morphospace)
+names(binned_morphospace)
+
+## Printing the object as a dispRity class
+binned_morphospace
+```
+
+> These objects will gradually contain more information when completing the following steps in the disparity-through-time analysis.
+
+### Bootstrapping the data
+
+Once we obtain our different time subsets, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
+The bootstrapping allows us to make each subset more robust to outliers and the rarefaction allows us to compare subsets with the same number of taxa to remove sampling biases (i.e.
+more taxa in one subset than the others).
+The `boot.matrix` function bootstraps the `dispRity` object and the `rarefaction` option within performs rarefaction.
+
+```{r}
+## Bootstrapping each time subset 100 times (default)
+(boot_bin_morphospace <- boot.matrix(binned_morphospace))
+
+## Getting the minimum number of rows (i.e. taxa) in the time subsets
+min(size.subsets(boot_bin_morphospace))
+
+## Bootstrapping each time subset 100 times and rarefying them 
+(rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
+    rarefaction = 6))
+```
+
+### Calculating disparity
+
+We can now calculate the disparity within each time subsets along with some confidence intervals generated by the pseudoreplication step above (bootstraps/rarefaction).
+Disparity can be calculated in many ways and this package allows users to come up with their own disparity metrics.
+For more details, please refer to the [`dispRity` metric section](#disparity-metrics).
+
+In this example, we are going to calculate the spread of the data in each time subset by calculating disparity as the sum of the variance of each dimension of the morphospace in each time subset using the `dispRity` function.
+Thus, in this example, disparity is defined by the multi-dimensional variance of each time subset (i.e. the spread of the taxa within the morphospace).
+Note that this metric comes with a caveat (not solved here) since it ignores covariances among the dimensions of the morphospace.
+We use this here because it is a standard metric used in disparity-through-time analysis [@Wills1994].
+
+```{r}
+## Calculating disparity for the bootstrapped data
+(boot_disparity <- dispRity(boot_bin_morphospace, metric = c(sum, variances)))
+
+## Calculating disparity for the rarefied data
+(rare_disparity <- dispRity(rare_bin_morphospace, metric = c(sum, variances)))
+```
+
+The `dispRity` function does not actually display the calculated disparity values but rather only the properties of the disparity object (size, subsets, metric, etc.).
+To display the actual calculated scores, we need to summarise the disparity object using the S3 method `summary` that is applied to a `dispRity` object (see `?summary.dispRity` for more details).
+
+As for any `R` package, you can refer to the help files for each individual function for more details.
+
+```{r}
+## Summarising the disparity results
+summary(boot_disparity)
+summary(rare_disparity)
+```
+> The summary.dispRity function comes with many options on which values to calculate (central tendency and quantiles) and on how many digits to display.
+Refer to the function's manual for more details.
+
+### Plotting the results
+
+It is sometimes easier to visualise the results in a plot than in a table.
+For that we can use the `plot` S3 function to plot the `dispRity` objects (see `?plot.dispRity` for more details).
+
+```{r, fig.width=12, fig.height=6}
+## Graphical options
+quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
+
+## Plotting the bootstrapped and rarefied results
+plot(boot_disparity, type = "continuous", main = "bootstrapped results")
+plot(rare_disparity, type = "continuous", main = "rarefied results")
+```
+
+Testing differences
+-------------------
+
+Finally, to draw some valid conclusions from these results, we can apply some statistical tests.
+We can test, for example, if mammalian disparity changed significantly through time over the last 100 million years.
+To do so, we can compare the means of each time-bin in a sequential manner to see whether the disparity in bin *n* is equal to the disparity in bin *n+1*, and whether this is in turn equal to the disparity in bin *n+2*, etc.
+Because our data is temporally autocorrelated (i.e.
+what happens in bin *n+1* depends on what happened in bin *n*) and pseudoreplicated (i.e.
+each bootstrap draw creates non-independent time subsets because they are all based on the same time subsets), we apply a non-parametric mean comparison: the `wilcox.test`.
+Also, we need to apply a p-value correction (e.g. Bonferroni correction) to correct for multiple testing (see `?p.adjust` for more details).
+
+```{r}
+## Testing the differences between bins in the bootstrapped dataset.
+test.dispRity(boot_disparity, test = wilcox.test, comparison = "sequential",
+    correction = "bonferroni")
+
+## Testing the differences between bins in the rarefied dataset.
+test.dispRity(rare_disparity, test = wilcox.test, comparison = "sequential",
+    correction = "bonferroni")
+```
+
+Here our results show significant changes in disparity through time between all time bins (all p-values < 0.05).
+However, when looking at the rarefied results, there is no significant difference between the time bins in the Palaeogene (60-40 to 40-20 Mya), suggesting that the differences detected in the first test might just be due to the differences in number of taxa sampled (13 or 6 taxa) in each time bin.
+
+<!--chapter:end:06_palaeo-demo.Rmd-->
+
+---
+title: "dispRity ecology demo"
+author: "Thomas Guillerme"
+date: "`r Sys.Date()`"
+output: rmarkdown::html_vignette
+bibliography: References.bib
+bst: sysbio.bst
+vignette: >
+  %\VignetteIndexEntry{dispRity ecology demo}
+  %\VignetteEngine{knitr::rmarkdown}
+  \usepackage[utf8]{inputenc}
+---
+
+# Ecology demo
+
+This is an example of typical disparity analysis that can be performed in ecology.
+
+## Data
+For this example, we will use the famous `iris` inbuilt data set
+
+```{r}
+data(iris)
+```
+
+This data contains petal and sepal length for 150 individual plants sorted into three species.
+
+```{r}
+## Separating the species
+species <- iris[,5]
+## Which species?
+unique(species)
+
+## Separating the petal/sepal length
+measurements <- iris[,1:4]
+head(measurements)
+```
+
+We can then ordinate the data using a PCA (`prcomp` function) thus defining our four dimensional space as the poetically named petal-space.
+
+```{r}
+## Ordinating the data
+ordination <- prcomp(measurements)
+
+## The petal-space
+petal_space <- ordination$x
+
+## Adding the elements names to the petal-space (the individuals IDs)
+rownames(petal_space) <- 1:nrow(petal_space)
+```
+
+## Classic analysis
+A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different species are distributed in the petal-space.
+
+```{r, fig.width=6, fig.height=6}
+## Measuring the variance on each axis
+variances <- apply(petal_space, 2, var)
+variances <- variances/sum(variances)
+
+## Graphical option
+par(bty = "n")
+
+## A classic 2D ordination plot
+plot(petal_space[, 1], petal_space[, 2], col = species,
+    xlab = paste0("PC 1 (", round(variances[1], 2), ")"),
+    ylab = paste0("PC 2 (", round(variances[2], 2), ")"))
+```
+
+This shows the distribution of the different species in the petal-space along the two first axis of variation.
+This is a pretty standard way to visualise the multidimensional space and further analysis might be necessary to test wether the groups are different such as a linear discriminant analysis (LDA).
+However, in this case we are ignoring the two other dimensions of the ordination!
+If we look at the two other axis we see a totally different result:
+
+```{r, fig.width=6, fig.height=6}
+## Plotting the two second axis of the petal-space
+plot(petal_space[, 3], petal_space[, 4], col = species,
+    xlab = paste0("PC 3 (", round(variances[3], 2), ")"),
+    ylab = paste0("PC 4 (", round(variances[4], 2), ")"))
+```
+
+Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e.
+the values on the first dimension do not represent a continuous trait (e.g.
+petal length), instead they just represent the ordinations of correlations between the data and some factors.
+
+Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.
+
+## A multidimensional approach with `dispRity`
+
+The first step is to create different subsets that represent subsets of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
+Each of these subsets will contain only the individuals of a specific species.
+
+```{r}
+## Creating the table that contain the elements and their attributes
+petal_subsets <- custom.subsets(petal_space, group = list(
+                                "setosa" = which(species == "setosa"),
+                                "versicolor" = which(species == "versicolor"),
+                                "virginica" = which(species == "virginica")))
+
+## Visualising the dispRity object content
+petal_subsets
+```
+
+This created a `dispRity` object (more about that [here](#guts)) with three subsets corresponding to each subspecies.
+
+### Bootstrapping the data
+We can the bootstrap the subsets to be able test the robustness of the measured disparity to outliers.
+We can do that using the default options of `boot.matrix` (more about that [here](#bootstraps-and-rarefactions)):
+
+```{r}
+## Bootstrapping the data
+(petal_bootstrapped <- boot.matrix(petal_subsets))
+```
+
+### Calculating disparity
+
+Disparity can be calculated in many ways, therefore the `dispRity` function allows users to define their own measure of disparity.
+For more details on measuring disparity, see the [dispRity metrics section](#disparity-metrics).
+
+In this example, we are going to define disparity as the median distance between the different individuals and the centroid of the ordinated space.
+High values of disparity will indicate a generally high spread of points from this centroid (i.e. on average, the individuals are far apart in the ordinated space).
+We can define the metrics easily in the `dispRity` function by feeding them to the `metric` argument.
+Here we are going to feed the functions `stats::median` and `dispRity::centroids` which calculates distances between elements and their centroid.
+
+```{r}
+## Calculating disparity as the median distance between each elements and
+## the centroid of the petal-space
+(petal_disparity <- dispRity(petal_bootstrapped, metric = c(median, centroids)))
+```
+
+### Summarising the results (plot)
+Similarly to the `custom.subsets` and `boot.matrix` function, `dispRity` displays a `dispRity` object.
+But we are definitely more interested in actually look at the calculated values.
+
+First we can summarise the data in a table by simply using `summary`:
+
+```{r}
+## Displaying the summary of the calculated disparity
+summary(petal_disparity)
+```
+
+We can also plot the results in a similar way:
+
+```{r, fig.width=6, fig.height=6}
+## Graphical options
+par(bty = "n")
+
+## Plotting the disparity in the petal_space
+plot(petal_disparity)
+```
+
+Now contrary to simply plotting the two first axis of the PCA where we saw that the species have a different position in the two first petal-space, we can now also see that they occupy this space clearly differently!
+
+### Testing hypothesis
+
+Finally we can test our hypothesis that we guessed from the disparity plot (that some groups occupy different volume of the petal-space) by using the `test.dispRity` option.
+
+```{r}
+## Fitting a linear model to our data
+disparity_lm <- test.dispRity(petal_disparity, test = lm, comparisons = "all")
+
+## Testing whether there is a difference in disparity between the different species
+summary(aov(disparity_lm))
+
+## Post-hoc testing of the differences between species (corrected for multiple tests)
+test.dispRity(petal_disparity, test = t.test, correction = "bonferroni")
+```
+
+We can now see that there is a significant difference in petal-space occupancy between all species of iris.
+
+#### Setting up a multidimensional null-hypothesis
+
+One other series of test can be done on the shape of the petal-space.
+Using a MCMC permutation test we can simulate a petal-space with specific properties and see if our observed petal-space matches these properties (similarly to @diaz2016global):
+
+```{r, fig.width=6, fig.height=6}
+## Testing against a uniform distribution
+disparity_uniform <- null.test(petal_disparity, replicates = 200,
+    null.distrib = runif, scale = FALSE)
+plot(disparity_uniform)
+```
+
+```{r, fig.width=6, fig.height=6}
+## Testing against a normal distribution
+disparity_normal <- null.test(petal_disparity, replicates = 200,
+    null.distrib = rnorm, scale = TRUE)
+plot(disparity_normal)
+```
+
+In both cases we can see that our petal-space is not entirely normal or uniform.
+This is expected because of the simplicity of these parameters.
+
+<!--chapter:end:07_ecology-demo.Rmd-->
+
+---
+title: "Future directions and references"
+author: "Thomas Guillerme"
+date: "`r Sys.Date()`"
+output: rmarkdown::html_vignette
+bibliography: References.bib
+bst: sysbio.bst
+vignette: >
+  %\VignetteIndexEntry{future directions}
+  %\VignetteEngine{knitr::rmarkdown}
+  \usepackage[utf8]{inputenc}
+---
+
+# Future directions
+
+## More tests!
+
+Some more tests are being developed such as a `sequential.test` to run sequential linear models for testing hypothesis in disparity through time or a `model.test` developed in collaboration with [Mark Puttick](https://puttickbiology.wordpress.com/) to fit modes of evolution to disparity curves.
+Stay tuned!
+
+## More rarefactions!
+
+I am working on including several "clever" bootstrap and rarefaction algorithms for palaeobiology studies based on geological sampling (following:and Butler et al.
+2012 *Evolution* and Kotric and Knoll 2015 *Palaeobiology*).
+
+## Faster disparity calculations
+
+I am slowly implementing parallel disparity calculation as well as C implementations of some disparity metrics to increase significantly improve the speed of the `dispRity` function.
+
+## More modularity
+
+I am equally slowly developing functions to allow more of the options in the package to be modular (in the same way as the `metric` argument in `dispRity`).
+The next arguments to benefit this increased modularity will be `time.subsets`'s `model` argument and `boot.matrix`'s `type` argument.
+
+## More suggestions?
+
+More generally, I welcome any suggestions or collaborations for developing some additional specific ideas.
+Feel free to contact me via email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
+
+<!-- 
+* Beck, R.
+M., & Lee, M.
+S.
+(2014).
+Ancient dates or accelerated rates?
+Morphological clocks and the antiquity of placental mammals.
+Proceedings of the Royal Society of London B: Biological Sciences, 281(1793), 20141278.
+* Cooper, N., & Guillerme, T. (in prep.). Coming soonish!.
+* Diaz, S., Kattge, J., Cornelissen, J.H., Wright, I.J., Lavorel, S., Dray, S., Reu, B., Kleyer, M., Wirth, C., Prentice, I.C.
+and Garnier, E.
+(2016).
+The global spectrum of plant form and function.
+Nature, 529(7585), 167.
+* Donohue, I., Petchey, O.L., Montoya, J.M., Jackson, A.L., McNally, L., Viana, M., Healy, K., Lurgi, M., O'Connor, N.E.
+and Emmerson, M.C, (2013).
+On the dimensionality of ecological stability.
+Ecology letters, 16(4), 421-429.
+* Wills, M. A., Briggs, D. E., & Fortey, R. A. (1994). Disparity as an evolutionary index: a comparison of Cambrian and Recent arthropods. Paleobiology, 20(2), 93-130. -->
+
+<!--chapter:end:08_future_directions.Rmd-->

### 2017-11-20 paper/response_to_reviewers.tex
Responses to reviewer 3

@@
 \item{\textcolor{blue}{Other minor additions the author may also wish to consider are the permutation test from Brusatte et al.
-(2014, Current Biology, 24, 2386-), a means of adding the geologic timescale to time series plots (e.g., using the geoscale package)}
+(2014, Current Biology, 24, 2386-), a means of adding the geologic timescale to time series plots (e.g., using the geoscale package)}}
 
@@
 The author may also want to cite use of non-metric multidimensional scaling (e.g., Liow 2004, American Naturalist, 164, 431–).}}
+\label{PCO_cites}
 
-%TG: check paper and cite
+I have changed this sentence to:
+
+\textit{These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix (Lloyd 2016 and references therein; e.g.
+in Close et al.
+2015), or as outputs from an ordination, whether it being a principal components analysis (PCA, Hotelling 1933; e.g.
+in Zelditch et al.
+2012), a metric scaling (PCO, PCoA, Torgerson 1958; e.g.
+in Brusatte et al.
+2008) or a non-metric scaling (MDS, NMDS, Shepard 1962; e.g.
+in Liow 2004; Donohue et al.
+2013).} lines @@@
 
@@
 
-%TG: done this down
+I've change this sentence to:
+
+\textit{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best it their needs (i.e.
+data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.}  lines @@@
 
@@
 
-%Implement a crown/stem phylo function
+Great suggestion! I've added an utility function \texttt{crown.stem} allowing to separate stem and crown species.
+The function is now briefly mentioned in the package description paper and detailed in the package itself.
 
@@
 I've fixed this typo.
-
-\item{\textcolor{blue}{L324 - Figure 4 caption should be "elements per subsample" and "50\% and 95\%".}}
-
-%Check! fixed and already
 
@@
 For example, disparity is defined twice as an object on lines 255 and 261.
-On line 261 it could be defined as DisparityB, and subsequently used without overwriting the object produced from a different function in line 255.}}
-
-% %TG: rename objects in the example
-
-\item{\textcolor{blue}{Relatedly, one should not name objects the same name as functions.
+On line 261 it could be defined as DisparityB, and subsequently used without overwriting the object produced from a different function in line 255.
+Relatedly, one should not name objects the same name as functions.
 This is simply bad news.
@@
 (The author will field requests for help from users, so this can be considered a pitfall avoided beforehand.)  This is discussed some here: https://stackoverflow.com/questions/6135868/in-r-what-exactly-is-the-problem-with-having-variables-with-the-same-name-as-ba}}
+
+I have changed the variable names in the package description manuscript to not be redundant.
+I've also changed the variable names in the package's help files to not override other functions/objects.
 
@@
 \end{itemize}
+
+The ANOVA problem is totally a garbage-in-garbage-out problem and I agree with this reviewer that this might be a serious problem down the line.
+
+Unfortunately, I do not have any proper solution to propose to this problem and I guess multidimensional distributions comparisons methods needs yet to be developed!
+
+The reason for these crazy high degrees of freedom come from the bootstrapping pseudo-replication: effectively, the test data contains all the bootstrap data.
+One solution is to use the concatenate = FALSE option in test.dispRity, performing the test for each bootstrap replicate rather than pooling the data.
+This methods outputs a distribution of test that can be used to assess the type I error.
+However, the pooling of the p-value is not satisfactory either.
+
+Some methods exists such as NPMANOVA (from vegan::adonis) but is performed on the wrong kind of matrix.
+
+One of the reasons for this functionality in the package is to make the test.dispRity function modular like the other main functions: i.e.
+it \textit{can} apply a \textit{lm} class function to the data but the reason for doing that might be dubious (or not, if the users don't bootstrap the matrix for example).
+
+%TG: I therefore removed the example in the manuscript and the manual to not mislead users into running flawed analysis.
+However, I've left the architecture implemented for adventurous (or clever) users.
+% Remove the ANOVA example
+% Remove mentions to it on bootstrapped data in the manuals
+% Replace by more savy example
+% Allow to use NPMANOVA instead
 
@@
 
+See my response to Reviewer 2 - Minor suggestion \ref{PCO_cites}.
+
 % %TG: source orginal papers
@@
 %TG: negative egein values explnations
+% Check Legendre and Legendre pp.449
 

### 2017-11-20 paper/dispRity_paper_highlighted.tex
Responses to reviewer 3

@@
 of the variables).}
-These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix \citep[][and references therein]{lloyd2016estimating}, or as outputs from a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, or a multidimensional scaling \citep[MDS;][]{DonohueDim}, among others.
+\hl{These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix (}
+\citealt{lloyd2016estimating}
+\hl{and references therein; e.g. in}
+\citealt{Close2015}
+\hl{), or as outputs from an ordination, whether it being a principal components analysis (PCA,}
+\citealt{PCA}
+\hl{; e.g. in}
+\citealt{zelditch2012geometric}
+\hl{), a metric scaling (PCO, PCoA,}
+\citealt{PCO}
+\hl{; e.g. in}
+\citealt{Brusatte12092008}
+\hl{) or a non-metric scaling (MDS, NMDS,}
+\citealt{MDS}
+\hl{; e.g. in}
+\citealt{Liow2004,DonohueDim}
+\hl{).}
 The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
@@
 In practice, however, this is hampered by existing software implementations.
-Although many software packages exist for multidimensional analysis \citep[e.g.][]{bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric}, package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric and then enforce this on users by allowing little to no flexibility.
+Although many software packages exist for multidimensional analysis \citep[e.g.][]{navarro2003MDA,bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric},
+\hl{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best it their needs (i.e.
+data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.}
 For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric
@@
 
-\noindent These two groups can be defined as follows (where the integers are the row number of the species in the matrix):
+\noindent These two groups can be defined using one the package's utility functions, \texttt{crown.stem} that selects :
 
-\noindent \texttt{> mammal\_groups <- list(\textquotedbl crown\textquotedbl = c(16, 19:41, 45:50), \textquotedbl stem\textquotedbl = c(1:15, 17:18, 42:44))}
+\noindent \texttt{> mammal\_groups <- crown.stem(BeckLee\_tree)}
 
@@
 
-\noindent \texttt{> subsets <- time.subsets(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
-\noindent \texttt{> bootstraps <- boot.matrix(subsets)}\\
-\noindent \texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\
+\noindent \texttt{> \hl{time\_subsets} <- time.subsets(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
+\noindent \texttt{> \hl{bootstraps2} <- boot.matrix(\hl{time\_subsets})}\\
+\noindent \texttt{> \hl{disparity2} <- dispRity(\hl{bootstraps2}, metric = c(sum, variances))}\\
 
@@
 
-\noindent \texttt{> summary(disparity)}
+\noindent \texttt{> summary(\hl{disparity2})}
 
@@
 
-\noindent \texttt{> plot(disparity, type = \textquotedbl continuous\textquotedbl)}
+\noindent \texttt{> plot(\hl{disparity2}, type = \textquotedbl continuous\textquotedbl)}
 
@@
 
-\noindent \texttt{> anova(test.dispRity(disparity, test = lm, comparisons = \textquotedbl all\textquotedbl))}
+\noindent \texttt{> anova(test.dispRity(disparity2, test = lm, comparisons = \textquotedbl all\textquotedbl))}
 

### 2017-11-20 paper/response_to_reviewers.tex
Updated reviewers comments

@@
 Furthermore, the manuscript is lacking a table summarising the metrics of disparity that this package can work with (which I suspect it was supposed to have).}}
+\label{define_disparity}
 
@@
 This requires the definition of a proxy for space occupancy: the disparity metric (or index; Hopkins and Gerber, 2017) which can be measured in a multitude of ways.
-For example, one could use a metric based on the variance or the range of each axis of space (Wills, 2001; Ciampaglio et al., 2001) , a distance (e.g.
-Euclidean) measured between observations (Foote, 1993, 1996) , a more direct approximation of the hyper volume (Cornwell et al., 2006; Donohue et al., 2013) , or many more (e.g.
+For example, one could use a metric based on the variance or the range of each axis of space (Wills, 2001; Ciampaglio et al., 2001), a distance (e.g.
+Euclidean) measured between observations (Foote, 1993, 1996), a more direct approximation of the hyper volume (Cornwell et al., 2006; Donohue et al., 2013), or many more (e.g.
 Navarro, 2003).} lines @@@
@@
 \item{\textcolor{blue}{Line 84 – Explain what the Procrustes variance is. And when is it appropriate and when is not appropriate? Be more specific.}}
+\label{tone_down}
 
@@
 % Allow to use NPMANOVA instead
+%
+% Dissimilarity analyses: ANOVA using dissimilarities, ANOSIM, MRPP, BIOENV, Mantel and partial Mantel tests.
+% Ordination and environment: vector fitting, centroid fitting and smooth surface fitting, adding species scores as weighted averages, adding convex hull, SD ellipses, arrows etc. to ordination.
+% Ordination: support and meta functions for NMDS, redundancy analysis, constrained correspondence analysis, constrained analysis of proximities (all three with partial analysis),
 
@@
 Cleaning up some of these details now will make it easier to remain dedicated later.}}
+
+%TODO: add [[1]] and [[2]] names
+
+I have changed the \texttt{test.dispRity} to now output the name of the statistic and the parameter in the output list (e.g.
+\texttt{statistic: t} and \texttt{parameter: df} for the example provided by the reviewer).
+I have also highlighted in the manuscript one of the features of the function allowing to output the results in a format more familiar to the user using the option \texttt{details = TRUE} (returning the raw test results):
+
+\textit{Note that by default the function only outputs the test's statistic, parameter (if parametric) and the p value.
+However, the raw test results can also be output using the option} \texttt{details = TRUE} \textit{in the function above.} lines @@@
 
@@
 There are frequent references to ecology and related disciplines where similar approaches are used.
-Yet, those in these broader fields will struggle to follow what is presented as the manuscript exclusively uses "morphospaces" and related terminology that will only be relevant to a specific subset of readers.}
-
-\textcolor{blue}{If you want to aim for a more general audience, consider rewriting the manuscript in more general terminology; samples in rows, variables in columns, for example.
+Yet, those in these broader fields will struggle to follow what is presented as the manuscript exclusively uses "morphospaces" and related terminology that will only be relevant to a specific subset of readers.If you want to aim for a more general audience, consider rewriting the manuscript in more general terminology; samples in rows, variables in columns, for example.
 Perhaps include a table to relate terminology used in morphometry with that more commonly encountered in community ecology.
@@
 
-% %TG: adding a table of terms correspondance
+I have worked towards making the manuscript more useful for non-morphologists.
+I have generalised the input description as suggest by this reviewer, minor comment \ref{rows}; added a glossary table and a brief example using ecological data.
+%TODO: add Diaz's example?
 % %TG: adding an ecology example (removing manual and vignette description)
@@
 
-See my response to Reviewer 2 - Minor suggestion \ref{PCO_cites}.
-
-% %TG: source orginal papers
+I thank this reviewer for sharing the excellent Legendre \& Legendre reference (that was definitely a reference I was missing) and I've accurately credited the statistical methods throughout the manuscript - see my response to Reviewer 2, minor suggestion \ref{PCO_cites}.
 
@@
 
+I apologise for the unfounded critique and I have toned down this sentence (see Reviewer 1, minor suggestion \ref{tone_down}).
+% Additionally, I have added more functionalities using vegan and geiger %TODO!
 % %TG: Tone down the critics to other packages
@@
 \label{nebulous}
+
+I have tried to define disparity with more examples and references to be more familiar to a broader range of readers (See Reviewer 1, major suggestion \ref{define_disparity}).
 
@@
 
+%TOD: add a note to the paper and the package (warnings if negative eigen values).
 %TG: negative egein values explnations
@@
 I've changed the sentence to:
+\label{rows}
 
@@
 
-%TODO
+I went with Legendre \& Legendre 1998's definition of ordinations (pp.
+388) in the three major distinct categories: (1) principal components analysis (PCA), (2), principal coordinates and metric scaling (PCO, PCoA) and non metric dimensional scaling (NMDS, MDS), ignoring correspondence and factor analysis for clarity.
+See reviewer 2, minor suggestion \ref{PCO_cites}.
 
@@
 
-%TODO I agree with the last point of this comment (that comparing CI's overlap is a terrible method!) -> change to "has been done in variable ways, sometimes leading to inaccuracies,"
+I have changed this sentence by citing Anderson 2001 \textit{Austral Ecol.} and Manly 1997 \textit{Randomization, bootstrap and Monte Carlo methods in biology}.
+I also totally agree with the poor choice of simply comparing CIs; unfortunately this method is not that uncommon common in palaeobiology.
+I've removed the citation (to avoid pointing-and-shaming) and added a caveat to this way of comparing disparity:
+
+\textit{[...] an equal variety of statistical tests such as non-parametric multivariate analyses of variance (NPMANOVA, Anderson 2001 ; e.g.
+in Brusatte et al.
+2008) multidimensional permutation tests (Manly 1997 ; e.g.
+in Díaz et al.
+2016) or even simply by looking at the confidence interval overlaps between disparity measurements (which, when not based on rigorous statistic methods can often leads to serious inaccuracy).}
 

### 2017-11-20 paper/dispRity_paper_highlighted.tex
Updated reviewers comments

@@
 \hl{, or many more }
-\citep[e.g.][]{Navarrro2003MDA}.
+\citep[e.g.][]{navarro2003MDA}.
 
-Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance \citep[e.g.][]{Brusatte12092008}, multidimensional permutation tests \citep[e.g.][]{diaz2016global} or even simply by looking at the confidence interval overlaps between disparity measurements \citep[e.g.][]{halliday2016eutherian}.
+Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance (
+\hl{NPMANOVA, }
+\citealt{NPMANOVA}
+\hl{; e.g. in }
+\citealt{Brusatte12092008}
+\hl{)}
+multidimensional permutation tests
+\hl{(}
+\citealt{ManlyPermutations}
+\hl{; e.g. in }
+\citealt{diaz2016global}
+\hl{)}
+or even simply by looking at the confidence interval overlaps between disparity measurements
+\hl{().}
 In summary, there are many different ways to perform each step of a multidimensional analysis, making analyses of complexity ever more complex.
@@
 (see Manuals and Vignettes below).
-Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow} and Table \ref{Tab:main_fun}).
+Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow}).% and Table \ref{Tab:main_fun}).
 \hl{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements} \citep[e.g.
@@
 
+% \begin{table}
+%     \begin{tabular}{ll}
+%         \hline
+%         Function & Description \\ 
+%         \hline
+%         \texttt{geomorph.ordination} & Imports \hl{shape} data from the \texttt{geomorph} package. \\
+%         \texttt{Claddis.ordination} & Imports \hl{discrete morphological} data from the \texttt{Claddis} package. \\
+%         \texttt{custom.subsets} & Separates data into custom subsets. \\
+%         \texttt{time.subsets} & Separates data into time subsets. \\
+%         \texttt{boot.matrix} & Bootstraps and rarefies a matrix or a \disp object. \\
+%         \disp & Calculates disparity from an matrix or a \disp object. \\
+%         \texttt{summary.dispRity} & Summarises a \disp object. \\
+%         \texttt{plot.dispRity} & Plots a \disp object. \\
+%         \texttt{test.dispRity} & Applies statistical tests to a \disp object.\\
+%         \texttt{dispRity.per.group} & Pipeline for \hl{disparity-within-groups} analysis. \\
+%         \texttt{dispRity.through.time} & Pipeline for disparity-through-time analysis. \\
+%         \hline
+%     \end{tabular}
+%     \caption{Main functions in the \disp package.}
+%     \label{Tab:main_fun}
+% \end{table}
+
+\begin{landscape}
 \begin{table}
-    \begin{tabular}{ll}
+    \begin{tabular}{p{4cm}p{5cm}p{6cm}p{6cm}}
         \hline
-        Function & Description \\ 
+        in this manuscript & in \texttt{dispRity} & in palaeobiology & in ecology \\ 
         \hline
-        \texttt{geomorph.ordination} & Imports \hl{shape} data from the \texttt{geomorph} package. \\
-        \texttt{Claddis.ordination} & Imports \hl{discrete morphological} data from the \texttt{Claddis} package. \\
-        \texttt{custom.subsets} & Separates data into custom subsets. \\
-        \texttt{time.subsets} & Separates data into time subsets. \\
-        \texttt{boot.matrix} & Bootstraps and rarefies a matrix or a \disp object. \\
-        \disp & Calculates disparity from an matrix or a \disp object. \\
-        \texttt{summary.dispRity} & Summarises a \disp object. \\
-        \texttt{plot.dispRity} & Plots a \disp object. \\
-        \texttt{test.dispRity} & Applies statistical tests to a \disp object.\\
-        \texttt{dispRity.per.group} & Pipeline for \hl{disparity-within-groups} analysis. \\
-        \texttt{dispRity.through.time} & Pipeline for disparity-through-time analysis. \\
+        multidimensional space & matrix ($n\times k$) & morphospace, traitspace, etc. & ecospace, function-space, etc. \\
+        elements & rows ($n$) & taxa, specimen, etc. & taxa, field sites, environments, etc. \\
+        dimensions & columns ($k$) & ordination scores, distances, etc. & ordination scores, distances, etc. \\
+        subsets & matrix ($m \times k$, with $m \leq n$) & e.g. every element in an strata or sharing the same ancestor & e.g. elements living in the same environment \\
+        disparity & a dimension-level 1 or 2 function$^1$ & disparity: e.g.
+the sum of variances \cite{Wills2001}, the average pairwise distance between taxa \cite{Foote01071994}, etc.
+& dissimilarity: e.g.
+ellipsoid volume \cite{DonohueDim}, convex hull volume \cite{cornwell2006trait}, etc.
+\\
         \hline
     \end{tabular}
-    \caption{Main functions in the \disp package.}
-    \label{Tab:main_fun}
+    \caption{Glossary and equivalences between this manuscript, the \texttt{dispRity} package and terms commonly used in palaeobiology or ecology. $^1$See Fig. \ref{Fig:levels}.}
+    \label{Tab:terms}
 \end{table}
+\end{landscape}
 
@@
 $^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA).}}
-    \label{Tab:metrics}
+    % \label{Tab:metrics}
 \end{table}
@@
 $^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA).
+\label{Tab:metrics}
 
@@
 \noindent \texttt{[[1]]}\\
-\noindent \texttt{             statistic}\\
+\noindent \texttt{          statistic: W}\\
 \noindent \texttt{crown : stem     10000}\\
@@
 \bigskip
-As indicated by the p value, there is a significant difference in disparity between the groups.
+As indicated by the \textit{p} value, there is a significant difference in disparity between the groups.
+\hl{Note that by default the function only outputs the test's statistic, parameter (if parametric) and the \textit{p} value.
+However, the raw test results can also be output using the option \texttt{details = TRUE} in the function above.}
 Relating back to our question: yes, crown mammals display a higher diversity in morphologies than their stem counterparts (in this example and dataset!).

### 2017-11-21 paper/response_to_reviewers.tex
response update

@@
 
-As Reviewer 4 (major comment \ref{nebulous}) also underlines: ``disparity is a broad, nebulous concept''.
+As reviewer 4 (major comment \ref{nebulous}) also underlines: ``disparity is a broad, nebulous concept''.
 This observation was the genesis for this package and I am currently working on a collaborative review to define the borders (or the absence thereof) of this concept.
@@
 Perhaps adding an additional multipanel figure exhibitng some simulated spaces.}}
+\label{simulations}
 
@@
 I agree, there are many more metrics commonly used for disparity analysis not implemented in this package.
-This was a conscious choice: due to the really high number of metrics available (especially when combining different ones), I designed the package to be modular hence allowing users to implement their own metrics (see response to Reviewer 1 minor comment \ref{table_missing}).
+This was a conscious choice: due to the really high number of metrics available (especially when combining different ones), I designed the package to be modular hence allowing users to implement their own metrics (see response to reviewer 1 minor comment \ref{table_missing}).
 I have precise this using the hypercube volume example in the manuscript:
@@
 
-See Reviewer 1 minor comment \ref{abstract_typo}.
+See reviewer 1 minor comment \ref{abstract_typo}.
 
@@
 
-Sincerest apologies! See Reviewer 1 minor comment \ref{table_missing}.
+Sincerest apologies! See reviewer 1 minor comment \ref{table_missing}.
 
@@
 
-I've removed this last sentence (see Reviewer 1 minor comment \ref{remove_last_sentence}).
+I've removed this last sentence (see reviewer 1 minor comment \ref{remove_last_sentence}).
 
@@
 
-I thank this reviewer for sharing the excellent Legendre \& Legendre reference (that was definitely a reference I was missing) and I've accurately credited the statistical methods throughout the manuscript - see my response to Reviewer 2, minor suggestion \ref{PCO_cites}.
+I thank this reviewer for sharing the excellent Legendre \& Legendre reference (that was definitely a reference I was missing) and I've accurately credited the statistical methods throughout the manuscript - see my response to reviewer 2, minor suggestion \ref{PCO_cites}.
 
@@
 
-I apologise for the unfounded critique and I have toned down this sentence (see Reviewer 1, minor suggestion \ref{tone_down}).
+I apologise for the unfounded critique and I have toned down this sentence (see reviewer 1, minor suggestion \ref{tone_down}).
 % Additionally, I have added more functionalities using vegan and geiger %TODO!
@@
 
-I have tried to define disparity with more examples and references to be more familiar to a broader range of readers (See Reviewer 1, major suggestion \ref{define_disparity}).
+I have tried to define disparity with more examples and references to be more familiar to a broader range of readers (See reviewer 1, major suggestion \ref{define_disparity}).
 
@@
 
-I've fixed these typos (see also Reviewer 2 minor comment \ref{altitude}).
+I've fixed these typos (see also reviewer 2 minor comment \ref{altitude}).
 
@@
 The solution is a confidence ellipse, which is often quite different to the interval implied by separate confidence interval bars for each axis.}}
+
+This metric (the sum of each ordination axis variance) is commonly used in palaeobiology (since Foote 1991 and Wills et al.
+1994) probably due to a ``citation diffusion effect'' (i.e.
+future authors replicating convincing methods) and partly due to its pretty good behaviour with palaeontological data (e.g.
+missing data, Ciampaglio 2001).
+However, as this reviewer high
 
@@
 
-%TODO cut down the manual and vignette part + simulation part: just link to it?
+I have removed the simulations part as suggested by this reviewer.
+However, following the encouraging suggestions of reviewer 2 (major suggestion \ref{simulations}) I have developed in much more details the simulation algorithms in the package manual.
 

### 2017-11-21 paper/dispRity_paper_highlighted.tex
response update

@@
 In this example, the multidimensional space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
-Additionally, we can define disparity as the sum of the variances on each dimension \citep{Wills1994} that will represent an aspect of the the volume of the morphospace (i.e.
+Additionally, we can define disparity as the sum of the variances on each dimension \citep{foote1991morphological,Wills1994} that will represent an aspect of the the volume of the morphospace (i.e.
 a large sum of variances approximates a large morphospace and vice versa).

### 2017-12-05 paper/dispRity_paper_highlighted.tex
Updated response with new changed models

@@
     \item \hl{``proximity''} where the data chosen along the branch is either the descendant or the ancestor depending on branch length
-    \item \hl{``punctuated'' where the data chosen along the branch is both the descendant and the ancestor with an even probability}
-    \item \hl{``gradual'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length}
+    \item \hl{``equal.split'' where the data chosen along the branch is both the descendant and the ancestor with an even probability}
+    \item \hl{``gradual.split'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length}
 \end{itemize}
 
-\hl{Note that the four first models are "fixed": the selected data is always either the one of the descendant or the ancestor.
-The two last models are "probabilistic": the data from both the descendant and the ancestor is used with an associate probability.
+\hl{Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
+The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
 These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).

### 2017-12-05 inst/gitbook/03_specific-tutorials.Rmd
updated manual
updated gitbook

@@
 
+ * **Punctuated models**
    *  `acctran` where the data chosen along the branch is always the one of the descendant
@@
    *  `proximity` where the data chosen along the branch is either the descendant or the ancestor depending on branch length
- *  `punctuated` where the data chosen along the branch is both the descendant and the ancestor with an even probability
- *  `gradual` where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length
-
-> Note that the four first models are "fixed": the selected data is always either the one of the descendant or the ancestor.
-The two last models are "probabilistic": the data from both the descendant and the ancestor is used with an associate probability.
+
+ * **Gradual models**
+   *  `equal.split` where the data chosen along the branch is both the descendant and the ancestor with an even probability
+   *  `gradual.split` where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length
+
+> Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
+In other words, changes along the branches always occur at either ends of it.
+The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
 These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants.
@@
 
+### NPMANOVA in `dispRity`
+
+One often useful test to apply to multidimensional data is the permutational multivariate analysis of variance based on distance matrices `vegan::adonis`.
+
+```{r}
+library(vegan)
+data(dune)
+data(dune.env)
+
+## Beta diversity distance matrix
+betad <- betadiver(dune, "z")
+
+## Create a pairwise distance matrix
+make.pairwise.dist.matrix <- function(matrix) {
+    if(class(matrix) == "matrix" || class(matrix) == "data.frame") {
+        ## Simply calculate the distances
+        return(dist(matrix))
+    } else {
+        ## Transform the distances into a matrix
+        pair_distances <- matrix
+        pairwise_dist <- matrix(NA, nrow = nrow(matrix), ncol = nrow(matrix))
+        colnames(pairwise_dist) <- rownames(pairwise_dist) <- rownames(matrix)
+        diag(pairwise_dist) <- 0
+        pairwise_dist[lower.tri(pairwise_dist)] <- pair_distances
+        pairwise_dist <- as.dist(pairwise_dist)
+    }
+}
+pairwise_dist <- make.pairwise.dist.matrix(dune)
+
+## Create the factor group
+factors <- data.frame(factor = dune.env$Management)
+
+## Testing adonis
+test1 <- adonis(betad ~ factor, factors)
+test2 <- adonis(pairwise_dist ~ factor, factors)
+
+set.seed(1)
+## Check if it works for mammals data set
+data(BeckLee_mat50)
+data(BeckLee_tree)
+## Crown stem group
+crown_stem <- crown.stem(BeckLee_tree, inc.nodes = FALSE)
+## Dummy group
+group_sample <- sample(1:50, 25)
+group1 <- rownames(BeckLee_mat50)[group_sample]
+group2 <- rownames(BeckLee_mat50)[-group_sample]
+dummy_group <- list(group1 = group1, group2 = group2)
+## Equal groups
+equal_group <- list(groupA = rownames(BeckLee_mat50), groupB = rownames(BeckLee_mat50))
+
+## Create the subsets
+subset_crown_stem <- custom.subsets(BeckLee_mat50, group = crown_stem)
+subset_dummy_group <- custom.subsets(BeckLee_mat50, group = dummy_group)
+subset_equal_group <- custom.subsets(BeckLee_mat50, group = equal_group)
+
+## Calculate disparity
+disp_crown_stem <- dispRity(subset_crown_stem, metric = c(median, centroids))
+disp_dummy_group <- dispRity(subset_dummy_group, metric = c(median, centroids))
+disp_equal_group <- dispRity(subset_equal_group, metric = c(median, centroids))
+
+summary(disp_crown_stem)
+summary(disp_dummy_group)
+summary(disp_equal_group)
+
+## Function for creating the factors table
+make.factors <- function(data) {
+    ## Extracting the factors
+    factors <- names(sort(unlist(lapply(data$subsets, function(X) return(X$elements)))))
+    ## Trimming the factors names
+    group_names <- names(data$subsets)
+    for(group in 1:length(group_names)) {
+        factors[grep(group_names[group], factors)] <- group_names[group]
+    }
+    return(data.frame(factor = factors, row.names = rownames(data$matrix)))
+}
+
+adonis.dispRity <- function(data, ...) {
+  ## Create the factor table
+  factors <- make.factors(data)
+
+  ## Run adonis
+  return(adonis(make.pairwise.dist.matrix(data$matrix)~factor, factors, ...))
+}
+
+```
+
 ## Disparity as a distribution

### 2017-12-05 inst/gitbook/04_making-up-data.Rmd
updated gitbook

@@
 
-The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;OReilly2017].
+The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;@OReilly2017].
 

### 2017-12-05 inst/gitbook/05_guts.Rmd
updated gitbook

@@
 
- * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"punctuated"`, `"gradual"`).
+ * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"equal.split"`, `"gradual.split"`).
 This element generated only once via `time.subsets()` and `custom.subsets()`.

### 2017-12-19 paper/dispRity_paper_highlighted.tex
Updated paper

@@
     \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
-        name & description & dim level & definition & source \\
+        name & description & dim & definition & source \\
         \hline
@@
         \texttt{centroids} & The distance between each element and a fixed point$^1$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & This package \\
-        \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
-        \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
+        \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
+        \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
         \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & This package \\
-        \texttt{ellipse. volume}$^2$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
+        \texttt{ellipse.
+volume}$^2$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & This package; based on \cite{DonohueDim}\\
         \texttt{mode.val} & The modal value & 1 & NA & This package\\
         \texttt{n.ball. volume} & The hyper-spherical ($n$-ball) volume & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R$ & This package\\
-        \texttt{pairwise. dist} & The pairwise distances between elements & 2 & NA & \texttt{stats::dist}\\        
+        \texttt{pairwise. dist} & The pairwise distances between elements & 2 & NA & \texttt{vegan::vegdist} \citep{oksanen2007vegan}\\        
         \texttt{radius} & The radius of each dimensions & 2 & $|\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|$ & This package\\
         \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & This package \\
-        \texttt{span.tree. length} & The minimal spanning tree length & 1 & $\sum(\mathrm{branch\ length})$ & \texttt{vegan::spantree}\\        
+        \texttt{span.tree. length} & The minimal spanning tree length & 1 & $\sum(\mathrm{branch\ length})$ & \texttt{vegan::spantree} \citep{oksanen2007vegan}\\
         \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & This package \\
@@
 
-Some tests are implemented within the package such as the Bhattacharrya Coefficient \citep[\texttt{bhatt.coeff};][]{Bhattacharyya,GuillermeCooper} or a permutation test based on null hypothesised multidimensional space following \cite{diaz2016global} (\texttt{null.test}).
+Some tests are implemented within the package such as the Bhattacharrya Coefficient \citep[\texttt{bhatt.coeff};][]{Bhattacharyya,GuillermeCooper}, a permutation test based on null hypothesised multidimensional space following \cite{diaz2016global} (\texttt{null.test})
+\hl{as well as a wrapper for the \texttt{vegan::adonis}}
+ \citep{oksanen2007vegan}
+\hl{and \texttt{geiger::dtt}}
+ \citep{geiger2008}
+\hl{functions (respectively \texttt{adonis.dispRity} and \texttt{dtt.dispRity} - allowing \texttt{dispRity} objects to be directly used with any disparity metric).}
 Additionally, this function also allows additional arguments such as \texttt{rarefaction} (as described above) or \texttt{correction} to adjust p-values when using multiple parametric tests.
@@
 From the plot and the summary, disparity seems to change (increase) through time from the Cretaceous to the Eocene.
-Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, an ANOVA (Table \ref{Tab:anova}):
+Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA \citep[PERMANOVA; ][]{permanova} through the \texttt{adonis.disprity} function that is a wrapper of the \texttt{vegan::adonis} function \citep{oksanen2007vegan} for \disp objects:
 
-\noindent \texttt{> anova(test.dispRity(disparity2, test = lm, comparisons = \textquotedbl all\textquotedbl))}
+\noindent \texttt{> test.dispRity(disparity2, test = adonis.dispRity)}
 
@@
 \centering
-\begin{tabular}{lrrrrr}
+\begin{tabular}{lrrrrrr}
   \hline
- & Df & Sum Sq & Mean Sq & F value & Pr($>$F) \\ 
+              & Df & Sum Sq & Mean Sq & F Model & R2   & Pr($>$F) \\ 
   \hline
-subsets & 2 & 3.14 & 1.57 & 177.84 & 0.0000 \\ 
-  Residuals & 297 & 2.62 & 0.01 &  &  \\ 
+    time      & 2  &  7.50 & 3.75     & 2.06    & 0.11 & <0.01 *** \\ 
+    Residuals & 34 & 61.82 & 1.81     &         & 0.89 & \\ 
+    Total     & 36 & 69.32 &          &         & 1.00 & \\ 
    \hline
 \end{tabular}
-\caption{ANOVA performed on a \disp object.}
+\caption{Permanova output from \texttt{adonis.dispRity}:\\
+Call: vegan::adonis(formula = dist(matrix) \texttildelow time, data = disparity2, method = "euclidean"); Permutation: free; Number of permutations: 999; Terms added sequentially (first to last).
+Signif.
+codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1 ` ' 1.}
 \label{Tab:anova}
@@
 Many thanks to Natalie Cooper for encouraging the development of this package and helping with the writing of this paper and the package manuals.
-Thanks to Martin Brazeau and Andrew Jackson for specific implementation ideas 
-\hl{and to Michael Collyer, Gavin Simpson and two other anonymous reviewers for constructive comments on the earlier version of this manuscript}. %TG: add Graeme Lloyd, Dave Bapst and Emma Sherratt
+Thanks to David Bapst, Martin Brazeau, Andrew Jackson, Graeme Lloyd and Emma Sherratt for specific implementation ideas 
+\hl{and to Michael Collyer, Gavin Simpson and two other anonymous reviewers for constructive and encouraging comments on the manuscript}. %TG: add Graeme Lloyd, Dave Bapst and Emma Sherratt
 I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau 

### 2017-12-20 inst/gitbook/07_ecology-demo.Rmd
v0.5 gitbook

@@
 output: rmarkdown::html_vignette
-bibliography: References.bib
+bibliography: [References.bib, packages.bib]
 bst: sysbio.bst
@@
 ```{r}
-## Fitting a linear model to our data
-disparity_lm <- test.dispRity(petal_disparity, test = lm, comparisons = "all")
-
-## Testing whether there is a difference in disparity between the different species
-summary(aov(disparity_lm))
+## Running a PERMANOVA
+test.dispRity(petal_disparity, test = adonis.dispRity)
 

### 2017-12-20 R/test.dispRity.R

@@ -210,7 +210,7 @@ test.dispRity <- function(data, test, comparisons = "pairwise", rarefaction = NU
 
     ## correction
     check.method(correction, c("holm", "hochberg", "hommel", "bonferroni", "BH", "BY", "fdr", "none"), "Correction methods")
-    if(!is.null(data$call$bootstrap) && correction == "none" && length(data$subsets) > 2) {
+    if(!is.null(data$call$bootstrap) && correction == "none" && length(data$subsets) > 2 && match_call$test != "adonis.dispRity") {
         warning("Multiple p-values will be calculated without adjustment!\nThis will inflate Type I error!")
     }
 

### 2017-12-20 inst/gitbook/01_glossary.Rmd

@@
 date: "`r Sys.Date()`"
+bibliography: [References.bib, packages.bib]
 output:
@@
 
--   **Multidimensional space**. The mathematical multidimensional object that will be analysed with this package.
+-   **Multidimensional space** (or just space). The mathematical multidimensional object that will be analysed with this package.
     In morphometrics, this is often referred to as the morphospace.
@@
     For example, if our space is composed of birds and mammals (the elements) and 50 principal components of variation (the dimensions), we can create two subsets containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
+
+-   **Disparity**.
+A metric expressing the similarities/dissimilarities of the elements within the space or a summarising the space dimensions.
+For example the pairwise distances between elements or the range of each dimensions.
+
+## Glossary equivalences in palaeobiology and ecology
+
+In this manual | In `dispRity` | E.g. in palaeobiology | E.g. in ecology
+---------------|---------------|------------------|-------------
+the multidimensional space | a `matrix` object ($n\times k$) | a morphospace | a function-space
+elements | rows ($n$) | taxa | field experiments
+dimensions | columns ($k$) | morphological characters | communities' compositions
+subsets | a `matrix` ($m \times k$, with $m \leq n$) | time series | experimental treatments
+disparity | a `function` | sum of variances | ellipsoid volume

### 2017-12-20 inst/gitbook/02_getting-started.Rmd
updated gitbook

@@
 
-Disparity can be estimated from pretty much any matrix.
-Classically, however, it is measured from ordinated matrices.
-These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).
+**Any matrix object in `R`.**
+Disparity can be estimated from pretty much any matrix as long as rows represent the elements and columns the dimensions.
+These matrices can be observations, pairwise differences between elements, ordinations, etc...
+
+## Ordinated matrices
+
+Classically, when a high number of variables is used, disparity is calculated from ordinated matrices.
+These can be any type of ordinations (PCO, PCA, PCoA, MDS, etc.) as long as elements are the rows (taxa, countries, field experiments) and the dimensions are the columns.
 
@@
 ## Creating the two groups (crown versus stem) as a list
-mammal_groups <- crown.stem(BeckLee_tree, inc.nodes = TRUE)
+mammal_groups <- crown.stem(BeckLee_tree, inc.nodes = FALSE)
 

### 2017-12-20 paper/response_to_reviewers.tex
Minor fixes

@@
 
-I am very grateful to the four referees for their helpful and constructive comments, which we believe have helped us to significantly improve the \texttt{dispRity} package and its associated description paper.
-I have taken all of their comments on board, and respond to their points below.
+I am very grateful to the four referees for their helpful and constructive comments, which I believe have helped me to significantly improve the \texttt{dispRity} package and this associated package description paper.
+I have taken all of their comments on board, and respond to their points in details below.
 For improving clarity in this document, reviewers comments are displayed in blue, my responses in normal text and the changes to the manuscript in italic.
@@
 I.e. no additional data should be added to the space \textit{a posteriori}.
-In practice, the \textit{time.subsamples} function can only select between two information: the ancestral node (DELTRAN) and the descendant node/tip (ACCTRAN) along the edge since only the two of them are real objects in the morphospace.
+In practice, the \textit{time.subset} function can only select between two information: the ancestral node (DELTRAN) and the descendant node/tip (ACCTRAN) along the edge since only the two of them are real objects in the morphospace.
 It is therefore not possible to create an ancestral states reconstruction along the edge after running the ordination.
@@
 I have modified the name of the two previous ``punctuated'' and ``gradual'' algorithms to ``random'' and ``proximity'' to reflect more accurately how they select either the ACCTRAN or DELTRAN information.
-I've added two more accurate models (now named ``punctuated'' and ``gradual'') to reflect the changes along the branches.
+I've added two more accurate models (now named ``equal.split'' and ``gradual.split'') to reflect the changes along the branches.
 These two algorithm now allows to use both the ACCTRAN and DELTRAN information with a linked probability to use either of them per bootstrap draw.
-For example, the ``punctuated'' model will have a 0.5 probability of using either node information for each bootstrap draw; the ``gradual'' model transforms the branch length were the slice occurs into a probability of drawing either of the nodes information (e.g.
+For example, the ``equal.split'' model will have a 0.5 probability of using either node information for each bootstrap draw; the ``gradual.split'' model transforms the branch length were the slice occurs into a probability of drawing either of the nodes information (e.g.
 if the slices occurs 3/4 along the branch from A to B, the model will draw the information from A and B with a probability of respectively 0.25 and 0.75).
@@
     \item \textit{``proximity'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length}
-    \item \textit{``punctuated'' where the data chosen along the branch is both the descendant and the ancestor with an even probability}
-    \item \textit{``gradual'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length}
+    \item \textit{``equal.split'' where the data chosen along the branch is both the descendant and the ancestor with an even probability}
+    \item \textit{``gradual.split'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length}
 \end{itemize}
 
-\textit{Note that the four first models are "fixed": the selected data is always either the one of the descendant or the ancestor.
-The two last models are "probabilistic": the data from both the descendant and the ancestor is used with an associate probability.
-These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).} lines @@@
+\textit{Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
+The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
+These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).
+Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).} lines @@@
 
@@
 
-%TODO: add [[1]] and [[2]] names
-
 I have changed the \texttt{test.dispRity} to now output the name of the statistic and the parameter in the output list (e.g.

### 2017-12-20 paper/dispRity_paper_highlighted.tex
Minor fixes

@@
         It is then possible to \hl{measure divers aspects of the distribution of some observation (e.g. species) in this space}.
-        For example, if studying morphology, one can create a morphospace for two groups of species, measure \hl{the multidimensional volume occupied by each of these groups and then test whether these two volumes are significantly different or not}.
+        For example, if studying morphology, one can create a morphospace for two groups of species, measure \hl{the volume occupied by each of these groups and then test whether these two volumes are significantly different or not}.
 
@@
 One solution to this problem is to analyse these data in a multivariate framework \citep[e.g.][]{price2015predation,diaz2016global}.
-Such analyses aim to capture the complex multidimensionality of biological data, while still providing outputs that are interpretable in our physical world restricted to three dimensions.
-These multivariate analyses can be used to investigate changes in morphological diversity through time \citep{Close2015}, competitive replacement scenarios \citep{Brusatte12092008}, relationships among form and function \citep{diaz2016global} and even to describe the entirety of possible shapes for a group of organisms \citep{raup1966geometric}.
+Such analyses aim to capture the complex multidimensionality of biological data, while still providing outputs that are interpretable.
+These multivariate analyses can be used to investigate changes in morphological diversity through time \citep[e.g][]{Close2015}, competitive replacement scenarios \citep[e.g.][]{Brusatte12092008}, relationships among form and function \citep[e.g.][]{diaz2016global} and even to describe the entirety of possible shapes for a group of organisms \citep[e.g.][]{raup1966geometric}.
 The biological variables in such analyses are equally diverse, including morphological traits (discrete traits like the presence or absence of a character, e.g.
-\citealt{brusattedinosaur2012}; and continuous traits such as lengths, e.g.
+\citealt{Close2015}; and continuous traits such as lengths, e.g.
 \citealt{price2015predation}), life history traits \citep[e.g.][]{diaz2016global}, or even ecosystem properties \citep[e.g.][]{DonohueDim}.
@@
 The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
-For example, when using morphological traits the multivariate space will be a morphospace, when using ecological traits it may be referred to as an ecospace or trait space.
+For example, when using morphological traits the space will be a morphospace, when using ecological traits it may be referred to as an ecospace, etc.
 
@@
 \hl{)}
-or even simply by looking at the confidence interval overlaps between disparity measurements
+or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements
 \hl{().}
@@
 % Problems with this variety
-In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods \citep[such as][for biogeography]{matzke2013biogeobears}.
+In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods.
 In practice, however, this is hampered by existing software implementations.
-Although many software packages exist for multidimensional analysis \citep[e.g.][]{navarro2003MDA,bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric},
+Although many software packages exist for multidimensional analysis \citep[e.g.][]{navarro2003MDA,bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,geiger2008,adams2013geomorph,Claddis,adams2017geometric},
 \hl{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best it their needs (i.e.
@@
 % Here comes the solution!
-The aim of the \disp package is to solve such problems by providing a completely flexible framework for studying multidimensional data.
+The aim of the \disp package is avoid such problems by providing a completely flexible framework for studying multidimensional data.
 This package is based on a highly modular architecture where each crucial decision in multidimensional analysis (which data, which metric and which test) can be provided by the user.
@@
 In brief, the package takes a \texttt{matrix} object (the multidimensional space), 
-\hl{calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation}
-(see Manuals and Vignettes below).
-Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow}).% and Table \ref{Tab:main_fun}).
+\hl{calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation}.
+Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow}).
 \hl{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements} \citep[e.g.
@@
 \end{figure}
-
-% \begin{table}
-%     \begin{tabular}{ll}
-%         \hline
-%         Function & Description \\ 
-%         \hline
-%         \texttt{geomorph.ordination} & Imports \hl{shape} data from the \texttt{geomorph} package. \\
-%         \texttt{Claddis.ordination} & Imports \hl{discrete morphological} data from the \texttt{Claddis} package. \\
-%         \texttt{custom.subsets} & Separates data into custom subsets. \\
-%         \texttt{time.subsets} & Separates data into time subsets. \\
-%         \texttt{boot.matrix} & Bootstraps and rarefies a matrix or a \disp object. \\
-%         \disp & Calculates disparity from an matrix or a \disp object. \\
-%         \texttt{summary.dispRity} & Summarises a \disp object. \\
-%         \texttt{plot.dispRity} & Plots a \disp object. \\
-%         \texttt{test.dispRity} & Applies statistical tests to a \disp object.\\
-%         \texttt{dispRity.per.group} & Pipeline for \hl{disparity-within-groups} analysis. \\
-%         \texttt{dispRity.through.time} & Pipeline for disparity-through-time analysis. \\
-%         \hline
-%     \end{tabular}
-%     \caption{Main functions in the \disp package.}
-%     \label{Tab:main_fun}
-% \end{table}
 
@@
 it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid).
-$^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA).
+$^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO (\textit{not} PCA).
 \label{Tab:metrics}
@@
 The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
-These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).
-Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).}
+There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).
+See Guillerme \& Cooper \textit{in review} for full description.}
 
@@
 The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity} (provided disparity was calculated).
-These results can then be analysed using the \texttt{test.dispRity} function for comparing subsets or testing hypotheses.
+These results can also be analysed using the \texttt{test.dispRity} function for comparing subsets or testing hypotheses.
 
@@
 the rarefaction level).
-The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base R.
+The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base \texttt{R}.
 % Note that if the disparity has been calculated on fully rarefied data (using \texttt{boot.matrix(data, rarefaction = TRUE)} prior to calculating disparity), the rarefaction curves for the whole multidimensional space (or its sub-samples) can be plotted using \texttt{rarefaction = ``plot''}.
@@
 first against second, second against third, etc.); (4) `all'' for comparing all the subsets simultaneously (i.e.
-\texttt{disparity $\mathtt{\sim}$ subsets} - this can be used for \texttt{lm} or \texttt{glm}) or (5) any list of pairs (or more) of subsets to compare.
+\texttt{disparity $\mathtt{\sim}$ subsets} - this can be used for \texttt{lm} class functions) or (5) any list of pairs (or more) of subsets to compare.
 
@@
 
-\subsection{Typical disparity among groups analysis}
+\subsection{Example}
+
+The \texttt{R} script for the following example is available in the supplementary materials.
+
+\subsubsection{Typical disparity among groups analysis}
 One typical question with such analysis would be to test whether two groups of species have a different disparity.
@@
 
-\noindent These two groups can be defined using one the package's utility functions, \texttt{crown.stem} that selects :
+\hl{\noindent These two groups can be defined using one the package's utility functions, \texttt{crown.stem} that separate the crown and stem species given a phylogeny (allowing to ignore the nodes or not):}
 
-\noindent \texttt{> mammal\_groups <- crown.stem(BeckLee\_tree)}
+\noindent
+\hl{\texttt{> mammal\_groups <- crown.stem(BeckLee\_tree, inc.nodes = FALSE)}}
 
@@
 
-\noindent \texttt{> disparity <- dispRity.per.group(data = BeckLee\_mat50, group = mammal\_groups, metric = c(sum, variances))}
+\noindent
+\texttt{> disparity <- dispRity.per.group(data = BeckLee\_mat50,}\\
+\texttt{group = mammal\_groups, metric = c(sum, variances))}
 
@@
 
-\noindent \texttt{> subsets <- custom.subsets(data = BeckLee\_mat50, group = mammal\_groups)}\\
-\noindent \texttt{> bootstraps <- boot.matrix(subsets)}\\
-\noindent \texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\
+\noindent
+\texttt{> subsets <- custom.subsets(data = BeckLee\_mat50, group = mammal\_groups)}\\
+\noindent
+\texttt{> bootstraps <- boot.matrix(subsets)}\\
+\noindent
+\texttt{> \hl{disparity2} <- dispRity(bootstraps, metric = c(sum, variances))}\\
 
@@
 
-\noindent \texttt{> disparity}\\
+\noindent \texttt{> \hl{disparity2}}\\
 \noindent \texttt{ ---- dispRity object ---- }\\
@@
 
-\noindent \texttt{> summary(disparity)}
+\noindent \texttt{> summary(\hl{disparity2)}}
 
@@
 \caption{Summarising a \disp object (disparity per groups).
-$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 95\% are the confidence intervals.}
+$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and
+\hl{97.5}
+\% are the confidence intervals.}
 \label{Tab:summary_group}
@@
 
-\noindent \texttt{> plot(disparity)}
+\noindent \texttt{> plot(\hl{disparity2)}}
 
@@
 
-\noindent \texttt{> test.dispRity(disparity, test = wilcox.test)}\\
+\noindent \texttt{> test.dispRity(\hl{disparity2}, test = wilcox.test)}\\
 
@@
 Using the same definition of the multidimensional space and disparity as in the previous example, we can measure, for example, changes in disparity through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).
+\hl{Note that stratigraphic times can be generated automatically using the \texttt{get.bin.ages} utility function.}
 
-\noindent \texttt{> time\_bins <- c(100.5, 66, 56, 33.9)}
+\noindent
+\texttt{> time\_bins <- c(100.5, 66, 56, 33.9)}
 
@@
 
-\noindent \texttt{> disparity <- dispRity.through.time(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, metric = c(sum, variances))}
+\noindent
+\texttt{> \hl{disparity3} <- dispRity.through.time(data = BeckLee\_mat50,}\\
+\texttt{tree = BeckLee\_tree, time = time\_bins, metric = c(sum, variances))}
 
@@
 
-\noindent \texttt{> \hl{time\_subsets} <- time.subsets(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
+\noindent \texttt{> \hl{time\_subsets} <- time.subsets(data = BeckLee\_mat50, tree = BeckLee\_tree,}\\ \texttt{time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
 \noindent \texttt{> \hl{bootstraps2} <- boot.matrix(\hl{time\_subsets})}\\
-\noindent \texttt{> \hl{disparity2} <- dispRity(\hl{bootstraps2}, metric = c(sum, variances))}\\
+\noindent \texttt{> \hl{disparity4} <- dispRity(\hl{bootstraps2}, metric = c(sum, variances))}\\
 
@@
 
-\noindent \texttt{> summary(\hl{disparity2})}
+\noindent \texttt{> summary(\hl{disparity4})}
 
@@
 
-\noindent \texttt{> plot(\hl{disparity2}, type = \textquotedbl continuous\textquotedbl)}
+\noindent \texttt{> plot(\hl{disparity4}, type = \textquotedbl continuous\textquotedbl)}
 
@@
 
-\noindent \texttt{> test.dispRity(disparity2, test = adonis.dispRity)}
+\noindent \texttt{> test.dispRity(\hl{disparity4, test = adonis.dispRity)}}
 
@@
 \end{tabular}
-\caption{Permanova output from \texttt{adonis.dispRity}:\\
+\caption{
+\hl{Permanova output from \texttt{adonis.dispRity}:\\
 Call: vegan::adonis(formula = dist(matrix) \texttildelow  time, data = disparity2, method = "euclidean"); Permutation: free; Number of permutations: 999; Terms added sequentially (first to last).
@@
 codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1 ` ' 1.}
+}
 \label{Tab:anova}
@@
 To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
+\hl{Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test}.
 
@@
 \subsection{Data simulations}
-This package also contains functions for simulating random discrete morphological matrices (\texttt{sim.morpho}) or random multidimensional spaces (\texttt{space.maker}).
-These functions are based on a similar modular architecture as that used by the \disp functions, allowing users to provide their own distribution parameters for the simulations.
-For example, \texttt{stats::rnorm} can be provided as an argument for drawing normal characters rates with \texttt{sim.morpho} or normally distributed multidimensional space with \texttt{space.maker}.
-The discrete morphological data simulations are based on protocols from \cite{GuillermeCooper}, \cite{OReilly20160081} and \cite{puttick2017uncertain}. 
-The multidimensional space simulations are based on the methods from \cite{diaz2016global}.
-Both functionalities are described in more details in the package manual.
+\hl{This package also contains functions for simulating random discrete morphological matrices (\texttt{sim.morpho}) or random multidimensional spaces (\texttt{space.maker}).}
+\hl{These functions are based on a similar modular architecture as that used by the \texttt{dispRity} functions, allowing users to provide their own distribution parameters for the simulations.}
+\hl{For example, \texttt{stats::rnorm} can be provided as an argument for drawing normal characters rates with \texttt{sim.morpho} or normally distributed multidimensional space with \texttt{space.maker}.}
+\hl{The discrete morphological data simulations are based on protocols from }
+\cite{GuillermeCooper}, \cite{OReilly20160081} and \cite{puttick2017uncertain}.
+\hl{The multidimensional space simulations are based on the methods from}
+\cite{diaz2016global}.
+\hl{Both functionalities are described in more details in the package manual.}
 
@@
 Many thanks to Natalie Cooper for encouraging the development of this package and helping with the writing of this paper and the package manuals.
-Thanks to David Bapst, Martin Brazeau, Andrew Jackson, Graeme Lloyd and Emma Sherratt for specific implementation ideas 
-\hl{and to Michael Collyer, Gavin Simpson and two other anonymous reviewers for constructive and encouraging comments on the manuscript}. %TG: add Graeme Lloyd, Dave Bapst and Emma Sherratt
-I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau 
+\hl{Thanks to David Bapst, Martin Brazeau, Andrew Jackson, Graeme Lloyd and Emma Sherratt for specific implementation ideas  and to Michael Collyer, Gavin Simpson and two other anonymous reviewers for constructive and encouraging comments on the manuscript}.
+I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 - 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau 
 \hl{and from the Discovery Project Grant DP170103227 awarded to Vera Weisbecker}.
@@
 \section{Funding}
-This work was funded by the European Research Council under the European Union's Seventh Framework Programme (FP/2007–2013)/ERC Grant Agreement number 311092 
+This work was funded by the European Research Council under the European Union's Seventh Framework Programme (FP/2007 - 2013)/ERC Grant Agreement number 311092 
 \hl{and the Discovery Project Grant number DP170103227}.

### 2017-12-24 paper/dispRity_paper_highlighted.tex
Final(ish) revised manuscript

@@
         One such toolkit is based on creating a multidimensional space where the variables are the \hl{axes}.
-        It is then possible to \hl{measure divers aspects of the distribution of some observation (e.g. species) in this space}.
-        For example, if studying morphology, one can create a morphospace for two groups of species, measure \hl{the volume occupied by each of these groups and then test whether these two volumes are significantly different or not}.
+        It is then possible to \hl{measure divers aspects of the distribution of some observation (e.g. species) in this space.
+        For example, if studying morphology, one can create a morphospace for two groups of species, measure the volume occupied by each of these groups and then test whether these two volumes are significantly different or not}.
 
@@
 
-\hl{It is then possible to measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupies more space than group B?'').
+\hl{One can then measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupies more space than group B?'').
 This requires the definition of a proxy for space occupancy: the disparity metric}
@@
 \citealt{diaz2016global}
-\hl{)}
-or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements
-\hl{().}
+\hl{) or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements.}
 In summary, there are many different ways to perform each step of a multidimensional analysis, making analyses of complexity ever more complex.
@@
 It implements many commonly used metrics and definitions of disparity, as well as providing a simple interface for users to implement their own disparity functions.
-The package is described here for the use of discrete morphological data diversity analysis but can be generalised to any type of multidimensional data (see the package vignette for more examples).
-
-\section{Description}
-In brief, the package takes a \texttt{matrix} object (the multidimensional space), 
-\hl{calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation}.
-Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow}).
-\hl{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements} \citep[e.g.
-the multidimensional space can be a distance matrix][]{Close2015}.
-
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=1\textwidth]{workflowsvg.pdf} 
-\caption{\disp package workflow: rectangles represent matrices; ellipses represent functions; plain black arrows indicate input/output; dashed grey arrows indicate output (though the summary, plot, and test function cannot be applied if no disparity has been calculated).}
-\label{Fig:workflow}
-\end{figure}
+The package is described here for the use of discrete morphological data disparity analysis but can be generalised to any type of multidimensional data
+\hl{(see the glossary Table}
+\ref{Tab:terms}
+\hl{for terms' equivalences between palaeobiology and ecology).}
 
@@
 \end{landscape}
+
+\section{Description}
+In brief, the package takes a \texttt{matrix} object (the multidimensional space), 
+\hl{calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation}.
+Some additional functions modify the space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow}).
+\hl{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements} \citep[e.g.
+the space can be a distance matrix][]{Close2015}.
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{workflowsvg.pdf} 
+\caption{\disp package workflow: rectangles represent matrices; ellipses represent functions; plain black arrows indicate input/output; dashed grey arrows indicate output (though the summary, plot, and test function cannot be applied if no disparity has been calculated).}
+\label{Fig:workflow}
+\end{figure}
 
@@
 
-The disparity metrics can be functions from other packages, defined by the user or from the \disp package (see Table \ref{Tab:metrics} for the currently implemented metrics).
-When multiple functions are passed to the \texttt{metric} argument, the functions are sorted by dimension-level and applied in decreasing order to the data.
-For example, if the metric is defined as \texttt{metric = c(\hl{prod}, ranges)} \hl{(the hyperbox volume)}, the \texttt{ranges} function (dimension-level 2) is first applied to data \hl{(calculating the range of each dimension)} and the function \hl{\texttt{prod}} is then applied to the results (e.g.
-\hl{\texttt{prod(ranges(data))}}).
-\hl{If one needs to define the metric as something more complex, say the hypercube volume, one can directly pass a function description to the \texttt{metric} argument (e.g.
-\texttt{metric = function(x) min(ranges(x)) * ncol(x)}).}
-There is no limitation on the metric dimensions only that at least one dimension-level 2 or 1 function must be present in the metric definition (i.e.
-disparity can only be defined as distribution or a single value).
+The disparity metrics can be functions from other packages, defined by the user or from the \disp package (see Table \ref{Tab:metrics}).
+When multiple functions are passed to the \texttt{metric} argument, they are sorted by dimension-level and applied in decreasing order to the data.
+For example, if the metric is defined as \texttt{metric = c(\hl{prod}, ranges)} \hl{(the hypercube volume)}, the \texttt{ranges} function (dimension-level 2) is first applied to data \hl{(the range of each dimensions)} and the function \hl{\texttt{prod}} is then applied to the results (\hl{\texttt{prod(ranges(data))}}).
+\hl{One can also directly pass a function description to the \texttt{metric} argument (e.g. \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} for the average squared pairwise distance).}
 \hl{Note that this function also allows to work on only a subset of dimensions via the \texttt{dimensions} argument (e.g. if only the $n$ first dimensions must be considered).}
@@
 \subsection{Splitting the multidimensional space into subsets}
-Prior to calculating disparity, the multidimensional space can be subdivided into subsets, typically to be compared to each other.
+Prior to calculating disparity, the space can be subdivided into subsets, typically to be compared to each other.
 For example, one may be interested in seeing how the disparity of a specific subset of the space compares to another (e.g.
@@
 In essence, the original space corresponds to the overall space (e.g. a morphospace contains all the observed morphologies).
-Conversely, subsets correspond to parts of the space with pooled characteristics (i.e. only a subset of the total morphospace).
+Conversely, subsets correspond to parts of the space with pooled characteristics.
 
@@
 \hl{The first argument is the matrix}
-defining the multidimensional space and a list of elements to group into different subsets.
-The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time subsets for example,
-\hl{see below}
-) and which subsets to create: (1) discrete time subsets (or time-binning) or (2) continuous time subsets (or time-slicing).
+defining the space and a list of elements to group into different subsets.
+The second also takes a matrix and arguments giving the age of the taxa 
+\hl{(a dated phylogeny of the elements present in the morphospace - see below)}
+and which subsets to create: (1) discrete time subsets (or time-binning) or (2) continuous time subsets (or time-slicing).
 
@@
 The time-slicing method works by using a phylogeny and looking at which taxa are present at any specific point in time.
-This method thus requires the nodes to be part of the multidimensional space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
+This method thus requires the nodes to be part of the space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
 When a slice occurs not on a tip or a node, \hl{six} methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
 \begin{itemize}
-    \item ``acctran'' (accelerated transformation) where the data chosen along the branch is always the one of the descendant
-    \item ``deltran'' (delayed transformation) where the data chosen along the branch is always the one of the ancestor
+    \item ``acctran'' where the data chosen along the branch is always the one of the descendant
+    \item ``deltran'' where the data chosen along the branch is always the one of the ancestor
     \item \hl{``random''} where the data chosen along the branch is randomly chosen between the descendant or the ancestor
@@
 
-Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the multidimensional space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).
-
-% In the case of disparity being a dimension-level 1 metric, it is important to note that the distribution resulting from a bootstrap will be a pseudo-replicated one and thus might not be suitable for some specific parametric statistical tests.
-% Conversely, it is possible to use a dimension-level 2 metric on a bootstrapped multidimensional space to avoid this problem: the disparity metric is then measured on a pseudo-replicated multidimensional space rather than the metric itself being pseudo-replicated.
+Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).
 
@@
 The \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow users to set which central tendency and which quantiles should be represented.
-% as common arguments which central tendency (e.g.
-if central tendency is the mean and disparity was calculated as the sum of the variances for each dimensions, the central disparity tendency would be the mean of the sum of the variances) and the quantiles to represent (in the same example, if the quantile chosen is the 50th, these functions will also display 50\% of the sum of variances around the mean).
 Additionally, the \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves (in the case of disparity-through-time analysis for example) and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
@@
 The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base \texttt{R}.
-% Note that if the disparity has been calculated on fully rarefied data (using \texttt{boot.matrix(data, rarefaction = TRUE)} prior to calculating disparity), the rarefaction curves for the whole multidimensional space (or its sub-samples) can be plotted using \texttt{rarefaction = ``plot''}.
 
@@
 
-Conversely, in the case of species diversity, one can also look at another independent multivariate trait: the diversity of morphologies \citep[or disparity;][]{ruta2013}.
+Conversely, in the case of species diversity, one can also look at another independent multivariate trait: the diversity of morphologies \citep[or disparity;][]{foote1991morphological}.
 Using disparity, it is then also possible to assess whether one ecosystem or/and time period display more morphological variation.
@@
 
-The following example is based on a classical palaeobiology morphological disparity analysis.
+The following example is based on a classical morphological disparity analysis.
 Note that more examples are available in the package manual (\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).
@@
     \item \texttt{BeckLee\_mat50}: an ordinated matrix for 50 mammals based on the distance between discrete morphological characters.
-    \item \texttt{BeckLee\_mat99}: the same matrix \texttt{BeckLee\_mat50} but also containing the reconstruction of their 49 ancestors.
+    \item \texttt{BeckLee\_mat99}: the same matrix \texttt{BeckLee\_mat50} with the reconstruction of their 49 ancestors.
     \item \texttt{BeckLee\_tree}: a chronogram with the 50 mammal species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
@@
 
-In this example, the multidimensional space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
+In this example, the space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
 Additionally, we can define disparity as the sum of the variances on each dimension \citep{foote1991morphological,Wills1994} that will represent an aspect of the the volume of the morphospace (i.e.
@@
 
-The \texttt{R} script for the following example is available in the supplementary materials.
+\hl{The \texttt{R} script for the following example is available in the supplementary materials.}
 
@@
 \noindent Which allows a finer tuning of the optional arguments in each function.
-The three arguments here are defined as follows: \texttt{data = BeckLee\_mat50} is our definition of the multidimensional space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity.
+The three arguments here are defined as follows: \texttt{data = BeckLee\_mat50} is our space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity.
 
@@
 question to this observation could be to test whether this difference is due to an overall change in disparity through time or not.
-In other words, did disparity increase through time in mammals globally or is it only between stem and crown mammals.
 Using the same definition of the multidimensional space and disparity as in the previous example, we can measure, for example, changes in disparity through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).
@@
 \caption{Summarising a \disp object (disparity through time).
-$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 95\% are the confidence intervals.}
+$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 97.5\% are the confidence intervals.}
 \label{Tab:summary_time}
@@
 From the plot and the summary, disparity seems to change (increase) through time from the Cretaceous to the Eocene.
-Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA \citep[PERMANOVA; ][]{permanova} through the \texttt{adonis.disprity} function that is a wrapper of the \texttt{vegan::adonis} function \citep{oksanen2007vegan} for \disp objects:
+\hl{Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA}
+ \citep[PERMANOVA; ][]{permanova}
+\hl{through the \texttt{adonis.disprity} function that is a wrapper of the \texttt{vegan::adonis} function}
+ \citep{oksanen2007vegan}
+\hl{ for \texttt{dispRity} objects:}
 
@@
 To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
-\hl{Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test}.
+\hl{Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test.
+Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.}
 
@@
 \subsection{Manuals and vignette}
-Supplementary information about the package is available on the project page (\url{https://TGuillerme/dispRity}).
-Information concerning each function can be found in R in the individual function manuals and a package manual with more detailed information is available online (\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).
+\hl{Supplementary information concerning the package and each function can be found in \texttt{R}, on the project page (}
+\url{https://TGuillerme/dispRity}
+\hl{) or in the online manual (}
+\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}
+\hl{).}
 This manual contains substantially more information and detailed examples including a tutorial for a ``classic'' disparity analysis in palaeobiology as well as an introduction to the use of this package in ecology or other disciplines.
@@
 \hl{These functions are based on a similar modular architecture as that used by the \texttt{dispRity} functions, allowing users to provide their own distribution parameters for the simulations.}
-\hl{For example, \texttt{stats::rnorm} can be provided as an argument for drawing normal characters rates with \texttt{sim.morpho} or normally distributed multidimensional space with \texttt{space.maker}.}
+\hl{For example, \texttt{stats::rnorm} can be provided as an argument for drawing normal characters rates with \texttt{sim.morpho} or normally distributed spaces with \texttt{space.maker}.}
 \hl{The discrete morphological data simulations are based on protocols from }
 \cite{GuillermeCooper}, \cite{OReilly20160081} and \cite{puttick2017uncertain}.
-\hl{The multidimensional space simulations are based on the methods from}
+\hl{The space simulations are based on the methods from}
 \cite{diaz2016global}.

### 2017-12-24 paper/response_to_reviewers.tex
reviewer 1 responded

@@
 species) in this space.
-For example, if studying morphology, one can create a morphospace for two groups of species, measure the multidimensional volume occupied by each of these groups and then test whether these two volumes are significantly different or not.} line @@@
+For example, if studying morphology, one can create a morphospace for two groups of species, measure the volume occupied by each of these groups and then test whether these two volumes are significantly different or not.} lines 6-10.
 
@@
 
-%TODO improve disparity definition in the abstract
-
+Apologies for the omission of the table summarising the metrics which is now properly included in the manuscript (see reviewer 1, minor comment \ref{table_missing})
 As reviewer 4 (major comment \ref{nebulous}) also underlines: ``disparity is a broad, nebulous concept''.
@@
 
-\textit{It is then possible to measure how the observations are distributed within this space to answer related questions (e.g.
-“does group A occupies more space than group B?”).
+\textit{One can then measure how the observations are distributed within this space to answer related questions (e.g.
+``does group A occupies more space than group B?'').
 This requires the definition of a proxy for space occupancy: the disparity metric (or index; Hopkins and Gerber, 2017) which can be measured in a multitude of ways.
@@
 Euclidean) measured between observations (Foote, 1993, 1996), a more direct approximation of the hyper volume (Cornwell et al., 2006; Donohue et al., 2013), or many more (e.g.
-Navarro, 2003).} lines @@@
+Navarro, 2003).} lines 60-67.
 
@@
 
-%TODO
-%TG: add more functions cross over with vegan and geiger
+I have added two wrapper functions to use the popular functions \texttt{vegan::adonis} and \texttt{geiger::dtt} to the package (\texttt{dispRity.adonis} and \texttt{dispRity.dtt}).
+Also, I have generalised the distances calculations using the \texttt{vegan::vegdist} function that allows for more distances than \texttt{stats::dist} to be calculated.
+
+I've updated the manual and the example to describe these new functions:
+
+\textit{Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA (PERMANOVA; Anderson, 2001b) through the adonis.disprity function that is a wrapper of the vegan::adonis function (Oksanen et al., 2007) for dispRity objects [TABLE 4].
+To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
+Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test.
+Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.} lines 351-361.
 
@@
 
-%TODO
-%Increase keywords
+I have broaden the keywords within the limits allowed by manuscript central for MEE publications.
 
@@
 
-\textit{For example, in the excellent and widely used} \texttt{geomorph} \textit{package, morphological disparity analysis uses the} \texttt{morphol.disparity} \textit{function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance), and uses permutation tests(Zelditch et al., 2012; Adams and Otarola-Castillo, 2013; Adams et al., 2017).
+\textit{For example, in the excellent and widely used} \texttt{geomorph} \textit{package, morphological disparity analysis uses the} \texttt{morphol.disparity} \textit{function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance), and uses permutation tests (Zelditch et al., 2012; Adams and Otarola-Castillo, 2013; Adams et al., 2017).
 This is ideal for testing volume based hypothesis (e.g.
 does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g.
-does groups A and B occupy the same location?).} lines @@@
+does groups A and B occupy the same location?).} lines 85-93.
 
@@
 this metric could be used as a proxy for the average space occupancy per unit element - e.g.
-per taxon).
+per taxa).
 This would not be doable directly by using the implemented \texttt{ellipse.volume} function but would require the user to use a function like:\\
@@
 I meant both matrices (as this reviewers rightly points out they are suitable for ordination).
-I have made this more clear in the figure.
+I have made this more clear in the figure caption.
 On a more general note, the package now allows users to input any matrix (although the package does not provide warranty that the provided matrix is meaningful to the user's specific analysis).
+I have added an example to the \texttt{dispRity} manual based on the \texttt{eurodist R} base dataset.
 
@@
 
-I've added ``\textit{see below}'' to the sentence (line @@@).
-Following this comment, I've also changed the code since, indeed, a phylogeny should not always be required!
-The function can now works without it as long as First/Last occurrence data are provided for each taxa.
+I've modified the sentence in bracket to
+``\textit{a dated phylogeny of the elements present in the morphospace - see below}'' (line @@@).
+
+I've also changed the code to allow non-phylogenetic dating (e.g.
+just adding the First/Last Occurrence Dates of all the taxa) but I've only mentioned it in the function manual and the vignette due to space restriction in this manuscript.
 
@@
 in Díaz et al.
-2016) or even simply by looking at the confidence interval overlaps between disparity measurements (which, when not based on rigorous statistic methods can often leads to serious inaccuracy).}
+2016) or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements.}
 

### 2017-12-27 paper/response_to_reviewers.tex
Reviewer 2 responded

@@
 Reviewer 4 (minor comment \ref{remove_simulations}) suggested completely removing this section which conflicts directly with this suggestion.
-The simulation protocols implemented here are already developed (and tested) in details in previous publications with new novel features attached to future publication.
+The simulation protocols implemented here are already developed (and tested) in details in previous publications along with novel features that are part of future publications.
 For example the discrete data simulation protocol is described and detailed in Guillerme \& Cooper 2016 \textit{Mol Phyl Evol}, O’Reilly et al.
@@
 2016 \textit{Nature} and will be used along with additional simulation features not implemented yet (including shape evolution) as part of a project currently in progress.
-I am therefore inclined to follow reviewer's 4 suggestion rather than this one.
+Due to the space constraint for application manuscripts in MEE, I am inclined to follow reviewer's 4 suggestion rather than this one.
 
@@
 The manual now has a chapter entirely dedicated to the data simulation functions with emphasis on both the methods used and the algorithms implemented (and illustrated with advanced examples).
-%TODO: check typos in there.
 
@@
 
-I have however modified the available models to better describe this limitation and I have added two new models that would approximate ancestral reconstruction along edges
-I have modified the name of the two previous ``punctuated'' and ``gradual'' algorithms to ``random'' and ``proximity'' to reflect more accurately how they select either the ACCTRAN or DELTRAN information.
+I have however modified the available models to better describe this limitation and I have added two new models that would approximate ancestral reconstruction along edges.
+I have modified the name of the two previous ``punctuated'' and ``gradual'' models to ``random'' and ``proximity'' to reflect more accurately how they select either the ACCTRAN or DELTRAN information.
 I've added two more accurate models (now named ``equal.split'' and ``gradual.split'') to reflect the changes along the branches.
@@
 These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).
-Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).} lines @@@
+Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).} lines 166-180.
 
@@
 
-I also agree with this comment: this method creates a trade-off between the number of slices and the type II errors: %TOCHEK@@@
+I also agree with this comment: this method is a trade-off between the number of slices.
 (1) too few slices decrease resolution and veer towards the time binning (which could increase "significance" by comparing too different time subsamples);
-(2) too many slices will increase the resolution be will create much more pseudo replication and thus increase type II error.
-
-A more detailed description and test of these models might be out of the scope of this package description paper but is currently in development for a special issue paper to be submitted to \textit{Palaeontology} as part as a special symposium for the Palaeontological Association annual meeting 2017 in collaboration with Natalie Cooper (\url{http://www.palass.org/meetings-events/annual-meeting/2017/annual-meeting-2017-london-symposium-abstracts}).
+(2) too many slices will increase the resolution be will create much more pseudo replication and thus increase type I error.
 
@@
 
-\textit{Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).} line @@@
+\textit{Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).} line 177-180.
+
+A more detailed description and test of these models might be out of the scope of this package description paper but is currently in development for a special issue paper to be submitted to \textit{Palaeontology} as part as a special symposium for the Palaeontological Association annual meeting 2017 in collaboration with Natalie Cooper.
+The full (un-reviewed an un-compiled) manuscript is available on Natalie's Github page (\url{https://github.com/nhcooper123/time-slice/blob/master/manuscript/time-slice.tex}).
 
@@
 
-I've reflected these changes in the manual with a subsection in the chapter ``Specific tutorials'' and in manuscript as follows:
+I've reflected these changes in the manual with a subsection in the chapter ``Specific tutorials - 4.8'' (based on a distance matrix) and in manuscript as follows:
 
 \textit{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements (e.g.
-the multidimensional space can be a distance matrixl; Close et al., 2015).} @@@
+the multidimensional space can be a distance matrixl; Close et al., 2015).} lines 109-111.
 
@@
 This was a conscious choice: due to the really high number of metrics available (especially when combining different ones), I designed the package to be modular hence allowing users to implement their own metrics (see response to reviewer 1 minor comment \ref{table_missing}).
-I have precise this using the hypercube volume example in the manuscript:
+I have add the hypercube volume description as an example of user defined metric in the manuscript:
 
-\textit{When multiple functions are passed to the} \texttt{metric} \textit{argument, the functions are sorted by dimension-level and applied in decreasing order to the data.
-For example, if the metric is defined as} \texttt{metric = c(prod, ranges)} \textit{(the hyperbox volume), the} \texttt{ranges} \textit{function (dimension-level 2) is first applied to data (calculating the range of each dimension) and the function} \texttt{prod} \textit{is then applied to the results (e.g.} \texttt{prod(ranges(data))}\textit{).
-If one needs to define the metric as something more complex, say the hypercube volume, one can directly pass a function description to the} \texttt{metric} \textit{argument (e.g.} \texttt{metric = function(x) min(ranges(x)) * ncol(x)}\textit{).}
+\textit{When multiple functions are passed to the} \texttt{metric} \textit{argument, they are sorted by dimension-level and applied in decreasing order to the data.
+For example, if the metric is defined as} \texttt{metric = c(prod, ranges)} \textit{(the hypercube volume), the} \texttt{ranges} \textit{function (dimension-level 2) is first applied to data (the range of each dimensions) and the function} \texttt{prod} \textit{is then applied to the results (}\texttt{prod(ranges(data))}\textit{).
+One can also directly pass a function description to the} \texttt{metric}\textit{ argument (e.g.} \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} \textit{ for the average squared pairwise distance).} lines 127-132.
 
-I have however added several additional ones to the newest version of the package including the minimal spanning tree length (\texttt{span.tree.length} - as suggested), the ancestral distance which is equivalent to the distance from centroid but were the ``centroid'' are the coordinates of each element's ancestors (\texttt{ancestral.dist}), the pairwise distances (\texttt{pairwise.dist}), the radius which equivalent to the distance from centroids but for the dimensions (i.e.
-the distance between each dimensions coordinates and the mean dimension's coordinates - \texttt{radius}) and the $n$-ball volume which is an extension of the \texttt{ellipse.volume} function for a sphere (the true $n$-ball volume) or an ellipsoid (\texttt{n.ball.volume}).
+I have however added several additional new metrics ones to the newest version of the package including: (1) the minimal spanning tree length (\texttt{span.tree.length} - as suggested), (2) the ancestral distance which is equivalent to the distance from centroid but were the ``centroid'' are the coordinates of each element's ancestors (\texttt{ancestral.dist}), (3) the pairwise distances (\texttt{pairwise.dist} - based on \texttt{vegan:vegdist}), (4) the radius which equivalent to the distance from centroids but for the dimensions (i.e.
+the distance between each dimensions coordinates and the mean dimension's coordinates - \texttt{radius}) and (5) the $n$-ball volume which is an extension of the \texttt{ellipse.volume} function for a sphere (the true $n$-ball volume) or an ellipsoid (\texttt{n.ball.volume} - see Table 2 in the new version of the manuscript).
 
@@
 This is again an excellent suggestion!
-I will develop more complex rarefaction algorithms based on sampling and occurrence based on methods applied empirically by Kotric and Knoll 2015 \textit{Palaeobiology} and Butler et al 2012 \textit{Evolution}.
-I've added a note
-
-% %TG: TODO maybe!
+I will implement the possibility to pass a probability matrix to the \texttt{boot.type} argument in the \texttt{boot.matrix} function where the probabilities can be reflecting occurences, regions, etc.
+I will also implement the ``fair'' algorithm mentioned in Kotric and Knoll (2015, Paleobiology, 41, 68–).
 
@@
 
-I have added an illustrative example in the \texttt{plot.dispRity} manual on how to add disparity data (continuous or discrete) to the \texttt{geoscalePlot} and \texttt{geoscaleBox} functions thus allowing users to make ready-to-publish figures.
+Similarly to previous comment, I will also add the permutation test based on the \texttt{PAST} package in a future version of the package.
+Furthermore, I have added an illustrative example in the \texttt{plot.dispRity} manual on how to add disparity data (continuous or discrete) to the \texttt{geoscalePlot} and \texttt{geoscaleBox} functions thus allowing users to make nice ready-to-publish figures using the \texttt{geoscale} package.
 
@@
 
-\textit{Note that this function also allows to work on only a subset of dimensions via the} \texttt{dimensions} \textit{argument (e.g. if only the $n$ first dimensions must be considered).} l. @@@:
+\textit{Note that this function also allows to work on only a subset of dimensions via the} \texttt{dimensions} \textit{argument (e.g.
+if only the $n$ first dimensions must be considered).} lines 132-133.
 

### 2017-12-27 paper/dispRity_paper_highlighted.tex
Reviewer 2 responded

@@
 The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
-There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).
+There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).
 See Guillerme \& Cooper \textit{in review} for full description.}

### 2017-12-27 paper/response_to_reviewers.tex
Reviewer 3 done

@@
 Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test.
-Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.} lines 351-361.
+Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.} lines 355-365.
 
@@
 in Liow 2004; Donohue et al.
-2013).} lines @@@
+2013).} lines 51-56
 
@@
 \textit{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best it their needs (i.e.
-data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.}  lines @@@
+data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.} lines 82-85.
 
@@
 
-I agree and have changed it to ``\textit{all the observed morphologies}'' line @@@.
+I agree and have changed it to ``\textit{all the observed morphologies}'' line 147-148
 
@@
 
-I have changed the start of the sentence to ``\textit{The first argument is the matrix}'' line @@@.
+I have changed the start of the sentence to ``\textit{The first argument is the matrix}'' line 151.
 
@@
 I've modified the sentence in bracket to
-``\textit{a dated phylogeny of the elements present in the morphospace - see below}'' (line @@@).
+``\textit{a dated phylogeny of the elements present in the morphospace - see below}'' line 153.
 
@@
 
-\item{\textcolor{blue}{The author presents a simple example and defers much of the many details users will encounter to his impermanent sources (vignettes, webpages), which I believe to be wise, as the R package will evolve.
+% \item{\textcolor{blue}{The author presents a simple example and defers much of the many details users will encounter to his impermanent sources (vignettes, webpages), which I believe to be wise, as the R package will evolve.
 I believe the purpose of a manuscript like this is illustrate its purpose and utility rather than serve as an instructional manual.
@@
 
-\item{\textcolor{blue}{The author acknowledged assistance with writing (suspect English is a second language) and the writing was great in the paper, but less so in the abstract.
+% \item{\textcolor{blue}{The author acknowledged assistance with writing (suspect English is a second language) and the writing was great in the paper, but less so in the abstract.
 Having someone double check the grammar in the abstract would be beneficial.
@@
 
-I have changed the variable names in the package description manuscript to not be redundant.
+I have changed the variable names in the package description manuscript to be non-redundant.
 I've also changed the variable names in the package's help files to not override other functions/objects.
@@
 
-% %TG: ANOVA problem
-
-% As a minimum, this analysis should be called something different than " ANOVA performed on a dispRity object", which is rather terse, and I would be loathe to suggest that it tests the null hypothesis that disparity does not change through time, as suggested by the author.
-In fact, it simply doesn't.
-I came up with this simple example to illustrate why, according to the dispRity guidance:
-
-% n <- 40
-% p <- 10
-% g <- as.data.frame(as.factor(rep(1:2, n)[1:n])) # generate groups
-% Y <- matrix(rnorm(n*p), n, p)# generate random data with no pattern
-% rownames(Y) <- rownames(g)
-% cs <- custom.subsamples(Y, g) # customized subsamples
-% bd <- boot.matrix(cs, bootstraps = 100) # bootstrapped data
-% sv <- dispRity(bd, metric = c(sum, variances)) # sum_variances, as per directions
-
-% anova(test.dispRity(sv, lm, "all")) # ALWAYS. A. SIGNIFICANT! RESULT!
-
-% To state this plainly, garbage in equals a pattern out.  This happened with every permutation I ran and the result was usually a huge effect size similar to that reported in the manuscript.
+\label{ANOVA}
 
@@
 
-The ANOVA problem is totally a garbage-in-garbage-out problem and I agree with this reviewer that this might be a serious problem down the line.
+I agree that this ANOVA is a garbage-in-garbage-out test and used it more for illustrating the package's possibilities rather than show-casing disparity analysis best (worst?) practices.
+Following this reviewer's comments with which I totally agree, I have completely removed the ANOVA example (from the paper, the manuals and the vignette), replacing it in the paper with a PERMANOVA (based on \texttt{vegan::adonis}) which is more conventional.
+As this reviewer points out, the ANOVA is indeed performed on the bootstrapped data rather than the observed data which, of course, artificially (and wrongly) increases the statistical power.
+Although the ANOVA is not mentioned any more, I have however left the ability for \texttt{test.dispRity} to deal with \texttt{lm} type tests (with a warning being triggered).
 
-Unfortunately, I do not have any proper solution to propose to this problem and I guess multidimensional distributions comparisons methods needs yet to be developed!
+I have implemented the reviewers' suggestion similary to the ``percentile bootstrap'' idea, allowing \texttt{test.dispRity} to work on either the pooled bootstrap distribution through the \texttt{concatenate = TRUE} argument (this was indeed erroneous in the previous ANOVA example but could be justifiable in other cases; e.g.
+when comparing distribution overlap probabilities) or by applying the test for each bootstrap pseudo-replicate with \texttt{concatenate = FALSE}.
+I've clarified this in the manuscript:
 
-The reason for these crazy high degrees of freedom come from the bootstrapping pseudo-replication: effectively, the test data contains all the bootstrap data.
-One solution is to use the concatenate = FALSE option in test.dispRity, performing the test for each bootstrap replicate rather than pooling the data.
-This methods outputs a distribution of test that can be used to assess the type I error.
-However, the pooling of the p-value is not satisfactory either.
+\textit{Additionally, the test is here performed on the pooled bootstrapped pseudo-replications which can increase the type I error.
+It is possible to compare each bootstrap in a pairwise way without pooling the data by using the} \texttt{concatenate = FALSE} \textit{argument.
+The results will then be a distribution of statistic and p values.} lines 319-323.
 
-Some methods exists such as NPMANOVA (from vegan::adonis) but is performed on the wrong kind of matrix.
-
-One of the reasons for this functionality in the package is to make the test.dispRity function modular like the other main functions: i.e.
-it \textit{can} apply a \textit{lm} class function to the data but the reason for doing that might be dubious (or not, if the users don't bootstrap the matrix for example).
-
-%TG: I therefore removed the example in the manuscript and the manual to not mislead users into running flawed analysis.
-However, I've left the architecture implemented for adventurous (or clever) users.
-% Remove the ANOVA example
-% Remove mentions to it on bootstrapped data in the manuals
-% Replace by more savy example
-% Allow to use NPMANOVA instead
-%
-% Dissimilarity analyses: ANOVA using dissimilarities, ANOSIM, MRPP, BIOENV, Mantel and partial Mantel tests.
-% Ordination and environment: vector fitting, centroid fitting and smooth surface fitting, adding species scores as weighted averages, adding convex hull, SD ellipses, arrows etc. to ordination.
-% Ordination: support and meta functions for NMDS, redundancy analysis, constrained correspondence analysis, constrained analysis of proximities (all three with partial analysis),
+I aim to implement more multidimensional test in future versions of the package (or that users develop their own implementations!).
 
@@
 \textit{Note that by default the function only outputs the test's statistic, parameter (if parametric) and the p value.
-However, the raw test results can also be output using the option} \texttt{details = TRUE} \textit{in the function above.} lines @@@
+However, the raw test results can also be output using the option} \texttt{details = TRUE} \textit{in the function above.} lines 318-319.
+
+Furthermore, I have developed the example for the function \texttt{geomorph.ordination} highlighting how to perform similar analysis with both packages (as shown in the reviewer's attached markdown example).
 

### 2017-12-27 paper/dispRity_paper_highlighted.tex
Reviewer 3 done

@@
 through time).
-In essence, the original space corresponds to the overall space (e.g. a morphospace contains all the observed morphologies).
+In essence, the original space corresponds to the overall space (e.g. a morphospace contains \hl{all the observed morphologies}).
 Conversely, subsets correspond to parts of the space with pooled characteristics.
@@
 first against second, second against third, etc.); (4) `all'' for comparing all the subsets simultaneously (i.e.
-\texttt{disparity $\mathtt{\sim}$ subsets} - this can be used for \texttt{lm} class functions) or (5) any list of pairs (or more) of subsets to compare.
+\texttt{disparity $\mathtt{\sim}$ subsets}) or (5) any list of pairs (or more) of subsets to compare.
 
@@
 \hl{Note that by default the function only outputs the test's statistic, parameter (if parametric) and the \textit{p} value.
-However, the raw test results can also be output using the option \texttt{details = TRUE} in the function above.}
+However, the raw test results can also be output using the option \texttt{details = TRUE} in the function above.
+Additionally, the test is here performed on the pooled bootstrapped pseudo-replications which can increase the type I error.
+It is possible to compare each bootstrap in a pairwise way without pooling the data by using the \texttt{concatenate = FALSE} argument.
+The results will then be a distribution of statistic and \textit{p} values.}
 Relating back to our question: yes, crown mammals display a higher diversity in morphologies than their stem counterparts (in this example and dataset!).

### 2018-01-04 paper/response_to_reviewers.tex

@@
 Similarly, the R package ‘diverse’ has a function called disparity, and the ‘geiger’ function disparity implements the disparity-through-time. }}
+\label{add_geiger_vegan}
 
@@
 Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test.
-Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.} lines 355-365.
+Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.} lines 359-369.
 
@@
 These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).
-Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).} lines 166-180.
+Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).} lines 169-183.
 
@@
 
-\textit{Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).} line 177-180.
+\textit{Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).} line 180-183.
 
@@
 For example, if the metric is defined as} \texttt{metric = c(prod, ranges)} \textit{(the hypercube volume), the} \texttt{ranges} \textit{function (dimension-level 2) is first applied to data (the range of each dimensions) and the function} \texttt{prod} \textit{is then applied to the results (}\texttt{prod(ranges(data))}\textit{).
-One can also directly pass a function description to the} \texttt{metric}\textit{ argument (e.g.} \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} \textit{ for the average squared pairwise distance).} lines 127-132.
+One can also directly pass a function description to the} \texttt{metric}\textit{ argument (e.g.} \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} \textit{ for the average squared pairwise distance).} lines 130-135.
 
@@
 \textit{Note that this function also allows to work on only a subset of dimensions via the} \texttt{dimensions} \textit{argument (e.g.
-if only the $n$ first dimensions must be considered).} lines 132-133.
+if only the $n$ first dimensions must be considered).} lines 134-135.
 
@@
 
-I agree and have changed it to ``\textit{all the observed morphologies}'' line 147-148
+I agree and have changed it to ``\textit{all the observed morphologies}'' line 150
 
@@
 
-I have changed the start of the sentence to ``\textit{The first argument is the matrix}'' line 151.
+I have changed the start of the sentence to ``\textit{The first argument is the matrix}'' line 153.
 
@@
 I've modified the sentence in bracket to
-``\textit{a dated phylogeny of the elements present in the morphospace - see below}'' line 153.
+``\textit{a dated phylogeny of the elements present in the morphospace - see below}'' line 155.
 
@@
 
-\label{ANOVA}
-
 \textcolor{blue}{I can see no reason to perform ANOVA like this.
@@
 
+\label{ANOVA}
+
 I agree that this ANOVA is a garbage-in-garbage-out test and used it more for illustrating the package's possibilities rather than show-casing disparity analysis best (worst?) practices.
@@
 It is possible to compare each bootstrap in a pairwise way without pooling the data by using the} \texttt{concatenate = FALSE} \textit{argument.
-The results will then be a distribution of statistic and p values.} lines 319-323.
+The results will then be a distribution of statistic and p values.} lines 323-327.
 
@@
 \textit{Note that by default the function only outputs the test's statistic, parameter (if parametric) and the p value.
-However, the raw test results can also be output using the option} \texttt{details = TRUE} \textit{in the function above.} lines 318-319.
+However, the raw test results can also be output using the option} \texttt{details = TRUE} \textit{in the function above.} lines 319-320.
 
@@
 
-I have worked towards making the manuscript more useful for non-morphologists.
-I have generalised the input description as suggest by this reviewer, minor comment \ref{rows}; added a glossary table and a brief example using ecological data.
-%TODO: add Diaz's example?
-% %TG: adding an ecology example (removing manual and vignette description)
+Due to space constraint for this format of MEE manuscript, I have left the references to ecology in the introduction of the manuscript but removed any further references.
+However, I have expanded the ecological example in the vignette which now include a test similar to the one in Diaz \textit{et al} 2016 (Science).
+I therefore hope that the mention of ecology in the introduction will encourage ecology-focused researcher to explore the manual.
+
+\textit{The following example is based on the discrete morphological data described above. More examples, including based on other types of data are available in the package vignette.} lines 271-273
+
+Additionally, I've also added a Glossary table as suggested and generalised disparity data description as suggest by this reviewer's minor comment \ref{rows}.
 
@@
 I apologise for the unfounded critique and I have toned down this sentence (see reviewer 1, minor suggestion \ref{tone_down}).
-% Additionally, I have added more functionalities using vegan and geiger %TODO!
-% %TG: Tone down the critics to other packages
+Following the points raised by this reviewer, I have added functionalities from other packages (namely \texttt{vegan::adonis}) using defaults more appropriate to disparity analysis in palaeobiology (see reviewer 1 major comment \ref{add_geiger_vegan}).
 
@@
 I have tried to define disparity with more examples and references to be more familiar to a broader range of readers (See reviewer 1, major suggestion \ref{define_disparity}).
-
-%TG: explain disparity
 
@@
 
-%TOD: add a note to the paper and the package (warnings if negative eigen values).
-%TG: negative egein values explnations
-% Check Legendre and Legendre pp.449
+The package does not deal with negative eigen values in any special way, this task being left to the user to deal with.
+In fact, the package simply takes as input a matrix and only considers rows as elements and columns as dimensions; it ignores whether it is an ordinated matrix or not and this, along with the subsequent tasks is left to the users to deal with appropriately.
+
+I have added a warning about that in the manuscript, in the vignette and the package manual (for the functions that allow the matrix as input).
+
+\textit{The matrix is always considered as the multidimensional space and no correction is applied to it (e.g.
+if ordinated and with negative eigen values, corrections should be applied prior to using the} \texttt{dispRity} \textit{package.)} lines 111-113
 
@@
 missing data, Ciampaglio 2001).
-However, as this reviewer high
+I now cite these three references lines 293-294 as a justification.
 
-%TG: justify from palaeo litterature? + add a note "One of the caveats of such method is that it implicitely ignores the covariance between the ordination axis."
+However, as this reviewer highlights, this metric suffers from several problems such as the fact that it ignores the covariance between the dimensions.
+This can be a problem when the disparity metric is used for an exact proxy of the multidimensional volume (which, as this reviewer proposes, would be closer to the ellipsoid volume).
+However as a standard and comparable disparity metric, the sum of each dimension's variance is mathematically equal to the mean squared pairwise distance, a metric heavily used through it's implementation in \texttt{geiger} (Harmon \textit{et al} 2008 Bioinformatics).
 
@@
 
-%TODO
+I have removed this statement altogether and now directly applied the test, the reader can now read directly to the results of the test (table 5) and the conclusions lines 365-366.
 
@@
 
-%TODO correct/explain this 300 subsamples + add a posthoc t-test.
+The high value of the samples, residuals and degrees of freedom comes from the test being performed on the pooled bootstrap values rather than the raw observations.
+I have now removed any mentions to this dodge anova and replaced it with a more traditional PERMANOVA instead (see reviewer 1 major comment \ref{add_geiger_vegan} and reviewer 3, major comment \ref{ANOVA}).
 

### 2018-01-04 inst/gitbook/02_getting-started.Rmd

@@
 These matrices can be observations, pairwise differences between elements, ordinations, etc...
+
+> Note that whether the matrix is an ordinated one or of any other type, it is always considered as the multidimensional space to analysis and no corrections are applied to it (e.g.
+if ordinated and with negative eigen values, no correction is used).
+Always make sure the input is the right multidimensional space to analyse!
 

### 2018-01-04 inst/gitbook/03_specific-tutorials.Rmd

--- 2017-11-16 inst/gitbook/03_specific-tutorials.Rmd
@@
 ## Loading the data
-data(BeckLee_mat50) ; data(BeckLee_mat99)
-data(BeckLee_tree) ; data(BeckLee_ages)
+data(BeckLee_mat50)
+data(BeckLee_mat99)
+data(BeckLee_tree)
+data(BeckLee_ages)
 ```
@@
 
+If you want to generate time subsets based on stratigraphy, the package proposes a useful functions to do it for you: `get.bin.ages` (check out the function's manual in `R`)!
+
 ## Customised subsets
@@
 ```{r, eval=TRUE}
-## Creating the two groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44))
+## Creating the two groups (crown and stems)
+mammal_groups <- crown.stem(BeckLee_tree, inc.nodes = FALSE)
 
@@
 
-Elements can easily be assigned to different groups if necessary!
+Like in this example, you can use the utility function `crown.stem` that allows to automatically separate the crown and stems taxa given a phylogenetic tree.
+Also, elements can easily be assigned to different groups if necessary!
 
@@
 ## Creating the three groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44).
+weird_groups <- list("even" = seq(from = 1, to = 49, by = 2),
+                      "odd" = seq(from = 2, to = 50, by = 2),
                       "all" = c(1:50))
 ```
+
+The `custom.subsets` function can also take a phylogeny (as a `phylo` object) as an argument to create groups as clades:
+
+```{r, eval=FALSE}
+## Creating groups as clades
+custom.subsets(BeckLee_mat50, group = BeckLee_tree)
+```
+
+This automatically creates 49 (the number of nodes) groups containing between two and 50 (the number of tips) elements.
 
@@
 
-Note that the `test.dispRity` algorithm deals with some classical test outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
+Note that the `test.dispRity` algorithm deals with some classical test outputs and summarises the test output.
 It is, however, possible to get the full detailed output by using the options `details = TRUE`.
@@
 
-It is also possible to apply some more *complex* tests that have their own output classes (like `stats::lm`).
-
-The results can then be analysed as usual using the associated `summary` S3 method:
-
-```{r}
-## Performing and linear model applied to the same data
-(slice_lm <- test.dispRity(disparity_time_slices, test = lm,
-                            comparisons = "all"))
-
-## The output is a regular `lm` output
-class(slice_lm)
-
-## This output can be summarised using summary
-summary(slice_lm)
-```
-
-Of course, due to the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
+Because of the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
 The only condition is that the test can be applied to at least two distributions.
-In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm` for example).
+In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm`-type functions for example).
+
+### NPMANOVA in `dispRity`
+
+One often useful test to apply to multidimensional data is the permutational multivariate analysis of variance based on distance matrices `vegan::adonis`.
+This can be done on `dispRity` objects using the `adonis.dispRity` wrapper function.
+Basically, this function takes the exact same arguments as `adonis` and a `dispRity` object for data and performs a PERMANOVA based on the distance matrix of the multidimensional space (unless the multidimensional space was already defined as a distance matrix).
+The `adonis.dispRity` function uses the information from the `dispRity` object to generate default formulas:
+
+  * If the object contains customised subsets, it applies the default formula `matrix ~ group` testing the effect of `group` as a predictor on `matrix` (called from the `dispRity` object as `data$matrix` see [`dispRitu` object details](#The-dispRity-object-content))
+  * If the object contains time subsets, it applies the default formula `matrix ~ time` testing the effect of `time` as a predictor (were the different levels of `time` are the different time slices/bins)
+
+```{r}
+set.seed(1)
+## Generating a random character matrix
+character_matrix <- sim.morpho(rtree(20), 50, rates = c(rnorm, 1, 0))
+
+## Calculating the distance matrix
+distance_matrix <- as.matrix(dist(character_matrix))
+
+## Creating two groups
+random_groups <- list("group1" = 1:10, "group2" = 11:20)
+
+## Generating a dispRity object
+random_disparity <- custom.subsets(distance_matrix, random_groups)
+
+## Running a default NPMANOVA
+adonis.dispRity(random_disparity)
+```
+
+Of course, it is possible to pass customised formulas if the disparity object contains more more groups.
+In that case the predictors must correspond to the names of the groups explained data must be set as `matrix`:
+
+```{r}
+## Creating two groups with two states each
+groups <- as.data.frame(matrix(data = c(rep(1,10), rep(2,10), rep(c(1,2), 10)),
+         nrow = 20, ncol = 2, dimnames = list(paste0("t", 1:20), c("g1", "g2"))))
+
+## Creating the dispRity object
+multi_groups <- custom.subsets(distance_matrix, groups)
+
+## Running the NPMANOVA
+adonis.dispRity(multi_groups, matrix ~ g1 + g2)
+```
+
+Finally, it is possible to use objects generated by `time.subsets`.
+In this case, `adonis.dispRity` will applied the `matrix ~ time` formula by default:
+
+```{r}
+## Creating time series
+time_subsets <- time.subsets(BeckLee_mat50, BeckLee_tree, 
+     method = "discrete", inc.nodes = FALSE, time = c(100, 85, 65, 0),
+     FADLAD = BeckLee_ages)
+
+## Running the NPMANOVA with time as a predictor
+adonis.dispRity(time_subsets)
+```
+
+Note that the function warns you that the input data was transformed into a distance matrix.
+This is reflected in the Call part of the output (`formula = dist(matrix) ~ time`).
+
+To use each time subset as a separate predictor, you can use the `matrix ~ time.subsets` formula; this is equivalent to `matrix ~ first_time_subset + second_time_subset + ...`:
+
+```{r}
+## Running the NPMANOVA with each time bin as a predictor
+adonis.dispRity(time_subsets, matrix ~ time.subsets)
+```
+
+### `geiger::dtt` model fitting in `dispRity`
+
+The `dtt` function from the `geiger` package is also often used to compare a trait's disparity observed in living taxa to the disparity of a simulated trait based on a given phylogeny.
+The `dispRity` package proposes a wrapper function for `geiger::dtt`, `dtt.dispRity` that allows the use of any disparity metric.
+Unfortunately, this implementation is slower that `geiger::dtt` (so if you're using the metrics implemented in `geiger` prefer the original version) and, as the original function, is limited to ultrametric trees (only living taxa!)...
+
+```{r}
+require(geiger)
+geiger_data <- get(data(geospiza))
+
+## Calculate the disparity of the dataset using the sum of variance
+dispRity_dtt <- dtt.dispRity(data = geiger_data$dat, metric = c(sum, variances),
+                             tree = geiger_data$phy, nsim = 100)
+
+## Plotting the results
+plot(dispRity_dtt, fig.width=8, fig.height=8)
+```
+
+Note that, like in the original `dtt` function, it is possible to change the evolutionary model (see `?geiger::sim.char` documentation).
+
+### null morphospace testing with `null.test`
+
+This test is equivalent to the test performed in @diaz2016global.
+It compares the disparity measured in the observed space to the disparity measured in a set of simulated spaces.
+These simulated spaces can be built with based on the hypothesis assumptions: for example, we can test whether our space is normal.
+
+```{r}
+set.seed(123)
+## A "normal" multidimensional space with 50 dimensions and 10 elements
+normal_space <- matrix(rnorm(1000), ncol = 50)
+
+## Calculating the disparity as the average pairwise distances
+obs_disparity <- dispRity(normal_space, metric = c(mean, pairwise.dist))
+
+## Testing against 100 randomly generated normal spaces
+(results <- null.test(obs_disparity, replicates = 100, null.distrib = rnorm))
+```
+
+Here the results show that disparity measured in our observed space is not significantly different than the one measured in a normal space.
+We can then propose that our observed space is normal!
+
+These results have an attributed `dispRity` and `randtest` class and can be plotted as `randtest` objects using the `dispRity` S3 `plot` method:
+
+```{r, fig.width=8, fig.height=8}
+## Plotting the results
+plot(results, main = "Is this space normal?")
+```
+
+For more details on generating spaces see the [`space.maker`](#Simulating-multidimensional-spaces) function tutorial.
 

### 2018-01-04 inst/gitbook/04_making-up-data.Rmd

@@
 
-The `dispRity` package also offers some advanced data simulation features to allow to test hypothesis, explore ordinate-spaces or metrics properties or simply playing around with data!
+The `dispRity` package also offers some advanced data simulation features to allow to test hypothesis, explore multidimensional spaces, metrics properties or simply playing around with data!
 All the following functions are based on the same modular architecture of the package and therefore can be used with most of the functions of the package.
@@
 The function `sim.morpho` allows to simulate discrete morphological data matrices (sometimes referred to as "cladistic" matrices).
-It allows to evolve multiple discrete characters on a given phylogenetic trees, given different models, rates, and states.
+It allows to evolve multiple discrete characters on a given phylogenetic tree, given different evolutionary models, rates, and states.
 It even allows to include proper inapplicable data to make datasets as messy as in real life!
 
-In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
+In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model and a function from which to draw the rates.
 The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
@@
 my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
-    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)
+    0.15), rates = c(rgamma, shape = 0.5, rate = 1), invariant = FALSE)
 
 ## The first few lines of the matrix
-my_matrix[1:5, 1:10]
+my_matrix[1:10, 1:10]
 
@@
 Note that this example produces a tree with a great consistency index and an identical topology to the random coalescent tree!
-Nearly too good to be true...
 
@@
 
-The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;@OReilly2017].
+The protocol implemented here to generate discrete morphological matrices is based on the ones developed in @GuillermeCooper @OReilly2016 @puttick2017uncertain and @OReilly2017.
 
  * The first `tree` argument will be the tree on which to "evolve" the characters and therefore requires branch length.
- You can generate quick and easy random Yule trees using `ape::rtree(number_of_taxa)` but I would advise to use more realistic trees for more realistic simulations based on more realistic models (really realistic then) using the function `tree.bd` from the [`diversitree`](http://www.zoology.ubc.ca/prog/diversitree/) package [@fitzjohndiversitree2012].
- * The second argument, `character` is the number of characters. Pretty straight forward.
- * The third, `states` is the proportion of characters states above two (yes, the minimum number of states is two).
+ You can generate quick and easy random Yule trees using `ape::rtree(number_of_taxa)` but I would advise to use more realistic trees for more realistic simulations based on more realistic models (really realistic then) using the function `tree.bd` from the [`diversitree`](http://www.zoology.ubc.ca/prog/diversitree/) package [@fitzjohndiversitree2012] for example.
+ * The second argument, `character` is the number of characters.
+ * The third, `states` is the proportion of characters states above two.
 This argument intakes the proportion of *n*-states characters, for example `states = c(0.5,0.3,0.2)` will generate 50% of binary-state characters, 30% of three-state characters and 20% of four-state characters.
 There is no limit in the number of state characters proportion as long as the total makes up 100%.
- * The forth, `model` is the evolutionary model for generating the character(s). More about this below.
+ * The fourth, `model` is the evolutionary model for generating the character(s). More about this below.
  * The fifth and sixth, `rates` and `substitution` are the model parameters described below as well.
@@
 It uses the four molecular states (A,C,G,T) with a unique overall evolutionary rate and a biased substitution rate towards transitions (A <-> G or C <-> T) against transvertions (A <-> C and G <-> T).
-After evolving the nucleotide, this model transforms them into binary states by converting the purines (A and G) into state 0 and the pyrimidines (C and T) into state 1.
+After evolving the nucleotides, this model transforms them into binary states by converting the purines (A and G) into state 0 and the pyrimidines (C and T) into state 1.
 This method is based on the `phyclust::seq.gen.HKY` function and was first proposed by @OReilly2016.
@@
 The models can take the following parameters:
-(1) `rates` is the evolutionary rate (i.e. the rate of changes along a branch: the evolutionary speed) and (2) `substitution` is the frequency of changes between one state or another.
+(1) `rates` is the evolutionary rate (i.e. the rate of changes along a branch; the evolutionary speed) and (2) `substitution` is the frequency of changes between one state or another.
 For example if a character can have high probability of changing (the *evolutionary* rate) with, each time a change occurs a probability of changing from state *X* to state *Y* (the *substitution* rate).
-Note that in the `"ER"` model, the substitution rate is ignore because... by definition this (substitution) rate is equal!
+Note that in the `"ER"` model, the substitution rate is ignore because by definition this rate is equal!
 
@@
 
-This type of characters can be added to the simulated matrices using the `apply.NA` function/
+This type of characters can be added to the simulated matrices using the `apply.NA` function.
 It takes, as arguments, the `matrix`, the source of inapplicability (`NAs` - more below), the `tree` used to generate the matrix and the two same `invariant` and `verbose` arguments as defined above.
@@
 not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all parameters combination end up creating "good" matrices.
-The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
+The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 10 and no scaling ($\beta$ = 1) with a rate of 50.
 
@@
 ## tree
-my_tree <- rcoal(15)
+my_tree <- rcoal(30)
 ## matrix
-morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
-    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
+morpho_mat <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
+    0.15), rates = c(rgamma, shape = 10, rate = 50), invariant = FALSE)
+
 check.morpho(morpho_mat, my_tree)
@@
 ## Simulating multidimensional spaces
-Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
-This function allows users to simulate multidimensional spaces with a certain number of properties.
-For example, it is possible to design a multidimensional space with a specific distribution on each axis, a correlation between the axes and a specific cumulative variance per axis.
+Another way to simulate data is to directly simulate a multidimensional space with the `space.maker` function.
+This function allows users to simulate spaces with a certain number of properties.
+For example, it is possible to design a space with a specific distribution on each axis, a correlation between the axes and a specific cumulative variance per axis.
 This can be useful for creating ordinated spaces for null hypothesis, for example if you're using the function `null.test` [@diaz2016global].
@@
 This function takes as arguments the number of elements (data points - `elements` argument) and dimensions (`dimensions` argument) to create the space and the distribution functions to be used for each axis.
-The distributions are passed through the `distribution` argument as... modular functions!
+The distributions are passed through the `distribution` argument as modular functions.
 You can either pass a single distribution function for all the axes (for example `distribution = runif` for all the axis being uniform) or a specific distribution function for each specific axis (for example `distribution = c(runif, rnorm, rgamma))` for the first axis being uniform, the second normal and the third gamma).
@@
                            arguments = list(list(min = 0, max = 0), NULL, NULL))
+head(plane_space)
 
@@
 ellipse_space <- space.maker(2500, 3, rnorm, cor.matrix = cor_matrix)
+head(ellipse_space)
 
@@
                                  scree = c(0.7, 0.2, 0.1))
+head(cylindrical_space)
 ```
+
+See below for the visualisation of the spaces.
 
@@
 For example, the `random.circle` function is a personalised one implemented in `dispRity`.
-This function allows to create circles based on basic trigonometry allowing to axis to covary to produce circle coordinates.
+This function allows to create circles based on trigonometry allowing two axis to covary to produce circle coordinates.
 By default, this function generates two sets of coordinates with a `distribution` argument and a minimum and maximum boundary (`inner` and `outer` respectively) to create nice sharp edges to the circle.
@@
      arguments = list(list(mean = 0), list(runif, inner = 0.5, outer = 1)))
+head(doughnut_space)
 ```
+
+See below for the visualisation of the "doughnut" space.
 
@@
 legend("topleft", legend = c("observed", "simulated"),
-        pch = c(20,21), col = c("red", "black"))
+        pch = c(20,21), col = c("red", "black"), bg = "white")
 ```
@@
 plot(test_norm1, main = "Normal (0,1)")
-plot(test_norm2, main = paste0("Normal (", round(obs_mu_sd_glob[[1]], digit = 3),
-                              ",", round(obs_mu_sd_glob[[2]], digit = 3), ")"))
-plot(test_norm3, main = "Normal (variable + correlation)")
+plot(test_norm2, main = "Normal (global observed\nmean and sd)")
+plot(test_norm3, main = "Normal (variable mean\nand sd with correlation)")
 

### 2018-01-04 paper/dispRity_paper.tex

@@
 \usepackage{dcolumn}
+\usepackage{hyphenat}
+\usepackage[T1]{fontenc}
+\usepackage{textcomp}
 \newcolumntype{d}[1]{D{.}{.}{#1}}
@@
         \item Biological data is multivariate in essence: many traits in organisms co-vary with each other in space and time.
-        This causes biologists to either reduce these variables in a manageable number of variables or, increasingly, to use multivariate toolkits.
-        One such toolkit is based on creating a multidimensional space where the variables are the axis of this space.
-        It is then possible to use a summary metrics for measuring some properties of this space for some group of species and compare them.
-        For example, if studying morphology, one can create a morphospace for two groups of species, measure some aspect of these groups (such as their spread in the space - often termed disparity) and then test whether they have a different disparity.
+        This causes biologists to either reduce these to a manageable number of variables or, increasingly, to use multivariate toolkits.
+        One such toolkit is based on creating a multidimensional space where the variables are the axes.
+        It is then possible to measure divers aspects of the distribution of some observation (e.g. species) in this space.
+        For example, if studying morphology, one can create a morphospace for two groups of species, measure the volume occupied by each of these groups and then test whether these two volumes are significantly different or not.
 
@@
         However, the definition of the space, metric and test is often dependent on the software/package's author's point of view or specific question.
-        This can unfortunately hamper researchers to apply different methods that suits best their specific questions.
+        This can unfortunately hamper researchers ability to apply different methods that best suits their specific questions.
 
@@
         This package allows users to define each step of the analysis (whether it is the space, the metric or the test) through a highly modular architecture where each definition can be passed as a function.
-        It also provides a tidy interface through the \disp object, allowing to easily run reproducible multivariate analysis.
+        It also provides a tidy interface through the \disp object, allowing users to easily run reproducible multivariate analysis.
 
@@
         Furthermore, the package contains some simulation tools (e.g to simulate complex multidimensional space or morphological data).
-        Finally it also contains a suit of utilities to work with \disp objects aimed at helping users to develop their own multidimensional metrics and/or tests.
+        Finally it also contains a suite of utilities to work with \disp objects aimed at helping users to develop their own multidimensional metrics and/or tests.
 
@@
 % Variety of multidimensional spaces, metrics and tests
-In all these analysis, each set of multivariate traits forms a multidimensional space, usually represented as matrices where the columns are the dimensions (or some transformation thereof) and rows are the elements (often species or individuals) within this space.
-These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix \citep{Close2015}, or as outputs from a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, or a multidimensional scaling \citep[MDS;][]{DonohueDim}, among others.
+In all these analysis, each set of multivariate traits forms a multidimensional space.
+This space is represented as a matrix where rows are regarded as samples or observations (e.g.
+a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g.
+an embedding, scaling, ordination, etc.
+of the variables).
+These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix (\citealt{lloyd2016estimating} and references therein; e.g.
+in \citealt{Close2015}), or as outputs from an ordination, whether it being a principal components analysis (PCA,
+\citealt{PCA}; e.g. in \citealt{zelditch2012geometric} ), a metric scaling (PCO, PCoA, \citealt{PCO} ; e.g. in \citealt{Brusatte12092008} ) or a non-metric scaling (MDS, NMDS, \citealt{MDS}
+; e.g. in \citealt{Liow2004,DonohueDim}).
 The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
-For example, when using morphological traits the multivariate space will be a morphospace, when using ecological traits it may be referred to as an ecospace or trait space.
-Once the multidimensional space is defined it is then possible to measure a specific aspect of it.
-This specific measurement (hereafter referred to as a ``disparity metric'') can in turn be defined in many different ways depending on the specific question being investigated \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015}.
-Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance \citep[e.g.][]{Brusatte12092008}, multidimensional permutation tests \citep[e.g.][]{diaz2016global} or even simply by looking at the confidence interval overlaps between disparity measurements \citep[e.g.][]{halliday2016eutherian}.
+For example, when using morphological traits the space will be a morphospace, when using ecological traits it may be referred to as an ecospace, etc.
+
+One can then measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupies more space than group B?'').
+This requires the definition of a proxy for space occupancy: the disparity metric \citep[or index;][]{Hopkins2017} which can be measured in a multitude of ways.
+For example, one could use a metric based on the variance or the range of each axis of space \citep{Wills2001, Ciampaglio2001}, a distance (e.g.
+Euclidean) measured between observations \citep{foote1993contributions,Foote29111996}, a more direct approximation of the hyper volume \citep{cornwell2006trait,DonohueDim}, or many more \citep[e.g.][]{navarro2003MDA}.
+
+Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance (
+NPMANOVA, \citealt{NPMANOVA}; e.g.
+in \citealt{Brusatte12092008}) multidimensional permutation tests ( \citealt{ManlyPermutations}; e.g.
+in \citealt{diaz2016global}) or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements.
 In summary, there are many different ways to perform each step of a multidimensional analysis, making analyses of complexity ever more complex.
@@
 % Problems with this variety
-In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods \citep[such as][for biogeography]{matzke2013biogeobears}.
+In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods.
 In practice, however, this is hampered by existing software implementations.
-Although many software packages exist for multidimensional analysis \citep[e.g.][]{bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric}, package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric and then enforce this on users by allowing little to no flexibility.
-For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of morphometric data, the disparity metric as the Procrustes variance, and uses a permutation tests to test hypotheses \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
-This is ideal in some cases, but in others may not be the most appropriate analysis.
+Although many software packages exist for multidimensional analysis \citep[e.g.][]{navarro2003MDA,bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,geiger2008,adams2013geomorph,Claddis,adams2017geometric},
+package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best it their needs (i.e.
+data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.
+For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric
+as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance)
+, and uses permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
+This is ideal for testing volume based hypothesis (e.g.
+does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g.
+does groups A and B occupy the same location?).
 This can lead to inappropriate analyses by users confined by the existing software implementations. %Oh no!
@@
 % Here comes the solution!
-The aim of the \disp package is to solve such problems by providing a completely flexible framework for studying multidimensional data.
+The aim of the \disp package is avoid such problems by providing a completely flexible framework for studying multidimensional data.
 This package is based on a highly modular architecture where each crucial decision in multidimensional analysis (which data, which metric and which test) can be provided by the user.
 It implements many commonly used metrics and definitions of disparity, as well as providing a simple interface for users to implement their own disparity functions.
-The package is described here for the use of discrete morphological data diversity analysis but can be generalised to any type of multidimensional data (see the package vignette for more examples).
+The package is described here for the use of discrete morphological data disparity analysis but can be generalised to any type of multidimensional data
+(see the glossary Table \ref{Tab:terms} for terms' equivalences between palaeobiology and ecology).
+
+\begin{landscape}
+\begin{table}
+    \begin{tabular}{p{4cm}p{5cm}p{6cm}p{6cm}}
+        \hline
+        in this manuscript & in \texttt{dispRity} & in palaeobiology & in ecology \\ 
+        \hline
+        multidimensional space & matrix ($n\times k$) & morphospace, traitspace, etc. & ecospace, function-space, etc. \\
+        elements & rows ($n$) & taxa, specimen, etc. & taxa, field sites, environments, etc. \\
+        dimensions & columns ($k$) & ordination scores, distances, etc. & ordination scores, distances, etc. \\
+        subsets & matrix ($m \times k$, with $m \leq n$) & e.g. every element in an strata or sharing the same ancestor & e.g. elements living in the same environment \\
+        disparity & a dimension-level 1 or 2 function$^1$ & disparity: e.g.
+the sum of variances \cite{Wills2001}, the average pairwise distance between taxa \cite{Foote01071994}, etc.
+& dissimilarity: e.g.
+ellipsoid volume \cite{DonohueDim}, convex hull volume \cite{cornwell2006trait}, etc.
+\\
+        \hline
+    \end{tabular}
+    \caption{Glossary and equivalences between this manuscript, the \texttt{dispRity} package and terms commonly used in palaeobiology or ecology. $^1$See Fig. \ref{Fig:levels}.}
+    \label{Tab:terms}
+\end{table}
+\end{landscape}
 
 \section{Description}
-In brief, the package takes a \texttt{matrix} object (the multidimensional space), calculates some aspect of it (the disparity metric) and allows users to analyse the resulting \disp object (see Manuals and Vignettes below).
-Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow} and Table \ref{Tab:main_fun}).
+In brief, the package takes a \texttt{matrix} object (the multidimensional space), 
+calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation.
+Some additional functions modify the space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow}).
+Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements \citep[e.g. the space can be a distance matrix][]{Close2015}.
+The matrix is always considered as the multidimensional space and no correction is applied to it (e.g.
+if ordinated and with negative eigen values, corrections should be applied prior to using the \texttt{dispRity} package.)
 
@@
 \end{figure}
-
-\begin{table}
-    \begin{tabular}{ll}
-        \hline
-        Function & Description \\ 
-        \hline
-        \texttt{geomorph.ordination} & Imports data from the \texttt{geomorph} package. \\
-        \texttt{Claddis.ordination} & Imports data from the \texttt{Claddis} package. \\
-        \texttt{custom.subsamples} & Separates data into custom subsamples. \\
-        \texttt{time.subsamples} & Separates data into time subsamples. \\
-        \texttt{boot.matrix} & Bootstraps and rarefies a matrix or a \disp object. \\
-        \disp & Calculates disparity from an matrix or a \disp object. \\
-        \texttt{summary.dispRity} & Summarises a \disp object. \\
-        \texttt{plot.dispRity} & Plots a \disp object. \\
-        \texttt{test.dispRity} & Applies statistical tests to a \disp object.\\
-        \texttt{dispRity.per.group} & Pipeline for disparity within groups analysis. \\
-        \texttt{dispRity.through.time} & Pipeline for disparity-through-time analysis. \\
-        \hline
-    \end{tabular}
-    \caption{Main functions in the \disp package.}
-    \label{Tab:main_fun}
-\end{table}
 
@@
    \includegraphics[width=0.8\textwidth]{../inst/gitbook/dispRity_fun.pdf} 
-\caption{Illustration of the different metric dimension-levels in the \disp package.
-In this example, each cell correspond to a single value (e.g.
-a 8 $\times$ 7 matrix or a vector of eight elements).}.
+\caption{Illustration of the different metric dimension-levels in the \disp package. In this example, each cell correspond to a single value (e.g. a 8 $\times$ 7 matrix or a vector of eight elements).
+A dimension-level 3 matrix would be a metric resulting in a matrix (e.g.
+the function \texttt{stats::cor} to calculate the correlation between each dimension), a dimension-level 2 metric would result in a vector (i.e.
+a distribution, e.g.
+\texttt{dispRity::variances} which calculates the variance within each dimension) and a dimension-level 1 metric would result in a single value (e.g.
+\texttt{stats::sd} which calculates the standard deviation of the input matrix)}.
 \label{Fig:levels}
@@
 
-The disparity metrics can be functions from other packages, defined by the user or from the \disp package (see Table \ref{Tab:metrics} for the currently implemented metrics).
-When multiple functions are passed to the \texttt{metric} argument, the functions are sorted by dimension-level and applied in decreasing order to the data.
-For example, if the metric is defined as \texttt{metric = c(mean, range)}, the \texttt{range} function (dimension-level 2) is first applied to the data and the function \texttt{mean} is then applied to the result (e.g.
-\texttt{mean(range(data))}).
-There is no limitation on the metric dimensions only that at least one dimension-level 2 or 1 function must be present in the metric definition (i.e.
-disparity can only be defined as distribution or a single value).
+The disparity metrics can be functions from other packages, defined by the user or from the \disp package (see Table \ref{Tab:metrics}).
+When multiple functions are passed to the \texttt{metric} argument, they are sorted by dimension-level and applied in decreasing order to the data.
+For example, if the metric is defined as \texttt{metric = c(prod, ranges)} (the hypercube volume), the \texttt{ranges} function (dimension-level 2) is first applied to data (the range of each dimensions) and the function \texttt{prod} is then applied to the results (\texttt{prod(ranges(data))}).
+One can also directly pass a function description to the \texttt{metric} argument (e.g. \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} for the average squared pairwise distance).
+Note that this function also allows to work on only a subset of dimensions via the \texttt{dimensions} argument (e.g. if only the $n$ first dimensions must be considered).
 
-% \begin{table}
-% \resizebox{\textwidth}{!}
-% {%
-%     \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
-%         name & description & dimen-sion level & definition & source \\
-%         \hline
-%         \texttt{ellipse. volume}$^1$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
-%         \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
-%         \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \\
-%         \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & \texttt{dispRity::diagonal} \\
-%         \texttt{mode.val} & The modal value & 1 & NA & \texttt{dispRity::mode.val}\\
-%         \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & \texttt{dispRity::ranges} \\
-%         \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & \texttt{dispRity::variances} \\
-%         \texttt{centroids} & The distance between each element and a fixed point$^2$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & \texttt{dispRity::centroids} \\
-%     \end{tabular}
-% }%
-%     \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean.
-$^1$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA); $^2$ by default that point is the centroid of the elements.}}
+\begin{table}
+\resizebox{\textwidth}{!}
+{%
+    \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
+        name & description & dim & definition & source \\
+        \hline
+        \texttt{ancestral. dist} & The distance between an element and its ancestor & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}$ & This package \\
+        \texttt{centroids} & The distance between each element and a fixed point$^1$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & This package \\
+        \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
+        \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
+        \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & This package \\
+        \texttt{ellipse.
+volume}$^2$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & This package; based on \cite{DonohueDim}\\
+        \texttt{mode.val} & The modal value & 1 & NA & This package\\
+        \texttt{n.ball. volume} & The hyper-spherical ($n$-ball) volume & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R$ & This package\\
+        \texttt{pairwise. dist} & The pairwise distances between elements & 2 & NA & \texttt{vegan::vegdist} \citep{oksanen2007vegan}\\        
+        \texttt{radius} & The radius of each dimensions & 2 & $|\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|$ & This package\\
+        \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & This package \\
+        \texttt{span.tree. length} & The minimal spanning tree length & 1 & $\sum(\mathrm{branch\ length})$ & \texttt{vegan::spantree} \citep{oksanen2007vegan}\\
+        \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & This package \\
+    \end{tabular}
+}%
+    % \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g.
+it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}$ - for a hyper-ellipsoid).
+$^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA).}}
     % \label{Tab:metrics}
-% \end{table}
+\end{table}
+
+\textbf{Table 1 (caption):} Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g.
+it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid).
+$^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO (\textit{not} PCA).
+\label{Tab:metrics}
+
+%TODO: Add the proper sources for the convhulln and vegan packages
 
 \subsection{Splitting the multidimensional space into subsets}
-Prior to calculating disparity, the multidimensional space can be subdivided into subsamples, typically to be compared to each other.
-For example, one may be interested in seeing how the disparity of a specific subsample of the space compares to another (e.g.
-difference among groups) or, how different subsamples change sequentially (e.g.
+Prior to calculating disparity, the space can be subdivided into subsets, typically to be compared to each other.
+For example, one may be interested in seeing how the disparity of a specific subset of the space compares to another (e.g.
+difference among groups) or, how different subsets change sequentially (e.g.
 through time).
-In essence, the original space corresponds to the overall space (e.g. a morphospace contains all the possible observed morphologies).
-Conversely, subsamples correspond to parts of the space with pooled characteristics (i.e. only a subset of the total morphospace).
+In essence, the original space corresponds to the overall space (e.g. a morphospace contains all the observed morphologies).
+Conversely, subsets correspond to parts of the space with pooled characteristics.
 
-This splitting can be done using the \texttt{custom.subsamples} or \texttt{time.subsamples} functions.
-The first takes as argument the matrix defining the multidimensional space and a list of elements to group into different subsamples.
-The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time subsamples for example) and which subsamples to create: (1) discrete time subsamples (or time-binning) or (2) continuous time subsamples (or time-slicing).
+This splitting can be done using the \texttt{custom.subsets} or \texttt{time.subsets} functions.
+The first argument is the matrix defining the space and a list of elements to group into different subsets.
+The second also takes a matrix and arguments giving the age of the taxa (a dated phylogeny of the elements present in the morphospace - see below) and which subsets to create: (1) discrete time subsets (or time-binning) or (2) continuous time subsets (or time-slicing).
 
@@
 The time-slicing method works by using a phylogeny and looking at which taxa are present at any specific point in time.
-This method thus requires the nodes to be part of the multidimensional space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
-When a slice occurs not on a tip or a node, four methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
+This method thus requires the nodes to be part of the space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
+When a slice occurs not on a tip or a node, six methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
 \begin{itemize}
-    \item ``acctran'' (accelerated transformation) where the data chosen along the branch is always the one of the descendant
-    \item ``deltran'' (delayed transformation) where the data chosen along the branch is always the one of the ancestor
-    \item ``punctuated'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor
-    \item ``gradual'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length
+    \item ``acctran'' where the data chosen along the branch is always the one of the descendant
+    \item ``deltran'' where the data chosen along the branch is always the one of the ancestor
+    \item ``random'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor
+    \item ``proximity'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length
+    \item ``equal.split'' where the data chosen along the branch is both the descendant and the ancestor with an even probability
+    \item ``gradual.split'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length
 \end{itemize}
+
+Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
+The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
+There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).
+See Guillerme \& Cooper \textit{in review} for full description.
 
@@
 Disparity measurement can be influenced by sampling \citep{Butler2012}.
-To take this source of bias into account, it is common practice to bootstrap the multidimensional space (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different subsamples).
+To take this source of bias into account, it is common practice to bootstrap the multidimensional space (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different subsets).
 Additionally, if disparity is defined as a dimension-level 1 metric, it can be useful to measure it on bootstrapped data to obtain a distribution on which to perform statistical analyses.
 
-Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the multidimensional space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).
-
-% In the case of disparity being a dimension-level 1 metric, it is important to note that the distribution resulting from a bootstrap will be a pseudo-replicated one and thus might not be suitable for some specific parametric statistical tests.
-% Conversely, it is possible to use a dimension-level 2 metric on a bootstrapped multidimensional space to avoid this problem: the disparity metric is then measured on a pseudo-replicated multidimensional space rather than the metric itself being pseudo-replicated.
+Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).
 
 Similarly, rarefaction can be achieved through the same \texttt{boot.matrix} function.
-Rarefaction can be useful when looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of subsamples with the same number of elements in each.
+Rarefaction can be useful when looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of subsets with the same number of elements in each.
 In practice, rarefaction limits the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).
@@
 The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity} (provided disparity was calculated).
-These results can then be analysed using the \texttt{test.dispRity} function for comparing subsamples or testing hypotheses.
+These results can also be analysed using the \texttt{test.dispRity} function for comparing subsets or testing hypotheses.
 
@@
 the rarefaction level).
-The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base R.
-% Note that if the disparity has been calculated on fully rarefied data (using \texttt{boot.matrix(data, rarefaction = TRUE)} prior to calculating disparity), the rarefaction curves for the whole multidimensional space (or its sub-samples) can be plotted using \texttt{rarefaction = ``plot''}.
+The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base \texttt{R}.
 
 \subsubsection{Testing hypotheses}
-The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data (e.g. by comparing the different subsamples to each other).
+The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data (e.g. by comparing the different subsets to each other).
 Similarly to the \disp function described above, this function can take any test defined by the user or from other R packages.
-The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsamples: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subsample to all the others; (3) ``sequential'' for comparing subsamples sequentially (e.g.
-first against second, second against third, etc.); (4) `all'' for comparing all the subsamples simultaneously (i.e.
-\texttt{disparity $\mathtt{\sim}$ subsamples} - this can be used for \texttt{lm} or \texttt{glm}) or (5) any list of pairs (or more) of subsamples to compare.
+The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsets: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subset to all the others; (3) ``sequential'' for comparing subsets sequentially (e.g.
+first against second, second against third, etc.); (4) `all'' for comparing all the subsets simultaneously (i.e.
+\texttt{disparity $\mathtt{\sim}$ subsets}) or (5) any list of pairs (or more) of subsets to compare.
 
-Some tests are implemented within the package such as the Bhattacharrya Coefficient \citep[\texttt{bhatt.coeff};][]{Bhattacharyya,GuillermeCooper} or a permutation test based on null hypothesised multidimensional space following \cite{diaz2016global} (\texttt{null.test}).
+Some tests are implemented within the package such as the Bhattacharrya Coefficient \citep[\texttt{bhatt.coeff};][]{Bhattacharyya,GuillermeCooper}, a permutation test based on null hypothesised multidimensional space following \cite{diaz2016global} (\texttt{null.test})
+as well as a wrapper for the \texttt{vegan::adonis} \citep{oksanen2007vegan} and \texttt{geiger::dtt} \citep{geiger2008} functions (respectively \texttt{adonis.dispRity} and \texttt{dtt.dispRity} - allowing \texttt{dispRity} objects to be directly used with any disparity metric).
 Additionally, this function also allows additional arguments such as \texttt{rarefaction} (as described above) or \texttt{correction} to adjust p-values when using multiple parametric tests.
@@
 \section{Examples}
-Multivariate analysis can be really useful for looking multiple aspects of organisms diversity together.
-For example, one classic univariate trait used in ecology and evolution can be the sheer number of species present in a ecosystem or/and through time.
+Multivariate analysis can be really useful for looking at multiple aspects of organisms diversity together.
+For example, one classic univariate trait used in ecology and evolution can be the number of species present in a ecosystem or/and through time.
 This trait can be useful to test whether some ecosystems or/and time periods have more species diversity than others.
-In this case one can then test some hypothesis such as: environments with bigger altitudes gradients contain more species.
+In this case one can then test some hypothesis such as: environments with longer altitudinal gradients contain more species.
 
-Conversely, in the case of species diversity, one can also look at another independent multivariate trait: the diversity of morphologies \citep[or disparity;][]{ruta2013}.
+Conversely, in the case of species diversity, one can also look at another independent multivariate trait: the diversity of morphologies \citep[or disparity;][]{foote1991morphological}.
 Using disparity, it is then also possible to assess whether one ecosystem or/and time period display more morphological variation.
-Following the previous example, if disparity is also higher with bigger altitudes gradients, one could argue that the species richness is caused by an increased number of morphologies adapted to more ecological niches (or something akin).
+Following the previous example, if disparity is also higher with longer altitudinal gradients, one could argue that the species richness is caused by an increased number of morphologies adapted to more ecological niches (or something akin).
 
-The following example is based on a classical palaeobiology morphological disparity analysis.
+The following example is based on a classical morphological disparity analysis.
 Note that more examples are available in the package manual (\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).
@@
 
-In this example, the multidimensional space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
-Additionally, we can define disparity as the sum of the variances on each dimension \citep{Wills1994} that will represent an aspect of the the volume of the morphospace (i.e.
+In this example, the space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
+Additionally, we can define disparity as the sum of the variances on each dimension \citep{foote1991morphological,Wills1994} that will represent an aspect of the the volume of the morphospace (i.e.
 a large sum of variances approximates a large morphospace and vice versa).
 
-\subsection{Typical disparity among groups analysis}
+\subsection{Example}
+
+The following example is based on the discrete morphological data described above. More examples, including based on other types of data are available in the package vignette.
+
+\subsubsection{Typical disparity among groups analysis}
 One typical question with such analysis would be to test whether two groups of species have a different disparity.
-For example, using the data described above, we can test whether the crown mammals (the living mammals and their descendants) are more diverse in term of morphology than the stem mammals (the fossil mammals with no living descendants).
+For example, using the data described above, we can test whether the crown mammals are more diverse in term of morphology than the stem ones.
 In other words, whether the approximation of the volume within the morphospace (the disparity as defined above) is different in crown or stem mammals.
 
-\noindent These two groups can be defined as follows (where the integers are the row number of the species in the matrix):
+\noindent These two groups can be defined using one the package's utility functions, \texttt{crown.stem} that separate the crown and stem species given a phylogeny (allowing to ignore the nodes or not):
 
-\noindent \texttt{> mammal\_groups <- list('crown' = c(16, 19:41, 45:50), 'stem' = c(1:15, 17:18, 42:44))}
+\noindent
+\texttt{> mammal\_groups <- crown.stem(BeckLee\_tree, inc.nodes = FALSE)}
 
@@
 
-\noindent \texttt{> subsamples <- custom.subsamples(data = BeckLee\_mat50, group = mammal\_groups)}\\
-\noindent \texttt{> bootstraps <- boot.matrix(subsamples)}\\
-\noindent \texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\
+\noindent
+\texttt{> subsets <- custom.subsets(data = BeckLee\_mat50, group = mammal\_groups)}\\
+\noindent
+\texttt{> bootstraps <- boot.matrix(subsets)}\\
+\noindent
+\texttt{> disparity2 <- dispRity(bootstraps, metric = c(sum, variances))}\\
 
 \noindent Which allows a finer tuning of the optional arguments in each function.
-The three arguments here are defined as follows: \texttt{data = BeckLee\_mat50} is our definition of the multidimensional space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity.
+The three arguments here are defined as follows: \texttt{data = BeckLee\_mat50} is our space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity \citep{foote1991morphological,Wills1994,Ciampaglio2001}.
 
@@
 
-\noindent \texttt{> disparity}\\
+\noindent \texttt{> disparity2}\\
 \noindent \texttt{ ---- dispRity object ---- }\\
-\noindent \texttt{2 customised subsamples for 50 elements with 48 dimensions:}\\
+\noindent \texttt{2 customised subsets for 50 elements with 48 dimensions:}\\
           \texttt{crown, stem.}\\
@@
 % mammal_groups <- list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44))
-% subsamples <- custom.subsamples(data = BeckLee_mat50, group = mammal_groups)
-% bootstraps <- boot.matrix(subsamples)
+% subsets <- custom.subsets(data = BeckLee_mat50, group = mammal_groups)
+% bootstraps <- boot.matrix(subsets)
 % disparity <- dispRity(bootstraps, metric = c(sum, variances))
 
-\noindent \texttt{> summary(disparity)}
+\noindent \texttt{> summary(disparity2)}
 
@@
 \caption{Summarising a \disp object (disparity per groups).
-$n$ is the number of elements per subsamples, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 95\% are the confidence intervals.}
+$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 97.5
+\% are the confidence intervals.}
 \label{Tab:summary_group}
@@
 
-\noindent \texttt{> plot(disparity)}
+\noindent \texttt{> plot(disparity2)}
 
@@
 
-\noindent \texttt{> test.dispRity(disparity, test = wilcox.test)}\\
+\noindent \texttt{> test.dispRity(disparity2, test = wilcox.test)}\\
 
 \noindent \texttt{[[1]]}\\
-\noindent \texttt{             statistic}\\
+\noindent \texttt{          statistic: W}\\
 \noindent \texttt{crown : stem     10000}\\
@@
 \bigskip
-As indicated by the p value, there is a significant difference in disparity between the groups.
+As indicated by the \textit{p} value, there is a significant difference in disparity between the groups.
+Note that by default the function only outputs the test's statistic, parameter (if parametric) and the \textit{p} value.
+However, the raw test results can also be output using the option \texttt{details = TRUE} in the function above.
+Additionally, the test is here performed on the pooled bootstrapped pseudo-replications which can increase the type I error.
+It is possible to compare each bootstrap in a pairwise way without pooling the data by using the \texttt{concatenate = FALSE} argument.
+The results will then be a distribution of statistic and \textit{p} values.
 Relating back to our question: yes, crown mammals display a higher diversity in morphologies than their stem counterparts (in this example and dataset!).
@@
 \subsubsection{Typical disparity-through-time analysis}
-A following up question to this observation could be to test whether this difference is due to an overall change in disparity through time or not.
-In other words, did disparity increased through time in mammals globally or is it only between stem and crown mammals.
+A subsequent question to this observation could be to test whether this difference is due to an overall change in disparity through time or not.
 Using the same definition of the multidimensional space and disparity as in the previous example, we can measure, for example, changes in disparity through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).
+Note that stratigraphic times can be generated automatically using the \texttt{get.bin.ages} utility function.
 
-\noindent \texttt{> time\_bins <- c(100.5, 66, 56, 33.9)}
+\noindent
+\texttt{> time\_bins <- c(100.5, 66, 56, 33.9)}
 
@@
 
-\noindent \texttt{> disparity <- dispRity.through.time(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, metric = c(sum, variances))}
+\noindent
+\texttt{> disparity3 <- dispRity.through.time(data = BeckLee\_mat50,}\\
+\texttt{tree = BeckLee\_tree, time = time\_bins, metric = c(sum, variances))}
 
@@
 
-\noindent \texttt{> subsamples <- time.subsamples(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, method = 'discrete')}\\
-\noindent \texttt{> bootstraps <- boot.matrix(subsamples)}\\
-\noindent \texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\
+\noindent \texttt{> time\_subsets <- time.subsets(data = BeckLee\_mat50, tree = BeckLee\_tree,}\\ \texttt{time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
+\noindent \texttt{> bootstraps2 <- boot.matrix(time\_subsets)}\\
+\noindent \texttt{> disparity4 <- dispRity(bootstraps2, metric = c(sum, variances))}\\
 
@@
 
-\noindent \texttt{> summary(disparity)}
+\noindent \texttt{> summary(disparity4)}
 
@@
 \caption{Summarising a \disp object (disparity through time).
-$n$ is the number of elements per subsamples, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 95\% are the confidence intervals.}
+$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 97.5\% are the confidence intervals.}
 \label{Tab:summary_time}
@@
 
-\noindent \texttt{> plot(disparity, type = 'continuous')}
+\noindent \texttt{> plot(disparity4, type = \textquotedbl continuous\textquotedbl)}
 
@@
 
-From the plot and the summary, disparity seems to change (increase) through time from the Cretaceous to the Eocene.
-Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, an ANOVA (Table \ref{Tab:anova}):
+Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA  \citep[PERMANOVA; ][]{permanova} through the \texttt{adonis.disprity} function that is a wrapper of the \texttt{vegan::adonis} function  \citep{oksanen2007vegan} for \texttt{dispRity} objects:
 
-\noindent \texttt{> anova(test.dispRity(disparity, test = lm, comparisons = 'all'))}
+\noindent \texttt{> test.dispRity(disparity4, test = adonis.dispRity)}
 
@@
 \centering
-\begin{tabular}{lrrrrr}
+\begin{tabular}{lrrrrrr}
   \hline
- & Df & Sum Sq & Mean Sq & F value & Pr($>$F) \\ 
+              & Df & Sum Sq & Mean Sq & F Model & R2   & Pr($>$F) \\ 
   \hline
-subsamples & 2 & 3.14 & 1.57 & 177.84 & 0.0000 \\ 
-  Residuals & 297 & 2.62 & 0.01 &  &  \\ 
+    time      & 2  &  7.50 & 3.75     & 2.06    & 0.11 & <0.01 *** \\ 
+    Residuals & 34 & 61.82 & 1.81     &         & 0.89 & \\
+    Total     & 36 & 69.32 &          &         & 1.00 & \\ 
    \hline
 \end{tabular}
-\caption{ANOVA performed on a \disp object.}
+\caption{
+Permanova output from \texttt{adonis.dispRity}:\\
+Call: vegan::adonis(formula = dist(matrix) \texttildelow time, data = disparity2, method = "euclidean"); Permutation: free; Number of permutations: 999; Terms added sequentially (first to last).
+Signif.
+codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1 ` ' 1.
+}
 \label{Tab:anova}
@@
 To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
+Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test.
+Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.
 
@@
 \subsection{Manuals and vignette}
-Supplementary information about the package is available on the project page (\url{https://TGuillerme/dispRity}).
-Information concerning each function can be found in R in the individual function manuals and a package manual with more detailed information is available online (\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).
+Supplementary information concerning the package and each function can be found in \texttt{R}, on the project page (
+\url{https://TGuillerme/dispRity}) or in the online manual (\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).
 This manual contains substantially more information and detailed examples including a tutorial for a ``classic'' disparity analysis in palaeobiology as well as an introduction to the use of this package in ecology or other disciplines.
@@
 This package also contains functions for simulating random discrete morphological matrices (\texttt{sim.morpho}) or random multidimensional spaces (\texttt{space.maker}).
-These functions are based on a similar modular architecture as that used by the \disp functions, allowing users to provide their own distribution parameters for the simulations.
-For example, \texttt{stats::rnorm} can be provided as an argument for drawing normal characters rates with \texttt{sim.morpho} or normally distributed multidimensional space with \texttt{space.maker}.
+These functions are based on a similar modular architecture as that used by the \texttt{dispRity} functions, allowing users to provide their own distribution parameters for the simulations.
+For example, \texttt{stats::rnorm} can be provided as an argument for drawing normal characters rates with \texttt{sim.morpho} or normally distributed spaces with \texttt{space.maker}.
 The discrete morphological data simulations are based on protocols from \cite{GuillermeCooper}, \cite{OReilly20160081} and \cite{puttick2017uncertain}.
-The multidimensional space simulations are based on the methods from \cite{diaz2016global}.
+The space simulations are based on the methods from \cite{diaz2016global}.
 Both functionalities are described in more details in the package manual.
@@
 Many thanks to Natalie Cooper for encouraging the development of this package and helping with the writing of this paper and the package manuals.
-Thanks to Martin Brazeau and Andrew Jackson for specific implementation ideas. %TG: add Graeme Lloyd, Dave Bapst and Emma Sherratt
-I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.
+Thanks to David Bapst, Martin Brazeau, Andrew Jackson, Graeme Lloyd and Emma Sherratt for specific implementation ideas  and to Michael Collyer, Gavin Simpson and two other anonymous reviewers for constructive and encouraging comments on the manuscript.
+I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 - 2013)/ERC Grant Agreement number 311092 awarded to Martin D.
+Brazeau and from the Discovery Project Grant DP170103227 awarded to Vera Weisbecker.
 
 \section{Funding}
-This work was funded by the European Research Council under the European Union's Seventh Framework Programme (FP/2007–2013)/ERC Grant Agreement number 311092.
+This work was funded by the European Research Council under the European Union's Seventh Framework Programme (FP/2007 - 2013)/ERC Grant Agreement number 311092 and the Discovery Project Grant number DP170103227.
 

### 2018-01-04 paper/dispRity_paper_highlighted.tex

@@
 the space can be a distance matrix][]{Close2015}.
+\hl{The matrix is always considered as the multidimensional space and no correction is applied to it (e.g.
+if ordinated and with negative eigen values, corrections should be applied prior to using the \texttt{dispRity} package.)}
 
@@
 
-\hl{The \texttt{R} script for the following example is available in the supplementary materials.}
+\hl{The following example is based on the discrete morphological data described above. More examples, including based on other types of data are available in the package vignette.}
 
@@
 \noindent Which allows a finer tuning of the optional arguments in each function.
-The three arguments here are defined as follows: \texttt{data = BeckLee\_mat50} is our space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity.
+The three arguments here are defined as follows: \texttt{data = BeckLee\_mat50} is our space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity \citep{foote1991morphological,Wills1994,Ciampaglio2001}.
 
@@
 
-From the plot and the summary, disparity seems to change (increase) through time from the Cretaceous to the Eocene.
 \hl{Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA}

### 2018-01-04 paper/response_to_reviewers.tex
Revert "v0.5.1"

--- 2017-12-27 paper/response_to_reviewers.tex
@@
 
-I am very grateful to the four referees for their helpful and constructive comments, which I believe have helped me to significantly improve the \texttt{dispRity} package and this associated package description paper.
-I have taken all of their comments on board, and respond to their points in details below.
+I am very grateful to the four referees for their helpful and constructive comments, which we believe have helped us to significantly improve the \texttt{dispRity} package and its associated description paper.
+I have taken all of their comments on board, and respond to their points below.
 For improving clarity in this document, reviewers comments are displayed in blue, my responses in normal text and the changes to the manuscript in italic.
@@
 species) in this space.
-For example, if studying morphology, one can create a morphospace for two groups of species, measure the volume occupied by each of these groups and then test whether these two volumes are significantly different or not.} lines 6-10.
+For example, if studying morphology, one can create a morphospace for two groups of species, measure the multidimensional volume occupied by each of these groups and then test whether these two volumes are significantly different or not.} line @@@
 
@@
 
-Apologies for the omission of the table summarising the metrics which is now properly included in the manuscript (see reviewer 1, minor comment \ref{table_missing})
+%TODO improve disparity definition in the abstract
+
 As reviewer 4 (major comment \ref{nebulous}) also underlines: ``disparity is a broad, nebulous concept''.
@@
 
-\textit{One can then measure how the observations are distributed within this space to answer related questions (e.g.
-``does group A occupies more space than group B?'').
+\textit{It is then possible to measure how the observations are distributed within this space to answer related questions (e.g.
+“does group A occupies more space than group B?”).
 This requires the definition of a proxy for space occupancy: the disparity metric (or index; Hopkins and Gerber, 2017) which can be measured in a multitude of ways.
@@
 Euclidean) measured between observations (Foote, 1993, 1996), a more direct approximation of the hyper volume (Cornwell et al., 2006; Donohue et al., 2013), or many more (e.g.
-Navarro, 2003).} lines 60-67.
+Navarro, 2003).} lines @@@
 
@@
 
-I have added two wrapper functions to use the popular functions \texttt{vegan::adonis} and \texttt{geiger::dtt} to the package (\texttt{dispRity.adonis} and \texttt{dispRity.dtt}).
-Also, I have generalised the distances calculations using the \texttt{vegan::vegdist} function that allows for more distances than \texttt{stats::dist} to be calculated.
-
-I've updated the manual and the example to describe these new functions:
-
-\textit{Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA (PERMANOVA; Anderson, 2001b) through the adonis.disprity function that is a wrapper of the vegan::adonis function (Oksanen et al., 2007) for dispRity objects [TABLE 4].
-To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
-Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test.
-Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.} lines 351-361.
+%TODO
+%TG: add more functions cross over with vegan and geiger
 
@@
 
-I have broaden the keywords within the limits allowed by manuscript central for MEE publications.
+%TODO
+%Increase keywords
 
@@
 
-\textit{For example, in the excellent and widely used} \texttt{geomorph} \textit{package, morphological disparity analysis uses the} \texttt{morphol.disparity} \textit{function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance), and uses permutation tests (Zelditch et al., 2012; Adams and Otarola-Castillo, 2013; Adams et al., 2017).
+\textit{For example, in the excellent and widely used} \texttt{geomorph} \textit{package, morphological disparity analysis uses the} \texttt{morphol.disparity} \textit{function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance), and uses permutation tests(Zelditch et al., 2012; Adams and Otarola-Castillo, 2013; Adams et al., 2017).
 This is ideal for testing volume based hypothesis (e.g.
 does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g.
-does groups A and B occupy the same location?).} lines 85-93.
+does groups A and B occupy the same location?).} lines @@@
 
@@
 this metric could be used as a proxy for the average space occupancy per unit element - e.g.
-per taxa).
+per taxon).
 This would not be doable directly by using the implemented \texttt{ellipse.volume} function but would require the user to use a function like:\\
@@
 I meant both matrices (as this reviewers rightly points out they are suitable for ordination).
-I have made this more clear in the figure caption.
+I have made this more clear in the figure.
 On a more general note, the package now allows users to input any matrix (although the package does not provide warranty that the provided matrix is meaningful to the user's specific analysis).
-I have added an example to the \texttt{dispRity} manual based on the \texttt{eurodist R} base dataset.
 
@@
 Reviewer 4 (minor comment \ref{remove_simulations}) suggested completely removing this section which conflicts directly with this suggestion.
-The simulation protocols implemented here are already developed (and tested) in details in previous publications along with novel features that are part of future publications.
+The simulation protocols implemented here are already developed (and tested) in details in previous publications with new novel features attached to future publication.
 For example the discrete data simulation protocol is described and detailed in Guillerme \& Cooper 2016 \textit{Mol Phyl Evol}, O’Reilly et al.
@@
 2016 \textit{Nature} and will be used along with additional simulation features not implemented yet (including shape evolution) as part of a project currently in progress.
-Due to the space constraint for application manuscripts in MEE, I am inclined to follow reviewer's 4 suggestion rather than this one.
+I am therefore inclined to follow reviewer's 4 suggestion rather than this one.
 
@@
 The manual now has a chapter entirely dedicated to the data simulation functions with emphasis on both the methods used and the algorithms implemented (and illustrated with advanced examples).
+%TODO: check typos in there.
 
@@
 
-I have however modified the available models to better describe this limitation and I have added two new models that would approximate ancestral reconstruction along edges.
-I have modified the name of the two previous ``punctuated'' and ``gradual'' models to ``random'' and ``proximity'' to reflect more accurately how they select either the ACCTRAN or DELTRAN information.
+I have however modified the available models to better describe this limitation and I have added two new models that would approximate ancestral reconstruction along edges
+I have modified the name of the two previous ``punctuated'' and ``gradual'' algorithms to ``random'' and ``proximity'' to reflect more accurately how they select either the ACCTRAN or DELTRAN information.
 I've added two more accurate models (now named ``equal.split'' and ``gradual.split'') to reflect the changes along the branches.
@@
 These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).
-Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).} lines 166-180.
+Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).} lines @@@
 
@@
 
-I also agree with this comment: this method is a trade-off between the number of slices.
+I also agree with this comment: this method creates a trade-off between the number of slices and the type II errors: %TOCHEK@@@
 (1) too few slices decrease resolution and veer towards the time binning (which could increase "significance" by comparing too different time subsamples);
-(2) too many slices will increase the resolution be will create much more pseudo replication and thus increase type I error.
+(2) too many slices will increase the resolution be will create much more pseudo replication and thus increase type II error.
+
+A more detailed description and test of these models might be out of the scope of this package description paper but is currently in development for a special issue paper to be submitted to \textit{Palaeontology} as part as a special symposium for the Palaeontological Association annual meeting 2017 in collaboration with Natalie Cooper (\url{http://www.palass.org/meetings-events/annual-meeting/2017/annual-meeting-2017-london-symposium-abstracts}).
 
@@
 
-\textit{Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).} line 177-180.
-
-A more detailed description and test of these models might be out of the scope of this package description paper but is currently in development for a special issue paper to be submitted to \textit{Palaeontology} as part as a special symposium for the Palaeontological Association annual meeting 2017 in collaboration with Natalie Cooper.
-The full (un-reviewed an un-compiled) manuscript is available on Natalie's Github page (\url{https://github.com/nhcooper123/time-slice/blob/master/manuscript/time-slice.tex}).
+\textit{Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).} line @@@
 
@@
 
-I've reflected these changes in the manual with a subsection in the chapter ``Specific tutorials - 4.8'' (based on a distance matrix) and in manuscript as follows:
+I've reflected these changes in the manual with a subsection in the chapter ``Specific tutorials'' and in manuscript as follows:
 
 \textit{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements (e.g.
-the multidimensional space can be a distance matrixl; Close et al., 2015).} lines 109-111.
+the multidimensional space can be a distance matrixl; Close et al., 2015).} @@@
 
@@
 This was a conscious choice: due to the really high number of metrics available (especially when combining different ones), I designed the package to be modular hence allowing users to implement their own metrics (see response to reviewer 1 minor comment \ref{table_missing}).
-I have add the hypercube volume description as an example of user defined metric in the manuscript:
+I have precise this using the hypercube volume example in the manuscript:
 
-\textit{When multiple functions are passed to the} \texttt{metric} \textit{argument, they are sorted by dimension-level and applied in decreasing order to the data.
-For example, if the metric is defined as} \texttt{metric = c(prod, ranges)} \textit{(the hypercube volume), the} \texttt{ranges} \textit{function (dimension-level 2) is first applied to data (the range of each dimensions) and the function} \texttt{prod} \textit{is then applied to the results (}\texttt{prod(ranges(data))}\textit{).
-One can also directly pass a function description to the} \texttt{metric}\textit{ argument (e.g.} \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} \textit{ for the average squared pairwise distance).} lines 127-132.
+\textit{When multiple functions are passed to the} \texttt{metric} \textit{argument, the functions are sorted by dimension-level and applied in decreasing order to the data.
+For example, if the metric is defined as} \texttt{metric = c(prod, ranges)} \textit{(the hyperbox volume), the} \texttt{ranges} \textit{function (dimension-level 2) is first applied to data (calculating the range of each dimension) and the function} \texttt{prod} \textit{is then applied to the results (e.g.} \texttt{prod(ranges(data))}\textit{).
+If one needs to define the metric as something more complex, say the hypercube volume, one can directly pass a function description to the} \texttt{metric} \textit{argument (e.g.} \texttt{metric = function(x) min(ranges(x)) * ncol(x)}\textit{).}
 
-I have however added several additional new metrics ones to the newest version of the package including: (1) the minimal spanning tree length (\texttt{span.tree.length} - as suggested), (2) the ancestral distance which is equivalent to the distance from centroid but were the ``centroid'' are the coordinates of each element's ancestors (\texttt{ancestral.dist}), (3) the pairwise distances (\texttt{pairwise.dist} - based on \texttt{vegan:vegdist}), (4) the radius which equivalent to the distance from centroids but for the dimensions (i.e.
-the distance between each dimensions coordinates and the mean dimension's coordinates - \texttt{radius}) and (5) the $n$-ball volume which is an extension of the \texttt{ellipse.volume} function for a sphere (the true $n$-ball volume) or an ellipsoid (\texttt{n.ball.volume} - see Table 2 in the new version of the manuscript).
+I have however added several additional ones to the newest version of the package including the minimal spanning tree length (\texttt{span.tree.length} - as suggested), the ancestral distance which is equivalent to the distance from centroid but were the ``centroid'' are the coordinates of each element's ancestors (\texttt{ancestral.dist}), the pairwise distances (\texttt{pairwise.dist}), the radius which equivalent to the distance from centroids but for the dimensions (i.e.
+the distance between each dimensions coordinates and the mean dimension's coordinates - \texttt{radius}) and the $n$-ball volume which is an extension of the \texttt{ellipse.volume} function for a sphere (the true $n$-ball volume) or an ellipsoid (\texttt{n.ball.volume}).
 
@@
 This is again an excellent suggestion!
-I will implement the possibility to pass a probability matrix to the \texttt{boot.type} argument in the \texttt{boot.matrix} function where the probabilities can be reflecting occurences, regions, etc.
-I will also implement the ``fair'' algorithm mentioned in Kotric and Knoll (2015, Paleobiology, 41, 68–).
+I will develop more complex rarefaction algorithms based on sampling and occurrence based on methods applied empirically by Kotric and Knoll 2015 \textit{Palaeobiology} and Butler et al 2012 \textit{Evolution}.
+I've added a note
+
+% %TG: TODO maybe!
 
@@
 
-Similarly to previous comment, I will also add the permutation test based on the \texttt{PAST} package in a future version of the package.
-Furthermore, I have added an illustrative example in the \texttt{plot.dispRity} manual on how to add disparity data (continuous or discrete) to the \texttt{geoscalePlot} and \texttt{geoscaleBox} functions thus allowing users to make nice ready-to-publish figures using the \texttt{geoscale} package.
+I have added an illustrative example in the \texttt{plot.dispRity} manual on how to add disparity data (continuous or discrete) to the \texttt{geoscalePlot} and \texttt{geoscaleBox} functions thus allowing users to make ready-to-publish figures.
 
@@
 
-\textit{Note that this function also allows to work on only a subset of dimensions via the} \texttt{dimensions} \textit{argument (e.g.
-if only the $n$ first dimensions must be considered).} lines 132-133.
+\textit{Note that this function also allows to work on only a subset of dimensions via the} \texttt{dimensions} \textit{argument (e.g. if only the $n$ first dimensions must be considered).} l. @@@:
 
@@
 
-I've modified the sentence in bracket to
-``\textit{a dated phylogeny of the elements present in the morphospace - see below}'' (line @@@).
-
-I've also changed the code to allow non-phylogenetic dating (e.g.
-just adding the First/Last Occurrence Dates of all the taxa) but I've only mentioned it in the function manual and the vignette due to space restriction in this manuscript.
+I've added ``\textit{see below}'' to the sentence (line @@@).
+Following this comment, I've also changed the code since, indeed, a phylogeny should not always be required!
+The function can now works without it as long as First/Last occurrence data are provided for each taxa.
 
@@
 in Díaz et al.
-2016) or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements.}
+2016) or even simply by looking at the confidence interval overlaps between disparity measurements (which, when not based on rigorous statistic methods can often leads to serious inaccuracy).}
 

### 2018-01-04 paper/dispRity_paper_highlighted.tex
Revert "v0.5.1"

--- 2017-12-27 paper/dispRity_paper_highlighted.tex
@@
         One such toolkit is based on creating a multidimensional space where the variables are the \hl{axes}.
-        It is then possible to \hl{measure divers aspects of the distribution of some observation (e.g. species) in this space.
-        For example, if studying morphology, one can create a morphospace for two groups of species, measure the volume occupied by each of these groups and then test whether these two volumes are significantly different or not}.
+        It is then possible to \hl{measure divers aspects of the distribution of some observation (e.g. species) in this space}.
+        For example, if studying morphology, one can create a morphospace for two groups of species, measure \hl{the multidimensional volume occupied by each of these groups and then test whether these two volumes are significantly different or not}.
 
@@
 One solution to this problem is to analyse these data in a multivariate framework \citep[e.g.][]{price2015predation,diaz2016global}.
-Such analyses aim to capture the complex multidimensionality of biological data, while still providing outputs that are interpretable.
-These multivariate analyses can be used to investigate changes in morphological diversity through time \citep[e.g][]{Close2015}, competitive replacement scenarios \citep[e.g.][]{Brusatte12092008}, relationships among form and function \citep[e.g.][]{diaz2016global} and even to describe the entirety of possible shapes for a group of organisms \citep[e.g.][]{raup1966geometric}.
+Such analyses aim to capture the complex multidimensionality of biological data, while still providing outputs that are interpretable in our physical world restricted to three dimensions.
+These multivariate analyses can be used to investigate changes in morphological diversity through time \citep{Close2015}, competitive replacement scenarios \citep{Brusatte12092008}, relationships among form and function \citep{diaz2016global} and even to describe the entirety of possible shapes for a group of organisms \citep{raup1966geometric}.
 The biological variables in such analyses are equally diverse, including morphological traits (discrete traits like the presence or absence of a character, e.g.
-\citealt{Close2015}; and continuous traits such as lengths, e.g.
+\citealt{brusattedinosaur2012}; and continuous traits such as lengths, e.g.
 \citealt{price2015predation}), life history traits \citep[e.g.][]{diaz2016global}, or even ecosystem properties \citep[e.g.][]{DonohueDim}.
@@
 The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
-For example, when using morphological traits the space will be a morphospace, when using ecological traits it may be referred to as an ecospace, etc.
+For example, when using morphological traits the multivariate space will be a morphospace, when using ecological traits it may be referred to as an ecospace or trait space.
 
-\hl{One can then measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupies more space than group B?'').
+\hl{It is then possible to measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupies more space than group B?'').
 This requires the definition of a proxy for space occupancy: the disparity metric}
@@
 \citealt{diaz2016global}
-\hl{) or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements.}
+\hl{)}
+or even simply by looking at the confidence interval overlaps between disparity measurements
+\hl{().}
 In summary, there are many different ways to perform each step of a multidimensional analysis, making analyses of complexity ever more complex.
@@
 % Problems with this variety
-In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods.
+In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods \citep[such as][for biogeography]{matzke2013biogeobears}.
 In practice, however, this is hampered by existing software implementations.
-Although many software packages exist for multidimensional analysis \citep[e.g.][]{navarro2003MDA,bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,geiger2008,adams2013geomorph,Claddis,adams2017geometric},
+Although many software packages exist for multidimensional analysis \citep[e.g.][]{navarro2003MDA,bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric},
 \hl{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best it their needs (i.e.
@@
 % Here comes the solution!
-The aim of the \disp package is avoid such problems by providing a completely flexible framework for studying multidimensional data.
+The aim of the \disp package is to solve such problems by providing a completely flexible framework for studying multidimensional data.
 This package is based on a highly modular architecture where each crucial decision in multidimensional analysis (which data, which metric and which test) can be provided by the user.
 It implements many commonly used metrics and definitions of disparity, as well as providing a simple interface for users to implement their own disparity functions.
-The package is described here for the use of discrete morphological data disparity analysis but can be generalised to any type of multidimensional data
-\hl{(see the glossary Table}
-\ref{Tab:terms}
-\hl{for terms' equivalences between palaeobiology and ecology).}
+The package is described here for the use of discrete morphological data diversity analysis but can be generalised to any type of multidimensional data (see the package vignette for more examples).
+
+\section{Description}
+In brief, the package takes a \texttt{matrix} object (the multidimensional space), 
+\hl{calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation}
+(see Manuals and Vignettes below).
+Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow}).
+\hl{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements} \citep[e.g.
+the multidimensional space can be a distance matrix][]{Close2015}.
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{workflowsvg.pdf} 
+\caption{\disp package workflow: rectangles represent matrices; ellipses represent functions; plain black arrows indicate input/output; dashed grey arrows indicate output (though the summary, plot, and test function cannot be applied if no disparity has been calculated).}
+\label{Fig:workflow}
+\end{figure}
+
+% \begin{table}
+%     \begin{tabular}{ll}
+%         \hline
+%         Function & Description \\ 
+%         \hline
+%         \texttt{geomorph.ordination} & Imports \hl{shape} data from the \texttt{geomorph} package. \\
+%         \texttt{Claddis.ordination} & Imports \hl{discrete morphological} data from the \texttt{Claddis} package. \\
+%         \texttt{custom.subsets} & Separates data into custom subsets. \\
+%         \texttt{time.subsets} & Separates data into time subsets. \\
+%         \texttt{boot.matrix} & Bootstraps and rarefies a matrix or a \disp object. \\
+%         \disp & Calculates disparity from an matrix or a \disp object. \\
+%         \texttt{summary.dispRity} & Summarises a \disp object. \\
+%         \texttt{plot.dispRity} & Plots a \disp object. \\
+%         \texttt{test.dispRity} & Applies statistical tests to a \disp object.\\
+%         \texttt{dispRity.per.group} & Pipeline for \hl{disparity-within-groups} analysis. \\
+%         \texttt{dispRity.through.time} & Pipeline for disparity-through-time analysis. \\
+%         \hline
+%     \end{tabular}
+%     \caption{Main functions in the \disp package.}
+%     \label{Tab:main_fun}
+% \end{table}
 
@@
 \end{landscape}
-
-\section{Description}
-In brief, the package takes a \texttt{matrix} object (the multidimensional space), 
-\hl{calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation}.
-Some additional functions modify the space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow}).
-\hl{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements} \citep[e.g.
-the space can be a distance matrix][]{Close2015}.
-
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=1\textwidth]{workflowsvg.pdf} 
-\caption{\disp package workflow: rectangles represent matrices; ellipses represent functions; plain black arrows indicate input/output; dashed grey arrows indicate output (though the summary, plot, and test function cannot be applied if no disparity has been calculated).}
-\label{Fig:workflow}
-\end{figure}
 
@@
 
-The disparity metrics can be functions from other packages, defined by the user or from the \disp package (see Table \ref{Tab:metrics}).
-When multiple functions are passed to the \texttt{metric} argument, they are sorted by dimension-level and applied in decreasing order to the data.
-For example, if the metric is defined as \texttt{metric = c(\hl{prod}, ranges)} \hl{(the hypercube volume)}, the \texttt{ranges} function (dimension-level 2) is first applied to data \hl{(the range of each dimensions)} and the function \hl{\texttt{prod}} is then applied to the results (\hl{\texttt{prod(ranges(data))}}).
-\hl{One can also directly pass a function description to the \texttt{metric} argument (e.g. \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} for the average squared pairwise distance).}
+The disparity metrics can be functions from other packages, defined by the user or from the \disp package (see Table \ref{Tab:metrics} for the currently implemented metrics).
+When multiple functions are passed to the \texttt{metric} argument, the functions are sorted by dimension-level and applied in decreasing order to the data.
+For example, if the metric is defined as \texttt{metric = c(\hl{prod}, ranges)} \hl{(the hyperbox volume)}, the \texttt{ranges} function (dimension-level 2) is first applied to data \hl{(calculating the range of each dimension)} and the function \hl{\texttt{prod}} is then applied to the results (e.g.
+\hl{\texttt{prod(ranges(data))}}).
+\hl{If one needs to define the metric as something more complex, say the hypercube volume, one can directly pass a function description to the \texttt{metric} argument (e.g.
+\texttt{metric = function(x) min(ranges(x)) * ncol(x)}).}
+There is no limitation on the metric dimensions only that at least one dimension-level 2 or 1 function must be present in the metric definition (i.e.
+disparity can only be defined as distribution or a single value).
 \hl{Note that this function also allows to work on only a subset of dimensions via the \texttt{dimensions} argument (e.g. if only the $n$ first dimensions must be considered).}
@@
 it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid).
-$^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO (\textit{not} PCA).
+$^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA).
 \label{Tab:metrics}
@@
 \subsection{Splitting the multidimensional space into subsets}
-Prior to calculating disparity, the space can be subdivided into subsets, typically to be compared to each other.
+Prior to calculating disparity, the multidimensional space can be subdivided into subsets, typically to be compared to each other.
 For example, one may be interested in seeing how the disparity of a specific subset of the space compares to another (e.g.
@@
 In essence, the original space corresponds to the overall space (e.g. a morphospace contains all the observed morphologies).
-Conversely, subsets correspond to parts of the space with pooled characteristics.
+Conversely, subsets correspond to parts of the space with pooled characteristics (i.e. only a subset of the total morphospace).
 
@@
 \hl{The first argument is the matrix}
-defining the space and a list of elements to group into different subsets.
-The second also takes a matrix and arguments giving the age of the taxa 
-\hl{(a dated phylogeny of the elements present in the morphospace - see below)}
-and which subsets to create: (1) discrete time subsets (or time-binning) or (2) continuous time subsets (or time-slicing).
+defining the multidimensional space and a list of elements to group into different subsets.
+The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time subsets for example,
+\hl{see below}
+) and which subsets to create: (1) discrete time subsets (or time-binning) or (2) continuous time subsets (or time-slicing).
 
@@
 The time-slicing method works by using a phylogeny and looking at which taxa are present at any specific point in time.
-This method thus requires the nodes to be part of the space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
+This method thus requires the nodes to be part of the multidimensional space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
 When a slice occurs not on a tip or a node, \hl{six} methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
 \begin{itemize}
-    \item ``acctran'' where the data chosen along the branch is always the one of the descendant
-    \item ``deltran'' where the data chosen along the branch is always the one of the ancestor
+    \item ``acctran'' (accelerated transformation) where the data chosen along the branch is always the one of the descendant
+    \item ``deltran'' (delayed transformation) where the data chosen along the branch is always the one of the ancestor
     \item \hl{``random''} where the data chosen along the branch is randomly chosen between the descendant or the ancestor
@@
 The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
-There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).
-See Guillerme \& Cooper \textit{in review} for full description.}
+These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).
+Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).
+See Guillerme \& Cooper \textit{in review} for full description}
 
@@
 
-Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).
+Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the multidimensional space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).
+
+% In the case of disparity being a dimension-level 1 metric, it is important to note that the distribution resulting from a bootstrap will be a pseudo-replicated one and thus might not be suitable for some specific parametric statistical tests.
+% Conversely, it is possible to use a dimension-level 2 metric on a bootstrapped multidimensional space to avoid this problem: the disparity metric is then measured on a pseudo-replicated multidimensional space rather than the metric itself being pseudo-replicated.
 
@@
 The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity} (provided disparity was calculated).
-These results can also be analysed using the \texttt{test.dispRity} function for comparing subsets or testing hypotheses.
+These results can then be analysed using the \texttt{test.dispRity} function for comparing subsets or testing hypotheses.
 
@@
 The \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow users to set which central tendency and which quantiles should be represented.
+% as common arguments which central tendency (e.g.
+if central tendency is the mean and disparity was calculated as the sum of the variances for each dimensions, the central disparity tendency would be the mean of the sum of the variances) and the quantiles to represent (in the same example, if the quantile chosen is the 50th, these functions will also display 50\% of the sum of variances around the mean).
 Additionally, the \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves (in the case of disparity-through-time analysis for example) and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
@@
 the rarefaction level).
-The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base \texttt{R}.
+The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base R.
+% Note that if the disparity has been calculated on fully rarefied data (using \texttt{boot.matrix(data, rarefaction = TRUE)} prior to calculating disparity), the rarefaction curves for the whole multidimensional space (or its sub-samples) can be plotted using \texttt{rarefaction = ``plot''}.
 
@@
 first against second, second against third, etc.); (4) `all'' for comparing all the subsets simultaneously (i.e.
-\texttt{disparity $\mathtt{\sim}$ subsets} - this can be used for \texttt{lm} class functions) or (5) any list of pairs (or more) of subsets to compare.
+\texttt{disparity $\mathtt{\sim}$ subsets} - this can be used for \texttt{lm} or \texttt{glm}) or (5) any list of pairs (or more) of subsets to compare.
 
@@
 
-Conversely, in the case of species diversity, one can also look at another independent multivariate trait: the diversity of morphologies \citep[or disparity;][]{foote1991morphological}.
+Conversely, in the case of species diversity, one can also look at another independent multivariate trait: the diversity of morphologies \citep[or disparity;][]{ruta2013}.
 Using disparity, it is then also possible to assess whether one ecosystem or/and time period display more morphological variation.
@@
 
-The following example is based on a classical morphological disparity analysis.
+The following example is based on a classical palaeobiology morphological disparity analysis.
 Note that more examples are available in the package manual (\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).
@@
     \item \texttt{BeckLee\_mat50}: an ordinated matrix for 50 mammals based on the distance between discrete morphological characters.
-    \item \texttt{BeckLee\_mat99}: the same matrix \texttt{BeckLee\_mat50} with the reconstruction of their 49 ancestors.
+    \item \texttt{BeckLee\_mat99}: the same matrix \texttt{BeckLee\_mat50} but also containing the reconstruction of their 49 ancestors.
     \item \texttt{BeckLee\_tree}: a chronogram with the 50 mammal species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
@@
 
-In this example, the space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
+In this example, the multidimensional space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
 Additionally, we can define disparity as the sum of the variances on each dimension \citep{foote1991morphological,Wills1994} that will represent an aspect of the the volume of the morphospace (i.e.
@@
 
-\hl{The \texttt{R} script for the following example is available in the supplementary materials.}
+The \texttt{R} script for the following example is available in the supplementary materials.
 
@@
 \noindent Which allows a finer tuning of the optional arguments in each function.
-The three arguments here are defined as follows: \texttt{data = BeckLee\_mat50} is our space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity.
+The three arguments here are defined as follows: \texttt{data = BeckLee\_mat50} is our definition of the multidimensional space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity.
 
@@
 question to this observation could be to test whether this difference is due to an overall change in disparity through time or not.
+In other words, did disparity increase through time in mammals globally or is it only between stem and crown mammals.
 Using the same definition of the multidimensional space and disparity as in the previous example, we can measure, for example, changes in disparity through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).
@@
 \caption{Summarising a \disp object (disparity through time).
-$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 97.5\% are the confidence intervals.}
+$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 95\% are the confidence intervals.}
 \label{Tab:summary_time}
@@
 From the plot and the summary, disparity seems to change (increase) through time from the Cretaceous to the Eocene.
-\hl{Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA}
- \citep[PERMANOVA; ][]{permanova}
-\hl{through the \texttt{adonis.disprity} function that is a wrapper of the \texttt{vegan::adonis} function}
- \citep{oksanen2007vegan}
-\hl{ for \texttt{dispRity} objects:}
+Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA \citep[PERMANOVA; ][]{permanova} through the \texttt{adonis.disprity} function that is a wrapper of the \texttt{vegan::adonis} function \citep{oksanen2007vegan} for \disp objects:
 
@@
 To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
-\hl{Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test.
-Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.}
+\hl{Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test}.
 
@@
 \subsection{Manuals and vignette}
-\hl{Supplementary information concerning the package and each function can be found in \texttt{R}, on the project page (}
-\url{https://TGuillerme/dispRity}
-\hl{) or in the online manual (}
-\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}
-\hl{).}
+Supplementary information about the package is available on the project page (\url{https://TGuillerme/dispRity}).
+Information concerning each function can be found in R in the individual function manuals and a package manual with more detailed information is available online (\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).
 This manual contains substantially more information and detailed examples including a tutorial for a ``classic'' disparity analysis in palaeobiology as well as an introduction to the use of this package in ecology or other disciplines.
@@
 \hl{These functions are based on a similar modular architecture as that used by the \texttt{dispRity} functions, allowing users to provide their own distribution parameters for the simulations.}
-\hl{For example, \texttt{stats::rnorm} can be provided as an argument for drawing normal characters rates with \texttt{sim.morpho} or normally distributed spaces with \texttt{space.maker}.}
+\hl{For example, \texttt{stats::rnorm} can be provided as an argument for drawing normal characters rates with \texttt{sim.morpho} or normally distributed multidimensional space with \texttt{space.maker}.}
 \hl{The discrete morphological data simulations are based on protocols from }
 \cite{GuillermeCooper}, \cite{OReilly20160081} and \cite{puttick2017uncertain}.
-\hl{The space simulations are based on the methods from}
+\hl{The multidimensional space simulations are based on the methods from}
 \cite{diaz2016global}.

### 2018-01-04 R/custom.subsets.R

@@ -3,7 +3,7 @@
 #'
 #' @description Splits the data into a customized subsets list.
 #'
-#' @param data A \code{matrix} (see details).
+#' @param data A \code{matrix}.
 #' @param group Either a \code{list} of row numbers or names to be used as different groups or a \code{data.frame} with the same \eqn{k} elements as in \code{data} as rownames. If \code{group} is a \code{phylo} object matching \code{data}, groups are automatically generated as clades.
 #'
 #' @return
@@ -15,8 +15,6 @@
 #' Use \link{summary.dispRity} to summarise the \code{dispRity} object.
 #' 
 #' @details
-#' The data is considered as the multidimensional space with rows as elements and columns as dimensions and is not transformed (e.g. if ordinated with negative eigen values, no correction is applied to the matrix).
-#' 
 #' Note that every element from the input data can be assigned to multiple groups!
 #'
 #' @examples

### 2018-01-11 paper/response_to_reviewers.tex
Response to R1 checked

--- 2018-01-04 paper/response_to_reviewers.tex
@@
 I am very grateful to the four referees for their helpful and constructive comments, which I believe have helped me to significantly improve the \texttt{dispRity} package and this associated package description paper.
-I have taken all of their comments on board, and respond to their points in details below.
-For improving clarity in this document, reviewers comments are displayed in blue, my responses in normal text and the changes to the manuscript in italic.
-Also, I've uploaded two versions of the revised manuscript: both have the exact same content but one has all the changes and additions to the main text highlighted in yellow to help with reviewing.
+I have taken all of their comments on board (though some will be implemented in later versions), and respond to their points in details below.
+For improving clarity in this document, reviewers' comments are displayed in blue, my responses in normal text and the changes to the manuscript in italic.
+Additionally, I've uploaded two versions of the revised manuscript: both have the exact same content but one has all the changes and additions to the main text highlighted in yellow to help with reviewing.
+All the changes implemented in the package and the vignette are available on the release branch under version 0.5.1.
 
@@
 
-\textit{It is then possible to measure divers aspects of the distribution of some observation (e.g.
+\textit{It is then possible to measure diverse aspects of the distribution of some observation (e.g.
 species) in this space.
@@
 
-Apologies for the omission of the table summarising the metrics which is now properly included in the manuscript (see reviewer 1, minor comment \ref{table_missing})
+Apologies for the omission of the table summarising the metrics which is now properly included in the manuscript (see reviewer 1, minor comment \ref{table_missing}).
 As reviewer 4 (major comment \ref{nebulous}) also underlines: ``disparity is a broad, nebulous concept''.
@@
 \textit{One can then measure how the observations are distributed within this space to answer related questions (e.g.
-``does group A occupies more space than group B?'').
+``does group A occupy more space than group B?'').
 This requires the definition of a proxy for space occupancy: the disparity metric (or index; Hopkins and Gerber, 2017) which can be measured in a multitude of ways.
@@
 To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
-Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test.
+Note that in this case, the function outputs different warnings on the usage of such test and the eventual data not used in the test.
 Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.} lines 359-369.
@@
 Using this disparity metric is appropriate for testing volume based hypothesis (e.g.
-Does group A occupies more morphospace than group B?) but does not allow to test position based hypothesis (e.g.
-Does group A occupies a different part of the morphospace than group B).
+Does group A occupy more morphospace than group B?) but does not allow to test position based hypothesis (e.g.
+Does group A occupy a different part of the morphospace than group B).
 For example, one might find out that group A and B have the same volume but that doesn't tell us if they occupy the same position in the morphospace.
@@
 I sincerely apologise for the omission of the Table! It is now included in the new version of the manuscript with the currently implemented metrics description.
-I would also like to underline that although not all usual disparity metrics are not represented in the table and in the package (by far!) the idea of the \texttt{dispRity} package and specifically the \texttt{metric} argument in the \texttt{dispRity} function is to allow users to come up with their \textit{own} metric.
+I would also like to underline that although not all usual disparity metrics are represented in the table and in the package (by far!) the idea of the \texttt{dispRity} package and specifically the \texttt{metric} argument in the \texttt{dispRity} function is to allow users to come up with their \textit{own} metric.
 
@@
 These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).
-Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).} lines 169-183.
+Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).} lines 169-183.
 
@@
 
-A more detailed description and test of these models might be out of the scope of this package description paper but is currently in development for a special issue paper to be submitted to \textit{Palaeontology} as part as a special symposium for the Palaeontological Association annual meeting 2017 in collaboration with Natalie Cooper.
+A more detailed description and test of these models might be out of the scope of this package description paper but is currently in development for a special issue paper in \textit{Palaeontology} (in review) as part as a special symposium for the Palaeontological Association annual meeting 2017 in collaboration with Natalie Cooper.
 The full (un-reviewed an un-compiled) manuscript is available on Natalie's Github page (\url{https://github.com/nhcooper123/time-slice/blob/master/manuscript/time-slice.tex}).
@@
 \textit{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements (e.g.
-the multidimensional space can be a distance matrixl; Close et al., 2015).} lines 109-111.
+the multidimensional space can be a distance matrix; Close et al., 2015).} lines 109-111.
 
@@
 
-I have however added several additional new metrics ones to the newest version of the package including: (1) the minimal spanning tree length (\texttt{span.tree.length} - as suggested), (2) the ancestral distance which is equivalent to the distance from centroid but were the ``centroid'' are the coordinates of each element's ancestors (\texttt{ancestral.dist}), (3) the pairwise distances (\texttt{pairwise.dist} - based on \texttt{vegan:vegdist}), (4) the radius which equivalent to the distance from centroids but for the dimensions (i.e.
+I have however added several additional metrics to the newest version of the package including: (1) the minimal spanning tree length (\texttt{span.tree.length} - as suggested), (2) the ancestral distance which is equivalent to the distance from centroid but were the ``centroid'' are the coordinates of each element's ancestors (\texttt{ancestral.dist}), (3) the pairwise distances (\texttt{pairwise.dist} - based on \texttt{vegan::vegdist}), (4) the radius, equivalent to the distance from centroids but for the dimensions rather than the elements (i.e.
 the distance between each dimensions coordinates and the mean dimension's coordinates - \texttt{radius}) and (5) the $n$-ball volume which is an extension of the \texttt{ellipse.volume} function for a sphere (the true $n$-ball volume) or an ellipsoid (\texttt{n.ball.volume} - see Table 2 in the new version of the manuscript).
@@
 
-This is again an excellent suggestion!
-I will implement the possibility to pass a probability matrix to the \texttt{boot.type} argument in the \texttt{boot.matrix} function where the probabilities can be reflecting occurences, regions, etc.
+This is an excellent suggestion for future direction for the package!
+I have raised an issue on my GitHub page for implementing the possibility to pass a probability matrix to the \texttt{boot.type} argument in the \texttt{boot.matrix} function where the probabilities can be reflecting occurences, regions, etc.
 I will also implement the ``fair'' algorithm mentioned in Kotric and Knoll (2015, Paleobiology, 41, 68–).

### 2018-01-11 paper/dispRity_paper.tex
Response to R1 checked

--- 2018-01-04 paper/dispRity_paper.tex
@@
         One such toolkit is based on creating a multidimensional space where the variables are the axes.
-        It is then possible to measure divers aspects of the distribution of some observation (e.g. species) in this space.
+        It is then possible to measure diverse aspects of the distribution of some observation (e.g. species) in this space.
         For example, if studying morphology, one can create a morphospace for two groups of species, measure the volume occupied by each of these groups and then test whether these two volumes are significantly different or not.
@@
 
-One can then measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupies more space than group B?'').
+One can then measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupy more space than group B?'').
 This requires the definition of a proxy for space occupancy: the disparity metric \citep[or index;][]{Hopkins2017} which can be measured in a multitude of ways.
@@
 To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
-Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test.
+Note that in this case, the function outputs different warnings on the usage of such test and the eventual data not used in the test.
 Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.

### 2018-01-11 paper/dispRity_paper_highlighted.tex
Response to R1 checked

--- 2018-01-04 paper/dispRity_paper_highlighted.tex
@@
         One such toolkit is based on creating a multidimensional space where the variables are the \hl{axes}.
-        It is then possible to \hl{measure divers aspects of the distribution of some observation (e.g. species) in this space.
+        It is then possible to \hl{measure diverse aspects of the distribution of some observation (e.g. species) in this space.
         For example, if studying morphology, one can create a morphospace for two groups of species, measure the volume occupied by each of these groups and then test whether these two volumes are significantly different or not}.
@@
 
-\hl{One can then measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupies more space than group B?'').
+\hl{One can then measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupy more space than group B?'').
 This requires the definition of a proxy for space occupancy: the disparity metric}
@@
 To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
-\hl{Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test.
+\hl{Note that in this case, the function outputs different warnings on the usage of such test and the eventual data not used in the test.
 Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.}

### 2018-01-11 paper/response_to_reviewers.tex
Response to R2 checked

@@
 
-\textit{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best it their needs (i.e.
+\textit{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best fit their needs (i.e.
 data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.} lines 82-85.

### 2018-01-11 paper/dispRity_paper.tex
Response to R2 checked

@@
 Although many software packages exist for multidimensional analysis \citep[e.g.][]{navarro2003MDA,bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,geiger2008,adams2013geomorph,Claddis,adams2017geometric},
-package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best it their needs (i.e.
+package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best fit their needs (i.e.
 data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.

### 2018-01-11 paper/dispRity_paper_highlighted.tex
Response to R2 checked

@@
 Although many software packages exist for multidimensional analysis \citep[e.g.][]{navarro2003MDA,bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,geiger2008,adams2013geomorph,Claddis,adams2017geometric},
-\hl{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best it their needs (i.e.
+\hl{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best fit their needs (i.e.
 data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.}

### 2018-01-15 paper/response_to_reviewers.tex
paper up

@@
 Note that in this case, the function outputs different warnings on the usage of such test and the eventual data not used in the test.
-Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.} lines 359-369.
+Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.} lines 333-342.
 
@@
 
-\textit{Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
-The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
-These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).
-Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).} lines 169-183.
+\textit{Note that the four first models are a proxy for punctuated evolution and the two last ones for gradual evolution.
+There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.
+See Guillerme \& Cooper \textit{in review} for full description.} lines 167-176.
 
@@
 
-\textit{Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).} line 180-183.
+\textit{There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.} line 173-175.
 
@@
 For example, if the metric is defined as} \texttt{metric = c(prod, ranges)} \textit{(the hypercube volume), the} \texttt{ranges} \textit{function (dimension-level 2) is first applied to data (the range of each dimensions) and the function} \texttt{prod} \textit{is then applied to the results (}\texttt{prod(ranges(data))}\textit{).
-One can also directly pass a function description to the} \texttt{metric}\textit{ argument (e.g.} \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} \textit{ for the average squared pairwise distance).} lines 130-135.
+One can also directly pass a function description to the} \texttt{metric}\textit{ argument (e.g.} \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} \textit{ for the average squared pairwise distance).} lines 127-134.
 
@@
 
-I've fixed these typos.
+I've removed this sentence.
 
@@
 It is possible to compare each bootstrap in a pairwise way without pooling the data by using the} \texttt{concatenate = FALSE} \textit{argument.
-The results will then be a distribution of statistic and p values.} lines 323-327.
+The results will then be a distribution of statistic and p values.} lines 298-302.
 
@@
 \textit{Note that by default the function only outputs the test's statistic, parameter (if parametric) and the p value.
-However, the raw test results can also be output using the option} \texttt{details = TRUE} \textit{in the function above.} lines 319-320.
+However, the raw test results can also be output using the option} \texttt{details = TRUE} \textit{in the function above.} lines 296-298.
 
@@
 
-\textit{The following example is based on the discrete morphological data described above. More examples, including based on other types of data are available in the package vignette.} lines 271-273
+\textit{Note that more examples are available in the package manual (...).} lines 229-231
 
@@
 an embedding, scaling, ordination, etc.
-of the variables).}
+of the variables).} lines 48-50
 
@@
 in Díaz et al.
-2016) or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements.}
+2016) or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements.} lines 69-73
 
@@
 I have removed the simulations part as suggested by this reviewer.
-However, following the encouraging suggestions of reviewer 2 (major suggestion \ref{simulations}) I have developed in much more details the simulation algorithms in the package manual.
+However, following the encouraging suggestions of reviewer 2 (major suggestion \ref{simulations}).
+I have developed in much more details the simulation algorithms in the package manual.
 

### 2018-01-15 paper/dispRity_paper.tex
paper up

@@
 
-\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$}\\
+\noindent {\normalsize \sc Thomas Guillerme$^12$$^,$$^*$}\\
 \noindent {\small \it 
-$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\\}
+$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\
+$^1$School of Biological Sciences, University of Queensland, St. Lucia, Australia.\\}
+
 \end{center}
@@
 }%
-    % \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g.
+    \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g.
 it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}$ - for a hyper-ellipsoid).
 $^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA).}}
-    % \label{Tab:metrics}
+    \label{Tab:metrics}
 \end{table}
@@
 \label{Tab:metrics}
-
-%TODO: Add the proper sources for the convhulln and vegan packages
 
@@
 This method thus requires the nodes to be part of the space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
-When a slice occurs not on a tip or a node, six methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
+When a slice occurs not on a tip or a node, six methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice.
+For each method, the selected data is: %TODO HL
 \begin{itemize}
-    \item ``acctran'' where the data chosen along the branch is always the one of the descendant
-    \item ``deltran'' where the data chosen along the branch is always the one of the ancestor
-    \item ``random'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor
-    \item ``proximity'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length
-    \item ``equal.split'' where the data chosen along the branch is both the descendant and the ancestor with an even probability
-    \item ``gradual.split'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length
+    \item ``acctran'': always the descendant
+    \item ``deltran'': always the ancestor
+    \item ``random'': randomly chosen between the descendant or the ancestor
+    \item ``proximity'': either the descendant or the ancestor depending on branch length
+    \item ``equal.split'': both the descendant and the ancestor with an even probability
+    \item ``gradual.split'': both the descendant and the ancestor with a probability depending on branch length
 \end{itemize}
 
-Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
-The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
-There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).
+Note that the four first models are a proxy for punctuated evolution and the two last ones for gradual evolution.
+There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.
 See Guillerme \& Cooper \textit{in review} for full description.
@@
 Disparity measurement can be influenced by sampling \citep{Butler2012}.
-To take this source of bias into account, it is common practice to bootstrap the multidimensional space (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different subsets).
+To take this source of bias into account, one can bootstrap the multidimensional space or/and to rarefy the data.
 Additionally, if disparity is defined as a dimension-level 1 metric, it can be useful to measure it on bootstrapped data to obtain a distribution on which to perform statistical analyses.
 
-Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).
+Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where $n = 1$ (somewhat akin to jackknife resampling).
 
 Similarly, rarefaction can be achieved through the same \texttt{boot.matrix} function.
-Rarefaction can be useful when looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of subsets with the same number of elements in each.
 In practice, rarefaction limits the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).
@@
 \subsection{Interpreting results}
-The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity} (provided disparity was calculated).
+The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity}.
 These results can also be analysed using the \texttt{test.dispRity} function for comparing subsets or testing hypotheses.
@@
 The \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow users to set which central tendency and which quantiles should be represented.
-Additionally, the \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves (in the case of disparity-through-time analysis for example) and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
+Additionally, the \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
 Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e.
@@
 \subsubsection{Testing hypotheses}
-The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data (e.g. by comparing the different subsets to each other).
+The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data.
 Similarly to the \disp function described above, this function can take any test defined by the user or from other R packages.
@@
 Multivariate analysis can be really useful for looking at multiple aspects of organisms diversity together.
-For example, one classic univariate trait used in ecology and evolution can be the number of species present in a ecosystem or/and through time.
-This trait can be useful to test whether some ecosystems or/and time periods have more species diversity than others.
-In this case one can then test some hypothesis such as: environments with longer altitudinal gradients contain more species.
-
-Conversely, in the case of species diversity, one can also look at another independent multivariate trait: the diversity of morphologies \citep[or disparity;][]{foote1991morphological}.
+For example, one can also look the diversity of morphologies \citep[or disparity;][]{foote1991morphological}.
 Using disparity, it is then also possible to assess whether one ecosystem or/and time period display more morphological variation.
-Following the previous example, if disparity is also higher with longer altitudinal gradients, one could argue that the species richness is caused by an increased number of morphologies adapted to more ecological niches (or something akin).
-
 The following example is based on a classical morphological disparity analysis.
@@
 In this example, the space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
-Additionally, we can define disparity as the sum of the variances on each dimension \citep{foote1991morphological,Wills1994} that will represent an aspect of the the volume of the morphospace (i.e.
-a large sum of variances approximates a large morphospace and vice versa).
-
-\subsection{Example}
-
-The following example is based on the discrete morphological data described above. More examples, including based on other types of data are available in the package vignette.
+Additionally, we can define disparity as the sum of the variances on each dimension \citep{foote1991morphological,Wills1994} that will represent an aspect of the the volume of the morphospace.
 
@@
 
-\section{Funding}
-This work was funded by the European Research Council under the European Union's Seventh Framework Programme (FP/2007 - 2013)/ERC Grant Agreement number 311092 and the Discovery Project Grant number DP170103227.
-
 \bibliographystyle{sysbio}

### 2018-01-15 paper/dispRity_paper_highlighted.tex
paper up

@@
 This method thus requires the nodes to be part of the space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
-When a slice occurs not on a tip or a node, \hl{six} methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
+When a slice occurs not on a tip or a node, \hl{six} methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice.
+\hl{For each method, the selected data is:}
 \begin{itemize}
-    \item ``acctran'' where the data chosen along the branch is always the one of the descendant
-    \item ``deltran'' where the data chosen along the branch is always the one of the ancestor
-    \item \hl{``random''} where the data chosen along the branch is randomly chosen between the descendant or the ancestor
-    \item \hl{``proximity''} where the data chosen along the branch is either the descendant or the ancestor depending on branch length
-    \item \hl{``equal.split'' where the data chosen along the branch is both the descendant and the ancestor with an even probability}
-    \item \hl{``gradual.split'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length}
+    \item ``acctran'': always the descendant
+    \item ``deltran'': always the ancestor
+    \item \hl{``random''}: randomly chosen between the descendant or the ancestor
+    \item \hl{``proximity''}: either the descendant or the ancestor depending on branch length
+    \item \hl{``equal.split'': both the descendant and the ancestor with an even probability}
+    \item \hl{``gradual.split'': both the descendant and the ancestor with a probability depending on branch length}
 \end{itemize}
 
-\hl{Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
-The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
-There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).
+\hl{Note that the four first models are a proxy for punctuated evolution and the two last ones for gradual evolution.
+There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.
 See Guillerme \& Cooper \textit{in review} for full description.}
@@
 Disparity measurement can be influenced by sampling \citep{Butler2012}.
-To take this source of bias into account, it is common practice to bootstrap the multidimensional space (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different subsets).
+To take this source of bias into account, one can bootstrap the multidimensional space or/and to rarefy the data.
 Additionally, if disparity is defined as a dimension-level 1 metric, it can be useful to measure it on bootstrapped data to obtain a distribution on which to perform statistical analyses.
 
-Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).
+Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where \hl{$n = 1$} (somewhat akin to jackknife resampling
 
 Similarly, rarefaction can be achieved through the same \texttt{boot.matrix} function.
-Rarefaction can be useful when looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of subsets with the same number of elements in each.
 In practice, rarefaction limits the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).
@@
 \section{Examples}
-Multivariate analysis can be really useful for looking
- \hl{at} 
- multiple aspects of organisms diversity together.
-For example, one classic univariate trait used in ecology and evolution can be the number of species present in a ecosystem or/and through time.
-This trait can be useful to test whether some ecosystems or/and time periods have more species diversity than others.
-In this case one can then test some hypothesis such as: environments with
- \hl{longer altitudinal} 
- gradients contain more species.
-
-Conversely, in the case of species diversity, one can also look at another independent multivariate trait: the diversity of morphologies \citep[or disparity;][]{foote1991morphological}.
+Multivariate analysis can be really useful for looking at multiple aspects of organisms diversity together.
+For example, one can also look the diversity of morphologies \citep[or disparity;][]{foote1991morphological}.
 Using disparity, it is then also possible to assess whether one ecosystem or/and time period display more morphological variation.
-Following the previous example, if disparity is also higher with
- \hl{longer altitudinal} 
- gradients, one could argue that the species richness is caused by an increased number of morphologies adapted to more ecological niches (or something akin).
-
 The following example is based on a classical morphological disparity analysis.
-Note that more examples are available in the package manual (\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).
+\hl{Note that more examples are available in the package manual (}\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).
 
@@
 In this example, the space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
-Additionally, we can define disparity as the sum of the variances on each dimension \citep{foote1991morphological,Wills1994} that will represent an aspect of the the volume of the morphospace (i.e.
-a large sum of variances approximates a large morphospace and vice versa).
-
-\subsection{Example}
-
-\hl{The following example is based on the discrete morphological data described above. More examples, including based on other types of data are available in the package vignette.}
+Additionally, we can define disparity as the sum of the variances on each dimension \citep{foote1991morphological,Wills1994} that will represent an aspect of the the volume of the morphospace.
 

### 2018-02-12 inst/gitbook/06_palaeo-demo.Rmd
updated gitbook

@@
 
-> These objects will gradually contain more information when completing the following steps in the disparity-through-time analysis.
+> These objects will gradual.splitly contain more information when completing the following steps in the disparity-through-time analysis.
 

### 2018-02-12 inst/gitbook/index.Rmd

@@
 documentclass: book
+bibliography: [References.bib, packages.bib]
 link-citations: yes
@@
 Multidimensional spaces can be ordinated matrices from MDS, PCA, PCO, PCoA but the package is *not* restricted to any type of matrices!
-This manual is based on the version `0.4.2`.
+This manual is based on the version `0.5.0`.
 
@@
 
-Note this uses the `release` branch (version`0.4`). <!-- @@@ -->
+Note this uses the `release` branch (version`0.5`). <!-- @@@ -->
 For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.

### 2018-02-12 paper/example_script.Rmd

---
title: "Example script the `dispRity` paper"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

This is the raw `R` script for running the examples present in the paper "`dispRity`: a modular `R` package for measuring disparity".

## Loading the package and the data

Loading the packages:

```{r}
## Installing (if needed) and loading devtools
if(!require(devtools)) install.packages("devtools")
library(devtools)

## Installing (if needed) and loading dispRity
if(!require(dispRity)) install_github("TGuillerme/dispRity", ref = "release")
library(devtools)
```

Loading the datasets:

```{r}
## 50*48 PCO matrix
data(BeckLee_mat50)

## 99*97 PCO matrix
data(BeckLee_mat99)

## 50 taxa tree
data(BeckLee_tree)

## FADLAD data
data(BeckLee_ages)
```

## Typical disparity among groups analysis

The wrapper version:

```{r}
## Creating the groups (crown and stem - ignoring the nodes)
mammal_groups <- crown.stem(BeckLee_tree, inc.nodes = FALSE)

## Measuring disparity as the sum of variances
disparity <- dispRity.per.group(data = BeckLee_mat50, group = mammal_groups,
                                metric = c(sum, variances))
```

The detailed version:

```{r, fig.width = 8, fig.height = 8}
## Creating custom subsets
subsets <- custom.subsets(data = BeckLee_mat50, group = mammal_groups)

## Bootstrapping the data
bootstraps <- boot.matrix(subsets)

## Measuring disparity as the sum of variance
disparity2 <- dispRity(bootstraps, metric = c(sum, variances))

## Printing the results
disparity2

## Summarising the results
summary(disparity2)

## Plotting the results
plot(disparity2)

## Testing the differences
test.dispRity(disparity2, test = wilcox.test)
```

## Typical disparity-through-time analysis

The wrapper version

```{r}
## Get the time bins
time_bins <- c(100.5, 66, 56, 33.9)

## Calculate disparity through time (sum of variances)
disparity3 <- dispRity.through.time(data = BeckLee_mat50, tree =  BeckLee_tree,
                                   time = time_bins, metric = c(sum, variances))
```

```{r, fig.width = 8, fig.height = 8}
## Creating the time subsets
time_subsets <- time.subsets(data = BeckLee_mat50, tree = BeckLee_tree,
                             time = time_bins, method = "discrete")

## Bootstrapping the data
bootstraps2 <- boot.matrix(time_subsets)

## Measuring disparity as the sum of variance
disparity4 <- dispRity(bootstraps2, metric = c(sum, variances))

## Summarising the results
summary(disparity4)

## Plotting the results
plot(disparity4, type = "continuous")

## Testing the effect if time
test.dispRity(disparity4, test = adonis.dispRity)
```

### 2018-02-22 paper/response_to_reviewers2.tex
Compiled responses

--- 2018-01-15 paper/response_to_reviewers.tex
@@
 
-\section{Reviewer 1}
+\section{Reviewer 2}
 
@@
 
-\item{\textcolor{blue}{While I understand that the author is writing a package that has many diverse capabilities and is presented to be useful to a wide range of disciplines, it is too vague in too many places to be helpful in its current state.
-For example, the line in the abstract “It is then possible to use a summary metrics for measuring some properties of this space for some group of species and compare them.”.
-“Some properties” is too vague without any qualifiers to be useful.
-But I believe that with revision, taking into account the issues of clarity, this will be an important addition to the available analytical tools for Biologists.}}
+\item{\textcolor{blue}{- In table 1 the singular of strata is stratum.}}
 
-I have changed the terms used in the abstract and proposed more specific examples for clarity:
+\item{\textcolor{blue}{- Rows in Tables 1 and 2 are not well distinguished, i.e., I had to look at multiple columns to know where text broke over a line versus a new row.}}
 
-\textit{It is then possible to measure diverse aspects of the distribution of some observation (e.g.
-species) in this space.
-For example, if studying morphology, one can create a morphospace for two groups of species, measure the volume occupied by each of these groups and then test whether these two volumes are significantly different or not.} lines 6-10.
+\item{\textcolor{blue}{L187 - Sentence seems to peter out. Close parenthesis and period?}}
 
-The changes have also been made throughout the manuscript as generally suggested by all four reviewers.
+\item{\textcolor{blue}{L304 - Probably shouldn't use exclamation marks in a scientific paper unless they are factorials.}}
 
-\item{\textcolor{blue}{My main concern is that what disparity actually means is left to the reader’s imagination – it is not defined or discussed anywhere.
-Disparity simply means lack of similarity or equality.
-While the author is probably considering the dispersion of points in a multidimensional space as a measure of dissimilarity, it is important to realise the ambiguity of the word and more explanation is needed early in the introduction (and foremost in the abstract).
-Specifically, the lines 58-61 need to be revised, as they simply don’t convey to the reader what is being presented here as disparity.
-Say what “specific aspect of it [multidimensional space]” you are referring to.
-Also, the different definitions of disparity alluded to on lines 61-63 should be summarised.
-Furthermore, the manuscript is lacking a table summarising the metrics of disparity that this package can work with (which I suspect it was supposed to have).}}
-\label{define_disparity}
-
-Apologies for the omission of the table summarising the metrics which is now properly included in the manuscript (see reviewer 1, minor comment \ref{table_missing}).
-As reviewer 4 (major comment \ref{nebulous}) also underlines: ``disparity is a broad, nebulous concept''.
-This observation was the genesis for this package and I am currently working on a collaborative review to define the borders (or the absence thereof) of this concept.
-I feel that explaining this would be out of the scope of this software description paper.
-However, I have tried to give specific examples each time I refer to disparity as a way to measure these ``specific aspect of the space'' and I have also added a clearer definition of what I mean by disparity throughout this manuscript.
-
-I rewrote the part in the introducing disparity as follows:
-
-\textit{One can then measure how the observations are distributed within this space to answer related questions (e.g.
-``does group A occupy more space than group B?'').
-This requires the definition of a proxy for space occupancy: the disparity metric (or index; Hopkins and Gerber, 2017) which can be measured in a multitude of ways.
-For example, one could use a metric based on the variance or the range of each axis of space (Wills, 2001; Ciampaglio et al., 2001), a distance (e.g.
-Euclidean) measured between observations (Foote, 1993, 1996), a more direct approximation of the hyper volume (Cornwell et al., 2006; Donohue et al., 2013), or many more (e.g.
-Navarro, 2003).} lines 60-67.
-
-\item{\textcolor{blue}{Another concern I have is about how this R package sits within the current analytical toolkit available.
-How does this new package relate to existing packages available for R?
-Claddis and geomorph are mentioned and discussed because they deal with bespoke data formats that need to be dealt with prior to analysis.
-But are there others?
-How does the package relate to the ‘vegan’ R package for multivariate analysis of ecological communities?
-Within there are metrics for community dissimilarity, metrics from ordinations etc.
-Vegan is a popular package for analysis of multivariate datasets and seems to be relevant here with similar functionalities but not discussed.
-Similarly, the R package ‘diverse’ has a function called disparity, and the ‘geiger’ function disparity implements the disparity-through-time. }}
-\label{add_geiger_vegan}
-
-I have added two wrapper functions to use the popular functions \texttt{vegan::adonis} and \texttt{geiger::dtt} to the package (\texttt{dispRity.adonis} and \texttt{dispRity.dtt}).
-Also, I have generalised the distances calculations using the \texttt{vegan::vegdist} function that allows for more distances than \texttt{stats::dist} to be calculated.
-
-I've updated the manual and the example to describe these new functions:
-
-\textit{Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA (PERMANOVA; Anderson, 2001b) through the adonis.disprity function that is a wrapper of the vegan::adonis function (Oksanen et al., 2007) for dispRity objects [TABLE 4].
-To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
-Note that in this case, the function outputs different warnings on the usage of such test and the eventual data not used in the test.
-Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.} lines 333-342.
+I've removed the exclamation mark.
 
@@
 
-\subsection{Reviewer 1 - Minor comments}
+\section{Reviewer 3}
 
@@
 
-\item{\textcolor{blue}{The paper keywords are narrow and suggest a very particular use of disparity analyses (morphological disparity).
-The paper suggests there are broader applications and so these should be revised to include the diverse applications that the paper suggests.
-}}
+\item{\textcolor{blue}{[pushing the package on the CRAN]}}
 
-I have broaden the keywords within the limits allowed by manuscript central for MEE publications.
+\item{\textcolor{blue}{Table 2 seems to be lacking a caption and part of it just seems to have wandered into the text on page 12. }}
 
-\item{\textcolor{blue}{Line 4 – changes ``in'' to ``to''. Line 6 – changes ``axis'' to ``axes''}}
-\label{abstract_typo}
+The caption is part of the table in the LaTeX document.
+However, due to the table size, it doesn't fit in the page and was thus written separately.
 
-I have fixed these typos.
+@@@ Force it to be under the table
 
-\item{\textcolor{blue}{Line 84 – Explain what the Procrustes variance is. And when is it appropriate and when is not appropriate? Be more specific.}}
-\label{tone_down}
+\item{\textcolor{blue}{I wondered, as pairwise.dist just vectorizes a distance matrix from vegan::vegdist, if these functions might be more so necessary internal functions that the author chose to externalize for users?
+If that is the case, maybe some careful wording - more than just citing the original sources - is needed to not portray all of these as features of the new package.
+One suggestion is to offer, e.g., convhulln$vol rather than convhull.volume as the function that calculates hull volume, in Table 2, and indicate that convhull.volume similarly returns the $vol part of a convhull object, so one does not have to sift through a list.
+(The idea is to do this with all cloned functions.) 
+Alternatively, write new bare functions (maybe the C++ coding could be improved for geometry::convhull?).
+Regardless of solution, the current help page for Disparity Metrics states things like, " Both convhull functions are based on the convhulln function."  They are not based on this function, they ARE this function.
+Some effort should be made to either better credit original sources or avoid cloning functions, altogether.}}
 
-In \texttt{geomorph}, the procrustes variance is defined as the sum of the diagonal covariance matrix between each elements divided by the number of elements (or effectively the sum of the squared ordination values divided by the number of elements - see example in \texttt{R} for \texttt{?dispRity::geomorph.ordination}).
-Using this disparity metric is appropriate for testing volume based hypothesis (e.g.
-Does group A occupy more morphospace than group B?) but does not allow to test position based hypothesis (e.g.
-Does group A occupy a different part of the morphospace than group B).
-For example, one might find out that group A and B have the same volume but that doesn't tell us if they occupy the same position in the morphospace.
-
-I've highlighted this in the text:
-
-\textit{For example, in the excellent and widely used} \texttt{geomorph} \textit{package, morphological disparity analysis uses the} \texttt{morphol.disparity} \textit{function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance), and uses permutation tests (Zelditch et al., 2012; Adams and Otarola-Castillo, 2013; Adams et al., 2017).
-This is ideal for testing volume based hypothesis (e.g.
-does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g.
-does groups A and B occupy the same location?).} lines 85-93.
-
-\item{\textcolor{blue}{Lines 98-100 – Suggest changing this to be specific: “calculates a disparity metric from the space”, and “analyse the resulting disparity object through hypothesis testing and visualisation”}}
-
-I have changed the sentence following the reviewer's suggestion.
-
-\item{\textcolor{blue}{Line 116 – “Table ??” = should have been caught during proofreading.
-Furthermore, there is no table included in the manuscript that details the implemented metrics, and this is an important part of the paper sadly omitted in this submission.}}
-\label{table_missing}
-
-I sincerely apologise for the omission of the Table! It is now included in the new version of the manuscript with the currently implemented metrics description.
-I would also like to underline that although not all usual disparity metrics are represented in the table and in the package (by far!) the idea of the \texttt{dispRity} package and specifically the \texttt{metric} argument in the \texttt{dispRity} function is to allow users to come up with their \textit{own} metric.
-
-For example, if one decides to use the relative ellipsoid hyper volume to the number of elements per group as a disparity metric (i.e.
-this metric could be used as a proxy for the average space occupancy per unit element - e.g.
-per taxa).
-This would not be doable directly by using the implemented \texttt{ellipse.volume} function but would require the user to use a function like:\\
-\texttt{dispRity(my\_data, metric = function(X) ellipse.volume(X)/nrows(X))}\\
-
-\item{\textcolor{blue}{Lines 356-357 – This final sentence of the conclusion section is vague and does not add to the manuscript.}}
-\label{remove_last_sentence}
-
-I removed the last sentence.
-
-\item{\textcolor{blue}{Figure 1 – Procrustes is a proper noun and requires a capital p.
-Is the Procrustes matrix a “Procrustes distance matrix” or a “Procrustes residuals matrix”?
-i.e.
-is it the shape variables after a Procrustes superimposition (latter) or the distances between individuals after a Procrustes superimposition (former)?
-Both are suitable inputs for ordination.}}
-
-I meant both matrices (as this reviewers rightly points out they are suitable for ordination).
-I have made this more clear in the figure caption.
-On a more general note, the package now allows users to input any matrix (although the package does not provide warranty that the provided matrix is meaningful to the user's specific analysis).
-I have added an example to the \texttt{dispRity} manual based on the \texttt{eurodist R} base dataset.
-
-\item{\textcolor{blue}{Figure 2 makes little sense without a summary of the kinds of disparity metrics available to the user, and needs more explanation with examples for each dimension level.}}
-
-I've added a longer explanation of the different dimension-levels in the caption of the figure:
-
-\textit{A dimension-level 3 matrix would be a metric resulting in a matrix (e.g.
-the function \texttt{stats::cor} to calculate the correlation between each dimension), a dimension-level 2 metric would result in a vector (i.e.
-a distribution, e.g.
-\texttt{dispRity::variances} which calculates the variance within each dimension) and a dimension-level 1 metric would result in a single value (e.g.
-\texttt{stats::sd} which calculates the standard deviation of the input matrix).}
-
-\item{\textcolor{blue}{Table 1 – change to disparity-within-groups with hyphens as DTT below.
-Also add “shape” before data for geomorph.ordination, and perhaps a similar qualifier for Claddis.ordination.}}
-
-I've added hyphens to disparity-within-group and ``\textit{shape data}'' and ``\textit{discrete morphological data}'' for the two ordinations functions in the table.
+\item{\textcolor{blue}{Finally, there were some English irregularities (e.g., data is the plural form of datum) that could be fixed with some proofing from a native English-speaking scientist.
+(I would not trust this to a copy-editor over a colleague, if technical/scientific writing should be precise).}}
 
 \end{enumerate}
-
-% ~~~~~~~~~~~~~~~~~~~~~~~~
-%
-% REVIEWER 2
-%
-% ~~~~~~~~~~~~~~~~~~~~~~~~
-
-\section{Reviewer 2}
-
-\begin{enumerate}
-\item{\textcolor{blue}{Firstly, features available in the package but not currently fully explained/highlighted.
-Really the only major omission here is a more detailed description of the tools avilable under the "Data Simulations" section of the manuscript.
-I suspect many users will find this the *most* useful feature of the package and the author should consider unpacking this section a bit more.
-Perhaps adding an additional multipanel figure exhibitng some simulated spaces.}}
-\label{simulations}
-
-Reviewer 4 (minor comment \ref{remove_simulations}) suggested completely removing this section which conflicts directly with this suggestion.
-The simulation protocols implemented here are already developed (and tested) in details in previous publications along with novel features that are part of future publications.
-For example the discrete data simulation protocol is described and detailed in Guillerme \& Cooper 2016 \textit{Mol Phyl Evol}, O’Reilly et al.
-2016 \textit{Biol Let}, O'Reilly et al.
-2017 \textit{Palaeontology} and Guillerme \& Brazeau \textit{in review}.
-The implemented ordinated space simulations are theoretically described in Diaz et al.
-2016 \textit{Nature} and will be used along with additional simulation features not implemented yet (including shape evolution) as part of a project currently in progress.
-Due to the space constraint for application manuscripts in MEE, I am inclined to follow reviewer's 4 suggestion rather than this one.
-
-However, I have heavily edited the online manual to highlight the simulation functionalities implemented in the \texttt{dispRity} package: \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/making-stuff-up.html}.
-The manual now has a chapter entirely dedicated to the data simulation functions with emphasis on both the methods used and the algorithms implemented (and illustrated with advanced examples).
-
-\item{\textcolor{blue}{Secondly, problematic implementations.
-Here there is again only one major issue, and that concerns the time-slicing options.
-Initially it should be noted by the author how useful this feature could be in generating time series of disparity that are long enough for time series statistics to be applied.
-Currently most palaeobiologists are stuck using binned data (e.g., Figure 4) which tends to lead to time series of insufficient length (here N = 3) to explore, for example, correlations with climate proxies (or some other hypothesised driver of morphological diversity).
-Time-slicing could revolutionise the field by opening up data sets to such analyses and an example alongside the binning in Figure 4 should be considered.
-However, the current implemented options are, in my view, deeply problematic.
-Specifically, dispRity offers four options (L145-156), but in effect these all boil down to treating a point along a branch as either identical to the descendant (end of branch) or the ancestor (start of branch).
-However, I think most biologists would treat such a value as more likely to be intermediate between these extremes and possessing some form of uncertainty depending on the generating model's assumptions (e.g., Brownian motion).
-I don't mind offering a punctuated model (descendant value propagated back along branch) but some form of intermediate value should also exist.
-Without this option data are likely to be biased, e.g., towards generating more "jagged" time series and perhaps more significant slice-to-slice changes (using the packages "test" functions) than are realistic (ie.e, type I errors).}}
-
-The standard approach in palaeobiology relies on the following pipeline: (i) coding characters for tips; (ii) running ancestral states reconstructions for the nodes; (iii) ordinating the resulting matrix (usually via a PCO); (iv) measuring disparity from there (see Lloyd 2016 \textit{Biol J Lin Soc} for an excellent description of this pipeline).
-I feel that the reviewer's suggestion of proposing more accurate models for ancestral states reconstructions and will technically not be doable after step (iii).
-In fact, the ordination, to be meaningful as a ``morphospace'' should always be done on the full dataset.
-I.e. no additional data should be added to the space \textit{a posteriori}.
-In practice, the \textit{time.subset} function can only select between two information: the ancestral node (DELTRAN) and the descendant node/tip (ACCTRAN) along the edge since only the two of them are real objects in the morphospace.
-It is therefore not possible to create an ancestral states reconstruction along the edge after running the ordination.
-
-I have however modified the available models to better describe this limitation and I have added two new models that would approximate ancestral reconstruction along edges.
-I have modified the name of the two previous ``punctuated'' and ``gradual'' models to ``random'' and ``proximity'' to reflect more accurately how they select either the ACCTRAN or DELTRAN information.
-I've added two more accurate models (now named ``equal.split'' and ``gradual.split'') to reflect the changes along the branches.
-These two algorithm now allows to use both the ACCTRAN and DELTRAN information with a linked probability to use either of them per bootstrap draw.
-For example, the ``equal.split'' model will have a 0.5 probability of using either node information for each bootstrap draw; the ``gradual.split'' model transforms the branch length were the slice occurs into a probability of drawing either of the nodes information (e.g.
-if the slices occurs 3/4 along the branch from A to B, the model will draw the information from A and B with a probability of respectively 0.25 and 0.75).
-For example, the median disparity over 1000 bootstrap pseudo-replication along the edge (AB) will integrate 0.75 information from A and 0.25 from B thus approximating the ``intermediate'' between A and B (when averaged ).
-
-I've added these models descriptions in the manual and in the manuscript:
-
-\begin{itemize}
-    \item \textit{``random'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor}
-    \item \textit{``proximity'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length}
-    \item \textit{``equal.split'' where the data chosen along the branch is both the descendant and the ancestor with an even probability}
-    \item \textit{``gradual.split'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length}
-\end{itemize}
-
-\textit{Note that the four first models are a proxy for punctuated evolution and the two last ones for gradual evolution.
-There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.
-See Guillerme \& Cooper \textit{in review} for full description.} lines 167-176.
-
-\item{\textcolor{blue}{An additional more minor implementation issue here concerns the "subsamples".
-Firstly, I would urge the author to reconsider this name as it implies subsampling (e.g., rarefaction) and not what is intended, i.e., subsets of the data (the term used on L124) or groups of taxa.
-However, a more pressing issue is that I do not see any restrictions on overlap between these "subsamples" and indeed in the case of time-binning this is suggested as a "pro" (L138).
-Although in cases such as time-binning it is obviously sensible to allow taxa to appear in multiple "subsamples" this changes the assumptions of a statistical test looking for differences between samples, i.e., as overlap increases so would type II errors.
-I suggest the author consider at least a warning to users in such circumstances.}}
-
-I've changed ``subsample'' to ``subset'' throughout the package, the manual and the paper.
-
-I also agree with this comment: this method is a trade-off between the number of slices.
-(1) too few slices decrease resolution and veer towards the time binning (which could increase "significance" by comparing too different time subsamples);
-(2) too many slices will increase the resolution be will create much more pseudo replication and thus increase type I error.
-
-I've added a warning in the manual and the manuscript:
-
-\textit{There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.} line 173-175.
-
-A more detailed description and test of these models might be out of the scope of this package description paper but is currently in development for a special issue paper in \textit{Palaeontology} (in review) as part as a special symposium for the Palaeontological Association annual meeting 2017 in collaboration with Natalie Cooper.
-The full (un-reviewed an un-compiled) manuscript is available on Natalie's Github page (\url{https://github.com/nhcooper123/time-slice/blob/master/manuscript/time-slice.tex}).
-
-\item{\textcolor{blue}{Finally, additions the author could consider adding to the package.
-These are mostly minor, but more numerous.
-Some of these are simple caveats for limitations to the package.
-For example, disparity metrics can be generated from, for example, a discrete-character taxon matrix directly, or the distance matrix given to the ordination (e.g., the weighted mean pairwise distance metric used in Close et al.
-2015, cited).
-Because the package only deals with matrices generated by the ordination itself these metrics are beyond its scope, although are of potential value to disparity analysis more broadly.
-Specifically, the Close et al.
-(2015) metric can be applied to an incomplete matrix, but principal coordinates/classic multidimensional scaling requires a complete distance matrix and hence taxa would have to be pruned to generate an ordination, but can be retained if using a pre-ordination metric.
-In a similar vein, although the package allows bootstrapping of the space it could also be useful for the raw data itself to be bootstrapped, e.g., the set of measurements used to generate a principal components analysis.
-It might be helpful for the author just to make it clear (with a sentence or two) that these are things a user might want to attempt, but the package cannot (reasonably) do.}}
-
-I have updated the package's structure to be able to deal with any type of matrix as long as rows are the elements (e.g. taxa).
-It is now possible to perform the whole pipelined analysis (including bootstrapping) on non-ordinated matrices.
-
-I've reflected these changes in the manual with a subsection in the chapter ``Specific tutorials - 4.8'' (based on a distance matrix) and in manuscript as follows:
-
-\textit{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements (e.g.
-the multidimensional space can be a distance matrix; Close et al., 2015).} lines 109-111.
-
-\item{\textcolor{blue}{Picking up on the first point from above, there are also additional post-ordination disparity metrics that have been used in the literature but are not available here.
-For example, hypercube volume, minimum spanning tree length etc.
-I would also urge the author to cite the work of Mike Foote, who made some early important contributions to disparity in general, and has written papers that discuss some of these potential metrics (e.g., Foote 1997, Annual Review of Ecology and Systematics, 28, 129–).
-I believe Wills (2001, cited) also discusses a long list of metrics.}}
-
-I agree, there are many more metrics commonly used for disparity analysis not implemented in this package.
-This was a conscious choice: due to the really high number of metrics available (especially when combining different ones), I designed the package to be modular hence allowing users to implement their own metrics (see response to reviewer 1 minor comment \ref{table_missing}).
-I have add the hypercube volume description as an example of user defined metric in the manuscript:
-
-\textit{When multiple functions are passed to the} \texttt{metric} \textit{argument, they are sorted by dimension-level and applied in decreasing order to the data.
-For example, if the metric is defined as} \texttt{metric = c(prod, ranges)} \textit{(the hypercube volume), the} \texttt{ranges} \textit{function (dimension-level 2) is first applied to data (the range of each dimensions) and the function} \texttt{prod} \textit{is then applied to the results (}\texttt{prod(ranges(data))}\textit{).
-One can also directly pass a function description to the} \texttt{metric}\textit{ argument (e.g.} \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} \textit{ for the average squared pairwise distance).} lines 127-134.
-
-I have however added several additional metrics to the newest version of the package including: (1) the minimal spanning tree length (\texttt{span.tree.length} - as suggested), (2) the ancestral distance which is equivalent to the distance from centroid but were the ``centroid'' are the coordinates of each element's ancestors (\texttt{ancestral.dist}), (3) the pairwise distances (\texttt{pairwise.dist} - based on \texttt{vegan::vegdist}), (4) the radius, equivalent to the distance from centroids but for the dimensions rather than the elements (i.e.
-the distance between each dimensions coordinates and the mean dimension's coordinates - \texttt{radius}) and (5) the $n$-ball volume which is an extension of the \texttt{ellipse.volume} function for a sphere (the true $n$-ball volume) or an ellipsoid (\texttt{n.ball.volume} - see Table 2 in the new version of the manuscript).
-
-\item{\textcolor{blue}{The package also employs a rarefaction function, but rarefaction can be problematic in the sense that rarefaction curves can sometimes cross such that the sampling level used to make comparisons can effect which of two or more subsets is considered the most disparate.
-This is why algorithms designed to sample "fairly" rather than equally have been explored, see for example Kotric and Knoll (2015, Paleobiology, 41, 68–).
-A user may also wish to rarefy by some external data source, such as number of occurrences, or geographic regions.
-These are obviously major additions, but are something the author might wish to add in future.}}
-
-This is an excellent suggestion for future direction for the package!
-I have raised an issue on my GitHub page for implementing the possibility to pass a probability matrix to the \texttt{boot.type} argument in the \texttt{boot.matrix} function where the probabilities can be reflecting occurences, regions, etc.
-I will also implement the ``fair'' algorithm mentioned in Kotric and Knoll (2015, Paleobiology, 41, 68–).
-
-\item{\textcolor{blue}{Other minor additions the author may also wish to consider are the permutation test from Brusatte et al.
-(2014, Current Biology, 24, 2386-), a means of adding the geologic timescale to time series plots (e.g., using the geoscale package)}}
-
-Similarly to previous comment, I will also add the permutation test based on the \texttt{PAST} package in a future version of the package.
-Furthermore, I have added an illustrative example in the \texttt{plot.dispRity} manual on how to add disparity data (continuous or discrete) to the \texttt{geoscalePlot} and \texttt{geoscaleBox} functions thus allowing users to make nice ready-to-publish figures using the \texttt{geoscale} package.
-
-\item{\textcolor{blue}{and options for the user to only use a limited set of ordination axes (matrix columns), e.g., to avoid using higher axes with negative eigenvalues.}}
-
-This option is implemented in the \texttt{boot.matrix} and \texttt{dispRity} functions allowing users to set a \texttt{dimensions} argument to either a proportion or a value of dimensions to use (e.g.
-\texttt{0.95} or \texttt{10} to use only the 95\% of the dimensions or the 10 first ones respectively).
-I've added a line highlighting this feature that is indeed often used:
-
-\textit{Note that this function also allows to work on only a subset of dimensions via the} \texttt{dimensions} \textit{argument (e.g.
-if only the $n$ first dimensions must be considered).} lines 134-135.
-
-\end{enumerate}
-
-\subsection{Reviewer 2 - Minor comments}
-
-\begin{enumerate}
-
-\item{\textcolor{blue}{L4 - "in" should be "to"? "of variables" is then redundant and can be removed. L6 - "are the axis of this space" could just be "are the axes".}}
-
-See reviewer 1 minor comment \ref{abstract_typo}.
-
-\item{\textcolor{blue}{L7 - "a" here should be removed. L17 - "suits best" should be "best suits". L23 - "allowing to" should be "allowing users to". L28 - "suit" should be "suite".}}
-
-I've fixed these typos.
-
-\item{\textcolor{blue}{L54 - Technically a principal coordinates analysis *is* classic multidimensional scaling (CMDS).
-The author may also want to cite use of non-metric multidimensional scaling (e.g., Liow 2004, American Naturalist, 164, 431–).}}
-\label{PCO_cites}
-
-I have changed this sentence to:
-
-\textit{These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix (Lloyd 2016 and references therein; e.g.
-in Close et al.
-2015), or as outputs from an ordination, whether it being a principal components analysis (PCA, Hotelling 1933; e.g.
-in Zelditch et al.
-2012), a metric scaling (PCO, PCoA, Torgerson 1958; e.g.
-in Brusatte et al.
-2008) or a non-metric scaling (MDS, NMDS, Shepard 1962; e.g.
-in Liow 2004; Donohue et al.
-2013).} lines 51-56
-
-\item{\textcolor{blue}{L80 - I'm not sure I would use the term "enforce" here.
-I think many of us write packages that serve our own needs, but the great thing about languages like R is that it is flexible and code can be borrowed and appropriated etc.
-Essentially I think the author may unintentionally disparage other package authors here.
-Besides which, I do not feel that dispRity is competing with other packages, rather it is supplemental to them.
-I.e., to repeat my comment from above I see dispRity as sitting at the end of a range of data analysis pipelines with other packages sitting in front of it.}}
-
-I've change this sentence to:
-
-\textit{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best fit their needs (i.e.
-data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.} lines 82-85.
-
-\item{\textcolor{blue}{L84 - Redundant "a".}}
-
-I've fixed this typo.
-
-\item{\textcolor{blue}{L116 - "??"!}}
-
-Sincerest apologies! See reviewer 1 minor comment \ref{table_missing}.
-
-\item{\textcolor{blue}{L130 - I don't think the term "possible observed" makes sense.
-Either the author means all the observed morphologies *or* all the possible morphologies.
-I suggest the former is safer as I think it likely that possible morphologies introduce sufficient complexity we can never really say this is true.
-For example, if we alter the number of elements in our sample (removing a species say) the ordination/morphospace we produce will differ, but the range of possible morphologies should not.}}
-
-I agree and have changed it to ``\textit{all the observed morphologies}'' line 150
-
-\item{\textcolor{blue}{L134 - Should probably be "an argument".}}
-
-I have changed the start of the sentence to ``\textit{The first argument is the matrix}'' line 153.
-
-\item{\textcolor{blue}{L136 - It is not immediately obvious to me why a phylogeny is required here.
-This only becomes clearer later (time-slicing) so maybe a "heads up" parenthetic (i.e., "see below") could be added here.}}
-
-I've modified the sentence in bracket to
-``\textit{a dated phylogeny of the elements present in the morphospace - see below}'' line 155.
-
-I've also changed the code to allow non-phylogenetic dating (e.g.
-just adding the First/Last Occurrence Dates of all the taxa) but I've only mentioned it in the function manual and the vignette due to space restriction in this manuscript.
-
-\item{\textcolor{blue}{L208 - "looking at"? L213 - Should be "altitudinal"?}}
-\label{altitude}
-
-I've removed this sentence.
-
-\item{\textcolor{blue}{L245-247 - I don't think these definitions for crown and stem are quite right.
-However, for the audience of MEE I suspect there is no need to define them so these parenthetic statements can just be cut.}}
-
-I've removed the inaccurate definitions in parenthesis.
-
-\item{\textcolor{blue}{L252 - It might be more helpful to define these using the phylogeny as this is a more generally applicable approach than manually delimiting row numbers (where human error can easily creep in or the inclusion of exclusion of data can mess things up).
-I.e., something that identifies descendants of the crown node and then a simple setdiff() between these and all possible names to get the non-crown (stem) taxa).}}
-
-Great suggestion! I've added an utility function \texttt{crown.stem} allowing to separate stem and crown species.
-The function is now briefly mentioned in the package description paper and detailed in the package itself.
-
-\item{\textcolor{blue}{L296 - Should be "increase".}}
-
-I've fixed this typo.
-
-\item{\textcolor{blue}{L357 - Should be "for such analyses".}}
-
-I've removed this last sentence (see reviewer 1 minor comment \ref{remove_last_sentence}).
-
-\end{enumerate}
-
-% ~~~~~~~~~~~~~~~~~~~~~~~~
-%
-% REVIEWER 3
-%
-% ~~~~~~~~~~~~~~~~~~~~~~~~
-
-\section{Reviewer 3 (Michael Collyer)}
-
-\begin{enumerate}
-
-% \item{\textcolor{blue}{The author presents a simple example and defers much of the many details users will encounter to his impermanent sources (vignettes, webpages), which I believe to be wise, as the R package will evolve.
-I believe the purpose of a manuscript like this is illustrate its purpose and utility rather than serve as an instructional manual.
-I believe the author has found the appropriate balance of instructional content between this manuscript and his other sources.}}
-
-% \item{\textcolor{blue}{The author acknowledged assistance with writing (suspect English is a second language) and the writing was great in the paper, but less so in the abstract.
-Having someone double check the grammar in the abstract would be beneficial.
-Also, line 116 refers to table ??.}}
-
-\item{\textcolor{blue}{In the examples, replacing objects is not a good idea.
-(This is also a problem in the help file examples in the R package.).
-Remember that novice R users will use this software and might develop the habit of overwriting objects in the global environment, as updates.
-It is possible to update objects with new names and retain the history during analyses, which is preferred in most cases and essential in others (for example, when a different analysis requires going back to a previously defined object.)  I recommend providing unique names.
-For example, disparity is defined twice as an object on lines 255 and 261.
-On line 261 it could be defined as DisparityB, and subsequently used without overwriting the object produced from a different function in line 255.
-Relatedly, one should not name objects the same name as functions.
-This is simply bad news.
-It is not something to encourage users to do.
-Using "Disparity" rather than the function name, disparity, accomplishes the same thing without potentially causing problems down the line.
-(The author will field requests for help from users, so this can be considered a pitfall avoided beforehand.)  This is discussed some here: https://stackoverflow.com/questions/6135868/in-r-what-exactly-is-the-problem-with-having-variables-with-the-same-name-as-ba}}
-
-I have changed the variable names in the package description manuscript to be non-redundant.
-I've also changed the variable names in the package's help files to not override other functions/objects.
-
-\item{\textcolor{blue}{It took me a bit of sleuthing for how the ANOVA on line 326 was performed. 
-If I have one major concern, it is here. 
-Rather than develop an, e.g., anova.test.dispRity S3 generic function, the author was able to use anova.lm by developing a way to coerce the output into a class "lm" object. 
-This is clever but perhaps dangerous. 
-For example, if one compares the results in Tables 3 and 4, they seem largely disparate. 
-I would expect a statistically significant result, as one 95\% CI from Table 3 only slightly overlaps with the others, but an effect size of F = 177.84?? 
-This is highly irregular, and I believe it is because of what the "lm" coercion does and what is actually tested here. 
-My discomfort should have been readily apparent. 
-The ANOVA df signals that what is performed is a parametric ANOVA on the bootstrapped group disparities across bootstrap iterations, as if these values are independent observations. 
-The danger here must be appreciated.
-Increasing the number of bootstrap iterations effectively and artificially increases statistical power (which I would argue should not be an objective). 
-The ANOVA statistics are not based on the data but on the bootstrap results, which is not at all conventional (I could not find a source to confirm this approach). 
-The normal procedure is to calculate a statistic, like an F statistic, in every bootstrap iteration and then draw inference from the distribution of that statistic, much like the "percentile bootstrap" (discussed some here:% \url{https://stats.stackexchange.com/questions/20701/computing-p-value-using-bootstrap-with-r/277391#277391}).
-I believe that an ANOVA consistent with the confidence interval approach in Table 3 would bootstrap values in every permutation representing a null model (e.g., the values sampled come from a population with consistent mean and variance), calculate an F statistic based on among- and within-group disparities, then draw inference from the observed F value in the distribution of random outcomes.}}
-
-\textcolor{blue}{I can see no reason to perform ANOVA like this.
-I'm not sure what it helps.
-I stopped here in terms of my sleuthing but I wonder what other analytical short-cuts like this could cause problems.
-This is important!
-As the author noted, and I agree, there is a paucity of development in this area.
-One of the reasons might be that it takes time to bench test methods.
-Surveying our functions in geomorph will reveal many methodological papers cited along with functions.
-We release functions once we feel they are theoretically sound.
-Introducing methods that have not been vetted will lead users to make dangerously incorrect inferences.
-This is a difficult job but it has to be done.
-While I'm excited to see this package released, to release it with known flawed analytical functions introduces a much larger mess to clean up later.
-My recommendations are these:}
-
-\begin{itemize}
-\item \textcolor{blue}{1) Remove these ANOVA results from the manuscript, which do not enhance the confidence intervals much at all.}
-
-\item \textcolor{blue}{2) Consider removing underdeveloped functions or arguments from your package until you have had sufficient time to test them.
-Release them along with subsequent manuscripts that show you have tested type 1 error rates and statistical power.}
-
-\item \textcolor{blue}{3) Or (and less recommended), put very stark and clear warnings that statistical results produced by some of these functions have not been tested and the user bares some risk making inferences from them.}
-
-\end{itemize}
-
-\label{ANOVA}
-
-I agree that this ANOVA is a garbage-in-garbage-out test and used it more for illustrating the package's possibilities rather than show-casing disparity analysis best (worst?) practices.
-Following this reviewer's comments with which I totally agree, I have completely removed the ANOVA example (from the paper, the manuals and the vignette), replacing it in the paper with a PERMANOVA (based on \texttt{vegan::adonis}) which is more conventional.
-As this reviewer points out, the ANOVA is indeed performed on the bootstrapped data rather than the observed data which, of course, artificially (and wrongly) increases the statistical power.
-Although the ANOVA is not mentioned any more, I have however left the ability for \texttt{test.dispRity} to deal with \texttt{lm} type tests (with a warning being triggered).
-
-I have implemented the reviewers' suggestion similary to the ``percentile bootstrap'' idea, allowing \texttt{test.dispRity} to work on either the pooled bootstrap distribution through the \texttt{concatenate = TRUE} argument (this was indeed erroneous in the previous ANOVA example but could be justifiable in other cases; e.g.
-when comparing distribution overlap probabilities) or by applying the test for each bootstrap pseudo-replicate with \texttt{concatenate = FALSE}.
-I've clarified this in the manuscript:
-
-\textit{Additionally, the test is here performed on the pooled bootstrapped pseudo-replications which can increase the type I error.
-It is possible to compare each bootstrap in a pairwise way without pooling the data by using the} \texttt{concatenate = FALSE} \textit{argument.
-The results will then be a distribution of statistic and p values.} lines 298-302.
-
-I aim to implement more multidimensional test in future versions of the package (or that users develop their own implementations!).
-
-\item{\textcolor{blue}{I also have a recommendation for the package, itself, based on my experiences. 
-The author should strive to make arguments and function details clearer to users, especially as using dispRity requires a lot of integration between different functions. 
-Also, it is a bit of an art, but one to be practiced, to use print.\_ S3 generic functions. 
-I tried to compare some of the dispRity functions to morphol.disparity in geomorph for known data sets. 
-Below are results (which show that t.tests on disparities seem to produce similar results to morphol.disparity, as expected). 
-I use this juxtaposition to demonstrate a subtle difference. 
-A novice user will have no idea what [[1]] and [[2]] means as output, and even though I consider myself fairly experienced with R, I had to go run a t.test on some data to remind myself that "statistic" is the t-stat and "parameter" means the degrees of freedom.
-I'm not sure what the confidence intervals on df should inspire but I respect that in this case - unlike my criticism above - it appears that the bootstrapped outcomes are on all values provided by a t.test in each iteration.
-Regardless, this complexity begs for direction. 
-Although the geomorph results by comparison were tedious to achieve, I can attest that it saved a lot of time spent working with users to understand results. 
-The author should remember that once this manuscript is published, this R package becomes a dedication to the researchers who use it. 
-Cleaning up some of these details now will make it easier to remain dedicated later.}}
-
-I have changed the \texttt{test.dispRity} to now output the name of the statistic and the parameter in the output list (e.g.
-\texttt{statistic: t} and \texttt{parameter: df} for the example provided by the reviewer).
-I have also highlighted in the manuscript one of the features of the function allowing to output the results in a format more familiar to the user using the option \texttt{details = TRUE} (returning the raw test results):
-
-\textit{Note that by default the function only outputs the test's statistic, parameter (if parametric) and the p value.
-However, the raw test results can also be output using the option} \texttt{details = TRUE} \textit{in the function above.} lines 296-298.
-
-Furthermore, I have developed the example for the function \texttt{geomorph.ordination} highlighting how to perform similar analysis with both packages (as shown in the reviewer's attached markdown example).
-
-\end{enumerate}
-
-% I hope these comments are helpful and I look forward to seeing this package fully developed!
-My recommendation would be to accept this manuscript with minor revision, if not for the one major issue I had with the ANOVA method.
-I hope I have made my concern clear and it is otherwise clear that I believe this manuscript would be a nice contribution to MEE upon revision.
-
-% Sincerely,
-% Michael Collyer
-
-% test.dispRity(disparity_var, test = t.test, comparisons = "pairwise",
-% +               concatenate = FALSE, correction = "bonferroni",
-% +               conc.quantiles = c(mean, c(95)))
-% [[1]]
-%                                                   statistic       2.5%     97.5%
-% plethodon$species.Jord : plethodon$species.Teyah -0.06351374 -0.6744747 0.5683461
-
-% [[2]]
-%                                                 parameter     2.5%    97.5%
-% plethodon$species.Jord : plethodon$species.Teyah  43.15056 36.62182 45.99247
-
-% [[3]]
-%  p.value      2.5%     97.5% 
-% 0.7956904 0.4691708 0.9976864
-
-% morphol.disparity(Y~groups , ~groups, data = gdf)
-%  |=================================================================================================================| 100%
-
-% Call:
-% morphol.disparity(f1 = Y ~ groups, groups = ~groups, data = gdf)
-
-% Randomized Residual Permutation Procedure Used
-% 1000 Permutations
-
-% Procrustes variances for defined groups
-%       Jord       Teyah 
-% 0.004149811 0.004234284
-
-% Pairwise absolute differences between variances
-%              Jord        Teyah
-% Jord  0.000000e+00 8.447277e-05
-% Teyah 8.447277e-05 0.000000e+00
-
-% P-Values
-%       Jord Teyah
-% Jord  1.000 0.876
-% Teyah 0.876 1.000
-
-% ~~~~~~~~~~~~~~~~~~~~~~~~
-%
-% REVIEWER 4
-%
-% ~~~~~~~~~~~~~~~~~~~~~~~~
-
-\section{Reviewer 4 (Gavin Simpson)}
-
-\begin{enumerate}
-
-\item{\textcolor{blue}{I have several concerns regarding the content in the manuscript, however, primarily how it is pitched.
-It seems unclear, to me at least, who the intended audience is?
-At various points I get the impression that the author intends for the package to be useful, and therefore, presents it, to readers beyond those studying morphometry etc.
-There are frequent references to ecology and related disciplines where similar approaches are used.
-Yet, those in these broader fields will struggle to follow what is presented as the manuscript exclusively uses "morphospaces" and related terminology that will only be relevant to a specific subset of readers.If you want to aim for a more general audience, consider rewriting the manuscript in more general terminology; samples in rows, variables in columns, for example.
-Perhaps include a table to relate terminology used in morphometry with that more commonly encountered in community ecology.
-Consider adding an ecological example and removing, if length is an issue, content on the manual and vignette --- we don't really need a paragraph on that, just a sentence to highlight it exists.}}
-
-Due to space constraint for this format of MEE manuscript, I have left the references to ecology in the introduction of the manuscript but removed any further references.
-However, I have expanded the ecological example in the vignette which now include a test similar to the one in Diaz \textit{et al} 2016 (Science).
-I therefore hope that the mention of ecology in the introduction will encourage ecology-focused researcher to explore the manual.
-
-\textit{Note that more examples are available in the package manual (...).} lines 229-231
-
-Additionally, I've also added a Glossary table as suggested and generalised disparity data description as suggest by this reviewer's minor comment \ref{rows}.
-
-\item{\textcolor{blue}{The referencing is poor; why cite Close et al for the concept of pairwise distances?
-I'm pretty sure PCO existed before Brusatte et al, and PCA certainly before 2012, etc.
-Throughout I was frustrated with this; the referencing ignores a century of work outside this specific field.
-I appreciate there merit in citing relevant literature for your field, but that doesn't absolve one of a duty to accurately attribute ideas to those that came up with them.
-If you are unfamiliar with developments in ideas beyond your particular field, I would suggest Legendre \& Legendre's *Numerical Ecology* textbook as a good source for original work relevant to what is presented here.}}
-
-I thank this reviewer for sharing the excellent Legendre \& Legendre reference (that was definitely a reference I was missing) and I've accurately credited the statistical methods throughout the manuscript - see my response to reviewer 2, minor suggestion \ref{PCO_cites}.
-
-\item{\textcolor{blue}{The comparison or critique of packages that, in the eyes of the author, restrict users grated a little with me.
-As a developer of one of the packages named, I feel that particular critique is unwarranted for several reasons;
-\begin{itemize}
-\item i) **vegan** doesn't restrict you at all, as with anything in R you are free to pick and choose ordination methods, dissimilarity measures etc and then do what you want with the scores,
-\item ii) **vegan** doesn't provide disparity measures because it was never intended for such usage, but it is general enough that a user could do whatever they want,
-\item and iii) like any software, the **dispaRity** author's software included, you have to provide defaults and **vegan**'s defaults reflect the intended audience of community ecology; that's a feature, as it is of **dispaRity**.
-As such the argument set up on page 6 is some straw man and unnecessary to motivate your package.
-\end{itemize}
-}}
-
-I apologise for the unfounded critique and I have toned down this sentence (see reviewer 1, minor suggestion \ref{tone_down}).
-Following the points raised by this reviewer, I have added functionalities from other packages (namely \texttt{vegan::adonis}) using defaults more appropriate to disparity analysis in palaeobiology (see reviewer 1 major comment \ref{add_geiger_vegan}).
-
-\item{\textcolor{blue}{You don't really define disparity anywhere, except in very loose terms.
-This needs making more explicit; I appreciate that disparity is a broad, nebulous concept, but you need to help the reader that may not be familiar with your subject area.
-Either that or as mentioned above, refocus the paper on those readers already versed in this particular type of analysis.}}
-\label{nebulous}
-
-I have tried to define disparity with more examples and references to be more familiar to a broader range of readers (See reviewer 1, major suggestion \ref{define_disparity}).
-
-\item{\textcolor{blue}{Something that I did miss, which is at the forefront of my mind as a lot of work has gone into **vegan** in this area, is how you handle negative eigenvalues and their eigenvectors?
-Typically, one needs to account for these issues, which readily arise when embedding data in an Euclidean space (via PCO) using a non-metric dissimilarity.
-Does **dispaRity** handle such problems, and if so how?
-If not, is this because those steps are left up to the user to deal with?
-If so, there are some issues that cannot be corrected by the user without strongly biasing the results of the analysis.
-This is one area where there is the potential for the ideal of being accommodating of users can cause problems downstream if you are not in control of issues like non-metric dissimilarities.
-An example where it needs specific handling is in multivariate dispersions (distances around centroids), but even simple ordinations based on dissimilarities require special handling.}}
-
-The package does not deal with negative eigen values in any special way, this task being left to the user to deal with.
-In fact, the package simply takes as input a matrix and only considers rows as elements and columns as dimensions; it ignores whether it is an ordinated matrix or not and this, along with the subsequent tasks is left to the users to deal with appropriately.
-
-I have added a warning about that in the manuscript, in the vignette and the package manual (for the functions that allow the matrix as input).
-
-\textit{The matrix is always considered as the multidimensional space and no correction is applied to it (e.g.
-if ordinated and with negative eigen values, corrections should be applied prior to using the} \texttt{dispRity} \textit{package.)} lines 111-113
-
-\end{enumerate}
-
-\subsection{Reviewer 4 - Minor comments}
-
-\begin{enumerate}
-
-\item{\textcolor{blue}{L17: insert "ability" between "researchers" and "to", and swap order of "suits" and "best" to read "best suits".}}
-
-I've fixed these typos.
-
-\item{\textcolor{blue}{L50: Here is an opportunity to bridge to the community ecology field and make more general statements regarding multivariate data; rows and samples or observations, columns are variables or some function of variables such as an embedding (PCO) or an ordination.}}
-
-I've changed the sentence to:
-\label{rows}
-
-\textit{In all these analysis, each set of multivariate traits forms a multidimensional space.
-This space is represented as a matrix where rows are regarded as samples or observations (e.g.
-a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g.
-an embedding, scaling, ordination, etc.
-of the variables).} lines 48-50
-
-\item{\textcolor{blue}{L51--54: here referencing is very poor.
-Please cite relevant original research.
-Also, expand on what you mean by MDS; is this different from PCO, which also goes by this name --- do you mean NMDS here?}}
-
-I went with Legendre \& Legendre 1998's definition of ordinations (pp.
-388) in the three major distinct categories: (1) principal components analysis (PCA), (2), principal coordinates and metric scaling (PCO, PCoA) and non metric dimensional scaling (NMDS, MDS), ignoring correspondence and factor analysis for clarity.
-See reviewer 2, minor suggestion \ref{PCO_cites}.
-
-\item{\textcolor{blue}{L65--64 here are additional areas where referencing is poor. Consider Marti Anderson's work on multivariate ANOVA \& dispersions as earlier work, for example:
-\begin{itemize}
-\item Anderson, M. J. A new method for non-parametric multivariate analysis of variance. Austral Ecol. 26, 32–46 (2001)
-\item Anderson, M. J. Distance-based tests for homogeneity of multivariate dispersions. Biometrics 62, 245–253 (2006).
-\item Anderson, M. J., Ellingsen, K. E. \& McArdle, B. H. Multivariate dispersion as a measure of beta diversity. Ecol. Lett. 9, 683–693 (2006).
-\end{itemize}
-}}
-\textcolor{blue}{For permutation tests, significant work has been done in this area by Cajo ter Braak, principally in his Canoco software and related publications.
-Also comparing overlap in CIs is a terrible way to perform inference; for 95\% confidence intervals, if the intervals for two estimated means are as close as possible but not overlapping, the p value of the difference means test is 0.005, not 0.05.
-To achieve 0.05 you can allow considerable overlap, but how much is tricky to discern by eye.
-As such you are at risk of making poor choices on the basis of perceived significance.}
-
-I have changed this sentence by citing Anderson 2001 \textit{Austral Ecol.} and Manly 1997 \textit{Randomization, bootstrap and Monte Carlo methods in biology}.
-I also totally agree with the poor choice of simply comparing CIs; unfortunately this method is not that uncommon common in palaeobiology.
-I've removed the citation (to avoid pointing-and-shaming) and added a caveat to this way of comparing disparity:
-
-\textit{[...] an equal variety of statistical tests such as non-parametric multivariate analyses of variance (NPMANOVA, Anderson 2001 ; e.g.
-in Brusatte et al.
-2008) multidimensional permutation tests (Manly 1997 ; e.g.
-in Díaz et al.
-2016) or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements.} lines 69-73
-
-\item{\textcolor{blue}{L116 There is either a missing reference or a missing table, I presume the latter as no existing table matches the textual description in and around this line.}}
-
-Sincerest apologies! See reviewer 1 minor comment \ref{table_missing}.
-
-\item{\textcolor{blue}{L134 Needs "an" between "as" and "argument". L210 delete "sheer". L213 "longer" not "bigger" and "altitude" not "altitudes". L218 "longer" not "bigger"}}
-
-I've fixed these typos (see also reviewer 2 minor comment \ref{altitude}).
-
-\item{\textcolor{blue}{L252 and throughout the code: insert straight quotes.
-There is a good chance these curly quotes will get replaced with some non-ASCII character and be impossible to paste into R.}}
-
-I've converted all quotes in quoted code lines as ASCII straight quotes.
-
-\item{\textcolor{blue}{In the example around L236--249, can you justify the metric used here?
-Is the sum of variances along components of an ordination a valid measure of the dispersion in multivariate space of samples?
-You don't, or I didn't see, reference to original work that justifies this, and it's not something I have come across despite implementing some approaches for testing multivariate dispersion in the **vegan** package.
-Doesn't this suffer the same problem as confidence bars in an x-y scatterplot where the bars have been defined with reference to only one axis at a time?
-The solution is a confidence ellipse, which is often quite different to the interval implied by separate confidence interval bars for each axis.}}
-
-This metric (the sum of each ordination axis variance) is commonly used in palaeobiology (since Foote 1991 and Wills et al.
-1994) probably due to a ``citation diffusion effect'' (i.e.
-future authors replicating convincing methods) and partly due to its pretty good behaviour with palaeontological data (e.g.
-missing data, Ciampaglio 2001).
-I now cite these three references lines 293-294 as a justification.
-
-However, as this reviewer highlights, this metric suffers from several problems such as the fact that it ignores the covariance between the dimensions.
-This can be a problem when the disparity metric is used for an exact proxy of the multidimensional volume (which, as this reviewer proposes, would be closer to the ellipsoid volume).
-However as a standard and comparable disparity metric, the sum of each dimension's variance is mathematically equal to the mean squared pairwise distance, a metric heavily used through it's implementation in \texttt{geiger} (Harmon \textit{et al} 2008 Bioinformatics).
-
-\item{\textcolor{blue}{L294 replace "following up" with "subsequent"}}
-
-I've changed the sentence following this suggestion.
-
-\item{\textcolor{blue}{L323 This statement seems overly vague.
-What level of confidence is there in this test?
-The 95\% intervals overlap throughout, and although you don't show then, one can infer that say the 0.2 probability quantile of subsample 2 overlaps with the 0.975 quantile of sub sample 1.}}
-
-I have removed this statement altogether and now directly applied the test, the reader can now read directly to the results of the test (table 5) and the conclusions lines 365-366.
-
-\item{\textcolor{blue}{L326 Where do the 300 samples come from as implied by the residual degrees of freedom.
-Table 3 suggests 15 + 9 + 13 = 37 samples in this analysis.
-I'm clearly missing something here.
-Also, this is an omnibus test, which if we believe the printed values (see previous re DF) only says there is some effect of `subsamples`.
-Can this analysis be extended to point estimates for mean variance (again assuming you can justify this sum of individual variances method) and post-hoc comparisons of pairwise `subsample` differences, which would be quite useful?}}
-
-The high value of the samples, residuals and degrees of freedom comes from the test being performed on the pooled bootstrap values rather than the raw observations.
-I have now removed any mentions to this dodge anova and replaced it with a more traditional PERMANOVA instead (see reviewer 1 major comment \ref{add_geiger_vegan} and reviewer 3, major comment \ref{ANOVA}).
-
-\item{\textcolor{blue}{If you need to save space to add an ecological example, cut the *Manuals and vignettes* section and also the *Data simulations* section.
-The later is somewhat tangential to the main thrust of the paper and you cover this in the vignette/bookdown resources.}}
-\label{remove_simulations}
-
-I have removed the simulations part as suggested by this reviewer.
-However, following the encouraging suggestions of reviewer 2 (major suggestion \ref{simulations}).
-I have developed in much more details the simulation algorithms in the package manual.
-
-\end{enumerate}
-
-\end{document}

### 2018-02-22 paper/dispRity_paper.tex
Compiled responses

@@
 % 2961w - 30p
-
-%---------------------------------------------
-%
-%       THIS IS THE HIGHLIGHTED VERSION.
-%       For the final version, carefully remove:
-%       '\protect'
-%       '\hl{' (don't forget the '}'
-%
-%---------------------------------------------
 
@@
 The results will then be a distribution of statistic and \textit{p} values.
-Relating back to our question: yes, crown mammals display a higher diversity in morphologies than their stem counterparts (in this example and dataset!).
+Relating back to our question: yes, crown mammals display a higher diversity in morphologies than their stem counterparts (in this example and dataset).
 

### 2018-02-23 inst/gitbook/03_specific-tutorials.Rmd
Revert "v0.5.1"
gitbook updates

@@
 2 | `centroids`<sup>1</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
-1 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
+1 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln$area` |
+1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln$vol` |
 1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
@@
 1 | `n.ball.volume` | The hyper-spherical (*n*-ball) volume | `dispRity` |
-2 | `pairwise.dist` | The pairwise distances between elements | [`stats`](https://cran.r-project.org/web/packages/stats/index.html)`::dist` |
+2 | `pairwise.dist` | The pairwise distances between elements | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::vegist` |
 2 | `radius` | The radius of each dimensions | `dispRity` |
@@
 
-The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
+The convex hull based functions are a call to the `geometry::convhulln` function with the `"FA"` option (computes total area and volume).
+Also note that they are really sensitive to the size of the dataset.
 
@@
 
-The pairwise distances uses the function `stats::dist` and can take the normal `dist` options:
+The pairwise distances uses the function `vegan::vegdist` and can take the normal `vegdist` options:
 
@@
 ```
+
+Note that this function is a direct call to `vegan::vegdist(matrix, method = method, diag = FALSE, upper = FALSE, ...)`.
 

### 2018-02-28 paper/response_to_reviewers2.tex
Paper update

@@
 
-\textbf{RE: MEE-17-08-729 }\\
+\textbf{RE: MEE-18-01-025}\\
 \bigskip
@@
 
+I havefixed this typo.
+
 \item{\textcolor{blue}{- Rows in Tables 1 and 2 are not well distinguished, i.e., I had to look at multiple columns to know where text broke over a line versus a new row.}}
+
+I have added horizontal lines to this table.
 
@@
 
+I am not entirely sure what this reviewer refers to.
+I have double checked punctuation in this paragraph.
+
 \item{\textcolor{blue}{L304 - Probably shouldn't use exclamation marks in a scientific paper unless they are factorials.}}
 
-I've removed the exclamation mark.
+I have removed the exclamation mark.
 
@@
 
-\item{\textcolor{blue}{Finally, there were some English irregularities (e.g., data is the plural form of datum) that could be fixed with some proofing from a native English-speaking scientist.
-(I would not trust this to a copy-editor over a colleague, if technical/scientific writing should be precise).}}
+I have now specifically made clear that \texttt{convhull.surface}, \texttt{convhull.volume} and \texttt{pairwise.dist} are just duplicates of their associated original functions, both in the manuscript, package and vignettes.
+
+\item{\textcolor{blue}{Finally, there were some English irregularities (e.g., data is the plural form of datum) that could be fixed with some proofing from a native English-speaking scientist..}}
+
+A colleague has double checked the manuscript for typos.
 
 \end{enumerate}
+
+\end{document}

### 2018-02-28 paper/dispRity_paper.tex
Paper update

@@
 
-        \item There are as many definition of these multidimensional spaces, metrics and tests as there are questions that can be tackled with such methods.
+        \item There are as many definitions of these multidimensional spaces, metrics and tests as there are questions that can be tackled with such methods.
         Many of these methods are implemented in specific software or R packages.
         However, the definition of the space, metric and test is often dependent on the software/package's author's point of view or specific question.
-        This can unfortunately hamper researchers ability to apply different methods that best suits their specific questions.
+        This can unfortunately hamper researchers' ability to apply different methods that best suits their specific questions.
 
@@
 
-        \item The \disp package also comes with an extend manual regularly updated following users questions or suggestions.
+        \item The \disp package also comes with an extend manual regularly updated following users' questions or suggestions.
         Furthermore, the package contains some simulation tools (e.g to simulate complex multidimensional space or morphological data).
-        Finally it also contains a suite of utilities to work with \disp objects aimed at helping users to develop their own multidimensional metrics and/or tests.
+        Finally, it also contains a suite of utilities to work with \disp objects aimed at helping users to develop their own multidimensional metrics and/or tests.
 
@@
 
-% Variety of multidimensional spaces, metrics and tests
 In all these analysis, each set of multivariate traits forms a multidimensional space.
@@
 The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
-For example, when using morphological traits the space will be a morphospace, when using ecological traits it may be referred to as an ecospace, etc.
+For example, when using morphological traits, the space will be a morphospace, when using ecological traits it may be referred to as an ecospace, etc.
 
@@
 
-% Problems with this variety
 In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods.
@@
 does groups A and B occupy the same location?).
-This can lead to inappropriate analyses by users confined by the existing software implementations. %Oh no!
+This can lead to inappropriate analyses by users confined by the existing software implementations.
 
-% Here comes the solution!
 The aim of the \disp package is avoid such problems by providing a completely flexible framework for studying multidimensional data.
@@
         \hline
-        in this manuscript & in \texttt{dispRity} & in palaeobiology & in ecology \\ 
+        \textbf{in this manuscript} & \textbf{in} \texttt{\textbf{dispRity}} & \textbf{in palaeobiology} & \textbf{in ecology} \\ 
         \hline
         multidimensional space & matrix ($n\times k$) & morphospace, traitspace, etc. & ecospace, function-space, etc. \\
+        \hline
         elements & rows ($n$) & taxa, specimen, etc. & taxa, field sites, environments, etc. \\
+        \hline
         dimensions & columns ($k$) & ordination scores, distances, etc. & ordination scores, distances, etc. \\
-        subsets & matrix ($m \times k$, with $m \leq n$) & e.g. every element in an strata or sharing the same ancestor & e.g. elements living in the same environment \\
+        \hline
+        subsets & matrix ($m \times k$, with $m \leq n$) & e.g. every element in a stratum or sharing the same ancestor & e.g. elements living in the same environment \\
+        \hline
         disparity & a dimension-level 1 or 2 function$^1$ & disparity: e.g.
-the sum of variances \cite{Wills2001}, the average pairwise distance between taxa \cite{Foote01071994}, etc.
+the sum of variances \cite{Wills2001}, the average pairwise distances between taxa \cite{Foote01071994}, etc.
 & dissimilarity: e.g.
@@
    \includegraphics[width=0.8\textwidth]{../inst/gitbook/dispRity_fun.pdf} 
-\caption{Illustration of the different metric dimension-levels in the \disp package. In this example, each cell correspond to a single value (e.g. a 8 $\times$ 7 matrix or a vector of eight elements).
+\caption{Illustration of the different metric dimension-levels in the \disp package.
+In this example, each cell corresponds to a single value (e.g.
+a 8 $\times$ 7 matrix or a vector of eight elements).
 A dimension-level 3 matrix would be a metric resulting in a matrix (e.g.
@@
     \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
-        name & description & dim & definition & source \\
+        \textbf{name} & \textbf{description} & \textbf{dim} & \textbf{definition} & \textbf{source} \\
         \hline
         \texttt{ancestral. dist} & The distance between an element and its ancestor & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}$ & This package \\
+        \hline
         \texttt{centroids} & The distance between each element and a fixed point$^1$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & This package \\
+        \hline
         \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
+        \hline
         \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
+        \hline
         \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & This package \\
+        \hline
         \texttt{ellipse.
 volume}$^2$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & This package; based on \cite{DonohueDim}\\
+        \hline
         \texttt{mode.val} & The modal value & 1 & NA & This package\\
+        \hline
         \texttt{n.ball. volume} & The hyper-spherical ($n$-ball) volume & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R$ & This package\\
+        \hline
         \texttt{pairwise. dist} & The pairwise distances between elements & 2 & NA & \texttt{vegan::vegdist} \citep{oksanen2007vegan}\\        
+        \hline
         \texttt{radius} & The radius of each dimensions & 2 & $|\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|$ & This package\\
+        \hline
         \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & This package \\
+        \hline
         \texttt{span.tree. length} & The minimal spanning tree length & 1 & $\sum(\mathrm{branch\ length})$ & \texttt{vegan::spantree} \citep{oksanen2007vegan}\\
+        \hline
         \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & This package \\
@@
 When a slice occurs not on a tip or a node, six methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice.
-For each method, the selected data is: %TODO HL
+For each method, the selected data is:
+
 \begin{itemize}
@@
 
-Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where $n = 1$ (somewhat akin to jackknife resampling).
+Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms: (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data); and (2) the ``single'' algorithm where $n = 1$ (somewhat akin to jackknife resampling).
 
@@
 \section{Examples}
-Multivariate analysis can be really useful for looking at multiple aspects of organisms diversity together.
+Multivariate analysis can be really useful for looking at multiple aspects of organisms' diversity together.
 For example, one can also look the diversity of morphologies \citep[or disparity;][]{foote1991morphological}.
@@
 To visualise the actual disparity values, one can use the \texttt{summary} or/and \texttt{plot} options (Table \ref{Tab:summary_group} and Fig. \ref{Fig:plot_group}):
-
-% mammal_groups <- list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44))
-% subsets <- custom.subsets(data = BeckLee_mat50, group = mammal_groups)
-% bootstraps <- boot.matrix(subsets)
-% disparity <- dispRity(bootstraps, metric = c(sum, variances))
 
@@
 
-Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA  \citep[PERMANOVA; ][]{permanova} through the \texttt{adonis.disprity} function that is a wrapper of the \texttt{vegan::adonis} function  \citep{oksanen2007vegan} for \texttt{dispRity} objects:
+Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, multivariate permutation ANOVA  \citep[PERMANOVA; ][]{permanova} through the \texttt{adonis.disprity} function that is a wrapper of the \texttt{vegan::adonis} function  \citep{oksanen2007vegan} for \texttt{dispRity} objects:
 
@@
 
-To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
+To answer our specific question above: yes, there is an effect of time on morphological disparity (an increase) in this dataset.
 Note that in this case, the function outputs different warnings on the usage of such test and the eventual data not used in the test.
@@
 
-% \subsection{Further directions} TG: maybe this section is useless
-% Further functionalities and improvement on current functionalities will be implemented in future versions.
-% In fact, optimised architecture and a C implementations are being currently developed for faster calculations and lower memory footprints for most computational heavy functions (e.g.
-\disp or \texttt{time.subsets}).
-% New functionalities currently being developed will include time sequential linear models (for accurately measuring directionality in disparity changes between time slices) or mode of disparity evolution testing (allowing to fit different models such as Brownian Motion or Ornstein-Uhlenbeck to disparity curves).
-
 \section{Conclusion}
@@
 The \disp object allows users to easily pipeline disparity analysis from the data input (the matrix) to publication standard results (tables, plots, hypothesis testing).
-
-% \section{Data availability and reproducibility}
-% All the code and the data for using the package and its vignette with examples are available in the R package help files or on GitHub \url{github.com/TGuillerme/dispRity}.
 

### 2018-03-01 inst/gitbook/05_guts.Rmd
changed scale to rescale

@@
 
-#### `scale.dispRity`
-This is the S3 method of `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object.
+#### `rescale.dispRity`
+This is the modified S3 method for `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object and can take optional arguments (for example the rescaling by dividing by a maximum value).
 
@@
 ## Scaling the same disparity values
-head(summary(scale(disparity, scale = TRUE)))
+head(summary(rescale.dispRity(disparity, scale = TRUE)))
 
 ## Scaling and centering:
-head(summary(scale(disparity, scale = TRUE, center = TRUE)))
+head(summary(rescale.dispRity(disparity, scale = TRUE, center = TRUE)))
+
+## Rescaling the value by dividing by a maximum value
+head(summary(rescale.dispRity(disparity, max = 10)))
 ```

### 2018-03-01 R/dispRity-package.R

@@ -11,13 +11,13 @@
 #' @keywords disparity, ordination, phylogeny, cladistic, morphometric, ecology
 #'
 #' @import ape
+#' @import stats
 #' @importFrom paleotree timeSliceTree 
 #' @importFrom geometry convhulln 
 #' @importFrom ade4 randtest as.randtest 
 #' @importFrom grDevices colorRampPalette grey 
 #' @importFrom caper comparative.data 
 #' @importFrom graphics axis boxplot hist image lines mtext par plot points polygon text legend
-#' @importFrom stats bw.nrd0 coef glm p.adjust quantile rnorm var median cmdscale as.dist density optim bartlett.test
 #' @importFrom utils combn data capture.output
 #' @importFrom phyclust gen.seq.HKY 
 #' @importFrom phangorn dist.hamming NJ RF.dist CI RI optim.parsimony parsimony

### 2018-03-01 inst/gitbook/03_specific-tutorials.Rmd
Updated book
changed summary to proper S3 method

@@
 ```
-> Because there is only one subset (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with rounding).
+> Because there is only one subset (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with digits).
 
@@
 > Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) that often results in near 0 disparity values.
+I strongly recommend reading [this really intuitive explanation](https://beta.observablehq.com/@tophtucker/theres-plenty-of-room-in-the-corners) from [Toph Tucker](https://github.com/tophtucker).
 
@@
  * `cent.tend` for the central tendency to use for summarising the results (default is `median`)
- * rounding` option corresponding to the number of decimal places to print (default is `2`)
+ * digits` option corresponding to the number of decimal places to print (default is `2`)
  * `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
@@
 
-## Printing the details of the object and rounding the values to the 5th decimal place
-summary(disparity_time_slices, recall = TRUE, rounding = 5)
+## Printing the details of the object and digits the values to the 5th decimal place
+summary(disparity_time_slices, recall = TRUE, digits = 5)
 ```

### 2018-03-01 R/summary.dispRity.R

@@ -3,10 +3,11 @@
 #' @description Creates a summary of a \code{dispRity} object.
 #'
 #' @param object A \code{dispRity} object.
+#' @param ... Additional arguments to be passed to \code{\link{summary}}.
 #' @param quantiles The quantiles to display (default is \code{quantiles = c(50, 95)}; is ignored if the \code{dispRity} object is not bootstrapped).
 #' @param cent.tend A function for summarising the bootstrapped disparity values (default is \code{\link[stats]{median}}).
 #' @param recall \code{logical} value specifying whether to recall the \code{dispRity} parameters input (default = \code{FALSE}).
-#' @param rounding Optional, a value for rounding the values in the output table (default = 2).
+#' @param digits Optional, a value for digits the values in the output table (default = 2).
 # ' @param results Optional, in the case of summarising a \code{\link{sequential.test}} which results to display (default = "coefficients")
 #'
 #' @return
@@ -23,7 +24,7 @@
 #' ## Summarising the results
 #' summary(disparity) # default
 #' ## Using different options
-#' summary(disparity, quantiles = 75, cent.tend = mean, rounding = 8,
+#' summary(disparity, quantiles = 75, cent.tend = mean, digits = 8,
 #'      recall = TRUE)
 #' 
 #' @seealso \code{\link{dispRity}}, \code{\link{plot.dispRity}}.
@@ -48,7 +49,7 @@
 # recall <- FALSE
 # match_call <- list() ; match_call$cent.tend <- "median"
 
-summary.dispRity <- function(object, quantiles = c(50, 95), cent.tend = median, recall = FALSE, rounding){#, results = "coefficients") {
+summary.dispRity <- function(object, ..., quantiles = c(50, 95), cent.tend = median, recall = FALSE, digits){#, results = "coefficients") {
     data <- object
 
     #----------------------
@@ -72,12 +73,12 @@ summary.dispRity <- function(object, quantiles = c(50, 95), cent.tend = median, 
     #recall
     check.class(recall, "logical")
 
-    #rounding
-    if(missing(rounding)) {
+    #digits
+    if(missing(digits)) {
         #Set to default (see below)
-        rounding <- "default"
+        digits <- "default"
     } else {
-        check.class(rounding, "numeric")
+        check.class(digits, "numeric")
     }
 
     #DATA
@@ -115,7 +116,7 @@ summary.dispRity <- function(object, quantiles = c(50, 95), cent.tend = median, 
[…]

### 2018-03-01 R/summary.dispRity_fun.R

@@ -64,27 +64,27 @@ get.digit <- function(column) {
 }
 
 
-round.column <- function(column, rounding) {
-    ## Get the rounding value
-    rounding <- ifelse(rounding != "default", rounding, get.digit(as.numeric(column)))
+round.column <- function(column, digits) {
+    ## Get the digits value
+    digits <- ifelse(digits != "default", digits, get.digit(as.numeric(column)))
     ## Round the table
     if(class(column) != "group") {
-        return(round(as.numeric(column), digits = rounding))
+        return(round(as.numeric(column), digits = digits))
     } else {
-        return(round(as.numeric(as.character(column)), digits = rounding))
+        return(round(as.numeric(as.character(column)), digits = digits))
     }
 }
 
 
-## Function for rounding the results
-rounding.fun <- function(results_table, rounding, model.test = FALSE) {
+## Function for digits the results
+digits.fun <- function(results_table, digits, model.test = FALSE) {
 
     ## model.test
     start_column <- ifelse(model.test, 1, 3)
 
-    ## Apply the rounding
+    ## Apply the digits
     rounded_table <- as.matrix(results_table[,c(start_column:ncol(results_table))])
-    rounded_table <- apply(rounded_table, 2, round.column, rounding)
+    rounded_table <- apply(rounded_table, 2, round.column, digits)
     results_table[,c(start_column:ncol(results_table))] <- rounded_table
 
     return(results_table)
@@ -169,7 +169,7 @@ match.parameters <- function(one_param, full_param) {
 #     return(results_table)
 # }
 
-# summary.seq.test <- function(data, quantiles, cent.tend, recall, rounding, results, match_call) {
+# summary.seq.test <- function(data, quantiles, cent.tend, recall, digits, results, match_call) {
     
 #     ## SANITIZING
 #     ## quantiles

### 2018-03-01 inst/gitbook/02_getting-started.Rmd
Changed time.subsets and merge.subsets to chrono.subsets and combine.subsets

@@
 Should you rarefy the data?
-See [`time.subsets`](#time-slicing), [`custom.subsets`](#customised-subsets), [`boot.matrix`](#bootstraps-and-rarefactions) and [`dispRity.metric`](#disparity-metrics) for more details of the defaults used in each of these functions.
+See [`chrono.subsets`](#time-slicing), [`custom.subsets`](#customised-subsets), [`boot.matrix`](#bootstraps-and-rarefactions) and [`dispRity.metric`](#disparity-metrics) for more details of the defaults used in each of these functions.
 Note that any of these default arguments can be changed within the `disparity.through.time` or `disparity.per.group` functions.

### 2018-03-01 inst/gitbook/03_specific-tutorials.Rmd
Changed time.subsets and merge.subsets to chrono.subsets and combine.subsets

@@
 
-The function `time.subsets` allows users to divide the matrix into different time subsets or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
+The function `chrono.subsets` allows users to divide the matrix into different time subsets or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
 Each subset generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.
@@
 ## Generating three time bins containing the taxa present every 40 Ma
-time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+chrono.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
                 time = c(120, 80, 40, 0))
@@
 ## Automatically generate three equal length bins:
-time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+chrono.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
                 time = 3)
@@
 ## Generating time bins including taxa that might span between them
-time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+chrono.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
                 time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
@@
 ## Generating four time slices every 40 million years under a model of proximity evolution
-time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
+chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
     method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
@@
 ## Generating four time slices automatically
-time.subsets(data = BeckLee_mat99, tree = BeckLee_tree,
+chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree,
     method = "continuous", model = "proximity", time = 4, FADLAD = BeckLee_ages)
@@
 
-Of course, one could directly supply the subsets generated above (using `time.subsets` or `custom.subsets`) to this function.
+Of course, one could directly supply the subsets generated above (using `chrono.subsets` or `custom.subsets`) to this function.
 
@@
 ## Creating time slice subsets
-time_slices <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
+time_slices <- chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
                                method = "continuous", model = "proximity", 
@@
 ## Creating time slice subsets
-time_slices <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
+time_slices <- chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
     method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
@@
 ## Creating time bin subsets
-time_bins <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
+time_bins <- chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
     method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages,
@@
     cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
-    ylab = c("Disparity", "Diversity"), time.subsets = FALSE,
+    ylab = c("Disparity", "Diversity"), chrono.subsets = FALSE,
     xlab = "Time (in in units from past to present)", observed = TRUE,
@@
 
-Finally, it is possible to use objects generated by `time.subsets`.
+Finally, it is possible to use objects generated by `chrono.subsets`.
 In this case, `adonis.dispRity` will applied the `matrix ~ time` formula by default:
@@
 ## Creating time series
-time_subsets <- time.subsets(BeckLee_mat50, BeckLee_tree, 
+time_subsets <- chrono.subsets(BeckLee_mat50, BeckLee_tree, 
      method = "discrete", inc.nodes = FALSE, time = c(100, 85, 65, 0),
@@
 
-To use each time subset as a separate predictor, you can use the `matrix ~ time.subsets` formula; this is equivalent to `matrix ~ first_time_subset + second_time_subset + ...`:
+To use each time subset as a separate predictor, you can use the `matrix ~ chrono.subsets` formula; this is equivalent to `matrix ~ first_time_subset + second_time_subset + ...`:
 
@@
 ## Running the NPMANOVA with each time bin as a predictor
-adonis.dispRity(time_subsets, matrix ~ time.subsets)
+adonis.dispRity(time_subsets, matrix ~ chrono.subsets)
 ```

### 2018-03-01 inst/gitbook/06_palaeo-demo.Rmd
Changed time.subsets and merge.subsets to chrono.subsets and combine.subsets

@@
 
-The `dispRity` package provides a `time.subsets` function that allows users to split the morphospace into time slices (using `method = continuous`) or into time bins (using `method = discrete`).
+The `dispRity` package provides a `chrono.subsets` function that allows users to split the morphospace into time slices (using `method = continuous`) or into time bins (using `method = discrete`).
 In this example, we are going to split the morphospace into five equal time bins of 20 million years long from 100 million years ago to the present.
@@
 
-## Splitting the morphospace using the time.subsets function
-(binned_morphospace <- time.subsets(data = BeckLee_mat50, tree = BeckLee_tree,
+## Splitting the morphospace using the chrono.subsets function
+(binned_morphospace <- chrono.subsets(data = BeckLee_mat50, tree = BeckLee_tree,
     method = "discrete", time = time_bins, inc.nodes = FALSE,

### 2018-03-01 R/custom.subsets.R

@@ -44,7 +44,7 @@
 #' custom.subsets(BeckLee_mat50, group = BeckLee_tree)
 #' 
 #'
-#' @seealso \code{\link{time.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}, \code{\link{crown.stem}}.
+#' @seealso \code{\link{chrono.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}, \code{\link{crown.stem}}.
 #'
 #' @author Thomas Guillerme
 

### 2018-03-01 R/dispRity-package.R

@@ -69,7 +69,7 @@ NULL
 #'
 #' And example of a \code{dispRity} object.
 #'
-#' This matrix is based on the \code{\link{BeckLee}} dataset and split into seven continuous subsets (\code{\link{time.subsets}}).
+#' This matrix is based on the \code{\link{BeckLee}} dataset and split into seven continuous subsets (\code{\link{chrono.subsets}}).
 #' It was bootstrapped 100 times (\code{\link{boot.matrix}}) with four rarefaction levels.
 #' Disparity was calculated as the \code{\link[stats]{median}} of the \code{\link{centroids}} (\code{\link{dispRity}}).
 #'

### 2018-03-01 tests/testthat/test-dispRity.utilities.R

@@ -175,7 +175,7 @@ test_that("matrix.dispRity", {
 test_that("get.subsets", {
     data(BeckLee_mat99)
     data(BeckLee_tree)
-    subsets_full <- time.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
+    subsets_full <- chrono.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
     bootstrapped_data <- boot.matrix(subsets_full, bootstraps = 10, rarefaction = c(3, 5))
     disparity_data <- dispRity(bootstrapped_data, variances)
 
@@ -246,7 +246,7 @@ test_that("get.subsets", {
 ## extract.dispRity
 test_that("extract.dispRity", {
     data(BeckLee_mat99) ; data(BeckLee_tree) 
-    subsets_full <- time.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
+    subsets_full <- chrono.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
     bootstrapped_data <- boot.matrix(subsets_full, bootstraps = 10, rarefaction = c(3, 5))
     data <- dispRity(bootstrapped_data, c(sum,variances))
 
@@ -357,7 +357,7 @@ test_that("rescale.dispRity", {
 
 test_that("sort.dispRity", {
     data(BeckLee_mat99) ; data(BeckLee_tree) 
-    subsets <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous", time = 5, model = "acctran")
+    subsets <- chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous", time = 5, model = "acctran")
     data <- dispRity(subsets, metric = mean)
 
     expect_error(
@@ -382,8 +382,8 @@ test_that("sort.dispRity", {
 
 
 
-## merge.subsets
-test_that("merge.subsets", {
+## combine.subsets
+test_that("combine.subsets", {
     data(disparity)
     data_test1 <- disparity
     expect_warning(data_test2 <- custom.subsets(matrix(rnorm(120), 40), group = list("a" = c(1:5), "b" = c(6:10), "c" = c(11:20), "d" = c(21:24), "e" = c(25:30), "f" = c(31:40))))
@@ -403,43 +403,43 @@ test_that("merge.subsets", {
 
[…]

### 2018-03-01 paper/dispRity_paper.tex
fixed time.subsets to chrono.subsets in paper

@@
 
-This splitting can be done using the \texttt{custom.subsets} or \texttt{time.subsets} functions.
+This splitting can be done using the \texttt{custom.subsets} or \texttt{chrono.subsets} functions.
 The first argument is the matrix defining the space and a list of elements to group into different subsets.
@@
 
-\noindent \texttt{> time\_subsets <- time.subsets(data = BeckLee\_mat50, tree = BeckLee\_tree,}\\ \texttt{time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
+\noindent \texttt{> time\_subsets <- chrono.subsets(data = BeckLee\_mat50, tree = BeckLee\_tree,}\\ \texttt{time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
 \noindent \texttt{> bootstraps2 <- boot.matrix(time\_subsets)}\\

### 2018-03-01 inst/gitbook/02_getting-started.Rmd
v1.0.0
# Conflicts:
#	NEWS.md
#	R/boot.matrix.R
#	R/dispRity.R
#	inst/gitbook/_book/details-of-specific-functions.html
#	inst/gitbook/_book/dispRity_manual.pdf
#	inst/gitbook/_book/dispRity_manual.tex
#	inst/gitbook/_book/ecology-demo.html
#	inst/gitbook/_book/future-directions.html
#	inst/gitbook/_book/getting-started-with-disprity.html
#	inst/gitbook/_book/glossary.html
#	inst/gitbook/_book/index.html
#	inst/gitbook/_book/making-stuff-up.html
#
inst/gitbook/_book/palaeobiology-demo-disparity-through-time-and-within-
groups.html
#	inst/gitbook/_book/references.html
#	inst/gitbook/_book/search

@@
 These matrices can be observations, pairwise differences between elements, ordinations, etc...
-
-> Note that whether the matrix is an ordinated one or of any other type, it is always considered as the multidimensional space to analysis and no corrections are applied to it (e.g.
-if ordinated and with negative eigen values, no correction is used).
-Always make sure the input is the right multidimensional space to analyse!
 

### 2018-03-01 inst/gitbook/05_guts.Rmd
v1.0.0
# Conflicts:
#	NEWS.md
#	R/boot.matrix.R
#	R/dispRity.R
#	inst/gitbook/_book/details-of-specific-functions.html
#	inst/gitbook/_book/dispRity_manual.pdf
#	inst/gitbook/_book/dispRity_manual.tex
#	inst/gitbook/_book/ecology-demo.html
#	inst/gitbook/_book/future-directions.html
#	inst/gitbook/_book/getting-started-with-disprity.html
#	inst/gitbook/_book/glossary.html
#	inst/gitbook/_book/index.html
#	inst/gitbook/_book/making-stuff-up.html
#
inst/gitbook/_book/palaeobiology-demo-disparity-through-time-and-within-
groups.html
#	inst/gitbook/_book/references.html
#	inst/gitbook/_book/search

@@
 
-#### `get.subsets.dispRity`
+#### `get.subsets`
 This function creates a dispRity object that contains only elements from one specific subsets.
@@
 ## Extracting all the data for the crown mammals
-(crown_mammals <- get.subsets.dispRity(disp_crown_stemBS, "Group.crown"))
+(crown_mammals <- get.subsets(disp_crown_stemBS, "Group.crown"))
 
@@
 ```
+
+#### `combine.subsets`
+This function allows to merge different subsets.
+If the
+
+```{r, eval=FALSE}
+## Combine the two first subsets in the dispRity data example
+combine.subsets(disparity, c(1,2))
+```
+
+Note that the computed values (bootstrapped data + disparity metric) are not merge.
 
@@
  * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"equal.split"`, `"gradual.split"`).
-This element generated only once via `time.subsets()` and `custom.subsets()`.
+This element generated only once via `chrono.subsets()` and `custom.subsets()`.
  * `$call$dimensions`: either a single `numeric` value indicating how many dimensions to use or a vector of `numeric` values indicating which specific dimensions to use.

### 2018-03-01 R/Claddis.ordination.R

@@ -16,7 +16,7 @@
 #' Claddis.ordination(Claddis::Michaux1989)
 #' }
 #'
-#' @seealso \code{\link[Claddis]{MorphDistMatrix}}, \code{\link[Claddis]{ReadMorphNexus}}, \code{\link[Claddis]{MakeMorphMatrix}}, \code{\link[stats]{cmdscale}}, \code{\link{custom.subsets}}, \code{\link{time.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}.
+#' @seealso \code{\link[Claddis]{MorphDistMatrix}}, \code{\link[Claddis]{ReadMorphNexus}}, \code{\link[Claddis]{MakeMorphMatrix}}, \code{\link[stats]{cmdscale}}, \code{\link{custom.subsets}}, \code{\link{chrono.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}.
 #' 
 #' @author Thomas Guillerme
 #' @export
@@ -28,7 +28,7 @@ Claddis.ordination <- function(data, distance = "Gower", transform = "arcsine_sq
     ## Data
     error_msg <- paste(match_call$data, "does not contain a matrix.\nUse Claddis::ReadMorphNexus to generate the proper data format.")
     check.class(data, "list", msg = error_msg)
-    ## Must have at least one matrix
+    ## Must have at least one matrix
     if(!any(names(data) %in% "matrix")) {
         stop(error_msg, call. = FALSE)
     }
@@ -58,7 +58,7 @@ Claddis.ordination <- function(data, distance = "Gower", transform = "arcsine_sq
     ## Transforming the Claddis data
 
     ## Compute the distance
-    distance <- MorphDistMatrix.support(data, distance = distance) #TG: Change to the proper version
+    distance <- MorphDistMatrix.support(data, distance = distance)
 
     ## Ordinate the matrix
     ordination <- stats::cmdscale(distance, k = k, ...)

### 2018-03-01 R/custom.subsets.R

@@ -42,7 +42,7 @@
 #' custom.subsets(BeckLee_mat50, group = BeckLee_tree)
 #' 
 #'
-#' @seealso \code{\link{time.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}, \code{\link{crown.stem}}.
+#' @seealso \code{\link{chrono.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}, \code{\link{crown.stem}}.
 #'
 #' @author Thomas Guillerme
 

### 2018-03-01 R/dispRity-package.R

@@ -11,20 +11,21 @@
 #' @keywords disparity, ordination, phylogeny, cladistic, morphometric, ecology
 #'
 #' @import ape
+#' @import stats
 #' @importFrom paleotree timeSliceTree 
-#' @importFrom snow makeCluster stopCluster parLapply 
 #' @importFrom geometry convhulln 
-# @importFrom hypervolume hypervolume estimate_bandwidth get_volume 
 #' @importFrom ade4 randtest as.randtest 
 #' @importFrom grDevices colorRampPalette grey 
 #' @importFrom caper comparative.data 
 #' @importFrom graphics axis boxplot hist image lines mtext par plot points polygon text legend
-#' @importFrom stats bw.nrd0 coef dist glm p.adjust quantile rnorm var median runif cmdscale
 #' @importFrom utils combn data capture.output
 #' @importFrom phyclust gen.seq.HKY 
 #' @importFrom phangorn dist.hamming NJ RF.dist CI RI optim.parsimony parsimony
-#' @importFrom RCurl getURL url.exists 
-#' 
+#' @importFrom vegan adonis vegdist
+#' @importFrom geiger dtt ratematrix sim.char
+#' @importFrom parallel parLapply detectCores makeCluster clusterExport stopCluster
+##' @importFrom Claddis MorphDistMatrixFast
+
 
 NULL
 
@@ -68,7 +69,7 @@ NULL
 #'
 #' And example of a \code{dispRity} object.
 #'
-#' This matrix is based on the \code{\link{BeckLee}} dataset and split into seven continuous subsets (\code{\link{time.subsets}}).
+#' This matrix is based on the \code{\link{BeckLee}} dataset and split into seven continuous subsets (\code{\link{chrono.subsets}}).
 #' It was bootstrapped 100 times (\code{\link{boot.matrix}}) with four rarefaction levels.
 #' Disparity was calculated as the \code{\link[stats]{median}} of the \code{\link{centroids}} (\code{\link{dispRity}}).
 #'

### 2018-03-01 R/dispRity.utilities.R

@@ -159,7 +159,7 @@ matrix.dispRity <- function(data, subsets, rarefaction, bootstrap){
 ## DEBUG
 # source("sanitizing.R")
 # data(BeckLee_mat99) ; data(BeckLee_tree) 
-# subsets_full <- time.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
+# subsets_full <- chrono.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
 # bootstrapped_data <- boot.matrix(subsets_full, bootstraps = 10, rarefaction = c(3, 5))
 # disparity_data <- dispRity(bootstrapped_data, variances)
 # get.subsets(bootstrapped_data, subsets = "66.75552") # 1 subsets for 23 elements
@@ -216,7 +216,7 @@ get.subsets <- function(data, subsets) {
 ## DEBUG
 # source("sanitizing.R")
 # data(BeckLee_mat99) ; data(BeckLee_tree) 
-# subsets_full <- time.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
+# subsets_full <- chrono.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
 # bootstrapped_data <- boot.matrix(subsets_full, bootstraps = 10, rarefaction = c(3, 5))
 # data <- dispRity(bootstrapped_data, c(sum,variances))
 # extract.dispRity(data, observed = FALSE, rarefaction = 5,subsets = 2)
@@ -228,7 +228,7 @@ extract.dispRity <- function(data, subsets, observed = TRUE, rarefaction = FALSE
     
     ## Data
     check.class(data, "dispRity")
-    ## Data must have disparity values
+    ## Data must have disparity values
     if(is.null(data$call$disparity)) {
         stop("dispRity object does not contain disparity values.")
     }
@@ -244,15 +244,14 @@ extract.dispRity <- function(data, subsets, observed = TRUE, rarefaction = FALSE
     }
 
     ## Rarefaction
-    if(!observed) {
-        if(rarefaction != FALSE) {
-            check.class(rarefaction, c("numeric", "integer"))
-            check.length(rarefaction, 1, errorif = FALSE, msg = "Only one rarefaction level can be used.")
[…]

### 2018-03-01 R/plot.dispRity_fun.R

@@ -546,7 +546,7 @@ plot.randtest <- function (data_sub, nclass = 10, coeff = 1, ...) {
     xlim <- data_sub$plot$xlim
     ylim <- c(0, max(data_sub$plot$hist$count))
 
-    ## Plotting the simulated data
+    ## Plotting the simulated data
     plot(data_sub$plot$hist, xlim = xlim, col = grey(0.8), ...)
 
     ## Adding the observed data
@@ -556,3 +556,85 @@ plot.randtest <- function (data_sub, nclass = 10, coeff = 1, ...) {
     ## Adding the legend (test results)
     legend("topleft", bty = "n", legend = c("p-value", round(data_sub$pvalue, 5)), cex = 0.7, adj = 0.2)
 }
+
+## Plotting model tests results
+# plot.model.test.support <- function(data, col, ylab, ylim, ...) {
+
+#         ## Extracting the weighted aicc
+#         weight_aicc <- data$aic.models[,3]
+
+#         ## Ordering the weighted aicc
+#         ordered_aicc <- weight_aicc[order(weight_aicc, decreasing = TRUE)]
+
+#         ## Plot
+#         plotcoords <- barplot(ordered_aicc, col = col, ylim = ylim,  ...)
+#         # plotcoords <- barplot(ordered_aicc, col = col, ylab = "Akaike weights") ; warning("DEBUG model.test plot")
+# }
+
+# ~~~~~~~~~~
+# sequential.test plots
+# ~~~~~~~~~~
+
+
+#Plot sequential.test shortcut
+# if(length(class(data)) == 2) {
+#     if(class(data)[[1]] == "dispRity" && class(data)[[2]] == "seq.test") {
+
+#         #lines.args sanitizing
+#         if(!is.null(lines.args)) check.class(lines.args, "list")
+
+#         #token.args sanitizing
+#         if(!is.null(token.args)) check.class(token.args, "list")
+
+#         #Creating the table results
+#         results_out <- summary.seq.test(data, quantiles, cent.tend, recall, digits = 10, results = "coefficients", match_call = list(cent.tend = NULL))
+
+#         #Checking if distribution
+#         is_distribution <- ifelse(length(data$models[[1]]) == 1, FALSE, TRUE)
+
+#         #significance sanitizing
+#         if(is_distribution == TRUE) {
+#             if(class(significance) == "character") {
+#                 if(significance != "cent.tend") {stop("significance argument must be either 'cent.tend' or a single 'numeric' value.")}
+#                 significance = 1
+#             } else {
[…]

### 2018-03-01 R/summary.dispRity.R

@@ -2,11 +2,12 @@
 #'
 #' @description Creates a summary of a \code{dispRity} object.
 #'
-#' @param data A \code{dispRity} object.
+#' @param object A \code{dispRity} object.
+#' @param ... Additional arguments to be passed to \code{\link{summary}}.
 #' @param quantiles The quantiles to display (default is \code{quantiles = c(50, 95)}; is ignored if the \code{dispRity} object is not bootstrapped).
 #' @param cent.tend A function for summarising the bootstrapped disparity values (default is \code{\link[stats]{median}}).
 #' @param recall \code{logical} value specifying whether to recall the \code{dispRity} parameters input (default = \code{FALSE}).
-#' @param rounding Optional, a value for rounding the values in the output table (default = 2).
+#' @param digits Optional, a value for digits the values in the output table (default = 2).
 # ' @param results Optional, in the case of summarising a \code{\link{sequential.test}} which results to display (default = "coefficients")
 #'
 #' @return
@@ -23,7 +24,7 @@
 #' ## Summarising the results
 #' summary(disparity) # default
 #' ## Using different options
-#' summary(disparity, quantiles = 75, cent.tend = mean, rounding = 8,
+#' summary(disparity, quantiles = 75, cent.tend = mean, digits = 8,
 #'      recall = TRUE)
 #' 
 #' @seealso \code{\link{dispRity}}, \code{\link{plot.dispRity}}.
@@ -48,7 +49,8 @@
 # recall <- FALSE
 # match_call <- list() ; match_call$cent.tend <- "median"
 
-summary.dispRity <- function(data, quantiles = c(50, 95), cent.tend = median, recall = FALSE, rounding){#, results = "coefficients") {
+summary.dispRity <- function(object, ..., quantiles = c(50, 95), cent.tend = median, recall = FALSE, digits){#, results = "coefficients") {
+    data <- object
 
     #----------------------
     # SANITIZING
@@ -71,19 +73,19 @@ summary.dispRity <- function(data, quantiles = c(50, 95), cent.tend = median, re
     #recall
     check.class(recall, "logical")
 
-    #rounding
-    if(missing(rounding)) {
+    #digits
+    if(missing(digits)) {
         #Set to default (see below)
-        rounding <- "default"
+        digits <- "default"
     } else {
-        check.class(rounding, "numeric")
+        check.class(digits, "numeric")
     }
 
     #DATA
     #must be class dispRity
[…]

### 2018-03-01 tests/testthat/test-boot.matrix.R

@@ -254,7 +254,7 @@ test_that("verbose bootstrap works", {
         c("Bootstrapping", ".", "Done."))
 
     ## Verbose works with single elements subsets
-    data1 <- time.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous", time = c(139, 60), model = "gradual.split", inc.nodes = TRUE, BeckLee_ages, verbose = FALSE, t0 = FALSE)
+    data1 <- chrono.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous", time = c(139, 60), model = "gradual.split", inc.nodes = TRUE, BeckLee_ages, verbose = FALSE, t0 = FALSE)
 
     data1$subsets$`139`$elements <- matrix(data1$subsets$`139`$elements[-1,], nrow = 1)
 
@@ -281,7 +281,7 @@ test_that("Boot.matrix works with small, empty/subsets", {
     data <- test_data$ord_data_tips_nodes
     FADLAD <- test_data$FADLAD_data
 
-    silent <- capture_warnings(data <- time.subsets(data, tree, model = "deltran", method = "continuous", time = c(140, 138, 130, 120, 100)))
+    silent <- capture_warnings(data <- chrono.subsets(data, tree, model = "deltran", method = "continuous", time = c(140, 138, 130, 120, 100)))
 
     warnings <- capture_warnings(test <- boot.matrix(data))
     expect_equal(warnings, "The following subsets have less than 3 elements: 140, 138, 130.\nThis might effect the bootstrap/rarefaction output.")
@@ -309,8 +309,8 @@ test_that("boot.matrix deals with probabilities subsets", {
     data(BeckLee_tree)
     
 
-    data1 <- time.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous", time = c(100, 60), model = "gradual.split", inc.nodes = TRUE, BeckLee_ages, verbose = FALSE, t0 = FALSE)
-    data2 <- time.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous", time = c(100, 60), model = "proximity", inc.nodes = TRUE, BeckLee_ages, verbose = FALSE, t0 = FALSE)
+    data1 <- chrono.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous", time = c(100, 60), model = "gradual.split", inc.nodes = TRUE, BeckLee_ages, verbose = FALSE, t0 = FALSE)
[…]

### 2018-03-01 tests/testthat/test-dispRity.R

@@ -5,7 +5,7 @@ context("dispRity")
 data(BeckLee_mat50)
 data(BeckLee_tree)
 data_boot <- boot.matrix(BeckLee_mat50, bootstraps = 11, rarefaction = c(5,6))
-data_subsets_simple <- time.subsets(BeckLee_mat50, tree = BeckLee_tree,  method = "discrete", time = c(120,80,40,20))
+data_subsets_simple <- chrono.subsets(BeckLee_mat50, tree = BeckLee_tree,  method = "discrete", time = c(120,80,40,20))
 data_subsets_boot <- boot.matrix(data_subsets_simple, bootstraps = 11, rarefaction = c(5,6))
 metric = c(sum, variances)
 verbose = TRUE
@@ -388,7 +388,7 @@ test_that("Example works", {
     bootstrapped_data <- boot.matrix(customised_subsets, bootstraps=100)
     sum_of_ranges <- dispRity(bootstrapped_data, metric=c(sum, ranges))
 
-    ex3 <- summary(sum_of_ranges, rounding = 2)
+    ex3 <- summary(sum_of_ranges, digits = 2)
 
     expect_is(
         ex3
@@ -420,7 +420,7 @@ test_that("dispRity works with small, empty/subsets", {
     data <- test_data$ord_data_tips_nodes
     FADLAD <- test_data$FADLAD_data
 
-    silent <- capture_warnings(data <- time.subsets(data, tree, model = "deltran", method = "continuous", time = c(140, 138, 130, 120, 100)))
+    silent <- capture_warnings(data <- chrono.subsets(data, tree, model = "deltran", method = "continuous", time = c(140, 138, 130, 120, 100)))
     silent <- capture_warnings(data <- boot.matrix(data))
 
     warnings <- capture_warnings(test <- dispRity(data, metric = c(sum, variances), verbose = TRUE))
@@ -440,8 +440,8 @@ test_that("dispRity deals with probabilities subsets", {
     data(BeckLee_tree)
     
 
-    data1 <- time.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous", time = c(100, 60), model = "gradual.split", inc.nodes = TRUE, BeckLee_ages, verbose = FALSE, t0 = FALSE)
[…]

### 2018-03-01 tests/testthat/test-model.test.R

@@ -1,0 +1,66 @@
+# #TESTING null.test
+
+# context("model.test")
+
+# ## Select model data
+# load("model_test_data.Rda")
+# data <- model_test_data
+
+# test_that("simple models work", {
+#     set.seed(1)
+#     test <- model.test(data, model = "BM", pool.variance = NULL, time.split = NULL, fixed.optima = FALSE, control.list = list(fnscale = -1), verbose = FALSE)
+
+#     expect_equal(class(test), c("dispRity", "model.test"))
+#     expect_equal(names(test), c("aic.models", "full.details", "call"))
+#     expect_is(test[[1]], c("matrix"))
+#     expect_is(test[[2]], c("list"))
+#     expect_equal(round(test[[2]][[1]]$value, digit = 4), 10.0892)
+
+#     set.seed(1)
+#     test <- model.test(data, model = "Stasis", pool.variance = NULL, time.split = NULL, fixed.optima = FALSE, control.list = list(fnscale = -1), verbose = FALSE)
+
+#     expect_equal(class(test), c("dispRity", "model.test"))
+#     expect_equal(names(test), c("aic.models", "full.details", "call"))
+#     expect_is(test[[1]], c("matrix"))
+#     expect_is(test[[2]], c("list"))
+#     expect_equal(round(test[[2]][[1]]$value, digit = 4), round(-15.11164, digit = 4))
+
+#     set.seed(1)
+#     test <- model.test(data, model = "Trend", pool.variance = NULL, time.split = NULL, fixed.optima = FALSE, control.list = list(fnscale = -1), verbose = FALSE)
+
+#     expect_equal(class(test), c("dispRity", "model.test"))
+#     expect_equal(names(test), c("aic.models", "full.details", "call"))
+#     expect_is(test[[1]], c("matrix"))
+#     expect_is(test[[2]], c("list"))
+#     expect_equal(round(test[[2]][[1]]$value, digit = 4), round(12.77702, digit = 4))
+
+#     set.seed(1)
+#     test <- model.test(data, model = "OU", pool.variance = NULL, time.split = NULL, fixed.optima = FALSE, control.list = list(fnscale = -1), verbose = FALSE)
+
+#     expect_equal(class(test), c("dispRity", "model.test"))
+#     expect_equal(names(test), c("aic.models", "full.details", "call"))
+#     expect_is(test[[1]], c("matrix"))
[…]

### 2018-03-02 paper/dispRity_paper.tex
acknowledgements up

@@
 Many thanks to Natalie Cooper for encouraging the development of this package and helping with the writing of this paper and the package manuals.
-Thanks to David Bapst, Martin Brazeau, Andrew Jackson, Graeme Lloyd and Emma Sherratt for specific implementation ideas and to Michael Collyer, Gavin Simpson and two other anonymous reviewers for constructive and encouraging comments on the manuscript.
+Thanks to David Bapst, Rompy Chompee, Martin Brazeau, Andrew Jackson, Graeme Lloyd and Emma Sherratt for specific implementation ideas and to Michael Collyer, Gavin Simpson and two other anonymous reviewers for constructive and encouraging comments on the manuscript.
 I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 - 2013)/ERC Grant Agreement number 311092 awarded to Martin D.

### 2018-03-15 paper/response_to_reviewers2.tex
updated paper

@@
 \bigskip
-Dear Dr Vamosi,\\
+Dear Editor,\\
 \bigskip
 
-I am very grateful to the four referees for their helpful and constructive comments, which I believe have helped me to significantly improve the \texttt{dispRity} package and this associated package description paper.
-I have taken all of their comments on board (though some will be implemented in later versions), and respond to their points in details below.
-For improving clarity in this document, reviewers' comments are displayed in blue, my responses in normal text and the changes to the manuscript in italic.
-Additionally, I've uploaded two versions of the revised manuscript: both have the exact same content but one has all the changes and additions to the main text highlighted in yellow to help with reviewing.
-All the changes implemented in the package and the vignette are available on the release branch under version 0.5.1.
-
+I have taken all the reviewers comments on board (except the CRAN release - see below) and updated the manuscript and the package accordingly (the new released version is now 1.0.2).
+Please find my detailed response below.
+To save some words, I have also cut down some text in the section describing the time-slice models and directly refer to the new publication detailing the models (Guillerme and Cooper 2018, Palaeontology 10.1111/pala.12364 - in the ``Splitting the multidimensional space into subsets'' section in this manuscript).
 % ~~~~~~~~~~~~~~~~~~~~~~~~
@@
 
-I havefixed this typo.
+I have fixed this typo.
 
@@
 
-I am not entirely sure what this reviewer refers to.
+I am not sure what this reviewer refers to.
 I have double checked punctuation in this paragraph.
@@
 
-\item{\textcolor{blue}{[pushing the package on the CRAN]}}
+\item{\textcolor{blue}{To not publish this package on CRAN would be a large mistake, in my opinion.}}
+
+I do understand the reviewer's concern and believe that having the package on the CRAN would allow greater visibility for this package.
+However, I do not wish to submit the package to the CRAN.
+In fact, one main draw back for packages on CRAN is that they do not allow to easily add/fix features regularly (the waiting period between two submissions should be at least 2 month).
+Furthermore, many features in the \texttt{dispRity} package cannot pass the CRAN checks, including the unit testing, the gitbook vignette or the use of functions implemented in other packages (or packages versions) not released on the CRAN.
+Therefore, a CRAN version of the \texttt{dispRity} package would contain less features than the one currently on GitHub.
+
+\item{\textcolor{blue}{First, CRAN's archival system makes it possible to go back in time and install previous versions of package.
+This is extremely useful, even if a reason to do that does not seem obvious.}}
+
+The \texttt{dispRity} releases are archived regularly on ZENODO (since the last three years - \url{https://zenodo.org/search?page=1&size=20&q=conceptrecid:%22593021%22&sort=-version&all_versions=True}).
+Each major release thus has an attached and citeable DOI and can be downloaded as ``time capsules'' (i.e. the version of the package as it was at the time of release).
+
+\item{\textcolor{blue}{Second, CRAN checks all dependencies and reverse dependencies on package updates, and makes sure packages are compliant with R updates.
+This is important if this package is supposed to work with other packages, which is rather true in this case.
+Finally (but perhaps not comprehensively), publishing packages on CRAN means performing CRAN checks that are much more involved than the excellent tool offered by devtools, guaranteeing compatibility with multiple operating systems.}}
+
+The \texttt{dispRity} package is constantly tested through Travis (\url{https://travis-ci.org/TGuillerme/dispRity/branches}).
+These tests include the normal CRAN checks (dependencies, compilation on different OS and different architectures) but also a thorough unit test of the functions themselves (currently at 91\% and I am aiming for a full coverage in the future - \url{https://codecov.io/gh/TGuillerme/dispRity}).
+
+Additionally, the package meets MEE policies on publishing code (\url{(http://besjournals.onlinelibrary.wiley.com/hub/journal/10.1111/(ISSN)2041-210X/journal-resources/policy-on-publishing-code.html}).
 
@@
 However, due to the table size, it doesn't fit in the page and was thus written separately.
-
-@@@ Force it to be under the table
+I have now forced the table to appear on the page before the caption.
 
@@
 If that is the case, maybe some careful wording - more than just citing the original sources - is needed to not portray all of these as features of the new package.
-One suggestion is to offer, e.g., convhulln$vol rather than convhull.volume as the function that calculates hull volume, in Table 2, and indicate that convhull.volume similarly returns the $vol part of a convhull object, so one does not have to sift through a list.
+One suggestion is to offer, e.g., convhulln\$vol rather than convhull.volume as the function that calculates hull volume, in Table 2, and indicate that convhull.volume similarly returns the \$vol part of a convhull object, so one does not have to sift through a list.
 (The idea is to do this with all cloned functions.) 
@@
 I have now specifically made clear that \texttt{convhull.surface}, \texttt{convhull.volume} and \texttt{pairwise.dist} are just duplicates of their associated original functions, both in the manuscript, package and vignettes.
+I will develop more C implementations for the package metrics and core functions in the future.
 
-\item{\textcolor{blue}{Finally, there were some English irregularities (e.g., data is the plural form of datum) that could be fixed with some proofing from a native English-speaking scientist..}}
+\item{\textcolor{blue}{Finally, there were some English irregularities (e.g., data is the plural form of datum) that could be fixed with some proofing from a native English-speaking scientist.}}
 

### 2018-03-16 inst/gitbook/03_specific-tutorials.Rmd
v1.0.0
# Conflicts:
#	NEWS.md
#	R/boot.matrix.R
#	R/dispRity.R
#	inst/gitbook/_book/details-of-specific-functions.html
#	inst/gitbook/_book/dispRity_manual.pdf
#	inst/gitbook/_book/dispRity_manual.tex
#	inst/gitbook/_book/ecology-demo.html
#	inst/gitbook/_book/future-directions.html
#	inst/gitbook/_book/getting-started-with-disprity.html
#	inst/gitbook/_book/glossary.html
#	inst/gitbook/_book/index.html
#	inst/gitbook/_book/making-stuff-up.html
#
inst/gitbook/_book/palaeobiology-demo-disparity-through-time-and-within-
groups.html
#	inst/gitbook/_book/references.html
#	inst/gitbook/_book/search

@@
 
-For the time-slicing method details see Cooper and Guillerme (in prep.). <!-- @@@ Change cite appropriately! -->
+For the time-slicing method details see @time-slice. <!-- @@@ Change cite appropriately! -->
 <!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
@@
 
-Note that the `test.dispRity` algorithm deals with some classical test outputs and summarises the test output.
+Note that the `test.dispRity` algorithm deals with some classical test outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
 It is, however, possible to get the full detailed output by using the options `details = TRUE`.

### 2018-03-16 paper/dispRity_paper.tex

--- 2018-03-01 paper/dispRity_paper.tex
@@
 \usepackage{textcomp}
+\usepackage{float}
+\restylefloat{table}
 \newcolumntype{d}[1]{D{.}{.}{#1}}
@@
 
-\noindent {\normalsize \sc Thomas Guillerme$^12$$^,$$^*$}\\
+\noindent {\normalsize \sc Thomas Guillerme$^{1,2,*}$}\\
 \noindent {\small \it 
-$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\
-$^1$School of Biological Sciences, University of Queensland, St. Lucia, Australia.\\}
+$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\\
+$^2$School of Biological Sciences, University of Queensland, St. Lucia, Australia.\\}
 
@@
         Many of these methods are implemented in specific software or R packages.
-        However, the definition of the space, metric and test is often dependent on the software/package's author's point of view or specific question.
+        However, the definition of the space, metric and test is often dependent on the software/package authors points of view or specific questions.
         This can unfortunately hamper researchers' ability to apply different methods that best suits their specific questions.
 
-        \item Here I present the \disp package, a flexible R package for running multidimensional analysis.
-        This package allows users to define each step of the analysis (whether it is the space, the metric or the test) through a highly modular architecture where each definition can be passed as a function.
+        \item Here I present the \disp package, a flexible \texttt{R} package for performing multidimensional analysis.
+        It allows users to define each step of the analysis (whether it is the space, the metric or the test) through a highly modular architecture where each definition can be passed as a function.
         It also provides a tidy interface through the \disp object, allowing users to easily run reproducible multivariate analysis.
@@
         Furthermore, the package contains some simulation tools (e.g to simulate complex multidimensional space or morphological data).
-        Finally, it also contains a suite of utilities to work with \disp objects aimed at helping users to develop their own multidimensional metrics and/or tests.
+        Finally, it also contains a suite of utility functions to work with \disp objects aimed at helping users to develop their own multidimensional metrics and/or tests.
 
@@
 Such analyses aim to capture the complex multidimensionality of biological data, while still providing outputs that are interpretable.
-These multivariate analyses can be used to investigate changes in morphological diversity through time \citep[e.g][]{Close2015}, competitive replacement scenarios \citep[e.g.][]{Brusatte12092008}, relationships among form and function \citep[e.g.][]{diaz2016global} and even to describe the entirety of possible shapes for a group of organisms \citep[e.g.][]{raup1966geometric}.
+These multivariate analyses can be used to investigate changes in morphological diversity through time \citep[e.g.][]{Close2015}, competitive replacement scenarios \citep[e.g.][]{Brusatte12092008}, relationships among form and function \citep[e.g.][]{diaz2016global} and even to describe the entirety of possible shapes for a group of organisms \citep[e.g.][]{raup1966geometric}.
 The biological variables in such analyses are equally diverse, including morphological traits (discrete traits like the presence or absence of a character, e.g.
-\citealt{Close2015}; and continuous traits such as lengths, e.g.
+\citealt{Close2015}; or continuous traits such as lengths, e.g.
 \citealt{price2015predation}), life history traits \citep[e.g.][]{diaz2016global}, or even ecosystem properties \citep[e.g.][]{DonohueDim}.
@@
 This space is represented as a matrix where rows are regarded as samples or observations (e.g.
-a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g.
-an embedding, scaling, ordination, etc.
-of the variables).
+specimens, field sites, etc.) and columns are variables or some transformation thereof (e.g.
+embedding, scaling, ordination, etc.).
 These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix (\citealt{lloyd2016estimating} and references therein; e.g.
@@
 
-Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance (
-NPMANOVA, \citealt{NPMANOVA}; e.g.
-in \citealt{Brusatte12092008}) multidimensional permutation tests ( \citealt{ManlyPermutations}; e.g.
+Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance (NPMANOVA, \citealt{NPMANOVA}; e.g.
+in \citealt{Brusatte12092008}) multidimensional permutation tests (\citealt{ManlyPermutations}; e.g.
 in \citealt{diaz2016global}) or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements.
@@
 This is ideal for testing volume based hypothesis (e.g.
-does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g.
-does groups A and B occupy the same location?).
+``does groups A and B have the same volume?''), but in other cases may not be appropriate in non-volume based hypothesis (e.g.
+``do they occupy the same location?'').
 This can lead to inappropriate analyses by users confined by the existing software implementations.
 
-The aim of the \disp package is avoid such problems by providing a completely flexible framework for studying multidimensional data.
-This package is based on a highly modular architecture where each crucial decision in multidimensional analysis (which data, which metric and which test) can be provided by the user.
-It implements many commonly used metrics and definitions of disparity, as well as providing a simple interface for users to implement their own disparity functions.
-The package is described here for the use of discrete morphological data disparity analysis but can be generalised to any type of multidimensional data
-(see the glossary Table \ref{Tab:terms} for terms' equivalences between palaeobiology and ecology).
+The aim of the \disp package is avoid such problems by providing a flexible framework for studying multidimensional data.
+This package is based on a modular architecture where each decision in multidimensional analysis (which data, metric and test) can be specified by the user.
+It implements many commonly used disparity metrics, as well as providing a simple interface for users to implement their own disparity functions.
+The package is described here for the use of discrete morphological data disparity analysis but can be generalised to any type of multidimensional data (see the glossary Table \ref{Tab:terms}).
 
@@
         disparity & a dimension-level 1 or 2 function$^1$ & disparity: e.g.
-the sum of variances \cite{Wills2001}, the average pairwise distances between taxa \cite{Foote01071994}, etc.
+the sum of variances \citep{Wills2001}, the average pairwise distances between taxa \citep{Foote01071994}, etc.
 & dissimilarity: e.g.
-ellipsoid volume \cite{DonohueDim}, convex hull volume \cite{cornwell2006trait}, etc.
+ellipsoid volume \citep{DonohueDim}, convex hull volume \citep{cornwell2006trait}, etc.
 \\
@@
 \section{Description}
-In brief, the package takes a \texttt{matrix} object (the multidimensional space), 
-calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation.
+In brief, the package takes a \texttt{matrix} object (the multidimensional space), calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation.
 Some additional functions modify the space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow}).
 Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements \citep[e.g. the space can be a distance matrix][]{Close2015}.
-The matrix is always considered as the multidimensional space and no correction is applied to it (e.g.
-if ordinated and with negative eigen values, corrections should be applied prior to using the \texttt{dispRity} package.)
+The matrix is always considered as the final multidimensional space to analyse and no correction is applied to it (e.g.
+potential corrections should be applied prior to using the \texttt{dispRity} package).
 
@@
 \subsection{Measuring disparity}
-The \disp function measures disparity of any multidimensional space (of class \texttt{matrix} or \texttt{data.frame}) where the columns correspond to the dimensions and the rows correspond to the elements present in the space.
-The disparity metric (i.e.
-the aspect of interest of the space) is passed through the \texttt{metric} argument and is defined by the user as any function or combination of functions that can either transform the matrix into:
+The \disp function measures disparity from a \texttt{matrix} where the columns correspond to the dimensions and the rows correspond to the elements present in the space.
+The disparity metric is passed through the \texttt{metric} argument and is defined by the user as one or more function(s) that can either transform the matrix into:
 
@@
 \centering
-   \includegraphics[width=0.8\textwidth]{../inst/gitbook/dispRity_fun.pdf} 
+   \includegraphics[width=0.8\textwidth]{dispRity_fun.pdf} 
 \caption{Illustration of the different metric dimension-levels in the \disp package.
@@
 a 8 $\times$ 7 matrix or a vector of eight elements).
-A dimension-level 3 matrix would be a metric resulting in a matrix (e.g.
-the function \texttt{stats::cor} to calculate the correlation between each dimension), a dimension-level 2 metric would result in a vector (i.e.
+A dimension-level 3 matrix would be a metric output a matrix (e.g.
+the function \texttt{stats::cor} to calculate the correlation between each dimension), a dimension-level 2 metric would output a vector (i.e.
 a distribution, e.g.
-\texttt{dispRity::variances} which calculates the variance within each dimension) and a dimension-level 1 metric would result in a single value (e.g.
+\texttt{dispRity::variances} which calculates the variance within each dimension) and a dimension-level 1 metric would output a single value (e.g.
 \texttt{stats::sd} which calculates the standard deviation of the input matrix)}.
@@
 
-The disparity metrics can be functions from other packages, defined by the user or from the \disp package (see Table \ref{Tab:metrics}).
+The disparity metrics can be any \texttt{R} function (see Table 2 for metrics implemented in the package).
 When multiple functions are passed to the \texttt{metric} argument, they are sorted by dimension-level and applied in decreasing order to the data.
-For example, if the metric is defined as \texttt{metric = c(prod, ranges)} (the hypercube volume), the \texttt{ranges} function (dimension-level 2) is first applied to data (the range of each dimensions) and the function \texttt{prod} is then applied to the results (\texttt{prod(ranges(data))}).
+For example, if the metric is defined as \texttt{metric = c(prod, ranges)} (the hypercube volume), the \texttt{ranges} function (dimension-level 2) is first applied to data and the function \texttt{prod} is then applied to the results (\texttt{prod(ranges(data))}).
 One can also directly pass a function description to the \texttt{metric} argument (e.g. \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} for the average squared pairwise distance).
@@
 
-\begin{table}
+\begin{table}[H]
 \resizebox{\textwidth}{!}
@@
 }%
-    \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g.
+    \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is a function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g.
 it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}$ - for a hyper-ellipsoid).
@@
 
-\textbf{Table 1 (caption):} Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g.
+\textbf{Table 2 (caption):} Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g.
 it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid).
@@
 Prior to calculating disparity, the space can be subdivided into subsets, typically to be compared to each other.
-For example, one may be interested in seeing how the disparity of a specific subset of the space compares to another (e.g.
-difference among groups) or, how different subsets change sequentially (e.g.
-through time).
-In essence, the original space corresponds to the overall space (e.g. a morphospace contains all the observed morphologies).
-Conversely, subsets correspond to parts of the space with pooled characteristics.
+For example, one may compare the disparity of a specific subset of the space to another or, how different subsets change sequentially (e.g. through time).
+The original space corresponds to the overall space (e.g. a morphospace contains all the observed morphologies).
+Subsets correspond to parts of the space with pooled characteristics.
 
 This splitting can be done using the \texttt{custom.subsets} or \texttt{chrono.subsets} functions.
-The first argument is the matrix defining the space and a list of elements to group into different subsets.
+The first function takes a matrix defining the space and a list of elements defining the subsets.
 The second also takes a matrix and arguments giving the age of the taxa (a dated phylogeny of the elements present in the morphospace - see below) and which subsets to create: (1) discrete time subsets (or time-binning) or (2) continuous time subsets (or time-slicing).
@@
 This method thus requires the nodes to be part of the space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
-When a slice occurs not on a tip or a node, six methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice.
-For each method, the selected data is:
-
-\begin{itemize}
-    \item ``acctran'': always the descendant
-    \item ``deltran'': always the ancestor
-    \item ``random'': randomly chosen between the descendant or the ancestor
-    \item ``proximity'': either the descendant or the ancestor depending on branch length
-    \item ``equal.split'': both the descendant and the ancestor with an even probability
-    \item ``gradual.split'': both the descendant and the ancestor with a probability depending on branch length
-\end{itemize}
-
-Note that the four first models are a proxy for punctuated evolution and the two last ones for gradual evolution.
-There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.
-See Guillerme \& Cooper \textit{in review} for full description.
+When a slice occurs not on a tip or a node, six methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
+``acctran'', ``deltran'', ``random'' and ``proximity'' as proxy for punctuated evolution models; and ``equal.split'', ``gradual.split'' as a proxy for gradual evolution.
+See \cite{time-slice} for full description of the method.
+Note that there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.
 
@@
 
-Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms: (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data); and (2) the ``single'' algorithm where $n = 1$ (somewhat akin to jackknife resampling).
+Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms: (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data); and (2) the ``single'' algorithm where $n = 1$ (similar to jackknife).
 
@@
 The \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow users to set which central tendency and which quantiles should be represented.
-Additionally, the \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
+The \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
 Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e.
@@
 The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsets: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subset to all the others; (3) ``sequential'' for comparing subsets sequentially (e.g.
-first against second, second against third, etc.); (4) `all'' for comparing all the subsets simultaneously (i.e.
-\texttt{disparity $\mathtt{\sim}$ subsets}) or (5) any list of pairs (or more) of subsets to compare.
+first against second, second against third, etc.); (4) ``all'' for comparing all the subsets simultaneously (i.e.
+\texttt{disparity $\mathtt{\sim}$ subsets}) or (5) any list of pairs of subsets to compare.
 
-Some tests are implemented within the package such as the Bhattacharrya Coefficient \citep[\texttt{bhatt.coeff};][]{Bhattacharyya,GuillermeCooper}, a permutation test based on null hypothesised multidimensional space following \cite{diaz2016global} (\texttt{null.test})
-as well as a wrapper for the \texttt{vegan::adonis} \citep{oksanen2007vegan} and \texttt{geiger::dtt} \citep{geiger2008} functions (respectively \texttt{adonis.dispRity} and \texttt{dtt.dispRity} - allowing \texttt{dispRity} objects to be directly used with any disparity metric).
-Additionally, this function also allows additional arguments such as \texttt{rarefaction} (as described above) or \texttt{correction} to adjust p-values when using multiple parametric tests.
+Some tests are implemented within the package such as the Bhattacharrya Coefficient \citep[\texttt{bhatt.coeff};][]{Bhattacharyya,GuillermeCooper}, a permutation test based on null hypothesised multidimensional space following \cite{ManlyPermutations,diaz2016global} (\texttt{null.test})
+as well as a wrapper for the \texttt{vegan::adonis} \citep{oksanen2007vegan} and \texttt{geiger::dtt} \citep{geiger2008} functions (respectively \texttt{adonis.dispRity} and \texttt{dtt.dispRity}).
+This function also allows additional arguments such as \texttt{rarefaction} (as described above) or \texttt{correction} to adjust p-values when using multiple parametric tests.
 
@@
     \item \texttt{BeckLee\_ages}: the first and last occurrence data for 14 of the mammal species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
-    \item \texttt{disparity}: a pre-made \disp object based on the data above.
+    \item \texttt{disparity}: a pre-analysed \disp object based on the data above.
 \end{itemize}
@@
 For example, using the data described above, we can test whether the crown mammals are more diverse in term of morphology than the stem ones.
-In other words, whether the approximation of the volume within the morphospace (the disparity as defined above) is different in crown or stem mammals.
-
-\noindent These two groups can be defined using one the package's utility functions, \texttt{crown.stem} that separate the crown and stem species given a phylogeny (allowing to ignore the nodes or not):
+In other words, whether the approximation of the volume within the morphospace is different in crown or stem mammals.
+These two groups can be defined using one the package's utility functions, \texttt{crown.stem} that separate the crown and stem species given a phylogeny (allowing to ignore the nodes or not):
 
@@
 \bigskip
-As indicated, the \disp object contains two customised subsets (called ``crown'' and ``stem'') from a morphospace made of 50 elements for 48 dimensions.
+As indicated, the \disp object contains two customised subsets from a morphospace made of 50 elements for 48 dimensions.
 The \disp object also displays information on the number and method of the bootstrap replicates as well as the definition of disparity.
@@
 \caption{Summarising a \disp object (disparity per groups).
-$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 97.5
-\% are the confidence intervals.}
+$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of variances) and the 2.5, 25, 75 and 97.5\% are the confidence intervals.}
 \label{Tab:summary_group}
@@
 
-As we can see from the summary table (Table \ref{Tab:summary_group}) and the plot (\ref{Fig:plot_group}), there seems to be a significant difference in morphospace volume occupied between the two groups.
+As we can see from the summary table (Table \ref{Tab:summary_group}) and the plot (Fig.
+\ref{Fig:plot_group}), there seems to be a significant difference in morphospace volume occupied between the two groups.
 It is possible to test this hypothesis by using, for example, a non-parametric Wilcoxon test (\texttt{stats::wilcox.test}):
@@
 It is possible to compare each bootstrap in a pairwise way without pooling the data by using the \texttt{concatenate = FALSE} argument.
-The results will then be a distribution of statistic and \textit{p} values.
+The results will then be a distribution of statistics and \textit{p} values.
 Relating back to our question: yes, crown mammals display a higher diversity in morphologies than their stem counterparts (in this example and dataset).
@@
 \noindent The arguments \texttt{data = BeckLee\_mat50} and \texttt{metric = c(sum, variances)} are the same as in the example above.
-However, in this type of analysis, we also need to have additional arguments: the \texttt{time = time\_bins} indicates the boundaries of the different time bins, the \texttt{tree = BeckLee\_tree} argument provides information on the age of each element and \texttt{method = 'discrete'} indicates that the data is time-binned.
-
-Again, the resulting \disp object can be summarised and plotted (Table \ref{Tab:summary_time} and Fig. \ref{Fig:plot_time}):
+However, in this type of analysis, we also need to have additional arguments: the \texttt{time = time\_bins} indicates the boundaries of the different time bins, the \texttt{tree = BeckLee\_tree} argument provides information on the age of each element and \texttt{method = "discrete"} indicates that the data is time-binned.
+The resulting \disp object can be summarised and plotted (Table \ref{Tab:summary_time} and Fig. \ref{Fig:plot_time}):
 
@@
 \caption{Summarising a \disp object (disparity through time).
-$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 97.5\% are the confidence intervals.}
+$n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of variances) and the 2.5, 25, 75 and 97.5\% are the confidence intervals.}
 \label{Tab:summary_time}
@@
 
-Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, multivariate permutation ANOVA  \citep[PERMANOVA; ][]{permanova} through the \texttt{adonis.disprity} function that is a wrapper of the \texttt{vegan::adonis} function  \citep{oksanen2007vegan} for \texttt{dispRity} objects:
+Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, multivariate permutation ANOVA  \citep[PERMANOVA; ][]{NPMANOVA} through the \texttt{adonis.dispRity} function that is a wrapper of the \texttt{vegan::adonis} function  \citep{oksanen2007vegan} for \texttt{dispRity} objects:
 
@@
 
-To answer our specific question above: yes, there is an effect of time on morphological disparity (an increase) in this dataset.
+To answer our specific question above: yes, there is an effect of time on morphological disparity (an increase) in this dataset (Table \ref{Tab:anova}).
 Note that in this case, the function outputs different warnings on the usage of such test and the eventual data not used in the test.
@@
 \section{Conclusion}
-The \disp package is based on a highly modular architecture allowing researchers to simply define both their multidimensional space and their disparity metric to efficiently analyse multivariate data.
-The \disp object allows users to easily pipeline disparity analysis from the data input (the matrix) to publication standard results (tables, plots, hypothesis testing).
+The \disp package is based on a modular architecture allowing researchers to simply define both their multidimensional space and their disparity metric to efficiently analyse multivariate data.
+The \disp object allows users to pipeline disparity analysis from the data input (the matrix) to publication standard results (tables, plots, hypothesis testing).
+
+\section{Package location}
+The \disp package is available at \url{https://github.com/TGuillerme/dispRity}.
+The latest released versions can be directly loaded in \texttt{R} using the following command:
+
+\noindent \texttt{> if(!require(devtools)) install.packages("devtools")}\\
+\noindent \texttt{> install\_github("TGuillerme/dispRity", ref = "release")}
 
 \section{Acknowledgments}
-Many thanks to Natalie Cooper for encouraging the development of this package and helping with the writing of this paper and the package manuals.
-Thanks to David Bapst, Martin Brazeau, Andrew Jackson, Graeme Lloyd and Emma Sherratt for specific implementation ideas and to Michael Collyer, Gavin Simpson and two other anonymous reviewers for constructive and encouraging comments on the manuscript.
-I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 - 2013)/ERC Grant Agreement number 311092 awarded to Martin D.
-Brazeau and from the Discovery Project Grant DP170103227 awarded to Vera Weisbecker.
+Many thanks to Natalie Cooper for encouraging and helping with the writing of this paper and the package manuals.
+Thanks to David Bapst, Martin Brazeau, Rompy Chompee, Andrew Jackson, Graeme Lloyd and Emma Sherratt and to Michael Collyer, Gavin Simpson and two other anonymous reviewers for comments on the package and manuscript.
+I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 - 2013)/ERC Grant Agreement number 311092 awarded to Martin Brazeau and from the Discovery Project Grant DP170103227 awarded to Vera Weisbecker.
 

### 2018-03-19 inst/gitbook/02_getting-started.Rmd
Matt's suggestion for gitbook

--- 2018-03-01 inst/gitbook/02_getting-started.Rmd
@@
 
-## Performing a simple dispRity analysis
+## Performing a simple dispRity analysis {#simpleanalysis}
 
@@
 ```
+<!-- 
+## Minimalist example
+
+This is the minimal and simplest analysis you can perform in `dispRity`, regardless of the question, data, metric and test.
+I strongly suggest to read [Simple Analysis section](#simpleanalysis) or the [specific tutorials](#specific-tutorial) for more info.
+
+ 1. Collect data prior to using the `dispRity` package
+
+```{r}
+## Generating a random space by combining two matrices
+## from two different random normal distributions
+distribution1 <- matrix(rnorm(25, mean = 1), 5, 5, dimnames = list(letters[1:5] ))
+distribution2 <- matrix(rnorm(25, mean = 5), 5, 5, dimnames = list(letters[6:10]))
+random_space <- rbind(distribution1, distribution2)
+```
+
+ 2. Load the package
+
+```{r}
+library(dispRity)
+```
+
+ 3. Create two subsets (the two different distributions)
+
+```{r}
+## Creating different subsets for each distribution
+(two_subsets <- custom.subsets(random_space, group = list(letters[1:5], letters[6:10])))
+```
+
+ 4. Bootstrap the subsets
+
+```{r}
+## Bootstrap the data
+(bootstrapped_subsets <- boot.matrix(two_subsets))
+```
+
+ 5. Calculate disparity as the median distance from each group's element to their centroid
+
+```{r}
+## Calculate disparity
+(disparity <- dispRity(bootstrapped_subsets, metric = c(median, centroids)))
+```
+
+ 6. Summarising the results
+
+```{r, fig.width=7, fig.height=7}
+## Summarising the results
+summary(disparity)
+
+## Plotting them
+plot(disparity)
+```
+
+ 7. Testing whether the groups are different
+
+```{r}
+## Testing if the mean bootstrap values are different between groups
+test.dispRity(disparity, test = t.test)
+``` -->

### 2018-03-19 inst/gitbook/index.Rmd
changed summary to proper S3 method
Updated book
Matt's suggestion for gitbook

@@
 Multidimensional spaces can be ordinated matrices from MDS, PCA, PCO, PCoA but the package is *not* restricted to any type of matrices!
-This manual is based on the version `0.5.0`.
+This manual is based on the version `1.0.0`.
 
@@
 Note that this citation is only temporary (but can still be used!).
-A proper description of the package is currently in review in Methods in Ecology and Evolution and should be re-submitted this winter.
+A proper description of the package is currently in review in Methods in Ecology and Evolution.
 Upon acceptance (hopefully!), a proper DOI will be released for this manual and the package description.

### 2018-03-22 inst/gitbook/02_getting-started.Rmd
Minor changes to push for vignette

--- 2018-03-01 inst/gitbook/02_getting-started.Rmd
@@
 
+> Note that whether the matrix is an ordinated one or of any other type, it is always considered as the multidimensional space to analysis and no corrections are applied to it (e.g.
+if ordinated and with negative eigen values, no correction is used).
+Always make sure the input is the right multidimensional space to analyse!
+
 ## Ordinated matrices
@@
 These can be any type of ordinations (PCO, PCA, PCoA, MDS, etc.) as long as elements are the rows (taxa, countries, field experiments) and the dimensions are the columns.
-
-### Ordination matrices from `Claddis`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
-if(!require(Claddis)) install.packages("Claddis")
-```
-
-`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
-For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:
-
-```{r}
-require(Claddis)
-
-## Ordinating the example data from Claddis
-Claddis.ordination(Michaux1989) 
-```
-
-Note that several options are available, namely which type of distance should be computed.
-See more info in the function manual (`?Claddis.ordination`).
-Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.
 
@@
 
+### Ordination matrices from `Claddis` {#Claddis-ordination}
+
+> This functionality is not available on the CRAN version of the package (1.0.3)<!-- versiontagCRAN --> due to CRAN compatibility issues see [to CRAN or not to CRAN?](#noCRAN) for more information and on how to install the GitHub (more complete) version of the package (1.1)<!-- versiontagGH -->.
+
+```{r, eval = TRUE, echo = FALSE, message = FALSE}
+## Checking if Claddis is installed BUILD ONLY
+if(!require(Claddis)) install.packages("Claddis")
+```
+
+`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
+For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:
+
+```{r}
+require(Claddis)
+
+## Ordinating the example data from Claddis
+Claddis.ordination(Michaux1989) 
+```
+
+Note that several options are available, namely which type of distance should be computed.
+See more info in the function manual (`?Claddis.ordination`).
+Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.
+
 ### Other kinds of ordination matrices

### 2018-03-22 inst/gitbook/index.Rmd
Minor changes to push for vignette

@@
 
-You can install this package easily if you use the latest version of `R` (> 3.4.0) and `devtools`.
+You can install this package easily, directly from the CRAN:
+
+```{r, eval = FALSE}
+install.packages("dispRity")
+```
+
+Alternatively, for the most up to data version and some functionalities not compatible with the CRAN, you can use the package through GitHub using `devtool` (see [to CRAN or not to CRAN?](#noCRAN) for more details):
 
@@
 install_github("TGuillerme/dispRity", ref = "release")
+```
 
-## Loading the package
-library(dispRity)
-```
+Anyways to upload it, you can use
 
@@
 
-Note this uses the `release` branch (version`0.5`). <!-- @@@ -->
+Note this uses the `release` branch (1.1)<!-- versiontagGH -->.
 For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
@@
 
-## Why not CRAN?
+## To CRAN or not to CRAN? {#noCRAN}
 
-This package is not available on CRAN.
-This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
-However, the package follows the strict (and useful!) CRAN standards via Travis.
+This package can be installed via [the CRAN]() or via [GitHub](https://github.com/TGuillerme/dispRity).
+
+ * The CRAN version is a more "strict" version of the package that does not include some functionalities (such as [`Claddis.ordination`](#Claddis-ordination) or this gitbook vignette) because of compatibility issues with the CRAN requirement.
+
+ * The GitHub version is a more "reactive" version (fed by user's suggestions and comments) and can contain more functionalities (see list below).
+This version exists on a `"release"` branch which is tested as strictly as the CRAN version with Travis and codecov.
+The `"master"` branch contains always the latest "piping hot", not yet release version.
+It might (though rarely) be bugged.
 
@@
 [![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
-[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/master/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
+[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
 [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
 <!-- badges out end   -->
+
+Note however, that apart from the functions listed below, both versions of the package have the same tested and implemented functions.
+
+Functions not available on the CRAN version (1.0.3)<!-- versiontagCRAN -->:
+
+ * [`Claddis.ordination`](#Claddis-ordination)
 

### 2018-03-22 paper/dispRity_paper.tex
Minor changes to push for vignette

@@
 \centering
-   \includegraphics[width=0.8\textwidth]{dispRity_fun.pdf} 
+   \includegraphics[width=0.8\textwidth]{dispRity_fun.pdf}  %../inst/gitbook/dispRity_fun.pdf} 
 \caption{Illustration of the different metric dimension-levels in the \disp package.
@@
 \section{Package location}
-The \disp package is available at \url{https://github.com/TGuillerme/dispRity}.
-The latest released versions can be directly loaded in \texttt{R} using the following command:
-
-\noindent \texttt{> if(!require(devtools)) install.packages("devtools")}\\
-\noindent \texttt{> install\_github("TGuillerme/dispRity", ref = "release")}
+The \disp package is available on the CRAN or at \url{https://github.com/TGuillerme/dispRity} with more associated information.
 

### 2018-04-17 inst/gitbook/02_getting-started.Rmd
updated gitbook compiler with added cleaning

@@
 
-> This functionality is not available on the CRAN version of the package (1.0.3)<!-- versiontagCRAN --> due to CRAN compatibility issues see [to CRAN or not to CRAN?](#noCRAN) for more information and on how to install the GitHub (more complete) version of the package (1.1)<!-- versiontagGH -->.
+> This functionality is not available on the CRAN version of the package (`r version_CRAN`) due to CRAN compatibility issues see [to CRAN or not to CRAN?](#noCRAN) for more information and on how to install the GitHub (more complete) version of the package (`r version_release`).
 

### 2018-04-17 inst/gitbook/index.Rmd
updated gitbook compiler with added cleaning

@@
 
+```{r, eval = TRUE, echo = FALSE}
+## master version
+version_raw <- RCurl::getURL("https://raw.githubusercontent.com/TGuillerme/dispRity/master/DESCRIPTION")
+version_master <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
+
+## release version
+version_raw <- RCurl::getURL("https://raw.githubusercontent.com/TGuillerme/dispRity/release/DESCRIPTION")
+version_release <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
+
+## CRAN version
+version_raw <- RCurl::getURL("https://cran.r-project.org/web/packages/dispRity/index.html")
+version_CRAN <- strsplit(strsplit(version_raw, split = "\n<td>Version:</td>\n<td>")[[1]][2], split = "</td>\n</tr>\n<tr>\n<td>Depends:")[[1]][1]
+```
+
 # `dispRity`
@@
 Multidimensional spaces can be ordinated matrices from MDS, PCA, PCO, PCoA but the package is *not* restricted to any type of matrices!
-This manual is based on the version `1.0.0`.
+This manual is based on the version `r version_release`.
 
@@
 
-Note this uses the `release` branch (1.1)<!-- versiontagGH -->.
+Note this uses the `release` branch (`r version_release`).
 For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
@@
 
-Functions not available on the CRAN version (1.0.3)<!-- versiontagCRAN -->:
+Functions not available on the CRAN version (`r version_CRAN`):
 

### 2018-04-20 paper/dispRity_paper.tex
Updated paper and figures in eps

@@
 \usepackage{natbib}
-
-%Packages
-
-\usepackage{xcolor}
-\usepackage{color,soul}
 \usepackage{pdflscape}
-\usepackage{fixltx2e}
-\usepackage{textcomp}
 \usepackage{fullpage}
-\usepackage{float}
-\usepackage{latexsym}
 \usepackage{url}
 \usepackage{epsfig}
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{bm}
-\usepackage{array}
-\usepackage[version=3]{mhchem}
-\usepackage{ifthen}
 \usepackage{caption}
 \usepackage{hyperref}
-\usepackage{amsthm}
-\usepackage{amstext}
 \usepackage{enumerate}
-\usepackage[osf]{mathpazo}
 \usepackage{dcolumn}
 \usepackage{lineno}
-\usepackage{dcolumn}
-\usepackage{hyphenat}
 \usepackage[T1]{fontenc}
@@
 
-\begin{landscape}
 \begin{table}
@@
 \end{table}
-\end{landscape}
 
@@
 \centering
-   \includegraphics[width=1\textwidth]{workflowsvg.pdf} 
+   %\includegraphics[width=1\textwidth]{workflowsvg.eps} 
 \caption{\disp package workflow: rectangles represent matrices; ellipses represent functions; plain black arrows indicate input/output; dashed grey arrows indicate output (though the summary, plot, and test function cannot be applied if no disparity has been calculated).}
@@
 \centering
-   \includegraphics[width=0.8\textwidth]{dispRity_fun.pdf}  %../inst/gitbook/dispRity_fun.pdf} 
+   %\includegraphics[width=0.8\textwidth]{dispRity_fun.eps}  %../inst/gitbook/dispRity_fun.pdf} 
 \caption{Illustration of the different metric dimension-levels in the \disp package.
@@
 $^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO (\textit{not} PCA).
-\label{Tab:metrics}
 
@@
 \centering
-   \includegraphics[width=1\textwidth]{plot_example_group.pdf} 
+   %\includegraphics[width=1\textwidth]{plot_example_group.eps} 
 \caption{\disp plot of disparity differences between groups.}
@@
 \centering
-   \includegraphics[width=1\textwidth]{plot_example_time.pdf} 
+   %\includegraphics[width=1\textwidth]{plot_example_time.eps} 
 \caption{\disp plot of disparity-through-time.
@@
 \section{Package location}
-The \disp package is available on the CRAN or at \url{https://github.com/TGuillerme/dispRity} with more associated information.
+The \disp package is available on the CRAN at \url{https://cran.r-project.org/web/packages/dispRity/index.html} or on GitHub at \url{https://github.com/TGuillerme/dispRity} with more associated information.
+All the versions of the package are archived on ZENODO with associated DOI \url{https://zenodo.org/record/1186467#.WtfbGsi-kW8}.
 

### 2018-05-13 inst/gitbook/03_specific-tutorials.Rmd
Merge branch 'master' into model-test2
# Conflicts:
#	R/est.VCV.R
#	R/est.mean.R
#	R/input.param.R
#	R/make.dispRity.models.R
#	R/misc.model.test.fun.R
#	R/model.test.R
#	R/model.test.lik.R
#	R/model.test.sim.R
#	R/opt.mode.R
#	man/model.test.Rd
#	tests/testthat/test-model.test.R

--- 2018-03-01 inst/gitbook/03_specific-tutorials.Rmd
@@
 
+Because there is only one subset (i.e. one matrix) in the dispRity object, the operations below are the equivalent of `metric(dummy_space)` (with rounding).
+
 ```{r}
@@
 ```
-> Because there is only one subset (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with digits).
+> WARNING: in such dummy space, this gives the estimation of the ellipsoid volume, not the real ellipsoid volume! See the cautionary note in `?ellipse.volume`.
 

### 2018-05-13 inst/gitbook/06_palaeo-demo.Rmd
Merge branch 'master' into model-test2
# Conflicts:
#	R/est.VCV.R
#	R/est.mean.R
#	R/input.param.R
#	R/make.dispRity.models.R
#	R/misc.model.test.fun.R
#	R/model.test.R
#	R/model.test.lik.R
#	R/model.test.sim.R
#	R/opt.mode.R
#	man/model.test.Rd
#	tests/testthat/test-model.test.R

@@
 
-This demo aims to give quick overview of the `dispRity` package (v.0.4) for palaeobiology analyses of disparity, including disparity through time analyses.
+This demo aims to give quick overview of the `dispRity` package (v.`r version_release`) for palaeobiology analyses of disparity, including disparity through time analyses.
 

### 2018-05-13 R/Claddis.ordination.R

@@ -17,7 +17,7 @@
 #' Claddis.ordination(Claddis::Michaux1989)
 #' }
 #'
-#' @seealso \code{\link[Claddis]{MorphDistMatrix}}, \code{\link[Claddis]{ReadMorphNexus}}, \code{\link[Claddis]{MakeMorphMatrix}}, \code{\link[stats]{cmdscale}}, \code{\link{custom.subsets}}, \code{\link{time.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}.
+#' @seealso \code{\link[Claddis]{MorphDistMatrix}}, \code{\link[Claddis]{ReadMorphNexus}}, \code{\link[Claddis]{MakeMorphMatrix}}, \code{\link[stats]{cmdscale}}, \code{\link{custom.subsets}}, \code{\link{chrono.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}.
 #' 
 #' @author Thomas Guillerme
 #' @export
@@ -29,7 +29,7 @@ Claddis.ordination <- function(data, distance = "Gower", transform = "arcsine_sq
     ## Data
     error_msg <- paste0("data does not contain a matrix.\nUse Claddis::ReadMorphNexus to generate the proper data format.")
     check.class(data, "list", msg = error_msg)
-    ## Must have at least one matrix
+    ## Must have at least one matrix
     if(!any(names(data) %in% "matrix")) {
         stop(error_msg, call. = FALSE)
     }
@@ -59,7 +59,7 @@ Claddis.ordination <- function(data, distance = "Gower", transform = "arcsine_sq
     ## Transforming the Claddis data
 
     ## Compute the distance
-    distance <- MorphDistMatrix.support(data, distance = distance) #TG: Change to the proper version
+    distance <- MorphDistMatrix.support(data, distance = distance)
 
     ## Ordinate the matrix
     ordination <- stats::cmdscale(distance, k = k, add = add, ...)

### 2018-05-13 R/dispRity.utilities.R

@@ -159,7 +159,7 @@ matrix.dispRity <- function(data, subsets, rarefaction, bootstrap){
 ## DEBUG
 # source("sanitizing.R")
 # data(BeckLee_mat99) ; data(BeckLee_tree) 
-# subsets_full <- time.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
+# subsets_full <- chrono.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
 # bootstrapped_data <- boot.matrix(subsets_full, bootstraps = 10, rarefaction = c(3, 5))
 # disparity_data <- dispRity(bootstrapped_data, variances)
 # get.subsets(bootstrapped_data, subsets = "66.75552") # 1 subsets for 23 elements
@@ -216,7 +216,7 @@ get.subsets <- function(data, subsets) {
 ## DEBUG
 # source("sanitizing.R")
 # data(BeckLee_mat99) ; data(BeckLee_tree) 
-# subsets_full <- time.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
+# subsets_full <- chrono.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
 # bootstrapped_data <- boot.matrix(subsets_full, bootstraps = 10, rarefaction = c(3, 5))
 # data <- dispRity(bootstrapped_data, c(sum,variances))
 # extract.dispRity(data, observed = FALSE, rarefaction = 5,subsets = 2)
@@ -228,7 +228,7 @@ extract.dispRity <- function(data, subsets, observed = TRUE, rarefaction = FALSE
     
     ## Data
     check.class(data, "dispRity")
-    ## Data must have disparity values
+    ## Data must have disparity values
     if(is.null(data$call$disparity)) {
         stop("dispRity object does not contain disparity values.")
     }
@@ -267,7 +267,7 @@ extract.dispRity <- function(data, subsets, observed = TRUE, rarefaction = FALSE
     }
 }
 
-#' @title Scaling and centering disparity results.
+#' @title Rescaling and centering disparity results.
 #'
 #' @description Scales or/and centers the disparity measurements.
 #'
@@ -283,9 +283,9 @@ extract.dispRity <- function(data, subsets, observed = TRUE, rarefaction = FALSE
 #' 
 #' ## Scaling the data
[…]

### 2018-05-13 R/get.bin.ages.R

@@ -16,10 +16,10 @@
 #' stratigraphy <- get.bin.ages(BeckLee_tree)
 #' 
 #' ## Making stratigraphic time subsets
-#' time.subsets(BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+#' chrono.subsets(BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
 #'                 time = stratigraphy)
 #'
-#' @seealso \code{\link{time.subsets}}
+#' @seealso \code{\link{chrono.subsets}}
 #' 
 #' @author Thomas Guillerme
 #' @import geoscale timescales
@@ -79,7 +79,7 @@ get.bin.ages <- function(tree, what = "End", type = "Age", ICS = 2015) {
 
     ## Remove eventual recent strats for trees not containing living taxa
     if(node_depth < root_time) {
-        ## Correct recent if tree contains only fossils
+        ## Correct recent if tree contains only fossils
         time_ro_recent <- abs(node_depth - tree$root.time)
         recent <- which(stratigraphy$End[strats] < min(node.depth.edgelength(tree) + time_ro_recent))
     } else {

### 2018-05-13 R/plot.dispRity.R

@@ -2,8 +2,9 @@
 #'
 #' @description Plots a \code{dispRity} object.
 #'
-#' @param data A \code{dispRity} object.
-#' @param type Either \code{"continuous"} (\code{"c"}), \code{"box"} (\code{"b"}), \code{"line"} (\code{"l"}) or \code{"polygon"} (\code{"p"}). When unspecified, is set to \code{"continuous"} if \code{\link{time.subsets}} is used with \code{method = "continuous"}, else is set to \code{"box"}. See details.
+#' @param x A \code{dispRity} object.
+#' @param ... Any optional arguments to be passed to \code{\link[graphics]{plot}}.
+#' @param type Either \code{"continuous"} (\code{"c"}), \code{"box"} (\code{"b"}), \code{"line"} (\code{"l"}) or \code{"polygon"} (\code{"p"}). When unspecified, is set to \code{"continuous"} if \code{\link{chrono.subsets}} is used with \code{method = "continuous"}, else is set to \code{"box"}. See details.
 #' @param quantiles The quantiles to display (default is \code{quantiles = c(50, 95)}; is ignored if the \code{dispRity} object is not bootstrapped).
 #' @param cent.tend A function for summarising the bootstrapped disparity values (default is \code{\link[stats]{median}}).
 #' @param rarefaction Either \code{NULL} (default) or \code{FALSE} for not using the rarefaction scores; a \code{numeric} value of the level of rarefaction to plot; or \code{TRUE} for plotting the rarefaction curves.
@@ -12,7 +13,7 @@
 #' @param xlab Optional, a \code{character} string for the caption of the x axis.
 #' @param ylab Optional, one or two (if \code{elements = TRUE}) \code{character} string(s) for the caption of the y axis.
 #' @param col Optional, some \code{character} string(s) for the colour of the graph.
-#' @param time.subsets \code{logical} whether to handle continuous data from the \code{time.subsets} function as time (in Ma). When this option is set to TRUE for other \code{type} options, the names of the subsets are used for the x axis labels.
+#' @param chrono.subsets \code{logical} whether to handle continuous data from the \code{chrono.subsets} function as time (in Ma). When this option is set to TRUE for other \code{type} options, the names of the subsets are used for the x axis labels.
[…]

### 2018-05-13 R/print.dispRity.R

@@ -18,7 +18,7 @@
 #' ## Displaying the full object
 #' print.dispRity(disparity, all = TRUE)
 #'
-#' @seealso \code{\link{cust.subsets}}, \code{\link{time.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}.
+#' @seealso \code{\link{custom.subsets}}, \code{\link{chrono.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}.
 #'
 #' @author Thomas Guillerme
 
@@ -26,19 +26,19 @@
 # warning("DEBUG dispRity.R")
 # library(dispRity)
 # source("sanitizing.R")
-# source("dispRity.R")
+# source("dispRity.R")
 # source("dispRity_fun.R")
 # source("dispRity.metric.R")
 # source("make.dispRity.R")
 # source("fetch.dispRity.R")
 # source("boot.matrix.R") ; source("boot.matrix_fun.R")
-# source("time.subsets.R") ; source("time.subsets_fun.R")
-# source("cust.subsets.R") ; source("cust.subsets_fun.R")
+# source("chrono.subsets.R") ; source("chrono.subsets_fun.R")
+# source("custom.subsets.R") ; source("custom.subsets_fun.R")
 # data(BeckLee_mat50)
 # data(BeckLee_tree)
 # data_simple <- BeckLee_mat50
 # data_boot <- boot.matrix(BeckLee_mat50, bootstraps = 11, rarefaction = c(5,6))
-# data_subsets_simple <- time.subsets(BeckLee_mat50, tree = BeckLee_tree,  method = "discrete", time = c(120,80,40,20))
+# data_subsets_simple <- chrono.subsets(BeckLee_mat50, tree = BeckLee_tree,  method = "discrete", time = c(120,80,40,20))
 # data_subsets_boot <- boot.matrix(data_subsets_simple, bootstraps = 11, rarefaction = c(5,6))
 # data <- dispRity(data_subsets_boot, metric = c(variances))
 
@@ -72,21 +72,37 @@ print.dispRity <- function(x, all = FALSE, ...) {
                 }
                 return()
             } 
-            if(class(x)[2] == "model.test") {
-                cat("Disparity evolution model fitting:\n")
-                print(x$aic.models)
+            # if(class(x)[2] == "model.test") {
+            #     cat("Disparity evolution model fitting:\n")
+            #     print(x$aic.models)
+            #     return()
+            # }
+
[…]

### 2018-05-13 R/slice.tree_fun.R

@@ -3,7 +3,7 @@
 get.node.ID <- function(node, tree) {
     ## Check if it's a node
     is_node <- which(tree$node.label == node)
-    ## Return the node or the tip
+    ## Return the node or the tip
     return(ifelse(length(is_node) > 0, is_node + Ntip(tree), which(tree$tip.label == node)))
 }
 
@@ -50,7 +50,7 @@ get.branch.length <- function(node1, node2, tree, full_tree) {
             anc_subID <- get.node.ID(ancestor[1], tree)
         }
 
-        ## Get the distance between the slice and the closest ancestor (in the subtree)
+        ## Get the distance between the slice and the closest ancestor (in the subtree)
         distance1 <- tree$edge.length[which(tree$edge[,1] == anc_subID & tree$edge[,2] == node2_ID)]
 
         ## Get it's ID in the full tree
@@ -141,7 +141,7 @@ slice.tree_DELTRAN <- function(tree, tip, tree_slice) {
     while(exists("offspring_node")) {
 
         if(offspring_node == tip) {
-            ## The offspring is just the tip
+            ## The offspring is just the tip
             return(parent_node)
         }
 
@@ -158,7 +158,7 @@ slice.tree_DELTRAN <- function(tree, tip, tree_slice) {
         precision <- num.decimals(c(off_nod_age, age))
         ## If precision are different and not equal to at least 1
         #if(length(precision) > 1 && sort(precision)[1] != 1) {
-        ## Round the node ages
+        ## Round the node ages
         num_digits <- ifelse(precision[1] < precision[2], precision[1], precision[2])
         off_nod_age <- round(off_nod_age, digits = num_digits)
         age <- round(age, digits = num_digits)

### 2018-05-13 R/summary.dispRity.R

@@ -2,11 +2,12 @@
 #'
 #' @description Creates a summary of a \code{dispRity} object.
 #'
-#' @param data A \code{dispRity} object.
+#' @param object A \code{dispRity} object.
+#' @param ... Additional arguments to be passed to \code{\link{summary}}.
 #' @param quantiles The quantiles to display (default is \code{quantiles = c(50, 95)}; is ignored if the \code{dispRity} object is not bootstrapped).
 #' @param cent.tend A function for summarising the bootstrapped disparity values (default is \code{\link[stats]{median}}).
 #' @param recall \code{logical} value specifying whether to recall the \code{dispRity} parameters input (default = \code{FALSE}).
-#' @param rounding Optional, a value for rounding the values in the output table (default = 2).
+#' @param digits Optional, a value for digits the values in the output table (default = 2).
 # ' @param results Optional, in the case of summarising a \code{\link{sequential.test}} which results to display (default = "coefficients")
 #'
 #' @return
@@ -23,7 +24,7 @@
 #' ## Summarising the results
 #' summary(disparity) # default
 #' ## Using different options
-#' summary(disparity, quantiles = 75, cent.tend = mean, rounding = 8,
+#' summary(disparity, quantiles = 75, cent.tend = mean, digits = 8,
 #'      recall = TRUE)
 #' 
 #' @seealso \code{\link{dispRity}}, \code{\link{plot.dispRity}}.
@@ -48,7 +49,8 @@
 # recall <- FALSE
 # match_call <- list() ; match_call$cent.tend <- "median"
 
-summary.dispRity <- function(data, quantiles = c(50, 95), cent.tend = median, recall = FALSE, rounding){#, results = "coefficients") {
+summary.dispRity <- function(object, ..., quantiles = c(50, 95), cent.tend = median, recall = FALSE, digits){#, results = "coefficients") {
+    data <- object
 
     #----------------------
     # SANITIZING
@@ -71,12 +73,12 @@ summary.dispRity <- function(data, quantiles = c(50, 95), cent.tend = median, re
     #recall
     check.class(recall, "logical")
 
-    #rounding
-    if(missing(rounding)) {
+    #digits
+    if(missing(digits)) {
         #Set to default (see below)
-        rounding <- "default"
+        digits <- "default"
     } else {
-        check.class(rounding, "numeric")
+        check.class(digits, "numeric")
     }
 
     #DATA
[…]

### 2018-05-13 R/tree.age.R

@@ -16,7 +16,7 @@
 #' ## A random tree with the distance since the root.
 #' tree.age(rtree(10), order = 'present')
 #'
-#' @seealso \code{\link{slice.tree}}, \code{\link{time.subsets}}.
+#' @seealso \code{\link{slice.tree}}, \code{\link{chrono.subsets}}.
 #'
 #' @author Thomas Guillerme
 

### 2018-05-13 R/zzz.R

@@ -1,15 +1,4 @@
 # ## Package version checking
-# .onLoad <- function(libname = find.package("dispRity"), pkgname = "dispRity") {
-
-#     ## Check connection
-#     connection <- RCurl::url.exists("https://raw.githubusercontent.com/TGuillerme/dispRity/release/DESCRIPTION")
-
-#     ## Check version
-#     if(connection) {
-#         last_release <- as.numeric_version(strsplit(strsplit(RCurl::getURL("https://raw.githubusercontent.com/TGuillerme/dispRity/release/DESCRIPTION", followlocation = TRUE), split = "\nVersion: ")[[1]][2], split = "\nDate:")[[1]][1])
-#         current_version <- utils::packageVersion("dispRity")
-#         if(current_version < last_release) {
-#             packageStartupMessage(paste("A newer released version of dispRity (", last_release, ") is available on GitHub.\nTo get the most up to date version, you can use:\n    devtools::install_github(\"TGuillerme/dispRity\", branch = \"release\")", sep = ""))
-#         }
-#     }
+# .onAttach <- function(libname = find.package("dispRity"), pkgname = "dispRity") {
+#     packageStartupMessage(paste0("       --- dispRity package ---\nThis is the latest Github released version (1.1.1).\n"))
 # }

### 2018-05-13 inst/gitbook/compiling_gitbook.sh

@@ -1,5 +1,9 @@
 #!/bin/sh
 
+## Remove files before hand
+rm -R _book/
+rm -R _bookdown_files/
+
 ## Compile the html
 R -e 'bookdown::render_book("index.Rmd", "bookdown::gitbook")' # HTML version
 
@@ -16,9 +20,4 @@ sed 's/<!-- badges out start/<!-- badges out start -->/g' index.Rmd > index.tmp
 sed 's/badges out end   -->/<!-- badges out end   -->/g' index.tmp > index.Rmd
 rm index.tmp
 
-
-echo "#################################"
-echo "#"
-echo "# SOME FIGURES MIGHT NEED UPDATE"
-echo "#"
-echo "#################################"
+rm -R _bookdown_files/

### 2018-05-13 tests/testthat/test-char.diff.R

@@ -23,6 +23,12 @@ test_that("internal function char.diff", {
     matrix <- list(as.character(A), as.character(B))
     expect_equal(char.diff_R(matrix[[1]], matrix[[2]]), char.diff(matrix))
 
+    ## char.diff_R
+    A <- c(1,2,3,1,1,3)
+    B <- c(NA,2,3,NA,1,3)
+    matrix <- list(as.character(A), as.character(B))
+    expect_equal(char.diff_R(matrix[[1]], matrix[[2]]), char.diff(matrix))
+
 })
 
 test_that("char.diff pair", {
@@ -131,6 +137,16 @@ test_that("char.diff plot (graphic)", {
     ## Pairwise comparisons in a morphological matrix
     morpho_matrix <- matrix(sample(c(0,1), 100, replace = TRUE), 10)
 
+    ## Errors
+    expect_error(plot.char.diff("bob", type = "density"))
+    expect_warning(expect_error(plot.char.diff(morpho_matrix, type = "density", legend.pos = 1)))
+    expect_warning(expect_error(plot.char.diff(morpho_matrix, type = "density", main = c("main", "bob"))))
+    expect_warning(expect_error(plot.char.diff(morpho_matrix, type = "density", col = "blue")))
+    expect_warning(expect_error(plot.char.diff(morpho_matrix, type = "matrix", col = "blue")))
+    expect_warning(expect_error(plot.char.diff(morpho_matrix, type = "density", xlab = c("main", "bob"))))
+    expect_warning(expect_error(plot.char.diff(morpho_matrix, type = "density", ylab = c("main", "bob"))))
+
+
     ## Plotting a matrix
     capture_warnings(test <- plot.char.diff(morpho_matrix))
     expect_equal(names(test), c("rect", "text"))
@@ -139,6 +155,13 @@ test_that("char.diff plot (graphic)", {
     ## Plotting the density profile of a char.diff object
     capture_warnings(char.diff_matrix <- char.diff(morpho_matrix))
     test <- plot(char.diff_matrix, type = "density")
+    expect_equal(names(test), c("rect", "text"))
+    expect_equal(unique(unlist(lapply(test, lapply, class))), "numeric")
+
+    ## With NA
+    morpho_matrix[, 1] <- NA
+    warn <- capture_warnings(test <- plot.char.diff(morpho_matrix))
+    expect_equal(warn, "NAs introduced by coercion")
     expect_equal(names(test), c("rect", "text"))
[…]

### 2018-05-13 tests/testthat/test-dispRity.utilities.R

@@ -175,7 +175,7 @@ test_that("matrix.dispRity", {
 test_that("get.subsets", {
     data(BeckLee_mat99)
     data(BeckLee_tree)
-    subsets_full <- time.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
+    subsets_full <- chrono.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
     bootstrapped_data <- boot.matrix(subsets_full, bootstraps = 10, rarefaction = c(3, 5))
     disparity_data <- dispRity(bootstrapped_data, variances)
 
@@ -246,7 +246,7 @@ test_that("get.subsets", {
 ## extract.dispRity
 test_that("extract.dispRity", {
     data(BeckLee_mat99) ; data(BeckLee_tree) 
-    subsets_full <- time.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
+    subsets_full <- chrono.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous",time = 5, model = "acctran")
     bootstrapped_data <- boot.matrix(subsets_full, bootstraps = 10, rarefaction = c(3, 5))
     data <- dispRity(bootstrapped_data, c(sum,variances))
 
@@ -314,7 +314,7 @@ test_that("extract.dispRity", {
 
 })
 
-test_that("scale.dispRity", {
+test_that("rescale.dispRity", {
     data(BeckLee_mat50)
     groups <- as.data.frame(matrix(data = c(rep(1, nrow(BeckLee_mat50)/2), rep(2, nrow(BeckLee_mat50)/2)), nrow = nrow(BeckLee_mat50), ncol = 1, dimnames = list(rownames(BeckLee_mat50))))
     customised_subsets <- custom.subsets(BeckLee_mat50, groups)
@@ -322,31 +322,31 @@ test_that("scale.dispRity", {
     data <- dispRity(bootstrapped_data, metric = c(sum, centroids))
 
     expect_error(
-        scale.dispRity(bootstrapped_data)
+        rescale.dispRity(bootstrapped_data)
         )
     expect_error(
-        scale.dispRity(data, scale = "yes")
+        rescale.dispRity(data, scale = "yes")
         )
     expect_error(
-        scale.dispRity(data, center = "yes")
+        rescale.dispRity(data, center = "yes")
         )
     expect_error(
[…]

### 2018-05-13 tests/testthat/test-null.test.R

@@ -170,7 +170,7 @@ test_that("null.test example works", {
     expect_is(results, c("dispRity", "randtest"))
 
     for(test in 1:length(results)) {
-        expect_equal(round(results[[test]][[1]], digits = 6), summary(sum_variances, rounding = 6)$obs[test])
+        expect_equal(round(results[[test]][[1]], digits = 6), summary(sum_variances, digits = 6)$obs[test])
         expect_equal(results[[test]][[5]], 0.09090909)
     }
 

### 2018-05-13 tests/testthat/test-test.dispRity.R

@@ -411,7 +411,7 @@ test_that("adonis and dtt works fine", {
     data(BeckLee_ages)
 
     groups <- dispRity(custom.subsets(BeckLee_mat99, group = crown.stem(BeckLee_tree)), metric = c(ranges))
-    time <- dispRity(time.subsets(BeckLee_mat99, tree = BeckLee_tree, method = "continuous", model = "acctran", FADLAD = BeckLee_ages, time = 10), metric = c(ranges))
+    time <- dispRity(chrono.subsets(BeckLee_mat99, tree = BeckLee_tree, method = "continuous", model = "acctran", FADLAD = BeckLee_ages, time = 10), metric = c(ranges))
     
    expect_warning(test_group_adonis <- test.dispRity(data = groups, test = adonis.dispRity))
    expect_warning(test_group_adonis2 <- test.dispRity(groups, test = vegan::adonis))
@@ -420,7 +420,4 @@ test_that("adonis and dtt works fine", {
     expect_is(test_group_adonis, "adonis")
     expect_is(test_group_adonis2, "adonis")
     expect_is(test_time_adonis, "adonis")
-
-
-
 })

### 2018-05-30 inst/gitbook/03_specific-tutorials.Rmd
Merged master into model-test2
# Conflicts:
#	DESCRIPTION
#	NAMESPACE
#	NEWS.md
#	R/adonis.dispRity.R
#	R/boot.matrix.R
#	R/clean.data.R
#	R/dispRity.metric.R
#	R/dispRity.utilities.R
#	R/est.VCV.R
#	R/est.mean.R
#	R/input.param.R
#	R/make.metric.R
#	R/misc.model.test.fun.R
#	R/model.test.R
#	R/model.test.lik.R
#	R/model.test.sim.R
#	R/opt.mode.R
#	R/plot.dispRity.R
#	R/plot.dispRity_fun.R
#	R/print.dispRity.R
#	R/summary.dispRity.R
#	README.md
#	inst/gitbook/03_specific-tutorials.Rmd
#	man/model.test.Rd
#	man/model.test.sim.Rd
#	paper/References.bib
#	paper/dispRity_paper.tex
#	paper/dispRity

@@
 
-One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
+One other argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
 When missing, all dimensions from the ordinated matrix are used.
@@
 boot.matrix(BeckLee_mat50, dimensions = 10)
+```
+
+It is also possible to specify the sampling probability in the bootstrap for each elements.
+This can be useful for weighting analysis for example (i.e. giving more importance to specific elements).
+These probabilities can be passed to the `prob` argument individually with a vector with the elements names or with a matrix with the rownames as elements names.
+The elements with no specified probability will be assigned a probability of 1 (or 1/maximum weight if the argument is weights rather than probabilities).
+
+```{r, eval=TRUE}
+## Attributing a weight of 0 to Cimolestes and 10 to Maelestes
+boot.matrix(BeckLee_mat50, prob = c("Cimolestes" = 0, "Maelestes" = 10))
 ```

### 2018-05-30 R/boot.matrix.R

@@ -8,6 +8,7 @@
 #' @param dimensions Optional, a \code{numeric} value or proportion of the dimensions to keep.
 #' @param verbose A \code{logical} value indicating whether to be verbose or not.
 #' @param boot.type The bootstrap algorithm to use (\code{default = "full"}; see details).
+#' @param prob Optional, a \code{matrix} or a \code{vector} of probabilities for each element to be selected during the bootstrap procedure. The \code{matrix} or the \code{vector} must have a rownames or names attribute that corresponds to the elements in \code{data}.
 #' 
 #' @return
 #' This function outputs a \code{dispRity} object containing:
@@ -21,11 +22,18 @@
 #' \code{data}: The data is considered as the multidimensional space and is not transformed (e.g. if ordinated with negative eigen values, no correction is applied to the matrix).
 #' 
 #' \code{rarefaction}: when the input is \code{numeric}, the number of elements is set to the value(s) for each bootstrap. If some subsets have fewer elements than the rarefaction value, the subsets is not rarefied.
+#' 
 #' \code{boot.type}: the different bootstrap algorithms are:
 #' \itemize{
 #'   \item \code{"full"}: resamples all the rows of the matrix and replaces them with a new random sample of rows (with \code{replace = TRUE}, meaning all the elements can be duplicated in each bootstrap).
 #'   \item \code{"single"}: resamples only one row of the matrix and replaces it with a new randomly sampled row (with \code{replace = FALSE}, meaning that only one element can be duplicated in each bootstrap).
 #' }
+#' 
+#' \code{prob}: This option allows to attribute specific probability to each element to be drawn.
+#' A probability of 0 will never sample the element, a probability of 1 will sample.
+#' This can also be useful for weighting elements: an element with a weight of 10 will be sampled ten times more.
+#' If the argument is a \code{matrix}, it must have rownames attrbiutes corresponding to the element names.
+#' If the argument is a \code{vector}, it must have names attributes corresponding to the element names.
 #'
 #' @seealso \code{\link{cust.subsets}}, \code{\link{chrono.subsets}}, \code{\link{dispRity}}.
 #'
@@ -42,6 +50,8 @@
 #' boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(7, 10, 11))
 #' ## Bootstrapping an ordinated matrix with only 3 dimensions
 #' boot.matrix(BeckLee_mat50, bootstraps = 20, dimensions = 3)
[…]

### 2018-05-30 R/chrono.subsets.R

@@ -117,7 +117,7 @@ chrono.subsets <- function(data, tree, method, time, model, inc.nodes = FALSE, F
     if(!missing(tree)) {
         check.class(tree, "phylo")
         ## tree must be dated
-        if(length(tree$root.time) != 1) stop("Tree must be a dated tree with a $root.time element.")
+        if(length(tree$root.time) != 1) stop("Tree must be a dated tree with a $root.time element.", call. = FALSE)
         ## tree.age_tree variable declaration
         tree.age_tree <- tree.age(tree)
     }
@@ -139,14 +139,14 @@ chrono.subsets <- function(data, tree, method, time, model, inc.nodes = FALSE, F
 
     ## If the tree is missing, the method can intake a star tree (i.e. no phylogeny)
     if(missing(tree)) {
-        if(missing(FADLAD)) stop("If no phylogeny is provided, all elements must be present in the FADLAD argument.")
-        if(method == "continuous") stop("If no phylogeny is provided, method must be \"discrete\".")
+        if(missing(FADLAD)) stop("If no phylogeny is provided, all elements must be present in the FADLAD argument.", call. = FALSE)
+        if(method == "continuous") stop("If no phylogeny is provided, method must be \"discrete\".", call. = FALSE)
         tree_was_missing <- TRUE
 
         ## Checking FADLAD disponibilities
         names_data <- rownames(data)
         ## All names must be present in the data
-        if(!all(names_data %in% rownames(FADLAD))) stop("If no phylogeny is provided, all elements must be present in the FADLAD argument.")
+        if(!all(names_data %in% rownames(FADLAD))) stop("If no phylogeny is provided, all elements must be present in the FADLAD argument.", call. = FALSE)
         ## Generating the star tree
         tree <- stree(nrow_data, tip.label = names_data)
         tree$root.time <- max(FADLAD)
@@ -170,7 +170,7 @@ chrono.subsets <- function(data, tree, method, time, model, inc.nodes = FALSE, F
     ## If time is a single value create the time vector by sampling evenly from just after the tree root time (1%) to the present
     if(length(time) == 1) {
         ## time must be at least three if discrete
-        if(time < 2) stop("Time must be greater or equal than 2.")
+        if(time < 2) stop("Time must be greater or equal than 2.", call. = FALSE)
 
         if(missing(tree)) {
             ## Set tmax
@@ -198,7 +198,7 @@ chrono.subsets <- function(data, tree, method, time, model, inc.nodes = FALSE, F
                 }
[…]

### 2018-05-30 R/model.test.sim.R

@@ -11,6 +11,7 @@
 #' @param variance The variance of each sample of class numeric. If one number is supplied this is the variance for all samples in the sequence. If a vector of equal length to the time.span vector is supplied, this is used for the variance of each sample in the sequence
 #' @param sample.size The sample.size of each sample of class numeric. If one number is supplied this is the sample.size for all samples in the sequence. If a vector of equal length to the time.span vector is supplied, this is used for the sample.size of each sample in the sequence
 #' @param parameters \code{list}  The model parameters used for simulations: ancestral.state (default = 0.01), sigma.squared (default = 1), alpha (default = 1), optima.1, optima.2, optima.3 (default = 0.15), theta.1, theta.2, theta.3 (default = 1), omega (default = 1), trend (default = 0.5), eb.rate (default = -0.1). 
+#' @param fixed.optima A \code{logical} value, whether to use an estimated optimum value in OU models (\code{FALSE} - default), or whether to set the OU optimum to the ancestral value (\code{TRUE}).
 #' @param model.rank If a \code{dispRity.sim} is supplied, which model is used for simulation. The rank refers to the order of models as specified by AICc, so if rank=1 (default) the best-fitting model is used for simulation
 #' @return A list of class \code{dispRity.sim}. Each list element contains the simulated central_tendency, as well as the variance, sample_size, and subsets useds to simulate the data.
 #'
@@ -36,94 +37,94 @@
 model.test.sim <- function(sim=1, model, time.split=NULL, time.span=100, variance=1, sample.size=100, parameters=list(), fixed.optima=FALSE, model.rank=1) {
     
     if(length(class(model)) == 2 && class(model)[2] == "model.test") {
-    		
-    		test.p <- TRUE
-		empirical.model <- model    		
-    		
-    		model.rank <- order(model[[1]][,1])[model.rank]
-		model.details <- model[[2]][model.rank][[1]]	
-		model.name <- rownames(model[[1]])[model.rank]
-    		parameters <- model[[2]][[model.rank]]$par
-    		
-    		time.span <- model$model.data$subsets
-    		variance <- model$model.data$variance
+            
+        test.p <- TRUE
+        empirical.model <- model            
+            
+        model.rank <- order(model[[1]][,1])[model.rank]
[…]

### 2018-05-30 R/plot.dispRity.R

@@ -202,7 +202,7 @@ plot.dispRity <- function(x, ..., type, quantiles = c(50, 95), cent.tend = media
                 }
                 ## Are quantiles proper proportions
                 if(any(quantiles < 0) | any(quantiles > 100)) {
-                    stop("quantiles(s) must be any value between 0 and 100.")
+                    stop("quantiles(s) must be any value between 0 and 100.", call. = FALSE)
                 }
                 quantiles_n <- length(quantiles)
 
@@ -210,7 +210,7 @@ plot.dispRity <- function(x, ..., type, quantiles = c(50, 95), cent.tend = media
                 check.class(cent.tend, "function")
                 ## The function must work
                 if(make.metric(cent.tend, silent = TRUE) != "level1") {
-                    stop("cent.tend argument must be a function that outputs a single numeric value.")
+                    stop("cent.tend argument must be a function that outputs a single numeric value.", call. = FALSE)
                 }
 
 
@@ -264,7 +264,7 @@ plot.dispRity <- function(x, ..., type, quantiles = c(50, 95), cent.tend = media
     ## must be class dispRity
     check.class(data, "dispRity")
     ## must have one element called dispRity
-    if(is.na(match("disparity", names(data)))) stop("Data must be a 'dispRity' object.")
+    if(is.na(match("disparity", names(data)))) stop("Data must be a 'dispRity' object.", call. = FALSE)
     ## Check if disparity is a value or a distribution
     is_distribution <- ifelse(length(data$disparity[[1]]$elements) != 1, TRUE, FALSE)
     ## Check the bootstraps
@@ -282,7 +282,7 @@ plot.dispRity <- function(x, ..., type, quantiles = c(50, 95), cent.tend = media
         }
         ## Are quantiles proper proportions
         if(any(quantiles < 0) | any(quantiles > 100)) {
-            stop("quantiles(s) must be any value between 0 and 100.")
+            stop("quantiles(s) must be any value between 0 and 100.", call. = FALSE)
         }
     }
 
@@ -291,7 +291,7 @@ plot.dispRity <- function(x, ..., type, quantiles = c(50, 95), cent.tend = media
     check.class(cent.tend, "function")
     ## The function must work
     if(make.metric(cent.tend, silent = TRUE) != "level1") {
-        stop("cent.tend argument must be a function that outputs a single numeric value.")
+        stop("cent.tend argument must be a function that outputs a single numeric value.", call. = FALSE)
[…]

### 2018-05-30 tests/testthat/test-model.test.R

@@ -10,7 +10,7 @@ test_that("simple models work", {
     test <- model.test(data, model = "BM", pool.variance = NULL, time.split = NULL, fixed.optima = FALSE, control.list = list(fnscale = -1), verbose = FALSE)
 
     expect_equal(class(test), c("dispRity", "model.test"))
-    expect_equal(names(test), c("aic.models", "full.details", "call"))
+    expect_equal(names(test), c("aic.models", "full.details", "call", "model.data", "fixed.optima"))
     expect_is(test[[1]], c("matrix"))
     expect_is(test[[2]], c("list"))
     expect_equal(round(test[[2]][[1]]$value, digit = 4), 8.2029)
@@ -19,7 +19,7 @@ test_that("simple models work", {
     test <- model.test(data, model = "Stasis", pool.variance = NULL, time.split = NULL, fixed.optima = FALSE, control.list = list(fnscale = -1), verbose = FALSE)
 
     expect_equal(class(test), c("dispRity", "model.test"))
-    expect_equal(names(test), c("aic.models", "full.details", "call"))
+    expect_equal(names(test), c("aic.models", "full.details", "call", "model.data", "fixed.optima"))
     expect_is(test[[1]], c("matrix"))
     expect_is(test[[2]], c("list"))
     expect_equal(round(test[[2]][[1]]$value, digit = 4), round(-14.7086, digit = 4))
@@ -28,7 +28,7 @@ test_that("simple models work", {
     test <- model.test(data, model = "Trend", pool.variance = NULL, time.split = NULL, fixed.optima = FALSE, control.list = list(fnscale = -1), verbose = FALSE)
 
     expect_equal(class(test), c("dispRity", "model.test"))
-    expect_equal(names(test), c("aic.models", "full.details", "call"))
+    expect_equal(names(test), c("aic.models", "full.details", "call", "model.data", "fixed.optima"))
     expect_is(test[[1]], c("matrix"))
     expect_is(test[[2]], c("list"))
     expect_equal(round(test[[2]][[1]]$value, digit = 4), round(10.5916, digit = 4))
@@ -37,7 +37,7 @@ test_that("simple models work", {
     test <- model.test(data, model = "OU", pool.variance = NULL, time.split = NULL, fixed.optima = FALSE, control.list = list(fnscale = -1), verbose = FALSE)
 
     expect_equal(class(test), c("dispRity", "model.test"))
[…]

### 2018-05-31 inst/gitbook/03_specific-tutorials.Rmd
draft for the model test vignette

@@
 
+### Disparity model fitting with `model.test`
+
+[The model.test does blabalbalb]
+
+First let's create a fine scaled disparity through time dataset:
+
+```{r}
+data(BeckLee_mat99)
+data(BeckLee_tree)
+
+## Making a time slicing analysis every million years
+slice120 <- chrono.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous", model = "proximity",
+                       time = seq(from = 0, to = 120, by = 1))
+## Calculating disparity as the sum of variance on the bootstrapped data
+disparity_data <- dispRity(boot.matrix(slice120), metric = c(sum, variances))
+```
+
+We can then check which mode of changes of disparity-through-time best fits the observed data.
+Here we are going to tests 7 different models:
+
+ * A simple Brownian Motion ("BM")
+ * A simple Ornstein-Uhlenbeck process ("OU")
+ * An early burst model ("EB") which is a rapid increase in variance followed by an OU model
+ * A multiple OU model ("multi.OU") which is an OU model but with varying alpha parameters (the sometimes called "selection" parameter)
+ * An OU model followed by a Trend model (the Trend model is an OU model with increasing mean through time)
+ * A Trend model followed by an OU model.
+
+Note that the models can be any combination of the based implemented models which are `"BM"`, `"OU"`, `"EB"` and `"Trend"`.
+For example, one could test a model `c("EB", "Trend", "OU", "BM", "OU", "BM")` for a (weird) model with an early burst, followed by a trend and a cycle of variance constrain (the OU parts) and releases (the BM parts).
+
+```{r}
+## The list of models to test
+models_to_test <- list("BM", "OU", "EB", "multi.OU", "Trend", c("OU", "Trend"), c("Trend", "OU"))
+
+## Testing all the models on the observed data (with fixed optima)
+model_test_output <- model.test(disparity_data, models_to_test,
+                                time.split = 66, fixed.optima = TRUE)
+```
+
+We can now visualise the results of the model using the normal S3 associated methods (`print`, `summary` and `plot`):
+
+```{r, fig.width=8, fig.height=8}
+## The model fitting results
+model_test_output
+
+## The summary table of all the models
+summary(model_test_output)
+
+## The visualisation of the models AICs
+plot(model_test_output)
+```
+
+We can then test whether out model is significant based on simulation replicates:
+
+```{r}
+## Simulate 1000 datasets for the best fitting model ("Trend") and compare it to the observed result
+model_simulation <- model.test.sim(sim = 1000, model = model_test_output)
+```
+
+Again, these results can be visualised easily as before:
+
+```{r}
+## The simulation results
+model_simulation
+
+# ## The summary table of the simulations
+# summary(model_simulation)
+
+# ## Plotting the simulations
+
+# #### plot for simulated data versus real data - could this be incorporated into the plot functions?
+# outputs.sim <- sapply(model_simulation$simulation.data, function(x) x$central_tendency)
+
+# plot(disparity_data, ylim=c(0, max(c(outputs.sim))))
+
+# min.max <- t(apply(outputs.sim, 1, range))
+
+# time.span <- as.numeric(outputs.sim[[4]][[4]])
+
+# polygon(x=c(time.span, rev(time.span), time.span, rev(time.span)), y=c(min.max[,1], rev(min.max[,2]), min.max[,1], rev(min.max[,2])), col="#ff000030", border=F)
+```
+
+```{r}
+
+# #### re-analyse simulated data (just the first five for demonstration)
+# five.models <- sapply(model.test.sim.output$simulation.data[1:5], function(x) model.test(data=x, model=models.to.test, time.split=66.02, fixed.optima=TRUE, pool.variance=FALSE, verbose=FALSE)[[1]][,1])
+
+# # model support (split between multi.OU and Trend) - it seems very difficult to choose between the two (at least with this miniscule sample size).
+We could maybe repeat this with x1000 or x10000 for publication
+# table(rownames(five.models)[apply(five.models, 2, which.min)])
+```
+
 ## Disparity as a distribution

### 2018-07-12 inst/gitbook/index.Rmd
added cites to the manual

@@
 
-You can cite both the package or this manual with the following citation:
+To cite the package, this manual or some specific functionalities, you can use the following references:
 
-> Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646
+The package main paper:
 
-<!-- > Guillerme, T. & Cooper, N. (2016). The dispRity R package manual. Fighsare XXXX -->
+> Guillerme T. dispRity: A modular R package for measuring disparity. Methods Ecol Evol. 2018;9:1755–1763. [doi.org/10.1111/2041-210X.13022](https://doi.org/10.1111/2041-210X.13022).
 
-Note that this citation is only temporary (but can still be used!).
-A proper description of the package is currently in review in Methods in Ecology and Evolution.
-Upon acceptance (hopefully!), a proper DOI will be released for this manual and the package description.
+The package manual (regularly updated!):
+
+> Guillerme, T. & Cooper, N. (2018): dispRity manual. figshare. Preprint. [10.6084/m9.figshare.6187337.v1](https://doi.org/10.6084/m9.figshare.6187337.v1).
+
+The time-slicing method implemented in `chrono.subsets` (unfortunately not Open Access, but you can still get a [free copy from here](https://tguillerme.github.io/papers/Guillerme&Cooper-2018-Palaeontology.pdf)):
+
+> Guillerme, T.
+and Cooper, N.
+(2018), Time for a rethink: time sub-sampling methods in disparity-through-time analyses.
+Palaeontology, 61: 481-493.
+[doi:10.1111/pala.12364](https://scholar.google.co.uk/scholar?hl=en&as_sdt=0%2C5&q=Guillerme%2C+T.+and+Cooper%2C+N.+%282018%29%2C+Time+for+a+rethink%3A+time+sub-sampling+methods+in+disparity-through-time+analyses.+Palaeontology%2C&btnG=).
+
+<!-- ## Associated papers -->
+
+<!-- Here are some papers that have used the `dispRity` package: -->

### 2018-07-19 inst/gitbook/01_glossary.Rmd
Updated book with model.test part

@@
 date: "`r Sys.Date()`"
-bibliography: [References.bib, packages.bib]
+bibliography: [../References.bib, ../packages.bib]
 output:

### 2018-07-19 inst/gitbook/03_specific-tutorials.Rmd
Updated book with model.test part

--- 2018-05-30 inst/gitbook/03_specific-tutorials.Rmd
@@
 title: "Details of specific functions"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
+author: "Thomas Guillerme (guillert@tcd.ie), Mark Puttick (marknputtick@gmail.com) and  Natalie Cooper (natalie.cooper@nhm.ac.uk)"
 date: "`r Sys.Date()`"
-bibliography: [References.bib, packages.bib]
+bibliography: [../References.bib, ../packages.bib]
 output:
@@
 
-For the time-slicing method details see @time-slice. <!-- @@@ Change cite appropriately! -->
-<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
+For the time-slicing method details see @time-slice.
 For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:
@@
 
+## Fitting modes of evolution to disparity data
+
+The code used for these models is based on those developed by Gene Hunt [@hunt2006fitting; @hunt2012measuring; @hunt2015simple].
+So we acknowledge and thank Gene Hunt for developing these models and writing the original R code that served as inspiration for these models.
+
+### Simple modes of disparity change through time
+
+#### `model.test`
+
+Changes in disparity-through-time can follow a range of models, such as random walks, stasis, constrained evolution, trends, or an early burst model of evolution.
+We will start with by fitting the simplest modes of evolution to our data.
+For example we may have a null expectation of time-invariant change in disparity in which values fluctuate with a variance around the mean - this would be best describe by a Stasis model:
+
+```{r}
+## Loading premade disparity data
+data(BeckLee_disparity)
+disp_time <- model.test(data = BeckLee_disparity, model = "Stasis")
+```
+
+We can see the standard output from ```model.test```.
+The first output message tells us it has tested for equal variances in each sample.
+The model uses Bartlett's test of equal variances to assess if variances are equal, so if p > 0.05 then variance is treated as the same for all samples, but if (p < 0.05) then each bin variance is unique.
+Here we have p < 0.05, so variance is not pooled between samples.
+
+By default ```model.test``` will use Bartlett's test to assess for homogeneity of variances, and then use this to decide to pool variances or not.
+This is ignored if the argument ```pool.variance``` in ```model.test``` is changed from the default ```NULL``` to ```TRUE``` or ```FALSE```.
+For example, to ignore Bartlett's test and pool variances manually we would do the following:
+
+```{r}
+disp_time_pooled <- model.test(data = BeckLee_disparity, model = "Stasis", pool.variance = TRUE)
+```
+
+However, unless you have good reason to choose otherwise it is recommended to use the default of ```pool.variance = NULL```:
+
+```{r}
+disp_time <- model.test(data = BeckLee_disparity, model = "Stasis", pool.variance = NULL)
+disp_time
+```
+
+The remaining output gives us the log-likelihood of the Stasis model of -59.501 (you may notice this change when we pooled variances above).
+The output also gives us the small sample Akaike Information Criterion (AICc), the delta AICc (the distance from the best fitting model), and the AICc weights (~the relative support of this model compared to all models, scaled to one).
+
+These are all metrics of relative fit, so when we test a single model they are not useful.
+By using the function summary in ```dispRity``` we can see the maximum likelihood estimates of the model parameters:
+
+```{r}
+summary(disp_time)
+```
+
+So we again see the AICc, delta AICc, AICc weight, and the log-likelihood we saw previously.
+We now also see the number of parameters from the model (2: theta and omega), and their estimates so the variance (omega = 0.01) and the mean (theta.1 = 3.4).
+
+The ```model.test``` function is designed to test relative model fit, so we need to test more than one model to make relative comparisons.
+So let's compare to the fit of the Stasis model to another model with two parameters: the Brownian motion.
+Brownian motion assumes a constant mean that is equal to the ancestral estimate of the sequence, and the variance around this mean increases linearly with time.
+The easier way to compare these models is to simply add `"BM"` to the ```models``` vector argument:
+
+```{r}
+disp_time <- model.test(data = BeckLee_disparity, model = c("Stasis", "BM"))
+disp_time
+```
+
+Et voilà! Here we can see by the log-likelihood, AICc, delta AICc, and AICc weight Brownian motion has a much better relative fit to these data than the Stasis model.
+Brownian motion has a relative AICc fit 366 units better than Stasis, and virtually has a AICc weight of 1.
+
+We can also all the information about the relative fit of models alongside the maximum likelihood estimates of model parameters using the summary function
+
+```{r}
+summary(disp_time)
+```
+
+Not that because the parameters per models differ, the summary includes `NA` for inapplicable parameters per models (e.g.
+the theta and omega parameters from the Stasis models are inapplicable for a Brownian motion model).
+
+We can plot the relative fit of our models using the ```plot``` function
+
+```{r plot1, fig.align="center", fig.cap = "relative fit (AICc weight) of Stasis and Brownian models of disparity through time", echo = TRUE, fig.height = 6, fig.width = 6}
+plot(disp_time)
+```
+
+Here we see and overwhelming support for the Brownian motion model.
+
+Alternatively, we could test all available models single modes: Stasis, Brownian motion, Ornstein-Uhlenbeck (evolution constrained to an optima), Trend (increasing or decreasing mean through time), and Early Burst (exponentially decreasing rate through time)
+
+```{r}
+disp_time <- model.test(data = BeckLee_disparity, model = c("Stasis", "BM", "OU", "Trend", "EB"))
+summary(disp_time)
+```
+
+These models indicate support for a Trend model, and we can plot the relative support of all model AICc weights
+
+```{r plot2, fig.align="center", fig.cap = "relative fit (AICc weight) of various modes of evolution", echo = TRUE, fig.height = 6, fig.width = 6}
+plot(disp_time)
+```
+
+Is this a trend of increasing or decreasing disparity through time? One way to find out is to look at the summary function for the Trend model:
+
+```{r}
+summary(disp_time)["Trend",]
+```
+
+This show a positive trend (0.01) of increasing disparity through time.
+
+### Plot and run simulation tests in a single step
+
+#### `model.test.wrapper`
+
+Patterns of evolution can be fit using ```model.test```, but the ```model.test.wrapper``` fits the same models as `model.test` as well as running predictive tests and plots.
+
+The predictive tests use the maximum likelihood estimates of model parameters to simulate a number of datasets (default = 1000), and analyse whether this is significantly different to the empirical input data using the Rank Envelope test [@murrell2018global].
+Finally we can plot the empirical data, simulated data, and the Rank Envelope test p values.
+This can all be done using the function ```model.test.wrapper```, and we will set the argument ```show.p = TRUE``` so _p_ values from the Rank Envelope test are printed on the plot:
+
+```{r plot3, fig.align="center", fig.cap = "Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for Trend, OU, BM, EB, and Stasis models", echo = TRUE, fig.height = 18, fig.width = 12}
+disp_time <- model.test.wrapper(data = BeckLee_disparity, model = c("Stasis", "BM", "OU", "Trend", "EB"),
+                                show.p = TRUE)
+disp_time
+```
+
+From this plot we can see the empirical estimates of disparity through time (pink) compared to the predictive data based upon the simulations using the estimated parameters from each model.
+There is no significant differences between the empirical data and simulated data, except for the Early Burst model.
+
+Trend is the best-fitting model but the plot suggests the OU model also follows a trend-like pattern.
+This is because the optima for the OU model (5.7) is different to the ancestral state 2.835 and outside the observed value.
+This is potentially unrealistic, and one way to alleviate this issue is to set the optima of the OU model to equal the ancestral estimate - this is the normal practice for OU models in comparative phylogenetics.
+To set the optima to the ancestral value we change the argument ```fixed.optima = TRUE```:
+
+```{r plot4, fig.align="center", fig.cap = "Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for Trend, OU, BM, EB, and Stasis models with the optima of the OU model set to equal the ancestral value", echo = TRUE, fig.height = 18, fig.width = 12}
+disp_time <- model.test.wrapper(data = BeckLee_disparity, model = c("Stasis", "BM", "OU", "Trend", "EB"),
+                                show.p = TRUE, fixed.optima = TRUE)
+disp_time
+```
+
+The relative fit of the OU model is decreased by constraining the fit of the optima to equal the ancestral state value.
+In fact as the OU attraction parameter (alpha) is zero, the model is equal to a Brownian motion model but is penalised by having an extra parameter. 
+Note that indeed, the plots of the BM model and the OU model look nearly identical.
+
+### Multiple modes of evolution (time shifts)
+
+As well as fitting a single model to a sequence of disparity values we can also allow for the mode of evolution to shift at a single or multiple points in time.
+The timing of a shift in mode can be based on an a prior expectation, such as a mass extinction event, or the model can test multiple points to allow to find time shift point with the highest likelihood.
+
+Models can be fit using `model.test` but it can be more convenient to use `model.test.wrapper`.
+Here we will compare the relative fit of Brownian motion, Trend, Ornstein-Uhlenbeck and a multi-mode Ornstein Uhlenbck model in which the optima changes at 66 million years ago, the Cretaceous-Palaeogene boundary.
+
+For example, we could be testing the hypothesis that the extinction of non-avian dinosaurs allowed mammals to go from scurrying in the undergrowth (low optima/low disparity) to dominating all habitats (high optima/high disparity).
+We will constrain the optima of OU model in the first time begin (i.e, pre-66 Mya) to equal the ancestral value:
+
+```{r plot5, fig.align="center", fig.cap = "Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for BM, Trend, OU, and multi OU models with a shift in optima allowed at 66 Ma", echo = TRUE, fig.height = 12, fig.width = 12}
+disp_time <- model.test.wrapper(data = BeckLee_disparity, model = c("BM", "Trend", "OU", "multi.OU"),
+                                time.split = 66, pool.variance = NULL, show.p = TRUE,
+                                fixed.optima = TRUE)
+disp_time
+```
+
+The multi-OU model shows an increase an optima at the Cretaceous-Palaeogene boundary, indicating a shift in disparity.
+However, this model does not fit as well as a model in which there is an increasing trend through time.
+We can also fit a model in which the we specify a heterogeneous model but we do not give a `time.split`.
+In this instance the model will test all splits that have at least 10 time slices on either side of the split.
+That's 102 potential time shifts in this example dataset so be warned, the following code will estimate 105 models!
+
+```{r, eval = FALSE}
+## An example of a time split model in which all potential splits are tested
+## WARNING: this will take between 20 minutes and half and hour to run!
+disp_time <- model.test.wrapper(data = BeckLee_disparity, model = c("BM", "Trend", "OU", "multi.OU"),
+                                show.p = TRUE, fixed.optima = TRUE)
+```
+
+As well as specifying a multi-OU model we can run any combination of models.
+For example we could fit a model at the Cretaceous-Palaeogene boundary that goes from an OU to a BM model, a Trend to an OU model, a Stasis to a Trend model or any combination you want to use.
+The only model that can't be used in combination is a multi-OU model.
+
+These can be introduced by changing the input for the models into a list, and supplying a vector with the two models.
+This is easier to see with an example:
+
+```{r plot6, fig.align="center", fig.cap = "Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for a variety of models with a shift in optima allowed at 66 Ma", echo = TRUE, fig.height = 18, fig.width = 12}
+## The models to test
+my_models <- list(c("BM", "OU"),
+                  c("Stasis", "OU"),
+                  c("BM", "Stasis"),
+                  c("OU", "Trend"),
+                  c("Stasis", "BM"))
+
+## Testing the models
+disp_time <- model.test.wrapper(data = BeckLee_disparity, model = my_models, time.split = 66,
+                                show.p = TRUE, fixed.optima = TRUE)
+disp_time
+```
+
+### `model.test.sim`
+
+Note that all the models above where run using the `model.test.wrapper` function that is a... wrapping function!
+In practice, this function runs two main functions from the `dispRity` package and then plots the results:
+
+ * `model.test` and
+ * `model.test.sim`
+
+The `model.test.sim` allows to simulate disparity evolution given a `dispRity` object input (as in `model.test.wrapper`) or given a model and its specification.
+For example, it is possible to simulate a simple Brownian motion model (or any of the other models or models combination described above):
+
+```{r}
+## A simple BM model
+model_simulation <- model.test.sim(sim = 1000, model = "BM", time.span = 50, variance = 0.1,
+                                   sample.size = 100, parameters = list(ancestral.state = 0))
+model_simulation
+```
+
+This will simulate 1000 Brownian motions for 50 units of time with 100 sampled elements, a variance of 0.1 and an ancestral state of 0.
+We can also pass multiple models in the same way we did it for `model.test`
+This model can then be summarised and plotted as most `dispRity` objects:
+
+```{r plot7, fig.align="center", fig.cap = "A simulated Brownian motion", echo = TRUE, fig.height = 6, fig.width = 6}
+## Displaying the 5 first rows of the summary
+head(summary(model_simulation))
+
+## Plotting the simulations
+plot(model_simulation)
+```
+
+Note that these functions can take all the arguments that can be passed to `plot`, `summary`, `plot.dispRity` and `summary.dispRity`.
+
+#### Simulating tested models
+
+Maybe more interestingly though, it is possible to pass the output of `model.test` directly to `model.test.sim` to simulate the models that fits the data the best and calculate the Rank Envelope test _p_ value.
+Let's see that using the simple example from the start:
+
+```{r}
+## Fitting multiple models on the data set
+disp_time <- model.test(data = BeckLee_disparity, model = c("Stasis", "BM", "OU", "Trend", "EB"))
+summary(disp_time)
+```
+
+As seen before, the Trend model fitted this dataset the best.
+To simulate what 1000 Trend models would look like using the same parameters as the ones estimated with `model.test` (here the ancestral state being 2.839, the sigma squared beeing 0.002 and the trend of 0.01), we can simply pass this model to `model.test.sim`:
+
+```{r}
+## Simulating 1000 Trend model with the observed parameters
+sim_trend <- model.test.sim(sim = 1000, model = disp_time)
+sim_trend
+```
+
+By default, the model simulated is the one with the lowest AICc (`model.rank = 1`) but it is possible to choose any ranked model, for example, the OU (second one):
+
+```{r}
+## Simulating 1000 OU model with the observed parameters
+sim_OU <- model.test.sim(sim = 1000, model = disp_time, model.rank = 2)
+sim_OU
+```
+
+And as the example above, the simulated data can be plotted or summarised:
+
+```{r}
+head(summary(sim_trend))
+head(summary(sim_OU))
+```
+
+```{r plot8, fig.align="center", fig.cap = "The best fitted model (Trend) and the observed disparity through time", echo = TRUE, fig.height = 6, fig.width = 6}
+## The trend model with some graphical options
+plot(sim_trend, xlab = "Time (Mya)", ylab = "sum of variances",
+    col = c("#F65205", "#F38336", "#F7B27E"))
+
+## Adding the observed disparity through time
+plot(BeckLee_disparity, add = TRUE, col = c("#3E9CBA", "#98D4CF90", "#BFE4E390"))
+```
+
 ## Disparity as a distribution

### 2018-08-21 inst/gitbook/07_ecology-demo.Rmd

@@
 output: rmarkdown::html_vignette
-bibliography: [References.bib, packages.bib]
+bibliography: [../References.bib, ../packages.bib]
 bst: sysbio.bst

### 2018-08-21 inst/gitbook/08_future_directions.Rmd
Merge branch 'master' into model-test2
# Conflicts:
#	R/est.VCV.R
#	R/est.mean.R
#	R/input.param.R
#	R/make.dispRity.models.R
#	R/misc.model.test.fun.R
#	R/model.test.R
#	R/model.test.lik.R
#	R/model.test.sim.R
#	R/opt.mode.R
#	man/model.test.Rd
#	tests/testthat/test-model.test.R

--- 2017-11-17 inst/gitbook/08_future_directions.Rmd
@@
 output: rmarkdown::html_vignette
-bibliography: References.bib
+bibliography: [../References.bib, ../packages.bib]
 bst: sysbio.bst
@@
 I am equally slowly developing functions to allow more of the options in the package to be modular (in the same way as the `metric` argument in `dispRity`).
-The next arguments to benefit this increased modularity will be `time.subsets`'s `model` argument and `boot.matrix`'s `type` argument.
+The next arguments to benefit this increased modularity will be `chrono.subsets`'s `model` argument and `boot.matrix`'s `type` argument.
 

### 2018-08-21 inst/gitbook/index.Rmd

@@
 title: "dispRity manual"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
+author: "Thomas Guillerme (guillert@tcd.ie), Mark Puttick (marknputtick@gmail.com) and  Natalie Cooper (natalie.cooper@nhm.ac.uk)"
 date: "`r Sys.Date()`"
@@
 documentclass: book
-bibliography: [References.bib, packages.bib]
+bibliography: [../References.bib, ../packages.bib]
 link-citations: yes
@@
 Palaeontology, 61: 481-493.
-[doi:10.1111/pala.12364](https://scholar.google.co.uk/scholar?hl=en&as_sdt=0%2C5&q=Guillerme%2C+T.+and+Cooper%2C+N.+%282018%29%2C+Time+for+a+rethink%3A+time+sub-sampling+methods+in+disparity-through-time+analyses.+Palaeontology%2C&btnG=).
+[doi:10.1111/pala.12364](https://onlinelibrary.wiley.com/doi/abs/10.1111/pala.12364).
 

### 2018-09-27 inst/gitbook/dispRity_manual.Rmd
Updated vignette

@@
 documentclass: book
+bibliography: [../References.bib, ../packages.bib]
 link-citations: yes
@@
 Multidimensional spaces can be ordinated matrices from MDS, PCA, PCO, PCoA but the package is *not* restricted to any type of matrices!
-This manual is based on the version `0.4.2`.
+This manual is based on the version `r version_release`.
 
@@
 install_github("TGuillerme/dispRity", ref = "release")
-
-## Loading the package
-library(dispRity)
-```
+```
+
+Anyways to upload it, you can use
 
@@
 
-Note this uses the `release` branch (version`0.4`). <!-- @@@ -->
+Note this uses the `release` branch (`r version_release`).
 For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
@@
 
-## Why not CRAN?
-
-This package is not available on CRAN.
-This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
-However, the package follows the strict (and useful!) CRAN standards via Travis.
-
-<!-- badges out start -->
+## To CRAN or not to CRAN? {#noCRAN}
+
+This package can be installed via [the CRAN](https://cran.r-project.org/package=dispRity) or via [GitHub](https://github.com/TGuillerme/dispRity).
+
+ * The CRAN version is a more "strict" version of the package that does not include some functionalities (such as [`Claddis.ordination`](#Claddis-ordination) or this gitbook vignette) because of compatibility issues with the CRAN requirement.
+
+ * The GitHub version is a more "reactive" version (fed by user's suggestions and comments) and can contain more functionalities (see list below).
+This version exists on a `"release"` branch which is tested as strictly as the CRAN version with Travis and codecov.
+The `"master"` branch contains always the latest "piping hot", not yet release version.
+It might (though rarely) be bugged.
+
+<!-- badges out start
 [![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
-[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/master/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
+[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
 [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
-<!-- badges out end   -->
+badges out end   -->
+
+Note however, that apart from the functions listed below, both versions of the package have the same tested and implemented functions.
+
+Functions not available on the CRAN version (`r version_CRAN`):
+
+ * [`Claddis.ordination`](#Claddis-ordination)
+
+### So which version do I choose? {#version}
+
+There are always three version of the package available:
+
+  * The CRAN one
+  * The GitHub `release` one
+  * The GitHub `master` one
+
+The differences between the CRAN one and the GitHub `release` or `master` ones is explained just above.
+For the the GitHub version, the differences are that the `release` one is more stable (i.e.
+more rarely modified) and the `master` one is more live one (i.e.
+bug fixes and new functionalities are added as they come).
+
+If you want the latest-latest version of the package I suggest using the GitHub `master` one, especially if you recently emailed me reporting a minor bug or wanting a new functionality!
+Note however that *it can happen* that the `master` version can sometimes be bugged (especially when there are major `R` and `R` packages updates), however, the status of the package state on both the `release` and the `master` version is constantly displayed on the [`README` page](https://github.com/TGuillerme/dispRity/) of the package with the nice badges displaying these different (and constantly tested) information
+
+ * The build status: [![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=master)](https://travis-ci.org/TGuillerme/dispRity) which tests whether the package compiles (i.e.
+can be installed on your computer).
+ * The code coverage: [![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/master/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity) which indicates the percentage of lines in the code that have been tested for expected behaviour.
+ * The project status: [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active) which indicates whether this package is maintained regularly or forgotten in the abyss of internet.
+ * The development version: [![develVersion](https://img.shields.io/badge/devel%20version-1.2.1-green.svg?style=flat)](https://github.com/TGuillerme/dispRity) which indicates which version of the package you are about to install.
+
+Note that when we (the `dispRity` package collaborators!) are developing some more major versions, including some more complex features for example, they will never be released on either the the versions described above.
+Therefore you can rest assure that you will never use a badly bugged version of `dispRity`! Enjoy!
 
@@
 
-You can cite both the package or this manual with the following citation:
-
-> Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646
-
-<!-- > Guillerme, T. & Cooper, N. (2016). The dispRity R package manual. Fighsare XXXX -->
-
-Note that this citation is only temporary (but can still be used!).
-A proper description of the package is currently in review in Methods in Ecology and Evolution and should be re-submitted this winter.
-Upon acceptance (hopefully!), a proper DOI will be released for this manual and the package description.
+To cite the package, this manual or some specific functionalities, you can use the following references:
+
+The package main paper:
+
+> Guillerme T. dispRity: A modular R package for measuring disparity. Methods Ecol Evol. 2018;9:1755–1763. [doi.org/10.1111/2041-210X.13022](https://doi.org/10.1111/2041-210X.13022).
+
+The package manual (regularly updated!):
+
+> Guillerme, T. & Cooper, N. (2018): dispRity manual. figshare. Preprint. [10.6084/m9.figshare.6187337.v1](https://doi.org/10.6084/m9.figshare.6187337.v1).
+
+The time-slicing method implemented in `chrono.subsets` (unfortunately not Open Access, but you can still get a [free copy from here](https://tguillerme.github.io/papers/Guillerme&Cooper-2018-Palaeontology.pdf)):
+
+> Guillerme, T.
+and Cooper, N.
+(2018), Time for a rethink: time sub-sampling methods in disparity-through-time analyses.
+Palaeontology, 61: 481-493.
+[doi:10.1111/pala.12364](https://onlinelibrary.wiley.com/doi/abs/10.1111/pala.12364).
+
+<!-- ## Associated papers -->
+
+<!-- Here are some papers that have used the `dispRity` package: -->
 
@@
 date: "`r Sys.Date()`"
+bibliography: [../References.bib, ../packages.bib]
 output:
@@
     For example, if our space is composed of birds and mammals (the elements) and 50 principal components of variation (the dimensions), we can create two subsets containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
+
+-   **Disparity**.
+A metric expressing the similarities/dissimilarities of the elements within the space or a summarising the space dimensions.
+For example the pairwise distances between elements or the range of each dimensions.
+
+## Glossary equivalences in palaeobiology and ecology
+
+In this manual | In `dispRity` | E.g. in palaeobiology | E.g. in ecology
+---------------|---------------|------------------|-------------
+the multidimensional space | a `matrix` object ($n\times k$) | a morphospace | a function-space
+elements | rows ($n$) | taxa | field experiments
+dimensions | columns ($k$) | morphological characters | communities' compositions
+subsets | a `matrix` ($m \times k$, with $m \leq n$) | time series | experimental treatments
+disparity | a `function` | sum of variances | ellipsoid volume
 
@@
 
-Disparity can be estimated from pretty much any matrix.
-Classically, however, it is measured from ordinated matrices.
-These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).
-
-### Ordination matrices from `Claddis`
+**Any matrix object in `R`.**
+Disparity can be estimated from pretty much any matrix as long as rows represent the elements and columns the dimensions.
+These matrices can be observations, pairwise differences between elements, ordinations, etc...
+
+> Note that whether the matrix is an ordinated one or of any other type, it is always considered as the multidimensional space to analysis and no corrections are applied to it (e.g.
+if ordinated and with negative eigen values, no correction is used).
+Always make sure the input is the right multidimensional space to analyse!
+
+## Ordinated matrices
+
+Classically, when a high number of variables is used, disparity is calculated from ordinated matrices.
+These can be any type of ordinations (PCO, PCA, PCoA, MDS, etc.) as long as elements are the rows (taxa, countries, field experiments) and the dimensions are the columns.
+
+### Ordination matrices from `geomorph`
+
+```{r, eval = TRUE, echo = FALSE, message = FALSE}
+## Checking if geomorph is installed BUILD ONLY
+if(!require(geomorph)) install.packages("geomorph")
+```
+
+You can also easily use data from `geomorph` using the `geomorph.ordination` function.
+This function simply takes Procrustes aligned data and performs an ordination:
+
+```{r}
+require(geomorph)
+
+## Loading the plethodon dataset
+data(plethodon)
+
+## Performing a Procrustes transform on the landmarks
+procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)
+
+## Ordinating this data
+geomorph.ordination(procrustes)[1:5,1:5]
+```
+
+Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
+Additionally you can give the function a `geomorph.data.frame` object.
+If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsets)!
+
+```{r}
+## Using a geomorph.data.frame
+geomorph_df <- geomorph.data.frame(procrustes,
+     species = plethodon$species, site = plethodon$site)
+
+## Ordinating this data and making a dispRity object
+geomorph.ordination(geomorph_df)
+```
+
+More about these `dispRity` objects below!
+
+### Ordination matrices from `Claddis` {#Claddis-ordination}
+
+> This functionality is not available on the CRAN version of the package (`r version_CRAN`) due to CRAN compatibility issues see [to CRAN or not to CRAN?](#noCRAN) for more information and on how to install the GitHub (more complete) version of the package (`r version_release`).
 
@@
 
-### Ordination matrices from `geomorph`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if geomorph is installed BUILD ONLY
-if(!require(geomorph)) install.packages("geomorph")
-```
-
-You can also easily use data from `geomorph` using the `geomorph.ordination` function.
-This function simply takes Procrustes aligned data and performs an ordination:
-
-```{r}
-require(geomorph)
-
-## Loading the plethodon dataset
-data(plethodon)
-
-## Performing a Procrustes transform on the landmarks
-procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)
-
-## Ordinating this data
-geomorph.ordination(procrustes)[1:5,1:5]
-```
-
-Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
-Additionally you can give the function a `geomorph.data.frame` object.
-If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsets)!
-
-```{r}
-## Using a geomorph.data.frame
-geomorph_df <- geomorph.data.frame(procrustes,
-     species = plethodon$species, site = plethodon$site)
-
-## Ordinating this data and making a dispRity object
-geomorph.ordination(geomorph_df)
-```
-
-More about these `dispRity` objects below!
-
 ### Other kinds of ordination matrices
@@
 title: "Details of specific functions"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
+author: "Thomas Guillerme (guillert@tcd.ie), Mark Puttick (marknputtick@gmail.com) and  Natalie Cooper (natalie.cooper@nhm.ac.uk)"
 date: "`r Sys.Date()`"
+bibliography: [../References.bib, ../packages.bib]
 output:
@@
 
-For the time-slicing method details see Cooper and Guillerme (in prep.). <!-- @@@ Change cite appropriately! -->
-<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
+For the time-slicing method details see @time-slice.
 For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:
@@
 ## Generating four time slices automatically
-time.subsets(data = BeckLee_mat99, tree = BeckLee_tree,
+chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree,
     method = "continuous", model = "proximity", time = 4, FADLAD = BeckLee_ages)
 ```
+
+If you want to generate time subsets based on stratigraphy, the package proposes a useful functions to do it for you: `get.bin.ages` (check out the function's manual in `R`)!
 
@@
 
-Of course, one could directly supply the subsets generated above (using `time.subsets` or `custom.subsets`) to this function.
+It is also possible to specify the sampling probability in the bootstrap for each elements.
+This can be useful for weighting analysis for example (i.e. giving more importance to specific elements).
+These probabilities can be passed to the `prob` argument individually with a vector with the elements names or with a matrix with the rownames as elements names.
+The elements with no specified probability will be assigned a probability of 1 (or 1/maximum weight if the argument is weights rather than probabilities).
+
+```{r, eval=TRUE}
+## Attributing a weight of 0 to Cimolestes and 10 to Maelestes
+boot.matrix(BeckLee_mat50, prob = c("Cimolestes" = 0, "Maelestes" = 10))
+```
+
+Of course, one could directly supply the subsets generated above (using `chrono.subsets` or `custom.subsets`) to this function.
 
@@
 ```
-> Because there is only one subset (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with rounding).
+> WARNING: in such dummy space, this gives the estimation of the ellipsoid volume, not the real ellipsoid volume! See the cautionary note in `?ellipse.volume`.
 
@@
 
-The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
+The convex hull based functions are a call to the `geometry::convhulln` function with the `"FA"` option (computes total area and volume).
+Also note that they are really sensitive to the size of the dataset.
 
 > Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) that often results in near 0 disparity values.
+I strongly recommend reading [this really intuitive explanation](https://beta.observablehq.com/@tophtucker/theres-plenty-of-room-in-the-corners) from [Toph Tucker](https://github.com/tophtucker).
 
@@
 
-It is also possible to apply some more *complex* tests that have their own output classes (like `stats::lm`).
-
-The results can then be analysed as usual using the associated `summary` S3 method:
-
-```{r}
-## Performing and linear model applied to the same data
-(slice_lm <- test.dispRity(disparity_time_slices, test = lm,
-                            comparisons = "all"))
-
-## The output is a regular `lm` output
-class(slice_lm)
-
-## This output can be summarised using summary
-summary(slice_lm)
-```
-
-Of course, due to the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
+Because of the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
 The only condition is that the test can be applied to at least two distributions.
-In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm` for example).
+In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm`-type functions for example).
+
+### NPMANOVA in `dispRity`
+
+One often useful test to apply to multidimensional data is the permutational multivariate analysis of variance based on distance matrices `vegan::adonis`.
+This can be done on `dispRity` objects using the `adonis.dispRity` wrapper function.
+Basically, this function takes the exact same arguments as `adonis` and a `dispRity` object for data and performs a PERMANOVA based on the distance matrix of the multidimensional space (unless the multidimensional space was already defined as a distance matrix).
+The `adonis.dispRity` function uses the information from the `dispRity` object to generate default formulas:
+
+  * If the object contains customised subsets, it applies the default formula `matrix ~ group` testing the effect of `group` as a predictor on `matrix` (called from the `dispRity` object as `data$matrix` see [`dispRitu` object details](#The-dispRity-object-content))
+  * If the object contains time subsets, it applies the default formula `matrix ~ time` testing the effect of `time` as a predictor (were the different levels of `time` are the different time slices/bins)
+
+```{r}
+set.seed(1)
+## Generating a random character matrix
+character_matrix <- sim.morpho(rtree(20), 50, rates = c(rnorm, 1, 0))
+
+## Calculating the distance matrix
+distance_matrix <- as.matrix(dist(character_matrix))
+
+## Creating two groups
+random_groups <- list("group1" = 1:10, "group2" = 11:20)
+
+## Generating a dispRity object
+random_disparity <- custom.subsets(distance_matrix, random_groups)
+
+## Running a default NPMANOVA
+adonis.dispRity(random_disparity)
+```
+
+Of course, it is possible to pass customised formulas if the disparity object contains more more groups.
+In that case the predictors must correspond to the names of the groups explained data must be set as `matrix`:
+
+```{r}
+## Creating two groups with two states each
+groups <- as.data.frame(matrix(data = c(rep(1,10), rep(2,10), rep(c(1,2), 10)),
+         nrow = 20, ncol = 2, dimnames = list(paste0("t", 1:20), c("g1", "g2"))))
+
+## Creating the dispRity object
+multi_groups <- custom.subsets(distance_matrix, groups)
+
+## Running the NPMANOVA
+adonis.dispRity(multi_groups, matrix ~ g1 + g2)
+```
+
+Finally, it is possible to use objects generated by `chrono.subsets`.
+In this case, `adonis.dispRity` will applied the `matrix ~ time` formula by default:
+
+```{r}
+## Creating time series
+time_subsets <- chrono.subsets(BeckLee_mat50, BeckLee_tree, 
+     method = "discrete", inc.nodes = FALSE, time = c(100, 85, 65, 0),
+     FADLAD = BeckLee_ages)
+
+## Running the NPMANOVA with time as a predictor
+adonis.dispRity(time_subsets)
+```
+
+Note that the function warns you that the input data was transformed into a distance matrix.
+This is reflected in the Call part of the output (`formula = dist(matrix) ~ time`).
+
+To use each time subset as a separate predictor, you can use the `matrix ~ chrono.subsets` formula; this is equivalent to `matrix ~ first_time_subset + second_time_subset + ...`:
+
+```{r}
+## Running the NPMANOVA with each time bin as a predictor
+adonis.dispRity(time_subsets, matrix ~ chrono.subsets)
+```
+
+### `geiger::dtt` model fitting in `dispRity`
+
+The `dtt` function from the `geiger` package is also often used to compare a trait's disparity observed in living taxa to the disparity of a simulated trait based on a given phylogeny.
+The `dispRity` package proposes a wrapper function for `geiger::dtt`, `dtt.dispRity` that allows the use of any disparity metric.
+Unfortunately, this implementation is slower that `geiger::dtt` (so if you're using the metrics implemented in `geiger` prefer the original version) and, as the original function, is limited to ultrametric trees (only living taxa!)...
+
+```{r}
+require(geiger)
+geiger_data <- get(data(geospiza))
+
+## Calculate the disparity of the dataset using the sum of variance
+dispRity_dtt <- dtt.dispRity(data = geiger_data$dat, metric = c(sum, variances),
+                             tree = geiger_data$phy, nsim = 100)
+
+## Plotting the results
+plot(dispRity_dtt, fig.width=8, fig.height=8)
+```
+
+Note that, like in the original `dtt` function, it is possible to change the evolutionary model (see `?geiger::sim.char` documentation).
+
+### null morphospace testing with `null.test`
+
+This test is equivalent to the test performed in @diaz2016global.
+It compares the disparity measured in the observed space to the disparity measured in a set of simulated spaces.
+These simulated spaces can be built with based on the hypothesis assumptions: for example, we can test whether our space is normal.
+
+```{r}
+set.seed(123)
+## A "normal" multidimensional space with 50 dimensions and 10 elements
+normal_space <- matrix(rnorm(1000), ncol = 50)
+
+## Calculating the disparity as the average pairwise distances
+obs_disparity <- dispRity(normal_space, metric = c(mean, pairwise.dist))
+
+## Testing against 100 randomly generated normal spaces
+(results <- null.test(obs_disparity, replicates = 100, null.distrib = rnorm))
+```
+
+Here the results show that disparity measured in our observed space is not significantly different than the one measured in a normal space.
+We can then propose that our observed space is normal!
+
+These results have an attributed `dispRity` and `randtest` class and can be plotted as `randtest` objects using the `dispRity` S3 `plot` method:
+
+```{r, fig.width=8, fig.height=8}
+## Plotting the results
+plot(results, main = "Is this space normal?")
+```
+
+For more details on generating spaces see the [`space.maker`](#Simulating-multidimensional-spaces) function tutorial.
+
+## Fitting modes of evolution to disparity data
+
+The code used for these models is based on those developed by Gene Hunt [@hunt2006fitting; @hunt2012measuring; @hunt2015simple].
+So we acknowledge and thank Gene Hunt for developing these models and writing the original R code that served as inspiration for these models.
+
+### Simple modes of disparity change through time
+
+#### `model.test`
+
+Changes in disparity-through-time can follow a range of models, such as random walks, stasis, constrained evolution, trends, or an early burst model of evolution.
+We will start with by fitting the simplest modes of evolution to our data.
+For example we may have a null expectation of time-invariant change in disparity in which values fluctuate with a variance around the mean - this would be best describe by a Stasis model:
+
+```{r}
+## Loading premade disparity data
+data(BeckLee_disparity)
+disp_time <- model.test(data = BeckLee_disparity, model = "Stasis")
+```
+
+We can see the standard output from ```model.test```.
+The first output message tells us it has tested for equal variances in each sample.
+The model uses Bartlett's test of equal variances to assess if variances are equal, so if p > 0.05 then variance is treated as the same for all samples, but if (p < 0.05) then each bin variance is unique.
+Here we have p < 0.05, so variance is not pooled between samples.
+
+By default ```model.test``` will use Bartlett's test to assess for homogeneity of variances, and then use this to decide to pool variances or not.
+This is ignored if the argument ```pool.variance``` in ```model.test``` is changed from the default ```NULL``` to ```TRUE``` or ```FALSE```.
+For example, to ignore Bartlett's test and pool variances manually we would do the following:
+
+```{r}
+disp_time_pooled <- model.test(data = BeckLee_disparity, model = "Stasis", pool.variance = TRUE)
+```
+
+However, unless you have good reason to choose otherwise it is recommended to use the default of ```pool.variance = NULL```:
+
+```{r}
+disp_time <- model.test(data = BeckLee_disparity, model = "Stasis", pool.variance = NULL)
+disp_time
+```
+
+The remaining output gives us the log-likelihood of the Stasis model of -59.501 (you may notice this change when we pooled variances above).
+The output also gives us the small sample Akaike Information Criterion (AICc), the delta AICc (the distance from the best fitting model), and the AICc weights (~the relative support of this model compared to all models, scaled to one).
+
+These are all metrics of relative fit, so when we test a single model they are not useful.
+By using the function summary in ```dispRity``` we can see the maximum likelihood estimates of the model parameters:
+
+```{r}
+summary(disp_time)
+```
+
+So we again see the AICc, delta AICc, AICc weight, and the log-likelihood we saw previously.
+We now also see the number of parameters from the model (2: theta and omega), and their estimates so the variance (omega = 0.01) and the mean (theta.1 = 3.4).
+
+The ```model.test``` function is designed to test relative model fit, so we need to test more than one model to make relative comparisons.
+So let's compare to the fit of the Stasis model to another model with two parameters: the Brownian motion.
+Brownian motion assumes a constant mean that is equal to the ancestral estimate of the sequence, and the variance around this mean increases linearly with time.
+The easier way to compare these models is to simply add `"BM"` to the ```models``` vector argument:
+
+```{r}
+disp_time <- model.test(data = BeckLee_disparity, model = c("Stasis", "BM"))
+disp_time
+```
+
+Et voilà! Here we can see by the log-likelihood, AICc, delta AICc, and AICc weight Brownian motion has a much better relative fit to these data than the Stasis model.
+Brownian motion has a relative AICc fit 366 units better than Stasis, and virtually has a AICc weight of 1.
+
+We can also all the information about the relative fit of models alongside the maximum likelihood estimates of model parameters using the summary function
+
+```{r}
+summary(disp_time)
+```
+
+Not that because the parameters per models differ, the summary includes `NA` for inapplicable parameters per models (e.g.
+the theta and omega parameters from the Stasis models are inapplicable for a Brownian motion model).
+
+We can plot the relative fit of our models using the ```plot``` function
+
+```{r plot1, fig.align="center", fig.cap = "relative fit (AICc weight) of Stasis and Brownian models of disparity through time", echo = TRUE, fig.height = 6, fig.width = 6}
+plot(disp_time)
+```
+
+Here we see and overwhelming support for the Brownian motion model.
+
+Alternatively, we could test all available models single modes: Stasis, Brownian motion, Ornstein-Uhlenbeck (evolution constrained to an optima), Trend (increasing or decreasing mean through time), and Early Burst (exponentially decreasing rate through time)
+
+```{r}
+disp_time <- model.test(data = BeckLee_disparity, model = c("Stasis", "BM", "OU", "Trend", "EB"))
+summary(disp_time)
+```
+
+These models indicate support for a Trend model, and we can plot the relative support of all model AICc weights
+
+```{r plot2, fig.align="center", fig.cap = "relative fit (AICc weight) of various modes of evolution", echo = TRUE, fig.height = 6, fig.width = 6}
+plot(disp_time)
+```
+
+Is this a trend of increasing or decreasing disparity through time? One way to find out is to look at the summary function for the Trend model:
+
+```{r}
+summary(disp_time)["Trend",]
+```
+
+This show a positive trend (0.01) of increasing disparity through time.
+
+### Plot and run simulation tests in a single step
+
+#### `model.test.wrapper`
+
+Patterns of evolution can be fit using ```model.test```, but the ```model.test.wrapper``` fits the same models as `model.test` as well as running predictive tests and plots.
+
+The predictive tests use the maximum likelihood estimates of model parameters to simulate a number of datasets (default = 1000), and analyse whether this is significantly different to the empirical input data using the Rank Envelope test [@murrell2018global].
+Finally we can plot the empirical data, simulated data, and the Rank Envelope test p values.
+This can all be done using the function ```model.test.wrapper```, and we will set the argument ```show.p = TRUE``` so _p_ values from the Rank Envelope test are printed on the plot:
+
+```{r plot3, fig.align="center", fig.cap = "Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for Trend, OU, BM, EB, and Stasis models", echo = TRUE, fig.height = 18, fig.width = 12}
+disp_time <- model.test.wrapper(data = BeckLee_disparity, model = c("Stasis", "BM", "OU", "Trend", "EB"),
+                                show.p = TRUE)
+disp_time
+```
+
+From this plot we can see the empirical estimates of disparity through time (pink) compared to the predictive data based upon the simulations using the estimated parameters from each model.
+There is no significant differences between the empirical data and simulated data, except for the Early Burst model.
+
+Trend is the best-fitting model but the plot suggests the OU model also follows a trend-like pattern.
+This is because the optima for the OU model (5.7) is different to the ancestral state 2.835 and outside the observed value.
+This is potentially unrealistic, and one way to alleviate this issue is to set the optima of the OU model to equal the ancestral estimate - this is the normal practice for OU models in comparative phylogenetics.
+To set the optima to the ancestral value we change the argument ```fixed.optima = TRUE```:
+
+```{r plot4, fig.align="center", fig.cap = "Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for Trend, OU, BM, EB, and Stasis models with the optima of the OU model set to equal the ancestral value", echo = TRUE, fig.height = 18, fig.width = 12}
+disp_time <- model.test.wrapper(data = BeckLee_disparity, model = c("Stasis", "BM", "OU", "Trend", "EB"),
+                                show.p = TRUE, fixed.optima = TRUE)
+disp_time
+```
+
+The relative fit of the OU model is decreased by constraining the fit of the optima to equal the ancestral state value.
+In fact as the OU attraction parameter (alpha) is zero, the model is equal to a Brownian motion model but is penalised by having an extra parameter. 
+Note that indeed, the plots of the BM model and the OU model look nearly identical.
+
+### Multiple modes of evolution (time shifts)
+
+As well as fitting a single model to a sequence of disparity values we can also allow for the mode of evolution to shift at a single or multiple points in time.
+The timing of a shift in mode can be based on an a prior expectation, such as a mass extinction event, or the model can test multiple points to allow to find time shift point with the highest likelihood.
+
+Models can be fit using `model.test` but it can be more convenient to use `model.test.wrapper`.
+Here we will compare the relative fit of Brownian motion, Trend, Ornstein-Uhlenbeck and a multi-mode Ornstein Uhlenbck model in which the optima changes at 66 million years ago, the Cretaceous-Palaeogene boundary.
+
+For example, we could be testing the hypothesis that the extinction of non-avian dinosaurs allowed mammals to go from scurrying in the undergrowth (low optima/low disparity) to dominating all habitats (high optima/high disparity).
+We will constrain the optima of OU model in the first time begin (i.e, pre-66 Mya) to equal the ancestral value:
+
+```{r plot5, fig.align="center", fig.cap = "Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for BM, Trend, OU, and multi OU models with a shift in optima allowed at 66 Ma", echo = TRUE, fig.height = 12, fig.width = 12}
+disp_time <- model.test.wrapper(data = BeckLee_disparity, model = c("BM", "Trend", "OU", "multi.OU"),
+                                time.split = 66, pool.variance = NULL, show.p = TRUE,
+                                fixed.optima = TRUE)
+disp_time
+```
+
+The multi-OU model shows an increase an optima at the Cretaceous-Palaeogene boundary, indicating a shift in disparity.
+However, this model does not fit as well as a model in which there is an increasing trend through time.
+We can also fit a model in which the we specify a heterogeneous model but we do not give a `time.split`.
+In this instance the model will test all splits that have at least 10 time slices on either side of the split.
+That's 102 potential time shifts in this example dataset so be warned, the following code will estimate 105 models!
+
+```{r, eval = FALSE}
+## An example of a time split model in which all potential splits are tested
+## WARNING: this will take between 20 minutes and half and hour to run!
+disp_time <- model.test.wrapper(data = BeckLee_disparity, model = c("BM", "Trend", "OU", "multi.OU"),
+                                show.p = TRUE, fixed.optima = TRUE)
+```
+
+As well as specifying a multi-OU model we can run any combination of models.
+For example we could fit a model at the Cretaceous-Palaeogene boundary that goes from an OU to a BM model, a Trend to an OU model, a Stasis to a Trend model or any combination you want to use.
+The only model that can't be used in combination is a multi-OU model.
+
+These can be introduced by changing the input for the models into a list, and supplying a vector with the two models.
+This is easier to see with an example:
+
+```{r plot6, fig.align="center", fig.cap = "Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for a variety of models with a shift in optima allowed at 66 Ma", echo = TRUE, fig.height = 18, fig.width = 12}
+## The models to test
+my_models <- list(c("BM", "OU"),
+                  c("Stasis", "OU"),
+                  c("BM", "Stasis"),
+                  c("OU", "Trend"),
+                  c("Stasis", "BM"))
+
+## Testing the models
+disp_time <- model.test.wrapper(data = BeckLee_disparity, model = my_models, time.split = 66,
+                                show.p = TRUE, fixed.optima = TRUE)
+disp_time
+```
+
+### `model.test.sim`
+
+Note that all the models above where run using the `model.test.wrapper` function that is a... wrapping function!
+In practice, this function runs two main functions from the `dispRity` package and then plots the results:
+
+ * `model.test` and
+ * `model.test.sim`
+
+The `model.test.sim` allows to simulate disparity evolution given a `dispRity` object input (as in `model.test.wrapper`) or given a model and its specification.
+For example, it is possible to simulate a simple Brownian motion model (or any of the other models or models combination described above):
+
+```{r}
+## A simple BM model
+model_simulation <- model.test.sim(sim = 1000, model = "BM", time.span = 50, variance = 0.1,
+                                   sample.size = 100, parameters = list(ancestral.state = 0))
+model_simulation
+```
+
+This will simulate 1000 Brownian motions for 50 units of time with 100 sampled elements, a variance of 0.1 and an ancestral state of 0.
+We can also pass multiple models in the same way we did it for `model.test`
+This model can then be summarised and plotted as most `dispRity` objects:
+
+```{r plot7, fig.align="center", fig.cap = "A simulated Brownian motion", echo = TRUE, fig.height = 6, fig.width = 6}
+## Displaying the 5 first rows of the summary
+head(summary(model_simulation))
+
+## Plotting the simulations
+plot(model_simulation)
+```
+
+Note that these functions can take all the arguments that can be passed to `plot`, `summary`, `plot.dispRity` and `summary.dispRity`.
+
+#### Simulating tested models
+
+Maybe more interestingly though, it is possible to pass the output of `model.test` directly to `model.test.sim` to simulate the models that fits the data the best and calculate the Rank Envelope test _p_ value.
+Let's see that using the simple example from the start:
+
+```{r}
+## Fitting multiple models on the data set
+disp_time <- model.test(data = BeckLee_disparity, model = c("Stasis", "BM", "OU", "Trend", "EB"))
+summary(disp_time)
+```
+
+As seen before, the Trend model fitted this dataset the best.
+To simulate what 1000 Trend models would look like using the same parameters as the ones estimated with `model.test` (here the ancestral state being 2.839, the sigma squared beeing 0.002 and the trend of 0.01), we can simply pass this model to `model.test.sim`:
+
+```{r}
+## Simulating 1000 Trend model with the observed parameters
+sim_trend <- model.test.sim(sim = 1000, model = disp_time)
+sim_trend
+```
+
+By default, the model simulated is the one with the lowest AICc (`model.rank = 1`) but it is possible to choose any ranked model, for example, the OU (second one):
+
+```{r}
+## Simulating 1000 OU model with the observed parameters
+sim_OU <- model.test.sim(sim = 1000, model = disp_time, model.rank = 2)
+sim_OU
+```
+
+And as the example above, the simulated data can be plotted or summarised:
+
+```{r}
+head(summary(sim_trend))
+head(summary(sim_OU))
+```
+
+```{r plot8, fig.align="center", fig.cap = "The best fitted model (Trend) and the observed disparity through time", echo = TRUE, fig.height = 6, fig.width = 6}
+## The trend model with some graphical options
+plot(sim_trend, xlab = "Time (Mya)", ylab = "sum of variances",
+    col = c("#F65205", "#F38336", "#F7B27E"))
+
+## Adding the observed disparity through time
+plot(BeckLee_disparity, add = TRUE, col = c("#3E9CBA", "#98D4CF90", "#BFE4E390"))
+```
 
@@
 
-The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;OReilly2017].
+The protocol implemented here to generate discrete morphological matrices is based on the ones developed in @GuillermeCooper @OReilly2016 @puttick2017uncertain and @OReilly2017.
 
  * The first `tree` argument will be the tree on which to "evolve" the characters and therefore requires branch length.
- You can generate quick and easy random Yule trees using `ape::rtree(number_of_taxa)` but I would advise to use more realistic trees for more realistic simulations based on more realistic models (really realistic then) using the function `tree.bd` from the [`diversitree`](http://www.zoology.ubc.ca/prog/diversitree/) package [@fitzjohndiversitree2012].
- * The second argument, `character` is the number of characters. Pretty straight forward.
- * The third, `states` is the proportion of characters states above two (yes, the minimum number of states is two).
+ You can generate quick and easy random Yule trees using `ape::rtree(number_of_taxa)` but I would advise to use more realistic trees for more realistic simulations based on more realistic models (really realistic then) using the function `tree.bd` from the [`diversitree`](http://www.zoology.ubc.ca/prog/diversitree/) package [@fitzjohndiversitree2012] for example.
+ * The second argument, `character` is the number of characters.
+ * The third, `states` is the proportion of characters states above two.
 This argument intakes the proportion of *n*-states characters, for example `states = c(0.5,0.3,0.2)` will generate 50% of binary-state characters, 30% of three-state characters and 20% of four-state characters.
 There is no limit in the number of state characters proportion as long as the total makes up 100%.
- * The forth, `model` is the evolutionary model for generating the character(s). More about this below.
+ * The fourth, `model` is the evolutionary model for generating the character(s). More about this below.
  * The fifth and sixth, `rates` and `substitution` are the model parameters described below as well.
@@
                                  scree = c(0.7, 0.2, 0.1))
-```
+head(cylindrical_space)
+```
+
+See below for the visualisation of the spaces.
 
@@
      arguments = list(list(mean = 0), list(runif, inner = 0.5, outer = 1)))
-```
+head(doughnut_space)
+```
+
+See below for the visualisation of the "doughnut" space.
 
@@
 
- * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"punctuated"`, `"gradual"`).
-This element generated only once via `time.subsets()` and `custom.subsets()`.
+ * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"equal.split"`, `"gradual.split"`).
+This element generated only once via `chrono.subsets()` and `custom.subsets()`.
  * `$call$dimensions`: either a single `numeric` value indicating how many dimensions to use or a vector of `numeric` values indicating which specific dimensions to use.
@@
 <!--chapter:end:08_future_directions.Rmd-->
+
+`r if (knitr:::is_html_output()) '# References {-}'`
+
+```{r include=FALSE}
+# generate a BibTeX database automatically for some R packages
+knitr::write_bib(c(
+  .packages(), 'bookdown', 'knitr', 'rmarkdown'
+), 'packages.bib')
+```
+
+<!--chapter:end:09_references.Rmd-->

### 2018-09-27 inst/gitbook/dispRity_manual.tex
Updated vignette

\documentclass[]{book}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={dispRity manual},
            pdfauthor={Thomas Guillerme (guillert@tcd.ie), Mark Puttick (marknputtick@gmail.com) and Natalie Cooper (natalie.cooper@nhm.ac.uk)},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{dispRity manual}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Thomas Guillerme
(\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}), Mark
Puttick
(\href{mailto:marknputtick@gmail.com}{\nolinkurl{marknputtick@gmail.com}})
and Natalie Cooper
(\href{mailto:natalie.cooper@nhm.ac.uk}{\nolinkurl{natalie.cooper@nhm.ac.uk}})}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{2018-09-27}

\usepackage{booktabs}

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}{Lemma}[chapter]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{corollary}{Corollary}[chapter]
\newtheorem{proposition}{Proposition}[chapter]
\theoremstyle{definition}
\newtheorem{example}{Example}[chapter]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[chapter]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\maketitle

{
\setcounter{tocdepth}{1}
\tableofcontents
}
\chapter{\texorpdfstring{\texttt{dispRity}}{dispRity}}\label{disprity}

This is a package for measuring disparity in \texttt{R}. It allows users
to summarise matrices as representations as multidimensional spaces into
a single value or distribution describing a specific aspect of this
multidimensional space (the disparity). Multidimensional spaces can be
ordinated matrices from MDS, PCA, PCO, PCoA but the package is
\emph{not} restricted to any type of matrices! This manual is based on
the version 1.2.

\section{\texorpdfstring{What is
\texttt{dispRity}?}{What is dispRity?}}\label{what-is-disprity}

This is a modular package for measuring disparity in \texttt{R}. It
allows users to summarise ordinated matrices (e.g.~MDS, PCA, PCO, PCoA)
to perform some multidimensional analysis. Typically, these analysis are
used in palaeobiology and evolutionary biology to study the changes in
morphology through time. However, there are many more applications in
ecology, evolution and beyond.

\subsection{Modular?}\label{modular}

Because their exist a multitude of ways to measure disparity, each
adapted to every specific question, this package uses an easy to modify
modular architecture. In coding, each module is simply a function or a
modification of a function that can be passed to the main functions of
the package to tweak it to your proper needs! In practice, you will
notice throughout this manual that some function can take other
functions as arguments: the modular architecture of this package allows
you to use any function for these arguments (with some restrictions
explained for each specific cases). This will allow you to finely tune
your multidimensional analysis to the needs of your specific question!

\section{Installing and running the
package}\label{installing-and-running-the-package}

You can install this package easily, directly from the CRAN:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{install.packages}\NormalTok{(}\StringTok{"dispRity"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Alternatively, for the most up to data version and some functionalities
not compatible with the CRAN, you can use the package through GitHub
using \texttt{devtool} (see \protect\hyperlink{noCRAN}{to CRAN or not to
CRAN?} for more details):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Checking if devtools is already installed}
\ControlFlowTok{if}\NormalTok{(}\OperatorTok{!}\KeywordTok{require}\NormalTok{(devtools)) }\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}

\NormalTok{## Installing the latest released version directly from GitHub}
\KeywordTok{install_github}\NormalTok{(}\StringTok{"TGuillerme/dispRity"}\NormalTok{, }\DataTypeTok{ref =} \StringTok{"release"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Anyways to upload it, you can use

Note this uses the \texttt{release} branch (1.2). For the piping-hot
(but potentially unstable) version, you can change the argument
\texttt{ref\ =\ release} to \texttt{ref\ =\ master}. \texttt{dispRity}
depends mainly on the \texttt{ape} package and uses functions from
several other packages (\texttt{ade4}, \texttt{geometry},
\texttt{grDevices}, \texttt{hypervolume}, \texttt{paleotree},
\texttt{snow}, \texttt{Claddis}, \texttt{geomorph} and \texttt{RCurl}).

\hypertarget{noCRAN}{\section{To CRAN or not to CRAN?}\label{noCRAN}}

This package can be installed via
\href{https://cran.r-project.org/package=dispRity}{the CRAN} or via
\href{https://github.com/TGuillerme/dispRity}{GitHub}.

\begin{itemize}
\item
  The CRAN version is a more ``strict'' version of the package that does
  not include some functionalities (such as
  \protect\hyperlink{Claddis-ordination}{\texttt{Claddis.ordination}} or
  this gitbook vignette) because of compatibility issues with the CRAN
  requirement.
\item
  The GitHub version is a more ``reactive'' version (fed by user's
  suggestions and comments) and can contain more functionalities (see
  list below). This version exists on a \texttt{"release"} branch which
  is tested as strictly as the CRAN version with Travis and codecov. The
  \texttt{"master"} branch contains always the latest ``piping hot'',
  not yet release version. It might (though rarely) be bugged.
\end{itemize}

Note however, that apart from the functions listed below, both versions
of the package have the same tested and implemented functions.

Functions not available on the CRAN version (1.2):

\begin{itemize}
\tightlist
\item
  \protect\hyperlink{Claddis-ordination}{\texttt{Claddis.ordination}}
\end{itemize}

\subsection{So which version do I choose?}\label{version}

There are always three version of the package available:

\begin{itemize}
\tightlist
\item
  The CRAN one
\item
  The GitHub \texttt{release} one
\item
  The GitHub \texttt{master} one
\end{itemize}

The differences between the CRAN one and the GitHub \texttt{release} or
\texttt{master} ones is explained just above. For the the GitHub
version, the differences are that the \texttt{release} one is more
stable (i.e.~more rarely modified) and the \texttt{master} one is more
live one (i.e.~bug fixes and new functionalities are added as they
come).

If you want the latest-latest version of the package I suggest using the
GitHub \texttt{master} one, especially if you recently emailed me
reporting a minor bug or wanting a new functionality! Note however that
\emph{it can happen} that the \texttt{master} version can sometimes be
bugged (especially when there are major \texttt{R} and \texttt{R}
packages updates), however, the status of the package state on both the
\texttt{release} and the \texttt{master} version is constantly displayed
on the \href{https://github.com/TGuillerme/dispRity/}{\texttt{README}
page} of the package with the nice badges displaying these different
(and constantly tested) information

\begin{itemize}
\tightlist
\item
  The build status:
  \href{https://travis-ci.org/TGuillerme/dispRity}{\includegraphics{https://travis-ci.org/TGuillerme/dispRity.svg?branch=master}}
  which tests whether the package compiles (i.e.~can be installed on
  your computer).
\item
  The code coverage:
  \href{https://codecov.io/gh/TGuillerme/dispRity}{\includegraphics{https://codecov.io/gh/TGuillerme/dispRity/branch/master/graph/badge.svg}}
  which indicates the percentage of lines in the code that have been
  tested for expected behaviour.
\item
  The project status:
  \href{http://www.repostatus.org/\#active}{\includegraphics{http://www.repostatus.org/badges/latest/active.svg}}
  which indicates whether this package is maintained regularly or
  forgotten in the abyss of internet.
\item
  The development version:
  \href{https://github.com/TGuillerme/dispRity}{\includegraphics{https://img.shields.io/badge/devel\%20version-1.2.1-green.svg?style=flat}}
  which indicates which version of the package you are about to install.
\end{itemize}

Note that when we (the \texttt{dispRity} package collaborators!) are
developing some more major versions, including some more complex
features for example, they will never be released on either the the
versions described above. Therefore you can rest assure that you will
never use a badly bugged version of \texttt{dispRity}! Enjoy!

\section{Help}\label{help}

If you need help with the package, hopefully the following manual will
be useful. However, parts of this package are still in development and
some other parts are probably not covered. Thus if you have suggestions
or comments on on what has already been developed or will be developed,
please send me an email
(\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}) or if you
are a GitHub user, directly create an issue on the
\href{https://github.com/TGuillerme/dispRity}{GitHub page}.

\section{Citations}\label{citations}

To cite the package, this manual or some specific functionalities, you
can use the following references:

The package main paper:

\begin{quote}
Guillerme T. dispRity: A modular R package for measuring disparity.
Methods Ecol Evol. 2018;9:1755--1763.
\href{https://doi.org/10.1111/2041-210X.13022}{doi.org/10.1111/2041-210X.13022}.
\end{quote}

The package manual (regularly updated!):

\begin{quote}
Guillerme, T. \& Cooper, N. (2018): dispRity manual. figshare. Preprint.
\href{https://doi.org/10.6084/m9.figshare.6187337.v1}{10.6084/m9.figshare.6187337.v1}.
\end{quote}

The time-slicing method implemented in \texttt{chrono.subsets}
(unfortunately not Open Access, but you can still get a
\href{https://tguillerme.github.io/papers/Guillerme\&Cooper-2018-Palaeontology.pdf}{free
copy from here}):

\begin{quote}
Guillerme, T. and Cooper, N. (2018), Time for a rethink: time
sub-sampling methods in disparity-through-time analyses. Palaeontology,
61: 481-493.
\href{https://onlinelibrary.wiley.com/doi/abs/10.1111/pala.12364}{doi:10.1111/pala.12364}.
\end{quote}

\chapter{Glossary}\label{glossary}

\begin{itemize}
\item
  \textbf{Multidimensional space} (or just space). The mathematical
  multidimensional object that will be analysed with this package. In
  morphometrics, this is often referred to as the morphospace. However
  it may also be referred to as the cladisto-space for cladistic data or
  the eco-space for ecological data etc. In practice, this term
  designates a matrix where the columns represent the dimensions of the
  space (often -- but not necessarily - \textgreater{} 3!) and the rows
  represent the elements within this space.
\item
  \textbf{Elements}. The rows of the multidimensional space matrix.
  Elements can be taxa, field sites, countries etc.
\item
  \textbf{Dimensions}. The columns of the multidimensional space matrix.
  The dimensions can be referred to as axes of variation, or principal
  components, for ordinated spaces obtained from a PCA for example.
\item
  \textbf{Subsets}. Subsets of the multidimensional space. A subset (or
  subsets) contains the same number of dimensions as the space but may
  contain a smaller subset of elements. For example, if our space is
  composed of birds and mammals (the elements) and 50 principal
  components of variation (the dimensions), we can create two subsets
  containing just mammals or birds, but with the same 50 dimensions, to
  compare disparity in the two clades.
\item
  \textbf{Disparity}. A metric expressing the
  similarities/dissimilarities of the elements within the space or a
  summarising the space dimensions. For example the pairwise distances
  between elements or the range of each dimensions.
\end{itemize}

\section{Glossary equivalences in palaeobiology and
ecology}\label{glossary-equivalences-in-palaeobiology-and-ecology}

\begin{longtable}[]{@{}llll@{}}
\toprule
\begin{minipage}[b]{0.20\columnwidth}\raggedright\strut
In this manual\strut
\end{minipage} & \begin{minipage}[b]{0.20\columnwidth}\raggedright\strut
In \texttt{dispRity}\strut
\end{minipage} & \begin{minipage}[b]{0.23\columnwidth}\raggedright\strut
E.g. in palaeobiology\strut
\end{minipage} & \begin{minipage}[b]{0.17\columnwidth}\raggedright\strut
E.g. in ecology\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
the multidimensional space\strut
\end{minipage} & \begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
a \texttt{matrix} object (\(n\times k\))\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
a morphospace\strut
\end{minipage} & \begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
a function-space\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
elements\strut
\end{minipage} & \begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
rows (\(n\))\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
taxa\strut
\end{minipage} & \begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
field experiments\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
dimensions\strut
\end{minipage} & \begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
columns (\(k\))\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
morphological characters\strut
\end{minipage} & \begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
communities' compositions\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
subsets\strut
\end{minipage} & \begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
a \texttt{matrix} (\(m \times k\), with \(m \leq n\))\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
time series\strut
\end{minipage} & \begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
experimental treatments\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
disparity\strut
\end{minipage} & \begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
a \texttt{function}\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
sum of variances\strut
\end{minipage} & \begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
ellipsoid volume\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

\chapter{\texorpdfstring{Getting started with
\texttt{dispRity}}{Getting started with dispRity}}\label{getting-started-with-disprity}

\section{\texorpdfstring{What sort of data does \texttt{dispRity} work
with?}{What sort of data does dispRity work with?}}\label{what-sort-of-data-does-disprity-work-with}

\textbf{Any matrix object in \texttt{R}.} Disparity can be estimated
from pretty much any matrix as long as rows represent the elements and
columns the dimensions. These matrices can be observations, pairwise
differences between elements, ordinations, etc\ldots{}

\begin{quote}
Note that whether the matrix is an ordinated one or of any other type,
it is always considered as the multidimensional space to analysis and no
corrections are applied to it (e.g.~if ordinated and with negative eigen
values, no correction is used). Always make sure the input is the right
multidimensional space to analyse!
\end{quote}

\section{Ordinated matrices}\label{ordinated-matrices}

Classically, when a high number of variables is used, disparity is
calculated from ordinated matrices. These can be any type of ordinations
(PCO, PCA, PCoA, MDS, etc.) as long as elements are the rows (taxa,
countries, field experiments) and the dimensions are the columns.

\subsection{\texorpdfstring{Ordination matrices from
\texttt{geomorph}}{Ordination matrices from geomorph}}\label{ordination-matrices-from-geomorph}

You can also easily use data from \texttt{geomorph} using the
\texttt{geomorph.ordination} function. This function simply takes
Procrustes aligned data and performs an ordination:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{require}\NormalTok{(geomorph)}

\NormalTok{## Loading the plethodon dataset}
\KeywordTok{data}\NormalTok{(plethodon)}

\NormalTok{## Performing a Procrustes transform on the landmarks}
\NormalTok{procrustes <-}\StringTok{ }\KeywordTok{gpagen}\NormalTok{(plethodon}\OperatorTok{$}\NormalTok{land, }\DataTypeTok{PrinAxes =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{print.progress =} \OtherTok{FALSE}\NormalTok{)}

\NormalTok{## Ordinating this data}
\KeywordTok{geomorph.ordination}\NormalTok{(procrustes)[}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{,}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                PC1        PC2           PC3          PC4          PC5
## [1,] -0.0369931363 0.05118247 -0.0016971082 -0.003128809 -0.010936371
## [2,] -0.0007493738 0.05942082  0.0001371715 -0.002768680 -0.008117383
## [3,]  0.0056004654 0.07419599 -0.0052612103 -0.005034566 -0.002746592
## [4,] -0.0134808572 0.06463959 -0.0458436015 -0.007887369  0.009816827
## [5,] -0.0334696244 0.06863518  0.0136292041  0.007359409  0.022347225
\end{verbatim}

Options for the ordination (from \texttt{?prcomp}) can be directly
passed to this function to perform customised ordinations. Additionally
you can give the function a \texttt{geomorph.data.frame} object. If the
latter contains sorting information (i.e.~factors), they can be directly
used to make a customised \texttt{dispRity} object
\protect\hyperlink{customised-subsets}{customised \texttt{dispRity}
object}!

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Using a geomorph.data.frame}
\NormalTok{geomorph_df <-}\StringTok{ }\KeywordTok{geomorph.data.frame}\NormalTok{(procrustes,}
     \DataTypeTok{species =}\NormalTok{ plethodon}\OperatorTok{$}\NormalTok{species, }\DataTypeTok{site =}\NormalTok{ plethodon}\OperatorTok{$}\NormalTok{site)}

\NormalTok{## Ordinating this data and making a dispRity object}
\KeywordTok{geomorph.ordination}\NormalTok{(geomorph_df)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 4 customised subsets for 40 elements:
##     species.Jord, species.Teyah, site.Allo, site.Symp.
\end{verbatim}

More about these \texttt{dispRity} objects below!

\hypertarget{Claddis-ordination}{\subsection{\texorpdfstring{Ordination
matrices from
\texttt{Claddis}}{Ordination matrices from Claddis}}\label{Claddis-ordination}}

\begin{quote}
This functionality is not available on the CRAN version of the package
(1.2) due to CRAN compatibility issues see \protect\hyperlink{noCRAN}{to
CRAN or not to CRAN?} for more information and on how to install the
GitHub (more complete) version of the package (1.2).
\end{quote}

\texttt{dispRity} package can easily take data from \texttt{Claddis}
using the \texttt{Claddis.ordination} function. For this, simply input a
matrix in the \texttt{Claddis} format to the function and it will
automatically calculate and ordinate the distances among taxa:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{require}\NormalTok{(Claddis)}

\NormalTok{## Ordinating the example data from Claddis}
\KeywordTok{Claddis.ordination}\NormalTok{(Michaux1989) }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                      [,1]          [,2]       [,3]
## Ancilla      0.000000e+00  4.154578e-01  0.2534942
## Turrancilla -5.106645e-01  1.956921e-16 -0.2534942
## Ancillista   5.106645e-01  6.523071e-17 -0.2534942
## Amalda      -2.004476e-16 -4.154578e-01  0.2534942
\end{verbatim}

Note that several options are available, namely which type of distance
should be computed. See more info in the function manual
(\texttt{?Claddis.ordination}). Alternatively, it is of course also
possible to manual calculate the ordination matrix using the functions
\texttt{Claddis::MorphDistMatrix} and \texttt{stats::cmdscale}.

\subsection{Other kinds of ordination
matrices}\label{other-kinds-of-ordination-matrices}

If you are not using the packages mentioned above (\texttt{Claddis} and
\texttt{geomorph}) you can easily make your own ordination matrices by
using the following functions from the \texttt{stats} package. Here is
how to do it for the following types of matrices:

\begin{itemize}
\tightlist
\item
  Multivariate matrices (principal components analysis; PCA)
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## A multivariate matrix}
\KeywordTok{head}\NormalTok{(USArrests)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##            Murder Assault UrbanPop Rape
## Alabama      13.2     236       58 21.2
## Alaska       10.0     263       48 44.5
## Arizona       8.1     294       80 31.0
## Arkansas      8.8     190       50 19.5
## California    9.0     276       91 40.6
## Colorado      7.9     204       78 38.7
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Ordinating the matrix using `prcomp` }
\NormalTok{ordination <-}\StringTok{ }\KeywordTok{prcomp}\NormalTok{(USArrests)}

\NormalTok{## Selecting the ordinated matrix}
\NormalTok{ordinated_matrix <-}\StringTok{ }\NormalTok{ordination}\OperatorTok{$}\NormalTok{x}
\KeywordTok{head}\NormalTok{(ordinated_matrix)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                  PC1        PC2        PC3        PC4
## Alabama     64.80216 -11.448007 -2.4949328 -2.4079009
## Alaska      92.82745 -17.982943 20.1265749  4.0940470
## Arizona    124.06822   8.830403 -1.6874484  4.3536852
## Arkansas    18.34004 -16.703911  0.2101894  0.5209936
## California 107.42295  22.520070  6.7458730  2.8118259
## Colorado    34.97599  13.719584 12.2793628  1.7214637
\end{verbatim}

This results in a ordinated matrix with US states as elements and four
dimensions (PC 1 to 4). For an alternative method, see the
\texttt{?princomp} function.

\begin{itemize}
\tightlist
\item
  Distance matrices (classical multidimensional scaling; MDS)
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## A matrix of distances between cities}
\KeywordTok{str}\NormalTok{(eurodist)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  'dist' num [1:210] 3313 2963 3175 3339 2762 ...
##  - attr(*, "Size")= num 21
##  - attr(*, "Labels")= chr [1:21] "Athens" "Barcelona" "Brussels" "Calais" ...
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Ordinating the matrix using cmdscale() with k = 5 dimensions }
\NormalTok{ordinated_matrix <-}\StringTok{ }\KeywordTok{cmdscale}\NormalTok{(eurodist, }\DataTypeTok{k =} \DecValTok{5}\NormalTok{)}
\KeywordTok{head}\NormalTok{(ordinated_matrix)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                 [,1]      [,2]       [,3]       [,4]       [,5]
## Athens    2290.27468 1798.8029   53.79314 -103.82696 -156.95511
## Barcelona -825.38279  546.8115 -113.85842   84.58583  291.44076
## Brussels    59.18334 -367.0814  177.55291   38.79751  -95.62045
## Calais     -82.84597 -429.9147  300.19274  106.35369 -180.44614
## Cherbourg -352.49943 -290.9084  457.35294  111.44915 -417.49668
## Cologne    293.68963 -405.3119  360.09323 -636.20238  159.39266
\end{verbatim}

This results in a ordinated matrix with European cities as elements and
five dimensions.

Of course any other method for creating the ordination matrix is totally
valid, you can also not use any ordination at all! The only requirements
for the \texttt{dispRity} functions is that the input is a matrix with
elements as rows and dimensions as columns.

\section{Performing a simple dispRity analysis}\label{simpleanalysis}

Two \texttt{dispRity} functions allow users to run an analysis pipeline
simply by inputting an ordination matrix. These functions allow users to
either calculate the disparity through time
(\texttt{dispRity.through.time}) or the disparity of user-defined groups
(\texttt{dispRity.per.group}).

\textbf{IMPORTANT}

Note that \texttt{disparity.through.time} and
\texttt{disparity.per.group} are wrapper functions (i.e.~they
incorporate lots of other functions) that allow users to run a basic
disparity-through-time, or disparity among groups, analysis without too
much effort. As such they use a lot of default options. These are
described in the help files for the functions that are used to make the
wrapper functions, and not described in the help files for
\texttt{disparity.through.time} and \texttt{disparity.per.group}. These
defaults are good enough for \textbf{data exploration}, but for a proper
analysis you should consider the \textbf{best parameters for your
question and data}. For example, which metric should you use? How many
bootstraps do you require? What model of evolution is most appropriate
if you are time slicing? Should you rarefy the data? See
\protect\hyperlink{time-slicing}{\texttt{chrono.subsets}},
\protect\hyperlink{customised-subsets}{\texttt{custom.subsets}},
\protect\hyperlink{bootstraps-and-rarefactions}{\texttt{boot.matrix}}
and \protect\hyperlink{disparity-metrics}{\texttt{dispRity.metric}} for
more details of the defaults used in each of these functions. Note that
any of these default arguments can be changed within the
\texttt{disparity.through.time} or \texttt{disparity.per.group}
functions.

\hypertarget{example-data}{\subsection{Example
data}\label{example-data}}

To illustrate these functions, we will use data from
\citet{beckancient2014}. This dataset contains an ordinated matrix of 50
discrete characters from mammals (\texttt{BeckLee\_mat50}), another
matrix of the same 50 mammals and the estimated discrete data characters
of their descendants (thus 50 + 49 rows, \texttt{BeckLee\_mat99}), a
dataframe containing the ages of each taxon in the dataset
(\texttt{BeckLee\_ages}) and finally a phylogenetic tree with the
relationships among the 50 mammals (\texttt{BeckLee\_tree}).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading the ordinated matrices}
\KeywordTok{data}\NormalTok{(BeckLee_mat50)}
\KeywordTok{data}\NormalTok{(BeckLee_mat99)}

\NormalTok{## The first five taxa and dimensions of the 50 taxa matrix}
\KeywordTok{head}\NormalTok{(BeckLee_mat50[, }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{])}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                    [,1]          [,2]        [,3]       [,4]      [,5]
## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The first five taxa and dimensions of the 99 taxa + ancestors matrix}
\NormalTok{BeckLee_mat99[}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{98}\NormalTok{, }\DecValTok{99}\NormalTok{), }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                   [,1]       [,2]       [,3]        [,4]        [,5]
## Cimolestes -0.60824375 -0.0323683 0.08458885 -0.43384481 -0.30536875
## Maelestes  -0.57302058 -0.2840361 0.01308847 -0.12588477  0.06123611
## n48        -0.05529018  0.4799330 0.04118477  0.04944912 -0.35588301
## n49        -0.13067785  0.4478168 0.11956268  0.13800340 -0.32227852
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading a list of first and last occurrence dates for the fossils}
\KeywordTok{data}\NormalTok{(BeckLee_ages)}
\KeywordTok{head}\NormalTok{(BeckLee_ages)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##             FAD  LAD
## Adapis     37.2 36.8
## Asioryctes 83.6 72.1
## Leptictis  33.9 33.3
## Miacis     49.0 46.7
## Mimotona   61.6 59.2
## Notharctus 50.2 47.0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading and plotting the phylogeny}
\KeywordTok{data}\NormalTok{(BeckLee_tree)}
\KeywordTok{plot}\NormalTok{(BeckLee_tree, }\DataTypeTok{cex =} \FloatTok{0.8}\NormalTok{) }
\KeywordTok{axisPhylo}\NormalTok{(}\DataTypeTok{root =} \DecValTok{140}\NormalTok{)}
\KeywordTok{nodelabels}\NormalTok{(}\DataTypeTok{cex =} \FloatTok{0.5}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-12-1.pdf}

Of course you can use your own data as detailed in the
\protect\hyperlink{What-sort-of-data-does-dispRity-work-with}{previous
chapter}.

\subsection{Disparity through time}\label{disparity-through-time}

The \texttt{dispRity.through.time} function calculates disparity through
time, a common analysis in palaeontology. This function (and the
following one) uses an analysis pipeline with a lot of default
parameters to make the analysis as simple as possible. Of course all the
defaults can be changed if required, more on this later.

For a disparity through time analysis, you will need:

\begin{itemize}
\tightlist
\item
  An ordinated matrix (we covered that above)
\item
  A phylogenetic tree: this must be a \texttt{phylo} object (from the
  \texttt{ape} package) and needs a \texttt{root.time} element. To give
  your tree a root time (i.e.~an age for the root), you can simply
  do\textbackslash{}
  \texttt{my\_tree\$root.time\ \textless{}-\ my\_age}.
\item
  The required number of time subsets (here \texttt{time\ =\ 3})
\item
  Your favourite disparity metric (here the sum of variances)
\end{itemize}

Using the Beck and Lee (2014) data described
\protect\hyperlink{example-data}{above}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring disparity through time}
\NormalTok{disparity_data <-}\StringTok{ }\KeywordTok{dispRity.through.time}\NormalTok{(BeckLee_mat50, BeckLee_tree,}
                                        \DataTypeTok{time =} \DecValTok{3}\NormalTok{, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances))}
\end{Highlighting}
\end{Shaded}

This generates a \texttt{dispRity} object (see
\protect\hyperlink{guts}{here} for technical details). When displayed,
these \texttt{dispRity} objects provide us with information on the
operations done to the matrix:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Print the disparity_data object}
\NormalTok{disparity_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 3 discrete time subsets for 50 elements with 48 dimensions:
##     133.51104 - 89.00736, 89.00736 - 44.50368, 44.50368 - 0.
## Data was bootstrapped 100 times (method:"full").
## Disparity was calculated as: metric.
\end{verbatim}

We asked for three subsets (evenly spread across the age of the tree),
the data was bootstrapped 100 times (default) and the metric used was
the sum of variances.

We can now summarise or plot the \texttt{disparity\_data} object, or
perform statistical tests on it (e.g.~a simple \texttt{lm}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Summarising disparity through time}
\KeywordTok{summary}\NormalTok{(disparity_data)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                subsets  n   obs bs.median  2.5%   25%   75% 97.5%
## 1 133.51104 - 89.00736  5 1.575     1.305 0.729 1.118 1.420 1.509
## 2  89.00736 - 44.50368 29 1.922     1.867 1.775 1.830 1.889 1.922
## 3         44.50368 - 0 16 1.990     1.871 1.716 1.831 1.914 1.942
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Plotting the results}
\KeywordTok{plot}\NormalTok{(disparity_data, }\DataTypeTok{type =} \StringTok{"continuous"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-15-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Testing for an difference among the time bins}
\NormalTok{disp_lm <-}\StringTok{ }\KeywordTok{test.dispRity}\NormalTok{(disparity_data, }\DataTypeTok{test =}\NormalTok{ lm, }\DataTypeTok{comparisons =} \StringTok{"all"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in test.dispRity(disparity_data, test = lm, comparisons = "all"): Multiple p-values will be calculated without adjustment!
## This can inflate Type I error!
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(disp_lm)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Call:
## test(formula = data ~ subsets, data = data)
## 
## Residuals:
##      Min       1Q   Median       3Q      Max 
## -0.56623 -0.04160  0.01049  0.05507  0.31886 
## 
## Coefficients:
##                            Estimate Std. Error t value Pr(>|t|)    
## (Intercept)                 1.25647    0.01270   98.97   <2e-16 ***
## subsets44.50368 - 0         0.60863    0.01795   33.90   <2e-16 ***
## subsets89.00736 - 44.50368  0.60169    0.01795   33.51   <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 0.127 on 297 degrees of freedom
## Multiple R-squared:  0.8361, Adjusted R-squared:  0.835 
## F-statistic: 757.5 on 2 and 297 DF,  p-value: < 2.2e-16
\end{verbatim}

Please refer to the \protect\hyperlink{specific-tutorial}{specific
tutorials} for (much!) more information on the nuts and bolts of the
package. You can also directly explore the specific function help files
within R and navigate to related functions.

\hypertarget{disparity-among-groups}{\subsection{Disparity among
groups}\label{disparity-among-groups}}

The \texttt{dispRity.per.group} function is used if you are interested
in looking at disparity among groups rather than through time. For
example, you could ask if there is a difference in disparity between two
groups?

To perform such an analysis, you will need:

\begin{itemize}
\tightlist
\item
  An matrix with rows as elements and columns as dimensions (always!)
\item
  A list of group members: this list should be a list of numeric vectors
  or names corresponding to the row names in the matrix. For example
  \texttt{list("a"\ =\ c(1,2),\ "b"\ =\ c(3,4))} will create a group
  \emph{a} containing elements 1 and 2 from the matrix and a group
  \emph{b} containing elements 3 and 4. Note that elements can be
  present in multiple groups at once.
\item
  Your favourite disparity metric (here the sum of variances)
\end{itemize}

Using the \citet{beckancient2014} data described
\protect\hyperlink{example-data}{above}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating the two groups (crown versus stem) as a list}
\NormalTok{mammal_groups <-}\StringTok{ }\KeywordTok{crown.stem}\NormalTok{(BeckLee_tree, }\DataTypeTok{inc.nodes =} \OtherTok{FALSE}\NormalTok{)}

\NormalTok{## Measuring disparity for each group}
\NormalTok{disparity_data <-}\StringTok{ }\KeywordTok{dispRity.per.group}\NormalTok{(BeckLee_mat50, }\DataTypeTok{group =}\NormalTok{ mammal_groups,}
                                     \DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances))}
\end{Highlighting}
\end{Shaded}

We can display the disparity of both groups by simply looking at the
output variable (\texttt{disparity\_data}) and then summarising the
\texttt{disparity\_data} object and plotting it, and/or by performing a
statistical test to compare disparity across the groups (here a Wilcoxon
test).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Print the disparity_data object}
\NormalTok{disparity_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 2 customised subsets for 50 elements with 48 dimensions:
##     crown, stem.
## Data was bootstrapped 100 times (method:"full").
## Disparity was calculated as: metric.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Summarising disparity in the different groups}
\KeywordTok{summary}\NormalTok{(disparity_data)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs bs.median  2.5%   25%   75% 97.5%
## 1   crown 30 1.995     1.933 1.883 1.917 1.951 1.983
## 2    stem 20 1.715     1.635 1.535 1.610 1.664 1.691
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Plotting the results}
\KeywordTok{plot}\NormalTok{(disparity_data)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-17-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Testing for a difference between the groups}
\KeywordTok{test.dispRity}\NormalTok{(disparity_data, }\DataTypeTok{test =}\NormalTok{ wilcox.test, }\DataTypeTok{details =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## $`crown : stem`
## $`crown : stem`[[1]]
## 
##  Wilcoxon rank sum test with continuity correction
## 
## data:  dots[[1L]][[1L]] and dots[[2L]][[1L]]
## W = 10000, p-value < 2.2e-16
## alternative hypothesis: true location shift is not equal to 0
\end{verbatim}

\chapter{Details of specific
functions}\label{details-of-specific-functions}

The following section contains information specific to some functions.
If any of your questions are not covered in these sections, please refer
to the function help files in R, send me an email
(\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}), or raise
an issue on GitHub. The several tutorials below describe specific
functionalities of certain functions; please always refer to the
function help files for the full function documentation!

Before each section, make sure you loaded the \citet{beckancient2014}
data (see \protect\hyperlink{example-data}{example data} for more
details).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading the data}
\KeywordTok{data}\NormalTok{(BeckLee_mat50)}
\KeywordTok{data}\NormalTok{(BeckLee_mat99)}
\KeywordTok{data}\NormalTok{(BeckLee_tree)}
\KeywordTok{data}\NormalTok{(BeckLee_ages)}
\end{Highlighting}
\end{Shaded}

\hypertarget{time-slicing}{\section{Time slicing}\label{time-slicing}}

The function \texttt{chrono.subsets} allows users to divide the matrix
into different time subsets or slices given a dated phylogeny that
contains all the elements (i.e.~taxa) from the matrix. Each subset
generated by this function will then contain all the elements present at
a specific point in time or during a specific period in time.

Two types of time subsets can be performed by using the \texttt{method}
option:

\begin{itemize}
\tightlist
\item
  Discrete time subsets (or time-binning) using
  \texttt{method\ =\ discrete}
\item
  Continuous time subsets (or time-slicing) using
  \texttt{method\ =\ continuous}
\end{itemize}

For the time-slicing method details see \citet{time-slice}. For both
methods, the function takes the \texttt{time} argument which can be a
vector of \texttt{numeric} values for:

\begin{itemize}
\tightlist
\item
  Defining the boundaries of the time bins (when
  \texttt{method\ =\ discrete})
\item
  Defining the time slices (when \texttt{method\ =\ continuous})
\end{itemize}

Otherwise, the \texttt{time} argument can be set as a single
\texttt{numeric} value for automatically generating a given number of
equidistant time-bins/slices. Additionally, it is also possible to input
a dataframe containing the first and last occurrence data (FAD/LAD) for
taxa that span over a longer time than the given tips/nodes age, so taxa
can appear in more than one time bin/slice.

Here is an example for \texttt{method\ =\ discrete}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Generating three time bins containing the taxa present every 40 Ma}
\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat50, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }\DataTypeTok{method =} \StringTok{"discrete"}\NormalTok{,}
                \DataTypeTok{time =} \KeywordTok{c}\NormalTok{(}\DecValTok{120}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{0}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 3 discrete time subsets for 50 elements:
##     120 - 80, 80 - 40, 40 - 0.
\end{verbatim}

Note that we can also generate equivalent results by just telling the
function that we want three time-bins as follow:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Automatically generate three equal length bins:}
\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat50, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }\DataTypeTok{method =} \StringTok{"discrete"}\NormalTok{,}
                \DataTypeTok{time =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 3 discrete time subsets for 50 elements:
##     133.51104 - 89.00736, 89.00736 - 44.50368, 44.50368 - 0.
\end{verbatim}

In this example, the taxa were split inside each time-bin according to
their age. However, the taxa here are considered as single points in
time. It is totally possible that some taxa could have had longer
longevity and that they exist in multiple time bins. In this case, it is
possible to include them in more than one bin by providing a table of
first and last occurrence dates (FAD/LAD). This table should have the
taxa names as row names and two columns for respectively the first and
last occurrence age:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Displaying the table of first and last occurrence dates for each taxa}
\KeywordTok{head}\NormalTok{(BeckLee_ages)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##             FAD  LAD
## Adapis     37.2 36.8
## Asioryctes 83.6 72.1
## Leptictis  33.9 33.3
## Miacis     49.0 46.7
## Mimotona   61.6 59.2
## Notharctus 50.2 47.0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Generating time bins including taxa that might span between them}
\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat50, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }\DataTypeTok{method =} \StringTok{"discrete"}\NormalTok{,}
                \DataTypeTok{time =} \KeywordTok{c}\NormalTok{(}\DecValTok{120}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{0}\NormalTok{), }\DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 3 discrete time subsets for 50 elements:
##     120 - 80, 80 - 40, 40 - 0.
\end{verbatim}

When using this method, the oldest boundary of the first bin (or the
first slice, see below) is automatically generated as the root age plus
1\% of the tree length, as long as at least three elements/taxa are
present at that point in time. The algorithm adds an extra 1\% tree
length until reaching the required minimum of three elements. It is also
possible to include nodes in each bin by using
\texttt{inc.nodes\ =\ TRUE} and providing a matrix that contains the
ordinated distance among tips \emph{and} nodes.

For the time-slicing method (\texttt{method\ =\ continuous}), the idea
is fairly similar. This option, however, requires a matrix that contains
the ordinated distance among taxa \emph{and} nodes and an extra argument
describing the assumed evolutionary model (via the \texttt{model}
argument). This model argument is used when the time slice occurs along
a branch of the tree rather than on a tip or a node, meaning that a
decision must be made about what the value for the branch should be. The
model can be one of the following:

\begin{itemize}
\item
  \textbf{Punctuated models}
\item
  \texttt{acctran} where the data chosen along the branch is always the
  one of the descendant
\item
  \texttt{deltran} where the data chosen along the branch is always the
  one of the ancestor
\item
  \texttt{random} where the data chosen along the branch is randomly
  chosen between the descendant or the ancestor
\item
  \texttt{proximity} where the data chosen along the branch is either
  the descendant or the ancestor depending on branch length
\item
  \textbf{Gradual models}
\item
  \texttt{equal.split} where the data chosen along the branch is both
  the descendant and the ancestor with an even probability
\item
  \texttt{gradual.split} where the data chosen along the branch is both
  the descendant and the ancestor with a probability depending on branch
  length
\end{itemize}

\begin{quote}
Note that the four first models are a proxy for punctuated evolution:
the selected data is always either the one of the descendant or the
ancestor. In other words, changes along the branches always occur at
either ends of it. The two last models are a proxy for gradual
evolution: the data from both the descendant and the ancestor is used
with an associate probability. These later models perform better when
bootstrapped, effectively approximating the ``intermediate'' state
between and the ancestor and the descendants.
\end{quote}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Generating four time slices every 40 million years under a model of proximity evolution}
\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat99, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }
    \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{, }\DataTypeTok{time =} \KeywordTok{c}\NormalTok{(}\DecValTok{120}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{0}\NormalTok{),}
    \DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 4 continuous (proximity) time subsets for 99 elements:
##     120, 80, 40, 0.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Generating four time slices automatically}
\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat99, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree,}
    \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{, }\DataTypeTok{time =} \DecValTok{4}\NormalTok{, }\DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 4 continuous (proximity) time subsets for 99 elements:
##     133.51104, 89.00736, 44.50368, 0.
\end{verbatim}

If you want to generate time subsets based on stratigraphy, the package
proposes a useful functions to do it for you: \texttt{get.bin.ages}
(check out the function's manual in \texttt{R})!

\hypertarget{customised-subsets}{\section{Customised
subsets}\label{customised-subsets}}

Another way of separating elements into different categories is to use
customised subsets as briefly explained
\protect\hyperlink{disparity-among-groups}{above}. This function simply
takes the list of elements to put in each group (whether they are the
actual element names or their position in the matrix).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating the two groups (crown and stems)}
\NormalTok{mammal_groups <-}\StringTok{ }\KeywordTok{crown.stem}\NormalTok{(BeckLee_tree, }\DataTypeTok{inc.nodes =} \OtherTok{FALSE}\NormalTok{)}

\NormalTok{## Separating the dataset into two different groups}
\KeywordTok{custom.subsets}\NormalTok{(BeckLee_mat50, }\DataTypeTok{group =}\NormalTok{ mammal_groups)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 2 customised subsets for 50 elements:
##     crown, stem.
\end{verbatim}

Like in this example, you can use the utility function
\texttt{crown.stem} that allows to automatically separate the crown and
stems taxa given a phylogenetic tree. Also, elements can easily be
assigned to different groups if necessary!

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating the three groups as a list}
\NormalTok{weird_groups <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\StringTok{"even"}\NormalTok{ =}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{1}\NormalTok{, }\DataTypeTok{to =} \DecValTok{49}\NormalTok{, }\DataTypeTok{by =} \DecValTok{2}\NormalTok{),}
                      \StringTok{"odd"}\NormalTok{ =}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{2}\NormalTok{, }\DataTypeTok{to =} \DecValTok{50}\NormalTok{, }\DataTypeTok{by =} \DecValTok{2}\NormalTok{),}
                      \StringTok{"all"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{50}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

The \texttt{custom.subsets} function can also take a phylogeny (as a
\texttt{phylo} object) as an argument to create groups as clades:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating groups as clades}
\KeywordTok{custom.subsets}\NormalTok{(BeckLee_mat50, }\DataTypeTok{group =}\NormalTok{ BeckLee_tree)}
\end{Highlighting}
\end{Shaded}

This automatically creates 49 (the number of nodes) groups containing
between two and 50 (the number of tips) elements.

\hypertarget{bootstraps-and-rarefactions}{\section{Bootstraps and
rarefactions}\label{bootstraps-and-rarefactions}}

One important step in analysing ordinated matrices is to
pseudo-replicate the data to see how robust the results are, and how
sensitive they are to outliers in the dataset. This can be achieved
using the function \texttt{boot.matrix} to bootstrap and/or rarefy the
data. The default options will bootstrap the matrix 100 times without
rarefaction using the ``full'' bootstrap method (see below):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Default bootstrapping}
\KeywordTok{boot.matrix}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat50)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 50 elements with 48 dimensions.
## Data was bootstrapped 100 times (method:"full").
\end{verbatim}

The number of bootstrap replicates can be defined using the
\texttt{bootstraps} option. The method can be modified by controlling
which bootstrap algorithm to use through the \texttt{boot.type}
argument. Currently two algorithms are implemented:

\begin{itemize}
\tightlist
\item
  \texttt{full} where the bootstrapping is entirely stochastic (\emph{n}
  elements are replaced by any \emph{m} elements drawn from the data)
\item
  \texttt{single} where only one random element is replaced by one other
  random element for each pseudo-replicate
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Bootstrapping with the single bootstrap method}
\KeywordTok{boot.matrix}\NormalTok{(BeckLee_mat50, }\DataTypeTok{boot.type =} \StringTok{"single"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 50 elements with 48 dimensions.
## Data was bootstrapped 100 times (method:"single").
\end{verbatim}

This function also allows users to rarefy the data using the
\texttt{rarefaction} argument. Rarefaction allows users to limit the
number of elements to be drawn at each bootstrap replication. This is
useful if, for example, one is interested in looking at the effect of
reducing the number of elements on the results of an analysis.

This can be achieved by using the \texttt{rarefaction} option that draws
only \emph{n-x} at each bootstrap replicate (where \emph{x} is the
number of elements not sampled). The default argument is \texttt{FALSE}
but it can be set to \texttt{TRUE} to fully rarefy the data (i.e.~remove
\emph{x} elements for the number of pseudo-replicates, where \emph{x}
varies from the maximum number of elements present in each subset to a
minimum of three elements). It can also be set to one or more
\texttt{numeric} values to only rarefy to the corresponding number of
elements.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Bootstrapping with the full rarefaction}
\KeywordTok{boot.matrix}\NormalTok{(BeckLee_mat50, }\DataTypeTok{bootstraps =} \DecValTok{20}\NormalTok{, }\DataTypeTok{rarefaction =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 50 elements with 48 dimensions.
## Data was bootstrapped 20 times (method:"full") and fully rarefied.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Or with a set number of rarefaction levels}
\KeywordTok{boot.matrix}\NormalTok{(BeckLee_mat50, }\DataTypeTok{bootstraps =} \DecValTok{20}\NormalTok{, }\DataTypeTok{rarefaction =} \KeywordTok{c}\NormalTok{(}\DecValTok{6}\OperatorTok{:}\DecValTok{8}\NormalTok{, }\DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 50 elements with 48 dimensions.
## Data was bootstrapped 20 times (method:"full") and rarefied to 6, 7, 8, 3 elements.
\end{verbatim}

One other argument is \texttt{dimensions} that specifies how many
dimensions from the matrix should be used for further analysis. When
missing, all dimensions from the ordinated matrix are used.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Using the first 50% of the dimensions}
\KeywordTok{boot.matrix}\NormalTok{(BeckLee_mat50, }\DataTypeTok{dimensions =} \FloatTok{0.5}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 50 elements with 24 dimensions.
## Data was bootstrapped 100 times (method:"full").
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Using the first 10 dimensions}
\KeywordTok{boot.matrix}\NormalTok{(BeckLee_mat50, }\DataTypeTok{dimensions =} \DecValTok{10}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 50 elements with 10 dimensions.
## Data was bootstrapped 100 times (method:"full").
\end{verbatim}

It is also possible to specify the sampling probability in the bootstrap
for each elements. This can be useful for weighting analysis for example
(i.e.~giving more importance to specific elements). These probabilities
can be passed to the \texttt{prob} argument individually with a vector
with the elements names or with a matrix with the rownames as elements
names. The elements with no specified probability will be assigned a
probability of 1 (or 1/maximum weight if the argument is weights rather
than probabilities).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Attributing a weight of 0 to Cimolestes and 10 to Maelestes}
\KeywordTok{boot.matrix}\NormalTok{(BeckLee_mat50, }\DataTypeTok{prob =} \KeywordTok{c}\NormalTok{(}\StringTok{"Cimolestes"}\NormalTok{ =}\StringTok{ }\DecValTok{0}\NormalTok{, }\StringTok{"Maelestes"}\NormalTok{ =}\StringTok{ }\DecValTok{10}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 50 elements with 48 dimensions.
## Data was bootstrapped 100 times (method:"full").
\end{verbatim}

Of course, one could directly supply the subsets generated above (using
\texttt{chrono.subsets} or \texttt{custom.subsets}) to this function.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating subsets of crown and stem mammals}
\NormalTok{crown_stem <-}\StringTok{ }\KeywordTok{custom.subsets}\NormalTok{(BeckLee_mat50,}
                                \DataTypeTok{group =} \KeywordTok{list}\NormalTok{(}\StringTok{"crown"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{16}\NormalTok{, }\DecValTok{19}\OperatorTok{:}\DecValTok{41}\NormalTok{, }\DecValTok{45}\OperatorTok{:}\DecValTok{50}\NormalTok{), }
                                             \StringTok{"stem"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{15}\NormalTok{, }\DecValTok{17}\OperatorTok{:}\DecValTok{18}\NormalTok{, }\DecValTok{42}\OperatorTok{:}\DecValTok{44}\NormalTok{)))}
\NormalTok{## Bootstrapping and rarefying these groups}
\KeywordTok{boot.matrix}\NormalTok{(crown_stem, }\DataTypeTok{bootstraps =} \DecValTok{200}\NormalTok{, }\DataTypeTok{rarefaction =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 2 customised subsets for 50 elements with 48 dimensions:
##     crown, stem.
## Data was bootstrapped 200 times (method:"full") and fully rarefied.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating time slice subsets}
\NormalTok{time_slices <-}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat99, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }
                               \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{, }
                               \DataTypeTok{time =} \KeywordTok{c}\NormalTok{(}\DecValTok{120}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{0}\NormalTok{),}
                               \DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages)}

\NormalTok{## Bootstrapping the time slice subsets}
\KeywordTok{boot.matrix}\NormalTok{(time_slices, }\DataTypeTok{bootstraps =} \DecValTok{100}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 4 continuous (proximity) time subsets for 99 elements with 97 dimensions:
##     120, 80, 40, 0.
## Data was bootstrapped 100 times (method:"full").
\end{verbatim}

\hypertarget{disparity-metrics}{\section{Disparity
metrics}\label{disparity-metrics}}

There are many ways of measuring disparity! In brief, disparity is a
summary metric that will represent an aspect of an ordinated space
(e.g.~a MDS, PCA, PCO, PCoA). For example, one can look at ellipsoid
hyper-volume of the ordinated space (Donohue \emph{et al.} 2013), the
sum and the product of the ranges and variances (Wills \emph{et al.}
1994) or the median position of the elements relative to their centroid
(Wills \emph{et al.} 1994). Of course, there are many more examples of
metrics one can use for describing some aspect of the ordinated space,
with some performing better than other ones at particular descriptive
tasks, and some being more generalist.

Because of this great diversity of metrics, the package
\texttt{dispRity} does not have one way to measure disparity but rather
proposes to facilitate users in defining their own disparity metric that
will best suit their particular analysis. In fact, the core function of
the package, \texttt{dispRity}, allows the user to define any metric
with the \texttt{metric} argument. However the \texttt{metric} argument
has to follow certain rules:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  It must be composed from one to three \texttt{function} objects;
\item
  The function(s) must take as a first argument a \texttt{matrix} or a
  \texttt{vector};
\item
  The function(s) must be of one of the three dimension-levels described
  below;
\item
  At least one of the functions must be of dimension-level 1 or 2 (see
  below).
\end{enumerate}

\subsection{The function
dimension-levels}\label{the-function-dimension-levels}

The metric function dimension-levels determine the ``dimensionality of
decomposition'' of the input matrix. In other words, each
dimension-level designates the dimensions of the output, i.e.~either
three (a \texttt{matrix}); two (a \texttt{vector}); or one (a single
\texttt{numeric} value) dimension.

\begin{figure}
\centering
\includegraphics{dispRity_fun.png}
\caption{Illustration of the different dimension-levels of functions
with an input \texttt{matrix}}
\end{figure}

\subsubsection{Dimension-level 1
functions}\label{dimension-level-1-functions}

A dimension-level 1 function will decompose a \texttt{matrix} or a
\texttt{vector} into a single value:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating a dummy matrix}
\NormalTok{dummy_matrix <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\KeywordTok{rnorm}\NormalTok{(}\DecValTok{12}\NormalTok{), }\DecValTok{4}\NormalTok{, }\DecValTok{3}\NormalTok{)}

\NormalTok{## Example of dimension-level 1 functions}
\KeywordTok{mean}\NormalTok{(dummy_matrix)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.4719271
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{median}\NormalTok{(dummy_matrix)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.4816107
\end{verbatim}

Any summary metric such as mean or median are good examples of
dimension-level 1 functions as they reduce the matrix to a single
dimension (i.e.~one value).

\subsubsection{Dimension-level 2
functions}\label{dimension-level-2-functions}

A dimension-level 2 function will decompose a \texttt{matrix} into a
\texttt{vector}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Defining the function as the product of rows}
\NormalTok{prod.rows <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(matrix) }\KeywordTok{apply}\NormalTok{(matrix, }\DecValTok{1}\NormalTok{, prod)}

\NormalTok{## A dimension-level 2 metric}
\KeywordTok{prod.rows}\NormalTok{(dummy_matrix)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] -0.1693259  0.1151639  0.4023232  0.2070852
\end{verbatim}

Several dimension-level 2 functions are implemented in \texttt{dispRity}
(see \texttt{?dispRity.metric}) such as the \texttt{variances} or
\texttt{ranges} functions that calculate the variance or the range of
each dimension of the ordinated matrix respectively.

\subsubsection{Dimension-level 3
functions}\label{dimension-level-3-functions}

Finally a dimension-level 3 function will transform the matrix into
another matrix. Note that the dimension of the output matrix doesn't
need to match the the input matrix:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## A dimension-level 3 metric}
\KeywordTok{var}\NormalTok{(dummy_matrix)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##           [,1]      [,2]      [,3]
## [1,] 0.2275789 0.2927228 0.2720648
## [2,] 0.2927228 0.6394468 0.6796798
## [3,] 0.2720648 0.6796798 1.9455929
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## A dimension-level 3 metric with a forced matrix output}
\KeywordTok{as.matrix}\NormalTok{(}\KeywordTok{dist}\NormalTok{(dummy_matrix))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##          1        2         3         4
## 1 0.000000 2.284299 3.1469895 3.2820238
## 2 2.284299 0.000000 1.8652681 1.9829745
## 3 3.146990 1.865268 0.0000000 0.6684051
## 4 3.282024 1.982975 0.6684051 0.0000000
\end{verbatim}

\subsection{\texorpdfstring{\texttt{make.metric}}{make.metric}}\label{make.metric}

Of course, functions can be more complex and involve multiple operations
such as the \texttt{centroids} function (see \texttt{?dispRity.metric})
that calculates the Euclidean distance between each element and the
centroid of the ordinated space. The \texttt{make.metric} function
implemented in \texttt{dispRity} is designed to help test and find the
dimension-level of the functions. This function tests:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  If your function can deal with a \texttt{matrix} or a \texttt{vector}
  as an input;
\item
  Your function's dimension-level according to its output
  (dimension-level 1, 2 or 3, see above);
\item
  Whether the function can be implemented in the \texttt{dispRity}
  function (the function is fed into a \texttt{lapply} loop).
\end{enumerate}

For example, let's see if the functions described above are the right
dimension-levels:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Which dimension-level is the mean function? And can it be used in dispRity?}
\KeywordTok{make.metric}\NormalTok{(mean)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## mean outputs a single value.
## mean is detected as being a dimension-level 1 function.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Which dimension-level is the prod.rows function? And can it be used in dispRity?}
\KeywordTok{make.metric}\NormalTok{(prod.rows)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## prod.rows outputs a matrix object.
## prod.rows is detected as being a dimension-level 2 function.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Which dimension-level is the var function? And can it be used in dispRity?}
\KeywordTok{make.metric}\NormalTok{(var)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## var outputs a matrix object.
## var is detected as being a dimension-level 3 function.
## Additional dimension-level 2 and/or 1 function(s) will be needed.
\end{verbatim}

A non verbose version of the function is also available. This can be
done using the option \texttt{silent\ =\ TRUE} and will simply output
the dimension-level of the metric.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Testing whether mean is dimension-level 1}
\ControlFlowTok{if}\NormalTok{(}\KeywordTok{make.metric}\NormalTok{(mean, }\DataTypeTok{silent =} \OtherTok{TRUE}\NormalTok{) }\OperatorTok{!=}\StringTok{ "level1"}\NormalTok{) \{}
    \KeywordTok{message}\NormalTok{(}\StringTok{"The metric is not dimension-level 1."}\NormalTok{)}
\NormalTok{\}}
\NormalTok{## Testing whether var is dimension-level 1}
\ControlFlowTok{if}\NormalTok{(}\KeywordTok{make.metric}\NormalTok{(var, }\DataTypeTok{silent =} \OtherTok{TRUE}\NormalTok{) }\OperatorTok{!=}\StringTok{ "level1"}\NormalTok{) \{}
    \KeywordTok{message}\NormalTok{(}\StringTok{"The metric is not dimension-level 1."}\NormalTok{)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## The metric is not dimension-level 1.
\end{verbatim}

\subsection{\texorpdfstring{Metrics in the \texttt{dispRity}
function}{Metrics in the dispRity function}}\label{metrics-in-the-disprity-function}

Using this metric structure, we can easily use any disparity metric in
the \texttt{dispRity} function as follows:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring disparity as the standard deviation of all the values of the}
\NormalTok{## ordinated matrix (dimension-level 1 function).}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(BeckLee_mat50, }\DataTypeTok{metric =}\NormalTok{ sd))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 50 0.201
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring disparity as the standard deviation of the variance of each axis of}
\NormalTok{## the ordinated matrix (dimension-level 1 and 2 functions).}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(BeckLee_mat50, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sd, variances)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 50 0.028
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring disparity as the standard deviation of the variance of each axis of}
\NormalTok{## the variance covariance matrix (dimension-level 1, 2 and 3 functions).}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(BeckLee_mat50, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sd, variances, var)), }\DataTypeTok{round =} \DecValTok{10}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n obs
## 1       1 50   0
\end{verbatim}

Note that the order of each function in the metric argument does not
matter, the \texttt{dispRity} function will automatically detect the
function dimension-levels (using \texttt{make.metric}) and apply them to
the data in decreasing order (dimension-level 3 \textgreater{} 2
\textgreater{} 1).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Disparity as the standard deviation of the variance of each axis of the}
\NormalTok{## variance covariance matrix:}
\NormalTok{disparity1 <-}\StringTok{ }\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(BeckLee_mat50, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sd, variances, var)),}
                      \DataTypeTok{round =} \DecValTok{10}\NormalTok{)}

\NormalTok{## Same as above but using a different function order for the metric argument}
\NormalTok{disparity2 <-}\StringTok{ }\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(BeckLee_mat50, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(variances, sd, var)),}
                      \DataTypeTok{round =} \DecValTok{10}\NormalTok{)}

\NormalTok{## Both ways output the same disparity values:}
\NormalTok{disparity1 }\OperatorTok{==}\StringTok{ }\NormalTok{disparity2}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      subsets    n  obs
## [1,]    TRUE TRUE TRUE
\end{verbatim}

In these examples, we considered disparity to be a single value. For
example, in the previous example, we defined disparity as the standard
deviation of the variances of each column of the variance/covariance
matrix (\texttt{metric\ =\ c(variances,\ sd,\ var)}). It is, however,
possible to calculate
\protect\hyperlink{disparity-as-a-distribution}{disparity as a
distribution}.

\subsection{\texorpdfstring{Metrics implemented in
\texttt{dispRity}}{Metrics implemented in dispRity}}\label{metrics-implemented-in-disprity}

Several disparity metrics are implemented in the \texttt{dispRity}
package. The detailed list can be found in \texttt{?dispRity.metric}
along with some description of each metric.

\begin{longtable}[]{@{}llll@{}}
\toprule
\begin{minipage}[b]{0.08\columnwidth}\raggedright\strut
Level\strut
\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\raggedright\strut
Name\strut
\end{minipage} & \begin{minipage}[b]{0.61\columnwidth}\raggedright\strut
Description\strut
\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\raggedright\strut
Source\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
2\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{ancestral.dist}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The distance between an element and its ancestor\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\texttt{dispRity}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
2\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{centroids}1\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The distance between each element and the centroid of the ordinated
space\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\texttt{dispRity}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
1\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{convhull.surface}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The surface of the convex hull formed by all the elements\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\href{https://cran.r-project.org/web/packages/geometry/index.html}{\texttt{geometry}}\texttt{::convhulln\$area}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
1\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{convhull.volume}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The volume of the convex hull formed by all the elements\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\href{https://cran.r-project.org/web/packages/geometry/index.html}{\texttt{geometry}}\texttt{::convhulln\$vol}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
1\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{diagonal}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The longest distance in the ordinated space (like the diagonal in two
dimensions)\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\texttt{dispRity}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
1\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{ellipse.volume}1\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The volume of the ellipsoid of the space\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
Donohue \emph{et al.} (2013)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
1\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{mode.val}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The modal value\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\texttt{dispRity}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
1\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{n.ball.volume}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The hyper-spherical (\emph{n}-ball) volume\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\texttt{dispRity}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
2\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{pairwise.dist}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The pairwise distances between elements\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\href{https://cran.r-project.org/web/packages/vegan/index.html}{\texttt{vegan}}\texttt{::vegist}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
2\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{radius}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The radius of each dimensions\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\texttt{dispRity}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
2\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{ranges}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The range of each dimension\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\texttt{dispRity}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
1\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{span.tree.length}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The minimal spanning tree length\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\href{https://cran.r-project.org/web/packages/vegan/index.html}{\texttt{vegan}}\texttt{::spantree}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
2\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
\texttt{variances}\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
The variance of each dimension\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
\texttt{dispRity}\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

1: Note that by default, the centroid is the centroid of the elements.
It can, however, be fixed to a different value by using the
\texttt{centroid} argument
\texttt{centroids(space,\ centroid\ =\ rep(0,\ ncol(space)))}, for
example the origin of the ordinated space.

2: This function uses an estimation of the eigenvalue that only works
for MDS or PCoA ordinations (\emph{not} PCA).

\subsection{Equations and
implementations}\label{equations-and-implementations}

Some of the functions described below are implemented in the
\texttt{dispRity} package and do not require any other packages to
calculate
(\href{https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R}{see
implementation here}).

\begin{equation}
    ancestral.dist = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}
\end{equation}

\begin{equation}
    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
\end{equation}

\begin{equation}
    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
\end{equation}

\begin{equation}
    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
\end{equation}

\begin{equation}
    n.ball.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R
\end{equation}

\begin{equation}
    radius = |\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|
\end{equation}

\begin{equation}
    ranges = |max(k_i) - min(k_i)|
\end{equation}

\begin{equation}
    variances = \sigma^{2}{k_i}
\end{equation}

\begin{equation}
    span.tree.length = \sum(\mathrm{branch\ length})
\end{equation}

Where \emph{k} is the number of dimensions, \emph{n} the number of
elements, \(\Gamma\) is the Gamma distribution, \(\lambda_i\) is the
eigenvalue of each dimensions, \(\sigma^{2}\) is their variance and
\(Centroid_{k}\) is their mean, \(Ancestor_{n}\) is the coordinates of
the ancestor of element \(n\), \(f(\mathbf{v}k)\) is function to select
one value from the vector \(\mathbf{v}\) of the dimension \(k\)
(e.g.~it's maximum, minimum, mean, etc.), \emph{R} is the radius of the
sphere or the product of the radii of each dimensions
(\(\displaystyle\prod_{i=1}^{k}R_{i}\) - for a hyper-ellipsoid).

\subsection{Using the different disparity
metrics}\label{using-the-different-disparity-metrics}

Here is a brief demonstration of the main metrics implemented in
\texttt{dispRity}. First, we will create a dummy/simulated ordinated
space using the \texttt{space.maker} utility function (more about that
\protect\hyperlink{space.maker}{here}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating a 10*5 normal space}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{dummy_space <-}\StringTok{ }\KeywordTok{space.maker}\NormalTok{(}\DecValTok{10}\NormalTok{, }\DecValTok{5}\NormalTok{, rnorm)}
\end{Highlighting}
\end{Shaded}

We will use this simulated space to demonstrate the different metrics.

\subsubsection{Volumes and surface
metrics}\label{volumes-and-surface-metrics}

The functions \texttt{ellipse.volume}, \texttt{convhull.surface},
\texttt{convhull.volume} and \texttt{n.ball.volume} all measure the
surface or the volume of the ordinated space occupied:

Because there is only one subset (i.e.~one matrix) in the dispRity
object, the operations below are the equivalent of
\texttt{metric(dummy\_space)} (with rounding).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the ellipsoid volume}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ ellipse.volume))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 10 257.8
\end{verbatim}

\begin{quote}
WARNING: in such dummy space, this gives the estimation of the ellipsoid
volume, not the real ellipsoid volume! See the cautionary note in
\texttt{?ellipse.volume}.
\end{quote}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the convex hull surface}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ convhull.surface))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 10 11.91
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the convex hull volume}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ convhull.volume))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 10 1.031
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the convex hull volume}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ n.ball.volume))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n  obs
## 1       1 10 4.43
\end{verbatim}

The convex hull based functions are a call to the
\texttt{geometry::convhulln} function with the \texttt{"FA"} option
(computes total area and volume). Also note that they are really
sensitive to the size of the dataset.

\begin{quote}
Cautionary note: measuring volumes in a high number of dimensions can be
strongly affected by the
\href{https://en.wikipedia.org/wiki/Curse_of_dimensionality}{curse of
dimensionality} that often results in near 0 disparity values. I
strongly recommend reading
\href{https://beta.observablehq.com/@tophtucker/theres-plenty-of-room-in-the-corners}{this
really intuitive explanation} from
\href{https://github.com/tophtucker}{Toph Tucker}.
\end{quote}

\subsubsection{Ranges, variances, radius, pairwise distance, modal value
and
diagonal}\label{ranges-variances-radius-pairwise-distance-modal-value-and-diagonal}

The functions \texttt{ranges}, \texttt{variances} \texttt{radius},
\texttt{pairwise.dist}, \texttt{mode.val} and \texttt{diagonal} all
measure properties of the ordinated space based on its dimensional
properties (they are also less affected by the ``curse of
dimensionality''):

\texttt{ranges}, \texttt{variances} and \texttt{radius} work on the same
principle and measure the range/variance/radius of each dimension:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the ranges of each dimension in the ordinated space}
\KeywordTok{ranges}\NormalTok{(dummy_space)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 2.430909 3.726481 2.908329 2.735739 1.588603
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity as the distribution of these ranges}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ ranges))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n obs.median  2.5%   25%   75% 97.5%
## 1       1 10      2.431 1.673 2.431 2.908 3.645
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity as the sum and the product of these ranges}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, ranges)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 10 13.39
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(prod, ranges)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 10 114.5
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the variances of each dimension in the ordinated space}
\KeywordTok{variances}\NormalTok{(dummy_space)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.6093144 1.1438620 0.9131859 0.6537768 0.3549372
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity as the distribution of these variances}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ variances))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n obs.median 2.5%   25%   75% 97.5%
## 1       1 10      0.609 0.38 0.609 0.913 1.121
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity as the sum and the product of these variances}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 10 3.675
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(prod, variances)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 10 0.148
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the radius of each dimension in the ordinated space}
\KeywordTok{radius}\NormalTok{(dummy_space)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 1.4630780 2.4635449 1.8556785 1.4977898 0.8416318
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## By default the radius is the maximum distance from the centre of}
\NormalTok{## the dimension. It can however be changed to any function:}
\KeywordTok{radius}\NormalTok{(dummy_space, }\DataTypeTok{type =}\NormalTok{ min)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.05144054 0.14099827 0.02212226 0.17453525 0.23044528
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{radius}\NormalTok{(dummy_space, }\DataTypeTok{type =}\NormalTok{ mean)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.6233501 0.7784888 0.7118713 0.6253263 0.5194332
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity as the mean average radius}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(mean, radius), }\DataTypeTok{type =}\NormalTok{ mean))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 10 0.652
\end{verbatim}

The pairwise distances uses the function \texttt{vegan::vegdist} and can
take the normal \texttt{vegdist} options:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The average pairwise euclidean distance}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(mean, pairwise.dist)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 10 2.539
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The distribution of the Manhattan distances}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ pairwise.dist, }\DataTypeTok{method =} \StringTok{"manhattan"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n obs.median  2.5%   25%   75% 97.5%
## 1       1 10      3.619 2.566 3.335 5.672  9.63
\end{verbatim}

Note that this function is a direct call to
\texttt{vegan::vegdist(matrix,\ method\ =\ method,\ diag\ =\ FALSE,\ upper\ =\ FALSE,\ ...)}.

The \texttt{diagonal} function measures the multidimensional diagonal of
the whole space (i.e.~in our case the longest Euclidean distance in our
five dimensional space). The \texttt{mode.val} function measures the
modal value of the matrix:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the ordinated space's diagonal}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ diagonal))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 10 3.659
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating the modal value of the matrix}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ mode.val))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 10 -2.21
\end{verbatim}

\begin{quote}
This metric is only a Euclidean diagonal (mathematically valid) if the
dimensions within the space are all orthogonal!
\end{quote}

\subsubsection{Centroids and ancestral distance
metrics}\label{centroids-and-ancestral-distance-metrics}

The \texttt{centroids} metric allows users to measure the position of
the different elements compared to a fixed point in the ordinated space.
By default, this function measures the distance between each element and
their centroid (centre point):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The distribution of the distances between each element and their centroid}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ centroids))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n obs.median  2.5%   25%   75% 97.5%
## 1       1 10      2.214 0.788 1.267 1.993 3.167
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Disparity as the median value of these distances}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(median, centroids)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs
## 1       1 10 1.435
\end{verbatim}

It is however possible to fix the coordinates of the centroid to a
specific point in the ordinated space, as long as it has the correct
number of dimensions:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The distance between each element and the origin of the ordinated space}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ centroids, }\DataTypeTok{centroid =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n obs.median  2.5% 25%   75% 97.5%
## 1       1 10      2.323 0.785 1.2 2.044 3.176
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Disparity as the distance between each element and a specific point in space}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ centroids, }\DataTypeTok{centroid =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{,}\DecValTok{4}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n obs.median  2.5%   25%   75% 97.5%
## 1       1 10      4.675 4.293 5.032 6.155 6.957
\end{verbatim}

The \texttt{ancestral.dist} metric works on a similar principle as the
\texttt{centroids} function but changes the centroid to be the
coordinates of each element's ancestor. Therefore this functions needs a
tree and node coordinates as additional arguments:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## A generating a random tree with node labels}
\NormalTok{tree <-}\StringTok{ }\KeywordTok{rtree}\NormalTok{(}\DecValTok{5}\NormalTok{) ; tree}\OperatorTok{$}\NormalTok{node.label <-}\StringTok{ }\KeywordTok{paste0}\NormalTok{(}\StringTok{"n"}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{)}
\NormalTok{## Adding the tip and node names to the matrix}
\NormalTok{dummy_space2 <-}\StringTok{ }\NormalTok{dummy_space[}\OperatorTok{-}\DecValTok{1}\NormalTok{,]}
\KeywordTok{rownames}\NormalTok{(dummy_space2) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(tree}\OperatorTok{$}\NormalTok{tip.label, tree}\OperatorTok{$}\NormalTok{node.label)}

\NormalTok{## Calculating all the ancestral nodes}
\NormalTok{all_anc_centroids <-}\StringTok{ }\KeywordTok{nodes.coordinates}\NormalTok{(dummy_space2, tree, }\DataTypeTok{full =} \OtherTok{TRUE}\NormalTok{)}

\NormalTok{## Calculating the distances from the ancestral nodes}
\NormalTok{ancestral_dist <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(dummy_space2, }\DataTypeTok{metric =}\NormalTok{ ancestral.dist,}
                           \DataTypeTok{nodes.coords =}\NormalTok{ all_anc_centroids)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in fun(matrix, ...): Missing tree and full argument for nodes.coordinates.
## See ?nodes.coordinates manual.
## The centroids function was applied instead.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity as the sum of the distances from all the ancestral nodes}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(ancestral_dist, }\DataTypeTok{metric =}\NormalTok{ sum))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets n   obs
## 1       1 9 36.56
\end{verbatim}

\subsubsection{Minimal spanning tree
length}\label{minimal-spanning-tree-length}

The \texttt{span.tree.length} uses the \texttt{vegan::spantree} function
to heuristically calculate the minimum spanning tree (the shortest
multidimensional tree connecting each elements) and calculates its
length as the sum of every branch lengths.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The length of the minimal spanning tree}
\KeywordTok{summary}\NormalTok{(}\KeywordTok{dispRity}\NormalTok{(dummy_space, }\DataTypeTok{metric =}\NormalTok{ span.tree.length))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n  obs
## 1       1 10 15.4
\end{verbatim}

Note that because the solution is heuristic, this metric can take a long
time to compute for big matrices.

\section{Summarising dispRity data
(plots)}\label{summarising-disprity-data-plots}

Because of its architecture, printing \texttt{dispRity} objects only
summarises their content but does not print the disparity value measured
or associated analysis (more about this
\protect\hyperlink{manipulating-dispRity-objects}{here}). To actually
see what is in a dispRity object, one can either use the
\texttt{summary} function for visualising the data in a table or
\texttt{plot} to have a graphical representation of the results.

\subsection{\texorpdfstring{Summarising \texttt{dispRity}
data}{Summarising dispRity data}}\label{summarising-disprity-data}

This function is an S3 function (\texttt{summary.dispRity}) allowing
users to summarise the content of \texttt{dispRity} objects that contain
disparity calculations.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Example data from previous sections}
\NormalTok{crown_stem <-}\StringTok{ }\KeywordTok{custom.subsets}\NormalTok{(BeckLee_mat50,}
                                \DataTypeTok{group =} \KeywordTok{list}\NormalTok{(}\StringTok{"crown"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{16}\NormalTok{, }\DecValTok{19}\OperatorTok{:}\DecValTok{41}\NormalTok{, }\DecValTok{45}\OperatorTok{:}\DecValTok{50}\NormalTok{), }
                                             \StringTok{"stem"}\NormalTok{ =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{15}\NormalTok{, }\DecValTok{17}\OperatorTok{:}\DecValTok{18}\NormalTok{, }\DecValTok{42}\OperatorTok{:}\DecValTok{44}\NormalTok{)))}
\NormalTok{## Bootstrapping and rarefying these groups}
\NormalTok{boot_crown_stem <-}\StringTok{ }\KeywordTok{boot.matrix}\NormalTok{(crown_stem, }\DataTypeTok{bootstraps =} \DecValTok{100}\NormalTok{, }\DataTypeTok{rarefaction =} \OtherTok{TRUE}\NormalTok{)}
\NormalTok{## Calculate disparity}
\NormalTok{disparity_crown_stem <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(boot_crown_stem, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances))}

\NormalTok{## Creating time slice subsets}
\NormalTok{time_slices <-}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat99, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }
    \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{, }\DataTypeTok{time =} \KeywordTok{c}\NormalTok{(}\DecValTok{120}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{0}\NormalTok{),}
    \DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages)}
\NormalTok{## Bootstrapping the time slice subsets}
\NormalTok{boot_time_slices <-}\StringTok{ }\KeywordTok{boot.matrix}\NormalTok{(time_slices, }\DataTypeTok{bootstraps =} \DecValTok{100}\NormalTok{)}
\NormalTok{## Calculate disparity}
\NormalTok{disparity_time_slices <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(boot_time_slices, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances))}

\NormalTok{## Creating time bin subsets}
\NormalTok{time_bins <-}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat99, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree, }
    \DataTypeTok{method =} \StringTok{"discrete"}\NormalTok{, }\DataTypeTok{time =} \KeywordTok{c}\NormalTok{(}\DecValTok{120}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{0}\NormalTok{), }\DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages,}
    \DataTypeTok{inc.nodes =} \OtherTok{TRUE}\NormalTok{)}
\NormalTok{## Bootstrapping the time bin subsets}
\NormalTok{boot_time_bins <-}\StringTok{ }\KeywordTok{boot.matrix}\NormalTok{(time_bins, }\DataTypeTok{bootstraps =} \DecValTok{100}\NormalTok{)}
\NormalTok{## Calculate disparity}
\NormalTok{disparity_time_bins <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(boot_time_bins, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances))}
\end{Highlighting}
\end{Shaded}

These objects are easy to summarise as follows:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Default summary}
\KeywordTok{summary}\NormalTok{(disparity_time_slices)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs bs.median  2.5%   25%   75% 97.5%
## 1     120  5 2.823     2.295 1.398 2.037 2.563 2.767
## 2      80 19 3.233     3.065 2.785 2.973 3.133 3.266
## 3      40 15 3.359     3.149 2.764 2.996 3.285 3.443
## 4       0 10 4.055     3.685 3.169 3.453 3.760 3.961
\end{verbatim}

Information about the number of elements in each subset and the observed
(i.e.~non-bootstrapped) disparity are also calculated. This is
specifically handy when rarefying the data for example:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(disparity_crown_stem))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs bs.median  2.5%   25%   75% 97.5%
## 1   crown 30 1.995     1.933 1.873 1.915 1.943 1.972
## 2   crown 29    NA     1.935 1.861 1.906 1.953 1.976
## 3   crown 28    NA     1.929 1.872 1.910 1.945 1.970
## 4   crown 27    NA     1.933 1.852 1.905 1.947 1.981
## 5   crown 26    NA     1.932 1.867 1.910 1.948 1.971
## 6   crown 25    NA     1.929 1.847 1.913 1.949 1.975
\end{verbatim}

The summary functions can also take various options such as:

\begin{itemize}
\tightlist
\item
  \texttt{quantile} values for the confidence interval levels (by
  default, the 50 and 95 quantiles are calculated)
\item
  \texttt{cent.tend} for the central tendency to use for summarising the
  results (default is \texttt{median})
\item
  digits\texttt{option\ corresponding\ to\ the\ number\ of\ decimal\ places\ to\ print\ (default\ is}2`)
\item
  \texttt{recall} option for printing the call of the \texttt{dispRity}
  object as well (default is \texttt{FALSE})
\end{itemize}

These options can easily be changed from the defaults as follows:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Same as above but using the 88th quantile and the standard deviation as the summary }
\KeywordTok{summary}\NormalTok{(disparity_time_slices, }\DataTypeTok{quantile =} \DecValTok{88}\NormalTok{, }\DataTypeTok{cent.tend =}\NormalTok{ sd)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs bs.sd  2.5%   25%   75% 97.5%
## 1     120  5 2.823 0.380 1.398 2.037 2.563 2.767
## 2      80 19 3.233 0.127 2.785 2.973 3.133 3.266
## 3      40 15 3.359 0.189 2.764 2.996 3.285 3.443
## 4       0 10 4.055 0.205 3.169 3.453 3.760 3.961
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Printing the details of the object and digits the values to the 5th decimal place}
\KeywordTok{summary}\NormalTok{(disparity_time_slices, }\DataTypeTok{recall =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{digits =} \DecValTok{5}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 4 continuous (proximity) time subsets for 99 elements with 97 dimensions:
##     120, 80, 40, 0.
## Data was bootstrapped 100 times (method:"full").
## Disparity was calculated as: c(sum, variances).
\end{verbatim}

\begin{verbatim}
##   subsets  n     obs bs.median    2.5%     25%     75%   97.5%
## 1     120  5 2.82292   2.29516 1.39758 2.03734 2.56317 2.76690
## 2      80 19 3.23312   3.06472 2.78542 2.97332 3.13308 3.26617
## 3      40 15 3.35947   3.14918 2.76395 2.99563 3.28541 3.44267
## 4       0 10 4.05457   3.68545 3.16864 3.45263 3.76034 3.96114
\end{verbatim}

Note that the summary table is a \texttt{data.frame}, hence it is as
easy to modify as any dataframe using \texttt{dplyr}. You can also
export it in \texttt{csv} format using \texttt{write.csv} or
\texttt{write\_csv} or even directly export into \texttt{LaTeX} format
using the following;

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading the xtable package}
\KeywordTok{require}\NormalTok{(xtable)}
\NormalTok{## Converting the table in LaTeX}
\KeywordTok{xtable}\NormalTok{(}\KeywordTok{summary}\NormalTok{(disparity_time_slices))}
\end{Highlighting}
\end{Shaded}

\subsection{\texorpdfstring{Plotting \texttt{dispRity}
data}{Plotting dispRity data}}\label{plotting-disprity-data}

An alternative (and more fun!) way to display the calculated disparity
is to plot the results using the S3 method \texttt{plot.dispRity}. This
function takes the same options as \texttt{summary.dispRity} along with
various graphical options described in the function help files (see
\texttt{?plot.dispRity}).

The plots can be of four different types:

\begin{itemize}
\tightlist
\item
  \texttt{continuous} for displaying continuous disparity curves
\item
  \texttt{box}, \texttt{lines}, and \texttt{polygons} to display
  discrete disparity results in respectively a boxplot, confidence
  interval lines, and confidence interval polygons.
\end{itemize}

\begin{quote}
This argument can be left empty. In this case, the algorithm will
automatically detect the type of subsets from the \texttt{dispRity}
object and plot accordingly.
\end{quote}

It is also possible to display the number of elements in each subset (as
a horizontal dotted line) using the option \texttt{elements\ =\ TRUE}.
Additionally, when the data is rarefied, one can indicate which level of
rarefaction to display (i.e.~only display the results for a certain
number of elements) by using the \texttt{rarefaction} argument.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical parameters}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting continuous disparity results}
\KeywordTok{plot}\NormalTok{(disparity_time_slices, }\DataTypeTok{type =} \StringTok{"continuous"}\NormalTok{)}

\NormalTok{## Plotting discrete disparity results}
\KeywordTok{plot}\NormalTok{(disparity_crown_stem, }\DataTypeTok{type =} \StringTok{"box"}\NormalTok{)}

\NormalTok{## As above but using lines for the rarefaction level of 20 elements only}
\KeywordTok{plot}\NormalTok{(disparity_crown_stem, }\DataTypeTok{type =} \StringTok{"line"}\NormalTok{, }\DataTypeTok{rarefaction =} \DecValTok{20}\NormalTok{)}

\NormalTok{## As above but using polygons while also displaying the number of elements}
\KeywordTok{plot}\NormalTok{(disparity_crown_stem, }\DataTypeTok{type =} \StringTok{"polygon"}\NormalTok{, }\DataTypeTok{elements =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-54-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Resetting graphical parameters}
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

Since \texttt{plot.dispRity} uses the arguments from the generic
\texttt{plot} method, it is of course possible to change pretty much
everything using the regular plot arguments:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting the results with some classic options from plot}
\KeywordTok{plot}\NormalTok{(disparity_time_slices, }\DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\StringTok{"blue"}\NormalTok{, }\StringTok{"orange"}\NormalTok{, }\StringTok{"green"}\NormalTok{),}
    \DataTypeTok{ylab =} \KeywordTok{c}\NormalTok{(}\StringTok{"Some measurement"}\NormalTok{), }\DataTypeTok{xlab =} \StringTok{"Some other measurement"}\NormalTok{,}
    \DataTypeTok{main =} \StringTok{"Many options..."}\NormalTok{, }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{5}\NormalTok{, }\DecValTok{0}\NormalTok{), }\DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{, }\DecValTok{0}\NormalTok{))}

\NormalTok{## Adding a legend}
\KeywordTok{legend}\NormalTok{(}\StringTok{"topleft"}\NormalTok{, }\DataTypeTok{legend =} \KeywordTok{c}\NormalTok{(}\StringTok{"Central tendency"}\NormalTok{,}
                             \StringTok{"Confidence interval 1"}\NormalTok{,}
                            \StringTok{"Confidence interval 2"}\NormalTok{),}
      \DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\StringTok{"blue"}\NormalTok{, }\StringTok{"orange"}\NormalTok{, }\StringTok{"green"}\NormalTok{), }\DataTypeTok{pch =} \DecValTok{19}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-55-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Resetting graphical parameters}
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

In addition to the classic \texttt{plot} arguments, the function can
also take arguments that are specific to \texttt{plot.dispRity} like
adding the number of elements or rarefaction level (as described above),
and also changing the values of the quantiles to plot as well as the
central tendency.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting the results with some plot.dispRity arguments}
\KeywordTok{plot}\NormalTok{(disparity_time_slices, }\DataTypeTok{quantile =} \KeywordTok{c}\NormalTok{(}\KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{10}\NormalTok{, }\DataTypeTok{to =} \DecValTok{100}\NormalTok{, }\DataTypeTok{by =} \DecValTok{10}\NormalTok{)),}
    \DataTypeTok{cent.tend =}\NormalTok{ sd, }\DataTypeTok{type =} \StringTok{"c"}\NormalTok{, }\DataTypeTok{elements =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\StringTok{"black"}\NormalTok{, }\KeywordTok{rainbow}\NormalTok{(}\DecValTok{10}\NormalTok{)),}
    \DataTypeTok{ylab =} \KeywordTok{c}\NormalTok{(}\StringTok{"Disparity"}\NormalTok{, }\StringTok{"Diversity"}\NormalTok{), }\DataTypeTok{chrono.subsets =} \OtherTok{FALSE}\NormalTok{,}
    \DataTypeTok{xlab =} \StringTok{"Time (in in units from past to present)"}\NormalTok{, }\DataTypeTok{observed =} \OtherTok{TRUE}\NormalTok{,}
    \DataTypeTok{main =} \StringTok{"Many more options..."}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in plot.window(...): "quantile" is not a graphical parameter
\end{verbatim}

\begin{verbatim}
## Warning in plot.xy(xy, type, ...): "quantile" is not a graphical parameter
\end{verbatim}

\begin{verbatim}
## Warning in axis(side = side, at = at, labels = labels, ...): "quantile" is
## not a graphical parameter

## Warning in axis(side = side, at = at, labels = labels, ...): "quantile" is
## not a graphical parameter
\end{verbatim}

\begin{verbatim}
## Warning in box(...): "quantile" is not a graphical parameter
\end{verbatim}

\begin{verbatim}
## Warning in title(...): "quantile" is not a graphical parameter
\end{verbatim}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-56-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Resetting graphical parameters}
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

\begin{quote}
Note that the argument \texttt{observed\ =\ TRUE} allows to plot the
disparity values calculated from the non-bootstrapped data as crosses on
the plot.
\end{quote}

For comparing results, it is also possible to add a plot to the existent
plot by using \texttt{add\ =\ TRUE}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting the continuous disparity with a fixed y axis}
\KeywordTok{plot}\NormalTok{(disparity_time_slices, }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{4}\NormalTok{))}
\NormalTok{## Adding the discrete data}
\KeywordTok{plot}\NormalTok{(disparity_time_bins, }\DataTypeTok{type =} \StringTok{"line"}\NormalTok{, }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{4}\NormalTok{), }\DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{,}
    \DataTypeTok{add =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-57-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Resetting graphical parameters}
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

Finally, if your data has been fully rarefied, it is also possible to
easily look at rarefaction curves by using the
\texttt{rarefaction\ =\ TRUE} argument:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting the rarefaction curves}
\KeywordTok{plot}\NormalTok{(disparity_crown_stem, }\DataTypeTok{rarefaction =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-58-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Resetting graphical parameters}
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

\section{Testing disparity
hypotheses}\label{testing-disparity-hypotheses}

The \texttt{dispRity} package allows users to apply statistical tests to
the calculated disparity to test various hypotheses. The function
\texttt{test.dispRity} works in a similar way to the \texttt{dispRity}
function: it takes a \texttt{dispRity} object, a \texttt{test} and a
\texttt{comparisons} argument.

The \texttt{comparisons} argument indicates the way the test should be
applied to the data:

\begin{itemize}
\tightlist
\item
  \texttt{pairwise} (default): to compare each subset in a pairwise
  manner
\item
  \texttt{referential}: to compare each subset to the first subset
\item
  \texttt{sequential}: to compare each subset to the following subset
\item
  \texttt{all}: to compare all the subsets together (like in analysis of
  variance)
\end{itemize}

It is also possible to input a list of pairs of \texttt{numeric} values
or \texttt{characters} matching the subset names to create personalised
tests. Some other tests implemented in \texttt{dispRity} such as the
\texttt{dispRity::null.test} have a specific way they are applied to the
data and therefore ignore the \texttt{comparisons} argument.

The \texttt{test} argument can be any statistical or non-statistical
test to apply to the disparity object. It can be a common statistical
test function (e.g. \texttt{stats::t.test}), a function implemented in
\texttt{dispRity} (e.g.~see \texttt{?null.test}) or any function defined
by the user.

This function also allows users to correct for Type I error inflation
(false positives) when using multiple comparisons via the
\texttt{correction} argument. This argument can be empty (no correction
applied) or can contain one of the corrections from the
\texttt{stats::p.adjust} function (see \texttt{?p.adjust}).

Note that the \texttt{test.dispRity} algorithm deals with some classical
test outputs and summarises the test output. It is, however, possible to
get the full detailed output by using the options
\texttt{details\ =\ TRUE}.

Here we are using the variables generated in the
\protect\hyperlink{summarising-dispRity-data-plots}{section above}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## T-test to test for a difference in disparity between crown and stem mammals}
\KeywordTok{test.dispRity}\NormalTok{(disparity_crown_stem, }\DataTypeTok{test =}\NormalTok{ t.test)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [[1]]
##              statistic: t
## crown : stem     66.67693
## 
## [[2]]
##              parameter: df
## crown : stem      169.1446
## 
## [[3]]
##                    p.value
## crown : stem 2.275095e-123
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Performing the same test but with the detailed t.test output}
\KeywordTok{test.dispRity}\NormalTok{(disparity_crown_stem, }\DataTypeTok{test =}\NormalTok{ t.test, }\DataTypeTok{details =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## $`crown : stem`
## $`crown : stem`[[1]]
## 
##  Welch Two Sample t-test
## 
## data:  dots[[1L]][[1L]] and dots[[2L]][[1L]]
## t = 66.677, df = 169.14, p-value < 2.2e-16
## alternative hypothesis: true difference in means is not equal to 0
## 95 percent confidence interval:
##  0.2894967 0.3071619
## sample estimates:
## mean of x mean of y 
##  1.928128  1.629799
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Wilcoxon test applied to time sliced disparity with sequential comparisons,}
\NormalTok{## with Bonferroni correction}
\KeywordTok{test.dispRity}\NormalTok{(disparity_time_slices, }\DataTypeTok{test =}\NormalTok{ wilcox.test,}
              \DataTypeTok{comparisons =} \StringTok{"sequential"}\NormalTok{, }\DataTypeTok{correction =} \StringTok{"bonferroni"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [[1]]
##          statistic: W
## 120 : 80           27
## 80 : 40          3466
## 40 : 0            455
## 
## [[2]]
##               p.value
## 120 : 80 1.720545e-33
## 80 : 40  5.370214e-04
## 40 : 0   3.599349e-28
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring the overlap between distributions in the time bins (using the}
\NormalTok{## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)}
\KeywordTok{test.dispRity}\NormalTok{(disparity_time_bins, }\DataTypeTok{test =}\NormalTok{ bhatt.coeff)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in test.dispRity(disparity_time_bins, test = bhatt.coeff): Multiple p-values will be calculated without adjustment!
## This can inflate Type I error!
\end{verbatim}

\begin{verbatim}
##                    bhatt.coeff
## 120 - 80 : 80 - 40    0.000000
## 120 - 80 : 40 - 0     0.000000
## 80 - 40 : 40 - 0      0.490448
\end{verbatim}

Because of the modular design of the package, tests can always be made
by the user (the same way disparity metrics can be user made). The only
condition is that the test can be applied to at least two distributions.
In practice, the \texttt{test.dispRity} function will pass the
calculated disparity data (distributions) to the provided function in
either pairs of distributions (if the \texttt{comparisons} argument is
set to \texttt{pairwise}, \texttt{referential} or \texttt{sequential})
or a table containing all the distributions
(\texttt{comparisons\ =\ all}; this should be in the same format as data
passed to \texttt{lm}-type functions for example).

\subsection{\texorpdfstring{NPMANOVA in
\texttt{dispRity}}{NPMANOVA in dispRity}}\label{npmanova-in-disprity}

One often useful test to apply to multidimensional data is the
permutational multivariate analysis of variance based on distance
matrices \texttt{vegan::adonis}. This can be done on \texttt{dispRity}
objects using the \texttt{adonis.dispRity} wrapper function. Basically,
this function takes the exact same arguments as \texttt{adonis} and a
\texttt{dispRity} object for data and performs a PERMANOVA based on the
distance matrix of the multidimensional space (unless the
multidimensional space was already defined as a distance matrix). The
\texttt{adonis.dispRity} function uses the information from the
\texttt{dispRity} object to generate default formulas:

\begin{itemize}
\tightlist
\item
  If the object contains customised subsets, it applies the default
  formula \texttt{matrix\ \textasciitilde{}~group} testing the effect of
  \texttt{group} as a predictor on \texttt{matrix} (called from the
  \texttt{dispRity} object as \texttt{data\$matrix} see
  \protect\hyperlink{The-dispRity-object-content}{\texttt{dispRitu}
  object details})
\item
  If the object contains time subsets, it applies the default formula
  \texttt{matrix\ \textasciitilde{}\ time} testing the effect of
  \texttt{time} as a predictor (were the different levels of
  \texttt{time} are the different time slices/bins)
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{## Generating a random character matrix}
\NormalTok{character_matrix <-}\StringTok{ }\KeywordTok{sim.morpho}\NormalTok{(}\KeywordTok{rtree}\NormalTok{(}\DecValTok{20}\NormalTok{), }\DecValTok{50}\NormalTok{, }\DataTypeTok{rates =} \KeywordTok{c}\NormalTok{(rnorm, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{))}

\NormalTok{## Calculating the distance matrix}
\NormalTok{distance_matrix <-}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(}\KeywordTok{dist}\NormalTok{(character_matrix))}

\NormalTok{## Creating two groups}
\NormalTok{random_groups <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\StringTok{"group1"}\NormalTok{ =}\StringTok{ }\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{, }\StringTok{"group2"}\NormalTok{ =}\StringTok{ }\DecValTok{11}\OperatorTok{:}\DecValTok{20}\NormalTok{)}

\NormalTok{## Generating a dispRity object}
\NormalTok{random_disparity <-}\StringTok{ }\KeywordTok{custom.subsets}\NormalTok{(distance_matrix, random_groups)}

\NormalTok{## Running a default NPMANOVA}
\KeywordTok{adonis.dispRity}\NormalTok{(random_disparity)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Call:
## vegan::adonis(formula = matrix ~ group, data = random_disparity,      method = "euclidean") 
## 
## Permutation: free
## Number of permutations: 999
## 
## Terms added sequentially (first to last)
## 
##           Df SumsOfSqs MeanSqs F.Model      R2 Pr(>F)  
## group      1      16.5  16.500  1.3904 0.07171  0.095 .
## Residuals 18     213.6  11.867         0.92829         
## Total     19     230.1                 1.00000         
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

Of course, it is possible to pass customised formulas if the disparity
object contains more more groups. In that case the predictors must
correspond to the names of the groups explained data must be set as
\texttt{matrix}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating two groups with two states each}
\NormalTok{groups <-}\StringTok{ }\KeywordTok{as.data.frame}\NormalTok{(}\KeywordTok{matrix}\NormalTok{(}\DataTypeTok{data =} \KeywordTok{c}\NormalTok{(}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{10}\NormalTok{), }\KeywordTok{rep}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{10}\NormalTok{), }\KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{), }\DecValTok{10}\NormalTok{)),}
         \DataTypeTok{nrow =} \DecValTok{20}\NormalTok{, }\DataTypeTok{ncol =} \DecValTok{2}\NormalTok{, }\DataTypeTok{dimnames =} \KeywordTok{list}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(}\StringTok{"t"}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{20}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\StringTok{"g1"}\NormalTok{, }\StringTok{"g2"}\NormalTok{))))}

\NormalTok{## Creating the dispRity object}
\NormalTok{multi_groups <-}\StringTok{ }\KeywordTok{custom.subsets}\NormalTok{(distance_matrix, groups)}

\NormalTok{## Running the NPMANOVA}
\KeywordTok{adonis.dispRity}\NormalTok{(multi_groups, matrix }\OperatorTok{~}\StringTok{ }\NormalTok{g1 }\OperatorTok{+}\StringTok{ }\NormalTok{g2)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Call:
## vegan::adonis(formula = matrix ~ g1 + g2, data = multi_groups,      method = "euclidean") 
## 
## Permutation: free
## Number of permutations: 999
## 
## Terms added sequentially (first to last)
## 
##           Df SumsOfSqs MeanSqs F.Model      R2 Pr(>F)
## g1         1      16.5  16.500 1.34921 0.07171  0.111
## g2         1       5.7   5.700 0.46609 0.02477  0.998
## Residuals 17     207.9  12.229         0.90352       
## Total     19     230.1                 1.00000
\end{verbatim}

Finally, it is possible to use objects generated by
\texttt{chrono.subsets}. In this case, \texttt{adonis.dispRity} will
applied the \texttt{matrix\ \textasciitilde{}\ time} formula by default:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating time series}
\NormalTok{time_subsets <-}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(BeckLee_mat50, BeckLee_tree, }
     \DataTypeTok{method =} \StringTok{"discrete"}\NormalTok{, }\DataTypeTok{inc.nodes =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{time =} \KeywordTok{c}\NormalTok{(}\DecValTok{100}\NormalTok{, }\DecValTok{85}\NormalTok{, }\DecValTok{65}\NormalTok{, }\DecValTok{0}\NormalTok{),}
     \DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages)}

\NormalTok{## Running the NPMANOVA with time as a predictor}
\KeywordTok{adonis.dispRity}\NormalTok{(time_subsets)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in adonis.dispRity(time_subsets): The input data for adonis.dispRity was not a distance matrix.
## The results are thus based on the distance matrix for the input data (i.e. dist(data$matrix)).
## Make sure that this is the desired methodological approach!
\end{verbatim}

\begin{verbatim}
## 
## Call:
## vegan::adonis(formula = dist(matrix) ~ time, data = time_subsets,      method = "euclidean") 
## 
## Permutation: free
## Number of permutations: 999
## 
## Terms added sequentially (first to last)
## 
##           Df SumsOfSqs MeanSqs F.Model      R2 Pr(>F)    
## time       2     8.049  4.0245  2.1303 0.08311  0.001 ***
## Residuals 47    88.792  1.8892         0.91689           
## Total     49    96.841                 1.00000           
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

Note that the function warns you that the input data was transformed
into a distance matrix. This is reflected in the Call part of the output
(\texttt{formula\ =\ dist(matrix)\ \textasciitilde{}\ time}).

To use each time subset as a separate predictor, you can use the
\texttt{matrix\ \textasciitilde{}\ chrono.subsets} formula; this is
equivalent to
\texttt{matrix\ \textasciitilde{}\ first\_time\_subset\ +\ second\_time\_subset\ +\ ...}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Running the NPMANOVA with each time bin as a predictor}
\KeywordTok{adonis.dispRity}\NormalTok{(time_subsets, matrix }\OperatorTok{~}\StringTok{ }\NormalTok{chrono.subsets)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in adonis.dispRity(time_subsets, matrix ~ chrono.subsets): The input data for adonis.dispRity was not a distance matrix.
## The results are thus based on the distance matrix for the input data (i.e. dist(data$matrix)).
## Make sure that this is the desired methodological approach!
\end{verbatim}

\begin{verbatim}
## 
## Call:
## vegan::adonis(formula = dist(matrix) ~ chrono.subsets, data = time_subsets,      method = "euclidean") 
## 
## Permutation: free
## Number of permutations: 999
## 
## Terms added sequentially (first to last)
## 
##           Df SumsOfSqs MeanSqs F.Model      R2 Pr(>F)    
## t100to85   1     3.090  3.0897  1.6354 0.03190  0.002 ** 
## t85to65    1     4.959  4.9593  2.6251 0.05121  0.001 ***
## Residuals 47    88.792  1.8892         0.91689           
## Total     49    96.841                 1.00000           
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\subsection{\texorpdfstring{\texttt{geiger::dtt} model fitting in
\texttt{dispRity}}{geiger::dtt model fitting in dispRity}}\label{geigerdtt-model-fitting-in-disprity}

The \texttt{dtt} function from the \texttt{geiger} package is also often
used to compare a trait's disparity observed in living taxa to the
disparity of a simulated trait based on a given phylogeny. The
\texttt{dispRity} package proposes a wrapper function for
\texttt{geiger::dtt}, \texttt{dtt.dispRity} that allows the use of any
disparity metric. Unfortunately, this implementation is slower that
\texttt{geiger::dtt} (so if you're using the metrics implemented in
\texttt{geiger} prefer the original version) and, as the original
function, is limited to ultrametric trees (only living taxa!)\ldots{}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{require}\NormalTok{(geiger)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Loading required package: geiger
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{geiger_data <-}\StringTok{ }\KeywordTok{get}\NormalTok{(}\KeywordTok{data}\NormalTok{(geospiza))}

\NormalTok{## Calculate the disparity of the dataset using the sum of variance}
\NormalTok{dispRity_dtt <-}\StringTok{ }\KeywordTok{dtt.dispRity}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ geiger_data}\OperatorTok{$}\NormalTok{dat, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
                             \DataTypeTok{tree =}\NormalTok{ geiger_data}\OperatorTok{$}\NormalTok{phy, }\DataTypeTok{nsim =} \DecValTok{100}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in dtt.dispRity(data = geiger_data$dat, metric = c(sum,
## variances), : The following tip(s) was not present in the data: olivacea.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Plotting the results}
\KeywordTok{plot}\NormalTok{(dispRity_dtt, }\DataTypeTok{fig.width=}\DecValTok{8}\NormalTok{, }\DataTypeTok{fig.height=}\DecValTok{8}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-64-1.pdf}

Note that, like in the original \texttt{dtt} function, it is possible to
change the evolutionary model (see \texttt{?geiger::sim.char}
documentation).

\subsection{\texorpdfstring{null morphospace testing with
\texttt{null.test}}{null morphospace testing with null.test}}\label{null-morphospace-testing-with-null.test}

This test is equivalent to the test performed in \citet{diaz2016global}.
It compares the disparity measured in the observed space to the
disparity measured in a set of simulated spaces. These simulated spaces
can be built with based on the hypothesis assumptions: for example, we
can test whether our space is normal.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
\NormalTok{## A "normal" multidimensional space with 50 dimensions and 10 elements}
\NormalTok{normal_space <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\KeywordTok{rnorm}\NormalTok{(}\DecValTok{1000}\NormalTok{), }\DataTypeTok{ncol =} \DecValTok{50}\NormalTok{)}

\NormalTok{## Calculating the disparity as the average pairwise distances}
\NormalTok{obs_disparity <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(normal_space, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(mean, pairwise.dist))}

\NormalTok{## Testing against 100 randomly generated normal spaces}
\NormalTok{(results <-}\StringTok{ }\KeywordTok{null.test}\NormalTok{(obs_disparity, }\DataTypeTok{replicates =} \DecValTok{100}\NormalTok{, }\DataTypeTok{null.distrib =}\NormalTok{ rnorm))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Monte-Carlo test
## Call: [1] "dispRity::null.test"
## 
## Observation: 9.910536 
## 
## Based on 100 replicates
## Simulated p-value: 0.7524752 
## Alternative hypothesis: two-sided 
## 
##     Std.Obs Expectation    Variance 
## -0.24885893  9.96420000  0.04650127
\end{verbatim}

Here the results show that disparity measured in our observed space is
not significantly different than the one measured in a normal space. We
can then propose that our observed space is normal!

These results have an attributed \texttt{dispRity} and \texttt{randtest}
class and can be plotted as \texttt{randtest} objects using the
\texttt{dispRity} S3 \texttt{plot} method:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Plotting the results}
\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{main =} \StringTok{"Is this space normal?"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-66-1.pdf}

For more details on generating spaces see the
\protect\hyperlink{Simulating-multidimensional-spaces}{\texttt{space.maker}}
function tutorial.

\section{Fitting modes of evolution to disparity
data}\label{fitting-modes-of-evolution-to-disparity-data}

The code used for these models is based on those developed by Gene Hunt
\citep{hunt2006fitting, hunt2012measuring, hunt2015simple}. So we
acknowledge and thank Gene Hunt for developing these models and writing
the original R code that served as inspiration for these models.

\subsection{Simple modes of disparity change through
time}\label{simple-modes-of-disparity-change-through-time}

\subsubsection{\texorpdfstring{\texttt{model.test}}{model.test}}\label{model.test}

Changes in disparity-through-time can follow a range of models, such as
random walks, stasis, constrained evolution, trends, or an early burst
model of evolution. We will start with by fitting the simplest modes of
evolution to our data. For example we may have a null expectation of
time-invariant change in disparity in which values fluctuate with a
variance around the mean - this would be best describe by a Stasis
model:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading premade disparity data}
\KeywordTok{data}\NormalTok{(BeckLee_disparity)}
\NormalTok{disp_time <-}\StringTok{ }\KeywordTok{model.test}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_disparity, }\DataTypeTok{model =} \StringTok{"Stasis"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Evidence of equal variance (Bartlett's test of equal variances p = 0).
## Variance is not pooled.
## Running Stasis model...Done. Log-likelihood = -59.501
\end{verbatim}

We can see the standard output from \texttt{model.test}. The first
output message tells us it has tested for equal variances in each
sample. The model uses Bartlett's test of equal variances to assess if
variances are equal, so if p \textgreater{} 0.05 then variance is
treated as the same for all samples, but if (p \textless{} 0.05) then
each bin variance is unique. Here we have p \textless{} 0.05, so
variance is not pooled between samples.

By default \texttt{model.test} will use Bartlett's test to assess for
homogeneity of variances, and then use this to decide to pool variances
or not. This is ignored if the argument \texttt{pool.variance} in
\texttt{model.test} is changed from the default \texttt{NULL} to
\texttt{TRUE} or \texttt{FALSE}. For example, to ignore Bartlett's test
and pool variances manually we would do the following:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{disp_time_pooled <-}\StringTok{ }\KeywordTok{model.test}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_disparity, }\DataTypeTok{model =} \StringTok{"Stasis"}\NormalTok{, }\DataTypeTok{pool.variance =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Running Stasis model...Done. Log-likelihood = -58.233
\end{verbatim}

However, unless you have good reason to choose otherwise it is
recommended to use the default of \texttt{pool.variance\ =\ NULL}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{disp_time <-}\StringTok{ }\KeywordTok{model.test}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_disparity, }\DataTypeTok{model =} \StringTok{"Stasis"}\NormalTok{, }\DataTypeTok{pool.variance =} \OtherTok{NULL}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Evidence of equal variance (Bartlett's test of equal variances p = 0).
## Variance is not pooled.
## Running Stasis model...Done. Log-likelihood = -59.501
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{disp_time}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Disparity evolution model fitting:
## Call: model.test(data = BeckLee_disparity, model = "Stasis", pool.variance = NULL) 
## 
##            aicc delta_aicc weight_aicc
## Stasis 123.1027          0           1
## 
## Use x$full.details for displaying the models details
## or summary(x) for summarising them.
\end{verbatim}

The remaining output gives us the log-likelihood of the Stasis model of
-59.501 (you may notice this change when we pooled variances above). The
output also gives us the small sample Akaike Information Criterion
(AICc), the delta AICc (the distance from the best fitting model), and
the AICc weights (\textasciitilde{}the relative support of this model
compared to all models, scaled to one).

These are all metrics of relative fit, so when we test a single model
they are not useful. By using the function summary in \texttt{dispRity}
we can see the maximum likelihood estimates of the model parameters:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(disp_time)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##         aicc delta_aicc weight_aicc log.lik param theta.1 omega
## Stasis 123.1          0           1   -59.5     2     3.4   0.1
\end{verbatim}

So we again see the AICc, delta AICc, AICc weight, and the
log-likelihood we saw previously. We now also see the number of
parameters from the model (2: theta and omega), and their estimates so
the variance (omega = 0.01) and the mean (theta.1 = 3.4).

The \texttt{model.test} function is designed to test relative model fit,
so we need to test more than one model to make relative comparisons. So
let's compare to the fit of the Stasis model to another model with two
parameters: the Brownian motion. Brownian motion assumes a constant mean
that is equal to the ancestral estimate of the sequence, and the
variance around this mean increases linearly with time. The easier way
to compare these models is to simply add \texttt{"BM"} to the
\texttt{models} vector argument:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{disp_time <-}\StringTok{ }\KeywordTok{model.test}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_disparity, }\DataTypeTok{model =} \KeywordTok{c}\NormalTok{(}\StringTok{"Stasis"}\NormalTok{, }\StringTok{"BM"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Evidence of equal variance (Bartlett's test of equal variances p = 0).
## Variance is not pooled.
## Running Stasis model...Done. Log-likelihood = -59.501
## Running BM model...Done. Log-likelihood = 123.938
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{disp_time}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Disparity evolution model fitting:
## Call: model.test(data = BeckLee_disparity, model = c("Stasis", "BM")) 
## 
##             aicc delta_aicc  weight_aicc
## Stasis  123.1027   366.8774 2.155677e-80
## BM     -243.7747     0.0000 1.000000e+00
## 
## Use x$full.details for displaying the models details
## or summary(x) for summarising them.
\end{verbatim}

Et voilà! Here we can see by the log-likelihood, AICc, delta AICc, and
AICc weight Brownian motion has a much better relative fit to these data
than the Stasis model. Brownian motion has a relative AICc fit 366 units
better than Stasis, and virtually has a AICc weight of 1.

We can also all the information about the relative fit of models
alongside the maximum likelihood estimates of model parameters using the
summary function

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(disp_time)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##        aicc delta_aicc weight_aicc log.lik param theta.1 omega
## Stasis  123      366.9           0   -59.5     2   3.403  0.15
## BM     -244        0.0           1   123.9     2      NA    NA
##        ancestral state sigma squared
## Stasis              NA            NA
## BM               2.858         0.003
\end{verbatim}

Not that because the parameters per models differ, the summary includes
\texttt{NA} for inapplicable parameters per models (e.g.~the theta and
omega parameters from the Stasis models are inapplicable for a Brownian
motion model).

We can plot the relative fit of our models using the \texttt{plot}
function

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot}\NormalTok{(disp_time)}
\end{Highlighting}
\end{Shaded}

\begin{figure}

{\centering \includegraphics{dispRity_manual_files/figure-latex/plot1-1}

}

\caption{relative fit (AICc weight) of Stasis and Brownian models of disparity through time}\label{fig:plot1}
\end{figure}

Here we see and overwhelming support for the Brownian motion model.

Alternatively, we could test all available models single modes: Stasis,
Brownian motion, Ornstein-Uhlenbeck (evolution constrained to an
optima), Trend (increasing or decreasing mean through time), and Early
Burst (exponentially decreasing rate through time)

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{disp_time <-}\StringTok{ }\KeywordTok{model.test}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_disparity, }\DataTypeTok{model =} \KeywordTok{c}\NormalTok{(}\StringTok{"Stasis"}\NormalTok{, }\StringTok{"BM"}\NormalTok{, }\StringTok{"OU"}\NormalTok{, }\StringTok{"Trend"}\NormalTok{, }\StringTok{"EB"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Evidence of equal variance (Bartlett's test of equal variances p = 0).
## Variance is not pooled.
## Running Stasis model...Done. Log-likelihood = -59.501
## Running BM model...Done. Log-likelihood = 123.938
## Running OU model...Done. Log-likelihood = 126.431
## Running Trend model...Done. Log-likelihood = 126.361
## Running EB model...Done. Log-likelihood = 113.081
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(disp_time)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##        aicc delta_aicc weight_aicc log.lik param theta.1 omega
## Stasis  123      369.6       0.000   -59.5     2   3.403  0.15
## BM     -244        2.7       0.157   123.9     2      NA    NA
## OU     -245        2.0       0.227   126.4     4      NA    NA
## Trend  -247        0.0       0.617   126.4     3      NA    NA
## EB     -220       26.6       0.000   113.1     3      NA    NA
##        ancestral state sigma squared alpha optima.1 trend     eb
## Stasis              NA            NA    NA       NA    NA     NA
## BM               2.858         0.003    NA       NA    NA     NA
## OU               2.835         0.002 0.004    5.707    NA     NA
## Trend            2.839         0.002    NA       NA  0.01     NA
## EB               4.055         0.002    NA       NA    NA -0.014
\end{verbatim}

These models indicate support for a Trend model, and we can plot the
relative support of all model AICc weights

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot}\NormalTok{(disp_time)}
\end{Highlighting}
\end{Shaded}

\begin{figure}

{\centering \includegraphics{dispRity_manual_files/figure-latex/plot2-1}

}

\caption{relative fit (AICc weight) of various modes of evolution}\label{fig:plot2}
\end{figure}

Is this a trend of increasing or decreasing disparity through time? One
way to find out is to look at the summary function for the Trend model:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(disp_time)[}\StringTok{"Trend"}\NormalTok{,]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##            aicc      delta_aicc     weight_aicc         log.lik 
##        -247.000           0.000           0.617         126.400 
##           param         theta.1           omega ancestral state 
##           3.000              NA              NA           2.839 
##   sigma squared           alpha        optima.1           trend 
##           0.002              NA              NA           0.010 
##              eb 
##              NA
\end{verbatim}

This show a positive trend (0.01) of increasing disparity through time.

\subsection{Plot and run simulation tests in a single
step}\label{plot-and-run-simulation-tests-in-a-single-step}

\subsubsection{\texorpdfstring{\texttt{model.test.wrapper}}{model.test.wrapper}}\label{model.test.wrapper}

Patterns of evolution can be fit using \texttt{model.test}, but the
\texttt{model.test.wrapper} fits the same models as \texttt{model.test}
as well as running predictive tests and plots.

The predictive tests use the maximum likelihood estimates of model
parameters to simulate a number of datasets (default = 1000), and
analyse whether this is significantly different to the empirical input
data using the Rank Envelope test \citep{murrell2018global}. Finally we
can plot the empirical data, simulated data, and the Rank Envelope test
p values. This can all be done using the function
\texttt{model.test.wrapper}, and we will set the argument
\texttt{show.p\ =\ TRUE} so \emph{p} values from the Rank Envelope test
are printed on the plot:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{disp_time <-}\StringTok{ }\KeywordTok{model.test.wrapper}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_disparity, }\DataTypeTok{model =} \KeywordTok{c}\NormalTok{(}\StringTok{"Stasis"}\NormalTok{, }\StringTok{"BM"}\NormalTok{, }\StringTok{"OU"}\NormalTok{, }\StringTok{"Trend"}\NormalTok{, }\StringTok{"EB"}\NormalTok{),}
                                \DataTypeTok{show.p =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Evidence of equal variance (Bartlett's test of equal variances p = 0).
## Variance is not pooled.
## Running Stasis model...Done. Log-likelihood = -59.501
## Running BM model...Done. Log-likelihood = 123.938
## Running OU model...Done. Log-likelihood = 126.431
## Running Trend model...Done. Log-likelihood = 126.361
## Running EB model...Done. Log-likelihood = 113.081
\end{verbatim}

\begin{figure}

{\centering \includegraphics{dispRity_manual_files/figure-latex/plot3-1}

}

\caption{Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for Trend, OU, BM, EB, and Stasis models}\label{fig:plot3}
\end{figure}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{disp_time}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##        aicc delta_aicc weight_aicc log.lik param theta.1 omega
## Trend  -247        0.0       0.617   126.4     3      NA    NA
## OU     -245        2.0       0.227   126.4     4      NA    NA
## BM     -244        2.7       0.157   123.9     2      NA    NA
## EB     -220       26.6       0.000   113.1     3      NA    NA
## Stasis  123      369.6       0.000   -59.5     2   3.403  0.15
##        ancestral state sigma squared alpha optima.1 trend     eb
## Trend            2.839         0.002    NA       NA  0.01     NA
## OU               2.835         0.002 0.004    5.707    NA     NA
## BM               2.858         0.003    NA       NA    NA     NA
## EB               4.055         0.002    NA       NA    NA -0.014
## Stasis              NA            NA    NA       NA    NA     NA
##        median p value lower p value upper p value
## Trend      0.97902098   0.978021978     0.9800200
## OU         0.93406593   0.934065934     0.9340659
## BM         0.33266733   0.322677323     0.3426573
## EB         0.06293706   0.000999001     0.1248751
## Stasis     1.00000000   1.000000000     1.0000000
\end{verbatim}

From this plot we can see the empirical estimates of disparity through
time (pink) compared to the predictive data based upon the simulations
using the estimated parameters from each model. There is no significant
differences between the empirical data and simulated data, except for
the Early Burst model.

Trend is the best-fitting model but the plot suggests the OU model also
follows a trend-like pattern. This is because the optima for the OU
model (5.7) is different to the ancestral state 2.835 and outside the
observed value. This is potentially unrealistic, and one way to
alleviate this issue is to set the optima of the OU model to equal the
ancestral estimate - this is the normal practice for OU models in
comparative phylogenetics. To set the optima to the ancestral value we
change the argument \texttt{fixed.optima\ =\ TRUE}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{disp_time <-}\StringTok{ }\KeywordTok{model.test.wrapper}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_disparity, }\DataTypeTok{model =} \KeywordTok{c}\NormalTok{(}\StringTok{"Stasis"}\NormalTok{, }\StringTok{"BM"}\NormalTok{, }\StringTok{"OU"}\NormalTok{, }\StringTok{"Trend"}\NormalTok{, }\StringTok{"EB"}\NormalTok{),}
                                \DataTypeTok{show.p =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{fixed.optima =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Evidence of equal variance (Bartlett's test of equal variances p = 0).
## Variance is not pooled.
## Running Stasis model...Done. Log-likelihood = -59.501
## Running BM model...Done. Log-likelihood = 123.938
## Running OU model...Done. Log-likelihood = 123.938
## Running Trend model...Done. Log-likelihood = 126.361
## Running EB model...Done. Log-likelihood = 113.081
\end{verbatim}

\begin{figure}

{\centering \includegraphics{dispRity_manual_files/figure-latex/plot4-1}

}

\caption{Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for Trend, OU, BM, EB, and Stasis models with the optima of the OU model set to equal the ancestral value}\label{fig:plot4}
\end{figure}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{disp_time}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##        aicc delta_aicc weight_aicc log.lik param theta.1 omega
## Trend  -247        0.0       0.745   126.4     3      NA    NA
## BM     -244        2.7       0.189   123.9     2      NA    NA
## OU     -242        4.8       0.066   123.9     3      NA    NA
## EB     -220       26.6       0.000   113.1     3      NA    NA
## Stasis  123      369.6       0.000   -59.5     2   3.403  0.15
##        ancestral state sigma squared alpha trend     eb median p value
## Trend            2.839         0.002    NA  0.01     NA     0.97402597
## BM               2.858         0.003    NA    NA     NA     0.29870130
## OU               2.858         0.003     0    NA     NA     0.35864136
## EB               4.055         0.002    NA    NA -0.014     0.06393606
## Stasis              NA            NA    NA    NA     NA     1.00000000
##        lower p value upper p value
## Trend    0.973026973     0.9750250
## BM       0.285714286     0.3116883
## OU       0.347652348     0.3696304
## EB       0.000999001     0.1268731
## Stasis   1.000000000     1.0000000
\end{verbatim}

The relative fit of the OU model is decreased by constraining the fit of
the optima to equal the ancestral state value. In fact as the OU
attraction parameter (alpha) is zero, the model is equal to a Brownian
motion model but is penalised by having an extra parameter. Note that
indeed, the plots of the BM model and the OU model look nearly
identical.

\subsection{Multiple modes of evolution (time
shifts)}\label{multiple-modes-of-evolution-time-shifts}

As well as fitting a single model to a sequence of disparity values we
can also allow for the mode of evolution to shift at a single or
multiple points in time. The timing of a shift in mode can be based on
an a prior expectation, such as a mass extinction event, or the model
can test multiple points to allow to find time shift point with the
highest likelihood.

Models can be fit using \texttt{model.test} but it can be more
convenient to use \texttt{model.test.wrapper}. Here we will compare the
relative fit of Brownian motion, Trend, Ornstein-Uhlenbeck and a
multi-mode Ornstein Uhlenbck model in which the optima changes at 66
million years ago, the Cretaceous-Palaeogene boundary.

For example, we could be testing the hypothesis that the extinction of
non-avian dinosaurs allowed mammals to go from scurrying in the
undergrowth (low optima/low disparity) to dominating all habitats (high
optima/high disparity). We will constrain the optima of OU model in the
first time begin (i.e, pre-66 Mya) to equal the ancestral value:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{disp_time <-}\StringTok{ }\KeywordTok{model.test.wrapper}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_disparity, }\DataTypeTok{model =} \KeywordTok{c}\NormalTok{(}\StringTok{"BM"}\NormalTok{, }\StringTok{"Trend"}\NormalTok{, }\StringTok{"OU"}\NormalTok{, }\StringTok{"multi.OU"}\NormalTok{),}
                                \DataTypeTok{time.split =} \DecValTok{66}\NormalTok{, }\DataTypeTok{pool.variance =} \OtherTok{NULL}\NormalTok{, }\DataTypeTok{show.p =} \OtherTok{TRUE}\NormalTok{,}
                                \DataTypeTok{fixed.optima =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Evidence of equal variance (Bartlett's test of equal variances p = 0).
## Variance is not pooled.
## Running BM model...Done. Log-likelihood = 123.938
## Running Trend model...Done. Log-likelihood = 126.361
## Running OU model...Done. Log-likelihood = 123.938
## Running multi.OU model...Done. Log-likelihood = 126.469
\end{verbatim}

\begin{figure}

{\centering \includegraphics{dispRity_manual_files/figure-latex/plot5-1}

}

\caption{Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for BM, Trend, OU, and multi OU models with a shift in optima allowed at 66 Ma}\label{fig:plot5}
\end{figure}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{disp_time}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##          aicc delta_aicc weight_aicc log.lik param ancestral state
## Trend    -247      0.000       0.580   126.4     3           2.839
## multi.OU -245      1.922       0.222   126.5     4           2.836
## BM       -244      2.742       0.147   123.9     2           2.858
## OU       -242      4.845       0.051   123.9     3           2.858
##          sigma squared trend alpha optima.2 median p value lower p value
## Trend            0.002  0.01    NA       NA      0.9920080     0.9920080
## multi.OU         0.002    NA 0.005    5.526      0.6308691     0.6263736
## BM               0.003    NA    NA       NA      0.3506494     0.3396603
## OU               0.003    NA 0.000       NA      0.3666334     0.3576424
##          upper p value
## Trend        0.9920080
## multi.OU     0.6353646
## BM           0.3616384
## OU           0.3756244
\end{verbatim}

The multi-OU model shows an increase an optima at the
Cretaceous-Palaeogene boundary, indicating a shift in disparity.
However, this model does not fit as well as a model in which there is an
increasing trend through time. We can also fit a model in which the we
specify a heterogeneous model but we do not give a \texttt{time.split}.
In this instance the model will test all splits that have at least 10
time slices on either side of the split. That's 102 potential time
shifts in this example dataset so be warned, the following code will
estimate 105 models!

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## An example of a time split model in which all potential splits are tested}
\NormalTok{## }\AlertTok{WARNING}\NormalTok{: this will take between 20 minutes and half and hour to run!}
\NormalTok{disp_time <-}\StringTok{ }\KeywordTok{model.test.wrapper}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_disparity, }\DataTypeTok{model =} \KeywordTok{c}\NormalTok{(}\StringTok{"BM"}\NormalTok{, }\StringTok{"Trend"}\NormalTok{, }\StringTok{"OU"}\NormalTok{, }\StringTok{"multi.OU"}\NormalTok{),}
                                \DataTypeTok{show.p =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{fixed.optima =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

As well as specifying a multi-OU model we can run any combination of
models. For example we could fit a model at the Cretaceous-Palaeogene
boundary that goes from an OU to a BM model, a Trend to an OU model, a
Stasis to a Trend model or any combination you want to use. The only
model that can't be used in combination is a multi-OU model.

These can be introduced by changing the input for the models into a
list, and supplying a vector with the two models. This is easier to see
with an example:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The models to test}
\NormalTok{my_models <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"BM"}\NormalTok{, }\StringTok{"OU"}\NormalTok{),}
                  \KeywordTok{c}\NormalTok{(}\StringTok{"Stasis"}\NormalTok{, }\StringTok{"OU"}\NormalTok{),}
                  \KeywordTok{c}\NormalTok{(}\StringTok{"BM"}\NormalTok{, }\StringTok{"Stasis"}\NormalTok{),}
                  \KeywordTok{c}\NormalTok{(}\StringTok{"OU"}\NormalTok{, }\StringTok{"Trend"}\NormalTok{),}
                  \KeywordTok{c}\NormalTok{(}\StringTok{"Stasis"}\NormalTok{, }\StringTok{"BM"}\NormalTok{))}

\NormalTok{## Testing the models}
\NormalTok{disp_time <-}\StringTok{ }\KeywordTok{model.test.wrapper}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_disparity, }\DataTypeTok{model =}\NormalTok{ my_models, }\DataTypeTok{time.split =} \DecValTok{66}\NormalTok{,}
                                \DataTypeTok{show.p =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{fixed.optima =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Evidence of equal variance (Bartlett's test of equal variances p = 0).
## Variance is not pooled.
## Running BM:OU model...Done. Log-likelihood = 115.367
## Running Stasis:OU model...Done. Log-likelihood = 80.16
## Running BM:Stasis model...Done. Log-likelihood = 34.278
## Running OU:Trend model...Done. Log-likelihood = 118.166
## Running Stasis:BM model...Done. Log-likelihood = 80.16
\end{verbatim}

\begin{figure}

{\centering \includegraphics{dispRity_manual_files/figure-latex/plot6-1}

}

\caption{Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for a variety of models with a shift in optima allowed at 66 Ma}\label{fig:plot6}
\end{figure}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{disp_time}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##           aicc delta_aicc weight_aicc log.lik param ancestral state
## OU:Trend  -228        0.0       0.943   118.2     4           3.051
## BM:OU     -222        5.6       0.057   115.4     4           3.052
## Stasis:BM -154       73.9       0.000    80.2     3              NA
## Stasis:OU -150       78.2       0.000    80.2     5              NA
## BM:Stasis  -60      167.8       0.000    34.3     4           2.858
##           sigma squared alpha optima.1 theta.1 omega trend median p value
## OU:Trend          0.003 0.039       NA      NA    NA 0.015      0.4840160
## BM:OU             0.003 0.000    4.055      NA    NA    NA      0.7857143
## Stasis:BM         0.004    NA       NA   3.109 0.025    NA      0.9860140
## Stasis:OU         0.004 0.000    4.055   3.109 0.025    NA      0.9635365
## BM:Stasis         0.002    NA       NA   3.648 0.113    NA      0.9615385
##           lower p value upper p value
## OU:Trend      0.4685315     0.4995005
## BM:OU         0.7792208     0.7922078
## Stasis:BM     0.9850150     0.9870130
## Stasis:OU     0.9610390     0.9660340
## BM:Stasis     0.9560440     0.9670330
\end{verbatim}

\subsection{\texorpdfstring{\texttt{model.test.sim}}{model.test.sim}}\label{model.test.sim}

Note that all the models above where run using the
\texttt{model.test.wrapper} function that is a\ldots{} wrapping
function! In practice, this function runs two main functions from the
\texttt{dispRity} package and then plots the results:

\begin{itemize}
\tightlist
\item
  \texttt{model.test} and
\item
  \texttt{model.test.sim}
\end{itemize}

The \texttt{model.test.sim} allows to simulate disparity evolution given
a \texttt{dispRity} object input (as in \texttt{model.test.wrapper}) or
given a model and its specification. For example, it is possible to
simulate a simple Brownian motion model (or any of the other models or
models combination described above):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## A simple BM model}
\NormalTok{model_simulation <-}\StringTok{ }\KeywordTok{model.test.sim}\NormalTok{(}\DataTypeTok{sim =} \DecValTok{1000}\NormalTok{, }\DataTypeTok{model =} \StringTok{"BM"}\NormalTok{, }\DataTypeTok{time.span =} \DecValTok{50}\NormalTok{, }\DataTypeTok{variance =} \FloatTok{0.1}\NormalTok{,}
                                   \DataTypeTok{sample.size =} \DecValTok{100}\NormalTok{, }\DataTypeTok{parameters =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{ancestral.state =} \DecValTok{0}\NormalTok{))}
\NormalTok{model_simulation}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Disparity evolution model simulation:
## Call: model.test.sim(sim = 1000, model = "BM", time.span = 50, variance = 0.1, sample.size = 100, parameters = list(ancestral.state = 0)) 
## 
## Model simulated (1000 times):
## [1] "BM"
\end{verbatim}

This will simulate 1000 Brownian motions for 50 units of time with 100
sampled elements, a variance of 0.1 and an ancestral state of 0. We can
also pass multiple models in the same way we did it for
\texttt{model.test} This model can then be summarised and plotted as
most \texttt{dispRity} objects:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Displaying the 5 first rows of the summary}
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(model_simulation))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets   n var      median      2.5%       25%       75%    97.5%
## 1      50 100 0.1 -0.09312806 -1.718481 -0.712478 0.5854576 1.772700
## 2      49 100 0.1 -0.01881396 -2.697017 -0.967495 0.9105681 2.712452
## 3      48 100 0.1 -0.06914146 -3.443629 -1.205211 1.1171606 3.185465
## 4      47 100 0.1 -0.01584249 -3.897544 -1.474240 1.3762412 3.639836
## 5      46 100 0.1 -0.04118287 -4.402526 -1.496096 1.5347916 4.239536
## 6      45 100 0.1 -0.17175251 -4.764863 -1.697076 1.5894314 4.349500
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Plotting the simulations}
\KeywordTok{plot}\NormalTok{(model_simulation)}
\end{Highlighting}
\end{Shaded}

\begin{figure}

{\centering \includegraphics{dispRity_manual_files/figure-latex/plot7-1}

}

\caption{A simulated Brownian motion}\label{fig:plot7}
\end{figure}

Note that these functions can take all the arguments that can be passed
to \texttt{plot}, \texttt{summary}, \texttt{plot.dispRity} and
\texttt{summary.dispRity}.

\subsubsection{Simulating tested models}\label{simulating-tested-models}

Maybe more interestingly though, it is possible to pass the output of
\texttt{model.test} directly to \texttt{model.test.sim} to simulate the
models that fits the data the best and calculate the Rank Envelope test
\emph{p} value. Let's see that using the simple example from the start:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Fitting multiple models on the data set}
\NormalTok{disp_time <-}\StringTok{ }\KeywordTok{model.test}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_disparity, }\DataTypeTok{model =} \KeywordTok{c}\NormalTok{(}\StringTok{"Stasis"}\NormalTok{, }\StringTok{"BM"}\NormalTok{, }\StringTok{"OU"}\NormalTok{, }\StringTok{"Trend"}\NormalTok{, }\StringTok{"EB"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Evidence of equal variance (Bartlett's test of equal variances p = 0).
## Variance is not pooled.
## Running Stasis model...Done. Log-likelihood = -59.501
## Running BM model...Done. Log-likelihood = 123.938
## Running OU model...Done. Log-likelihood = 126.431
## Running Trend model...Done. Log-likelihood = 126.361
## Running EB model...Done. Log-likelihood = 113.081
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(disp_time)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##        aicc delta_aicc weight_aicc log.lik param theta.1 omega
## Stasis  123      369.6       0.000   -59.5     2   3.403  0.15
## BM     -244        2.7       0.157   123.9     2      NA    NA
## OU     -245        2.0       0.227   126.4     4      NA    NA
## Trend  -247        0.0       0.617   126.4     3      NA    NA
## EB     -220       26.6       0.000   113.1     3      NA    NA
##        ancestral state sigma squared alpha optima.1 trend     eb
## Stasis              NA            NA    NA       NA    NA     NA
## BM               2.858         0.003    NA       NA    NA     NA
## OU               2.835         0.002 0.004    5.707    NA     NA
## Trend            2.839         0.002    NA       NA  0.01     NA
## EB               4.055         0.002    NA       NA    NA -0.014
\end{verbatim}

As seen before, the Trend model fitted this dataset the best. To
simulate what 1000 Trend models would look like using the same
parameters as the ones estimated with \texttt{model.test} (here the
ancestral state being 2.839, the sigma squared beeing 0.002 and the
trend of 0.01), we can simply pass this model to
\texttt{model.test.sim}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Simulating 1000 Trend model with the observed parameters}
\NormalTok{sim_trend <-}\StringTok{ }\KeywordTok{model.test.sim}\NormalTok{(}\DataTypeTok{sim =} \DecValTok{1000}\NormalTok{, }\DataTypeTok{model =}\NormalTok{ disp_time)}
\NormalTok{sim_trend}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Disparity evolution model simulation:
## Call: model.test.sim(sim = 1000, model = disp_time) 
## 
## Model simulated (1000 times):
##       aicc log.lik param ancestral state sigma squared trend
## Trend -247   126.4     3           2.839         0.002  0.01
## 
## Rank envelope test
##  p-value of the test: 0.987013 (ties method: midrank)
##  p-interval         : (0.987013, 0.987013)
\end{verbatim}

By default, the model simulated is the one with the lowest AICc
(\texttt{model.rank\ =\ 1}) but it is possible to choose any ranked
model, for example, the OU (second one):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Simulating 1000 OU model with the observed parameters}
\NormalTok{sim_OU <-}\StringTok{ }\KeywordTok{model.test.sim}\NormalTok{(}\DataTypeTok{sim =} \DecValTok{1000}\NormalTok{, }\DataTypeTok{model =}\NormalTok{ disp_time, }\DataTypeTok{model.rank =} \DecValTok{2}\NormalTok{)}
\NormalTok{sim_OU}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Disparity evolution model simulation:
## Call: model.test.sim(sim = 1000, model = disp_time, model.rank = 2) 
## 
## Model simulated (1000 times):
##    aicc log.lik param ancestral state sigma squared alpha optima.1
## OU -245   126.4     4           2.835         0.002 0.004    5.707
## 
## Rank envelope test
##  p-value of the test: 0.8971029 (ties method: midrank)
##  p-interval         : (0.8941059, 0.9000999)
\end{verbatim}

And as the example above, the simulated data can be plotted or
summarised:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(sim_trend))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets n        var   median     2.5%      25%      75%    97.5%
## 1     120 5 0.06056490 2.837869 2.614855 2.760862 2.908493 3.059927
## 2     119 5 0.07453663 2.858233 2.598562 2.756114 2.945448 3.106885
## 3     118 6 0.07556947 2.856971 2.589279 2.769942 2.941288 3.129721
## 4     117 6 0.07556947 2.870782 2.584536 2.777017 2.965582 3.147397
## 5     116 6 0.07556947 2.869144 2.589171 2.769528 2.978955 3.173132
## 6     115 7 0.06590243 2.883672 2.593479 2.790683 2.978769 3.179178
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(sim_OU))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets n        var   median     2.5%      25%      75%    97.5%
## 1     120 5 0.06056490 2.832068 2.618984 2.759091 2.908017 3.051097
## 2     119 5 0.07453663 2.849006 2.592719 2.759753 2.940149 3.115460
## 3     118 6 0.07556947 2.858304 2.579423 2.767239 2.946567 3.125425
## 4     117 6 0.07556947 2.872844 2.585720 2.782702 2.971421 3.142925
## 5     116 6 0.07556947 2.881609 2.599897 2.779863 2.985630 3.168541
## 6     115 7 0.06590243 2.897410 2.602577 2.790346 2.998301 3.182402
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The trend model with some graphical options}
\KeywordTok{plot}\NormalTok{(sim_trend, }\DataTypeTok{xlab =} \StringTok{"Time (Mya)"}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{"sum of variances"}\NormalTok{,}
    \DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\StringTok{"#F65205"}\NormalTok{, }\StringTok{"#F38336"}\NormalTok{, }\StringTok{"#F7B27E"}\NormalTok{))}

\NormalTok{## Adding the observed disparity through time}
\KeywordTok{plot}\NormalTok{(BeckLee_disparity, }\DataTypeTok{add =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\StringTok{"#3E9CBA"}\NormalTok{, }\StringTok{"#98D4CF90"}\NormalTok{, }\StringTok{"#BFE4E390"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{figure}

{\centering \includegraphics{dispRity_manual_files/figure-latex/plot8-1}

}

\caption{The best fitted model (Trend) and the observed disparity through time}\label{fig:plot8}
\end{figure}

\hypertarget{disparity-as-a-distribution}{\section{Disparity as a
distribution}\label{disparity-as-a-distribution}}

Disparity is often regarded as a summary value of the position of the
all elements in the ordinated space. For example, the sum of variances,
the product of ranges or the median distance between the elements and
their centroid will summarise disparity as a single value. This value
can be pseudo-replicated (bootstrapped) to obtain a distribution of the
summary metric with estimated error. However, another way to perform
disparity analysis is to use the \emph{whole distribution} rather than
just a summary metric (e.g.~the variances or the ranges).

This is possible in the \texttt{dispRity} package by calculating
disparity as a dimension-level 2 metric only! Let's have a look using
our \protect\hyperlink{summarising-dispRity-data-plots}{previous
example} of bootstrapped time slices but by measuring the distances
between each taxon and their centroid as disparity.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring disparity as a whole distribution}
\NormalTok{disparity_centroids <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(boot_time_slices, }\DataTypeTok{metric =}\NormalTok{ centroids)}
\end{Highlighting}
\end{Shaded}

The resulting disparity object is of dimension-level 2, so it can easily
be transformed into a dimension-level 1 object by, for example,
measuring the median distance of all these distributions:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring median disparity in each time slice}
\NormalTok{disparity_centroids_median <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(disparity_centroids, }\DataTypeTok{metric =}\NormalTok{ median)}
\end{Highlighting}
\end{Shaded}

And we can now compare the differences between these methods:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Summarising both disparity measurements:}
\NormalTok{## The distributions:}
\KeywordTok{summary}\NormalTok{(disparity_centroids)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n obs.median bs.median  2.5%   25%   75% 97.5%
## 1     120  5      1.539     1.287 0.536 1.092 1.553 1.864
## 2      80 19      1.846     1.688 1.409 1.582 1.814 1.945
## 3      40 15      1.892     1.685 1.327 1.563 1.839 2.062
## 4       0 10      1.855     1.801 1.313 1.680 1.962 2.104
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The summary of the distributions (as median)}
\KeywordTok{summary}\NormalTok{(disparity_centroids_median)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   subsets  n   obs bs.median  2.5%   25%   75% 97.5%
## 1     120  5 1.508     1.287 0.535 0.943 1.477 1.508
## 2      80 19 1.790     1.682 1.534 1.645 1.716 1.791
## 3      40 15 1.689     1.678 1.552 1.635 1.717 1.822
## 4       0 10 1.910     1.804 1.577 1.761 1.855 1.909
\end{verbatim}

We can see that the summary message for the distribution is slightly
different than before. Here \texttt{summary} also displays the observed
central tendency (i.e.~the central tendency of the measured
distributions). Note that, as expected, this central tendency is the
same in both metrics!

Another, maybe more intuitive way, to compare both approaches for
measuring disparity is to plot the distributions:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical parameters}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{, }\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}

\NormalTok{## Plotting both disparity measurements}
\KeywordTok{plot}\NormalTok{(disparity_centroids, }\DataTypeTok{ylab =} \StringTok{"Distribution of all the distances"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(disparity_centroids_median,}
     \DataTypeTok{ylab =} \StringTok{"Distribution of the medians of all the distances"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-84-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

We can then test for differences in the resulting distributions using
\texttt{test.dispRity} and the \texttt{bhatt.coeff} test as described
above.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Probability of overlap in the distribution of medians}
\KeywordTok{test.dispRity}\NormalTok{(disparity_centroids_median, }\DataTypeTok{test =}\NormalTok{ bhatt.coeff)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in test.dispRity(disparity_centroids_median, test = bhatt.coeff): Multiple p-values will be calculated without adjustment!
## This can inflate Type I error!
\end{verbatim}

\begin{verbatim}
##          bhatt.coeff
## 120 : 80   0.1905968
## 120 : 40   0.1936492
## 120 : 0    0.1381682
## 80 : 40    0.9416139
## 80 : 0     0.6076888
## 40 : 0     0.7124871
\end{verbatim}

In this case, we are looking at the probability of overlap of the
distribution of median distances from centroids among each pair of time
slices. In other words, we are measuring whether the medians from each
bootstrap pseudo-replicate for each time slice overlap. But of course,
we might be interested in the actual distribution of the distances from
the centroid rather than simply their central tendencies. This can be
problematic depending on the research question asked since we are
effectively comparing non-independent medians distributions (because of
the pseudo-replication).

One solution, therefore, is to look at the full distribution:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Probability of overlap for the full distributions}
\KeywordTok{test.dispRity}\NormalTok{(disparity_centroids, }\DataTypeTok{test =}\NormalTok{ bhatt.coeff)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in test.dispRity(disparity_centroids, test = bhatt.coeff): Multiple p-values will be calculated without adjustment!
## This can inflate Type I error!
\end{verbatim}

\begin{verbatim}
##          bhatt.coeff
## 120 : 80   0.6286805
## 120 : 40   0.6631398
## 120 : 0    0.5819930
## 80 : 40    0.9397416
## 80 : 0     0.8578102
## 40 : 0     0.9361860
\end{verbatim}

These results show the actual overlap among all the measured distances
from centroids concatenated across all the bootstraps. For example, when
comparing the slices 120 and 80, we are effectively comparing the 5
\(\times\) 100 distances (the distances of the five elements in slice
120 bootstrapped 100 times) to the 19 \(\times\) 100 distances from
slice 80. However, this can also be problematic for some specific tests
since the \emph{n} \(\times\) 100 distances are also pseudo-replicates
and thus are still not independent.

A second solution is to compare the distributions to each other
\emph{for each replicate}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Boostrapped probability of overlap for the full distributions}
\KeywordTok{test.dispRity}\NormalTok{(disparity_centroids, }\DataTypeTok{test =}\NormalTok{ bhatt.coeff, }\DataTypeTok{concatenate =} \OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in test.dispRity(disparity_centroids, test = bhatt.coeff, concatenate = FALSE): Multiple p-values will be calculated without adjustment!
## This can inflate Type I error!
\end{verbatim}

\begin{verbatim}
##          bhatt.coeff      2.5%       25%       75%     97.5%
## 120 : 80   0.2507287 0.0000000 0.1450953 0.3829003 0.5431424
## 120 : 40   0.3184407 0.0000000 0.2000000 0.4312174 0.7651213
## 120 : 0    0.2368001 0.0000000 0.0000000 0.3464102 0.5791641
## 80 : 40    0.5739203 0.2252672 0.4678055 0.6965587 0.8184853
## 80 : 0     0.4649973 0.1382736 0.3688643 0.5757552 0.7292754
## 40 : 0     0.5418625 0.1954287 0.4385665 0.6767986 0.8082483
\end{verbatim}

These results show the median overlap among pairs of distributions in
the first column (\texttt{bhatt.coeff}) and then the distribution of
these overlaps among each pair of bootstraps. In other words, when two
distributions are compared, they are now compared for each bootstrap
pseudo-replicate, thus effectively creating a distribution of
probabilities of overlap. For example, when comparing the slices 120 and
80, we have a mean probability of overlap of 0.28 and a probability
between 0.18 and 0.43 in 50\% of the pseudo-replicates. Note that the
quantiles and central tendencies can be modified via the
\texttt{conc.quantiles} option.

\section{Disparity from other
matrices}\label{disparity-from-other-matrices}

In the example so far, disparity was measured from an ordinated
multidimensional space (i.e.~a PCO of the distances between taxa based
on discrete morphological characters). This is a common approach in
palaeobiology, morphometrics or ecology but ordinated matrices are not
mandatory for the \texttt{dispRity} package! It is totally possible to
perform the same analysis detailed above using other types of matrices
as long as your elements are rows in your matrix.

For example, we can use the data set \texttt{eurodist}, an \texttt{R}
inbuilt dataset that contains the distances (in km) between European
cities. We can check for example, if Northern European cities are closer
to each other than Southern ones:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Making the eurodist data set into a matrix (rather than "dist" object)}
\NormalTok{eurodist <-}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(eurodist)}
\NormalTok{eurodist[}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##           Athens Barcelona Brussels Calais Cherbourg
## Athens         0      3313     2963   3175      3339
## Barcelona   3313         0     1318   1326      1294
## Brussels    2963      1318        0    204       583
## Calais      3175      1326      204      0       460
## Cherbourg   3339      1294      583    460         0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The two groups of cities}
\NormalTok{Northern <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"Brussels"}\NormalTok{, }\StringTok{"Calais"}\NormalTok{, }\StringTok{"Cherbourg"}\NormalTok{, }\StringTok{"Cologne"}\NormalTok{, }\StringTok{"Copenhagen"}\NormalTok{,}
              \StringTok{"Hamburg"}\NormalTok{, }\StringTok{"Hook of Holland"}\NormalTok{, }\StringTok{"Paris"}\NormalTok{, }\StringTok{"Stockholm"}\NormalTok{)}
\NormalTok{Southern <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"Athens"}\NormalTok{, }\StringTok{"Barcelona"}\NormalTok{, }\StringTok{"Geneva"}\NormalTok{, }\StringTok{"Gibraltar"}\NormalTok{, }\StringTok{"Lisbon"}\NormalTok{, }\StringTok{"Lyons"}\NormalTok{,}
              \StringTok{"Madrid"}\NormalTok{, }\StringTok{"Marseilles"}\NormalTok{, }\StringTok{"Milan"}\NormalTok{, }\StringTok{"Munich"}\NormalTok{, }\StringTok{"Rome"}\NormalTok{, }\StringTok{"Vienna"}\NormalTok{)}

\NormalTok{## Creating the subset dispRity object}
\NormalTok{eurodist_subsets <-}\StringTok{ }\KeywordTok{custom.subsets}\NormalTok{(eurodist, }\DataTypeTok{group =} \KeywordTok{list}\NormalTok{(}\StringTok{"Northern"}\NormalTok{ =}\StringTok{ }\NormalTok{Northern,}
                                                        \StringTok{"Southern"}\NormalTok{ =}\StringTok{ }\NormalTok{Southern))}

\NormalTok{## Bootstrapping and rarefying to 9 elements (the number of Northern cities)}
\NormalTok{eurodist_bs <-}\StringTok{ }\KeywordTok{boot.matrix}\NormalTok{(eurodist_subsets, }\DataTypeTok{rarefaction =} \DecValTok{9}\NormalTok{)}

\NormalTok{## Measuring disparity as the median distance from group's centroid}
\NormalTok{euro_disp <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(eurodist_bs, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(median, centroids))}

\NormalTok{## Testing the differences using a simple wilcox.test}
\NormalTok{euro_diff <-}\StringTok{ }\KeywordTok{test.dispRity}\NormalTok{(euro_disp, }\DataTypeTok{test =}\NormalTok{ wilcox.test)}
\NormalTok{euro_diff_rar <-}\StringTok{ }\KeywordTok{test.dispRity}\NormalTok{(euro_disp, }\DataTypeTok{test =}\NormalTok{ wilcox.test, }\DataTypeTok{rarefaction =} \DecValTok{9}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We can compare this approach to an ordination one:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Ordinating the eurodist matrix}
\NormalTok{euro_ord <-}\StringTok{ }\KeywordTok{cmdscale}\NormalTok{(eurodist, }\DataTypeTok{k =} \KeywordTok{nrow}\NormalTok{(eurodist) }\OperatorTok{-}\StringTok{ }\DecValTok{2}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in cmdscale(eurodist, k = nrow(eurodist) - 2): only 11 of the first
## 19 eigenvalues are > 0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity on the bootstrapped and rarefied subset data}
\NormalTok{euro_ord_disp <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(}\KeywordTok{boot.matrix}\NormalTok{(}\KeywordTok{custom.subsets}\NormalTok{(euro_ord, }\DataTypeTok{group =}
        \KeywordTok{list}\NormalTok{(}\StringTok{"Northern"}\NormalTok{ =}\StringTok{ }\NormalTok{Northern, }\StringTok{"Southern"}\NormalTok{ =}\StringTok{ }\NormalTok{Southern)), }\DataTypeTok{rarefaction =} \DecValTok{9}\NormalTok{),}
        \DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(median, centroids))}

\NormalTok{## Testing the differences using a simple wilcox.test}
\NormalTok{euro_ord_diff <-}\StringTok{ }\KeywordTok{test.dispRity}\NormalTok{(euro_ord_disp, }\DataTypeTok{test =}\NormalTok{ wilcox.test)}
\NormalTok{euro_ord_diff_rar <-}\StringTok{ }\KeywordTok{test.dispRity}\NormalTok{(euro_ord_disp, }\DataTypeTok{test =}\NormalTok{ wilcox.test, }\DataTypeTok{rarefaction =} \DecValTok{9}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

And visualise the differences:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Plotting the differences}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{), }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}
\NormalTok{## Plotting the normal disparity}
\KeywordTok{plot}\NormalTok{(euro_disp, }\DataTypeTok{main =} \StringTok{"Distance differences"}\NormalTok{)}
\NormalTok{## Adding the p-value}
\KeywordTok{text}\NormalTok{(}\FloatTok{1.5}\NormalTok{, }\DecValTok{4000}\NormalTok{, }\KeywordTok{paste0}\NormalTok{(}\StringTok{"p="}\NormalTok{,}\KeywordTok{round}\NormalTok{(euro_diff[[}\DecValTok{2}\NormalTok{]][[}\DecValTok{1}\NormalTok{]], }\DataTypeTok{digit =} \DecValTok{5}\NormalTok{)))}
\NormalTok{## Plotting the rarefied disparity}
\KeywordTok{plot}\NormalTok{(euro_disp, }\DataTypeTok{rarefaction =} \DecValTok{9}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Distance differences (rarefied)"}\NormalTok{)}
\NormalTok{## Adding the p-value}
\KeywordTok{text}\NormalTok{(}\FloatTok{1.5}\NormalTok{, }\DecValTok{4000}\NormalTok{, }\KeywordTok{paste0}\NormalTok{(}\StringTok{"p="}\NormalTok{,}\KeywordTok{round}\NormalTok{(euro_diff_rar[[}\DecValTok{2}\NormalTok{]][[}\DecValTok{1}\NormalTok{]], }\DataTypeTok{digit =} \DecValTok{5}\NormalTok{)))}

\NormalTok{## Plotting the ordinated disparity}
\KeywordTok{plot}\NormalTok{(euro_ord_disp, }\DataTypeTok{main =} \StringTok{"Ordinated differences"}\NormalTok{)}
\NormalTok{## Adding the p-value}
\KeywordTok{text}\NormalTok{(}\FloatTok{1.5}\NormalTok{, }\DecValTok{1400}\NormalTok{, }\KeywordTok{paste0}\NormalTok{(}\StringTok{"p="}\NormalTok{,}\KeywordTok{round}\NormalTok{(euro_ord_diff[[}\DecValTok{2}\NormalTok{]][[}\DecValTok{1}\NormalTok{]], }\DataTypeTok{digit =} \DecValTok{5}\NormalTok{) ))}
\NormalTok{## Plotting the rarefied disparity}
\KeywordTok{plot}\NormalTok{(euro_ord_disp, }\DataTypeTok{rarefaction =} \DecValTok{9}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Ordinated differences (rarefied)"}\NormalTok{)}
\NormalTok{## Adding the p-value}
\KeywordTok{text}\NormalTok{(}\FloatTok{1.5}\NormalTok{, }\DecValTok{1400}\NormalTok{, }\KeywordTok{paste0}\NormalTok{(}\StringTok{"p="}\NormalTok{,}\KeywordTok{round}\NormalTok{(euro_ord_diff_rar[[}\DecValTok{2}\NormalTok{]][[}\DecValTok{1}\NormalTok{]], }\DataTypeTok{digit =} \DecValTok{5}\NormalTok{) ))}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-90-1.pdf}

As expected, the results are pretty similar in pattern but different in
terms of scale. The median centroids distance is expressed in km in the
``Distance differences'' plots and in Euclidean units of variation in
the ``Ordinated differences'' plots.

\chapter{Making stuff up!}\label{making-stuff-up}

The \texttt{dispRity} package also offers some advanced data simulation
features to allow to test hypothesis, explore multidimensional spaces,
metrics properties or simply playing around with data! All the following
functions are based on the same modular architecture of the package and
therefore can be used with most of the functions of the package.

\section{Simulating discrete morphological
data}\label{simulating-discrete-morphological-data}

The function \texttt{sim.morpho} allows to simulate discrete
morphological data matrices (sometimes referred to as ``cladistic''
matrices). It allows to evolve multiple discrete characters on a given
phylogenetic tree, given different evolutionary models, rates, and
states. It even allows to include proper inapplicable data to make
datasets as messy as in real life!

In brief, the function \texttt{sim.morpho} takes a phylogenetic tree,
the number of required characters, the evolutionary model and a function
from which to draw the rates. The package also contains a function for
quickly checking the matrix's phylogenetic signal (as defined in
systematics not phylogenetic comparative methods) using parsimony. The
methods are described in details below

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{3}\NormalTok{)}
\NormalTok{## Simulating a starting tree with 15 taxa as a random coalescent tree}
\NormalTok{my_tree <-}\StringTok{ }\KeywordTok{rcoal}\NormalTok{(}\DecValTok{15}\NormalTok{)}

\NormalTok{## Generating a matrix with 100 characters (85% binary and 15% three state) and}
\NormalTok{## an equal rates model with a gamma rate distribution (0.5, 1) with no }
\NormalTok{## invariant characters.}
\NormalTok{my_matrix <-}\StringTok{ }\KeywordTok{sim.morpho}\NormalTok{(}\DataTypeTok{tree =}\NormalTok{ my_tree, }\DataTypeTok{characters =} \DecValTok{100}\NormalTok{, }\DataTypeTok{states =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.85}\NormalTok{,}
    \FloatTok{0.15}\NormalTok{), }\DataTypeTok{rates =} \KeywordTok{c}\NormalTok{(rgamma, }\DataTypeTok{shape =} \FloatTok{0.5}\NormalTok{, }\DataTypeTok{rate =} \DecValTok{1}\NormalTok{), }\DataTypeTok{invariant =} \OtherTok{FALSE}\NormalTok{)}

\NormalTok{## The first few lines of the matrix}
\NormalTok{my_matrix[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##     [,1] [,2] [,3] [,4] [,5] [,6] [,7] [,8] [,9] [,10]
## t15 "1"  "1"  "0"  "1"  "1"  "2"  "1"  "1"  "0"  "0"  
## t12 "1"  "1"  "0"  "1"  "1"  "2"  "1"  "1"  "0"  "0"  
## t14 "1"  "1"  "0"  "1"  "1"  "2"  "1"  "1"  "0"  "0"  
## t6  "1"  "1"  "0"  "1"  "1"  "2"  "1"  "1"  "0"  "0"  
## t3  "0"  "0"  "1"  "0"  "1"  "0"  "0"  "2"  "1"  "0"  
## t2  "0"  "0"  "1"  "0"  "1"  "0"  "0"  "2"  "1"  "0"  
## t11 "0"  "0"  "1"  "0"  "1"  "0"  "0"  "2"  "1"  "1"  
## t7  "0"  "0"  "1"  "0"  "1"  "0"  "0"  "2"  "1"  "0"  
## t1  "0"  "0"  "1"  "0"  "0"  "0"  "0"  "2"  "0"  "0"  
## t5  "0"  "0"  "1"  "0"  "0"  "0"  "0"  "2"  "0"  "0"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Checking the matrix properties with a quick Maximum Parsimony tree search}
\KeywordTok{check.morpho}\NormalTok{(my_matrix, my_tree)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                                     
## Maximum parsimony        139.0000000
## Consistency index          0.7625899
## Retention index            0.8881356
## Robinson-Foulds distance   0.0000000
\end{verbatim}

Note that this example produces a tree with a great consistency index
and an identical topology to the random coalescent tree!

\subsection{A more detailed
description}\label{a-more-detailed-description}

The protocol implemented here to generate discrete morphological
matrices is based on the ones developed in \citet{GuillermeCooper}
\citet{OReilly2016} \citet{puttick2017uncertain} and
\citet{OReilly2017}.

\begin{itemize}
\tightlist
\item
  The first \texttt{tree} argument will be the tree on which to
  ``evolve'' the characters and therefore requires branch length. You
  can generate quick and easy random Yule trees using
  \texttt{ape::rtree(number\_of\_taxa)} but I would advise to use more
  realistic trees for more realistic simulations based on more realistic
  models (really realistic then) using the function \texttt{tree.bd}
  from the
  \href{http://www.zoology.ubc.ca/prog/diversitree/}{\texttt{diversitree}}
  package \citep{fitzjohndiversitree2012} for example.
\item
  The second argument, \texttt{character} is the number of characters.
\item
  The third, \texttt{states} is the proportion of characters states
  above two. This argument intakes the proportion of \emph{n}-states
  characters, for example \texttt{states\ =\ c(0.5,0.3,0.2)} will
  generate 50\% of binary-state characters, 30\% of three-state
  characters and 20\% of four-state characters. There is no limit in the
  number of state characters proportion as long as the total makes up
  100\%.
\item
  The fourth, \texttt{model} is the evolutionary model for generating
  the character(s). More about this below.
\item
  The fifth and sixth, \texttt{rates} and \texttt{substitution} are the
  model parameters described below as well.
\item
  Finally, the two logical arguments, are self explanatory:
  \texttt{invariant} whether to allow invariant characters
  (i.e.~characters that don't change) and \texttt{verbose} whether to
  print the simulation progress on your console.
\end{itemize}

\subsubsection{Available evolutionary
models}\label{available-evolutionary-models}

There are currently three evolutionary models implemented in
\texttt{sim.morpho} but more will come in the future. Note also that
they allow fine tuning parameters making them pretty plastic!

\begin{itemize}
\tightlist
\item
  \texttt{"ER"}: this model allows any number of character states and is
  based on the Mk model \citep{lewisa2001}. It assumes a unique overall
  evolutionary rate equal substitution rate between character states.
  This model is based on the \texttt{ape::rTraitDisc} function.
\item
  \texttt{"HKY"}: this is binary state character model based on the
  molecular HKY model \citep{HKY85}. It uses the four molecular states
  (A,C,G,T) with a unique overall evolutionary rate and a biased
  substitution rate towards transitions (A \textless{}-\textgreater{} G
  or C \textless{}-\textgreater{} T) against transvertions (A
  \textless{}-\textgreater{} C and G \textless{}-\textgreater{} T).
  After evolving the nucleotides, this model transforms them into binary
  states by converting the purines (A and G) into state 0 and the
  pyrimidines (C and T) into state 1. This method is based on the
  \texttt{phyclust::seq.gen.HKY} function and was first proposed by
  \citet{OReilly2016}.
\item
  \texttt{"MIXED"}: this model uses a random (uniform) mix between both
  the \texttt{"ER"} and the \texttt{"HKY"} models.
\end{itemize}

The models can take the following parameters: (1) \texttt{rates} is the
evolutionary rate (i.e.~the rate of changes along a branch; the
evolutionary speed) and (2) \texttt{substitution} is the frequency of
changes between one state or another. For example if a character can
have high probability of changing (the \emph{evolutionary} rate) with,
each time a change occurs a probability of changing from state \emph{X}
to state \emph{Y} (the \emph{substitution} rate). Note that in the
\texttt{"ER"} model, the substitution rate is ignore because by
definition this rate is equal!

The parameters arguments \texttt{rates} and \texttt{substitution} takes
a distributions from which to draw the parameters values for each
character. For example, if you want an \texttt{"HKY"} model with an
evolutionary rate (i.e.~speed) drawn from a uniform distribution bounded
between 0.001 and 0.005, you can define it as
\texttt{rates\ =\ c(runif,\ min\ =\ 0.001,\ max\ =\ 0.005)},
\texttt{runif} being the function for random draws from a uniform
distribution and \texttt{max} and \texttt{min} being the distribution
parameters. These distributions should always be passed in the format
\texttt{c(random\_distribution\_function,\ distribution\_parameters)}
with the names of the distribution parameters arguments.

\subsubsection{Checking the results}\label{checking-the-results}

An additional function, \texttt{check.morpho} runs a quick Maximum
Parsimony tree search using the \texttt{phangorn} parsimony algorithm.
It quickly calculates the parsimony score, the consistency and retention
indices and, if a tree is provided (e.g.~the tree used to generate the
matrix) it calculates the Robinson-Foulds distance between the most
parsimonious tree and the provided tree to determine how different they
are.

\subsubsection{Adding inapplicable
characters}\label{adding-inapplicable-characters}

Once a matrix is generated, it is possible to apply inapplicable
characters to it for increasing realism! Inapplicable characters are
commonly designated as \texttt{NA} or simply \texttt{-}. They differ
from missing characters \texttt{?} in their nature by being inapplicable
rather than unknown. For example, considering a binary character defined
as ``colour of the tail'' with the following states ``blue'' and
``red''; on a taxa with no tail, the character should be coded as
inapplicable (``\texttt{-}'') since the state of the character ``colour
of tail'' is \emph{known}: it's neither ``blue'' or ``red'', it's just
not there! It contrasts with coding it as missing (``\texttt{?}'' - also
called as ambiguous) where the state is \emph{unknown}, for example, the
taxon of interest is a fossil where the tail has no colour preserved or
is not present at all due to bad conservation!

This type of characters can be added to the simulated matrices using the
\texttt{apply.NA} function. It takes, as arguments, the \texttt{matrix},
the source of inapplicability (\texttt{NAs} - more below), the
\texttt{tree} used to generate the matrix and the two same
\texttt{invariant} and \texttt{verbose} arguments as defined above. The
\texttt{NAs} argument allows two types of sources of inapplicability:

\begin{itemize}
\tightlist
\item
  \texttt{"character"} where the inapplicability is due to the character
  (e.g.~coding a character tail for species with no tail). In practice,
  the algorithm chooses a character \emph{X} as the underlying character
  (e.g. ``presence and absence of tail''), arbitrarily chooses one of
  the states as ``absent'' (e.g.~0 = absent) and changes in the next
  character \emph{Y} any state next to character \emph{X} state 0 into
  an inapplicable token (``\texttt{-}''). This simulates the
  inapplicability induced by coding the characters (i.e.~not always
  biological).
\item
  \texttt{"clade"} where the inapplicability is due to evolutionary
  history (e.g.~a clade loosing its tail). In practice, the algorithm
  chooses a random clade in the tree and a random character \emph{Z} and
  replaces the state of the taxa present in the clade by the
  inapplicable token (``\texttt{-}''). This simulates the
  inapplicability induced by evolutionary biology (e.g.~the lose of a
  feature in a clade).
\end{itemize}

To apply these sources of inapplicability, simply repeat the number of
inapplicable sources for the desired number of characters with
inapplicable data.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Generating 5 "character" NAs and 10 "clade" NAs}
\NormalTok{my_matrix_NA <-}\StringTok{ }\KeywordTok{apply.NA}\NormalTok{(my_matrix, }\DataTypeTok{tree =}\NormalTok{ my_tree,}
                         \DataTypeTok{NAs =} \KeywordTok{c}\NormalTok{(}\KeywordTok{rep}\NormalTok{(}\StringTok{"character"}\NormalTok{, }\DecValTok{5}\NormalTok{), }\KeywordTok{rep}\NormalTok{(}\StringTok{"clade"}\NormalTok{, }\DecValTok{10}\NormalTok{)))}

\NormalTok{## The first few lines of the resulting matrix}
\NormalTok{my_matrix_NA[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{, }\DecValTok{90}\OperatorTok{:}\DecValTok{100}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##     [,1] [,2] [,3] [,4] [,5] [,6] [,7] [,8] [,9] [,10] [,11]
## t15 "0"  "1"  "0"  "1"  "0"  "0"  "0"  "1"  "0"  "1"   "0"  
## t12 "1"  "1"  "0"  "1"  "0"  "0"  "0"  "1"  "0"  "1"   "0"  
## t14 "0"  "0"  "0"  "1"  "0"  "0"  "0"  "1"  "0"  "1"   "0"  
## t6  "0"  "0"  "0"  "1"  "0"  "0"  "0"  "1"  "0"  "1"   "0"  
## t3  "0"  "0"  "1"  "0"  "-"  "0"  "0"  "0"  "-"  "0"   "1"  
## t2  "0"  "0"  "1"  "0"  "-"  "0"  "0"  "0"  "-"  "0"   "1"  
## t11 "0"  "0"  "1"  "0"  "-"  "0"  "0"  "0"  "-"  "0"   "1"  
## t7  "0"  "0"  "1"  "0"  "-"  "0"  "0"  "0"  "-"  "0"   "1"  
## t1  "0"  "0"  "0"  "0"  "-"  "0"  "0"  "0"  "-"  "2"   "1"  
## t5  "0"  "0"  "0"  "0"  "-"  "0"  "0"  "0"  "-"  "0"   "1"
\end{verbatim}

\subsection{Parameters for a realistic(ish)
matrix}\label{parameters-for-a-realisticish-matrix}

There are many parameters that can create a ``realistic'' matrix
(i.e.~not too different from the input tree with a consistency and
retention index close to what is seen in the literature) but because of
the randomness of the matrix generation not all parameters combination
end up creating ``good'' matrices. The following parameters however,
seem to generate fairly ``realist'' matrices with a starting coalescent
tree, equal rates model with 0.85 binary characters and 0.15 three state
characters, a gamma distribution with a shape parameter (\(\alpha\)) of
10 and no scaling (\(\beta\) = 1) with a rate of 50.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
\NormalTok{## tree}
\NormalTok{my_tree <-}\StringTok{ }\KeywordTok{rcoal}\NormalTok{(}\DecValTok{30}\NormalTok{)}
\NormalTok{## matrix}
\NormalTok{morpho_mat <-}\StringTok{ }\KeywordTok{sim.morpho}\NormalTok{(}\DataTypeTok{tree =}\NormalTok{ my_tree, }\DataTypeTok{characters =} \DecValTok{100}\NormalTok{, }\DataTypeTok{states =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.85}\NormalTok{,}
    \FloatTok{0.15}\NormalTok{), }\DataTypeTok{rates =} \KeywordTok{c}\NormalTok{(rgamma, }\DataTypeTok{shape =} \DecValTok{10}\NormalTok{, }\DataTypeTok{rate =} \DecValTok{50}\NormalTok{), }\DataTypeTok{invariant =} \OtherTok{FALSE}\NormalTok{)}

\KeywordTok{check.morpho}\NormalTok{(morpho_mat, my_tree)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                                     
## Maximum parsimony        133.0000000
## Consistency index          0.7819549
## Retention index            0.9597222
## Robinson-Foulds distance   4.0000000
\end{verbatim}

\section{Simulating multidimensional
spaces}\label{simulating-multidimensional-spaces}

Another way to simulate data is to directly simulate a multidimensional
space with the \texttt{space.maker} function. This function allows users
to simulate spaces with a certain number of properties. For example, it
is possible to design a space with a specific distribution on each axis,
a correlation between the axes and a specific cumulative variance per
axis. This can be useful for creating ordinated spaces for null
hypothesis, for example if you're using the function \texttt{null.test}
\citep{diaz2016global}.

This function takes as arguments the number of elements (data points -
\texttt{elements} argument)~and dimensions (\texttt{dimensions}
argument) to create the space and the distribution functions to be used
for each axis. The distributions are passed through the
\texttt{distribution} argument as modular functions. You can either pass
a single distribution function for all the axes (for example
\texttt{distribution\ =\ runif} for all the axis being uniform) or a
specific distribution function for each specific axis (for example
\texttt{distribution\ =\ c(runif,\ rnorm,\ rgamma))} for the first axis
being uniform, the second normal and the third gamma). You can of course
use your very own functions or use the ones implemented in
\texttt{dispRity} for more complex ones (see below). Specific optional
arguments for each of these distributions can be passed as a list via
the \texttt{arguments} argument.

Furthermore, it is possible to add a correlation matrix to add a
correlation between the axis via the \texttt{cor.matrix} argument or
even a vector of proportion of variance to be bear by each axis via the
\texttt{scree} argument to simulate realistic ordinated spaces.

Here is a simple two dimensional example:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## A square space}
\NormalTok{square_space <-}\StringTok{ }\KeywordTok{space.maker}\NormalTok{(}\DecValTok{100}\NormalTok{, }\DecValTok{2}\NormalTok{, runif)}

\NormalTok{## The resulting 2D matrix}
\KeywordTok{head}\NormalTok{(square_space)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##           [,1]       [,2]
## [1,] 0.3322704 0.73741888
## [2,] 0.5157255 0.33160759
## [3,] 0.3716847 0.62284731
## [4,] 0.4367013 0.76535328
## [5,] 0.7424290 0.07764732
## [6,] 0.0111013 0.65066885
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Visualising the space}
\KeywordTok{plot}\NormalTok{(square_space, }\DataTypeTok{pch =} \DecValTok{20}\NormalTok{, }\DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Uniform 2D space"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-94-1.pdf}

Of course, more complex spaces can be created by changing the
distributions, their arguments or adding a correlation matrix or a
cumulative variance vector:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## A plane space: uniform with one dimensions equal to 0}
\NormalTok{plane_space <-}\StringTok{ }\KeywordTok{space.maker}\NormalTok{(}\DecValTok{2500}\NormalTok{, }\DecValTok{3}\NormalTok{, }\KeywordTok{c}\NormalTok{(runif, runif, runif),}
                           \DataTypeTok{arguments =} \KeywordTok{list}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{min =} \DecValTok{0}\NormalTok{, }\DataTypeTok{max =} \DecValTok{0}\NormalTok{), }\OtherTok{NULL}\NormalTok{, }\OtherTok{NULL}\NormalTok{))}
\KeywordTok{head}\NormalTok{(plane_space)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      [,1]       [,2]       [,3]
## [1,]    0 0.18294640 0.32682604
## [2,]    0 0.67243526 0.77182827
## [3,]    0 0.56028603 0.06603386
## [4,]    0 0.80409259 0.19401581
## [5,]    0 0.31732158 0.16460302
## [6,]    0 0.09244429 0.99873117
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Correlation matrix for a 3D space}
\NormalTok{(cor_matrix <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\KeywordTok{cbind}\NormalTok{(}\DecValTok{1}\NormalTok{, }\FloatTok{0.8}\NormalTok{, }\FloatTok{0.2}\NormalTok{, }\FloatTok{0.8}\NormalTok{, }\DecValTok{1}\NormalTok{, }\FloatTok{0.7}\NormalTok{, }\FloatTok{0.2}\NormalTok{, }\FloatTok{0.7}\NormalTok{, }\DecValTok{1}\NormalTok{), }\DataTypeTok{nrow =} \DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      [,1] [,2] [,3]
## [1,]  1.0  0.8  0.2
## [2,]  0.8  1.0  0.7
## [3,]  0.2  0.7  1.0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## An ellipsoid space (normal space with correlation)}
\NormalTok{ellipse_space <-}\StringTok{ }\KeywordTok{space.maker}\NormalTok{(}\DecValTok{2500}\NormalTok{, }\DecValTok{3}\NormalTok{, rnorm, }\DataTypeTok{cor.matrix =}\NormalTok{ cor_matrix)}
\KeywordTok{head}\NormalTok{(ellipse_space)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##            [,1]       [,2]       [,3]
## [1,] -0.5080285 -0.7014059 -1.0286500
## [2,]  0.7250674  0.9498248  0.7220638
## [3,]  0.4476787  0.2102857 -0.3835125
## [4,] -1.3252827 -0.5060518  0.2211104
## [5,] -0.9193155 -1.3861888 -1.6151720
## [6,] -0.7217653 -0.1967257  0.2672965
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## A cylindrical space with decreasing axes variance}
\NormalTok{cylindrical_space <-}\StringTok{ }\KeywordTok{space.maker}\NormalTok{(}\DecValTok{2500}\NormalTok{, }\DecValTok{3}\NormalTok{, }\KeywordTok{c}\NormalTok{(rnorm, rnorm, runif),}
                                 \DataTypeTok{scree =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.7}\NormalTok{, }\FloatTok{0.2}\NormalTok{, }\FloatTok{0.1}\NormalTok{))}
\KeywordTok{head}\NormalTok{(cylindrical_space)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##             [,1]        [,2]         [,3]
## [1,] -0.07935545  0.10932383 0.0017309562
## [2,] -0.20619425 -0.14809591 0.0008416656
## [3,] -0.20133258  0.05766707 0.0041888464
## [4,]  0.69412491 -0.47106729 0.0036967387
## [5,]  0.41300602  0.23924098 0.0077537936
## [6,]  0.53369953  0.02882589 0.0073476805
\end{verbatim}

See below for the visualisation of the spaces.

\subsection{Personalised dimensions
distributions}\label{personalised-dimensions-distributions}

Following the modular architecture of the package, it is of course
possible to pass home made distribution functions to the
\texttt{distribution} argument. For example, the \texttt{random.circle}
function is a personalised one implemented in \texttt{dispRity}. This
function allows to create circles based on trigonometry allowing two
axis to covary to produce circle coordinates. By default, this function
generates two sets of coordinates with a \texttt{distribution} argument
and a minimum and maximum boundary (\texttt{inner} and \texttt{outer}
respectively) to create nice sharp edges to the circle. The maximum
boundary is equivalent to the radius of the circle (it removes
coordinates beyond the circle radius) and the minimum is equivalent to
the radius of a smaller circle with no data (it removes coordinates
below this inner circle radius).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Generating coordinates for a normal circle with a upper boundary of 1}
\NormalTok{circle <-}\StringTok{ }\KeywordTok{random.circle}\NormalTok{(}\DecValTok{1000}\NormalTok{, rnorm, }\DataTypeTok{inner =} \DecValTok{0}\NormalTok{, }\DataTypeTok{outer =} \DecValTok{1}\NormalTok{)}

\NormalTok{## Plotting the circle}
\KeywordTok{plot}\NormalTok{(circle, }\DataTypeTok{xlab =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{"y"}\NormalTok{, }\DataTypeTok{main =} \StringTok{"A normal circle"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-96-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating doughnut space (a spherical space with a hole)}
\NormalTok{doughnut_space <-}\StringTok{ }\KeywordTok{space.maker}\NormalTok{(}\DecValTok{5000}\NormalTok{, }\DecValTok{3}\NormalTok{, }\KeywordTok{c}\NormalTok{(rnorm, random.circle),}
     \DataTypeTok{arguments =} \KeywordTok{list}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{mean =} \DecValTok{0}\NormalTok{), }\KeywordTok{list}\NormalTok{(runif, }\DataTypeTok{inner =} \FloatTok{0.5}\NormalTok{, }\DataTypeTok{outer =} \DecValTok{1}\NormalTok{)))}
\KeywordTok{head}\NormalTok{(doughnut_space)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##            [,1]        [,2]       [,3]
## [1,]  1.0692065 -0.34286592 -0.4077478
## [2,]  0.3898210 -0.06393936  0.5212298
## [3,] -0.9878641  0.36223496  0.4519167
## [4,]  0.1031183  0.08781562  0.9176893
## [5,]  1.1351951 -0.34076407  0.8712020
## [6,]  0.1769964 -0.77904629  0.5770467
\end{verbatim}

See below for the visualisation of the ``doughnut'' space.

\subsection{Visualising the space}\label{visualising-the-space}

I suggest using the excellent \texttt{scatterplot3d} package to play
around and visualise the simulated spaces:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =}\NormalTok{ (}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{)), }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}
\NormalTok{## Visualising 3D spaces}
\KeywordTok{require}\NormalTok{(scatterplot3d)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Loading required package: scatterplot3d
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The plane space}
\KeywordTok{scatterplot3d}\NormalTok{(plane_space, }\DataTypeTok{pch =} \DecValTok{20}\NormalTok{, }\DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{, }\DataTypeTok{zlab =} \StringTok{""}\NormalTok{,}
              \DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(}\OperatorTok{-}\FloatTok{0.5}\NormalTok{, }\FloatTok{0.5}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Plane space"}\NormalTok{)}

\NormalTok{## The ellipsoid space}
\KeywordTok{scatterplot3d}\NormalTok{(ellipse_space, }\DataTypeTok{pch =} \DecValTok{20}\NormalTok{, }\DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{, }\DataTypeTok{zlab =} \StringTok{""}\NormalTok{,}
              \DataTypeTok{main =} \StringTok{"Normal ellipsoid space"}\NormalTok{)}

\NormalTok{## A cylindrical space with a decreasing variance per axis}
\KeywordTok{scatterplot3d}\NormalTok{(cylindrical_space, }\DataTypeTok{pch =} \DecValTok{20}\NormalTok{, }\DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{, }\DataTypeTok{zlab =} \StringTok{""}\NormalTok{,}
              \DataTypeTok{main =} \StringTok{"Normal cylindrical space"}\NormalTok{)}
\NormalTok{## Axes have different orders of magnitude}

\NormalTok{## Plotting the doughnut space}
\KeywordTok{scatterplot3d}\NormalTok{(doughnut_space[,}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{1}\NormalTok{,}\DecValTok{3}\NormalTok{)], }\DataTypeTok{pch =} \DecValTok{20}\NormalTok{, }\DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{,}
              \DataTypeTok{zlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Doughnut space"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-97-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(op)}
\end{Highlighting}
\end{Shaded}

\subsection{Generating realistic
spaces}\label{generating-realistic-spaces}

It is possible to generate ``realistic'' spaces by simply extracting the
parameters of an existing space and scaling it up to the simulated
space. For example, we can extract the parameters of the
\texttt{BeckLee\_mat50} ordinated space and simulate a similar space.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading the data}
\KeywordTok{data}\NormalTok{(BeckLee_mat50)}

\NormalTok{## Number of dimensions}
\NormalTok{obs_dim <-}\StringTok{ }\KeywordTok{ncol}\NormalTok{(BeckLee_mat50)}

\NormalTok{## Observed correlation between the dimensions}
\NormalTok{obs_correlations <-}\StringTok{ }\KeywordTok{cor}\NormalTok{(BeckLee_mat50)}

\NormalTok{## Observed mean and standard deviation per axis}
\NormalTok{obs_mu_sd_axis <-}\StringTok{ }\KeywordTok{mapply}\NormalTok{(}\ControlFlowTok{function}\NormalTok{(x,y) }\KeywordTok{list}\NormalTok{(}\StringTok{"mean"}\NormalTok{ =}\StringTok{ }\NormalTok{x, }\StringTok{"sd"}\NormalTok{ =}\StringTok{ }\NormalTok{y),}
                         \KeywordTok{as.list}\NormalTok{(}\KeywordTok{apply}\NormalTok{(BeckLee_mat50, }\DecValTok{2}\NormalTok{, mean)),}
                         \KeywordTok{as.list}\NormalTok{(}\KeywordTok{apply}\NormalTok{(BeckLee_mat50, }\DecValTok{2}\NormalTok{, sd)), }\DataTypeTok{SIMPLIFY =} \OtherTok{FALSE}\NormalTok{)}

\NormalTok{## Observed overall mean and standard deviation}
\NormalTok{obs_mu_sd_glob <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\StringTok{"mean"}\NormalTok{ =}\StringTok{ }\KeywordTok{mean}\NormalTok{(BeckLee_mat50), }\StringTok{"sd"}\NormalTok{ =}\StringTok{ }\KeywordTok{sd}\NormalTok{(BeckLee_mat50))}

\NormalTok{## Scaled observed variance per axis (scree plot)}
\NormalTok{obs_scree <-}\StringTok{ }\KeywordTok{variances}\NormalTok{(BeckLee_mat50)}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(}\KeywordTok{variances}\NormalTok{(BeckLee_mat50))}

\NormalTok{## Generating our simulated space}
\NormalTok{simulated_space <-}\StringTok{ }\KeywordTok{space.maker}\NormalTok{(}\DecValTok{1000}\NormalTok{, }\DataTypeTok{dimensions =}\NormalTok{ obs_dim, }
                               \DataTypeTok{distribution =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{list}\NormalTok{(rnorm), obs_dim),}
                               \DataTypeTok{arguments =}\NormalTok{ obs_mu_sd_axis,}
                               \DataTypeTok{cor.matrix =}\NormalTok{ obs_correlations)}

\NormalTok{## Visualising the fit of our data in the space (in the two first dimensions)}
\KeywordTok{plot}\NormalTok{(simulated_space[,}\DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{], }\DataTypeTok{xlab =} \StringTok{"PC1"}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{"PC2"}\NormalTok{)}
\KeywordTok{points}\NormalTok{(BeckLee_mat50[,}\DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{], }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{, }\DataTypeTok{pch =} \DecValTok{20}\NormalTok{)}
\KeywordTok{legend}\NormalTok{(}\StringTok{"topleft"}\NormalTok{, }\DataTypeTok{legend =} \KeywordTok{c}\NormalTok{(}\StringTok{"observed"}\NormalTok{, }\StringTok{"simulated"}\NormalTok{),}
        \DataTypeTok{pch =} \KeywordTok{c}\NormalTok{(}\DecValTok{20}\NormalTok{,}\DecValTok{21}\NormalTok{), }\DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\StringTok{"red"}\NormalTok{, }\StringTok{"black"}\NormalTok{), }\DataTypeTok{bg =} \StringTok{"white"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-98-1.pdf}

It is now possible to simulate a space using these observed arguments to
test several hypothesis:

\begin{itemize}
\tightlist
\item
  Is the space uniform or normal?
\item
  If the space is normal, is the mean and variance global or specific
  for each axis?
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring disparity as the sum of variance}
\NormalTok{observed_disp <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(BeckLee_mat50, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(median, centroids))}

\NormalTok{## Is the space uniform?}
\NormalTok{test_unif <-}\StringTok{ }\KeywordTok{null.test}\NormalTok{(observed_disp, }\DataTypeTok{null.distrib =}\NormalTok{ runif)}

\NormalTok{## Is the space normal with a mean of 0 and a sd of 1?}
\NormalTok{test_norm1 <-}\StringTok{ }\KeywordTok{null.test}\NormalTok{(observed_disp, }\DataTypeTok{null.distrib =}\NormalTok{ rnorm)}

\NormalTok{## Is the space normal with the observed mean and sd and cumulative variance}
\NormalTok{test_norm2 <-}\StringTok{ }\KeywordTok{null.test}\NormalTok{(observed_disp, }\DataTypeTok{null.distrib =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{list}\NormalTok{(rnorm), obs_dim),}
                        \DataTypeTok{null.args =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{list}\NormalTok{(obs_mu_sd_glob), obs_dim),}
                        \DataTypeTok{null.scree =}\NormalTok{ obs_scree)}

\NormalTok{## Is the space multiple normal with multiple means and sds and a correlation?}
\NormalTok{test_norm3 <-}\StringTok{ }\KeywordTok{null.test}\NormalTok{(observed_disp, }\DataTypeTok{null.distrib =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{list}\NormalTok{(rnorm), obs_dim),}
                        \DataTypeTok{null.args =}\NormalTok{ obs_mu_sd_axis, }\DataTypeTok{null.cor =}\NormalTok{ obs_correlations)}

\NormalTok{## Graphical options}
\NormalTok{op <-}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =}\NormalTok{ (}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{)), }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}
\NormalTok{## Plotting the results}
\KeywordTok{plot}\NormalTok{(test_unif, }\DataTypeTok{main =} \StringTok{"Uniform (0,1)"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(test_norm1, }\DataTypeTok{main =} \StringTok{"Normal (0,1)"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(test_norm2, }\DataTypeTok{main =} \StringTok{"Normal (global observed}\CharTok{\textbackslash{}n}\StringTok{mean and sd)"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(test_norm3, }\DataTypeTok{main =} \StringTok{"Normal (variable mean}\CharTok{\textbackslash{}n}\StringTok{and sd with correlation)"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-99-1.pdf}

If we measure disparity as the median distance from the morphospace
centroid, we can explain the distribution of the data as normal with the
variable observed mean and standard deviation and with a correlation
between the dimensions.

\chapter{\texorpdfstring{The guts of the \texttt{dispRity}
package}{The guts of the dispRity package}}\label{the-guts-of-the-disprity-package}

\section{\texorpdfstring{Manipulating \texttt{dispRity}
objects}{Manipulating dispRity objects}}\label{manipulating-disprity-objects}

Disparity analysis involves a lot of manipulation of many matrices
(especially when bootstrapping) which can be impractical to visualise
and will quickly overwhelm your \texttt{R} console. Even the simple Beck
and Lee 2014 example above produces an object with \textgreater{} 72
lines of lists of lists of matrices!

Therefore \texttt{dispRity} uses a specific class of object called a
\texttt{dispRity} object. These objects allow users to use S3 method
functions such as \texttt{summary.dispRity}, \texttt{plot.dispRity} and
\texttt{print.dispRity}. \texttt{dispRity} also contains various utility
functions that manipulate the \texttt{dispRity} object (e.g.
\texttt{sort.dispRity}, \texttt{extract.dispRity} see the full list in
the next section). These functions modify the \texttt{dispRity} object
without having to delve into its complex structure! The full structure
of a \texttt{dispRity} object is detailed
\href{https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md}{here}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading the example data}
\KeywordTok{data}\NormalTok{(disparity)}

\NormalTok{## What is the class of the median_centroids object?}
\KeywordTok{class}\NormalTok{(disparity)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "dispRity"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## What does the object contain?}
\KeywordTok{names}\NormalTok{(disparity)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "matrix"    "call"      "subsets"   "disparity"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Summarising it using the S3 method print.dispRity}
\NormalTok{disparity}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 7 continuous (acctran) time subsets for 99 elements with 97 dimensions:
##      90, 80, 70, 60, 50 ...
## Data was bootstrapped 100 times (method:"full") and rarefied to 20, 15, 10, 5 elements.
## Disparity was calculated as: c(median, centroids).
\end{verbatim}

Note that it is always possible to recall the full object using the
argument \texttt{all\ =\ TRUE} in \texttt{print.dispRity}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Display the full object}
\KeywordTok{print}\NormalTok{(disparity, }\DataTypeTok{all =} \OtherTok{TRUE}\NormalTok{)}
\NormalTok{## This is more nearly ~ 5000 lines on my 13 inch laptop screen!}
\end{Highlighting}
\end{Shaded}

\section{\texorpdfstring{\texttt{dispRity}
utilities}{dispRity utilities}}\label{disprity-utilities}

The package also provides some utility functions to facilitate
multidimensional analysis.

\subsection{\texorpdfstring{\texttt{dispRity} object utilities
}{dispRity object utilities }}\label{disprity-object-utilities}

The first set of utilities are functions for manipulating
\texttt{dispRity} objects:

\subsubsection{\texorpdfstring{\texttt{make.dispRity}}{make.dispRity}}\label{make.disprity}

This function creates empty \texttt{dispRity} objects.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating an empty dispRity object}
\KeywordTok{make.dispRity}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Empty dispRity object.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating an "empty" dispRity object with a matrix}
\NormalTok{(disparity_obj <-}\StringTok{ }\KeywordTok{make.dispRity}\NormalTok{(}\KeywordTok{matrix}\NormalTok{(}\KeywordTok{rnorm}\NormalTok{(}\DecValTok{20}\NormalTok{), }\DecValTok{5}\NormalTok{, }\DecValTok{4}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## Contains only a matrix 5x4.
\end{verbatim}

\subsubsection{\texorpdfstring{\texttt{fill.dispRity}}{fill.dispRity}}\label{fill.disprity}

This function initialises a \texttt{dispRity} object and generates its
call properties.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The dispRity object's call is indeed empty}
\NormalTok{disparity_obj}\OperatorTok{$}\NormalTok{call}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## list()
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Filling an empty disparity object (that needs to contain at least a matrix)}
\NormalTok{(disparity_obj <-}\StringTok{ }\KeywordTok{fill.dispRity}\NormalTok{(disparity_obj))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 elements with 4 dimensions.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The dipRity object has now the correct minimal attributes}
\NormalTok{disparity_obj}\OperatorTok{$}\NormalTok{call}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## $dimensions
## [1] 4
\end{verbatim}

\subsubsection{\texorpdfstring{\texttt{matrix.dispRity}}{matrix.dispRity}}\label{matrix.disprity}

This function extracts a specific matrix from a disparity object. The
matrix can be one of the bootstrapped matrices or/and a rarefied matrix.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Extracting the matrix containing the coordinates of the elements at time 50}
\KeywordTok{str}\NormalTok{(}\KeywordTok{matrix.dispRity}\NormalTok{(disparity, }\StringTok{"50"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  num [1:18, 1:97] -0.1038 0.2844 0.2848 0.0927 0.1619 ...
##  - attr(*, "dimnames")=List of 2
##   ..$ : chr [1:18] "Leptictis" "Dasypodidae" "n24" "Potamogalinae" ...
##   ..$ : NULL
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Extracting the 3rd bootstrapped matrix with the 2nd rarefaction level}
\NormalTok{## (15 elements) from the second group (80 Mya)}
\KeywordTok{str}\NormalTok{(}\KeywordTok{matrix.dispRity}\NormalTok{(disparity, }\DataTypeTok{subsets =} \DecValTok{1}\NormalTok{, }\DataTypeTok{bootstrap =} \DecValTok{3}\NormalTok{, }\DataTypeTok{rarefaction =} \DecValTok{2}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  num [1:15, 1:97] -0.7161 0.3496 -0.573 -0.0445 -0.1427 ...
##  - attr(*, "dimnames")=List of 2
##   ..$ : chr [1:15] "n7" "n34" "Maelestes" "n20" ...
##   ..$ : NULL
\end{verbatim}

\subsubsection{\texorpdfstring{\texttt{get.subsets}}{get.subsets}}\label{get.subsets}

This function creates a dispRity object that contains only elements from
one specific subsets.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Extracting all the data for the crown mammals}
\NormalTok{(crown_mammals <-}\StringTok{ }\KeywordTok{get.subsets}\NormalTok{(disp_crown_stemBS, }\StringTok{"Group.crown"}\NormalTok{))}

\NormalTok{## The object keeps the properties of the parent object but is composed of only one subsets}
\KeywordTok{length}\NormalTok{(crown_mammals}\OperatorTok{$}\NormalTok{subsets)}
\end{Highlighting}
\end{Shaded}

\subsubsection{\texorpdfstring{\texttt{combine.subsets}}{combine.subsets}}\label{combine.subsets}

This function allows to merge different subsets. If the

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Combine the two first subsets in the dispRity data example}
\KeywordTok{combine.subsets}\NormalTok{(disparity, }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

Note that the computed values (bootstrapped data + disparity metric) are
not merge.

\subsubsection{\texorpdfstring{\texttt{extract.dispRity}}{extract.dispRity}}\label{extract.disprity}

This function extracts the calculated disparity values of a specific
matrix.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Extracting the observed disparity (default)}
\KeywordTok{extract.dispRity}\NormalTok{(disparity)}

\NormalTok{## Extracting the disparity from the bootstrapped values from the}
\NormalTok{## 10th rarefaction level from the second subsets (80 Mya)}
\KeywordTok{extract.dispRity}\NormalTok{(disparity, }\DataTypeTok{observed =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{subsets =} \DecValTok{2}\NormalTok{, }\DataTypeTok{rarefaction =} \DecValTok{10}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\subsubsection{\texorpdfstring{\texttt{rescale.dispRity}}{rescale.dispRity}}\label{rescale.disprity}

This is the modified S3 method for \texttt{scale} (scaling and/or
centring) that can be applied to the disparity data of a
\texttt{dispRity} object and can take optional arguments (for example
the rescaling by dividing by a maximum value).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Getting the disparity values of the time subsets}
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(disparity))}

\NormalTok{## Scaling the same disparity values}
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(}\KeywordTok{rescale.dispRity}\NormalTok{(disparity, }\DataTypeTok{scale =} \OtherTok{TRUE}\NormalTok{)))}

\NormalTok{## Scaling and centering:}
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(}\KeywordTok{rescale.dispRity}\NormalTok{(disparity, }\DataTypeTok{scale =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{center =} \OtherTok{TRUE}\NormalTok{)))}

\NormalTok{## Rescaling the value by dividing by a maximum value}
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(}\KeywordTok{rescale.dispRity}\NormalTok{(disparity, }\DataTypeTok{max =} \DecValTok{10}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\subsubsection{\texorpdfstring{\texttt{sort.dispRity}}{sort.dispRity}}\label{sort.disprity}

This is the S3 method of \texttt{sort} for sorting the subsets
alphabetically (default) or following a specific pattern.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Sorting the disparity subsets in inverse alphabetic order}
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(}\KeywordTok{sort}\NormalTok{(disparity, }\DataTypeTok{decreasing =} \OtherTok{TRUE}\NormalTok{)))}

\NormalTok{## Customised sorting}
\KeywordTok{head}\NormalTok{(}\KeywordTok{summary}\NormalTok{(}\KeywordTok{sort}\NormalTok{(disparity, }\DataTypeTok{sort =} \KeywordTok{c}\NormalTok{(}\DecValTok{7}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{3}\NormalTok{, }\DecValTok{4}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{))))}
\end{Highlighting}
\end{Shaded}

\section{\texorpdfstring{The \texttt{dispRity} object
content}{The dispRity object content}}\label{the-disprity-object-content}

The functions above are utilities to easily and safely access different
elements in the \texttt{dispRity} object. Alternatively, of course, each
elements can be accessed manually. Here is an explanation on how it
works. The \texttt{dispRity} object is a \texttt{list} of two to four
elements, each of which are detailed below:

\begin{itemize}
\tightlist
\item
  \texttt{\$matrix}: an object of class \texttt{matrix}, the full
  multidimensional space.
\item
  \texttt{\$call}: an object of class \texttt{list} containing
  information on the \texttt{dispRity} object content.
\item
  \texttt{\$subsets}: an object of class \texttt{list} containing the
  subsets of the multidimensional space.
\item
  \texttt{\$disparity}: an object of class \texttt{list} containing the
  disparity values.
\end{itemize}

The \texttt{dispRity} object is loosely based on \texttt{C} structure
objects. In fact, it is composed of one unique instance of a matrix (the
multidimensional space) upon which the metric function is called via
``pointers'' to only a certain number of elements and/or dimensions of
this matrix. This allows for: (1) faster and easily tractable execution
time: the metric functions are called through apply family function and
can be parallelised; and (2) a really low memory footprint: at any time,
only one matrix is present in the \texttt{R} environment rather than
multiple copies of it for each subset.

\subsection{\texorpdfstring{\texttt{\$matrix}}{\$matrix}}\label{matrix}

This is the multidimensional space, stored in the \texttt{R} environment
as a \texttt{matrix} object. It requires row names but not column names.
By default, if the row names are missing, \texttt{dispRity} function
will arbitrarily generate them in numeric order (i.e.
\texttt{rownames(matrix)\ \textless{}-\ 1:nrow(matrix)}). This element
of the \texttt{dispRity} object is never modified.

\subsection{\texorpdfstring{\texttt{\$call}}{\$call}}\label{call}

This element contains the information on the \texttt{dispRity} object
content. It is a \texttt{list} that can contain the following:

\begin{itemize}
\tightlist
\item
  \texttt{\$call\$subsets}: a vector of \texttt{character} with
  information on the subsets type (either \texttt{"continuous"},
  \texttt{"discrete"} or \texttt{"custom"}) and their eventual model
  (\texttt{"acctran"}, \texttt{"deltran"}, \texttt{"random"},
  \texttt{"proximity"}, \texttt{"equal.split"},
  \texttt{"gradual.split"}). This element generated only once via
  \texttt{chrono.subsets()} and \texttt{custom.subsets()}.
\item
  \texttt{\$call\$dimensions}: either a single \texttt{numeric} value
  indicating how many dimensions to use or a vector of \texttt{numeric}
  values indicating which specific dimensions to use. This element is by
  default the number of columns in \texttt{\$matrix} but can be modified
  through \texttt{boot.matrix()} or \texttt{dispRity()}.
\item
  \texttt{\$call\$bootstrap}: this is a \texttt{list} containing three
  elements:

  \begin{itemize}
  \tightlist
  \item
    \texttt{{[}{[}1{]}{]}}: the number of bootstrap replicates
    (\texttt{numeric})
  \item
    \texttt{{[}{[}2{]}{]}}: the bootstrap method (\texttt{character})
  \item
    \texttt{{[}{[}3{]}{]}}: the rarefaction levels (\texttt{numeric}
    vector)
  \end{itemize}
\item
  \texttt{\$call\$disparity}: this is a \texttt{list} containing one
  element, \texttt{\$metric}, that is a \texttt{list} containing the
  different functions passed to the \texttt{metric} argument in
  \texttt{dispRity}. These are \texttt{call} elements and get modified
  each time the \texttt{dispRity} function is used (the first element is
  the first metric(s), the second, the second metric(s), etc.).
\end{itemize}

\subsection{\texorpdfstring{\texttt{\$subsets}}{\$subsets}}\label{subsets}

This element contain the eventual subsets of the multidimensional space.
It is a \texttt{list} of subset names. Each subset name is in turn a
\texttt{list} of at least one element called \texttt{elements} which is
in turn a \texttt{matrix}. This \texttt{elements} matrix is the raw
(observed) elements in the subsets. The \texttt{elements} matrix is
composed of \texttt{numeric} values in one column and \emph{n} rows (the
number of elements in the subset). Each of these values are a
``pointer'' (\texttt{C} inspired) to the element of the
\texttt{\$matrix}. For example, lets assume a \texttt{dispRity} object
called \texttt{disparity}, composed of at least one subsets called
\texttt{sub1}:

\begin{verbatim}
 disparity$subsets$sub1$elements
      [,1]
 [1,]    5
 [2,]    4
 [3,]    6
 [4,]    7
\end{verbatim}

The values in the matrix ``point'' to the elements in \texttt{\$matrix}:
here, the multidimensional space with only the 4th, 5th, 6th and 7th
elements. The following elements in \texttt{diparity\$subsets\$sub1}
will correspond to the same ``pointers'' but drawn from the bootstrap
replicates. The columns will correspond to different bootstrap
replicates. For example:

\begin{verbatim}
 disparity$subsets$sub1[[2]]
      [,1] [,2] [,3] [,4]
 [1,]   57   43   70    4
 [2,]   43   44    4    4
 [3,]   42   84   44    1
 [4,]   84    7    2   10
\end{verbatim}

This signifies that we have four bootstrap pseudo-replicates pointing
each time to four elements in \texttt{\$matrix}. The next element
(\texttt{{[}{[}3{]}{]}}) will be the same for the eventual first
rarefaction level (i.e.~the resulting bootstrap matrix will have
\emph{m} rows where \emph{m} is the number of elements for this
rarefaction level). The next element after that (\texttt{{[}{[}4{]}{]}})
will be the same for with an other rarefaction level and so
forth\ldots{}

\subsection{\texorpdfstring{\texttt{\$disparity}}{\$disparity}}\label{disparity}

The \texttt{\$disparity} element is identical to the \texttt{\$subsets}
element structure (a list of list(s) containing matrices) but the
matrices don't contain ``pointers'' to \texttt{\$matrix} but the
disparity result of the disparity metric applied to the ``pointers''.
For example, in our first example (\texttt{\$elements}) from above, if
the disparity metric is of dimensions level 1, we would have:

\begin{verbatim}
 disparity$disparity$sub1$elements
      [,1]
 [1,]    1.82
\end{verbatim}

This is the observed disparity (1.82) for the subset called
\texttt{sub1}. If the disparity metric is of dimension level 2 (say the
function \texttt{range} that outputs two values), we would have:

\begin{verbatim}
 disparity$disparity$sub1$elements
      [,1]
 [1,]    0.82
 [2,]    2.82
\end{verbatim}

The following elements in the list follow the same logic as before: rows
are disparity values (one row for a dimension level 1 metric, multiple
for a dimensions level 2 metric) and columns are the bootstrap
replicates (the bootstrap with all elements followed by the eventual
rarefaction levels). For example for the bootstrap without rarefaction
(second element of the list):

\begin{verbatim}
 disparity$disparity$sub1[[2]]
         [,1]     [,2]     [,3]     [,4]
[1,] 1.744668 1.777418 1.781624 1.739679 
\end{verbatim}

\chapter{Palaeobiology demo: disparity-through-time and within
groups}\label{palaeobiology-demo-disparity-through-time-and-within-groups}

This demo aims to give quick overview of the \texttt{dispRity} package
(v.1.2) for palaeobiology analyses of disparity, including disparity
through time analyses.

This demo showcases a typical disparity-through-time analysis: we are
going to test whether the disparity changed through time in a subset of
eutherian mammals from the last 100 million years using a dataset from
Beck and Lee (2014).

\section{Before starting}\label{before-starting}

\subsection{The morphospace}\label{the-morphospace}

In this example, we are going to use a subset of the data from
\citet{beckancient2014}. See the
\protect\hyperlink{example-data}{example data} description for more
details. Briefly, this dataset contains an ordinated matrix of 50
discrete characters from mammals (\texttt{BeckLee\_mat50}), another
matrix of the same 50 mammals and the estimated discrete data characters
of their descendants (thus 50 + 49 rows, \texttt{BeckLee\_mat99}), a
dataframe containing the ages of each taxon in the dataset
(\texttt{BeckLee\_ages}) and finally a phylogenetic tree with the
relationships among the 50 mammals (\texttt{BeckLee\_tree}). The
ordinated matrix will represent our full morphospace, i.e.~all the
mammalian morphologies that ever existed through time (for this
dataset).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Loading demo and the package data}
\KeywordTok{library}\NormalTok{(dispRity)}

\NormalTok{## Setting the random seed for repeatability}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}

\NormalTok{## Loading the ordinated matrix/morphospace:}
\KeywordTok{data}\NormalTok{(BeckLee_mat50)}
\KeywordTok{head}\NormalTok{(BeckLee_mat50[,}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{])}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                    [,1]          [,2]        [,3]       [,4]      [,5]
## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{dim}\NormalTok{(BeckLee_mat50)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 50 48
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## The morphospace contains 50 taxa and has 48 dimensions (or axes)}

\NormalTok{## Showing a list of first and last occurrences data for some fossils}
\KeywordTok{data}\NormalTok{(BeckLee_ages)}
\KeywordTok{head}\NormalTok{(BeckLee_ages)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##             FAD  LAD
## Adapis     37.2 36.8
## Asioryctes 83.6 72.1
## Leptictis  33.9 33.3
## Miacis     49.0 46.7
## Mimotona   61.6 59.2
## Notharctus 50.2 47.0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Plotting a phylogeny}
\KeywordTok{data}\NormalTok{(BeckLee_tree)}
\KeywordTok{plot}\NormalTok{(BeckLee_tree, }\DataTypeTok{cex =} \FloatTok{0.7}\NormalTok{)}
\KeywordTok{axisPhylo}\NormalTok{(}\DataTypeTok{root =} \DecValTok{140}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-110-1.pdf}

\begin{quote}
You can have an even nicer looking tree if you use the \texttt{strap}
package!
\end{quote}

\begin{Shaded}
\begin{Highlighting}[]
\ControlFlowTok{if}\NormalTok{(}\OperatorTok{!}\KeywordTok{require}\NormalTok{(strap)) }\KeywordTok{install.packages}\NormalTok{(}\StringTok{"strap"}\NormalTok{)}
\NormalTok{strap}\OperatorTok{::}\KeywordTok{geoscalePhylo}\NormalTok{(BeckLee_tree, }\DataTypeTok{cex.tip =} \FloatTok{0.7}\NormalTok{, }\DataTypeTok{cex.ts =} \FloatTok{0.6}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-111-1.pdf}

\section{A disparity-through-time
analysis}\label{a-disparity-through-time-analysis}

\subsection{Splitting the morphospace through
time}\label{splitting-the-morphospace-through-time}

One of the crucial steps in disparity-through-time analysis is to split
the full morphospace into smaller time subsets that contain the total
number of morphologies at certain points in time (time-slicing) or
during certain periods in time (time-binning). Basically, the full
morphospace represents the total number of morphologies across all time
and will be greater than any of the time subsets of the morphospace.

The \texttt{dispRity} package provides a \texttt{chrono.subsets}
function that allows users to split the morphospace into time slices
(using \texttt{method\ =\ continuous}) or into time bins (using
\texttt{method\ =\ discrete}). In this example, we are going to split
the morphospace into five equal time bins of 20 million years long from
100 million years ago to the present. We will also provide to the
function a table containing the first and last occurrences dates for
some fossils to take into account that some fossils might occur in
several of our different time bins.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating the vector of time bins ages}
\NormalTok{(time_bins <-}\StringTok{ }\KeywordTok{rev}\NormalTok{(}\KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{0}\NormalTok{, }\DataTypeTok{to =} \DecValTok{100}\NormalTok{, }\DataTypeTok{by =} \DecValTok{20}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 100  80  60  40  20   0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Splitting the morphospace using the chrono.subsets function}
\NormalTok{(binned_morphospace <-}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee_mat50, }\DataTypeTok{tree =}\NormalTok{ BeckLee_tree,}
    \DataTypeTok{method =} \StringTok{"discrete"}\NormalTok{, }\DataTypeTok{time =}\NormalTok{ time_bins, }\DataTypeTok{inc.nodes =} \OtherTok{FALSE}\NormalTok{,}
    \DataTypeTok{FADLAD =}\NormalTok{ BeckLee_ages))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 discrete time subsets for 50 elements:
##     100 - 80, 80 - 60, 60 - 40, 40 - 20, 20 - 0.
\end{verbatim}

The output object is a \texttt{dispRity} object (see more about that
\protect\hyperlink{The-guts-of-the-dispRity-package}{here}. In brief,
however, \texttt{dispRity} objects are lists of different elements
(i.e.~disparity results, morphospace time subsets, morphospace
attributes, etc.) that display only a summary of the object when calling
the object to avoiding filling the \texttt{R} console with superfluous
output.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Printing the class of the object}
\KeywordTok{class}\NormalTok{(binned_morphospace)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "dispRity"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Printing the content of the object}
\KeywordTok{str}\NormalTok{(binned_morphospace)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## List of 3
##  $ matrix : num [1:50, 1:48] -0.532 -0.409 -0.692 -0.68 -0.739 ...
##   ..- attr(*, "dimnames")=List of 2
##   .. ..$ : chr [1:50] "Cimolestes" "Maelestes" "Batodon" "Bulaklestes" ...
##   .. ..$ : NULL
##  $ call   :List of 1
##   ..$ subsets: chr "discrete"
##  $ subsets:List of 5
##   ..$ 100 - 80:List of 1
##   .. ..$ elements: int [1:8, 1] 5 4 6 8 43 10 11 42
##   ..$ 80 - 60 :List of 1
##   .. ..$ elements: int [1:15, 1] 7 8 9 1 2 3 12 13 14 44 ...
##   ..$ 60 - 40 :List of 1
##   .. ..$ elements: int [1:13, 1] 41 49 24 25 26 27 28 21 22 19 ...
##   ..$ 40 - 20 :List of 1
##   .. ..$ elements: int [1:6, 1] 15 39 40 35 23 47
##   ..$ 20 - 0  :List of 1
##   .. ..$ elements: int [1:10, 1] 36 37 38 32 33 34 50 48 29 30
##  - attr(*, "class")= chr "dispRity"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{names}\NormalTok{(binned_morphospace)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "matrix"  "call"    "subsets"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Printing the object as a dispRity class}
\NormalTok{binned_morphospace}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 discrete time subsets for 50 elements:
##     100 - 80, 80 - 60, 60 - 40, 40 - 20, 20 - 0.
\end{verbatim}

\begin{quote}
These objects will gradual.splitly contain more information when
completing the following steps in the disparity-through-time analysis.
\end{quote}

\subsection{Bootstrapping the data}\label{bootstrapping-the-data}

Once we obtain our different time subsets, we can bootstrap and rarefy
them (i.e.~pseudo-replicating the data). The bootstrapping allows us to
make each subset more robust to outliers and the rarefaction allows us
to compare subsets with the same number of taxa to remove sampling
biases (i.e.~more taxa in one subset than the others). The
\texttt{boot.matrix} function bootstraps the \texttt{dispRity} object
and the \texttt{rarefaction} option within performs rarefaction.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Bootstrapping each time subset 100 times (default)}
\NormalTok{(boot_bin_morphospace <-}\StringTok{ }\KeywordTok{boot.matrix}\NormalTok{(binned_morphospace))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 discrete time subsets for 50 elements with 48 dimensions:
##     100 - 80, 80 - 60, 60 - 40, 40 - 20, 20 - 0.
## Data was bootstrapped 100 times (method:"full").
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Getting the minimum number of rows (i.e. taxa) in the time subsets}
\KeywordTok{min}\NormalTok{(}\KeywordTok{size.subsets}\NormalTok{(boot_bin_morphospace))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 6
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Bootstrapping each time subset 100 times and rarefying them }
\NormalTok{(rare_bin_morphospace <-}\StringTok{ }\KeywordTok{boot.matrix}\NormalTok{(binned_morphospace, }\DataTypeTok{bootstraps =} \DecValTok{100}\NormalTok{,}
    \DataTypeTok{rarefaction =} \DecValTok{6}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 discrete time subsets for 50 elements with 48 dimensions:
##     100 - 80, 80 - 60, 60 - 40, 40 - 20, 20 - 0.
## Data was bootstrapped 100 times (method:"full") and rarefied to 6 elements.
\end{verbatim}

\subsection{Calculating disparity}\label{calculating-disparity}

We can now calculate the disparity within each time subsets along with
some confidence intervals generated by the pseudoreplication step above
(bootstraps/rarefaction). Disparity can be calculated in many ways and
this package allows users to come up with their own disparity metrics.
For more details, please refer to the
\protect\hyperlink{disparity-metrics}{\texttt{dispRity} metric section}.

In this example, we are going to calculate the spread of the data in
each time subset by calculating disparity as the sum of the variance of
each dimension of the morphospace in each time subset using the
\texttt{dispRity} function. Thus, in this example, disparity is defined
by the multi-dimensional variance of each time subset (i.e.~the spread
of the taxa within the morphospace). Note that this metric comes with a
caveat (not solved here) since it ignores covariances among the
dimensions of the morphospace. We use this here because it is a standard
metric used in disparity-through-time analysis \citep{Wills1994}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity for the bootstrapped data}
\NormalTok{(boot_disparity <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(boot_bin_morphospace, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 discrete time subsets for 50 elements with 48 dimensions:
##     100 - 80, 80 - 60, 60 - 40, 40 - 20, 20 - 0.
## Data was bootstrapped 100 times (method:"full").
## Disparity was calculated as: c(sum, variances).
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity for the rarefied data}
\NormalTok{(rare_disparity <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(rare_bin_morphospace, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 5 discrete time subsets for 50 elements with 48 dimensions:
##     100 - 80, 80 - 60, 60 - 40, 40 - 20, 20 - 0.
## Data was bootstrapped 100 times (method:"full") and rarefied to 6 elements.
## Disparity was calculated as: c(sum, variances).
\end{verbatim}

The \texttt{dispRity} function does not actually display the calculated
disparity values but rather only the properties of the disparity object
(size, subsets, metric, etc.). To display the actual calculated scores,
we need to summarise the disparity object using the S3 method
\texttt{summary} that is applied to a \texttt{dispRity} object (see
\texttt{?summary.dispRity} for more details).

As for any \texttt{R} package, you can refer to the help files for each
individual function for more details.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Summarising the disparity results}
\KeywordTok{summary}\NormalTok{(boot_disparity)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##    subsets  n   obs bs.median  2.5%   25%   75% 97.5%
## 1 100 - 80  8 1.675     1.488 1.087 1.389 1.568 1.648
## 2  80 - 60 15 1.782     1.679 1.538 1.631 1.728 1.792
## 3  60 - 40 13 1.913     1.772 1.607 1.734 1.826 1.886
## 4  40 - 20  6 2.022     1.707 1.212 1.537 1.822 1.942
## 5   20 - 0 10 1.971     1.794 1.598 1.716 1.842 1.890
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(rare_disparity)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##    subsets  n   obs bs.median  2.5%   25%   75% 97.5%
## 1 100 - 80  8 1.675     1.484 1.194 1.400 1.547 1.636
## 2 100 - 80  6    NA     1.477 0.993 1.361 1.569 1.698
## 3  80 - 60 15 1.782     1.674 1.517 1.600 1.725 1.793
## 4  80 - 60  6    NA     1.655 1.299 1.532 1.754 1.882
## 5  60 - 40 13 1.913     1.767 1.601 1.714 1.829 1.861
## 6  60 - 40  6    NA     1.787 1.314 1.672 1.879 1.984
## 7  40 - 20  6 2.022     1.736 1.281 1.603 1.822 1.948
## 8   20 - 0 10 1.971     1.807 1.595 1.729 1.856 1.917
## 9   20 - 0  6    NA     1.790 1.435 1.718 1.873 1.995
\end{verbatim}

\begin{quote}
The summary.dispRity function comes with many options on which values to
calculate (central tendency and quantiles) and on how many digits to
display. Refer to the function's manual for more details.
\end{quote}

\subsection{Plotting the results}\label{plotting-the-results}

It is sometimes easier to visualise the results in a plot than in a
table. For that we can use the \texttt{plot} S3 function to plot the
\texttt{dispRity} objects (see \texttt{?plot.dispRity} for more
details).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\KeywordTok{quartz}\NormalTok{(}\DataTypeTok{width =} \DecValTok{10}\NormalTok{, }\DataTypeTok{height =} \DecValTok{5}\NormalTok{) ; }\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =}\NormalTok{ (}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{)), }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting the bootstrapped and rarefied results}
\KeywordTok{plot}\NormalTok{(boot_disparity, }\DataTypeTok{type =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{main =} \StringTok{"bootstrapped results"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(rare_disparity, }\DataTypeTok{type =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{main =} \StringTok{"rarefied results"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-117-1.pdf}

\section{Testing differences}\label{testing-differences}

Finally, to draw some valid conclusions from these results, we can apply
some statistical tests. We can test, for example, if mammalian disparity
changed significantly through time over the last 100 million years. To
do so, we can compare the means of each time-bin in a sequential manner
to see whether the disparity in bin \emph{n} is equal to the disparity
in bin \emph{n+1}, and whether this is in turn equal to the disparity in
bin \emph{n+2}, etc. Because our data is temporally autocorrelated
(i.e.~what happens in bin \emph{n+1} depends on what happened in bin
\emph{n}) and pseudoreplicated (i.e.~each bootstrap draw creates
non-independent time subsets because they are all based on the same time
subsets), we apply a non-parametric mean comparison: the
\texttt{wilcox.test}. Also, we need to apply a p-value correction
(e.g.~Bonferroni correction) to correct for multiple testing (see
\texttt{?p.adjust} for more details).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Testing the differences between bins in the bootstrapped dataset.}
\KeywordTok{test.dispRity}\NormalTok{(boot_disparity, }\DataTypeTok{test =}\NormalTok{ wilcox.test, }\DataTypeTok{comparison =} \StringTok{"sequential"}\NormalTok{,}
    \DataTypeTok{correction =} \StringTok{"bonferroni"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [[1]]
##                    statistic: W
## 100 - 80 : 80 - 60          471
## 80 - 60 : 60 - 40          1562
## 60 - 40 : 40 - 20          6250
## 40 - 20 : 20 - 0           3725
## 
## [[2]]
##                         p.value
## 100 - 80 : 80 - 60 7.427563e-28
## 80 - 60 : 60 - 40  1.798899e-16
## 60 - 40 : 40 - 20  9.061511e-03
## 40 - 20 : 20 - 0   7.379715e-03
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Testing the differences between bins in the rarefied dataset.}
\KeywordTok{test.dispRity}\NormalTok{(rare_disparity, }\DataTypeTok{test =}\NormalTok{ wilcox.test, }\DataTypeTok{comparison =} \StringTok{"sequential"}\NormalTok{,}
    \DataTypeTok{correction =} \StringTok{"bonferroni"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [[1]]
##                    statistic: W
## 100 - 80 : 80 - 60          662
## 80 - 60 : 60 - 40          1814
## 60 - 40 : 40 - 20          5752
## 40 - 20 : 20 - 0           3621
## 
## [[2]]
##                         p.value
## 100 - 80 : 80 - 60 1.214988e-25
## 80 - 60 : 60 - 40  2.823697e-14
## 60 - 40 : 40 - 20  2.653018e-01
## 40 - 20 : 20 - 0   3.026079e-03
\end{verbatim}

Here our results show significant changes in disparity through time
between all time bins (all p-values \textless{} 0.05). However, when
looking at the rarefied results, there is no significant difference
between the time bins in the Palaeogene (60-40 to 40-20 Mya), suggesting
that the differences detected in the first test might just be due to the
differences in number of taxa sampled (13 or 6 taxa) in each time bin.

\chapter{Ecology demo}\label{ecology-demo}

This is an example of typical disparity analysis that can be performed
in ecology.

\section{Data}\label{data}

For this example, we will use the famous \texttt{iris} inbuilt data set

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data}\NormalTok{(iris)}
\end{Highlighting}
\end{Shaded}

This data contains petal and sepal length for 150 individual plants
sorted into three species.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Separating the species}
\NormalTok{species <-}\StringTok{ }\NormalTok{iris[,}\DecValTok{5}\NormalTok{]}
\NormalTok{## Which species?}
\KeywordTok{unique}\NormalTok{(species)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] setosa     versicolor virginica 
## Levels: setosa versicolor virginica
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Separating the petal/sepal length}
\NormalTok{measurements <-}\StringTok{ }\NormalTok{iris[,}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{]}
\KeywordTok{head}\NormalTok{(measurements)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   Sepal.Length Sepal.Width Petal.Length Petal.Width
## 1          5.1         3.5          1.4         0.2
## 2          4.9         3.0          1.4         0.2
## 3          4.7         3.2          1.3         0.2
## 4          4.6         3.1          1.5         0.2
## 5          5.0         3.6          1.4         0.2
## 6          5.4         3.9          1.7         0.4
\end{verbatim}

We can then ordinate the data using a PCA (\texttt{prcomp} function)
thus defining our four dimensional space as the poetically named
petal-space.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Ordinating the data}
\NormalTok{ordination <-}\StringTok{ }\KeywordTok{prcomp}\NormalTok{(measurements)}

\NormalTok{## The petal-space}
\NormalTok{petal_space <-}\StringTok{ }\NormalTok{ordination}\OperatorTok{$}\NormalTok{x}

\NormalTok{## Adding the elements names to the petal-space (the individuals IDs)}
\KeywordTok{rownames}\NormalTok{(petal_space) <-}\StringTok{ }\DecValTok{1}\OperatorTok{:}\KeywordTok{nrow}\NormalTok{(petal_space)}
\end{Highlighting}
\end{Shaded}

\section{Classic analysis}\label{classic-analysis}

A classical way to represent this ordinated data would be to use two
dimensional plots to look at how the different species are distributed
in the petal-space.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Measuring the variance on each axis}
\NormalTok{variances <-}\StringTok{ }\KeywordTok{apply}\NormalTok{(petal_space, }\DecValTok{2}\NormalTok{, var)}
\NormalTok{variances <-}\StringTok{ }\NormalTok{variances}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(variances)}

\NormalTok{## Graphical option}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## A classic 2D ordination plot}
\KeywordTok{plot}\NormalTok{(petal_space[, }\DecValTok{1}\NormalTok{], petal_space[, }\DecValTok{2}\NormalTok{], }\DataTypeTok{col =}\NormalTok{ species,}
    \DataTypeTok{xlab =} \KeywordTok{paste0}\NormalTok{(}\StringTok{"PC 1 ("}\NormalTok{, }\KeywordTok{round}\NormalTok{(variances[}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{), }\StringTok{")"}\NormalTok{),}
    \DataTypeTok{ylab =} \KeywordTok{paste0}\NormalTok{(}\StringTok{"PC 2 ("}\NormalTok{, }\KeywordTok{round}\NormalTok{(variances[}\DecValTok{2}\NormalTok{], }\DecValTok{2}\NormalTok{), }\StringTok{")"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-122-1.pdf}

This shows the distribution of the different species in the petal-space
along the two first axis of variation. This is a pretty standard way to
visualise the multidimensional space and further analysis might be
necessary to test wether the groups are different such as a linear
discriminant analysis (LDA). However, in this case we are ignoring the
two other dimensions of the ordination! If we look at the two other axis
we see a totally different result:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Plotting the two second axis of the petal-space}
\KeywordTok{plot}\NormalTok{(petal_space[, }\DecValTok{3}\NormalTok{], petal_space[, }\DecValTok{4}\NormalTok{], }\DataTypeTok{col =}\NormalTok{ species,}
    \DataTypeTok{xlab =} \KeywordTok{paste0}\NormalTok{(}\StringTok{"PC 3 ("}\NormalTok{, }\KeywordTok{round}\NormalTok{(variances[}\DecValTok{3}\NormalTok{], }\DecValTok{2}\NormalTok{), }\StringTok{")"}\NormalTok{),}
    \DataTypeTok{ylab =} \KeywordTok{paste0}\NormalTok{(}\StringTok{"PC 4 ("}\NormalTok{, }\KeywordTok{round}\NormalTok{(variances[}\DecValTok{4}\NormalTok{], }\DecValTok{2}\NormalTok{), }\StringTok{")"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-123-1.pdf}

Additionally, these two represented dimensions do not represent a
biological reality \emph{per se}; i.e.~the values on the first dimension
do not represent a continuous trait (e.g.~petal length), instead they
just represent the ordinations of correlations between the data and some
factors.

Therefore, we might want to approach this problem without getting stuck
in only two dimensions and consider the whole dataset as a
\emph{n}-dimensional object.

\section{\texorpdfstring{A multidimensional approach with
\texttt{dispRity}}{A multidimensional approach with dispRity}}\label{a-multidimensional-approach-with-disprity}

The first step is to create different subsets that represent subsets of
the ordinated space (i.e.~sub-regions within the \emph{n}-dimensional
object). Each of these subsets will contain only the individuals of a
specific species.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Creating the table that contain the elements and their attributes}
\NormalTok{petal_subsets <-}\StringTok{ }\KeywordTok{custom.subsets}\NormalTok{(petal_space, }\DataTypeTok{group =} \KeywordTok{list}\NormalTok{(}
                                \StringTok{"setosa"}\NormalTok{ =}\StringTok{ }\KeywordTok{which}\NormalTok{(species }\OperatorTok{==}\StringTok{ "setosa"}\NormalTok{),}
                                \StringTok{"versicolor"}\NormalTok{ =}\StringTok{ }\KeywordTok{which}\NormalTok{(species }\OperatorTok{==}\StringTok{ "versicolor"}\NormalTok{),}
                                \StringTok{"virginica"}\NormalTok{ =}\StringTok{ }\KeywordTok{which}\NormalTok{(species }\OperatorTok{==}\StringTok{ "virginica"}\NormalTok{)))}

\NormalTok{## Visualising the dispRity object content}
\NormalTok{petal_subsets}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 3 customised subsets for 150 elements:
##     setosa, versicolor, virginica.
\end{verbatim}

This created a \texttt{dispRity} object (more about that
\protect\hyperlink{guts}{here}) with three subsets corresponding to each
subspecies.

\subsection{Bootstrapping the data}\label{bootstrapping-the-data-1}

We can the bootstrap the subsets to be able test the robustness of the
measured disparity to outliers. We can do that using the default options
of \texttt{boot.matrix} (more about that
\protect\hyperlink{bootstraps-and-rarefactions}{here}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Bootstrapping the data}
\NormalTok{(petal_bootstrapped <-}\StringTok{ }\KeywordTok{boot.matrix}\NormalTok{(petal_subsets))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 3 customised subsets for 150 elements with 4 dimensions:
##     setosa, versicolor, virginica.
## Data was bootstrapped 100 times (method:"full").
\end{verbatim}

\subsection{Calculating disparity}\label{calculating-disparity-1}

Disparity can be calculated in many ways, therefore the
\texttt{dispRity} function allows users to define their own measure of
disparity. For more details on measuring disparity, see the
\protect\hyperlink{disparity-metrics}{dispRity metrics section}.

In this example, we are going to define disparity as the median distance
between the different individuals and the centroid of the ordinated
space. High values of disparity will indicate a generally high spread of
points from this centroid (i.e.~on average, the individuals are far
apart in the ordinated space). We can define the metrics easily in the
\texttt{dispRity} function by feeding them to the \texttt{metric}
argument. Here we are going to feed the functions \texttt{stats::median}
and \texttt{dispRity::centroids} which calculates distances between
elements and their centroid.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Calculating disparity as the median distance between each elements and}
\NormalTok{## the centroid of the petal-space}
\NormalTok{(petal_disparity <-}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(petal_bootstrapped, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(median, centroids)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dispRity object ---- 
## 3 customised subsets for 150 elements with 4 dimensions:
##     setosa, versicolor, virginica.
## Data was bootstrapped 100 times (method:"full").
## Disparity was calculated as: c(median, centroids).
\end{verbatim}

\subsection{Summarising the results
(plot)}\label{summarising-the-results-plot}

Similarly to the \texttt{custom.subsets} and \texttt{boot.matrix}
function, \texttt{dispRity} displays a \texttt{dispRity} object. But we
are definitely more interested in actually look at the calculated
values.

First we can summarise the data in a table by simply using
\texttt{summary}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Displaying the summary of the calculated disparity}
\KeywordTok{summary}\NormalTok{(petal_disparity)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      subsets  n   obs bs.median  2.5%   25%   75% 97.5%
## 1     setosa 50 0.421     0.419 0.342 0.401 0.445 0.503
## 2 versicolor 50 0.693     0.657 0.530 0.622 0.692 0.733
## 3  virginica 50 0.785     0.745 0.577 0.690 0.797 0.880
\end{verbatim}

We can also plot the results in a similar way:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Graphical options}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}

\NormalTok{## Plotting the disparity in the petal_space}
\KeywordTok{plot}\NormalTok{(petal_disparity)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-128-1.pdf}

Now contrary to simply plotting the two first axis of the PCA where we
saw that the species have a different position in the two first
petal-space, we can now also see that they occupy this space clearly
differently!

\subsection{Testing hypothesis}\label{testing-hypothesis}

Finally we can test our hypothesis that we guessed from the disparity
plot (that some groups occupy different volume of the petal-space) by
using the \texttt{test.dispRity} option.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Running a PERMANOVA}
\KeywordTok{test.dispRity}\NormalTok{(petal_disparity, }\DataTypeTok{test =}\NormalTok{ adonis.dispRity)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in test.dispRity(petal_disparity, test = adonis.dispRity): adonis.dispRity test will be applied to the data matrix, not to the calculated disparity.
## See ?adonis.dispRity for more details.
\end{verbatim}

\begin{verbatim}
## Warning in adonis.dispRity(data, ...): The input data for adonis.dispRity was not a distance matrix.
## The results are thus based on the distance matrix for the input data (i.e. dist(data$matrix)).
## Make sure that this is the desired methodological approach!
\end{verbatim}

\begin{verbatim}
## 
## Call:
## vegan::adonis(formula = dist(matrix) ~ group, data = data, method = "euclidean") 
## 
## Permutation: free
## Number of permutations: 999
## 
## Terms added sequentially (first to last)
## 
##            Df SumsOfSqs MeanSqs F.Model      R2 Pr(>F)    
## group       2    592.07 296.037  487.33 0.86894  0.001 ***
## Residuals 147     89.30   0.607         0.13106           
## Total     149    681.37                 1.00000           
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Post-hoc testing of the differences between species (corrected for multiple tests)}
\KeywordTok{test.dispRity}\NormalTok{(petal_disparity, }\DataTypeTok{test =}\NormalTok{ t.test, }\DataTypeTok{correction =} \StringTok{"bonferroni"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [[1]]
##                        statistic: t
## setosa : versicolor      -33.714480
## setosa : virginica       -34.257797
## versicolor : virginica    -8.595829
## 
## [[2]]
##                        parameter: df
## setosa : versicolor         189.0217
## setosa : virginica          150.1371
## versicolor : virginica      171.2387
## 
## [[3]]
##                             p.value
## setosa : versicolor    2.107278e-81
## setosa : virginica     2.252270e-72
## versicolor : virginica 1.506037e-14
\end{verbatim}

We can now see that there is a significant difference in petal-space
occupancy between all species of iris.

\subsubsection{Setting up a multidimensional
null-hypothesis}\label{setting-up-a-multidimensional-null-hypothesis}

One other series of test can be done on the shape of the petal-space.
Using a MCMC permutation test we can simulate a petal-space with
specific properties and see if our observed petal-space matches these
properties (similarly to \citet{diaz2016global}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Testing against a uniform distribution}
\NormalTok{disparity_uniform <-}\StringTok{ }\KeywordTok{null.test}\NormalTok{(petal_disparity, }\DataTypeTok{replicates =} \DecValTok{200}\NormalTok{,}
    \DataTypeTok{null.distrib =}\NormalTok{ runif, }\DataTypeTok{scale =} \OtherTok{FALSE}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(disparity_uniform)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-130-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Testing against a normal distribution}
\NormalTok{disparity_normal <-}\StringTok{ }\KeywordTok{null.test}\NormalTok{(petal_disparity, }\DataTypeTok{replicates =} \DecValTok{200}\NormalTok{,}
    \DataTypeTok{null.distrib =}\NormalTok{ rnorm, }\DataTypeTok{scale =} \OtherTok{TRUE}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(disparity_normal)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dispRity_manual_files/figure-latex/unnamed-chunk-131-1.pdf}

In both cases we can see that our petal-space is not entirely normal or
uniform. This is expected because of the simplicity of these parameters.

\chapter{Future directions}\label{future-directions}

\section{More tests!}\label{more-tests}

Some more tests are being developed such as a \texttt{sequential.test}
to run sequential linear models for testing hypothesis in disparity
through time or a \texttt{model.test} developed in collaboration with
\href{https://puttickbiology.wordpress.com/}{Mark Puttick} to fit modes
of evolution to disparity curves. Stay tuned!

\section{More rarefactions!}\label{more-rarefactions}

I am working on including several ``clever'' bootstrap and rarefaction
algorithms for palaeobiology studies based on geological sampling
(following:and Butler et al. 2012 \emph{Evolution} and Kotric and Knoll
2015 \emph{Palaeobiology}).

\section{Faster disparity
calculations}\label{faster-disparity-calculations}

I am slowly implementing parallel disparity calculation as well as C
implementations of some disparity metrics to increase significantly
improve the speed of the \texttt{dispRity} function.

\section{More modularity}\label{more-modularity}

I am equally slowly developing functions to allow more of the options in
the package to be modular (in the same way as the \texttt{metric}
argument in \texttt{dispRity}). The next arguments to benefit this
increased modularity will be \texttt{chrono.subsets}'s \texttt{model}
argument and \texttt{boot.matrix}'s \texttt{type} argument.

\section{More suggestions?}\label{more-suggestions}

More generally, I welcome any suggestions or collaborations for
developing some additional specific ideas. Feel free to contact me via
email (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}) or if
you are a GitHub user, directly create an issue on the
\href{https://github.com/TGuillerme/dispRity}{GitHub page}.

\bibliography{../References.bib,../packages.bib}

\end{document}

### 2018-09-27 inst/gitbook/index.Rmd
Updated vignette

@@
 
-This package can be installed via [the CRAN]() or via [GitHub](https://github.com/TGuillerme/dispRity).
+This package can be installed via [the CRAN](https://cran.r-project.org/package=dispRity) or via [GitHub](https://github.com/TGuillerme/dispRity).
 
@@
  * [`Claddis.ordination`](#Claddis-ordination)
+
+### So which version do I choose? {#version}
+
+There are always three version of the package available:
+
+  * The CRAN one
+  * The GitHub `release` one
+  * The GitHub `master` one
+
+The differences between the CRAN one and the GitHub `release` or `master` ones is explained just above.
+For the the GitHub version, the differences are that the `release` one is more stable (i.e.
+more rarely modified) and the `master` one is more live one (i.e.
+bug fixes and new functionalities are added as they come).
+
+If you want the latest-latest version of the package I suggest using the GitHub `master` one, especially if you recently emailed me reporting a minor bug or wanting a new functionality!
+Note however that *it can happen* that the `master` version can sometimes be bugged (especially when there are major `R` and `R` packages updates), however, the status of the package state on both the `release` and the `master` version is constantly displayed on the [`README` page](https://github.com/TGuillerme/dispRity/) of the package with the nice badges displaying these different (and constantly tested) information
+
+ * The build status: [![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=master)](https://travis-ci.org/TGuillerme/dispRity) which tests whether the package compiles (i.e.
+can be installed on your computer).
+ * The code coverage: [![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/master/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity) which indicates the percentage of lines in the code that have been tested for expected behaviour.
+ * The project status: [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active) which indicates whether this package is maintained regularly or forgotten in the abyss of internet.
+ * The development version: [![develVersion](https://img.shields.io/badge/devel%20version-1.2.1-green.svg?style=flat)](https://github.com/TGuillerme/dispRity) which indicates which version of the package you are about to install.
+
+Note that when we (the `dispRity` package collaborators!) are developing some more major versions, including some more complex features for example, they will never be released on either the the versions described above.
+Therefore you can rest assure that you will never use a badly bugged version of `dispRity`! Enjoy!
 

### 2018-10-12 inst/gitbook/03_specific-tutorials.Rmd
Minor changes in the gitbook

@@
 
-These models indicate support for a Trend model, and we can plot the relative support of all model AICc weights
+These models indicate support for a Trend model, and we can plot the relative support of all model AICc weights.
 
@@
 ```
+
+> Note that although AIC values are indicator of model best fit, it is also important to look at the parameters themselves.
+For example OU can be really well supported but with an alpha parameter really close to 0, making it effectively a BM model [@Cooper2016].
 

### 2019-01-18 R/dispRity_fun.R
fixed metric(...) bug and updated gitbook

@@ -1,5 +1,5 @@
 #FUNCTIONS FOR DISPRITY
-get.dispRity.metric.handle <- function(metric, match_call) {
+get.dispRity.metric.handle <- function(metric, match_call, ...) {
     level3.fun <- NULL
     level2.fun <- NULL
     level1.fun <- NULL
@@ -17,7 +17,7 @@ get.dispRity.metric.handle <- function(metric, match_call) {
             metric <- metric[[1]]
         }
         ## Which level is the metric?
-        level <- make.metric(metric, silent = TRUE)
+        level <- make.metric(metric, silent = TRUE, ...)
         if(level == "level3") {
             stop(paste(as.expression(match_call$metric), " metric must contain at least a dimension-level 1 or a dimension-level 2 metric.\nFor more information, see ?make.metric.", sep = ""))
         } else {

### 2019-03-06 inst/gitbook/index.Rmd
Updated book + fixed caddis bug

--- 2018-08-21 inst/gitbook/index.Rmd
@@
 
-Anyways to upload it, you can use
-
 ```{r, eval = TRUE, echo = FALSE, message = FALSE}
@@
 
-This package can be installed via [the CRAN]() or via [GitHub](https://github.com/TGuillerme/dispRity).
+This package can be installed via [the CRAN](https://cran.r-project.org/package=dispRity) or via [GitHub](https://github.com/TGuillerme/dispRity).
 
- * The CRAN version is a more "strict" version of the package that does not include some functionalities (such as [`Claddis.ordination`](#Claddis-ordination) or this gitbook vignette) because of compatibility issues with the CRAN requirement.
-
+ * The CRAN version is a more "strict" version of the package that does not include some functionalities such as this gitbook vignette because of compatibility issues with the CRAN requirement.
  * The GitHub version is a more "reactive" version (fed by user's suggestions and comments) and can contain more functionalities (see list below).
@@
 
-<!-- badges out start -->
-[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
-[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
-[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
-<!-- badges out end   -->
-
 Note however, that apart from the functions listed below, both versions of the package have the same tested and implemented functions.
 
-Functions not available on the CRAN version (`r version_CRAN`):
+### So which version do I choose? {#version}
 
- * [`Claddis.ordination`](#Claddis-ordination)
+There are always three version of the package available:
+
+  * The CRAN one
+  * The GitHub `release` one
+  * The GitHub `master` one
+
+The differences between the CRAN one and the GitHub `release` or `master` ones is explained just above.
+For the the GitHub version, the differences are that the `release` one is more stable (i.e.
+more rarely modified) and the `master` one is more live one (i.e.
+bug fixes and new functionalities are added as they come).
+
+If you want the latest-latest version of the package I suggest using the GitHub `master` one, especially if you recently emailed me reporting a minor bug or wanting a new functionality!
+Note however that *it can happen* that the `master` version can sometimes be bugged (especially when there are major `R` and `R` packages updates), however, the status of the package state on both the `release` and the `master` version is constantly displayed on the [`README` page](https://github.com/TGuillerme/dispRity/) of the package with the nice badges displaying these different (and constantly tested) information.
+
+Note that when we (the `dispRity` package collaborators!) are developing some more major versions, including some more complex features for example, they will never be released on either the the versions described above.
+Therefore you can rest assure that you will never use a badly bugged version of `dispRity`! Enjoy!
 

### 2019-03-06 R/Claddis.ordination.R

@@ -14,7 +14,7 @@
 #' require(Claddis)
 #' 
 #' ## Creating an ordination of the distance matrix of Claddis example data
-#' Claddis.ordination(Claddis::Michaux1989)
+#' Claddis.ordination(Michaux1989)
 #' }
 #'
 #' @seealso \code{\link[Claddis]{MorphDistMatrix}}, \code{\link[Claddis]{ReadMorphNexus}}, \code{\link[Claddis]{MakeMorphMatrix}}, \code{\link[stats]{cmdscale}}, \code{\link{custom.subsets}}, \code{\link{chrono.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity}}.
@@ -30,11 +30,12 @@ Claddis.ordination <- function(data, distance = "MORD", ..., k, add = TRUE, arg.
     error_msg <- paste0("data does not contain a matrix.\nUse Claddis::ReadMorphNexus to generate the proper data format.")
     check.class(data, "list", msg = error_msg)
     ## Must have at least one matrix
-    if(!any(names(data) %in% "matrix")) {
+    # if(!any(names(data) %in% "Matrix")) {
+    if(length(grep("Matrix", names(data))) == 0) {
         stop.call("", error_msg)
     }
     ## Matrix must be a matrix
-    check.class(data$matrix, "matrix", msg = error_msg)
+    check.class(data$Matrix_1$Matrix, "matrix", msg = error_msg)
 
     ## Distance
     distances_available <- c("GC", "GED", "RED", "MORD")
@@ -45,14 +46,14 @@ Claddis.ordination <- function(data, distance = "MORD", ..., k, add = TRUE, arg.
         arg.cmdscale <- list()
     }
     ## k
-    max_k <- (nrow(data$matrix) -1)
+    max_k <- (nrow(data$Matrix_1$Matrix) -1)
     if(missing(k)) {
         arg.cmdscale$k <- max_k
     } else {
         check.class(k, "numeric")
         check.length(k, 1, " must be a single numeric value.")
         if(k > max_k) {
-            stop.call("", paste0("k cannot be greater than the number of rows in data - 1 (data has", max_k, "rows)."))
+            stop.call("", paste0("k cannot be greater than the number of rows in data - 1 (data has ", max_k, " rows)."))
         }
     }
     ## add
@@ -62,10 +63,15 @@ Claddis.ordination <- function(data, distance = "MORD", ..., k, add = TRUE, arg.
[…]

### 2019-03-06 inst/gitbook/03_specific-tutorials.Rmd
fixed metric(...) bug and updated gitbook
Updated 1.2.3

@@
 
+## The ancestral distances distributions
+summary(ancestral_dist)
+
 ## Calculating disparity as the sum of the distances from all the ancestral nodes

### 2019-03-06 inst/gitbook/index.Rmd
Updated 1.2.3

@@
 
+Anyways to upload it, you can use
+
 ```{r, eval = TRUE, echo = FALSE, message = FALSE}
@@
 
- * The CRAN version is a more "strict" version of the package that does not include some functionalities such as this gitbook vignette because of compatibility issues with the CRAN requirement.
+ * The CRAN version is a more "strict" version of the package that does not include some functionalities (such as [`Claddis.ordination`](#Claddis-ordination) or this gitbook vignette) because of compatibility issues with the CRAN requirement.
+
  * The GitHub version is a more "reactive" version (fed by user's suggestions and comments) and can contain more functionalities (see list below).
@@
 
+<!-- badges out start 
+[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
+[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
+[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
+<-- badges out end   -->
+
 Note however, that apart from the functions listed below, both versions of the package have the same tested and implemented functions.
+
+Functions not available on the CRAN version (`r version_CRAN`):
+
+ * [`Claddis.ordination`](#Claddis-ordination)
 
@@
 Note however that *it can happen* that the `master` version can sometimes be bugged (especially when there are major `R` and `R` packages updates), however, the status of the package state on both the `release` and the `master` version is constantly displayed on the [`README` page](https://github.com/TGuillerme/dispRity/) of the package with the nice badges displaying these different (and constantly tested) information.
+
+<!-- badges out start
+ * The build status: [![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=master)](https://travis-ci.org/TGuillerme/dispRity) which tests whether the package compiles (i.e.
+can be installed on your computer).
+ * The code coverage: [![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/master/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity) which indicates the percentage of lines in the code that have been tested for expected behaviour.
+ * The project status: [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active) which indicates whether this package is maintained regularly or forgotten in the abyss of internet.
+ * The development version: [![develVersion](https://img.shields.io/badge/devel%20version-1.2.1-green.svg?style=flat)](https://github.com/TGuillerme/dispRity) which indicates which version of the package you are about to install.
+
+<-- badges out end   -->
 

### 2019-03-06 R/Claddis.ordination.R

@@ -3,10 +3,15 @@
 #' @description Takes Claddis data and computes both the distance and the ordination matrix 
 #'
 #' @param data Data from Claddis::ReadMorphNexus.
-#' @param distance Distance type to be computed by \code{\link[Claddis]{MorphDistMatrix}}. Can be either \code{"Gower"}, \code{"GED"}, \code{"Max"}, \code{"Comp"}
-#' @param transform Whether to transform the proportional distances (for Gower and Max). Options are \code{"none"}, \code{"sqrt"}, or \code{"arcsine_sqrt"} (the default).
+#' @param distance Distance type to be computed by \code{\link[Claddis]{MorphDistMatrix}}. Can be either \code{"GC"}, \code{"GED"}, \code{"RED"}, \code{"MORD"}
+#' @param ... Any optional arguments to be passed to \code{\link[Claddis]{MorphDistMatrix}}.
 #' @param k The number of dimensions in the ordination. If left empty, the number of dimensions is set to number of rows - 1.
+<<<<<<< HEAD
 #' @param ... Any optional arguments to be passed to \code{\link[stats]{cmdscale}}.
+=======
+#' @param add whether to use the Cailliez correction for negative eigen values (\code{add = TRUE}; default - see \code{\link[stats]{cmdscale}}) or not (\code{add = FALSE}).
+#' @param arg.cmdscale Any optional arguments to be passed to \code{\link[stats]{cmdscale}} (as a named list such as \code{list(x.ret = TRUE)}).
+>>>>>>> master
 #' 
 #' @examples
 #' \dontrun{
@@ -21,7 +26,11 @@
 #' @author Thomas Guillerme
 #' @export
 
+<<<<<<< HEAD
 Claddis.ordination <- function(data, distance = "Gower", transform = "arcsine_sqrt", k, ...) {
+=======
+Claddis.ordination <- function(data, distance = "MORD", ..., k, add = TRUE, arg.cmdscale) {
+>>>>>>> master
     match_call <- match.call()
     ## Sanitizing
 
@@ -30,38 +39,57 @@ Claddis.ordination <- function(data, distance = "Gower", transform = "arcsine_sq
     check.class(data, "list", msg = error_msg)
     ## Must have at least one matrix
     if(!any(names(data) %in% "matrix")) {
-        stop(error_msg, call. = FALSE)
+        stop.call("", error_msg)
     }
     ## Matrix must be a matrix
     check.class(data$matrix, "matrix", msg = error_msg)
 
     ## Distance
[…]

### 2019-03-06 R/boot.matrix.R

@@ -91,6 +91,12 @@ boot.matrix <- function(data, bootstraps = 100, rarefaction = FALSE, dimensions,
     if(class(data) != "dispRity") {
         ## Data must be a matrix
         check.class(data, "matrix")
+
+        ## Check whether it is a distance matrix
+        if(check.dist.matrix(data, just.check = TRUE)) {
+            warning("boot.matrix is applied on what seems to be a distance matrix.\nThe resulting matrices won't be distance matrices anymore!", call. = FALSE)
+        }
+
         ## Creating the dispRity object
         dispRity_object <- make.dispRity(data = data)
         #dispRity_object$subsets$origin$elements <- seq(1:nrow(data))
@@ -103,10 +109,10 @@ boot.matrix <- function(data, bootstraps = 100, rarefaction = FALSE, dimensions,
         ## With the correct names
         data_names <- names(data)
         if(is.null(data_names)) {
-            stop(paste(as.expression(match_call$data), "must be either a matrix or an output from the chrono.subsets or custom.subsets functions."), call. = FALSE)
+            stop.call(match_call$data, " must be either a matrix or an output from the chrono.subsets or custom.subsets functions.")
         } else {
             if(data_names[[1]] != "matrix" | data_names[[2]] != "call" | data_names[[3]] != "subsets") {
-                stop(paste(as.expression(match_call$data), "must be either a matrix or an output from the chrono.subsets or custom.subsets functions."), call. = FALSE)
+                stop.call(match_call$data, "must be either a matrix or an output from the chrono.subsets or custom.subsets functions.")
             }
         }
 
@@ -130,8 +136,7 @@ boot.matrix <- function(data, bootstraps = 100, rarefaction = FALSE, dimensions,
 
     if(!missing(prob)) {
         if(probabilistic_subsets) {
-            stop(paste0(as.expression(match_call$data), " was generated using a gradual time-slicing (", data$call$subsets[2], ").\nThe prob option is not yet implemented for this case."), call. = FALSE)
-            #warning(paste0(match_call$data, " was generated using a gradual time-slicing (", data$call$subsets[2], ").\nThe probabilities generated by chrono.subsets() are multiplied by prob." ))
+            stop.call(match_call$data, paste0(" was generated using a gradual time-slicing (", data$call$subsets[2], ").\nThe prob option is not yet implemented for this case."))
         } else {
             probabilistic_subsets <- TRUE
             ## Check if prob is the right class
[…]

### 2019-03-06 R/dispRity.wrapper.R

@@ -15,8 +15,6 @@
 #' As such it has a lot of defaults described in the functions that make up the analysis. 
 #' See \code{\link{chrono.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity.metric}}, \code{\link{summary.dispRity}}, \code{\link{plot.dispRity}} for more details of the defaults used in each of these functions.
 #' Note that any of these defaults can be changed within the \code{disparity.through.time} function.
-#' For more details see the \code{dispRity} manual.
-#'
 #' 
 #' @return
 #' A \code{dispRity} object that can be passed to \code{summary} or \code{plot}.
@@ -28,6 +26,11 @@
 #' ## Run a simple disparity through time analysis (with three time bins)
 #' result <- dispRity.through.time(BeckLee_mat50, BeckLee_tree, 3)
 #' summary(result) ; plot(result)
+#' 
+#' ## This is equivalent to run the following decomposed code
+#' dispRity(boot.matrix(chrono.subsets(BeckLee_mat50, BeckLee_tree, time = 3, method = "discrete"),
+#'                      bootstraps = 100),
+#'          metric = c(median, centroids))
 #' 
 #' @seealso \code{\link{chrono.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity.metric}}, \code{\link{summary.dispRity}}, \code{\link{plot.dispRity}}.
 #'
@@ -66,8 +69,7 @@ dispRity.through.time <- function(data, tree, time, metric = c(median, centroids
 #' As such it has a lot of defaults described in the functions that make up the analysis. 
 #' See \code{\link{custom.subsets}}, \code{\link{boot.matrix}}, \code{\link{dispRity.metric}}, \code{\link{summary.dispRity}}, \code{\link{plot.dispRity}} for more details of the defaults used in each of these functions.
 #' Note that any of these defaults can be changed within the \code{disparity.through.time} function.
-#' For more details see the \code{dispRity} manual.
-#' 
+#'
 #' @return
 #' A \code{dispRity} object that can be passed to \code{summary} or \code{plot}.
 #'
@@ -79,6 +81,12 @@ dispRity.through.time <- function(data, tree, time, metric = c(median, centroids
 #' result <- dispRity.per.group(BeckLee_mat50, list(crown = c(16, 19:41, 45:50),
 #'                              stem = c(1:15, 17:18, 42:44)))
 #' summary(result) ; plot(result)
+#' 
[…]

### 2019-03-06 R/get.bin.ages.R

@@ -31,7 +31,7 @@ get.bin.ages <- function(tree, what = "End", type = "Age", ICS = 2015) {
     ## Tree
     # check.class(tree, "phylo")
     if(is.null(tree$root.time)) {
-        stop("The tree must have a root time element (tree$root.time).", call. = FALSE)
+        stop.call("", "The tree must have a root time element (tree$root.time).")
     }
 
     ## What
@@ -45,7 +45,7 @@ get.bin.ages <- function(tree, what = "End", type = "Age", ICS = 2015) {
     check.length(ICS, 1, errorif = FALSE, msg = " must be a year from 2008 to the present.")
     ICS_available <- match(paste0("ICS", ICS), names(geoscale::timescales))
     if(is.na(ICS_available)) {
-        stop(paste0("No stratigraphic data found for ", paste0("ICS", ICS), ".\nAvailable years are: ", paste(names(geoscale::timescales), collapse = ", "), "."), call. = FALSE)
+        stop.call("", paste0("No stratigraphic data found for ", paste0("ICS", ICS), ".\nAvailable years are: ", paste(names(geoscale::timescales), collapse = ", "), "."))
     } else {
         ## Selecting the ICS
         stratigraphy <- geoscale::timescales[[paste0("ICS", ICS)]]

### 2019-03-06 R/make.metric.R

@@ -58,14 +58,14 @@ make.metric <- function(fun, ..., silent = FALSE) {
 
     ## Testing the metric
     test <- NULL
-    op <- options(warn = 2)
+    op <- options(warn = -1)
     test <- try(fun(matrix, ...), silent = TRUE)
     #try(test <- fun(matrix), silent = TRUE) ; warning("DEBUG")
     options(op)
 
     if(any(test == "try-error") || is.na(test)) {
         if(silent != TRUE) {
-            stop(paste("The provided metric function generated an error or a warning!\nDoes the following work?\n    ", as.expression(match_call$fun),"(matrix(rnorm(20), 5,4))\nThe problem may also come from the optional arguments (...) in ", as.expression(match_call$fun), ".", sep = ""), call. = FALSE)
+            stop.call(match_call$fun, paste0("(matrix(rnorm(20), 5,4))\nThe problem may also come from the optional arguments (...) in ", as.expression(match_call$fun), "."), "The provided metric function generated an error or a warning!\nDoes the following work?\n    ")
         }
     } else {
 
@@ -99,7 +99,7 @@ make.metric <- function(fun, ..., silent = FALSE) {
             } else {
                 ## Function provides a wrong output
                 if(silent != TRUE) {
-                    stop(paste("The provided function did not output a matrix or a numeric vector!\nDoes the following output a matrix or a numeric vector?\n    ", as.expression(match_call$fun),"(matrix(rnorm(20), 5,4))\nThe problem may also come from the optional arguments (...).", sep = ""), call. = FALSE)
+                    stop.call(match_call$fun, paste0("(matrix(rnorm(20), 5,4))\nThe problem may also come from the optional arguments (...) in ", as.expression(match_call$fun), "."), "The provided metric function generated an error or a warning!\nDoes the following work?\n    ")
                 }
             }
         }

### 2019-03-06 R/model.test.R

@@ -186,7 +186,7 @@ model.test <- function(data, model, pool.variance = NULL, time.split = NULL, fix
             } else {
 
                 ## Can't run multi.OU
-                stop("Fewer than 30 samples are available.\nThe \"multi.OU\" model cannot be run with time.split = NULL.\n", call. = FALSE)
+                stop.call("", "Fewer than 30 samples are available.\nThe \"multi.OU\" model cannot be run with time.split = NULL.")
             }
         
         } else {

### 2019-03-06 R/null.test.R

@@ -75,13 +75,14 @@ null.test <- function(data, replicates = 100, null.distrib, null.args = NULL, nu
 
     ## Sanitizing
     check.class(data, "dispRity")
-    if(is.null(data$call$disparity)) stop("Disparity has not been calculated yet.\nUse the dispRity() function to do so.\n", sep = "", call. = FALSE)
+    if(is.null(data$call$disparity)) {
+        stop.call("", "Disparity has not been calculated yet.\nUse the dispRity() function to do so.")
+    }
 
     ## is_distribution?
     is_distribution <- ifelse(length(data$disparity[[1]]$elements) != 1, TRUE, FALSE)
     if(is_distribution) {
-        stop(paste("null.test cannot take disparity distributions yet.\nTry averaging disparity to a dimension-level 1 using:\n",
-            paste("  ", as.expression(match_call$data), " <- dispRity(", as.expression(match_call$data), ", metric = median)\n", sep = "")), call. = FALSE)
+        stop.call(match_call$data, paste0(" <- dispRity(", as.expression(match_call$data), ", metric = median)\n"), msg.pre = "null.test cannot take disparity distributions yet.\nTry averaging disparity to a dimension-level 1 using:\n    ")
     }
 
     ## replicates
@@ -97,15 +98,19 @@ null.test <- function(data, replicates = 100, null.distrib, null.args = NULL, nu
 
     ## NULL TESTING
 
+    ## Handle the disparity metrics
+    metric <- get.metric.from.call(data, what = "fun")
+    args <- get.metric.from.call(data, what = "args")
+
     ## Generating the null models
     if(length(data$subsets) != 1) {
         ## Subdivide the data per subsets
         sub_data <- lapply(as.list(seq(1:length(data$subsets))), function(X) get.subsets(data, X))
         ## Apply the data to all subsets
-        null_models_results <- lapply(sub_data, make.null.model, replicates, null.distrib, null.args, null.cor, null.scree, scale)
+        null_models_results <- lapply(sub_data, make.null.model, replicates, null.distrib, null.args, null.cor, null.scree, scale, metric, args)
     } else {
         ## Apply the data to one subsets
-        null_models_results <- make.null.model(data, replicates, null.distrib, null.args, null.cor, null.scree, scale)
+        null_models_results <- make.null.model(data, replicates, null.distrib, null.args, null.cor, null.scree, scale, metric, args)
     }
 
     ## testing the null hypothesis
[…]

### 2019-03-06 R/plot.dispRity.R

@@ -14,7 +14,7 @@
 #' @param ylab Optional, one or two (if \code{elements = TRUE}) \code{character} string(s) for the caption of the y axis.
 #' @param col Optional, some \code{character} string(s) for the colour of the graph.
 #' @param chrono.subsets \code{logical} whether to handle continuous data from the \code{chrono.subsets} function as time (in Ma). When this option is set to TRUE for other \code{type} options, the names of the subsets are used for the x axis labels.
-#' @param observed \code{logical} whether to add the observed values on the plot as crosses (default is \code{FALSE}).
+#' @param observed \code{logical} whether to add the observed values on the plot as crosses (default is \code{FALSE}) or a \code{list} of any of the graphical arguments \code{"col"}, \code{"pch"} and/or \code{"cex"}.
 #' @param add \code{logical} whether to add the new plot an existing one (default is \code{FALSE}).
 #' @param density the density of shading lines to be passed to \code{\link[graphics]{polygon}}. Is ignored if \code{type = "box"} or \code{type = "line"}.
 #' @param element.pch optional, if \code{elements = TRUE}, the point type to represent them (default are squares: \code{element.pch = 15})
@@ -63,6 +63,12 @@
 #' 
 #' ## Rarefactions plots
 #' plot(disparity, rarefaction = TRUE)
+#' 
+#' ## Observed data
+#' plot(disparity, observed = TRUE)
+#'
+#' ## Observed data with graphical details
+#' plot(disparity, observed = list("pch" = 19, col = "blue", cex = 4))
 #' 
 #' \dontrun{
 #' ## Geoscale plots
@@ -116,7 +122,7 @@
 # token.args=NULL
 
 # data(disparity)
-# data <- dispRity
+# data <- disparity
 # type = "line"
 # elements = TRUE
 # ylim = c(0, 5)
@@ -205,7 +211,7 @@ plot.dispRity <- function(x, ..., type, quantiles = c(50, 95), cent.tend = media
                 }
                 ## Are quantiles proper proportions
                 if(any(quantiles < 0) | any(quantiles > 100)) {
-                    stop("quantiles(s) must be any value between 0 and 100.", call. = FALSE)
+                    stop.call("", "quantiles(s) must be any value between 0 and 100.")
                 }
                 quantiles_n <- length(quantiles)
 
[…]

### 2019-03-06 R/test.dispRity.R

@@ -102,9 +102,13 @@ test.dispRity <- function(data, test, comparisons = "pairwise", rarefaction = NU
     ## must be class dispRity...
     check.class(data, "dispRity")
     ## ...and have disparity data
-    if(is.null(data$call$disparity)) stop("Disparity has not been calculated yet.\nUse the dispRity() function to do so.\n", sep = "", call. = FALSE)
+    if(is.null(data$call$disparity)) {
+        stop.call("", "Disparity has not been calculated yet.\nUse the dispRity() function to do so.\n")
+    }
     ## ...and must have more than one subsets
-    if(length(data$subsets) == 1) stop(paste(as.expression(match_call$data), "must have more than one subset."), call. = FALSE)
+    if(length(data$subsets) == 1){
+        stop.call(match_call$data, " must have more than one subset.")
+    }
 
     ## Check if disparity is a value or a distribution
     is_distribution <- ifelse(length(data$disparity[[1]]$elements) != 1, TRUE, FALSE)
@@ -113,10 +117,14 @@ test.dispRity <- function(data, test, comparisons = "pairwise", rarefaction = NU
     is_bootstrapped <- ifelse(!is.null(data$call$bootstrap), TRUE, FALSE)
 
     ## Stop if disparity is not a distribution, nor bootstrapped
-    if(!is_bootstrapped & !is_distribution) stop(paste(as.expression(match_call$data), "is neither a distribution nor bootstrapped: impossible to compare single values."), call. = FALSE)
+    if(!is_bootstrapped & !is_distribution){
+        stop.call(match_call$data, " is neither a distribution nor bootstrapped: impossible to compare single values.")
+    }
     
     ## Stop if disparity is not bootstrapped and rarefaction is required
-    if(!is_bootstrapped & !is.null(rarefaction)) stop("Impossible to use a rarefaction level for non-bootstrapped data.", call. = FALSE)
+    if(!is_bootstrapped & !is.null(rarefaction)) {
+        stop.call("", "Impossible to use a rarefaction level for non-bootstrapped data.")
+    }
 
     ## Test
     ## must be a single function
@@ -137,17 +145,21 @@ test.dispRity <- function(data, test, comparisons = "pairwise", rarefaction = NU
         
         ## must be pairs
         if(length(unlist(comparisons))%%2 != 0) {
[…]

### 2019-03-06 tests/testthat/make.data/make.data.R

@@ -107,6 +107,7 @@ McClean_data <- list("ordination"=McClean_ordination, "treatment"=treatment, "de
 #####################################
 
 ## Test data
+set.seed(42)
 data(BeckLee_mat99) ; data(BeckLee_ages) ; data(BeckLee_tree)
 data_bootstrapped <- boot.matrix(chrono.subsets(BeckLee_mat99, BeckLee_tree, method = "continuous", rev(seq(from = 0, to = 120, by = 5)), model = "proximity"))
 model_test_data <- dispRity(data_bootstrapped, c(sum, variances))

### 2019-03-06 tests/testthat/test-custom.subsets.R

@@ -218,3 +218,16 @@ test_that("clade subsets works", {
     expect_equal(nrow(without_nodes$subsets[[1]]$elements), nrow(BeckLee_mat50))
     expect_equal(nrow(with_nodes$subsets[[1]]$elements), nrow(BeckLee_mat99))
 })
+
+
+
+test_that("custom.subsets detects distance matrices", {
+    non_dist <- matrix(1:100, 10, 10)
+    rownames(non_dist) <- letters[1:10]
+    is_dist <- as.matrix(dist(non_dist))
+
+    expect_warning(custom.subsets(is_dist, group = list(letters[1:5], letters[6:10])))
+    msg <- capture_warnings(custom.subsets(is_dist, group = list(letters[1:5], letters[6:10])))
+    expect_equal(msg, "custom.subsets is applied on what seems to be a distance matrix.\nThe resulting matrices won't be distance matrices anymore!")
+})
+

### 2019-03-06 tests/testthat/test-print.dispRity.R

@@ -110,16 +110,16 @@ test_that("normal printing", {
 
 test_that("randtest printing", {
     set.seed(1)
-    obs_disparity <- dispRity(BeckLee_mat50, metric = ellipse.volume)
+    obs_disparity <- dispRity(BeckLee_mat50, metric = ellipse.volume, dimensions = 5)
     test <- null.test(obs_disparity, replicates = 100, null.distrib = rnorm)
 
     expect_equal(capture.output(test),
         c("Monte-Carlo test", "Call: [1] \"dispRity::null.test\"",
-        "", "Observation: 6.634e-07 ", 
+        "", "Observation: 0.01698412 ", 
         "", "Based on 100 replicates", 
-        "Simulated p-value: 0.03960396 ", "Alternative hypothesis: two-sided ", 
-        "", "      Std.Obs   Expectation      Variance ",
-        "-1.400160e-01  3.141577e+29  5.034313e+60 "))
+        "Simulated p-value: 0.05940594 ", "Alternative hypothesis: two-sided ", 
+        "", "    Std.Obs Expectation    Variance ",
+        "  -2.023035    5.094270    6.298782 "))
 
     expect_equal(capture.output(print.dispRity(test, all = TRUE)),
         c(
@@ -128,14 +128,14 @@ test_that("randtest printing", {
         "Call: ade4::as.randtest(sim = null_models_results, obs = summary(data, ",
         "    digits = 10)[, 3], alter = alter)"                                  ,
         ""                                                                       ,
-        "Observation: 6.634e-07 "                                                ,
+        "Observation: 0.01698412 "                                                ,
         ""                                                                       ,
         "Based on 100 replicates"                                                ,
-        "Simulated p-value: 0.03960396 "                                         ,
+        "Simulated p-value: 0.05940594 "                                         ,
         "Alternative hypothesis: two-sided "                                     ,
         ""                                                                       ,
-        "      Std.Obs   Expectation      Variance "                             ,
-        "-1.400160e-01  3.141577e+29  5.034313e+60 "                             ,
+        "    Std.Obs Expectation    Variance "                             ,
+        "  -2.023035    5.094270    6.298782 "                             ,
         ""
         ))
 

### 2019-03-06 tests/testthat/test-sanitizing.R

@@ -99,3 +99,50 @@ test_that("check.method works", {
     expect_error(check.method("c", methods))
     expect_null(check.method(c("a", "c"), methods))
 })
+
+## Test distance
+test_that("check.dist.matrix works", {
+    non_dist <- matrix(1:9, 3, 3)
+    is_dist <- as.matrix(dist(non_dist))
+
+    expect_error(check.dist.matrix(non_dist, just.check = "blabla"))
+    expect_error(check.dist.matrix(non_dist, just.check = FALSE))
+    expect_false(check.dist.matrix(non_dist, just.check = TRUE))
+    expect_true(check.dist.matrix(is_dist, just.check = TRUE))
+    test <- check.dist.matrix(non_dist, just.check = FALSE, method = "euclidean")
+    expect_equal(class(test), "list")
+    expect_equal(class(test[[1]]), "dist")
+    expect_false(test[[2]])
+    test <- check.dist.matrix(is_dist, just.check = FALSE, method = "euclidean")
+    expect_equal(class(test), "list")
+    expect_equal(class(test[[1]]), "dist")
+    expect_true(test[[2]])
+    expect_true(all(test[[1]] == dist(non_dist)))
+})
+
+
+## Stop call
+test_that("stop.call works", {
+
+    return.match.call <- function(character, numeric, fun, formula, matrix) {
+        match_call <- match.call()
+        return(match_call)
+    }
+
+    my_matrix <- matrix(NA)
+    call <- return.match.call("a", 1, mean, ~ a + b, my_matrix)
+
+    test <- capture_error(stop.call(call$character, " works."))
+    expect_equal(test[[1]], "a works.")
+    test <- capture_error(stop.call(call$numeric, " works."))
+    expect_equal(test[[1]], "1 works.")
+    test <- capture_error(stop.call(call$fun, " works."))
+    expect_equal(test[[1]], "mean works.")
+    test <- capture_error(stop.call(call$formula, " works."))
+    expect_equal(test[[1]], "~a + b works.")
+    test <- capture_error(stop.call(call$matrix, " works."))
+    expect_equal(test[[1]], "my_matrix works.")
+
+    test <- capture_error(stop.call(call$character, " works.", "look: "))
+    expect_equal(test[[1]], "look: a works.")
+})

### 2019-03-06 inst/gitbook/dispRity_manual.Rmd
bugged

@@
 
- * The CRAN version is a more "strict" version of the package that does not include some functionalities (such as [`Claddis.ordination`](#Claddis-ordination) or this gitbook vignette) because of compatibility issues with the CRAN requirement.
-
+ * The CRAN version is a more "strict" version of the package that does not include some functionalities such as this gitbook vignette because of compatibility issues with the CRAN requirement.
  * The GitHub version is a more "reactive" version (fed by user's suggestions and comments) and can contain more functionalities (see list below).
@@
 
-<!-- badges out start
-[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
-[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
-[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
-badges out end   -->
-
 Note however, that apart from the functions listed below, both versions of the package have the same tested and implemented functions.
-
-Functions not available on the CRAN version (`r version_CRAN`):
-
- * [`Claddis.ordination`](#Claddis-ordination)
 
@@
 If you want the latest-latest version of the package I suggest using the GitHub `master` one, especially if you recently emailed me reporting a minor bug or wanting a new functionality!
-Note however that *it can happen* that the `master` version can sometimes be bugged (especially when there are major `R` and `R` packages updates), however, the status of the package state on both the `release` and the `master` version is constantly displayed on the [`README` page](https://github.com/TGuillerme/dispRity/) of the package with the nice badges displaying these different (and constantly tested) information
-
- * The build status: [![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=master)](https://travis-ci.org/TGuillerme/dispRity) which tests whether the package compiles (i.e.
-can be installed on your computer).
- * The code coverage: [![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/master/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity) which indicates the percentage of lines in the code that have been tested for expected behaviour.
- * The project status: [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active) which indicates whether this package is maintained regularly or forgotten in the abyss of internet.
- * The development version: [![develVersion](https://img.shields.io/badge/devel%20version-1.2.1-green.svg?style=flat)](https://github.com/TGuillerme/dispRity) which indicates which version of the package you are about to install.
+Note however that *it can happen* that the `master` version can sometimes be bugged (especially when there are major `R` and `R` packages updates), however, the status of the package state on both the `release` and the `master` version is constantly displayed on the [`README` page](https://github.com/TGuillerme/dispRity/) of the package with the nice badges displaying these different (and constantly tested) information.
+
+### So which version do I choose? {#version}
+
+There are always three version of the package available:
+
+  * The CRAN one
+  * The GitHub `release` one
+  * The GitHub `master` one
+
+The differences between the CRAN one and the GitHub `release` or `master` ones is explained just above.
+For the the GitHub version, the differences are that the `release` one is more stable (i.e.
+more rarely modified) and the `master` one is more live one (i.e.
+bug fixes and new functionalities are added as they come).
+
+If you want the latest-latest version of the package I suggest using the GitHub `master` one, especially if you recently emailed me reporting a minor bug or wanting a new functionality!
+Note however that *it can happen* that the `master` version can sometimes be bugged (especially when there are major `R` and `R` packages updates), however, the status of the package state on both the `release` and the `master` version is constantly displayed on the [`README` page](https://github.com/TGuillerme/dispRity/) of the package with the nice badges displaying these different (and constantly tested) information.
 
@@
 
-The `dispRity` package also offers some advanced data simulation features to allow to test hypothesis, explore multidimensional spaces, metrics properties or simply playing around with data!
+The `dispRity` package also offers some advanced data simulation features to allow to test hypothesis, explore ordinate-spaces or metrics properties or simply playing around with data!
 All the following functions are based on the same modular architecture of the package and therefore can be used with most of the functions of the package.
@@
 The function `sim.morpho` allows to simulate discrete morphological data matrices (sometimes referred to as "cladistic" matrices).
-It allows to evolve multiple discrete characters on a given phylogenetic tree, given different evolutionary models, rates, and states.
+It allows to evolve multiple discrete characters on a given phylogenetic trees, given different models, rates, and states.
 It even allows to include proper inapplicable data to make datasets as messy as in real life!
 
-In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model and a function from which to draw the rates.
+In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
 The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
@@
 my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
-    0.15), rates = c(rgamma, shape = 0.5, rate = 1), invariant = FALSE)
+    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)
 
 ## The first few lines of the matrix
-my_matrix[1:10, 1:10]
+my_matrix[1:5, 1:10]
 
@@
 Note that this example produces a tree with a great consistency index and an identical topology to the random coalescent tree!
+Nearly too good to be true...
 
@@
 
-The protocol implemented here to generate discrete morphological matrices is based on the ones developed in @GuillermeCooper @OReilly2016 @puttick2017uncertain and @OReilly2017.
+The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;@OReilly2017].
 
  * The first `tree` argument will be the tree on which to "evolve" the characters and therefore requires branch length.
- You can generate quick and easy random Yule trees using `ape::rtree(number_of_taxa)` but I would advise to use more realistic trees for more realistic simulations based on more realistic models (really realistic then) using the function `tree.bd` from the [`diversitree`](http://www.zoology.ubc.ca/prog/diversitree/) package [@fitzjohndiversitree2012] for example.
- * The second argument, `character` is the number of characters.
- * The third, `states` is the proportion of characters states above two.
+ You can generate quick and easy random Yule trees using `ape::rtree(number_of_taxa)` but I would advise to use more realistic trees for more realistic simulations based on more realistic models (really realistic then) using the function `tree.bd` from the [`diversitree`](http://www.zoology.ubc.ca/prog/diversitree/) package [@fitzjohndiversitree2012].
+ * The second argument, `character` is the number of characters. Pretty straight forward.
+ * The third, `states` is the proportion of characters states above two (yes, the minimum number of states is two).
 This argument intakes the proportion of *n*-states characters, for example `states = c(0.5,0.3,0.2)` will generate 50% of binary-state characters, 30% of three-state characters and 20% of four-state characters.
 There is no limit in the number of state characters proportion as long as the total makes up 100%.
- * The fourth, `model` is the evolutionary model for generating the character(s). More about this below.
+ * The forth, `model` is the evolutionary model for generating the character(s). More about this below.
  * The fifth and sixth, `rates` and `substitution` are the model parameters described below as well.
@@
 It uses the four molecular states (A,C,G,T) with a unique overall evolutionary rate and a biased substitution rate towards transitions (A <-> G or C <-> T) against transvertions (A <-> C and G <-> T).
-After evolving the nucleotides, this model transforms them into binary states by converting the purines (A and G) into state 0 and the pyrimidines (C and T) into state 1.
+After evolving the nucleotide, this model transforms them into binary states by converting the purines (A and G) into state 0 and the pyrimidines (C and T) into state 1.
 This method is based on the `phyclust::seq.gen.HKY` function and was first proposed by @OReilly2016.
@@
 The models can take the following parameters:
-(1) `rates` is the evolutionary rate (i.e. the rate of changes along a branch; the evolutionary speed) and (2) `substitution` is the frequency of changes between one state or another.
+(1) `rates` is the evolutionary rate (i.e. the rate of changes along a branch: the evolutionary speed) and (2) `substitution` is the frequency of changes between one state or another.
 For example if a character can have high probability of changing (the *evolutionary* rate) with, each time a change occurs a probability of changing from state *X* to state *Y* (the *substitution* rate).
-Note that in the `"ER"` model, the substitution rate is ignore because by definition this rate is equal!
+Note that in the `"ER"` model, the substitution rate is ignore because... by definition this (substitution) rate is equal!
 
@@
 
-This type of characters can be added to the simulated matrices using the `apply.NA` function.
+This type of characters can be added to the simulated matrices using the `apply.NA` function/
 It takes, as arguments, the `matrix`, the source of inapplicability (`NAs` - more below), the `tree` used to generate the matrix and the two same `invariant` and `verbose` arguments as defined above.
@@
 not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all parameters combination end up creating "good" matrices.
-The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 10 and no scaling ($\beta$ = 1) with a rate of 50.
+The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
 
@@
 ## tree
-my_tree <- rcoal(30)
+my_tree <- rcoal(15)
 ## matrix
-morpho_mat <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
-    0.15), rates = c(rgamma, shape = 10, rate = 50), invariant = FALSE)
-
+morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
+    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
 check.morpho(morpho_mat, my_tree)
@@
 ## Simulating multidimensional spaces
-Another way to simulate data is to directly simulate a multidimensional space with the `space.maker` function.
-This function allows users to simulate spaces with a certain number of properties.
-For example, it is possible to design a space with a specific distribution on each axis, a correlation between the axes and a specific cumulative variance per axis.
+Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
+This function allows users to simulate multidimensional spaces with a certain number of properties.
+For example, it is possible to design a multidimensional space with a specific distribution on each axis, a correlation between the axes and a specific cumulative variance per axis.
 This can be useful for creating ordinated spaces for null hypothesis, for example if you're using the function `null.test` [@diaz2016global].
@@
 This function takes as arguments the number of elements (data points - `elements` argument) and dimensions (`dimensions` argument) to create the space and the distribution functions to be used for each axis.
-The distributions are passed through the `distribution` argument as modular functions.
+The distributions are passed through the `distribution` argument as... modular functions!
 You can either pass a single distribution function for all the axes (for example `distribution = runif` for all the axis being uniform) or a specific distribution function for each specific axis (for example `distribution = c(runif, rnorm, rgamma))` for the first axis being uniform, the second normal and the third gamma).
@@
                            arguments = list(list(min = 0, max = 0), NULL, NULL))
-head(plane_space)
 
@@
 ellipse_space <- space.maker(2500, 3, rnorm, cor.matrix = cor_matrix)
-head(ellipse_space)
 
@@
                                  scree = c(0.7, 0.2, 0.1))
-head(cylindrical_space)
-```
-
-See below for the visualisation of the spaces.
+```
 
@@
 For example, the `random.circle` function is a personalised one implemented in `dispRity`.
-This function allows to create circles based on trigonometry allowing two axis to covary to produce circle coordinates.
+This function allows to create circles based on basic trigonometry allowing to axis to covary to produce circle coordinates.
 By default, this function generates two sets of coordinates with a `distribution` argument and a minimum and maximum boundary (`inner` and `outer` respectively) to create nice sharp edges to the circle.
@@
      arguments = list(list(mean = 0), list(runif, inner = 0.5, outer = 1)))
-head(doughnut_space)
-```
-
-See below for the visualisation of the "doughnut" space.
+```
 
@@
 legend("topleft", legend = c("observed", "simulated"),
-        pch = c(20,21), col = c("red", "black"), bg = "white")
+        pch = c(20,21), col = c("red", "black"))
 ```
@@
 plot(test_norm1, main = "Normal (0,1)")
-plot(test_norm2, main = "Normal (global observed\nmean and sd)")
-plot(test_norm3, main = "Normal (variable mean\nand sd with correlation)")
+plot(test_norm2, main = paste0("Normal (", round(obs_mu_sd_glob[[1]], digit = 3),
+                              ",", round(obs_mu_sd_glob[[2]], digit = 3), ")"))
+plot(test_norm3, main = "Normal (variable + correlation)")
 
@@
 
-# Future directions
-
-## More tests!
-
-Some more tests are being developed such as a `sequential.test` to run sequential linear models for testing hypothesis in disparity through time or a `model.test` developed in collaboration with [Mark Puttick](https://puttickbiology.wordpress.com/) to fit modes of evolution to disparity curves.
-Stay tuned!
-
-## More rarefactions!
-
-I am working on including several "clever" bootstrap and rarefaction algorithms for palaeobiology studies based on geological sampling (following:and Butler et al.
-2012 *Evolution* and Kotric and Knoll 2015 *Palaeobiology*).
-
-## Faster disparity calculations
-
-I am slowly implementing parallel disparity calculation as well as C implementations of some disparity metrics to increase significantly improve the speed of the `dispRity` function.
-
-## More modularity
-
-I am equally slowly developing functions to allow more of the options in the package to be modular (in the same way as the `metric` argument in `dispRity`).
-The next arguments to benefit this increased modularity will be `chrono.subsets`'s `model` argument and `boot.matrix`'s `type` argument.
-
-## More suggestions?
-
-More generally, I welcome any suggestions or collaborations for developing some additional specific ideas.
-Feel free to contact me via email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
+# Developing `dispRity`
+
+Everyone is more than welcome to help developing this package, whether it is through developing functions (metrics, tests, plots, etc...), the manual (more examples, typos, etc...) or completely new functionalities!
+
+## More functions
+
+## More manuals
 

### 2019-03-06 inst/gitbook/index.Rmd
bugged

--- 2019-03-06 inst/gitbook/index.Rmd
@@
 
-Note that when we (the `dispRity` package collaborators!) are developing some more major versions, including some more complex features for example, they will never be released on either the the versions described above.
-Therefore you can rest assure that you will never use a badly bugged version of `dispRity`! Enjoy!
-
 ## Help

### 2019-04-10 inst/gitbook/03_specific-tutorials.Rmd
updated gitbook

--- 2018-10-12 inst/gitbook/03_specific-tutorials.Rmd
@@
 1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
+2 | `displacements` | The ratio between the distance from a reference and the distance from the centroid | `dispRity` |
 1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
@@
 1 | `n.ball.volume` | The hyper-spherical (*n*-ball) volume | `dispRity` |
+2 | `neighbours` | The distance to specific neighbours (e.g. the nearest neighbours - by default) | `dispRity` |
 2 | `pairwise.dist` | The pairwise distances between elements | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::vegist` |
+2 | `quantiles` | The *n*th quantile range per axis | `dispRity` |
 2 | `radius` | The radius of each dimensions | `dispRity` |
@@
     diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
+\end{equation}
+
+\begin{equation}
+    displacements = \frac{\sqrt{\sum_{i=1}^{n}{({k}_{n}-Reference_{k})^2}}}{\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}}
 \end{equation}
@@
 $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.),
-*R* is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid).
+*R* is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid),
+and $Reference_{k}$ is an arbitrary point's coordinates (usually 0).
 
@@
 
-#### Ranges, variances, radius, pairwise distance, modal value and diagonal
+#### Ranges, variances, quantiles, radius, pairwise distance, neighbours, modal value and diagonal
 
@@
 
-`ranges`, `variances` and `radius` work on the same principle and measure the range/variance/radius of each dimension:
+`ranges`, `variances` `quantiles` and `radius` work on the same principle and measure the range/variance/radius of each dimension:
 
@@
 summary(dispRity(dummy_space, metric = c(prod, variances)))
+
+## Calculating the quantiles of each dimension in the ordinated space
+quantiles(dummy_space)
+
+## Calculating disparity as the distribution of these variances
+summary(dispRity(dummy_space, metric = quantiles))
+
+## By default, the quantile calculated is the 95% (i.e. 95% of the data on each axis)
+## this can be changed using the option quantile:
+summary(dispRity(dummy_space, metric = quantiles, quantile = 50))
 
@@
 
-The pairwise distances uses the function `vegan::vegdist` and can take the normal `vegdist` options:
+The pairwise distances and the neighbours distances uses the function `vegan::vegdist` and can take the normal `vegdist` options:
 
@@
 summary(dispRity(dummy_space, metric = pairwise.dist, method = "manhattan"))
+
+## The average nearest neighbour distances
+summary(dispRity(dummy_space, metric = neighbours))
+
+## The average furthest neighbour manhattan distances
+summary(dispRity(dummy_space, metric = neighbours, which = max, method = "manhattan"))
 ```
@@
 
-#### Centroids and ancestral distance metrics
+#### Centroids, displacements, and ancestral distance metrics
 
@@
 ## The distance between each element and the origin of the ordinated space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))
+summary(dispRity(dummy_space, metric = centroids, centroid = 0))
 
@@
 summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
+```
+
+The `displacements` distance is the ratio between the `centroids` distance and the `centroids` distance with `centroid = 0`.
+Note that it is possible to measure a ratio from another point than `0` using the `reference` argument.
+It gives indication of the relative displacement of elements in the multidimensional space: a score  >1 signifies a displacement *away* from the reference.
+A score of >1 signifies a displacement *towards* the reference.
+
+```{r}
+## The relative displacement of the group in space to the centre
+summary(dispRity(dummy_space, metric = displacements))
+
+## The relative displacement of the group to an arbitrary point
+summary(dispRity(dummy_space, metric = displacements, reference = c(0,1,2,3,4)))
 ```
@@
                            nodes.coords = all_anc_centroids)
+
+## The ancestral distances distributions
+summary(ancestral_dist)
 
@@
 
- * `quantile` values for the confidence interval levels (by default, the 50 and 95 quantiles are calculated)
+ * `quantiles` values for the confidence interval levels (by default, the 50 and 95 quantiles are calculated)
  * `cent.tend` for the central tendency to use for summarising the results (default is `median`)
@@
 ## Same as above but using the 88th quantile and the standard deviation as the summary 
-summary(disparity_time_slices, quantile = 88, cent.tend = sd)
+summary(disparity_time_slices, quantiles = 88, cent.tend = sd)
 
@@
 ## Plotting the results with some plot.dispRity arguments
-plot(disparity_time_slices, quantile = c(seq(from = 10, to = 100, by = 10)),
+plot(disparity_time_slices, quantiles = c(seq(from = 10, to = 100, by = 10)),
     cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),

### 2019-07-10 inst/gitbook/05_other-functionalities.Rmd
man pre-update

---
title: "other functionalities"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: [../References.bib, ../packages.bib]
bst: sysbio.bst
vignette: >
  %\VignetteIndexEntry{other functionalities}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Other functionalities

The `dispRity` package also contains several other functions that are not specific to multidimensional analysis but that are often used by `dispRity` internal functions.
However, we decided to make these functions also available at a user level since they can be handy for certain specific operations!
You'll find a brief description of each of them (alphabetically) here:

## `clean.data`

This is a rather useful function that allows matching a `matrix` or a `data.frame` to a tree (`phylo`) or a distribution of trees (`multiPhylo`).
This function outputs the cleaned data and trees (if cleaning was needed) and a list of dropped rows and tips.

```{r}
## Generating a trees with labels from a to e
dummy_tree <- rtree(5, tip.label = LETTERS[1:5])

## Generating a matrix with rows from b to f
dummy_data <- matrix(1, 5, 2, dimnames = list(LETTERS[2:6], c("var1", "var2")))

##Cleaning the trees and the data
(cleaned <- clean.data(data = dummy_data, tree = dummy_tree))
```

## `crown.stem`

This function quiet handily separates tips from a phylogeny between crown members (the living taxa and their descendants) and their stem members (the fossil taxa without any living relatives).

```{r}
data(BeckLee_tree)
## Diving both crow and stem species
(crown.stem(BeckLee_tree, inc.nodes = FALSE))
```

Note that it is possible to include or exclude nodes from the output.
To see a more applied example: this function is used in [chapter 03: specific tutorials](#custom-subsets).

## `get.bin.ages`

This function is similar than the `crown.stem` one as it is based on a tree but this one outputs the stratigraphic bins ages that the tree is covering.
This can be useful to generate precise bin ages for the `chrono.subsets` function:

```{r}
get.bin.ages(BeckLee_tree)
```

Note that this function outputs the stratigraphic age limits by default but this can be customisable by specifying the `type` of data (e.g. `type = "Eon"` for eons).
The function also intakes several optional arguments such as whether to output the startm end, range or midpoint of the stratigraphy or the year of reference of the International Commission of Stratigraphy.
To see a more applied example: this function is used in [chapter 03: specific tutorials](#chrono-subsets).

## `pair.plot`

This utility function allows to plot a matrix image of pairwise comparisons.
This can be useful when getting pairwise comparisons and if you'd like to see at a glance which pairs of comparisons have high or low values.

```{r}
## Random data
data <- matrix(data = runif(42), ncol = 2)

## Plotting the first column as a pairwise comparisons
pair.plot(data, what = 1, col = c("orange", "blue"), legend = TRUE, diag = 1)
```

Here blue squares are ones that have a high value and orange ones the ones that have low values.
Note that the values plotted correspond the first column of the data as designated by `what = 1`.

It is also possible to add some tokens or symbols to quickly highlight to specific cells, for example which elements in the data are below a certain value:

```{r}
## The same plot as before without the diagonal being the maximal observed value
pair.plot(data, what = 1, col = c("orange", "blue"), legend = TRUE, diag = "max")
## Highlighting with an asterisk which squares have a value below 0.2
pair.plot(data, what = 1, binary = 0.2, add = "*", cex = 2)
```

This function can also be used as a binary display when running a series of pairwise t-tests.
For example, the following script runs a wilcoxon test between the time-slices from the `disparity` example dataset and displays in black which pairs of slices have a p-value below 0.05:

```{r}
## Loading disparity data
data(disparity)

## Testing the pairwise difference between slices
tests <- test.dispRity(disparity, test = wilcox.test, correction = "bonferroni")

## Plotting the significance
pair.plot(as.data.frame(tests), what = "p.value", binary = 0.05)
```

## `reduce.matrix`

This function allows to reduce columns or rows of a matrix to make sure that there is enough overlap for further analysis.
This is particularly useful if you are going to use distance matrices since it uses the `vegan::vegdist` function to test whether distances can be calculated or not.

For example, if we have a patchy matrix like so (where the black squares represent available data):

```{r}
set.seed(1)
## A 10*5 matrix
na_matrix <- matrix(rnorm(50), 10, 5)
## Making sure some rows don't overlap
na_matrix[1, 1:2] <- NA
na_matrix[2, 3:5] <- NA
## Adding 50% NAs
na_matrix[sample(1:50, 25)] <- NA
## Illustrating the gappy matrix
image(t(na_matrix), col = "black")
```

We can use the `reduce.matrix` to double check whether any rows cannot be compared.
The functions needs as an input the type of distance that will be used, say a `"gower"` distance:

```{r}
## Reducing the matrix by row
(reduction <- reduce.matrix(na_matrix, distance = "gower"))
```

We can not remove the rows 1 and 9 and see if that improved the overlap:

```{r}
image(t(na_matrix[-as.numeric(reduction$rows.to.remove), ]), col = "black")
```

## `slice.tree`

This function is a modification of the `paleotree::timeSliceTree` function that allows to make slices through a phylogenetic tree.
Compared to the `paleotree::timeSliceTree`, this function allows a model to decide which tip or node to use when slicing through a branch (whereas `paleotree::timeSliceTree` always choose the first available tip alphabetically).
The models for choosing which tip or node are the same as the ones used in the `chrono.subsets` and are described in [chapter 03: specific tutorials](#chrono-subsets).

The function works by using at least a tree, a slice age and a model:

```{r}
set.seed(1)
## Generate a random ultrametric tree
tree <- rcoal(20)
## Add some node labels
tree$node.label <- letters[1:19]
## Add its root time
tree$root.time <- max(tree.age(tree)$ages)

## Slicing the tree at age 0.75
tree_75 <- slice.tree(tree, age = 0.75, "acctran")

## Showing both trees
par(mfrow = c(1,2))
plot(tree, main = "original tree")
axisPhylo() ; nodelabels(tree$node.label, cex = 0.8)
abline(v = (max(tree.age(tree)$ages) - 0.75), col = "red")
plot(tree_75, main = "sliced tree")
```

## `tree.age`

This function allows to quickly calculate the ages of each tips and nodes present in a tree.

```{r}
set.seed(1)
tree <- rtree(10)
## The tree age from a 10 tip tree
tree.age(tree)
```

It also allows to set the age of the root of the tree:

```{r}
## The ages starting from -100 units
tree.age(tree, age = 100)
```

Usually tree age is calculated from the present to the past (e.g. in million years ago) but it is possible to reverse it using the `order = present` option:

```{r}
## The ages in terms of tip/node height
tree.age(tree, order = "present")
```

### 2019-07-10 inst/gitbook/09_future_directions.Rmd
man pre-update

--- 2017-11-17 inst/gitbook/dispRity_manual.Rmd
@@
---- 
-title: "dispRity manual"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-site: bookdown::bookdown_site
-output: bookdown::gitbook
-documentclass: book
-link-citations: yes
-github-repo: TGuillerme/dispRity
-description: "dispRity R package vignette"
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-# `dispRity`
-
-This is a package for measuring disparity in `R`.
-It allows users to summarise matrices as representations as multidimensional spaces into a single value or distribution describing a specific aspect of this multidimensional space (the disparity).
-Multidimensional spaces can be ordinated matrices from MDS, PCA, PCO, PCoA but the package is *not* restricted to any type of matrices!
-This manual is based on the version `0.4.2`.
-
-## What is `dispRity`?
-This is a modular package for measuring disparity in `R`.
-It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
-Typically, these analysis are used in palaeobiology and evolutionary biology to study the changes in morphology through time.
-However, there are many more applications in ecology, evolution and beyond.
-
-### Modular?
-Because their exist a multitude of ways to measure disparity, each adapted to every specific question, this package uses an easy to modify modular architecture.
-In coding, each module is simply a function or a modification of a function that can be passed to the main functions of the package to tweak it to your proper needs!
-In practice, you will notice throughout this manual that some function can take other functions as arguments: the modular architecture of this package allows you to use any function for these arguments (with some restrictions explained for each specific cases).
-This will allow you to finely tune your multidimensional analysis to the needs of your specific question!
-
-## Installing and running the package
-
-You can install this package easily if you use the latest version of `R` (> 3.4.0) and `devtools`.
-
-```{r, eval=FALSE}
-## Checking if devtools is already installed
-if(!require(devtools)) install.packages("devtools")
-
-## Installing the latest released version directly from GitHub
-install_github("TGuillerme/dispRity", ref = "release")
-
-## Loading the package
-library(dispRity)
-```
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## loading the package and setting up the start seed.
-library(dispRity)
-## Setting a random seed for repeatability
-set.seed(123)
-```
-
-Note this uses the `release` branch (version`0.4`). <!-- @@@ -->
-For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
-`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, `Claddis`, `geomorph` and `RCurl`).
-
-## Why not CRAN?
-
-This package is not available on CRAN.
-This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
-However, the package follows the strict (and useful!) CRAN standards via Travis.
-
-<!-- badges out start -->
-[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
-[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/master/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
-[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
-<!-- badges out end   -->
-
-## Help
-
-If you need help with the package, hopefully the following manual will be useful.
-However, parts of this package are still in development and some other parts are probably not covered.
-Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
-
-## Citations
-
-You can cite both the package or this manual with the following citation:
-
-> Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646
-
-<!-- > Guillerme, T. & Cooper, N. (2016). The dispRity R package manual. Fighsare XXXX -->
-
-Note that this citation is only temporary (but can still be used!).
-A proper description of the package is currently in review in Methods in Ecology and Evolution and should be re-submitted this winter.
-Upon acceptance (hopefully!), a proper DOI will be released for this manual and the package description.
-
-<!--chapter:end:index.Rmd-->
-
----
-title: "Glossary"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Glossary
-
--   **Multidimensional space**. The mathematical multidimensional object that will be analysed with this package.
-    In morphometrics, this is often referred to as the morphospace.
-    However it may also be referred to as the cladisto-space for cladistic data or the eco-space for ecological data etc.
-    In practice, this term designates a matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
-
--   **Elements**. The rows of the multidimensional space matrix. Elements can be taxa, field sites, countries etc.
-
--   **Dimensions**.
-The columns of the multidimensional space matrix.
-The dimensions can be referred to as axes of variation, or principal components, for ordinated spaces obtained from a PCA for example.
-
--   **Subsets**. Subsets of the multidimensional space.
-    A subset (or subsets) contains the same number of dimensions as the space but may contain a smaller subset of elements.
-    For example, if our space is composed of birds and mammals (the elements) and 50 principal components of variation (the dimensions), we can create two subsets containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
-
-<!--chapter:end:01_glossary.Rmd-->
-
----
-title: "Getting started with dispRity"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Getting started with `dispRity`
-
-## What sort of data does `dispRity` work with?
-
-Disparity can be estimated from pretty much any matrix.
-Classically, however, it is measured from ordinated matrices.
-These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).
-
-### Ordination matrices from `Claddis`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
-if(!require(Claddis)) install.packages("Claddis")
-```
-
-`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
-For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:
-
-```{r}
-require(Claddis)
-
-## Ordinating the example data from Claddis
-Claddis.ordination(Michaux1989) 
-```
-
-Note that several options are available, namely which type of distance should be computed.
-See more info in the function manual (`?Claddis.ordination`).
-Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.
-
-### Ordination matrices from `geomorph`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if geomorph is installed BUILD ONLY
-if(!require(geomorph)) install.packages("geomorph")
-```
-
-You can also easily use data from `geomorph` using the `geomorph.ordination` function.
-This function simply takes Procrustes aligned data and performs an ordination:
-
-```{r}
-require(geomorph)
-
-## Loading the plethodon dataset
-data(plethodon)
-
-## Performing a Procrustes transform on the landmarks
-procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)
-
-## Ordinating this data
-geomorph.ordination(procrustes)[1:5,1:5]
-```
-
-Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
-Additionally you can give the function a `geomorph.data.frame` object.
-If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsets)!
-
-```{r}
-## Using a geomorph.data.frame
-geomorph_df <- geomorph.data.frame(procrustes,
-     species = plethodon$species, site = plethodon$site)
-
-## Ordinating this data and making a dispRity object
-geomorph.ordination(geomorph_df)
-```
-
-More about these `dispRity` objects below!
-
-### Other kinds of ordination matrices
-
-If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordination matrices by using the following functions from the `stats` package.
-Here is how to do it for the following types of matrices:
-
- * Multivariate matrices (principal components analysis; PCA)
-
-```{r}
-## A multivariate matrix
-head(USArrests)
-
-## Ordinating the matrix using `prcomp` 
-ordination <- prcomp(USArrests)
-
-## Selecting the ordinated matrix
-ordinated_matrix <- ordination$x
-head(ordinated_matrix)
-```
-  This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, see the `?princomp` function.
-
- * Distance matrices (classical multidimensional scaling; MDS)
-
-```{r}
-## A matrix of distances between cities
-str(eurodist)
-
-## Ordinating the matrix using cmdscale() with k = 5 dimensions 
-ordinated_matrix <- cmdscale(eurodist, k = 5)
-head(ordinated_matrix)
-```
-
-  This results in a ordinated matrix with European cities as elements and five dimensions.
-
-Of course any other method for creating the ordination matrix is totally valid, you can also not use any ordination at all!
-The only requirements for the `dispRity` functions is that the input is a matrix with elements as rows and dimensions as columns.
-
-## Performing a simple dispRity analysis
-
-Two `dispRity` functions allow users to run an analysis pipeline simply by inputting an ordination matrix.
-These functions allow users to either calculate the disparity through time (`dispRity.through.time`) or the disparity of user-defined groups (`dispRity.per.group`).
-
-**IMPORTANT**
-
-Note that `disparity.through.time` and `disparity.per.group` are wrapper functions (i.e.
-they incorporate lots of other functions) that allow users to run a basic disparity-through-time, or disparity among groups, analysis without too much effort.
-As such they use a lot of default options. 
-These are described in the help files for the functions that are used to make the wrapper functions, and not described in the help files for `disparity.through.time` and `disparity.per.group`. 
-These defaults are good enough for **data exploration**, but for a proper analysis you should consider the **best parameters for your question and data**. 
-For example, which metric should you use?
-How many bootstraps do you require?
-What model of evolution is most appropriate if you are time slicing?
-Should you rarefy the data?
-See [`time.subsets`](#time-slicing), [`custom.subsets`](#customised-subsets), [`boot.matrix`](#bootstraps-and-rarefactions) and [`dispRity.metric`](#disparity-metrics) for more details of the defaults used in each of these functions.
-Note that any of these default arguments can be changed within the `disparity.through.time` or `disparity.per.group` functions.
-
-### Example data
-
-To illustrate these functions, we will use data from @beckancient2014.
-This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
-
-```{r, fig.width=7, fig.height=7}
-## Loading the ordinated matrices
-data(BeckLee_mat50)
-data(BeckLee_mat99)
-
-## The first five taxa and dimensions of the 50 taxa matrix
-head(BeckLee_mat50[, 1:5])
-
-## The first five taxa and dimensions of the 99 taxa + ancestors matrix
-BeckLee_mat99[c(1, 2, 98, 99), 1:5]
-
-## Loading a list of first and last occurrence dates for the fossils
-data(BeckLee_ages)
-head(BeckLee_ages)
-
-## Loading and plotting the phylogeny
-data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.8) 
-axisPhylo(root = 140)
-nodelabels(cex = 0.5)
-```
-
-Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).
-
-### Disparity through time
-
-The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
-This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
-Of course all the defaults can be changed if required, more on this later.
-
-For a disparity through time analysis, you will need:
-
-  * An ordinated matrix (we covered that above)
-  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element.
-To give your tree a root time (i.e.
-an age for the root), you can simply do\\ `my_tree$root.time <- my_age`.
-  * The required number of time subsets (here `time = 3`)
-  * Your favourite disparity metric (here the sum of variances)
-
-Using the Beck and Lee (2014) data described [above](#example-data):
-
-```{r}
-## Measuring disparity through time
-disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
-                                        time = 3, metric = c(sum, variances))
-```
-
-This generates a `dispRity` object (see [here](#guts) for technical details).
-When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:
-
-```{r}
-## Print the disparity_data object
-disparity_data
-```
-
-We asked for three subsets (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
-
-We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):
-
-```{r, fig.width=7, fig.height=7}
-## Summarising disparity through time
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data, type = "continuous")
-
-## Testing for an difference among the time bins
-disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
-summary(disp_lm)
-```
-
-Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
-You can also directly explore the specific function help files within R and navigate to related functions.
-
-### Disparity among groups
-
-The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
-For example, you could ask if there is a difference in disparity between two groups?
-
-To perform such an analysis, you will need:
-
- * An matrix with rows as elements and columns as dimensions (always!)
- * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix.
-For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4.
-Note that elements can be present in multiple groups at once.
- * Your favourite disparity metric (here the sum of variances)
-
-Using the @beckancient2014 data described [above](#example-data):
-
-```{r}
-## Creating the two groups (crown versus stem) as a list
-mammal_groups <- crown.stem(BeckLee_tree, inc.nodes = TRUE)
-
-## Measuring disparity for each group
-disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
-                                     metric = c(sum, variances))
-```
-
-We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).
-
-```{r, fig.width=7, fig.height=7}
-## Print the disparity_data object
-disparity_data
-
-## Summarising disparity in the different groups
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data)
-
-## Testing for a difference between the groups
-test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
-```
-
-<!--chapter:end:02_getting-started.Rmd-->
-
----
-title: "Details of specific functions"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Details of specific functions
-
-The following section contains information specific to some functions.
-If any of your questions are not covered in these sections, please refer to the function help files in R, send me an email (<guillert@tcd.ie>), or raise an issue on GitHub.
-The several tutorials below describe specific functionalities of certain functions; please always refer to the function help files for the full function documentation!
-
-Before each section, make sure you loaded the @beckancient2014 data (see [example data](#example-data) for more details).
-
-```{r}
-## Loading the data
-data(BeckLee_mat50) ; data(BeckLee_mat99)
-data(BeckLee_tree) ; data(BeckLee_ages)
-```
-
-## Time slicing
-
-The function `time.subsets` allows users to divide the matrix into different time subsets or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
-Each subset generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.
-
-Two types of time subsets can be performed by using the `method` option:
-
- *  Discrete time subsets (or time-binning) using `method = discrete`
- *  Continuous time subsets (or time-slicing) using `method = continuous`
-
-For the time-slicing method details see Cooper and Guillerme (in prep.). <!-- @@@ Change cite appropriately! -->
-<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
-For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:
-
- *  Defining the boundaries of the time bins (when `method = discrete`)
- *  Defining the time slices (when `method = continuous`)
-
-Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
-Additionally, it is also possible to input a dataframe containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age, so taxa can appear in more than one time bin/slice.
-
-Here is an example for `method = discrete`:
-
-```{r, eval=TRUE}
-## Generating three time bins containing the taxa present every 40 Ma
-time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0))
-```
-
-Note that we can also generate equivalent results by just telling the function that we want three time-bins as follow:
-
-```{r, eval=TRUE}
-## Automatically generate three equal length bins:
-time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = 3)
-```
-
-In this example, the taxa were split inside each time-bin according to their age.
-However, the taxa here are considered as single points in time.
-It is totally possible that some taxa could have had longer longevity and that they exist in multiple time bins.
-In this case, it is possible to include them in more than one bin by providing a table of first and last occurrence dates (FAD/LAD).
-This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:
-
-```{r, eval=TRUE}
-## Displaying the table of first and last occurrence dates for each taxa
-head(BeckLee_ages)
-
-## Generating time bins including taxa that might span between them
-time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
-```
-
-When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as the root age plus 1\% of the tree length, as long as at least three elements/taxa are present at that point in time.
-The algorithm adds an extra 1\% tree length until reaching the required minimum of three elements.
-It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance among tips *and* nodes.
-
-For the time-slicing method (`method = continuous`), the idea is fairly similar.
-This option, however, requires a matrix that contains the ordinated distance among taxa *and* nodes and an extra argument describing the assumed evolutionary model (via the `model` argument).
-This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node, meaning that a decision must be made about what the value for the branch should be.
-The model can be one of the following:
-
- *  `acctran` where the data chosen along the branch is always the one of the descendant
- *  `deltran` where the data chosen along the branch is always the one of the ancestor
- *  `random` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
- *  `proximity` where the data chosen along the branch is either the descendant or the ancestor depending on branch length
- *  `punctuated` where the data chosen along the branch is both the descendant and the ancestor with an even probability
- *  `gradual` where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length
-
-> Note that the four first models are "fixed": the selected data is always either the one of the descendant or the ancestor.
-The two last models are "probabilistic": the data from both the descendant and the ancestor is used with an associate probability.
-These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants.
-
-```{r, eval=TRUE}
-## Generating four time slices every 40 million years under a model of proximity evolution
-time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-
-## Generating four time slices automatically
-time.subsets(data = BeckLee_mat99, tree = BeckLee_tree,
-    method = "continuous", model = "proximity", time = 4, FADLAD = BeckLee_ages)
-```
-
-## Customised subsets
-
-Another way of separating elements into different categories is to use customised subsets as briefly explained [above](#disparity-among-groups).
-This function simply takes the list of elements to put in each group (whether they are the actual element names or their position in the matrix).
-
-```{r, eval=TRUE}
-## Creating the two groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44))
-
-## Separating the dataset into two different groups
-custom.subsets(BeckLee_mat50, group = mammal_groups)
-```
-
-Elements can easily be assigned to different groups if necessary!
-
-```{r, eval=FALSE}
-## Creating the three groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44).
-                      "all" = c(1:50))
-```
-
-## Bootstraps and rarefactions
-
-One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results are, and how sensitive they are to outliers in the dataset.
-This can be achieved using the function `boot.matrix` to bootstrap and/or rarefy the data.
-The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):
-
-```{r, eval=TRUE}
-## Default bootstrapping
-boot.matrix(data = BeckLee_mat50)
-```
-
-The number of bootstrap replicates can be defined using the `bootstraps` option.
-The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
-Currently two algorithms are implemented:
-
- * `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any *m* elements drawn from the data)
- * `single` where only one random element is replaced by one other random element for each pseudo-replicate
-
-```{r, eval=TRUE}
-## Bootstrapping with the single bootstrap method
-boot.matrix(BeckLee_mat50, boot.type = "single")
-```
-
-This function also allows users to rarefy the data using the `rarefaction` argument.
-Rarefaction allows users to limit the number of elements to be drawn at each bootstrap replication.
-This is useful if, for example, one is interested in looking at the effect of reducing the number of elements on the results of an analysis.
-
-This can be achieved by using the `rarefaction` option that draws only *n-x* at each bootstrap replicate (where *x* is the number of elements not sampled).
-The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subset to a minimum of three elements).
-It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
-
-```{r, eval=TRUE}
-## Bootstrapping with the full rarefaction
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
-
-## Or with a set number of rarefaction levels
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8, 3))
-```
-
-One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
-When missing, all dimensions from the ordinated matrix are used.
-
-```{r, eval=TRUE}
-## Using the first 50% of the dimensions
-boot.matrix(BeckLee_mat50, dimensions = 0.5)
-
-## Using the first 10 dimensions
-boot.matrix(BeckLee_mat50, dimensions = 10)
-```
-
-Of course, one could directly supply the subsets generated above (using `time.subsets` or `custom.subsets`) to this function.
-
-```{r, eval=TRUE}
-## Creating subsets of crown and stem mammals
-crown_stem <- custom.subsets(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot.matrix(crown_stem, bootstraps = 200, rarefaction = TRUE)
-
-## Creating time slice subsets
-time_slices <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-                               method = "continuous", model = "proximity", 
-                               time = c(120, 80, 40, 0),
-                               FADLAD = BeckLee_ages)
-
-## Bootstrapping the time slice subsets
-boot.matrix(time_slices, bootstraps = 100)
-```
-
-## Disparity metrics
-
-There are many ways of measuring disparity!
-In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
-For example, one can look at ellipsoid hyper-volume of the ordinated space (Donohue _et al._ 2013), the sum and the product of the ranges and variances (Wills _et al._ 1994) or the median position of the elements relative to their centroid (Wills _et al._ 1994).
-Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.
-
-Because of this great diversity of metrics, the package `dispRity` does not have one way to measure disparity but rather proposes to facilitate users in defining their own disparity metric that will best suit their particular analysis.
-In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
-However the `metric` argument has to follow certain rules:
-
-1.  It must be composed from one to three `function` objects;
-2.  The function(s) must take as a first argument a `matrix` or a `vector`;
-3.  The function(s) must be of one of the three dimension-levels described below;
-4.  At least one of the functions must be of dimension-level 1 or 2 (see below).
-
-### The function dimension-levels
-
-The metric function dimension-levels determine the "dimensionality of decomposition" of the input matrix.
-In other words, each dimension-level designates the dimensions of the output, i.e. either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimension.
-
-![Illustration of the different dimension-levels of functions with an input `matrix`](dispRity_fun.png)
-
-#### Dimension-level 1 functions
-
-A dimension-level 1 function will decompose a `matrix` or a `vector` into a single value:
-
-```{r}
-## Creating a dummy matrix
-dummy_matrix <- matrix(rnorm(12), 4, 3)
-
-## Example of dimension-level 1 functions
-mean(dummy_matrix)
-median(dummy_matrix)
-```
-
-Any summary metric such as mean or median are good examples of dimension-level 1 functions as they reduce the matrix to a single dimension (i.e. one value).
-
-#### Dimension-level 2 functions
-
-A dimension-level 2 function will decompose a `matrix` into a `vector`.
-
-```{r}
-## Defining the function as the product of rows
-prod.rows <- function(matrix) apply(matrix, 1, prod)
-
-## A dimension-level 2 metric
-prod.rows(dummy_matrix)
-```
-
-Several dimension-level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculate the variance or the range of each dimension of the ordinated matrix respectively.
-
-#### Dimension-level 3 functions
-
-Finally a dimension-level 3 function will transform the matrix into another matrix.
-Note that the dimension of the output matrix doesn't need to match the the input matrix:
-
-```{r}
-## A dimension-level 3 metric
-var(dummy_matrix)
-
-## A dimension-level 3 metric with a forced matrix output
-as.matrix(dist(dummy_matrix))
-```
-
-### `make.metric`
-
-Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the Euclidean distance between each element and the centroid of the ordinated space.
-The `make.metric` function implemented in `dispRity` is designed to help test and find the dimension-level of the functions.
-This function tests:
-
-1.  If your function can deal with a `matrix` or a `vector` as an input;
-2.  Your function's dimension-level according to its output (dimension-level 1, 2 or 3, see above);
-3.  Whether the function can be implemented in the `dispRity` function (the function is fed into a `lapply` loop).
-
-For example, let's see if the functions described above are the right dimension-levels:
-
-```{r}
-## Which dimension-level is the mean function? And can it be used in dispRity?
-make.metric(mean)
-
-## Which dimension-level is the prod.rows function? And can it be used in dispRity?
-make.metric(prod.rows)
-
-## Which dimension-level is the var function? And can it be used in dispRity?
-make.metric(var)
-```
-
-A non verbose version of the function is also available.
-This can be done using the option `silent = TRUE` and will simply output the dimension-level of the metric.
-
-```{r}
-## Testing whether mean is dimension-level 1
-if(make.metric(mean, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-## Testing whether var is dimension-level 1
-if(make.metric(var, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-```
-
-### Metrics in the `dispRity` function
-
-Using this metric structure, we can easily use any disparity metric in the `dispRity` function as follows:
-
-```{r}
-## Measuring disparity as the standard deviation of all the values of the
-## ordinated matrix (dimension-level 1 function).
-summary(dispRity(BeckLee_mat50, metric = sd))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the ordinated matrix (dimension-level 1 and 2 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the variance covariance matrix (dimension-level 1, 2 and 3 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
-```
-
-Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > 2 > 1).
-
-```{r}
-## Disparity as the standard deviation of the variance of each axis of the
-## variance covariance matrix:
-disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)),
-                      round = 10)
-
-## Same as above but using a different function order for the metric argument
-disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
-                      round = 10)
-
-## Both ways output the same disparity values:
-disparity1 == disparity2
-```
-
-In these examples, we considered disparity to be a single value.
-For example, in the previous example, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
-It is, however, possible to calculate [disparity as a distribution](#disparity-as-a-distribution).
-
-### Metrics implemented in `dispRity`
-
-Several disparity metrics are implemented in the `dispRity` package.
-The detailed list can be found in `?dispRity.metric` along with some description of each metric.
-
-Level | Name | Description | Source | 
-------|------|---------------------------------------------------|--------|
-2 | `ancestral.dist` | The distance between an element and its ancestor | `dispRity` |
-2 | `centroids`<sup>1</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
-1 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
-1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
-1 | `mode.val` | The modal value | `dispRity` |
-1 | `n.ball.volume` | The hyper-spherical (*n*-ball) volume | `dispRity` |
-2 | `pairwise.dist` | The pairwise distances between elements | [`stats`](https://cran.r-project.org/web/packages/stats/index.html)`::dist` |
-2 | `radius` | The radius of each dimensions | `dispRity` |
-2 | `ranges` | The range of each dimension | `dispRity` |
-1 | `span.tree.length` | The minimal spanning tree length | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree` |
-2 | `variances` | The variance of each dimension | `dispRity` |
-
-1: Note that by default, the centroid is the centroid of the elements.
-It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
-
-2: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (*not* PCA).
-
-### Equations and implementations
-Some of the functions described below are implemented in the `dispRity` package and do not require any other packages to calculate ([see implementation here](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R)).
-
-\begin{equation}
-    ancestral.dist = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}
-\end{equation}
-
-\begin{equation}
-    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
-\end{equation}
-
-\begin{equation}
-    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
-\end{equation}
-
-\begin{equation}
-    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
-\end{equation}
-
-\begin{equation}
-    n.ball.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R
-\end{equation}
-
-\begin{equation}
-    radius = |\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|
-\end{equation}
-
-\begin{equation}
-    ranges = |max(k_i) - min(k_i)|
-\end{equation}
-
-\begin{equation}
-    variances = \sigma^{2}{k_i}
-\end{equation}
-
-\begin{equation}
-    span.tree.length = \sum(\mathrm{branch\ length})
-\end{equation}
-
-Where *k* is the number of dimensions,
-*n* the number of elements,
-$\Gamma$ is the Gamma distribution,
-$\lambda_i$ is the eigenvalue of each dimensions,
-$\sigma^{2}$ is their variance and
-$Centroid_{k}$ is their mean,
-$Ancestor_{n}$ is the coordinates of the ancestor of element $n$,
-$f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.),
-*R* is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid).
-
-### Using the different disparity metrics
-Here is a brief demonstration of the main metrics implemented in `dispRity`.
-First, we will create a dummy/simulated ordinated space using the `space.maker` utility function (more about that [here](#space.maker):
-
-```{r}
-## Creating a 10*5 normal space
-set.seed(1)
-dummy_space <- space.maker(10, 5, rnorm)
-```
-
-We will use this simulated space to demonstrate the different metrics.
-
-#### Volumes and surface metrics
-The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `n.ball.volume`  all measure the surface or the volume of the ordinated space occupied:
-
-```{r}
-## Calculating the ellipsoid volume
-summary(dispRity(dummy_space, metric = ellipse.volume))
-```
-> Because there is only one subset (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with rounding).
-
-```{r}
-## Calculating the convex hull surface
-summary(dispRity(dummy_space, metric = convhull.surface))
-
-## Calculating the convex hull volume
-summary(dispRity(dummy_space, metric = convhull.volume))
-
-## Calculating the convex hull volume
-summary(dispRity(dummy_space, metric = n.ball.volume))
-```
-
-The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
-
-> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) that often results in near 0 disparity values.
-
-#### Ranges, variances, radius, pairwise distance, modal value and diagonal
-
-The functions `ranges`, `variances` `radius`, `pairwise.dist`, `mode.val` and `diagonal` all measure properties of the ordinated space based on its dimensional properties (they are also less affected by the "curse of dimensionality"):
-
-`ranges`, `variances` and `radius` work on the same principle and measure the range/variance/radius of each dimension:
-
-```{r}
-## Calculating the ranges of each dimension in the ordinated space
-ranges(dummy_space)
-
-## Calculating disparity as the distribution of these ranges
-summary(dispRity(dummy_space, metric = ranges))
-
-## Calculating disparity as the sum and the product of these ranges
-summary(dispRity(dummy_space, metric = c(sum, ranges)))
-summary(dispRity(dummy_space, metric = c(prod, ranges)))
-
-## Calculating the variances of each dimension in the ordinated space
-variances(dummy_space)
-
-## Calculating disparity as the distribution of these variances
-summary(dispRity(dummy_space, metric = variances))
-
-## Calculating disparity as the sum and the product of these variances
-summary(dispRity(dummy_space, metric = c(sum, variances)))
-summary(dispRity(dummy_space, metric = c(prod, variances)))
-
-## Calculating the radius of each dimension in the ordinated space
-radius(dummy_space)
-
-## By default the radius is the maximum distance from the centre of
-## the dimension. It can however be changed to any function:
-radius(dummy_space, type = min)
-radius(dummy_space, type = mean)
-
-## Calculating disparity as the mean average radius
-summary(dispRity(dummy_space, metric = c(mean, radius), type = mean))
-```
-
-The pairwise distances uses the function `stats::dist` and can take the normal `dist` options:
-
-```{r}
-## The average pairwise euclidean distance
-summary(dispRity(dummy_space, metric = c(mean, pairwise.dist)))
-
-## The distribution of the Manhattan distances
-summary(dispRity(dummy_space, metric = pairwise.dist, method = "manhattan"))
-```
-
-The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest Euclidean distance in our five dimensional space).
-The `mode.val` function measures the modal value of the matrix:
-
-```{r}
-## Calculating the ordinated space's diagonal
-summary(dispRity(dummy_space, metric = diagonal))
-
-## Calculating the modal value of the matrix
-summary(dispRity(dummy_space, metric = mode.val))
-```
-
-> This metric is only a Euclidean diagonal (mathematically valid) if the dimensions within the space are all orthogonal!
-
-#### Centroids and ancestral distance metrics
-
-The `centroids` metric allows users to measure the position of the different elements compared to a fixed point in the ordinated space.
-By default, this function measures the distance between each element and their centroid (centre point):
-
-```{r}
-## The distribution of the distances between each element and their centroid
-summary(dispRity(dummy_space, metric = centroids))
-
-## Disparity as the median value of these distances
-summary(dispRity(dummy_space, metric = c(median, centroids)))
-```
-
-It is however possible to fix the coordinates of the centroid to a specific point in the ordinated space, as long as it has the correct number of dimensions:
-
-```{r}
-## The distance between each element and the origin of the ordinated space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))
-
-## Disparity as the distance between each element and a specific point in space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
-```
-
-The `ancestral.dist` metric works on a similar principle as the `centroids` function but changes the centroid to be the coordinates of each element's ancestor.
-Therefore this functions needs a tree and node coordinates as additional arguments:
-
-```{r}
-## A generating a random tree with node labels
-tree <- rtree(5) ; tree$node.label <- paste0("n", 1:4)
-## Adding the tip and node names to the matrix
-dummy_space2 <- dummy_space[-1,]
-rownames(dummy_space2) <- c(tree$tip.label, tree$node.label)
-
-## Calculating all the ancestral nodes
-all_anc_centroids <- nodes.coordinates(dummy_space2, tree, full = TRUE)
-
-## Calculating the distances from the ancestral nodes
-ancestral_dist <- dispRity(dummy_space2, metric = ancestral.dist,
-                           nodes.coords = all_anc_centroids)
-
-## Calculating disparity as the sum of the distances from all the ancestral nodes
-summary(dispRity(ancestral_dist, metric = sum))
-```
-
-#### Minimal spanning tree length
-
-The `span.tree.length` uses the `vegan::spantree` function to heuristically calculate the minimum spanning tree (the shortest multidimensional tree connecting each elements) and calculates its length as the sum of every branch lengths.
-
-```{r}
-## The length of the minimal spanning tree
-summary(dispRity(dummy_space, metric = span.tree.length))
-```
-
-Note that because the solution is heuristic, this metric can take a long time to compute for big matrices.
-
-## Summarising dispRity data (plots)
-
-Because of its architecture, printing `dispRity` objects only summarises their content but does not print the disparity value measured or associated analysis (more about this [here](#manipulating-dispRity-objects)).
-To actually see what is in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphical representation of the results.
-
-### Summarising `dispRity` data
-This function is an S3 function (`summary.dispRity`) allowing users to summarise the content of `dispRity` objects that contain disparity calculations.
-
-```{r}
-## Example data from previous sections
-crown_stem <- custom.subsets(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
-## Calculate disparity
-disparity_crown_stem <- dispRity(boot_crown_stem, metric = c(sum, variances))
-
-## Creating time slice subsets
-time_slices <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-## Bootstrapping the time slice subsets
-boot_time_slices <- boot.matrix(time_slices, bootstraps = 100)
-## Calculate disparity
-disparity_time_slices <- dispRity(boot_time_slices, metric = c(sum, variances))
-
-## Creating time bin subsets
-time_bins <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages,
-    inc.nodes = TRUE)
-## Bootstrapping the time bin subsets
-boot_time_bins <- boot.matrix(time_bins, bootstraps = 100)
-## Calculate disparity
-disparity_time_bins <- dispRity(boot_time_bins, metric = c(sum, variances))
-```
-
-These objects are easy to summarise as follows:
-
-```{r}
-## Default summary
-summary(disparity_time_slices)
-```
-
-Information about the number of elements in each subset and the observed (i.e. non-bootstrapped) disparity are also calculated.
-This is specifically handy when rarefying the data for example:
-
-```{r}
-head(summary(disparity_crown_stem))
-```
-
-The summary functions can also take various options such as:
-
- * `quantile` values for the confidence interval levels (by default, the 50 and 95 quantiles are calculated)
- * `cent.tend` for the central tendency to use for summarising the results (default is `median`)
- * rounding` option corresponding to the number of decimal places to print (default is `2`)
- * `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
-
-These options can easily be changed from the defaults as follows:
-
-```{r}
-## Same as above but using the 88th quantile and the standard deviation as the summary 
-summary(disparity_time_slices, quantile = 88, cent.tend = sd)
-
-## Printing the details of the object and rounding the values to the 5th decimal place
-summary(disparity_time_slices, recall = TRUE, rounding = 5)
-```
-
-Note that the summary table is a `data.frame`, hence it is as easy to modify as any dataframe using `dplyr`.
-You can also export it in `csv` format using `write.csv` or `write_csv` or even directly export into `LaTeX` format using the following;
-
-```{r, eval=FALSE}
-## Loading the xtable package
-require(xtable)
-## Converting the table in LaTeX
-xtable(summary(disparity_time_slices))
-```
-
-### Plotting `dispRity` data
-An alternative (and more fun!) way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
-This function takes the same options as `summary.dispRity` along with various graphical options described in the function help files (see `?plot.dispRity`).
-
-The plots can be of four different types:
-
- * `continuous` for displaying continuous disparity curves
- * `box`, `lines`, and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.
-
-> This argument can be left empty. In this case, the algorithm will automatically detect the type of subsets from the `dispRity` object and plot accordingly.
-
-It is also possible to display the number of elements in each subset (as a horizontal dotted line) using the option `elements = TRUE`.
-Additionally, when the data is rarefied, one can indicate which level of rarefaction to display (i.e. only display the results for a certain number of elements) by using the `rarefaction` argument.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(mfrow = c(2, 2), bty = "n")
-
-## Plotting continuous disparity results
-plot(disparity_time_slices, type = "continuous")
-
-## Plotting discrete disparity results
-plot(disparity_crown_stem, type = "box")
-
-## As above but using lines for the rarefaction level of 20 elements only
-plot(disparity_crown_stem, type = "line", rarefaction = 20)
-
-## As above but using polygons while also displaying the number of elements
-plot(disparity_crown_stem, type = "polygon", elements = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Since `plot.dispRity` uses the arguments from the generic `plot` method, it is of course possible to change pretty much everything using the regular plot arguments:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some classic options from plot
-plot(disparity_time_slices, col = c("blue", "orange", "green"),
-    ylab = c("Some measurement"), xlab = "Some other measurement",
-    main = "Many options...", ylim = c(5, 0), xlim = c(4, 0))
-
-## Adding a legend
-legend("topleft", legend = c("Central tendency",
-                             "Confidence interval 1",
-                            "Confidence interval 2"),
-      col = c("blue", "orange", "green"), pch = 19)
-
-## Resetting graphical parameters
-par(op)
-```
-
-In addition to the classic `plot` arguments, the function can also take arguments that are specific to `plot.dispRity` like adding the number of elements or rarefaction level (as described above), and also changing the values of the quantiles to plot as well as the central tendency.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some plot.dispRity arguments
-plot(disparity_time_slices, quantile = c(seq(from = 10, to = 100, by = 10)),
-    cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
-    ylab = c("Disparity", "Diversity"), time.subsets = FALSE,
-    xlab = "Time (in in units from past to present)", observed = TRUE,
-    main = "Many more options...")
-
-## Resetting graphical parameters
-par(op)
-```
-
-> Note that the argument `observed = TRUE` allows to plot the disparity values calculated from the non-bootstrapped data as crosses on the plot.
-
-For comparing results, it is also possible to add a plot to the existent plot by using `add = TRUE`:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the continuous disparity with a fixed y axis
-plot(disparity_time_slices, ylim = c(1, 4))
-## Adding the discrete data
-plot(disparity_time_bins, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
-    add = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Finally, if your data has been fully rarefied, it is also possible to easily look at rarefaction curves by using the `rarefaction = TRUE` argument:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the rarefaction curves
-plot(disparity_crown_stem, rarefaction = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-## Testing disparity hypotheses
-
-The `dispRity` package allows users to apply statistical tests to the calculated disparity to test various hypotheses.
-The function `test.dispRity` works in a similar way to the `dispRity` function: it takes a `dispRity` object, a `test` and a `comparisons` argument.
-
-The `comparisons` argument indicates the way the test should be applied to the data:
-
- * `pairwise` (default): to compare each subset in a pairwise manner
- * `referential`: to compare each subset to the first subset
- * `sequential`: to compare each subset to the following subset
- * `all`: to compare all the subsets together (like in analysis of variance)
-
-It is also possible to input a list of pairs of `numeric` values or `characters` matching the subset names to create personalised tests.
-Some other tests implemented in `dispRity` such as the `dispRity::null.test` have a specific way they are applied to the data and therefore ignore the `comparisons` argument. 
-<!-- Add sequential test one day! -->
-
-The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
-It can be a common statistical test function (e.g. `stats::t.test`), a function implemented in `dispRity` (e.g. see `?null.test`) or any function defined by the user.
-
-This function also allows users to correct for Type I error inflation (false positives) when using multiple comparisons via the `correction` argument. 
-This argument can be empty (no correction applied) or can contain one of the corrections from the `stats::p.adjust` function (see `?p.adjust`).
-
-Note that the `test.dispRity` algorithm deals with some classical test outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
-It is, however, possible to get the full detailed output by using the options `details = TRUE`.
-
-Here we are using the variables generated in the [section above](#summarising-dispRity-data-plots):
-
-```{r}
-## T-test to test for a difference in disparity between crown and stem mammals
-test.dispRity(disparity_crown_stem, test = t.test)
-
-## Performing the same test but with the detailed t.test output
-test.dispRity(disparity_crown_stem, test = t.test, details = TRUE)
-
-## Wilcoxon test applied to time sliced disparity with sequential comparisons,
-## with Bonferroni correction
-test.dispRity(disparity_time_slices, test = wilcox.test,
-              comparisons = "sequential", correction = "bonferroni")
-
-## Measuring the overlap between distributions in the time bins (using the
-## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)
-test.dispRity(disparity_time_bins, test = bhatt.coeff)
-```
-
-It is also possible to apply some more *complex* tests that have their own output classes (like `stats::lm`).
-
-The results can then be analysed as usual using the associated `summary` S3 method:
-
-```{r}
-## Performing and linear model applied to the same data
-(slice_lm <- test.dispRity(disparity_time_slices, test = lm,
-                            comparisons = "all"))
-
-## The output is a regular `lm` output
-class(slice_lm)
-
-## This output can be summarised using summary
-summary(slice_lm)
-```
-
-Of course, due to the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
-The only condition is that the test can be applied to at least two distributions.
-In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm` for example).
-
-## Disparity as a distribution
-
-Disparity is often regarded as a summary value of the position of the all elements in the ordinated space.
-For example, the sum of variances, the product of ranges or the median distance between the elements and their centroid will summarise disparity as a single value.
-This value can be pseudo-replicated (bootstrapped) to obtain a distribution of the summary metric with estimated error.
-However, another way to perform disparity analysis is to use the *whole distribution* rather than just a summary metric (e.g. the variances or the ranges).
-
-This is possible in the `dispRity` package by calculating disparity as a dimension-level 2 metric only!
-Let's have a look using our [previous example](#summarising-dispRity-data-plots) of bootstrapped time slices but by measuring the distances between each taxon and their centroid as disparity.
-
-```{r}
-## Measuring disparity as a whole distribution
-disparity_centroids <- dispRity(boot_time_slices, metric = centroids)
-```
-
-The resulting disparity object is of dimension-level 2, so it can easily be transformed into a dimension-level 1 object by, for example, measuring the median distance of all these distributions:
-
-```{r}
-## Measuring median disparity in each time slice
-disparity_centroids_median <- dispRity(disparity_centroids, metric = median)
-```
-
-And we can now compare the differences between these methods:
-
-```{r}
-## Summarising both disparity measurements:
-## The distributions:
-summary(disparity_centroids)
-## The summary of the distributions (as median)
-summary(disparity_centroids_median)
-```
-
-We can see that the summary message for the distribution is slightly different than before.
-Here `summary` also displays the observed central tendency (i.e. the central tendency of the measured distributions).
-Note that, as expected, this central tendency is the same in both metrics!
-
-Another, maybe more intuitive way, to compare both approaches for measuring disparity is to plot the distributions:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(bty = "n", mfrow = c(1, 2))
-
-## Plotting both disparity measurements
-plot(disparity_centroids, ylab = "Distribution of all the distances")
-plot(disparity_centroids_median,
-     ylab = "Distribution of the medians of all the distances")
-
-par(op)
-```
-
-We can then test for differences in the resulting distributions using `test.dispRity` and the `bhatt.coeff` test as described above.
-
-```{r}
-## Probability of overlap in the distribution of medians
-test.dispRity(disparity_centroids_median, test = bhatt.coeff)
-```
-
-In this case, we are looking at the probability of overlap of the distribution of median distances from centroids among each pair of time slices.
-In other words, we are measuring whether the medians from each bootstrap pseudo-replicate for each time slice overlap.
-But of course, we might be interested in the actual distribution of the distances from the centroid rather than simply their central tendencies.
-This can be problematic depending on the research question asked since we are effectively comparing non-independent medians distributions (because of the pseudo-replication).
-
-One solution, therefore, is to look at the full distribution:
-
-```{r}
-## Probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff)
-```
-
-These results show the actual overlap among all the measured distances from centroids concatenated across all the bootstraps.
-For example, when comparing the slices 120 and 80, we are effectively comparing the 5 $\times$ 100 distances (the distances of the five elements in slice 120 bootstrapped 100 times) to the 19 $\times$ 100 distances from slice 80.
-However, this can also be problematic for some specific tests since the _n_ $\times$ 100 distances are also pseudo-replicates and thus are still not independent.
-
-A second solution is to compare the distributions to each other _for each replicate_:
-
-```{r}
-## Boostrapped probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff, concatenate = FALSE)
-```
-
-These results show the median overlap among pairs of distributions in the first column (`bhatt.coeff`) and then the distribution of these overlaps among each pair of bootstraps.
-In other words, when two distributions are compared, they are now compared for each bootstrap pseudo-replicate, thus effectively creating a distribution of probabilities of overlap.
-For example, when comparing the slices 120 and 80, we have a mean probability of overlap of 0.28 and a probability between 0.18 and 0.43 in 50\% of the pseudo-replicates.
-Note that the quantiles and central tendencies can be modified via the `conc.quantiles` option.
-
-## Disparity from other matrices
-
-In the example so far, disparity was measured from an ordinated multidimensional space (i.e. a PCO of the distances between taxa based on discrete morphological characters).
-This is a common approach in palaeobiology, morphometrics or ecology but ordinated matrices are not mandatory for the `dispRity` package!
-It is totally possible to perform the same analysis detailed above using other types of matrices as long as your elements are rows in your matrix.
-
-For example, we can use the data set `eurodist`, an `R` inbuilt dataset that contains the distances (in km) between European cities.
-We can check for example, if Northern European cities are closer to each other than Southern ones:
-
-```{r}
-## Making the eurodist data set into a matrix (rather than "dist" object)
-eurodist <- as.matrix(eurodist)
-eurodist[1:5, 1:5]
-
-## The two groups of cities
-Northern <- c("Brussels", "Calais", "Cherbourg", "Cologne", "Copenhagen",
-              "Hamburg", "Hook of Holland", "Paris", "Stockholm")
-Southern <- c("Athens", "Barcelona", "Geneva", "Gibraltar", "Lisbon", "Lyons",
-              "Madrid", "Marseilles", "Milan", "Munich", "Rome", "Vienna")
-
-## Creating the subset dispRity object
-eurodist_subsets <- custom.subsets(eurodist, group = list("Northern" = Northern,
-                                                        "Southern" = Southern))
-
-## Bootstrapping and rarefying to 9 elements (the number of Northern cities)
-eurodist_bs <- boot.matrix(eurodist_subsets, rarefaction = 9)
-
-## Measuring disparity as the median distance from group's centroid
-euro_disp <- dispRity(eurodist_bs, metric = c(median, centroids))
-
-## Testing the differences using a simple wilcox.test
-euro_diff <- test.dispRity(euro_disp, test = wilcox.test)
-euro_diff_rar <- test.dispRity(euro_disp, test = wilcox.test, rarefaction = 9)
-```
-
-We can compare this approach to an ordination one:
-
-```{r}
-## Ordinating the eurodist matrix
-euro_ord <- cmdscale(eurodist, k = nrow(eurodist) - 2)
-
-## Calculating disparity on the bootstrapped and rarefied subset data
-euro_ord_disp <- dispRity(boot.matrix(custom.subsets(euro_ord, group =
-        list("Northern" = Northern, "Southern" = Southern)), rarefaction = 9),
-        metric = c(median, centroids))
-
-## Testing the differences using a simple wilcox.test
-euro_ord_diff <- test.dispRity(euro_ord_disp, test = wilcox.test)
-euro_ord_diff_rar <- test.dispRity(euro_ord_disp, test = wilcox.test, rarefaction = 9)
-```
-
-And visualise the differences:
-
-```{r, fig.width=12, fig.height=12}
-## Plotting the differences
-par(mfrow = c(2,2), bty = "n")
-## Plotting the normal disparity
-plot(euro_disp, main = "Distance differences")
-## Adding the p-value
-text(1.5, 4000, paste0("p=",round(euro_diff[[2]][[1]], digit = 5)))
-## Plotting the rarefied disparity
-plot(euro_disp, rarefaction = 9, main = "Distance differences (rarefied)")
-## Adding the p-value
-text(1.5, 4000, paste0("p=",round(euro_diff_rar[[2]][[1]], digit = 5)))
-
-## Plotting the ordinated disparity
-plot(euro_ord_disp, main = "Ordinated differences")
-## Adding the p-value
-text(1.5, 1400, paste0("p=",round(euro_ord_diff[[2]][[1]], digit = 5) ))
-## Plotting the rarefied disparity
-plot(euro_ord_disp, rarefaction = 9, main = "Ordinated differences (rarefied)")
-## Adding the p-value
-text(1.5, 1400, paste0("p=",round(euro_ord_diff_rar[[2]][[1]], digit = 5) ))
-
-```
-
-As expected, the results are pretty similar in pattern but different in terms of scale.
-The median centroids distance is expressed in km in the "Distance differences" plots and in Euclidean units of variation in the "Ordinated differences" plots.
-
-<!--chapter:end:03_specific-tutorials.Rmd-->
-
----
-title: "Data simulation features"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Making stuff up!
-
-The `dispRity` package also offers some advanced data simulation features to allow to test hypothesis, explore ordinate-spaces or metrics properties or simply playing around with data!
-All the following functions are based on the same modular architecture of the package and therefore can be used with most of the functions of the package.
-
-## Simulating discrete morphological data
-
-The function `sim.morpho` allows to simulate discrete morphological data matrices (sometimes referred to as "cladistic" matrices).
-It allows to evolve multiple discrete characters on a given phylogenetic trees, given different models, rates, and states.
-It even allows to include proper inapplicable data to make datasets as messy as in real life!
-
-In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
-The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
-The methods are described in details below
-
-```{r}
-set.seed(3)
-## Simulating a starting tree with 15 taxa as a random coalescent tree
-my_tree <- rcoal(15)
-
-## Generating a matrix with 100 characters (85% binary and 15% three state) and
-## an equal rates model with a gamma rate distribution (0.5, 1) with no 
-## invariant characters.
-my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
-    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)
-
-## The first few lines of the matrix
-my_matrix[1:5, 1:10]
-
-## Checking the matrix properties with a quick Maximum Parsimony tree search
-check.morpho(my_matrix, my_tree)
-```
-
-Note that this example produces a tree with a great consistency index and an identical topology to the random coalescent tree!
-Nearly too good to be true...
-
-### A more detailed description
-
-The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;OReilly2017].
-
- * The first `tree` argument will be the tree on which to "evolve" the characters and therefore requires branch length.
- You can generate quick and easy random Yule trees using `ape::rtree(number_of_taxa)` but I would advise to use more realistic trees for more realistic simulations based on more realistic models (really realistic then) using the function `tree.bd` from the [`diversitree`](http://www.zoology.ubc.ca/prog/diversitree/) package [@fitzjohndiversitree2012].
- * The second argument, `character` is the number of characters. Pretty straight forward.
- * The third, `states` is the proportion of characters states above two (yes, the minimum number of states is two).
-This argument intakes the proportion of *n*-states characters, for example `states = c(0.5,0.3,0.2)` will generate 50% of binary-state characters, 30% of three-state characters and 20% of four-state characters.
-There is no limit in the number of state characters proportion as long as the total makes up 100%.
- * The forth, `model` is the evolutionary model for generating the character(s). More about this below.
- * The fifth and sixth, `rates` and `substitution` are the model parameters described below as well.
- * Finally, the two logical arguments, are self explanatory: `invariant` whether to allow invariant characters (i.e.
-characters that don't change) and `verbose` whether to print the simulation progress on your console.
-
-#### Available evolutionary models
-
-There are currently three evolutionary models implemented in `sim.morpho` but more will come in the future.
-Note also that they allow fine tuning parameters making them pretty plastic!
-
- * `"ER"`: this model allows any number of character states and is based on the Mk model [@lewisa2001].
-It assumes a unique overall evolutionary rate equal substitution rate between character states.
-This model is based on the `ape::rTraitDisc` function.
- * `"HKY"`: this is binary state character model based on the molecular HKY model [@HKY85].
-It uses the four molecular states (A,C,G,T) with a unique overall evolutionary rate and a biased substitution rate towards transitions (A <-> G or C <-> T) against transvertions (A <-> C and G <-> T).
-After evolving the nucleotide, this model transforms them into binary states by converting the purines (A and G) into state 0 and the pyrimidines (C and T) into state 1.
-This method is based on the `phyclust::seq.gen.HKY` function and was first proposed by @OReilly2016.
- * `"MIXED"`: this model uses a random (uniform) mix between both the `"ER"` and the `"HKY"` models.
-
-The models can take the following parameters:
-(1) `rates` is the evolutionary rate (i.e. the rate of changes along a branch: the evolutionary speed) and (2) `substitution` is the frequency of changes between one state or another.
-For example if a character can have high probability of changing (the *evolutionary* rate) with, each time a change occurs a probability of changing from state *X* to state *Y* (the *substitution* rate).
-Note that in the `"ER"` model, the substitution rate is ignore because... by definition this (substitution) rate is equal!
-
-The parameters arguments `rates` and `substitution` takes a distributions from which to draw the parameters values for each character.
-For example, if you want an `"HKY"` model with an evolutionary rate (i.e.
-speed) drawn from a uniform distribution bounded between 0.001 and 0.005, you can define it as `rates = c(runif, min = 0.001, max = 0.005)`, `runif` being the function for random draws from a uniform distribution and `max` and `min` being the distribution parameters.
-These distributions should always be passed in the format `c(random_distribution_function, distribution_parameters)` with the names of the distribution parameters arguments.
-
-#### Checking the results
-
-An additional function, `check.morpho` runs a quick Maximum Parsimony tree search using the `phangorn` parsimony algorithm.
-It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g.
-the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree to determine how different they are.
-
-#### Adding inapplicable characters
-
-Once a matrix is generated, it is possible to apply inapplicable characters to it for increasing realism!
-Inapplicable characters are commonly designated as `NA` or simply `-`.
-They differ from missing characters `?` in their nature by being inapplicable rather than unknown.
-For example, considering a binary character defined as "colour of the tail" with the following states "blue" and "red"; on a taxa with no tail, the character should be coded as inapplicable ("`-`") since the state of the character "colour of tail" is *known*: it's neither "blue" or "red", it's just not there!
-It contrasts with coding it as missing ("`?`" - also called as ambiguous) where the state is *unknown*, for example, the taxon of interest is a fossil where the tail has no colour preserved or is not present at all due to bad conservation!
-
-This type of characters can be added to the simulated matrices using the `apply.NA` function/
-It takes, as arguments, the `matrix`, the source of inapplicability (`NAs` - more below), the `tree` used to generate the matrix and the two same `invariant` and `verbose` arguments as defined above.
-The `NAs` argument allows two types of sources of inapplicability:
-
- * `"character"` where the inapplicability is due to the character (e.g. coding a character tail for species with no tail).
- In practice, the algorithm chooses a character *X* as the underlying character (e.g.
-"presence and absence of tail"), arbitrarily chooses one of the states as "absent" (e.g.
-0 = absent) and changes in the next character *Y* any state next to character *X* state 0 into an inapplicable token ("`-`").
- This simulates the inapplicability induced by coding the characters (i.e. not always biological).
- * `"clade"` where the inapplicability is due to evolutionary history (e.g. a clade loosing its tail).
- In practice, the algorithm chooses a random clade in the tree and a random character *Z* and replaces the state of the taxa present in the clade by the inapplicable token ("`-`").
- This simulates the inapplicability induced by evolutionary biology (e.g. the lose of a feature in a clade).
-
-To apply these sources of inapplicability, simply repeat the number of inapplicable sources for the desired number of characters with inapplicable data.
-
-```{r}
-## Generating 5 "character" NAs and 10 "clade" NAs
-my_matrix_NA <- apply.NA(my_matrix, tree = my_tree,
-                         NAs = c(rep("character", 5), rep("clade", 10)))
-
-## The first few lines of the resulting matrix
-my_matrix_NA[1:10, 90:100]
-```
-<!--TG: Add inapplicable paper here -->
-
-### Parameters for a realistic(ish) matrix
-
-There are many parameters that can create a "realistic" matrix (i.e.
-not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all parameters combination end up creating "good" matrices.
-The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
-
-```{r}
-set.seed(0)
-## tree
-my_tree <- rcoal(15)
-## matrix
-morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
-    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
-check.morpho(morpho_mat, my_tree)
-```
-
-<!-- ## Simulating continuous morphological data -->
-<!-- TG: Add tutorial for evolving "morpho" shapes -->
-
-## Simulating multidimensional spaces
-Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
-This function allows users to simulate multidimensional spaces with a certain number of properties.
-For example, it is possible to design a multidimensional space with a specific distribution on each axis, a correlation between the axes and a specific cumulative variance per axis.
-This can be useful for creating ordinated spaces for null hypothesis, for example if you're using the function `null.test` [@diaz2016global].
-
-This function takes as arguments the number of elements (data points - `elements` argument) and dimensions (`dimensions` argument) to create the space and the distribution functions to be used for each axis.
-The distributions are passed through the `distribution` argument as... modular functions!
-You can either pass a single distribution function for all the axes (for example `distribution = runif` for all the axis being uniform) or a specific distribution function for each specific axis (for example `distribution = c(runif, rnorm, rgamma))` for the first axis being uniform, the second normal and the third gamma).
-You can of course use your very own functions or use the ones implemented in `dispRity` for more complex ones (see below).
-Specific optional arguments for each of these distributions can be passed as a list via the `arguments` argument.
-
-Furthermore, it is possible to add a correlation matrix to add a correlation between the axis via the `cor.matrix` argument or even a vector of proportion of variance to be bear by each axis via the `scree` argument to simulate realistic ordinated spaces.
-
-Here is a simple two dimensional example:
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(bty = "n")
-
-## A square space
-square_space <- space.maker(100, 2, runif)
-
-## The resulting 2D matrix
-head(square_space)
-
-## Visualising the space
-plot(square_space, pch = 20, xlab = "", ylab = "", main = "Uniform 2D space")
-```
-
-Of course, more complex spaces can be created by changing the distributions, their arguments or adding a correlation matrix or a cumulative variance vector:
-
-```{r}
-## A plane space: uniform with one dimensions equal to 0
-plane_space <- space.maker(2500, 3, c(runif, runif, runif),
-                           arguments = list(list(min = 0, max = 0), NULL, NULL))
-
-## Correlation matrix for a 3D space
-(cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3))
-
-## An ellipsoid space (normal space with correlation)
-ellipse_space <- space.maker(2500, 3, rnorm, cor.matrix = cor_matrix)
-
-## A cylindrical space with decreasing axes variance
-cylindrical_space <- space.maker(2500, 3, c(rnorm, rnorm, runif),
-                                 scree = c(0.7, 0.2, 0.1))
-```
-
-### Personalised dimensions distributions
-
-Following the modular architecture of the package, it is of course possible to pass home made distribution functions to the `distribution` argument.
-For example, the `random.circle` function is a personalised one implemented in `dispRity`.
-This function allows to create circles based on basic trigonometry allowing to axis to covary to produce circle coordinates.
-By default, this function generates two sets of coordinates with a `distribution` argument and a minimum and maximum boundary (`inner` and `outer` respectively) to create nice sharp edges to the circle.
-The maximum boundary is equivalent to the radius of the circle (it removes coordinates beyond the circle radius) and the minimum is equivalent to the radius of a smaller circle with no data (it removes coordinates below this inner circle radius).
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(bty = "n")
-
-## Generating coordinates for a normal circle with a upper boundary of 1
-circle <- random.circle(1000, rnorm, inner = 0, outer = 1)
-
-## Plotting the circle
-plot(circle, xlab = "x", ylab = "y", main = "A normal circle")
-
-## Creating doughnut space (a spherical space with a hole)
-doughnut_space <- space.maker(5000, 3, c(rnorm, random.circle),
-     arguments = list(list(mean = 0), list(runif, inner = 0.5, outer = 1)))
-```
-
-### Visualising the space
-
-I suggest using the excellent `scatterplot3d` package to play around and visualise the simulated spaces:
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(mfrow = (c(2, 2)), bty = "n")
-## Visualising 3D spaces
-require(scatterplot3d)
-
-## The plane space
-scatterplot3d(plane_space, pch = 20, xlab = "", ylab = "", zlab = "",
-              xlim = c(-0.5, 0.5), main = "Plane space")
-
-## The ellipsoid space
-scatterplot3d(ellipse_space, pch = 20, xlab = "", ylab = "", zlab = "",
-              main = "Normal ellipsoid space")
-
-## A cylindrical space with a decreasing variance per axis
-scatterplot3d(cylindrical_space, pch = 20, xlab = "", ylab = "", zlab = "",
-              main = "Normal cylindrical space")
-## Axes have different orders of magnitude
-
-## Plotting the doughnut space
-scatterplot3d(doughnut_space[,c(2,1,3)], pch = 20, xlab = "", ylab = "",
-              zlab = "", main = "Doughnut space")
-par(op)
-```
-
-### Generating realistic spaces
-
-It is possible to generate "realistic" spaces by simply extracting the parameters of an existing space and scaling it up to the simulated space.
-For example, we can extract the parameters of the `BeckLee_mat50` ordinated space and simulate a similar space.
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Loading the data
-data(BeckLee_mat50)
-
-## Number of dimensions
-obs_dim <- ncol(BeckLee_mat50)
-
-## Observed correlation between the dimensions
-obs_correlations <- cor(BeckLee_mat50)
-
-## Observed mean and standard deviation per axis
-obs_mu_sd_axis <- mapply(function(x,y) list("mean" = x, "sd" = y),
-                         as.list(apply(BeckLee_mat50, 2, mean)),
-                         as.list(apply(BeckLee_mat50, 2, sd)), SIMPLIFY = FALSE)
-
-## Observed overall mean and standard deviation
-obs_mu_sd_glob <- list("mean" = mean(BeckLee_mat50), "sd" = sd(BeckLee_mat50))
-
-## Scaled observed variance per axis (scree plot)
-obs_scree <- variances(BeckLee_mat50)/sum(variances(BeckLee_mat50))
-
-## Generating our simulated space
-simulated_space <- space.maker(1000, dimensions = obs_dim, 
-                               distribution = rep(list(rnorm), obs_dim),
-                               arguments = obs_mu_sd_axis,
-                               cor.matrix = obs_correlations)
-
-## Visualising the fit of our data in the space (in the two first dimensions)
-plot(simulated_space[,1:2], xlab = "PC1", ylab = "PC2")
-points(BeckLee_mat50[,1:2], col = "red", pch = 20)
-legend("topleft", legend = c("observed", "simulated"),
-        pch = c(20,21), col = c("red", "black"))
-```
-
-It is now possible to simulate a space using these observed arguments to test several hypothesis:
-
- * Is the space uniform or normal?
- * If the space is normal, is the mean and variance global or specific for each axis?
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Measuring disparity as the sum of variance
-observed_disp <- dispRity(BeckLee_mat50, metric = c(median, centroids))
-
-## Is the space uniform?
-test_unif <- null.test(observed_disp, null.distrib = runif)
-
-## Is the space normal with a mean of 0 and a sd of 1?
-test_norm1 <- null.test(observed_disp, null.distrib = rnorm)
-
-## Is the space normal with the observed mean and sd and cumulative variance
-test_norm2 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
-                        null.args = rep(list(obs_mu_sd_glob), obs_dim),
-                        null.scree = obs_scree)
-
-## Is the space multiple normal with multiple means and sds and a correlation?
-test_norm3 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
-                        null.args = obs_mu_sd_axis, null.cor = obs_correlations)
-
-## Graphical options
-op <- par(mfrow = (c(2, 2)), bty = "n")
-## Plotting the results
-plot(test_unif, main = "Uniform (0,1)")
-plot(test_norm1, main = "Normal (0,1)")
-plot(test_norm2, main = paste0("Normal (", round(obs_mu_sd_glob[[1]], digit = 3),
-                              ",", round(obs_mu_sd_glob[[2]], digit = 3), ")"))
-plot(test_norm3, main = "Normal (variable + correlation)")
-
-```
-
-If we measure disparity as the median distance from the morphospace centroid, we can explain the distribution of the data as normal with the variable observed mean and standard deviation and with a correlation between the dimensions.
-
-<!--chapter:end:04_making-up-data.Rmd-->
-
----
-title: "The guts of the dispRity package"
-author: "Thomas Guillerme (guillert@tcd.ie)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# The guts of the `dispRity` package
-
-## Manipulating `dispRity` objects
-
-Disparity analysis involves a lot of manipulation of many matrices (especially when bootstrapping) which can be impractical to visualise and will quickly overwhelm your `R` console.
-Even the simple Beck and Lee 2014 example above produces an object with > 72 lines of lists of lists of matrices!
-
-Therefore `dispRity` uses a specific class of object called a `dispRity` object.
-These objects allow users to use S3 method functions such as `summary.dispRity`, `plot.dispRity` and `print.dispRity`. 
-`dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g. `sort.dispRity`, `extract.dispRity` see the full list in the next section).
-These functions modify the `dispRity` object without having to delve into its complex structure!
-The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).
-
-```{r}
-## Loading the example data
-data(disparity)
-
-## What is the class of the median_centroids object?
-class(disparity)
-
-## What does the object contain?
-names(disparity)
-
-## Summarising it using the S3 method print.dispRity
-disparity
-```
-
-Note that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:
-
-```{r, eval=FALSE}
-## Display the full object
-print(disparity, all = TRUE)
-## This is more nearly ~ 5000 lines on my 13 inch laptop screen!
-```
-
-## `dispRity` utilities
-
-The package also provides some utility functions to facilitate multidimensional analysis.
-
-### `dispRity` object utilities  <a name="dispRity.utilities"></a>
-The first set of utilities are functions for manipulating `dispRity` objects:
-
-#### `make.dispRity`
-This function creates empty `dispRity` objects.
-
-```{r}
-## Creating an empty dispRity object
-make.dispRity()
-
-## Creating an "empty" dispRity object with a matrix
-(disparity_obj <- make.dispRity(matrix(rnorm(20), 5, 4)))
-```
-
-####  `fill.dispRity`
-This function initialises a `dispRity` object and generates its call properties.
-
-```{r}
-## The dispRity object's call is indeed empty
-disparity_obj$call
-
-## Filling an empty disparity object (that needs to contain at least a matrix)
-(disparity_obj <- fill.dispRity(disparity_obj))
-
-## The dipRity object has now the correct minimal attributes
-disparity_obj$call
-```
-
-####  `matrix.dispRity`
-This function extracts a specific matrix from a disparity object.
-The matrix can be one of the bootstrapped matrices or/and a rarefied matrix.
-
-```{r}
-## Extracting the matrix containing the coordinates of the elements at time 50
-str(matrix.dispRity(disparity, "50"))
-
-## Extracting the 3rd bootstrapped matrix with the 2nd rarefaction level
-## (15 elements) from the second group (80 Mya)
-str(matrix.dispRity(disparity, subsets = 1, bootstrap = 3, rarefaction = 2))
-```
-
-#### `get.subsets.dispRity`
-This function creates a dispRity object that contains only elements from one specific subsets.
-
-```{r, eval=FALSE}
-## Extracting all the data for the crown mammals
-(crown_mammals <- get.subsets.dispRity(disp_crown_stemBS, "Group.crown"))
-
-## The object keeps the properties of the parent object but is composed of only one subsets
-length(crown_mammals$subsets)
-```
-
-#### `extract.dispRity`
-This function extracts the calculated disparity values of a specific matrix.
-
-```{r, eval=FALSE}
-## Extracting the observed disparity (default)
-extract.dispRity(disparity)
-
-## Extracting the disparity from the bootstrapped values from the
-## 10th rarefaction level from the second subsets (80 Mya)
-extract.dispRity(disparity, observed = FALSE, subsets = 2, rarefaction = 10)
-```
-
-#### `scale.dispRity`
-This is the S3 method of `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object.
-
-```{r, eval=FALSE}
-## Getting the disparity values of the time subsets
-head(summary(disparity))
-
-## Scaling the same disparity values
-head(summary(scale(disparity, scale = TRUE)))
-
-## Scaling and centering:
-head(summary(scale(disparity, scale = TRUE, center = TRUE)))
-```
-
-#### `sort.dispRity`
-This is the S3 method of `sort` for sorting the subsets alphabetically (default) or following a specific pattern.
-
-```{r, eval=FALSE}
-## Sorting the disparity subsets in inverse alphabetic order
-head(summary(sort(disparity, decreasing = TRUE)))
-
-## Customised sorting
-head(summary(sort(disparity, sort = c(7, 1, 3, 4, 5, 2, 6))))
-```
-
-## The `dispRity` object content
-The functions above are utilities to easily and safely access different elements in the `dispRity` object.
-Alternatively, of course, each elements can be accessed manually.
-Here is an explanation on how it works.
-The `dispRity` object is a `list` of two to four elements, each of which are detailed below:
-
- * `$matrix`: an object of class `matrix`, the full multidimensional space.
- * `$call`: an object of class `list` containing information on the `dispRity` object content.
- * `$subsets`: an object of class `list` containing the subsets of the multidimensional space.
- * `$disparity`: an object of class `list` containing the disparity values.
-
-The `dispRity` object is loosely based on `C` structure objects.
-In fact, it is composed of one unique instance of a matrix (the multidimensional space) upon which the metric function is called via "pointers" to only a certain number of elements and/or dimensions of this matrix.
-This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint:  at any time, only one matrix is present in the `R` environment rather than multiple copies of it for each subset.
-
-### `$matrix`
-This is the multidimensional space, stored in the `R` environment as a `matrix` object.
-It requires row names but not column names.
-By default, if the row names are missing, `dispRity` function will arbitrarily generate them in numeric order (i.e. `rownames(matrix) <- 1:nrow(matrix)`).
-This element of the `dispRity` object is never modified.
-
-### `$call`
-This element contains the information on the `dispRity` object content.
-It is a `list` that can contain the following:
-
- * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"punctuated"`, `"gradual"`).
-This element generated only once via `time.subsets()` and `custom.subsets()`.
- * `$call$dimensions`: either a single `numeric` value indicating how many dimensions to use or a vector of `numeric` values indicating which specific dimensions to use.
-This element is by default the number of columns in `$matrix` but can be modified through `boot.matrix()` or `dispRity()`.
- * `$call$bootstrap`: this is a `list` containing three elements:
-    * `[[1]]`: the number of bootstrap replicates (`numeric`)
-    * `[[2]]`: the bootstrap method (`character`)
-    * `[[3]]`: the rarefaction levels (`numeric` vector)
- * `$call$disparity`: this is a `list` containing one element, `$metric`, that is a `list` containing the different functions passed to the `metric` argument in `dispRity`.
-These are `call` elements and get modified each time the `dispRity` function is used (the first element is the first metric(s), the second, the second metric(s), etc.).
-
-### `$subsets`
-This element contain the eventual subsets of the multidimensional space.
-It is a `list` of subset names.
-Each subset name is in turn a `list` of at least one element called `elements` which is in turn a `matrix`.
-This `elements` matrix is the raw (observed) elements in the subsets.
-The `elements` matrix is composed of `numeric` values in one column and _n_ rows (the number of elements in the subset).
-Each of these values are a "pointer" (`C` inspired) to the element of the `$matrix`.
-For example, lets assume a `dispRity` object called `disparity`, composed of at least one subsets called `sub1`:
-
-```
- disparity$subsets$sub1$elements
-      [,1]
- [1,]    5
- [2,]    4
- [3,]    6
- [4,]    7
-```
-
-The values in the matrix "point" to the elements in `$matrix`: here, the multidimensional space with only the 4th, 5th, 6th and 7th elements.
-The following elements in `diparity$subsets$sub1` will correspond to the same "pointers" but drawn from the bootstrap replicates.
-The columns will correspond to different bootstrap replicates.
-For example:
-
-```
- disparity$subsets$sub1[[2]]
-      [,1] [,2] [,3] [,4]
- [1,]   57   43   70    4
- [2,]   43   44    4    4
- [3,]   42   84   44    1
- [4,]   84    7    2   10
-```
-This signifies that we have four bootstrap pseudo-replicates pointing each time to four elements in `$matrix`.
-The next element (`[[3]]`) will be the same for the eventual first rarefaction level (i.e.
-the resulting bootstrap matrix will have _m_ rows where _m_ is the number of elements for this rarefaction level).
-The next element after that (`[[4]]`) will be the same for with an other rarefaction level and so forth...
-
-### `$disparity`
-The `$disparity` element is identical to the `$subsets` element structure (a list of list(s) containing matrices) but the matrices don't contain "pointers" to `$matrix` but the disparity result of the disparity metric applied to the "pointers".
-For example, in our first example (`$elements`) from above, if the disparity metric is of dimensions level 1, we would have:
-
-```
- disparity$disparity$sub1$elements
-      [,1]
- [1,]    1.82
-```
-
-This is the observed disparity (1.82) for the subset called `sub1`.
-If the disparity metric is of dimension level 2 (say the function `range` that outputs two values), we would have:
-
-```
- disparity$disparity$sub1$elements
-      [,1]
- [1,]    0.82
- [2,]    2.82
-```
-
-The following elements in the list follow the same logic as before: rows are disparity values (one row for a dimension level 1 metric, multiple for a dimensions level 2 metric) and columns are the bootstrap replicates (the bootstrap with all elements followed by the eventual rarefaction levels).
-For example for the bootstrap without rarefaction (second element of the list):
-
-```
- disparity$disparity$sub1[[2]]
-         [,1]     [,2]     [,3]     [,4]
-[1,] 1.744668 1.777418 1.781624 1.739679 
-```
-
-<!-- ## Modularity
-
-### Making your own metrics
-
-### Making your own tests
-
-## Where is this going?
- -->
-<!-- ## Running `dispRity` in parallel
-
-The computationally intensive function `dispRity` have a `parallel` option to speed up their calculations.
-
-This option requires the package `snow` and takes arguments that are to be passed to  -->
-
-<!--chapter:end:05_guts.Rmd-->
-
----
-title: "Palaeobiology demo: disparity-through-time and within groups"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Palaeobiology demo: disparity-through-time and within groups
-
-This demo aims to give quick overview of the `dispRity` package (v.0.4) for palaeobiology analyses of disparity, including disparity through time analyses.
-
-This demo showcases a typical disparity-through-time analysis: we are going to test whether the disparity changed through time in a subset of eutherian mammals from the last 100 million years using a dataset from Beck and Lee (2014).
-
-## Before starting
-
-### The morphospace
-
-In this example, we are going to use a subset of the data from @beckancient2014.
-See the [example data](#example-data) description for more details.
-Briefly, this dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
-The ordinated matrix will represent our full morphospace, i.e. all the mammalian morphologies that ever existed through time (for this dataset).
-
-```{r, fig.width=6, fig.height=6}
-## Loading demo and the package data
-library(dispRity)
-
-## Setting the random seed for repeatability
-set.seed(123)
-
-## Loading the ordinated matrix/morphospace:
-data(BeckLee_mat50)
-head(BeckLee_mat50[,1:5])
-
-dim(BeckLee_mat50)
-## The morphospace contains 50 taxa and has 48 dimensions (or axes)
-
-## Showing a list of first and last occurrences data for some fossils
-data(BeckLee_ages)
-head(BeckLee_ages)
-
-## Plotting a phylogeny
-data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.7)
-axisPhylo(root = 140)
-```
-
-> You can have an even nicer looking tree if you use the `strap` package!
-
-```{r, eval = TRUE}
-if(!require(strap)) install.packages("strap")
-strap::geoscalePhylo(BeckLee_tree, cex.tip = 0.7, cex.ts = 0.6)
-```
-
-## A disparity-through-time analysis
-
-### Splitting the morphospace through time
-
-One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time subsets that contain the total number of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
-Basically, the full morphospace represents the total number of morphologies across all time and will be greater than any of the time subsets of the morphospace.
-
-The `dispRity` package provides a `time.subsets` function that allows users to split the morphospace into time slices (using `method = continuous`) or into time bins (using `method = discrete`).
-In this example, we are going to split the morphospace into five equal time bins of 20 million years long from 100 million years ago to the present.
-We will also provide to the function a table containing the first and last occurrences dates for some fossils to take into account that some fossils might occur in several of our different time bins.
-
-```{r}
-## Creating the vector of time bins ages
-(time_bins <- rev(seq(from = 0, to = 100, by = 20)))
-
-## Splitting the morphospace using the time.subsets function
-(binned_morphospace <- time.subsets(data = BeckLee_mat50, tree = BeckLee_tree,
-    method = "discrete", time = time_bins, inc.nodes = FALSE,
-    FADLAD = BeckLee_ages))
-```
-
-The output object is a `dispRity` object (see more about that [here](#The-guts-of-the-dispRity-package).
-In brief, however, `dispRity` objects are lists of different elements (i.e.
-disparity results, morphospace time subsets, morphospace attributes, etc.) that display only a summary of the object when calling the object to avoiding filling the `R` console with superfluous output.
-
-```{r}
-## Printing the class of the object
-class(binned_morphospace)
-
-## Printing the content of the object
-str(binned_morphospace)
-names(binned_morphospace)
-
-## Printing the object as a dispRity class
-binned_morphospace
-```
-
-> These objects will gradually contain more information when completing the following steps in the disparity-through-time analysis.
-
-### Bootstrapping the data
-
-Once we obtain our different time subsets, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
-The bootstrapping allows us to make each subset more robust to outliers and the rarefaction allows us to compare subsets with the same number of taxa to remove sampling biases (i.e.
-more taxa in one subset than the others).
-The `boot.matrix` function bootstraps the `dispRity` object and the `rarefaction` option within performs rarefaction.
-
-```{r}
-## Bootstrapping each time subset 100 times (default)
-(boot_bin_morphospace <- boot.matrix(binned_morphospace))
-
-## Getting the minimum number of rows (i.e. taxa) in the time subsets
-min(size.subsets(boot_bin_morphospace))
-
-## Bootstrapping each time subset 100 times and rarefying them 
-(rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
-    rarefaction = 6))
-```
-
-### Calculating disparity
-
-We can now calculate the disparity within each time subsets along with some confidence intervals generated by the pseudoreplication step above (bootstraps/rarefaction).
-Disparity can be calculated in many ways and this package allows users to come up with their own disparity metrics.
-For more details, please refer to the [`dispRity` metric section](#disparity-metrics).
-
-In this example, we are going to calculate the spread of the data in each time subset by calculating disparity as the sum of the variance of each dimension of the morphospace in each time subset using the `dispRity` function.
-Thus, in this example, disparity is defined by the multi-dimensional variance of each time subset (i.e. the spread of the taxa within the morphospace).
-Note that this metric comes with a caveat (not solved here) since it ignores covariances among the dimensions of the morphospace.
-We use this here because it is a standard metric used in disparity-through-time analysis [@Wills1994].
-
-```{r}
-## Calculating disparity for the bootstrapped data
-(boot_disparity <- dispRity(boot_bin_morphospace, metric = c(sum, variances)))
-
-## Calculating disparity for the rarefied data
-(rare_disparity <- dispRity(rare_bin_morphospace, metric = c(sum, variances)))
-```
-
-The `dispRity` function does not actually display the calculated disparity values but rather only the properties of the disparity object (size, subsets, metric, etc.).
-To display the actual calculated scores, we need to summarise the disparity object using the S3 method `summary` that is applied to a `dispRity` object (see `?summary.dispRity` for more details).
-
-As for any `R` package, you can refer to the help files for each individual function for more details.
-
-```{r}
-## Summarising the disparity results
-summary(boot_disparity)
-summary(rare_disparity)
-```
-> The summary.dispRity function comes with many options on which values to calculate (central tendency and quantiles) and on how many digits to display.
-Refer to the function's manual for more details.
-
-### Plotting the results
-
-It is sometimes easier to visualise the results in a plot than in a table.
-For that we can use the `plot` S3 function to plot the `dispRity` objects (see `?plot.dispRity` for more details).
-
-```{r, fig.width=12, fig.height=6}
-## Graphical options
-quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
-
-## Plotting the bootstrapped and rarefied results
-plot(boot_disparity, type = "continuous", main = "bootstrapped results")
-plot(rare_disparity, type = "continuous", main = "rarefied results")
-```
-
-Testing differences
--------------------
-
-Finally, to draw some valid conclusions from these results, we can apply some statistical tests.
-We can test, for example, if mammalian disparity changed significantly through time over the last 100 million years.
-To do so, we can compare the means of each time-bin in a sequential manner to see whether the disparity in bin *n* is equal to the disparity in bin *n+1*, and whether this is in turn equal to the disparity in bin *n+2*, etc.
-Because our data is temporally autocorrelated (i.e.
-what happens in bin *n+1* depends on what happened in bin *n*) and pseudoreplicated (i.e.
-each bootstrap draw creates non-independent time subsets because they are all based on the same time subsets), we apply a non-parametric mean comparison: the `wilcox.test`.
-Also, we need to apply a p-value correction (e.g. Bonferroni correction) to correct for multiple testing (see `?p.adjust` for more details).
-
-```{r}
-## Testing the differences between bins in the bootstrapped dataset.
-test.dispRity(boot_disparity, test = wilcox.test, comparison = "sequential",
-    correction = "bonferroni")
-
-## Testing the differences between bins in the rarefied dataset.
-test.dispRity(rare_disparity, test = wilcox.test, comparison = "sequential",
-    correction = "bonferroni")
-```
-
-Here our results show significant changes in disparity through time between all time bins (all p-values < 0.05).
-However, when looking at the rarefied results, there is no significant difference between the time bins in the Palaeogene (60-40 to 40-20 Mya), suggesting that the differences detected in the first test might just be due to the differences in number of taxa sampled (13 or 6 taxa) in each time bin.
-
-<!--chapter:end:06_palaeo-demo.Rmd-->
-
----
-title: "dispRity ecology demo"
-author: "Thomas Guillerme"
-date: "`r Sys.Date()`"
-output: rmarkdown::html_vignette
-bibliography: References.bib
-bst: sysbio.bst
-vignette: >
-  %\VignetteIndexEntry{dispRity ecology demo}
-  %\VignetteEngine{knitr::rmarkdown}
-  \usepackage[utf8]{inputenc}
----
-
-# Ecology demo
-
-This is an example of typical disparity analysis that can be performed in ecology.
-
-## Data
-For this example, we will use the famous `iris` inbuilt data set
-
-```{r}
-data(iris)
-```
-
-This data contains petal and sepal length for 150 individual plants sorted into three species.
-
-```{r}
-## Separating the species
-species <- iris[,5]
-## Which species?
-unique(species)
-
-## Separating the petal/sepal length
-measurements <- iris[,1:4]
-head(measurements)
-```
-
-We can then ordinate the data using a PCA (`prcomp` function) thus defining our four dimensional space as the poetically named petal-space.
-
-```{r}
-## Ordinating the data
-ordination <- prcomp(measurements)
-
-## The petal-space
-petal_space <- ordination$x
-
-## Adding the elements names to the petal-space (the individuals IDs)
-rownames(petal_space) <- 1:nrow(petal_space)
-```
-
-## Classic analysis
-A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different species are distributed in the petal-space.
-
-```{r, fig.width=6, fig.height=6}
-## Measuring the variance on each axis
-variances <- apply(petal_space, 2, var)
-variances <- variances/sum(variances)
-
-## Graphical option
-par(bty = "n")
-
-## A classic 2D ordination plot
-plot(petal_space[, 1], petal_space[, 2], col = species,
-    xlab = paste0("PC 1 (", round(variances[1], 2), ")"),
-    ylab = paste0("PC 2 (", round(variances[2], 2), ")"))
-```
-
-This shows the distribution of the different species in the petal-space along the two first axis of variation.
-This is a pretty standard way to visualise the multidimensional space and further analysis might be necessary to test wether the groups are different such as a linear discriminant analysis (LDA).
-However, in this case we are ignoring the two other dimensions of the ordination!
-If we look at the two other axis we see a totally different result:
-
-```{r, fig.width=6, fig.height=6}
-## Plotting the two second axis of the petal-space
-plot(petal_space[, 3], petal_space[, 4], col = species,
-    xlab = paste0("PC 3 (", round(variances[3], 2), ")"),
-    ylab = paste0("PC 4 (", round(variances[4], 2), ")"))
-```
-
-Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e.
-the values on the first dimension do not represent a continuous trait (e.g.
-petal length), instead they just represent the ordinations of correlations between the data and some factors.
-
-Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.
-
-## A multidimensional approach with `dispRity`
-
-The first step is to create different subsets that represent subsets of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
-Each of these subsets will contain only the individuals of a specific species.
-
-```{r}
-## Creating the table that contain the elements and their attributes
-petal_subsets <- custom.subsets(petal_space, group = list(
-                                "setosa" = which(species == "setosa"),
-                                "versicolor" = which(species == "versicolor"),
-                                "virginica" = which(species == "virginica")))
-
-## Visualising the dispRity object content
-petal_subsets
-```
-
-This created a `dispRity` object (more about that [here](#guts)) with three subsets corresponding to each subspecies.
-
-### Bootstrapping the data
-We can the bootstrap the subsets to be able test the robustness of the measured disparity to outliers.
-We can do that using the default options of `boot.matrix` (more about that [here](#bootstraps-and-rarefactions)):
-
-```{r}
-## Bootstrapping the data
-(petal_bootstrapped <- boot.matrix(petal_subsets))
-```
-
-### Calculating disparity
-
-Disparity can be calculated in many ways, therefore the `dispRity` function allows users to define their own measure of disparity.
-For more details on measuring disparity, see the [dispRity metrics section](#disparity-metrics).
-
-In this example, we are going to define disparity as the median distance between the different individuals and the centroid of the ordinated space.
-High values of disparity will indicate a generally high spread of points from this centroid (i.e. on average, the individuals are far apart in the ordinated space).
-We can define the metrics easily in the `dispRity` function by feeding them to the `metric` argument.
-Here we are going to feed the functions `stats::median` and `dispRity::centroids` which calculates distances between elements and their centroid.
-
-```{r}
-## Calculating disparity as the median distance between each elements and
-## the centroid of the petal-space
-(petal_disparity <- dispRity(petal_bootstrapped, metric = c(median, centroids)))
-```
-
-### Summarising the results (plot)
-Similarly to the `custom.subsets` and `boot.matrix` function, `dispRity` displays a `dispRity` object.
-But we are definitely more interested in actually look at the calculated values.
-
-First we can summarise the data in a table by simply using `summary`:
-
-```{r}
-## Displaying the summary of the calculated disparity
-summary(petal_disparity)
-```
-
-We can also plot the results in a similar way:
-
-```{r, fig.width=6, fig.height=6}
-## Graphical options
-par(bty = "n")
-
-## Plotting the disparity in the petal_space
-plot(petal_disparity)
-```
-
-Now contrary to simply plotting the two first axis of the PCA where we saw that the species have a different position in the two first petal-space, we can now also see that they occupy this space clearly differently!
-
-### Testing hypothesis
-
-Finally we can test our hypothesis that we guessed from the disparity plot (that some groups occupy different volume of the petal-space) by using the `test.dispRity` option.
-
-```{r}
-## Fitting a linear model to our data
-disparity_lm <- test.dispRity(petal_disparity, test = lm, comparisons = "all")
-
-## Testing whether there is a difference in disparity between the different species
-summary(aov(disparity_lm))
-
-## Post-hoc testing of the differences between species (corrected for multiple tests)
-test.dispRity(petal_disparity, test = t.test, correction = "bonferroni")
-```
-
-We can now see that there is a significant difference in petal-space occupancy between all species of iris.
-
-#### Setting up a multidimensional null-hypothesis
-
-One other series of test can be done on the shape of the petal-space.
-Using a MCMC permutation test we can simulate a petal-space with specific properties and see if our observed petal-space matches these properties (similarly to @diaz2016global):
-
-```{r, fig.width=6, fig.height=6}
-## Testing against a uniform distribution
-disparity_uniform <- null.test(petal_disparity, replicates = 200,
-    null.distrib = runif, scale = FALSE)
-plot(disparity_uniform)
-```
-
-```{r, fig.width=6, fig.height=6}
-## Testing against a normal distribution
-disparity_normal <- null.test(petal_disparity, replicates = 200,
-    null.distrib = rnorm, scale = TRUE)
-plot(disparity_normal)
-```
-
-In both cases we can see that our petal-space is not entirely normal or uniform.
-This is expected because of the simplicity of these parameters.
-
-<!--chapter:end:07_ecology-demo.Rmd-->
-
 ---
@@
 
-# Future directions
+# Developing `dispRity`
 
-## More tests!
+Everyone is more than welcome to help developing this package, whether it is through developing functions (metrics, tests, plots, etc...), the manual (more examples, typos, etc...) or completely new functionalities!
 
-Some more tests are being developed such as a `sequential.test` to run sequential linear models for testing hypothesis in disparity through time or a `model.test` developed in collaboration with [Mark Puttick](https://puttickbiology.wordpress.com/) to fit modes of evolution to disparity curves.
-Stay tuned!
+## More functions
 
-## More rarefactions!
-
-I am working on including several "clever" bootstrap and rarefaction algorithms for palaeobiology studies based on geological sampling (following:and Butler et al.
-2012 *Evolution* and Kotric and Knoll 2015 *Palaeobiology*).
-
-## Faster disparity calculations
-
-I am slowly implementing parallel disparity calculation as well as C implementations of some disparity metrics to increase significantly improve the speed of the `dispRity` function.
-
-## More modularity
-
-I am equally slowly developing functions to allow more of the options in the package to be modular (in the same way as the `metric` argument in `dispRity`).
-The next arguments to benefit this increased modularity will be `time.subsets`'s `model` argument and `boot.matrix`'s `type` argument.
-
-## More suggestions?
-
-More generally, I welcome any suggestions or collaborations for developing some additional specific ideas.
-Feel free to contact me via email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
+## More manuals
 
@@
 * Wills, M. A., Briggs, D. E., & Fortey, R. A. (1994). Disparity as an evolutionary index: a comparison of Cambrian and Recent arthropods. Paleobiology, 20(2), 93-130. -->
-
-<!--chapter:end:08_future_directions.Rmd-->

### 2019-07-24 inst/gitbook/03_specific-tutorials.Rmd
man pre-update
Updated span.tree.length

@@
 
-## Time slicing
+## Time slicing {#chrono-subsets}
 
@@
 
-## Customised subsets
+## Customised subsets {#custom-subsets}
 
@@
 \begin{equation}
-    span.tree.length = \sum(\mathrm{branch\ length})
+    span.tree.length = \mathrm{branch\ length}
 \end{equation}
@@
 
-#### Centroids, displacements, and ancestral distance metrics
+#### Centroids, displacements, and ancestral distance metrics {#centroids}
 
@@
 ## The length of the minimal spanning tree
-summary(dispRity(dummy_space, metric = span.tree.length))
+summary(dispRity(dummy_space, metric = c(sum, span.tree.length)))
 ```

### 2019-07-24 tests/testthat/test-dispRity.metric.R

@@ -4,8 +4,8 @@ context("dispRity.metric")
 
 test_that("dimension generic", {
     expect_equal(capture_output(dimension.level3.fun()), "No implemented Dimension level 3 functions implemented in dispRity!\nYou can create your own by using: ?make.metric")
-    expect_equal(capture_output(dimension.level2.fun()), "Dimension level 2 functions implemented in dispRity:\n?ancestral.dist\n?centroids\n?displacements\n?neighbours\n?pairwise.dist\n?ranges\n?radius\n?variances")
-    expect_equal(capture_output(dimension.level1.fun()), "Dimension level 1 functions implemented in dispRity:\n?convhull.surface\n?convhull.volume\n?diagonal\n?ellipse.volume\n?mode.val\n?n.ball.volume\n?span.tree.length")
+    expect_equal(capture_output(dimension.level2.fun()), "Dimension level 2 functions implemented in dispRity:\n?ancestral.dist\n?centroids\n?displacements\n?neighbours\n?pairwise.dist\n?ranges\n?radius\n?variances\n?span.tree.length")
+    expect_equal(capture_output(dimension.level1.fun()), "Dimension level 1 functions implemented in dispRity:\n?convhull.surface\n?convhull.volume\n?diagonal\n?ellipse.volume\n?mode.val\n?n.ball.volume")
 })
 
 
@@ -398,13 +398,14 @@ test_that("span.tree.length", {
     set.seed(1)
     matrix <- matrix(rnorm(50), 10, 5)
 
-    test <- round(span.tree.length(matrix), digit = 5)
+    test <- round(span.tree.length(matrix), digit = 1)
+    expect_equal(test, c(1.6, 1.4, 3.1, 1.5, 1.5, 1.3, 1.7, 1.7, 1.6))
 
-    expect_equal(test, 15.40008)
+    test2 <- round(sum(span.tree.length(matrix)), digit = 5)
+    expect_equal(test2, 15.40008)
 
     dist <- as.matrix(dist(matrix))
-
-    expect_equal(test, round(span.tree.length(dist), digit = 5))
+    expect_equal(test, round(span.tree.length(dist), digit = 1))
 
     ## Working for a non-distance matrix
     set.seed(1)
@@ -501,3 +502,42 @@ test_that("neighbours", {
         round(c(5.943374, 4.515470, 4.008678, 5.943374, 5.059321), digits = 5)
         )
 })
+
+test_that("neighbours", {
[…]

### 2019-07-24 inst/gitbook/03_specific-tutorials.Rmd
Updated metrics

@@
 2 | `ranges` | The range of each dimension | `dispRity` |
-1 | `span.tree.length` | The minimal spanning tree length | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree` |
+2 | `span.tree.length` | The minimal spanning tree length | [`ape`](https://cran.r-project.org/web/packages/vegan/index.html) (but see also [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree`) |
 2 | `variances` | The variance of each dimension | `dispRity` |

### 2019-07-24 R/reduce.matrix.R

@@ -27,8 +27,6 @@
 #' ## Reducing the matrix by columns (and being verbose)
 #' reduce.matrix(na_matrix, by.row = FALSE, verbose = TRUE)
 #' 
-#' @seealso
-#' 
 #' @author Thomas Guillerme
 #' @export
 

### 2019-07-25 inst/gitbook/03_specific-tutorials.Rmd
Updated metrics table

@@
 1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
+1 | `func.div` | The functional divergence | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FDiv` but without abundance)|
+1 | `func.eve` | The functional evenness | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FEve` but without abundance)|
 1 | `mode.val` | The modal value | `dispRity` |

### 2019-07-31 inst/gitbook/03_specific-tutorials.Rmd

--- 2019-07-24 inst/gitbook/03_specific-tutorials.Rmd
@@
 1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
+1 | `func.div` | The functional divergence (the ratio of deviation from the centroid) | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FDiv` but without abundance)|
+1 | `func.eve` | The functional evenness (the minimal spanning tree distances evenness) | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FEve` but without abundance)|
 1 | `mode.val` | The modal value | `dispRity` |
@@
 Note that because the solution is heuristic, this metric can take a long time to compute for big matrices.
+
+#### Functional divergence and evenness
+
+The `func.div` and `func.eve` functions are based on the `FD::dpFD` package.
+They are the equivalent to `FD::dpFD(matrix)$FDiv` and `FD::dpFD(matrix)$FEve` but a bit faster (since they don't deal with abundance data).
+They are pretty straightforward to use:
+
+```{r}
+## The ratio of deviation from the centroid 
+summary(dispRity(dummy_space, metric = func.div))
+
+## The minimal spanning tree distances evenness
+summary(dispRity(dummy_space, metric = func.eve))
+
+## The minimal spanning tree manhanttan distances evenness
+summary(dispRity(dummy_space, metric = func.eve, method = "manhattan"))
+```
 

### 2019-08-06 inst/gitbook/05_other-functionalities.Rmd
slide.nodes!

@@
 
+## `slide.nodes`
+
+This function allows to slide nodes along a tree!
+In other words it allows to change the branch length leading to a node without modifying the overall tree shape.
+This can be useful to add some value to 0 branch lengths for example.
+
+The function works by taking a node (or a list of nodes), a tree and a sliding value.
+The node will be moved "up" (towards the tips) for the given sliding value.
+You can move the node "down" (towards the roots) using a negative value.
+
+```{r}
+set.seed(42)
+## Generating simple coalescent tree
+tree <- rcoal(5)
+
+## Sliding node 8 up and down
+tree_slide_up <- slide.nodes(8, tree, slide = 0.075)
+tree_slide_down <- slide.nodes(8, tree, slide = -0.075)
+
+## Display the results
+par(mfrow = c(3,1))
+plot(tree, main = "original tree") ; axisPhylo() ; nodelabels()
+plot(tree_slide_up, main = "slide up!") ; axisPhylo() ; nodelabels()
+plot(tree_slide_down, main = "slide down!") ; axisPhylo() ; nodelabels()
+```
+
 ## `tree.age`

### 2019-08-07 inst/gitbook/05_other-functionalities.Rmd
Added vignette

@@
 
-## `slide.nodes`
+## `slide.nodes` and `remove.zero.brlen`
 
@@
 plot(tree_slide_down, main = "slide down!") ; axisPhylo() ; nodelabels()
+```
+
+The `remove.zero.brlen` is a "clever" wrapping function that uses the `slide.nodes` function to stochastically remove zero branch lengths across a whole tree.
+This function will slide nodes up or down in successive postorder traversals (i.e.
+going down the tree clade by clade) in order to minimise the number of nodes to slide while making sure there are no silly negative branch lengths produced!
+By default it is trying to slide the nodes using 1% of the minimum branch length to avoid changing the topology too much.
+
+```{r}
+set.seed(42)
+## Generating a tree
+tree <- rtree(20)
+
+## Adding some zero branch lengths (5)
+tree$edge.length[sample(1:Nedge(tree), 5)] <- 0
+
+## And now removing these zero branch lengths!
+tree_no_zero <- remove.zero.brlen(tree)
+
+## Exaggerating the removal (to make it visible)
+tree_exaggerated <- remove.zero.brlen(tree, slide = 1)
+
+## Check the differences
+any(tree$edge.length == 0)
+any(tree_no_zero$edge.length == 0)
+any(tree_exaggerated$edge.length == 0)
+
+## Display the results
+par(mfrow = c(3,1))
+plot(tree, main = "with zero edges")
+plot(tree_no_zero, main = "without zero edges!")
+plot(tree_exaggerated, main = "with longer edges")
 ```

### 2019-08-12 inst/gitbook/03_specific-tutorials.Rmd
added plot preview function

@@
 
+Note that all the options above are plotting disparity objects for which a disparity metric *has been calculated*.
+This makes totally sense for `dispRity` objects but sometimes it might be interesting to look at what the trait-space looks like before measuring the disparity.
+This can be done by plotting `dispRity` objects with no calculated disparity!
+
+For example, we might be interested in looking at how the distribution of elements change as a function of the distributions of different sub-settings.
+For example custom subsets *vs.* time subsets:
+
+```{r, fig.width=8, fig.height=4}
+## Making the different subsets
+cust_subsets <- custom.subsets(BeckLee_mat99, crown.stem(BeckLee_tree, inc.nodes = TRUE))
+time_subsets <- chrono.subsets(BeckLee_mat99, tree = BeckLee_tree, method = "discrete", time = 5)
+
+## Note that no disarity has been calculated here:
+is.null(cust_subsets$disparity)
+is.null(time_subsets$disparity)
+
+## But we can still plot both spaces by using the default plot functions
+par(mfrow = c(1,2))
+## Default plotting
+plot(cust_subsets)
+## Plotting with more arguments
+plot(time_subsets, dimensions = c(4,5), main = "Some \"low\" dimensions")
+```
+
+> DISCLAIMER: This functionality can be handy for exploring the data (e.g.
+to visually check whether the subset attribution worked) but it might be misleading on how the data is *actually* distributed in the multidimensional space!
+Groups that don't overlap on two set dimensions can totally overlap in all other dimensions!
+
 ## Testing disparity hypotheses

### 2019-08-15 inst/gitbook/07_ecology-demo.Rmd
updated book

@@
 ## Measuring the variance on each axis
-variances <- apply(petal_space, 2, var)
-variances <- variances/sum(variances)
+axis_variances <- apply(petal_space, 2, var)
+axis_variances <- axis_variances/sum(axis_variances)
 
@@
 plot(petal_space[, 1], petal_space[, 2], col = species,
-    xlab = paste0("PC 1 (", round(variances[1], 2), ")"),
-    ylab = paste0("PC 2 (", round(variances[2], 2), ")"))
+    xlab = paste0("PC 1 (", round(axis_variances[1], 2), ")"),
+    ylab = paste0("PC 2 (", round(axis_variances[2], 2), ")"))
 ```
@@
 plot(petal_space[, 3], petal_space[, 4], col = species,
-    xlab = paste0("PC 3 (", round(variances[3], 2), ")"),
-    ylab = paste0("PC 4 (", round(variances[4], 2), ")"))
+    xlab = paste0("PC 3 (", round(axis_variances[3], 2), ")"),
+    ylab = paste0("PC 4 (", round(axis_variances[4], 2), ")"))
 ```

### 2019-08-15 inst/gitbook/index.Rmd
updated book

--- 2019-03-06 inst/gitbook/index.Rmd
@@
 version_master <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
+version_master <- 1.3
 
@@
 version_release <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
+version_release <- 1.3
 
@@
 version_CRAN <- strsplit(strsplit(version_raw, split = "\n<td>Version:</td>\n<td>")[[1]][2], split = "</td>\n</tr>\n<tr>\n<td>Depends:")[[1]][1]
+version_CRAN <- 1.3
 ```

### 2019-08-15 inst/gitbook/06_guts.Rmd

--- 2017-11-17 inst/gitbook/dispRity_manual.Rmd
@@
---- 
-title: "dispRity manual"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-site: bookdown::bookdown_site
-output: bookdown::gitbook
-documentclass: book
-link-citations: yes
-github-repo: TGuillerme/dispRity
-description: "dispRity R package vignette"
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-# `dispRity`
-
-This is a package for measuring disparity in `R`.
-It allows users to summarise matrices as representations as multidimensional spaces into a single value or distribution describing a specific aspect of this multidimensional space (the disparity).
-Multidimensional spaces can be ordinated matrices from MDS, PCA, PCO, PCoA but the package is *not* restricted to any type of matrices!
-This manual is based on the version `0.4.2`.
-
-## What is `dispRity`?
-This is a modular package for measuring disparity in `R`.
-It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
-Typically, these analysis are used in palaeobiology and evolutionary biology to study the changes in morphology through time.
-However, there are many more applications in ecology, evolution and beyond.
-
-### Modular?
-Because their exist a multitude of ways to measure disparity, each adapted to every specific question, this package uses an easy to modify modular architecture.
-In coding, each module is simply a function or a modification of a function that can be passed to the main functions of the package to tweak it to your proper needs!
-In practice, you will notice throughout this manual that some function can take other functions as arguments: the modular architecture of this package allows you to use any function for these arguments (with some restrictions explained for each specific cases).
-This will allow you to finely tune your multidimensional analysis to the needs of your specific question!
-
-## Installing and running the package
-
-You can install this package easily if you use the latest version of `R` (> 3.4.0) and `devtools`.
-
-```{r, eval=FALSE}
-## Checking if devtools is already installed
-if(!require(devtools)) install.packages("devtools")
-
-## Installing the latest released version directly from GitHub
-install_github("TGuillerme/dispRity", ref = "release")
-
-## Loading the package
-library(dispRity)
-```
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## loading the package and setting up the start seed.
-library(dispRity)
-## Setting a random seed for repeatability
-set.seed(123)
-```
-
-Note this uses the `release` branch (version`0.4`). <!-- @@@ -->
-For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
-`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, `Claddis`, `geomorph` and `RCurl`).
-
-## Why not CRAN?
-
-This package is not available on CRAN.
-This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
-However, the package follows the strict (and useful!) CRAN standards via Travis.
-
-<!-- badges out start -->
-[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
-[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/master/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
-[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
-<!-- badges out end   -->
-
-## Help
-
-If you need help with the package, hopefully the following manual will be useful.
-However, parts of this package are still in development and some other parts are probably not covered.
-Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
-
-## Citations
-
-You can cite both the package or this manual with the following citation:
-
-> Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646
-
-<!-- > Guillerme, T. & Cooper, N. (2016). The dispRity R package manual. Fighsare XXXX -->
-
-Note that this citation is only temporary (but can still be used!).
-A proper description of the package is currently in review in Methods in Ecology and Evolution and should be re-submitted this winter.
-Upon acceptance (hopefully!), a proper DOI will be released for this manual and the package description.
-
-<!--chapter:end:index.Rmd-->
-
----
-title: "Glossary"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Glossary
-
--   **Multidimensional space**. The mathematical multidimensional object that will be analysed with this package.
-    In morphometrics, this is often referred to as the morphospace.
-    However it may also be referred to as the cladisto-space for cladistic data or the eco-space for ecological data etc.
-    In practice, this term designates a matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
-
--   **Elements**. The rows of the multidimensional space matrix. Elements can be taxa, field sites, countries etc.
-
--   **Dimensions**.
-The columns of the multidimensional space matrix.
-The dimensions can be referred to as axes of variation, or principal components, for ordinated spaces obtained from a PCA for example.
-
--   **Subsets**. Subsets of the multidimensional space.
-    A subset (or subsets) contains the same number of dimensions as the space but may contain a smaller subset of elements.
-    For example, if our space is composed of birds and mammals (the elements) and 50 principal components of variation (the dimensions), we can create two subsets containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
-
-<!--chapter:end:01_glossary.Rmd-->
-
----
-title: "Getting started with dispRity"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Getting started with `dispRity`
-
-## What sort of data does `dispRity` work with?
-
-Disparity can be estimated from pretty much any matrix.
-Classically, however, it is measured from ordinated matrices.
-These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).
-
-### Ordination matrices from `Claddis`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
-if(!require(Claddis)) install.packages("Claddis")
-```
-
-`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
-For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:
-
-```{r}
-require(Claddis)
-
-## Ordinating the example data from Claddis
-Claddis.ordination(Michaux1989) 
-```
-
-Note that several options are available, namely which type of distance should be computed.
-See more info in the function manual (`?Claddis.ordination`).
-Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.
-
-### Ordination matrices from `geomorph`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if geomorph is installed BUILD ONLY
-if(!require(geomorph)) install.packages("geomorph")
-```
-
-You can also easily use data from `geomorph` using the `geomorph.ordination` function.
-This function simply takes Procrustes aligned data and performs an ordination:
-
-```{r}
-require(geomorph)
-
-## Loading the plethodon dataset
-data(plethodon)
-
-## Performing a Procrustes transform on the landmarks
-procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)
-
-## Ordinating this data
-geomorph.ordination(procrustes)[1:5,1:5]
-```
-
-Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
-Additionally you can give the function a `geomorph.data.frame` object.
-If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsets)!
-
-```{r}
-## Using a geomorph.data.frame
-geomorph_df <- geomorph.data.frame(procrustes,
-     species = plethodon$species, site = plethodon$site)
-
-## Ordinating this data and making a dispRity object
-geomorph.ordination(geomorph_df)
-```
-
-More about these `dispRity` objects below!
-
-### Other kinds of ordination matrices
-
-If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordination matrices by using the following functions from the `stats` package.
-Here is how to do it for the following types of matrices:
-
- * Multivariate matrices (principal components analysis; PCA)
-
-```{r}
-## A multivariate matrix
-head(USArrests)
-
-## Ordinating the matrix using `prcomp` 
-ordination <- prcomp(USArrests)
-
-## Selecting the ordinated matrix
-ordinated_matrix <- ordination$x
-head(ordinated_matrix)
-```
-  This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, see the `?princomp` function.
-
- * Distance matrices (classical multidimensional scaling; MDS)
-
-```{r}
-## A matrix of distances between cities
-str(eurodist)
-
-## Ordinating the matrix using cmdscale() with k = 5 dimensions 
-ordinated_matrix <- cmdscale(eurodist, k = 5)
-head(ordinated_matrix)
-```
-
-  This results in a ordinated matrix with European cities as elements and five dimensions.
-
-Of course any other method for creating the ordination matrix is totally valid, you can also not use any ordination at all!
-The only requirements for the `dispRity` functions is that the input is a matrix with elements as rows and dimensions as columns.
-
-## Performing a simple dispRity analysis
-
-Two `dispRity` functions allow users to run an analysis pipeline simply by inputting an ordination matrix.
-These functions allow users to either calculate the disparity through time (`dispRity.through.time`) or the disparity of user-defined groups (`dispRity.per.group`).
-
-**IMPORTANT**
-
-Note that `disparity.through.time` and `disparity.per.group` are wrapper functions (i.e.
-they incorporate lots of other functions) that allow users to run a basic disparity-through-time, or disparity among groups, analysis without too much effort.
-As such they use a lot of default options. 
-These are described in the help files for the functions that are used to make the wrapper functions, and not described in the help files for `disparity.through.time` and `disparity.per.group`. 
-These defaults are good enough for **data exploration**, but for a proper analysis you should consider the **best parameters for your question and data**. 
-For example, which metric should you use?
-How many bootstraps do you require?
-What model of evolution is most appropriate if you are time slicing?
-Should you rarefy the data?
-See [`time.subsets`](#time-slicing), [`custom.subsets`](#customised-subsets), [`boot.matrix`](#bootstraps-and-rarefactions) and [`dispRity.metric`](#disparity-metrics) for more details of the defaults used in each of these functions.
-Note that any of these default arguments can be changed within the `disparity.through.time` or `disparity.per.group` functions.
-
-### Example data
-
-To illustrate these functions, we will use data from @beckancient2014.
-This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
-
-```{r, fig.width=7, fig.height=7}
-## Loading the ordinated matrices
-data(BeckLee_mat50)
-data(BeckLee_mat99)
-
-## The first five taxa and dimensions of the 50 taxa matrix
-head(BeckLee_mat50[, 1:5])
-
-## The first five taxa and dimensions of the 99 taxa + ancestors matrix
-BeckLee_mat99[c(1, 2, 98, 99), 1:5]
-
-## Loading a list of first and last occurrence dates for the fossils
-data(BeckLee_ages)
-head(BeckLee_ages)
-
-## Loading and plotting the phylogeny
-data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.8) 
-axisPhylo(root = 140)
-nodelabels(cex = 0.5)
-```
-
-Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).
-
-### Disparity through time
-
-The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
-This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
-Of course all the defaults can be changed if required, more on this later.
-
-For a disparity through time analysis, you will need:
-
-  * An ordinated matrix (we covered that above)
-  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element.
-To give your tree a root time (i.e.
-an age for the root), you can simply do\\ `my_tree$root.time <- my_age`.
-  * The required number of time subsets (here `time = 3`)
-  * Your favourite disparity metric (here the sum of variances)
-
-Using the Beck and Lee (2014) data described [above](#example-data):
-
-```{r}
-## Measuring disparity through time
-disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
-                                        time = 3, metric = c(sum, variances))
-```
-
-This generates a `dispRity` object (see [here](#guts) for technical details).
-When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:
-
-```{r}
-## Print the disparity_data object
-disparity_data
-```
-
-We asked for three subsets (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
-
-We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):
-
-```{r, fig.width=7, fig.height=7}
-## Summarising disparity through time
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data, type = "continuous")
-
-## Testing for an difference among the time bins
-disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
-summary(disp_lm)
-```
-
-Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
-You can also directly explore the specific function help files within R and navigate to related functions.
-
-### Disparity among groups
-
-The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
-For example, you could ask if there is a difference in disparity between two groups?
-
-To perform such an analysis, you will need:
-
- * An matrix with rows as elements and columns as dimensions (always!)
- * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix.
-For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4.
-Note that elements can be present in multiple groups at once.
- * Your favourite disparity metric (here the sum of variances)
-
-Using the @beckancient2014 data described [above](#example-data):
-
-```{r}
-## Creating the two groups (crown versus stem) as a list
-mammal_groups <- crown.stem(BeckLee_tree, inc.nodes = TRUE)
-
-## Measuring disparity for each group
-disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
-                                     metric = c(sum, variances))
-```
-
-We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).
-
-```{r, fig.width=7, fig.height=7}
-## Print the disparity_data object
-disparity_data
-
-## Summarising disparity in the different groups
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data)
-
-## Testing for a difference between the groups
-test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
-```
-
-<!--chapter:end:02_getting-started.Rmd-->
-
----
-title: "Details of specific functions"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Details of specific functions
-
-The following section contains information specific to some functions.
-If any of your questions are not covered in these sections, please refer to the function help files in R, send me an email (<guillert@tcd.ie>), or raise an issue on GitHub.
-The several tutorials below describe specific functionalities of certain functions; please always refer to the function help files for the full function documentation!
-
-Before each section, make sure you loaded the @beckancient2014 data (see [example data](#example-data) for more details).
-
-```{r}
-## Loading the data
-data(BeckLee_mat50) ; data(BeckLee_mat99)
-data(BeckLee_tree) ; data(BeckLee_ages)
-```
-
-## Time slicing
-
-The function `time.subsets` allows users to divide the matrix into different time subsets or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
-Each subset generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.
-
-Two types of time subsets can be performed by using the `method` option:
-
- *  Discrete time subsets (or time-binning) using `method = discrete`
- *  Continuous time subsets (or time-slicing) using `method = continuous`
-
-For the time-slicing method details see Cooper and Guillerme (in prep.). <!-- @@@ Change cite appropriately! -->
-<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
-For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:
-
- *  Defining the boundaries of the time bins (when `method = discrete`)
- *  Defining the time slices (when `method = continuous`)
-
-Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
-Additionally, it is also possible to input a dataframe containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age, so taxa can appear in more than one time bin/slice.
-
-Here is an example for `method = discrete`:
-
-```{r, eval=TRUE}
-## Generating three time bins containing the taxa present every 40 Ma
-time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0))
-```
-
-Note that we can also generate equivalent results by just telling the function that we want three time-bins as follow:
-
-```{r, eval=TRUE}
-## Automatically generate three equal length bins:
-time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = 3)
-```
-
-In this example, the taxa were split inside each time-bin according to their age.
-However, the taxa here are considered as single points in time.
-It is totally possible that some taxa could have had longer longevity and that they exist in multiple time bins.
-In this case, it is possible to include them in more than one bin by providing a table of first and last occurrence dates (FAD/LAD).
-This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:
-
-```{r, eval=TRUE}
-## Displaying the table of first and last occurrence dates for each taxa
-head(BeckLee_ages)
-
-## Generating time bins including taxa that might span between them
-time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
-```
-
-When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as the root age plus 1\% of the tree length, as long as at least three elements/taxa are present at that point in time.
-The algorithm adds an extra 1\% tree length until reaching the required minimum of three elements.
-It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance among tips *and* nodes.
-
-For the time-slicing method (`method = continuous`), the idea is fairly similar.
-This option, however, requires a matrix that contains the ordinated distance among taxa *and* nodes and an extra argument describing the assumed evolutionary model (via the `model` argument).
-This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node, meaning that a decision must be made about what the value for the branch should be.
-The model can be one of the following:
-
- *  `acctran` where the data chosen along the branch is always the one of the descendant
- *  `deltran` where the data chosen along the branch is always the one of the ancestor
- *  `random` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
- *  `proximity` where the data chosen along the branch is either the descendant or the ancestor depending on branch length
- *  `punctuated` where the data chosen along the branch is both the descendant and the ancestor with an even probability
- *  `gradual` where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length
-
-> Note that the four first models are "fixed": the selected data is always either the one of the descendant or the ancestor.
-The two last models are "probabilistic": the data from both the descendant and the ancestor is used with an associate probability.
-These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants.
-
-```{r, eval=TRUE}
-## Generating four time slices every 40 million years under a model of proximity evolution
-time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-
-## Generating four time slices automatically
-time.subsets(data = BeckLee_mat99, tree = BeckLee_tree,
-    method = "continuous", model = "proximity", time = 4, FADLAD = BeckLee_ages)
-```
-
-## Customised subsets
-
-Another way of separating elements into different categories is to use customised subsets as briefly explained [above](#disparity-among-groups).
-This function simply takes the list of elements to put in each group (whether they are the actual element names or their position in the matrix).
-
-```{r, eval=TRUE}
-## Creating the two groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44))
-
-## Separating the dataset into two different groups
-custom.subsets(BeckLee_mat50, group = mammal_groups)
-```
-
-Elements can easily be assigned to different groups if necessary!
-
-```{r, eval=FALSE}
-## Creating the three groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44).
-                      "all" = c(1:50))
-```
-
-## Bootstraps and rarefactions
-
-One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results are, and how sensitive they are to outliers in the dataset.
-This can be achieved using the function `boot.matrix` to bootstrap and/or rarefy the data.
-The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):
-
-```{r, eval=TRUE}
-## Default bootstrapping
-boot.matrix(data = BeckLee_mat50)
-```
-
-The number of bootstrap replicates can be defined using the `bootstraps` option.
-The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
-Currently two algorithms are implemented:
-
- * `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any *m* elements drawn from the data)
- * `single` where only one random element is replaced by one other random element for each pseudo-replicate
-
-```{r, eval=TRUE}
-## Bootstrapping with the single bootstrap method
-boot.matrix(BeckLee_mat50, boot.type = "single")
-```
-
-This function also allows users to rarefy the data using the `rarefaction` argument.
-Rarefaction allows users to limit the number of elements to be drawn at each bootstrap replication.
-This is useful if, for example, one is interested in looking at the effect of reducing the number of elements on the results of an analysis.
-
-This can be achieved by using the `rarefaction` option that draws only *n-x* at each bootstrap replicate (where *x* is the number of elements not sampled).
-The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subset to a minimum of three elements).
-It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
-
-```{r, eval=TRUE}
-## Bootstrapping with the full rarefaction
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
-
-## Or with a set number of rarefaction levels
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8, 3))
-```
-
-One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
-When missing, all dimensions from the ordinated matrix are used.
-
-```{r, eval=TRUE}
-## Using the first 50% of the dimensions
-boot.matrix(BeckLee_mat50, dimensions = 0.5)
-
-## Using the first 10 dimensions
-boot.matrix(BeckLee_mat50, dimensions = 10)
-```
-
-Of course, one could directly supply the subsets generated above (using `time.subsets` or `custom.subsets`) to this function.
-
-```{r, eval=TRUE}
-## Creating subsets of crown and stem mammals
-crown_stem <- custom.subsets(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot.matrix(crown_stem, bootstraps = 200, rarefaction = TRUE)
-
-## Creating time slice subsets
-time_slices <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-                               method = "continuous", model = "proximity", 
-                               time = c(120, 80, 40, 0),
-                               FADLAD = BeckLee_ages)
-
-## Bootstrapping the time slice subsets
-boot.matrix(time_slices, bootstraps = 100)
-```
-
-## Disparity metrics
-
-There are many ways of measuring disparity!
-In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
-For example, one can look at ellipsoid hyper-volume of the ordinated space (Donohue _et al._ 2013), the sum and the product of the ranges and variances (Wills _et al._ 1994) or the median position of the elements relative to their centroid (Wills _et al._ 1994).
-Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.
-
-Because of this great diversity of metrics, the package `dispRity` does not have one way to measure disparity but rather proposes to facilitate users in defining their own disparity metric that will best suit their particular analysis.
-In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
-However the `metric` argument has to follow certain rules:
-
-1.  It must be composed from one to three `function` objects;
-2.  The function(s) must take as a first argument a `matrix` or a `vector`;
-3.  The function(s) must be of one of the three dimension-levels described below;
-4.  At least one of the functions must be of dimension-level 1 or 2 (see below).
-
-### The function dimension-levels
-
-The metric function dimension-levels determine the "dimensionality of decomposition" of the input matrix.
-In other words, each dimension-level designates the dimensions of the output, i.e. either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimension.
-
-![Illustration of the different dimension-levels of functions with an input `matrix`](dispRity_fun.png)
-
-#### Dimension-level 1 functions
-
-A dimension-level 1 function will decompose a `matrix` or a `vector` into a single value:
-
-```{r}
-## Creating a dummy matrix
-dummy_matrix <- matrix(rnorm(12), 4, 3)
-
-## Example of dimension-level 1 functions
-mean(dummy_matrix)
-median(dummy_matrix)
-```
-
-Any summary metric such as mean or median are good examples of dimension-level 1 functions as they reduce the matrix to a single dimension (i.e. one value).
-
-#### Dimension-level 2 functions
-
-A dimension-level 2 function will decompose a `matrix` into a `vector`.
-
-```{r}
-## Defining the function as the product of rows
-prod.rows <- function(matrix) apply(matrix, 1, prod)
-
-## A dimension-level 2 metric
-prod.rows(dummy_matrix)
-```
-
-Several dimension-level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculate the variance or the range of each dimension of the ordinated matrix respectively.
-
-#### Dimension-level 3 functions
-
-Finally a dimension-level 3 function will transform the matrix into another matrix.
-Note that the dimension of the output matrix doesn't need to match the the input matrix:
-
-```{r}
-## A dimension-level 3 metric
-var(dummy_matrix)
-
-## A dimension-level 3 metric with a forced matrix output
-as.matrix(dist(dummy_matrix))
-```
-
-### `make.metric`
-
-Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the Euclidean distance between each element and the centroid of the ordinated space.
-The `make.metric` function implemented in `dispRity` is designed to help test and find the dimension-level of the functions.
-This function tests:
-
-1.  If your function can deal with a `matrix` or a `vector` as an input;
-2.  Your function's dimension-level according to its output (dimension-level 1, 2 or 3, see above);
-3.  Whether the function can be implemented in the `dispRity` function (the function is fed into a `lapply` loop).
-
-For example, let's see if the functions described above are the right dimension-levels:
-
-```{r}
-## Which dimension-level is the mean function? And can it be used in dispRity?
-make.metric(mean)
-
-## Which dimension-level is the prod.rows function? And can it be used in dispRity?
-make.metric(prod.rows)
-
-## Which dimension-level is the var function? And can it be used in dispRity?
-make.metric(var)
-```
-
-A non verbose version of the function is also available.
-This can be done using the option `silent = TRUE` and will simply output the dimension-level of the metric.
-
-```{r}
-## Testing whether mean is dimension-level 1
-if(make.metric(mean, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-## Testing whether var is dimension-level 1
-if(make.metric(var, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-```
-
-### Metrics in the `dispRity` function
-
-Using this metric structure, we can easily use any disparity metric in the `dispRity` function as follows:
-
-```{r}
-## Measuring disparity as the standard deviation of all the values of the
-## ordinated matrix (dimension-level 1 function).
-summary(dispRity(BeckLee_mat50, metric = sd))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the ordinated matrix (dimension-level 1 and 2 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the variance covariance matrix (dimension-level 1, 2 and 3 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
-```
-
-Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > 2 > 1).
-
-```{r}
-## Disparity as the standard deviation of the variance of each axis of the
-## variance covariance matrix:
-disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)),
-                      round = 10)
-
-## Same as above but using a different function order for the metric argument
-disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
-                      round = 10)
-
-## Both ways output the same disparity values:
-disparity1 == disparity2
-```
-
-In these examples, we considered disparity to be a single value.
-For example, in the previous example, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
-It is, however, possible to calculate [disparity as a distribution](#disparity-as-a-distribution).
-
-### Metrics implemented in `dispRity`
-
-Several disparity metrics are implemented in the `dispRity` package.
-The detailed list can be found in `?dispRity.metric` along with some description of each metric.
-
-Level | Name | Description | Source | 
-------|------|---------------------------------------------------|--------|
-2 | `ancestral.dist` | The distance between an element and its ancestor | `dispRity` |
-2 | `centroids`<sup>1</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
-1 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
-1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
-1 | `mode.val` | The modal value | `dispRity` |
-1 | `n.ball.volume` | The hyper-spherical (*n*-ball) volume | `dispRity` |
-2 | `pairwise.dist` | The pairwise distances between elements | [`stats`](https://cran.r-project.org/web/packages/stats/index.html)`::dist` |
-2 | `radius` | The radius of each dimensions | `dispRity` |
-2 | `ranges` | The range of each dimension | `dispRity` |
-1 | `span.tree.length` | The minimal spanning tree length | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree` |
-2 | `variances` | The variance of each dimension | `dispRity` |
-
-1: Note that by default, the centroid is the centroid of the elements.
-It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
-
-2: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (*not* PCA).
-
-### Equations and implementations
-Some of the functions described below are implemented in the `dispRity` package and do not require any other packages to calculate ([see implementation here](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R)).
-
-\begin{equation}
-    ancestral.dist = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}
-\end{equation}
-
-\begin{equation}
-    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
-\end{equation}
-
-\begin{equation}
-    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
-\end{equation}
-
-\begin{equation}
-    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
-\end{equation}
-
-\begin{equation}
-    n.ball.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R
-\end{equation}
-
-\begin{equation}
-    radius = |\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|
-\end{equation}
-
-\begin{equation}
-    ranges = |max(k_i) - min(k_i)|
-\end{equation}
-
-\begin{equation}
-    variances = \sigma^{2}{k_i}
-\end{equation}
-
-\begin{equation}
-    span.tree.length = \sum(\mathrm{branch\ length})
-\end{equation}
-
-Where *k* is the number of dimensions,
-*n* the number of elements,
-$\Gamma$ is the Gamma distribution,
-$\lambda_i$ is the eigenvalue of each dimensions,
-$\sigma^{2}$ is their variance and
-$Centroid_{k}$ is their mean,
-$Ancestor_{n}$ is the coordinates of the ancestor of element $n$,
-$f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.),
-*R* is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid).
-
-### Using the different disparity metrics
-Here is a brief demonstration of the main metrics implemented in `dispRity`.
-First, we will create a dummy/simulated ordinated space using the `space.maker` utility function (more about that [here](#space.maker):
-
-```{r}
-## Creating a 10*5 normal space
-set.seed(1)
-dummy_space <- space.maker(10, 5, rnorm)
-```
-
-We will use this simulated space to demonstrate the different metrics.
-
-#### Volumes and surface metrics
-The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `n.ball.volume`  all measure the surface or the volume of the ordinated space occupied:
-
-```{r}
-## Calculating the ellipsoid volume
-summary(dispRity(dummy_space, metric = ellipse.volume))
-```
-> Because there is only one subset (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with rounding).
-
-```{r}
-## Calculating the convex hull surface
-summary(dispRity(dummy_space, metric = convhull.surface))
-
-## Calculating the convex hull volume
-summary(dispRity(dummy_space, metric = convhull.volume))
-
-## Calculating the convex hull volume
-summary(dispRity(dummy_space, metric = n.ball.volume))
-```
-
-The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
-
-> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) that often results in near 0 disparity values.
-
-#### Ranges, variances, radius, pairwise distance, modal value and diagonal
-
-The functions `ranges`, `variances` `radius`, `pairwise.dist`, `mode.val` and `diagonal` all measure properties of the ordinated space based on its dimensional properties (they are also less affected by the "curse of dimensionality"):
-
-`ranges`, `variances` and `radius` work on the same principle and measure the range/variance/radius of each dimension:
-
-```{r}
-## Calculating the ranges of each dimension in the ordinated space
-ranges(dummy_space)
-
-## Calculating disparity as the distribution of these ranges
-summary(dispRity(dummy_space, metric = ranges))
-
-## Calculating disparity as the sum and the product of these ranges
-summary(dispRity(dummy_space, metric = c(sum, ranges)))
-summary(dispRity(dummy_space, metric = c(prod, ranges)))
-
-## Calculating the variances of each dimension in the ordinated space
-variances(dummy_space)
-
-## Calculating disparity as the distribution of these variances
-summary(dispRity(dummy_space, metric = variances))
-
-## Calculating disparity as the sum and the product of these variances
-summary(dispRity(dummy_space, metric = c(sum, variances)))
-summary(dispRity(dummy_space, metric = c(prod, variances)))
-
-## Calculating the radius of each dimension in the ordinated space
-radius(dummy_space)
-
-## By default the radius is the maximum distance from the centre of
-## the dimension. It can however be changed to any function:
-radius(dummy_space, type = min)
-radius(dummy_space, type = mean)
-
-## Calculating disparity as the mean average radius
-summary(dispRity(dummy_space, metric = c(mean, radius), type = mean))
-```
-
-The pairwise distances uses the function `stats::dist` and can take the normal `dist` options:
-
-```{r}
-## The average pairwise euclidean distance
-summary(dispRity(dummy_space, metric = c(mean, pairwise.dist)))
-
-## The distribution of the Manhattan distances
-summary(dispRity(dummy_space, metric = pairwise.dist, method = "manhattan"))
-```
-
-The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest Euclidean distance in our five dimensional space).
-The `mode.val` function measures the modal value of the matrix:
-
-```{r}
-## Calculating the ordinated space's diagonal
-summary(dispRity(dummy_space, metric = diagonal))
-
-## Calculating the modal value of the matrix
-summary(dispRity(dummy_space, metric = mode.val))
-```
-
-> This metric is only a Euclidean diagonal (mathematically valid) if the dimensions within the space are all orthogonal!
-
-#### Centroids and ancestral distance metrics
-
-The `centroids` metric allows users to measure the position of the different elements compared to a fixed point in the ordinated space.
-By default, this function measures the distance between each element and their centroid (centre point):
-
-```{r}
-## The distribution of the distances between each element and their centroid
-summary(dispRity(dummy_space, metric = centroids))
-
-## Disparity as the median value of these distances
-summary(dispRity(dummy_space, metric = c(median, centroids)))
-```
-
-It is however possible to fix the coordinates of the centroid to a specific point in the ordinated space, as long as it has the correct number of dimensions:
-
-```{r}
-## The distance between each element and the origin of the ordinated space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))
-
-## Disparity as the distance between each element and a specific point in space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
-```
-
-The `ancestral.dist` metric works on a similar principle as the `centroids` function but changes the centroid to be the coordinates of each element's ancestor.
-Therefore this functions needs a tree and node coordinates as additional arguments:
-
-```{r}
-## A generating a random tree with node labels
-tree <- rtree(5) ; tree$node.label <- paste0("n", 1:4)
-## Adding the tip and node names to the matrix
-dummy_space2 <- dummy_space[-1,]
-rownames(dummy_space2) <- c(tree$tip.label, tree$node.label)
-
-## Calculating all the ancestral nodes
-all_anc_centroids <- nodes.coordinates(dummy_space2, tree, full = TRUE)
-
-## Calculating the distances from the ancestral nodes
-ancestral_dist <- dispRity(dummy_space2, metric = ancestral.dist,
-                           nodes.coords = all_anc_centroids)
-
-## Calculating disparity as the sum of the distances from all the ancestral nodes
-summary(dispRity(ancestral_dist, metric = sum))
-```
-
-#### Minimal spanning tree length
-
-The `span.tree.length` uses the `vegan::spantree` function to heuristically calculate the minimum spanning tree (the shortest multidimensional tree connecting each elements) and calculates its length as the sum of every branch lengths.
-
-```{r}
-## The length of the minimal spanning tree
-summary(dispRity(dummy_space, metric = span.tree.length))
-```
-
-Note that because the solution is heuristic, this metric can take a long time to compute for big matrices.
-
-## Summarising dispRity data (plots)
-
-Because of its architecture, printing `dispRity` objects only summarises their content but does not print the disparity value measured or associated analysis (more about this [here](#manipulating-dispRity-objects)).
-To actually see what is in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphical representation of the results.
-
-### Summarising `dispRity` data
-This function is an S3 function (`summary.dispRity`) allowing users to summarise the content of `dispRity` objects that contain disparity calculations.
-
-```{r}
-## Example data from previous sections
-crown_stem <- custom.subsets(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
-## Calculate disparity
-disparity_crown_stem <- dispRity(boot_crown_stem, metric = c(sum, variances))
-
-## Creating time slice subsets
-time_slices <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-## Bootstrapping the time slice subsets
-boot_time_slices <- boot.matrix(time_slices, bootstraps = 100)
-## Calculate disparity
-disparity_time_slices <- dispRity(boot_time_slices, metric = c(sum, variances))
-
-## Creating time bin subsets
-time_bins <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages,
-    inc.nodes = TRUE)
-## Bootstrapping the time bin subsets
-boot_time_bins <- boot.matrix(time_bins, bootstraps = 100)
-## Calculate disparity
-disparity_time_bins <- dispRity(boot_time_bins, metric = c(sum, variances))
-```
-
-These objects are easy to summarise as follows:
-
-```{r}
-## Default summary
-summary(disparity_time_slices)
-```
-
-Information about the number of elements in each subset and the observed (i.e. non-bootstrapped) disparity are also calculated.
-This is specifically handy when rarefying the data for example:
-
-```{r}
-head(summary(disparity_crown_stem))
-```
-
-The summary functions can also take various options such as:
-
- * `quantile` values for the confidence interval levels (by default, the 50 and 95 quantiles are calculated)
- * `cent.tend` for the central tendency to use for summarising the results (default is `median`)
- * rounding` option corresponding to the number of decimal places to print (default is `2`)
- * `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
-
-These options can easily be changed from the defaults as follows:
-
-```{r}
-## Same as above but using the 88th quantile and the standard deviation as the summary 
-summary(disparity_time_slices, quantile = 88, cent.tend = sd)
-
-## Printing the details of the object and rounding the values to the 5th decimal place
-summary(disparity_time_slices, recall = TRUE, rounding = 5)
-```
-
-Note that the summary table is a `data.frame`, hence it is as easy to modify as any dataframe using `dplyr`.
-You can also export it in `csv` format using `write.csv` or `write_csv` or even directly export into `LaTeX` format using the following;
-
-```{r, eval=FALSE}
-## Loading the xtable package
-require(xtable)
-## Converting the table in LaTeX
-xtable(summary(disparity_time_slices))
-```
-
-### Plotting `dispRity` data
-An alternative (and more fun!) way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
-This function takes the same options as `summary.dispRity` along with various graphical options described in the function help files (see `?plot.dispRity`).
-
-The plots can be of four different types:
-
- * `continuous` for displaying continuous disparity curves
- * `box`, `lines`, and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.
-
-> This argument can be left empty. In this case, the algorithm will automatically detect the type of subsets from the `dispRity` object and plot accordingly.
-
-It is also possible to display the number of elements in each subset (as a horizontal dotted line) using the option `elements = TRUE`.
-Additionally, when the data is rarefied, one can indicate which level of rarefaction to display (i.e. only display the results for a certain number of elements) by using the `rarefaction` argument.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(mfrow = c(2, 2), bty = "n")
-
-## Plotting continuous disparity results
-plot(disparity_time_slices, type = "continuous")
-
-## Plotting discrete disparity results
-plot(disparity_crown_stem, type = "box")
-
-## As above but using lines for the rarefaction level of 20 elements only
-plot(disparity_crown_stem, type = "line", rarefaction = 20)
-
-## As above but using polygons while also displaying the number of elements
-plot(disparity_crown_stem, type = "polygon", elements = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Since `plot.dispRity` uses the arguments from the generic `plot` method, it is of course possible to change pretty much everything using the regular plot arguments:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some classic options from plot
-plot(disparity_time_slices, col = c("blue", "orange", "green"),
-    ylab = c("Some measurement"), xlab = "Some other measurement",
-    main = "Many options...", ylim = c(5, 0), xlim = c(4, 0))
-
-## Adding a legend
-legend("topleft", legend = c("Central tendency",
-                             "Confidence interval 1",
-                            "Confidence interval 2"),
-      col = c("blue", "orange", "green"), pch = 19)
-
-## Resetting graphical parameters
-par(op)
-```
-
-In addition to the classic `plot` arguments, the function can also take arguments that are specific to `plot.dispRity` like adding the number of elements or rarefaction level (as described above), and also changing the values of the quantiles to plot as well as the central tendency.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some plot.dispRity arguments
-plot(disparity_time_slices, quantile = c(seq(from = 10, to = 100, by = 10)),
-    cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
-    ylab = c("Disparity", "Diversity"), time.subsets = FALSE,
-    xlab = "Time (in in units from past to present)", observed = TRUE,
-    main = "Many more options...")
-
-## Resetting graphical parameters
-par(op)
-```
-
-> Note that the argument `observed = TRUE` allows to plot the disparity values calculated from the non-bootstrapped data as crosses on the plot.
-
-For comparing results, it is also possible to add a plot to the existent plot by using `add = TRUE`:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the continuous disparity with a fixed y axis
-plot(disparity_time_slices, ylim = c(1, 4))
-## Adding the discrete data
-plot(disparity_time_bins, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
-    add = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Finally, if your data has been fully rarefied, it is also possible to easily look at rarefaction curves by using the `rarefaction = TRUE` argument:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the rarefaction curves
-plot(disparity_crown_stem, rarefaction = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-## Testing disparity hypotheses
-
-The `dispRity` package allows users to apply statistical tests to the calculated disparity to test various hypotheses.
-The function `test.dispRity` works in a similar way to the `dispRity` function: it takes a `dispRity` object, a `test` and a `comparisons` argument.
-
-The `comparisons` argument indicates the way the test should be applied to the data:
-
- * `pairwise` (default): to compare each subset in a pairwise manner
- * `referential`: to compare each subset to the first subset
- * `sequential`: to compare each subset to the following subset
- * `all`: to compare all the subsets together (like in analysis of variance)
-
-It is also possible to input a list of pairs of `numeric` values or `characters` matching the subset names to create personalised tests.
-Some other tests implemented in `dispRity` such as the `dispRity::null.test` have a specific way they are applied to the data and therefore ignore the `comparisons` argument. 
-<!-- Add sequential test one day! -->
-
-The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
-It can be a common statistical test function (e.g. `stats::t.test`), a function implemented in `dispRity` (e.g. see `?null.test`) or any function defined by the user.
-
-This function also allows users to correct for Type I error inflation (false positives) when using multiple comparisons via the `correction` argument. 
-This argument can be empty (no correction applied) or can contain one of the corrections from the `stats::p.adjust` function (see `?p.adjust`).
-
-Note that the `test.dispRity` algorithm deals with some classical test outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
-It is, however, possible to get the full detailed output by using the options `details = TRUE`.
-
-Here we are using the variables generated in the [section above](#summarising-dispRity-data-plots):
-
-```{r}
-## T-test to test for a difference in disparity between crown and stem mammals
-test.dispRity(disparity_crown_stem, test = t.test)
-
-## Performing the same test but with the detailed t.test output
-test.dispRity(disparity_crown_stem, test = t.test, details = TRUE)
-
-## Wilcoxon test applied to time sliced disparity with sequential comparisons,
-## with Bonferroni correction
-test.dispRity(disparity_time_slices, test = wilcox.test,
-              comparisons = "sequential", correction = "bonferroni")
-
-## Measuring the overlap between distributions in the time bins (using the
-## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)
-test.dispRity(disparity_time_bins, test = bhatt.coeff)
-```
-
-It is also possible to apply some more *complex* tests that have their own output classes (like `stats::lm`).
-
-The results can then be analysed as usual using the associated `summary` S3 method:
-
-```{r}
-## Performing and linear model applied to the same data
-(slice_lm <- test.dispRity(disparity_time_slices, test = lm,
-                            comparisons = "all"))
-
-## The output is a regular `lm` output
-class(slice_lm)
-
-## This output can be summarised using summary
-summary(slice_lm)
-```
-
-Of course, due to the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
-The only condition is that the test can be applied to at least two distributions.
-In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm` for example).
-
-## Disparity as a distribution
-
-Disparity is often regarded as a summary value of the position of the all elements in the ordinated space.
-For example, the sum of variances, the product of ranges or the median distance between the elements and their centroid will summarise disparity as a single value.
-This value can be pseudo-replicated (bootstrapped) to obtain a distribution of the summary metric with estimated error.
-However, another way to perform disparity analysis is to use the *whole distribution* rather than just a summary metric (e.g. the variances or the ranges).
-
-This is possible in the `dispRity` package by calculating disparity as a dimension-level 2 metric only!
-Let's have a look using our [previous example](#summarising-dispRity-data-plots) of bootstrapped time slices but by measuring the distances between each taxon and their centroid as disparity.
-
-```{r}
-## Measuring disparity as a whole distribution
-disparity_centroids <- dispRity(boot_time_slices, metric = centroids)
-```
-
-The resulting disparity object is of dimension-level 2, so it can easily be transformed into a dimension-level 1 object by, for example, measuring the median distance of all these distributions:
-
-```{r}
-## Measuring median disparity in each time slice
-disparity_centroids_median <- dispRity(disparity_centroids, metric = median)
-```
-
-And we can now compare the differences between these methods:
-
-```{r}
-## Summarising both disparity measurements:
-## The distributions:
-summary(disparity_centroids)
-## The summary of the distributions (as median)
-summary(disparity_centroids_median)
-```
-
-We can see that the summary message for the distribution is slightly different than before.
-Here `summary` also displays the observed central tendency (i.e. the central tendency of the measured distributions).
-Note that, as expected, this central tendency is the same in both metrics!
-
-Another, maybe more intuitive way, to compare both approaches for measuring disparity is to plot the distributions:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(bty = "n", mfrow = c(1, 2))
-
-## Plotting both disparity measurements
-plot(disparity_centroids, ylab = "Distribution of all the distances")
-plot(disparity_centroids_median,
-     ylab = "Distribution of the medians of all the distances")
-
-par(op)
-```
-
-We can then test for differences in the resulting distributions using `test.dispRity` and the `bhatt.coeff` test as described above.
-
-```{r}
-## Probability of overlap in the distribution of medians
-test.dispRity(disparity_centroids_median, test = bhatt.coeff)
-```
-
-In this case, we are looking at the probability of overlap of the distribution of median distances from centroids among each pair of time slices.
-In other words, we are measuring whether the medians from each bootstrap pseudo-replicate for each time slice overlap.
-But of course, we might be interested in the actual distribution of the distances from the centroid rather than simply their central tendencies.
-This can be problematic depending on the research question asked since we are effectively comparing non-independent medians distributions (because of the pseudo-replication).
-
-One solution, therefore, is to look at the full distribution:
-
-```{r}
-## Probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff)
-```
-
-These results show the actual overlap among all the measured distances from centroids concatenated across all the bootstraps.
-For example, when comparing the slices 120 and 80, we are effectively comparing the 5 $\times$ 100 distances (the distances of the five elements in slice 120 bootstrapped 100 times) to the 19 $\times$ 100 distances from slice 80.
-However, this can also be problematic for some specific tests since the _n_ $\times$ 100 distances are also pseudo-replicates and thus are still not independent.
-
-A second solution is to compare the distributions to each other _for each replicate_:
-
-```{r}
-## Boostrapped probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff, concatenate = FALSE)
-```
-
-These results show the median overlap among pairs of distributions in the first column (`bhatt.coeff`) and then the distribution of these overlaps among each pair of bootstraps.
-In other words, when two distributions are compared, they are now compared for each bootstrap pseudo-replicate, thus effectively creating a distribution of probabilities of overlap.
-For example, when comparing the slices 120 and 80, we have a mean probability of overlap of 0.28 and a probability between 0.18 and 0.43 in 50\% of the pseudo-replicates.
-Note that the quantiles and central tendencies can be modified via the `conc.quantiles` option.
-
-## Disparity from other matrices
-
-In the example so far, disparity was measured from an ordinated multidimensional space (i.e. a PCO of the distances between taxa based on discrete morphological characters).
-This is a common approach in palaeobiology, morphometrics or ecology but ordinated matrices are not mandatory for the `dispRity` package!
-It is totally possible to perform the same analysis detailed above using other types of matrices as long as your elements are rows in your matrix.
-
-For example, we can use the data set `eurodist`, an `R` inbuilt dataset that contains the distances (in km) between European cities.
-We can check for example, if Northern European cities are closer to each other than Southern ones:
-
-```{r}
-## Making the eurodist data set into a matrix (rather than "dist" object)
-eurodist <- as.matrix(eurodist)
-eurodist[1:5, 1:5]
-
-## The two groups of cities
-Northern <- c("Brussels", "Calais", "Cherbourg", "Cologne", "Copenhagen",
-              "Hamburg", "Hook of Holland", "Paris", "Stockholm")
-Southern <- c("Athens", "Barcelona", "Geneva", "Gibraltar", "Lisbon", "Lyons",
-              "Madrid", "Marseilles", "Milan", "Munich", "Rome", "Vienna")
-
-## Creating the subset dispRity object
-eurodist_subsets <- custom.subsets(eurodist, group = list("Northern" = Northern,
-                                                        "Southern" = Southern))
-
-## Bootstrapping and rarefying to 9 elements (the number of Northern cities)
-eurodist_bs <- boot.matrix(eurodist_subsets, rarefaction = 9)
-
-## Measuring disparity as the median distance from group's centroid
-euro_disp <- dispRity(eurodist_bs, metric = c(median, centroids))
-
-## Testing the differences using a simple wilcox.test
-euro_diff <- test.dispRity(euro_disp, test = wilcox.test)
-euro_diff_rar <- test.dispRity(euro_disp, test = wilcox.test, rarefaction = 9)
-```
-
-We can compare this approach to an ordination one:
-
-```{r}
-## Ordinating the eurodist matrix
-euro_ord <- cmdscale(eurodist, k = nrow(eurodist) - 2)
-
-## Calculating disparity on the bootstrapped and rarefied subset data
-euro_ord_disp <- dispRity(boot.matrix(custom.subsets(euro_ord, group =
-        list("Northern" = Northern, "Southern" = Southern)), rarefaction = 9),
-        metric = c(median, centroids))
-
-## Testing the differences using a simple wilcox.test
-euro_ord_diff <- test.dispRity(euro_ord_disp, test = wilcox.test)
-euro_ord_diff_rar <- test.dispRity(euro_ord_disp, test = wilcox.test, rarefaction = 9)
-```
-
-And visualise the differences:
-
-```{r, fig.width=12, fig.height=12}
-## Plotting the differences
-par(mfrow = c(2,2), bty = "n")
-## Plotting the normal disparity
-plot(euro_disp, main = "Distance differences")
-## Adding the p-value
-text(1.5, 4000, paste0("p=",round(euro_diff[[2]][[1]], digit = 5)))
-## Plotting the rarefied disparity
-plot(euro_disp, rarefaction = 9, main = "Distance differences (rarefied)")
-## Adding the p-value
-text(1.5, 4000, paste0("p=",round(euro_diff_rar[[2]][[1]], digit = 5)))
-
-## Plotting the ordinated disparity
-plot(euro_ord_disp, main = "Ordinated differences")
-## Adding the p-value
-text(1.5, 1400, paste0("p=",round(euro_ord_diff[[2]][[1]], digit = 5) ))
-## Plotting the rarefied disparity
-plot(euro_ord_disp, rarefaction = 9, main = "Ordinated differences (rarefied)")
-## Adding the p-value
-text(1.5, 1400, paste0("p=",round(euro_ord_diff_rar[[2]][[1]], digit = 5) ))
-
-```
-
-As expected, the results are pretty similar in pattern but different in terms of scale.
-The median centroids distance is expressed in km in the "Distance differences" plots and in Euclidean units of variation in the "Ordinated differences" plots.
-
-<!--chapter:end:03_specific-tutorials.Rmd-->
-
----
-title: "Data simulation features"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Making stuff up!
-
-The `dispRity` package also offers some advanced data simulation features to allow to test hypothesis, explore ordinate-spaces or metrics properties or simply playing around with data!
-All the following functions are based on the same modular architecture of the package and therefore can be used with most of the functions of the package.
-
-## Simulating discrete morphological data
-
-The function `sim.morpho` allows to simulate discrete morphological data matrices (sometimes referred to as "cladistic" matrices).
-It allows to evolve multiple discrete characters on a given phylogenetic trees, given different models, rates, and states.
-It even allows to include proper inapplicable data to make datasets as messy as in real life!
-
-In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
-The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
-The methods are described in details below
-
-```{r}
-set.seed(3)
-## Simulating a starting tree with 15 taxa as a random coalescent tree
-my_tree <- rcoal(15)
-
-## Generating a matrix with 100 characters (85% binary and 15% three state) and
-## an equal rates model with a gamma rate distribution (0.5, 1) with no 
-## invariant characters.
-my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
-    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)
-
-## The first few lines of the matrix
-my_matrix[1:5, 1:10]
-
-## Checking the matrix properties with a quick Maximum Parsimony tree search
-check.morpho(my_matrix, my_tree)
-```
-
-Note that this example produces a tree with a great consistency index and an identical topology to the random coalescent tree!
-Nearly too good to be true...
-
-### A more detailed description
-
-The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;OReilly2017].
-
- * The first `tree` argument will be the tree on which to "evolve" the characters and therefore requires branch length.
- You can generate quick and easy random Yule trees using `ape::rtree(number_of_taxa)` but I would advise to use more realistic trees for more realistic simulations based on more realistic models (really realistic then) using the function `tree.bd` from the [`diversitree`](http://www.zoology.ubc.ca/prog/diversitree/) package [@fitzjohndiversitree2012].
- * The second argument, `character` is the number of characters. Pretty straight forward.
- * The third, `states` is the proportion of characters states above two (yes, the minimum number of states is two).
-This argument intakes the proportion of *n*-states characters, for example `states = c(0.5,0.3,0.2)` will generate 50% of binary-state characters, 30% of three-state characters and 20% of four-state characters.
-There is no limit in the number of state characters proportion as long as the total makes up 100%.
- * The forth, `model` is the evolutionary model for generating the character(s). More about this below.
- * The fifth and sixth, `rates` and `substitution` are the model parameters described below as well.
- * Finally, the two logical arguments, are self explanatory: `invariant` whether to allow invariant characters (i.e.
-characters that don't change) and `verbose` whether to print the simulation progress on your console.
-
-#### Available evolutionary models
-
-There are currently three evolutionary models implemented in `sim.morpho` but more will come in the future.
-Note also that they allow fine tuning parameters making them pretty plastic!
-
- * `"ER"`: this model allows any number of character states and is based on the Mk model [@lewisa2001].
-It assumes a unique overall evolutionary rate equal substitution rate between character states.
-This model is based on the `ape::rTraitDisc` function.
- * `"HKY"`: this is binary state character model based on the molecular HKY model [@HKY85].
-It uses the four molecular states (A,C,G,T) with a unique overall evolutionary rate and a biased substitution rate towards transitions (A <-> G or C <-> T) against transvertions (A <-> C and G <-> T).
-After evolving the nucleotide, this model transforms them into binary states by converting the purines (A and G) into state 0 and the pyrimidines (C and T) into state 1.
-This method is based on the `phyclust::seq.gen.HKY` function and was first proposed by @OReilly2016.
- * `"MIXED"`: this model uses a random (uniform) mix between both the `"ER"` and the `"HKY"` models.
-
-The models can take the following parameters:
-(1) `rates` is the evolutionary rate (i.e. the rate of changes along a branch: the evolutionary speed) and (2) `substitution` is the frequency of changes between one state or another.
-For example if a character can have high probability of changing (the *evolutionary* rate) with, each time a change occurs a probability of changing from state *X* to state *Y* (the *substitution* rate).
-Note that in the `"ER"` model, the substitution rate is ignore because... by definition this (substitution) rate is equal!
-
-The parameters arguments `rates` and `substitution` takes a distributions from which to draw the parameters values for each character.
-For example, if you want an `"HKY"` model with an evolutionary rate (i.e.
-speed) drawn from a uniform distribution bounded between 0.001 and 0.005, you can define it as `rates = c(runif, min = 0.001, max = 0.005)`, `runif` being the function for random draws from a uniform distribution and `max` and `min` being the distribution parameters.
-These distributions should always be passed in the format `c(random_distribution_function, distribution_parameters)` with the names of the distribution parameters arguments.
-
-#### Checking the results
-
-An additional function, `check.morpho` runs a quick Maximum Parsimony tree search using the `phangorn` parsimony algorithm.
-It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g.
-the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree to determine how different they are.
-
-#### Adding inapplicable characters
-
-Once a matrix is generated, it is possible to apply inapplicable characters to it for increasing realism!
-Inapplicable characters are commonly designated as `NA` or simply `-`.
-They differ from missing characters `?` in their nature by being inapplicable rather than unknown.
-For example, considering a binary character defined as "colour of the tail" with the following states "blue" and "red"; on a taxa with no tail, the character should be coded as inapplicable ("`-`") since the state of the character "colour of tail" is *known*: it's neither "blue" or "red", it's just not there!
-It contrasts with coding it as missing ("`?`" - also called as ambiguous) where the state is *unknown*, for example, the taxon of interest is a fossil where the tail has no colour preserved or is not present at all due to bad conservation!
-
-This type of characters can be added to the simulated matrices using the `apply.NA` function/
-It takes, as arguments, the `matrix`, the source of inapplicability (`NAs` - more below), the `tree` used to generate the matrix and the two same `invariant` and `verbose` arguments as defined above.
-The `NAs` argument allows two types of sources of inapplicability:
-
- * `"character"` where the inapplicability is due to the character (e.g. coding a character tail for species with no tail).
- In practice, the algorithm chooses a character *X* as the underlying character (e.g.
-"presence and absence of tail"), arbitrarily chooses one of the states as "absent" (e.g.
-0 = absent) and changes in the next character *Y* any state next to character *X* state 0 into an inapplicable token ("`-`").
- This simulates the inapplicability induced by coding the characters (i.e. not always biological).
- * `"clade"` where the inapplicability is due to evolutionary history (e.g. a clade loosing its tail).
- In practice, the algorithm chooses a random clade in the tree and a random character *Z* and replaces the state of the taxa present in the clade by the inapplicable token ("`-`").
- This simulates the inapplicability induced by evolutionary biology (e.g. the lose of a feature in a clade).
-
-To apply these sources of inapplicability, simply repeat the number of inapplicable sources for the desired number of characters with inapplicable data.
-
-```{r}
-## Generating 5 "character" NAs and 10 "clade" NAs
-my_matrix_NA <- apply.NA(my_matrix, tree = my_tree,
-                         NAs = c(rep("character", 5), rep("clade", 10)))
-
-## The first few lines of the resulting matrix
-my_matrix_NA[1:10, 90:100]
-```
-<!--TG: Add inapplicable paper here -->
-
-### Parameters for a realistic(ish) matrix
-
-There are many parameters that can create a "realistic" matrix (i.e.
-not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all parameters combination end up creating "good" matrices.
-The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
-
-```{r}
-set.seed(0)
-## tree
-my_tree <- rcoal(15)
-## matrix
-morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
-    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
-check.morpho(morpho_mat, my_tree)
-```
-
-<!-- ## Simulating continuous morphological data -->
-<!-- TG: Add tutorial for evolving "morpho" shapes -->
-
-## Simulating multidimensional spaces
-Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
-This function allows users to simulate multidimensional spaces with a certain number of properties.
-For example, it is possible to design a multidimensional space with a specific distribution on each axis, a correlation between the axes and a specific cumulative variance per axis.
-This can be useful for creating ordinated spaces for null hypothesis, for example if you're using the function `null.test` [@diaz2016global].
-
-This function takes as arguments the number of elements (data points - `elements` argument) and dimensions (`dimensions` argument) to create the space and the distribution functions to be used for each axis.
-The distributions are passed through the `distribution` argument as... modular functions!
-You can either pass a single distribution function for all the axes (for example `distribution = runif` for all the axis being uniform) or a specific distribution function for each specific axis (for example `distribution = c(runif, rnorm, rgamma))` for the first axis being uniform, the second normal and the third gamma).
-You can of course use your very own functions or use the ones implemented in `dispRity` for more complex ones (see below).
-Specific optional arguments for each of these distributions can be passed as a list via the `arguments` argument.
-
-Furthermore, it is possible to add a correlation matrix to add a correlation between the axis via the `cor.matrix` argument or even a vector of proportion of variance to be bear by each axis via the `scree` argument to simulate realistic ordinated spaces.
-
-Here is a simple two dimensional example:
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(bty = "n")
-
-## A square space
-square_space <- space.maker(100, 2, runif)
-
-## The resulting 2D matrix
-head(square_space)
-
-## Visualising the space
-plot(square_space, pch = 20, xlab = "", ylab = "", main = "Uniform 2D space")
-```
-
-Of course, more complex spaces can be created by changing the distributions, their arguments or adding a correlation matrix or a cumulative variance vector:
-
-```{r}
-## A plane space: uniform with one dimensions equal to 0
-plane_space <- space.maker(2500, 3, c(runif, runif, runif),
-                           arguments = list(list(min = 0, max = 0), NULL, NULL))
-
-## Correlation matrix for a 3D space
-(cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3))
-
-## An ellipsoid space (normal space with correlation)
-ellipse_space <- space.maker(2500, 3, rnorm, cor.matrix = cor_matrix)
-
-## A cylindrical space with decreasing axes variance
-cylindrical_space <- space.maker(2500, 3, c(rnorm, rnorm, runif),
-                                 scree = c(0.7, 0.2, 0.1))
-```
-
-### Personalised dimensions distributions
-
-Following the modular architecture of the package, it is of course possible to pass home made distribution functions to the `distribution` argument.
-For example, the `random.circle` function is a personalised one implemented in `dispRity`.
-This function allows to create circles based on basic trigonometry allowing to axis to covary to produce circle coordinates.
-By default, this function generates two sets of coordinates with a `distribution` argument and a minimum and maximum boundary (`inner` and `outer` respectively) to create nice sharp edges to the circle.
-The maximum boundary is equivalent to the radius of the circle (it removes coordinates beyond the circle radius) and the minimum is equivalent to the radius of a smaller circle with no data (it removes coordinates below this inner circle radius).
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(bty = "n")
-
-## Generating coordinates for a normal circle with a upper boundary of 1
-circle <- random.circle(1000, rnorm, inner = 0, outer = 1)
-
-## Plotting the circle
-plot(circle, xlab = "x", ylab = "y", main = "A normal circle")
-
-## Creating doughnut space (a spherical space with a hole)
-doughnut_space <- space.maker(5000, 3, c(rnorm, random.circle),
-     arguments = list(list(mean = 0), list(runif, inner = 0.5, outer = 1)))
-```
-
-### Visualising the space
-
-I suggest using the excellent `scatterplot3d` package to play around and visualise the simulated spaces:
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(mfrow = (c(2, 2)), bty = "n")
-## Visualising 3D spaces
-require(scatterplot3d)
-
-## The plane space
-scatterplot3d(plane_space, pch = 20, xlab = "", ylab = "", zlab = "",
-              xlim = c(-0.5, 0.5), main = "Plane space")
-
-## The ellipsoid space
-scatterplot3d(ellipse_space, pch = 20, xlab = "", ylab = "", zlab = "",
-              main = "Normal ellipsoid space")
-
-## A cylindrical space with a decreasing variance per axis
-scatterplot3d(cylindrical_space, pch = 20, xlab = "", ylab = "", zlab = "",
-              main = "Normal cylindrical space")
-## Axes have different orders of magnitude
-
-## Plotting the doughnut space
-scatterplot3d(doughnut_space[,c(2,1,3)], pch = 20, xlab = "", ylab = "",
-              zlab = "", main = "Doughnut space")
-par(op)
-```
-
-### Generating realistic spaces
-
-It is possible to generate "realistic" spaces by simply extracting the parameters of an existing space and scaling it up to the simulated space.
-For example, we can extract the parameters of the `BeckLee_mat50` ordinated space and simulate a similar space.
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Loading the data
-data(BeckLee_mat50)
-
-## Number of dimensions
-obs_dim <- ncol(BeckLee_mat50)
-
-## Observed correlation between the dimensions
-obs_correlations <- cor(BeckLee_mat50)
-
-## Observed mean and standard deviation per axis
-obs_mu_sd_axis <- mapply(function(x,y) list("mean" = x, "sd" = y),
-                         as.list(apply(BeckLee_mat50, 2, mean)),
-                         as.list(apply(BeckLee_mat50, 2, sd)), SIMPLIFY = FALSE)
-
-## Observed overall mean and standard deviation
-obs_mu_sd_glob <- list("mean" = mean(BeckLee_mat50), "sd" = sd(BeckLee_mat50))
-
-## Scaled observed variance per axis (scree plot)
-obs_scree <- variances(BeckLee_mat50)/sum(variances(BeckLee_mat50))
-
-## Generating our simulated space
-simulated_space <- space.maker(1000, dimensions = obs_dim, 
-                               distribution = rep(list(rnorm), obs_dim),
-                               arguments = obs_mu_sd_axis,
-                               cor.matrix = obs_correlations)
-
-## Visualising the fit of our data in the space (in the two first dimensions)
-plot(simulated_space[,1:2], xlab = "PC1", ylab = "PC2")
-points(BeckLee_mat50[,1:2], col = "red", pch = 20)
-legend("topleft", legend = c("observed", "simulated"),
-        pch = c(20,21), col = c("red", "black"))
-```
-
-It is now possible to simulate a space using these observed arguments to test several hypothesis:
-
- * Is the space uniform or normal?
- * If the space is normal, is the mean and variance global or specific for each axis?
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Measuring disparity as the sum of variance
-observed_disp <- dispRity(BeckLee_mat50, metric = c(median, centroids))
-
-## Is the space uniform?
-test_unif <- null.test(observed_disp, null.distrib = runif)
-
-## Is the space normal with a mean of 0 and a sd of 1?
-test_norm1 <- null.test(observed_disp, null.distrib = rnorm)
-
-## Is the space normal with the observed mean and sd and cumulative variance
-test_norm2 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
-                        null.args = rep(list(obs_mu_sd_glob), obs_dim),
-                        null.scree = obs_scree)
-
-## Is the space multiple normal with multiple means and sds and a correlation?
-test_norm3 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
-                        null.args = obs_mu_sd_axis, null.cor = obs_correlations)
-
-## Graphical options
-op <- par(mfrow = (c(2, 2)), bty = "n")
-## Plotting the results
-plot(test_unif, main = "Uniform (0,1)")
-plot(test_norm1, main = "Normal (0,1)")
-plot(test_norm2, main = paste0("Normal (", round(obs_mu_sd_glob[[1]], digit = 3),
-                              ",", round(obs_mu_sd_glob[[2]], digit = 3), ")"))
-plot(test_norm3, main = "Normal (variable + correlation)")
-
-```
-
-If we measure disparity as the median distance from the morphospace centroid, we can explain the distribution of the data as normal with the variable observed mean and standard deviation and with a correlation between the dimensions.
-
-<!--chapter:end:04_making-up-data.Rmd-->
-
 ---
@@
 This option requires the package `snow` and takes arguments that are to be passed to  -->
-
-<!--chapter:end:05_guts.Rmd-->
-
----
-title: "Palaeobiology demo: disparity-through-time and within groups"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Palaeobiology demo: disparity-through-time and within groups
-
-This demo aims to give quick overview of the `dispRity` package (v.0.4) for palaeobiology analyses of disparity, including disparity through time analyses.
-
-This demo showcases a typical disparity-through-time analysis: we are going to test whether the disparity changed through time in a subset of eutherian mammals from the last 100 million years using a dataset from Beck and Lee (2014).
-
-## Before starting
-
-### The morphospace
-
-In this example, we are going to use a subset of the data from @beckancient2014.
-See the [example data](#example-data) description for more details.
-Briefly, this dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
-The ordinated matrix will represent our full morphospace, i.e. all the mammalian morphologies that ever existed through time (for this dataset).
-
-```{r, fig.width=6, fig.height=6}
-## Loading demo and the package data
-library(dispRity)
-
-## Setting the random seed for repeatability
-set.seed(123)
-
-## Loading the ordinated matrix/morphospace:
-data(BeckLee_mat50)
-head(BeckLee_mat50[,1:5])
-
-dim(BeckLee_mat50)
-## The morphospace contains 50 taxa and has 48 dimensions (or axes)
-
-## Showing a list of first and last occurrences data for some fossils
-data(BeckLee_ages)
-head(BeckLee_ages)
-
-## Plotting a phylogeny
-data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.7)
-axisPhylo(root = 140)
-```
-
-> You can have an even nicer looking tree if you use the `strap` package!
-
-```{r, eval = TRUE}
-if(!require(strap)) install.packages("strap")
-strap::geoscalePhylo(BeckLee_tree, cex.tip = 0.7, cex.ts = 0.6)
-```
-
-## A disparity-through-time analysis
-
-### Splitting the morphospace through time
-
-One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time subsets that contain the total number of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
-Basically, the full morphospace represents the total number of morphologies across all time and will be greater than any of the time subsets of the morphospace.
-
-The `dispRity` package provides a `time.subsets` function that allows users to split the morphospace into time slices (using `method = continuous`) or into time bins (using `method = discrete`).
-In this example, we are going to split the morphospace into five equal time bins of 20 million years long from 100 million years ago to the present.
-We will also provide to the function a table containing the first and last occurrences dates for some fossils to take into account that some fossils might occur in several of our different time bins.
-
-```{r}
-## Creating the vector of time bins ages
-(time_bins <- rev(seq(from = 0, to = 100, by = 20)))
-
-## Splitting the morphospace using the time.subsets function
-(binned_morphospace <- time.subsets(data = BeckLee_mat50, tree = BeckLee_tree,
-    method = "discrete", time = time_bins, inc.nodes = FALSE,
-    FADLAD = BeckLee_ages))
-```
-
-The output object is a `dispRity` object (see more about that [here](#The-guts-of-the-dispRity-package).
-In brief, however, `dispRity` objects are lists of different elements (i.e.
-disparity results, morphospace time subsets, morphospace attributes, etc.) that display only a summary of the object when calling the object to avoiding filling the `R` console with superfluous output.
-
-```{r}
-## Printing the class of the object
-class(binned_morphospace)
-
-## Printing the content of the object
-str(binned_morphospace)
-names(binned_morphospace)
-
-## Printing the object as a dispRity class
-binned_morphospace
-```
-
-> These objects will gradually contain more information when completing the following steps in the disparity-through-time analysis.
-
-### Bootstrapping the data
-
-Once we obtain our different time subsets, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
-The bootstrapping allows us to make each subset more robust to outliers and the rarefaction allows us to compare subsets with the same number of taxa to remove sampling biases (i.e.
-more taxa in one subset than the others).
-The `boot.matrix` function bootstraps the `dispRity` object and the `rarefaction` option within performs rarefaction.
-
-```{r}
-## Bootstrapping each time subset 100 times (default)
-(boot_bin_morphospace <- boot.matrix(binned_morphospace))
-
-## Getting the minimum number of rows (i.e. taxa) in the time subsets
-min(size.subsets(boot_bin_morphospace))
-
-## Bootstrapping each time subset 100 times and rarefying them 
-(rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
-    rarefaction = 6))
-```
-
-### Calculating disparity
-
-We can now calculate the disparity within each time subsets along with some confidence intervals generated by the pseudoreplication step above (bootstraps/rarefaction).
-Disparity can be calculated in many ways and this package allows users to come up with their own disparity metrics.
-For more details, please refer to the [`dispRity` metric section](#disparity-metrics).
-
-In this example, we are going to calculate the spread of the data in each time subset by calculating disparity as the sum of the variance of each dimension of the morphospace in each time subset using the `dispRity` function.
-Thus, in this example, disparity is defined by the multi-dimensional variance of each time subset (i.e. the spread of the taxa within the morphospace).
-Note that this metric comes with a caveat (not solved here) since it ignores covariances among the dimensions of the morphospace.
-We use this here because it is a standard metric used in disparity-through-time analysis [@Wills1994].
-
-```{r}
-## Calculating disparity for the bootstrapped data
-(boot_disparity <- dispRity(boot_bin_morphospace, metric = c(sum, variances)))
-
-## Calculating disparity for the rarefied data
-(rare_disparity <- dispRity(rare_bin_morphospace, metric = c(sum, variances)))
-```
-
-The `dispRity` function does not actually display the calculated disparity values but rather only the properties of the disparity object (size, subsets, metric, etc.).
-To display the actual calculated scores, we need to summarise the disparity object using the S3 method `summary` that is applied to a `dispRity` object (see `?summary.dispRity` for more details).
-
-As for any `R` package, you can refer to the help files for each individual function for more details.
-
-```{r}
-## Summarising the disparity results
-summary(boot_disparity)
-summary(rare_disparity)
-```
-> The summary.dispRity function comes with many options on which values to calculate (central tendency and quantiles) and on how many digits to display.
-Refer to the function's manual for more details.
-
-### Plotting the results
-
-It is sometimes easier to visualise the results in a plot than in a table.
-For that we can use the `plot` S3 function to plot the `dispRity` objects (see `?plot.dispRity` for more details).
-
-```{r, fig.width=12, fig.height=6}
-## Graphical options
-quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
-
-## Plotting the bootstrapped and rarefied results
-plot(boot_disparity, type = "continuous", main = "bootstrapped results")
-plot(rare_disparity, type = "continuous", main = "rarefied results")
-```
-
-Testing differences
--------------------
-
-Finally, to draw some valid conclusions from these results, we can apply some statistical tests.
-We can test, for example, if mammalian disparity changed significantly through time over the last 100 million years.
-To do so, we can compare the means of each time-bin in a sequential manner to see whether the disparity in bin *n* is equal to the disparity in bin *n+1*, and whether this is in turn equal to the disparity in bin *n+2*, etc.
-Because our data is temporally autocorrelated (i.e.
-what happens in bin *n+1* depends on what happened in bin *n*) and pseudoreplicated (i.e.
-each bootstrap draw creates non-independent time subsets because they are all based on the same time subsets), we apply a non-parametric mean comparison: the `wilcox.test`.
-Also, we need to apply a p-value correction (e.g. Bonferroni correction) to correct for multiple testing (see `?p.adjust` for more details).
-
-```{r}
-## Testing the differences between bins in the bootstrapped dataset.
-test.dispRity(boot_disparity, test = wilcox.test, comparison = "sequential",
-    correction = "bonferroni")
-
-## Testing the differences between bins in the rarefied dataset.
-test.dispRity(rare_disparity, test = wilcox.test, comparison = "sequential",
-    correction = "bonferroni")
-```
-
-Here our results show significant changes in disparity through time between all time bins (all p-values < 0.05).
-However, when looking at the rarefied results, there is no significant difference between the time bins in the Palaeogene (60-40 to 40-20 Mya), suggesting that the differences detected in the first test might just be due to the differences in number of taxa sampled (13 or 6 taxa) in each time bin.
-
-<!--chapter:end:06_palaeo-demo.Rmd-->
-
----
-title: "dispRity ecology demo"
-author: "Thomas Guillerme"
-date: "`r Sys.Date()`"
-output: rmarkdown::html_vignette
-bibliography: References.bib
-bst: sysbio.bst
-vignette: >
-  %\VignetteIndexEntry{dispRity ecology demo}
-  %\VignetteEngine{knitr::rmarkdown}
-  \usepackage[utf8]{inputenc}
----
-
-# Ecology demo
-
-This is an example of typical disparity analysis that can be performed in ecology.
-
-## Data
-For this example, we will use the famous `iris` inbuilt data set
-
-```{r}
-data(iris)
-```
-
-This data contains petal and sepal length for 150 individual plants sorted into three species.
-
-```{r}
-## Separating the species
-species <- iris[,5]
-## Which species?
-unique(species)
-
-## Separating the petal/sepal length
-measurements <- iris[,1:4]
-head(measurements)
-```
-
-We can then ordinate the data using a PCA (`prcomp` function) thus defining our four dimensional space as the poetically named petal-space.
-
-```{r}
-## Ordinating the data
-ordination <- prcomp(measurements)
-
-## The petal-space
-petal_space <- ordination$x
-
-## Adding the elements names to the petal-space (the individuals IDs)
-rownames(petal_space) <- 1:nrow(petal_space)
-```
-
-## Classic analysis
-A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different species are distributed in the petal-space.
-
-```{r, fig.width=6, fig.height=6}
-## Measuring the variance on each axis
-variances <- apply(petal_space, 2, var)
-variances <- variances/sum(variances)
-
-## Graphical option
-par(bty = "n")
-
-## A classic 2D ordination plot
-plot(petal_space[, 1], petal_space[, 2], col = species,
-    xlab = paste0("PC 1 (", round(variances[1], 2), ")"),
-    ylab = paste0("PC 2 (", round(variances[2], 2), ")"))
-```
-
-This shows the distribution of the different species in the petal-space along the two first axis of variation.
-This is a pretty standard way to visualise the multidimensional space and further analysis might be necessary to test wether the groups are different such as a linear discriminant analysis (LDA).
-However, in this case we are ignoring the two other dimensions of the ordination!
-If we look at the two other axis we see a totally different result:
-
-```{r, fig.width=6, fig.height=6}
-## Plotting the two second axis of the petal-space
-plot(petal_space[, 3], petal_space[, 4], col = species,
-    xlab = paste0("PC 3 (", round(variances[3], 2), ")"),
-    ylab = paste0("PC 4 (", round(variances[4], 2), ")"))
-```
-
-Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e.
-the values on the first dimension do not represent a continuous trait (e.g.
-petal length), instead they just represent the ordinations of correlations between the data and some factors.
-
-Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.
-
-## A multidimensional approach with `dispRity`
-
-The first step is to create different subsets that represent subsets of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
-Each of these subsets will contain only the individuals of a specific species.
-
-```{r}
-## Creating the table that contain the elements and their attributes
-petal_subsets <- custom.subsets(petal_space, group = list(
-                                "setosa" = which(species == "setosa"),
-                                "versicolor" = which(species == "versicolor"),
-                                "virginica" = which(species == "virginica")))
-
-## Visualising the dispRity object content
-petal_subsets
-```
-
-This created a `dispRity` object (more about that [here](#guts)) with three subsets corresponding to each subspecies.
-
-### Bootstrapping the data
-We can the bootstrap the subsets to be able test the robustness of the measured disparity to outliers.
-We can do that using the default options of `boot.matrix` (more about that [here](#bootstraps-and-rarefactions)):
-
-```{r}
-## Bootstrapping the data
-(petal_bootstrapped <- boot.matrix(petal_subsets))
-```
-
-### Calculating disparity
-
-Disparity can be calculated in many ways, therefore the `dispRity` function allows users to define their own measure of disparity.
-For more details on measuring disparity, see the [dispRity metrics section](#disparity-metrics).
-
-In this example, we are going to define disparity as the median distance between the different individuals and the centroid of the ordinated space.
-High values of disparity will indicate a generally high spread of points from this centroid (i.e. on average, the individuals are far apart in the ordinated space).
-We can define the metrics easily in the `dispRity` function by feeding them to the `metric` argument.
-Here we are going to feed the functions `stats::median` and `dispRity::centroids` which calculates distances between elements and their centroid.
-
-```{r}
-## Calculating disparity as the median distance between each elements and
-## the centroid of the petal-space
-(petal_disparity <- dispRity(petal_bootstrapped, metric = c(median, centroids)))
-```
-
-### Summarising the results (plot)
-Similarly to the `custom.subsets` and `boot.matrix` function, `dispRity` displays a `dispRity` object.
-But we are definitely more interested in actually look at the calculated values.
-
-First we can summarise the data in a table by simply using `summary`:
-
-```{r}
-## Displaying the summary of the calculated disparity
-summary(petal_disparity)
-```
-
-We can also plot the results in a similar way:
-
-```{r, fig.width=6, fig.height=6}
-## Graphical options
-par(bty = "n")
-
-## Plotting the disparity in the petal_space
-plot(petal_disparity)
-```
-
-Now contrary to simply plotting the two first axis of the PCA where we saw that the species have a different position in the two first petal-space, we can now also see that they occupy this space clearly differently!
-
-### Testing hypothesis
-
-Finally we can test our hypothesis that we guessed from the disparity plot (that some groups occupy different volume of the petal-space) by using the `test.dispRity` option.
-
-```{r}
-## Fitting a linear model to our data
-disparity_lm <- test.dispRity(petal_disparity, test = lm, comparisons = "all")
-
-## Testing whether there is a difference in disparity between the different species
-summary(aov(disparity_lm))
-
-## Post-hoc testing of the differences between species (corrected for multiple tests)
-test.dispRity(petal_disparity, test = t.test, correction = "bonferroni")
-```
-
-We can now see that there is a significant difference in petal-space occupancy between all species of iris.
-
-#### Setting up a multidimensional null-hypothesis
-
-One other series of test can be done on the shape of the petal-space.
-Using a MCMC permutation test we can simulate a petal-space with specific properties and see if our observed petal-space matches these properties (similarly to @diaz2016global):
-
-```{r, fig.width=6, fig.height=6}
-## Testing against a uniform distribution
-disparity_uniform <- null.test(petal_disparity, replicates = 200,
-    null.distrib = runif, scale = FALSE)
-plot(disparity_uniform)
-```
-
-```{r, fig.width=6, fig.height=6}
-## Testing against a normal distribution
-disparity_normal <- null.test(petal_disparity, replicates = 200,
-    null.distrib = rnorm, scale = TRUE)
-plot(disparity_normal)
-```
-
-In both cases we can see that our petal-space is not entirely normal or uniform.
-This is expected because of the simplicity of these parameters.
-
-<!--chapter:end:07_ecology-demo.Rmd-->
-
----
-title: "Future directions and references"
-author: "Thomas Guillerme"
-date: "`r Sys.Date()`"
-output: rmarkdown::html_vignette
-bibliography: References.bib
-bst: sysbio.bst
-vignette: >
-  %\VignetteIndexEntry{future directions}
-  %\VignetteEngine{knitr::rmarkdown}
-  \usepackage[utf8]{inputenc}
----
-
-# Future directions
-
-## More tests!
-
-Some more tests are being developed such as a `sequential.test` to run sequential linear models for testing hypothesis in disparity through time or a `model.test` developed in collaboration with [Mark Puttick](https://puttickbiology.wordpress.com/) to fit modes of evolution to disparity curves.
-Stay tuned!
-
-## More rarefactions!
-
-I am working on including several "clever" bootstrap and rarefaction algorithms for palaeobiology studies based on geological sampling (following:and Butler et al.
-2012 *Evolution* and Kotric and Knoll 2015 *Palaeobiology*).
-
-## Faster disparity calculations
-
-I am slowly implementing parallel disparity calculation as well as C implementations of some disparity metrics to increase significantly improve the speed of the `dispRity` function.
-
-## More modularity
-
-I am equally slowly developing functions to allow more of the options in the package to be modular (in the same way as the `metric` argument in `dispRity`).
-The next arguments to benefit this increased modularity will be `time.subsets`'s `model` argument and `boot.matrix`'s `type` argument.
-
-## More suggestions?
-
-More generally, I welcome any suggestions or collaborations for developing some additional specific ideas.
-Feel free to contact me via email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
-
-<!-- 
-* Beck, R.
-M., & Lee, M.
-S.
-(2014).
-Ancient dates or accelerated rates?
-Morphological clocks and the antiquity of placental mammals.
-Proceedings of the Royal Society of London B: Biological Sciences, 281(1793), 20141278.
-* Cooper, N., & Guillerme, T. (in prep.). Coming soonish!.
-* Diaz, S., Kattge, J., Cornelissen, J.H., Wright, I.J., Lavorel, S., Dray, S., Reu, B., Kleyer, M., Wirth, C., Prentice, I.C.
-and Garnier, E.
-(2016).
-The global spectrum of plant form and function.
-Nature, 529(7585), 167.
-* Donohue, I., Petchey, O.L., Montoya, J.M., Jackson, A.L., McNally, L., Viana, M., Healy, K., Lurgi, M., O'Connor, N.E.
-and Emmerson, M.C, (2013).
-On the dimensionality of ecological stability.
-Ecology letters, 16(4), 421-429.
-* Wills, M. A., Briggs, D. E., & Fortey, R. A. (1994). Disparity as an evolutionary index: a comparison of Cambrian and Recent arthropods. Paleobiology, 20(2), 93-130. -->
-
-<!--chapter:end:08_future_directions.Rmd-->

### 2019-09-09 inst/gitbook/03_specific-tutorials.Rmd
added examples for dispRity utilities matrix.dispRity

@@
 
+If you have subsets in your `dispRity` object, you can also use the `matrix.dispRity` (see [utilities](#utilities)) and `colMeans` to get the centre of a specific subgroup.
+For example
+
+```{r, warning = FALSE}
+## Create a custom subsets object
+dummy_groups <- custom.subsets(dummy_space, group = list("group1" = 1:5, "group2" = 6:10))
+summary(dispRity(dummy_groups, metric = centroids,
+                 centroid = colMeans(matrix.dispRity(dummy_groups, "group1"))))
+```
+
 The `displacements` distance is the ratio between the `centroids` distance and the `centroids` distance with `centroid = 0`.

### 2019-10-14 inst/gitbook/03_specific-tutorials.Rmd
Updated gitbook with moms links

@@
 
-## Disparity metrics
+## Disparity metrics {#disparity-metrics}
 
@@
 Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.
-
-Because of this great diversity of metrics, the package `dispRity` does not have one way to measure disparity but rather proposes to facilitate users in defining their own disparity metric that will best suit their particular analysis.
+Check out this pre-print on selecting the best metric for your specific question on [biorXiv](https://www.biorxiv.org/content/10.1101/801571v1).
+<!-- biorXiv preprint -->
+You can also use the  [`moms` shiny app](https://tguillerme.shinyapps.io/moms/) to test which metric captures which aspect of traitspace occupancy regarding your specific space and your specific question.
+
+Regardless, and because of this great diversity of metrics, the package `dispRity` does not have one way to measure disparity but rather proposes to facilitate users in defining their own disparity metric that will best suit their particular analysis.
 In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
@@
 
+You can find more informations on the vast variety of metrics that you can use in your analysis in this [preprint](https://www.biorxiv.org/content/10.1101/801571v1).
+<!-- biorXiv preprint -->
+
 ### Equations and implementations

### 2019-11-01 inst/gitbook/03_specific-tutorials.Rmd
Changed crown.stem deprecated in gitbook (not pushed)

@@
 crown_stem <- custom.subsets(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
+                                group = crown.stem(BeckLee_tree, inc.nodes = FALSE))
 ## Bootstrapping and rarefying these groups

### 2020-02-14 inst/gitbook/03_specific-tutorials.Rmd

@@
 2 | `ancestral.dist` | The distance between an element and its ancestor | `dispRity` |
+2 | `angles` | The angle of main variation of each dimensions | `dispRity` |
 2 | `centroids`<sup>1</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
@@
 
+#### Orientation: angles
+
+The `angle` performs a least square regression (via the `lm` function) and returns slope of the main axis of variation for each dimension.
+This slope can be converted into different units, `"slope"`, `"degree"` (the default) and `"radian"`.
+This can be changed through the `unit` argument.
+By default, the angle is measured from the slope 0 (the horizontal line in a 2D plot) but this can be changed through the `base` argument (using the defined `unit`):
+
+```{r}
+## The distribution of each angles in degrees for each main axis in the matrix
+summary(dispRity(dummy_space, metric = angles))
+
+## The distribution of slopes deviating from the 1:1 slope:
+summary(dispRity(dummy_space, metric = angles, unit = "slope", base = 1))
+```
+
 ## Summarising dispRity data (plots)

### 2020-02-27 inst/gitbook/05_other-functionalities.Rmd
Added `char.diff` to the manual

@@
 You'll find a brief description of each of them (alphabetically) here:
+
+## `char.diff`
+
+This is yet another function for calculating distance matrices.
+There are many functions for calculating pairwise distance matrices in `R` (`stats::dist`, `vegan::vegdist`, `cluster::daisy` or `Claddis::MorphDistMatrix`) but this one is the `dispRity` one.
+It is slightly different to the ones mentioned above (though not that dissimilar from `Claddis::MorphDistMatrix`) in the fact that it focuses on comparing discrete morphological characters and tries to solve all the problems linked to these kind of matrices (especially dealing with special tokens).
+
+The function intakes a matrix with either `numeric` or `integer` (`NA` included) or matrices with `character` that are indeed `integers` (e.g.`"0"` and `"1"`).
+It then uses a [bitwise operations](https://en.wikipedia.org/wiki/Bitwise_operations_in_C) architecture implemented in `C` that renders the function pretty fast and pretty modular.
+This bitwise operations translates the character states into binary values.
+This way, `0` becomes `1`, `1` becomes `2`, `2` becomes `4`, `3` becomes `8`, etc...
+Specifically it can handle any rules specific to special tokens (i.e.
+symbols) for discrete morphological characters.
+For example, should you treat missing values `"?"` as `NA` (ignoring them) or as any possible character state (e.g.
+`c("0", "1")`?)?
+And how to treat characters with a ampersand (`"&"`)?
+`char.diff` can answer to all these questions!
+
+Let's start by a basic binary matrix 4\*3 with random `integer`:
+
+```{r}
+## A random binary matrix
+matrix_binary <- matrix(sample(c(0,1), 12, replace = TRUE), ncol = 4,
+                        dimnames = list(letters[1:3], LETTERS[1:4]))
+```
+
+By default, `char.diff` measures the hamming distance between characters:
+
+```{r}
+## The hamming distance between characters
+(differences <- char.diff(matrix_binary))
+```
+
+Note that the results is just a pairwise distance (dissimilarity) matrix with some special dual class `matrix` and `char.diff`.
+This means it can easily be plotted via the disparity package:
+
+```{r}
+## Visualising the matrix
+plot(differences)
+```
+
+You can check all the numerous plotting options in the `?plot.char.diff` manual (it won't be developed here).
+
+The `char.diff` function has much more options however (see all of them in the `?char.diff` manual) for example to measure different differences (via `method`) or making the comparison work per row (for a distance matrix between the rows):
+
+```{r}
+## Euclidean distance between rows
+char.diff(matrix_binary, by.col = FALSE, method = "euclidean")
+```
+
+We can however make it more interesting by playing with the different rules to play with different tokens.
+First let's create a matrix with morphological characters as numeric `character`s:
+
+```{r}
+## A random character matrix
+(matrix_character <- matrix(sample(c("0","1","2"), 30, replace = TRUE), ncol = 5,
+                           dimnames = list(letters[1:6], LETTERS[1:5])))
+
+## The hamming difference between columns
+char.diff(matrix_character)
+```
+
+Here the characters are automatically converted into bitwise integers to be compared efficiently.
+We can now add some more special tokens like `"?"` or `"0/1"` for uncertainties between state `"0"` and `"1"` but not `"2"`:
+
+```{r}
+## Adding uncertain characters
+matrix_character[sample(1:30, 8)] <- "0/1"
+
+## Adding missing data
+(matrix_character[sample(1:30, 5)] <- "?")
+
+## The hamming difference between columns including the special characters
+char.diff(matrix_character)
+```
+
+Note here that it detected the default behaviours for the special tokens `"?"` and `"/"`: `"?"` are treated as `NA` (not compared) and `"/"` are treated as both states (e.g.
+`"0/1"` is treated as `"0"` and as `"1"`).
+We can specify both the special tokens and the special behaviours to consider via `special.tokens` and `special.behaviours`.
+The `special.tokens` are `missing = "?"`, `inapplicable = "-"`, `uncertainty = "\"` and `polymorphism = "&"` meaning we don't have to modify them for now.
+However, say we want to change the behaviour for `"?"` and treat them as all possible characters and treat `"/"` as only the character `"0"` (as an integer) we can specify them giving a behaviour function:
+
+```{r}
+## Specifying some special behaviours
+my_special_behaviours <- list(missing = function(x,y) return(y),
+                              uncertainty = function(x,y) return(as.integer(0)))
+
+## Passing these special behaviours to the char.diff function
+char.diff(matrix_character, special.behaviour = my_special_behaviours)
+```
+
+The results are quiet different as before! Note that you can also specify some really specific behaviours for any type of special token.
+
+```{r}
+## Adding weird tokens to the matrix
+matrix_character[sample(1:30, 8)] <- "%"
+
+## Specify the new token and the new behaviour
+char.diff(matrix_character, special.tokens = c(weird_one = "%"),
+                            special.behaviours = list(
+                                weird_one = function(x,y) return(as.integer(42)))
+                            )
+```
+
+Of course the results can be quiet surprising then... But that's the essence of the modularity. You can see more options in the function manual `?char.diff`!
 

### 2020-04-17 inst/gitbook/03_specific-tutorials.Rmd
added deviations metric

@@
 
+<!-- TODO: Sort by group of metrics -->
+
 Several disparity metrics are implemented in the `dispRity` package.
@@
 1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln$vol` |
+2 | `deviations` | The minimal distance between each element and a hyperplane | `dispRity` |
 1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
@@
 \begin{equation}
+    deviations = \frac{|Ax + By + ... + Nm + Intercept|}{\sqrt{A^2 + B^2 + ... + N^2}}
+\end{equation}
+
+\begin{equation}
     displacements = \frac{\sqrt{\sum_{i=1}^{n}{({k}_{n}-Reference_{k})^2}}}{\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}}
@@
 ```
+
+<!-- TODO: add significance -->
 

### 2020-04-17 R/dispRity.metric.R

@@ -1,5 +1,5 @@
 #' @name dispRity.metric
-#' @aliases dimension.level3.fun dimension.level2.fun dimension.level1.fun variances ranges centroids mode.val ellipse.volume convhull.surface convhull.volume diagonal ancestral.dist pairwise.dist span.tree.length n.ball.volume radius neighbours displacements quantiles func.eve func.div angles
+#' @aliases dimension.level3.fun dimension.level2.fun dimension.level1.fun variances ranges centroids mode.val ellipse.volume convhull.surface convhull.volume diagonal ancestral.dist pairwise.dist span.tree.length n.ball.volume radius neighbours displacements quantiles func.eve func.div angles deviations
 #' @title Disparity metrics
 #'
 #' @description Different implemented disparity metrics.
@@ -51,10 +51,11 @@
 #' \itemize{
 #'   \item \code{ancestral.dist}: calculates the distance between each tip and node and their ancestral. This function needs either (1) \code{matrix}/\code{list} from \code{\link{nodes.coordinates}}; or a \code{tree} (\code{"phylo"}) and \code{full} (\code{"logical"}) argument to calculate the node coordinates for the direct descendants (\code{full = FALSE}) or all descendants down to the root (\code{full = TRUE}). NOTE: distance is calculated as \code{"euclidean"} by default, this can be changed using the \code{method} argument.
 #' 
-#'   \item \code{angles}: calculates the angles of the main axis of variation per dimension in a \code{matrix}. The angles are calculated using the least square algorithm from the \code{\link[stats]{lm}} function. The unit of the angle can be changed through the \code{unit} argument (either \code{"degree"} (default), \code{radian} or \code{slope}) and a base angle to measure the angle from can be passed through the \code{base} argument (by default \code{base = 0}, measuring the angle from the horizontal line (not that the \code{base} argument has to be passed in the same unit as \code{unit}).
[…]

### 2020-04-22 inst/gitbook/01_glossary.Rmd
Updated gitbook until section 3

@@
 ---------------|---------------|------------------|-------------
-the multidimensional space | a `matrix` object ($n\times k$) | a morphospace | a function-space
+the multidimensional space | a `matrix` object ($n\times d$) | a morphospace | a function-space
 elements | rows ($n$) | taxa | field experiments
-dimensions | columns ($k$) | morphological characters | communities' compositions
-subsets | a `matrix` ($m \times k$, with $m \leq n$) | time series | experimental treatments
+dimensions | columns ($d$) | morphological characters | communities' compositions
+subsets | a `matrix` ($m \times d$, with $m \leq n$) | time series | experimental treatments
 disparity | a `function` | sum of variances | ellipsoid volume

### 2020-04-22 inst/gitbook/02_getting-started.Rmd
Updated gitbook until section 3

@@
 
+Since version `1.4` it is also possible to include a `"list"` containing matrices.
+These matrices need to have the same dimensions and rownames but can contain different values.
+This is especially useful for modelling uncertainty (see [here](#multi.input) for more details).
+
 ## Ordinated matrices
@@
 These can be any type of ordinations (PCO, PCA, PCoA, MDS, etc.) as long as elements are the rows (taxa, countries, field experiments) and the dimensions are the columns.
+However, note that this is not required from any of the functions in this package.
+You can also use distance matrices or any other matrix type that suits your question and your analysis!
 
@@
 
-> This functionality is not available on the CRAN version of the package (`r version_CRAN`) due to CRAN compatibility issues see [to CRAN or not to CRAN?](#noCRAN) for more information and on how to install the GitHub (more complete) version of the package (`r version_release`).
-
 ```{r, eval = TRUE, echo = FALSE, message = FALSE}
@@
 
-`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
+`dispRity` package can also easily take data from the `Claddis` package using the `Claddis.ordination` function.
 For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:
@@
 
-Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).
+Of course you can use your own data as detailed in the [previous section](#What-sort-of-data-does-dispRity-work-with).
 
@@
  * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix.
-For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4.
+For example `list("A" = c(1,2), "B" = c(3,4))` will create a group _A_ containing elements 1 and 2 from the matrix and a group _B_ containing elements 3 and 4.
 Note that elements can be present in multiple groups at once.

### 2020-04-22 inst/gitbook/index.Rmd
Updated gitbook until section 3

--- 2019-03-06 inst/gitbook/index.Rmd
@@
 version_master <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
+version_master <- 1.4
 
@@
 version_release <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
+version_release <- 1.4
 
@@
 version_CRAN <- strsplit(strsplit(version_raw, split = "\n<td>Version:</td>\n<td>")[[1]][2], split = "</td>\n</tr>\n<tr>\n<td>Depends:")[[1]][1]
+version_CRAN <- 1.4
 ```
@@
 
-This is a package for measuring disparity in `R`.
+This is a package for measuring disparity (aka multidimensional space occupancy) in `R`.
 It allows users to summarise matrices as representations as multidimensional spaces into a single value or distribution describing a specific aspect of this multidimensional space (the disparity).

### 2020-04-24 inst/gitbook/03_specific-tutorials.Rmd
book update (up until section 4)

@@
 The following section contains information specific to some functions.
-If any of your questions are not covered in these sections, please refer to the function help files in R, send me an email (<guillert@tcd.ie>), or raise an issue on GitHub.
+If any of your questions are not covered in these sections, please refer to the function help files in `R`, send me an email (<guillert@tcd.ie>), or raise an issue on [GitHub](https://github.com/TGuillerme/dispRity/issues).
 The several tutorials below describe specific functionalities of certain functions; please always refer to the function help files for the full function documentation!
@@
 
-Here is an example for `method = discrete`:
+### Time-binning
+
+Here is an example for the time binning method (`method = discrete`):
 
@@
 
+If you want to generate time subsets based on stratigraphy, the package proposes a useful functions to do it for you: `get.bin.ages` (check out the function's manual in `R`)!
+
+### Time-slicing
+
 For the time-slicing method (`method = continuous`), the idea is fairly similar.
@@
 
+More details about the differences between these methods can be found in @time-slice.
+
 ```{r, eval=TRUE}
@@
 ```
-
-If you want to generate time subsets based on stratigraphy, the package proposes a useful functions to do it for you: `get.bin.ages` (check out the function's manual in `R`)!
 
@@
 \begin{equation}
-    ancestral.dist = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}
+    ancestral.dist = \sqrt{\sum_{i=1}^{n}{({d}_{n}-Ancestor_{n})^2}}
 \end{equation}
@@
 \begin{equation}
-    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
+    centroids = \sqrt{\sum_{i=1}^{n}{({d}_{n}-Centroid_{d})^2}}
 \end{equation}
@@
 \begin{equation}
-    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
+    diagonal = \sqrt{\sum_{i=1}^{d}|max(d_i) - min(k_i)|}
 \end{equation}
@@
 \begin{equation}
-    displacements = \frac{\sqrt{\sum_{i=1}^{n}{({k}_{n}-Reference_{k})^2}}}{\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}}
+    displacements = \frac{\sqrt{\sum_{i=1}^{n}{({d}_{n}-Reference_{d})^2}}}{\sqrt{\sum_{i=1}^{n}{({d}_{n}-Centroid_{k})^2}}}
 \end{equation}
@@
 \begin{equation}
-    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
+    ellipse.volume = \frac{\pi^{d/2}}{\Gamma(\frac{d}{2}+1)}\displaystyle\prod_{i=1}^{d} (\lambda_{i}^{0.5})
 \end{equation}
@@
 \begin{equation}
-    n.ball.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R
+    n.ball.volume = \frac{\pi^{d/2}}{\Gamma(\frac{d}{2}+1)}\displaystyle\prod_{i=1}^{d} R
 \end{equation}
@@
 \begin{equation}
-    radius = |\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|
+    radius = |\frac{\sum_{i=1}^{n}d_i}{n} - f(\mathbf{v}d)|
 \end{equation}
@@
 \begin{equation}
-    ranges = |max(k_i) - min(k_i)|
+    ranges = |max(d_i) - min(d_i)|
 \end{equation}
@@
 \begin{equation}
-    variances = \sigma^{2}{k_i}
+    variances = \sigma^{2}{d_i}
 \end{equation}
@@
 
-Where *k* is the number of dimensions,
+Where *d* is the number of dimensions,
 *n* the number of elements,
@@
 
-#### Orientation: angles
-
-The `angle` performs a least square regression (via the `lm` function) and returns slope of the main axis of variation for each dimension.
+#### Orientation: angles and deviations
+
+The `angles` performs a least square regression (via the `lm` function) and returns slope of the main axis of variation for each dimension.
 This slope can be converted into different units, `"slope"`, `"degree"` (the default) and `"radian"`.
@@
 
-<!-- TODO: add significance -->
+The `deviations` function is based on a similar algorithm as above but measures the deviation from the main axis (or hyperplane) of variation.
+In other words, it finds the least square line (for a 2D dataset), plane (for a 3D dataset) or hyperplane (for a >3D dataset) and measures the shortest distances between every points and the line/plane/hyperplane.
+By default, the hyperplane is fitted using the least square algorithm from `stats::glm`:
+
+```{r}
+## The distribution of the deviation of each point from the least square hyperplane
+summary(dispRity(dummy_space), metric = deviations))
+```
+
+It is also possible to specify the hyperplane equation through the `hyperplane` equation.
+The equation must contain the intercept first and then all the slopes and is interpreted as $intercept + Ax + By + ...
++ Nd = 0$.
+For example, a 2 line defined as beta + intercept (e.g.
+$y = 2x + 1$) should be defined as `hyperplane = c(1, 2, 1)` ($2x - y + 1 = 0$).
+
+```{r}
+## The distribution of the deviation of each point from a slope (with only the two first dimensions)
+summary(dispRity(dummy_space[, c(1:2)]), metric = deviations, hyperplane = c(1, 2, -1)))
+```
+
+Since both the functions `angles` and `deviations` effectively run a `lm` or `glm` to estimate slopes or hyperplanes, it is possible to use the option `significant = TRUE` to only consider slopes or intercepts that have a slope significantly different than zero using an `aov` with a significant threshold of $p = 0.05$.
+Note that depending on your dataset, using and `aov` could be completely inappropriate!
+In doubt, it's probably better to enter your `base` (for `angles`) or your `hyperplane` (for `deviations`) manually so you're sure you know what the function is measuring.
 
@@
 
-The plots can be of four different types:
-
+The plots can be of five different types:
+
+ * `preview` for a 2d preview of the trait-space. 
  * `continuous` for displaying continuous disparity curves
@@
 
+### `type = preview`
+
 Note that all the options above are plotting disparity objects for which a disparity metric *has been calculated*.
@@
 
-## Note that no disarity has been calculated here:
+## Note that no disparity has been calculated here:
 is.null(cust_subsets$disparity)
@@
 Groups that don't overlap on two set dimensions can totally overlap in all other dimensions!
+
+For `dispRity` objects that do contain disparity data, the default option is to plot your disparity data.
+However you can always force the `preview` option using the following:
+
+```{r}
+par(mfrow = c(2,1))
+## Default plotting
+plot(disparity_time_slices, main = "Disparity through time")
+## Plotting with more arguments
+plot(disparity_time_slices, type = "preview", main = "Two first dimensions of the trait space")
+```
 
@@
 The median centroids distance is expressed in km in the "Distance differences" plots and in Euclidean units of variation in the "Ordinated differences" plots.
+
+## Disparity from multiple matrices (and multiple trees!) {#multi.input}
+
+Since the version `1.4` of this package, it is possible to use multiple trees and multiple matrices in `dispRity` objects.
+To use multiple matrices, this is rather easy: just supply a list of matrices to any of the `dispRity` functions and, as long as they have the same size and the same rownames they will be handled as a distribution of matrices.
+
+```{r}
+set.seed(1)
+## Creating 3 matrices with 4 dimensions and 10 elements each (called t1, t2, t3, etc...)
+matrix_list <- replicate(3, matrix(rnorm(40), 10, 4, dimnames = list(paste0("t", 1:10))),
+                         simplify = FALSE)
+class(matrix_list) # This is a list of matrices
+
+## Measuring some disparity metric on one of the matrices
+summary(dispRity(matrix_list[[1]], metric = c(sum, variances)))
+
+## Measuring the same disparity metric on the three matrices
+summary(dispRity(matrix_list, metric = c(sum, variances)))
+```
+
+As you can see, when measuring the sum of variances on multiple matrices, we now have a distribution of sum of variances rather than a single observed value.
+
+Similarly as running disparity analysis using multiple matrices, you can run the `chrono.subsets` function using multiple trees.
+This can be useful if you want to use a tree posterior distribution rather than a single consensus tree.
+These trees can be passed to `chrono.subsets` as a `"multiPhylo"` object (with the same node and tip labels in each tree).
+First let's define a function to generate multiple trees with the same labels and root ages:
+
+```{r}
+set.seed(1)
+## Matches the trees and the matrices
+## A bunch of trees
+make.tree <- function(n, fun = rtree) {
+    ## Make the tree
+    tree <- fun(n)
+    tree <- chronos(tree, quiet = TRUE,
+                    calibration = makeChronosCalib(tree, age.min = 10, age.max = 10))
+    class(tree) <- "phylo"
+    ## Add the node labels
+    tree$node.label <- paste0("n", 1:Nnode(tree))
+    ## Add the root time
+    tree$root.time <- max(tree.age(tree)$ages)
+    return(tree)
+}
+trees <- replicate(3, make.tree(10), simplify = FALSE)
+class(trees) <- "multiPhylo"
+trees
+```
+
+We can now simulate some ancestral states for the matrices in the example above to have multiple matrices associated with the multiple trees.
+
+```{r}
+## A function for running the ancestral states estimations
+do.ace <- function(tree, matrix) {
+    ## Run one ace
+    fun.ace <- function(character, tree) {
+        results <- ace(character, phy = tree)$ace
+        names(results) <- paste0("n", 1:Nnode(tree))
+        return(results)
+    }
+    ## Run all ace
+    return(rbind(matrix, apply(matrix, 2, fun.ace, tree = tree)))
+}
+
+## All matrices
+matrices <- mapply(do.ace, trees, matrix_list, SIMPLIFY = FALSE)
+```
+
+Let's first see an example of time-slicing with one matrix and multiple trees.
+This assumes that your tip values (observed) and node values (estimated) are fixed with no error on them.
+It also assumes that the nodes in the matrix always corresponds to the node in the trees (in other words, the tree topologies are fixed):
+
+```{r}
+## Making three "proximity" time slices across one tree
+one_tree <- chrono.subsets(matrices[[1]], trees[[1]],
+                           method = "continuous", model = "proximity", time = 3)
+## Making three "proximity" time slices across the three trees
+three_tree <- chrono.subsets(matrices[[1]], trees,
+                             method = "continuous", model = "proximity", time = 3)
+## Measuring disparity as the sum of variances and summarising it
+summary(dispRity(one_tree, metric = c(sum, variances)))
+summary(dispRity(three_tree, metric = c(sum, variances)))
+```
+
+This results show the effect of considering a tree distribution: in the first case (`one_tree`) the time slice at 3.95 Mya has a sum of variances of 2.9 but this values goes down to 0.256 in the second case (`three_tree`) which is due to the differences in branch lengths distributions:
+
+```{r, print = FALSE}
+par(mfrow = c(3,1))
+slices <- c(7.9, 3.95, 0)
+fun.plot <- function(tree) {
+  plot(tree)
+  nodelabels(tree$node.label, cex = 0.8)
+  axisPhylo()
+  abline(v = tree$root.time - slices)
+}
+silent <- lapply(trees, fun.plot)
+```
+
+Note that in this example, the nodes are actually even different in each tree! The node `n4` for example, is not direct descendent of `t4` and `t6` in all trees!
+To fix that, it is possible to input a list of trees and a list of matrices that correspond to each tree in `chrono.subsets` by using the `bind.data = TRUE` option.
+In this case, the matrices need to all have the same row names and the trees all need the same labels as before:
+
+```{r}
+## Making three "proximity" time slices across three trees and three bound matrices
+bound_data <- chrono.subsets(matrices, trees,
+                             method = "continuous", model = "proximity",
+                             time = 3, bind.data = TRUE)
+## Making three "proximity" time slices across three trees and three matrices
+unbound_data <- chrono.subsets(matrices, trees,
+                               method = "continuous", model = "proximity",
+                               time = 3, bind.data = FALSE)
+
+## Measuring disparity as the sum of variances and summarising it
+summary(dispRity(bound_data, metric = c(sum, variances)))
+summary(dispRity(unbound_data, metric = c(sum, variances)))
+```
+
+Note here that the results are again rather different: with the bound data, the slices are done across the three trees and each of their corresponding matrix (resulting in three observation) which is more accurate than the previous results from `three_trees` above.
+With the unbound data, the slices are done across the three trees and applied to the three matrices (resulting in 9 observations).
+As we've seen before, this is incorrect in this case since the trees don't have the same topology (so the nodes selected by a slice through the second tree are not equivalent to the nodes in the first matrix) but it can be useful if the topology is fixed to integrate both uncertainty in branch length (slicing through different trees) and uncertainty from, say, ancestral states estimations (applying the slices on different matrices).

### 2020-04-24 inst/gitbook/index.Rmd
book update (up until section 4)

@@
 
-## To CRAN or not to CRAN? {#noCRAN}
-
-This package can be installed via [the CRAN](https://cran.r-project.org/package=dispRity) or via [GitHub](https://github.com/TGuillerme/dispRity).
-
- * The CRAN version is a more "strict" version of the package that does not include some functionalities such as this gitbook vignette because of compatibility issues with the CRAN requirement.
- * The GitHub version is a more "reactive" version (fed by user's suggestions and comments) and can contain more functionalities (see list below).
-This version exists on a `"release"` branch which is tested as strictly as the CRAN version with Travis and codecov.
-The `"master"` branch contains always the latest "piping hot", not yet release version.
-It might (though rarely) be bugged.
-
-Note however, that apart from the functions listed below, both versions of the package have the same tested and implemented functions.
-
-### So which version do I choose? {#version}
+## Which version do I choose? {#version}
 
@@
 
+### Why is it important to cite us?
+
+Aside from how science works (if you're using a method from a specific, cite that specific paper to refer to that specific method), why is it important to also cite the package and the manual?
+
+All the people involve in making the `dispRity` package happen do it enthusiastically and freely and most amazingly not asking anything in return!
+I created the package with this idea in mind and I am still sticking to it.
+However, academia (the institutions and people producing science around the globe) is unfortunately not optimal at many level (some might even say "broken"): high impact papers attract big grants that attract high impact papers and big grants again, all this along with livelihood, permanent position and job security.
+Unfortunately however, method development has a hard time to catch up with the current publish or perish system: constantly updating the `dispRity` package and this manual is hugely time consuming (but really fun!) and that is not even taking into account maintenance and helping users.
+Although I do truly believe that this time spent doing these things modestly help the scientific endeavour, it does not contribute to our paper list!
+
+Therefore, by citing the package and this manual, you help provide visibility to other workers and you might help them in their work!
+And you directly contribute in making this project fun for all the people involved and most of all, free, updated and independent from the publish and perish system!
+
+Thank you!
 <!-- ## Associated papers -->

### 2020-05-01 inst/gitbook/06_guts.Rmd
added examples for dispRity utilities matrix.dispRity
gitbook updates

@@
 
-## `dispRity` utilities
+## `dispRity` utilities {#utilities}
 
@@
 ### `dispRity` object utilities  <a name="dispRity.utilities"></a>
+
 The first set of utilities are functions for manipulating `dispRity` objects:
@@
 #### `make.dispRity`
+
 This function creates empty `dispRity` objects.
@@
 ####  `fill.dispRity`
+
 This function initialises a `dispRity` object and generates its call properties.
@@
 ####  `matrix.dispRity`
+
 This function extracts a specific matrix from a disparity object.
@@
 #### `get.subsets`
+
 This function creates a dispRity object that contains only elements from one specific subsets.
@@
 #### `combine.subsets`
+
 This function allows to merge different subsets.
-If the
 
@@
 #### `extract.dispRity`
+
 This function extracts the calculated disparity values of a specific matrix.
@@
 #### `rescale.dispRity`
+
 This is the modified S3 method for `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object and can take optional arguments (for example the rescaling by dividing by a maximum value).
@@
 #### `sort.dispRity`
+
 This is the S3 method of `sort` for sorting the subsets alphabetically (default) or following a specific pattern.
@@
 ## The `dispRity` object content
+
 The functions above are utilities to easily and safely access different elements in the `dispRity` object.
@@
 
- * `$matrix`: an object of class `matrix`, the full multidimensional space.
+ * `$matrix`: an object of class `list` that contains at least one object of class `matrix`: the full multidimensional space.
  * `$call`: an object of class `list` containing information on the `dispRity` object content.
@@
 In fact, it is composed of one unique instance of a matrix (the multidimensional space) upon which the metric function is called via "pointers" to only a certain number of elements and/or dimensions of this matrix.
-This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint:  at any time, only one matrix is present in the `R` environment rather than multiple copies of it for each subset.
+This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint: at any time, only one matrix (or list of matrices) is present in the `R` environment rather than multiple copies of it for each subset.
 
 ### `$matrix`
-This is the multidimensional space, stored in the `R` environment as a `matrix` object.
-It requires row names but not column names.
+
+This is the multidimensional space, stored in the `R` environment as a `list` object containing one or more `matrix` objects.
+Each `matrix` requires row names but not column names (optional).
 By default, if the row names are missing, `dispRity` function will arbitrarily generate them in numeric order (i.e. `rownames(matrix) <- 1:nrow(matrix)`).
@@
 ### `$call`
+
 This element contains the information on the `dispRity` object content.
@@
 
- * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"equal.split"`, `"gradual.split"`).
+ * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`), their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"equal.split"`, `"gradual.split"`) and eventual information about the trees and matrices used through `chrono.subsets`.
 This element generated only once via `chrono.subsets()` and `custom.subsets()`.
@@
 ### `$subsets`
+
 This element contain the eventual subsets of the multidimensional space.
@@
 
+When a probabilistic model was used to select the elements (models that have the `"split"` suffix, e.g.
+`chrono.subsets(..., model = "gradual.split")`), the `$elements` is a matrix containing a pair of elements of the matrix and a probability for sampling the first element in that list:
+
+```
+ disparity$subsets$sub1$elements
+      [,1] [,2]       [,3]
+ [1,]   73   36 0.01871893
+ [2,]   74   37 0.02555876
+ [3,]   33   38 0.85679821
+```
+
+In this example, you can read the table row by row as: "there is a probability of `0.018` for sampling element `73` and a probability of `0.82` (`1-0.018`) of sampling element `36`".
+
 ### `$disparity`
+
 The `$disparity` element is identical to the `$subsets` element structure (a list of list(s) containing matrices) but the matrices don't contain "pointers" to `$matrix` but the disparity result of the disparity metric applied to the "pointers".
@@
 ```
-
-<!-- ## Modularity
-
-### Making your own metrics
-
-### Making your own tests
-
-## Where is this going?
- -->
-<!-- ## Running `dispRity` in parallel
-
-The computationally intensive function `dispRity` have a `parallel` option to speed up their calculations.
-
-This option requires the package `snow` and takes arguments that are to be passed to  -->

### 2020-05-01 inst/gitbook/06_guts.Rmd
Revert "gitbook updates"
This reverts commit e41ca75f88719d79e031e170e8ef0df6b29f7ff7.

--- 2019-08-15 inst/gitbook/06_guts.Rmd
@@
 In fact, it is composed of one unique instance of a matrix (the multidimensional space) upon which the metric function is called via "pointers" to only a certain number of elements and/or dimensions of this matrix.
-This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint:  at any time, only one matrix is present in the `R` environment rather than multiple copies of it for each subset.
+This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint: at any time, only one matrix (or list of matrices) is present in the `R` environment rather than multiple copies of it for each subset.
 
 ### `$matrix`
-This is the multidimensional space, stored in the `R` environment as a `matrix` object.
-It requires row names but not column names.
+This is the multidimensional space, stored in the `R` environment as a `list` object containing one or more `matrix` objects.
+Each `matrix` requires row names but not column names (optional).
 By default, if the row names are missing, `dispRity` function will arbitrarily generate them in numeric order (i.e. `rownames(matrix) <- 1:nrow(matrix)`).
@@
 
+When a probabilistic model was used to select the elements (models that have the `"split"` suffix, e.g.
+`chrono.subsets(..., model = "gradual.split")`), the `$elements` is a matrix containing a pair of elements of the matrix and a probability for sampling the first element in that list:
+
+```
+ disparity$subsets$sub1$elements
+      [,1] [,2]       [,3]
+ [1,]   73   36 0.01871893
+ [2,]   74   37 0.02555876
+ [3,]   33   38 0.85679821
+```
+
+In this example, you can read the table row by row as: "there is a probability of `0.018` for sampling element `73` and a probability of `0.82` (`1-0.018`) of sampling element `36`".
+
 ### `$disparity`

### 2020-05-01 inst/gitbook/04_making-up-data.Rmd
updated gitbook for 1.4

@@
 It allows to evolve multiple discrete characters on a given phylogenetic trees, given different models, rates, and states.
-It even allows to include proper inapplicable data to make datasets as messy as in real life!
+It even allows to include "proper" inapplicable data to make datasets as messy as in real life!
 
@@
 Inapplicable characters are commonly designated as `NA` or simply `-`.
-They differ from missing characters `?` in their nature by being inapplicable rather than unknown.
+They differ from missing characters `?` in their nature by being inapplicable rather than unknown[see @brazeau2018 for more details].
 For example, considering a binary character defined as "colour of the tail" with the following states "blue" and "red"; on a taxa with no tail, the character should be coded as inapplicable ("`-`") since the state of the character "colour of tail" is *known*: it's neither "blue" or "red", it's just not there!
@@
 ```
-<!--TG: Add inapplicable paper here -->
 
@@
 plot(test_norm3, main = "Normal (variable + correlation)")
-
 ```

### 2020-05-04 tests/testthat/test-print.dispRity.R
update make.metric with data.dim option

@@ -124,9 +124,9 @@ test_that("randtest printing", {
         c("Monte-Carlo test", "Call: [1] \"dispRity::null.test\"",
         "", "Observation: 0.02463073 ", 
         "", "Based on 100 replicates", 
-        "Simulated p-value: 0.05940594 ", "Alternative hypothesis: two-sided ", 
+        "Simulated p-value: 0.07920792 ", "Alternative hypothesis: two-sided ", 
         "", "    Std.Obs Expectation    Variance ",
-        "  -2.019988    5.094270    6.298782 "))
+        "  -1.842882    5.338100    8.313070 "))
 
   
     ## Running the test on multiple subsets (may take some time!)
@@ -154,7 +154,7 @@ test_that("randtest printing", {
         "Alternative hypothesis: two-sided "  ,
         ""                                    ,
         "    Std.Obs Expectation    Variance ",
-        "  -32.23852    47.82240     2.01022 ",
+        " -34.623901   47.841400    1.744226 ",
         ""                                    ,
         "$V1.2"                               ,
         "Monte-Carlo test"                    ,
@@ -167,7 +167,7 @@ test_that("randtest printing", {
         "Alternative hypothesis: two-sided "  ,
         ""                                    ,
         "    Std.Obs Expectation    Variance ",
-        " -33.801715   47.929400    1.819248 ",
+        " -33.531312   48.068600    1.860014 ",
         ""                                    ,
         "$V1.3"                               ,
         "Monte-Carlo test"                    ,
@@ -180,7 +180,7 @@ test_that("randtest printing", {
         "Alternative hypothesis: two-sided "  ,
         ""                                    ,
         "    Std.Obs Expectation    Variance ",
-        "  -33.55446    48.01460     1.84056 ",
+        " -35.684854   48.116700    1.634665 ",
         ""                                    ,
         "$V1.4"                               ,
         "Monte-Carlo test"                    ,
@@ -193,7 +193,7 @@ test_that("randtest printing", {
         "Alternative hypothesis: two-sided "  ,
         ""                                    ,
         "    Std.Obs Expectation    Variance ",
-        " -30.305293   47.884300    2.235166 ",
+        " -32.322690   48.021700    1.976796 ",
         ""                                    ,
         "NULL" 
         )

### 2020-05-04 inst/gitbook/index.Rmd
updated manual

@@
 title: "dispRity manual"
-author: "Thomas Guillerme (guillert@tcd.ie), Mark Puttick (marknputtick@gmail.com) and  Natalie Cooper (natalie.cooper@nhm.ac.uk)"
+author: "Thomas Guillerme (guillert@tcd.ie),\nMark Puttick (marknputtick@gmail.com) and\nNatalie Cooper (natalie.cooper@nhm.ac.uk)"
 date: "`r Sys.Date()`"

### 2020-05-05 inst/gitbook/02_getting-started.Rmd

@@
 ## Performing a Procrustes transform on the landmarks
-procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)
+procrustes <- gpagen(plethodon$land, PrinAxes = FALSE,
+                     print.progress = FALSE)
 
@@
 disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
-                                        time = 3, metric = c(sum, variances))
+                                        metric = c(sum, variances),
+                                        time = 3)
 ```
@@
 ## Testing for an difference among the time bins
-disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
+disp_lm <- test.dispRity(disparity_data, test = lm,
+                         comparisons = "all")
 summary(disp_lm)
@@
 ## Measuring disparity for each group
-disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
+disparity_data <- dispRity.per.group(BeckLee_mat50,
+                                     group = mammal_groups,
                                      metric = c(sum, variances))

### 2020-05-05 inst/gitbook/03_specific-tutorials.Rmd

@@
 ## Generating three time bins containing the taxa present every 40 Ma
-chrono.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+chrono.subsets(data = BeckLee_mat50, tree = BeckLee_tree,
+               method = "discrete",
                time = c(120, 80, 40, 0))
@@
 ## Automatically generate three equal length bins:
-chrono.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+chrono.subsets(data = BeckLee_mat50, tree = BeckLee_tree,
+               method = "discrete",
                time = 3)
@@
 ```{r, eval=TRUE}
-## Displaying the table of first and last occurrence dates for each taxa
+## Displaying the table of first and last occurrence dates
+## for each taxa
 head(BeckLee_ages)
@@
 ## Generating time bins including taxa that might span between them
-chrono.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+chrono.subsets(data = BeckLee_mat50, tree = BeckLee_tree,
+               method = "discrete",
                time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
@@
 ```{r, eval=TRUE}
-## Generating four time slices every 40 million years under a model of proximity evolution
+## Generating four time slices every 40 million years
+## under a model of proximity evolution
 chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
+               method = "continuous", model = "proximity",
+               time = c(120, 80, 40, 0),
                FADLAD = BeckLee_ages)
@@
 chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree,
-    method = "continuous", model = "proximity", time = 4, FADLAD = BeckLee_ages)
+               method = "continuous", model = "proximity",
+               time = 4, FADLAD = BeckLee_ages)
 ```
@@
 ## Bootstrapping with the full rarefaction
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
+boot.matrix(BeckLee_mat50, bootstraps = 20,
+            rarefaction = TRUE)
 
 ## Or with a set number of rarefaction levels
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8, 3))
+boot.matrix(BeckLee_mat50, bootstraps = 20,
+            rarefaction = c(6:8, 3))
 ```
@@
 ## Attributing a weight of 0 to Cimolestes and 10 to Maelestes
-boot.matrix(BeckLee_mat50, prob = c("Cimolestes" = 0, "Maelestes" = 10))
+boot.matrix(BeckLee_mat50,
+            prob = c("Cimolestes" = 0, "Maelestes" = 10))
 ```
@@
 crown_stem <- custom.subsets(BeckLee_mat50,
-                                group = crown.stem(BeckLee_tree, inc.nodes = FALSE))
+                             group = crown.stem(BeckLee_tree,
+                                                inc.nodes = FALSE))
 ## Bootstrapping and rarefying these groups
@@
 ## Creating time slice subsets
-time_slices <- chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-                               method = "continuous", model = "proximity", 
+time_slices <- chrono.subsets(data = BeckLee_mat99,
+                              tree = BeckLee_tree, 
+                              method = "continuous",
+                              model = "proximity", 
                               time = c(120, 80, 40, 0),
@@
 ```{r}
-## Which dimension-level is the mean function? And can it be used in dispRity?
+## Which dimension-level is the mean function?
+## And can it be used in dispRity?
 make.metric(mean)
 
-## Which dimension-level is the prod.rows function? And can it be used in dispRity?
+## Which dimension-level is the prod.rows function?
+## And can it be used in dispRity?
 make.metric(prod.rows)
 
-## Which dimension-level is the var function? And can it be used in dispRity?
+## Which dimension-level is the var function?
+## And can it be used in dispRity?
 make.metric(var)
@@
 ```{r}
-## Measuring disparity as the standard deviation of all the values of the
+## Measuring disparity as the standard deviation
+## of all the values of the
 ## ordinated matrix (dimension-level 1 function).
@@
 
-## Measuring disparity as the standard deviation of the variance of each axis of
+## Measuring disparity as the standard deviation
+## of the variance of each axis of
 ## the ordinated matrix (dimension-level 1 and 2 functions).
@@
 
-## Measuring disparity as the standard deviation of the variance of each axis of
+## Measuring disparity as the standard deviation
+## of the variance of each axis of
 ## the variance covariance matrix (dimension-level 1, 2 and 3 functions).
@@
 ## variance covariance matrix:
-disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)),
+disparity1 <- summary(dispRity(BeckLee_mat50,
+                               metric = c(sd, variances, var)),
                       round = 10)
@@
 ## Same as above but using a different function order for the metric argument
-disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
+disparity2 <- summary(dispRity(BeckLee_mat50,
+                               metric = c(variances, sd, var)),
                       round = 10)
@@
 dummy_space <- space.maker(10, 5, rnorm)
+rownames(dummy_space) <- 1:10
 ```
@@
 
-## Calculating the variances of each dimension in the ordinated space
+## Calculating the variances of each dimension in the
+## ordinated space
 variances(dummy_space)
@@
 
-## Calculating disparity as the sum and the product of these variances
+## Calculating disparity as the sum and
+## the product of these variances
 summary(dispRity(dummy_space, metric = c(sum, variances)))
@@
 
-## Calculating the quantiles of each dimension in the ordinated space
+## Calculating the quantiles of each dimension
+## in the ordinated space
 quantiles(dummy_space)
@@
 
-## By default, the quantile calculated is the 95% (i.e. 95% of the data on each axis)
+## By default, the quantile calculated is the 95%
+## (i.e. 95% of the data on each axis)
 ## this can be changed using the option quantile:
@@
 ## Calculating disparity as the mean average radius
-summary(dispRity(dummy_space, metric = c(mean, radius), type = mean))
+summary(dispRity(dummy_space,
+                 metric = c(mean, radius),
+                 type = mean))
 ```
@@
 ## The distribution of the Manhattan distances
-summary(dispRity(dummy_space, metric = pairwise.dist, method = "manhattan"))
+summary(dispRity(dummy_space, metric = pairwise.dist,
+                 method = "manhattan"))
 
@@
 ## The average furthest neighbour manhattan distances
-summary(dispRity(dummy_space, metric = neighbours, which = max, method = "manhattan"))
+summary(dispRity(dummy_space, metric = neighbours,
+                 which = max, method = "manhattan"))
 ```
@@
 ```{r}
-## The distance between each element and the origin of the ordinated space
+## The distance between each element and the origin
+## of the ordinated space
 summary(dispRity(dummy_space, metric = centroids, centroid = 0))
 
-## Disparity as the distance between each element and a specific point in space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
+## Disparity as the distance between each element
+## and a specific point in space
+summary(dispRity(dummy_space, metric = centroids,
+                 centroid = c(0,1,2,3,4)))
 ```
@@
 ## Create a custom subsets object
-dummy_groups <- custom.subsets(dummy_space, group = list("group1" = 1:5, "group2" = 6:10))
+dummy_groups <- custom.subsets(dummy_space,
+                               group = list("group1" = 1:5,
+                                            "group2" = 6:10))
 summary(dispRity(dummy_groups, metric = centroids,
@@
 ## The relative displacement of the group to an arbitrary point
-summary(dispRity(dummy_space, metric = displacements, reference = c(0,1,2,3,4)))
+summary(dispRity(dummy_space, metric = displacements,
+                 reference = c(0,1,2,3,4)))
 ```
@@
 ## Calculating all the ancestral nodes
-all_anc_centroids <- nodes.coordinates(dummy_space2, tree, full = TRUE)
+all_anc_centroids <- nodes.coordinates(dummy_space2, tree,
+                                       full = TRUE)
 
@@
 ## The minimal spanning tree manhanttan distances evenness
-summary(dispRity(dummy_space, metric = func.eve, method = "manhattan"))
+summary(dispRity(dummy_space, metric = func.eve,
+                 method = "manhattan"))
 ```
@@
 ```{r}
-## The distribution of each angles in degrees for each main axis in the matrix
+## The distribution of each angles in degrees for each
+## main axis in the matrix
 summary(dispRity(dummy_space, metric = angles))
@@
 ## The distribution of slopes deviating from the 1:1 slope:
-summary(dispRity(dummy_space, metric = angles, unit = "slope", base = 1))
+summary(dispRity(dummy_space, metric = angles, unit = "slope",
+                 base = 1))
 ```
@@
 ```{r}
-## The distribution of the deviation of each point from the least square hyperplane
-summary(dispRity(dummy_space), metric = deviations))
+## The distribution of the deviation of each point
+## from the least square hyperplane
+summary(dispRity(dummy_space, metric = deviations))
 ```
@@
 ```{r}
-## The distribution of the deviation of each point from a slope (with only the two first dimensions)
-summary(dispRity(dummy_space[, c(1:2)]), metric = deviations, hyperplane = c(1, 2, -1)))
+## The distribution of the deviation of each point
+## from a slope (with only the two first dimensions)
+summary(dispRity(dummy_space[, c(1:2)], metric = deviations,
+                 hyperplane = c(1, 2, -1)))
 ```
@@
 crown_stem <- custom.subsets(BeckLee_mat50,
-                                group = crown.stem(BeckLee_tree, inc.nodes = FALSE))
+                             group = crown.stem(BeckLee_tree,
+                             inc.nodes = FALSE))
 ## Bootstrapping and rarefying these groups
-boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
+boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100,
+                               rarefaction = TRUE)
 ## Calculate disparity
-disparity_crown_stem <- dispRity(boot_crown_stem, metric = c(sum, variances))
+disparity_crown_stem <- dispRity(boot_crown_stem,
+                                 metric = c(sum, variances))
 
 ## Creating time slice subsets
-time_slices <- chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
+time_slices <- chrono.subsets(data = BeckLee_mat99,
+                              tree = BeckLee_tree,
+                              method = "continuous",
+                              model = "proximity",
+                              time = c(120, 80, 40, 0),
                               FADLAD = BeckLee_ages)
@@
 ## Calculate disparity
-disparity_time_slices <- dispRity(boot_time_slices, metric = c(sum, variances))
+disparity_time_slices <- dispRity(boot_time_slices,
+                                  metric = c(sum, variances))
 
 ## Creating time bin subsets
-time_bins <- chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages,
+time_bins <- chrono.subsets(data = BeckLee_mat99,
+                            tree = BeckLee_tree, 
+                            method = "discrete",
+                            time = c(120, 80, 40, 0),
+                            FADLAD = BeckLee_ages,
                             inc.nodes = TRUE)
@@
 ## Calculate disparity
-disparity_time_bins <- dispRity(boot_time_bins, metric = c(sum, variances))
+disparity_time_bins <- dispRity(boot_time_bins,
+                                metric = c(sum, variances))
 ```
@@
 ## Plotting the results with some plot.dispRity arguments
-plot(disparity_time_slices, quantiles = c(seq(from = 10, to = 100, by = 10)),
-    cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
+plot(disparity_time_slices,
+    quantiles = c(seq(from = 10, to = 100, by = 10)),
+    cent.tend = sd, type = "c", elements = TRUE,
+    col = c("black", rainbow(10)),
     ylab = c("Disparity", "Diversity"), chrono.subsets = FALSE,
-    xlab = "Time (in in units from past to present)", observed = TRUE,
+    xlab = "Time (in in units from past to present)",
+    observed = TRUE,
     main = "Many more options...")
@@
 ## Adding the discrete data
-plot(disparity_time_bins, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
-    add = TRUE)
+plot(disparity_time_bins, type = "line", ylim = c(1, 4),
+     xlab = "", ylab = "", add = TRUE)
 
@@
 ## Making the different subsets
-cust_subsets <- custom.subsets(BeckLee_mat99, crown.stem(BeckLee_tree, inc.nodes = TRUE))
-time_subsets <- chrono.subsets(BeckLee_mat99, tree = BeckLee_tree, method = "discrete", time = 5)
+cust_subsets <- custom.subsets(BeckLee_mat99,
+                               crown.stem(BeckLee_tree,
+                                          inc.nodes = TRUE))
+time_subsets <- chrono.subsets(BeckLee_mat99,
+                               tree = BeckLee_tree,
+                               method = "discrete",
+                               time = 5)
 
@@
 ## Plotting with more arguments
-plot(time_subsets, dimensions = c(4,5), main = "Some \"low\" dimensions")
+plot(time_subsets, dimensions = c(4,5),
+     main = "Some \"low\" dimensions")
 ```
@@
 ## Plotting with more arguments
-plot(disparity_time_slices, type = "preview", main = "Two first dimensions of the trait space")
+plot(disparity_time_slices, type = "preview",
+     main = "Two first dimensions of the trait space")
 ```
@@
 ## Generating a random character matrix
-character_matrix <- sim.morpho(rtree(20), 50, rates = c(rnorm, 1, 0))
+character_matrix <- sim.morpho(rtree(20), 50,
+                               rates = c(rnorm, 1, 0))
 
@@
 ## Creating two groups with two states each
-groups <- as.data.frame(matrix(data = c(rep(1,10), rep(2,10), rep(c(1,2), 10)),
-         nrow = 20, ncol = 2, dimnames = list(paste0("t", 1:20), c("g1", "g2"))))
+groups <- as.data.frame(matrix(data = c(rep(1,10),
+                                        rep(2,10),
+                                        rep(c(1,2), 10)),
+                        nrow = 20, ncol = 2,
+                        dimnames = list(paste0("t", 1:20),
+                                        c("g1", "g2"))))
 
@@
 time_subsets <- chrono.subsets(BeckLee_mat50, BeckLee_tree,
-     method = "discrete", inc.nodes = FALSE, time = c(100, 85, 65, 0),
+                               method = "discrete",
+                               inc.nodes = FALSE,
+                               time = c(100, 85, 65, 0),
                                FADLAD = BeckLee_ages)
@@
 ## Calculate the disparity of the dataset using the sum of variance
-dispRity_dtt <- dtt.dispRity(data = geiger_data$dat, metric = c(sum, variances),
-                             tree = geiger_data$phy, nsim = 100)
+dispRity_dtt <- dtt.dispRity(data = geiger_data$dat,
+                             metric = c(sum, variances),
+                             tree = geiger_data$phy,
+                             nsim = 100)
 
@@
 ## Calculating the disparity as the average pairwise distances
-obs_disparity <- dispRity(normal_space, metric = c(mean, pairwise.dist))
+obs_disparity <- dispRity(normal_space,
+                          metric = c(mean, pairwise.dist))
 
 ## Testing against 100 randomly generated normal spaces
-(results <- null.test(obs_disparity, replicates = 100, null.distrib = rnorm))
+(results <- null.test(obs_disparity, replicates = 100,
+                      null.distrib = rnorm))
 ```
@@
 ```{r}
-disp_time_pooled <- model.test(data = BeckLee_disparity, model = "Stasis", pool.variance = TRUE)
+disp_time_pooled <- model.test(data = BeckLee_disparity,
+                               model = "Stasis",
+                               pool.variance = TRUE)
 ```
@@
 ```{r}
-disp_time <- model.test(data = BeckLee_disparity, model = "Stasis", pool.variance = NULL)
+disp_time <- model.test(data = BeckLee_disparity,
+                        model = "Stasis",
+                        pool.variance = NULL)
 disp_time
@@
 ```{r}
-disp_time <- model.test(data = BeckLee_disparity, model = c("Stasis", "BM"))
+disp_time <- model.test(data = BeckLee_disparity,
+                        model = c("Stasis", "BM"))
 disp_time
@@
 ```{r}
-disp_time <- model.test(data = BeckLee_disparity, model = c("Stasis", "BM", "OU", "Trend", "EB"))
+disp_time <- model.test(data = BeckLee_disparity,
+                  model = c("Stasis", "BM", "OU", "Trend", "EB"))
 summary(disp_time)
@@
 ```{r plot3, fig.align="center", fig.cap = "Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for Trend, OU, BM, EB, and Stasis models", echo = TRUE, fig.height = 18, fig.width = 12}
-disp_time <- model.test.wrapper(data = BeckLee_disparity, model = c("Stasis", "BM", "OU", "Trend", "EB"),
+disp_time <- model.test.wrapper(data = BeckLee_disparity,
+                    model = c("Stasis", "BM", "OU", "Trend", "EB"),
                                 show.p = TRUE)
@@
 ```{r plot4, fig.align="center", fig.cap = "Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for Trend, OU, BM, EB, and Stasis models with the optima of the OU model set to equal the ancestral value", echo = TRUE, fig.height = 18, fig.width = 12}
-disp_time <- model.test.wrapper(data = BeckLee_disparity, model = c("Stasis", "BM", "OU", "Trend", "EB"),
+disp_time <- model.test.wrapper(data = BeckLee_disparity,
+                    model = c("Stasis", "BM", "OU", "Trend", "EB"),
                                 show.p = TRUE, fixed.optima = TRUE)
@@
 ```{r plot5, fig.align="center", fig.cap = "Empirical disparity through time (pink), simulate data based on estimated model parameters (grey), delta AICc, and range of p values from the Rank Envelope test for BM, Trend, OU, and multi OU models with a shift in optima allowed at 66 Ma", echo = TRUE, fig.height = 12, fig.width = 12}
-disp_time <- model.test.wrapper(data = BeckLee_disparity, model = c("BM", "Trend", "OU", "multi.OU"),
-                                time.split = 66, pool.variance = NULL, show.p = TRUE,
+disp_time <- model.test.wrapper(data = BeckLee_disparity,
+                        model = c("BM", "Trend", "OU", "multi.OU"),
+                                time.split = 66,
+                                pool.variance = NULL,
+                                show.p = TRUE,
                                 fixed.optima = TRUE)
@@
 ## WARNING: this will take between 20 minutes and half and hour to run!
-disp_time <- model.test.wrapper(data = BeckLee_disparity, model = c("BM", "Trend", "OU", "multi.OU"),
+disp_time <- model.test.wrapper(data = BeckLee_disparity,
+                        model = c("BM", "Trend", "OU", "multi.OU"),
                                 show.p = TRUE, fixed.optima = TRUE)
@@
 ## Testing the models
-disp_time <- model.test.wrapper(data = BeckLee_disparity, model = my_models, time.split = 66,
+disp_time <- model.test.wrapper(data = BeckLee_disparity,
+                                model = my_models, time.split = 66,
                                 show.p = TRUE, fixed.optima = TRUE)
@@
 ## A simple BM model
-model_simulation <- model.test.sim(sim = 1000, model = "BM", time.span = 50, variance = 0.1,
-                                   sample.size = 100, parameters = list(ancestral.state = 0))
+model_simulation <- model.test.sim(sim = 1000, model = "BM",
+                                   time.span = 50, variance = 0.1,
+                                   sample.size = 100,
+                                   parameters = list(ancestral.state = 0))
 model_simulation
@@
 ## Fitting multiple models on the data set
-disp_time <- model.test(data = BeckLee_disparity, model = c("Stasis", "BM", "OU", "Trend", "EB"))
+disp_time <- model.test(data = BeckLee_disparity,
+                    model = c("Stasis", "BM", "OU", "Trend", "EB"))
 summary(disp_time)
@@
 ## Simulating 1000 OU model with the observed parameters
-sim_OU <- model.test.sim(sim = 1000, model = disp_time, model.rank = 2)
+sim_OU <- model.test.sim(sim = 1000, model = disp_time,
+                         model.rank = 2)
 sim_OU
@@
 ## Measuring disparity as a whole distribution
-disparity_centroids <- dispRity(boot_time_slices, metric = centroids)
+disparity_centroids <- dispRity(boot_time_slices, 
+                                metric = centroids)
 ```
@@
 ## Measuring median disparity in each time slice
-disparity_centroids_median <- dispRity(disparity_centroids, metric = median)
+disparity_centroids_median <- dispRity(disparity_centroids,
+                                       metric = median)
 ```
@@
 ## Plotting both disparity measurements
-plot(disparity_centroids, ylab = "Distribution of all the distances")
+plot(disparity_centroids,
+     ylab = "Distribution of all the distances")
 plot(disparity_centroids_median,
@@
 ## Boostrapped probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff, concatenate = FALSE)
+test.dispRity(disparity_centroids, test = bhatt.coeff,
+              concatenate = FALSE)
 ```
@@
 one_tree <- chrono.subsets(matrices[[1]], trees[[1]],
-                           method = "continuous", model = "proximity", time = 3)
+                           method = "continuous",
+                           model = "proximity", time = 3)
 ## Making three "proximity" time slices across the three trees
 three_tree <- chrono.subsets(matrices[[1]], trees,
-                             method = "continuous", model = "proximity", time = 3)
+                             method = "continuous",
+                             model = "proximity", time = 3)
 ## Measuring disparity as the sum of variances and summarising it
@@
 bound_data <- chrono.subsets(matrices, trees,
-                             method = "continuous", model = "proximity",
-                             time = 3, bind.data = TRUE)
+                             method = "continuous",
+                             model = "proximity",
+                             time = 3,
+                             bind.data = TRUE)
 ## Making three "proximity" time slices across three trees and three matrices
 unbound_data <- chrono.subsets(matrices, trees,
-                               method = "continuous", model = "proximity",
-                               time = 3, bind.data = FALSE)
+                               method = "continuous",
+                               model = "proximity",
+                               time = 3,
+                               bind.data = FALSE)
 

### 2020-07-06 inst/gitbook/index.Rmd
minor plot functions update

@@
 
-Aside from how science works (if you're using a method from a specific, cite that specific paper to refer to that specific method), why is it important to also cite the package and the manual?
+Aside from how science works (if you're using a method from a specific paper, cite that specific paper to refer to that specific method), why is it important to also cite the package and the manual?
 
-All the people involve in making the `dispRity` package happen do it enthusiastically and freely and most amazingly not asking anything in return!
+All the people involve in making the `dispRity` package happened to do it enthusiastically, freely and most amazingly without asking anything in return!
 I created the package with this idea in mind and I am still sticking to it.

### 2020-07-06 R/plot.dispRity.R

@@ -154,7 +154,8 @@ plot.dispRity <- function(x, ..., type, quantiles = c(50, 95), cent.tend = media
             ## Set up the plotting window
             ## Open the multiple plots
             if(length_data != 1) {
-                op_tmp <- par(mfrow = c(ceiling(sqrt(length_data)), round(sqrt(length_data))))
+                plot_size <- ifelse(length_data == 3, 4, length_data)
+                op_tmp <- par(mfrow = c(ceiling(sqrt(plot_size)), round(sqrt(plot_size))))
 
                 ## Rarefaction plots
                 for(model in 1:length_data) {
@@ -333,8 +334,11 @@ plot.dispRity <- function(x, ..., type, quantiles = c(50, 95), cent.tend = media
             n_plots <- length(unique(group_plot))
 
             if(n_plots > 1){
+                ## Correct the number of plots if only 3
+                plot_size <- ifelse(n_plots == 3, 4, n_plots)
+
                 ## Setting up plotting window
-                op_tmp <- par(mfrow = c(ceiling(sqrt(n_plots)),floor(sqrt(n_plots))))
+                op_tmp <- par(mfrow = c(ceiling(sqrt(plot_size)),floor(sqrt(plot_size))))
             }
 
             ## Separating the data
@@ -671,7 +675,8 @@ plot.dispRity <- function(x, ..., type, quantiles = c(50, 95), cent.tend = media
         n_plots <- length(data$subsets)
 
         ## Open the multiple plots
-        op_tmp <- par(mfrow = c(ceiling(sqrt(n_plots)),round(sqrt(n_plots))))
+        plot_size <- ifelse(n_plots == 3, 4, n_plots)
+        op_tmp <- par(mfrow = c(ceiling(sqrt(plot_size)),round(sqrt(plot_size))))
 
         ## Rarefaction plots
 

### 2020-07-13 inst/gitbook/03_specific-tutorials.Rmd
Update 03_specific-tutorials.Rmd

@@
 
+### Which disparity metric to choose?
+
+The disparity metric that gives the most consistent results is the following one:
+
+```{r}
+best.metric <- function() return(42)
+```
+Joke aside, this is a legitimate question that has no simple answer: **it depends on the dataset and question at hand**.
+Thoughts on which metric to choose can be find in @moms and @Guillerme2020 but again, will ultimately depend on the question and dataset.
+The question should help figuring out which type of metric is desired: for example, in the question "does the extinction released niches for mammals to evolve", the metric in interest should probably pick up a change in size in the trait space (the release could result in some expansion of the mammalian morphospace); or if the question is "does group X compete with group Y", maybe the metric of interested should pick up changes in position (group X can be displaced by group Y).
+
+In order to visualise what signal different disparity metrics are picking, you can use the [`moms`](https://tguillerme.shinyapps.io/moms/) that come with a detailed [manual](https://rawcdn.githack.com/TGuillerme/moms/master/inst/moms_vignette.html) on how to use it.
+
+Alternatively, you can use the `test.metric` function:
+
+#### `test.metric`
+
+This function allows to test whether a metric picks different changes in disparity.
+It intakes the space on which to test the metric, the disparity metric and the type of changes to apply gradually to the space.
+Basically this is a type of biased data rarefaction (or non-biased for `"random"`) to see how the metric reacts to specific changes in trait space.
+
+```{r}
+# Creating a 2D uniform space
+example_space <- space.maker(300, 2, runif)
+
+## Testing the product of ranges metric on the example space
+example_test <- test.metric(example_space, metric = c(prod, ranges),
+                           shifts = c("random", "size")) 
+```
+
+By default, the test runs three replicates of space reduction as described in @moms by gradually removing 10% of the data points following the different algorithms from @moms (here the `"random"` reduction and the `"size"`) reduction, resulting in a `dispRity` object that can be summarised or plotted.
+The number of replicates can be changed using the `replicates` option.
+Still by default, the function then runs a linear model on the simulated data to measure some potential trend in the changes in disparity. 
+The model can be changed using the `model` option.
+Finally, the function runs 10 reductions by default from keeping 10% of the data (removing 90%) and way up to keeping 100% of the data (removing 0%).
+This can be changed using the `steps` option.
+A good disparity metric for your dataset will typically have no trend in the `"random"` reduction (the metric is ideally not affected by sample size) but should have a trend for the reduction of interest.
+
+```{r}
+## The results as a dispRity object
+example_test
+
+## Summarising these results
+summary(example_test)
+
+## Or visualising them
+plot(example_test)
+```
+
 ## Summarising dispRity data (plots)

### 2020-07-14 inst/gitbook/03_specific-tutorials.Rmd
updated gitbook

@@
 
-### Time-slicing
+### Time-slicing {#time-slicing}
 
@@
 
-#### `test.metric`
+#### `test.metric` {#test-metric}
 
@@
 
-### NPMANOVA in `dispRity`
+### NPMANOVA in `dispRity` {#adonis}
 
@@
 
-### `geiger::dtt` model fitting in `dispRity`
+### `geiger::dtt` model fitting in `dispRity` {#dtt}
 
@@
 
-### null morphospace testing with `null.test`
+### null morphospace testing with `null.test` {#null-test}
 
@@
 
-## Fitting modes of evolution to disparity data
+## Fitting modes of evolution to disparity data {#model-fitting}
 
@@
 
-## Disparity as a distribution
+## Disparity as a distribution {#disparity-distribution}
 
@@
 
-## Disparity from other matrices
+## Disparity from other matrices {#other-matrices}
 

### 2020-07-14 inst/gitbook/08_palaeo-demo.Rmd
updated gitbook

--- 2017-11-17 inst/gitbook/dispRity_manual.Rmd
@@
---- 
-title: "dispRity manual"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-site: bookdown::bookdown_site
-output: bookdown::gitbook
-documentclass: book
-link-citations: yes
-github-repo: TGuillerme/dispRity
-description: "dispRity R package vignette"
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-# `dispRity`
-
-This is a package for measuring disparity in `R`.
-It allows users to summarise matrices as representations as multidimensional spaces into a single value or distribution describing a specific aspect of this multidimensional space (the disparity).
-Multidimensional spaces can be ordinated matrices from MDS, PCA, PCO, PCoA but the package is *not* restricted to any type of matrices!
-This manual is based on the version `0.4.2`.
-
-## What is `dispRity`?
-This is a modular package for measuring disparity in `R`.
-It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
-Typically, these analysis are used in palaeobiology and evolutionary biology to study the changes in morphology through time.
-However, there are many more applications in ecology, evolution and beyond.
-
-### Modular?
-Because their exist a multitude of ways to measure disparity, each adapted to every specific question, this package uses an easy to modify modular architecture.
-In coding, each module is simply a function or a modification of a function that can be passed to the main functions of the package to tweak it to your proper needs!
-In practice, you will notice throughout this manual that some function can take other functions as arguments: the modular architecture of this package allows you to use any function for these arguments (with some restrictions explained for each specific cases).
-This will allow you to finely tune your multidimensional analysis to the needs of your specific question!
-
-## Installing and running the package
-
-You can install this package easily if you use the latest version of `R` (> 3.4.0) and `devtools`.
-
-```{r, eval=FALSE}
-## Checking if devtools is already installed
-if(!require(devtools)) install.packages("devtools")
-
-## Installing the latest released version directly from GitHub
-install_github("TGuillerme/dispRity", ref = "release")
-
-## Loading the package
-library(dispRity)
-```
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## loading the package and setting up the start seed.
-library(dispRity)
-## Setting a random seed for repeatability
-set.seed(123)
-```
-
-Note this uses the `release` branch (version`0.4`). <!-- @@@ -->
-For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
-`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, `Claddis`, `geomorph` and `RCurl`).
-
-## Why not CRAN?
-
-This package is not available on CRAN.
-This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
-However, the package follows the strict (and useful!) CRAN standards via Travis.
-
-<!-- badges out start -->
-[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
-[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/master/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
-[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
-<!-- badges out end   -->
-
-## Help
-
-If you need help with the package, hopefully the following manual will be useful.
-However, parts of this package are still in development and some other parts are probably not covered.
-Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
-
-## Citations
-
-You can cite both the package or this manual with the following citation:
-
-> Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646
-
-<!-- > Guillerme, T. & Cooper, N. (2016). The dispRity R package manual. Fighsare XXXX -->
-
-Note that this citation is only temporary (but can still be used!).
-A proper description of the package is currently in review in Methods in Ecology and Evolution and should be re-submitted this winter.
-Upon acceptance (hopefully!), a proper DOI will be released for this manual and the package description.
-
-<!--chapter:end:index.Rmd-->
-
----
-title: "Glossary"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Glossary
-
--   **Multidimensional space**. The mathematical multidimensional object that will be analysed with this package.
-    In morphometrics, this is often referred to as the morphospace.
-    However it may also be referred to as the cladisto-space for cladistic data or the eco-space for ecological data etc.
-    In practice, this term designates a matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.
-
--   **Elements**. The rows of the multidimensional space matrix. Elements can be taxa, field sites, countries etc.
-
--   **Dimensions**.
-The columns of the multidimensional space matrix.
-The dimensions can be referred to as axes of variation, or principal components, for ordinated spaces obtained from a PCA for example.
-
--   **Subsets**. Subsets of the multidimensional space.
-    A subset (or subsets) contains the same number of dimensions as the space but may contain a smaller subset of elements.
-    For example, if our space is composed of birds and mammals (the elements) and 50 principal components of variation (the dimensions), we can create two subsets containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.
-
-<!--chapter:end:01_glossary.Rmd-->
-
----
-title: "Getting started with dispRity"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Getting started with `dispRity`
-
-## What sort of data does `dispRity` work with?
-
-Disparity can be estimated from pretty much any matrix.
-Classically, however, it is measured from ordinated matrices.
-These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).
-
-### Ordination matrices from `Claddis`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if Claddis is installed BUILD ONLY
-if(!require(Claddis)) install.packages("Claddis")
-```
-
-`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
-For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:
-
-```{r}
-require(Claddis)
-
-## Ordinating the example data from Claddis
-Claddis.ordination(Michaux1989) 
-```
-
-Note that several options are available, namely which type of distance should be computed.
-See more info in the function manual (`?Claddis.ordination`).
-Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.
-
-### Ordination matrices from `geomorph`
-
-```{r, eval = TRUE, echo = FALSE, message = FALSE}
-## Checking if geomorph is installed BUILD ONLY
-if(!require(geomorph)) install.packages("geomorph")
-```
-
-You can also easily use data from `geomorph` using the `geomorph.ordination` function.
-This function simply takes Procrustes aligned data and performs an ordination:
-
-```{r}
-require(geomorph)
-
-## Loading the plethodon dataset
-data(plethodon)
-
-## Performing a Procrustes transform on the landmarks
-procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)
-
-## Ordinating this data
-geomorph.ordination(procrustes)[1:5,1:5]
-```
-
-Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
-Additionally you can give the function a `geomorph.data.frame` object.
-If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsets)!
-
-```{r}
-## Using a geomorph.data.frame
-geomorph_df <- geomorph.data.frame(procrustes,
-     species = plethodon$species, site = plethodon$site)
-
-## Ordinating this data and making a dispRity object
-geomorph.ordination(geomorph_df)
-```
-
-More about these `dispRity` objects below!
-
-### Other kinds of ordination matrices
-
-If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordination matrices by using the following functions from the `stats` package.
-Here is how to do it for the following types of matrices:
-
- * Multivariate matrices (principal components analysis; PCA)
-
-```{r}
-## A multivariate matrix
-head(USArrests)
-
-## Ordinating the matrix using `prcomp` 
-ordination <- prcomp(USArrests)
-
-## Selecting the ordinated matrix
-ordinated_matrix <- ordination$x
-head(ordinated_matrix)
-```
-  This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, see the `?princomp` function.
-
- * Distance matrices (classical multidimensional scaling; MDS)
-
-```{r}
-## A matrix of distances between cities
-str(eurodist)
-
-## Ordinating the matrix using cmdscale() with k = 5 dimensions 
-ordinated_matrix <- cmdscale(eurodist, k = 5)
-head(ordinated_matrix)
-```
-
-  This results in a ordinated matrix with European cities as elements and five dimensions.
-
-Of course any other method for creating the ordination matrix is totally valid, you can also not use any ordination at all!
-The only requirements for the `dispRity` functions is that the input is a matrix with elements as rows and dimensions as columns.
-
-## Performing a simple dispRity analysis
-
-Two `dispRity` functions allow users to run an analysis pipeline simply by inputting an ordination matrix.
-These functions allow users to either calculate the disparity through time (`dispRity.through.time`) or the disparity of user-defined groups (`dispRity.per.group`).
-
-**IMPORTANT**
-
-Note that `disparity.through.time` and `disparity.per.group` are wrapper functions (i.e.
-they incorporate lots of other functions) that allow users to run a basic disparity-through-time, or disparity among groups, analysis without too much effort.
-As such they use a lot of default options. 
-These are described in the help files for the functions that are used to make the wrapper functions, and not described in the help files for `disparity.through.time` and `disparity.per.group`. 
-These defaults are good enough for **data exploration**, but for a proper analysis you should consider the **best parameters for your question and data**. 
-For example, which metric should you use?
-How many bootstraps do you require?
-What model of evolution is most appropriate if you are time slicing?
-Should you rarefy the data?
-See [`time.subsets`](#time-slicing), [`custom.subsets`](#customised-subsets), [`boot.matrix`](#bootstraps-and-rarefactions) and [`dispRity.metric`](#disparity-metrics) for more details of the defaults used in each of these functions.
-Note that any of these default arguments can be changed within the `disparity.through.time` or `disparity.per.group` functions.
-
-### Example data
-
-To illustrate these functions, we will use data from @beckancient2014.
-This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
-
-```{r, fig.width=7, fig.height=7}
-## Loading the ordinated matrices
-data(BeckLee_mat50)
-data(BeckLee_mat99)
-
-## The first five taxa and dimensions of the 50 taxa matrix
-head(BeckLee_mat50[, 1:5])
-
-## The first five taxa and dimensions of the 99 taxa + ancestors matrix
-BeckLee_mat99[c(1, 2, 98, 99), 1:5]
-
-## Loading a list of first and last occurrence dates for the fossils
-data(BeckLee_ages)
-head(BeckLee_ages)
-
-## Loading and plotting the phylogeny
-data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.8) 
-axisPhylo(root = 140)
-nodelabels(cex = 0.5)
-```
-
-Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).
-
-### Disparity through time
-
-The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
-This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
-Of course all the defaults can be changed if required, more on this later.
-
-For a disparity through time analysis, you will need:
-
-  * An ordinated matrix (we covered that above)
-  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element.
-To give your tree a root time (i.e.
-an age for the root), you can simply do\\ `my_tree$root.time <- my_age`.
-  * The required number of time subsets (here `time = 3`)
-  * Your favourite disparity metric (here the sum of variances)
-
-Using the Beck and Lee (2014) data described [above](#example-data):
-
-```{r}
-## Measuring disparity through time
-disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
-                                        time = 3, metric = c(sum, variances))
-```
-
-This generates a `dispRity` object (see [here](#guts) for technical details).
-When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:
-
-```{r}
-## Print the disparity_data object
-disparity_data
-```
-
-We asked for three subsets (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.
-
-We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):
-
-```{r, fig.width=7, fig.height=7}
-## Summarising disparity through time
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data, type = "continuous")
-
-## Testing for an difference among the time bins
-disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
-summary(disp_lm)
-```
-
-Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
-You can also directly explore the specific function help files within R and navigate to related functions.
-
-### Disparity among groups
-
-The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
-For example, you could ask if there is a difference in disparity between two groups?
-
-To perform such an analysis, you will need:
-
- * An matrix with rows as elements and columns as dimensions (always!)
- * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix.
-For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4.
-Note that elements can be present in multiple groups at once.
- * Your favourite disparity metric (here the sum of variances)
-
-Using the @beckancient2014 data described [above](#example-data):
-
-```{r}
-## Creating the two groups (crown versus stem) as a list
-mammal_groups <- crown.stem(BeckLee_tree, inc.nodes = TRUE)
-
-## Measuring disparity for each group
-disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
-                                     metric = c(sum, variances))
-```
-
-We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).
-
-```{r, fig.width=7, fig.height=7}
-## Print the disparity_data object
-disparity_data
-
-## Summarising disparity in the different groups
-summary(disparity_data)
-
-## Plotting the results
-plot(disparity_data)
-
-## Testing for a difference between the groups
-test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
-```
-
-<!--chapter:end:02_getting-started.Rmd-->
-
----
-title: "Details of specific functions"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Details of specific functions
-
-The following section contains information specific to some functions.
-If any of your questions are not covered in these sections, please refer to the function help files in R, send me an email (<guillert@tcd.ie>), or raise an issue on GitHub.
-The several tutorials below describe specific functionalities of certain functions; please always refer to the function help files for the full function documentation!
-
-Before each section, make sure you loaded the @beckancient2014 data (see [example data](#example-data) for more details).
-
-```{r}
-## Loading the data
-data(BeckLee_mat50) ; data(BeckLee_mat99)
-data(BeckLee_tree) ; data(BeckLee_ages)
-```
-
-## Time slicing
-
-The function `time.subsets` allows users to divide the matrix into different time subsets or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
-Each subset generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.
-
-Two types of time subsets can be performed by using the `method` option:
-
- *  Discrete time subsets (or time-binning) using `method = discrete`
- *  Continuous time subsets (or time-slicing) using `method = continuous`
-
-For the time-slicing method details see Cooper and Guillerme (in prep.). <!-- @@@ Change cite appropriately! -->
-<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
-For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:
-
- *  Defining the boundaries of the time bins (when `method = discrete`)
- *  Defining the time slices (when `method = continuous`)
-
-Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
-Additionally, it is also possible to input a dataframe containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age, so taxa can appear in more than one time bin/slice.
-
-Here is an example for `method = discrete`:
-
-```{r, eval=TRUE}
-## Generating three time bins containing the taxa present every 40 Ma
-time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0))
-```
-
-Note that we can also generate equivalent results by just telling the function that we want three time-bins as follow:
-
-```{r, eval=TRUE}
-## Automatically generate three equal length bins:
-time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = 3)
-```
-
-In this example, the taxa were split inside each time-bin according to their age.
-However, the taxa here are considered as single points in time.
-It is totally possible that some taxa could have had longer longevity and that they exist in multiple time bins.
-In this case, it is possible to include them in more than one bin by providing a table of first and last occurrence dates (FAD/LAD).
-This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:
-
-```{r, eval=TRUE}
-## Displaying the table of first and last occurrence dates for each taxa
-head(BeckLee_ages)
-
-## Generating time bins including taxa that might span between them
-time.subsets(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-                time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
-```
-
-When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as the root age plus 1\% of the tree length, as long as at least three elements/taxa are present at that point in time.
-The algorithm adds an extra 1\% tree length until reaching the required minimum of three elements.
-It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance among tips *and* nodes.
-
-For the time-slicing method (`method = continuous`), the idea is fairly similar.
-This option, however, requires a matrix that contains the ordinated distance among taxa *and* nodes and an extra argument describing the assumed evolutionary model (via the `model` argument).
-This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node, meaning that a decision must be made about what the value for the branch should be.
-The model can be one of the following:
-
- *  `acctran` where the data chosen along the branch is always the one of the descendant
- *  `deltran` where the data chosen along the branch is always the one of the ancestor
- *  `random` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
- *  `proximity` where the data chosen along the branch is either the descendant or the ancestor depending on branch length
- *  `punctuated` where the data chosen along the branch is both the descendant and the ancestor with an even probability
- *  `gradual` where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length
-
-> Note that the four first models are "fixed": the selected data is always either the one of the descendant or the ancestor.
-The two last models are "probabilistic": the data from both the descendant and the ancestor is used with an associate probability.
-These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants.
-
-```{r, eval=TRUE}
-## Generating four time slices every 40 million years under a model of proximity evolution
-time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-
-## Generating four time slices automatically
-time.subsets(data = BeckLee_mat99, tree = BeckLee_tree,
-    method = "continuous", model = "proximity", time = 4, FADLAD = BeckLee_ages)
-```
-
-## Customised subsets
-
-Another way of separating elements into different categories is to use customised subsets as briefly explained [above](#disparity-among-groups).
-This function simply takes the list of elements to put in each group (whether they are the actual element names or their position in the matrix).
-
-```{r, eval=TRUE}
-## Creating the two groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44))
-
-## Separating the dataset into two different groups
-custom.subsets(BeckLee_mat50, group = mammal_groups)
-```
-
-Elements can easily be assigned to different groups if necessary!
-
-```{r, eval=FALSE}
-## Creating the three groups as a list
-mammal_groups <- list("crown" = c(16, 19:41, 45:50),
-                      "stem" = c(1:15, 17:18, 42:44).
-                      "all" = c(1:50))
-```
-
-## Bootstraps and rarefactions
-
-One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results are, and how sensitive they are to outliers in the dataset.
-This can be achieved using the function `boot.matrix` to bootstrap and/or rarefy the data.
-The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):
-
-```{r, eval=TRUE}
-## Default bootstrapping
-boot.matrix(data = BeckLee_mat50)
-```
-
-The number of bootstrap replicates can be defined using the `bootstraps` option.
-The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
-Currently two algorithms are implemented:
-
- * `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any *m* elements drawn from the data)
- * `single` where only one random element is replaced by one other random element for each pseudo-replicate
-
-```{r, eval=TRUE}
-## Bootstrapping with the single bootstrap method
-boot.matrix(BeckLee_mat50, boot.type = "single")
-```
-
-This function also allows users to rarefy the data using the `rarefaction` argument.
-Rarefaction allows users to limit the number of elements to be drawn at each bootstrap replication.
-This is useful if, for example, one is interested in looking at the effect of reducing the number of elements on the results of an analysis.
-
-This can be achieved by using the `rarefaction` option that draws only *n-x* at each bootstrap replicate (where *x* is the number of elements not sampled).
-The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e.
-remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subset to a minimum of three elements).
-It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
-
-```{r, eval=TRUE}
-## Bootstrapping with the full rarefaction
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)
-
-## Or with a set number of rarefaction levels
-boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8, 3))
-```
-
-One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
-When missing, all dimensions from the ordinated matrix are used.
-
-```{r, eval=TRUE}
-## Using the first 50% of the dimensions
-boot.matrix(BeckLee_mat50, dimensions = 0.5)
-
-## Using the first 10 dimensions
-boot.matrix(BeckLee_mat50, dimensions = 10)
-```
-
-Of course, one could directly supply the subsets generated above (using `time.subsets` or `custom.subsets`) to this function.
-
-```{r, eval=TRUE}
-## Creating subsets of crown and stem mammals
-crown_stem <- custom.subsets(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot.matrix(crown_stem, bootstraps = 200, rarefaction = TRUE)
-
-## Creating time slice subsets
-time_slices <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-                               method = "continuous", model = "proximity", 
-                               time = c(120, 80, 40, 0),
-                               FADLAD = BeckLee_ages)
-
-## Bootstrapping the time slice subsets
-boot.matrix(time_slices, bootstraps = 100)
-```
-
-## Disparity metrics
-
-There are many ways of measuring disparity!
-In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
-For example, one can look at ellipsoid hyper-volume of the ordinated space (Donohue _et al._ 2013), the sum and the product of the ranges and variances (Wills _et al._ 1994) or the median position of the elements relative to their centroid (Wills _et al._ 1994).
-Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.
-
-Because of this great diversity of metrics, the package `dispRity` does not have one way to measure disparity but rather proposes to facilitate users in defining their own disparity metric that will best suit their particular analysis.
-In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
-However the `metric` argument has to follow certain rules:
-
-1.  It must be composed from one to three `function` objects;
-2.  The function(s) must take as a first argument a `matrix` or a `vector`;
-3.  The function(s) must be of one of the three dimension-levels described below;
-4.  At least one of the functions must be of dimension-level 1 or 2 (see below).
-
-### The function dimension-levels
-
-The metric function dimension-levels determine the "dimensionality of decomposition" of the input matrix.
-In other words, each dimension-level designates the dimensions of the output, i.e. either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimension.
-
-![Illustration of the different dimension-levels of functions with an input `matrix`](dispRity_fun.png)
-
-#### Dimension-level 1 functions
-
-A dimension-level 1 function will decompose a `matrix` or a `vector` into a single value:
-
-```{r}
-## Creating a dummy matrix
-dummy_matrix <- matrix(rnorm(12), 4, 3)
-
-## Example of dimension-level 1 functions
-mean(dummy_matrix)
-median(dummy_matrix)
-```
-
-Any summary metric such as mean or median are good examples of dimension-level 1 functions as they reduce the matrix to a single dimension (i.e. one value).
-
-#### Dimension-level 2 functions
-
-A dimension-level 2 function will decompose a `matrix` into a `vector`.
-
-```{r}
-## Defining the function as the product of rows
-prod.rows <- function(matrix) apply(matrix, 1, prod)
-
-## A dimension-level 2 metric
-prod.rows(dummy_matrix)
-```
-
-Several dimension-level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculate the variance or the range of each dimension of the ordinated matrix respectively.
-
-#### Dimension-level 3 functions
-
-Finally a dimension-level 3 function will transform the matrix into another matrix.
-Note that the dimension of the output matrix doesn't need to match the the input matrix:
-
-```{r}
-## A dimension-level 3 metric
-var(dummy_matrix)
-
-## A dimension-level 3 metric with a forced matrix output
-as.matrix(dist(dummy_matrix))
-```
-
-### `make.metric`
-
-Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the Euclidean distance between each element and the centroid of the ordinated space.
-The `make.metric` function implemented in `dispRity` is designed to help test and find the dimension-level of the functions.
-This function tests:
-
-1.  If your function can deal with a `matrix` or a `vector` as an input;
-2.  Your function's dimension-level according to its output (dimension-level 1, 2 or 3, see above);
-3.  Whether the function can be implemented in the `dispRity` function (the function is fed into a `lapply` loop).
-
-For example, let's see if the functions described above are the right dimension-levels:
-
-```{r}
-## Which dimension-level is the mean function? And can it be used in dispRity?
-make.metric(mean)
-
-## Which dimension-level is the prod.rows function? And can it be used in dispRity?
-make.metric(prod.rows)
-
-## Which dimension-level is the var function? And can it be used in dispRity?
-make.metric(var)
-```
-
-A non verbose version of the function is also available.
-This can be done using the option `silent = TRUE` and will simply output the dimension-level of the metric.
-
-```{r}
-## Testing whether mean is dimension-level 1
-if(make.metric(mean, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-## Testing whether var is dimension-level 1
-if(make.metric(var, silent = TRUE) != "level1") {
-    message("The metric is not dimension-level 1.")
-}
-```
-
-### Metrics in the `dispRity` function
-
-Using this metric structure, we can easily use any disparity metric in the `dispRity` function as follows:
-
-```{r}
-## Measuring disparity as the standard deviation of all the values of the
-## ordinated matrix (dimension-level 1 function).
-summary(dispRity(BeckLee_mat50, metric = sd))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the ordinated matrix (dimension-level 1 and 2 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))
-
-## Measuring disparity as the standard deviation of the variance of each axis of
-## the variance covariance matrix (dimension-level 1, 2 and 3 functions).
-summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
-```
-
-Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > 2 > 1).
-
-```{r}
-## Disparity as the standard deviation of the variance of each axis of the
-## variance covariance matrix:
-disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)),
-                      round = 10)
-
-## Same as above but using a different function order for the metric argument
-disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
-                      round = 10)
-
-## Both ways output the same disparity values:
-disparity1 == disparity2
-```
-
-In these examples, we considered disparity to be a single value.
-For example, in the previous example, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
-It is, however, possible to calculate [disparity as a distribution](#disparity-as-a-distribution).
-
-### Metrics implemented in `dispRity`
-
-Several disparity metrics are implemented in the `dispRity` package.
-The detailed list can be found in `?dispRity.metric` along with some description of each metric.
-
-Level | Name | Description | Source | 
-------|------|---------------------------------------------------|--------|
-2 | `ancestral.dist` | The distance between an element and its ancestor | `dispRity` |
-2 | `centroids`<sup>1</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
-1 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
-1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
-1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
-1 | `mode.val` | The modal value | `dispRity` |
-1 | `n.ball.volume` | The hyper-spherical (*n*-ball) volume | `dispRity` |
-2 | `pairwise.dist` | The pairwise distances between elements | [`stats`](https://cran.r-project.org/web/packages/stats/index.html)`::dist` |
-2 | `radius` | The radius of each dimensions | `dispRity` |
-2 | `ranges` | The range of each dimension | `dispRity` |
-1 | `span.tree.length` | The minimal spanning tree length | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree` |
-2 | `variances` | The variance of each dimension | `dispRity` |
-
-1: Note that by default, the centroid is the centroid of the elements.
-It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
-
-2: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (*not* PCA).
-
-### Equations and implementations
-Some of the functions described below are implemented in the `dispRity` package and do not require any other packages to calculate ([see implementation here](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R)).
-
-\begin{equation}
-    ancestral.dist = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}
-\end{equation}
-
-\begin{equation}
-    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
-\end{equation}
-
-\begin{equation}
-    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
-\end{equation}
-
-\begin{equation}
-    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
-\end{equation}
-
-\begin{equation}
-    n.ball.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R
-\end{equation}
-
-\begin{equation}
-    radius = |\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|
-\end{equation}
-
-\begin{equation}
-    ranges = |max(k_i) - min(k_i)|
-\end{equation}
-
-\begin{equation}
-    variances = \sigma^{2}{k_i}
-\end{equation}
-
-\begin{equation}
-    span.tree.length = \sum(\mathrm{branch\ length})
-\end{equation}
-
-Where *k* is the number of dimensions,
-*n* the number of elements,
-$\Gamma$ is the Gamma distribution,
-$\lambda_i$ is the eigenvalue of each dimensions,
-$\sigma^{2}$ is their variance and
-$Centroid_{k}$ is their mean,
-$Ancestor_{n}$ is the coordinates of the ancestor of element $n$,
-$f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.),
-*R* is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid).
-
-### Using the different disparity metrics
-Here is a brief demonstration of the main metrics implemented in `dispRity`.
-First, we will create a dummy/simulated ordinated space using the `space.maker` utility function (more about that [here](#space.maker):
-
-```{r}
-## Creating a 10*5 normal space
-set.seed(1)
-dummy_space <- space.maker(10, 5, rnorm)
-```
-
-We will use this simulated space to demonstrate the different metrics.
-
-#### Volumes and surface metrics
-The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `n.ball.volume`  all measure the surface or the volume of the ordinated space occupied:
-
-```{r}
-## Calculating the ellipsoid volume
-summary(dispRity(dummy_space, metric = ellipse.volume))
-```
-> Because there is only one subset (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with rounding).
-
-```{r}
-## Calculating the convex hull surface
-summary(dispRity(dummy_space, metric = convhull.surface))
-
-## Calculating the convex hull volume
-summary(dispRity(dummy_space, metric = convhull.volume))
-
-## Calculating the convex hull volume
-summary(dispRity(dummy_space, metric = n.ball.volume))
-```
-
-The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
-
-> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) that often results in near 0 disparity values.
-
-#### Ranges, variances, radius, pairwise distance, modal value and diagonal
-
-The functions `ranges`, `variances` `radius`, `pairwise.dist`, `mode.val` and `diagonal` all measure properties of the ordinated space based on its dimensional properties (they are also less affected by the "curse of dimensionality"):
-
-`ranges`, `variances` and `radius` work on the same principle and measure the range/variance/radius of each dimension:
-
-```{r}
-## Calculating the ranges of each dimension in the ordinated space
-ranges(dummy_space)
-
-## Calculating disparity as the distribution of these ranges
-summary(dispRity(dummy_space, metric = ranges))
-
-## Calculating disparity as the sum and the product of these ranges
-summary(dispRity(dummy_space, metric = c(sum, ranges)))
-summary(dispRity(dummy_space, metric = c(prod, ranges)))
-
-## Calculating the variances of each dimension in the ordinated space
-variances(dummy_space)
-
-## Calculating disparity as the distribution of these variances
-summary(dispRity(dummy_space, metric = variances))
-
-## Calculating disparity as the sum and the product of these variances
-summary(dispRity(dummy_space, metric = c(sum, variances)))
-summary(dispRity(dummy_space, metric = c(prod, variances)))
-
-## Calculating the radius of each dimension in the ordinated space
-radius(dummy_space)
-
-## By default the radius is the maximum distance from the centre of
-## the dimension. It can however be changed to any function:
-radius(dummy_space, type = min)
-radius(dummy_space, type = mean)
-
-## Calculating disparity as the mean average radius
-summary(dispRity(dummy_space, metric = c(mean, radius), type = mean))
-```
-
-The pairwise distances uses the function `stats::dist` and can take the normal `dist` options:
-
-```{r}
-## The average pairwise euclidean distance
-summary(dispRity(dummy_space, metric = c(mean, pairwise.dist)))
-
-## The distribution of the Manhattan distances
-summary(dispRity(dummy_space, metric = pairwise.dist, method = "manhattan"))
-```
-
-The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest Euclidean distance in our five dimensional space).
-The `mode.val` function measures the modal value of the matrix:
-
-```{r}
-## Calculating the ordinated space's diagonal
-summary(dispRity(dummy_space, metric = diagonal))
-
-## Calculating the modal value of the matrix
-summary(dispRity(dummy_space, metric = mode.val))
-```
-
-> This metric is only a Euclidean diagonal (mathematically valid) if the dimensions within the space are all orthogonal!
-
-#### Centroids and ancestral distance metrics
-
-The `centroids` metric allows users to measure the position of the different elements compared to a fixed point in the ordinated space.
-By default, this function measures the distance between each element and their centroid (centre point):
-
-```{r}
-## The distribution of the distances between each element and their centroid
-summary(dispRity(dummy_space, metric = centroids))
-
-## Disparity as the median value of these distances
-summary(dispRity(dummy_space, metric = c(median, centroids)))
-```
-
-It is however possible to fix the coordinates of the centroid to a specific point in the ordinated space, as long as it has the correct number of dimensions:
-
-```{r}
-## The distance between each element and the origin of the ordinated space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))
-
-## Disparity as the distance between each element and a specific point in space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
-```
-
-The `ancestral.dist` metric works on a similar principle as the `centroids` function but changes the centroid to be the coordinates of each element's ancestor.
-Therefore this functions needs a tree and node coordinates as additional arguments:
-
-```{r}
-## A generating a random tree with node labels
-tree <- rtree(5) ; tree$node.label <- paste0("n", 1:4)
-## Adding the tip and node names to the matrix
-dummy_space2 <- dummy_space[-1,]
-rownames(dummy_space2) <- c(tree$tip.label, tree$node.label)
-
-## Calculating all the ancestral nodes
-all_anc_centroids <- nodes.coordinates(dummy_space2, tree, full = TRUE)
-
-## Calculating the distances from the ancestral nodes
-ancestral_dist <- dispRity(dummy_space2, metric = ancestral.dist,
-                           nodes.coords = all_anc_centroids)
-
-## Calculating disparity as the sum of the distances from all the ancestral nodes
-summary(dispRity(ancestral_dist, metric = sum))
-```
-
-#### Minimal spanning tree length
-
-The `span.tree.length` uses the `vegan::spantree` function to heuristically calculate the minimum spanning tree (the shortest multidimensional tree connecting each elements) and calculates its length as the sum of every branch lengths.
-
-```{r}
-## The length of the minimal spanning tree
-summary(dispRity(dummy_space, metric = span.tree.length))
-```
-
-Note that because the solution is heuristic, this metric can take a long time to compute for big matrices.
-
-## Summarising dispRity data (plots)
-
-Because of its architecture, printing `dispRity` objects only summarises their content but does not print the disparity value measured or associated analysis (more about this [here](#manipulating-dispRity-objects)).
-To actually see what is in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphical representation of the results.
-
-### Summarising `dispRity` data
-This function is an S3 function (`summary.dispRity`) allowing users to summarise the content of `dispRity` objects that contain disparity calculations.
-
-```{r}
-## Example data from previous sections
-crown_stem <- custom.subsets(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
-## Bootstrapping and rarefying these groups
-boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
-## Calculate disparity
-disparity_crown_stem <- dispRity(boot_crown_stem, metric = c(sum, variances))
-
-## Creating time slice subsets
-time_slices <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
-    FADLAD = BeckLee_ages)
-## Bootstrapping the time slice subsets
-boot_time_slices <- boot.matrix(time_slices, bootstraps = 100)
-## Calculate disparity
-disparity_time_slices <- dispRity(boot_time_slices, metric = c(sum, variances))
-
-## Creating time bin subsets
-time_bins <- time.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages,
-    inc.nodes = TRUE)
-## Bootstrapping the time bin subsets
-boot_time_bins <- boot.matrix(time_bins, bootstraps = 100)
-## Calculate disparity
-disparity_time_bins <- dispRity(boot_time_bins, metric = c(sum, variances))
-```
-
-These objects are easy to summarise as follows:
-
-```{r}
-## Default summary
-summary(disparity_time_slices)
-```
-
-Information about the number of elements in each subset and the observed (i.e. non-bootstrapped) disparity are also calculated.
-This is specifically handy when rarefying the data for example:
-
-```{r}
-head(summary(disparity_crown_stem))
-```
-
-The summary functions can also take various options such as:
-
- * `quantile` values for the confidence interval levels (by default, the 50 and 95 quantiles are calculated)
- * `cent.tend` for the central tendency to use for summarising the results (default is `median`)
- * rounding` option corresponding to the number of decimal places to print (default is `2`)
- * `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
-
-These options can easily be changed from the defaults as follows:
-
-```{r}
-## Same as above but using the 88th quantile and the standard deviation as the summary 
-summary(disparity_time_slices, quantile = 88, cent.tend = sd)
-
-## Printing the details of the object and rounding the values to the 5th decimal place
-summary(disparity_time_slices, recall = TRUE, rounding = 5)
-```
-
-Note that the summary table is a `data.frame`, hence it is as easy to modify as any dataframe using `dplyr`.
-You can also export it in `csv` format using `write.csv` or `write_csv` or even directly export into `LaTeX` format using the following;
-
-```{r, eval=FALSE}
-## Loading the xtable package
-require(xtable)
-## Converting the table in LaTeX
-xtable(summary(disparity_time_slices))
-```
-
-### Plotting `dispRity` data
-An alternative (and more fun!) way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
-This function takes the same options as `summary.dispRity` along with various graphical options described in the function help files (see `?plot.dispRity`).
-
-The plots can be of four different types:
-
- * `continuous` for displaying continuous disparity curves
- * `box`, `lines`, and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.
-
-> This argument can be left empty. In this case, the algorithm will automatically detect the type of subsets from the `dispRity` object and plot accordingly.
-
-It is also possible to display the number of elements in each subset (as a horizontal dotted line) using the option `elements = TRUE`.
-Additionally, when the data is rarefied, one can indicate which level of rarefaction to display (i.e. only display the results for a certain number of elements) by using the `rarefaction` argument.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(mfrow = c(2, 2), bty = "n")
-
-## Plotting continuous disparity results
-plot(disparity_time_slices, type = "continuous")
-
-## Plotting discrete disparity results
-plot(disparity_crown_stem, type = "box")
-
-## As above but using lines for the rarefaction level of 20 elements only
-plot(disparity_crown_stem, type = "line", rarefaction = 20)
-
-## As above but using polygons while also displaying the number of elements
-plot(disparity_crown_stem, type = "polygon", elements = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Since `plot.dispRity` uses the arguments from the generic `plot` method, it is of course possible to change pretty much everything using the regular plot arguments:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some classic options from plot
-plot(disparity_time_slices, col = c("blue", "orange", "green"),
-    ylab = c("Some measurement"), xlab = "Some other measurement",
-    main = "Many options...", ylim = c(5, 0), xlim = c(4, 0))
-
-## Adding a legend
-legend("topleft", legend = c("Central tendency",
-                             "Confidence interval 1",
-                            "Confidence interval 2"),
-      col = c("blue", "orange", "green"), pch = 19)
-
-## Resetting graphical parameters
-par(op)
-```
-
-In addition to the classic `plot` arguments, the function can also take arguments that are specific to `plot.dispRity` like adding the number of elements or rarefaction level (as described above), and also changing the values of the quantiles to plot as well as the central tendency.
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the results with some plot.dispRity arguments
-plot(disparity_time_slices, quantile = c(seq(from = 10, to = 100, by = 10)),
-    cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
-    ylab = c("Disparity", "Diversity"), time.subsets = FALSE,
-    xlab = "Time (in in units from past to present)", observed = TRUE,
-    main = "Many more options...")
-
-## Resetting graphical parameters
-par(op)
-```
-
-> Note that the argument `observed = TRUE` allows to plot the disparity values calculated from the non-bootstrapped data as crosses on the plot.
-
-For comparing results, it is also possible to add a plot to the existent plot by using `add = TRUE`:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the continuous disparity with a fixed y axis
-plot(disparity_time_slices, ylim = c(1, 4))
-## Adding the discrete data
-plot(disparity_time_bins, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
-    add = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-Finally, if your data has been fully rarefied, it is also possible to easily look at rarefaction curves by using the `rarefaction = TRUE` argument:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical options
-op <- par(bty = "n")
-
-## Plotting the rarefaction curves
-plot(disparity_crown_stem, rarefaction = TRUE)
-
-## Resetting graphical parameters
-par(op)
-```
-
-## Testing disparity hypotheses
-
-The `dispRity` package allows users to apply statistical tests to the calculated disparity to test various hypotheses.
-The function `test.dispRity` works in a similar way to the `dispRity` function: it takes a `dispRity` object, a `test` and a `comparisons` argument.
-
-The `comparisons` argument indicates the way the test should be applied to the data:
-
- * `pairwise` (default): to compare each subset in a pairwise manner
- * `referential`: to compare each subset to the first subset
- * `sequential`: to compare each subset to the following subset
- * `all`: to compare all the subsets together (like in analysis of variance)
-
-It is also possible to input a list of pairs of `numeric` values or `characters` matching the subset names to create personalised tests.
-Some other tests implemented in `dispRity` such as the `dispRity::null.test` have a specific way they are applied to the data and therefore ignore the `comparisons` argument. 
-<!-- Add sequential test one day! -->
-
-The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
-It can be a common statistical test function (e.g. `stats::t.test`), a function implemented in `dispRity` (e.g. see `?null.test`) or any function defined by the user.
-
-This function also allows users to correct for Type I error inflation (false positives) when using multiple comparisons via the `correction` argument. 
-This argument can be empty (no correction applied) or can contain one of the corrections from the `stats::p.adjust` function (see `?p.adjust`).
-
-Note that the `test.dispRity` algorithm deals with some classical test outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
-It is, however, possible to get the full detailed output by using the options `details = TRUE`.
-
-Here we are using the variables generated in the [section above](#summarising-dispRity-data-plots):
-
-```{r}
-## T-test to test for a difference in disparity between crown and stem mammals
-test.dispRity(disparity_crown_stem, test = t.test)
-
-## Performing the same test but with the detailed t.test output
-test.dispRity(disparity_crown_stem, test = t.test, details = TRUE)
-
-## Wilcoxon test applied to time sliced disparity with sequential comparisons,
-## with Bonferroni correction
-test.dispRity(disparity_time_slices, test = wilcox.test,
-              comparisons = "sequential", correction = "bonferroni")
-
-## Measuring the overlap between distributions in the time bins (using the
-## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)
-test.dispRity(disparity_time_bins, test = bhatt.coeff)
-```
-
-It is also possible to apply some more *complex* tests that have their own output classes (like `stats::lm`).
-
-The results can then be analysed as usual using the associated `summary` S3 method:
-
-```{r}
-## Performing and linear model applied to the same data
-(slice_lm <- test.dispRity(disparity_time_slices, test = lm,
-                            comparisons = "all"))
-
-## The output is a regular `lm` output
-class(slice_lm)
-
-## This output can be summarised using summary
-summary(slice_lm)
-```
-
-Of course, due to the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
-The only condition is that the test can be applied to at least two distributions.
-In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm` for example).
-
-## Disparity as a distribution
-
-Disparity is often regarded as a summary value of the position of the all elements in the ordinated space.
-For example, the sum of variances, the product of ranges or the median distance between the elements and their centroid will summarise disparity as a single value.
-This value can be pseudo-replicated (bootstrapped) to obtain a distribution of the summary metric with estimated error.
-However, another way to perform disparity analysis is to use the *whole distribution* rather than just a summary metric (e.g. the variances or the ranges).
-
-This is possible in the `dispRity` package by calculating disparity as a dimension-level 2 metric only!
-Let's have a look using our [previous example](#summarising-dispRity-data-plots) of bootstrapped time slices but by measuring the distances between each taxon and their centroid as disparity.
-
-```{r}
-## Measuring disparity as a whole distribution
-disparity_centroids <- dispRity(boot_time_slices, metric = centroids)
-```
-
-The resulting disparity object is of dimension-level 2, so it can easily be transformed into a dimension-level 1 object by, for example, measuring the median distance of all these distributions:
-
-```{r}
-## Measuring median disparity in each time slice
-disparity_centroids_median <- dispRity(disparity_centroids, metric = median)
-```
-
-And we can now compare the differences between these methods:
-
-```{r}
-## Summarising both disparity measurements:
-## The distributions:
-summary(disparity_centroids)
-## The summary of the distributions (as median)
-summary(disparity_centroids_median)
-```
-
-We can see that the summary message for the distribution is slightly different than before.
-Here `summary` also displays the observed central tendency (i.e. the central tendency of the measured distributions).
-Note that, as expected, this central tendency is the same in both metrics!
-
-Another, maybe more intuitive way, to compare both approaches for measuring disparity is to plot the distributions:
-
-```{r, fig.width=8, fig.height=8}
-## Graphical parameters
-op <- par(bty = "n", mfrow = c(1, 2))
-
-## Plotting both disparity measurements
-plot(disparity_centroids, ylab = "Distribution of all the distances")
-plot(disparity_centroids_median,
-     ylab = "Distribution of the medians of all the distances")
-
-par(op)
-```
-
-We can then test for differences in the resulting distributions using `test.dispRity` and the `bhatt.coeff` test as described above.
-
-```{r}
-## Probability of overlap in the distribution of medians
-test.dispRity(disparity_centroids_median, test = bhatt.coeff)
-```
-
-In this case, we are looking at the probability of overlap of the distribution of median distances from centroids among each pair of time slices.
-In other words, we are measuring whether the medians from each bootstrap pseudo-replicate for each time slice overlap.
-But of course, we might be interested in the actual distribution of the distances from the centroid rather than simply their central tendencies.
-This can be problematic depending on the research question asked since we are effectively comparing non-independent medians distributions (because of the pseudo-replication).
-
-One solution, therefore, is to look at the full distribution:
-
-```{r}
-## Probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff)
-```
-
-These results show the actual overlap among all the measured distances from centroids concatenated across all the bootstraps.
-For example, when comparing the slices 120 and 80, we are effectively comparing the 5 $\times$ 100 distances (the distances of the five elements in slice 120 bootstrapped 100 times) to the 19 $\times$ 100 distances from slice 80.
-However, this can also be problematic for some specific tests since the _n_ $\times$ 100 distances are also pseudo-replicates and thus are still not independent.
-
-A second solution is to compare the distributions to each other _for each replicate_:
-
-```{r}
-## Boostrapped probability of overlap for the full distributions
-test.dispRity(disparity_centroids, test = bhatt.coeff, concatenate = FALSE)
-```
-
-These results show the median overlap among pairs of distributions in the first column (`bhatt.coeff`) and then the distribution of these overlaps among each pair of bootstraps.
-In other words, when two distributions are compared, they are now compared for each bootstrap pseudo-replicate, thus effectively creating a distribution of probabilities of overlap.
-For example, when comparing the slices 120 and 80, we have a mean probability of overlap of 0.28 and a probability between 0.18 and 0.43 in 50\% of the pseudo-replicates.
-Note that the quantiles and central tendencies can be modified via the `conc.quantiles` option.
-
-## Disparity from other matrices
-
-In the example so far, disparity was measured from an ordinated multidimensional space (i.e. a PCO of the distances between taxa based on discrete morphological characters).
-This is a common approach in palaeobiology, morphometrics or ecology but ordinated matrices are not mandatory for the `dispRity` package!
-It is totally possible to perform the same analysis detailed above using other types of matrices as long as your elements are rows in your matrix.
-
-For example, we can use the data set `eurodist`, an `R` inbuilt dataset that contains the distances (in km) between European cities.
-We can check for example, if Northern European cities are closer to each other than Southern ones:
-
-```{r}
-## Making the eurodist data set into a matrix (rather than "dist" object)
-eurodist <- as.matrix(eurodist)
-eurodist[1:5, 1:5]
-
-## The two groups of cities
-Northern <- c("Brussels", "Calais", "Cherbourg", "Cologne", "Copenhagen",
-              "Hamburg", "Hook of Holland", "Paris", "Stockholm")
-Southern <- c("Athens", "Barcelona", "Geneva", "Gibraltar", "Lisbon", "Lyons",
-              "Madrid", "Marseilles", "Milan", "Munich", "Rome", "Vienna")
-
-## Creating the subset dispRity object
-eurodist_subsets <- custom.subsets(eurodist, group = list("Northern" = Northern,
-                                                        "Southern" = Southern))
-
-## Bootstrapping and rarefying to 9 elements (the number of Northern cities)
-eurodist_bs <- boot.matrix(eurodist_subsets, rarefaction = 9)
-
-## Measuring disparity as the median distance from group's centroid
-euro_disp <- dispRity(eurodist_bs, metric = c(median, centroids))
-
-## Testing the differences using a simple wilcox.test
-euro_diff <- test.dispRity(euro_disp, test = wilcox.test)
-euro_diff_rar <- test.dispRity(euro_disp, test = wilcox.test, rarefaction = 9)
-```
-
-We can compare this approach to an ordination one:
-
-```{r}
-## Ordinating the eurodist matrix
-euro_ord <- cmdscale(eurodist, k = nrow(eurodist) - 2)
-
-## Calculating disparity on the bootstrapped and rarefied subset data
-euro_ord_disp <- dispRity(boot.matrix(custom.subsets(euro_ord, group =
-        list("Northern" = Northern, "Southern" = Southern)), rarefaction = 9),
-        metric = c(median, centroids))
-
-## Testing the differences using a simple wilcox.test
-euro_ord_diff <- test.dispRity(euro_ord_disp, test = wilcox.test)
-euro_ord_diff_rar <- test.dispRity(euro_ord_disp, test = wilcox.test, rarefaction = 9)
-```
-
-And visualise the differences:
-
-```{r, fig.width=12, fig.height=12}
-## Plotting the differences
-par(mfrow = c(2,2), bty = "n")
-## Plotting the normal disparity
-plot(euro_disp, main = "Distance differences")
-## Adding the p-value
-text(1.5, 4000, paste0("p=",round(euro_diff[[2]][[1]], digit = 5)))
-## Plotting the rarefied disparity
-plot(euro_disp, rarefaction = 9, main = "Distance differences (rarefied)")
-## Adding the p-value
-text(1.5, 4000, paste0("p=",round(euro_diff_rar[[2]][[1]], digit = 5)))
-
-## Plotting the ordinated disparity
-plot(euro_ord_disp, main = "Ordinated differences")
-## Adding the p-value
-text(1.5, 1400, paste0("p=",round(euro_ord_diff[[2]][[1]], digit = 5) ))
-## Plotting the rarefied disparity
-plot(euro_ord_disp, rarefaction = 9, main = "Ordinated differences (rarefied)")
-## Adding the p-value
-text(1.5, 1400, paste0("p=",round(euro_ord_diff_rar[[2]][[1]], digit = 5) ))
-
-```
-
-As expected, the results are pretty similar in pattern but different in terms of scale.
-The median centroids distance is expressed in km in the "Distance differences" plots and in Euclidean units of variation in the "Ordinated differences" plots.
-
-<!--chapter:end:03_specific-tutorials.Rmd-->
-
----
-title: "Data simulation features"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Making stuff up!
-
-The `dispRity` package also offers some advanced data simulation features to allow to test hypothesis, explore ordinate-spaces or metrics properties or simply playing around with data!
-All the following functions are based on the same modular architecture of the package and therefore can be used with most of the functions of the package.
-
-## Simulating discrete morphological data
-
-The function `sim.morpho` allows to simulate discrete morphological data matrices (sometimes referred to as "cladistic" matrices).
-It allows to evolve multiple discrete characters on a given phylogenetic trees, given different models, rates, and states.
-It even allows to include proper inapplicable data to make datasets as messy as in real life!
-
-In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
-The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
-The methods are described in details below
-
-```{r}
-set.seed(3)
-## Simulating a starting tree with 15 taxa as a random coalescent tree
-my_tree <- rcoal(15)
-
-## Generating a matrix with 100 characters (85% binary and 15% three state) and
-## an equal rates model with a gamma rate distribution (0.5, 1) with no 
-## invariant characters.
-my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
-    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)
-
-## The first few lines of the matrix
-my_matrix[1:5, 1:10]
-
-## Checking the matrix properties with a quick Maximum Parsimony tree search
-check.morpho(my_matrix, my_tree)
-```
-
-Note that this example produces a tree with a great consistency index and an identical topology to the random coalescent tree!
-Nearly too good to be true...
-
-### A more detailed description
-
-The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;OReilly2017].
-
- * The first `tree` argument will be the tree on which to "evolve" the characters and therefore requires branch length.
- You can generate quick and easy random Yule trees using `ape::rtree(number_of_taxa)` but I would advise to use more realistic trees for more realistic simulations based on more realistic models (really realistic then) using the function `tree.bd` from the [`diversitree`](http://www.zoology.ubc.ca/prog/diversitree/) package [@fitzjohndiversitree2012].
- * The second argument, `character` is the number of characters. Pretty straight forward.
- * The third, `states` is the proportion of characters states above two (yes, the minimum number of states is two).
-This argument intakes the proportion of *n*-states characters, for example `states = c(0.5,0.3,0.2)` will generate 50% of binary-state characters, 30% of three-state characters and 20% of four-state characters.
-There is no limit in the number of state characters proportion as long as the total makes up 100%.
- * The forth, `model` is the evolutionary model for generating the character(s). More about this below.
- * The fifth and sixth, `rates` and `substitution` are the model parameters described below as well.
- * Finally, the two logical arguments, are self explanatory: `invariant` whether to allow invariant characters (i.e.
-characters that don't change) and `verbose` whether to print the simulation progress on your console.
-
-#### Available evolutionary models
-
-There are currently three evolutionary models implemented in `sim.morpho` but more will come in the future.
-Note also that they allow fine tuning parameters making them pretty plastic!
-
- * `"ER"`: this model allows any number of character states and is based on the Mk model [@lewisa2001].
-It assumes a unique overall evolutionary rate equal substitution rate between character states.
-This model is based on the `ape::rTraitDisc` function.
- * `"HKY"`: this is binary state character model based on the molecular HKY model [@HKY85].
-It uses the four molecular states (A,C,G,T) with a unique overall evolutionary rate and a biased substitution rate towards transitions (A <-> G or C <-> T) against transvertions (A <-> C and G <-> T).
-After evolving the nucleotide, this model transforms them into binary states by converting the purines (A and G) into state 0 and the pyrimidines (C and T) into state 1.
-This method is based on the `phyclust::seq.gen.HKY` function and was first proposed by @OReilly2016.
- * `"MIXED"`: this model uses a random (uniform) mix between both the `"ER"` and the `"HKY"` models.
-
-The models can take the following parameters:
-(1) `rates` is the evolutionary rate (i.e. the rate of changes along a branch: the evolutionary speed) and (2) `substitution` is the frequency of changes between one state or another.
-For example if a character can have high probability of changing (the *evolutionary* rate) with, each time a change occurs a probability of changing from state *X* to state *Y* (the *substitution* rate).
-Note that in the `"ER"` model, the substitution rate is ignore because... by definition this (substitution) rate is equal!
-
-The parameters arguments `rates` and `substitution` takes a distributions from which to draw the parameters values for each character.
-For example, if you want an `"HKY"` model with an evolutionary rate (i.e.
-speed) drawn from a uniform distribution bounded between 0.001 and 0.005, you can define it as `rates = c(runif, min = 0.001, max = 0.005)`, `runif` being the function for random draws from a uniform distribution and `max` and `min` being the distribution parameters.
-These distributions should always be passed in the format `c(random_distribution_function, distribution_parameters)` with the names of the distribution parameters arguments.
-
-#### Checking the results
-
-An additional function, `check.morpho` runs a quick Maximum Parsimony tree search using the `phangorn` parsimony algorithm.
-It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g.
-the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree to determine how different they are.
-
-#### Adding inapplicable characters
-
-Once a matrix is generated, it is possible to apply inapplicable characters to it for increasing realism!
-Inapplicable characters are commonly designated as `NA` or simply `-`.
-They differ from missing characters `?` in their nature by being inapplicable rather than unknown.
-For example, considering a binary character defined as "colour of the tail" with the following states "blue" and "red"; on a taxa with no tail, the character should be coded as inapplicable ("`-`") since the state of the character "colour of tail" is *known*: it's neither "blue" or "red", it's just not there!
-It contrasts with coding it as missing ("`?`" - also called as ambiguous) where the state is *unknown*, for example, the taxon of interest is a fossil where the tail has no colour preserved or is not present at all due to bad conservation!
-
-This type of characters can be added to the simulated matrices using the `apply.NA` function/
-It takes, as arguments, the `matrix`, the source of inapplicability (`NAs` - more below), the `tree` used to generate the matrix and the two same `invariant` and `verbose` arguments as defined above.
-The `NAs` argument allows two types of sources of inapplicability:
-
- * `"character"` where the inapplicability is due to the character (e.g. coding a character tail for species with no tail).
- In practice, the algorithm chooses a character *X* as the underlying character (e.g.
-"presence and absence of tail"), arbitrarily chooses one of the states as "absent" (e.g.
-0 = absent) and changes in the next character *Y* any state next to character *X* state 0 into an inapplicable token ("`-`").
- This simulates the inapplicability induced by coding the characters (i.e. not always biological).
- * `"clade"` where the inapplicability is due to evolutionary history (e.g. a clade loosing its tail).
- In practice, the algorithm chooses a random clade in the tree and a random character *Z* and replaces the state of the taxa present in the clade by the inapplicable token ("`-`").
- This simulates the inapplicability induced by evolutionary biology (e.g. the lose of a feature in a clade).
-
-To apply these sources of inapplicability, simply repeat the number of inapplicable sources for the desired number of characters with inapplicable data.
-
-```{r}
-## Generating 5 "character" NAs and 10 "clade" NAs
-my_matrix_NA <- apply.NA(my_matrix, tree = my_tree,
-                         NAs = c(rep("character", 5), rep("clade", 10)))
-
-## The first few lines of the resulting matrix
-my_matrix_NA[1:10, 90:100]
-```
-<!--TG: Add inapplicable paper here -->
-
-### Parameters for a realistic(ish) matrix
-
-There are many parameters that can create a "realistic" matrix (i.e.
-not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all parameters combination end up creating "good" matrices.
-The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.
-
-```{r}
-set.seed(0)
-## tree
-my_tree <- rcoal(15)
-## matrix
-morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
-    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
-check.morpho(morpho_mat, my_tree)
-```
-
-<!-- ## Simulating continuous morphological data -->
-<!-- TG: Add tutorial for evolving "morpho" shapes -->
-
-## Simulating multidimensional spaces
-Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
-This function allows users to simulate multidimensional spaces with a certain number of properties.
-For example, it is possible to design a multidimensional space with a specific distribution on each axis, a correlation between the axes and a specific cumulative variance per axis.
-This can be useful for creating ordinated spaces for null hypothesis, for example if you're using the function `null.test` [@diaz2016global].
-
-This function takes as arguments the number of elements (data points - `elements` argument) and dimensions (`dimensions` argument) to create the space and the distribution functions to be used for each axis.
-The distributions are passed through the `distribution` argument as... modular functions!
-You can either pass a single distribution function for all the axes (for example `distribution = runif` for all the axis being uniform) or a specific distribution function for each specific axis (for example `distribution = c(runif, rnorm, rgamma))` for the first axis being uniform, the second normal and the third gamma).
-You can of course use your very own functions or use the ones implemented in `dispRity` for more complex ones (see below).
-Specific optional arguments for each of these distributions can be passed as a list via the `arguments` argument.
-
-Furthermore, it is possible to add a correlation matrix to add a correlation between the axis via the `cor.matrix` argument or even a vector of proportion of variance to be bear by each axis via the `scree` argument to simulate realistic ordinated spaces.
-
-Here is a simple two dimensional example:
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(bty = "n")
-
-## A square space
-square_space <- space.maker(100, 2, runif)
-
-## The resulting 2D matrix
-head(square_space)
-
-## Visualising the space
-plot(square_space, pch = 20, xlab = "", ylab = "", main = "Uniform 2D space")
-```
-
-Of course, more complex spaces can be created by changing the distributions, their arguments or adding a correlation matrix or a cumulative variance vector:
-
-```{r}
-## A plane space: uniform with one dimensions equal to 0
-plane_space <- space.maker(2500, 3, c(runif, runif, runif),
-                           arguments = list(list(min = 0, max = 0), NULL, NULL))
-
-## Correlation matrix for a 3D space
-(cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3))
-
-## An ellipsoid space (normal space with correlation)
-ellipse_space <- space.maker(2500, 3, rnorm, cor.matrix = cor_matrix)
-
-## A cylindrical space with decreasing axes variance
-cylindrical_space <- space.maker(2500, 3, c(rnorm, rnorm, runif),
-                                 scree = c(0.7, 0.2, 0.1))
-```
-
-### Personalised dimensions distributions
-
-Following the modular architecture of the package, it is of course possible to pass home made distribution functions to the `distribution` argument.
-For example, the `random.circle` function is a personalised one implemented in `dispRity`.
-This function allows to create circles based on basic trigonometry allowing to axis to covary to produce circle coordinates.
-By default, this function generates two sets of coordinates with a `distribution` argument and a minimum and maximum boundary (`inner` and `outer` respectively) to create nice sharp edges to the circle.
-The maximum boundary is equivalent to the radius of the circle (it removes coordinates beyond the circle radius) and the minimum is equivalent to the radius of a smaller circle with no data (it removes coordinates below this inner circle radius).
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(bty = "n")
-
-## Generating coordinates for a normal circle with a upper boundary of 1
-circle <- random.circle(1000, rnorm, inner = 0, outer = 1)
-
-## Plotting the circle
-plot(circle, xlab = "x", ylab = "y", main = "A normal circle")
-
-## Creating doughnut space (a spherical space with a hole)
-doughnut_space <- space.maker(5000, 3, c(rnorm, random.circle),
-     arguments = list(list(mean = 0), list(runif, inner = 0.5, outer = 1)))
-```
-
-### Visualising the space
-
-I suggest using the excellent `scatterplot3d` package to play around and visualise the simulated spaces:
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Graphical options
-op <- par(mfrow = (c(2, 2)), bty = "n")
-## Visualising 3D spaces
-require(scatterplot3d)
-
-## The plane space
-scatterplot3d(plane_space, pch = 20, xlab = "", ylab = "", zlab = "",
-              xlim = c(-0.5, 0.5), main = "Plane space")
-
-## The ellipsoid space
-scatterplot3d(ellipse_space, pch = 20, xlab = "", ylab = "", zlab = "",
-              main = "Normal ellipsoid space")
-
-## A cylindrical space with a decreasing variance per axis
-scatterplot3d(cylindrical_space, pch = 20, xlab = "", ylab = "", zlab = "",
-              main = "Normal cylindrical space")
-## Axes have different orders of magnitude
-
-## Plotting the doughnut space
-scatterplot3d(doughnut_space[,c(2,1,3)], pch = 20, xlab = "", ylab = "",
-              zlab = "", main = "Doughnut space")
-par(op)
-```
-
-### Generating realistic spaces
-
-It is possible to generate "realistic" spaces by simply extracting the parameters of an existing space and scaling it up to the simulated space.
-For example, we can extract the parameters of the `BeckLee_mat50` ordinated space and simulate a similar space.
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Loading the data
-data(BeckLee_mat50)
-
-## Number of dimensions
-obs_dim <- ncol(BeckLee_mat50)
-
-## Observed correlation between the dimensions
-obs_correlations <- cor(BeckLee_mat50)
-
-## Observed mean and standard deviation per axis
-obs_mu_sd_axis <- mapply(function(x,y) list("mean" = x, "sd" = y),
-                         as.list(apply(BeckLee_mat50, 2, mean)),
-                         as.list(apply(BeckLee_mat50, 2, sd)), SIMPLIFY = FALSE)
-
-## Observed overall mean and standard deviation
-obs_mu_sd_glob <- list("mean" = mean(BeckLee_mat50), "sd" = sd(BeckLee_mat50))
-
-## Scaled observed variance per axis (scree plot)
-obs_scree <- variances(BeckLee_mat50)/sum(variances(BeckLee_mat50))
-
-## Generating our simulated space
-simulated_space <- space.maker(1000, dimensions = obs_dim, 
-                               distribution = rep(list(rnorm), obs_dim),
-                               arguments = obs_mu_sd_axis,
-                               cor.matrix = obs_correlations)
-
-## Visualising the fit of our data in the space (in the two first dimensions)
-plot(simulated_space[,1:2], xlab = "PC1", ylab = "PC2")
-points(BeckLee_mat50[,1:2], col = "red", pch = 20)
-legend("topleft", legend = c("observed", "simulated"),
-        pch = c(20,21), col = c("red", "black"))
-```
-
-It is now possible to simulate a space using these observed arguments to test several hypothesis:
-
- * Is the space uniform or normal?
- * If the space is normal, is the mean and variance global or specific for each axis?
-
-```{r, fig.width=8, fig.height=8, eval=TRUE}
-## Measuring disparity as the sum of variance
-observed_disp <- dispRity(BeckLee_mat50, metric = c(median, centroids))
-
-## Is the space uniform?
-test_unif <- null.test(observed_disp, null.distrib = runif)
-
-## Is the space normal with a mean of 0 and a sd of 1?
-test_norm1 <- null.test(observed_disp, null.distrib = rnorm)
-
-## Is the space normal with the observed mean and sd and cumulative variance
-test_norm2 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
-                        null.args = rep(list(obs_mu_sd_glob), obs_dim),
-                        null.scree = obs_scree)
-
-## Is the space multiple normal with multiple means and sds and a correlation?
-test_norm3 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
-                        null.args = obs_mu_sd_axis, null.cor = obs_correlations)
-
-## Graphical options
-op <- par(mfrow = (c(2, 2)), bty = "n")
-## Plotting the results
-plot(test_unif, main = "Uniform (0,1)")
-plot(test_norm1, main = "Normal (0,1)")
-plot(test_norm2, main = paste0("Normal (", round(obs_mu_sd_glob[[1]], digit = 3),
-                              ",", round(obs_mu_sd_glob[[2]], digit = 3), ")"))
-plot(test_norm3, main = "Normal (variable + correlation)")
-
-```
-
-If we measure disparity as the median distance from the morphospace centroid, we can explain the distribution of the data as normal with the variable observed mean and standard deviation and with a correlation between the dimensions.
-
-<!--chapter:end:04_making-up-data.Rmd-->
-
----
-title: "The guts of the dispRity package"
-author: "Thomas Guillerme (guillert@tcd.ie)"
-date: "`r Sys.Date()`"
-output:
-  html_document: default
-  pdf_document: default
----
-
-# The guts of the `dispRity` package
-
-## Manipulating `dispRity` objects
-
-Disparity analysis involves a lot of manipulation of many matrices (especially when bootstrapping) which can be impractical to visualise and will quickly overwhelm your `R` console.
-Even the simple Beck and Lee 2014 example above produces an object with > 72 lines of lists of lists of matrices!
-
-Therefore `dispRity` uses a specific class of object called a `dispRity` object.
-These objects allow users to use S3 method functions such as `summary.dispRity`, `plot.dispRity` and `print.dispRity`. 
-`dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g. `sort.dispRity`, `extract.dispRity` see the full list in the next section).
-These functions modify the `dispRity` object without having to delve into its complex structure!
-The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).
-
-```{r}
-## Loading the example data
-data(disparity)
-
-## What is the class of the median_centroids object?
-class(disparity)
-
-## What does the object contain?
-names(disparity)
-
-## Summarising it using the S3 method print.dispRity
-disparity
-```
-
-Note that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:
-
-```{r, eval=FALSE}
-## Display the full object
-print(disparity, all = TRUE)
-## This is more nearly ~ 5000 lines on my 13 inch laptop screen!
-```
-
-## `dispRity` utilities
-
-The package also provides some utility functions to facilitate multidimensional analysis.
-
-### `dispRity` object utilities  <a name="dispRity.utilities"></a>
-The first set of utilities are functions for manipulating `dispRity` objects:
-
-#### `make.dispRity`
-This function creates empty `dispRity` objects.
-
-```{r}
-## Creating an empty dispRity object
-make.dispRity()
-
-## Creating an "empty" dispRity object with a matrix
-(disparity_obj <- make.dispRity(matrix(rnorm(20), 5, 4)))
-```
-
-####  `fill.dispRity`
-This function initialises a `dispRity` object and generates its call properties.
-
-```{r}
-## The dispRity object's call is indeed empty
-disparity_obj$call
-
-## Filling an empty disparity object (that needs to contain at least a matrix)
-(disparity_obj <- fill.dispRity(disparity_obj))
-
-## The dipRity object has now the correct minimal attributes
-disparity_obj$call
-```
-
-####  `matrix.dispRity`
-This function extracts a specific matrix from a disparity object.
-The matrix can be one of the bootstrapped matrices or/and a rarefied matrix.
-
-```{r}
-## Extracting the matrix containing the coordinates of the elements at time 50
-str(matrix.dispRity(disparity, "50"))
-
-## Extracting the 3rd bootstrapped matrix with the 2nd rarefaction level
-## (15 elements) from the second group (80 Mya)
-str(matrix.dispRity(disparity, subsets = 1, bootstrap = 3, rarefaction = 2))
-```
-
-#### `get.subsets.dispRity`
-This function creates a dispRity object that contains only elements from one specific subsets.
-
-```{r, eval=FALSE}
-## Extracting all the data for the crown mammals
-(crown_mammals <- get.subsets.dispRity(disp_crown_stemBS, "Group.crown"))
-
-## The object keeps the properties of the parent object but is composed of only one subsets
-length(crown_mammals$subsets)
-```
-
-#### `extract.dispRity`
-This function extracts the calculated disparity values of a specific matrix.
-
-```{r, eval=FALSE}
-## Extracting the observed disparity (default)
-extract.dispRity(disparity)
-
-## Extracting the disparity from the bootstrapped values from the
-## 10th rarefaction level from the second subsets (80 Mya)
-extract.dispRity(disparity, observed = FALSE, subsets = 2, rarefaction = 10)
-```
-
-#### `scale.dispRity`
-This is the S3 method of `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object.
-
-```{r, eval=FALSE}
-## Getting the disparity values of the time subsets
-head(summary(disparity))
-
-## Scaling the same disparity values
-head(summary(scale(disparity, scale = TRUE)))
-
-## Scaling and centering:
-head(summary(scale(disparity, scale = TRUE, center = TRUE)))
-```
-
-#### `sort.dispRity`
-This is the S3 method of `sort` for sorting the subsets alphabetically (default) or following a specific pattern.
-
-```{r, eval=FALSE}
-## Sorting the disparity subsets in inverse alphabetic order
-head(summary(sort(disparity, decreasing = TRUE)))
-
-## Customised sorting
-head(summary(sort(disparity, sort = c(7, 1, 3, 4, 5, 2, 6))))
-```
-
-## The `dispRity` object content
-The functions above are utilities to easily and safely access different elements in the `dispRity` object.
-Alternatively, of course, each elements can be accessed manually.
-Here is an explanation on how it works.
-The `dispRity` object is a `list` of two to four elements, each of which are detailed below:
-
- * `$matrix`: an object of class `matrix`, the full multidimensional space.
- * `$call`: an object of class `list` containing information on the `dispRity` object content.
- * `$subsets`: an object of class `list` containing the subsets of the multidimensional space.
- * `$disparity`: an object of class `list` containing the disparity values.
-
-The `dispRity` object is loosely based on `C` structure objects.
-In fact, it is composed of one unique instance of a matrix (the multidimensional space) upon which the metric function is called via "pointers" to only a certain number of elements and/or dimensions of this matrix.
-This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint:  at any time, only one matrix is present in the `R` environment rather than multiple copies of it for each subset.
-
-### `$matrix`
-This is the multidimensional space, stored in the `R` environment as a `matrix` object.
-It requires row names but not column names.
-By default, if the row names are missing, `dispRity` function will arbitrarily generate them in numeric order (i.e. `rownames(matrix) <- 1:nrow(matrix)`).
-This element of the `dispRity` object is never modified.
-
-### `$call`
-This element contains the information on the `dispRity` object content.
-It is a `list` that can contain the following:
-
- * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"punctuated"`, `"gradual"`).
-This element generated only once via `time.subsets()` and `custom.subsets()`.
- * `$call$dimensions`: either a single `numeric` value indicating how many dimensions to use or a vector of `numeric` values indicating which specific dimensions to use.
-This element is by default the number of columns in `$matrix` but can be modified through `boot.matrix()` or `dispRity()`.
- * `$call$bootstrap`: this is a `list` containing three elements:
-    * `[[1]]`: the number of bootstrap replicates (`numeric`)
-    * `[[2]]`: the bootstrap method (`character`)
-    * `[[3]]`: the rarefaction levels (`numeric` vector)
- * `$call$disparity`: this is a `list` containing one element, `$metric`, that is a `list` containing the different functions passed to the `metric` argument in `dispRity`.
-These are `call` elements and get modified each time the `dispRity` function is used (the first element is the first metric(s), the second, the second metric(s), etc.).
-
-### `$subsets`
-This element contain the eventual subsets of the multidimensional space.
-It is a `list` of subset names.
-Each subset name is in turn a `list` of at least one element called `elements` which is in turn a `matrix`.
-This `elements` matrix is the raw (observed) elements in the subsets.
-The `elements` matrix is composed of `numeric` values in one column and _n_ rows (the number of elements in the subset).
-Each of these values are a "pointer" (`C` inspired) to the element of the `$matrix`.
-For example, lets assume a `dispRity` object called `disparity`, composed of at least one subsets called `sub1`:
-
-```
- disparity$subsets$sub1$elements
-      [,1]
- [1,]    5
- [2,]    4
- [3,]    6
- [4,]    7
-```
-
-The values in the matrix "point" to the elements in `$matrix`: here, the multidimensional space with only the 4th, 5th, 6th and 7th elements.
-The following elements in `diparity$subsets$sub1` will correspond to the same "pointers" but drawn from the bootstrap replicates.
-The columns will correspond to different bootstrap replicates.
-For example:
-
-```
- disparity$subsets$sub1[[2]]
-      [,1] [,2] [,3] [,4]
- [1,]   57   43   70    4
- [2,]   43   44    4    4
- [3,]   42   84   44    1
- [4,]   84    7    2   10
-```
-This signifies that we have four bootstrap pseudo-replicates pointing each time to four elements in `$matrix`.
-The next element (`[[3]]`) will be the same for the eventual first rarefaction level (i.e.
-the resulting bootstrap matrix will have _m_ rows where _m_ is the number of elements for this rarefaction level).
-The next element after that (`[[4]]`) will be the same for with an other rarefaction level and so forth...
-
-### `$disparity`
-The `$disparity` element is identical to the `$subsets` element structure (a list of list(s) containing matrices) but the matrices don't contain "pointers" to `$matrix` but the disparity result of the disparity metric applied to the "pointers".
-For example, in our first example (`$elements`) from above, if the disparity metric is of dimensions level 1, we would have:
-
-```
- disparity$disparity$sub1$elements
-      [,1]
- [1,]    1.82
-```
-
-This is the observed disparity (1.82) for the subset called `sub1`.
-If the disparity metric is of dimension level 2 (say the function `range` that outputs two values), we would have:
-
-```
- disparity$disparity$sub1$elements
-      [,1]
- [1,]    0.82
- [2,]    2.82
-```
-
-The following elements in the list follow the same logic as before: rows are disparity values (one row for a dimension level 1 metric, multiple for a dimensions level 2 metric) and columns are the bootstrap replicates (the bootstrap with all elements followed by the eventual rarefaction levels).
-For example for the bootstrap without rarefaction (second element of the list):
-
-```
- disparity$disparity$sub1[[2]]
-         [,1]     [,2]     [,3]     [,4]
-[1,] 1.744668 1.777418 1.781624 1.739679 
-```
-
-<!-- ## Modularity
-
-### Making your own metrics
-
-### Making your own tests
-
-## Where is this going?
- -->
-<!-- ## Running `dispRity` in parallel
-
-The computationally intensive function `dispRity` have a `parallel` option to speed up their calculations.
-
-This option requires the package `snow` and takes arguments that are to be passed to  -->
-
-<!--chapter:end:05_guts.Rmd-->
-
 ---
@@
 
-This demo aims to give quick overview of the `dispRity` package (v.0.4) for palaeobiology analyses of disparity, including disparity through time analyses.
+This demo aims to give quick overview of the `dispRity` package (v.`r version_release`) for palaeobiology analyses of disparity, including disparity through time analyses.
 
-This demo showcases a typical disparity-through-time analysis: we are going to test whether the disparity changed through time in a subset of eutherian mammals from the last 100 million years using a dataset from Beck and Lee (2014).
+This demo showcases a typical disparity-through-time analysis: we are going to test whether the disparity changed through time in a subset of eutherian mammals from the last 100 million years using a dataset from @beckancient2014.
 
@@
 See the [example data](#example-data) description for more details.
-Briefly, this dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
+Briefly, this dataset contains an ordinated matrix of the Gower distance between 50 mammals based (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
 The ordinated matrix will represent our full morphospace, i.e. all the mammalian morphologies that ever existed through time (for this dataset).
@@
 data(BeckLee_mat50)
+data(BeckLee_mat99)
 head(BeckLee_mat50[,1:5])
@@
 
+### Setting up your own data
+
+I greatly encourage you to follow along this tutorial with your very own data: it is more exciting and, ultimately, that's probably your objective.
+
+> What data can I use?
+
+You can use any type of morphospace in any dataset form (`"matrix"`, `"data.frame"`).
+Throughout this tutorial, you we assume you are using the (loose) morphospace definition from @Guillerme2020: any matrix were columns are traits and rows are observations (in a distance matrix, columns are still trait, i.e.
+"distance to species A", etc.).
+We won't cover it here but you can also use lists of matrices and list of trees.
+
+> How should I format my data for this tutorial?
+
+To go through this tutorial you will need:
+
+ * A matrix with tip data
+ * A phylogenetic tree 
+ * A matrix with tip and node data
+ * A table of first and last occurrences data (FADLAD)
+
+If you are missing any of these, fear not, here are a couple of functions to simulate the missing data, it will surely make your results look funky but it'll let you go through the tutorial.
+
+```{r}
+## Functions to get simulate a PCO looking like matrix from a tree
+i.need.a.matrix <- function(tree) {
+    matrix <- space.maker(elements = Ntip(tree), dimensions = Ntip(tree), distribution = rnorm,
+                          scree = rev(cumsum(rep(1/Ntip(tree), Ntip(tree)))))
+    rownames(matrix) <- tree$tip.label
+    return(matrix)
+}
+
+## Function to simulate a tree
+i.need.a.tree <- function(matrix) {
+    tree <- rtree(nrow(matrix))
+    tree$root.time <- max(tree.age(tree)$age)
+    tree$tip.label <- rownames(matrix)
+    tree$node.label <- paste0("n", 1:(nrow(matrix)-1))
+    return(tree)
+}
+
+## Function to simulate some "node" data
+i.need.node.data <- function(matrix, tree) {
+    matrix_node <- space.maker(elements = Nnode(tree), dimensions = ncol(matrix),
+                               distribution = rnorm, scree = apply(matrix, 2, var))
+    if(!is.null(tree$node.label)) {
+        rownames(matrix_node) <- tree$node.label
+    } else {
+        rownames(matrix_node) <- paste0("n", 1:(nrow(matrix)-1))
+    }
+    return(rbind(matrix, matrix_node))
+}
+
+## Function to simulate some "FADLAD" data
+i.need.FADLAD <- function(tree) {
+    tree_ages <- tree.age(tree)[1:Ntip(tree),]
+    return(data.frame(FAD = tree_ages[,1], LAD = tree_ages[,1], row.names = tree_ages[,2]))
+}
+```
+
+You can use these functions for the generating the data you need. For example
+
+```{r, eval = FALSE}
+## Aaaaah I don't have FADLAD data!
+my_FADLAD <- i.need.FADLAD(tree)
+## Sorted.
+```
+
+In the end this is what your data should be named to facilitate the rest of this tutorial (fill in yours here):
+
+```{r}
+## A matrix with tip data
+my_matrix <- BeckLee_mat50
+
+## A phylogenetic tree 
+my_tree <- BeckLee_tree
+
+## A matrix with tip and node data
+my_tip_node_matrix <- BeckLee_mat99
+
+## A table of first and last occurrences data (FADLAD)
+my_fadlad <- BeckLee_ages
+```
+
 ## A disparity-through-time analysis
@@
 
-## Splitting the morphospace using the time.subsets function
-(binned_morphospace <- time.subsets(data = BeckLee_mat50, tree = BeckLee_tree,
+## Splitting the morphospace using the chrono.subsets function
+(binned_morphospace <- chrono.subsets(data = my_matrix, tree = my_tree,
     method = "discrete", time = time_bins, inc.nodes = FALSE,
-    FADLAD = BeckLee_ages))
+    FADLAD = my_fadlad))
 ```
@@
 The output object is a `dispRity` object (see more about that [here](#The-guts-of-the-dispRity-package).
-In brief, however, `dispRity` objects are lists of different elements (i.e.
+In brief, `dispRity` objects are lists of different elements (i.e.
 disparity results, morphospace time subsets, morphospace attributes, etc.) that display only a summary of the object when calling the object to avoiding filling the `R` console with superfluous output.
+It also allows easy plotting/summarising/analysing for repeatability down the line but we will not go into this right now.
 
@@
 
+> Note how information is adding up to the `dispRity` object.
+
 ### Calculating disparity
@@
 Disparity can be calculated in many ways and this package allows users to come up with their own disparity metrics.
-For more details, please refer to the [`dispRity` metric section](#disparity-metrics).
+For more details, please refer to the [`dispRity` metric section](#disparity-metrics) (or directly use [`moms`](https://tguillerme.shinyapps.io/moms/)).
 
-In this example, we are going to calculate the spread of the data in each time subset by calculating disparity as the sum of the variance of each dimension of the morphospace in each time subset using the `dispRity` function.
-Thus, in this example, disparity is defined by the multi-dimensional variance of each time subset (i.e. the spread of the taxa within the morphospace).
-Note that this metric comes with a caveat (not solved here) since it ignores covariances among the dimensions of the morphospace.
-We use this here because it is a standard metric used in disparity-through-time analysis [@Wills1994].
+In this example, we are going to look at how the spread of the data in the morphospace through time.
+For that we are going to use the sum of the variance from each dimension of the morphospace in the morphospace.
+We highly recommend using a metric that makes sense for your specific analysis and for your specific dataset and not just because everyone uses it [@moms, @Guillerme2020]!
+
+> How can I be sure that the metric is the most appropriate for my morphospace and question?
+
+This is not a straightforward question but you can use the `test.metric` function to check your assumptions (more details [here](#disparity-metrics)): basically what `test.metric` does is modifying your morphospace using a null process of interest (e.g.
+changes in size) and checks whether your metric does indeed pick up that change.
+For example here, let see if the sum of variances picks up changes in size but not random changes:
+
+```{r}
+my_test <- test.metric(my_matrix, metric = c(sum, variances), shifts = c("random", "size"))
+summary(my_test)
+plot(my_test)
+```
+
+We see that changes in the inner size (see @moms for more details) is actually picked up by the sum of variances but not random changes or outer changes. Which is a good thing!
+
+As you've noted, the sum of variances is defined in `test.metric` as `c(sum, variances)`.
+This is a core bit of the `dispRity` package were you can define your own metric as a function or a set of functions.
+You can find more info about this in the [`dispRity` metric section](#disparity-metrics) but in brief, the `dispRity` package considers metrics by their "dimensions" level which corresponds to what they output.
+For example, the function `sum` is a dimension level 1 function because no matter the input it outputs a single value (the sum), `variances` on the other hand is a dimension level 2 function because it will output the variance of each column in a matrix (an example of a dimensions level 3 would be the function `var` that outputs a matrix).
+The `dispRity` package always automatically sorts the dimensions levels: it will always run dimensions level 3 > dimensions level 2 > and dimensions level 1.
+In this case both `c(sum, variances)` and `c(variances, sum)` will result in actually running `sum(variances(matrix))`.
+
+Anyways, let's calculate the sum of variances on our bootstrapped and rarefied morphospaces:
 
@@
 
-The `dispRity` function does not actually display the calculated disparity values but rather only the properties of the disparity object (size, subsets, metric, etc.).
 To display the actual calculated scores, we need to summarise the disparity object using the S3 method `summary` that is applied to a `dispRity` object (see `?summary.dispRity` for more details).
-
-As for any `R` package, you can refer to the help files for each individual function for more details.
+By the way, as for any `R` package, you can refer to the help files for each individual function for more details.
 
@@
 ```
-> The summary.dispRity function comes with many options on which values to calculate (central tendency and quantiles) and on how many digits to display.
+
+> The `summary.dispRity` function comes with many options on which values to calculate (central tendency and quantiles) and on how many digits to display.
 Refer to the function's manual for more details.
@@
 
-Testing differences
--------------------
+Nice. The curves look pretty similar.
+
+> Same as for the `summary.dispRity` function, check out the `plot.dispRity` manual for the many, many options available.
+
+### Testing differences
 
@@
 
-<!--chapter:end:06_palaeo-demo.Rmd-->
+## Some more advanced stuff
 
----
-title: "dispRity ecology demo"
-author: "Thomas Guillerme"
-date: "`r Sys.Date()`"
-output: rmarkdown::html_vignette
-bibliography: References.bib
-bst: sysbio.bst
-vignette: >
-  %\VignetteIndexEntry{dispRity ecology demo}
-  %\VignetteEngine{knitr::rmarkdown}
-  \usepackage[utf8]{inputenc}
----
+The previous section detailed some of the basic functionalities in the `dispRity` package but of course, you can do some much more advanced analysis, here is just a list of some specific tutorials from this manual that you might be interested in:
 
-# Ecology demo
+ * [Time slicing](#time-slicing): an alternative method to look at disparity through time that allows you to specify evolutionary models [@time-slice].
+ * [Many more disparity metrics](#disparity-metrics): there are many, many different things you might be interested to measure in your morphospace!
+This manual has some extended documentation on what to use (or check @moms).
+ * Many more ways to look at disparity: you can for example, [use distributions rather than point estimates](#disparity-distribution) for your disparity metric (e.g.
+the variances rather than the sum of variances); or calculate [disparity from non ordinated matrices](#other-matrices) or even [from multiple matrices and trees](#multi.input).
+ * And finally there are much more advanced statistical tests you might be interested in using, such as the [NPMANOVA](#adonis), the ["disparity-through-time test"](#dtt), using a [null model approach](#null-test) or some [model fitting](#model-fitting)...
 
-This is an example of typical disparity analysis that can be performed in ecology.
-
-## Data
-For this example, we will use the famous `iris` inbuilt data set
-
-```{r}
-data(iris)
-```
-
-This data contains petal and sepal length for 150 individual plants sorted into three species.
-
-```{r}
-## Separating the species
-species <- iris[,5]
-## Which species?
-unique(species)
-
-## Separating the petal/sepal length
-measurements <- iris[,1:4]
-head(measurements)
-```
-
-We can then ordinate the data using a PCA (`prcomp` function) thus defining our four dimensional space as the poetically named petal-space.
-
-```{r}
-## Ordinating the data
-ordination <- prcomp(measurements)
-
-## The petal-space
-petal_space <- ordination$x
-
-## Adding the elements names to the petal-space (the individuals IDs)
-rownames(petal_space) <- 1:nrow(petal_space)
-```
-
-## Classic analysis
-A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different species are distributed in the petal-space.
-
-```{r, fig.width=6, fig.height=6}
-## Measuring the variance on each axis
-variances <- apply(petal_space, 2, var)
-variances <- variances/sum(variances)
-
-## Graphical option
-par(bty = "n")
-
-## A classic 2D ordination plot
-plot(petal_space[, 1], petal_space[, 2], col = species,
-    xlab = paste0("PC 1 (", round(variances[1], 2), ")"),
-    ylab = paste0("PC 2 (", round(variances[2], 2), ")"))
-```
-
-This shows the distribution of the different species in the petal-space along the two first axis of variation.
-This is a pretty standard way to visualise the multidimensional space and further analysis might be necessary to test wether the groups are different such as a linear discriminant analysis (LDA).
-However, in this case we are ignoring the two other dimensions of the ordination!
-If we look at the two other axis we see a totally different result:
-
-```{r, fig.width=6, fig.height=6}
-## Plotting the two second axis of the petal-space
-plot(petal_space[, 3], petal_space[, 4], col = species,
-    xlab = paste0("PC 3 (", round(variances[3], 2), ")"),
-    ylab = paste0("PC 4 (", round(variances[4], 2), ")"))
-```
-
-Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e.
-the values on the first dimension do not represent a continuous trait (e.g.
-petal length), instead they just represent the ordinations of correlations between the data and some factors.
-
-Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.
-
-## A multidimensional approach with `dispRity`
-
-The first step is to create different subsets that represent subsets of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
-Each of these subsets will contain only the individuals of a specific species.
-
-```{r}
-## Creating the table that contain the elements and their attributes
-petal_subsets <- custom.subsets(petal_space, group = list(
-                                "setosa" = which(species == "setosa"),
-                                "versicolor" = which(species == "versicolor"),
-                                "virginica" = which(species == "virginica")))
-
-## Visualising the dispRity object content
-petal_subsets
-```
-
-This created a `dispRity` object (more about that [here](#guts)) with three subsets corresponding to each subspecies.
-
-### Bootstrapping the data
-We can the bootstrap the subsets to be able test the robustness of the measured disparity to outliers.
-We can do that using the default options of `boot.matrix` (more about that [here](#bootstraps-and-rarefactions)):
-
-```{r}
-## Bootstrapping the data
-(petal_bootstrapped <- boot.matrix(petal_subsets))
-```
-
-### Calculating disparity
-
-Disparity can be calculated in many ways, therefore the `dispRity` function allows users to define their own measure of disparity.
-For more details on measuring disparity, see the [dispRity metrics section](#disparity-metrics).
-
-In this example, we are going to define disparity as the median distance between the different individuals and the centroid of the ordinated space.
-High values of disparity will indicate a generally high spread of points from this centroid (i.e. on average, the individuals are far apart in the ordinated space).
-We can define the metrics easily in the `dispRity` function by feeding them to the `metric` argument.
-Here we are going to feed the functions `stats::median` and `dispRity::centroids` which calculates distances between elements and their centroid.
-
-```{r}
-## Calculating disparity as the median distance between each elements and
-## the centroid of the petal-space
-(petal_disparity <- dispRity(petal_bootstrapped, metric = c(median, centroids)))
-```
-
-### Summarising the results (plot)
-Similarly to the `custom.subsets` and `boot.matrix` function, `dispRity` displays a `dispRity` object.
-But we are definitely more interested in actually look at the calculated values.
-
-First we can summarise the data in a table by simply using `summary`:
-
-```{r}
-## Displaying the summary of the calculated disparity
-summary(petal_disparity)
-```
-
-We can also plot the results in a similar way:
-
-```{r, fig.width=6, fig.height=6}
-## Graphical options
-par(bty = "n")
-
-## Plotting the disparity in the petal_space
-plot(petal_disparity)
-```
-
-Now contrary to simply plotting the two first axis of the PCA where we saw that the species have a different position in the two first petal-space, we can now also see that they occupy this space clearly differently!
-
-### Testing hypothesis
-
-Finally we can test our hypothesis that we guessed from the disparity plot (that some groups occupy different volume of the petal-space) by using the `test.dispRity` option.
-
-```{r}
-## Fitting a linear model to our data
-disparity_lm <- test.dispRity(petal_disparity, test = lm, comparisons = "all")
-
-## Testing whether there is a difference in disparity between the different species
-summary(aov(disparity_lm))
-
-## Post-hoc testing of the differences between species (corrected for multiple tests)
-test.dispRity(petal_disparity, test = t.test, correction = "bonferroni")
-```
-
-We can now see that there is a significant difference in petal-space occupancy between all species of iris.
-
-#### Setting up a multidimensional null-hypothesis
-
-One other series of test can be done on the shape of the petal-space.
-Using a MCMC permutation test we can simulate a petal-space with specific properties and see if our observed petal-space matches these properties (similarly to @diaz2016global):
-
-```{r, fig.width=6, fig.height=6}
-## Testing against a uniform distribution
-disparity_uniform <- null.test(petal_disparity, replicates = 200,
-    null.distrib = runif, scale = FALSE)
-plot(disparity_uniform)
-```
-
-```{r, fig.width=6, fig.height=6}
-## Testing against a normal distribution
-disparity_normal <- null.test(petal_disparity, replicates = 200,
-    null.distrib = rnorm, scale = TRUE)
-plot(disparity_normal)
-```
-
-In both cases we can see that our petal-space is not entirely normal or uniform.
-This is expected because of the simplicity of these parameters.
-
-<!--chapter:end:07_ecology-demo.Rmd-->
-
----
-title: "Future directions and references"
-author: "Thomas Guillerme"
-date: "`r Sys.Date()`"
-output: rmarkdown::html_vignette
-bibliography: References.bib
-bst: sysbio.bst
-vignette: >
-  %\VignetteIndexEntry{future directions}
-  %\VignetteEngine{knitr::rmarkdown}
-  \usepackage[utf8]{inputenc}
----
-
-# Future directions
-
-## More tests!
-
-Some more tests are being developed such as a `sequential.test` to run sequential linear models for testing hypothesis in disparity through time or a `model.test` developed in collaboration with [Mark Puttick](https://puttickbiology.wordpress.com/) to fit modes of evolution to disparity curves.
-Stay tuned!
-
-## More rarefactions!
-
-I am working on including several "clever" bootstrap and rarefaction algorithms for palaeobiology studies based on geological sampling (following:and Butler et al.
-2012 *Evolution* and Kotric and Knoll 2015 *Palaeobiology*).
-
-## Faster disparity calculations
-
-I am slowly implementing parallel disparity calculation as well as C implementations of some disparity metrics to increase significantly improve the speed of the `dispRity` function.
-
-## More modularity
-
-I am equally slowly developing functions to allow more of the options in the package to be modular (in the same way as the `metric` argument in `dispRity`).
-The next arguments to benefit this increased modularity will be `time.subsets`'s `model` argument and `boot.matrix`'s `type` argument.
-
-## More suggestions?
-
-More generally, I welcome any suggestions or collaborations for developing some additional specific ideas.
-Feel free to contact me via email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
-
-<!-- 
-* Beck, R.
-M., & Lee, M.
-S.
-(2014).
-Ancient dates or accelerated rates?
-Morphological clocks and the antiquity of placental mammals.
-Proceedings of the Royal Society of London B: Biological Sciences, 281(1793), 20141278.
-* Cooper, N., & Guillerme, T. (in prep.). Coming soonish!.
-* Diaz, S., Kattge, J., Cornelissen, J.H., Wright, I.J., Lavorel, S., Dray, S., Reu, B., Kleyer, M., Wirth, C., Prentice, I.C.
-and Garnier, E.
-(2016).
-The global spectrum of plant form and function.
-Nature, 529(7585), 167.
-* Donohue, I., Petchey, O.L., Montoya, J.M., Jackson, A.L., McNally, L., Viana, M., Healy, K., Lurgi, M., O'Connor, N.E.
-and Emmerson, M.C, (2013).
-On the dimensionality of ecological stability.
-Ecology letters, 16(4), 421-429.
-* Wills, M. A., Briggs, D. E., & Fortey, R. A. (1994). Disparity as an evolutionary index: a comparison of Cambrian and Recent arthropods. Paleobiology, 20(2), 93-130. -->
-
-<!--chapter:end:08_future_directions.Rmd-->
+You can even come up with your own ideas, implementations and modifications of the package: the `dispRity` package is a modular and collaborative package and I encourage you to contact me (guillert@tcd.e) for any ideas you have about adding new features to the package (whether you have them already implemented or not)!

### 2020-07-14 inst/gitbook/09_gmm-demo.Rmd
updated gitbook

---
title: "Morphometric geometric demo: a between group analysis"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

# Palaeobiology demo: disparity-through-time and within groups

This demo aims to give quick overview of the `dispRity` package (v.`r version_release`) for palaeobiology analyses of disparity, including disparity through time analyses.

This demo showcases a typical between groups geometric morphometric analysis: we are going to test whether the disparity in two species of salamander (plethodons!) are different and in which ways they are different.

## Before starting

Here we are going to use the `geomorph` `plethodon` dataset that is a set of 12 2D landmark coordinates for 40 specimens from two species of salamanders.
This section will really quickly cover how to make a Procrustes sumperimposition analysis and create a `geomorph` data.frame to have data ready for the `dispRity` package.

```{r}
## Loading geomorph
library(geomorph)

## Loading the plethodon dataset
data(plethodon)

## Running a simple Procrustes superimposition
gpa_plethodon <- gpagen(plethodon$land)

## Making a geomorph data frame object with the species and sites attributes
gdf_plethodon <- geomorph.data.frame(gpa_plethodon,
                                     species = plethodon$species,
                                     site = plethodon$site)
```

You can of course use your very own landmark coordinates dataset (though you will have to do some modifications in the scripts that will come below - they will be easy though!).

```{r}
## You can replace the gdf_plethodon by your own geomorph data frame!
my_geomorph_data <- gdf_plethodon
```

### The morphospace

The first step of every disparity analysis is to define your morphospace.

> Note that this is actually not true at all and kept as a erroneous sentence: the first step of your disparity analysis should be to define your question!

Our question here will be: is there a difference in disparity between the different species of salamanders and between the different sites (allopatric and sympatric)?

OK, now we can go to the _second_ step of every disparity analysis: defining the morphospace.
Here we will define it with the ordination of all possible Procrustes superimposed plethodon landmark coordinates.
You can do this directly in `dispRity` using the `geomorph.ordination` function that can input a geomorph data frame:

```{r}
## The morphospace
morphospace <- geomorph.ordination(gdf_plethodon)
```

This automatically generates a `dispRity` object with the information of each groups. You can find more information about `dispRity` objects [here](#disprity-object) but basically it summarises the content of your object without spamming your R console and is associated with many utility functions like `summary` or `plot`. For example here you can quickly visualise the two first dimensions of your space using the `plot` function:

```{r}
## The dispRity object
morphospace

## Plotting the morphospace
plot(morphospace)
## Note that this only displays the two last groups (site.Allo and site.Symp) since they overlap!
```

The `dispRity` package function comes with a lot of documentation of examples so don't hesitate to type `plot.dispRity` to check more plotting options.

## Calculating disparity

Now that we have our morphospace, we can think about what we want to measure.
Two aspects of disparity that would be interesting for our question (is there a difference in disparity between the different species of salamanders and between the different sites?) would be the differences in size in the morphospace (do both groups occupy the same amount of morphospace) and position in the morphospace (do the do groups occupy the same position in the morphospace?).

To choose which metric would cover best these two aspects, please check the @moms paper and associated [app](https://tguillerme.shinyapps.io/moms/). Here we are going to use the procrustes variance (`geomorph::morphol.disparity`) for measuring the size of the trait space and the average displacements [@moms] for the position in the trait space.

```{r}
## Defining a the procrustes variance metric
## (as in geomorph::morphol.disparity)
proc.var <- function(matrix) {sum(matrix^2)/nrow(matrix)}
## The size metric
test_size <- test.metric(morphospace, metric = proc.var,
                         shifts = c("random", "size"))
plot(test_size)
summary(test_size)

## The position metric
test_position <- test.metric(morphospace, metric = c(mean, displacements),
                         shifts = c("random", "position"))
plot(test_position)
summary(test_position)
```

You can see [here](#test-metric) for more details on the `test.metric` function but basically these graphs are showing that there is a relation between changes in size and in position for each metric.
Note that there are some caveats here but the selection of the metric is just for the sake of the example!

Note also the format of defining the disparity metrics here using `metric = c(mean, displacements)` or `metric = proc.var`. This is a core bit of the `dispRity` package were you can define your own metric as a function or a set of functions. You can find more info about this in the [`dispRity` metric section](#disparity-metrics) but in brief, the `dispRity` package considers metrics by their "dimensions" level which corresponds to what they output. For example, the function `mean` is a dimension level 1 function because no matter the input it outputs a single value (the mean), `displacements` on the other hand is a dimension level 2 function because it will output the ratio between the distance from the centroid and from the centre of the trait space for each row in a matrix (an example of a dimensions level 3 would be the function `var` that outputs a matrix).
The `dispRity` package always automatically sorts the dimensions levels: it will always run dimensions level 3 > dimensions level 2 > and dimensions level 1. In this case both `c(mean, displacements)` and `c(mean, displacements)` will result in actually running `mean(displacements(matrix))`.
Alternatively you can define your metric prior to the disparity analysis like we did for the `proc.var` function.

Anyways, we can measure disparity using these two metrics on all the groups as follows:

```{r}
## Bootstrapped disparity
disparity_size <-  dispRity(boot.matrix(morphospace), metric = proc.var)
disparity_position <- dispRity(boot.matrix(morphospace), metric = c(mean, displacements))
```

Note that here we use the `boot.matrix` function for quickly bootstrapping the matrix.
This is not an essential step in this kind of analysis but it allows to "reduce" the effect of outliers and create a distribution of disparity measures (rather than single point estimates).

## Analyse the results

We can visualise the results using the `plot` function on the resulting disparity objects (or summarising them using `summary`):

```{r}
## Plotting the results
par(mfrow = c(1,2))
plot(disparity_size, main = "group sizes", las = 2, xlab = "")
plot(disparity_position, main = "group positions", las = 2, xlab = "")

## Summarising the results
summary(disparity_size)
summary(disparity_position)
```

Just from looking at the data, we can guess that there is not much difference in terms of morphospace occupancy and position for the species but there is on for the sites (allopatric or sympatric).
We can test it using a simple non-parametric mean difference test (e.g. `wilcox.test`) using the `dispRity` package.

```{r}
## Testing the differences
test.dispRity(disparity_size, test = wilcox.test, correction = "bonferroni")
test.dispRity(disparity_position, test = wilcox.test, correction = "bonferroni")
```

So by applying the tests we see a difference in terms of position between each groups and differences in size between groups but between the species.

### 2020-07-16 inst/gitbook/03_specific-tutorials.Rmd
book update

@@
 
-```{r}
+```{r, fig.height = 16, fig.width = 6}
 ## The results as a dispRity object
@@
     ylab = c("Some measurement"), xlab = "Some other measurement",
-    main = "Many options...", ylim = c(5, 0), xlim = c(4, 0))
+    main = "Many options...", ylim = c(10, 0), xlim = c(4, 0))
 
@@
 ## Plotting the continuous disparity with a fixed y axis
-plot(disparity_time_slices, ylim = c(1, 4))
+plot(disparity_time_slices, ylim = c(3, 9))
 ## Adding the discrete data
-plot(disparity_time_bins, type = "line", ylim = c(1, 4),
+plot(disparity_time_bins, type = "line", ylim = c(3, 9),
      xlab = "", ylab = "", add = TRUE)
@@
 
-```{r}
+```{r, fig.height = 16, fig.width = 8}
 par(mfrow = c(2,1))

### 2020-07-16 inst/gitbook/dispRity_manual.Rmd
book update

@@
 Note however that *it can happen* that the `master` version can sometimes be bugged (especially when there are major `R` and `R` packages updates), however, the status of the package state on both the `release` and the `master` version is constantly displayed on the [`README` page](https://github.com/TGuillerme/dispRity/) of the package with the nice badges displaying these different (and constantly tested) information.
-
-### So which version do I choose? {#version}
-
-There are always three version of the package available:
-
-  * The CRAN one
-  * The GitHub `release` one
-  * The GitHub `master` one
-
-The differences between the CRAN one and the GitHub `release` or `master` ones is explained just above.
-For the the GitHub version, the differences are that the `release` one is more stable (i.e.
-more rarely modified) and the `master` one is more live one (i.e.
-bug fixes and new functionalities are added as they come).
-
-If you want the latest-latest version of the package I suggest using the GitHub `master` one, especially if you recently emailed me reporting a minor bug or wanting a new functionality!
-Note however that *it can happen* that the `master` version can sometimes be bugged (especially when there are major `R` and `R` packages updates), however, the status of the package state on both the `release` and the `master` version is constantly displayed on the [`README` page](https://github.com/TGuillerme/dispRity/) of the package with the nice badges displaying these different (and constantly tested) information.
-
-Note that when we (the `dispRity` package collaborators!) are developing some more major versions, including some more complex features for example, they will never be released on either the the versions described above.
-Therefore you can rest assure that you will never use a badly bugged version of `dispRity`! Enjoy!
 
@@
 
+### Why is it important to cite us?
+
+Aside from how science works (if you're using a method from a specific paper, cite that specific paper to refer to that specific method), why is it important to also cite the package and the manual?
+
+All the people involve in making the `dispRity` package happened to do it enthusiastically, freely and most amazingly without asking anything in return!
+I created the package with this idea in mind and I am still sticking to it.
+However, academia (the institutions and people producing science around the globe) is unfortunately not optimal at many level (some might even say "broken"): high impact papers attract big grants that attract high impact papers and big grants again, all this along with livelihood, permanent position and job security.
+Unfortunately however, method development has a hard time to catch up with the current publish or perish system: constantly updating the `dispRity` package and this manual is hugely time consuming (but really fun!) and that is not even taking into account maintenance and helping users.
+Although I do truly believe that this time spent doing these things modestly help the scientific endeavour, it does not contribute to our paper list!
+
+Therefore, by citing the package and this manual, you help provide visibility to other workers and you might help them in their work!
+And you directly contribute in making this project fun for all the people involved and most of all, free, updated and independent from the publish and perish system!
+
+Thank you!
 <!-- ## Associated papers -->
@@
 It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance among tips *and* nodes.
+
+If you want to generate time subsets based on stratigraphy, the package proposes a useful functions to do it for you: `get.bin.ages` (check out the function's manual in `R`)!
+
+### Time-slicing {#time-slicing}
 
@@
 
+More details about the differences between these methods can be found in @time-slice.
+
 ```{r, eval=TRUE}
-## Generating four time slices every 40 million years under a model of proximity evolution
+## Generating four time slices every 40 million years
+## under a model of proximity evolution
 chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
+               method = "continuous", model = "proximity",
+               time = c(120, 80, 40, 0),
                FADLAD = BeckLee_ages)
@@
 chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree,
-    method = "continuous", model = "proximity", time = 4, FADLAD = BeckLee_ages)
-```
-
-If you want to generate time subsets based on stratigraphy, the package proposes a useful functions to do it for you: `get.bin.ages` (check out the function's manual in `R`)!
-
-## Customised subsets
+               method = "continuous", model = "proximity",
+               time = 4, FADLAD = BeckLee_ages)
+```
+
+## Customised subsets {#custom-subsets}
 
@@
 crown_stem <- custom.subsets(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
+                             group = crown.stem(BeckLee_tree,
+                                                inc.nodes = FALSE))
 ## Bootstrapping and rarefying these groups
@@
 1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln$vol` |
+2 | `deviations` | The minimal distance between each element and a hyperplane | `dispRity` |
 1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
+2 | `displacements` | The ratio between the distance from a reference and the distance from the centroid | `dispRity` |
 1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
+1 | `func.div` | The functional divergence (the ratio of deviation from the centroid) | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FDiv` but without abundance)|
+1 | `func.eve` | The functional evenness (the minimal spanning tree distances evenness) | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FEve` but without abundance)|
 1 | `mode.val` | The modal value | `dispRity` |
 1 | `n.ball.volume` | The hyper-spherical (*n*-ball) volume | `dispRity` |
+2 | `neighbours` | The distance to specific neighbours (e.g. the nearest neighbours - by default) | `dispRity` |
 2 | `pairwise.dist` | The pairwise distances between elements | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::vegist` |
+2 | `quantiles` | The *n*th quantile range per axis | `dispRity` |
 2 | `radius` | The radius of each dimensions | `dispRity` |
 2 | `ranges` | The range of each dimension | `dispRity` |
-1 | `span.tree.length` | The minimal spanning tree length | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree` |
+2 | `span.tree.length` | The minimal spanning tree length | [`ape`](https://cran.r-project.org/web/packages/vegan/index.html) (but see also [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree`) |
 2 | `variances` | The variance of each dimension | `dispRity` |
@@
 \begin{equation}
-    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
+    deviations = \frac{|Ax + By + ... + Nm + Intercept|}{\sqrt{A^2 + B^2 + ... + N^2}}
 \end{equation}
@@
 \begin{equation}
-    n.ball.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R
+    displacements = \frac{\sqrt{\sum_{i=1}^{n}{({d}_{n}-Reference_{d})^2}}}{\sqrt{\sum_{i=1}^{n}{({d}_{n}-Centroid_{k})^2}}}
 \end{equation}
@@
 \begin{equation}
-    radius = |\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|
+    ellipse.volume = \frac{\pi^{d/2}}{\Gamma(\frac{d}{2}+1)}\displaystyle\prod_{i=1}^{d} (\lambda_{i}^{0.5})
 \end{equation}
@@
 \begin{equation}
-    ranges = |max(k_i) - min(k_i)|
+    n.ball.volume = \frac{\pi^{d/2}}{\Gamma(\frac{d}{2}+1)}\displaystyle\prod_{i=1}^{d} R
 \end{equation}
@@
 \begin{equation}
-    variances = \sigma^{2}{k_i}
+    radius = |\frac{\sum_{i=1}^{n}d_i}{n} - f(\mathbf{v}d)|
 \end{equation}
@@
 \begin{equation}
-    span.tree.length = \sum(\mathrm{branch\ length})
+    ranges = |max(d_i) - min(d_i)|
 \end{equation}
 
-Where *k* is the number of dimensions,
+\begin{equation}
+    variances = \sigma^{2}{d_i}
+\end{equation}
+
+\begin{equation}
+    span.tree.length = \mathrm{branch\ length}
+\end{equation}
+
+Where *d* is the number of dimensions,
 *n* the number of elements,
@@
 
-## Calculating disparity as the sum and the product of these variances
+## Calculating disparity as the sum and
+## the product of these variances
 summary(dispRity(dummy_space, metric = c(sum, variances)))
 summary(dispRity(dummy_space, metric = c(prod, variances)))
+
+## Calculating the quantiles of each dimension
+## in the ordinated space
+quantiles(dummy_space)
+
+## Calculating disparity as the distribution of these variances
+summary(dispRity(dummy_space, metric = quantiles))
+
+## By default, the quantile calculated is the 95%
+## (i.e. 95% of the data on each axis)
+## this can be changed using the option quantile:
+summary(dispRity(dummy_space, metric = quantiles, quantile = 50))
 
@@
 ## Calculating disparity as the mean average radius
-summary(dispRity(dummy_space, metric = c(mean, radius), type = mean))
-```
-
-The pairwise distances uses the function `vegan::vegdist` and can take the normal `vegdist` options:
+summary(dispRity(dummy_space,
+                 metric = c(mean, radius),
+                 type = mean))
+```
+
+The pairwise distances and the neighbours distances uses the function `vegan::vegdist` and can take the normal `vegdist` options:
 
@@
 ## The distribution of the Manhattan distances
-summary(dispRity(dummy_space, metric = pairwise.dist, method = "manhattan"))
+summary(dispRity(dummy_space, metric = pairwise.dist,
+                 method = "manhattan"))
+
+## The average nearest neighbour distances
+summary(dispRity(dummy_space, metric = neighbours))
+
+## The average furthest neighbour manhattan distances
+summary(dispRity(dummy_space, metric = neighbours,
+                 which = max, method = "manhattan"))
 ```
@@
 
-#### Centroids and ancestral distance metrics
+#### Centroids, displacements, and ancestral distance metrics {#centroids}
 
@@
 ```{r}
-## The distance between each element and the origin of the ordinated space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))
-
-## Disparity as the distance between each element and a specific point in space
-summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
+## The distance between each element and the origin
+## of the ordinated space
+summary(dispRity(dummy_space, metric = centroids, centroid = 0))
+
+## Disparity as the distance between each element
+## and a specific point in space
+summary(dispRity(dummy_space, metric = centroids,
+                 centroid = c(0,1,2,3,4)))
+```
+
+If you have subsets in your `dispRity` object, you can also use the `matrix.dispRity` (see [utilities](#utilities)) and `colMeans` to get the centre of a specific subgroup.
+For example
+
+```{r, warning = FALSE}
+## Create a custom subsets object
+dummy_groups <- custom.subsets(dummy_space,
+                               group = list("group1" = 1:5,
+                                            "group2" = 6:10))
+summary(dispRity(dummy_groups, metric = centroids,
+    centroid = colMeans(matrix.dispRity(dummy_groups, "group1"))))
+```
+
+The `displacements` distance is the ratio between the `centroids` distance and the `centroids` distance with `centroid = 0`.
+Note that it is possible to measure a ratio from another point than `0` using the `reference` argument.
+It gives indication of the relative displacement of elements in the multidimensional space: a score  >1 signifies a displacement *away* from the reference.
+A score of >1 signifies a displacement *towards* the reference.
+
+```{r}
+## The relative displacement of the group in space to the centre
+summary(dispRity(dummy_space, metric = displacements))
+
+## The relative displacement of the group to an arbitrary point
+summary(dispRity(dummy_space, metric = displacements,
+                 reference = c(0,1,2,3,4)))
 ```
@@
 Note that because the solution is heuristic, this metric can take a long time to compute for big matrices.
+
+#### Functional divergence and evenness
+
+The `func.div` and `func.eve` functions are based on the `FD::dpFD` package.
+They are the equivalent to `FD::dpFD(matrix)$FDiv` and `FD::dpFD(matrix)$FEve` but a bit faster (since they don't deal with abundance data).
+They are pretty straightforward to use:
+
+```{r}
+## The ratio of deviation from the centroid 
+summary(dispRity(dummy_space, metric = func.div))
+
+## The minimal spanning tree distances evenness
+summary(dispRity(dummy_space, metric = func.eve))
+
+## The minimal spanning tree manhanttan distances evenness
+summary(dispRity(dummy_space, metric = func.eve,
+                 method = "manhattan"))
+```
+
+#### Orientation: angles and deviations
+
+The `angles` performs a least square regression (via the `lm` function) and returns slope of the main axis of variation for each dimension.
+This slope can be converted into different units, `"slope"`, `"degree"` (the default) and `"radian"`.
+This can be changed through the `unit` argument.
+By default, the angle is measured from the slope 0 (the horizontal line in a 2D plot) but this can be changed through the `base` argument (using the defined `unit`):
+
+```{r}
+## The distribution of each angles in degrees for each
+## main axis in the matrix
+summary(dispRity(dummy_space, metric = angles))
+
+## The distribution of slopes deviating from the 1:1 slope:
+summary(dispRity(dummy_space, metric = angles, unit = "slope",
+                 base = 1))
+```
+
+The `deviations` function is based on a similar algorithm as above but measures the deviation from the main axis (or hyperplane) of variation.
+In other words, it finds the least square line (for a 2D dataset), plane (for a 3D dataset) or hyperplane (for a >3D dataset) and measures the shortest distances between every points and the line/plane/hyperplane.
+By default, the hyperplane is fitted using the least square algorithm from `stats::glm`:
+
+```{r}
+## The distribution of the deviation of each point
+## from the least square hyperplane
+summary(dispRity(dummy_space, metric = deviations))
+```
+
+It is also possible to specify the hyperplane equation through the `hyperplane` equation.
+The equation must contain the intercept first and then all the slopes and is interpreted as $intercept + Ax + By + ...
++ Nd = 0$.
+For example, a 2 line defined as beta + intercept (e.g.
+$y = 2x + 1$) should be defined as `hyperplane = c(1, 2, 1)` ($2x - y + 1 = 0$).
+
+```{r}
+## The distribution of the deviation of each point
+## from a slope (with only the two first dimensions)
+summary(dispRity(dummy_space[, c(1:2)], metric = deviations,
+                 hyperplane = c(1, 2, -1)))
+```
+
+Since both the functions `angles` and `deviations` effectively run a `lm` or `glm` to estimate slopes or hyperplanes, it is possible to use the option `significant = TRUE` to only consider slopes or intercepts that have a slope significantly different than zero using an `aov` with a significant threshold of $p = 0.05$.
+Note that depending on your dataset, using and `aov` could be completely inappropriate!
+In doubt, it's probably better to enter your `base` (for `angles`) or your `hyperplane` (for `deviations`) manually so you're sure you know what the function is measuring.
+
+### Which disparity metric to choose?
+
+The disparity metric that gives the most consistent results is the following one:
+
+```{r}
+best.metric <- function() return(42)
+```
+Joke aside, this is a legitimate question that has no simple answer: **it depends on the dataset and question at hand**.
+Thoughts on which metric to choose can be find in @moms and @Guillerme2020 but again, will ultimately depend on the question and dataset.
+The question should help figuring out which type of metric is desired: for example, in the question "does the extinction released niches for mammals to evolve", the metric in interest should probably pick up a change in size in the trait space (the release could result in some expansion of the mammalian morphospace); or if the question is "does group X compete with group Y", maybe the metric of interested should pick up changes in position (group X can be displaced by group Y).
+
+In order to visualise what signal different disparity metrics are picking, you can use the [`moms`](https://tguillerme.shinyapps.io/moms/) that come with a detailed [manual](https://rawcdn.githack.com/TGuillerme/moms/master/inst/moms_vignette.html) on how to use it.
+
+Alternatively, you can use the `test.metric` function:
+
+#### `test.metric` {#test-metric}
+
+This function allows to test whether a metric picks different changes in disparity.
+It intakes the space on which to test the metric, the disparity metric and the type of changes to apply gradually to the space.
+Basically this is a type of biased data rarefaction (or non-biased for `"random"`) to see how the metric reacts to specific changes in trait space.
+
+```{r}
+# Creating a 2D uniform space
+example_space <- space.maker(300, 2, runif)
+
+## Testing the product of ranges metric on the example space
+example_test <- test.metric(example_space, metric = c(prod, ranges),
+                           shifts = c("random", "size")) 
+```
+
+By default, the test runs three replicates of space reduction as described in @moms by gradually removing 10% of the data points following the different algorithms from @moms (here the `"random"` reduction and the `"size"`) reduction, resulting in a `dispRity` object that can be summarised or plotted.
+The number of replicates can be changed using the `replicates` option.
+Still by default, the function then runs a linear model on the simulated data to measure some potential trend in the changes in disparity. 
+The model can be changed using the `model` option.
+Finally, the function runs 10 reductions by default from keeping 10% of the data (removing 90%) and way up to keeping 100% of the data (removing 0%).
+This can be changed using the `steps` option.
+A good disparity metric for your dataset will typically have no trend in the `"random"` reduction (the metric is ideally not affected by sample size) but should have a trend for the reduction of interest.
+
+```{r}
+## The results as a dispRity object
+example_test
+
+## Summarising these results
+summary(example_test)
+
+## Or visualising them
+plot(example_test)
+```
 
@@
 crown_stem <- custom.subsets(BeckLee_mat50,
-                                group = list("crown" = c(16, 19:41, 45:50), 
-                                             "stem" = c(1:15, 17:18, 42:44)))
+                             group = crown.stem(BeckLee_tree,
+                             inc.nodes = FALSE))
 ## Bootstrapping and rarefying these groups
-boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
+boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100,
+                               rarefaction = TRUE)
 ## Calculate disparity
-disparity_crown_stem <- dispRity(boot_crown_stem, metric = c(sum, variances))
+disparity_crown_stem <- dispRity(boot_crown_stem,
+                                 metric = c(sum, variances))
 
 ## Creating time slice subsets
-time_slices <- chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, 
-    method = "continuous", model = "proximity", time = c(120, 80, 40, 0),
+time_slices <- chrono.subsets(data = BeckLee_mat99,
+                              tree = BeckLee_tree,
+                              method = "continuous",
+                              model = "proximity",
+                              time = c(120, 80, 40, 0),
                               FADLAD = BeckLee_ages)
@@
 ## Plotting the results with some plot.dispRity arguments
-plot(disparity_time_slices, quantile = c(seq(from = 10, to = 100, by = 10)),
-    cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
+plot(disparity_time_slices,
+    quantiles = c(seq(from = 10, to = 100, by = 10)),
+    cent.tend = sd, type = "c", elements = TRUE,
+    col = c("black", rainbow(10)),
     ylab = c("Disparity", "Diversity"), chrono.subsets = FALSE,
-    xlab = "Time (in in units from past to present)", observed = TRUE,
+    xlab = "Time (in in units from past to present)",
+    observed = TRUE,
     main = "Many more options...")
@@
 ## Plotting the continuous disparity with a fixed y axis
-plot(disparity_time_slices, ylim = c(1, 4))
+plot(disparity_time_slices, ylim = c(3, 9))
 ## Adding the discrete data
-plot(disparity_time_bins, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
-    add = TRUE)
+plot(disparity_time_bins, type = "line", ylim = c(3, 9),
+     xlab = "", ylab = "", add = TRUE)
 
@@
 par(op)
+```
+
+### `type = preview`
+
+Note that all the options above are plotting disparity objects for which a disparity metric *has been calculated*.
+This makes totally sense for `dispRity` objects but sometimes it might be interesting to look at what the trait-space looks like before measuring the disparity.
+This can be done by plotting `dispRity` objects with no calculated disparity!
+
+For example, we might be interested in looking at how the distribution of elements change as a function of the distributions of different sub-settings.
+For example custom subsets *vs.* time subsets:
+
+```{r, fig.width=8, fig.height=4}
+## Making the different subsets
+cust_subsets <- custom.subsets(BeckLee_mat99,
+                               crown.stem(BeckLee_tree,
+                                          inc.nodes = TRUE))
+time_subsets <- chrono.subsets(BeckLee_mat99,
+                               tree = BeckLee_tree,
+                               method = "discrete",
+                               time = 5)
+
+## Note that no disparity has been calculated here:
+is.null(cust_subsets$disparity)
+is.null(time_subsets$disparity)
+
+## But we can still plot both spaces by using the default plot functions
+par(mfrow = c(1,2))
+## Default plotting
+plot(cust_subsets)
+## Plotting with more arguments
+plot(time_subsets, dimensions = c(4,5),
+     main = "Some \"low\" dimensions")
+```
+
+> DISCLAIMER: This functionality can be handy for exploring the data (e.g.
+to visually check whether the subset attribution worked) but it might be misleading on how the data is *actually* distributed in the multidimensional space!
+Groups that don't overlap on two set dimensions can totally overlap in all other dimensions!
+
+For `dispRity` objects that do contain disparity data, the default option is to plot your disparity data.
+However you can always force the `preview` option using the following:
+
+```{r}
+par(mfrow = c(2,1))
+## Default plotting
+plot(disparity_time_slices, main = "Disparity through time")
+## Plotting with more arguments
+plot(disparity_time_slices, type = "preview",
+     main = "Two first dimensions of the trait space")
 ```
@@
 
+## Disparity from multiple matrices (and multiple trees!) {#multi.input}
+
+Since the version `1.4` of this package, it is possible to use multiple trees and multiple matrices in `dispRity` objects.
+To use multiple matrices, this is rather easy: just supply a list of matrices to any of the `dispRity` functions and, as long as they have the same size and the same rownames they will be handled as a distribution of matrices.
+
+```{r}
+set.seed(1)
+## Creating 3 matrices with 4 dimensions and 10 elements each (called t1, t2, t3, etc...)
+matrix_list <- replicate(3, matrix(rnorm(40), 10, 4, dimnames = list(paste0("t", 1:10))),
+                         simplify = FALSE)
+class(matrix_list) # This is a list of matrices
+
+## Measuring some disparity metric on one of the matrices
+summary(dispRity(matrix_list[[1]], metric = c(sum, variances)))
+
+## Measuring the same disparity metric on the three matrices
+summary(dispRity(matrix_list, metric = c(sum, variances)))
+```
+
+As you can see, when measuring the sum of variances on multiple matrices, we now have a distribution of sum of variances rather than a single observed value.
+
+Similarly as running disparity analysis using multiple matrices, you can run the `chrono.subsets` function using multiple trees.
+This can be useful if you want to use a tree posterior distribution rather than a single consensus tree.
+These trees can be passed to `chrono.subsets` as a `"multiPhylo"` object (with the same node and tip labels in each tree).
+First let's define a function to generate multiple trees with the same labels and root ages:
+
+```{r}
+set.seed(1)
+## Matches the trees and the matrices
+## A bunch of trees
+make.tree <- function(n, fun = rtree) {
+    ## Make the tree
+    tree <- fun(n)
+    tree <- chronos(tree, quiet = TRUE,
+                    calibration = makeChronosCalib(tree, age.min = 10, age.max = 10))
+    class(tree) <- "phylo"
+    ## Add the node labels
+    tree$node.label <- paste0("n", 1:Nnode(tree))
+    ## Add the root time
+    tree$root.time <- max(tree.age(tree)$ages)
+    return(tree)
+}
+trees <- replicate(3, make.tree(10), simplify = FALSE)
+class(trees) <- "multiPhylo"
+trees
+```
+
+We can now simulate some ancestral states for the matrices in the example above to have multiple matrices associated with the multiple trees.
+
+```{r}
+## A function for running the ancestral states estimations
+do.ace <- function(tree, matrix) {
+    ## Run one ace
+    fun.ace <- function(character, tree) {
+        results <- ace(character, phy = tree)$ace
+        names(results) <- paste0("n", 1:Nnode(tree))
+        return(results)
+    }
+    ## Run all ace
+    return(rbind(matrix, apply(matrix, 2, fun.ace, tree = tree)))
+}
+
+## All matrices
+matrices <- mapply(do.ace, trees, matrix_list, SIMPLIFY = FALSE)
+```
+
+Let's first see an example of time-slicing with one matrix and multiple trees.
+This assumes that your tip values (observed) and node values (estimated) are fixed with no error on them.
+It also assumes that the nodes in the matrix always corresponds to the node in the trees (in other words, the tree topologies are fixed):
+
+```{r}
+## Making three "proximity" time slices across one tree
+one_tree <- chrono.subsets(matrices[[1]], trees[[1]],
+                           method = "continuous",
+                           model = "proximity", time = 3)
+## Making three "proximity" time slices across the three trees
+three_tree <- chrono.subsets(matrices[[1]], trees,
+                             method = "continuous",
+                             model = "proximity", time = 3)
+## Measuring disparity as the sum of variances and summarising it
+summary(dispRity(one_tree, metric = c(sum, variances)))
+summary(dispRity(three_tree, metric = c(sum, variances)))
+```
+
+This results show the effect of considering a tree distribution: in the first case (`one_tree`) the time slice at 3.95 Mya has a sum of variances of 2.9 but this values goes down to 0.256 in the second case (`three_tree`) which is due to the differences in branch lengths distributions:
+
+```{r, print = FALSE}
+par(mfrow = c(3,1))
+slices <- c(7.9, 3.95, 0)
+fun.plot <- function(tree) {
+  plot(tree)
+  nodelabels(tree$node.label, cex = 0.8)
+  axisPhylo()
+  abline(v = tree$root.time - slices)
+}
+silent <- lapply(trees, fun.plot)
+```
+
+Note that in this example, the nodes are actually even different in each tree! The node `n4` for example, is not direct descendent of `t4` and `t6` in all trees!
+To fix that, it is possible to input a list of trees and a list of matrices that correspond to each tree in `chrono.subsets` by using the `bind.data = TRUE` option.
+In this case, the matrices need to all have the same row names and the trees all need the same labels as before:
+
+```{r}
+## Making three "proximity" time slices across three trees and three bound matrices
+bound_data <- chrono.subsets(matrices, trees,
+                             method = "continuous",
+                             model = "proximity",
+                             time = 3,
+                             bind.data = TRUE)
+## Making three "proximity" time slices across three trees and three matrices
+unbound_data <- chrono.subsets(matrices, trees,
+                               method = "continuous",
+                               model = "proximity",
+                               time = 3,
+                               bind.data = FALSE)
+
+## Measuring disparity as the sum of variances and summarising it
+summary(dispRity(bound_data, metric = c(sum, variances)))
+summary(dispRity(unbound_data, metric = c(sum, variances)))
+```
+
+Note here that the results are again rather different: with the bound data, the slices are done across the three trees and each of their corresponding matrix (resulting in three observation) which is more accurate than the previous results from `three_trees` above.
+With the unbound data, the slices are done across the three trees and applied to the three matrices (resulting in 9 observations).
+As we've seen before, this is incorrect in this case since the trees don't have the same topology (so the nodes selected by a slice through the second tree are not equivalent to the nodes in the first matrix) but it can be useful if the topology is fixed to integrate both uncertainty in branch length (slicing through different trees) and uncertainty from, say, ancestral states estimations (applying the slices on different matrices).
+
 <!--chapter:end:03_specific-tutorials.Rmd-->
@@
 my_matrix_NA <- apply.NA(my_matrix, tree = my_tree,
-                         NAs = c(rep("character", 5), rep("clade", 10)))
+                         NAs = c(rep("character", 5),
+                                 rep("clade", 10)))
 
@@
 ## matrix
-morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
-    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
+morpho_mat <- sim.morpho(my_tree,
+                         characters = 100,
+                         model = "ER",
+                         rates = c(rgamma, rate = 100, shape = 5),
+                         invariant = FALSE)
 check.morpho(morpho_mat, my_tree)
@@
 ## Visualising the space
-plot(square_space, pch = 20, xlab = "", ylab = "", main = "Uniform 2D space")
+plot(square_space, pch = 20, xlab = "", ylab = "",
+     main = "Uniform 2D space")
 ```
@@
 plane_space <- space.maker(2500, 3, c(runif, runif, runif),
-                           arguments = list(list(min = 0, max = 0), NULL, NULL))
+                           arguments = list(list(min = 0, max = 0),
+                           NULL, NULL))
 
@@
 ## An ellipsoid space (normal space with correlation)
-ellipse_space <- space.maker(2500, 3, rnorm, cor.matrix = cor_matrix)
+ellipse_space <- space.maker(2500, 3, rnorm,
+                             cor.matrix = cor_matrix)
 
@@
 doughnut_space <- space.maker(5000, 3, c(rnorm, random.circle),
-     arguments = list(list(mean = 0), list(runif, inner = 0.5, outer = 1)))
+     arguments = list(list(mean = 0),
+                      list(runif, inner = 0.5, outer = 1)))
 ```
@@
 <!--chapter:end:04_making-up-data.Rmd-->
+
+---
+title: "other functionalities"
+author: "Thomas Guillerme"
+date: "`r Sys.Date()`"
+output: rmarkdown::html_vignette
+bibliography: [../References.bib, ../packages.bib]
+bst: sysbio.bst
+vignette: >
+  %\VignetteIndexEntry{other functionalities}
+  %\VignetteEngine{knitr::rmarkdown}
+  \usepackage[utf8]{inputenc}
+---
+
+# Other functionalities
+
+The `dispRity` package also contains several other functions that are not specific to multidimensional analysis but that are often used by `dispRity` internal functions.
+However, we decided to make these functions also available at a user level since they can be handy for certain specific operations!
+You'll find a brief description of each of them (alphabetically) here:
+
+## `char.diff`
+
+This is yet another function for calculating distance matrices.
+There are many functions for calculating pairwise distance matrices in `R` (`stats::dist`, `vegan::vegdist`, `cluster::daisy` or `Claddis::MorphDistMatrix`) but this one is the `dispRity` one.
+It is slightly different to the ones mentioned above (though not that dissimilar from `Claddis::MorphDistMatrix`) in the fact that it focuses on comparing discrete morphological characters and tries to solve all the problems linked to these kind of matrices (especially dealing with special tokens).
+
+The function intakes a matrix with either `numeric` or `integer` (`NA` included) or matrices with `character` that are indeed `integers` (e.g.`"0"` and `"1"`).
+It then uses a [bitwise operations](https://en.wikipedia.org/wiki/Bitwise_operations_in_C) architecture implemented in `C` that renders the function pretty fast and pretty modular.
+This bitwise operations translates the character states into binary values.
+This way, `0` becomes `1`, `1` becomes `2`, `2` becomes `4`, `3` becomes `8`, etc...
+Specifically it can handle any rules specific to special tokens (i.e.
+symbols) for discrete morphological characters.
+For example, should you treat missing values `"?"` as `NA` (ignoring them) or as any possible character state (e.g.
+`c("0", "1")`?)?
+And how to treat characters with a ampersand (`"&"`)?
+`char.diff` can answer to all these questions!
+
+Let's start by a basic binary matrix 4\*3 with random `integer`:
+
+```{r}
+## A random binary matrix
+matrix_binary <- matrix(sample(c(0,1), 12, replace = TRUE), ncol = 4,
+                        dimnames = list(letters[1:3], LETTERS[1:4]))
+```
+
+By default, `char.diff` measures the hamming distance between characters:
+
+```{r}
+## The hamming distance between characters
+(differences <- char.diff(matrix_binary))
+```
+
+Note that the results is just a pairwise distance (dissimilarity) matrix with some special dual class `matrix` and `char.diff`.
+This means it can easily be plotted via the disparity package:
+
+```{r}
+## Visualising the matrix
+plot(differences)
+```
+
+You can check all the numerous plotting options in the `?plot.char.diff` manual (it won't be developed here).
+
+The `char.diff` function has much more options however (see all of them in the `?char.diff` manual) for example to measure different differences (via `method`) or making the comparison work per row (for a distance matrix between the rows):
+
+```{r}
+## Euclidean distance between rows
+char.diff(matrix_binary, by.col = FALSE, method = "euclidean")
+```
+
+We can however make it more interesting by playing with the different rules to play with different tokens.
+First let's create a matrix with morphological characters as numeric `character`s:
+
+```{r}
+## A random character matrix
+(matrix_character <- matrix(sample(c("0","1","2"), 30, replace = TRUE), ncol = 5,
+                           dimnames = list(letters[1:6], LETTERS[1:5])))
+
+## The hamming difference between columns
+char.diff(matrix_character)
+```
+
+Here the characters are automatically converted into bitwise integers to be compared efficiently.
+We can now add some more special tokens like `"?"` or `"0/1"` for uncertainties between state `"0"` and `"1"` but not `"2"`:
+
+```{r}
+## Adding uncertain characters
+matrix_character[sample(1:30, 8)] <- "0/1"
+
+## Adding missing data
+(matrix_character[sample(1:30, 5)] <- "?")
+
+## The hamming difference between columns including the special characters
+char.diff(matrix_character)
+```
+
+Note here that it detected the default behaviours for the special tokens `"?"` and `"/"`: `"?"` are treated as `NA` (not compared) and `"/"` are treated as both states (e.g.
+`"0/1"` is treated as `"0"` and as `"1"`).
+We can specify both the special tokens and the special behaviours to consider via `special.tokens` and `special.behaviours`.
+The `special.tokens` are `missing = "?"`, `inapplicable = "-"`, `uncertainty = "\"` and `polymorphism = "&"` meaning we don't have to modify them for now.
+However, say we want to change the behaviour for `"?"` and treat them as all possible characters and treat `"/"` as only the character `"0"` (as an integer) we can specify them giving a behaviour function:
+
+```{r}
+## Specifying some special behaviours
+my_special_behaviours <- list(missing = function(x,y) return(y),
+                              uncertainty = function(x,y) return(as.integer(0)))
+
+## Passing these special behaviours to the char.diff function
+char.diff(matrix_character, special.behaviour = my_special_behaviours)
+```
+
+The results are quiet different as before! Note that you can also specify some really specific behaviours for any type of special token.
+
+```{r}
+## Adding weird tokens to the matrix
+matrix_character[sample(1:30, 8)] <- "%"
+
+## Specify the new token and the new behaviour
+char.diff(matrix_character, special.tokens = c(weird_one = "%"),
+                            special.behaviours = list(
+                                weird_one = function(x,y) return(as.integer(42)))
+                            )
+```
+
+Of course the results can be quiet surprising then... But that's the essence of the modularity. You can see more options in the function manual `?char.diff`!
+
+## `clean.data`
+
+This is a rather useful function that allows matching a `matrix` or a `data.frame` to a tree (`phylo`) or a distribution of trees (`multiPhylo`).
+This function outputs the cleaned data and trees (if cleaning was needed) and a list of dropped rows and tips.
+
+```{r}
+## Generating a trees with labels from a to e
+dummy_tree <- rtree(5, tip.label = LETTERS[1:5])
+
+## Generating a matrix with rows from b to f
+dummy_data <- matrix(1, 5, 2, dimnames = list(LETTERS[2:6], c("var1", "var2")))
+
+##Cleaning the trees and the data
+(cleaned <- clean.data(data = dummy_data, tree = dummy_tree))
+```
+
+## `crown.stem`
+
+This function quiet handily separates tips from a phylogeny between crown members (the living taxa and their descendants) and their stem members (the fossil taxa without any living relatives).
+
+```{r}
+data(BeckLee_tree)
+## Diving both crow and stem species
+(crown.stem(BeckLee_tree, inc.nodes = FALSE))
+```
+
+Note that it is possible to include or exclude nodes from the output.
+To see a more applied example: this function is used in [chapter 03: specific tutorials](#custom-subsets).
+
+## `get.bin.ages`
+
+This function is similar than the `crown.stem` one as it is based on a tree but this one outputs the stratigraphic bins ages that the tree is covering.
+This can be useful to generate precise bin ages for the `chrono.subsets` function:
+
+```{r}
+get.bin.ages(BeckLee_tree)
+```
+
+Note that this function outputs the stratigraphic age limits by default but this can be customisable by specifying the `type` of data (e.g. `type = "Eon"` for eons).
+The function also intakes several optional arguments such as whether to output the startm end, range or midpoint of the stratigraphy or the year of reference of the International Commission of Stratigraphy.
+To see a more applied example: this function is used in [chapter 03: specific tutorials](#chrono-subsets).
+
+## `pair.plot`
+
+This utility function allows to plot a matrix image of pairwise comparisons.
+This can be useful when getting pairwise comparisons and if you'd like to see at a glance which pairs of comparisons have high or low values.
+
+```{r}
+## Random data
+data <- matrix(data = runif(42), ncol = 2)
+
+## Plotting the first column as a pairwise comparisons
+pair.plot(data, what = 1, col = c("orange", "blue"), legend = TRUE, diag = 1)
+```
+
+Here blue squares are ones that have a high value and orange ones the ones that have low values.
+Note that the values plotted correspond the first column of the data as designated by `what = 1`.
+
+It is also possible to add some tokens or symbols to quickly highlight to specific cells, for example which elements in the data are below a certain value:
+
+```{r}
+## The same plot as before without the diagonal being the maximal observed value
+pair.plot(data, what = 1, col = c("orange", "blue"), legend = TRUE, diag = "max")
+## Highlighting with an asterisk which squares have a value below 0.2
+pair.plot(data, what = 1, binary = 0.2, add = "*", cex = 2)
+```
+
+This function can also be used as a binary display when running a series of pairwise t-tests.
+For example, the following script runs a wilcoxon test between the time-slices from the `disparity` example dataset and displays in black which pairs of slices have a p-value below 0.05:
+
+```{r}
+## Loading disparity data
+data(disparity)
+
+## Testing the pairwise difference between slices
+tests <- test.dispRity(disparity, test = wilcox.test, correction = "bonferroni")
+
+## Plotting the significance
+pair.plot(as.data.frame(tests), what = "p.value", binary = 0.05)
+```
+
+## `reduce.matrix`
+
+This function allows to reduce columns or rows of a matrix to make sure that there is enough overlap for further analysis.
+This is particularly useful if you are going to use distance matrices since it uses the `vegan::vegdist` function to test whether distances can be calculated or not.
+
+For example, if we have a patchy matrix like so (where the black squares represent available data):
+
+```{r}
+set.seed(1)
+## A 10*5 matrix
+na_matrix <- matrix(rnorm(50), 10, 5)
+## Making sure some rows don't overlap
+na_matrix[1, 1:2] <- NA
+na_matrix[2, 3:5] <- NA
+## Adding 50% NAs
+na_matrix[sample(1:50, 25)] <- NA
+## Illustrating the gappy matrix
+image(t(na_matrix), col = "black")
+```
+
+We can use the `reduce.matrix` to double check whether any rows cannot be compared.
+The functions needs as an input the type of distance that will be used, say a `"gower"` distance:
+
+```{r}
+## Reducing the matrix by row
+(reduction <- reduce.matrix(na_matrix, distance = "gower"))
+```
+
+We can not remove the rows 1 and 9 and see if that improved the overlap:
+
+```{r}
+image(t(na_matrix[-as.numeric(reduction$rows.to.remove), ]), col = "black")
+```
+
+## `slice.tree`
+
+This function is a modification of the `paleotree::timeSliceTree` function that allows to make slices through a phylogenetic tree.
+Compared to the `paleotree::timeSliceTree`, this function allows a model to decide which tip or node to use when slicing through a branch (whereas `paleotree::timeSliceTree` always choose the first available tip alphabetically).
+The models for choosing which tip or node are the same as the ones used in the `chrono.subsets` and are described in [chapter 03: specific tutorials](#chrono-subsets).
+
+The function works by using at least a tree, a slice age and a model:
+
+```{r}
+set.seed(1)
+## Generate a random ultrametric tree
+tree <- rcoal(20)
+## Add some node labels
+tree$node.label <- letters[1:19]
+## Add its root time
+tree$root.time <- max(tree.age(tree)$ages)
+
+## Slicing the tree at age 0.75
+tree_75 <- slice.tree(tree, age = 0.75, "acctran")
+
+## Showing both trees
+par(mfrow = c(1,2))
+plot(tree, main = "original tree")
+axisPhylo() ; nodelabels(tree$node.label, cex = 0.8)
+abline(v = (max(tree.age(tree)$ages) - 0.75), col = "red")
+plot(tree_75, main = "sliced tree")
+```
+
+## `slide.nodes` and `remove.zero.brlen`
+
+This function allows to slide nodes along a tree!
+In other words it allows to change the branch length leading to a node without modifying the overall tree shape.
+This can be useful to add some value to 0 branch lengths for example.
+
+The function works by taking a node (or a list of nodes), a tree and a sliding value.
+The node will be moved "up" (towards the tips) for the given sliding value.
+You can move the node "down" (towards the roots) using a negative value.
+
+```{r}
+set.seed(42)
+## Generating simple coalescent tree
+tree <- rcoal(5)
+
+## Sliding node 8 up and down
+tree_slide_up <- slide.nodes(8, tree, slide = 0.075)
+tree_slide_down <- slide.nodes(8, tree, slide = -0.075)
+
+## Display the results
+par(mfrow = c(3,1))
+plot(tree, main = "original tree") ; axisPhylo() ; nodelabels()
+plot(tree_slide_up, main = "slide up!") ; axisPhylo() ; nodelabels()
+plot(tree_slide_down, main = "slide down!") ; axisPhylo() ; nodelabels()
+```
+
+The `remove.zero.brlen` is a "clever" wrapping function that uses the `slide.nodes` function to stochastically remove zero branch lengths across a whole tree.
+This function will slide nodes up or down in successive postorder traversals (i.e.
+going down the tree clade by clade) in order to minimise the number of nodes to slide while making sure there are no silly negative branch lengths produced!
+By default it is trying to slide the nodes using 1% of the minimum branch length to avoid changing the topology too much.
+
+```{r}
+set.seed(42)
+## Generating a tree
+tree <- rtree(20)
+
+## Adding some zero branch lengths (5)
+tree$edge.length[sample(1:Nedge(tree), 5)] <- 0
+
+## And now removing these zero branch lengths!
+tree_no_zero <- remove.zero.brlen(tree)
+
+## Exaggerating the removal (to make it visible)
+tree_exaggerated <- remove.zero.brlen(tree, slide = 1)
+
+## Check the differences
+any(tree$edge.length == 0)
+any(tree_no_zero$edge.length == 0)
+any(tree_exaggerated$edge.length == 0)
+
+## Display the results
+par(mfrow = c(3,1))
+plot(tree, main = "with zero edges")
+plot(tree_no_zero, main = "without zero edges!")
+plot(tree_exaggerated, main = "with longer edges")
+```
+
+## `tree.age`
+
+This function allows to quickly calculate the ages of each tips and nodes present in a tree.
+
+```{r}
+set.seed(1)
+tree <- rtree(10)
+## The tree age from a 10 tip tree
+tree.age(tree)
+```
+
+It also allows to set the age of the root of the tree:
+
+```{r}
+## The ages starting from -100 units
+tree.age(tree, age = 100)
+```
+
+Usually tree age is calculated from the present to the past (e.g. in million years ago) but it is possible to reverse it using the `order = present` option:
+
+```{r}
+## The ages in terms of tip/node height
+tree.age(tree, order = "present")
+```
+
+<!--chapter:end:05_other-functionalities.Rmd-->
 
@@
 
-## The `dispRity` object content
+## The `dispRity` object content {#disprity-object}
+
 The functions above are utilities to easily and safely access different elements in the `dispRity` object.
@@
 
-<!-- ## Modularity
-
-### Making your own metrics
-
-### Making your own tests
-
-## Where is this going?
- -->
-<!-- ## Running `dispRity` in parallel
-
-The computationally intensive function `dispRity` have a `parallel` option to speed up their calculations.
-
-This option requires the package `snow` and takes arguments that are to be passed to  -->
-
-<!--chapter:end:05_guts.Rmd-->
-
----
-title: "Palaeobiology demo: disparity-through-time and within groups"
-author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
-date: "`r Sys.Date()`"
-bibliography: [../References.bib, ../packages.bib]
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Palaeobiology demo: disparity-through-time and within groups
-
-This demo aims to give quick overview of the `dispRity` package (v.`r version_release`) for palaeobiology analyses of disparity, including disparity through time analyses.
-
-This demo showcases a typical disparity-through-time analysis: we are going to test whether the disparity changed through time in a subset of eutherian mammals from the last 100 million years using a dataset from Beck and Lee (2014).
-
-## Before starting
-
-### The morphospace
-
-In this example, we are going to use a subset of the data from @beckancient2014.
-See the [example data](#example-data) description for more details.
-Briefly, this dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
-The ordinated matrix will represent our full morphospace, i.e. all the mammalian morphologies that ever existed through time (for this dataset).
-
-```{r, fig.width=6, fig.height=6}
-## Loading demo and the package data
-library(dispRity)
-
-## Setting the random seed for repeatability
-set.seed(123)
-
-## Loading the ordinated matrix/morphospace:
-data(BeckLee_mat50)
-head(BeckLee_mat50[,1:5])
-
-dim(BeckLee_mat50)
-## The morphospace contains 50 taxa and has 48 dimensions (or axes)
-
-## Showing a list of first and last occurrences data for some fossils
-data(BeckLee_ages)
-head(BeckLee_ages)
-
-## Plotting a phylogeny
-data(BeckLee_tree)
-plot(BeckLee_tree, cex = 0.7)
-axisPhylo(root = 140)
-```
-
-> You can have an even nicer looking tree if you use the `strap` package!
-
-```{r, eval = TRUE}
-if(!require(strap)) install.packages("strap")
-strap::geoscalePhylo(BeckLee_tree, cex.tip = 0.7, cex.ts = 0.6)
-```
-
-## A disparity-through-time analysis
-
-### Splitting the morphospace through time
-
-One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time subsets that contain the total number of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
-Basically, the full morphospace represents the total number of morphologies across all time and will be greater than any of the time subsets of the morphospace.
-
-The `dispRity` package provides a `chrono.subsets` function that allows users to split the morphospace into time slices (using `method = continuous`) or into time bins (using `method = discrete`).
-In this example, we are going to split the morphospace into five equal time bins of 20 million years long from 100 million years ago to the present.
-We will also provide to the function a table containing the first and last occurrences dates for some fossils to take into account that some fossils might occur in several of our different time bins.
-
-```{r}
-## Creating the vector of time bins ages
-(time_bins <- rev(seq(from = 0, to = 100, by = 20)))
-
-## Splitting the morphospace using the chrono.subsets function
-(binned_morphospace <- chrono.subsets(data = BeckLee_mat50, tree = BeckLee_tree,
-    method = "discrete", time = time_bins, inc.nodes = FALSE,
-    FADLAD = BeckLee_ages))
-```
-
-The output object is a `dispRity` object (see more about that [here](#The-guts-of-the-dispRity-package).
-In brief, however, `dispRity` objects are lists of different elements (i.e.
-disparity results, morphospace time subsets, morphospace attributes, etc.) that display only a summary of the object when calling the object to avoiding filling the `R` console with superfluous output.
-
-```{r}
-## Printing the class of the object
-class(binned_morphospace)
-
-## Printing the content of the object
-str(binned_morphospace)
-names(binned_morphospace)
-
-## Printing the object as a dispRity class
-binned_morphospace
-```
-
-> These objects will gradual.splitly contain more information when completing the following steps in the disparity-through-time analysis.
-
-### Bootstrapping the data
-
-Once we obtain our different time subsets, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
-The bootstrapping allows us to make each subset more robust to outliers and the rarefaction allows us to compare subsets with the same number of taxa to remove sampling biases (i.e.
-more taxa in one subset than the others).
-The `boot.matrix` function bootstraps the `dispRity` object and the `rarefaction` option within performs rarefaction.
-
-```{r}
-## Bootstrapping each time subset 100 times (default)
-(boot_bin_morphospace <- boot.matrix(binned_morphospace))
-
-## Getting the minimum number of rows (i.e. taxa) in the time subsets
-min(size.subsets(boot_bin_morphospace))
-
-## Bootstrapping each time subset 100 times and rarefying them 
-(rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
-    rarefaction = 6))
-```
-
-### Calculating disparity
-
-We can now calculate the disparity within each time subsets along with some confidence intervals generated by the pseudoreplication step above (bootstraps/rarefaction).
-Disparity can be calculated in many ways and this package allows users to come up with their own disparity metrics.
-For more details, please refer to the [`dispRity` metric section](#disparity-metrics).
-
-In this example, we are going to calculate the spread of the data in each time subset by calculating disparity as the sum of the variance of each dimension of the morphospace in each time subset using the `dispRity` function.
-Thus, in this example, disparity is defined by the multi-dimensional variance of each time subset (i.e. the spread of the taxa within the morphospace).
-Note that this metric comes with a caveat (not solved here) since it ignores covariances among the dimensions of the morphospace.
-We use this here because it is a standard metric used in disparity-through-time analysis [@Wills1994].
-
-```{r}
-## Calculating disparity for the bootstrapped data
-(boot_disparity <- dispRity(boot_bin_morphospace, metric = c(sum, variances)))
-
-## Calculating disparity for the rarefied data
-(rare_disparity <- dispRity(rare_bin_morphospace, metric = c(sum, variances)))
-```
-
-The `dispRity` function does not actually display the calculated disparity values but rather only the properties of the disparity object (size, subsets, metric, etc.).
-To display the actual calculated scores, we need to summarise the disparity object using the S3 method `summary` that is applied to a `dispRity` object (see `?summary.dispRity` for more details).
-
-As for any `R` package, you can refer to the help files for each individual function for more details.
-
-```{r}
-## Summarising the disparity results
-summary(boot_disparity)
-summary(rare_disparity)
-```
-> The summary.dispRity function comes with many options on which values to calculate (central tendency and quantiles) and on how many digits to display.
-Refer to the function's manual for more details.
-
-### Plotting the results
-
-It is sometimes easier to visualise the results in a plot than in a table.
-For that we can use the `plot` S3 function to plot the `dispRity` objects (see `?plot.dispRity` for more details).
-
-```{r, fig.width=12, fig.height=6}
-## Graphical options
-quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
-
-## Plotting the bootstrapped and rarefied results
-plot(boot_disparity, type = "continuous", main = "bootstrapped results")
-plot(rare_disparity, type = "continuous", main = "rarefied results")
-```
-
-Testing differences
--------------------
-
-Finally, to draw some valid conclusions from these results, we can apply some statistical tests.
-We can test, for example, if mammalian disparity changed significantly through time over the last 100 million years.
-To do so, we can compare the means of each time-bin in a sequential manner to see whether the disparity in bin *n* is equal to the disparity in bin *n+1*, and whether this is in turn equal to the disparity in bin *n+2*, etc.
-Because our data is temporally autocorrelated (i.e.
-what happens in bin *n+1* depends on what happened in bin *n*) and pseudoreplicated (i.e.
-each bootstrap draw creates non-independent time subsets because they are all based on the same time subsets), we apply a non-parametric mean comparison: the `wilcox.test`.
-Also, we need to apply a p-value correction (e.g. Bonferroni correction) to correct for multiple testing (see `?p.adjust` for more details).
-
-```{r}
-## Testing the differences between bins in the bootstrapped dataset.
-test.dispRity(boot_disparity, test = wilcox.test, comparison = "sequential",
-    correction = "bonferroni")
-
-## Testing the differences between bins in the rarefied dataset.
-test.dispRity(rare_disparity, test = wilcox.test, comparison = "sequential",
-    correction = "bonferroni")
-```
-
-Here our results show significant changes in disparity through time between all time bins (all p-values < 0.05).
-However, when looking at the rarefied results, there is no significant difference between the time bins in the Palaeogene (60-40 to 40-20 Mya), suggesting that the differences detected in the first test might just be due to the differences in number of taxa sampled (13 or 6 taxa) in each time bin.
-
-<!--chapter:end:06_palaeo-demo.Rmd-->
+<!--chapter:end:06_guts.Rmd-->
 
@@
 ---
-title: "Future directions and references"
-author: "Thomas Guillerme"
+title: "Palaeobiology demo: disparity-through-time and within groups"
+author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
 date: "`r Sys.Date()`"
-output: rmarkdown::html_vignette
 bibliography: [../References.bib, ../packages.bib]
-bst: sysbio.bst
-vignette: >
-  %\VignetteIndexEntry{future directions}
-  %\VignetteEngine{knitr::rmarkdown}
-  \usepackage[utf8]{inputenc}
+output:
+  html_document: default
+  pdf_document: default
 ---
 
-# Developing `dispRity`
-
-Everyone is more than welcome to help developing this package, whether it is through developing functions (metrics, tests, plots, etc...), the manual (more examples, typos, etc...) or completely new functionalities!
-
-## More functions
-
-## More manuals
-
-<!-- 
-* Beck, R.
-M., & Lee, M.
-S.
-(2014).
-Ancient dates or accelerated rates?
-Morphological clocks and the antiquity of placental mammals.
-Proceedings of the Royal Society of London B: Biological Sciences, 281(1793), 20141278.
-* Cooper, N., & Guillerme, T. (in prep.). Coming soonish!.
-* Diaz, S., Kattge, J., Cornelissen, J.H., Wright, I.J., Lavorel, S., Dray, S., Reu, B., Kleyer, M., Wirth, C., Prentice, I.C.
-and Garnier, E.
-(2016).
-The global spectrum of plant form and function.
-Nature, 529(7585), 167.
-* Donohue, I., Petchey, O.L., Montoya, J.M., Jackson, A.L., McNally, L., Viana, M., Healy, K., Lurgi, M., O'Connor, N.E.
-and Emmerson, M.C, (2013).
-On the dimensionality of ecological stability.
-Ecology letters, 16(4), 421-429.
-* Wills, M. A., Briggs, D. E., & Fortey, R. A. (1994). Disparity as an evolutionary index: a comparison of Cambrian and Recent arthropods. Paleobiology, 20(2), 93-130. -->
-
-<!--chapter:end:08_future_directions.Rmd-->
+# Palaeobiology demo: disparity-through-time and within groups
+
+This demo aims to give quick overview of the `dispRity` package (v.`r version_release`) for palaeobiology analyses of disparity, including disparity through time analyses.
+
+This demo showcases a typical disparity-through-time analysis: we are going to test whether the disparity changed through time in a subset of eutherian mammals from the last 100 million years using a dataset from @beckancient2014.
+
+## Before starting
+
+### The morphospace
+
+In this example, we are going to use a subset of the data from @beckancient2014.
+See the [example data](#example-data) description for more details.
+Briefly, this dataset contains an ordinated matrix of the Gower distance between 50 mammals based (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).
+The ordinated matrix will represent our full morphospace, i.e. all the mammalian morphologies that ever existed through time (for this dataset).
+
+```{r, fig.width=6, fig.height=6}
+## Loading demo and the package data
+library(dispRity)
+
+## Setting the random seed for repeatability
+set.seed(123)
+
+## Loading the ordinated matrix/morphospace:
+data(BeckLee_mat50)
+data(BeckLee_mat99)
+head(BeckLee_mat50[,1:5])
+
+dim(BeckLee_mat50)
+## The morphospace contains 50 taxa and has 48 dimensions (or axes)
+
+## Showing a list of first and last occurrences data for some fossils
+data(BeckLee_ages)
+head(BeckLee_ages)
+
+## Plotting a phylogeny
+data(BeckLee_tree)
+plot(BeckLee_tree, cex = 0.7)
+axisPhylo(root = 140)
+```
+
+> You can have an even nicer looking tree if you use the `strap` package!
+
+```{r, eval = TRUE}
+if(!require(strap)) install.packages("strap")
+strap::geoscalePhylo(BeckLee_tree, cex.tip = 0.7, cex.ts = 0.6)
+```
+
+### Setting up your own data
+
+I greatly encourage you to follow along this tutorial with your very own data: it is more exciting and, ultimately, that's probably your objective.
+
+> What data can I use?
+
+You can use any type of morphospace in any dataset form (`"matrix"`, `"data.frame"`).
+Throughout this tutorial, you we assume you are using the (loose) morphospace definition from @Guillerme2020: any matrix were columns are traits and rows are observations (in a distance matrix, columns are still trait, i.e.
+"distance to species A", etc.).
+We won't cover it here but you can also use lists of matrices and list of trees.
+
+> How should I format my data for this tutorial?
+
+To go through this tutorial you will need:
+
+ * A matrix with tip data
+ * A phylogenetic tree 
+ * A matrix with tip and node data
+ * A table of first and last occurrences data (FADLAD)
+
+If you are missing any of these, fear not, here are a couple of functions to simulate the missing data, it will surely make your results look funky but it'll let you go through the tutorial.
+
+```{r}
+## Functions to get simulate a PCO looking like matrix from a tree
+i.need.a.matrix <- function(tree) {
+    matrix <- space.maker(elements = Ntip(tree), dimensions = Ntip(tree), distribution = rnorm,
+                          scree = rev(cumsum(rep(1/Ntip(tree), Ntip(tree)))))
+    rownames(matrix) <- tree$tip.label
+    return(matrix)
+}
+
+## Function to simulate a tree
+i.need.a.tree <- function(matrix) {
+    tree <- rtree(nrow(matrix))
+    tree$root.time <- max(tree.age(tree)$age)
+    tree$tip.label <- rownames(matrix)
+    tree$node.label <- paste0("n", 1:(nrow(matrix)-1))
+    return(tree)
+}
+
+## Function to simulate some "node" data
+i.need.node.data <- function(matrix, tree) {
+    matrix_node <- space.maker(elements = Nnode(tree), dimensions = ncol(matrix),
+                               distribution = rnorm, scree = apply(matrix, 2, var))
+    if(!is.null(tree$node.label)) {
+        rownames(matrix_node) <- tree$node.label
+    } else {
+        rownames(matrix_node) <- paste0("n", 1:(nrow(matrix)-1))
+    }
+    return(rbind(matrix, matrix_node))
+}
+
+## Function to simulate some "FADLAD" data
+i.need.FADLAD <- function(tree) {
+    tree_ages <- tree.age(tree)[1:Ntip(tree),]
+    return(data.frame(FAD = tree_ages[,1], LAD = tree_ages[,1], row.names = tree_ages[,2]))
+}
+```
+
+You can use these functions for the generating the data you need. For example
+
+```{r, eval = FALSE}
+## Aaaaah I don't have FADLAD data!
+my_FADLAD <- i.need.FADLAD(tree)
+## Sorted.
+```
+
+In the end this is what your data should be named to facilitate the rest of this tutorial (fill in yours here):
+
+```{r}
+## A matrix with tip data
+my_matrix <- BeckLee_mat50
+
+## A phylogenetic tree 
+my_tree <- BeckLee_tree
+
+## A matrix with tip and node data
+my_tip_node_matrix <- BeckLee_mat99
+
+## A table of first and last occurrences data (FADLAD)
+my_fadlad <- BeckLee_ages
+```
+
+## A disparity-through-time analysis
+
+### Splitting the morphospace through time
+
+One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time subsets that contain the total number of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
+Basically, the full morphospace represents the total number of morphologies across all time and will be greater than any of the time subsets of the morphospace.
+
+The `dispRity` package provides a `chrono.subsets` function that allows users to split the morphospace into time slices (using `method = continuous`) or into time bins (using `method = discrete`).
+In this example, we are going to split the morphospace into five equal time bins of 20 million years long from 100 million years ago to the present.
+We will also provide to the function a table containing the first and last occurrences dates for some fossils to take into account that some fossils might occur in several of our different time bins.
+
+```{r}
+## Creating the vector of time bins ages
+(time_bins <- rev(seq(from = 0, to = 100, by = 20)))
+
+## Splitting the morphospace using the chrono.subsets function
+(binned_morphospace <- chrono.subsets(data = my_matrix, tree = my_tree,
+    method = "discrete", time = time_bins, inc.nodes = FALSE,
+    FADLAD = my_fadlad))
+```
+
+The output object is a `dispRity` object (see more about that [here](#The-guts-of-the-dispRity-package).
+In brief, `dispRity` objects are lists of different elements (i.e.
+disparity results, morphospace time subsets, morphospace attributes, etc.) that display only a summary of the object when calling the object to avoiding filling the `R` console with superfluous output.
+It also allows easy plotting/summarising/analysing for repeatability down the line but we will not go into this right now.
+
+```{r}
+## Printing the class of the object
+class(binned_morphospace)
+
+## Printing the content of the object
+str(binned_morphospace)
+names(binned_morphospace)
+
+## Printing the object as a dispRity class
+binned_morphospace
+```
+
+> These objects will gradually contain more information when completing the following steps in the disparity-through-time analysis.
+
+### Bootstrapping the data
+
+Once we obtain our different time subsets, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
+The bootstrapping allows us to make each subset more robust to outliers and the rarefaction allows us to compare subsets with the same number of taxa to remove sampling biases (i.e.
+more taxa in one subset than the others).
+The `boot.matrix` function bootstraps the `dispRity` object and the `rarefaction` option within performs rarefaction.
+
+```{r}
+## Bootstrapping each time subset 100 times (default)
+(boot_bin_morphospace <- boot.matrix(binned_morphospace))
+
+## Getting the minimum number of rows (i.e. taxa) in the time subsets
+min(size.subsets(boot_bin_morphospace))
+
+## Bootstrapping each time subset 100 times and rarefying them 
+(rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
+    rarefaction = 6))
+```
+
+> Note how information is adding up to the `dispRity` object.
+
+### Calculating disparity
+
+We can now calculate the disparity within each time subsets along with some confidence intervals generated by the pseudoreplication step above (bootstraps/rarefaction).
+Disparity can be calculated in many ways and this package allows users to come up with their own disparity metrics.
+For more details, please refer to the [`dispRity` metric section](#disparity-metrics) (or directly use [`moms`](https://tguillerme.shinyapps.io/moms/)).
+
+In this example, we are going to look at how the spread of the data in the morphospace through time.
+For that we are going to use the sum of the variance from each dimension of the morphospace in the morphospace.
+We highly recommend using a metric that makes sense for your specific analysis and for your specific dataset and not just because everyone uses it [@moms, @Guillerme2020]!
+
+> How can I be sure that the metric is the most appropriate for my morphospace and question?
+
+This is not a straightforward question but you can use the `test.metric` function to check your assumptions (more details [here](#disparity-metrics)): basically what `test.metric` does is modifying your morphospace using a null process of interest (e.g.
+changes in size) and checks whether your metric does indeed pick up that change.
+For example here, let see if the sum of variances picks up changes in size but not random changes:
+
+```{r}
+my_test <- test.metric(my_matrix, metric = c(sum, variances), shifts = c("random", "size"))
+summary(my_test)
+plot(my_test)
+```
+
+We see that changes in the inner size (see @moms for more details) is actually picked up by the sum of variances but not random changes or outer changes. Which is a good thing!
+
+As you've noted, the sum of variances is defined in `test.metric` as `c(sum, variances)`.
+This is a core bit of the `dispRity` package were you can define your own metric as a function or a set of functions.
+You can find more info about this in the [`dispRity` metric section](#disparity-metrics) but in brief, the `dispRity` package considers metrics by their "dimensions" level which corresponds to what they output.
+For example, the function `sum` is a dimension level 1 function because no matter the input it outputs a single value (the sum), `variances` on the other hand is a dimension level 2 function because it will output the variance of each column in a matrix (an example of a dimensions level 3 would be the function `var` that outputs a matrix).
+The `dispRity` package always automatically sorts the dimensions levels: it will always run dimensions level 3 > dimensions level 2 > and dimensions level 1.
+In this case both `c(sum, variances)` and `c(variances, sum)` will result in actually running `sum(variances(matrix))`.
+
+Anyways, let's calculate the sum of variances on our bootstrapped and rarefied morphospaces:
+
+```{r}
+## Calculating disparity for the bootstrapped data
+(boot_disparity <- dispRity(boot_bin_morphospace, metric = c(sum, variances)))
+
+## Calculating disparity for the rarefied data
+(rare_disparity <- dispRity(rare_bin_morphospace, metric = c(sum, variances)))
+```
+
+To display the actual calculated scores, we need to summarise the disparity object using the S3 method `summary` that is applied to a `dispRity` object (see `?summary.dispRity` for more details).
+By the way, as for any `R` package, you can refer to the help files for each individual function for more details.
+
+```{r}
+## Summarising the disparity results
+summary(boot_disparity)
+summary(rare_disparity)
+```
+
+> The `summary.dispRity` function comes with many options on which values to calculate (central tendency and quantiles) and on how many digits to display.
+Refer to the function's manual for more details.
+
+### Plotting the results
+
+It is sometimes easier to visualise the results in a plot than in a table.
+For that we can use the `plot` S3 function to plot the `dispRity` objects (see `?plot.dispRity` for more details).
+
+```{r, fig.width=12, fig.height=6}
+## Graphical options
+quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
+
+## Plotting the bootstrapped and rarefied results
+plot(boot_disparity, type = "continuous", main = "bootstrapped results")
+plot(rare_disparity, type = "continuous", main = "rarefied results")
+```
+
+Nice. The curves look pretty similar.
+
+> Same as for the `summary.dispRity` function, check out the `plot.dispRity` manual for the many, many options available.
+
+### Testing differences
+
+Finally, to draw some valid conclusions from these results, we can apply some statistical tests.
+We can test, for example, if mammalian disparity changed significantly through time over the last 100 million years.
+To do so, we can compare the means of each time-bin in a sequential manner to see whether the disparity in bin *n* is equal to the disparity in bin *n+1*, and whether this is in turn equal to the disparity in bin *n+2*, etc.
+Because our data is temporally autocorrelated (i.e.
+what happens in bin *n+1* depends on what happened in bin *n*) and pseudoreplicated (i.e.
+each bootstrap draw creates non-independent time subsets because they are all based on the same time subsets), we apply a non-parametric mean comparison: the `wilcox.test`.
+Also, we need to apply a p-value correction (e.g. Bonferroni correction) to correct for multiple testing (see `?p.adjust` for more details).
+
+```{r}
+## Testing the differences between bins in the bootstrapped dataset.
+test.dispRity(boot_disparity, test = wilcox.test, comparison = "sequential",
+    correction = "bonferroni")
+
+## Testing the differences between bins in the rarefied dataset.
+test.dispRity(rare_disparity, test = wilcox.test, comparison = "sequential",
+    correction = "bonferroni")
+```
+
+Here our results show significant changes in disparity through time between all time bins (all p-values < 0.05).
+However, when looking at the rarefied results, there is no significant difference between the time bins in the Palaeogene (60-40 to 40-20 Mya), suggesting that the differences detected in the first test might just be due to the differences in number of taxa sampled (13 or 6 taxa) in each time bin.
+
+## Some more advanced stuff
+
+The previous section detailed some of the basic functionalities in the `dispRity` package but of course, you can do some much more advanced analysis, here is just a list of some specific tutorials from this manual that you might be interested in:
+
+ * [Time slicing](#time-slicing): an alternative method to look at disparity through time that allows you to specify evolutionary models [@time-slice].
+ * [Many more disparity metrics](#disparity-metrics): there are many, many different things you might be interested to measure in your morphospace!
+This manual has some extended documentation on what to use (or check @moms).
+ * Many more ways to look at disparity: you can for example, [use distributions rather than point estimates](#disparity-distribution) for your disparity metric (e.g.
+the variances rather than the sum of variances); or calculate [disparity from non ordinated matrices](#other-matrices) or even [from multiple matrices and trees](#multi.input).
+ * And finally there are much more advanced statistical tests you might be interested in using, such as the [NPMANOVA](#adonis), the ["disparity-through-time test"](#dtt), using a [null model approach](#null-test) or some [model fitting](#model-fitting)...
+
+You can even come up with your own ideas, implementations and modifications of the package: the `dispRity` package is a modular and collaborative package and I encourage you to contact me (guillert@tcd.e) for any ideas you have about adding new features to the package (whether you have them already implemented or not)!
+
+<!--chapter:end:08_palaeo-demo.Rmd-->
+
+---
+title: "Morphometric geometric demo: a between group analysis"
+author: "Thomas Guillerme (guillert@tcd.ie)"
+date: "`r Sys.Date()`"
+bibliography: [../References.bib, ../packages.bib]
+output:
+  html_document: default
+  pdf_document: default
+---
+
+# Morphometric geometric demo: a between group analysis
+
+This demo aims to give quick overview of the `dispRity` package (v.`r version_release`) for palaeobiology analyses of disparity, including disparity through time analyses.
+
+This demo showcases a typical between groups geometric morphometric analysis: we are going to test whether the disparity in two species of salamander (plethodons!) are different and in which ways they are different.
+
+## Before starting
+
+Here we are going to use the `geomorph` `plethodon` dataset that is a set of 12 2D landmark coordinates for 40 specimens from two species of salamanders.
+This section will really quickly cover how to make a Procrustes sumperimposition analysis and create a `geomorph` data.frame to have data ready for the `dispRity` package.
+
+```{r}
+## Loading geomorph
+library(geomorph)
+
+## Loading the plethodon dataset
+data(plethodon)
+
+## Running a simple Procrustes superimposition
+gpa_plethodon <- gpagen(plethodon$land)
+
+## Making a geomorph data frame object with the species and sites attributes
+gdf_plethodon <- geomorph.data.frame(gpa_plethodon,
+                                     species = plethodon$species,
+                                     site = plethodon$site)
+```
+
+You can of course use your very own landmark coordinates dataset (though you will have to do some modifications in the scripts that will come below - they will be easy though!).
+
+```{r}
+## You can replace the gdf_plethodon by your own geomorph data frame!
+my_geomorph_data <- gdf_plethodon
+```
+
+### The morphospace
+
+The first step of every disparity analysis is to define your morphospace.
+
+> Note that this is actually not true at all and kept as a erroneous sentence: the first step of your disparity analysis should be to define your question!
+
+Our question here will be: is there a difference in disparity between the different species of salamanders and between the different sites (allopatric and sympatric)?
+
+OK, now we can go to the _second_ step of every disparity analysis: defining the morphospace.
+Here we will define it with the ordination of all possible Procrustes superimposed plethodon landmark coordinates.
+You can do this directly in `dispRity` using the `geomorph.ordination` function that can input a geomorph data frame:
+
+```{r}
+## The morphospace
+morphospace <- geomorph.ordination(gdf_plethodon)
+```
+
+This automatically generates a `dispRity` object with the information of each groups.
+You can find more information about `dispRity` objects [here](#disprity-object) but basically it summarises the content of your object without spamming your R console and is associated with many utility functions like `summary` or `plot`.
+For example here you can quickly visualise the two first dimensions of your space using the `plot` function:
+
+```{r}
+## The dispRity object
+morphospace
+
+## Plotting the morphospace
+plot(morphospace)
+## Note that this only displays the two last groups (site.Allo and site.Symp) since they overlap!
+```
+
+The `dispRity` package function comes with a lot of documentation of examples so don't hesitate to type `plot.dispRity` to check more plotting options.
+
+## Calculating disparity
+
+Now that we have our morphospace, we can think about what we want to measure.
+Two aspects of disparity that would be interesting for our question (is there a difference in disparity between the different species of salamanders and between the different sites?) would be the differences in size in the morphospace (do both groups occupy the same amount of morphospace) and position in the morphospace (do the do groups occupy the same position in the morphospace?).
+
+To choose which metric would cover best these two aspects, please check the @moms paper and associated [app](https://tguillerme.shinyapps.io/moms/).
+Here we are going to use the procrustes variance (`geomorph::morphol.disparity`) for measuring the size of the trait space and the average displacements [@moms] for the position in the trait space.
+
+```{r}
+## Defining a the procrustes variance metric
+## (as in geomorph::morphol.disparity)
+proc.var <- function(matrix) {sum(matrix^2)/nrow(matrix)}
+## The size metric
+test_size <- test.metric(morphospace, metric = proc.var,
+                         shifts = c("random", "size"))
+plot(test_size)
+summary(test_size)
+
+## The position metric
+test_position <- test.metric(morphospace, metric = c(mean, displacements),
+                         shifts = c("random", "position"))
+plot(test_position)
+summary(test_position)
+```
+
+You can see [here](#test-metric) for more details on the `test.metric` function but basically these graphs are showing that there is a relation between changes in size and in position for each metric.
+Note that there are some caveats here but the selection of the metric is just for the sake of the example!
+
+Note also the format of defining the disparity metrics here using `metric = c(mean, displacements)` or `metric = proc.var`.
+This is a core bit of the `dispRity` package were you can define your own metric as a function or a set of functions.
+You can find more info about this in the [`dispRity` metric section](#disparity-metrics) but in brief, the `dispRity` package considers metrics by their "dimensions" level which corresponds to what they output.
+For example, the function `mean` is a dimension level 1 function because no matter the input it outputs a single value (the mean), `displacements` on the other hand is a dimension level 2 function because it will output the ratio between the distance from the centroid and from the centre of the trait space for each row in a matrix (an example of a dimensions level 3 would be the function `var` that outputs a matrix).
+The `dispRity` package always automatically sorts the dimensions levels: it will always run dimensions level 3 > dimensions level 2 > and dimensions level 1.
+In this case both `c(mean, displacements)` and `c(mean, displacements)` will result in actually running `mean(displacements(matrix))`.
+Alternatively you can define your metric prior to the disparity analysis like we did for the `proc.var` function.
+
+Anyways, we can measure disparity using these two metrics on all the groups as follows:
+
+```{r}
+## Bootstrapped disparity
+disparity_size <-  dispRity(boot.matrix(morphospace), metric = proc.var)
+disparity_position <- dispRity(boot.matrix(morphospace), metric = c(mean, displacements))
+```
+
+Note that here we use the `boot.matrix` function for quickly bootstrapping the matrix.
+This is not an essential step in this kind of analysis but it allows to "reduce" the effect of outliers and create a distribution of disparity measures (rather than single point estimates).
+
+## Analyse the results
+
+We can visualise the results using the `plot` function on the resulting disparity objects (or summarising them using `summary`):
+
+```{r}
+## Plotting the results
+par(mfrow = c(1,2))
+plot(disparity_size, main = "group sizes", las = 2, xlab = "")
+plot(disparity_position, main = "group positions", las = 2, xlab = "")
+
+## Summarising the results
+summary(disparity_size)
+summary(disparity_position)
+```
+
+Just from looking at the data, we can guess that there is not much difference in terms of morphospace occupancy and position for the species but there is on for the sites (allopatric or sympatric).
+We can test it using a simple non-parametric mean difference test (e.g. `wilcox.test`) using the `dispRity` package.
+
+```{r}
+## Testing the differences
+test.dispRity(disparity_size, test = wilcox.test, correction = "bonferroni")
+test.dispRity(disparity_position, test = wilcox.test, correction = "bonferroni")
+```
+
+So by applying the tests we see a difference in terms of position between each groups and differences in size between groups but between the species.
+
+<!--chapter:end:09_gmm-demo.Rmd-->
 
@@
 
-<!--chapter:end:09_references.Rmd-->
+<!--chapter:end:10_references.Rmd-->

### 2020-08-18 inst/gitbook/05_other-functionalities.Rmd
Updating Claddis calls across all of dispRity.

@@
 This is yet another function for calculating distance matrices.
-There are many functions for calculating pairwise distance matrices in `R` (`stats::dist`, `vegan::vegdist`, `cluster::daisy` or `Claddis::MorphDistMatrix`) but this one is the `dispRity` one.
-It is slightly different to the ones mentioned above (though not that dissimilar from `Claddis::MorphDistMatrix`) in the fact that it focuses on comparing discrete morphological characters and tries to solve all the problems linked to these kind of matrices (especially dealing with special tokens).
+There are many functions for calculating pairwise distance matrices in `R` (`stats::dist`, `vegan::vegdist`, `cluster::daisy` or `Claddis::calculate_morphological_distances`) but this one is the `dispRity` one.
+It is slightly different to the ones mentioned above (though not that dissimilar from `Claddis::calculate_morphological_distances`) in the fact that it focuses on comparing discrete morphological characters and tries to solve all the problems linked to these kind of matrices (especially dealing with special tokens).
 

### 2020-08-19 inst/gitbook/02_getting-started.Rmd

@@
 ## Ordinating the example data from Claddis
-Claddis.ordination(Michaux1989) 
+Claddis.ordination(michaux_1989)
 ```
@@
 See more info in the function manual (`?Claddis.ordination`).
-Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.
+Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::calculate_morphological_distances` and `stats::cmdscale`.
 

### 2020-09-03 R/Claddis.ordination_fun.R
Updated 1.4.2 (Claddis 0.6 update)

@@ -3,27 +3,29 @@ convert.to.Claddis <- function(data) {
     output <- list()
     
     ## Topper
-    output$Topper$Head <- "List generated by dispRity::Claddis.ordination"
+    output$topper$head <- "List generated by dispRity::Claddis.ordination"
 
     ## Matrix_1
-    output$Matrix_1$BlockName <- NA
-    output$Matrix_1$Datatype <- "STANDARD"
-    output$Matrix_1$Matrix <- do.call(rbind, data)
-    output$Matrix_1$Ordering <- rep("unord", length(data[[1]]))
-    output$Matrix_1$Weights <- rep(1, length(data[[1]]))
+    output$matrix_1$block_name <- NA
+    output$matrix_1$datatype <- "STANDARD"
+    output$matrix_1$matrix <- do.call(rbind, data)
+    output$matrix_1$ordering <- rep("unord", length(data[[1]]))
+    output$matrix_1$character_weights <- rep(1, length(data[[1]]))
     op <- options(warn = -1)
-    output$Matrix_1$MinVals <- unlist(apply(output$Matrix_1$Matrix, 2, function(x) min(as.numeric(x), na.rm = TRUE)))
-    is_nas <- is.na(output$Matrix_1$MinVals)
-    output$Matrix_1$MaxVals <- unlist(apply(output$Matrix_1$Matrix, 2, function(x) max(as.numeric(x), na.rm = TRUE)))
-    output$Matrix_1$Characters$Symbols <- unique(as.numeric(unlist(data)[-which(is.na(unlist(data)))]))
+    output$matrix_1$minimum_values <- unlist(apply(output$matrix_1$matrix, 2, function(x) min(as.numeric(x), na.rm = TRUE)))
+    is_nas <- is.na(output$matrix_1$minimum_values)
+    output$matrix_1$maximum_values <- unlist(apply(output$matrix_1$matrix, 2, function(x) max(as.numeric(x), na.rm = TRUE)))
+    output$matrix_1$characters$symbols <- unique(as.numeric(unlist(data)[-which(is.na(unlist(data)))]))
     op <- options(warn = 0)
-    output$Matrix_1$Characters$Missing <- "?"
-    output$Matrix_1$Characters$Gap <- "-"
+    output$matrix_1$characters$missing <- "?"
+    output$matrix_1$characters$gap <- "-"
 
     ## Transform the missing data into "?"
[…]

### 2020-09-04 inst/gitbook/03_specific-tutorials.Rmd
Added manuals for group.dist

@@
 1 | `func.eve` | The functional evenness (the minimal spanning tree distances evenness) | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FEve` but without abundance)|
+1 | `group.dist` | The distance between two groups | `dispRity` |
 1 | `mode.val` | The modal value | `dispRity` |
@@
 In doubt, it's probably better to enter your `base` (for `angles`) or your `hyperplane` (for `deviations`) manually so you're sure you know what the function is measuring.
+
+#### Between group metrics
+
+The `group.dist` metric allows to measure the distance between two groups in the multidimensional space.
+This function needs to intake several groups and use the option `between.groups = TRUE` in the `dispRity` function.
+It calculates the vector normal distance (euclidean) between two groups and returns 0 if that distance is negative.
+Note that it is possible to set up which quantiles to consider for calculating the distances between groups.
+For example, one might be interested in only considering the 95% CI for each group.
+This can be done through the option `probs = c(0.025, 0.975)` that is passed to the `quantile` function.
+It is also possible to use this function to measure the distance between the groups centroids by calculating the 50% quantile (`probs = c(0.5)`).
+
+```{r}
+## Creating a dispRity object with two groups
+grouped_space <- custom.subsets(dummy_space,
+                      group = list(c(1:5), c(6:10)))
+
+## Measuring the minimum distance between both groups
+summary(dispRity(grouped_space, metric = group.dist,
+                 between.groups = TRUE))
+
+## Measuring the centroid distance between both groups
+summary(dispRity(grouped_space, metric = group.dist,
+                 between.groups = TRUE, probs = 0.5))
+
+## Measuring the distance between both group's 75% CI
+summary(dispRity(grouped_space, metric = group.dist,
+                 between.groups = TRUE, probs = c(0.25, 0.75)))
+```
 

### 2020-09-04 tests/testthat/test-dispRity.metric.R

@@ -5,7 +5,7 @@ context("dispRity.metric")
 test_that("dimension generic", {
     expect_equal(capture_output(dimension.level3.fun()), "No implemented Dimension level 3 functions implemented in dispRity!\nYou can create your own by using: ?make.metric")
     expect_equal(capture_output(dimension.level2.fun()), "Dimension level 2 functions implemented in dispRity:\n?ancestral.dist\n?angles\n?centroids\n?deviations\n?displacements\n?neighbours\n?pairwise.dist\n?ranges\n?radius\n?variances\n?span.tree.length")
-    expect_equal(capture_output(dimension.level1.fun()), "Dimension level 1 functions implemented in dispRity:\n?convhull.surface\n?convhull.volume\n?diagonal\n?ellipse.volume\n?func.div\n?func.eve\n?mode.val\n?n.ball.volume")
+    expect_equal(capture_output(dimension.level1.fun()), "Dimension level 1 functions implemented in dispRity:\n?convhull.surface\n?convhull.volume\n?diagonal\n?ellipse.volume\n?func.div\n?func.eve\n?group.dist\n?mode.val\n?n.ball.volume")
 })
 
 

### 2020-09-07 inst/gitbook/03_specific-tutorials.Rmd
Added between groups tutorial

@@
 
-### `make.metric`
+### Between groups metrics {#betweengroupmetricsexplain}
+
+One specific category of metrics in the `dispRity` package is the between groups metrics.
+As the name suggest, these metrics can be used to calculate the disparity between groups rather than within the groups.
+These metrics follow the same classifications as the "normal" (within group) metrics with dimension-level 1, 2 and 3 between groups metrics.
+However, at the difference of the "normal" metrics, their input arguments must be `matrix` and `matrix2` (and of course any other additional arguments).
+For example, this metric measures the difference in mean between two matrices:
+
+```{r}
+## A simple example
+mean.difference <- function(matrix, matrix2) {
+  mean(matrix) - mean(matrix2)
+}
+```
+
+You can find the list of implemented between groups metric [here](#betweengroupmetricslist) or design them yourself for your specific needs (potentially using [`make.metric`](#makemetric) for help).
+
+The function works by simply using the two available matrices, with no restriction in terms of dimensions (although you'd probably want both matrices to have the same number of dimensions)
+
+```{r}
+## A second matrix
+dummy_matrix2 <- matrix(runif(12), 4, 3)
+
+## The difference between groups
+mean.difference(dummy_matrix, dummy_matrix2)
+```
+
+Beyond this super simple example, it might probably be interesting to use this metric on `dispRity` objects, especially the ones from [`custom.subsets`](#custom-subsets) and [`chrono.subsets`](#chrono-subsets).
+In fact, the `dispRity` function allows to apply the between groups metric directly to the `dispRity` objects using the `between.groups = TRUE` option.
+For example:
+
+```{r}
+## Combining both matrices
+big_matrix <- rbind(dummy_matrix, dummy_matrix2)
+rownames(big_matrix) <- 1:8
+
+## Making a dispRity object with both groups
+grouped_matrix <- custom.subsets(big_matrix,
+                                 group = c(list(1:4), list(1:4)))
+
+## Calculating the mean difference between groups
+(mean_differences <- dispRity(grouped_matrix,
+                              metric = mean.difference,
+                              between.groups = TRUE))
+
+## Summarising the object
+summary(mean_differences)
+## Note how the summary table now indicates
+## the number of elements for each group
+```
+
+For `dispRity` objects generated by `custom.subsets`, the `dispRity` function will by default apply the metric on the groups in a pairwise fashion.
+For example, if the object contains multiple groups, all groups will be compared to each other:
+
+```{r}
+## A dispRity object with multiple groups
+grouped_matrix <- custom.subsets(big_matrix,
+                                 group = c("A" = list(1:4),
+                                           "B" = list(1:4),
+                                           "C" = list(2:6), 
+                                           "D" = list(1:8)))
+
+## Measuring disparity between all groups
+summary(dispRity(grouped_matrix, metric = mean.difference,
+                 between.groups = TRUE))
+```
+
+For `dispRity` objects generated by `chrono.subsets` (not shown here), the `dispRity` function will by default apply the metric on the groups in a serial way (group 1 vs.
+group 2, group 2 vs.
+group 3, group 3 vs.
+group 4, etc...).
+However, in both cases (for objects from `custom.subsets` or `chrono.subsets`) it is possible to manually specific the list of pairs of comparisons through their ID numbers:
+
+```{r}
+## Measuring disparity between specific groups
+summary(dispRity(grouped_matrix, metric = mean.difference,
+                 between.groups = list(c(1,3), c(3,1), c(4,1))))
+
+```
+
+Note that in any case, the order of the comparison can matter.
+In our example, it is obvious that `mean(matrix) - mean(matrix2)` is not the same as `mean(matrix2) - mean(matrix)`.
+
+### `make.metric` {#makemetric}
 
@@
 
-#### Between group metrics
-
+#### Between group metrics {#betweengroupmetricslist}
+
+You can find detailed explanation on how between group metrics work [here](#betweengroupmetricsexplain).
 The `group.dist` metric allows to measure the distance between two groups in the multidimensional space.

### 2020-09-11 inst/gitbook/03_specific-tutorials.Rmd
Added manual and test!
Nearly there!

@@
 2 | `pairwise.dist` | The pairwise distances between elements | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::vegist` |
+2 | `point.dist` | The distance between one group and the point of another group | `dispRity` |
 2 | `quantiles` | The *n*th quantile range per axis | `dispRity` |
@@
 You can find detailed explanation on how between group metrics work [here](#betweengroupmetricsexplain).
+
+##### `group.dist`
+
 The `group.dist` metric allows to measure the distance between two groups in the multidimensional space.
@@
 
+##### `point.dist`
+
+The metric measures the distance between the elements in one group (`matrix`) and a point calculated from a second group (`matrix2`).
+By default this point is the centroid but can be any point defined by a function passed to the `point` argument.
+For example, the centroid of `matrix2` is the mean of each column of that matrix so `point = colMeans` (default).
+This function also takes the `method` argument like previous one described above to measure either the `"euclidean"` (default) or the `"manhattan"` distances:
+
+```{r}
+## Measuring the distance between the elements of the first group
+## and the centroid of the second group
+summary(dispRity(grouped_space, metric = point.dist,
+                 between.groups = TRUE))
+
+## Measuring the distance between the elements of the second group
+## and the centroid of the first group
+summary(dispRity(grouped_space, metric = point.dist,
+                 between.groups = list(c(2,1)))))
+
+## Measuring the distance between the elements of the first group
+## a point defined as the standard deviation of each column
+## in the second group
+sd.point <- function(matrix2) {apply(matrix2, 2, sd)}
+summary(dispRity(grouped_space, metric = point.dist,
+                 point = sd.point, method = "manhattan",
+                 between.groups = TRUE))
+```
+
 ### Which disparity metric to choose?
@@
 ```
+
 Joke aside, this is a legitimate question that has no simple answer: **it depends on the dataset and question at hand**.

### 2020-09-18 inst/gitbook/03_specific-tutorials.Rmd
Update 03_specific-tutorials.Rmd

@@
 
-The remaining output gives us the log-likelihood of the Stasis model of -59.501 (you may notice this change when we pooled variances above).
+The remaining output gives us the log-likelihood of the Stasis model of `r summary(disp_time)[,"log.lik"]` (you may notice this change when we pooled variances above).
 The output also gives us the small sample Akaike Information Criterion (AICc), the delta AICc (the distance from the best fitting model), and the AICc weights (~the relative support of this model compared to all models, scaled to one).
@@
 So we again see the AICc, delta AICc, AICc weight, and the log-likelihood we saw previously.
-We now also see the number of parameters from the model (2: theta and omega), and their estimates so the variance (omega = 0.01) and the mean (theta.1 = 3.4).
+We now also see the number of parameters from the model (2: theta and omega), and their estimates so the variance (omega = `r summary(disp_time)[,"omega"]`) and the mean (theta.1 = `r summary(disp_time)[,"theta.1"]`).
 
@@
 Et voilà! Here we can see by the log-likelihood, AICc, delta AICc, and AICc weight Brownian motion has a much better relative fit to these data than the Stasis model.
-Brownian motion has a relative AICc fit 366 units better than Stasis, and virtually has a AICc weight of 1.
+Brownian motion has a relative AICc fit`r summary(disp_time)[,"delta_aicc"][[1]]` units better than Stasis, and has a AICc weight of 1.
 
@@
 
-This show a positive trend (0.01) of increasing disparity through time.
+This show a positive trend (`r unname(summary(disp_time)["Trend",]["trend"])`) of increasing disparity through time.
 
@@
 Trend is the best-fitting model but the plot suggests the OU model also follows a trend-like pattern.
-This is because the optima for the OU model (5.7) is different to the ancestral state 2.835 and outside the observed value.
+This is because the optima for the OU model (`r disp_time["OU", "optima.1"]`) is different to the ancestral state (`r disp_time["OU", "ancestral state"]`) and outside the observed value.
 This is potentially unrealistic, and one way to alleviate this issue is to set the optima of the OU model to equal the ancestral estimate - this is the normal practice for OU models in comparative phylogenetics.
@@
 As seen before, the Trend model fitted this dataset the best.
-To simulate what 1000 Trend models would look like using the same parameters as the ones estimated with `model.test` (here the ancestral state being 2.839, the sigma squared beeing 0.002 and the trend of 0.01), we can simply pass this model to `model.test.sim`:
+To simulate what 1000 Trend models would look like using the same parameters as the ones estimated with `model.test` (here the ancestral state being `r summary(disp_time)["Trend", "ancestral state"]`, the sigma squared being `r summary(disp_time)["Trend", "sigma squared"]` and the trend of `r summary(disp_time)["Trend", "trend"]`), we can simply pass this model to `model.test.sim`:
 

### 2020-09-25 inst/gitbook/03_specific-tutorials.Rmd

@@
 
+> DISCLAIMER: this method of analysing disparity has not been published yet and has not been peer reviewed.
+Caution should be used in interpreting these results: it is unclear what "a disparity curve fitting a Brownian motion" actually means biologically.
+
+As Malcolm said in Jurassic Park: "although the examples within this chapter all work and produce solid tested results (from an algorithm point of view), that doesn't mean you should use it" (or something along those lines).
+
 ### Simple modes of disparity change through time

### 2020-09-25 inst/gitbook/03_specific-tutorials.Rmd
Up gitbook

@@
 summary(dispRity(grouped_space, metric = point.dist,
-                 between.groups = list(c(2,1)))))
+                 between.groups = list(c(2,1))))
 

### 2020-09-25 inst/gitbook/03_specific-tutorials.Rmd
Updated book

@@
 ## Plotting with more arguments
-plot(time_subsets, dimensions = c(4,5),
+plot(time_subsets, specific.args = list(dimensions = c(1,2)),
      main = "Some \"low\" dimensions")
@@
 ## Plotting the results
-plot(dispRity_dtt, fig.width=8, fig.height=8)
+plot(dispRity_dtt)
 ```

### 2020-09-25 inst/gitbook/09_gmm-demo.Rmd
Updated book

@@
 
-# Palaeobiology demo: disparity-through-time and within groups
+# Morphometric geometric demo: a between group analysis
 
@@
 
-```{r}
+```{r, fig.width = 8, fig.height = 16}
 ## Defining a the procrustes variance metric

### 2020-11-02 inst/gitbook/08_palaeo-demo.Rmd
bug fix in paleo demo

@@
 ## Creating the vector of time bins ages
-(time_bins <- rev(seq(from = 0, to = 100, by = 20)))
+time_bins <- rev(seq(from = 0, to = 100, by = 20))
 
 ## Splitting the morphospace using the chrono.subsets function
-(binned_morphospace <- chrono.subsets(data = my_matrix, tree = my_tree,
+binned_morphospace <- chrono.subsets(data = my_matrix, tree = my_tree,
     method = "discrete", time = time_bins, inc.nodes = FALSE,
-    FADLAD = my_fadlad))
+    FADLAD = my_fadlad)
 ```
@@
 ```{r}
-## Bootstrapping each time subset 100 times (default)
-(boot_bin_morphospace <- boot.matrix(binned_morphospace))
-
 ## Getting the minimum number of rows (i.e. taxa) in the time subsets
-min(size.subsets(boot_bin_morphospace))
+minimum_size <- min(size.subsets(binned_morphospace))
 
 ## Bootstrapping each time subset 100 times and rarefying them 
-(rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
-    rarefaction = 6))
+rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
+    rarefaction = minimum_size)
 ```
@@
 
-```{r}
+```{r, fig.height = 16, fig.width = 8}
 my_test <- test.metric(my_matrix, metric = c(sum, variances), shifts = c("random", "size"))
@@
 ```{r}
-## Calculating disparity for the bootstrapped data
-(boot_disparity <- dispRity(boot_bin_morphospace, metric = c(sum, variances)))
-
-## Calculating disparity for the rarefied data
-(rare_disparity <- dispRity(rare_bin_morphospace, metric = c(sum, variances)))
+## Calculating disparity for the bootstrapped and rarefied data
+disparity <- dispRity(rare_bin_morphospace , metric = c(sum, variances))
 ```
@@
 ## Summarising the disparity results
-summary(boot_disparity)
-summary(rare_disparity)
+summary(disparity)
 ```
@@
 ## Plotting the bootstrapped and rarefied results
-plot(boot_disparity, type = "continuous", main = "bootstrapped results")
-plot(rare_disparity, type = "continuous", main = "rarefied results")
+plot(disparity, type = "continuous", main = "bootstrapped results")
+plot(disparity, type = "continuous", main = "rarefied results",
+     rarefaciton = minimum_size)
 ```
@@
 ## Testing the differences between bins in the bootstrapped dataset.
-test.dispRity(boot_disparity, test = wilcox.test, comparison = "sequential",
+test.dispRity(disparity, test = wilcox.test, comparison = "sequential",
     correction = "bonferroni")
@@
 ## Testing the differences between bins in the rarefied dataset.
-test.dispRity(rare_disparity, test = wilcox.test, comparison = "sequential",
-    correction = "bonferroni")
+test.dispRity(disparity, test = wilcox.test, comparison = "sequential",
+    correction = "bonferroni", rarefaction  = minimum_size)
 ```

### 2020-11-02 inst/gitbook/dispRity_manual.Rmd
bug fix in paleo demo

@@
 version_master <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
-version_master <- 1.4
+version_master <- 1.5
 
@@
 version_release <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
-version_release <- 1.4
+version_release <- 1.5
 
@@
 version_CRAN <- strsplit(strsplit(version_raw, split = "\n<td>Version:</td>\n<td>")[[1]][2], split = "</td>\n</tr>\n<tr>\n<td>Depends:")[[1]][1]
-version_CRAN <- 1.4
+version_CRAN <- 1.5
 ```
@@
 [doi:10.1111/pala.12364](https://onlinelibrary.wiley.com/doi/abs/10.1111/pala.12364).
+
+Furthermore, don't forget to cite `R`:
+
+> R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
+
+Bonus: you can also cite `ape` since the `dispRity` package heavily relies on it:
+
+> Paradis E. & Schliep K. 2019. ape 5.0: an environment for modern phylogenetics and evolutionary analyses in R. Bioinformatics 35: 526-528.
 
@@
 
+> Note that using the `rarefaction` argument also bootstraps the data.
+In these examples, the function bootstraps the data (without rarefaction) AND also bootstraps the data with the different rarefaction levels.
+
 One other argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
@@
 
+#### Between group metrics {#betweengroupmetricslist}
+
+You can find detailed explanation on how between group metrics work [here](#betweengroupmetricsexplain).
+
+##### `group.dist`
+
+The `group.dist` metric allows to measure the distance between two groups in the multidimensional space.
+This function needs to intake several groups and use the option `between.groups = TRUE` in the `dispRity` function.
+It calculates the vector normal distance (euclidean) between two groups and returns 0 if that distance is negative.
+Note that it is possible to set up which quantiles to consider for calculating the distances between groups.
+For example, one might be interested in only considering the 95% CI for each group.
+This can be done through the option `probs = c(0.025, 0.975)` that is passed to the `quantile` function.
+It is also possible to use this function to measure the distance between the groups centroids by calculating the 50% quantile (`probs = c(0.5)`).
+
+```{r}
+## Creating a dispRity object with two groups
+grouped_space <- custom.subsets(dummy_space,
+                      group = list(c(1:5), c(6:10)))
+
+## Measuring the minimum distance between both groups
+summary(dispRity(grouped_space, metric = group.dist,
+                 between.groups = TRUE))
+
+## Measuring the centroid distance between both groups
+summary(dispRity(grouped_space, metric = group.dist,
+                 between.groups = TRUE, probs = 0.5))
+
+## Measuring the distance between both group's 75% CI
+summary(dispRity(grouped_space, metric = group.dist,
+                 between.groups = TRUE, probs = c(0.25, 0.75)))
+```
+
+##### `point.dist`
+
+The metric measures the distance between the elements in one group (`matrix`) and a point calculated from a second group (`matrix2`).
+By default this point is the centroid but can be any point defined by a function passed to the `point` argument.
+For example, the centroid of `matrix2` is the mean of each column of that matrix so `point = colMeans` (default).
+This function also takes the `method` argument like previous one described above to measure either the `"euclidean"` (default) or the `"manhattan"` distances:
+
+```{r}
+## Measuring the distance between the elements of the first group
+## and the centroid of the second group
+summary(dispRity(grouped_space, metric = point.dist,
+                 between.groups = TRUE))
+
+## Measuring the distance between the elements of the second group
+## and the centroid of the first group
+summary(dispRity(grouped_space, metric = point.dist,
+                 between.groups = list(c(2,1))))
+
+## Measuring the distance between the elements of the first group
+## a point defined as the standard deviation of each column
+## in the second group
+sd.point <- function(matrix2) {apply(matrix2, 2, sd)}
+summary(dispRity(grouped_space, metric = point.dist,
+                 point = sd.point, method = "manhattan",
+                 between.groups = TRUE))
+```
+
 ### Which disparity metric to choose?
@@
 So we acknowledge and thank Gene Hunt for developing these models and writing the original R code that served as inspiration for these models.
+
+> DISCLAIMER: this method of analysing disparity has not been published yet and has not been peer reviewed.
+Caution should be used in interpreting these results: it is unclear what "a disparity curve fitting a Brownian motion" actually means biologically.
+
+As Malcolm said in Jurassic Park: "although the examples within this chapter all work and produce solid tested results (from an algorithm point of view), that doesn't mean you should use it" (or something along those lines).
 

### 2020-11-10 inst/gitbook/08_palaeo-demo.Rmd
fixed typo in the paleo demo

@@
 plot(disparity, type = "continuous", main = "rarefied results",
-     rarefaciton = minimum_size)
+     rarefaction = minimum_size)
 ```

### 2020-12-03 inst/gitbook/03_specific-tutorials.Rmd
bug fix in paleo demo

@@
 ```
+
+> Note that using the `rarefaction` argument also bootstraps the data.
+In these examples, the function bootstraps the data (without rarefaction) AND also bootstraps the data with the different rarefaction levels.
 
@@
     col = c("black", rainbow(10)),
-    ylab = c("Disparity", "Diversity"), chrono.subsets = FALSE,
+    ylab = c("Disparity", "Diversity"),
     xlab = "Time (in in units from past to present)",
@@
 ```{r}
-## Ordinating the eurodist matrix
-euro_ord <- cmdscale(eurodist, k = nrow(eurodist) - 2)
+## Ordinating the eurodist matrix (with 11 dimensions)
+euro_ord <- cmdscale(eurodist, k = 11)
 

### 2020-12-03 inst/gitbook/index.Rmd

@@
 
+Furthermore, don't forget to cite `R`:
+
+> R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
+
+Bonus: you can also cite `ape` since the `dispRity` package heavily relies on it:
+
+> Paradis E. & Schliep K. 2019. ape 5.0: an environment for modern phylogenetics and evolutionary analyses in R. Bioinformatics 35: 526-528.
+
 ### Why is it important to cite us?

### 2021-01-27 inst/gitbook/06_guts.Rmd
added n.subsets function

--- 2020-05-01 inst/gitbook/06_guts.Rmd
@@
 
+#### `n.subsets`
+
+This function simply counts the number of subsets in a `dispRity` object.
+
+```{r}
+## How many subsets are in this object?
+n.subsets(disparity)
+```
+
+#### `size.subsets`
+
+This function tells the number of elements in each subsets of a `dispRity` object.
+
+```{r}
+## How many elements are there in each subset?
+size.subsets(disparity)
+```
+
 #### `get.subsets`
@@
 
-## The `dispRity` object content
+## The `dispRity` object content {#disprity-object}
 

### 2021-02-11 inst/gitbook/03_specific-tutorials.Rmd
Updated manual

@@
 2 | `point.dist` | The distance between one group and the point of another group | `dispRity` |
+2 | `projections` | The distance *on* (projection) or *from* (rejection) an arbitrary vector | `dispRity` |
 2 | `quantiles` | The *n*th quantile range per axis | `dispRity` |
@@
     n.ball.volume = \frac{\pi^{d/2}}{\Gamma(\frac{d}{2}+1)}\displaystyle\prod_{i=1}^{d} R
+\end{equation}
+
+\begin{equation}
+    projection_{on} = \| \overrightarrow{i} \cdot \overrightarrow{b} \|
+\end{equation}
+\begin{equation}
+    projection_{from} = \| \overrightarrow{i} - \overrightarrow{i} \cdot \overrightarrow{b} \|
 \end{equation}
@@
 *R* is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid),
-and $Reference_{k}$ is an arbitrary point's coordinates (usually 0).
+$Reference_{k}$ is an arbitrary point's coordinates (usually 0),
+\overrightarrow{b} is the vector defined by (`(point1, point2)`),
+and \overrightarrow{i} is the vector defined by (`(point1, i)` where `i` is any row of the matrix).
 
@@
 In doubt, it's probably better to enter your `base` (for `angles`) or your `hyperplane` (for `deviations`) manually so you're sure you know what the function is measuring.
+
+#### Projections: elaboration and exploration
+
+The `projections` metric calculates the geometric projection and corresponding rejection of all the rows in a matrix on an arbitrary vector (respectively the distance *on* and the distance *from* that vector).
+The function is based on @aguilera2004's n-dimensional rotation algorithm to use linear algebra in mutidimensional spaces.
+The projection or rejection can be seen as respectively the elaboration and exploration scores on a trajectory (*sensu* @endler2005).
+
+By default, the vector (e.g. a trajectory, an axis), on which the data is projected is the one going from the centre of the space (coordinates 0,0, ...) and the centroid of the matrix.
+However, we advice you do define this axis to something more meaningful using the `point1` and `point2` options, to create the vector (the vector's norm will be `dist(point1, point2)` and its direction will be from `point1` towards `point2`).
+
+```{r}
+## The elaboration on the axis defined by the first and
+## second row in the dummy_space
+summary(dispRity(dummy_space, metric = projections,
+                              point1 = dummy_space[1,],
+                              point2 = dummy_space[2,]))
+
+## The exploration on the same axis
+summary(dispRity(dummy_space, metric = projections,
+                              point1 = dummy_space[1,],
+                              point2 = dummy_space[2,],
+                              measure = "distance"))
+```
+
+By default, the vector `(point1, point2)` is used as unit vector of the projections (i.e.
+the Euclidean distance between `(point1, point2)` is set to 1) meaning that a projection value (`"distance"` or `"position"`) of *X* means *X* times the distance between `point1` and `point2`.
+If you want use the unit vector of the input matrix or are using a space where Euclidean distances are non-sensical, you can remove this option using `scale = FALSE`:
+
+```{r}
+## The elaboration on the same axis using the dummy_space's
+## unit vector
+summary(dispRity(dummy_space, metric = projections,
+                              point1 = dummy_space[1,],
+                              point2 = dummy_space[2,],
+                              scale = FALSE))
+```
 

### 2021-03-02 inst/gitbook/03_specific-tutorials.Rmd
Added phylo.projections (not checked)
Not checked!

@@
 2 | `point.dist` | The distance between one group and the point of another group | `dispRity` |
+2 | `phylo.projections` | The `projections` metric but where the vector can be based on a tree | `dispRity` |
 2 | `projections` | The distance *on* (projection) or *from* (rejection) an arbitrary vector | `dispRity` |
@@
 
-#### Projections: elaboration and exploration
+#### Projections and phylo.projections: elaboration and exploration
 
@@
                               scale = FALSE))
+```
+
+The `phylo.projections` is the same as the `projections` metric but allows to determine the vector (`(point1, point2)`) using a tree rather than manually entering these points.
+The function intakes the exact same options as the `projections` function described above at the exception of `point1` and `point2`.
+Instead it takes a the argument `type` that designates the type of vector to draw from the data based on a phylogenetic tree `phy`.
+The argument `type` can be a pair of any of the following inputs:
+
+ * `"root"`: to automatically use the coordinates of the root of the tree (the first element in `phy$node.label`);
+ * `"ancestor"`: to automatically use the coordinates of the elements' (i.e. any row in the matrix) most recent ancestor;
+ * `"tips"`: to automatically use the coordinates from the centroid of all tips;
+ * `"nodes"`: to automatically use the coordinates from the centroid of all nodes;
+ * `"livings"`: to automatically use the coordinates from the centroid of all "living" tips (i.e. the tips that are the furthest away from the root);
+ * `"fossils"`: to automatically use the coordinates from the centroid of all "fossil" tips and nodes (i.e. not the "living" ones);
+ *  any numeric values that can be interpreted as `point1` and `point2` in `projections` (e.g. `0`, `c(0, 1.2, 3/4)`, etc.);
+ *  or a user defined function that with the inputs `matrix` and `phy` and `row` (the element's ID, i.e. the row number in `matrix`).
+
+For example, if you want to measure the projection of each element in the matrix (tips and nodes) on the axis from the root of the tree to each element's most recent ancestor, you can define the vector as `type = c("root", "ancestor")`.
+
+```{r}
+## Adding a extra row to dummy matrix (to match dummy_tree)
+tree_space <- rbind(dummy_space, root = rnorm(5))
+## Creating a random dummy tree (with labels matching the ones from tree_space)
+dummy_tree <- rtree(6)
+dummy_tree$tip.label <- rownames(tree_space)[1:6]
+dummy_tree$node.label <- rownames(tree_space)[rev(7:11)]
+
+## Measuring the disparity as the projection of each element
+## on its root-ancestor vector
+summary(dispRity(tree_space, metric = phylo.projections,
+                             phy    = dummy_tree,
+                             type   = c("root", "ancestor")))
+```
+
+Of course you can also use any other options from the projections function:
+
+```{r}
+## A user defined function that's returns the centroid of
+## the first three nodes
+fun.root <- function(matrix, phy, row = NULL) {
+  return(colMeans(matrix[phy$node.label[1:3], ]))
+}
+## Measuring the unscaled rejection from the vector from the
+## centroid of the three first nodes
+## to the coordinates of the first tip
+summary(dispRity(tree_space, metric  = phylo.projections,
+                             phy     = dummy_tree,
+                             measure = "distance",
+                             type    = list(fun.root,
+                                            tree_space[1, ])))
 ```

### 2021-03-26 inst/gitbook/03_specific-tutorials.Rmd
Update 03_specific-tutorials.Rmd

@@
 2 | `displacements` | The ratio between the distance from a reference and the distance from the centroid | `dispRity` |
+1 | `edge.length.tree` | The edge lengths of the elements on a tree | `ape` |
 1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
@@
 2 | `point.dist` | The distance between one group and the point of another group | `dispRity` |
-2 | `phylo.projections` | The `projections` metric but where the vector can be based on a tree | `dispRity` |
+2 | `projections.tree` | The `projections` metric but where the vector can be based on a tree | `dispRity` |
 2 | `projections` | The distance *on* (projection) or *from* (rejection) an arbitrary vector | `dispRity` |
@@
 
-#### Projections and phylo.projections: elaboration and exploration
+#### Projections and phylo projections: elaboration and exploration
 
@@
 
-The `phylo.projections` is the same as the `projections` metric but allows to determine the vector (`(point1, point2)`) using a tree rather than manually entering these points.
+The `projections.tree` is the same as the `projections` metric but allows to determine the vector (`(point1, point2)`) using a tree rather than manually entering these points.
 The function intakes the exact same options as the `projections` function described above at the exception of `point1` and `point2`.
@@
 ## on its root-ancestor vector
-summary(dispRity(tree_space, metric = phylo.projections,
+summary(dispRity(tree_space, metric = projections.tree,
                              phy    = dummy_tree,
@@
 ## to the coordinates of the first tip
-summary(dispRity(tree_space, metric  = phylo.projections,
+summary(dispRity(tree_space, metric  = projections.tree,
                              phy     = dummy_tree,
@@
 As we've seen before, this is incorrect in this case since the trees don't have the same topology (so the nodes selected by a slice through the second tree are not equivalent to the nodes in the first matrix) but it can be useful if the topology is fixed to integrate both uncertainty in branch length (slicing through different trees) and uncertainty from, say, ancestral states estimations (applying the slices on different matrices).
+
+## Disparity with trees: *dispRitree!*
+
+Since the package's version 1.5.10, trees can be directly att
+
+If phylogeny is going to be an important part of your disparity analyses, you can attach a tree or a list of trees (classes `"phylo"` or `"multiPhylo"`) to your `dispRity` object.
+These trees will then be fixed and used where possible throughout your analyses (e.g. when using `chrono.subsets` or disparity metrics based using the argument `tree`).
+You can attach trees in different ways, either voluntarily (using @@@) or throughout your analyses through specific functions (@@@).
+The tree labels must match the rownames in the data. If the tree has only tip labels, then only these need to be matched, if the tree or the data also has node labels, they must match as well.

### 2021-03-26 inst/gitbook/03_specific-tutorials.Rmd
added things to tutorial

@@
 
-Since the package's version 1.5.10, trees can be directly att
+Since the package's version 1.5.10, trees can be directly attached to `dispRity` objects.
+This allows any function in the package that has an input argument called "`tree`" to automatically intake the tree from the `dispRity` object.
+This is especially useful for disparity metrics that requires calculations based on a phylogenetic tree (e.g.
+`ancestral.dist` or `projections.tree`) and if phylogeny (or phylogenie\*s\*) are going to be an important part of your analyses.
+
+Trees are attached to `dispRity` object as soon as they are called in any function of the package (e.g.
+as an argument in `chrono.subsets` or in `dispRity`) and are stored in `my_dispRity_object$tree`.
+You can always manually attach, detach or modify the tree parts of a `dispRity` object using the utility functions `get.tree` (to access the trees), `remove.tree` (to remove it) and `add.tree` (to...
+add trees!).
+The only requirement for this to work is that the labels in the tree must match the ones in the data.
+If the tree has node labels, their node labels must also match the data.
+Similarly if the data has entries for node labels, they must be present in the tree.
+
+Here is a quick demo on how attaching trees to `dispRity` objects can work and make your life easy: for example here we will measure how the sum of branch length changes through time when time slicing through some demo data with a acctran split time slice model (see more info [here](#chrono-subsets)).
+
+```{r}
+## Loading some demo data:
+## An ordinated matrix with node and tip labels
+data(BeckLee_mat99)
+## The corresponding tree with tip and node labels
+data(BeckLee_tree)
+## A list of tips ages for the fossil data
+data(BeckLee_ages)
+
+## Time slicing through the tree using the equal split algorithm
+time_slices <- chrono.subsets(data   = BeckLee_mat99,
+                              tree   = BeckLee_tree,
+                              FADLAD = BeckLee_ages,
+                              method = "continuous",
+                              model  = "acctran",
+                              time   = 15)
+
+## We can visualise the resulting trait space with the phylogeny
+## (using the specific argument as follows)
+plot(time_slices, type = "preview",
+     specific.args = list(tree = TRUE))
+## Note that some nodes are never selected thus explaining the branches not reaching them.
+```
+
+And we can then measure disparity as the sum of the edge length at each time slice on the bootstrapped data:
+
+```{r}
+## Measuring the sum of the edge length per slice
+sum_edge_length <- dispRity(boot.matrix(time_slices), metric = c(sum, edge.length.tree))
+## Summarising and plotting
+summary(sum_edge_length)
+plot(sum_edge_length)
+```
+
+Of course this can be done with multiple trees and be combined with an approach using multiple matrices (see [here](#multi.input))!
 

### 2021-03-30 inst/gitbook/03_specific-tutorials.Rmd
updated ancestral.dist

@@
 
-#### Centroids, displacements, and ancestral distance metrics {#centroids}
+#### Centroids, displacements and ancestral distances metrics {#centroids}
 
@@
 
-The `ancestral.dist` metric works on a similar principle as the `centroids` function but changes the centroid to be the coordinates of each element's ancestor.
-Therefore this functions needs a tree and node coordinates as additional arguments:
+The `ancestral.dist` metric works on a similar principle as the `centroids` function but changes the centroid to be the coordinates of each element's ancestor (if `to.root = FALSE`; default) or to the root of the tree (`to.root = TRUE`).
+Therefore this function needs a matrix that contains tips and nodes and a tree as additional argument.
 
@@
 ## A generating a random tree with node labels
-tree <- rtree(5) ; tree$node.label <- paste0("n", 1:4)
+my_tree <- makeNodeLabel(rtree(5), prefix = "n")
 ## Adding the tip and node names to the matrix
@@
 
-## Calculating all the ancestral nodes
-all_anc_centroids <- nodes.coordinates(dummy_space2, tree,
-                                       full = TRUE)
-
 ## Calculating the distances from the ancestral nodes
 ancestral_dist <- dispRity(dummy_space2, metric = ancestral.dist,
-                           nodes.coords = all_anc_centroids)
+                           tree = my_tree)
 

### 2021-03-31 inst/gitbook/06_guts.Rmd
added book entries

@@
 
+#### `get.tree` `add.tree` and `remove.tree`
+
+These functions allow to manipulate the potential tree components of `dispRity` objects.
+
+```{r, eval = FALSE}
+## Getting the tree component of a dispRity object
+get.tree(disparity)
+
+## Removing the tree
+remove.tree(disparity)
+
+## Adding a tree
+add.tree(disparity, tree = BeckLee_tree)
+```
+
 ## The `dispRity` object content {#disprity-object}

### 2021-04-01 inst/gitbook/03_specific-tutorials.Rmd

@@
 Of course this can be done with multiple trees and be combined with an approach using multiple matrices (see [here](#multi.input))!
-
-If phylogeny is going to be an important part of your disparity analyses, you can attach a tree or a list of trees (classes `"phylo"` or `"multiPhylo"`) to your `dispRity` object.
-These trees will then be fixed and used where possible throughout your analyses (e.g. when using `chrono.subsets` or disparity metrics based using the argument `tree`).
-You can attach trees in different ways, either voluntarily (using @@@) or throughout your analyses through specific functions (@@@).
-The tree labels must match the rownames in the data. If the tree has only tip labels, then only these need to be matched, if the tree or the data also has node labels, they must match as well.

### 2021-04-13 inst/gitbook/index.Rmd
added waranty disclaimer

@@
 
+## `dispRity` is always changing, how do I know it's not broken?
+
+This is a really common a legitimate question in software development. Like `R` itself:
+
+> `dispRity` is free software and comes with ABSOLUTELY NO WARRANTY.
+
+So you are using it at your own risk.
+
+**HOWEVER**, there are two points that can be used as objective-ish markers on why it's OK to use `dispRity`.
+
+First, the package has been use in a number of [peer reviewed](https://scholar.google.co.uk/scholar?oi=bibs&hl=en&cites=13311379491028410826,7753828186872068057) publications (the majority of them independently) which could be taken as warranty.
+
+Second, I spend a lot of time and attention in making sure that every function in every version actually does what I think it is supposed to do.
+This is done through [CI; continuous integration development](https://en.wikipedia.org/wiki/Continuous_integration), the [CRAN check](https://cran.r-project.org/web/checks/check_results_dispRity.html), and [unit testing](https://en.wikipedia.org/wiki/Unit_testing).
+The two first checks (CRAN and CI) ensure that the version you are using is not bugged (the [CRAN check](https://cran.r-project.org/web/checks/check_results_dispRity.html) if you are using the CRAN version and the [Travis CI](https://travis-ci.org/TGuillerme/dispRity) if you are using a GitHub version).
+The third check, unit testing, is checking that every function is doing what it is supposed to do.
+For a real basic example, it is testing that the following expression should always return the same thing no matter what changes in the package.
+
+```
+> mean(c(1,2,3))
+[1] 2
+```
+Or, more formally:
+
+```{r}
+testthat::expect_equal(object = mean(c(1,2,3)),
+                       expected = 2)
+```
+
+You can always access what is actually tested in the [`test/testthat`](https://github.com/TGuillerme/dispRity/tree/master/tests/testthat) sub-folder.
+For example [here is how the core function `dispRity`](https://github.com/TGuillerme/dispRity/blob/master/tests/testthat/test-dispRity.R) is tested (through > 500 tests!).
+All these tests are run every time a change is made to the package and you can always see for yourself how much a single function is covered (i.e.
+what percentage of the function is actually covered by at least one test).
+You can always see the global coverage [here](https://app.codecov.io/gh/TGuillerme/dispRity) or the specific coverage for each function [here](https://codecov.io/gh/TGuillerme/dispRity/tree/master/R).
+
+Finally, this package is build on the shoulders of the whole open science philosophy so when bugs do occur and are caught by myself or the package users, they are quickly fixed and notified in the [`NEWS.md` file](https://github.com/TGuillerme/dispRity/blob/master/NEWS.md).
+
 ## Help

### 2021-04-16 inst/gitbook/03_specific-tutorials.Rmd

@@
 $Reference_{k}$ is an arbitrary point's coordinates (usually 0),
-\overrightarrow{b} is the vector defined by (`(point1, point2)`),
-and \overrightarrow{i} is the vector defined by (`(point1, i)` where `i` is any row of the matrix).
+$\overrightarrow{b}$ is the vector defined by (`(point1, point2)`),
+and $\overrightarrow{i}$ is the vector defined by (`(point1, i)` where `i` is any row of the matrix).
 
@@
 dummy_space2 <- dummy_space[-1,]
-rownames(dummy_space2) <- c(tree$tip.label, tree$node.label)
+rownames(dummy_space2) <- c(my_tree$tip.label,
+                            my_tree$node.label)
 
@@
 summary(dispRity(tree_space, metric = projections.tree,
-                             phy    = dummy_tree,
+                             tree   = dummy_tree,
                              type   = c("root", "ancestor")))
@@
 ## the first three nodes
-fun.root <- function(matrix, phy, row = NULL) {
-  return(colMeans(matrix[phy$node.label[1:3], ]))
+fun.root <- function(matrix, tree, row = NULL) {
+  return(colMeans(matrix[tree$node.label[1:3], ]))
 }
@@
 summary(dispRity(tree_space, metric  = projections.tree,
-                             phy     = dummy_tree,
+                             tree    = dummy_tree,
                              measure = "distance",

### 2021-04-16 inst/gitbook/index.Rmd
Update index.Rmd

@@
 version_master <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
-version_master <- 1.5
+version_master <- 1.6
 
@@
 version_release <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
-version_release <- 1.5
+version_release <- 1.6
 
@@
 version_CRAN <- strsplit(strsplit(version_raw, split = "\n<td>Version:</td>\n<td>")[[1]][2], split = "</td>\n</tr>\n<tr>\n<td>Depends:")[[1]][1]
-version_CRAN <- 1.5
+version_CRAN <- 1.6
 ```
@@
 Finally, this package is build on the shoulders of the whole open science philosophy so when bugs do occur and are caught by myself or the package users, they are quickly fixed and notified in the [`NEWS.md` file](https://github.com/TGuillerme/dispRity/blob/master/NEWS.md).
+And all the changes to the package are [public and annotated](https://github.com/TGuillerme/dispRity/commits/master) so there's that too...
 

### 2021-06-03 inst/gitbook/03_specific-tutorials.Rmd

@@
 Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.
-Check out this pre-print on selecting the best metric for your specific question on [biorXiv](https://www.biorxiv.org/content/10.1101/801571v1).
+Check out this paper on selecting the best metric for your specific question in [Ecology and Evolution](https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.6452).
 <!-- biorXiv preprint -->

### 2021-07-01 inst/gitbook/08_palaeo-demo.Rmd
Added warning for data simulations in demo

@@
 If you are missing any of these, fear not, here are a couple of functions to simulate the missing data, it will surely make your results look funky but it'll let you go through the tutorial.
+
+> **WARNING:** the data generated by the functions `i.need.a.matrix`, `i.need.a.tree`, `i.need.node.data` and `i.need.FADLAD` are used to **SIMULATE** data for this tutorial.
+This is _not_ to be used for publications or analysing real data!
+> If you need a data matrix, a phylogenetic tree or FADLAD data, (`i.need.a.matrix`, `i.need.a.tree` and `i.need.FADLAD`), you will actually need to collect data from the literature or the field!
+If you need node data, you will need to use ancestral states estimations (e.g.
+using `estimate_ancestral_states` from the [`Claddis` package](https://cran.r-project.org/web/packages/Claddis/index.html)).
 

### 2021-07-28 inst/gitbook/03_specific-tutorials.Rmd
typos up

@@
 Check out this paper on selecting the best metric for your specific question in [Ecology and Evolution](https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.6452).
-<!-- biorXiv preprint -->
 You can also use the  [`moms` shiny app](https://tguillerme.shinyapps.io/moms/) to test which metric captures which aspect of traitspace occupancy regarding your specific space and your specific question.
@@
 2 | `ranges` | The range of each dimension | `dispRity` |
-2 | `span.tree.length` | The minimal spanning tree length | [`ape`](https://cran.r-project.org/web/packages/vegan/index.html) (but see also [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree`) |
+2 | `span.tree.length` | The minimal spanning tree length | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree` |
 2 | `variances` | The variance of each dimension | `dispRity` |

### 2021-08-02 inst/gitbook/05_other-functionalities.Rmd
Updated 1.4.2 (Claddis 0.6 update)

@@
 ## Adding missing data
-(matrix_character[sample(1:30, 5)] <- "?")
+matrix_character[sample(1:30, 5)] <- "?"
+
+## This is what it looks like now
+matrix_character
 
@@
 
+## `select.axes`
+
+This function allows you to select which axes (or how many of them) are relevant in your trait space analyses.
+Usually, when the trait space is an ordination, workers select a certain number of axes to reduce the dimensionality of the dataset by removing axes that contain relatively little information.
+This is often done by selecting the axes from which the cumulative individual variance is lower than an arbitrary threshold.
+For example, all the axes that contain together 0.95 of the variance:
+
+```{r}
+## The USArrest example in R
+ordination <- princomp(USArrests, cor = TRUE)
+
+## The loading of each variable
+loadings(ordination)
+
+## Or the same operation but manually
+variances <- apply(ordination$scores, 2, var)
+scaled_variances <- variances/sum(variances)
+sumed_variances <- cumsum(scaled_variances)
+round(rbind(variances, scaled_variances, sumed_variances), 3)
+```
+
+In this example, you can see that the three first axes are required to have at least 0.95 of the variance.
+You can do that automatically in `dispRity` using the `select.axes` function.
+
+```{r}
+## Same operation automatised
+(selected <- select.axes(ordination))
+```
+
+This function does basically what the script above does and allows the results to be plotted or summarised into a table.
+
+```{r}
+## Summarising this info
+summary(selected)
+## Plotting it
+plot(selected)
+## Extracting the dimensions
+## (for the dispRity function for example)
+selected$dimensions
+```
+
+However, it might be interesting to not only consider the variance within the whole trait space but also among groups of specific interest.
+E.g.
+if the 95% of the variance is concentrated in the two first axes for the whole trait space, that does not automatically mean that it is the case for each subset in this space.
+Some subset might require more than the two first axes to express 95% of their variance!
+You can thus use the `select.axes` function to look at the results per group as well as through the whole trait space.
+
+> Note that you can always change the threshold value (default is 0.95). Here for example we set it to 0.9 (we arbitrarily decide that explain 90% of the variance is enough).
+
+```{r}
+## Creating some groups of stats
+states_groups <- list("Group1" = c("Mississippi","North Carolina",
+                                   "South Carolina", "Georgia", "Alabama",
+                                   "Alaska", "Tennessee", "Louisiana"),
+                      "Group2" = c("Florida", "New Mexico", "Michigan",
+                                   "Indiana", "Virginia", "Wyoming", "Montana",
+                                   "Maine", "Idaho", "New Hampshire", "Iowa"),
+                      "Group3" = c("Rhode Island", "New Jersey", "Hawaii", "Massachusetts"))
+## Running the same analyses but per groups
+selected <- select.axes(ordination, group = states_groups, threshold = 0.9)
+## Plotting the results
+plot(selected)
+```
+
+As you can see here, the whole space requires the three first axes to explain at least 90% of the variance (in fact, 95% as seen before).
+However, different groups have a different story!
+The Group 1 and 3 requires 4 dimensions whereas Group 2 requires only 1 dimensions (note how for Group 3, there is actually nearly no variance explained on the second axes)!
+Using this method, you can safely use the four axes returned by the function (`selected$dimensions`) so that every group has at least 90% of their variance explained in the trait space.
+
+If you've used the function if you've already done some grouping in your disparity analyses (e.g.
+using the function `custom.subsets` or `chrono.subsets`), you can use the generated `dispRity` to automatise this analyses:
+
+```{r}
+## Loading the dispRity package demo data
+data(demo_data)
+## A dispRity object with two groups
+demo_data$hopkins
+## Selecting axes on a dispRity object
+selected <- select.axes(demo_data$hopkins)
+plot(selected)
+## Displaying which axes are necessary for which group
+selected$dim.list
+## Note how the whole space needs only 16 axes
+## but both groups need 22 and 23 axes
+```
+
 ## `slice.tree`

### 2022-01-10 inst/gitbook/05_other-functionalities.Rmd
added match.tip.edge entry

@@
 To see a more applied example: this function is used in [chapter 03: specific tutorials](#chrono-subsets).
+
+## `match.tip.edge`
+
+This function matches a vector of discreet tip values with the edges connecting these tips in the `"phylo"` structure.
+This can be used to pull the branches of interest for some specific trait of some group of species or for colouring tree tips based on clades.
+
+For example, with the `charadriiformes` dataset, you can plot the tree with the branches coloured by clade.
+To work properly, the function requires the characteristics of the tip labels (e.g. the clade colour) to match the order of the tips in the tree:
+
+```{r}
+## Loading the charadriiformes data
+data(charadriiformes) 
+## Extracting the tree
+my_tree <- charadriiformes$tree
+## Extracting the data column that contains the clade assignments
+my_data <- charadriiformes$data[, "clade"]
+## Changing the levels names (the clade names) to colours
+levels(my_data) <- c("orange", "blue", "darkgreen")
+my_data <- as.character(my_data)
+## Matching the data rownames to the tip order in the tree
+my_data <- my_data[match(ladderize(my_tree)$tip.label, rownames(charadriiformes$data))]
+```
+
+We can then match this tip data to their common descending edges.
+We will also colour the edges that is not descendant directly from a common coloured tip in grey using `"replace.na = "grey"`.
+Note that these edges are usually the edges at the root of the tree that are the descendant edges from multiple clades.
+
+```{r}
+## Matching the tip colours (labels) to their descending edges in the tree
+## (and making the non-match edges grey)
+clade_edges <- match.tip.edge(my_data, my_tree, replace.na = "grey")
+
+## Plotting the results
+plot(ladderize(my_tree), show.tip.label = FALSE, edge.color = clade_edges)
+```
+
+But you can also use this option to only select some specific edges and modify them (for example making them all equal to one):
+
+```{r}
+## Adding a fixed edge length to the green clade
+my_tree_modif <- my_tree
+green_clade <- which(clade_edges == "darkgreen")
+my_tree_modif$edge.length[green_clade] <- 1
+plot(ladderize(my_tree_modif), show.tip.label = FALSE, edge.color = clade_edges)
+```
 

### 2022-01-18 inst/gitbook/03_specific-tutorials.Rmd
Manual update for v1.7 INIT

@@
 Of course this can be done with multiple trees and be combined with an approach using multiple matrices (see [here](#multi.input))!
+
+## Disparity of variance-covariance matrices (covar) {#covar}
+
+What is a variance-covariance matrix
+How to get them: MCMCglmm @MCMCglmm from @hadfield2010 (but see also @mulTree)
+
+### Creating a `dispRity` object with a `$covar` component
+
+### Visualising covar objects
+
+### Disparity analyses with a `$covar` component
+
+`as.covar`
+
+#### Elaborations and explorations
+
+But see probably paper
+
+Finally there are other tools for manipulating MCMCglmm objects. See other [MCMCglmm utilities](#MCMCglmm-utilities)

### 2022-01-19 inst/gitbook/05_other-functionalities.Rmd
Manual update for v1.7 INIT
added MCMCglmm utilities in the manual

@@
 ```
+
+## `MCMCglmm` utilities {#MCMCglmm-utilities}
+
+Since version `1.7`, the `dispRity` package contains several utility functions for manipulating `"MCMCglmm"` (that is, objects returned by the function `MCMCglmm::MCMCglmm`).
+These objects are a modification of the `mcmc` object (from the package `coda`) and can be sometimes cumbersome to manipulate because of the huge amount of data in it.
+You can use the functions `MCMCglmm.traits` for extracting the number of traits, `MCMCglmm.levels` for extracting the level names, `MCMCglmm.sample` for sampling posterior IDs and `MCMCglmm.covars` for extracting variance-covariance matrices.
+
+```{r}
+## Loading the charadriiformes data that contains a MCMCglmm object
+data(charadriiformes)]
+my_MCMCglmm <- charadriiformes$posteriors
+
+## Which traits where used in this model?
+MCMCglmm.traits(my_MCMCglmm)
+
+## Which levels where used for the model's random terms and/or residuals?
+MCMCglmm.levels(my_MCMCglmm)
+
+## The level names are converted for clarity but you can get them unconverted
+## (i.e. as they appear in the model)
+MCMCglmm.levels(my_MCMCglmm, convert = FALSE)
+
+## Sampling 2 random posteriors samples IDs
+(random_samples <- MCMCglmm.sample(my_MCMCglmm, n = 2))
+
+## Extracting these two random amples
+my_covars <- MCMCglmm.covars(my_MCMCglmm, sample = random_samples)
+```
+
+See more in the [`$covar` section](#covar) on what to do with these `"MCMCglmm"` objects.
 

### 2022-02-11 inst/gitbook/03_specific-tutorials.Rmd
Started the covar section wrtitng

@@
 
-What is a variance-covariance matrix
-How to get them: MCMCglmm @MCMCglmm from @hadfield2010 (but see also @mulTree)
+Variance-covariance matrices are sometimes a useful way to summarise multidimensional data.
+In fact, you can express the variation in your multidimensional dataset directly in terms of how your trait covary rather than simply the positions of your elements in the trait space.
+Furthermore, variance-covariance matrices can be estimated from multidimensional in sometimes more useful ways that simply looking at the the data in your trait space.
+This can be done by describing your data as hierarchical models like generalised linear mixed effect models (glmm).
+For example, you might have a multidimensional dataset where your observations have a nested structure (e.g. they are part of the same phylogeny).
+You can then analyse this data using a glmm with something like `my_data ~ observations + phylogeny + redisduals`.
+For more info on these models [start here](https://en.wikipedia.org/wiki/Generalized_linear_mixed_model).
+For more details on running these models, I suggest using the `MCMCglmm` package (@MCMCglmm) from @hadfield2010 (but see also @mulTree).
 
@@
 
+Once you have a trait space and variance-covariance matrices output from the `MCMCglmm` model, you can use the function `MCMCglmm.subsets` to create a `"dispRity"` object that contains the classic `"dispRity"` data (the matrix, the subsets, etc...) but also a the new `$covar` element:
+
+```{r}
+## Loading the charadriiformes data
+data(charadriiformes)
+```
+
+Here we using precaculated variance-covariance matrices from the charadriiformes dataset that contains a set of posteriors from a `MCMCglmm` model.
+The model here was `data ~ traits + clade specific phylogenetic effect + global phylogenetic effect + residuals`.
+We can retrieve the model information using the  [`MCMCglmm` utilities](#MCMCglmm-utilities) tools, namely the `MCMCglmm.levels` function to directly extract the terms names as used in the model and then build our `"dispRity"` object with the correct data, the posteriors and the correct term names:
+
+```{r}
+## The term names
+model_terms <- MCMCglmm.levels(charadriiformes$posteriors)[1:4]
+## Note that we're ignoring the 5th term of the model that's just the normal residuals
+
+## The dispRity object
+MCMCglmm.subsets(data = charadriiformes$data,
+                 posteriors = charadriiformes$posteriors,
+                 group = model_terms)
+```
+
+As you can see this creates a normal dispRity object with the information you are now familiar with.
+However, we can be more fancy and provide more understandable names for the groups and provide the underlying phylogenetic structure used:
+
+```{r}
+## A fancier dispRity object
+my_covar <- MCMCglmm.subsets(data = charadriiformes$data,
+                             posteriors = charadriiformes$posteriors,
+                             group = model_terms,
+                             tree = charadriiformes$tree,
+                             rename.groups = c(levels(charadriiformes$data$clade), "phylogeny"))
+## Note that the group names is contained in the clade column of the charadriiformes dataset as factors
+```
+
 ### Visualising covar objects
 
+One useful thing to do with these objects is then to visualise them in 2D.
+Here we can use the `covar.plot` function (that has many different options that just `plot.dispRity` for plotting covar objects) to plot the trait space, the 95% confidence interval ellipses of the variance-covariance matrices and the major axes from these ellipses.
+See the `?covar.plot` help page for all the options available:
+
+```{r}
+par(mfrow = c(2,2))
+## The traitspace
+covar.plot(my_covar, col = c("orange", "darkgreen", "blue"), main = "Trait space")
+## The traitspace's variance-covariance mean ellipses
+covar.plot(my_covar, col = c("orange", "darkgreen", "blue", "grey"), main = "Mean VCV ellipses",
+           points = FALSE, ellipses = mean) 
+## The traitspace's variance-covariance mean ellipses
+covar.plot(my_covar, col = c("orange", "darkgreen", "blue", "grey"), main = "Mean major axes",
+           points = FALSE, major.axes = mean)
+## A bit of everything
+covar.plot(my_covar, col = c("orange", "darkgreen", "blue", "grey"), main = "Ten random VCV matrices",
+           points = TRUE, major.axes = TRUE, points.cex = 1/3, n = 10, ellipses = TRUE)
+```
+
 ### Disparity analyses with a `$covar` component

### 2022-02-11 inst/gitbook/03_specific-tutorials.Rmd
finished covar bit (very brief)

@@
 covar.plot(my_covar, col = c("orange", "darkgreen", "blue", "grey"), main = "Ten random VCV matrices",
-           points = TRUE, major.axes = TRUE, points.cex = 1/3, n = 10, ellipses = TRUE)
+           points = TRUE, major.axes = TRUE, points.cex = 1/3, n = 10, ellipses = TRUE, legend = TRUE)
 ```
@@
 
-`as.covar`
-
-#### Elaborations and explorations
-
-But see probably paper
-
-Finally there are other tools for manipulating MCMCglmm objects. See other [MCMCglmm utilities](#MCMCglmm-utilities)
+You can then calculate disparity on the `"dispRity"` object like shown previously.
+For example, you can get the variances of the groups that where used in the model by using the normal `dispRity` function:
+
+```{r}
+summary(dispRity(my_covar, metric = variances))
+```
+
+However this is not applied on the variance-covariance matrices from the posteriors of the `MCMCglmm`.
+To do that, you need to modify the metric to be recognised as a "covar" metric using the `as.covar` function.
+This function transforms any disparity metric (or disparity metric style function) to be applied to the `$covar` part of a `"dispRity"` object.
+Basically this `$covar` part is a list containing, for each posterior sample `$VCV`, the variance-covariance matrix and `$loc`, it's optional location in the traitspace.
+
+```{r}
+## The first variance covariance matrix for the "gulls" group
+my_covar$covar[["gulls"]][[1]]
+```
+
+And this is how `as.covar` modifies the disparity metric:
+
+```{r}
+## Using the variances function on a VCV matrix
+variances(my_covar$covar[["gulls"]][[1]]$VCV)
+
+## The same but using it as a covar metric
+as.covar(variances)(my_covar$covar[["gulls"]][[1]])
+
+## The same but applied to the dispRity function
+summary(dispRity(my_covar, metric = as.covar(variances)))
+```

### 2022-02-15 inst/gitbook/03_specific-tutorials.Rmd
Minor updates on the gitbook

@@
 summary(dispRity(dummy_groups, metric = centroids,
-    centroid = colMeans(matrix.dispRity(dummy_groups, "group1"))))
+    centroid = colMeans(get.matrix(dummy_groups, "group1"))))
 ```
@@
 
-```{r, fig.height = 16, fig.width = 6}
+```{r, fig.height = 12, fig.width = 6}
 ## The results as a dispRity object
@@
  * `cent.tend` for the central tendency to use for summarising the results (default is `median`)
- * digits` option corresponding to the number of decimal places to print (default is `2`)
+ * `digits` option corresponding to the number of decimal places to print (default is `2`)
  * `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
@@
 ```
+
+### Graphical options
 

### 2022-02-15 inst/gitbook/08_palaeo-demo.Rmd
Minor updates on the gitbook

@@
 ```{r, fig.height = 16, fig.width = 8}
-my_test <- test.metric(my_matrix, metric = c(sum, variances), shifts = c("random", "size"))
+my_test <- test.metric(my_matrix, metric = c(sum, dispRity::variances), shifts = c("random", "size"))
 summary(my_test)
@@
 ## Calculating disparity for the bootstrapped and rarefied data
-disparity <- dispRity(rare_bin_morphospace , metric = c(sum, variances))
+disparity <- dispRity(rare_bin_morphospace , metric = c(sum, dispRity::variances))
 ```

### 2022-05-16 inst/gitbook/03_specific-tutorials.Rmd
sync

@@
 
-### Graphical options
+### Graphical options with `...`
+
+As mentioned above all the plots using `plot.dispRity` you can use the `...` options to add any type of graphical parameters recognised by `plot`.
+However, sometimes, plotting more advanced `"dispRity"` objects also calls other generic functions such as `lines`, `points` or `legend`.
+You can fine tune which specific function should be affected by `...` by using the syntax `<function>.<argument>` where `<function>` is usually the function to plot a specific element in the plot (e.g.
+`points`) and the `<argument>` is the specific argument you want to change for that function.
+For example, in a plot containing several elements, including circles (plotted internally with `points`), you can decide to colour everything in blue using the normal `col = "blue"` option.
+But you can also decide to _only_ colour the circles in blue using `points.col = "blue"`!
+
+Here is an example with multiple elements (lines and points) taken from the [disparity with trees](#dispRitree) section below:
+
+```{r}
+## Loading some demo data:
+## An ordinated matrix with node and tip labels
+data(BeckLee_mat99)
+## The corresponding tree with tip and node labels
+data(BeckLee_tree)
+## A list of tips ages for the fossil data
+data(BeckLee_ages)
+
+## Time slicing through the tree using the equal split algorithm
+time_slices <- chrono.subsets(data   = BeckLee_mat99,
+                              tree   = BeckLee_tree,
+                              FADLAD = BeckLee_ages,
+                              method = "continuous",
+                              model  = "acctran",
+                              time   = 15)
+
+par(mfrow = c(2,2))
+## The preview plot with the tree using only defaults
+plot(time_slices, type = "preview", specific.args = list(tree = TRUE))
+## The same plot but by applying general options
+plot(time_slices, type = "preview", specific.args = list(tree = TRUE),
+     col = "blue", main = "General options")
+## The same plot but by applying the colour only to the lines and change of shape only to the points
+plot(time_slices, type = "preview", specific.args = list(tree = TRUE),
+     lines.col = "blue", points.pch = 15, main = "Specific options")
+## And now without the legend
+plot(time_slices, type = "preview", specific.args = list(tree = TRUE),
+     lines.col = "blue", points.pch = 15, legend = FALSE)
+```
 
@@
 
-## Disparity with trees: *dispRitree!*
+## Disparity with trees: *dispRitree!* {#dispRitree}
 

### 2022-05-16 inst/gitbook/06_guts.Rmd
sync

@@
 
-####  `matrix.dispRity`
+####  `get.matrix`
 
@@
 ## Extracting the matrix containing the coordinates of the elements at time 50
-str(matrix.dispRity(disparity, "50"))
+str(get.matrix(disparity, "50"))
 
@@
 ## (15 elements) from the second group (80 Mya)
-str(matrix.dispRity(disparity, subsets = 1, bootstrap = 3, rarefaction = 2))
+str(get.matrix(disparity, subsets = 1, bootstrap = 3, rarefaction = 2))
 ```
@@
 
-#### `extract.dispRity`
+#### `get.disparity`
 
@@
 ## Extracting the observed disparity (default)
-extract.dispRity(disparity)
+get.disparity(disparity)
 
@@
 ## 10th rarefaction level from the second subsets (80 Mya)
-extract.dispRity(disparity, observed = FALSE, subsets = 2, rarefaction = 10)
+get.disparity(disparity, observed = FALSE, subsets = 2, rarefaction = 10)
 ```

### 2022-05-16 inst/gitbook/09_gmm-demo.Rmd
sync

@@
 
-```{r, fig.width = 8, fig.height = 16}
+```{r, fig.width = 8, fig.height = 16, eval = TRUE}
 ## Defining a the procrustes variance metric
@@
 proc.var <- function(matrix) {sum(matrix^2)/nrow(matrix)}
+```
+
+```{r, fig.width = 8, fig.height = 16, eval = FALSE}
+
 ## The size metric

### 2022-07-21 inst/gitbook/03_specific-tutorials.Rmd
vignette update

@@
 1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
+1 | `disalignment` | The rejection of the centroid of a matrix from the major axis of another (typically an `"as.covar"` metric) | `dispRity` |
 2 | `displacements` | The ratio between the distance from a reference and the distance from the centroid | `dispRity` |
@@
 2 | `point.dist` | The distance between one group and the point of another group | `dispRity` |
+2 | `projections` | The distance *on* (projection) or *from* (rejection) an arbitrary vector | `dispRity` |
+1 | `projections.between` | `projections` metric applied between groups | `dispRity` |
 2 | `projections.tree` | The `projections` metric but where the vector can be based on a tree | `dispRity` |
-2 | `projections` | The distance *on* (projection) or *from* (rejection) an arbitrary vector | `dispRity` |
 2 | `quantiles` | The *n*th quantile range per axis | `dispRity` |
@@
 
-#### Projections and phylo projections: elaboration and exploration
+#### Projections and phylo projections: elaboration and exploration {#projections}
 
@@
                  between.groups = TRUE))
+```
+
+##### `projections.between` and `disalignment`
+
+These two metrics are typically based on variance-covariance matrices from a `dispRity` object that has a `$covar` component (see more about that [here](#covar)).
+Both are based on the `projections` metric and can take the same optional arguments (more info [here](#projections)).
+The examples and explanations below are based on the default arguments but it is possible (and easy!) to change them.
+
+We are going to use the `charadriiformes` example for both metrics (see more about that [here](#covar)).
+
+```{r}
+## Loading the charadriiformes data
+data(charadriiformes)
+
+## Creating the dispRity object (see the #covar section in the manual for more info)
+my_covar <- MCMCglmm.subsets(n = 50,
+                             data = charadriiformes$data,
+                             posteriors = charadriiformes$posteriors,
+                             group = MCMCglmm.levels(charadriiformes$posteriors)[1:4],
+                             tree = charadriiformes$tree,
+                             rename.groups = c(levels(charadriiformes$data$clade), "phylogeny"))
+```
+
+The first metric, `projections.between` projects the major axis of one group (`matrix`) onto the major axis of another one (`matrix2`).
+For example we might want to know how some groups compare in terms of angle (orientation) to a base group:
+
+```{r}
+## Creating the list of groups to compare
+comparisons_list <- list(c("gulls", "phylogeny"),
+                         c("plovers", "phylogeny"),
+                         c("sandpipers", "phylogeny"))
+
+## Measuring the angles between each groups
+## (note that we set the metric as.covar, more on that in the #covar section below)
+groups_angles <- dispRity(data = my_covar,
+                          metric = as.covar(projections.between),
+                          between.groups = comparisons_list,
+                          measure = "degree")
+## And here are the angles in degrees:
+summary(groups_angles)
+```
+
+The second metric, `disalignment` rejects the centroid of a group (`matrix`) onto the major axis of another one (`matrix2`).
+This allows to measure wether the center of a group is aligned with the major axis of another.
+A disalignement value of 0 means that the groups are aligned. A higher disalignment value means the groups are more and more disaligned.
+We can use the same set of comparisons as in the `projections.between` examples to measure which group is most aligned (less disaligned) with the phylogenetic major axis:
+
+```{r}
+## Measuring the disalignement of each group
+groups_alignement <- dispRity(data = my_covar,
+                              metric = as.covar(disalignment),
+                              between.groups = comparisons_list)
+## And here are the groups alignment (0 = aligned)
+summary(groups_alignement)
 ```

### 2022-08-09 inst/gitbook/05_other-functionalities.Rmd
Minor updates on the gitbook
updated gitbook for 1.7

@@
 my_tree_modif$edge.length[green_clade] <- 1
-plot(ladderize(my_tree_modif), show.tip.label = FALSE, edge.color = clade_edges)
+plot(ladderize(my_tree_modif), show.tip.label = FALSE,
+     edge.color = clade_edges)
 ```
@@
 ## Loading the charadriiformes data that contains a MCMCglmm object
-data(charadriiformes)]
+data(charadriiformes)
 my_MCMCglmm <- charadriiformes$posteriors
@@
 ## Plotting the first column as a pairwise comparisons
-pair.plot(data, what = 1, col = c("orange", "blue"), legend = TRUE, diag = 1)
+pair.plot(data, what = 1, col = c("orange", "blue"), legend = TRUE,
+          diag = 1)
 ```
@@
 ```{r}
-## The same plot as before without the diagonal being the maximal observed value
-pair.plot(data, what = 1, col = c("orange", "blue"), legend = TRUE, diag = "max")
-## Highlighting with an asterisk which squares have a value below 0.2
+## The same plot as before without the diagonal being
+## the maximal observed value
+pair.plot(data, what = 1, col = c("orange", "blue"), legend = TRUE,
+          diag = "max")
+## Highlighting with an asterisk which squares have a value
+## below 0.2
 pair.plot(data, what = 1, binary = 0.2, add = "*", cex = 2)
@@
 ```{r}
-image(t(na_matrix[-as.numeric(reduction$rows.to.remove), ]), col = "black")
+image(t(na_matrix[-as.numeric(reduction$rows.to.remove), ]),
+      col = "black")
 ```

### 2022-08-09 inst/gitbook/03_specific-tutorials.Rmd

@@
 ## Testing whether mean is dimension-level 1
-if(make.metric(mean, silent = TRUE) != "level1") {
+if(make.metric(mean, silent = TRUE)$type != "level1") {
     message("The metric is not dimension-level 1.")
@@
 ## Testing whether var is dimension-level 1
-if(make.metric(var, silent = TRUE) != "level1") {
+if(make.metric(var, silent = TRUE)$type != "level1") {
     message("The metric is not dimension-level 1.")
@@
      col = "blue", main = "General options")
-## The same plot but by applying the colour only to the lines and change of shape only to the points
+## The same plot but by applying the colour only to the lines
+## and change of shape only to the points
 plot(time_slices, type = "preview", specific.args = list(tree = TRUE),

### 2022-08-09 inst/gitbook/index.Rmd

@@
 version_master <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
-version_master <- 1.6
+version_master <- 1.7
 
@@
 version_release <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
-version_release <- 1.6
+version_release <- 1.7
 
@@
 version_CRAN <- strsplit(strsplit(version_raw, split = "\n<td>Version:</td>\n<td>")[[1]][2], split = "</td>\n</tr>\n<tr>\n<td>Depends:")[[1]][1]
-version_CRAN <- 1.6
+version_CRAN <- 1.7
 ```

### 2022-09-20 inst/gitbook/05_other-functionalities.Rmd
added MCMCglmm.variance and updates on randtest.dispRity

@@
 You can use the functions `MCMCglmm.traits` for extracting the number of traits, `MCMCglmm.levels` for extracting the level names, `MCMCglmm.sample` for sampling posterior IDs and `MCMCglmm.covars` for extracting variance-covariance matrices.
+You can also quickly calculate the variance (or relative variance) for each terms in the model using `MCMCglmm.variance` (the variance is calculated as the sum of the diagonal of each variance-covariance matrix for each term).
 
@@
 
-## Extracting these two random amples
+## Extracting these two random samples
 my_covars <- MCMCglmm.covars(my_MCMCglmm, sample = random_samples)
+
+## Plotting the variance for each term in the model
+boxplot(MCMCglmm.variance(my_MCMCglmm), horizontal = TRUE, las = 1,
+        xlab = "Relative variance",
+        main = "Variance explained by each term")
 ```

### 2022-09-20 R/randtest.dispRity.R

@@ -73,6 +73,7 @@ randtest.dispRity <- function(data, subsets, metric, replicates = 100, resample 
     ## Distribution and subset
     data_class <- check.class(data, c("matrix", "dispRity"))
     inherits_metrics <- inherits_subsets <- FALSE
+    subsets_names <- NULL
 
     if(data_class == "dispRity") {
 
@@ -87,6 +88,11 @@ randtest.dispRity <- function(data, subsets, metric, replicates = 100, resample 
             ## Set the sampling population list
             sample_pop <- replicate(length(subsets), list(sample_pop))
         } else {
+            ## Get the subset names (if available)
+            if(!is.null(names(subsets))) {
+                subsets_names <- names(subsets)
+            }
+
             ## Split the subsets in observed and random
             subsets_obs  <- lapply(subsets, `[`, 1)
             subsets_rand <- lapply(subsets, function(x) unname(x[-1]))
@@ -226,9 +232,16 @@ randtest.dispRity <- function(data, subsets, metric, replicates = 100, resample 
         results <- mapply(function(results, call_summary) {results$call <- call_summary; return(results)}, results, call_summary, SIMPLIFY = FALSE)
 
         ## Add names to the results (if any)
-        if(!is.null(sub_names)) {
-            names(results) <- sub_names
+        if(!is.null(subsets_names)) {
+            ## Input names
+            names(results) <- subsets_names
+        } else {
+            ## Auto naming (if possible)
+            if(!is.null(sub_names)) {
+                names(results) <- sub_names
+            }
         }
+
         class(results) <- c("dispRity", "randtest")
         
         return(results)

### 2023-11-17 tests/testthat/test-dispRity.multi.R
Updated and finalised test
good to go for merging!

@@ -1,176 +1,116 @@
-# #context("dispRity.utilities")
+test_that("dispRity.multi.split", {
+    load("bound_test_data.rda")
+    trees <- bound_test_data$trees
+    matrices <- bound_test_data$matrices
 
-# # 1 - sanitizing, detect if the trees and matrices don't match. If they don't you need the same number of trees and matrices. Then output apply.multi.trees activator
-
-
-# # - test case Mario
-
-# # Email Mario:
-# # For cleaning the trees, you can use dispRity::clean.data
-# # For removing the 0 branch lengths, you can use dispRity::remove.zero.brlen (which shifts nodes to remove 0 values rather than adding branch lengths which can results in the trees not being ultrametric anymore)
-
-# matrix <- Claddis::read_nexus_matrix("~/R/data/Matrix_20230214", equalize_weights = TRUE)
-# tree <- read.nexus("~/R/data/sample100")
-# data <- Claddis::ordinate_cladistic_matrix(morpho_matrix)$vectors
-
-# ## Tips in the tree and data are the same
-
-
-# #Extract the vector matrix from the Claddis morphospace object to feed it to dispRity
-# morphosp = morphospace$vectors
-# rownames(morphosp) = c(tree$tip.label,tree$node.label)
-
-
-# ######POARS analysis
-# ####################
-
-# ## Time slices
-# time_slices <- seq(from = 0, to = 280, by = 10)
-
-# ## Gradual splits model
-# gradual_model <- chrono.subsets(morphosp, tree, time = time_slices, method = "continuous",
-#                                 model = "gradual.split", inc.nodes=TRUE)
+    ## Split just data
+    data <- fill.dispRity(make.dispRity(data = matrices))
+    test <- dispRity.multi.split(data)
+    expect_is(test, "list")
+    expect_equal(length(test), 3)
+    for(i in 1:3) {
+        expect_is(test[[i]], "dispRity")
+        expect_equal(length(test[[i]]$matrix), 1)
+        expect_null(test[[i]]$tree[[1]])
+    }
+    ## Three trees and one matrix
+    data <- fill.dispRity(make.dispRity(data = matrices[[1]]))
+    test <- dispRity.multi.split(data)
+    expect_is(test, "list")
+    expect_equal(length(test), 1)
+    expect_is(test[[1]], "dispRity")
[…]

### 2023-11-17 inst/gitbook/03_specific-tutorials.Rmd
Updated and finalised test
good to go for merging!
Added vignette for multi.ace

@@
 
- * `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any *m* elements drawn from the data)
- * `single` where only one random element is replaced by one other random element for each pseudo-replicate
+ * `"full"` where the bootstrapping is entirely stochastic (*n* elements are replaced by any *m* elements drawn from the data)
+ * `"single"` where only one random element is replaced by one other random element for each pseudo-replicate
+ * `"null"` where every element is resampled across the whole matrix (not just the subsets).
+I.e.
+for each subset of *n* elements, this algorithm resamples *n* elements across *ALL* subsets (not just the current one).
+If only one subset (or none) is used, this does the same as the `"full"` algorithm.
 
@@
 2 | `ranges` | The range of each dimension | `dispRity` |
+1 | `roundness` | The integral of the ranked scaled eigenvalues of a variance-covariance matrix | `dispRity`
 2 | `span.tree.length` | The minimal spanning tree length | [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html)`::spantree` |
@@
 
-You can find more informations on the vast variety of metrics that you can use in your analysis in this [preprint](https://www.biorxiv.org/content/10.1101/801571v1).
-<!-- biorXiv preprint -->
+You can find more informations on the vast variety of metrics that you can use in your analysis in this [paper](https://onlinelibrary.wiley.com/doi/10.1002/ece3.6452).
 
@@
     ranges = |max(d_i) - min(d_i)|
+\end{equation}
+
+\begin{equation}
+    roundness = \int_{i = 1}^{n}{\frac{\lambda_{i}}{\text{max}(\lambda)}}
 \end{equation}
@@
                                             tree_space[1, ])))
+```
+
+#### Roundness
+
+The roundness coefficient (or metric) ranges between 0 and 1 and expresses the distribution of and ellipse' major axis ranging from 1, a totally round ellipse (i.e.
+a circle) to 0 a totally flat ellipse (i.e.
+a line).
+A value of $0.5$ represents a regular ellipse where each major axis is half the size of the previous major axis.
+A value $> 0.5$ describes a pancake where the major axis distribution is convex (values close to 1 can be pictured in 3D as a cr\`{e}pes with the first two axis being rather big - a circle - and the third axis being particularly thin; values closer to $0.5$ can be pictured as flying saucers).
+Conversely, a value $< 0.5$ describes a cigar where the major axis distribution is concave (values close to 0 can be pictured in 3D as a spaghetti with the first axis rather big and the two next ones being small; values closer to $0.5$ can be pictured in 3D as a fat cigar).
+
+This is what it looks for example for three simulated variance-covariance matrices in 3D:
+
+```{r, eval = TRUE, echo = FALSE, message = FALSE}
+## Note that the plotting functions requires the ellipse, rgl and VCVtools packages (but not need to load them)
+require("ellipse")
+require("rgl")
+require("VCVtools")
+## To install them
+# install.packages(c("ellipse", "rgl"))
+# devtools::install_github("TGuillerme/VCVtools")
+
+## knitr RGL setup
+# knitr::knit_hooks$set(webgl = hook_webgl)
+options(rgl.useNULL = TRUE) # Suppress the separate window.
+
+## A 3D sphere
+sphere  <- VCVtools::make.VCV(shape = 1, dimensions = 3, min.thick = 0.01)
+pancake <- VCVtools::make.VCV(shape = 0.75, dimensions = 3, min.thick = 0.01)
+cigar   <- VCVtools::make.VCV(shape = 0.25, dimensions = 3, min.thick = 0.01)
+```
+
+```{r, fig.height = 4, fig.width = 4, echo = FALSE}
+# options(rgl.printRglwidget = TRUE)
+VCVtools::plot.VCV(sphere, dimensions = c(1,2,3), col.major.axes = c("orange", "blue", "darkgreen"), main = "Sphere's roundness: 1", lwd.major.axes = 3)
+rglwidget()
+```
+```{r, fig.height = 4, fig.width = 4, echo = FALSE}
+VCVtools::plot.VCV(pancake, dimensions = c(1,2,3), col.major.axes = c("orange", "blue", "darkgreen"), main = "Pancake's roundness: ~0.70", lwd.major.axes = 3)
+rglwidget()
+```
+```{r, fig.height = 4, fig.width = 4, echo = FALSE}
+VCVtools::plot.VCV(cigar, dimensions = c(1,2,3), col.major.axes = c("orange", "blue", "darkgreen"), main = "Cigar's roundness: ~0.25", lwd.major.axes = 3)
+rglwidget()
+```
+```{r, eval = TRUE, fig.height = 4, fig.width = 12, echo = FALSE}
+par(mfrow = c(1,3))
+VCVtools::plot.roundness(sphere, dimensions = c(1,2,3), col.lines = c("orange", "blue", "darkgreen"), main = "Sphere's roundness: 1")
+VCVtools::plot.roundness(pancake, dimensions = c(1,2,3), col.lines = c("orange", "blue", "darkgreen"), main = "Pancake's roundness: ~0.70")
+VCVtools::plot.roundness(cigar, dimensions = c(1,2,3), col.lines = c("orange", "blue", "darkgreen"), main = "Cigar's roundness: ~0.25")
 ```
@@
 
+Note that since the version `1.8` the trees and the matrices don't have to match allowing to run disparity analyses with variable matrices and trees.
+This can be useful when running ancestral states estimations from a tree distribution where not all trees have the same topology.
+
 ## Disparity with trees: *dispRitree!* {#dispRitree}

### 2023-11-17 inst/gitbook/index.Rmd
Up manual and vignettes

@@
 github-repo: TGuillerme/dispRity
-description: "dispRity R package vignette"
+description: "dispRity R package manual"
 ---

### 2023-11-24 tree_hackathon.Rmd
Create tree_hackathon.Rmd

# tree hackathon!

## Implementation idea

The idea would be that users do `chrono.subsets` with all their desired options to generate the time bins (this is already implemented and solid), and then they could use the function `get.tree` to extract the trees from the resulting `dispRity` objects.
In the case of the time bins, it would return a list of length number of time bins with each list element containing a `multiPhylo` object that itself contains at least one tree (the one in the time bin).

For example, if we have this tree:

```{r}
## Testing detect edges and get new tree
set.seed(1)
simple_tree <- rtree(5)
simple_tree$edge.length <- rep(1, Nedge(simple_tree))
simple_tree$root.time <- 4
simple_tree$node.label <- letters[1:4]
plot(simple_tree, show.tip.label = FALSE); axisPhylo()
nodelabels()
nodelabels(simple_tree$node.label, adj = -1, col = "blue")
edgelabels()
tiplabels()
tiplabels(simple_tree$tip.label, adj = -1, col = "blue")
abline(v = c(0, 1, 2, 3, 3.8), col = "grey", lty = 2)
```

We can have the pipeline looking something like:

```{r, eval = FALSE}
## Some data (to be ignored)
ignore_data <- matrix(1, ncol = 1, nrow = 9, dimnames = list(c(simple_tree$tip.label, simple_tree$node.label)))

## Using chrono.subsets to do the time bins
my_bins <- chrono.subsets(data = ignore_data, tree = simple_tree, method = "discrete", time = c(0, 1, 2, 3, 3.8))

## Get the trees
my_subtrees <- get.tree(my_bins)
```

You can then measure branch length for each tree using something like:

```{r}
my_branch_lengths_per_bin <- lapply(my_subtrees, lapply, function(x) sum(x$edge.length))
```

Using `chrono.subsets` and `get.tree` allows to generalise the method for any type of time bins (e.g. with tips or nodes spanning across time - FADLAD), and any number of trees (e.g. using a tree distribution rather than a single tree).

### Output format needed

For all that to work smoothly we want `my_subtrees` in this example to be something like:

```
my_subtrees
   |
   \---<first_bin_name>
   |    |
   |    \--<"multiPhylo"> = list of trees in the first bin for the first tree
   |                        (if only one subtree is in that bin, it's a multiPhylo of length 1)
   |
   \---<second_bin_name>
   |    |
   |    \--<"multiPhylo"> = list of trees in the first bin for the first tree
   |
   \--- etc.
```

Easy!

# What does `chrono.subsets` outputs

(this generalises to `custom.subsets` by the way - but ignore for this part of the implementation).

```
object
    |
    \---$matrix* = class:"list" (a list containing the orginal matrix/matrices)
    |   |
    |   \---[[1]]* = class:"matrix" (the matrix (trait-space))
    |   |
    |   \---[[...]] = class:"matrix" (any additional matrices)
    |
    |
    \---$tree* = class:"multiPhylo" (a list containing the attached tree(s) or NULL)
    |   |
    |   \---[[1]] = class:"phylo" (the first tree)
    |   |
    |   \---[[...]] = class:"phylo" (any additional trees)  
    |
    |
    \---$call* = class:"list" (details of the methods used)
    |   |
    |   \---$subsets = class:"character": some info about the subset type (discrete, continuous, custom)
    |   |
    |   \---more stuff. IGNORE
    |
    \---$subsets* = class:"list" (subsets as a list)
    |   |
    |   \---[[1]]* = class:"list" (first item in subsets list)
    |   |   |
    |   |   \---$elements* = class:"matrix" (one column matrix containing the elements within the first subset)
    |   |   |
    |   |   \---[[2]] = IGNORE (for bootstrap and rarefaction)
    |   |
    |   \---[[2]] = class:"list" (second item in subsets list)
    |   |   |
    |   |   \---$elements* = class:"matrix" (one column matrix containing the elements within the second subset)
    |   |   |
    |   |   \---[[2]] = IGNORE (for bootstrap and rarefaction)
    |   |
    |   \---[[...]] = class:"list" (the following subsets)
    |       |
    |       \---$elements* = class:"matrix" (a one column matrix containing the elements within this subset)
    |       |
    |       \---[[...]] = IGNORE (for bootstrap and rarefaction)
    |
    \--- more stuff. IGNORE
```

Basically most info about the subsets generated by `chrono.subsets` is either in `data$call$subsets` for the meta data and in `data$subsets` for the actually data.
The names of the subsets can be extracted using `name.subsets()`.
What is in each subset is in the `$elements` part, it's a matrix of usually one column with the ID of the elements.

For example:
```
data$subsets[[1]]
    $elements
         [,1]
    [1,]    1
    [2,]    2
    [3,]    3
```
Here the integers 1:3 in the column refer to the row numbers in the data. So you can access the sub-matrix using `data$matrix[[1]][data$subsets[[1]]$elements[, 1]]` or it's rownames using `rownames(data$matrix[[1]][data$subsets[[1]]$elements[, 1]])`.
But no need to worry here. This is handled by some other bits.

# Using these elements to get the sub trees

The idea is to use these elements though to access the parts of the subtree and generate it as an output.
So for example, if elements 1 to 3 are the tips `t4` and `t3` and the node `d` in the simple tree above, it should output the sub tree `(d(t4,t3))`.

I've implementing this in the following files:

 * testing: `dispRity/test/testthat/test-dispRity.utilities.R` l.725 ("get.tree with subsets")
 * code: `dispRity/R/dispRity.utilities.R`: l.539 `get.tree.R`
 * code: `dispRity/R/dispRity.utilities_fun.R`: l.137 `detect.edges`, `get.new.tree`, `get.one.tree.subset`.

The best way to load is to go to the root of the repo and use `devtools::load_all()` to get all the functions linked:

```{r}
setwd("dispRity/")
devtools::load_all()
```

So far I have:

1. find the edges connecting the elements (`detect.edges`) which is used in 2
2. using `get.new.tree` to get the tree using the elements as inputs which is used in 3
3. using `get.one.tree.subset` as a function to `lapply` across all subsets in `get.tree`

The problem though is that:

 * it doesn't cut branch lengths
 * it doesn't take age into account
 * it doesn't really works

# Working idea

1. Feed the list of elements (in our case element 9, 2)
2. Feed the boundary ages (1 - 0.2)
3. Have a function to detect the subtree with the element

    a.

tree$edge
[1,] 1 1
tree$nodelabel
"d"

    b. go back to the orginal tree
    c. find the sub tree(s) that go into the future
    d. cut these sub trees(s) at the younger age (use `slice.tree`)

### 2023-11-24 tree_hackathon.Rmd
Update tree_hackathon.Rmd

@@
 2. Feed the boundary ages (1 - 0.2)
-3. Have a function to detect the subtree with the element
-
-    a.
-
-tree$edge
-[1,] 1 1
-tree$nodelabel
-"d"
-
-    b. go back to the orginal tree
-    c. find the sub tree(s) that go into the future
-    d. cut these sub trees(s) at the younger age (use `slice.tree`)
+    * Edit the age boundaries to that the old age becomes the root of the tree and the young age becomes the cutting point
+3. Recreate the subtree spanning from the selected elements (descendant from them)
+    * Make sure that the elements names are kept
+4. Using `slice.tree` to cut the upper boundary (young age)
+5. Return the new subtree

### 2023-11-27 inst/gitbook/06_guts.Rmd
added MCMCglmm.variance and updates on randtest.dispRity
added example for get.tree in vignette

@@
 
+#### `name.subsets`
+
+This function gets you the names of the subsets in a `dispRity` object as a vector.
+
+```{r}
+## What are they called?
+name.subsets(disparity)
+```
+
 #### `size.subsets`
@@
 add.tree(disparity, tree = BeckLee_tree)
+```
+
+Note that `get.tree` can also be used to extract trees from different subsets (custom or continuous/discrete subsets).
+
+For example, if we have three time bins like in the example below we have three time bins and we can extract the subtrees for these three time bins in different ways using the option `subsets` and `to.root`:
+
+```{r}
+## Load the Beck & Lee 2014 data
+data(BeckLee_tree) ; data(BeckLee_mat99) ; data(BeckLee_ages)
+
+## Time binning (discrete method)
+## Generate two discrete time bins from 120 to 40 Ma every 20 Ma
+time_bins <- chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree,
+                            method = "discrete", time = c(120, 100, 80, 60),
+                            inc.nodes = TRUE, FADLAD = BeckLee_ages)
+
+## Getting the subtrees all the way to the root
+root_subsets <- get.tree(time_bins, subsets = TRUE)
+
+## Plotting the bin contents
+old_par <- par(mfrow = c(2,2))
+plot(BeckLee_tree, main = "original tree", show.tip.label = FALSE)
+axisPhylo()
+abline(v = BeckLee_tree$root.time - c(120, 100, 80, 60))
+for(i in 1:3) {
+     plot(root_subsets[[i]], main = names(root_subsets)[i],
+          show.tip.label = FALSE)
+     axisPhylo()
+}
+par(old_par)
+```
+
+But we can also extract the subtrees containing only branch lengths for the actual bins using `to.root = FALSE`:
+
+```{r}
+## Getting the subtrees all the way to the root
+bin_subsets <- get.tree(time_bins, subsets = TRUE, to.root = FALSE)
+
+## Plotting the bin contents
+old_par <- par(mfrow = c(2,2))
+plot(BeckLee_tree, main = "original tree", show.tip.label = FALSE)
+axisPhylo()
+abline(v = BeckLee_tree$root.time - c(120, 100, 80, 60))
+for(i in 1:3) {
+     plot(bin_subsets[[i]], main = names(bin_subsets)[i],
+          show.tip.label = FALSE)
+     axisPhylo()
+}
+par(old_par)
+```
+
+This can be useful for example for calculating the branch lengths in each bin:
+
+```{r}
+## How many cumulated phylogenetic diversity in each bin?
+lapply(bin_subsets, function(tree) sum(tree$edge.length))
 ```

### 2023-11-27 tests/testthat/test-dispRity.utilities.R

@@ -737,7 +737,7 @@ test_that("get.tree with subsets", {
     tiplabels()
     tiplabels(simple_tree$tip.label, adj = -1, col = "blue")
     abline(v = c(0, 1, 2, 3, 3.8), col = "grey", lty = 2)
-    dev.new()
+    # dev.new()
     tree <- simple_tree
 
     ## Detect edges

### 2023-12-06 inst/gitbook/03_specific-tutorials.Rmd
Up manual and vignettes
updated book

@@
 1 | `edge.length.tree` | The edge lengths of the elements on a tree | `ape` |
-1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
+1 | `ellipsoid.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
 1 | `func.div` | The functional divergence (the ratio of deviation from the centroid) | `dispRity` (similar to [`FD`](https://cran.r-project.org/web/packages/FD/index.html)`::dbFD$FDiv` but without abundance)|
@@
 \begin{equation}
-    ellipse.volume = \frac{\pi^{d/2}}{\Gamma(\frac{d}{2}+1)}\displaystyle\prod_{i=1}^{d} (\lambda_{i}^{0.5})
+    ellipsoid.volume = \frac{\pi^{d/2}}{\Gamma(\frac{d}{2}+1)}\displaystyle\prod_{i=1}^{d} (\lambda_{i}^{0.5})
 \end{equation}
@@
 #### Volumes and surface metrics
-The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `n.ball.volume`  all measure the surface or the volume of the ordinated space occupied:
+The functions `ellipsoid.volume`, `convhull.surface`, `convhull.volume` and `n.ball.volume`  all measure the surface or the volume of the ordinated space occupied:
 
@@
 ## Calculating the ellipsoid volume
-summary(dispRity(dummy_space, metric = ellipse.volume))
-```
-> WARNING: in such dummy space, this gives the estimation of the ellipsoid volume, not the real ellipsoid volume! See the cautionary note in `?ellipse.volume`.
+summary(dispRity(dummy_space, metric = ellipsoid.volume))
+```
+> WARNING: in such dummy space, this gives the estimation of the ellipsoid volume, not the real ellipsoid volume! See the cautionary note in `?ellipsoid.volume`.
 
@@
 ```{r}
-# Creating a 2D uniform space
+## Creating a 2D uniform space
 example_space <- space.maker(300, 2, runif)

### 2023-12-06 inst/gitbook/06_guts.Rmd
updated book

@@
 
-#### `rescale.dispRity`
+#### `scale.dispRity`
 
@@
 ## Scaling the same disparity values
-head(summary(rescale.dispRity(disparity, scale = TRUE)))
+head(summary(scale.dispRity(disparity, scale = TRUE)))
 
 ## Scaling and centering:
-head(summary(rescale.dispRity(disparity, scale = TRUE, center = TRUE)))
+head(summary(scale.dispRity(disparity, scale = TRUE, center = TRUE)))
 
 ## Rescaling the value by dividing by a maximum value
-head(summary(rescale.dispRity(disparity, max = 10)))
+head(summary(scale.dispRity(disparity, max = 10)))
 ```

### 2023-12-11 inst/gitbook/05_other-functionalities.Rmd

@@
 
-## `char.diff`
+## `char.diff` {#char.diff}
 
@@
 ```
+
+## `multi.ace`
+
+This function allows to run the `ape::ace` function (ancestral characters estimations) on multiple trees.
+In it's most basic structure (e.g.
+using all default arguments) this function is using a mix of `ape::ace` and `castor::asr_mk_model` depending on the data and the situation and is generally faster than both functions when applied to a list of trees.
+However, this function provides also some more complex and modular functionalities, especially appropriate when using discrete morphological character data.
+
+### Using different character tokens in different situations
+
+This data can be often coded in non-standard way with different character tokens having different meanings.
+For example, in some datasets the token `-` can mean "the trait is inapplicable" but this can be also coded by the more conventional `NA` or can mean "this trait is missing" (often coded `?`).
+This makes the meaning of specific tokens idiosyncratic to different matrices.
+For example we can have the following discrete morphological matrix with all the data encoded:
+
+```{r}
+set.seed(42)
+## A random tree with 10 tips
+tree <- rcoal(10)
+## Setting up the parameters
+my_rates = c(rgamma, rate = 10, shape = 5)
+
+## Generating a bunch of trees
+multiple_trees <- rmtree(5, 10)
+
+## A random Mk matrix (10*50)
+matrix_simple <- sim.morpho(tree, characters = 50, model = "ER", rates = my_rates,
+                            invariant = FALSE)
+matrix_simple[1:10, 1:10]
+```
+
+But of course, as mentioned above, in practice, such matrices have more nuance and can including missing characters, ambiguous characters, multi-state characters, inapplicable characters, etc...
+All these coded and defined by different authors using different tokens (or symbols).
+Let's give it a go and transform this simple data to something more messy:
+
+```{r}
+## Modify the matrix to contain missing and special data
+matrix_complex <- matrix_simple
+## Adding 50 random "-" tokens
+matrix_complex[sample(1:length(matrix_complex), 50)] <- "-"
+## Adding 50 random "?" tokens
+matrix_complex[sample(1:length(matrix_complex), 50)] <- "?"
+## Adding 50 random "0%2" tokens
+matrix_complex[sample(1:length(matrix_complex), 50)] <- "0%2"
+matrix_complex[1:10,1:10]
+```
+
+In `multi.ace` you can specify what all these tokens actually mean and how the code should interpret them.
+For example, `-` often means inapplicable data (i.e.
+the specimen does not have the coded feature, for example, the colour of the tail of a tailless bird); or `?` that often means missing data (i.e.
+it is unknown if the specimen has a tail or not since only the head was available).
+And more than the differences in meaning between these characters, different people treat these characters differently even if they have the same meaning for the token.
+For example, one might want to treat `-` as meaning "we don't know" (which will be treated by the algorithm as "any possible trait value") or "we know, and it's no possible" (which will be treated by the algorithm as `NA`).
+Because of this situation, `multi.ace` allows combining any special case marked with a special token to a special behaviour.
+For example we might want to create a special case called `"missing"` (i.e.
+the data is missing) that we want to denote using the token `"?"` and we can specify the algorithm to treat this `"missing"` cases (`"?"`) as treating the character token value as "any possible values".
+This behaviour can be hard coded by providing a function with the name of the behaviour.
+For example:
+
+```{r}
+## The specific token for the missing cases (note the "\\" for protecting the value)
+special.tokens <- c("missing" = "\\?")
+
+## The behaviour for the missing cases (?)
+special.behaviour <- list(missing <- function(x, y) return(y))
+## Where x is the input value (here "?") and y is all the possible normal values for the character
+```
+
+This example shows a very common case (and is actually used by default, more on that below) but this architecture allows for very modular combination of tokens and behaviours.
+For example, in our code above we introduced the token `"%"` which is very odd (to my knowledge) and might mean something very specific in our case.
+Say we want to call this case `"weirdtoken"` and mean that whenever this token is encountered in a character, it should be interpreted by the algorithm as the values 1 and 2, no matter what:
+
+```{r}
+## Set a list of extra special tokens
+my_spec_tokens <- c("weirdtoken" = "\\%")
+
+## Weird tokens are considered as state 0 and 3
+my_spec_behaviours <- list()
+my_spec_behaviours$weirdtoken <- function(x,y) return(c(1,2))
+```
+
+If you don't need/don't have any of this specific tokens, don't worry, most special but common tokens are handled by default as such:
+
+```{r}
+## The token for missing values:
+default_tokens <- c("missing"      = "\\?",
+## The token for inapplicable values:                    
+                    "inapplicable" = "\\-",
+## The token for polymorphisms:
+                    "polymorphism" = "\\&",
+## The token for uncertainties:
+                    "uncertanity"  = "\\/")
+```
+
+With the following associated default behaviours
+
+```{r}
+## Treating missing data as all data values
+default_behaviour <- list(missing      <- function(x,y) y,
+## Treating inapplicable data as all data values (like missing)    
+                          inapplicable <- function(x, y) y,
+## Treating polymorphisms as all values present:
+                          polymorphism <- function(x,y) strsplit(x, split = "\\&")[[1]],
+## Treating uncertainties as all values present (like polymorphisms):
+                          uncertanity  <- function(x,y) strsplit(x, split = "\\&")[[1]])
+```
+
+We can then use these token description along with our complex matrix and our list of trees to run the ancestral states estimations as follows:
+
+```{r}
+## Running ancestral states
+ancestral_states <- multi.ace(matrix_complex, multiple_trees,
+                              special.tokens = my_spec_tokens,
+                              special.behaviours = my_spec_behaviours,
+                              verbose = TRUE)
+
+## This outputs a list of ancestral parts of the matrices for each tree
+## For example, here's the first one:
+ancestral_states[[1]][1:9, 1:10]
+```
+
+Note that there are many different options that are not covered here.
+For example, you can use different models for each character via the `models` argument, you can specify how to handle uncertainties via the `threshold` argument, use a branch length modifier (`brlen.multiplier`), specify the type of output, etc...
+
+### Feeding the results to `char.diff` to get distance matrices
+
+Finally, after running your ancestral states estimations, it is not uncommon to then use these resulting data to calculate the distances between taxa and then ordinate the results to measure disparity.
+You can do that using the `char.diff` function [described above](#char.diff) but instead of measuring the distances between characters (columns) you can measure the distances between species (rows).
+You might notice that this function uses the same modular token and behaviour descriptions.
+That makes sense because they're using the same core C functions implemented in dispRity that greatly speed up distance calculations.
+
+```{r}
+## Running ancestral states
+## and outputing a list of combined matrices (tips and nodes)
+ancestral_states <- multi.ace(matrix_complex, multiple_trees,
+                              special.tokens = my_spec_tokens,
+                              special.behaviours = my_spec_behaviours,
+                              output = "combined.matrix",
+                              verbose = TRUE)
+```
+
+We can then feed these matrices directly to `char.diff`, say for calculating the "MORD" distance:
+
+```{r}
+## Measuring the distances between rows using the MORD distance
+distances <- lapply(ancestral_states, char.diff, method = "mord", by.col = FALSE)
+```
+
+And we now have a list of distances matrices with ancestral states estimated!

### 2024-01-24 inst/gitbook/05_other-functionalities.Rmd

@@
 
-This function allows to run the `ape::ace` function (ancestral characters estimations) on multiple trees.
+This function allows to run ancestral characters estimations on multiple trees.
 In it's most basic structure (e.g.
@@
 
-Finally, after running your ancestral states estimations, it is not uncommon to then use these resulting data to calculate the distances between taxa and then ordinate the results to measure disparity.
+After running your ancestral states estimations, it is not uncommon to then use these resulting data to calculate the distances between taxa and then ordinate the results to measure disparity.
 You can do that using the `char.diff` function [described above](#char.diff) but instead of measuring the distances between characters (columns) you can measure the distances between species (rows).
@@
 And we now have a list of distances matrices with ancestral states estimated!
+
+### Running ancestral states estimations for continuous characters
+
+You can also run `multi.ace` on continuous characters.
+The function detects any continuous characters as being of class `"numeric"` and runs them using the `ape::ace` function.
+
+```{r}
+set.seed(1)
+## Creating three coalescent trees
+my_trees <- replicate(3, rcoal(15), simplify = FALSE)
+## Adding node labels
+my_trees <- lapply(my_trees, makeNodeLabel)
+## Making into a multiPhylo object
+class(my_trees) <- "multiPhylo"
+
+## Creating a matrix of continuous characters
+data <- space.maker(elements = 15, dimensions = 5, distribution = rnorm,
+                    elements.name = my_trees[[1]]$tip.label)
+```
+
+With such data and trees you can easily run the `multi.ace` estimations. 
+By default, the estimations use the default arguments from `ape::ace`, knowingly a Brownian Motion (`model = "BM"`) with the REML method (`method = "REML"`; this method "first estimates the ancestral value at the root (aka, the phylogenetic mean), then the variance of the Brownian motion process is estimated by optimizing the residual log-likelihood" - from `?ape::ace`).
+
+```{r}
+## Running multi.ace on continuous data
+my_ancestral_states <- multi.ace(data, my_trees)
+## We end up with three matrices of node states estimates
+str(my_ancestral_states)
+```
+
+This results in three matrices with ancestral states for the nodes.
+When using continuous characters, however, you can output the results directly as a `dispRity` object that allows visualisation and other normal dispRity pipeline:
+
+```{r}
+## Running multi.ace on continuous data
+my_ancestral_states <- multi.ace(data, my_trees, output = "dispRity")
+## We end up with three matrices of node states estimates
+plot(my_ancestral_states)
+```
+
+You can also mix continuous and discrete characters together.
+By default the `multi.ace` detects which character is of which type and applies the correct estimations based on that.
+However you can always specify models or other details character per characters.
+
+```{r}
+## Adding two discrete characters
+data <- as.data.frame(data)
+data <- cbind(data, "new_char" = as.character(sample(1:2, 15, replace = TRUE)))
+data <- cbind(data, "new_char2" = as.character(sample(1:2, 15, replace = TRUE)))
+
+## Setting up different models for each characters
+## BM for all 5 continuous characters
+## and ER and ARD for the two discrete ones
+my_models <- c(rep("BM", 5), "ER", "ARD")
+
+## Running the estimation with the specified models
+my_ancestral_states <- multi.ace(data, my_trees, models = my_models)
+```
+
+Of course all the options discussed in the first part above also can apply here!

### 2024-02-19 inst/gitbook/05_other-functionalities.Rmd
added manual entry for set.root.time

@@
 
+## `set.root.time`
+
+This function can be used to easily add a `$root.time` element to `"phylo"` or `"multiPhylo"` objects.
+This `$root.time` element is used by `dispRity` and several packages (e.g.
+`Claddis` and `paleotree`) to scale the branch length units of a tree allowing them to be usually expressed in million of years (Mya).
+
+For example, on a standard random tree, no `$root.time` exist so the edge lengths are not expressed in any specific unit:
+
+```{r}
+## A random tree with no root.time
+my_tree <- rtree(10)
+my_tree$root.time # is NULL
+```
+
+You can add a root time by either manually setting it:
+
+```{r}
+## Adding an arbitrary root time
+my_tree_arbitrary <- my_tree
+## Setting the age of the root to 42
+my_tree_arbitrary$root.time <- 42
+```
+
+Or by calculating it automatically from the cumulated branch length information (making the youngest tip age 0 and the oldest the total age/depth of the tree)
+
+```{r}
+## Calculating the root time from the present
+my_tree_aged <- my_tree 
+my_tree_aged <- set.root.time(my_tree)
+```
+
+If you want the youngest tip to not be of age 0, you can define an arbitrary age for it and recalculate the age of the root from there using the `present` argument (say the youngest tip is 42 Mya old):
+
+```{r}
+## Caculating the root time from 42 Mya 
+my_tree_age <- set.root.time(my_tree, present = 42)
+```
+
+This function also works with a distribution of trees (`"multiPhylo"`).
+
 ## `slice.tree`

### 2024-03-22 inst/gitbook/05_other-functionalities.Rmd
Update 05_other-functionalities.Rmd

@@
 ## Treating uncertainties as all values present (like polymorphisms):
-                          uncertanity  <- function(x,y) strsplit(x, split = "\\&")[[1]])
+                          uncertanity  <- function(x,y) strsplit(x, split = "\\/")[[1]])
 ```

### 2024-08-05 inst/gitbook/03_specific-tutorials.Rmd
Update 03_specific-tutorials.Rmd

@@
 
-  * If the object contains customised subsets, it applies the default formula `matrix ~ group` testing the effect of `group` as a predictor on `matrix` (called from the `dispRity` object as `data$matrix` see [`dispRitu` object details](#The-dispRity-object-content))
+  * If the object contains customised subsets, it applies the default formula `matrix ~ group` testing the effect of `group` as a predictor on `matrix` (called from the `dispRity` object as `data$matrix` see [`dispRity` object details](#The-dispRity-object-content))
   * If the object contains time subsets, it applies the default formula `matrix ~ time` testing the effect of `time` as a predictor (were the different levels of `time` are the different time slices/bins)

### 2024-10-02 inst/gitbook/03_specific-tutorials.Rmd
Minor ups

--- 2023-12-06 inst/gitbook/03_specific-tutorials.Rmd
@@
 1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln$vol` |
+2 | `count.neighbours` | The number of neigbhours to each element in a specified radius | `dispRity` |
 2 | `deviations` | The minimal distance between each element and a hyperplane | `dispRity` |
@@
 
-#### Ranges, variances, quantiles, radius, pairwise distance, neighbours, modal value and diagonal
+#### Ranges, variances, quantiles, radius, pairwise distance, neighbours (and counting them), modal value and diagonal
 
@@
                  which = max, method = "manhattan"))
+
+## The overall number of neighbours per point
+summary(dispRity(dummy_space, metric = count.neighbours,
+                 relative = FALSE))
+
+## The relative number of neigbhours
+## two standard deviations of each element
+summary(dispRity(dummy_space, metric = count.neighbours,
+                 radius = function(x)(sd(x)*2),
+                 relative = TRUE))
 ```
@@
 
-One often useful test to apply to multidimensional data is the permutational multivariate analysis of variance based on distance matrices `vegan::adonis`.
+One often useful test to apply to multidimensional data is the permutational multivariate analysis of variance based on distance matrices `vegan::adonis2`.
 This can be done on `dispRity` objects using the `adonis.dispRity` wrapper function.

### 2024-10-03 inst/gitbook/03_specific-tutorials.Rmd
Update 03_specific-tutorials.Rmd

@@
 ```
+
+## Disparity and distances
+
+There are two ways to use distances in `dispRity`, either with your input data being directly a distance matrix or with your disparity metric involving some kind of distance calculations.
+
+### Disparity data is distance
+
+If your disparity data is a distance matrix, you can use the option `is.dist = TRUE` in `dispRity` to make sure that all the operations done on your data take into account the fact that your disparity data has distance properties.
+For example, if you bootstrap the data, this will automatically bootstrap both rows AND columns (i.e. so that the bootstrapped matrices are still distances).
+This also improves speed on some calculations if you use [disparity metrics](#disparity-metrics) directly implemented in the package by avoiding recalculating distances (the full list can be seen in `?dispRity.metric` - they are usually the metrics with `dist` in their name).
+
+### Disparity metric is distance
+
+On the other hand if your data is not a distance matrix but you are using a metric that uses some kind of distance calculations, you can use the option `dist.helper` to greatly speed up calculations.
+`dist.helper` can be either a pre-calculated distance matrix (or a list of distance matrices) or, better yet, a function to calculate distance matrices, like `stats::dist` or `vegan::vegdist`.
+This option directly stores the distance matrix separately in the RAM and allows the disparity metric to directly access it at every disparity calculation iteration, making it much faster.
+Note that if you provide a function for `dist.helper`, you can also provide any un-ambiguous optional argument to that function, for example `method = "euclidean"`.
+
+If you use a disparity metric implemented in `dispRity`, the `dist.helper` option is correctly loaded onto the RAM regardless of the argument you provide (a matrix, a list of matrix or any function to calculate a distance matrix).
+On the other hand, if you use your own function for the disparity metric, make sure that `dist.helper` exactly matches the internal distance calculation function.
+For example if you use the already implemented `pairwise.dist` metric all the following options will be using `dist.helper` optimally:
+
+```{r, eval = FALSE}
+## Using the dist function from stats (specifying it comes from stats)
+dispRity(my_data, metric = pairwise.dist, dist.helper = stats::dist)
+
+## Using the dist function from vegdist function (without specifying its origin)
+dispRity(my_data, metric = pairwise.dist, dist.helper = vegdist)
+
+## Using some pre-calculated distance with a generic function
+my_distance_matrix <- dist(my_distance_data)
+dispRity(my_data, metric = pairwise.dist, dist.helper = my_distance_matrix)
+
+## Using some pre-calculated distance with a user function defined elsewhere
+my_distance_matrix <- my.personalised.function(my_distance_data)
+dispRity(my_data, metric = pairwise.dist, dist.helper = my_distance_matrix)
+```
+
+However, if you use a homemade metric for calculating distances like this:
+
+```{r, eval = FALSE}
+## a personalised distance function
+my.sum.of.dist <- function(matrix) {
+  return(sum(dist(matrix)))
+}
+```
+
+The `dist.helper` will only work if you specify the function using the same syntax as in the user function:
+
+```{r, eval = FALSE}
+## The following uses the helper correctly (as in saves a lot of calculation time)
+dispRity(my_data, metric = my.sum.of.dist, dist.helper = dist)
+
+## These ones however, work but don't use the dist.helper (don't save time)
+## The dist.helper is not a function
+dispRity(my_data, metric = my.sum.of.dist, dist.helper = dist(my_data))
+## The dist.helper is not the correct function (should be dist)
+dispRity(my_data, metric = my.sum.of.dist, dist.helper = vegdist)
+## The dist.helper is not the correct function (should be just dist)
+dispRity(my_data, metric = my.sum.of.dist, dist.helper = stats::dist)
+```

### 2024-11-08 inst/gitbook/03_specific-tutorials.Rmd
draft add

@@
 boot.matrix(time_slices, bootstraps = 100)
+```
+
+### Bootstrapping dimensions
+
+In some cases, you might also be interested in bootstrapping dimensions rather than observations.
+I.e. bootstrapping the columns of a matrix rather than the rows.
+
+It's pretty easy! By default, `boot.matrix` uses the option `boot.by = "rows"` which you can toggle to `boot.by = "columns"`
+
+```{r, eval = TRUE}
+## Bootstrapping the observations (default)
+boot_obs <- boot.matrix(data = crown_stem, boot.by = "rows")
+
+## Bootstrapping the columns rather than the rows
+boot_dim <- boot.matrix(data = crown_stem, boot.by = "columns")
+```
+
+In these two examples, the first one `boot_obs` bootstraps the rows as showed before (default behaviour).
+But the second one, `boot_dim` bootstraps the dimensions.
+That means that for each
+
+```{r, eval = TRUE}
+## Measuring disparity and summarising
+summary(dispRity(boot_obs, metric = mean))
+summary(dispRity(boot_dim, metric = mean))
+```
+
+### Bootstrapping with fine tuned probabilities
+
+```{r, eval = TRUE}
+boot.matrix(data = BeckLee_mat50)
 ```
@@
 
-### Disparity data is distance
-
-If your disparity data is a distance matrix, you can use the option `is.dist = TRUE` in `dispRity` to make sure that all the operations done on your data take into account the fact that your disparity data has distance properties.
+### Disparity data is a distance
+
+If your disparity data is a distance matrix, you can use the option `dist.data = TRUE` in `dispRity` to make sure that all the operations done on your data take into account the fact that your disparity data has distance properties.
 For example, if you bootstrap the data, this will automatically bootstrap both rows AND columns (i.e. so that the bootstrapped matrices are still distances).
@@
 
-### Disparity metric is distance
+#### Subsets
+
+dist.data = TRUE
+
+#### Bootstrapping
+
+boot.by = "dist"
+
+### Disparity metric is a distance
 

### 2024-11-11 inst/gitbook/03_specific-tutorials.Rmd
Updated tutorials

@@
 ```
-
-It is also possible to specify the sampling probability in the bootstrap for each elements.
-This can be useful for weighting analysis for example (i.e. giving more importance to specific elements).
-These probabilities can be passed to the `prob` argument individually with a vector with the elements names or with a matrix with the rownames as elements names.
-The elements with no specified probability will be assigned a probability of 1 (or 1/maximum weight if the argument is weights rather than probabilities).
-
-```{r, eval=TRUE}
-## Attributing a weight of 0 to Cimolestes and 10 to Maelestes
-boot.matrix(BeckLee_mat50,
-            prob = c("Cimolestes" = 0, "Maelestes" = 10))
-```
-
 Of course, one could directly supply the subsets generated above (using `chrono.subsets` or `custom.subsets`) to this function.
@@
 
+### Bootstrapping with probabilities
+
+It is also possible to specify the sampling probability in the bootstrap for each elements.
+This can be useful for weighting analysis for example (i.e. giving more importance to specific elements).
+These probabilities can be passed to the `prob` argument individually with a vector with the elements names or with a matrix with the rownames as elements names.
+The elements with no specified probability will be assigned a probability of 1 (or 1/maximum weight if the argument is weights rather than probabilities).
+
+```{r, eval=TRUE}
+## Attributing a weight of 0 to Cimolestes and 10 to Maelestes
+boot.matrix(BeckLee_mat50,
+            prob = c("Cimolestes" = 0, "Maelestes" = 10))
+```
+
 ### Bootstrapping dimensions
@@
 ## Bootstrapping the observations (default)
+set.seed(1)
 boot_obs <- boot.matrix(data = crown_stem, boot.by = "rows")
@@
 ## Bootstrapping the columns rather than the rows
+set.seed(1)
 boot_dim <- boot.matrix(data = crown_stem, boot.by = "columns")
@@
 But the second one, `boot_dim` bootstraps the dimensions.
-That means that for each
+That means that for each bootstrap sample, the value calculated is actually obtained by reshuffling the dimensions (columns) rather than the observations (rows).
 
@@
 ## Measuring disparity and summarising
-summary(dispRity(boot_obs, metric = mean))
-summary(dispRity(boot_dim, metric = mean))
-```
-
-### Bootstrapping with fine tuned probabilities
-
-```{r, eval = TRUE}
-boot.matrix(data = BeckLee_mat50)
-```
+summary(dispRity(boot_obs, metric = sum))
+summary(dispRity(boot_dim, metric = sum))
+```
+
+Note here how the observed sum is the same (no bootstrapping) but the bootstrapping distributions are quiet different even though the same seed was used.
 
@@
 
-dist.data = TRUE
+By default, the `dispRity` package does not treat any matrix as a distance matrix.
+It will however try to guess whether your input data is a distance matrix or not.
+This means that if you input a distance matrix, you might get a warning letting you know the input matrix might not be treated correctly (e.g. when bootstrapping or subsetting).
+For the functions `dispRity`, `custom.subsets` and `chrono.subsets` you can simply toggle the option `dist.data = TRUE` to make sure you treat your input data as a distance matrix throughout your analysis.
+
+```{r}
+## Creating a distance matrix
+distance_data <- as.matrix(dist(BeckLee_mat50))
+
+## Measuring the diagonal of the distance matrix
+dispRity(distance_data, metric = diag, dist.data = TRUE)
+```
+
+If you use a pipeline of any of these functions, you only need to specify it once and the data will be treated as a distance matrix throughout.
+
+```{r}
+## Creating a distance matrix
+distance_data <- as.matrix(dist(BeckLee_mat50))
+
+## Creating two subsets specifying that the data is a distance matrix
+subsets <- custom.subsets(distance_data, group = list(c(1:5), c(6:10)), dist.data = TRUE)
+## Measuring disparity treating the data as distance matrices
+dispRity(subsets, metric = diag)
+
+## Measuring disparity treating the data as a normal matrix (toggling the option to FALSE)
+dispRity(subsets, metric = diag, dist.data = FALSE)
+## Note that a warning appears but the function still runs
+```
 
@@
 
-boot.by = "dist"
+The function `boot.matrix` also can deal with distance matrices by bootstrapping both rows and columns in a linked way (e.g.
+if a bootstrap pseudo-replicate draws the values 1, 2, and 5, it will select both columns 1, 2, and 5 and rows 1, 2, and 5 - keeping the distance structure of the data).
+You can do that by using the `boot.by = "dist"` function that will bootstrap the data in a distance matrix fashion:
+
+```{r}
+## Measuring the diagonal of a bootstrapped matrix
+boot.matrix(distance_data, boot.by = "dist")
+```
+
+Similarly to the `dispRity`, `custom.subsets` and `chrono.subsets` function above, the option to treat the input data as a distance matrix is recorded and recycled so there is no need to specify it each time.
 

### 2024-11-11 inst/gitbook/03_specific-tutorials.Rmd
Update 03_specific-tutorials.Rmd

@@
 ```
+<!-- 
+## Multi disparity: running disparity analyses on different data
+
+All the options and functions we've covered so far are running disparity analyses assuming that there is one unique trait space (although that trait space can be probabilistic and comprising multiple matrices).
+
+```{r}
+
+    tree <- rmtree(2, 10)
+    tree[[1]] <- makeNodeLabel(tree[[1]])
+    tree[[2]] <- makeNodeLabel(tree[[2]], prefix = "shnode")
+    tree[[1]]$root.time <- max(tree.age(tree[[1]])$ages)
+    tree[[2]]$root.time <- max(tree.age(tree[[2]])$ages)
+    data <- list(matrix(0, nrow = Ntip(tree[[1]]) + Nnode(tree[[1]]), dimnames = list(c(tree[[1]]$tip.label, tree[[1]]$node.label))),
+                 matrix(0, nrow = Ntip(tree[[2]]) + Nnode(tree[[2]]), dimnames = list(c(tree[[2]]$tip.label, tree[[2]]$node.label))))
+
+    ## Test working fine
+    test <- dispRity(data, metric = mean, tree = tree)
+    plot(test)
+    expect_equal(summary(test)$obs.median, 0)
+```
+
+ -->

### 2024-11-12 inst/gitbook/03_specific-tutorials.Rmd
book up

@@
 In these examples, the function bootstraps the data (without rarefaction) AND also bootstraps the data with the different rarefaction levels.
-
-One other argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
-When missing, all dimensions from the ordinated matrix are used.
-
-```{r, eval=TRUE}
-## Using the first 50% of the dimensions
-boot.matrix(BeckLee_mat50, dimensions = 0.5)
-
-## Using the first 10 dimensions
-boot.matrix(BeckLee_mat50, dimensions = 10)
-```
-Of course, one could directly supply the subsets generated above (using `chrono.subsets` or `custom.subsets`) to this function.
 
@@
 
-```{r, fig.width=8, fig.height=4}
+```{r, fig.width=12, fig.height=6}
 ## Making the different subsets
@@
 
-```{r, fig.height = 16, fig.width = 8}
+```{r, fig.height = 12, fig.width = 6}
 par(mfrow = c(2,1))
@@
 
-```{r}
+```{r, fig.height = 12, fig.width = 12}
 ## Loading some demo data:
@@
 
-```{r, fig.width=8, fig.height=8}
+```{r, fig.width=8, fig.height=4}
 ## Graphical parameters
@@
 
-```{r, print = FALSE}
+```{r, print = FALSE, fig.width = 4, fig.height = 12}
 par(mfrow = c(3,1))
@@
 For more details on running these models, I suggest using the `MCMCglmm` package (@MCMCglmm) from @hadfield2010 (but see also @mulTree).
+For an example use of this code, see @guillerme2023innovation.
 
@@
 
-```{r}
+```{r, fig.height = 12, fig.width = 12}
 par(mfrow = c(2,2))

### 2024-11-12 inst/gitbook/05_other-functionalities.Rmd
book up

@@
 
-```{r}
+```{r, fig.height = 12, fig.width = 4}
 set.seed(42)

### 2025-02-27 inst/gitbook/05_other-functionalities.Rmd
Vignette update. Read to be pushed

@@
 This function allows to run ancestral characters estimations on multiple trees.
-In it's most basic structure (e.g.
-using all default arguments) this function is using a mix of `ape::ace` and `castor::asr_mk_model` depending on the data and the situation and is generally faster than both functions when applied to a list of trees.
-However, this function provides also some more complex and modular functionalities, especially appropriate when using discrete morphological character data.
+In it's most basic structure (e.g. using all default arguments) this function is using a mix of `ape::ace` and `castor::asr_mk_model` depending on the data and the situation.
+If you're running only a single trait estimation, using these functions is more recommended rather that using `multi.ace`.
+However, this provides some more complex and modular functionalities which can make it more appropriate to use that their parent functions in the following cases:
 
-### Using different character tokens in different situations
+ * Inferring multiple traits from a `"matrix"` or a `"data.frame"` (these all continuous or discrete traits, or a mix of both!);
+ * Inferring traits using multiple trees;
+ * Using traits estimations distributions (samples) rather than point estimates;
+ * Inferring complex discrete characters;
+ * Running parallel estimations;
+
+The following tutorials cover some of the more advanced options in more details compared to the function manual (`?multi.ace`).
+
+### Complex discrete characters: using different tokens in different situations
 
@@
 Of course all the options discussed in the first part above also can apply here!
+
+### Getting sampled distributions instead of point estimates from the estimations
+
+You can also use `multi.ace` to sample distributions instead of getting point estimates.
+For example, by default, when using `multi.ace` on continuous characters, the output estimation is the average most likely value for the trait at every specific nodes.
+
+```{r}
+set.seed(1)
+## Creating a coalescent tree
+my_tree <- makeNodeLabel(rcoal(15))
+
+## Creating a matrix of continuous characters
+data <- space.maker(elements = 15, dimensions = 1, distribution = rnorm,
+                    elements.name = my_tree$tip.label)
+
+## Running the estimation with details
+point_estimates <- multi.ace(data, my_tree, estimation.details = list("CI95"))
+```
+
+For example for trait 1 for Node 1, the estimated trait value is `r unname(round(point_estimates$estimations[1,1], 3))` but with a confidence interval of `r unname(round(point_estimates$details[[1]]$CI95[1,1], 3))` and `r unname(round(point_estimates$details[[1]]$CI95[1,2], 3))` which is quiet the spread.
+To avoid treating the ancestral estimation as a point estimate (here the average of a large spread) we can also just get a number of samples to get an estimation of the distribution instead of the point estimate using the `sample` argument.
+
+```{r}
+## Getting 1000 samples from the distribution
+distribution_estimates <- multi.ace(data, my_tree, sample = 1000)
+```
+
+This results in 1000 matrices that are sampled by default uniformaly.
+We can visualise that for Node 1 that will be randomly sampled between the estimate's 95% confidence interval highlighted above.
+
+```{r, fig.height = 4, fig.width = 4}
+## Visualising the sampling of Node 1
+node_1_estimates <- unlist(lapply(distribution_estimates, function(x) x[1,1]))
+hist(node_1_estimates, main = "Distribution of estimates for Node 1")
+# Note that the two small bars are because of the
+# last and first bins should start/end at the actual 95 CI
+```
+
+Of course, we can change the type of function to use for sampling knowing the confidence interval of our estimations.
+We can specify this by providing a named list containing the following two elements:
+
+ 1. `fun` that is the function to use for the sampling (`rnorm` in the example below);
+ 2.
+`param` that contains a list of functions or variables on how to calculate the extra arguments for `fun`.
+For example for `rnorm`, the function needs 1) a `mean` argument (here `mean = mean`, to use the function `mean` to calculate...
+the mean); and 2) a `sd` argument (here `sd = approx.sd`, a function to approximate the stnadard deviation knowing the confidence interval - assuming normality of the data).
+
+```{r}
+## Function for approximating the standard deviation
+## from the range of a confidence interval.
+approx.sd <- function(CI)return(diff(range(CI))/4)
+
+## Getting 1000 samples from the distribution using a normal one
+distribution_estimates <- multi.ace(data, my_tree, sample = 1000,
+                            sample.fun = list(fun = rnorm,
+                                              param = list(mean = mean,
+                                                           sd = approx.sd)))
+```
+
+And again, we can visualise that:
+
+```{r, fig.height = 4, fig.width = 4}
+## Visualising the sampling of Node 1
+node_1_estimates <- unlist(lapply(distribution_estimates, function(x) x[1,1]))
+hist(node_1_estimates, main = "Distribution of estimates for Node 1")
+# Which is now a neat normal distribution
+```
+
+In this example we used the default and a specified `sample.fun` argument for a matrix with only one trait.
+However you can use any number of distributions for any number of traits: either just the one to be applied to all traits or a list of the same length as the number of traits with a list of `fun` and `param` for each.
+
+#### Sampling discrete characters
+
+You can also use the `sample` option on discrete characters.
+In that case it's even easier because it doesn't need a `sample.fun` argument.
+The character states estimates are just drawn from the scale likelihood estimate for each character (e.g. X% chance of being in state `0` and Y% chance of being in state `1` for a binary character).

### 2025-04-29 inst/gitbook/03_specific-tutorials.Rmd
minor fixes

@@
 Note that it is possible to measure a ratio from another point than `0` using the `reference` argument.
-It gives indication of the relative displacement of elements in the multidimensional space: a score  >1 signifies a displacement *away* from the reference.
-A score of >1 signifies a displacement *towards* the reference.
+It gives indication of the relative displacement of elements in the multidimensional space: a score > 1 signifies a displacement away from the reference.
+A score of < 1 signifies a displacement *towards* the reference.
 

### 2025-04-29 tests/testthat/test-multi.ace.R

@@ -502,6 +502,22 @@ test_that("sample.ace works", {
         expect_true(all(out[[i]] > ace$CI95[i,1]))
         expect_true(all(out[[i]] < ace$CI95[i,2]))
     }
+
+    # ## Visual check if samples actually resamples the distributions correctly:
+    # normal.sample <- list(fun = rnorm, param = list(mean = mean, sd = function(x)return(diff(range(x))/4)))
+    # uniform.sample <- list(fun = runif, param = list(min = min, max = max))
+    # norm_data <- rnorm(1000, sd = 8, mean = -3)
+    # unif_data <- runif(1000, min = 7, max = 70)
+    # test_norm <- list(CI95 = matrix(quantile(norm_data, prob = c(0.025, 0.975)), nrow = 1))
+    # test_unif <- list(CI95 = matrix(quantile(unif_data, prob = c(0.025, 0.975)), nrow = 1))
+    # res_norm <- sample.ace(test_norm, sample.fun = normal.sample, samples = 1000)[[1]]
+    # res_unif <- sample.ace(test_unif, sample.fun = uniform.sample, samples = 1000)[[1]]
+    # plot(density(norm_data))
+    # lines(density(res_norm), col = "red")
+    # plot(density(unif_data))
+    # lines(density(res_unif), col = "red")
+
+
 })
 
 test_that("multi.ace works with sample", {

### 2025-07-17 inst/gitbook/05_other-functionalities.Rmd

@@
 
+To play around with different sampling strategies, but if your estimations take a lot of time, you can save your estimations as a `"multi.ace"` object that you can then recycle and reuse very easily:
+
+```{r}
+## Saving the estimation without getting any estimate (raw):
+raw_estimates <- multi.ace(data, my_tree, output = "multi.ace`")
+
+## Sampling from the raw_estimates
+sampled_estimates <- multi.ace(raw_estimates, sample = 100)
+```
+
 #### Sampling discrete characters
